### 2022-08-01 02-cross-refs.Rmd
try child page

--- 
title: 'Cross-refs'
has_children: true
---

# Cross-references {#cross}

Cross-references make it easier for your readers to find and link to elements in your book.

## Chapters and sub-chapters

There are two steps to cross-reference any heading:

1. Label the heading: `# Hello world {#nice-label}`. 
    - Leave the label off if you like the automated heading generated based on your heading title: for example, `# Hello world` = `# Hello world {#hello-world}`.
    - To label an un-numbered heading, use: `# Hello world {-#nice-label}` or `{# Hello world .unnumbered}`.

1. Next, reference the labeled heading anywhere in the text using `\@ref(nice-label)`; for example, please see Chapter \@ref(cross). 
    - If you prefer text as the link instead of a numbered reference use: [any text you want can go here](#cross).

## Captioned figures and tables

Figures and tables *with captions* can also be cross-referenced from elsewhere in your book using `\@ref(fig:chunk-label)` and `\@ref(tab:chunk-label)`, respectively.

See Figure \@ref(fig:nice-fig).

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center', fig.alt='Plot with connected points showing that vapor pressure of mercury increases exponentially as temperature increases.'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Don't miss Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(pressure, 10), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

### 2022-08-01 03-parts.Rmd
try child page

---
title: 'parts'
parent: Cross-refs

---

# Parts

You can add parts to organize one or more book chapters together. Parts can be inserted at the top of an .Rmd file, before the first-level chapter heading in that same file.

Add a numbered part: `# (PART) Act one {-}` (followed by `# A chapter`)

Add an unnumbered part: `# (PART\*) Act one {-}` (followed by `# A chapter`)

Add an appendix as a special kind of un-numbered part: `# (APPENDIX) Other stuff {-}` (followed by `# A chapter`). Chapters in an appendix are prepended with letters instead of numbers.

### 2022-08-02 02-structure.Rmd
structure

# Overview {-}

<!-- This page is useful because by calling the indicator rmd files as child files we bypass the need to sequentially numbering them. An overview page is anywhays useful.  -->

## Test 1 {-}
This indicator is about ....

## Test 2 {-}
And this indicator is about...

```{r, child= c('indicators/indicatorTest.Rmd')}
```

```{r, child= c('indicators/indicatorTest2.Rmd')}
```

### 2022-08-02 index.Rmd
structure

--- 
title: "Indicators for Ecosystem Condition in Norway"
author: "Anders L. Kolstad"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
url: https://ninanor.github.io/ecosystemCondition/index.html
# cover-image: path to the social sharing image like images/cover.jpg
description: |
  This is a site for documenting the analyses related to the development and design of ecosystem condition indicators.
biblio-style: apalike
csl: chicago-fullnote-bibliography.csl
---

# Introduction {-}

On this site you may view the nitty gritty details for how cetain indicators for ecosystem condtion has been calculated.

To do:

Embed shiny app?

Link to author guidelines?

### 2022-08-02 indicators/indicatorTest.Rmd
structure

# This is an indicator {-#test1}

```{r}
plot(rnorm(10,10,10))
```

### 2022-08-02 indicators/indicatorTest.Rmd
template

@@
+# (PART\*) INDICATORS {-}
+
 # This is an indicator {-#test1}
+
+<br />
+
+```{r setup, include=FALSE}
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
+
+```{r, echo=F}
+Ecosystem <- "Skog og fjell"
+Egenskap  <- "Primærproduksjon"
+ECT       <- "Structural state characteristic"
+Contact   <- "Anders Kolstad"
+```
+
+```{r, echo=F}
+metaData <- data.frame(Ecosystem,
+                       "Økologisk egenskap" = Egenskap,
+                       "ECT class" = ECT)
+knitr::kable(metaData)
+```
+
+<br />
+<br />
+
+<hr />
+
+<!-- Documenting you work below. Dtaa can be stores on NINA server, harvested using API or similar, or if it is not too big, stored under /data/ on this repository -->
 

### 2022-08-02 indicators/template.Rmd
template

<!-- TEMPLATE -->

<!-- This is a template for how to document the indicator analyses. Make a copy of this file and sore it in the same folder (indicators/) before starting.  -->

<!-- Replace 'indicator name' with yo own indicator name  -->
# Indicator name {-}

<br />

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->

```{r, echo=F}
Ecosystem <- "" # e.g. "Skog og fjell"
Egenskap  <- "" # e.g. "Primærproduksjon"
ECT       <- "" # e.g. "Structural state characteristic"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these threehtml lines -->
<br />
<br />
<hr />

<!-- Document you work below. Data can be stored on NINA server, harvested using API or similar, or if it is not too big, stored under /data/ on this repository -->

Here's a nice plot,   Blablabla
```{r, fig.cap="This is my figure caption"}
plot(rnorm(10,10,10))
```

### 2022-08-03 indicators/workingFiles/template.Rmd
template

--- 2022-08-02 indicators/template.Rmd
@@
-<!-- TEMPLATE -->
+---
+title: "Template for indicator documentation"
+output:
+  pdf_document
+---
 
-<!-- This is a template for how to document the indicator analyses. Make a copy of this file and sore it in the same folder (indicators/) before starting.  -->
+<!-- This is a template for how to document the indicator analyses. 
+Make a copy of this file and sore it in the same folder (indicators/workingFiles/) 
+before starting. This is a working file. It is also a stand-alone Rmd file
+inside a bookdown project. Therefore, to knit the file for preview reasons,
+you need to type rmarkdown::render("indicators/workingFiles/template.Rmd") 
+in the console, as the the usual Ctrl + Alt + k will not work. When the analyses is complete,
+Anders Kolstad or someone else will perform a rapid code review, making sure 
+the analyses is repeatable by someone else. Be therefore very careful to annotate 
+and comment on every step in the analysis. When completed, Anders will make a copy
+of the working file and put it in the folder below (indicators/). After modifying it
+slightly (mainly removing the top YAML header) the output will become available 
+on the website (https://ninanor.github.io/ecosystemCondition/) -->
 
-<!-- Replace 'indicator name' with yo own indicator name  -->
+<!-- Replace 'indicator name' with yor own indicator name  -->
 # Indicator name {-}
@@
 ECT       <- "" # e.g. "Structural state characteristic"
+Contact   <- "" # e.g. "Anders Kolstad"
 ```
@@
 
-<!-- Don't remove these threehtml lines -->
+<!-- Don't remove these three html lines -->
 <br />

### 2022-08-03 02-structure.Rmd
Faktaark

@@
 
-## Test 1 {-}
+## Areal av isbreer / Glacier area {-}
 This indicator is about ....
 
-## Test 2 {-}
+```{r, echo=F}
+source("indicators/indicatorMetaData/Areal av isbreer.R")
+knitr::kable(t(out), 
+            format = "html",
+            caption = paste("Fakatark for tilstandindikatoren ", 
+                        out$Indikator, col.names = NULL))
+```
+
+## Indicator X {-}
 And this indicator is about...

### 2022-08-03 indicators/workingFiles/template.Rmd
Faktaark

@@
 output:
-  pdf_document
+  html_document
 ---
@@
 slightly (mainly removing the top YAML header) the output will become available 
-on the website (https://ninanor.github.io/ecosystemCondition/) -->
+on the website (https://ninanor.github.io/ecosystemCondition/). -->
 
@@
 ```
+
+add more detailed instruction about georeferencing ect
+...

### 2022-08-03 indicators/indicatorMetaData/Areal av isbreer.R

@@ -1,0 +1,33 @@
+# Areal av isbreer ----
+
+
+# indikatorens navn ----
+indikator <- "Areal av isbreer"
+
+# Utfylling av protokollen ----
+utfylling <- "Anders L. Kolstad"
+
+# Dato utfylt/revidert -----
+dato <- "08.12.2021"
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+#-------------------------------
+
+# Sett sammen til dataframe ----
+out <- data.frame(
+  "Indikator" = indikator,
+  "Utfylling av protokollen" = utfylling,
+  "Data utfylt/revidert" = dato
+)

### 2022-08-08 02-overview.Rmd
proj6
WW

--- 2022-08-03 02-structure.Rmd
@@
-# Overview {-}
-
-<!-- This page is useful because by calling the indicator rmd files as child files we bypass the need to sequentially numbering them. An overview page is anywhays useful.  -->
+# Overview of indicators {-}
 
 ## Areal av isbreer / Glacier area {-}
-This indicator is about ....
+This indicator reflects the relative loss of glacial area between tha last climatic normal persion and today.
 
@@
 And this indicator is about...
-
-```{r, child= c('indicators/indicatorTest.Rmd')}
-```
-
-```{r, child= c('indicators/indicatorTest2.Rmd')}
-```

### 2022-08-08 docs/template.Rmd
WW

---
title: "Template for indicator documentation"
output:
  html_document
---

<!-- This is a template for how to document the indicator analyses. 
Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name te file DRAFT_myIndicatorName.Rmd. 
Because this is also a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
you need to type rmarkdown::render("template.Rmd") in the console, 
as the the usual Ctrl + Alt + k will not work. When the analyses is complete,
Anders Kolstad, or someone else, will perform a rapid code review, making sure 
the analysis is repeatable by someone else. Be therefore very careful to annotate 
and comment on every step in the analysis. Se When completed, Anders will make a copy
of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') the output will become available 
on the website (https://ninanor.github.io/ecosystemCondition/). -->

<!-- Examples: For an example, see 'indicators/breareal-workingFile.Rmd'. Notice the degree of annotation.
Make sure also to not change the order, or modify the headers unless you really need to. This is because it
easier to read the web page as a whole of the indicators are presented using the same layout. 
To complete the exmaple, se 'breareal.Rmd' and notice how Anders has prepared the working file to become a
chapter in eth bookdown project. Anders will do this for you as patr of the code review.
-->

<!-- Replace 'indicator name' with yor own indicator name  -->
# Indicator name

<br />

_Author and date:_
<!-- Write your name here -->

```{r}
Sys.Date()
```

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "" # e.g. "Skog og fjell"
Egenskap  <- "" # e.g. "Primærproduksjon"
ECT       <- "" # e.g. "Structural state characteristic"
Contact   <- "" # e.g. "Anders Kolstad"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
<!-- Text here -->

## About the underlying data
<!-- Text here -->

### Representativity in time and space
<!-- Text here -->

### Original units
<!-- Text here -->

### Temporal coverage
<!-- Text here -->

### Aditional comments about the dataset
<!-- Text here -->

## Ecosystem characteristic
### Norwegain standard
<!-- Text here -->

### <!-- Text here -->
<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->

## Collinearities with other indicators
<!-- Text here -->

## Reference state and values
### Reference state
<!-- Text here -->

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
<!-- Text here -->

## Uncertainties
<!-- Text here -->

## References
<!-- Text here -->

## Analyses
### Data sets
<!-- Text here -->

#### Data set A
<!-- Text here -->

#### Data set B
<!-- Text here -->

#### Regions
<!-- In case you need to map the indicator value to regions, you can do that here -->
Importing a shape file with the regional delineation. 
```{r}
#reg <- st_read("data/regions.shp")
#st_crs(reg)
```

### Scaled indicator values
<!-- Text and analyses here -->

### Uncertainty
<!-- Text here -->

## Prepare export
<!-- Text here -->

### Eksport file (final product)
Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors.

### 2022-08-08 index.Rmd
rm com. rev. ch.

@@
 
-# Introduction {-}
+# About {-}
 
-On this site you may view the nitty gritty details for how cetain indicators for ecosystem condtion has been calculated.
+On this site you may view the nitty gritty details for how certain indicators for ecosystem condition for Norway has been calculated.
 
-To do:
+The web page is made to document the work related to a R&D project to develop indicators for ecosystem condition for three main ecosystems in Norway: wetlands, open areas below the forest line, and semi-natural land.
+These indicators are not yet used in real assessments, and may represent unfinished work pending future data availability.
+In addition,  we may chose to also present some analyses documenting indicators that _are_ used in existing assessments of ecosystem condition.
 
-Embed shiny app?
+## Community review {-}
 
-Link to author guidelines?
+We want our work to be transparent and credible.
+Therefor we encourage all types of feedback on the analyses of the indicators for ecological condition that you can read about on this web site.
+To submit your comments, please to to [*this github repository*](https://github.com/NINAnor/ecosystemCondition/issues) and open a new _issue_ and we will answer them there.
+
+<img src="images/newIssue.PNG" alt="" width="600"/>
+
+## Author guidelines {-}
+
+For those participating in the indicator project, detailed information about how to document your work will be communicated over email, but the main steps can be summarised as this:
+
+* Obtain a _go ahead_ signal from the coordination group to make sure you are free to start your work
+* Read the [`template.Rmd`](template.Rmd) file.
+Make a copy of it and store it, in the same (root) folder as `DRAFT_myIndicator.Rmd`.
+(For an example, see `DRAFT_breareal.Rmd`).
+Finish your documentation and analyses there.
+  + You can load data from internal NINA servers, from web hosts, of store smaller data set under `data`.
+  The `data` folder also contains some supporting data sets that you may want to use, such as a delineation of the five regions used in the ecosystem assessment (`data/regions.shp`), and an outline of Norway (`outlineOfNorway_EPSG25833.shp`).
+  + To preview you rmarkdown file as html, type `rmarkdown::render("DRAFT_myIndicator.Rmd")` in the console (the knitr shortcuts don't work as expected inside bookdown projects). 
+  + When filling out the template.Rmd file with your own work, try not to change the headers too much.
+  + Final result from the analyses should be written to the `indicators` folder.
+The preferred output is georeferenced data (rasters or shape files), andthese should be placed under `indators/indicatorMaps`.
+All georeference data should use EPSG:25833 - ETRS89 / UTM zone 33N.
+Other output, like data tables, can be stired under `indicators`.
+* Make a pull request to the main [*GitHub repository*](https://github.com/NINAnor/ecosystemCondition) to submit your file.
+* Anders or someone else in the project wil conduct a rapid code review, making sure the code is reproducible, interpretable and that it renders locally on the R Studio server.
+* When approved, the reviewer will make a copy of the DRAFT_ file, removing the DRAFT_ part of the name, as well as the YML header, and as well as updating the rmd_files part of _bookdown.yml.
+(For an example of such a file, see `breareal.Rmd`.)
+* To update the web site (this web site), an admin user needs to pull down the repo to R Studio server, compile the book there (Ctr+Shift+B), commit and push to main.
+The html version of the book will be in the `docs` folder.

### 2022-08-12 AndersTest.Rmd

--- 2022-08-08 docs/template.Rmd
@@
 
+This is a test
+
 <!-- This is a template for how to document the indicator analyses.
+
 Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
 Name te file DRAFT_myIndicatorName.Rmd.
-Because this is also a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
+
+Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
 you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work. When the analyses is complete,
-Anders Kolstad, or someone else, will perform a rapid code review, making sure 
+as the the usual Ctrl + Alt + k will not work.
+
+Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
+easier to read the web page as a whole of the indicators are presented using the same layout.
+
+When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
 the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. Se When completed, Anders will make a copy
+and comment on every step in the analysis. When completed, Anders will make a copy
 of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') the output will become available 
-on the website (https://ninanor.github.io/ecosystemCondition/). -->
+After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
+the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
 
-<!-- Examples: For an example, see 'indicators/breareal-workingFile.Rmd'. Notice the degree of annotation.
-Make sure also to not change the order, or modify the headers unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout. 
-To complete the exmaple, se 'breareal.Rmd' and notice how Anders has prepared the working file to become a
-chapter in eth bookdown project. Anders will do this for you as patr of the code review.
+<!-- Examples:
+
+For an example, see 'indicators/breareal-workingFile.Rmd' 
+and the subsequent 'breareal.Rmd' 
+which again, when rendered, produces this:
+https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
+
+Notice the degree of annotation.
+
 -->
@@
 #### Regions
-<!-- In case you need to map the indicator value to regions, you can do that here -->
+<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
 Importing a shape file with the regional delineation. 
 ```{r}
-#reg <- st_read("data/regions.shp")
+#reg <- sf::st_read("data/regions.shp")
 #st_crs(reg)
@@
 ### Eksport file (final product)
-Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors.
+<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2022-08-18 naturtype.Rmd
scaling and nat.type

# (PART\*) GENERAL TOPICS {-}

# On the application of the naturetype dataset {#naturtype}

<br />

```{r, include=FALSE}
library(knitr)
library(sf)
library(tmap)
library(ggplot2)
library(units)
library(dplyr)
knitr::opts_chunk$set(echo = TRUE)
```

Author: Anders L. Kolstad

Date: `r Sys.Date()`

Here I will investigate a specific data set, [Naturtyper - Miljødirektoratets instruks](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c) and try to evaluate its usability for designing indicators for ecosystem condition. This involves assessing both the spatial and temporal representativity, as well as conseptual alignment with the [Norwegian system for ecosystem condition assessments](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/).

The precision in field mapping will not be assess in itself. We assume some, or even considerable, sampling error, but this is inherent to all field data.

## Import data and general summary statistics

* Total mapped area
* Temporal trend
* Histogram of conditions (shiny?)

```{r}
# local path
path <- "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
dat <- sf::st_read(dsn = path)
names(dat)
```
Import outline of Norway
```{r}
nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
```

The dataset has 95k polygons, each with 36 variables:
```{r}
dim(dat)
```
It's therefore takes a little minute to render a plot, but this is the code to do it:

```{r, eval = T}
tmap_mode("view")
tm_shape(dat) + 
  tm_polygons(col="tilstand")+
  tm_shape(nor)+
  tm_polygons(alpha = 0,border.col = "black")
```

### Area
Calculating the area

```{r}
dat$area <- sf::st_area(dat)
summary(dat$area)
```

The smallest polygons are 20 m2, and the biggset is 8.5 km2

Sum of mapped area divided by Norwegian mainland area
```{r}
sum(dat$area)/sf::st_area(nor)
```

About 0.5% has been mapped.
It is therefore essential that these 0.5% are representative (of the remaining 99.5%).

## Overview of nature types
* What are the dominant ecosystem groups recorded. 
```{r}
ggplot(dat, aes(x = hovedøkosystem))+
  geom_bar()+
  coord_flip()
```

This project is concerned only with wetlands (våtmark), semi-natural areas (semi-naturlig mark) and naturally open areas below the forest line (naturlig åpne områder i lavlandet/under skoggrensa).
```{r}
dat$hovedøkosystem[dat$hovedøkosystem == "Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
```

```{r}
unique(dat$hovedøkosystem)
utvlagt <- c("Semi-naturlig mark",
             "Våtmark",
             "Naturlig åpne områder under skoggrensa")
dat2 <- dat[dat$hovedøkosystem %in% utvlagt,]
```

```{r}
temp <- tapply(dat2$area, dat2$hovedøkosystem, sum)
temp <- data.frame(eco = row.names(temp),
                   area = temp$temp)
ggplot(temp, aes(x = eco, y=area))+
  geom_bar(stat="identity")+
  coord_flip()
```

The same bias towards sami-natural areas are seen when looking at the area mapped, as when looking the number of polygons.

* What are the dominant nature types recorded.

```{r}

temp <- dat2[dat2$hovedøkosystem=="Våtmark",]
ggplot(temp, 
       aes(x = naturtype,
           y = area))+
  geom_bar(stat = "sum")+
  coord_flip()

temp <- tapply(temp$area, temp$naturtype, sum)
temp <- data.frame(nat = row.names(temp),
                   area = temp)

temp <- temp[order(temp$area),]
ggplot(temp[order(temp$area),], 
       aes(x = nat,
           y = area))+
  geom_bar(stat = "identity")+
  coord_flip()
```

* Can we group these into main ecosystem type categories?
* Are there any regional differences

## Environmental space

* Extract environmental data from the polygons and compare against reference data points.
* Uni-variate (e.g. _boreal hei_ plotted against elevation, oceanity, distance from roads, etc)

## Represntaivity in mapping units
Only a subset of nature types are mapped. How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?

## Conclution 1 
Conclude about which main ecosystem types that can reliably assess using the available mapping units?

What about suitability for defining reference values?

## Suitability 
In the subset of mapping units that can be said to represent an ecosystem group, which NiN or MdirPRAM variables should we use?

## Concrete examples
### Andel lynghei med minst to faser i kystlynghei

### Areal uten skadet og/eller død røsslyng i kystlynghei

### Aktuell bruksintensitet

### 2022-08-18 scalingFunctions.Rmd
scaling and nat.type

# Scaling functions

```{r, include=F}
library(ggplot2)
library(ggridges)
library(data.table)
theme_set(theme_bw(base_size = 12))
```

Indicator for ecological condition may be scaled against a reference value to become normalised. Actually, in SEEA EA, unscaled or raw indikcatr are not called indicators at all, but simply referred to as variables. But scaling variables against reference conditions can be done in many ways. Here are some exampels, for inspiration.

```{r}
set.seed(123)
```

Create random raw data for two variables
```{r}
dat <- data.frame(variable1 = rnorm(50, 10, 4), 
                  variable2 = rnorm(50, 30, 3))

temp <- data.table::melt(setDT(dat))
ggplot(temp,
       aes(x = value,
           y = variable,
           fill = stat(x)))+
  geom_density_ridges_gradient(scale = 3, size = 0.3)+
  scale_fill_gradient(low = 'red', high = 'green')
```

Lets say that the reference condition for variable1 to equal 16, and for variable 2 its 35

```{r}
dat$referenceHigh1 <- 16
dat$referenceHigh2 <- 35
```

## Linear, with a natural zero

We can scale our variable linearly, saying that all values above 12 represents 'perfect' condition. We also assume that the value zero is the worst possible condition for this variable. To do this we divide by the reference value, and truncate values above it.

```{r}
dat$indicator1 <- dat$variable1/dat$referenceHigh1
dat$indicator1[dat$indicator1>1] <- 1

dat$indicator2 <- dat$variable2/dat$referenceHigh2
dat$indicator2[dat$indicator2>1] <- 1
```

```{r}
temp <- melt(setDT(dat),
             measure.vars = c("indicator1", "indicator2"))
```

```{r}
ggplot(dat)+
  #geom_smooth(size=1.5, colour = "lightblue")+
  geom_point(
    aes(x = variable1, y = indicator1, fill = indicator1),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 12,
             linetype = "dashed",
             size =2)+

  geom_point(aes(x = variable2, y = indicator2, fill = indicator2),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 35,
             linetype = "dashed",
             size =2)+
  ylab("Indicator values")+
  xlab("Variable values")+

  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```

The two indicator become more comparable, at least around the reference value. But perhaps a variable value around 25 for variabel2 is qctually quite bad. Here we have assumed that the value zero is the worst possible value. That may not always be the case

## Linear, with defined lower limit
We can also say that the worst possible condition for thess variables are not zero, but for example say that it is 5 for variable1 and 28 for variable2.
```{r}
dat$referenceLow1 <- 5
dat$referenceLow2 <- 28
```

```{r}
dat$indicator1_LowHigh <- (dat$variable1-dat$referenceLow1)/(dat$referenceHigh1 - dat$referenceLow1)
dat$indicator1_LowHigh[dat$indicator1_LowHigh <0] <- 0
dat$indicator1_LowHigh[dat$indicator1_LowHigh >1] <- 1

dat$indicator2_LowHigh <- (dat$variable2-dat$referenceLow2)/(dat$referenceHigh2 - dat$referenceLow2)
dat$indicator2_LowHigh[dat$indicator2_LowHigh <0] <- 0
dat$indicator2_LowHigh[dat$indicator2_LowHigh >1] <- 1
```

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_LowHigh, fill = indicator1_LowHigh),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 16,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 5,
             linetype = "dashed",
             size =2)+

  geom_point(aes(x = variable2, y = indicator2_LowHigh, fill = indicator2_LowHigh),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 35,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 28,
             linetype = "dashed",
             size =2)+
  ylab("Indicator values")+
  xlab("Variable values")+

  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```

## Break point scaling
We can also define a threshold for good ecological condition. Lets for example say that for variable1, ecosystem function really declines when the variable shows values below 9.

```{r}
dat$thr1 <- 9
```

```{r}
dat$indicator1_breakPoint <- dat$indicator1_LowHigh
dat$indicator1_breakPoint[dat$variable1 < dat$thr1] <- (1.6*dat$indicator1_LowHigh[dat$variable1 < dat$thr1])

dat$indicator1_breakPoint[dat$variable1 > dat$thr1] <- (dat$variable1[dat$variable1 > dat$thr1]-
                                                          dat$thr1[dat$variable1 > dat$thr1])/
                                                        (dat$referenceHigh1[dat$variable1 > dat$thr1] - 
                                                           dat$thr1[dat$variable1 > dat$thr1])+0.6

```

```{r}
ggplot(dat)+
  geom_point(
    aes(x = variable1, y = indicator1_breakPoint, fill = indicator1_breakPoint),
             colour="black",
             size=10,
             alpha=.7,
             shape=21)+
  geom_vline(xintercept = 16,
             linetype = "dashed",
             size =2)+
  geom_vline(xintercept = 5,
             linetype = "dashed",
             size =2)+
  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+
  geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
               linetype="dashed",
               size=1.5)+

  ylab("Indicator values")+
  xlab("Variable values")+

  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
```

```{r}

```

### 2022-08-19 scalingFunctions.Rmd
scaling functions

@@
 
-Indicator for ecological condition may be scaled against a reference value to become normalised.
-Actually, in SEEA EA, unscaled or raw indikcatr are not called indicators at all, but simply referred to as variables.
-But scaling variables against reference conditions can be done in many ways.
-Here are some exampels, for inspiration.
+Author. Anders L. Kolstad
+
+Date: `r Sys.Date()`
+
+Indicator for ecological condition may be scaled against a reference value to become normalised and hence comparable.
+In SEEA EA, un-scaled or raw indicators are not called indicators at all, but simply referred to as variables, and I will use this distinction here.
+Scaling variables against reference conditions can be done in many ways.
+Here are some examples, for inspiration.
 
@@
 set.seed(123)
+```
+
+Define plotting aesthetics
+```{r}
+low <- "red"
+high <- "green"
+scale_fill_gradient(low = low, high = high)
+ridge <- geom_density_ridges_gradient(scale = 3, size = .3)
+myColour <- "black"
+mySize <- 10
+myAlpha <- .7
+myShape <- 21
 ```
@@
 temp <- data.table::melt(setDT(dat))
+
 ggplot(temp,
@@
            fill = stat(x)))+
-  geom_density_ridges_gradient(scale = 3, size = 0.3)+
-  scale_fill_gradient(low = 'red', high = 'green')
+  ridge+
+  scale_fill_gradient(low = low, high = high)
 ```
@@
 ggplot(dat)+
-  #geom_smooth(size=1.5, colour = "lightblue")+
   geom_point(
     aes(x = variable1, y = indicator1, fill = indicator1),
-             colour="black",
-             size=10,
-             alpha=.7,
-             shape=21)+
-  geom_vline(xintercept = 12,
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
              linetype = "dashed",
@@
   geom_point(aes(x = variable2, y = indicator2, fill = indicator2),
-             colour="black",
-             size=10,
-             alpha=.7,
-             shape=21)+
-  geom_vline(xintercept = 35,
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_vline(xintercept = dat$referenceHigh2,
              linetype = "dashed",
@@
 
-  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
+  scale_fill_gradient("Indicator values", low = low, high = high)
 ```
 
-The two indicator become more comparable, at least around the reference value.
-But perhaps a variable value around 25 for variabel2 is qctually quite bad.
-Here we have assumed that the value zero is the worst possible value.
-That may not always be the case
+To reverse what is good and what is bad condition: change the sign and add one.
+```{r}
+ggplot(dat)+
+  geom_point(
+    aes(x = variable1, y = indicator1*(-1)+1, fill = indicator1*(-1)+1),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
+             linetype = "dashed",
+             size =2)+
+
+  ylab("Indicator values")+
+  xlab("Variable values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+```
+
+The two indicators become more comparable now that they are normalised.
+At least this is true around the reference value.
+But perhaps a variable value around 25 for variabel2 is actually quite bad, and in the above example I have assumed that the value zero is the worst possible value.
+That may not always be the case.
 
 ## Linear, with defined lower limit
-We can also say that the worst possible condition for thess variables are not zero, but for example say that it is 5 for variable1 and 28 for variable2.
+We can also say that the worst possible condition for these variables is not zero, but for example say that it is 5 for variable1 and 28 for variable2.
 ```{r}
@@
 
+We can then normalise the data between these two reference values.
 ```{r}
@@
     aes(x = variable1, y = indicator1_LowHigh, fill = indicator1_LowHigh),
-             colour="black",
-             size=10,
-             alpha=.7,
-             shape=21)+
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
   geom_vline(xintercept = 16,
@@
   geom_point(aes(x = variable2, y = indicator2_LowHigh, fill = indicator2_LowHigh),
-             colour="black",
-             size=10,
-             alpha=.7,
-             shape=21)+
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
   geom_vline(xintercept = 35,
@@
 
-  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
+  scale_fill_gradient("Indicator values", low = low, high = high)
 ```
@@
 
+Then we can normalise again, using what I refer to as break point normalisation. For the maths, see [here](https://stats.stackexchange.com/questions/281162/scale-a-number-between-a-range)
 ```{r}
-dat$indicator1_breakPoint <- dat$indicator1_LowHigh
-dat$indicator1_breakPoint[dat$variable1 < dat$thr1] <- (1.6*dat$indicator1_LowHigh[dat$variable1 < dat$thr1])
+dat$indicator1_breakPoint <- ifelse(dat$variable1 < dat$thr1,
+                                    ((dat$variable1-dat$referenceLow1)/(dat$thr1 - dat$referenceLow1))*(0.6-0)+0,
+                                    ((dat$variable1-dat$thr1)/(dat$referenceHigh1 - dat$thr1))*(1-0.6)+0.6
+                                    )
 
-dat$indicator1_breakPoint[dat$variable1 > dat$thr1] <- (dat$variable1[dat$variable1 > dat$thr1]-
-                                                          dat$thr1[dat$variable1 > dat$thr1])/
-                                                        (dat$referenceHigh1[dat$variable1 > dat$thr1] - 
-                                                           dat$thr1[dat$variable1 > dat$thr1])+0.6
+dat$indicator1_breakPoint[dat$indicator1_breakPoint<0] <- 0
+dat$indicator1_breakPoint[dat$indicator1_breakPoint>1] <- 1
 
@@
     aes(x = variable1, y = indicator1_breakPoint, fill = indicator1_breakPoint),
-             colour="black",
-             size=10,
-             alpha=.7,
-             shape=21)+
-  geom_vline(xintercept = 16,
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
              linetype = "dashed",
              size =2)+
-  geom_vline(xintercept = 5,
+  geom_vline(xintercept = dat$referenceLow1,
              linetype = "dashed",
@@
 
-  scale_fill_gradient("Indicator values", low = 'red', high = 'green')
+  scale_fill_gradient("Indicator values", low = low, high = high)
+```
+
+## Two-sided indicator
+
+An indicator may have an optimum value somewhere in the middle of its range, rather than a max or min value. You can then define a two-sided indicator.
+Lets say that variable2 is optimal when its at a level of 30, and values either lower or high are both bad.
+
+```{r}
+dat$referenceMid2 <- 30
+```
+
+```{r}
+dat$indicator2_twoSided <- ifelse(dat$variable2<dat$referenceMid2,
+                                  (dat$variable2-dat$referenceLow2)/(dat$referenceMid2-dat$referenceLow2),
+                                  ((dat$variable2-dat$referenceMid2)/(dat$referenceHigh2-dat$referenceMid2))*(-1)+1
+                                  )
+
+# truncating
+dat$indicator2_twoSided[dat$indicator2_twoSided<0] <- 0
+dat$indicator2_twoSided[dat$indicator2_twoSided>1] <- 1
+```
+
+```{r}
+ggplot(dat)+
+  geom_point(
+    aes(x = variable2, y = indicator2_twoSided, fill = indicator2_twoSided),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_vline(xintercept = dat$referenceMid2[1],
+             linetype = "dashed",
+             size =2)+
+  geom_vline(xintercept = dat$referenceLow2[1],
+             linetype = "dashed",
+             size =2)+
+  geom_vline(xintercept = dat$referenceHigh2[1],
+             linetype = "dashed",
+             size =2)+
+
+  ylab("Indicator values")+
+  xlab("Variable values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+```
+
+## Sigmoid transformation
+There are many other, non-linear ways to normalise an indicator.
+The correct choice for scaling function varies between indicators and depends on the nature of the indicator and what it represents, but also on the uncertainty around the reference values.
+Personally, I think the sigmoid function often makes a lot a sense.
+It is less sensitive to changes around the min and max reference values.
+A slight deviation from the reference condition can often not be described as a true reduction in condition.
+This is true for example when reference communities (ecosystemns defines as having or representing good condition) themselves show variation in the variable values, yet the reference value is fixed.
+A linear scaling will in these cases lead to what is referred to as underestimation biaz (making thing look worse than they really are).
+
+The equation for the sigmoid transformation that I use here comes from [Oliver at al. (2021)](https://www.sciencedirect.com/science/article/pii/S1470160X21000066); Eq. 1, Supplemetary Information S2.
+
+```{r}
+dat$indicator1_sigmoid <- 100.68*(1-exp(-5*(dat$indicator1_LowHigh)^2.5))/100
+```
+
+Finding the transformed indicator value when the variable equals the threshold
+```{r}
+sigmoid_thr <- 100.68*(1-exp(-5*((dat$thr1[1]-dat$referenceLow1[1])/(dat$referenceHigh1[1]-dat$referenceLow1[1]))^2.5))/100
+
+```
+
+```{r}
+ggplot(dat)+
+  geom_point(
+    aes(x = variable1, y = indicator1_sigmoid, fill = indicator1_sigmoid),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+
+  geom_vline(xintercept = dat$referenceLow1[1],
+             linetype = "dashed",
+             size =2)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
+             linetype = "dashed",
+             size =2)+
+
+  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = sigmoid_thr),
+               linetype="solid",
+               size=1)+
+  geom_segment(aes(x = 0, y = sigmoid_thr, xend = thr1, yend = sigmoid_thr),
+               linetype="solid",
+               size=1)+
+
+  ylab("Indicator values")+
+  xlab("Variable values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+```
+
+The solid helperlines point to the indicator value where the variable value equals the threshold. Later we will shift the values so that the indicator value becomes 0.6 at this point.
+
+Also plotted against the linearly normalised indicator values:
+```{r}
+ggplot(dat)+
+  geom_point(
+    aes(x = indicator1_LowHigh, y = indicator1_sigmoid, fill = indicator1_sigmoid),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+
+  geom_vline(xintercept = 0.6,
+             linetype = "dashed",
+             size =2)+
+   geom_hline(yintercept = 0.6,
+             linetype = "dashed",
+             size =2)+
+
+  ylab("Sigmoid-transformed\nIndicator values")+
+  xlab("Linear indicatior values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+```
+
+### Accounting for the threshold value
+
+We can account for the  threshold for good ecological condition when applying sigmoid transformation like this
+```{r}
+dat$indicator1_sigmoid_thr <- ifelse(dat$variable1<dat$thr1,
+                                     ((dat$indicator1_sigmoid-0)/(sigmoid_thr-0))*(0.6-0)+0,
+                                     ((dat$indicator1_sigmoid-sigmoid_thr)/(1-sigmoid_thr))*(1-0.6)+0.6
+                                     )
+ggplot(dat)+
+  geom_point(
+    aes(x = variable1, y = indicator1_sigmoid),
+             fill   = "grey", 
+             colour = myColour,
+             size   = 5,
+             alpha  = myAlpha,
+             shape  = myShape)+
+
+  geom_point(
+    aes(x = variable1, y = indicator1_sigmoid_thr, fill = indicator1_sigmoid_thr),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+
+  geom_vline(xintercept = dat$referenceLow1[1],
+             linetype = "dashed",
+             size =1)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
+             linetype = "dashed",
+             size =1)+
+
+  geom_segment(aes(x = 0, y = sigmoid_thr, xend = thr1, yend = sigmoid_thr),
+               linetype="solid",
+               colour = "grey",
+               size=1)+
+
+  geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
+               linetype="solid",
+               size=1)+
+  geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
+               linetype="solid",
+               size=1)+
+
+  ylab("Indicator values")+
+  xlab("Variable values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+
+```
+
+## Exponential functoins
+Here are two example of exponential function:
+
+```{r}
+dat$indicator1_expConvex <- dat$indicator1_LowHigh^0.5
+dat$indicator1_expConcave <- dat$indicator1_LowHigh^2
 ```
@@
 
+ggplot(dat)+
+  geom_point(
+    aes(x = variable1, y = indicator1_expConvex, fill = indicator1_expConvex),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+  geom_point(
+    aes(x = variable1, y = indicator1_expConcave, fill = indicator1_expConcave),
+             colour = myColour,
+             size   = mySize,
+             alpha  = myAlpha,
+             shape  = myShape)+
+
+  geom_vline(xintercept = dat$referenceLow1[1],
+             linetype = "dashed",
+             size =1)+
+  geom_vline(xintercept = dat$referenceHigh1[1],
+             linetype = "dashed",
+             size =1)+
+  ylab("Indicator values")+
+  xlab("Variable values")+
+
+  scale_fill_gradient("Indicator values", low = low, high = high)
+
 ```

### 2022-09-14 naturtype.Rmd
hovedøkosystem and Nin match

@@
-# (PART\*) GENERAL TOPICS {-}
+# (PART\*) GENERAL TOPICS {.unnumbered}
 
@@
 library(ggplot2)
+library(data.table)
+library(stringr)
 library(units)
@@
 ```
+
+*Work in progress*
 
@@
 
-* Total mapped area
-* Temporal trend
-* Histogram of conditions (shiny?)
-
 ```{r}
 # local path
-path <- "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+#path <- "data/R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+
+# server path
+#path <- "/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+
+# temporary path 
+path <- "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+
 dat <- sf::st_read(dsn = path)
@@
 ```
+
+Fix non-valid polygons
+
+```{r}
+dat <- sf::st_make_valid(dat)
+```
+
 Import outline of Norway
+
 ```{r}
@@
 The dataset has 95k polygons, each with 36 variables:
+
 ```{r}
@@
 ```
-It's therefore takes a little minute to render a plot, but this is the code to do it:
 
-```{r, eval = T}
+It therefore takes a little minute to render a plot, but this is the code to do it:
+
+```{r, eval = F}
 tmap_mode("view")
@@
 ### Area
+
 Calculating the area
@@
 
-The smallest polygons are 20 m2, and the biggset is 8.5 km2
+The smallest polygons are 20 m2, and the biggest is 8.5 km2.
 
 Sum of mapped area divided by Norwegian mainland area
+
 ```{r}
@@
 
-About 0.5% has been mapped.
-It is therefore essential that these 0.5% are representative (of the remaining 99.5%).
+About 0.5% has been mapped. It is therefore essential that these 0.5% are representative (of the remaining 99.5%).
+
+### Temporal trend
+
+```{r}
+dat$kartleggingsår <- as.numeric(dat$kartleggingsår)
+```
+
+```{r}
+area_year <- as.data.frame(tapply(dat$area, dat$kartleggingsår, sum))
+names(area_year) <- "area"
+area_year$year <- row.names(area_year)
+area_year$area_km2 <- area_year$area/10^6
+```
+
+```{r area_year, fig.cap="Temproal trend in nature type mapping using Miljødirektoratets Instruks"}
+ggplot(area_year)+
+  geom_bar(aes(x = year, y = area_km2),
+           stat = "identity",
+           fill = "grey",
+           colour="black",
+           size = 2)+
+  theme_bw(base_size = 12)+
+  labs(x = "Year", y = "Total mapped area (km/2)")
+```
+
+The rate of nature type mapping is strongly positive last three years.
+There are some differences in the field mapping instructoin between the years.
+I will need to decide whether to include all years, or to perhaps exclude the first year.
+
+### Condition
+
+A quick overview of the conditions
+
+```{r tilstand_overview, fig.cap="The overal distribution of condition scores"}
+barplot(tapply(dat$area/10^6, factor(dat$tilstand), sum))
+```
+
+Most sites/polygons are in either good or moderately good condition.
+I'm not sure what thet first class represents.
+It seems like som polygons dont have a condition score.
+Looking at just the dataset, and also the online *faktaark* for some of these localities, does not give the answer.
+
+```{r, eval=F}
+View(dat[dat$tilstand=="",])
+```
+
+This figure also show that these cases are not restricted to just one field season.
+```{r}
+barplot(table(dat$kartleggingsår[dat$tilstand==""]), 
+        ylab="Number of localities without condition scores",
+        xlab="Year")
+```
+
+```{r}
+par(mar=c(5,20,1,1))
+barplot(table(dat$hovedøkosystem[dat$tilstand==""]),
+        horiz = T, las=2,
+        xlab="Number of localities without condition scores")
+
+```
+
+These cases are restricted to two main ecosystems, and one class where the main ecosystem is not recorded.
+Look at some of those cases it is clear that they are not relevant for our work here, I and I dont know why they are in the data set to begin with.
+
+```{r}
+dat <- dat[dat$hovedøkosystem!="Ingen",]
+```
+
+```{r}
+par(mar=c(5,20,1,1))
+barplot(table(dat$naturtype[dat$tilstand==""]),
+        horiz = T, las=2,
+        xlab="Number of localities without condition scores")
+
+```
+
+After excluding those without a main ecosystem, there are only six nature types left that don't have a condition score.
+
+Snøleieblokkmark and rabbeblokkmark are not assessed for condition status.
+Leirskredgrop,,leirravine and grotte were nature types in 2018 (not mapped in 2021), and was not assessed for condition scores.
+Isinnfrysingsmark is assessed for condition now, but not in 2018.
+We can therefore exclude these localities from our data set.
+
+```{r}
+dat <- dat[dat$tilstand!="",]
+```
+
+### Mosaic types
+
+The field mapping instruction include and option for delineating mosaic types. Let's investigate these cases.
+When an area displays a repeating pattern of mixed nature types that each are smaller than the minimum mapping unit MMU, these are grouped into as many overlapping polygons as there are uniqe nature types.
+Within the nature type polyogons, these will have the same distribution of NiN-grunntyper (online you can see the percentage distribution, but our data set only has the presence/absence data) but be assigned different nature types (nature types is the Environmental agencies classification).
+The condition scoring can be unique to each overlapping nature type in the mosaic.
+But we don't know the precise location of the NiN-grunntyper that are part of mosaic nature types.
+
+```{r}
+barplot(table(dat$mosaikk),
+        xlab = "Mosaic", ylab="Number of localities")
+```
+
+```{r}
+mos <- dat[dat$mosaikk=="Ja",]
+```
+
+```{r}
+par(mar=c(5,20,1,1))
+barplot(table(mos$hovedøkosystem),
+        horiz = T, las=2,
+        xlab="Number of mosaic localities")
+```
+
+Mosaic localities occur in many main ecosystems (and many nature types therein).
+
+Some conclusion here could be that
+
+1. Mosaic localities cannot be used to pin-point NiN grunntyper. 
+1.
+Mosaic localities can be used to extract condition scores for nature types, but these should not be ties to all the NiN grunntype listed in that polygon, because that will include some that belong to the other part of the mosaic.
+
+### NiN types across main ecosystems
+
+The NiN-types can be extracted from the column `ninkartleggingsenheter`, but it is oddly concatenated. Let's tease it apart.
+
+```{r}
+dat$ninkartleggingsenheter2 <- gsub("NA_", dat$ninkartleggingsenheter, replacement = "")
+dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, ".[CE].[\\d]{1,}", replacement = "")
+dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, "-.", replacement = "")
+uni <- function(x){paste(unique(unlist(strsplit(x, ","))), collapse = ",")}
+dat$ninkartleggingsenheter3 <- lapply(dat$ninkartleggingsenheter2, uni)
+n_uni <- function(x){length(unique(unlist(strsplit(x, ","))))}
+dat$n_ninkartleggingsenheter <- lapply(dat$ninkartleggingsenheter2, n_uni)
+dat$n_ninkartleggingsenheter <- as.numeric(dat$n_ninkartleggingsenheter)
+
+par(mfrow=c(1,3))
+barplot(table(dat$n_ninkartleggingsenheter),
+        xlab = "Number of NiN main types",
+        ylab = "Number of localities")
+barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Nei"]),
+        xlab = "Number of NiN main types\n(Mosaic localities excluded)")
+barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Ja"]),
+        xlab = "Number of NiN main types\n(Mosaic localities only)")
+
+```
+
+Mosaic localities have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (about 8000), so we need to see whats going on there.
+
+```{r}
+dat_melt <- tidyr::separate_rows(dat, ninkartleggingsenheter3)
+```
+
+```{r}
+ggplot(dat_melt[dat_melt$mosaikk!="Ja",])+
+  geom_bar(aes(x = ninkartleggingsenheter3))+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free")
+
+```
+
+There is a lot of miss-match between NiN main types and 'hovedøkosystem'. 
+Taking one example. T2 (åpne grunnledt mark) is not found in mountains, but there is one case where this occurs. If I view that case
+
+```{r, eval= FALSE}
+View(dat_melt[dat_melt$mosaikk!="Ja" & dat_melt$hovedøkosystem=="Fjell" & dat_melt$ninkartleggingsenheter3=="T2",])
+```
+
+..
+and find the online fact sheet for that locality, I see that it is indeed not a mosaic.
+It is a nature type called *Kalkfattig og intermediær fjellhei, leside og tundra* which is defined as strictly T3, but the field surveyor has recorded lots of main nature types.
+This is a mistake.
+Online I can see that the locality is 50% T3, but this information is not in the data set that we have available.
+And the order of the NiN types in the `ninkartleggingsenhete` column is not reflecting the commonness of the types ether.
+
+If there was a way to extract the defining NiN type from the `naturtype` column, that would be nice.
+If we exclude all localities that are not mosaics, but that have multiple NiN main types, we fisr need to know if there are not any `naturtype` which allow for multiple NiN main types.
+
+As this next figure show, these cases are not correlated to mapping year, and hence to any changes in the field protocol.
+```{r}
+temp <- dat[dat$n_ninkartleggingsenheter>1 & dat$mosaikk!="Ja",]
+barplot(table(temp$kartleggingsår), ylab="Number of localities ")
+```
+
+```{r}
+temp2 <- as.data.frame(table(temp$naturtype, temp$n_ninkartleggingsenheter))
+temp2 <- temp2[base::order(temp2$Freq, decreasing = T),][1:10,]
+par(mar=c(8,20,1,1))
+barplot(temp2$Freq, names.arg = temp2$Var1, las=2,
+        horiz = T, xlab = "Number of localities with >1 NiN main type")
+```
+
+'Hule eiker' is a special case because these are recorded as points and not polygons, and they can be found in any NiN main type and any *hovedøkosystem*.
+
+```{r}
+dat[dat$naturtype=="Hule eiker",][1:5,c("naturtype", "hovedøkosystem", "ninkartleggingsenheter")]
+```
+
+*Kystlynghei* often occurs as a moisaic, but the figure above has excluded the mosaic localities, and then *kystlynghei* is stricktly defined to T34.
+```{r}
+dat[dat$naturtype=="Kystlynghei",][1:5,c("naturtype", "hovedøkosystem", "ninkartleggingsenheter")]
+
+```
+
+The same is true for *naturbeitemark* which is restricted to T32.
+
+Maybe these should all have been recorded as mosaic localities.
+In any case, we cannot automatically extract the main NiN type from these rows.
+One option is to manually define all *naturbeitemark* as T32, and all *kystlynghei* as T34, etc.
+
+Lets look at how the NiN main types cover the different *hovedøkosystem* when we exclude *Hule eiker* and all localities that have more than one NiN main type.
+
+```{r}
+dat_red <- dat_melt[dat_melt$n_ninkartleggingsenheter==1 & dat_melt$naturtype != "Hule eiker",]
+```
+
+```{r}
+  ggplot(dat_red)+
+  geom_bar(aes(x = ninkartleggingsenheter3),
+           fill="grey",
+           colour="black")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free")+
+  labs(y = "Number of localities",
+       x = "NiN main types")
+
+```
+
+A reduced version just looking at the three target main ecosystems:
+```{r}
+target <- c("Semi-naturlig mark",
+            "Våtmark",
+            "Naturlig åpne områder under skoggrensa")
+```
+
+```{r}
+  ggplot(dat_red[dat_red$hovedøkosystem %in% target,])+
+  geom_bar(aes(x = ninkartleggingsenheter3),
+           fill="grey",
+           colour="black")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free",
+             ncol = 3)+
+  labs(y = "Number of localities",
+       x = "NiN main types")
+
+```
+
+Still some weird cases...
+
+```{r}
+dat_red_ordered <- dat_red[base::order(dat_red$ninkartleggingsenheter3),] %>%
+  dplyr::mutate(ninkartleggingsenheter_f=
+                  factor(dat_red_ordered$ninkartleggingsenheter3, 
+                      levels(dat_red_ordered$ninkartleggingsenheter3)))
+
+```
+
+```{r}
+ggplot(dat_red)+
+  geom_bar(aes(x = ninkartleggingsenheter3),
+           fill="grey",
+           colour="black")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free")
+
+```
 
 ## Overview of nature types
-* What are the dominant ecosystem groups recorded. 
+
+-   What are the dominant ecosystem groups recorded.
+
 ```{r}
@@
 This project is concerned only with wetlands (våtmark), semi-natural areas (semi-naturlig mark) and naturally open areas below the forest line (naturlig åpne områder i lavlandet/under skoggrensa).
+
 ```{r}
@@
 temp <- data.frame(eco = row.names(temp),
-                   area = temp$temp)
+                   area = temp)
 ggplot(temp, aes(x = eco, y=area))+
@@
 
-* What are the dominant nature types recorded.
+-   What are the dominant nature types recorded.
 
@@
 temp <- dat2[dat2$hovedøkosystem=="Våtmark",]
-ggplot(temp, 
-       aes(x = naturtype,
-           y = area))+
-  geom_bar(stat = "sum")+
-  coord_flip()
 
@@
 
-temp <- temp[order(temp$area),]
 ggplot(temp[order(temp$area),], 
@@
 
-* Can we group these into main ecosystem type categories?
-* Are there any regional differences
+-   Can we group these into main ecosystem type categories?
+-   Are there any regional differences
 
@@
 
-* Extract environmental data from the polygons and compare against reference data points.
-* Uni-variate (e.g. _boreal hei_ plotted against elevation, oceanity, distance from roads, etc)
+-   Extract environmental data from the polygons and compare against reference data points.
+-   Uni-variate (e.g. *boreal hei* plotted against elevation, oceanity, distance from roads, etc)
 
 ## Represntaivity in mapping units
+
 Only a subset of nature types are mapped. How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?
@@
 ## Conclution 1
+
 Conclude about which main ecosystem types that can reliably assess using the available mapping units?
@@
 ## Suitability
+
 In the subset of mapping units that can be said to represent an ecosystem group, which NiN or MdirPRAM variables should we use?
@@
 ## Concrete examples
+
 ### Andel lynghei med minst to faser i kystlynghei

### 2022-09-15 naturtype.Rmd
naturtype part 1

--- 2022-08-18 naturtype.Rmd
@@
-# (PART\*) GENERAL TOPICS {-}
+# (PART\*) GENERAL TOPICS {.unnumbered}
 
-# On the application of the naturetype dataset {#naturtype}
+# On the application of the naturetype dataset - Part 1 - Data exploration and cleaning {#naturtype1}
 
@@
 Here I will investigate a specific data set, [Naturtyper - Miljødirektoratets instruks](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c) and try to evaluate its usability for designing indicators for ecosystem condition.
-This involves assessing both the spatial and temporal representativity, as well as conseptual alignment with the [Norwegian system for ecosystem condition assessments](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/).
+This involves assessing both the spatial and temporal representation, as well as conseptual alignment with the [Norwegian system for ecosystem condition assessments](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/).
 
@@
 
+The first part of this analysis is simply to get an overview of the data and to clean it, making it ready for part 2.
+
 ## Import data and general summary statistics
+```{r import_nin_data, results="hide"}
+# local path
+#path <- "data/R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+# server path
+#path <- "/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
 
-* Total mapped area
-* Temporal trend
-* Histogram of conditions (shiny?)
+# temporary path 
+path <- "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+dat <- sf::st_read(dsn = path)
+
+```
+
+Fix non-valid polygons
 
 ```{r}
-# local path
-path <- "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
-dat <- sf::st_read(dsn = path)
-names(dat)
-```
-Import outline of Norway
-```{r}
-nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
+dat <- sf::st_make_valid(dat)
 ```
@@
 ```
-It's therefore takes a little minute to render a plot, but this is the code to do it:
 
-```{r, eval = T}
+It therefore takes a little minute to render a plot, but this is the code to do it:
+```{r, eval = F}
+nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
 tmap_mode("view")
@@
 
-### Area
-Calculating the area
-
+## Area
+Calculating the area for each polygon/locality
 ```{r}
@@
 ```
+The smallest polygons are 20 m2, and the biggest is 8.5 km2.
 
-The smallest polygons are 20 m2, and the biggset is 8.5 km2
-
-Sum of mapped area divided by Norwegian mainland area
+Sum of mapped area divided by Norwegian mainland area:
 ```{r}
+nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
 sum(dat$area)/sf::st_area(nor)
 ```
+About 0.5% of Norway has been mapped (not that a bigger area than this has been surveyed, but only a small fraction is delineated). It is therefore essential that these 0.5% are representative.
 
-About 0.5% has been mapped.
-It is therefore essential that these 0.5% are representative (of the remaining 99.5%).
-
-## Overview of nature types
-* What are the dominant ecosystem groups recorded. 
+## Temporal trend
 ```{r}
-ggplot(dat, aes(x = hovedøkosystem))+
-  geom_bar()+
-  coord_flip()
-```
-
-This project is concerned only with wetlands (våtmark), semi-natural areas (semi-naturlig mark) and naturally open areas below the forest line (naturlig åpne områder i lavlandet/under skoggrensa).
-```{r}
-dat$hovedøkosystem[dat$hovedøkosystem == "Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+dat$kartleggingsår <- as.numeric(dat$kartleggingsår)
 ```
@@
 ```{r}
-unique(dat$hovedøkosystem)
-utvlagt <- c("Semi-naturlig mark",
-             "Våtmark",
-             "Naturlig åpne områder under skoggrensa")
-dat2 <- dat[dat$hovedøkosystem %in% utvlagt,]
+area_year <- as.data.frame(tapply(dat$area, dat$kartleggingsår, sum))
+names(area_year) <- "area"
+area_year$year <- row.names(area_year)
+area_year$area_km2 <- area_year$area/10^6
+```
+
+```{r area_year, fig.cap="Temproal trend in nature type mapping using Miljødirektoratets Instruks"}
+ggplot(area_year)+
+  geom_bar(aes(x = year, y = area_km2),
+           stat = "identity",
+           fill = "grey",
+           colour="black",
+           size = 1.2)+
+  theme_bw(base_size = 12)+
+  labs(x = "Year", y = "Total mapped area (km/2)")
+```
+
+The rate of nature type mapping is strongly positive last three years.
+There are some differences in the field mapping instructions between the years.
+I will need to decide whether to include all years, or to perhaps exclude the first year.
+
+## Condition
+A quick overview of the condition scores 
+```{r tilstand_overview, fig.cap="The overal distribution of condition scores"}
+barplot(tapply(dat$area/10^6, factor(dat$tilstand), sum))
+```
+
+Most sites/polygons are in either good or moderately good condition.
+I'm not sure what the first class represents.
+It seems like some polygons don't have a condition score.
+Looking at just the data set, and also the online *faktaark* for some of these localities, does not give the answer:
+```{r, eval=F}
+View(dat[dat$tilstand=="",])
+```
+
+This figure show that these cases are not restricted to just one field season.
+```{r}
+barplot(table(dat$kartleggingsår[dat$tilstand==""]), 
+        ylab="Number of localities without condition scores",
+        xlab="Year")
 ```
@@
 ```{r}
-temp <- tapply(dat2$area, dat2$hovedøkosystem, sum)
-temp <- data.frame(eco = row.names(temp),
-                   area = temp$temp)
-ggplot(temp, aes(x = eco, y=area))+
-  geom_bar(stat="identity")+
-  coord_flip()
+par(mar=c(5,20,1,1))
+barplot(table(dat$hovedøkosystem[dat$tilstand==""]),
+        horiz = T, las=2,
+        xlab="Number of localities without condition scores")
+
 ```
 
-The same bias towards sami-natural areas are seen when looking at the area mapped, as when looking the number of polygons.
+These cases are restricted to two main ecosystems, and one class where the main ecosystem is not recorded.
+Looking at some of those cases it is clear that they are not relevant for our work here, I and I don't know why they are in the data set to begin with.
+```{r}
+par(mar=c(5,20,1,1))
+barplot(table(dat$naturtype[dat$tilstand==""]),
+        horiz = T, las=2,
+        xlab="Number of localities without condition scores")
 
-* What are the dominant nature types recorded.
+```
+
+There are only six nature types (if you exclude those that are not mapped) that don't have a condition score.
+
+`Snøleieblokkmark` and `rabbeblokkmark` do not have a protocol for assessing condition status.
+`Leirskredgrop`, `leirravine` and `grotte` were nature types in 2018 (not mapped in 2021), and was similarly not assessed for condition scores.
+`Isinnfrysingsmark` is assessed for condition now, but not in 2018.
+We can therefore exclude these localities from our data set.
+```{r}
+dat <- dat[dat$tilstand!="",]
+```
+
+## Mosaic types
+The field mapping instruction include and option for delineating mosaic types. Let's investigate these cases.
+
+When an area displays a repeating pattern of mixed nature types that each are smaller than the minimum mapping unit MMU, these are grouped into as many overlapping polygons as there are unique nature types.
+Within the nature type polygons, these will have the same distribution of NiN-grunntyper (online you can see the percentage distribution, but our data set only has the presence/absence data) but be assigned different nature types (nature types is the Environmental agencies classification).
+The condition scoring can be unique to each overlapping nature type in the mosaic.
+But we don't know the precise location of the NiN-grunntyper that are part of mosaic nature types.
 
 ```{r}
+barplot(table(dat$mosaikk),
+        xlab = "Mosaic", ylab="Number of localities")
+```
 
-temp <- dat2[dat2$hovedøkosystem=="Våtmark",]
-ggplot(temp, 
-       aes(x = naturtype,
-           y = area))+
-  geom_bar(stat = "sum")+
-  coord_flip()
+Fix duplicated `hovedøkosystem`
+```{r}
+#unique(dat$hovedøkosystem)
+dat$hovedøkosystem[dat$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+```
 
-temp <- tapply(temp$area, temp$naturtype, sum)
+```{r, fig.cap=Distribution of mosaic nature types across hovedøkosystem"}
+par(mar=c(5,20,1,1))
+barplot(table(dat$hovedøkosystem[dat$mosaikk=="Ja"]),
+        horiz = T, las=2,
+        xlab="Number of mosaic localities")
+```
+
+Mosaic localities occur in many main ecosystems (and many nature types therein).
+
+Some conclusion here could be that
+
+1. Mosaic localities CANNOT be used to pin-point NiN grunntyper (e.g. for remote sensing purposes).
+1.
+Mosaic localities CAN be used to extract condition scores for nature types, but these should not be tied to all the *NiN grunntype* listed in that polygon, because that will include some that belong to the other part(s) of the mosaic.
+
+## NiN types across main ecosystems
+The NiN main types can be extracted from the column `ninkartleggingsenheter`, but it is oddly concatenated. Let's tease it apart.
+
+```{r, fig.cap="The number of NiN main types within one locality should normally be one, except for mosaic localities."}
+#dat$ninkartleggingsenheter[1:10]
+dat$ninkartleggingsenheter2 <- gsub("NA_", dat$ninkartleggingsenheter, replacement = "")
+#dat$ninkartleggingsenheter2[1:30]
+dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, ".[CE].[\\d]{1,}", replacement = "")
+#dat$ninkartleggingsenheter2[1:30]
+dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, "-.", replacement = "")
+uni <- function(x){paste(unique(unlist(strsplit(x, ","))), collapse = ",")}
+dat$ninkartleggingsenheter3 <- lapply(dat$ninkartleggingsenheter2, uni)
+n_uni <- function(x){length(unique(unlist(strsplit(x, ","))))}
+dat$n_ninkartleggingsenheter <- lapply(dat$ninkartleggingsenheter2, n_uni)
+dat$n_ninkartleggingsenheter <- as.numeric(dat$n_ninkartleggingsenheter)
+
+par(mfrow=c(1,3))
+barplot(table(dat$n_ninkartleggingsenheter),
+        xlab = "Number of NiN main types",
+        ylab = "Number of localities")
+barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Nei"]),
+        xlab = "Number of NiN main types\n(Mosaic localities excluded)")
+barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Ja"]),
+        xlab = "Number of NiN main types\n(Mosaic localities only)")
+
+```
+
+Mosaic localities have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (about 8000 cases), so we need to see whats going on there.
+
+First I need to melt the data frame in order to tally the number of NiN min types within each `hovedøkosystem`.
+```{r}
+dat_melt <- tidyr::separate_rows(dat, ninkartleggingsenheter3)
+```
+
+```{r, fig.cap="Figure showing the number of localities for each NiN main type. One locality can have more than one NiN main type. Mosaic localities are excluded in this figure."}
+ggplot(dat_melt[dat_melt$mosaikk!="Ja",])+
+  geom_bar(aes(x = ninkartleggingsenheter3))+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free")
+
+```
+
+There is a lot of miss-match between NiN main types and *hovedøkosystem*. 
+Taking one example: T2 (åpne grunnledt mark) is not found in mountains, but there is one case where this occurs. If I view that case
+
+```{r, eval= FALSE}
+View(dat_melt[dat_melt$mosaikk!="Ja" & dat_melt$hovedøkosystem=="Fjell" & dat_melt$ninkartleggingsenheter3=="T2",])
+```
+
+..
+and find the online fact sheet for that locality, I see that it is indeed not a mosaic.
+It is a nature type called *Kalkfattig og intermediær fjellhei, leside og tundra* which is defined as strictly T3, but the field surveyor has recorded lots of main NiN types.
+This is a mistake.
+Online I can see that the locality is 50% T3, but this information is not in the data set that we have available.
+And the order of the NiN types in the `ninkartleggingsenheter` column is not reflecting the commonness of the types ether.
+
+If there was a way to extract the defining NiN type from the `naturtype` column, that would be nice.
+If we exclude all localities that are not mosaics, but that have multiple NiN main types, we first need to know if there are not any `naturtype` which allow for multiple NiN main types.
+
+As this next figure show, these cases are not correlated to mapping year, and hence to any changes in the field protocol.
+```{r}
+par(mar=c(2,6,2,2))
+temp <- dat[dat$n_ninkartleggingsenheter>1 & dat$mosaikk!="Ja",]
+barplot(table(temp$kartleggingsår), ylab="Number of localities with more than one NiN main type\n(mosaic localitie excluded) ")
+```
+
+```{r multipleNiN}
+temp2 <- as.data.frame(table(temp$naturtype, temp$n_ninkartleggingsenheter))
+temp2 <- temp2[base::order(temp2$Freq, decreasing = T),][1:10,]
+par(mar=c(8,20,1,1))
+barplot(temp2$Freq, names.arg = temp2$Var1, las=2,
+        horiz = T, xlab = "Number of localities with >1 NiN main type")
+```
+
+'Hule eiker' is a special case because these are recorded as points and not polygons, and they can be found in any NiN main type and any *hovedøkosystem*. 
+```{r, fig.cap="Table of the first 5 cases of Hule eiker, showing that this nature type can be found in several NiN main types and hovedøkosystem"}
+DT::datatable(
+  dat[dat$naturtype=="Hule eiker",][1:5,c("naturtype", "hovedøkosystem", "ninkartleggingsenheter")])
+
+```
+
+*Kystlynghei* often occurs as a mosaic, but figure \@ref(fig:multipleNiN) has excluded the mosaic localities, and then *kystlynghei* is strictly defined to T34.
+The same is true for *naturbeitemark* which is restricted to T32.
+Maybe these should all have been recorded as mosaic localities.
+In any case, we cannot automatically extract the main NiN type from these rows now.
+One option is to manually define all *naturbeitemark* as T32, and all *kystlynghei* as T34, but this is also not without risk.
+
+Lets look at how the NiN main types cover the different *hovedøkosystem* when we exclude *Hule eiker* and all localities that have more than one NiN main type.
+```{r}
+dat_red <- dat_melt[dat_melt$n_ninkartleggingsenheter==1 & 
+                      dat_melt$naturtype != "Hule eiker" &
+                      dat_melt$mosaikk == "Nei",]
+```
+
+```{r}
+ggplot(dat_red)+
+  geom_bar(aes(x = ninkartleggingsenheter3),
+           fill="grey",
+           colour="black")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  facet_wrap("hovedøkosystem",
+             scales = "free")+
+  labs(y = "Number of localities",
+       x = "NiN main types")
+
+```
+
+Let's focus in on our three targeted ecosystems.
+```{r}
+dat_red_tally <- stats::aggregate(data = dat_red,
+                                  area~hovedøkosystem+ninkartleggingsenheter3, FUN = length)
+names(dat_red_tally)[3] <- "count"
+```
+
+```{r}
+target <- c("Semi-naturlig mark",
+            "Våtmark",
+            "Naturlig åpne områder under skoggrensa")
+```
+
+```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type", fig.width= 8}
+ggplot(dat_red_tally[dat_red_tally$hovedøkosystem %in% target,],
+       aes(x = ninkartleggingsenheter3,
+               y = count))+
+  geom_bar(
+           fill="grey",
+           colour="black",
+           stat="identity")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  labs(y = "Number of localities",
+       x = "NiN main types")+
+  scale_y_continuous(position = "left",
+                   expand = expansion(mult=c(.0,.3)))+
+  geom_text(aes(label=count), hjust=-0.25)+
+    facet_wrap("hovedøkosystem",
+             scales = "free",
+             ncol = 3)
+```
+
+Still some weird cases which I will look at in turn below.
+
+**Naturlig åpne områder under skoggrensa:**
+1. V10 (semi-naturlig våteng)
+  + One case of Åpen flomfastmark wrongly associated with V10 whan it should be T18.
+  + **Exclude** this case
+1. T8 (fulglefjelleng og fugltopp)
+  + Ok
+1. T6 (Standberg)
+  + Ok
+1. T33 (Semi naturlig strandeng)
+  + These are cases where either the `naturtype` should have been `semi-naturlig strandeng` (and hence the hovedøkosystem would be Semi-nat), or the `naturtype` is correctly recorded as `Strandeng` but then the NiN type should be `T12`.
+  + **Exclude**
+1. T32 (Semi-naturlig eng)
+  + Obvious error
+  + **Exclude**
+1. T30 (Flomskogsmark)
+  + Obvious error
+  + **Exclude**
+1. T29 (Grus og steindominert strand og strandlinje)
+  + Ok
+1. T21 (Sanddynemark)
+  + Ok
+1. T20 (Isinnfrysingsmark)
+  + Ok. Rare nature type
+1. T2 (Åpen grunlendt mark)
+  + Ok
+1. T18 (Åpne flomfastmark)
+  + Ok
+1. T17 (Aktiv skredmark)
+  + Ok
+1. T15 (Fosse-eng)
+  + Ok
+1. T13 (Rasmark)
+  + Ok
+1. T12 (Strandeng)
+  + Ok
+1. T1 (Nakent berg)
+  + Ok
+
+```{r}
+#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "V10",])
+#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "T33",])
+#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "T32",])
+
+`%!in%` <- Negate(`%in%`)
+dat_red2 <- dat_red[
+  dat_red$hovedøkosystem!="Naturlig åpne områder under skoggrensa" |
+  dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 %!in% c("T30", "T32", "T33", "V10"),]
+```
+
+**Semi-naturlig mark**
+
+1. V9 (Semi-naturlig myr)
+  + Kystlynghei recorded as mire
+  + **Exclude**
+1. V10 (semi-nat våteng)
+  + Ok
+1. T41 (Oppdyrket mark med preg av semi-naturlig eng)
+  + Probably not always use correctly, but at least its the correct ``hovedkøosystem`
+  + Ok
+1. T40 (Sterkt endret fastmark med preg av semi-naturlig eng)
+  + Ok
+1. T4 (Fastmarksskogsmark)
+  + These should all be T32 (naturbeitemark, lauveng, hagemark, +++)
+  + **Exclude**
+1. T35 (Sterkt endret fastmark med løsmassedekke)
+  + Wrong NiN type given to eng-aktig sterkt endret fastmark (it should be T40)
+  + **Exclude**
+1. T31-34
+  + Ok
+1. T3 (Fjellhei, leside og tundra)
+  + Boreal hei (should've been T31)
+  + **Exclude**
+1. T12 (Strandeng)
+  + Should probably have been T33
+  + **Exclude**
+1. T1 (Nakent berg)
+  + Obvious mistake
+  + **Exclude**
+
+```{r}
+#unique(dat$hovedøkosystem)
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "V9",])
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T41",])
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T4",])
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T35",])
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T3",])
+#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T1",])
+
+dat_red3 <- dat_red2[
+  dat_red2$hovedøkosystem!="Semi-naturlig mark" |
+  dat_red2$hovedøkosystem=="Semi-naturlig mark" & dat_red2$ninkartleggingsenheter3 %!in% c("V9", "T4", "T35", "T3", "T12", "T1"),]
+```
+
+**Våtmark**
+
+Våtmakr doesn't distinguish between semi-natural and natural, so all V-types are correct, except perhaps the strongly modified V11-V13
+1. V11 (Torvtak)
+  + Obvious mistake
+  + **Exclude**
+1. V12 (Grøftet åpne torvmark)
+  + Obvious mistake
+  + **Exclude**
+1. V13 (Ny våtmark)
+  + Obvious mistake
+  + **Exclude**
+1. T4 (skog)
+  +   some that should've been T2 and some other strange things
+  + **Exclude**
+1. T30-T32
+  + Someone switching the semi-natural and natural equivalents
+  + **Exclude**
+1. L4 (Helofyttsump)
+  + Ok
+
+```{r}
+#unique(dat$hovedøkosystem)
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "T4",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "T32",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "L4",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V9",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V11",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V12",])
+#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V13",])
+dat_red4 <- dat_red3[
+  dat_red3$hovedøkosystem!="Våtmark" |
+  dat_red3$hovedøkosystem=="Våtmark" & dat_red3$ninkartleggingsenheter3 %!in% c("T4", "T30", "T31", "T32", "V11", "V12", "V13"),]
+```
+
+Let's make the plot again to see if it worked
+```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type (after data set cleaning)", fig.width= 8}
+dat_red_tally2 <- stats::aggregate(data = dat_red4,
+                                 area~hovedøkosystem+ninkartleggingsenheter3, FUN = length)
+names(dat_red_tally2)[3] <- "count"
+
+ggplot(dat_red_tally2[dat_red_tally2$hovedøkosystem %in% target,],
+       aes(x = ninkartleggingsenheter3,
+               y = count))+
+  geom_bar(
+           fill="grey",
+           colour="black",
+           stat="identity")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  labs(y = "Number of localities",
+       x = "NiN main types")+
+  scale_y_continuous(position = "left",
+                   expand = expansion(mult=c(.0,.3)))+
+  geom_text(aes(label=count), hjust=-0.25)+
+    facet_wrap("hovedøkosystem",
+             scales = "free",
+             ncol = 3)
+```
+
+## Export
+`dat_red4` now has one row per locality, and only localities with 1 NiN main type.
+`Hule eiker` and `Mosaic localities` are not included, but we will add these back in now.
+Depending on what the data set will be used for, these last two can be excluded again later.
+```{r}
+exp <- rbind(dat_red4,
+             dat[dat$mosaikk=="Ja",],
+             dat[dat$naturtype== "Hule eiker",])
+```
+
+To remove `hule eiker`, use this
+```{r, eval=F}
+exp[exp$naturtype != "Hule eiker",]
+```
+
+To remove mosaic localities, first consider if you want to remove all mosaic localities, or just those that include different NiN main types, because a mosaic locality may also consist of different sub units (*grunntype*) within the same NiN main type, or even the same *grunntype* throughout, but varying in some other parameter (e.g.
+a mosaic between old pine forest and old spruce forest, which are both T4 and could both be, say T4-C-2).
+
+To exclude mosaics in general:
+```{r, eval=F}
+exp[exp$mosaikk == "Nei",]
+```
+
+To exclude only mosaics that include multiple NiN main types:
+```{r, eval=F}
+exp[exp$n_ninkartleggingsenheter == 1]
+```
+
+I will save this as an rds file to my personal folder (the shape standard truncates field names, and gdb is not soo easy to use in R), but then manually move it to 
+`/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.shp`.
+This is simply because I cannot access that drive from RStudio Server at the moment.
+
+```{r, eval=F}
+exp$ninkartleggingsenheter3 <- unlist(exp$ninkartleggingsenheter3)
+#names(exp)[1] <- "ID" # Field name too long
+saveRDS(exp, "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Mdir_Naturtyper_cleanedALK.rds")
+
+```
+
+This is the code for importing the cleaned data back into R ()
+```{r, eval=F}
+#myData <- readRDS("/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Mdir_Naturtyper_cleanedALK.rds")
+myData <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+```
+
+## Overview of nature types
+
+```{r, fig.height=15, fig.width=8, fig.cap="An overview of the nature types in the data set"}
+temp <- tapply(exp$area, exp$naturtype, sum)
 temp <- data.frame(nat = row.names(temp),
@@
 
-temp <- temp[order(temp$area),]
-ggplot(temp[order(temp$area),], 
-       aes(x = nat,
-           y = area))+
-  geom_bar(stat = "identity")+
-  coord_flip()
+ggplot(exp[exp$hovedøkosystem %in% target,], 
+       aes(x = naturtype))+
+  geom_bar(fill = "grey20",
+           size=0
+           )+
+  theme_bw(base_size = 12)+
+  coord_flip()+
+  facet_grid(rows = vars(hovedøkosystem),
+             scales = "free",
+             space = "free")+
+  labs(y = "Number of localities",
+       x = "Nature type")
 ```
-
-* Can we group these into main ecosystem type categories?
-* Are there any regional differences
-
-## Environmental space
-
-* Extract environmental data from the polygons and compare against reference data points.
-* Uni-variate (e.g. _boreal hei_ plotted against elevation, oceanity, distance from roads, etc)
-
-## Represntaivity in mapping units
-Only a subset of nature types are mapped. How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?
-
-## Conclution 1 
-Conclude about which main ecosystem types that can reliably assess using the available mapping units?
-
-What about suitability for defining reference values?
-
-## Suitability 
-In the subset of mapping units that can be said to represent an ecosystem group, which NiN or MdirPRAM variables should we use?
-
-## Concrete examples
-### Andel lynghei med minst to faser i kystlynghei
-
-### Areal uten skadet og/eller død røsslyng i kystlynghei
-
-### Aktuell bruksintensitet

### 2022-09-15 naturtype2.Rmd
naturtype part 1

# On the application of the naturetype dataset - Part 2 - Representativity {#naturtype2}

<br />

```{r, include=FALSE}
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
```

*Work in progress*

Author: Anders L. Kolstad

Date: `r Sys.Date()`

## Environmental space

-   Extract environmental data from the polygons and compare against reference data points.
-   Uni-variate (e.g. *boreal hei* plotted against elevation, oceanity, distance from roads, etc)

## Represntaivity in mapping units

Only a subset of nature types are mapped. How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?

## Conclution 1

Conclude about which main ecosystem types that can reliably assess using the available mapping units?

What about suitability for defining reference values?

## Suitability

In the subset of mapping units that can be said to represent an ecosystem group, which NiN or MdirPRAM variables should we use?

## Concrete examples

### Andel lynghei med minst to faser i kystlynghei

### Areal uten skadet og/eller død røsslyng i kystlynghei

### Aktuell bruksintensitet

### 2022-09-15 naturtype.Rmd
naturtype part 1

@@
 
-```{r area_year, fig.cap="Temproal trend in nature type mapping using Miljødirektoratets Instruks"}
+```{r, fig.cap="Temproal trend in nature type mapping using Miljødirektoratets Instruks"}
 ggplot(area_year)+
@@
 A quick overview of the condition scores 
-```{r tilstand_overview, fig.cap="The overal distribution of condition scores"}
+```{r, fig.cap="The overal distribution of condition scores"}
 barplot(tapply(dat$area/10^6, factor(dat$tilstand), sum))
@@
 This figure show that these cases are not restricted to just one field season.
-```{r}
+```{r, fig.cap="Temporal trend in localities without condition scores."}
 barplot(table(dat$kartleggingsår[dat$tilstand==""]), 
@@
 
-```{r}
+```{r, fig.cap="Main ecosystems with localities missing condition scores."}
 par(mar=c(5,20,1,1))
@@
 Looking at some of those cases it is clear that they are not relevant for our work here, I and I don't know why they are in the data set to begin with.
-```{r}
+```{r, fig.cap="Nature types with localities missing condition scores."}
 par(mar=c(5,20,1,1))
@@
 
-```{r}
+```{r, fig.cap="The number of mosaic localities is relatively small."}
 barplot(table(dat$mosaikk),
@@
 
-```{r, fig.cap=Distribution of mosaic nature types across hovedøkosystem"}
+```{r, fig.cap="Distribution of mosaic nature types across hovedøkosystem"}
 par(mar=c(5,20,1,1))
@@
 
-```{r multipleNiN}
+```{r, fig.cap="The ten most common nature types and the number of localities with with multiple NiN main types"}
 temp2 <- as.data.frame(table(temp$naturtype, temp$n_ninkartleggingsenheter))
@@
 'Hule eiker' is a special case because these are recorded as points and not polygons, and they can be found in any NiN main type and any *hovedøkosystem*. 
-```{r, fig.cap="Table of the first 5 cases of Hule eiker, showing that this nature type can be found in several NiN main types and hovedøkosystem"}
+```{r, ="Table of the first 5 cases of Hule eiker, showing that this nature type can be found in several NiN main types and hovedøkosystem"}
 DT::datatable(
@@
 
-```{r}
+```{r, fig.cap="NiN main type across five main ecosystems (Hule eiker, mosaic localities, and localities with >1 NiN main type excluded)"}
 ggplot(dat_red)+
@@
 
-```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type", fig.width= 8}
+```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type"}
 ggplot(dat_red_tally[dat_red_tally$hovedøkosystem %in% target,],
@@
 **Naturlig åpne områder under skoggrensa:**
-1. V10 (semi-naturlig våteng)
+
+* V10 (semi-naturlig våteng)
     + One case of Åpen flomfastmark wrongly associated with V10 whan it should be T18.
     + **Exclude** this case
-1. T8 (fulglefjelleng og fugltopp)
+* T8 (fulglefjelleng og fugltopp)
     + Ok
-1. T6 (Standberg)
+* T6 (Standberg)
     + Ok
-1. T33 (Semi naturlig strandeng)
+* T33 (Semi naturlig strandeng)
     + These are cases where either the `naturtype` should have been `semi-naturlig strandeng` (and hence the hovedøkosystem would be Semi-nat), or the `naturtype` is correctly recorded as `Strandeng` but then the NiN type should be `T12`.
     + **Exclude**
-1. T32 (Semi-naturlig eng)
+* T32 (Semi-naturlig eng)
     + Obvious error
     + **Exclude**
-1. T30 (Flomskogsmark)
+* T30 (Flomskogsmark)
     + Obvious error
     + **Exclude**
-1. T29 (Grus og steindominert strand og strandlinje)
+* T29 (Grus og steindominert strand og strandlinje)
     + Ok
-1. T21 (Sanddynemark)
+* T21 (Sanddynemark)
     + Ok
-1. T20 (Isinnfrysingsmark)
+* T20 (Isinnfrysingsmark)
     + Ok. Rare nature type
-1. T2 (Åpen grunlendt mark)
+* T2 (Åpen grunlendt mark)
     + Ok
-1. T18 (Åpne flomfastmark)
+* T18 (Åpne flomfastmark)
     + Ok
-1. T17 (Aktiv skredmark)
+* T17 (Aktiv skredmark)
     + Ok
-1. T15 (Fosse-eng)
+* T15 (Fosse-eng)
     + Ok
-1. T13 (Rasmark)
+* T13 (Rasmark)
     + Ok
-1. T12 (Strandeng)
+* T12 (Strandeng)
     + Ok
-1. T1 (Nakent berg)
+* T1 (Nakent berg)
     + Ok
@@
   dat_red$hovedøkosystem!="Naturlig åpne områder under skoggrensa" |
-  dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 %!in% c("T30", "T32", "T33", "V10"),]
+  dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & 
+    dat_red$ninkartleggingsenheter3 %!in% c("T30", "T32", "T33", "V10"),]
 ```
@@
 
-1. V9 (Semi-naturlig myr)
+* V9 (Semi-naturlig myr)
    + Kystlynghei recorded as mire
    + **Exclude**
-1. V10 (semi-nat våteng)
+
+* V10 (semi-nat våteng)
    + Ok
-1. T41 (Oppdyrket mark med preg av semi-naturlig eng)
+
+* T41 (Oppdyrket mark med preg av semi-naturlig eng)
    + Probably not always use correctly, but at least its the correct ``hovedkøosystem`
    + Ok
-1. T40 (Sterkt endret fastmark med preg av semi-naturlig eng)
+
+* T40 (Sterkt endret fastmark med preg av semi-naturlig eng)
    + Ok
-1. T4 (Fastmarksskogsmark)
+
+* T4 (Fastmarksskogsmark)
    + These should all be T32 (naturbeitemark, lauveng, hagemark, +++)
    + **Exclude**
-1. T35 (Sterkt endret fastmark med løsmassedekke)
+
+* T35 (Sterkt endret fastmark med løsmassedekke)
    + Wrong NiN type given to eng-aktig sterkt endret fastmark (it should be T40)
    + **Exclude**
-1. T31-34
+
+* T31-34
    + Ok
-1. T3 (Fjellhei, leside og tundra)
+
+* T3 (Fjellhei, leside og tundra)
    + Boreal hei (should've been T31)
    + **Exclude**
-1. T12 (Strandeng)
+
+* T12 (Strandeng)
    + Should probably have been T33
    + **Exclude**
-1. T1 (Nakent berg)
+
+* T1 (Nakent berg)
    + Obvious mistake
@@
 
-Våtmakr doesn't distinguish between semi-natural and natural, so all V-types are correct, except perhaps the strongly modified V11-V13
-1. V11 (Torvtak)
+Våtmakr doesn't distinguish between semi-natural and natural, so all V-types are correct, except perhaps the strongly modified V11-V13.
+
+* V11 (Torvtak)
   + Obvious mistake
   + **Exclude**
-1. V12 (Grøftet åpne torvmark)
+
+* V12 (Grøftet åpne torvmark)
   + Obvious mistake
   + **Exclude**
-1. V13 (Ny våtmark)
+
+* V13 (Ny våtmark)
   + Obvious mistake
   + **Exclude**
-1. T4 (skog)
+
+* T4 (skog)
   +   some that should've been T2 and some other strange things
   + **Exclude**
-1. T30-T32
+
+* T30-T32
   + Someone switching the semi-natural and natural equivalents
   + **Exclude**
-1. L4 (Helofyttsump)
+
+* L4 (Helofyttsump)
     + Ok
@@
 Let's make the plot again to see if it worked
-```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type (after data set cleaning)", fig.width= 8}
+```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type (after data set cleaning)"}
 dat_red_tally2 <- stats::aggregate(data = dat_red4,
@@
 
-```{r, fig.height=15, fig.width=8, fig.cap="An overview of the nature types in the data set"}
+```{r, fig.cap="An overview of the nature types in the data set", fig.height=15}
 temp <- tapply(exp$area, exp$naturtype, sum)

### 2022-09-20 naturtype.Rmd

@@
 
-**Naturlig åpne områder under skoggrensa:**
+### Naturlig åpne områder under skoggrensa
 
@@
 
-**Semi-naturlig mark**
+### Semi-naturlig mark
 
@@
 
-**Våtmark**
+### Våtmark
 
@@
 
+### Areal 
+Same type of analysis as above, but focusing onthe accumalitive mapped rea.
+```{r, fig.cap="Total mapped are for each combination of main ecosystem and NiN main type (after data set cleaning)"}
+dat_red_tally_area <- stats::aggregate(data = dat_red4,
+                                 area~hovedøkosystem+ninkartleggingsenheter3, FUN = sum)
+names(dat_red_tally_area)[3] <- "area_m2"
+dat_red_tally_area$area_km2 <- drop_units(dat_red_tally_area$area_m2/10^6)
+ggplot(dat_red_tally_area[dat_red_tally_area$hovedøkosystem %in% target,],
+       aes(x = ninkartleggingsenheter3,
+               y = area_km2))+
+  geom_bar(
+           fill="grey",
+           colour="black",
+           stat="identity")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  labs(y = "Total mapped area (km/2)",
+       x = "NiN main types")+
+    facet_wrap("hovedøkosystem",
+             scales = "free",
+             ncol = 3)
+```
+
 ## Export
@@
 exp$ninkartleggingsenheter3 <- unlist(exp$ninkartleggingsenheter3)
-#names(exp)[1] <- "ID" # Field name too long
 saveRDS(exp, "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Mdir_Naturtyper_cleanedALK.rds")

### 2022-09-20 naturtype2.Rmd
V1

@@
 
+*Work in progress*
+
 ```{r, include=FALSE}
 library(knitr)
-
+library(sf)
+library(tmap)
+library(ggplot2)
+library(data.table)
+library(stringr)
+library(units)
+library(dplyr)
 knitr::opts_chunk$set(echo = TRUE)
 ```
-
-*Work in progress*
 
@@
 
+This chapter continues with the cleaned dataset exported at the end of the previous chapter:
+```{r, eval=T}
+myData <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+```
+
+## Representativity in mapping units
+Only a subset of nature types are mapped.
+How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?
+I will tsrt by looking at wetlands (våtmark).
+
+### Våtmark
+```{r}
+wet <- myData[myData$hovedøkosystem == "Våtmark",]
+any(wet[wet$naturtype == " Hule eiker",])
+```
+
+How many mosaic localities do we have?
+```{r}
+wet_Mos <- wet[wet$mosaikk == "Ja",]
+nrow(wet_Mos)
+```
+275, compared to 
+```{r}
+wet_noMos <- wet[wet$mosaikk != "Ja",]
+nrow(wet_noMos)
+```
+118550 *normal* localities.
+
+Let's plot the summed area of each NiN main type, excluding mosaic localities.
+```{r, fig.cap="Total mapped are for each NiN main type in the wetland ecosystem (excluding mosaics)"}
+temp <- stats::aggregate(data = wet_noMos,
+                                 area~ninkartleggingsenheter3, FUN = sum)
+
+names(temp)[2] <- "area_m2"
+temp$area_km2 <- drop_units(temp$area_m2/10^6)
+
+ggplot(temp,
+       aes(x = ninkartleggingsenheter3,
+               y = area_km2))+
+  geom_bar(
+           fill="grey",
+           colour="black",
+           stat="identity")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  labs(y = "Total mapped area (km/2)",
+       x = "NiN main types")
+```
+
+If we include mosaic localities we get some other main types included as well
+```{r, fig.cap="Total mapped are for each NiN main type in the wetland ecosystem (including mosaics)"}
+wet_long <- tidyr::separate_rows(wet, ninkartleggingsenheter3)
+wet_long_agg <- stats::aggregate(data = wet_long,
+                                 area~ninkartleggingsenheter3, FUN = sum)
+names(wet_long_agg)[2] <- "area_m2"
+wet_long_agg$area_km2 <- drop_units(wet_long_agg$area_m2/10^6)
+ggplot(wet_long_agg,
+       aes(x = ninkartleggingsenheter3,
+               y = area_km2))+
+  geom_bar(
+           fill="grey",
+           colour="black",
+           stat="identity")+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  labs(y = "Total mapped area (km/2)",
+       x = "NiN main types")
+```
+
+Especially T34 (kystlynghei) and T1 (nakent berg) commonly occur in mosaics with wetlands.
+
+V5, V6 are V7 are not represented at all. V5 is *Varm kilde* which we don't have in mainland Norway. V6 and V7 might be linked to the alpine ecosystem.
+
+### V1 Åpen jordvassmyr
+
+The field recorders are allowed to themselves select the NiN types (sub units of the NiN main types) that thir locality supposedy belongs to.
+This may differ from what NEA (Norwegian Environmental Agency) uses to define the nature type.
+Therefore, in the data set, there are a lot of nature types that presumably cover, at least partly, V1 *Åpen jordvassmyr*:
+```{r}
+unique(wet$naturtype[wet$ninkartleggingsenheter3 =="V1"])
+```
+If we  look at the field protocol, we get a different picture.
+
+![alt text](images/V1.PNG)
+
+Green areas means that the entire NiN space is included in the nature type (which is bold).
+So for V1, the highest KA level (level *i*) is covered by E10 (*Rik åpne jordvassmyr*) with three sub-classes.
+For the yellow areas, only parts of the NiN spaces is mapped, and the defining parameters are in bold followingthe nature type code.
+So for KA e-f for exmaple, this NiN space is covered by E10.1 (*Rik åpen sørlig jordvassmye*), but only in boreonemoral and south boreal sones (*BN/SB*).
+The red boxes indicate a third axis in this figure related to MF (myrflate).
+The two boxes on the right are partly covered, and the two to the right are not.
+Finally, there is one more NiN type (V1-C9) written at the bottom which is partly covered (yellow outline).
+
+So, NiN main type is actually, today, only mapped through one nature type, E10.
+Some of the nature types might be old refracted, types.
+The list of nature types linked to V1 is, however, not of much use to us.
+
 ## Environmental space
@@
 -   Uni-variate (e.g. *boreal hei* plotted against elevation, oceanity, distance from roads, etc)
-
-## Represntaivity in mapping units
-
-Only a subset of nature types are mapped. How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?
 

### 2022-09-21 indicators/insectindicators/tutorials/tutorial.Rmd
first jens commit

---
title: "Showcase some database functionality"
author: "Jens Å"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  NinaR::jensAnalysis:
    highlight: tango
    fig_caption: yes
    toc: yes
---

```{r, include = F}
# Some common packages, loading rmarkdown doesn't like the messages from tidyverse, so we don't include this in the document'
require(tidyverse)
require(DBI)
require(RPostgres)
require(ggplot2)
require(xtable)
require(NinaR)
```

```{r setup, include=FALSE}
# This is optional
# I choose the 'styler' package for tidying the code to preserve indentations
# I set the cutoff for code tidying to 60, but this doesn't currently work with styler.
# Set tidy = True to get the knitr default
# I want all figures as png and pdf in high quality in a subfolder called figure

knitr::opts_chunk$set(
  echo = TRUE,
  tidy = "styler",
  dev = c("png", "pdf"),
  dpi = 600,
  fig.path = "figure/"
)

options(
  xtable.comment = F,
  xtable.include.rownames = F,
  nina.logo.y.pos = 0.15
)
palette(ninaPalette())
```

```{r, include = F, eval = F}
# This connects to the gisdatabase with a DBI connection named `con`.
# Use for example dbGetQuery(con, "SELECT * FROM ....") to query the database
postgreSQLConnect(
  host = "gisdata-db.nina.no",
  dbname = "gisdata",
  username = "postgjest",
  password = "gjestpost"
)
```

```{r, include = F, eval = F}
#To log into something with your own user, it's good practice not to store passwords in the script.
##I store the username and passwords in a file in my home folder, that only the owner can read.
#Not the prettiest solution, since I read in the info in plain text, but we delete the objects in the same code chunk so it works. 
##Here I also specify that big integers should be converted to integers in R. 
##There are some functions that don't have support for bigints in R at the moment.

source("~/.rpgpass")

postgreSQLConnect(
  host = "gisdata-db.nina.no",
  dbname = "gisdata",
  username = username,
  password = password,
  bigint  = "integer"
)

rm(list = c("username", "password"))
```

Bla bla

### 2022-09-21 indicators/insectindicators/tutorials/tutorial.Rmd
test

@@
 
-```{r, include = F, eval = F}
-# This connects to the gisdatabase with a DBI connection named `con`.
-# Use for example dbGetQuery(con, "SELECT * FROM ....") to query the database
-postgreSQLConnect(
-  host = "gisdata-db.nina.no",
-  dbname = "gisdata",
-  username = "postgjest",
-  password = "gjestpost"
-)
+Bla bla
+
+```{r}
+library(Norimon)
 ```
-
-```{r, include = F, eval = F}
-#To log into something with your own user, it's good practice not to store passwords in the script.
-##I store the username and passwords in a file in my home folder, that only the owner can read.
-#Not the prettiest solution, since I read in the info in plain text, but we delete the objects in the same code chunk so it works. 
-##Here I also specify that big integers should be converted to integers in R. 
-##There are some functions that don't have support for bigints in R at the moment.
-
-source("~/.rpgpass")
-
-postgreSQLConnect(
-  host = "gisdata-db.nina.no",
-  dbname = "gisdata",
-  username = username,
-  password = password,
-  bigint  = "integer"
-)
-
-rm(list = c("username", "password"))
+```{r}
+connect_to_insect_db()
 ```
-
-Bla bla

### 2022-09-21 naturtype2.Rmd
excluding NiN-codes

@@
 
-### V1 Åpen jordvassmyr
+### Nature types in *Våtmark*
 
-The field recorders are allowed to themselves select the NiN types (sub units of the NiN main types) that thir locality supposedy belongs to.
-This may differ from what NEA (Norwegian Environmental Agency) uses to define the nature type.
-Therefore, in the data set, there are a lot of nature types that presumably cover, at least partly, V1 *Åpen jordvassmyr*:
 ```{r}
-unique(wet$naturtype[wet$ninkartleggingsenheter3 =="V1"])
+ntyp <- unique(wet$naturtype)
 ```
-If we  look at the field protocol, we get a different picture.
 
-![alt text](images/V1.PNG)
+```{r}
+years <- NULL
+for(i in 1:length(ntyp)){
+years[i] <- paste(unique(wet$kartleggingsår[wet$naturtype == ntyp[i]]), collapse = ", ")
+}
+```
 
-Green areas means that the entire NiN space is included in the nature type (which is bold).
-So for V1, the highest KA level (level *i*) is covered by E10 (*Rik åpne jordvassmyr*) with three sub-classes.
-For the yellow areas, only parts of the NiN spaces is mapped, and the defining parameters are in bold followingthe nature type code.
-So for KA e-f for exmaple, this NiN space is covered by E10.1 (*Rik åpen sørlig jordvassmye*), but only in boreonemoral and south boreal sones (*BN/SB*).
-The red boxes indicate a third axis in this figure related to MF (myrflate).
-The two boxes on the right are partly covered, and the two to the right are not.
-Finally, there is one more NiN type (V1-C9) written at the bottom which is partly covered (yellow outline).
+```{r}
+ntypDF <- data.frame(
+  "Nature_type" = ntyp,
+  "Year"        = years
+)
+```
 
-So, NiN main type is actually, today, only mapped through one nature type, E10.
-Some of the nature types might be old refracted, types.
-The list of nature types linked to V1 is, however, not of much use to us.
+```{r, fig.cap="List of nature types associated with the ecosystem Våtmark, showing the year when that nature type was mapped." }
+DT::datatable(ntypDF)
+```
+
+The way to do this I think is to look at each of these in turn and map the NiN main types and sub-types that they cover.
+First, I'll go over them and exclude those that do not satisfy our working definition of the main ecosystem as *Åpen våtmark*.
+This mainly implies excluding forested types.
+
+```{r}
+exclude <- c("Kalkrik myr- og sumpskogsmark",
+             #"Flommyr, myrkant og myrskogsmark",  # Only V9 is relevant. This type also includes V2 and V8
+             "Gammel fattig sumpskog",
+             "Rik gransumpskog",
+             "Grankildeskog",
+             "Rik svartorsumpskog",
+             "Rik gråorsumpskog",
+             "Rik svartorstrandskog",
+             "Saltpåvirket svartorstrandskog",
+             "Kilde-edellauvskog",
+             "Saltpåvirket strand- og sumpskogsmark",
+             "Svak kilde og kildeskogsmark",
+             "Rik vierstrandskog",
+             "Varmekjær kildelauvskog"
+             )
+```
+
+```{r}
+`%!in%` <- Negate(`%in%`)
+#wet_1 <- wet[wet$naturtype == ntyp[1],]
+wet_1 <- wet[wet$naturtype %!in% exclude,]
+
+wet_1_long <- tidyr::separate_rows(wet_1, ninbeskrivelsesvariabler, sep=",")
+wet_1_long <- tidyr::separate(wet_1_long, 
+                              col = ninbeskrivelsesvariabler,
+                              into = c("NiN_variable_code", "NiN_variable_value"),
+                              sep = "_",
+                              remove=F
+                              )
+wet_1_long$NiN_variable_value <- as.numeric(wet_1_long$NiN_variable_value)
+
+```
+
+Converting NiN_variable_value to numeric introduced NA's. 
+```{r}
+unique(wet_1_long$ninbeskrivelsesvariabler[is.na(wet_1_long$NiN_variable_value)])
+```
+
+These variables include LKM's (lokal kompleks miljøvariabel) which are nonrelevant to us here.
+
+4DL-SS-0 and 4DL-S-0 has to do with the total amount of coarse woody debris/logs.
+It was only recorded for one nature type in 2018, and then as a a biodiversity variable, and not a state variable.
+*X* therefore means *not relevant*.
+
+```{r}
+temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "4DL-SS-0_X" |
+                   wet_1_long$ninbeskrivelsesvariabler == "4DL-S-0_X" ,]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+7SD-NS and 7SD-0 was also only used for that same nature type, and then only for wooded localities. *X* means *not relevant*.
+
+```{r}
+temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7SD-NS_X" |
+                     wet_1_long$ninbeskrivelsesvariabler == "7SD-0_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened three times in 2018 (for that same nature type again). It's fine to exclude these.
+
+```{r}
+temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7TK_X" |
+                    wet_1_long$ninbeskrivelsesvariabler == "7SE_X" ,]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+7GR-GI (grøftingsintensitet) should not have the value X either, so ok to exclude these.
+```{r}
+temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7GR-GI_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+```{r}
+wet_1_long2 <- wet_1_long[!is.na(wet_1_long$NiN_variable_value),]
+```
+
+This leaves us with the following parameters.
+```{r}
+unique(wet_1_long2$NiN_variable_code)
+```
+We can probably exclude several of these as non-relevant.
+
+The PR- variables are the NEAs own variables (not NiN). We can exclude some of these:
+```{r}
+excl <- c("PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  # red list categories
+          "PRKA",   # Number of calcaroues species indicatrs 
+          #"PRSL",  # Slitasje/physical damage. This one we might actually keep
+          "PRHA",   # Number of habitat spesific species
+          #"PRTO",  # Torvuttak / peat extraction. This one we can keep also
+          #"PRTK",  # Spor av tunge kjøretøy / Heavy vehicles. Keep this.
+          "PRMY",   # Myrstruktur / mire structure. A biodiversity indicator
+          "PRSE-KA")
+```
+
+1AG-B is the total shrub cover
+7TK is *kjøreskade*
+1AG-A-G is *dekning gjenveksttrær*
+
+```{r}
+temp <- wet_1_long2[wet_1_long2$NiN_variable_code == "PRKA",]
+View(temp)
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+```{r}
+ntyp <- unique(wet_1_long$naturtype)
+```
+
+```{r}
+
+temp <- wet_1_long[wet_1_long$naturtype == ntyp[1],]
+table(temp$naturtype)
+
+ggplot(temp)+
+  geom_histogram(aes(x = NiN_variable_value))+
+  facet_wrap(vars(NiN_variable_code))
+```
+
+### Kalkrik myr- og sumpskogsmark
+Only mapped in 2018.
+Back then, the defining NiN variables were not clearly stated, but the defining text clearly show that this is forested land, and so it should not be included in our assessment since our operational definition of Våtmark is *Åpen våtmark / open wetlands*.
+
+```{r}
+ntypDF$Eligible[ntypDF$Nature_type == "Kalkrik myr- og sumpskogsmark"] <- "No"
+```
+
+### Slåttemyr
+Mapped in all years and covers the entire NiN space in V9. 
+```{r}
+ntypDF$Eligible[ntypDF$Nature_type == "Slåttemyr"] <- "Yes"
+ntypDF$NiN_mainType[ntypDF$Nature_type == "Slåttemyr"] <- "V9"
+ntypDF$NiN_mainType_coverage[ntypDF$Nature_type == "Slåttemyr"] <- "Complete"
+
+ntypDF$'Busksjikt-dekning (1AG-B)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
+ntypDF$'Grøftings-intensitet (7GR-GI)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
+ntypDF$'Dekning av gjenvekst-trær (1AG-A-G)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
+ntypDF$'Andel av vedvekster i feltsjiktet (1AR-C-L)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
+
+ntypDF$Remarks[ntypDF$Nature_type == "Slåttemyr"] <- "The list of NiN condition variables has changed over the years, and possibly also the different threshold values."
+```
 

### 2022-09-23 naturtype2.Rmd
nin variables

@@
 ```{r, eval=T}
-myData <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+dat <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
 ```
 
-## Representativity in mapping units
+## Naturetypes, condition variables, and their linkage to NiN main types
 Only a subset of nature types are mapped.
-How well do the mapped units represent, or cover, the range of nature types that is included for each main ecosystem group?
-I will tsrt by looking at wetlands (våtmark).
+How well do these mapped units/areas represent the range of nature types that is included for each main ecosystem group?
+And what condition variables do we have available for each of these?
+
+### Nature types
+Lets extract the unique nature types, their associated main ecosystem, and the year when these nature types were mapped.
+
+We will only include our target ecosystems, and we will exclude *Hule eiker*.
+```{r}
+target <- c("Semi-naturlig mark",
+            "Våtmark",
+            "Naturlig åpne områder under skoggrensa")
+dat <- dat[dat$hovedøkosystem %in% target,]
+dat <- dat[dat$naturtype != "Hule eiker",]
+```
+
+Get the unique nature types
+```{r}
+ntyp <- unique(dat$naturtype)
+```
+
+Extract the year when these were mapped
+```{r}
+years <- NULL
+for(i in 1:length(ntyp)){
+years[i] <- paste(unique(dat$kartleggingsår[dat$naturtype == ntyp[i]]), collapse = ", ")
+}
+```
+
+Extract the associated ecosystem
+```{r}
+eco <- NULL
+for(i in 1:length(ntyp)){
+eco[i] <- paste(unique(dat$hovedøkosystem[dat$naturtype == ntyp[i]]), collapse = ", ")
+}
+```
+
+Combine into one data frame
+```{r}
+ntypDF <- data.frame(
+  "Nature_type" = ntyp,
+  "Year"        = years,
+  "Ecosystem"   = eco
+)
+```
+
+We have 67 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *Open Wetland*.
+
+```{r}
+excl <- c("Kalkrik myr- og sumpskogsmark",
+             #"Flommyr, myrkant og myrskogsmark",  # Only V9 is relevant. This type also includes V2 and V8
+             "Gammel fattig sumpskog",
+             "Rik gransumpskog",
+             "Grankildeskog",
+             "Rik svartorsumpskog",
+             "Rik gråorsumpskog",
+             "Rik svartorstrandskog",
+             "Saltpåvirket svartorstrandskog",
+             "Kilde-edellauvskog",
+             "Saltpåvirket strand- og sumpskogsmark",
+             "Svak kilde og kildeskogsmark",
+             "Rik vierstrandskog",
+             "Varmekjær kildelauvskog"
+             )
+`%!in%` <- Negate(`%in%`)
+ntypDF <- ntypDF[ntypDF$Nature_type %!in% excl,]
+```
+
+```{r, fig.cap="List of nature types associated with the ecosystem Våtmark, showing the year when that nature type was mapped." }
+DT::datatable(ntypDF)
+```
+
+The way to do this I think is to look at each nature type in turn and map the NiN main types and sub-types that they cover. At the same time I can extract the NiN variables and values to see.
+
+### NiN variables
+
+```{r}
+dat2 <- dat[dat$naturtype %!in% exclude,]
+
+dat2L <- tidyr::separate_rows(dat2, ninbeskrivelsesvariabler, sep=",")
+dat2L <- tidyr::separate(dat2L, 
+                              col = ninbeskrivelsesvariabler,
+                              into = c("NiN_variable_code", "NiN_variable_value"),
+                              sep = "_",
+                              remove=F
+                              )
+dat2L$NiN_variable_value <- as.numeric(dat2L$NiN_variable_value)
+
+```
+```{r}
+#View(dat2L[280973,])
+#View(dat2[dat2$identifikasjon_lokalid =="NINFP2110037887",])
+```
+
+Converting NiN_variable_value to numeric introduced NA's. 
+```{r}
+sort(unique(dat2L$ninbeskrivelsesvariabler[is.na(dat2L$NiN_variable_value)]))
+```
+#### 4DL
+4DL-SS-0 and 4DL-S-0 has to do with the total amount of coarse woody debris/logs.
+It was only recorded for one nature type in 2018, and then as a a biodiversity variable, and not a state variable.
+*X* therefore means *not relevant*.
+
+#### 5AB and 5BY
+These area land use types (arealbruksklasser) and building types, respectively, and the values are not numeric, but categorical, and represents presence or absence of these land uses or objects.
+These are not suited for automatically determining ecosystem condition as this is done subjectively in the field.
+```{r}
+exclude <- c(exclude,
+             "5AB-0_DO",
+             "5AB-0_FO",
+             "5AB-0_KO",
+             "5AB-0_TO",
+             "5BY-0_BO",
+             "5BY-0_FB",
+             "5BY-0_IL",
+             "5BY-0_KB",
+             "5BY-0_SK",
+             "5BY-0_XB")
+```
+
+#### 7FA (fremmedartsinnslag)
+Should not have been allowed the value X, so Ok to exclude these.
+
+#### 7GR-GI
+7GR-GI (grøftingsintensitet) should not have the value X either, so ok to exclude these.
+```{r}
+exclude <- c(exclude, "7GR-GI_X")
+```
+
+#### 7JA jord-bruk
+
+* BA = Aktuell bruksintensitet
+* BT = Beitetrykk
+* GJ = Gjødsling
+* KU = Kystlyngheias utviklingsfaser
+    + BY = byggefase
+    + DE = degenereringsfase
+    + MO = moden fase
+    + PI = pionerfase
+
+NiN defines the possible values for these as numeric between 0 and 4, but there are 544 cases where it has been given then value *X*.
+
+First, is seems like the value range is shifted from 0-4 to be 1-5.
+```{r, fig.cap=""}
+dotchart(dat2L$NiN_variable_value[dat2L$NiN_variable_code == "7JB-KU-BY"], xlab="value", pch=3)
+```
+
+```{r}
+
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-KU-BY_X" |
+                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-DE_X" |
+                dat2L$ninbeskrivelsesvariabler == "7JB-KU-MO_X" |
+                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-PI_X",]
+table(temp$kartleggingsår, temp$naturtype)
+#View(dat2L[dat2L$identifikasjon_lokalid ==  "NINFP1810033954",])
+```
+
+The reason could be that this variable is used in the biodiversity assessment, which is not performed if the condition is *very poor*.
+```{r}
+table(temp$tilstand)
+```
+That is not the case. If the condition is scored as low, the variable does not even appear in the data set for that locality.
+We can also make a note that a big proportion of the total number of localities of *Kystlynghei* had very poor condition, and therefore this variable 7JB-KU was not recorded.
+```{r, fig.cap="The distribution of condition scores for Kystlynghei."}
+KU <- dat2[dat2$naturtype=="Kystlynghei",]
+#table(KU$tilstand)
+barplot(table(KU$tilstand))
+```
+
+If we look at some cases in more detail, to understand why any of these variables has been set as X (probably means they were left blank), for example like this:
+```{r}
+#View(dat2L[dat2L$NiN_variable_code == "7JB-KU-BY" & is.na(dat2L$NiN_variable_value),])
+#View(dat2L[dat2L$identifikasjon_lokalid =="NINFP1810002975",]) # None of the 7JB-KU variables have values
+#View(dat2L[dat2L$identifikasjon_lokalid =="NINFP2110057689",]) # Two of the 7JB-KU variables have values
+```
+ ...
+we can see that sometimes <4 but >0 of the variables/phases has been given a score, but sometimes none have.
+This makes it problematic to set these NA' to be zeros.
+When we also consider the fact that there is an obvious bias in that only localities with poor condition or better have these variables assessed in the first place, I will exclude all of these variables and treat them as *not suited*.
+
+```{r}
+exclude <- c(exclude, "7JB-KU-BY_X", "7JB-KU-DE_X", "7JB-KU-MO_X", "7JB-KU-PI_X")
+```
+
+#### 7RA-SJ
+*Rask gjenvekstsuksesjon i semi-naturlig og sterkt endret jordbruksmark inkludert våteng*
+There are some very few cases where this variable has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these.
+
+```{r}
+exclude <- c(exclude, "7RA-SJ_X")
+```
+
+#### 7SD
+7SD-NS and 7SD-0 was also only used for that same nature type, and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*. *X* means *not relevant*.
+```{r}
+exclude <- c(exclude, "7SD-0_X",     "7SD-NS_X")
+```
+
+#### 7TK and /SE
+7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to exclude these.
+```{r}
+exclude <- c(exclude, "7TK_X", "7SE_X")
+```
+
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7FA_X" ,]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+#### 7VR-RI
+*Reguleringsintensitet*
+This variable is defined as numeric between 1-5. There is, however, quite a few cases where it is X.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7VR-RI_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to exclude.
+
+#### LKMs and MdirPR-variables
+These variables include LKM's (lokal kompleks miljøvariabel) which are nonrelevant to us here. We can also exclude some of the PR- variables, which are the NEAs own variables (not NiN). 
+```{r}
+excl <- c("LKMKI",     
+          "LKMKI",     
+          "LKMKI",     
+          "LKMKI",     
+          "LKMKI",     
+          "LKMKI",     
+          "LKMSP", 
+          "PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  # red list categories
+          "PRKA",   # Number of calcaroues species indicators 
+          #"PRSL",  # Slitasje/physical damage. This one we might actually keep
+          "PRHA",   # Number of habitat spesific species
+          #"PRTO",  # Torvuttak / peat extraction. This one we can keep also
+          #"PRTK",  # Spor av tunge kjøretøy / Heavy vehicles. Keep this.
+          "PRMY",   # Myrstruktur / mire structure. A biodiversity indicator
+          "PRSE-KA",
+          "PRH", # no idea what this is
+          "PRVS_X" # should be numeric, but is X in 1 locality. It's 'Variasjon i vannsprutintensitet'.
+          )
+```
+
+-------------------------
+
+```{r}
+wet_1_long2 <- wet_1_long[!is.na(wet_1_long$NiN_variable_value),]
+```
+
+This leaves us with the following parameters.
+```{r}
+unique(wet_1_long2$NiN_variable_code)
+```
+We can probably exclude several of these as non-relevant.
+
+1AG-B is the total shrub cover
+7TK is *kjøreskade*
+1AG-A-G is *dekning gjenveksttrær*
+
+```{r}
+ntyp <- unique(wet_1_long$naturtype)
+```
+
+```{r}
+
+temp <- wet_1_long[wet_1_long$naturtype == ntyp[1],]
+table(temp$naturtype)
+
+ggplot(temp)+
+  geom_histogram(aes(x = NiN_variable_value))+
+  facet_wrap(vars(NiN_variable_code))
+```
+
+------------------------------------------
 
 ### Våtmark
+I will start by looking at wetlands (våtmark).
+
 ```{r}
@@
 
-### Nature types in *Våtmark*
-
-```{r}
-ntyp <- unique(wet$naturtype)
-```
-
-```{r}
-years <- NULL
-for(i in 1:length(ntyp)){
-years[i] <- paste(unique(wet$kartleggingsår[wet$naturtype == ntyp[i]]), collapse = ", ")
-}
-```
-
-```{r}
-ntypDF <- data.frame(
-  "Nature_type" = ntyp,
-  "Year"        = years
-)
-```
-
-```{r, fig.cap="List of nature types associated with the ecosystem Våtmark, showing the year when that nature type was mapped." }
-DT::datatable(ntypDF)
-```
-
-The way to do this I think is to look at each of these in turn and map the NiN main types and sub-types that they cover.
-First, I'll go over them and exclude those that do not satisfy our working definition of the main ecosystem as *Åpen våtmark*.
-This mainly implies excluding forested types.
-
-```{r}
-exclude <- c("Kalkrik myr- og sumpskogsmark",
-             #"Flommyr, myrkant og myrskogsmark",  # Only V9 is relevant. This type also includes V2 and V8
-             "Gammel fattig sumpskog",
-             "Rik gransumpskog",
-             "Grankildeskog",
-             "Rik svartorsumpskog",
-             "Rik gråorsumpskog",
-             "Rik svartorstrandskog",
-             "Saltpåvirket svartorstrandskog",
-             "Kilde-edellauvskog",
-             "Saltpåvirket strand- og sumpskogsmark",
-             "Svak kilde og kildeskogsmark",
-             "Rik vierstrandskog",
-             "Varmekjær kildelauvskog"
-             )
-```
-
-```{r}
-`%!in%` <- Negate(`%in%`)
-#wet_1 <- wet[wet$naturtype == ntyp[1],]
-wet_1 <- wet[wet$naturtype %!in% exclude,]
-
-wet_1_long <- tidyr::separate_rows(wet_1, ninbeskrivelsesvariabler, sep=",")
-wet_1_long <- tidyr::separate(wet_1_long, 
-                              col = ninbeskrivelsesvariabler,
-                              into = c("NiN_variable_code", "NiN_variable_value"),
-                              sep = "_",
-                              remove=F
-                              )
-wet_1_long$NiN_variable_value <- as.numeric(wet_1_long$NiN_variable_value)
-
-```
-
-Converting NiN_variable_value to numeric introduced NA's. 
-```{r}
-unique(wet_1_long$ninbeskrivelsesvariabler[is.na(wet_1_long$NiN_variable_value)])
-```
-
-These variables include LKM's (lokal kompleks miljøvariabel) which are nonrelevant to us here.
-
-4DL-SS-0 and 4DL-S-0 has to do with the total amount of coarse woody debris/logs.
-It was only recorded for one nature type in 2018, and then as a a biodiversity variable, and not a state variable.
-*X* therefore means *not relevant*.
-
-```{r}
-temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "4DL-SS-0_X" |
-                   wet_1_long$ninbeskrivelsesvariabler == "4DL-S-0_X" ,]
-table(temp$kartleggingsår, temp$naturtype)
-```
-
-7SD-NS and 7SD-0 was also only used for that same nature type, and then only for wooded localities. *X* means *not relevant*.
-
-```{r}
-temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7SD-NS_X" |
-                     wet_1_long$ninbeskrivelsesvariabler == "7SD-0_X",]
-table(temp$kartleggingsår, temp$naturtype)
-```
-
-7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened three times in 2018 (for that same nature type again). It's fine to exclude these.
-
-```{r}
-temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7TK_X" |
-                    wet_1_long$ninbeskrivelsesvariabler == "7SE_X" ,]
-table(temp$kartleggingsår, temp$naturtype)
-```
-
-7GR-GI (grøftingsintensitet) should not have the value X either, so ok to exclude these.
-```{r}
-temp <- wet_1_long[wet_1_long$ninbeskrivelsesvariabler == "7GR-GI_X",]
-table(temp$kartleggingsår, temp$naturtype)
-```
-
-```{r}
-wet_1_long2 <- wet_1_long[!is.na(wet_1_long$NiN_variable_value),]
-```
-
-This leaves us with the following parameters.
-```{r}
-unique(wet_1_long2$NiN_variable_code)
-```
-We can probably exclude several of these as non-relevant.
-
-The PR- variables are the NEAs own variables (not NiN). We can exclude some of these:
-```{r}
-excl <- c("PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  # red list categories
-          "PRKA",   # Number of calcaroues species indicatrs 
-          #"PRSL",  # Slitasje/physical damage. This one we might actually keep
-          "PRHA",   # Number of habitat spesific species
-          #"PRTO",  # Torvuttak / peat extraction. This one we can keep also
-          #"PRTK",  # Spor av tunge kjøretøy / Heavy vehicles. Keep this.
-          "PRMY",   # Myrstruktur / mire structure. A biodiversity indicator
-          "PRSE-KA")
-```
-
-1AG-B is the total shrub cover
-7TK is *kjøreskade*
-1AG-A-G is *dekning gjenveksttrær*
-
-```{r}
-temp <- wet_1_long2[wet_1_long2$NiN_variable_code == "PRKA",]
-View(temp)
-table(temp$kartleggingsår, temp$naturtype)
-```
-
-```{r}
-ntyp <- unique(wet_1_long$naturtype)
-```
-
-```{r}
-
-temp <- wet_1_long[wet_1_long$naturtype == ntyp[1],]
-table(temp$naturtype)
-
-ggplot(temp)+
-  geom_histogram(aes(x = NiN_variable_value))+
-  facet_wrap(vars(NiN_variable_code))
-```
-
 ### Kalkrik myr- og sumpskogsmark

### 2022-09-28 naturtype2.Rmd
nin codes

@@
 ```{r}
-excl <- c("Kalkrik myr- og sumpskogsmark",
+excl_nt <- c("Kalkrik myr- og sumpskogsmark",
              #"Flommyr, myrkant og myrskogsmark",  # Only V9 is relevant. This type also includes V2 and V8
@@
 `%!in%` <- Negate(`%in%`)
-ntypDF <- ntypDF[ntypDF$Nature_type %!in% excl,]
+ntypDF <- ntypDF[ntypDF$Nature_type %!in% excl_nt,]
 ```
@@
 ```{r}
-dat2 <- dat[dat$naturtype %!in% exclude,]
+dat2 <- dat[dat$naturtype %!in% excl,]
 
@@
 ```
+
+Checking some NA's produced
 ```{r}
@@
 ```
+No problem.
 
-Converting NiN_variable_value to numeric introduced NA's. 
+Here are all the Nin variable codes:
 ```{r}
-sort(unique(dat2L$ninbeskrivelsesvariabler[is.na(dat2L$NiN_variable_value)]))
+unique(sort(dat2L$NiN_variable_code))
 ```
+
+Converting NiN_variable_value to numeric also introduced NA's for these categories: 
+```{r}
+sort(unique(dat2L$NiN_variable_code[is.na(dat2L$NiN_variable_value)]))
+```
+When I now go through the variable codes sperately I will also judge if these NA's are real, or if they for example should be coded as zeros or something.
+
+#### 1AG-A
+"1AG-A-0"   "1AG-A-E"   "1AG-A-G"   "1AG-B"     "1AG-C"
+
+#### 1AR-C-L
+
+#### 3TO
+"3TO-HA"    "3TO-HE"    "3TO-HK"    "3TO-HN"   "3TO-HP"    "3TO-PA"    "3TO-TE"
+
 #### 4DL
@@
 
+#### 4TG
+"4TG-BL"    "4TG-EL"    "4TG-GF"
+
+#### 4TL
+"4TL-BS"    "4TL-HE"    "4TL-HL" "4TL-RB"    "4TL-SB"
+
 #### 5AB and 5BY
@@
 ```{r}
-exclude <- c(exclude,
-             "5AB-0_DO",
-             "5AB-0_FO",
-             "5AB-0_KO",
-             "5AB-0_TO",
-             "5BY-0_BO",
-             "5BY-0_FB",
-             "5BY-0_IL",
-             "5BY-0_KB",
-             "5BY-0_SK",
-             "5BY-0_XB")
+exclude <- c(
+             "5AB-0",
+             "5AB-DO-TT",
+             "5BY-0"
+             )
 ```
+
+#### 6SE
+
+#### 6SO
 
@@
 #### 7GR-GI
-7GR-GI (grøftingsintensitet) should not have the value X either, so ok to exclude these.
-```{r}
-exclude <- c(exclude, "7GR-GI_X")
-```
+7GR-GI (grøftingsintensitet) should not have the value X either, so OK to exclude these.
 
-#### 7JA jord-bruk
-
-* BA = Aktuell bruksintensitet
-* BT = Beitetrykk
-* GJ = Gjødsling
+#### 7JB-KU
 * KU = Kystlyngheias utviklingsfaser
@@
 
-NiN defines the possible values for these as numeric between 0 and 4, but there are 544 cases where it has been given then value *X*.
-
-First, is seems like the value range is shifted from 0-4 to be 1-5.
-```{r, fig.cap=""}
-dotchart(dat2L$NiN_variable_value[dat2L$NiN_variable_code == "7JB-KU-BY"], xlab="value", pch=3)
-```
-
+Let's look at 7JB-KU first.    
+NiN defines the possible values for these as numeric between 0 and 4 (shifted to become 1-5 in the dataset), but there are 544 cases where it has been given then value *X*.
 ```{r}
-
 temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-KU-BY_X" |
@@
 KU <- dat2[dat2$naturtype=="Kystlynghei",]
-#table(KU$tilstand)
 barplot(table(KU$tilstand))
@@
 
-If we look at some cases in more detail, to understand why any of these variables has been set as X (probably means they were left blank), for example like this:
+Each *kystlynghei* locality should have values for all four parameters (7BA-BY/DE/MO/PI), but if we look at some cases in more detail, to understand why some of these values have been set as X (probably means they were left blank), for example like this:
 ```{r}
@@
 we can see that sometimes <4 but >0 of the variables/phases has been given a score, but sometimes none have.
-This makes it problematic to set these NA' to be zeros.
+This makes it problematic to set these NA's to be zeros.
 When we also consider the fact that there is an obvious bias in that only localities with poor condition or better have these variables assessed in the first place, I will exclude all of these variables and treat them as *not suited*.
@@
 ```{r}
-exclude <- c(exclude, "7JB-KU-BY_X", "7JB-KU-DE_X", "7JB-KU-MO_X", "7JB-KU-PI_X")
+exclude <- c(exclude, 
+             "7JB-KU-BY", 
+             "7JB-KU-DE", 
+             "7JB-KU-MO", 
+             "7JB-KU-PI")
 ```
+
+#### 7JB-BA = Aktuell bruksintensitet
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BA_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+These are just a few cases, mostly from 2018. OK to treat as NA.
+
+#### 7JB-BT = Beitetrykk
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BT_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+OK to treat as NA's.
+
+#### 7JB-GJ = Gjødsling
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BT_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+OK to treat as NA's. This one, and some of the previous ones as well, are more like pressure indicators than condition indicators, but I wont worry too much about that at this stage.
+
+#### 7JB-HT
+"7JB-HT-SL" "7JB-HT-ST"
+
+#### 7RA-BH
+"7RA-BH"
 
@@
 *Rask gjenvekstsuksesjon i semi-naturlig og sterkt endret jordbruksmark inkludert våteng*
-There are some very few cases where this variable has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these.
+There are some very few cases where this variable has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these (i.e. allow them to be NA's).
+
+#### 7SD
+7SD-NS and 7SD-0 was also only used for one nature type, and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*.
+*X* therefore means *not relevant*, and we can also exclude the parameters from the dataset all together.
 
 ```{r}
-exclude <- c(exclude, "7RA-SJ_X")
-```
-
-#### 7SD
-7SD-NS and 7SD-0 was also only used for that same nature type, and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*. *X* means *not relevant*.
-```{r}
-exclude <- c(exclude, "7SD-0_X",     "7SD-NS_X")
+exclude <- c(exclude, 
+             "7SD-0",     
+             "7SD-NS")
 ```
@@
 7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to exclude these.
-```{r}
-exclude <- c(exclude, "7TK_X", "7SE_X")
-```
-
-```{r}
-temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7FA_X" ,]
-table(temp$kartleggingsår, temp$naturtype)
-```
 
@@
 
-I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to exclude.
+I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to treat as NA's
 
@@
 These variables include LKM's (lokal kompleks miljøvariabel) which are nonrelevant to us here. We can also exclude some of the PR- variables, which are the NEAs own variables (not NiN).
+
+* LKMKI
+    + Kalk (lime)
+    + Exclude
+* LKMSP
+    + Slåttemarkspreg
+    + Exclude
+* PRAK
+    + Antall NiN-kartleggingsenheter
+    + Used in the biodiversity assessment
+    + Exclude
+* PRAM
+    + Menneskeskapte objekter
+    + Related to 5BY and 5AB
+    + Exclude (se 5BY and 5AB)
+* PRH
+    + This is perhaps a mistake and should be PRHA or PRHT
+    + Exclude
+* PRHA
+    + Habitat spesifikke arter
+    + Used in the biodiversity assessment
+    + Exclude
+* PRHT
+    + Høsting av tresjiktet
+    + Include
+* PRKA
+    + Kalkindikatorer
+    + Used in the biodiversity assessment
+    + Exclude
+* PRKU
+    + Kystlyngheias utviklingsfaser
+    + Related to 7JB-KU
+    + Used in the biodiversity assessment
+    + Exclude
+* PRMY
+    + Myrstruktur
+    + Used in the biodiversity assessment
+    + Exclude
+* PRRL-CR/DD/EN/NT/VU
+    + Red list categories
+    + Used in the biodiversity assessment
+    + Exclude
+* PRSE-KA
+    + Strukturer, elementer og torvmarksformer (in 2018)
+    + See PRSE-PA
+* PRSE-PA
+    + Strukturer, elementer og torvmarksformer
+    + Used in the biodiversity assessment
+    + Exclude
+* PRSE-SH
+    + Strukturer, elementer og torvmarksformer (in 2018)
+    + See PRSE-PA
+* PRSL
+    + Slitasje
+    + Rlated to 7SE
+    + Include
+* PRTK
+    + Spor av tunge kjøretøyer
+    + Related to 7TK
+    + Include
+* PRTO
+    + Torvuttak
+    + Include
+* PRVS
+    + Variasjon i vannsprutintensitet
+    + Used in the biodiversity assessment
+    + Exclude
+
 ```{r}
-excl <- c("LKMKI",     
-          "LKMKI",     
-          "LKMKI",     
-          "LKMKI",     
-          "LKMKI",     
-          "LKMKI",     
+exclude <- c("LKMKI",     
           "LKMSP",
-          "PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  # red list categories
-          "PRKA",   # Number of calcaroues species indicators 
-          #"PRSL",  # Slitasje/physical damage. This one we might actually keep
-          "PRHA",   # Number of habitat spesific species
-          #"PRTO",  # Torvuttak / peat extraction. This one we can keep also
-          #"PRTK",  # Spor av tunge kjøretøy / Heavy vehicles. Keep this.
-          "PRMY",   # Myrstruktur / mire structure. A biodiversity indicator
+          "PRAK",
+          "PRAM",
+          "PRH",
+          "PRHA",
+          "PRKA",
+          "PRKU",
+          "PRMY",
+          "PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  
           "PRSE-KA",
-          "PRH", # no idea what this is
-          "PRVS_X" # should be numeric, but is X in 1 locality. It's 'Variasjon i vannsprutintensitet'.
+          "PRSE-PA",
+          "PRSE-SH",
+          "PRVS"
           )
 ```
+
+### Subset
+```{r}
+dat2L2 <- dat2L[!is.na(dat2L$NiN_variable_value),]
+dat2L2 <- dat2L2[dat2L2$NiN_variable_code %!in% exclude,]
+```
+
+Are the nature types or groups of nature types representative of the main NiN types?
+And, is the coverage of NiN main types representative of the Hovedøkosystem?
+If so then check for environmental bias.
+If OK to continue, then check for available NiN variables, especially reoccurring variables.
+
+```{r}
+ntypDF$hovedgruppe[ntypDF$Nature_type == "Slåttemark"] <- "T32"
+```
+
+```{r}
+ntypDF$coverage[ntypDF$Nature_type == "Slåttemark"] <- "Partial"
+```
+
+```{r}
+codes <- unique(dat2L2$NiN_variable_code)
+```
+
+Next, add 'codes' in seperate rows (wide) and cast to columns
 

### 2022-09-30 index.Rmd
Update index.Rmd

@@
 
-For those participating in the indicator project, detailed information about how to document your work will be communicated over email, but the main steps can be summarised as this:
+This project uses a type of standardised reporting where reproducible workflows (typically code) for calculating each indicator is presented on separate webpages within a shared website.
+The website is built from the GitHub repo using seperate RMarkdown files for each chapter.
+These files must be in English.
+The process of authoring and sharing/distributing these Rmarkdown-files is explained below.
+
+It is highly encouraged that the indicators are delivered as georeferenced maps, hereafter called indicator maps, and not a simple tables.
+There is a common structure that all indicator maps should follow.
+The indicator maps can be vector formats raster formats.
+For vector data, .Rdata or .rds files can be used, as for example shape files do not allow field names (i.e.
+column names) that are as long as the ones we use.
+All georeferenced data should use EPSG:25833 - ETRS89 / UTM zone 33N.
+
+Indicator maps should have the highest possible spatial resolution possible but weighing this against the need for spatial representativity.
+For example, you may need to aggregate x number of data points to get an average which is stable and representative for a given region.
+There is no need to spatially aggregate indicator maps any further, for example to produce a national estimate.
+
+The indicator values, reference values ect should be mapped to the following columns (vector data) or bands (raster data):
+* `v_YYYY` for variable value (i.e. un-scaled or raw indicator values) for year YYYY. 
+* `sd_YYYY` for the standard deviation associated with v_YYYY. Other measures for uncertainty is currently not supported. 
+* `referance_high` for the reference value, i.e. the value of the variable *v* under reference conditions. This value cannot vary between years.
+* `referance_low` for the lower limit reference value, i.e. the worst possible value of the variable *v*. This value cannot vary between years. If missing, this value is assumed to be zero.
+* `thr` for the threshold value defining *good condition* for variable *v*. If missing, this value will default to 0.6.
+* `i_YYYY` for the indicator value at year YYYY.
+This value depends on all the above values, but also the scaling function (.e.
+linear, exponential, two-sided, ect.).
+The scaling function needs to be presented in other documentation.
+
+The rest of the workflow is as follows.
 
@@
   The `data` folder also contains some supporting data sets that you may want to use, such as a delineation of the five regions used in the ecosystem assessment (`data/regions.shp`), and an outline of Norway (`outlineOfNorway_EPSG25833.shp`).
-  + To preview you rmarkdown file as html, type `rmarkdown::render("DRAFT_myIndicator.Rmd")` in the console (the knitr shortcuts don't work as expected inside bookdown projects). 
+  + To preview you rmarkdown file as html in R, type `rmarkdown::render('DRAFT_myIndicator.Rmd', output_format = "html_document", output_dir = "temp")` in the console (the knitr shortcuts don't work as expected inside bookdown projects).
   + When filling out the template.Rmd file with your own work, try not to change the headers too much.
   + Final result from the analyses should be written to the `indicators` folder.
-The preferred output is georeferenced data (rasters or shape files), andthese should be placed under `indators/indicatorMaps`.
-All georeference data should use EPSG:25833 - ETRS89 / UTM zone 33N.
-Other output, like data tables, can be stired under `indicators`.
+The preferred output is georeferenced data (rasters or shape files), and these should be placed under `indators/indicatorMaps`.
+Non-georeferenced data, like data tables, can be stored under `indicators`.
 * Make a pull request to the main [*GitHub repository*](https://github.com/NINAnor/ecosystemCondition) to submit your file.
+If you are not comfortable using GitHub, contac Anders to rrange sending the files via email.
 * Anders or someone else in the project wil conduct a rapid code review, making sure the code is reproducible, interpretable and that it renders locally on the R Studio server.

### 2022-10-06 naturtype2.Rmd
evaluate nin variables

@@
 ```{r, eval=T}
-dat <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+dat <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
 ```
@@
 
-We have 67 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *Open Wetland*.
+We have 66 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *Open Wetland*.
 
@@
 
-```{r, fig.cap="List of nature types associated with the ecosystem Våtmark, showing the year when that nature type was mapped." }
+```{r, fig.cap="List of nature types showing the years when that nature type was mapped." }
 DT::datatable(ntypDF)
@@
 ```{r}
-dat2 <- dat[dat$naturtype %!in% excl,]
+#Exclude non-relevant types
+dat2 <- dat[dat$naturtype %!in% excl_nt,]
 
+# Melt
 dat2L <- tidyr::separate_rows(dat2, ninbeskrivelsesvariabler, sep=",")
+
+# Split the code and teh value into seperate columns
 dat2L <- tidyr::separate(dat2L, 
@@
                               )
+# One NA produced here, but I check it, and it's fine.
+
+# Convert values to numeric. This causes some NA's which I will go though below
 dat2L$NiN_variable_value <- as.numeric(dat2L$NiN_variable_value)
@@
 ```
-
-Checking some NA's produced
-```{r}
-#View(dat2L[280973,])
-#View(dat2[dat2$identifikasjon_lokalid =="NINFP2110037887",])
-```
-No problem.
 
@@
 ```
-When I now go through the variable codes sperately I will also judge if these NA's are real, or if they for example should be coded as zeros or something.
+When I now go through the variable codes separately I will also judge if these NA's are real, or if they for example should be coded as zeros or something.
 
-#### 1AG-A
-"1AG-A-0"   "1AG-A-E"   "1AG-A-G"   "1AG-B"     "1AG-C"
+#### 1AG-A (sjiktdekningsvariabler)
+These are different variables describing the cover in different vegetation strata and/or species
 
-#### 1AR-C-L
+* 1AG-A-0 = Total tre canopy cover
+* 1AG-B = Total shrub cover
+* 1AG-C = Total field layer cover
+* 1AG-A-E = 'Overstandere'
+* 1AG-A-G = 'gjenveksttrær'
 
-#### 3TO
-"3TO-HA"    "3TO-HE"    "3TO-HK"    "3TO-HN"   "3TO-HP"    "3TO-PA"    "3TO-TE"
+#### 1AR-C-L (Andel vedvekster i fletsjiktet)
+A condition variable in some mire nature types
 
-#### 4DL
+#### 3TO (Torvmarksformer)
+These are defining variables, and not part of the condition assessment.
+```{r}
+exclude <- c("3TO-BØ",
+             "3TO-HA",
+             "3TO-HE",
+             "3TO-HK",
+             "3TO-HN",
+             "3TO-HP",
+             "3TO-PA",
+             "3TO-TE")
+```
+
+#### 4DL (Coarse dead wood)
 4DL-SS-0 and 4DL-S-0 has to do with the total amount of coarse woody debris/logs.
 It was only recorded for one nature type in 2018, and then as a a biodiversity variable, and not a state variable.
-*X* therefore means *not relevant*.
+Biodiversity variables were only recorded when condition was better than *very poor*.
+This bias means we cannot use these variables.
 
-#### 4TG
-"4TG-BL"    "4TG-EL"    "4TG-GF"
+```{r}
+exclude <- c(exclude,
+             "4DL-SS-0",
+             "4DL-S-0")
+```
 
-#### 4TL
-"4TL-BS"    "4TL-HE"    "4TL-HL" "4TL-RB"    "4TL-SB"
+#### 4TG (old trees)
+
+These three variables are used as biodiveristy variables in  2018 (exclude).
+```{r}
+exclude <- c(exclude,
+             "4TG-BL",
+             "4TG-EL",
+             "4TG-GF")
+```
+
+#### 4TL (trees with fire scars)
+These five variables are ised together with 4TG as biodiversity variables in 2018.
+```{r}
+exclude <- c(exclude,
+             "4TL-BS",
+             "4TL-HE",
+             "4TL-HL",
+             "4TL-RB",
+             "4TL-SB")
+```
 
@@
 ```{r}
-exclude <- c(
+exclude <- c(exclude,
              "5AB-0",
@@
 
-#### 6SE
-
-#### 6SO
+#### 6SE and 6SO (Bioclimatical sections and sones)
+Not relevant here as a condition variable.
+```{r}
+exclude <- c(exclude,
+             "6SE",
+             "6SO")
+```
 
@@
 
-#### 7JB-KU
-* KU = Kystlyngheias utviklingsfaser
-    + BY = byggefase
-    + DE = degenereringsfase
-    + MO = moden fase
-    + PI = pionerfase
+#### 7JB-KU (Kystlyngheias utviklingsfaser)
 
-Let's look at 7JB-KU first.    
+* BY = byggefase
+* DE = degenereringsfase
+* MO = moden fase
+* PI = pionerfase
+
+This variable is used as in the biodiversity assessment, and can therefore not be used in the condition assessment because localities with *very poor* conditions have not been mapped/assessed.
+```{r}
+exclude <- c(exclude, 
+             "7JB-KU-BY", 
+             "7JB-KU-DE", 
+             "7JB-KU-MO", 
+             "7JB-KU-PI")
+```
+
+I will nonetheless explore this variable a bit more.
 NiN defines the possible values for these as numeric between 0 and 4 (shifted to become 1-5 in the dataset), but there are 544 cases where it has been given then value *X*.
@@
 table(temp$kartleggingsår, temp$naturtype)
-#View(dat2L[dat2L$identifikasjon_lokalid ==  "NINFP1810033954",])
 ```
@@
 That is not the case. If the condition is scored as low, the variable does not even appear in the data set for that locality.
+
 We can also make a note that a big proportion of the total number of localities of *Kystlynghei* had very poor condition, and therefore this variable 7JB-KU was not recorded.
@@
 
-```{r}
-exclude <- c(exclude, 
-             "7JB-KU-BY", 
-             "7JB-KU-DE", 
-             "7JB-KU-MO", 
-             "7JB-KU-PI")
-```
+#### 7JB-BA = Aktuell bruksintensitet
+A common condition variable in semi-natural nature types. It's a borderline pressure indicator, but I will keep it in the data set for now at least.
 
-#### 7JB-BA = Aktuell bruksintensitet
+Overview of cases where the value is set as *X*.
 ```{r}
@@
 #### 7JB-BT = Beitetrykk
+Similar variable to the above.
 ```{r}
@@
 ```
-OK to treat as NA's.
+OK to treat X's as NA's.
 
 #### 7JB-GJ = Gjødsling
+Similar variable to the two above.
 ```{r}
@@
 ```
-OK to treat as NA's. This one, and some of the previous ones as well, are more like pressure indicators than condition indicators, but I wont worry too much about that at this stage.
+OK to treat as X's NA's.
 
-#### 7JB-HT
-"7JB-HT-SL" "7JB-HT-ST"
+#### 7JB-HT (Høsting av tresjiktet)
+Used for *Lauveng*in 2019 and 2020.
+```{r}
+temp <- dat2L[dat2L$NiN_variable_code == "7JB-HT-ST" | 
+               dat2L$NiN_variable_code == "7JB-HT-SL" ,]
+table(temp$kartleggingsår, temp$naturtype)
+```
+This is so marginal that I will exclude then already now.
+```{r}
+exclude <- c(exclude,
+             "7JB-HT-SL",
+             "7JB-HT-ST")
+```
 
-#### 7RA-BH
-"7RA-BH"
-
-#### 7RA-SJ
-*Rask gjenvekstsuksesjon i semi-naturlig og sterkt endret jordbruksmark inkludert våteng*
-There are some very few cases where this variable has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these (i.e. allow them to be NA's).
+#### 7RA-BH and 7RA-SJ (rask suksesjon i boreal hei og i i semi-naturlig og sterkt endret jordbruksmark inkludert våteng)
+Condition variables.
+There are some very few cases where 7RA-SJ has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these (i.e. allow them to be NA's).
 
@@
 #### 7TK and /SE
-7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to exclude these.
+7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to treat these as NA's.
 
@@
 ```
-
-I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to treat as NA's
+I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to treat as NA's.
 

### 2022-10-06 naturtype2.Rmd
mapp NiN main types

@@
 for(i in 1:length(ntyp)){
-years[i] <- paste(unique(dat$kartleggingsår[dat$naturtype == ntyp[i]]), collapse = ", ")
+years[i] <- paste(sort(unique(dat$kartleggingsår[dat$naturtype == ntyp[i]])), collapse = ", ")
 }
@@
 ```{r}
-exclude <- c("LKMKI",     
+exclude <- c(exclude,
+             "LKMKI",     
           "LKMSP",
@@
 dat2L2 <- dat2L2[dat2L2$NiN_variable_code %!in% exclude,]
+length(unique(dat2L2$NiN_variable_code))
+```
+Now we are down to 20 possible condition variables.
+
+I will also now exclude nature types that were only mapped in 2018 and/or and not after that.
+These nature types will not only not get more data in the future, but also they were mapped in a time when the methodology was quite unstable.
+
+```{r}
+ntypDF2 <- ntypDF[ntypDF$Year != "2018" &
+                  ntypDF$Year != "2019" & 
+                  ntypDF$Year != "2018, 2019"# no types were mappen in 2018 & 2019 only
+                  ,]
 ```
 
-Are the nature types or groups of nature types representative of the main NiN types?
-And, is the coverage of NiN main types representative of the Hovedøkosystem?
-If so then check for environmental bias.
-If OK to continue, then check for available NiN variables, especially reoccurring variables.
+### Fill inn ntypDF2 
+Next I want to expand ntypDF2 to include the NiN main types covered by the different nature types.
+I cannot use the field data for this, because there are so many mistakes in the data.
+I need to look at the definition of each nature type and get the NiN code from there.
+I could also extract the NiN sub types (grunntyper), but it would become too messy.
 
+(Split the value seperated by | into new column after and use to colour a stacked barplot)
+
+I will look at the definitions of the nature type in the first and last year when that type was mapped, but not for the years in between.
 ```{r}
-ntypDF$hovedgruppe[ntypDF$Nature_type == "Slåttemark"] <- "T32"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Aktiv skredmark"                                    ] <- "T17 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen flomfastmark"                                  ] <- "T18 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen grunnlendt kalkrik mark i boreonemoral sone"   ] <- "T2  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen grunnlendt kalkrik mark i sørboreal sone"      ] <- "T2  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Atlantisk høymyr"                                   ] <- "V3  | partial" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Boreal hei"                                         ] <- "T31 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eksentrisk høymyr"                                  ] <- "V3  | partial" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eng-aktig sterkt endret fastmark"                   ] <- "T41 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosse-eng"                                          ] <- "T15 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosseberg"                                          ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fossepåvirket berg"                                 ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fuglefjell-eng og fugletopp"                        ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Høgereligende og nordlig nedbørsmyr"                ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Isinnfrysingsmark"                                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kalkrik helofyttsump"                               ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kanthøymyr"                                         ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Konsentrisk høymyr"                                 ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kystlynghei"                                        ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Nakent tørkeutsatt kalkberg"                        ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øvre sandstrand uten pionervegetasjon"              ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øyblandingsmyr"                                     ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Palsmyr"                                            ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Platåhøymyr"                                        ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i mellomboreal sone"          ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen sørlig jordvannsmyr"                       ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sanddynemark"                                       ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig eng"                                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig myr"                                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig strandeng"                            ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig våteng"                               ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig etablert sanddynemark"                       ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig kaldkilde"                                   ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig nedbørsmyr"                                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig slåttemyr"                                   ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Strandeng"                                          ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Svært tørkeutsatt sørlig kalkberg"                  ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Terrengdekkende myr"                                ] <- ""
+
 ```
+
+Next, are the coverage of NiN main types representative of the Hovedøkosystem?
+Then I need to manually enter the NiN main types associated with each Hovedøkosystem and perhaps add them as blank rows in ntypDF2.
+
+After that, I will add 20 more columns, one for each nin variable, and indicate for each row/nature type which variables are recorded.
+
+Candidate nature types or groups of nature types that may be suitable to describe ecosystem condition for a NiN main type or Hovedøkosystem will be subject to further testing to look for environmental bias.
 

### 2022-10-06 naturtype2.Rmd
mapping Nin main types

@@
 ```{r}
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Aktiv skredmark"                                    ] <- "T17 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Aktiv skredmark"                                    ] <- "T17 | all (if including sub-types)"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen flomfastmark"                                  ] <- "T18 | all"
@@
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen grunnlendt kalkrik mark i sørboreal sone"      ] <- "T2  | calcareous"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Atlantisk høymyr"                                   ] <- "V3  | partial" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Atlantisk høymyr"                                   ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Boreal hei"                                         ] <- "T31 | all"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eksentrisk høymyr"                                  ] <- "V3  | partial" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eksentrisk høymyr"                                  ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eng-aktig sterkt endret fastmark"                   ] <- "T41 | all"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosse-eng"                                          ] <- "T15 | all"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosseberg"                                          ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fossepåvirket berg"                                 ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fuglefjell-eng og fugletopp"                        ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Høgereligende og nordlig nedbørsmyr"                ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Isinnfrysingsmark"                                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kalkrik helofyttsump"                               ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kanthøymyr"                                         ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Konsentrisk høymyr"                                 ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kystlynghei"                                        ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Nakent tørkeutsatt kalkberg"                        ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øvre sandstrand uten pionervegetasjon"              ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øyblandingsmyr"                                     ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Palsmyr"                                            ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Platåhøymyr"                                        ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i mellomboreal sone"          ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen sørlig jordvannsmyr"                       ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sanddynemark"                                       ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig eng"                                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig myr"                                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig strandeng"                            ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig våteng"                               ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig etablert sanddynemark"                       ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig kaldkilde"                                   ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig nedbørsmyr"                                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig slåttemyr"                                   ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Strandeng"                                          ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Svært tørkeutsatt sørlig kalkberg"                  ] <- ""
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Terrengdekkende myr"                                ] <- ""
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosseberg"                                          ] <- "T1  | extra wet"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fossepåvirket berg"                                 ] <- "T1  | extra wet"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fuglefjell-eng og fugletopp"                        ] <- "T8  | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Høgereligende og nordlig nedbørsmyr"                ] <- "V3  | all (if including sub-types)" # torvmarksformene are excluded
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Isinnfrysingsmark"                                  ] <- "T20 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kalkrik helofyttsump"                               ] <- "L4 | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kanthøymyr"                                         ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Konsentrisk høymyr"                                 ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kystlynghei"                                        ] <- "T34 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Nakent tørkeutsatt kalkberg"                        ] <- "T1  | calcareous and dry"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øvre sandstrand uten pionervegetasjon"              ] <- "T29 | sandy and vegetated"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øyblandingsmyr"                                     ] <- "V1  | partial" # also includes V3
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Palsmyr"                                            ] <- "V3  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Platåhøymyr"                                        ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i mellomboreal sone"          ] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen sørlig jordvannsmyr"                       ] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sanddynemark"                                       ] <- "T21 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig eng"                                  ] <- "T32 | all (if including sub-types)"
+# This might be called Kulturmarkseng in the 2018 protocol, but Semi-naturlig eng in the data set. 
+# Kulturmarkseng also includes V10
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig myr"                                  ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig strandeng"                            ] <- "T33 | all (-2018)" # could perhaps be used in 2018, but it's defined awkwardly
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig våteng"                               ] <- "V10 | all (-2018)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- "T17 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | partial"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig etablert sanddynemark"                       ] <- "T21 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig kaldkilde"                                   ] <- "V4  | southern"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig nedbørsmyr"                                  ] <- "V3  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig slåttemyr"                                   ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Strandeng"                                          ] <- "T12 | all (-2018)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Svært tørkeutsatt sørlig kalkberg"                  ] <- "T1  | calcareous and dry"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Terrengdekkende myr"                                ] <- "V3  | all (if including sub-types)"
 

### 2022-10-07 naturtype2.Rmd
plot area and number of locs

@@
 library(dplyr)
+library(tidyr)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-# Split the code and teh value into seperate columns
+# Split the code and the value into separate columns
 dat2L <- tidyr::separate(dat2L, 
@@
                   ntypDF$Year != "2019" & 
-                  ntypDF$Year != "2018, 2019"# no types were mappen in 2018 & 2019 only
+                  ntypDF$Year != "2018, 2019"# no types were mappend in 2018 & 2019 only
                   ,]
@@
 
-(Split the value seperated by | into new column after and use to colour a stacked barplot)
+I will look at the definitions of the nature type in the first and last year when that type was mapped, but not for the years in between.
 
-I will look at the definitions of the nature type in the first and last year when that type was mapped, but not for the years in between.
+#### Add NiN main type and degree of representativity
 ```{r}
@@
 
+Split this new column in two.
+```{r}
+ntypDF2 <- ntypDF2 %>%
+  tidyr::separate(col = hovedgruppe,
+                  into = c("NiN_mainType", "NiN_mainTypeCoverage"),
+                  #sep = "|",
+                  remove=T,
+                  extra = "merge"
+              )
+```
+
+#### Add NiN variables
+
+```{r}
+ntyp_vars <- as.data.frame(table(dat2L2$naturtype, dat2L2$NiN_variable_code))
+names(ntyp_vars) <- c("Nature_type", "NiN_variable_code", "NiN_variable_count")
+(ntyp_vars <- ntyp_vars[ntyp_vars$NiN_variable_count >0,])
+```
+
+Add a column for the total nmber of localities for each nature type and get the percentage of localities where each NiN has been recorded.
+```{r}
+#options(scipen=999) # suppress exp notation
+count <- as.data.frame(table(dat2$naturtype))
+names(count) <- c("Nature_type", "totalLocations")
+ntyp_vars$totalLocations <- count$totalLocations[match(ntyp_vars$Nature_type, count$Nature_type)]
+ntyp_vars$percentageUse <- round((ntyp_vars$NiN_variable_count/ntyp_vars$totalLocations)*100, 1)
+```
+
+Cast to get NiN codes as columns
+```{r}
+ntyp_vars_wide <- data.table::dcast(setDT(ntyp_vars),
+                                    Nature_type~NiN_variable_code,
+                                    value.var = "percentageUse")
+```
+
+This dat set contains the nature types that were only mappen in 2018 or 2019, and which we removed from ntypDF2. 
+```{r}
+ntyp_vars_wide$Nature_type[ntyp_vars_wide$Nature_type %!in% ntypDF2$Nature_type]
+```
+
+So we should remove them also here.
+```{r}
+ntyp_vars_wide <- ntyp_vars_wide[ntyp_vars_wide$Nature_type %in% ntypDF2$Nature_type]
+```
+
+Combine datasets
+```{r}
+ntyp_fill <- cbind(ntypDF2, ntyp_vars_wide[,-1][match(ntypDF2$Nature_type, ntyp_vars_wide$Nature_type)])
+```
+
+#### Add number of location and total area mapped
+```{r}
+dat2$km2 <- units::drop_units(dat2$area/10^6)
+num <- aggregate(data = dat2,
+              km2~naturtype,
+              FUN = sum)
+
+num$numberOfLocalities <- aggregate(data = dat2,
+              km2~naturtype,
+              FUN = length)[,2]
+#num <- num[num$naturtype %in% ntypDF2$Nature_type,]
+
+ntyp_fill <- cbind(ntyp_fill, num[,-1][match(ntyp_fill$Nature_type, num$naturtype),])
+```
+
+```{r, fig.cap="The total mapped area for nature types associated with three selected main ecosystems"}
+
+mySize <- 8
+gg_area <- ntyp_fill %>%
+  arrange(km2) %>%
+  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%   # This trick update the factor levels
+  ggplot( aes(x = Nature_type,
+                   y = km2)) +
+        geom_segment( aes(xend=Nature_type, yend=0)) +
+        geom_point( size=4, aes(group = Ecosystem,
+                                colour= Ecosystem)) +
+        coord_flip() +
+        theme_bw(base_size = mySize) +
+        xlab("")+
+        ylab(expression(km^2))+
+      theme(legend.position = "top",
+        legend.key.size =  unit(.05, 'cm'),
+            legend.background = element_rect(fill = "white", color = "black")
+            )
+
+gg_locs <- ntyp_fill %>%
+  arrange(km2) %>%
+  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%    # also sorted after km2
+  ggplot( aes(x = Nature_type,
+                   y = numberOfLocalities)) +
+        geom_segment( aes(xend=Nature_type, yend=0)) +
+        geom_point( size=4, aes(group = Ecosystem,
+                                colour= Ecosystem)) +
+        coord_flip() +
+        theme_bw(base_size = mySize) +
+        xlab("")+
+        ylab("Number of localities")+
+  theme(legend.position = "none",
+        axis.text.y = element_blank())
+
+egg::ggarrange(gg_area, gg_locs, 
+                        ncol = 2)
+
+```
+
+```{r, fig.cap="The total number of mapped localities for nature types associated with three selected main ecosystems"}
+ntyp_fill %>%
+  arrange(numberOfLocalities) %>%
+  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%   # This trick update the factor levels
+  ggplot( aes(x = Nature_type,
+               y = numberOfLocalities)) +
+    geom_segment( aes(xend=Nature_type, yend=0)) +
+    geom_point( size=4, aes(group = Ecosystem,
+                            colour= Ecosystem)) +
+    coord_flip() +
+    theme_bw() +
+    xlab("")+
+    ylab("Number of localities")+
+  theme(legend.position = c(0.55, 0.25),
+        legend.background = element_rect(fill = "white", color = "black"))
+
+```
+
 Next, are the coverage of NiN main types representative of the Hovedøkosystem?
 Then I need to manually enter the NiN main types associated with each Hovedøkosystem and perhaps add them as blank rows in ntypDF2.
-
-After that, I will add 20 more columns, one for each nin variable, and indicate for each row/nature type which variables are recorded.
 

### 2022-10-07 naturtype2.Rmd
new plot

@@
 library(dplyr)
+library(RColorBrewer)
+library(forcats)
 library(tidyr)
@@
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- "T17 | all (if including sub-types)"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | partial"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | special management"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- "V9  | all (if including sub-types)"
@@
 dat2$km2 <- units::drop_units(dat2$area/10^6)
-num <- aggregate(data = dat2,
-              km2~naturtype,
-              FUN = sum)
 
-num$numberOfLocalities <- aggregate(data = dat2,
-              km2~naturtype,
-              FUN = length)[,2]
-#num <- num[num$naturtype %in% ntypDF2$Nature_type,]
+num <- as.data.frame(dat2) %>%
+  group_by(naturtype)%>%
+  summarise(km2 = sum(km2),
+            numberOfLocalities = n())
 
@@
 ```
+
+## Plots
 
@@
 
-```{r, fig.cap="The total number of mapped localities for nature types associated with three selected main ecosystems"}
-ntyp_fill %>%
-  arrange(numberOfLocalities) %>%
-  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%   # This trick update the factor levels
-  ggplot( aes(x = Nature_type,
-               y = numberOfLocalities)) +
-    geom_segment( aes(xend=Nature_type, yend=0)) +
-    geom_point( size=4, aes(group = Ecosystem,
-                            colour= Ecosystem)) +
-    coord_flip() +
-    theme_bw() +
-    xlab("")+
-    ylab("Number of localities")+
-  theme(legend.position = c(0.55, 0.25),
-        legend.background = element_rect(fill = "white", color = "black"))
+### Area per NiN main type conditioned on ecosystem and adding empty main types
+First I will add the NiN main types that are not covered by any nature types.
+
+```{r}
+open <- c(
+  "T1",
+  "T2",
+  "T6",
+  "T8",
+  "T11",
+  "T12",
+  "T13",
+  "T15",
+  "T16",
+  "T17",
+  "T18",
+  "T20",
+  "T21",
+  "T23",
+  "T24",
+  "T25",
+  "T26",
+  "T27",
+  "T29"
+)
+
+semi <- c(
+  "T31",
+  "T32",
+  "T33",
+  "T34",
+  "T40",  # are these included
+  "T41"   # are these included
+)
+
+wetland <- c(
+  "V1",
+  "V3",
+  "V4",
+  "V6",
+  "V9",
+  "V10",
+  "L4"    # this will probably not be part of the assessments
+
+)
 
 ```
+
+Adding the non-mapped open types
+```{r}
+open[open %!in% ntyp_fill$NiN_mainType]
+```
+
+```{r}
+ntyp_fill2 <- ntyp_fill %>%
+  add_row(NiN_mainType = "T6", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T11", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T13", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T16", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T23", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T24", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T25", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T26", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T27", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+
+```
+
+Adding the non-mapped semi-natural types
+```{r}
+semi[semi %!in% ntyp_fill$NiN_mainType]
+```
+```{r}
+ntyp_fill2 <- ntyp_fill %>%
+  add_row(NiN_mainType = "T40", Ecosystem = "Semi-naturlig mark", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+```
+
+Adding the non-mapped wetland types
+```{r}
+wetland[wetland %!in% ntyp_fill$NiN_mainType]
+```
+This is Våtsnøleie, and my guess is that it IS mapped, but grouped with the aloine ecosystem. I'll therefore not add it in here, but mention it in the figure caption below.
+
+```{r, fig.cap="The areas mapped for each NiN main type, and the degree of spatial representativity (or coverage) of the mapping units (nature types). V6 Våtsnøleie og snøleiekilde is not included."}
+mySize <- 10
+ntyp_fill2 %>%
+  # combining two classes to get a better colour palette
+  mutate(NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "southern", "partial"),
+         NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "calcareous and dry", "calcareous")) %>%
+
+  group_by(NiN_mainType, NiN_mainTypeCoverage) %>%
+  mutate(km2 = sum(km2)) %>%
+  slice_head()%>%
+  select(Ecosystem,
+         NiN_mainType,
+         NiN_mainTypeCoverage,
+         km2)%>%
+  ungroup() %>%
+  mutate(NiN_mainType = fct_reorder(NiN_mainType, km2)) %>%
+  ggplot(aes(x = NiN_mainType,
+             y = km2,
+             fill = NiN_mainTypeCoverage))+
+    geom_bar(stat = "identity",
+             colour = "grey40",
+             position = "stack")+
+    theme_bw(base_size = mySize)+
+    coord_flip()+
+  labs(x = "NiN main type",
+       y = expression(km^2),
+       fill = "Spatial coverage")+
+  scale_fill_brewer(palette = "Set1")+
+  facet_wrap(vars(Ecosystem),
+             scales = "free",
+             labeller = label_wrap_gen(width=25))
+
+```
+
+### For each NiN main type, the proportion mapped with each NiN variable
+
+### For each ecosystem, the proportion mapped with each NiN variable
 

### 2022-10-10 naturtype2.Rmd
nin variables draft

@@
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fuglefjell-eng og fugletopp"                        ] <- "T8  | all"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- "T32 | all (if including sub-types)"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Høgereligende og nordlig nedbørsmyr"                ] <- "V3  | all (if including sub-types)" # torvmarksformene are excluded
@@
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kystlynghei"                                        ] <- "T34 | all"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- "T32 | all (if including sub-types)"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Nakent tørkeutsatt kalkberg"                        ] <- "T1  | calcareous and dry"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- "T32 | all (if including sub-types)"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øvre sandstrand uten pionervegetasjon"              ] <- "T29 | sandy and vegetated"
@@
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- "T17 | all (if including sub-types)"
-ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | special management"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | all (if including sub-types)"
 ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- "V9  | all (if including sub-types)"
@@
 ```{r}
+# should switch to pivot_wider
 ntyp_vars_wide <- data.table::dcast(setDT(ntyp_vars),
@@
 
-This dat set contains the nature types that were only mappen in 2018 or 2019, and which we removed from ntypDF2. 
+This data set contains the nature types that were only mapped in 2018 or 2019, and which we removed from ntypDF2. 
 ```{r}
@@
 
+From the figure above I take that *Naturlig åpne områder* is very poorly represented in general, and that there are many NiN main types that are completely missing from the dataset.
+The three most common types, however, have complete thematic coverage (in excluding the year 2018 and including all sub-types).
+I will need to investigate if there are some NiN variables for the nature types mapping T18 Åpen flomfastmark, T12 Strandeng and T21 Sanddynemark which we can use.
+
+For *Semi-naturlig mark* we have quite good thematic coverage for the thre main types.The other three types probably makes up a considerably smaller area, but note that T33 has some coverage, and is classes as *all (-2018)*.
+Of the thre main ecosystems, semi-naturlig mark dominates clearly.
+I will need to look for NiN variabels associated with the maooing units for these main NiN types.
+I can also possibly explore the distribution of the mapped polygons in environmental space and compare it agains AR5 as a reference.
+
+For *Våtmark* we have good thematic coverage for V3 nedbørsmyr and V9 semi-naturlig myr, but for V1 åpen jordvassmyr only calcareous localities are mapped and assessed.
+A question is then whether, for a given NiN variable, that calcareous localities can be representative for all mires (ppor and rich) or if this will intrucude too much bias in one way or another.
+
+I will give priority to *Naturlig åpne områder* because of the poor data availabiliy for this ecosystem in general.
+
 ### For each NiN main type, the proportion mapped with each NiN variable
+
+```{r}
+areaSum <- ntyp_fill2 %>%
+  group_by(Ecosystem) %>%
+  summarise(km2 = sum(km2)) %>%
+  tibble::add_column(.before=1,
+                     NiN_code = rep("Combined", 3))
+
+```
+
+```{r, fig.cap="Barplot showing the proportion area for which each NiN variable is recorded.
+The last letter of the NiN variable code refers to the ecosystem tye, and is not part of the variable name.
+The total mapped area for each main ecosystem (the facets) is shown as a dashed red line.
+"}
+ntyp_fill2 %>%
+  pivot_longer(cols = unique(ntyp_vars$NiN_variable_code)) %>%
+  drop_na(value) %>%
+  rename(NiN_code = name,
+         useFrequency = value) %>%
+  filter(useFrequency > 0) %>%
+  group_by(NiN_code, Ecosystem) %>%
+  summarise(km2 = sum(km2)) %>%
+  arrange(km2) %>%
+  mutate(lab = paste(NiN_code, substr(Ecosystem, start = 1, stop = 1)), sep = " ") %>%
+  group_by(Ecosystem) %>%
+  mutate(lab = forcats::fct_reorder(lab, km2)) %>%
+  ggplot(aes(x = lab,
+              y = km2))+
+  geom_bar(stat = "identity",
+           colour = "grey40",
+           fill = "grey80")+
+  geom_hline(data = areaSum,
+             aes(yintercept = km2),
+             colour = "red",
+             linetype = "dashed")+
+  facet_wrap(vars(Ecosystem),
+             scales = "free",
+             labeller = label_wrap_gen(width=25))+
+  labs(x = "",
+       y = expression(km^2))+
+  coord_flip()+
+  theme_bw(base_size = mySize)
+```
 

### 2022-10-10 naturtype2.Rmd
finish

@@
-# On the application of the naturetype dataset - Part 2 - Representativity {#naturtype2}
+# Nature types - Part 2
+
+On the application of the naturetype dataset - Part 2 - Representativity {#naturtype2}
 
 <br />
-
-*Work in progress*
 
@@
 
-#### 7TK and /SE
+#### 7TK and 7SE (kjørespor & slitasje)
 7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to treat these as NA's.
@@
 ```{r, fig.cap="Barplot showing the proportion area for which each NiN variable is recorded.
-The last letter of the NiN variable code refers to the ecosystem tye, and is not part of the variable name.
 The total mapped area for each main ecosystem (the facets) is shown as a dashed red line.
-"}
+The three most dominant Nin main types for each ecosystem is given a uniqe colour, and all the remaining are grouped as 'other'."}
 ntyp_fill2 %>%
@@
   filter(useFrequency > 0) %>%
-  group_by(NiN_code, Ecosystem) %>%
+  group_by(NiN_code, Ecosystem, NiN_mainType) %>%  # NiN main type could be replaced with Nature_type
   summarise(km2 = sum(km2)) %>%
-  arrange(km2) %>%
-  mutate(lab = paste(NiN_code, substr(Ecosystem, start = 1, stop = 1)), sep = " ") %>%
+
+  # create a truncated fill variable 
   group_by(Ecosystem) %>%
-  mutate(lab = forcats::fct_reorder(lab, km2)) %>%
-  ggplot(aes(x = lab,
-              y = km2))+
+  mutate(myFill = fct_lump(NiN_mainType, 3, w = km2)) %>%
+
+  # qick fix to sort variable across facets (dont work after adding 'Nature_type' in the first group_by)
+  #mutate(lab = paste(NiN_code, substr(Ecosystem, start = 1, stop = 1), sep = " ")) %>%
+  #ungroup()%>%
+  #mutate(lab = forcats::fct_reorder(lab, km2)) %>%
+
+  # plot
+  ggplot(aes(x = NiN_code,
+              y = km2,
+             fill = myFill
+                ))+
   geom_bar(stat = "identity",
            colour = "grey40",
-           fill = "grey80")+
+           #fill = "grey80",
+           position = "stack")+
   geom_hline(data = areaSum,
@@
   labs(x = "",
-       y = expression(km^2))+
+       y = expression(km^2),
+       fill = "")+
   coord_flip()+
+  scale_fill_brewer(palette = "Set1")+
   theme_bw(base_size = mySize)
@@
 
-### For each ecosystem, the proportion mapped with each NiN variable
+This figure, in combination with the above, points to he most obvious NiN variable candidates.
 
-Next, are the coverage of NiN main types representative of the Hovedøkosystem?
-Then I need to manually enter the NiN main types associated with each Hovedøkosystem and perhaps add them as blank rows in ntypDF2.
+7TK, /SE and 7FA are good candidates for *Naturlig åpne områder*. 7JA-BT is probably best classes as a pressures indikcator. Same with 7VR-RI with is recorded for all T18 localities.
 
-Candidate nature types or groups of nature types that may be suitable to describe ecosystem condition for a NiN main type or Hovedøkosystem will be subject to further testing to look for environmental bias.
+The same three variables are the best candidates also for *Semi-natural areas*.
 
-```{r}
-ntypDF$coverage[ntypDF$Nature_type == "Slåttemark"] <- "Partial"
-```
-
-```{r}
-codes <- unique(dat2L2$NiN_variable_code)
-```
-
-Next, add 'codes' in seperate rows (wide) and cast to columns
-
--------------------------
-
-```{r}
-wet_1_long2 <- wet_1_long[!is.na(wet_1_long$NiN_variable_value),]
-```
-
-This leaves us with the following parameters.
-```{r}
-unique(wet_1_long2$NiN_variable_code)
-```
-We can probably exclude several of these as non-relevant.
-
-1AG-B is the total shrub cover
-7TK is *kjøreskade*
-1AG-A-G is *dekning gjenveksttrær*
-
-```{r}
-ntyp <- unique(wet_1_long$naturtype)
-```
-
-```{r}
-
-temp <- wet_1_long[wet_1_long$naturtype == ntyp[1],]
-table(temp$naturtype)
-
-ggplot(temp)+
-  geom_histogram(aes(x = NiN_variable_value))+
-  facet_wrap(vars(NiN_variable_code))
-```
-
-------------------------------------------
-
-### Våtmark
-I will start by looking at wetlands (våtmark).
-
-```{r}
-wet <- myData[myData$hovedøkosystem == "Våtmark",]
-any(wet[wet$naturtype == " Hule eiker",])
-```
-
-How many mosaic localities do we have?
-```{r}
-wet_Mos <- wet[wet$mosaikk == "Ja",]
-nrow(wet_Mos)
-```
-275, compared to 
-```{r}
-wet_noMos <- wet[wet$mosaikk != "Ja",]
-nrow(wet_noMos)
-```
-118550 *normal* localities.
-
-Let's plot the summed area of each NiN main type, excluding mosaic localities.
-```{r, fig.cap="Total mapped are for each NiN main type in the wetland ecosystem (excluding mosaics)"}
-temp <- stats::aggregate(data = wet_noMos,
-                                 area~ninkartleggingsenheter3, FUN = sum)
-
-names(temp)[2] <- "area_m2"
-temp$area_km2 <- drop_units(temp$area_m2/10^6)
-
-ggplot(temp,
-       aes(x = ninkartleggingsenheter3,
-               y = area_km2))+
-  geom_bar(
-           fill="grey",
-           colour="black",
-           stat="identity")+
-  coord_flip()+
-  theme_bw(base_size = 12)+
-  labs(y = "Total mapped area (km/2)",
-       x = "NiN main types")
-```
-
-If we include mosaic localities we get some other main types included as well
-```{r, fig.cap="Total mapped are for each NiN main type in the wetland ecosystem (including mosaics)"}
-wet_long <- tidyr::separate_rows(wet, ninkartleggingsenheter3)
-wet_long_agg <- stats::aggregate(data = wet_long,
-                                 area~ninkartleggingsenheter3, FUN = sum)
-names(wet_long_agg)[2] <- "area_m2"
-wet_long_agg$area_km2 <- drop_units(wet_long_agg$area_m2/10^6)
-ggplot(wet_long_agg,
-       aes(x = ninkartleggingsenheter3,
-               y = area_km2))+
-  geom_bar(
-           fill="grey",
-           colour="black",
-           stat="identity")+
-  coord_flip()+
-  theme_bw(base_size = 12)+
-  labs(y = "Total mapped area (km/2)",
-       x = "NiN main types")
-```
-
-Especially T34 (kystlynghei) and T1 (nakent berg) commonly occur in mosaics with wetlands.
-
-V5, V6 are V7 are not represented at all. V5 is *Varm kilde* which we don't have in mainland Norway. V6 and V7 might be linked to the alpine ecosystem.
-
-### Kalkrik myr- og sumpskogsmark
-Only mapped in 2018.
-Back then, the defining NiN variables were not clearly stated, but the defining text clearly show that this is forested land, and so it should not be included in our assessment since our operational definition of Våtmark is *Åpen våtmark / open wetlands*.
-
-```{r}
-ntypDF$Eligible[ntypDF$Nature_type == "Kalkrik myr- og sumpskogsmark"] <- "No"
-```
-
-### Slåttemyr
-Mapped in all years and covers the entire NiN space in V9. 
-```{r}
-ntypDF$Eligible[ntypDF$Nature_type == "Slåttemyr"] <- "Yes"
-ntypDF$NiN_mainType[ntypDF$Nature_type == "Slåttemyr"] <- "V9"
-ntypDF$NiN_mainType_coverage[ntypDF$Nature_type == "Slåttemyr"] <- "Complete"
-
-ntypDF$'Busksjikt-dekning (1AG-B)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
-ntypDF$'Grøftings-intensitet (7GR-GI)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
-ntypDF$'Dekning av gjenvekst-trær (1AG-A-G)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
-ntypDF$'Andel av vedvekster i feltsjiktet (1AR-C-L)'[ntypDF$Nature_type == "Slåttemyr"] <- "Recorded in all years."
-
-ntypDF$Remarks[ntypDF$Nature_type == "Slåttemyr"] <- "The list of NiN condition variables has changed over the years, and possibly also the different threshold values."
-```
-
-## Environmental space
-
--   Extract environmental data from the polygons and compare against reference data points.
--   Uni-variate (e.g. *boreal hei* plotted against elevation, oceanity, distance from roads, etc)
-
-## Conclution 1
-
-Conclude about which main ecosystem types that can reliably assess using the available mapping units?
-
-What about suitability for defining reference values?
-
-## Suitability
-
-In the subset of mapping units that can be said to represent an ecosystem group, which NiN or MdirPRAM variables should we use?
-
-## Concrete examples
-
-### Andel lynghei med minst to faser i kystlynghei
-
-### Areal uten skadet og/eller død røsslyng i kystlynghei
-
-### Aktuell bruksintensitet
+For *Våtmark*, 7GR-Gi is the most common variables, but this is a pressure variable.
+But, perhaps it could be used as a surrogate indicator.
+/SE and 7TK+PRTK are other candidates.
+PRTO is only relevant for V3, but covers a relatively large area.

### 2022-10-11 naturtype2.Rmd
complete part 2

@@
-# Nature types - Part 2
+# Nature types - Part 2 {#naturtype2}
 
-On the application of the naturetype dataset - Part 2 - Representativity {#naturtype2}
+**On the application of the naturetype dataset - Part 2 - Representativity **
 
@@
 Date: `r Sys.Date()`
+
+<br />
 
@@
 
-We have 66 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *Open Wetland*.
+We have 66 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *open* wetland.
 
@@
 
-The way to do this I think is to look at each nature type in turn and map the NiN main types and sub-types that they cover. At the same time I can extract the NiN variables and values to see.
+The way to do this I think is to look at each nature type in turn and map the NiN main types and sub-types that they cover.
+At the same time I can extract the NiN variables and values to see how they distribute over the NiN main types.
 
@@
 
-# Convert values to numeric. This causes some NA's which I will go though below
+# Convert values to numeric. This causes some NA's which I will go through below
 dat2L$NiN_variable_value <- as.numeric(dat2L$NiN_variable_value)
@@
 
+> Biodiversity variables were only recorded when condition was better than *very poor*. This bias means we cannot use these variables.
+
 #### 4TG (old trees)
@@
 #### 4TL (trees with fire scars)
-These five variables are ised together with 4TG as biodiversity variables in 2018.
+These five variables are used together with 4TG as biodiversity variables in 2018.
 ```{r}
@@
 
-#### 5AB and 5BY
-These area land use types (arealbruksklasser) and building types, respectively, and the values are not numeric, but categorical, and represents presence or absence of these land uses or objects.
+#### 5AB and 5BY (arealbruksklasser og byggningstyper)
+These area land use types and building types, respectively, and the values are not numeric, but categorical, and represents presence or absence of these land uses or objects.
 These are not suited for automatically determining ecosystem condition as this is done subjectively in the field.
@@
 #### 6SE and 6SO (Bioclimatical sections and sones)
-Not relevant here as a condition variable.
+Not relevant here as condition variables.
 ```{r}
@@
 Should not have been allowed the value X, so Ok to exclude these.
+7FA also has potentially an issue related to how the NiN-app as build which doesn't allow for registering alien species that are not on the *fremmedartsliste*.
 
 #### 7GR-GI
-7GR-GI (grøftingsintensitet) should not have the value X either, so OK to exclude these.
+7GR-GI (grøftingsintensitet) should not have the value X either, so OK to exclude these NA's.
+But we keep the variable, even though its clearly a pressure indikator, we might want to use it as a surrogate for mire hydrology.
 
@@
 ```
-That is not the case. If the condition is scored as low, the variable does not even appear in the data set for that locality.
+That is not the case: If the condition is scored as low, the variable does not even appear in the data set for that locality.
 
-We can also make a note that a big proportion of the total number of localities of *Kystlynghei* had very poor condition, and therefore this variable 7JB-KU was not recorded.
+We can also make a note that a big proportion of the total number of localities of *Kystlynghei* had very poor condition, and therefore this variable 7JB-KU was not recorded in a large proportion of the localities.
 ```{r, fig.cap="The distribution of condition scores for Kystlynghei."}
@@
 
-#### 7JB-BA = Aktuell bruksintensitet
+#### 7JB-BA (Aktuell bruksintensitet)
 A common condition variable in semi-natural nature types. It's a borderline pressure indicator, but I will keep it in the data set for now at least.
@@
 
-#### 7JB-BT = Beitetrykk
+#### 7JB-BT (Beitetrykk)
 Similar variable to the above.
@@
 
-#### 7JB-GJ = Gjødsling
+#### 7JB-GJ (Gjødsling)
 Similar variable to the two above.
@@
 
+This variable is used as a condition indicator, but this could be questioned.
+Firstly, it is a pressure indicator.
+Secondly, many semi-natural areas where fertilized, also in the reference condition.
+The people that made the meadows would probably not agree that a productive field/meadow has poor condition.
+This condition variable is therefor more directly tuned towards the maintenance of biodiversity.
+
 #### 7JB-HT (Høsting av tresjiktet)
-Used for *Lauveng*in 2019 and 2020.
+Used for *Lauveng* in 2019 and 2020.
 ```{r}
@@
 ```
-This is so marginal that I will exclude then already now.
+This is so marginal that I will exclude these already now.
 ```{r}
@@
 
-#### 7SD
-7SD-NS and 7SD-0 was also only used for one nature type, and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*.
+#### 7SD-0 and 7SD- NS (Natur- og normalskogsdynamikk)
+7SD-NS and 7SD-0 was also only used for only one nature type (which is not forest), and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*.
 *X* therefore means *not relevant*, and we can also exclude the parameters from the dataset all together.
@@
 
-#### 7VR-RI
-*Reguleringsintensitet*
+#### 7VR-RI (Reguleringsintensitet)
 This variable is defined as numeric between 1-5. There is, however, quite a few cases where it is X.
@@
     + Related to 5BY and 5AB
-    + Exclude (se 5BY and 5AB)
+    + Exclude (see 5BY and 5AB)
 * PRH
@@
 
-I will also now exclude nature types that were only mapped in 2018 and/or and not after that.
+I will also now exclude nature types that were only mapped in 2018 and/or 2019 and not after that.
 These nature types will not only not get more data in the future, but also they were mapped in a time when the methodology was quite unstable.
@@
 I could also extract the NiN sub types (grunntyper), but it would become too messy.
+Therefor I will create a second column with a textual/categorical description of the degree of themaic coverage.
 
@@
 names(ntyp_vars) <- c("Nature_type", "NiN_variable_code", "NiN_variable_count")
-(ntyp_vars <- ntyp_vars[ntyp_vars$NiN_variable_count >0,])
+ntyp_vars <- ntyp_vars[ntyp_vars$NiN_variable_count >0,]
+head(ntyp_vars)
 ```
 
-Add a column for the total nmber of localities for each nature type and get the percentage of localities where each NiN has been recorded.
+Add a column for the total number of localities for each nature type and get the percentage of localities where each NiN has been recorded.
 ```{r}
@@
 
-This data set contains the nature types that were only mapped in 2018 or 2019, and which we removed from ntypDF2. 
+This data set contains the nature types that were only mapped in 2018 or 2019, and which we removed from ntypDF2:
 ```{r}
@@
 ```
-This is Våtsnøleie, and my guess is that it IS mapped, but grouped with the aloine ecosystem. I'll therefore not add it in here, but mention it in the figure caption below.
+V6 is *Våtsnøleie*, and my guess is that it *is* mapped, but grouped with the alpine ecosystem. I'll therefore not add it in here, but mention it in the figure caption below.
 
@@
 From the figure above I take that *Naturlig åpne områder* is very poorly represented in general, and that there are many NiN main types that are completely missing from the dataset.
-The three most common types, however, have complete thematic coverage (in excluding the year 2018 and including all sub-types).
+The three most common types, however, have complete thematic coverage (if excluding the year 2018 and including all sub-types).
 I will need to investigate if there are some NiN variables for the nature types mapping T18 Åpen flomfastmark, T12 Strandeng and T21 Sanddynemark which we can use.
 
-For *Semi-naturlig mark* we have quite good thematic coverage for the thre main types.The other three types probably makes up a considerably smaller area, but note that T33 has some coverage, and is classes as *all (-2018)*.
-Of the thre main ecosystems, semi-naturlig mark dominates clearly.
-I will need to look for NiN variabels associated with the maooing units for these main NiN types.
-I can also possibly explore the distribution of the mapped polygons in environmental space and compare it agains AR5 as a reference.
+For *Semi-naturlig mark* we have quite good thematic coverage for the three main types.
+The other three types probably makes up a considerably smaller area, but note that T33 has some coverage, and is classes as *all (-2018)*.
+Of the three main ecosystems, semi-naturlig mark dominates clearly.
+I will need to look for NiN variables associated with the mapping units for these main NiN types.
+I can also possibly explore the distribution of the mapped polygons in environmental space and compare it against AR5 as a reference, but I'm not sure how good AR5 is for anything whihc is not cropland.
 
 For *Våtmark* we have good thematic coverage for V3 nedbørsmyr and V9 semi-naturlig myr, but for V1 åpen jordvassmyr only calcareous localities are mapped and assessed.
-A question is then whether, for a given NiN variable, that calcareous localities can be representative for all mires (ppor and rich) or if this will intrucude too much bias in one way or another.
-
-I will give priority to *Naturlig åpne områder* because of the poor data availabiliy for this ecosystem in general.
+A question is then whether, for a given NiN variable, that calcareous localities can be representative for all mires (poor and rich) or if this will intruduce too much bias in one way or another.
 
@@
 The total mapped area for each main ecosystem (the facets) is shown as a dashed red line.
-The three most dominant Nin main types for each ecosystem is given a uniqe colour, and all the remaining are grouped as 'other'."}
+The three most dominant NiN main types for each ecosystem is given a uniqe colour, and all the remaining are grouped as 'other'."}
 ntyp_fill2 %>%
@@
 
+## Tentative conclusions and the way forward
 This figure, in combination with the above, points to he most obvious NiN variable candidates.
 
-7TK, /SE and 7FA are good candidates for *Naturlig åpne områder*. 7JA-BT is probably best classes as a pressures indikcator. Same with 7VR-RI with is recorded for all T18 localities.
+7SE and 7TK are good candidates for *Naturlig åpne områder*. 7JA-BT is probably more appropriately classed as a pressure indicator. Same with 7VR-RI with is recorded for all T18 localities. 
+7FA has potentially an issue related to how the NiN-app as build which doesn't allow for registering alien species that are not on the *fremmedartsliste*.
 
-The same three variables are the best candidates also for *Semi-natural areas*.
+The same two variables (7SE and 7TK) are the best candidates also for *Semi-natural areas*.
+7JB-BT is a pressure variable (*beitetrykk*).
+7RA-SJ/BH are related to *gjengoing* which will be covered by another indicator informed from remote sensing (LiDAR), and we therefor don't need to describe this with a field-based indicator as well.
 
-For *Våtmark*, 7GR-Gi is the most common variables, but this is a pressure variable.
+For *Våtmark*, 7GR-Gi is the most common variables, but this is a pressure variable (*grøfting*).
 But, perhaps it could be used as a surrogate indicator.
-/SE and 7TK+PRTK are other candidates.
+I think this is justifiable, especially since the relationship between the pressure (grøfting) and the condition (hydrology) is so well known.
+7SE and 7TK+PRTK are other candidates.
 PRTO is only relevant for V3, but covers a relatively large area.
+
+Next, I will focus in *Våtmark*-variables, since here we can use a new ecosystem delineation map as a reference to evaluate the environamlental and spatial bias in the dataset.
+This is not currently possible for the other two ecosystems.
+
+## Tables
+Here are some output tables with numbers the numbers underlying the above figures.
+
+```{r, fig.cap= "List of 45 nature types additionam data, including the proportion of localities for which there is data for each of the NiN variables."}
+DT::datatable(ntyp_fill2,
+              #extensions = "FixedColumns",
+  options = list(
+    scrollX = T,
+    scrollY=T,
+    pageLength = 10))
+```
+
+```{r, fig.cap= "List of unique combinatios of NiN variable codes and naturetypes with the summed area for which we have this data."}
+DT::datatable(
+  ntyp_fill2 %>%
+  pivot_longer(cols = unique(ntyp_vars$NiN_variable_code)) %>%
+  drop_na(value) %>%
+  rename(NiN_code = name,
+         useFrequency = value) %>%
+  filter(useFrequency > 0) %>%
+  group_by(NiN_code, Ecosystem, NiN_mainType) %>%  # NiN main type could be replaced with Nature_type
+  summarise(km2 = sum(km2)),
+  options = list(
+    scrollX = T,
+    scrollY=T,
+    pageLength = 10))
+```

### 2022-11-09 naturtype2.Rmd
typo

@@
             "Våtmark",
-            "Naturlig åpne områder under skoggrensa")
+            "Naturlig åpne områder under skoggresa")
 dat <- dat[dat$hovedøkosystem %in% target,]

### 2022-11-09 DRAFT_slitasjeOgKjørespor.Rmd
add chapter

--- 2022-08-12 AndersTest.Rmd
@@
 ---
-title: "Template for indicator documentation"
+title: "Slitasje og kjørespor"
 output:
@@
 
-This is a test
-
-<!-- This is a template for how to document the indicator analyses.
-
-Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name te file DRAFT_myIndicatorName.Rmd.
-
-Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
-you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work.
-
-Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout.
-
-When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
-the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. When completed, Anders will make a copy
-of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
-the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
-
-<!-- Examples:
-
-For an example, see 'indicators/breareal-workingFile.Rmd' 
-and the subsequent 'breareal.Rmd' 
-which again, when rendered, produces this:
-https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
-
-Notice the degree of annotation.
-
--->
-
 <!-- Replace 'indicator name' with yor own indicator name  -->
-# Indicator name
+# Slitasje og kjørespor
 
@@
 _Author and date:_
-<!-- Write your name here -->
+Anders L. Kolstad
 
@@
 library(knitr)
+library(sf)
 knitr::opts_chunk$set(echo = TRUE)
@@
 ```{r, echo=F}
-Ecosystem <- "" # e.g. "Skog og fjell"
-Egenskap  <- "" # e.g. "Primærproduksjon"
-ECT       <- "" # e.g. "Structural state characteristic"
-Contact   <- "" # e.g. "Anders Kolstad"
+Ecosystem <- "Våtmark, Naturlig åpne områder under skoggrensa, Semi-naturlig mark" 
+Egenskap  <- "Primærproduksjon (evt. Abiotiske fohold)"
+ECT       <- "Physical state characteristics" 
+Contact   <- "anders.kolstad@nina.no"
 ```
@@
 ### Data sets
-<!-- Text here -->
 
-#### Data set A
-<!-- Text here -->
+#### Nature type mapping
+This indicator uses the dataset `Naturtyper etter Miljødirektoratets Instruks`, which can be found [here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
+The dataset has already been [cleaned](#naturtype1) and [initial analyses on thematic representativity](#naturtype2) has been done.
 
-#### Data set B
-<!-- Text here -->
+```{r, eval=T}
+dat <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+```
+
+#### Vegetation zones (bioregions)
 

### 2022-11-09 naturtype2.Rmd
add lost nin types

--- 2022-10-11 naturtype2.Rmd
@@
 
-## Naturetypes, condition variables, and their linkage to NiN main types
+## Resarch question
 Only a subset of nature types are mapped.
@@
 We will only include our target ecosystems, and we will exclude *Hule eiker*.
+
+```{r}
+unique(dat$hovedøkosystem)
+```
+
 ```{r}
@@
 ```{r}
-ntyp_fill2 <- ntyp_fill %>%
+ntyp_fill2 <- ntyp_fill2 %>%
   add_row(NiN_mainType = "T40", Ecosystem = "Semi-naturlig mark", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
@@
 V6 is *Våtsnøleie*, and my guess is that it *is* mapped, but grouped with the alpine ecosystem. I'll therefore not add it in here, but mention it in the figure caption below.
+
+### Export summary dataset
+```{r, eval = F}
+saveRDS(ntyp_fill2, "data/naturetypes/natureType_summary.rds")
+```
 
@@
 I will need to look for NiN variables associated with the mapping units for these main NiN types.
-I can also possibly explore the distribution of the mapped polygons in environmental space and compare it against AR5 as a reference, but I'm not sure how good AR5 is for anything whihc is not cropland.
+I can also possibly explore the distribution of the mapped polygons in environmental space and compare it against AR5 as a reference, but I'm not sure how good AR5 is for anything which is not cropland.
 

### 2022-11-14 naturtype.Rmd
update title
merge chapters

--- 2022-09-15 naturtype.Rmd
@@
 
-# On the application of the naturetype dataset - Part 1 - Data exploration and cleaning {#naturtype1}
+# Nature types - Part 1 {#naturtype1}
+
+**On the application of the naturetype dataset - Part 1 - Data exploration and cleaning**
 
@@
 
-*Work in progress*
-
 Author: Anders L. Kolstad
@@
 
+<br />
+
 Here I will investigate a specific data set, [Naturtyper - Miljødirektoratets instruks](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c) and try to evaluate its usability for designing indicators for ecosystem condition.
-This involves assessing both the spatial and temporal representation, as well as conseptual alignment with the [Norwegian system for ecosystem condition assessments](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/).
+This involves assessing both the spatial and temporal representation, as well as conceptual alignment with the [Norwegian system for ecosystem condition assessments](https://www.regjeringen.no/no/dokumenter/fagsystem-for-fastsetting-av-god-okologisk-tilstand/id2558481/).
 
@@
 
-The first part of this analysis is simply to get an overview of the data and to clean it, making it ready for part 2.
+The first part of this analysis is simply to get an overview of the data, making it ready for part 2 where we look at the thematic coverage of nature types and how the NiN variables are used.
 
@@
 #path <- "data/R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+
 # server path
-#path <- "/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+path <- "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
 
 # temporary path 
-path <- "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+#path <- "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.g#db"
 dat <- sf::st_read(dsn = path)
@@
 
-Fix non-valid polygons
-
+Fix non-valid polygons:
 ```{r}
@@
 
+The largest polygon is a 
+```{r, echo=F}
+dat$naturtype[dat$area == max(dat$area)]
+```
+
+The smallest polygon is a 
+```{r, echo=F}
+dat$naturtype[dat$area == min(dat$area)]
+```
+
 Sum of mapped area divided by Norwegian mainland area:
 ```{r}
+#Import outline of mainland Norway
 nor <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
@@
 ```
-About 0.5% of Norway has been mapped (not that a bigger area than this has been surveyed, but only a small fraction is delineated). It is therefore essential that these 0.5% are representative.
+About 0.5% of Norway has been mapped (note that a bigger area than this has been surveyed, but only a small fraction is delineated). It is therefore essential that these 0.5% are representative.
 
@@
 `Isinnfrysingsmark` is assessed for condition now, but not in 2018.
-We can therefore exclude these localities from our data set.
+We can therefore exclude these localities from our data set:
 ```{r}
-dat <- dat[dat$tilstand!="",]
+dat_rm <- dat
+dat <- dat_rm[dat_rm$tilstand!="",]
+```
+
+This resulted in the deletion of `r nrow(dat_rm)-nrow(dat)` rows, or `r round(((nrow(dat_rm)-nrow(dat))/nrow(dat_rm))*100, 2)` % of the data.
+
+```{r, include=FALSE}
+rm(dat_rm)
 ```
@@
 ```
-
-Fix duplicated `hovedøkosystem`
+Mosaic localities occur in many main ecosystems (and many nature types therein). 
 ```{r}
+# Fix duplicated `hovedøkosystem`
 #unique(dat$hovedøkosystem)
@@
 
-Mosaic localities occur in many main ecosystems (and many nature types therein).
-
 Some conclusion here could be that
@@
 ## NiN types across main ecosystems
-The NiN main types can be extracted from the column `ninkartleggingsenheter`, but it is oddly concatenated. Let's tease it apart.
+The NiN main types can be extracted from the column `ninkartleggingsenheter`.
+This is the mapping NiN mapping units recorded in the field.
+The NiN main type makes up the first part of this mapping unit code.
+The variable is oddly concatenated.
+Let's tease it apart.
 
@@
 
-Mosaic localities have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (about 8000 cases), so we need to see whats going on there.
+Mosaic localities (right pane) have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (about 8000 cases).
+Most, if not all, nature types are defined within a single NiN main type, so we need to see whats going on there.
 
@@
 There is a lot of miss-match between NiN main types and *hovedøkosystem*. 
-Taking one example: T2 (åpne grunnledt mark) is not found in mountains, but there is one case where this occurs. If I view that case
+Taking one example: T2 (åpen grunnledt mark) is not found in mountains, but there is one case where this occurs. If I view that case
 
@@
 ..
-and find the online fact sheet for that locality, I see that it is indeed not a mosaic.
-It is a nature type called *Kalkfattig og intermediær fjellhei, leside og tundra* which is defined as strictly T3, but the field surveyor has recorded lots of main NiN types.
+and find the online fact sheet for that locality, I see that it is actually NOT a mosaic.
+It is a nature type called *Kalkfattig og intermediær fjellhei, leside og tundra* which is defined as strictly T3, but the field surveyor has recorded lots of NiN mapping units (and hence, main NiN types) in addition.
 This is a mistake.
 Online I can see that the locality is 50% T3, but this information is not in the data set that we have available.
-And the order of the NiN types in the `ninkartleggingsenheter` column is not reflecting the commonness of the types ether.
+Miljødirektoratet could consider adding this information to the downloadable dataset.
+The order of the NiN types in the `ninkartleggingsenheter` column is not reflecting the commonness of the types ether.
 
 If there was a way to extract the defining NiN type from the `naturtype` column, that would be nice.
+Miljødirektoratet may consider adding this as well.
 If we exclude all localities that are not mosaics, but that have multiple NiN main types, we first need to know if there are not any `naturtype` which allow for multiple NiN main types.
 
-As this next figure show, these cases are not correlated to mapping year, and hence to any changes in the field protocol.
-```{r}
+As this next figure show, these cases are not restricted to mapping year, and hence to any changes in the field protocol. Rather this phenomenon is increasing in commonness.
+```{r, fig.cap="Temporal trend in the recording of non-mosaic localities that are recorded as consisting of multiple NiN mapping units from different NiN main types."}
 par(mar=c(2,6,2,2))
@@
 
+Lets look at the most common nature types that are recorded this way.
 ```{r, fig.cap="The ten most common nature types and the number of localities with with multiple NiN main types"}
@@
 The same is true for *naturbeitemark* which is restricted to T32.
-Maybe these should all have been recorded as mosaic localities.
-In any case, we cannot automatically extract the main NiN type from these rows now.
-One option is to manually define all *naturbeitemark* as T32, and all *kystlynghei* as T34, but this is also not without risk.
+Maybe these localities should all have been recorded as mosaic localities. In any case, we cannot automatically extract the main NiN type from these rows now.
 
-Lets look at how the NiN main types cover the different *hovedøkosystem* when we exclude *Hule eiker* and all localities that have more than one NiN main type.
+The remaining option is to manually assign the defining NiN main type to each nature type.
+This data is not included in the data set, and would need to be sourced from the mapping protocols for all the different field seasons.
+
+For a deeper analyses into the causes for localities being mapped to NiN main types outside the definition of the nature type, see [Appendix 1].
+
+## Assigning NiN main types to Nature types
+The NiN main types need to be added manually to the nature types, as the field recorded data is error prone, and the defining NiN units for each nature type is not included in the data set.
+
+I will start by excluding all but the three target ecosystems.
+```{r}
+dat <- dat[dat$hovedøkosystem %in% target,]
+```
+
+I will also exclude `Hule eiker`, as this nature type is not easy to tie to a single main ecosystem
+
+```{r}
+dat <- dat[dat$naturtype != "Hule eiker",]
+```
+
+Get the unique nature types
+```{r}
+ntyp <- unique(dat$naturtype)
+```
+
+Extract the year when these were mapped
+```{r}
+years <- NULL
+for(i in 1:length(ntyp)){
+years[i] <- paste(sort(unique(dat$kartleggingsår[dat$naturtype == ntyp[i]])), collapse = ", ")
+}
+```
+
+Extract the associated ecosystem
+```{r}
+eco <- NULL
+for(i in 1:length(ntyp)){
+eco[i] <- paste(unique(dat$hovedøkosystem[dat$naturtype == ntyp[i]]), collapse = ", ")
+}
+```
+
+Combine into one data frame
+```{r}
+ntypDF <- data.frame(
+  "Nature_type" = ntyp,
+  "Year"        = years,
+  "Ecosystem"   = eco
+)
+```
+
+We have 66 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *open* wetland.
+
+```{r}
+excl_nt <- c("Kalkrik myr- og sumpskogsmark",
+             #"Flommyr, myrkant og myrskogsmark",  # Only V9 is relevant. This type also includes V2 and V8
+             "Gammel fattig sumpskog",
+             "Rik gransumpskog",
+             "Grankildeskog",
+             "Rik svartorsumpskog",
+             "Rik gråorsumpskog",
+             "Rik svartorstrandskog",
+             "Saltpåvirket svartorstrandskog",
+             "Kilde-edellauvskog",
+             "Saltpåvirket strand- og sumpskogsmark",
+             "Svak kilde og kildeskogsmark",
+             "Rik vierstrandskog",
+             "Varmekjær kildelauvskog"
+             )
+`%!in%` <- Negate(`%in%`)
+ntypDF <- ntypDF[ntypDF$Nature_type %!in% excl_nt,]
+```
+
+```{r, fig.cap="List of nature types showing the years when that nature type was mapped." }
+DT::datatable(ntypDF)
+```
+
+I will also now exclude nature types that were only mapped in 2018 and/or 2019 and not after that.
+These nature types will not only not get more data in the future, but also they were mapped in a time when the methodology was quite unstable.
+
+```{r}
+ntypDF2 <- ntypDF[ntypDF$Year != "2018" &
+                  ntypDF$Year != "2019" & 
+                  ntypDF$Year != "2018, 2019"# no types were mappend in 2018 & 2019 only
+                  ,]
+```
+
+Next I want to add the NiN main types.
+I need to look at the definition of each nature type and get the NiN code from there.
+I could also extract the NiN sub types (grunntyper), but it would become too messy.
+Therefor I will create a second column with a textual/categorical description of the degree of thematic coverage.
+
+I will look at the definitions of the nature type in the first and last year when that type was mapped, but not for the years in between.
+
+### Add NiN main type and degree of representativity
+```{r}
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Aktiv skredmark"                                    ] <- "T17 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen flomfastmark"                                  ] <- "T18 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen grunnlendt kalkrik mark i boreonemoral sone"   ] <- "T2  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Åpen grunnlendt kalkrik mark i sørboreal sone"      ] <- "T2  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Atlantisk høymyr"                                   ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Boreal hei"                                         ] <- "T31 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eksentrisk høymyr"                                  ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Eng-aktig sterkt endret fastmark"                   ] <- "T41 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosse-eng"                                          ] <- "T15 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fosseberg"                                          ] <- "T1  | extra wet"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fossepåvirket berg"                                 ] <- "T1  | extra wet"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Fuglefjell-eng og fugletopp"                        ] <- "T8  | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Hagemark"                                           ] <- "T32 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Høgereligende og nordlig nedbørsmyr"                ] <- "V3  | all (if including sub-types)" # torvmarksformene are excluded
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Isinnfrysingsmark"                                  ] <- "T20 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kalkrik helofyttsump"                               ] <- "L4 | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kanthøymyr"                                         ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Konsentrisk høymyr"                                 ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Kystlynghei"                                        ] <- "T34 | all"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Lauveng"                                            ] <- "T32 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Nakent tørkeutsatt kalkberg"                        ] <- "T1  | calcareous and dry"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Naturbeitemark"                                     ] <- "T32 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øvre sandstrand uten pionervegetasjon"              ] <- "T29 | sandy and vegetated"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Øyblandingsmyr"                                     ] <- "V1  | partial" # also includes V3
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Palsmyr"                                            ] <- "V3  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Platåhøymyr"                                        ] <- "V3  | all (if including sub-types)" # can also include smaller areas of V1
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i mellomboreal sone"          ] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen jordvannsmyr i nordboreal og lavalpin sone"] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Rik åpen sørlig jordvannsmyr"                       ] <- "V1  | calcareous"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sanddynemark"                                       ] <- "T21 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig eng"                                  ] <- "T32 | all (if including sub-types)"
+# This might be called Kulturmarkseng in the 2018 protocol, but Semi-naturlig eng in the data set. 
+# Kulturmarkseng also includes V10
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig myr"                                  ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig strandeng"                            ] <- "T33 | all (-2018)" # could perhaps be used in 2018, but it's defined awkwardly
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Semi-naturlig våteng"                               ] <- "V10 | all (-2018)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Silt og leirskred"                                  ] <- "T17 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemark"                                         ] <- "T32 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Slåttemyr"                                          ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig etablert sanddynemark"                       ] <- "T21 | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig kaldkilde"                                   ] <- "V4  | southern"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig nedbørsmyr"                                  ] <- "V3  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Sørlig slåttemyr"                                   ] <- "V9  | all (if including sub-types)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Strandeng"                                          ] <- "T12 | all (-2018)"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Svært tørkeutsatt sørlig kalkberg"                  ] <- "T1  | calcareous and dry"
+ntypDF2$hovedgruppe[ntypDF2$Nature_type == "Terrengdekkende myr"                                ] <- "V3  | all (if including sub-types)"
+
+```
+
+Split this new column in two.
+```{r}
+ntypDF2 <- ntypDF2 %>%
+  tidyr::separate(col = hovedgruppe,
+                  into = c("NiN_mainType", "NiN_mainTypeCoverage"),
+                  #sep = "|",
+                  remove=T,
+                  extra = "merge"
+              )
+```
+
+## Add NiN variables
+
+Preparing the data:
+```{r}
+#Exclude non-relevant forest wetland types
+dat2 <- dat[dat$naturtype %!in% excl_nt,]
+
+# Melt data
+dat2L <- tidyr::separate_rows(dat2, ninbeskrivelsesvariabler, sep=",")
+
+# Split the code and the value into separate columns
+dat2L <- tidyr::separate(dat2L, 
+                              col = ninbeskrivelsesvariabler,
+                              into = c("NiN_variable_code", "NiN_variable_value"),
+                              sep = "_",
+                              remove=F
+                              )
+# One NA produced here, but I check it, and it's fine.
+
+# Convert values to numeric. This causes some NA's which I will go through below
+dat2L$NiN_variable_value <- as.numeric(dat2L$NiN_variable_value)
+
+```
+
+Here are all the NiN variable codes:
+```{r}
+unique(sort(dat2L$NiN_variable_code))
+```
+
+Converting NiN_variable_value to numeric also introduced NA's for these categories: 
+```{r}
+sort(unique(dat2L$NiN_variable_code[is.na(dat2L$NiN_variable_value)]))
+```
+When I now go through the variable codes separately I will also judge if these NA's are real, or if they for example should be coded as zeros or something.
+I will exclude variables that are part of the biodiversity assessment of the localities, since these are not assessed for localities that have a very poor condition (this bias is not possible to correct).
+
+### 1AG-A (sjiktdekningsvariabler)
+These are different variables describing the cover in different vegetation strata and/or species
+
+* 1AG-A-0 = Total tre canopy cover
+* 1AG-B = Total shrub cover
+* 1AG-C = Total field layer cover
+* 1AG-A-E = 'Overstandere'
+* 1AG-A-G = 'gjenveksttrær'
+
+### 1AR-C-L (Andel vedvekster i fletsjiktet)
+A condition variable in some mire nature types
+
+### 3TO (Torvmarksformer)
+These are defining variables, and not part of the condition assessment.
+```{r}
+exclude <- c("3TO-BØ",
+             "3TO-HA",
+             "3TO-HE",
+             "3TO-HK",
+             "3TO-HN",
+             "3TO-HP",
+             "3TO-PA",
+             "3TO-TE")
+```
+
+### 4DL (Coarse dead wood)
+4DL-SS-0 and 4DL-S-0 has to do with the total amount of coarse woody debris/logs. It was only recorded for one nature type in 2018, and then as a a *biodiversity variable*, and not a state variable.
+
+```{r}
+exclude <- c(exclude,
+             "4DL-SS-0",
+             "4DL-S-0")
+```
+
+### 4TG (old trees)
+These three variables are used as *biodiversity variables* in  2018.
+```{r}
+exclude <- c(exclude,
+             "4TG-BL",
+             "4TG-EL",
+             "4TG-GF")
+```
+
+### 4TL (trees with fire scars)
+These five variables are used together with 4TG as *biodiversity variables* in 2018.
+```{r}
+exclude <- c(exclude,
+             "4TL-BS",
+             "4TL-HE",
+             "4TL-HL",
+             "4TL-RB",
+             "4TL-SB")
+```
+
+### 5AB and 5BY (arealbruksklasser og byggningstyper)
+These area land use types and building types, respectively, and the values are not numeric, but categorical, and represents presence or absence of these land uses or objects.
+These are not suited for automatically determining ecosystem condition as this is done subjectively in the field.
+```{r}
+exclude <- c(exclude,
+             "5AB-0",
+             "5AB-DO-TT",
+             "5BY-0"
+             )
+```
+
+### 6SE and 6SO (Bioclimatical sections and sones)
+Not relevant here as condition variables.
+```{r}
+exclude <- c(exclude,
+             "6SE",
+             "6SO")
+```
+
+### 7FA (fremmedartsinnslag)
+Should not have been allowed the value X, so Ok to keep these as NA.
+
+### 7GR-GI
+7GR-GI (grøftingsintensitet) should not have the value X either, so OK to exclude these NA's.
+But we keep the variable, even though its clearly a pressure indicator, we might want to use it as a surrogate for mire hydrology.
+
+### 7JB-BA (Aktuell bruksintensitet)
+A common condition/pressure variable in semi-natural nature types.
+
+Overview of cases where the value is set as *X*.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BA_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+These are just a few cases, mostly from 2018. OK to treat as NA.
+
+### 7JB-BT (Beitetrykk)
+Similar variable to the above.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BT_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+OK to treat X's as NA's.
+
+### 7JB-GJ (Gjødsling)
+Similar variable to the two above.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BT_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+OK to treat as X's NA's.
+
+This variable is used as a condition indicator, but it is perhaps strictly speaking a pressure indicator.
+Secondly, many semi-natural areas were fertilized, also in the reference condition.
+The people that made the meadows would probably not agree that a productive field/meadow has poor condition.
+This condition variable is therefor more directly tuned towards the maintenance of biodiversity.
+
+### 7JB-HT (Høsting av tresjiktet)
+Used for *Lauveng* in 2019 and 2020.
+```{r}
+temp <- dat2L[dat2L$NiN_variable_code == "7JB-HT-ST" | 
+               dat2L$NiN_variable_code == "7JB-HT-SL" ,]
+table(temp$kartleggingsår, temp$naturtype)
+```
+This is so marginal that I will exclude these already now.
+```{r}
+exclude <- c(exclude,
+             "7JB-HT-SL",
+             "7JB-HT-ST")
+```
+
+### 7JB-KU (Kystlyngheias utviklingsfaser)
+
+* BY = byggefase
+* DE = degenereringsfase
+* MO = moden fase
+* PI = pionerfase
+
+This variable is used as in the *biodiversity assessment*, and can therefore not be used in the condition assessment because localities with *very poor* conditions have not been mapped/assessed.
+```{r}
+exclude <- c(exclude, 
+             "7JB-KU-BY", 
+             "7JB-KU-DE", 
+             "7JB-KU-MO", 
+             "7JB-KU-PI")
+```
+
+I will nonetheless explore this variable a bit more since this variable has been suggested previously as a condition variable.
+NiN defines the possible values for these as numeric between 0 and 4 (shifted to become 1-5 in the data set), but there are 998 cases where it has been given then value *X*.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-KU-BY_X" |
+                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-DE_X" |
+                dat2L$ninbeskrivelsesvariabler == "7JB-KU-MO_X" |
+                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-PI_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+
+The reason could be that this variable is used in the biodiversity assessment, which is not performed if the condition is *very poor*, but the following table shows that the localities where this variable has been given the value X is spread across all condition scores:
+```{r}
+table(temp$tilstand)
+```
+
+We can also make a note that a big proportion of the total number of localities of *Kystlynghei* had very poor (svært redusert) condition, and therefore this variable 7JB-KU was not recorded in a large proportion of the localities.
+```{r, fig.cap="The distribution of condition scores for Kystlynghei."}
+KU <- dat2[dat2$naturtype=="Kystlynghei",]
+barplot(table(KU$tilstand))
+```
+
+The percentage of localities with very poor condition is `r round(nrow(KU[KU$tilstand=="Svært redusert",]) / nrow(KU), 2) *100`%.
+
+Each *kystlynghei* locality should have values for all four parameters (7BA-BY/DE/MO/PI), but if we look at some cases in more detail, to understand why some of these values have been set as X (probably means they were left blank), for example like this:
+```{r}
+#View(dat2L[dat2L$NiN_variable_code == "7JB-KU-BY" & is.na(dat2L$NiN_variable_value),])
+#View(dat2L[dat2L$identifikasjon_lokalid =="NINFP1810002975",]) # None of the 7JB-KU variables have values
+#View(dat2L[dat2L$identifikasjon_lokalid =="NINFP2110057689",]) # Two of the 7JB-KU variables have values
+```
+ ... we can see that sometimes <4 but >0 of the variables/phases has been given a score, but sometimes none have. This makes it problematic to set these NA's to be zeros.
+
+For this variable to be used in as an indicator for Ecological Condition in the future, the variable needs to be assessed for all localities, and the field protocol and/or field app needs to be revised to that it is not possible to record blank or NA values.
+
+### 7RA-BH and 7RA-SJ (rask suksesjon i boreal hei og i i semi-naturlig og sterkt endret jordbruksmark inkludert våteng)
+Condition variables.
+There are some very few cases where 7RA-SJ has the value X. It should be numeric 1-4, so the value X is a mistake here I think. We can exclude these (i.e. allow them to be NA's).
+
+### 7SD-0 and 7SD- NS (Natur- og normalskogsdynamikk)
+7SD-NS and 7SD-0 was also only used for only one nature type (which is not forest), and then only for wooded localities of *Flommyr, myrkant og myrskogsmark*.
+*X* therefore means *not relevant*, and we can also exclude the parameters from the dataset all together.
+
+```{r}
+exclude <- c(exclude, 
+             "7SD-0",     
+             "7SD-NS")
+```
+
+### 7TK and 7SE (kjørespor & slitasje)
+7TK (kjørespor) and 7SE (slitasje) are not allowed the value X, but it happened on some rare occasions (24) anyhow. It's fine to treat these as NA's.
+
+### 7VR-RI (Reguleringsintensitet)
+This variable is defined as numeric between 1-5. There are, however, quite a few cases where it is X.
+```{r}
+temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7VR-RI_X",]
+table(temp$kartleggingsår, temp$naturtype)
+```
+I cannot explain these, but in any case, this variable is not a good indicator for condition as it rather represents a pressure/driver. So OK to treat as NA's.
+
+*LKMs*  
+LKM's (lokal kompleks miljøvariabel) are not relevant to us here.
+
+### LKMKI
+    + Kalk (lime)
+    + Exclude
+
+### LKMSP
+    + Slåttemarkspreg
+    + Exclude
+
+*MdirPR-variables*
+These are Miljødirektoratets own variables (usually variations on existing NiN-variables)
+
+### PRAK
+    + Antall NiN-kartleggingsenheter
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRAM
+    + Menneskeskapte objekter
+    + Related to 5BY and 5AB
+    + Exclude (see 5BY and 5AB)
+
+### PRH
+    + This is perhaps a mistake and should be PRHA or PRHT
+    + Exclude
+
+### PRHA
+    + Habitat spesifikke arter
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRHT
+    + Høsting av tresjiktet
+
+### PRKA
+    + Kalkindikatorer
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRKU
+    + Kystlyngheias utviklingsfaser
+    + Related to 7JB-KU
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRMY
+    + Myrstruktur
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRRL-CR/DD/EN/NT/VU
+    + Red list categories
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRSE-KA
+    + Strukturer, elementer og torvmarksformer (in 2018)
+    + See PRSE-PA
+
+### PRSE-PA
+    + Strukturer, elementer og torvmarksformer
+    + Used in the biodiversity assessment
+    + Exclude
+
+### PRSE-SH
+    + Strukturer, elementer og torvmarksformer (in 2018)
+    + See PRSE-PA
+
+### PRSL
+    + Slitasje
+    + Related to 7SE
+
+### PRTK
+    + Spor av tunge kjøretøyer
+    + Related to 7TK
+
+### PRTO
+    + Torvuttak
+    + Strictly speaking a pressure variable
+
+### PRVS
+    + Variasjon i vannsprutintensitet
+    + Used in the biodiversity assessment
+    + Exclude
+
+```{r}
+exclude <- c(exclude,
+             "LKMKI",     
+          "LKMSP",
+          "PRAK",
+          "PRAM",
+          "PRH",
+          "PRHA",
+          "PRKA",
+          "PRKU",
+          "PRMY",
+          "PRRL-DD",   "PRRL-VU",   "PRRL-NT",   "PRRL-EN",   "PRRL-CR",  
+          "PRSE-KA",
+          "PRSE-PA",
+          "PRSE-SH",
+          "PRVS"
+          )
+```
+
+## Subset
+```{r}
+# Exclude NAs
+dat2L2 <- dat2L[!is.na(dat2L$NiN_variable_value),]
+
+# Exclude selected NiN variables 
+dat2L2 <- dat2L2[dat2L2$NiN_variable_code %!in% exclude,]
+```
+Now we are now down to `r length(unique(dat2L2$NiN_variable_code))` possible condition variables.
+
+`dat2L2` includes only the three target ecosystems.
+Hule eiker are excluded.
+Mosaics are included.
+Several non-relevant NiN variables are excluded.
+The localities are duplicated in as many rows as there are NiN variables for that locality.
+
+## Combine Naturetypes, NiN variables and  NiN main types
+
+```{r}
+ntyp_vars <- as.data.frame(table(dat2L2$naturtype, dat2L2$NiN_variable_code))
+names(ntyp_vars) <- c("Nature_type", "NiN_variable_code", "NiN_variable_count")
+ntyp_vars <- ntyp_vars[ntyp_vars$NiN_variable_count >0,]
+head(ntyp_vars)
+```
+
+Add a column for the total number of localities for each nature type and get the percentage of localities where each NiN main type has been recorded.
+```{r}
+#options(scipen=999) # suppress exp notation
+count <- as.data.frame(table(dat2$naturtype))
+names(count) <- c("Nature_type", "totalLocations")
+ntyp_vars$totalLocations <- count$totalLocations[match(ntyp_vars$Nature_type, count$Nature_type)]
+ntyp_vars$percentageUse <- round((ntyp_vars$NiN_variable_count/ntyp_vars$totalLocations)*100, 1)
+```
+
+Cast to get NiN codes as columns
+```{r}
+# should switch to pivot_wider
+ntyp_vars_wide <- data.table::dcast(setDT(ntyp_vars),
+                                    Nature_type~NiN_variable_code,
+                                    value.var = "percentageUse")
+```
+
+This data set contains the nature types that were only mapped in 2018 or 2019, and which we removed from ntypDF2. Let's exclude these nature types now
+```{r}
+ntyp_vars_wide <- ntyp_vars_wide[ntyp_vars_wide$Nature_type %in% ntypDF2$Nature_type]
+```
+
+Combine datasets
+```{r}
+ntyp_fill <- cbind(ntypDF2, ntyp_vars_wide[,-1][match(ntypDF2$Nature_type, ntyp_vars_wide$Nature_type)])
+```
+
+### Add number of location and total area mapped
+```{r}
+dat2$km2 <- units::drop_units(dat2$area/10^6)
+
+num <- as.data.frame(dat2) %>%
+  group_by(naturtype)%>%
+  summarise(km2 = sum(km2),
+            numberOfLocalities = n())
+
+ntyp_fill <- cbind(ntyp_fill, num[,-1][match(ntyp_fill$Nature_type, num$naturtype),])
+```
+
+## Plots
+
+### The total mapped area
+```{r, fig.cap="The total mapped area for nature types associated with three selected main ecosystems"}
+
+mySize <- 8
+gg_area <- ntyp_fill %>%
+  arrange(km2) %>%
+  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%   # This trick update the factor levels
+  ggplot( aes(x = Nature_type,
+                   y = km2)) +
+        geom_segment( aes(xend=Nature_type, yend=0)) +
+        geom_point( size=4, aes(group = Ecosystem,
+                                colour= Ecosystem)) +
+        coord_flip() +
+        theme_bw(base_size = mySize) +
+        xlab("")+
+        ylab(expression(km^2))+
+      theme(legend.position = "top",
+        legend.key.size =  unit(.05, 'cm'),
+            legend.background = element_rect(fill = "white", color = "black")
+            )
+
+gg_locs <- ntyp_fill %>%
+  arrange(km2) %>%
+  mutate(Nature_type=factor(Nature_type, levels=Nature_type)) %>%    # also sorted after km2
+  ggplot( aes(x = Nature_type,
+                   y = numberOfLocalities)) +
+        geom_segment( aes(xend=Nature_type, yend=0)) +
+        geom_point( size=4, aes(group = Ecosystem,
+                                colour= Ecosystem)) +
+        coord_flip() +
+        theme_bw(base_size = mySize) +
+        xlab("")+
+        ylab("Number of localities")+
+  theme(legend.position = "none",
+        axis.text.y = element_blank())
+
+egg::ggarrange(gg_area, gg_locs, 
+                        ncol = 2)
+
+```
+
+### Area of localities per NiN main type
+First I will add the NiN main types that are not covered by any nature types.
+
+```{r}
+open <- c(
+  "T1",
+  "T2",
+  "T6",
+  "T8",
+  "T11",
+  "T12",
+  "T13",
+  "T15",
+  "T16",
+  "T17",
+  "T18",
+  "T20",
+  "T21",
+  "T23",
+  "T24",
+  "T25",
+  "T26",
+  "T27",
+  "T29"
+)
+
+semi <- c(
+  "T31",
+  "T32",
+  "T33",
+  "T34",
+  "T40",  # are these included
+  "T41"   # are these included
+)
+
+wetland <- c(
+  "V1",
+  "V3",
+  "V4",
+  "V6",
+  "V9",
+  "V10",
+  "L4"    # this will probably not be part of the assessments
+
+)
+
+```
+
+Adding the non-mapped open types
+```{r}
+open[open %!in% ntyp_fill$NiN_mainType]
+```
+
+```{r}
+ntyp_fill2 <- ntyp_fill %>%
+  add_row(NiN_mainType = "T6", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T11", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T13", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T16", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T23", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T24", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T25", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T26", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T27", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+
+```
+
+Adding the non-mapped semi-natural types
+```{r}
+semi[semi %!in% ntyp_fill$NiN_mainType]
+```
+```{r}
+ntyp_fill2 <- ntyp_fill2 %>%
+  add_row(NiN_mainType = "T40", Ecosystem = "Semi-naturlig mark", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+```
+
+Adding the non-mapped wetland types
+```{r}
+wetland[wetland %!in% ntyp_fill$NiN_mainType]
+```
+V6 is *Våtsnøleie*, and my guess is that it *is* mapped, but grouped with the alpine ecosystem. I'll therefore not add it in here, but mention it in the figure caption below.
+
+```{r, fig.cap="The areas mapped for each NiN main type, and the degree of spatial representativity (or coverage) of the mapping units (nature types). V6 Våtsnøleie og snøleiekilde is not included."}
+mySize <- 10
+ntyp_fill2 %>%
+  # combining two classes to get a better colour palette
+  mutate(NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "southern", "partial"),
+         NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "calcareous and dry", "calcareous")) %>%
+
+  group_by(NiN_mainType, NiN_mainTypeCoverage) %>%
+  mutate(km2 = sum(km2)) %>%
+  slice_head()%>%
+  select(Ecosystem,
+         NiN_mainType,
+         NiN_mainTypeCoverage,
+         km2)%>%
+  ungroup() %>%
+  mutate(NiN_mainType = fct_reorder(NiN_mainType, km2)) %>%
+  ggplot(aes(x = NiN_mainType,
+             y = km2,
+             fill = NiN_mainTypeCoverage))+
+    geom_bar(stat = "identity",
+             colour = "grey40",
+             position = "stack")+
+    theme_bw(base_size = mySize)+
+    coord_flip()+
+  labs(x = "NiN main type",
+       y = expression(km^2),
+       fill = "Spatial coverage")+
+  scale_fill_brewer(palette = "Set1")+
+  facet_wrap(vars(Ecosystem),
+             scales = "free",
+             labeller = label_wrap_gen(width=25))
+
+```
+
+### Number of localities per NiN main type
+
+```{r, fig.cap="The number of localities mapped for each NiN main type, and the degree of spatial representativity (or coverage) of the mapping units (nature types).
+V6 Våtsnøleie og snøleiekilde is not included.
+NiN main types are arranged rougly by decreasing order, except for types that are not mapped at all, which are arranged on top."}
+mySize <- 10
+ntyp_fill2 %>%
+  # combining two classes to get a better colour palette
+  mutate(NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "southern", "partial"),
+         NiN_mainTypeCoverage = 
+           replace(NiN_mainTypeCoverage, NiN_mainTypeCoverage == "calcareous and dry", "calcareous")) %>%
+
+  group_by(NiN_mainType, NiN_mainTypeCoverage) %>%
+  mutate(count = sum(numberOfLocalities)) %>%
+  slice_head()%>%
+  select(Ecosystem,
+         NiN_mainType,
+         NiN_mainTypeCoverage,
+         count)%>%
+  ungroup() %>%
+  mutate(NiN_mainType = fct_reorder(NiN_mainType, count)) %>%
+  ggplot(aes(x = NiN_mainType,
+             y = count,
+             fill = NiN_mainTypeCoverage))+
+    geom_bar(stat = "identity",
+             colour = "grey40",
+             position = "stack")+
+    theme_bw(base_size = mySize)+
+    coord_flip()+
+  labs(x = "NiN main type",
+       y = "Number of localities",
+       fill = "Spatial coverage")+
+  scale_fill_brewer(palette = "Set1")+
+  facet_wrap(vars(Ecosystem),
+             scales = "free",
+             labeller = label_wrap_gen(width=25))
+
+```
+
+From the figure above I take that *Naturlig åpne områder* is very poorly represented in general as there are many NiN main types that are completely missing from the dataset.
+The three most common types by area, however, have complete thematic coverage (if excluding the year 2018 and including all sub-types).
+I will need to investigate especially if there are some NiN variables for the nature types mapping T18 Åpen flomfastmark, T12 Strandeng and T21 Sanddynemark which we can use.
+
+For *Semi-naturlig mark* we have quite good thematic coverage for the three main types (by area).
+The other three types probably makes up a considerably smaller area, but note that T33 has some coverage, and is classes as *all (-2018)*.
+Of the three main ecosystems, semi-naturlig mark dominates clearly in ters of area and in terms of number of localities.
+
+For *Våtmark* we have good thematic coverage for V3 nedbørsmyr and V9 semi-naturlig myr, but for V1 åpen jordvassmyr only calcareous localities are mapped and assessed.
+A question is then whether, for a given NiN variable, that calcareous localities can be representative for all mires (poor and rich) or if this will introduce too much bias in one way or another.
+Another important bias for this main ecosystem is related to the lower size limit for mapping units (MMU), which is quite big for V3.
+We need to think about whether small bogs are more or less prone to different pressures compared to big bogs.
+
+### Commonness of NiN variables
+
+I want to see how common the different NiN variables are, and if we can use this to select core variables (kjernevariabler).
+
+First I will get the total area mappe per main ecosystem
+```{r}
+areaSum <- ntyp_fill2 %>%
+  group_by(Ecosystem) %>%
+  summarise(km2 = sum(km2)) %>%
+  tibble::add_column(.before=1,
+                     NiN_code = rep("Combined", 3))
+
+```
+
+```{r, fig.cap="Barplot showing the proportion area for which each NiN variable is recorded.
+The total mapped area for each main ecosystem (the facets) is shown as a dashed red line.
+The three most dominant NiN main types for each ecosystem is given a uniqe colour, and all the remaining are grouped as 'other'."}
+ntyp_fill2 %>%
+  pivot_longer(cols = unique(ntyp_vars$NiN_variable_code)) %>%
+  drop_na(value) %>%
+  rename(NiN_code = name,
+         useFrequency = value) %>%
+  filter(useFrequency > 0) %>%
+  group_by(NiN_code, Ecosystem, NiN_mainType) %>%  # NiN main type could be replaced with Nature_type
+  summarise(km2 = sum(km2)) %>%
+
+  # create a truncated fill variable 
+  group_by(Ecosystem) %>%
+  mutate(myFill = fct_lump(NiN_mainType, 3, w = km2)) %>%
+
+  # qick fix to sort variable across facets (dont work after adding 'Nature_type' in the first group_by)
+  #mutate(lab = paste(NiN_code, substr(Ecosystem, start = 1, stop = 1), sep = " ")) %>%
+  #ungroup()%>%
+  #mutate(lab = forcats::fct_reorder(lab, km2)) %>%
+
+  # plot
+  ggplot(aes(x = NiN_code,
+              y = km2,
+             fill = myFill
+                ))+
+  geom_bar(stat = "identity",
+           colour = "grey40",
+           #fill = "grey80",
+           position = "stack")+
+  geom_hline(data = areaSum,
+             aes(yintercept = km2),
+             colour = "red",
+             linetype = "dashed")+
+  facet_wrap(vars(Ecosystem),
+             scales = "free",
+             labeller = label_wrap_gen(width=25))+
+  labs(x = "",
+       y = expression(km^2),
+       fill = "")+
+  coord_flip()+
+  scale_fill_brewer(palette = "Set1")+
+  theme_bw(base_size = mySize)
+```
+
+This figure, in combination with the above, points to he most obvious NiN variable candidates.
+
+7SE and 7TK are good candidates for *Naturlig åpne områder*. 7JA-BT is probably more appropriately classed as a pressure indicator. Same with 7VR-RI with is recorded for all T18 localities.
+
+The same two variables (7SE and 7TK) are the best candidates also for *Semi-natural areas*.
+7JB-BT is a pressure variable (*beitetrykk*).
+7RA-SJ/BH are related to *gjengoing* which will be covered by another indicator informed from remote sensing (LiDAR), and we therefor don't need to describe this with a field-based indicator as well.
+
+For *Våtmark*, 7GR-Gi is the most common variables, but this is a pressure variable (*grøfting*).
+But, perhaps it could be used as a surrogate indicator.
+I think this is justifiable, especially since the relationship between the pressure (grøfting) and the condition (hydrology) is so well known.
+7SE+PRSL and 7TK+PRTK are other good candidates.
+PRTO is only relevant for V3, but covers a relatively large area.
+It is, however,, more like a pressure indicator.
+
+## Tables
+Here are some output tables with numbers the numbers underlying the above figures.
+
+```{r, fig.cap= "List of 45 nature types additionam data, including the proportion of localities for which there is data for each of the NiN variables."}
+DT::datatable(ntyp_fill2,
+              #extensions = "FixedColumns",
+  options = list(
+    scrollX = T,
+    scrollY=T,
+    pageLength = 10))
+```
+
+```{r, fig.cap= "List of unique combinatios of NiN variable codes and naturetypes with the summed area for which we have this data."}
+DT::datatable(
+  ntyp_fill2 %>%
+  pivot_longer(cols = unique(ntyp_vars$NiN_variable_code)) %>%
+  drop_na(value) %>%
+  rename(NiN_code = name,
+         useFrequency = value) %>%
+  filter(useFrequency > 0) %>%
+  group_by(NiN_code, Ecosystem, NiN_mainType) %>%  # NiN main type could be replaced with Nature_type
+  summarise(km2 = sum(km2)),
+  options = list(
+    scrollX = T,
+    scrollY=T,
+    pageLength = 10))
+```
+
+## Export data
+Exporting the table with the summary statistics and other information that I compled for each nature type in the three main ecosystems.
+This can be merged with the main dataset in order to subset according to a specific research question.
+
+```{r, eval = F}
+saveRDS(ntyp_fill2, "data/naturetypes/natureType_summary.rds")
+```
+
+Note also that the main dataset still contains some localities where there is a mis-match between the nature type definition and the NiN mapping uits recorded in the field.
+Each researcher should decide to include or exclude these.
+
+The main data set also includes mosaic localities, that are less suitable to use as ground trouth data for remote sensing methods.
+
+# Appendix 1
+Here is a deeper look into the causes for why some localities are recorded with seemingly erroneous NiN mapping units (and hence NiN main types)
+
+Lets have a look at how the NiN main types (recorded in the field) cover the different *hovedøkosystem* when we exclude *Hule eiker* and all localities that have more than one NiN main type.
 ```{r}
@@
                       dat_melt$mosaikk == "Nei",]
-```
-
-```{r, fig.cap="NiN main type across five main ecosystems (Hule eiker, mosaic localities, and localities with >1 NiN main type excluded)"}
-ggplot(dat_red)+
-  geom_bar(aes(x = ninkartleggingsenheter3),
-           fill="grey",
-           colour="black")+
-  coord_flip()+
-  theme_bw(base_size = 12)+
-  facet_wrap("hovedøkosystem",
-             scales = "free")+
-  labs(y = "Number of localities",
-       x = "NiN main types")
-
 ```
@@
 
-Still some weird cases which I will look at in turn below.
+Still some weird cases which I will look at in turn below, trying to understand how they came to be recorded in this way.
 
-**Naturlig åpne områder under skoggrensa:**
+#### Naturlig åpne områder under skoggrensa
+Listing the NiN main types recorded for nature types belonging to *Naturlig åpne områder under skoggrensa*, but which clearly don't belong there:
 
@@
     + One case of Åpen flomfastmark wrongly associated with V10 whan it should be T18.
-    + **Exclude** this case
-* T8 (fulglefjelleng og fugltopp)
-    + Ok
-* T6 (Standberg)
-    + Ok
 * T33 (Semi naturlig strandeng)
     + These are cases where either the `naturtype` should have been `semi-naturlig strandeng` (and hence the hovedøkosystem would be Semi-nat), or the `naturtype` is correctly recorded as `Strandeng` but then the NiN type should be `T12`.
-    + **Exclude**
 * T32 (Semi-naturlig eng)
     + Obvious error
-    + **Exclude**
 * T30 (Flomskogsmark)
     + Obvious error
-    + **Exclude**
+
+Listing the remaining NiN main types:
+* T8 (fulglefjelleng og fugltopp)
+* T6 (Standberg)
 * T29 (Grus og steindominert strand og strandlinje)
-    + Ok
 * T21 (Sanddynemark)
-    + Ok
 * T20 (Isinnfrysingsmark)
-    + Ok. Rare nature type
 * T2 (Åpen grunlendt mark)
-    + Ok
 * T18 (Åpne flomfastmark)
-    + Ok
 * T17 (Aktiv skredmark)
-    + Ok
 * T15 (Fosse-eng)
-    + Ok
 * T13 (Rasmark)
-    + Ok
 * T12 (Strandeng)
-    + Ok
 * T1 (Nakent berg)
-    + Ok
 
 ```{r}
-#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "V10",])
-#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "T33",])
-#View(dat_red[dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & dat_red$ninkartleggingsenheter3 == "T32",])
-
 `%!in%` <- Negate(`%in%`)
@@
 
-**Semi-naturlig mark**
+This resulted in deleting `r nrow(dat_red) - nrow(dat_red2)`m localities.
+
+#### Semi-naturlig mark
+Listing the NiN main types recorded for nature types belonging to *Semi-naturlig mark*, but which clearly don't belong there:
 
@@
    + Kystlynghei recorded as mire
-   + **Exclude**
-
-* V10 (semi-nat våteng)
-   + Ok
-
-* T41 (Oppdyrket mark med preg av semi-naturlig eng)
-   + Probably not always use correctly, but at least its the correct ``hovedkøosystem`
-   + Ok
-
-* T40 (Sterkt endret fastmark med preg av semi-naturlig eng)
-   + Ok
-
 * T4 (Fastmarksskogsmark)
    + These should all be T32 (naturbeitemark, lauveng, hagemark, +++)
-   + **Exclude**
-
 * T35 (Sterkt endret fastmark med løsmassedekke)
    + Wrong NiN type given to eng-aktig sterkt endret fastmark (it should be T40)
-   + **Exclude**
-
-* T31-34
-   + Ok
-
 * T3 (Fjellhei, leside og tundra)
    + Boreal hei (should've been T31)
-   + **Exclude**
-
 * T12 (Strandeng)
    + Should probably have been T33
-   + **Exclude**
-
 * T1 (Nakent berg)
    + Obvious mistake
-   + **Exclude**
+
+Listing the remaining NiN-main types:
+* V10 (semi-nat våteng)
+* T41 (Oppdyrket mark med preg av semi-naturlig eng)
+* T40 (Sterkt endret fastmark med preg av semi-naturlig eng)
+* T31-34
 
 ```{r}
-#unique(dat$hovedøkosystem)
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "V9",])
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T41",])
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T4",])
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T35",])
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T3",])
-#View(dat_red[dat_red$hovedøkosystem=="Semi-naturlig mark" & dat_red$ninkartleggingsenheter3 == "T1",])
-
 dat_red3 <- dat_red2[
@@
 
-**Våtmark**
+This resulted in the deletion of `r nrow(dat_red2) - nrow(dat_red3)` localities.
 
-Våtmakr doesn't distinguish between semi-natural and natural, so all V-types are correct, except perhaps the strongly modified V11-V13.
+#### Våtmark
+Listing the NiN main types recorded for nature types belonging to *Våtmark*, but which clearly don't belong there.
+
+Våtmark doesn't distinguish between semi-natural and natural, so all V-types are correct, except the strongly modified V11-V13.
 
@@
   + Obvious mistake
-  + **Exclude**
-
 * V12 (Grøftet åpne torvmark)
   + Obvious mistake
-  + **Exclude**
-
 * V13 (Ny våtmark)
   + Obvious mistake
-  + **Exclude**
-
 * T4 (skog)
   +   some that should've been T2 and some other strange things
-  + **Exclude**
-
 * T30-T32
   + Someone switching the semi-natural and natural equivalents
-  + **Exclude**
-
-* L4 (Helofyttsump)
-    + Ok
 
 ```{r}
-#unique(dat$hovedøkosystem)
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "T4",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "T32",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "L4",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V9",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V11",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V12",])
-#w(dat_red[dat_red$hovedøkosystem=="Våtmark" & dat_red$ninkartleggingsenheter3 == "V13",])
 dat_red4 <- dat_red3[
@@
 ```
-
-Let's make the plot again to see if it worked
-```{r, fig.cap="Number of localities for each combination of main ecosystem and NiN main type (after data set cleaning)"}
-dat_red_tally2 <- stats::aggregate(data = dat_red4,
-                                 area~hovedøkosystem+ninkartleggingsenheter3, FUN = length)
-names(dat_red_tally2)[3] <- "count"
-
-ggplot(dat_red_tally2[dat_red_tally2$hovedøkosystem %in% target,],
-       aes(x = ninkartleggingsenheter3,
-               y = count))+
-  geom_bar(
-           fill="grey",
-           colour="black",
-           stat="identity")+
-  coord_flip()+
-  theme_bw(base_size = 12)+
-  labs(y = "Number of localities",
-       x = "NiN main types")+
-  scale_y_continuous(position = "left",
-                   expand = expansion(mult=c(.0,.3)))+
-  geom_text(aes(label=count), hjust=-0.25)+
-    facet_wrap("hovedøkosystem",
-             scales = "free",
-             ncol = 3)
-```
-
-## Export
-`dat_red4` now has one row per locality, and only localities with 1 NiN main type.
-`Hule eiker` and `Mosaic localities` are not included, but we will add these back in now.
-Depending on what the data set will be used for, these last two can be excluded again later.
-```{r}
-exp <- rbind(dat_red4,
-             dat[dat$mosaikk=="Ja",],
-             dat[dat$naturtype== "Hule eiker",])
-```
-
-To remove `hule eiker`, use this
-```{r, eval=F}
-exp[exp$naturtype != "Hule eiker",]
-```
-
-To remove mosaic localities, first consider if you want to remove all mosaic localities, or just those that include different NiN main types, because a mosaic locality may also consist of different sub units (*grunntype*) within the same NiN main type, or even the same *grunntype* throughout, but varying in some other parameter (e.g.
-a mosaic between old pine forest and old spruce forest, which are both T4 and could both be, say T4-C-2).
-
-To exclude mosaics in general:
-```{r, eval=F}
-exp[exp$mosaikk == "Nei",]
-```
-
-To exclude only mosaics that include multiple NiN main types:
-```{r, eval=F}
-exp[exp$n_ninkartleggingsenheter == 1]
-```
-
-I will save this as an rds file to my personal folder (the shape standard truncates field names, and gdb is not soo easy to use in R), but then manually move it to 
-`/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.shp`.
-This is simply because I cannot access that drive from RStudio Server at the moment.
-
-```{r, eval=F}
-exp$ninkartleggingsenheter3 <- unlist(exp$ninkartleggingsenheter3)
-#names(exp)[1] <- "ID" # Field name too long
-saveRDS(exp, "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Mdir_Naturtyper_cleanedALK.rds")
-
-```
-
-This is the code for importing the cleaned data back into R ()
-```{r, eval=F}
-#myData <- readRDS("/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Mdir_Naturtyper_cleanedALK.rds")
-myData <- readRDS("/data/P-Prosjekter/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
-```
-
-## Overview of nature types
-
-```{r, fig.cap="An overview of the nature types in the data set", fig.height=15}
-temp <- tapply(exp$area, exp$naturtype, sum)
-temp <- data.frame(nat = row.names(temp),
-                   area = temp)
-
-ggplot(exp[exp$hovedøkosystem %in% target,], 
-       aes(x = naturtype))+
-  geom_bar(fill = "grey20",
-           size=0
-           )+
-  theme_bw(base_size = 12)+
-  coord_flip()+
-  facet_grid(rows = vars(hovedøkosystem),
-             scales = "free",
-             space = "free")+
-  labs(y = "Number of localities",
-       x = "Nature type")
-```
+This resulted in the deletion of `r nrow(dat_red3) - nrow(dat_red4)` localities.

### 2022-11-15 DRAFT_rosslyng.Rmd
add røsslyng indicator

---
title: "DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei"
output:
  html_document
---

# DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei

<br />

_Author and date:_
Anders L. Kolstad

```{r}
Sys.Date()
```

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
library(sf)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "Semi-naturlig mark" # e.g. "Skog og fjell"
Egenskap  <- "Primærproduksjon" # e.g. "Primærproduksjon"
ECT       <- "Structual state characteristic" # e.g. "Structural state characteristic"
Contact   <- "Anders L. Kolstad" # e.g. "Anders Kolstad"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
Here I quickly go through the ANO data set for the years 2019-2021 to investigate the frequeny of poins that fall within coastal heathands (kystlynghei). This will decide wether we can use this data to inform an indikcator about the proportion of dead Calluna vulgaris (røsslyng).

## Analyses

#### ANO data
Data is downloaded from [here](https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054).

Imnport and list terms: 
```{r}
ano <- sf::st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                   layer = "ANO_SurveyPoint")
#st_crs(ano)
names(ano)

```

Table of the number of sampling points per year:
```{r}
table(ano$aar)
```

```{r, fig.cap="The distribution of ANO points that fall within different main ecosystems."}
par(mar=c(15,5,0,0))
barplot(table(ano$hovedoekosystem_punkt), las=2)
```

Sub-setting the data to only consider seni natural areas.
```{r}
snat <- ano[ano$hovedoekosystem_punkt == "semi_naturlig_mark",]
snat2 <- snat[!is.na(snat$GlobalID),]
```
230 points are semi-natural.

```{r}
table(snat$hovedtype_1m2)
```
Only 49 1m2 samples are kystlynghei.

```{r}
table(ano$hovedtype_250m2)
```
Only 26 of the 250m2 circles are kystlynghei.

## Conclusion
Because of the random allocation of sampling points, not enough points fall in coastal heathlands to allow us to create a trustworth indicator for this nature type. With a few more years of data we might end up with about 100 points, which is borderline what we need in order to calculate condition values at a regional level with some level of precition.

### 2022-11-28 DRAFT_slitasjeOgKjørespor.Rmd
update paths

@@
 library(sf)
+library(tmap)
+library(ggplot2)
 knitr::opts_chunk$set(echo = TRUE)
@@
 This indicator uses the dataset `Naturtyper etter Miljødirektoratets Instruks`, which can be found [here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
-The dataset has already been [cleaned](#naturtype1) and [initial analyses on thematic representativity](#naturtype2) has been done.
+See [here](#naturtype) for a detailed description of tha dataset.
 
 ```{r, eval=T}
-dat <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Mdir_Naturtyper_cleanedALK.rds")
+
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
+
+dat <- sf::st_read(dsn = path)
 ```
 
+We also have a separate summary file where the nature types are manually mapped to the correct NiN-main types.
+We can use this to find the nature types associated with the NiN main types that we are interested in.
+```{r}
+nt <- readRDS("data/naturetypes/natureType_summary.rds")
+```
+
+We are only interested in a few NiN main types and their mapping units (see [here](https://ninanor.github.io/ecosystemCondition/naturtype.html) for justification)
+```{r}
+main <- c("T12",
+          "T18",
+          "T21",
+          "T31",
+          "T33",
+          "T34",
+          "V1",
+          "V3")
+```
+
+```{r}
+nt2 <- nt[nt$NiN_mainType %in% main,]
+```
+
+Sub-setting the main data file:
+```{r}
+dat2 <- dat[dat$naturtype %in% nt2$Nature_type,]
+```
+
+Fix duplicated name for an ecosystem
+```{r}
+dat2$hovedøkosystem[dat2$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+```
+
+```{r, fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation."}
+ggplot(data = dat2, aes(x = naturtype, fill = hovedøkosystem))+
+  geom_bar()+
+  coord_flip()+
+  theme_bw(base_size = 12)+
+  guides(fill = "none")+
+  xlab("")+
+  ylab("Number of localities")
+```
+
+# GRUK
+
+This variable is also recorded in GRUK
+
 #### Vegetation zones (bioregions)
+I want to look at if the indicator values can be averages over smaller and more meaningfu areas (ecological homogeneous areas) than the regions.
+```{r}
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp",
+               "/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp")
+```
+
+```{r}
+soner <- sf::read_sf(path)
+```
+
+```{r}
+tm_shape(soner)+
+  tm_polygons(col = "NAVN")
+```
+
+It could make sense to combine some zones, but I will later split the sones by region, and then merge sones that are very uncommon in a spesific region.
 
 #### Regions
-<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
 Importing a shape file with the regional delineation. 
 ```{r}
-#reg <- sf::st_read("data/regions.shp")
+reg <- sf::st_read("data/regions.shp")
 #st_crs(reg)

### 2022-11-29 DRAFT_slitasjeOgKjørespor.Rmd
GRUK

@@
 
-This variable is also recorded in GRUK
+This variable is also recorded in GRUK.
+The nature type dataset I'm working on here includes this data already (presently only 2021 included).
+GRUK also records a related variable: % cover in 5m radii circles.
+This data is not published.In any case it is better to use the harmonized dataset.
 
@@
 tm_shape(soner)+
-  tm_polygons(col = "NAVN")
+  tm_polygons(col = "NAVN",
+              title = "Vegetation zone")
 ```
 
-It could make sense to combine some zones, but I will later split the sones by region, and then merge sones that are very uncommon in a spesific region.
+It could make sense to combine some zones, but I will later split the zones by region, and then merge sones that are very uncommon in a spesific region.
 

### 2022-11-30 DRAFT_slitasjeOgKjørespor.Rmd
prep naturetype dataset

@@
 library(ggplot2)
+library(tidyverse)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-dat <- sf::st_read(dsn = path)
+naturetypes <- sf::st_read(dsn = path)
 ```
@@
 ```{r}
-nt <- readRDS("data/naturetypes/natureType_summary.rds")
+naturetypes_summary <- readRDS("data/naturetypes/natureType_summary.rds")
 ```
@@
 ```{r}
-main <- c("T12",
+main_NiN_types <- c("T12",
           "T18",
@@
 ```{r}
-nt2 <- nt[nt$NiN_mainType %in% main,]
+naturetypes_summary <- naturetypes_summary[naturetypes_summary$NiN_mainType %in% main_NiN_types,]
 ```
@@
 ```{r}
-dat2 <- dat[dat$naturtype %in% nt2$Nature_type,]
+naturetypes <- naturetypes[naturetypes$naturtype %in% naturetypes_summary$Nature_type,]
 ```
@@
 ```{r}
-dat2$hovedøkosystem[dat2$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+naturetypes$hovedøkosystem[naturetypes$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
 ```
@@
 ```{r, fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation."}
-ggplot(data = dat2, aes(x = naturtype, fill = hovedøkosystem))+
+ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
   geom_bar()+
@@
 
-# GRUK
+Calculate area
+```{r}
+naturetypes$area <- sf::st_area(naturetypes)
+```
+
+The nin variables are combined into a single field which we need to spit apart.
+```{r}
+naturetypes <- tidyr::separate_rows(naturetypes,
+                                    ninbeskrivelsesvariabler,
+                                    sep=",") %>%
+  separate(col = ninbeskrivelsesvariabler,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F
+           )
+```
+
+Convert to numeric. Ignore NA's
+```{r}
+naturetypes$NiN_variable_value <- as.numeric(naturetypes$NiN_variable_value)
+```
+
+Keep only the relevant variables
+```{r}
+myVars <- c(
+  #"7FA",     #treated separately in a different indicator
+  "7SE", "7TK", "PRTK", "PRSL")
+naturetypes <- filter(naturetypes,
+                      NiN_variable_code %in% myVars)
+```
+
+Columnames starting with a number is problematic, so adding a prefix
+```{r}
+naturetypes$NiN_variable_code <- paste0("var_", naturetypes$NiN_variable_code)
+```
+
+Now I need to create a single row per locality with a new variable which is a product of the four variables "7SE", "7TK", "PRTK" and "PRSL".
+I will create a new data table where I calculate the new variable which I then paste back into the sf object.
+```{r}
+naturetypes_wide <- pivot_wider(naturetypes,
+                                names_from = NiN_variable_code,
+                                values_from = NiN_variable_value,
+                                id_cols = identifikasjon_lokalid)
+```
+
+The variables use slightly different scales.
+PRSL and PRTK use this 8 step scale:
+![](images/mdirPRscale.PNG)
+
+7TK and 7SE use a 4 step scale
+![](images/ninAR4b.PNG)
+
+I want to base this indicatr on whichever of the variables have the highest value (worst condition). This table shows that a loclity may have been assessed with both a PR variable and a NiN variable
+```{r}
+table(naturetypes_wide$var_7SE, naturetypes_wide$var_PRTK)
+```
+Therefore I need to reconcile these scales. The only way I can think of is to decrease the resolution of the PR variables to the ar4b scale. This is of course a shame to loose this detail.
+
+A locality will never have values for both 7SE and PRSL
+```{r}
+table(naturetypes_wide$var_7SE, naturetypes_wide$var_PRSL)
+```
+nor 7TK and PRTK
+```{r}
+table(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK)
+```
+
+```{r}
+naturetypes_wide$var_7TK <-  ifelse(is.na(naturetypes_wide$var_7TK),
+                              ifelse(naturetypes_wide$var_PRTK == 0, 0,
+                                ifelse(naturetypes_wide$var_PRTK < 3, 1,
+                                  ifelse(naturetypes_wide$var_PRTK < 6, 2, 3))),
+                            naturetypes_wide$var_7TK)
+```
+
+Checking that I successfully combined classes 1+2, 3+4+5 and 6+7.
+```{r}
+table(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK,
+      deparse.level = 2)
+```
+That looks fine. Doing the same for the other variable:
+```{r}
+naturetypes_wide$var_7SE <-  ifelse(is.na(naturetypes_wide$var_7SE),
+                              ifelse(naturetypes_wide$var_PRSL == 0, 0,
+                                ifelse(naturetypes_wide$var_PRSL < 3, 1,
+                                  ifelse(naturetypes_wide$var_PRSL < 6, 2, 3))),
+                            naturetypes_wide$var_7SE)
+```
+
+Remove the old PR variables
+```{r}
+naturetypes_wide <- select(naturetypes_wide,
+                           -var_PRTK,
+                           -var_PRSL)
+```
+
+There are some cells with X's. These are NA's.
+```{r}
+naturetypes_wide$var_7SE[naturetypes_wide$var_7SE == "X"] <- NA
+naturetypes_wide$var_7TK[naturetypes_wide$var_7TK == "X"] <- NA
+
+naturetypes_wide$var_7SE <- as.numeric(naturetypes_wide$var_7SE)
+naturetypes_wide$var_7TK <- as.numeric(naturetypes_wide$var_7TK)
+```
+
+7SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7SE),])` NA's, and 7TK has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7TK),])`.
+
+```{r, fig.cap="7SE scores in the naturetype dataset"}
+temp <- naturetypes_wide %>%
+  group_by(var_7SE) %>%
+  summarise(sum = n())
+
+ggplot(temp, aes(x = factor(var_7SE),
+                 y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "7SE score",
+       y = "Number of localities")
+```
+```{r, fig.cap="7TK scores in the nature type dataset"}
+temp <- naturetypes_wide %>%
+  group_by(var_7TK) %>%
+  summarise(sum = n())
+
+ggplot(temp, aes(x = factor(var_7TK),
+                 y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "7TK score",
+       y = "Number of localities")
+```
+
+```{r}
+naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
+```
+
+When both variables are NA, we get -Inf. Removing these
+```{r}
+naturetypes_wide <- naturetypes_wide[naturetypes_wide$slitasje>=0,]
+```
+
+```{r, fig.cap="Slitasje scores."}
+temp <- naturetypes_wide %>%
+  group_by(slitasje) %>%
+  summarise(sum = n())
+
+ggplot(temp, aes(x = factor(slitasje),
+                 y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "slitasje score",
+       y = "Number of localities")
+```
+
+I would also like to know how often 7TK was defining of the slitasje-indicator, and how often it was 7SE. I can do this by taking the difference
+
+```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score."}
+diff <- naturetypes_wide$var_7SE - 
+        naturetypes_wide$var_7TK
+
+diff <- ifelse(diff == 0, "7TK and 7SE",
+               ifelse(diff <0, "7TK", "7SE"))
+
+diff_tbl <- as.data.frame(table(diff))
+
+ggplot(diff_tbl, aes(x = diff, y = Freq))+
+  geom_bar(stat = "identity",
+           colour = "black",
+           fill = "grey")+
+  theme_bw(base_size = 12)+
+  labs(x = "Defining variable",
+       y = "Number of localities")
+```
+
+Looks like 7SE is more likely to be in a detrimental state.
+
+Now I will copy these slitasje-values into the sf object.
+```{r}
+naturetypes$slitasje <- naturetypes_wide$slitasje[match(naturetypes$identifikasjon_lokalid, naturetypes_wide$identifikasjon_lokalid)]
+#nrow(naturetypes[is.na(naturetypes$slitasje),])  # 13 cases
+naturetypes <- naturetypes[!is.na(naturetypes$slitasje),]
+```
+
+#### GRUK
 
@@
 This data is not published.In any case it is better to use the harmonized dataset.
+
+#### ANO
+
+Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically placed locations each with 18 sample poins.
+In each sample point a circle of 250 m2 is visualised, and the main ecosystem is recorded.
+Depnding oon the main ecosystem, certain NiN variables are also recorded.
+7SE is recorded for våtmark, but not semi-natural areas or naturally open areas.
+7TK is recorded in våtmark and naturally open areas only.
+
+Variable    | Våtmark | Naturlig åpne | Semi-nat.
+7SE         | X       |               |
+7TK         | X       | X             |
+
+It would be very nice to have 7SE recorded for naturally open areas. This variable is very relevant here.
+
+I think I will only use våtmark here since my approach will be to the 'worst value' of the variable 7SE and 7TK (se below).
+I think not having 7SE for naturally open areas will underestimate the degree of degredation in these areas.
+
+```{r}
+path <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
+
+ano <- sf::st_read(path,
+                   layer = "ANO_SurveyPoint") %>%
+  dplyr::filter(hovedoekosystem_punkt == "vaatmark")
+```
+
+Each point here is 250 square meters, but the data also contains information about how big a proportion of thi area is made up of the dominant main ecosystem.
+However, there are `r (NAS <- length(ano$andel_hovedoekosystem_punkt[is.na(ano$andel_hovedoekosystem_punkt)]))` NA's here, which is `r round(NAS/nrow(ano), 2)*100`% of the data.
+```{r}
+ano <- ano[!is.na(ano$andel_hovedoekosystem_punkt),]
+```
+
+```{r}
+summary(ano$andel_hovedoekosystem_punkt)
+```
+
+Where we have data, this is the distribution of values:
+```{r, warning=F}
+ggplot(ano, aes(x = andel_hovedoekosystem_punkt))+
+         geom_histogram(fill = "grey",
+                        colour="black",
+                        binwidth = 1)+
+  theme_bw(base_size = 12)+
+  xlab("Percentage cover of the Våtmark main ecosystem\n in the 250m2 circle")+
+  scale_x_continuous(limits = c(0,101),
+                     breaks = seq(0,100,10))
+```
+
+We can see that people tend to record the variable in steps of 5, and that most sample points are 100% belonging to the same main ecosystem.
+
+We want to use area weighting in this indicator, so we can use this percentage cover data to calculate the area.
+```{r}
+ano$area_of_våtmark <- (ano$andel_hovedoekosystem_punkt/100)*250
+```
+
+Let's now look at the distribution of the variables. First I need to slip the cell values to make them numeric.
+
+```{r}
+ano$'7SE' <- as.numeric(sub(pattern = "7SE_",
+                 replacement = "",
+                 x = ano$bv_7se))
+```
+
+The NA's is the case of blank cells. The field app should not allow users leaving this field blank. I will need to remove these rows.
+```{r}
+ano <- ano[!is.na(ano$`7SE`),]
+```
+
+```{r}
+ano$'7TK' <- as.numeric(sub(pattern = "7TK_",
+                 replacement = "",
+                 x = ano$bv_7tk))
+```
+
+```{r}
+par(mfrow=c(1,2))
+barplot(table(ano$`7TK`), xlab="7TK scores")
+barplot(table(ano$`7SE`), xlab="7SE scores")
+```
+
+### Combine datasets
+We need to combine the nature type dataset with th ano dataset
+
+```{r}
+naturetypes
+```
 

### 2022-11-30 DRAFT_slitasjeOgKjørespor.Rmd
combine ano and naturetypes

@@
 library(ggplot2)
+library(units)
 library(tidyverse)
@@
 Where we have data, this is the distribution of values:
-```{r, warning=F}
+```{r, warning=F, fig.cap="Percentage cover of the Våtmark main ecosystem in the 250m2 circle"}
 ggplot(ano, aes(x = andel_hovedoekosystem_punkt))+
@@
 
-We can see that people tend to record the variable in steps of 5, and that most sample points are 100% belonging to the same main ecosystem.
+We can see that people tend to record the variable in steps of 5%, and that most sample points are 100% belonging to the same main ecosystem.
 
 We want to use area weighting in this indicator, so we can use this percentage cover data to calculate the area.
+Note that both datasets use m2 as area units.
 ```{r}
-ano$area_of_våtmark <- (ano$andel_hovedoekosystem_punkt/100)*250
+ano$area <- (ano$andel_hovedoekosystem_punkt/100)*250
 ```
@@
 Let's now look at the distribution of the variables. First I need to slip the cell values to make them numeric.
-
 ```{r}
-ano$'7SE' <- as.numeric(sub(pattern = "7SE_",
+ano$var_7SE <- as.numeric(sub(pattern = "7SE_",
                  replacement = "",
@@
 ```{r}
-ano <- ano[!is.na(ano$`7SE`),]
+ano <- ano[!is.na(ano$var_7SE),]
 ```
@@
 ```{r}
-ano$'7TK' <- as.numeric(sub(pattern = "7TK_",
+ano$var_7TK <- as.numeric(sub(pattern = "7TK_",
                  replacement = "",
@@
 
+```{r, fig.cap="Distribution of 7TK and 7SE scores in the ANO data"}
+par(mfrow=c(1,2))
+barplot(table(ano$var_7TK), xlab="7TK scores")
+barplot(table(ano$var_7SE), xlab="7SE scores")
+```
+
+Combining these two variables into the indicator, same as for the nature type data.
 ```{r}
-par(mfrow=c(1,2))
-barplot(table(ano$`7TK`), xlab="7TK scores")
-barplot(table(ano$`7SE`), xlab="7SE scores")
+temp <- as.data.frame(ano)
+class(temp)
+temp$slitasje <- apply(temp[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
+ano$slitasje <- temp$slitasje
 ```
@@
 We need to combine the nature type dataset with th ano dataset
+I will tag the ano rows in the nature type column, and add data abot the main ecosystem
+```{r}
+ano$naturtype <- "ANO"
+ano$hovedøkosystem <- "Våtmark"
+ano$kartleggingsår <- ano$aar
+```
 
+I will loose the sf class and the crs when joining the datasets, so I will save the crs
 ```{r}
-naturetypes
+sf::st_crs(ano) == sf::st_crs(naturetypes) # TRUE
+myCRS <- st_crs(ano)
+```
+
+Fix class
+```{r}
+naturetypes$kartleggingsår <- as.numeric(naturetypes$kartleggingsår)
+naturetypes$area <- units::drop_units(naturetypes$area)
+```
+
+I can reduce the number of column to keep things a bit more tidy
+```{r}
+temp <- as.data.frame(ano)
+temp2 <- as.data.frame(naturetypes)
+slitasje_data <- full_join(select(temp,
+                                  GlobalID,
+                                  kartleggingsår,
+                                  hovedøkosystem,
+                                  area,
+                                  slitasje,
+                                  SHAPE), 
+                           select(temp2,
+                                  identifikasjon_lokalid,
+                                  hovedøkosystem,
+                                  kartleggingsår,
+                                  area,
+                                  slitasje,
+                                  SHAPE))
+# convert back to sf object
+slitasje_data <- st_as_sf(slitasje_data)
+st_crs(slitasje_data) <- myCRS
+```
+
+```{r, fig.cap="Slitasje scores."}
+temp <- slitasje_data %>%
+  group_by(slitasje) %>%
+  summarise(sum = n())
+
+ggplot(temp, aes(x = factor(slitasje),
+                 y = sum))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "slitasje score",
+       y = "Number of localities")
 ```
@@
 #### Vegetation zones (bioregions)
-I want to look at if the indicator values can be averages over smaller and more meaningfu areas (ecological homogeneous areas) than the regions.
+I want to look at if the indicator values can be averages over smaller and more meaningful areas (ecological homogeneous areas) than the 5 default regions.
 ```{r}
-
 path <- ifelse(dir == "C:", 

### 2022-11-30 DRAFT_slitasjeOgKjørespor.Rmd
text about agg strategy

@@
 ```{r}
-reg <- sf::st_read("data/regions.shp")
+#reg <- sf::st_read("data/regions.shp")
 #st_crs(reg)
@@
 <!-- Text and analyses here -->
+I can scale the indicator for each polygon, or I can chose to aggregate them first.
+If the scaled value is representative and precise at the polygon level, then I could scale at that level.
+However, the combined surveyed area is still so little of the total area of Norway, so that only producing indicator values for these leaves the indicator without much value for regional assessments.
+Later when we produce assessments and indicator values at the regional level, we cannot simply do an area weighting of the polygons in each region.
+This is because we don't want to assume that the polygons are representative outside of the mapped area.
+
+An option is to export one map with just the raw un-scaled slitasje variable at the polygon level, and another, scaled indicator map aggregated to either
+
+* each *dekningskart* (map of surveyed area) 
+* each vegetation zone (e.g. boreonemal zone in Østlandet; excluding zone with too little data points)
+* each vegetation region (aka bioregion, e.g. boreonemal zone and continental section in Østlandet)
+* each pixel of a give size, say 1 x 1 km
+
+The last option is problematic because we often will have too few points to be able to estimate  credible mean. 
+I can try the second option first, and see how many data points we get for each zone.
 

### 2022-12-14 DRAFT_slitasjeOgKjørespor.Rmd
cache

@@
 This indicator uses the dataset `Naturtyper etter Miljødirektoratets Instruks`, which can be found [here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
-See [here](#naturtype) for a detailed description of tha dataset.
+See [here](#naturtype) for a detailed description of the dataset.
 
+Importing the full dataset can be done like this 
 ```{r, eval=T}
@@
 naturetypes$hovedøkosystem[naturetypes$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+```
+
+Export subsetted data set
+```{r}
+saveRDS(naturetypes, "P:/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
 ```

### 2022-12-14 DRAFT_slitasjeOgKjørespor.Rmd
ano

--- 2022-11-30 DRAFT_slitasjeOgKjørespor.Rmd
@@
   html_document
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 <!-- Replace 'indicator name' with yor own indicator name  -->
+
 # Slitasje og kjørespor
@@
 
-_Author and date:_
-Anders L. Kolstad
+*Author and date:* Anders L. Kolstad
 
@@
 <!-- Load all you dependencies here -->
+
 ```{r setup, include=FALSE}
@@
 <!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
+
 ```{r, echo=F}
@@
 <!-- Don't remove these three html lines -->
-<br />
-<br />
+
+<br /> <br />
+
 <hr />
@@
 ## Introduction
+
 <!-- Text here -->
@@
 ## About the underlying data
+
 <!-- Text here -->
@@
 ### Representativity in time and space
+
 <!-- Text here -->
@@
 ### Original units
+
 <!-- Text here -->
@@
 ### Temporal coverage
+
 <!-- Text here -->
@@
 ### Aditional comments about the dataset
+
 <!-- Text here -->
@@
 ## Ecosystem characteristic
+
 ### Norwegain standard
+
 <!-- Text here -->
@@
 ### <!-- Text here -->
+
 <!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->
@@
 ## Collinearities with other indicators
+
 <!-- Text here -->
@@
 ## Reference state and values
+
 ### Reference state
+
 <!-- Text here -->
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+
 <!-- Text here -->
@@
 ## Uncertainties
+
 <!-- Text here -->
@@
 ## References
+
 <!-- Text here -->
@@
 ## Analyses
+
 ### Data sets
@@
 #### Nature type mapping
-This indicator uses the dataset `Naturtyper etter Miljødirektoratets Instruks`, which can be found [here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
-See [here](#naturtype) for a detailed description of tha dataset.
 
-```{r, eval=T}
+This indicator uses the dataset
+`Naturtyper etter Miljødirektoratets Instruks`, which can be found
+[here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
+See [here](#naturtype) for a detailed description of the dataset.
 
+Importing the full dataset can be done like this, but because it can
+take a few seconds due to its size, I will cache the data on the P
+server.
+
+```{r}
 dir <- substr(getwd(), 1,2)
+```
 
+```{r, eval=F}
 path <- ifelse(dir == "C:", 
@@
 
-We also have a separate summary file where the nature types are manually mapped to the correct NiN-main types.
-We can use this to find the nature types associated with the NiN main types that we are interested in.
+We also have a separate summary file where the nature types are manually
+mapped to the correct NiN-main types. We can use this to find the nature
+types associated with the NiN main types that we are interested in.
+
 ```{r}
@@
 
-We are only interested in a few NiN main types and their mapping units (see [here](https://ninanor.github.io/ecosystemCondition/naturtype.html) for justification)
+We are only interested in a few NiN main types and their mapping units
+(see [here](https://ninanor.github.io/ecosystemCondition/naturtype.html)
+for justification)
+
 ```{r}
@@
 Sub-setting the main data file:
-```{r}
+
+```{r, eval=F}
 naturetypes <- naturetypes[naturetypes$naturtype %in% naturetypes_summary$Nature_type,]
@@
 Fix duplicated name for an ecosystem
+
+```{r, eval=F}
+naturetypes$hovedøkosystem[naturetypes$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+```
+
+Export subsetted data set
+
+```{r, eval=F}
+saveRDS(naturetypes, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
+```
+
+Import back in
+
 ```{r}
-naturetypes$hovedøkosystem[naturetypes$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+naturetypes <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
+
 ```
@@
 Calculate area
+
 ```{r}
@@
 
-The nin variables are combined into a single field which we need to spit apart.
+The nin variables are combined into a single field which we need to spit
+apart.
+
 ```{r}
@@
 
-Convert to numeric. Ignore NA's
+Convert the value column to numeric. Ignore NA's
+
 ```{r}
@@
 Keep only the relevant variables
+
 ```{r}
@@
 
-Columnames starting with a number is problematic, so adding a prefix
+Colum names starting with a number is problematic, so adding a prefix
+
 ```{r}
@@
 
-Now I need to create a single row per locality with a new variable which is a product of the four variables "7SE", "7TK", "PRTK" and "PRSL".
-I will create a new data table where I calculate the new variable which I then paste back into the sf object.
+Now I need to create a single row per locality with a new variable which
+is a product of the four variables "7SE", "7TK", "PRTK" and "PRSL". I
+will create a new data table where I calculate the new variable which I
+then paste back into the sf object.
+
 ```{r}
 naturetypes_wide <- pivot_wider(naturetypes,
-                                names_from = NiN_variable_code,
-                                values_from = NiN_variable_value,
-                                id_cols = identifikasjon_lokalid)
+                                names_from = "NiN_variable_code",
+                                values_from = "NiN_variable_value",
+                                id_cols = "identifikasjon_lokalid")
+naturetypes_wide <- as.data.frame(naturetypes_wide)
+head(naturetypes_wide)
 ```
 
-The variables use slightly different scales.
-PRSL and PRTK use this 8 step scale:
-![](images/mdirPRscale.PNG)
+The variables use slightly different scales. PRSL and PRTK use this 8
+step scale:
 
-7TK and 7SE use a 4 step scale
-![](images/ninAR4b.PNG)
+![](images/mdirPRscale.PNG){width="400"}
 
-I want to base this indicatr on whichever of the variables have the highest value (worst condition). This table shows that a loclity may have been assessed with both a PR variable and a NiN variable
+7TK and 7SE use a 4 step scale ![](images/ninAR4b.PNG)
+
+I want to base this indicator on whichever of the variables have the
+highest value (worst condition). This table shows that a locality may
+have been assessed with both a PR variable and a NiN variable
+
 ```{r}
@@
 ```
-Therefore I need to reconcile these scales. The only way I can think of is to decrease the resolution of the PR variables to the ar4b scale. This is of course a shame to loose this detail.
+
+Therefore I need to reconcile these scales. The only way I can think of
+is to decrease the resolution of the PR variables to the ar4b scale.
+This is of course a shame to loose this detail.
 
 A locality will never have values for both 7SE and PRSL
+
 ```{r}
@@
 ```
+
 nor 7TK and PRTK
+
 ```{r}
@@
 Checking that I successfully combined classes 1+2, 3+4+5 and 6+7.
+
 ```{r}
@@
 ```
-That looks fine. Doing the same for the other variable:
+
+That looks fine. Now doing the same for the other variable PRSL:
+
 ```{r}
@@
 Remove the old PR variables
+
 ```{r}
@@
 
-There are some cells with X's. These are NA's.
+There are some very few cells with X's. These are NA's.
+
 ```{r}
@@
 
-7SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7SE),])` NA's, and 7TK has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7TK),])`.
+```{r}
+head(naturetypes_wide)
+anyDuplicated(naturetypes_wide$identifikasjon_lokalid)
+```
+
+7SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7SE),])`
+NA's, and 7TK has
+`r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7TK),])`.
 
@@
 ```
+
+The NA fraction is quite big. These are localities with 7TK recorded,
+but not 7SE.
+
 ```{r, fig.cap="7TK scores in the nature type dataset"}
@@
 
-```{r}
+Then I can combine these two variables into a composite variable called
+`slitasje`
+
+```{r, warning=F}
 naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
@@
 
-When both variables are NA, we get -Inf. Removing these
+When both variables are NA, we get -Inf. There were nine cases of this.
+Removing these now.
+
 ```{r}
@@
 
-I would also like to know how often 7TK was defining of the slitasje-indicator, and how often it was 7SE. I can do this by taking the difference
+I appears most localities are in good condition.
+
+I would also like to know how often 7TK was defining of the
+slitasje-indicator, and how often it was 7SE. I can do this by taking
+the difference
 
@@
 Now I will copy these slitasje-values into the sf object.
+
 ```{r}
@@
 
-This variable is also recorded in GRUK.
-The nature type dataset I'm working on here includes this data already (presently only 2021 included).
-GRUK also records a related variable: % cover in 5m radii circles.
-This data is not published.In any case it is better to use the harmonized dataset.
+This variable is also recorded in GRUK. The nature type dataset I'm
+working on here includes this data already (presently only 2021
+included). GRUK also records a related variable:
+`% cover in 5m radii circles`. This data is not published. In any case
+it is better to use the harmonized dataset.
 
@@
 
-Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically placed locations each with 18 sample poins.
-In each sample point a circle of 250 m2 is visualised, and the main ecosystem is recorded.
-Depnding oon the main ecosystem, certain NiN variables are also recorded.
-7SE is recorded for våtmark, but not semi-natural areas or naturally open areas.
-7TK is recorded in våtmark and naturally open areas only.
+Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
+placed locations each with 18 sample points. In each sample point a
+circle of 250 m2 is visualised, and the main ecosystem is recorded.
+Depending on the main ecosystem, certain NiN variables are also
+recorded. 7SE is recorded for våtmark, but not semi-natural areas or
+naturally open areas. 7TK is recorded in våtmark and naturally open
+areas only.
 
-Variable    | Våtmark | Naturlig åpne | Semi-nat.
-7SE         | X       |               |
-7TK         | X       | X             |
+| Variable | Våtmark | Naturlig åpne områder | Semi-naturlige områder |
+|----------|---------|-----------------------|------------------------|
+| 7SE      | X       | \-                    | \-                     |
+| 7TK      | X       | X                     | \-                     |
 
-It would be very nice to have 7SE recorded for naturally open areas. This variable is very relevant here.
+: Table showing which main ecosystems the two NiN variables 7SE and 7TK
+is recorded in withing the ANO dataset.
 
-I think I will only use våtmark here since my approach will be to the 'worst value' of the variable 7SE and 7TK (se below).
-I think not having 7SE for naturally open areas will underestimate the degree of degredation in these areas.
+It would be very nice to have 7SE recorded for naturally open areas.
+This variable is very relevant here.
+
+I think I will only use våtmark here since my approach will be to the
+'worst value' of the variable 7SE and 7TK (see below). I think not
+having 7SE for naturally open areas will underestimate the degree of
+degredation in these areas.
 
@@
 
-Each point here is 250 square meters, but the data also contains information about how big a proportion of thi area is made up of the dominant main ecosystem.
-However, there are `r (NAS <- length(ano$andel_hovedoekosystem_punkt[is.na(ano$andel_hovedoekosystem_punkt)]))` NA's here, which is `r round(NAS/nrow(ano), 2)*100`% of the data.
+```{r}
+table(ano$aar)
+```
+
+This dataset only contains data from year 2019 and 2021.
+
+Each point/row here is 250 square meters. The data also contains
+information about how big a proportion of this area is made up of the
+dominant main ecosystem. However, there are
+`r (NAS <- length(ano$andel_hovedoekosystem_punkt[is.na(ano$andel_hovedoekosystem_punkt)]))`
+NA's here, which is `r round(NAS/nrow(ano), 2)*100`% of the data.
+
+It appears the proportion of each circle that is made up of the dominant
+ecosystem was only recorded after year 2019. In fact, the main ecosystem
+was not recorded at all in 2019:
+
+```{r}
+table(ano$hovedtype_250m2, ano$aar)
+```
+
+I can remove the NA's, and thus the 2019 data.
+
 ```{r}
@@
 
+Let's look at the variation in the recorded proportion of ecosystem
+cover
+
 ```{r}
-summary(ano$andel_hovedoekosystem_punkt)
+par(mar=c(5,6,4,2))
+plot(ano$andel_hovedoekosystem_punkt[order(ano$andel_hovedoekosystem_punkt)],
+     ylab="Percentage of the 250 m2 area\ncovered by the main ecosystem")
 ```
 
-Where we have data, this is the distribution of values:
+The zero in there is an obvious mistake.
+
+```{r}
+ano <- ano[ano$andel_hovedoekosystem_punkt>20,]
+```
+
+Here's another plot of the distribution of the same variable:
+
 ```{r, warning=F, fig.cap="Percentage cover of the Våtmark main ecosystem in the 250m2 circle"}
@@
 
-We can see that people tend to record the variable in steps of 5%, and that most sample points are 100% belonging to the same main ecosystem.
+We can see that people tend to record the variable in steps of 5%, and
+that most sample points are 100% belonging to the same main ecosystem.
 
-We want to use area weighting in this indicator, so we can use this percentage cover data to calculate the area.
-Note that both datasets use m2 as area units.
+We want to use area weighting in this indicator, so we can use this
+percentage cover data to calculate the area. Note that both datasets use
+m2 as area units.
+
 ```{r}
@@
 
-Let's now look at the distribution of the variables. First I need to slip the cell values to make them numeric.
+Let's now look at the distribution of the variables. First I need to
+sepertae the variable name fro the values themsolves. Now the data looks
+like this:
+
+```{r}
+ano$bv_7se[1:6]
+```
+
+So I create a new variable prefixed var\_:
+
 ```{r}
@@
 
-The NA's is the case of blank cells. The field app should not allow users leaving this field blank. I will need to remove these rows.
+The NA's is the case of blank cells. The field app should not allow
+users leaving this field blank. I will need to remove these rows. There
+are fourteen cases:
+
+```{r}
+nrow(ano[is.na(ano$var_7SE),])
+```
+
 ```{r}
@@
 ```
+
+Same with the other variable 7TK
 
@@
 
+No NA's this time.
+
 ```{r, fig.cap="Distribution of 7TK and 7SE scores in the ANO data"}
@@
 
-Combining these two variables into the indicator, same as for the nature type data.
+Combining these two variables into the indicator, same as for the nature
+type data.
+
 ```{r}
 temp <- as.data.frame(ano)
-class(temp)
 temp$slitasje <- apply(temp[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
@@
 
+```{r, fig.cap="Distribution of slitasje scores in the ANO data"}
+barplot(table(ano$slitasje), xlab="Sliatsje scores")
+```
+
 ### Combine datasets
-We need to combine the nature type dataset with th ano dataset
-I will tag the ano rows in the nature type column, and add data abot the main ecosystem
+
+We need to combine the nature type dataset with the ANO dataset. I will
+add a column `origin` to show where the data comes from. I will also and
+add data about the main ecosystem.
+
 ```{r}
-ano$naturtype <- "ANO"
+ano$origin <- "ANO"
+naturetypes$origin <- "Nature type mapping"
 ano$hovedøkosystem <- "Våtmark"
@@
 
-I will loose the sf class and the crs when joining the datasets, so I will save the crs
+I will loose the sf class and the crs when joining the datasets, so I
+will save the crs
+
 ```{r}
@@
 Fix class
+
 ```{r}
@@
 
-I can reduce the number of column to keep things a bit more tidy
+I can reduce the number of columns to keep things a bit more tidy
+
 ```{r}
@@
                                   GlobalID,
+                                  origin,
                                   kartleggingsår,
@@
                                   identifikasjon_lokalid,
+                                  origin,
                                   hovedøkosystem,
@@
 
+Let's see the propotion of datapoints (not area) origination from each
+data set
+
+```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator."}
+barplot(table(slitasje_data$origin),
+        ylab="Number of localities")
+```
+
+So the ANO data is not very important here, bu it can become more
+important in the future.
+
 #### Vegetation zones (bioregions)
-I want to look at if the indicator values can be averages over smaller and more meaningful areas (ecological homogeneous areas) than the 5 default regions.
+
+I want to look at if the indicator values can be averages over smaller
+and more meaningful areas (ecological homogeneous areas) than the 5
+default regions.
+
 ```{r}
@@
 
-It could make sense to combine some zones, but I will later split the zones by region, and then merge sones that are very uncommon in a spesific region.
+It could make sense to combine some zones, but I will later split the
+zones by region, and then merge sones that are very uncommon in a
+spesific region.
 
 #### Regions
+
 Importing a shape file with the regional delineation.
+
 ```{r}
@@
 ### Scaled indicator values
+
 <!-- Text and analyses here -->
-I can scale the indicator for each polygon, or I can chose to aggregate them first.
-If the scaled value is representative and precise at the polygon level, then I could scale at that level.
-However, the combined surveyed area is still so little of the total area of Norway, so that only producing indicator values for these leaves the indicator without much value for regional assessments.
-Later when we produce assessments and indicator values at the regional level, we cannot simply do an area weighting of the polygons in each region.
-This is because we don't want to assume that the polygons are representative outside of the mapped area.
 
-An option is to export one map with just the raw un-scaled slitasje variable at the polygon level, and another, scaled indicator map aggregated to either
+I can scale the indicator for each polygon, or I can chose to aggregate
+them first. If the scaled value is representative and precise at the
+polygon level, then I could scale at that level. However, the combined
+surveyed area is still so little of the total area of Norway, so that
+only producing indicator values for these leaves the indicator without
+much value for regional assessments. Later when we produce assessments
+and indicator values at the regional level, we cannot simply do an area
+weighting of the polygons in each region. This is because we don't want
+to assume that the polygons are representative outside of the mapped
+area.
 
-* each *dekningskart* (map of surveyed area) 
-* each vegetation zone (e.g. boreonemal zone in Østlandet; excluding zone with too little data points)
-* each vegetation region (aka bioregion, e.g. boreonemal zone and continental section in Østlandet)
-* each pixel of a give size, say 1 x 1 km
+An option is to export one map with just the raw un-scaled slitasje
+variable at the polygon level, and another, scaled indicator map
+aggregated to either
 
-The last option is problematic because we often will have too few points to be able to estimate  credible mean. 
-I can try the second option first, and see how many data points we get for each zone.
+-   each *dekningskart* (map of surveyed area)
+-   each vegetation zone (e.g. boreonemal zone in Østlandet; excluding
+    zone with too little data points)
+-   each vegetation region (aka bioregion, e.g. boreonemal zone and
+    continental section in Østlandet)
+-   each pixel of a give size, say 1 x 1 km
+
+The last option is problematic because we often will have too few points
+to be able to estimate credible mean. I can try the second option first,
+and see how many data points we get for each zone.
 
 ### Uncertainty
+
 <!-- Text here -->
@@
 ## Prepare export
+
 <!-- Text here -->
@@
 ### Eksport file (final product)
+
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2022-12-14 DRAFT_slitasjeOgKjørespor.Rmd
text

@@
 
+#### Infrastructure index
+
 #### Vegetation zones (bioregions)
 
-I want to look at if the indicator values can be averages over smaller
-and more meaningful areas (ecological homogeneous areas) than the 5
-default regions.
+I want to look at wheter the indicator values can be averages over
+smaller and more meaningful areas (ecological homogeneous areas) than
+the 5 default regions. One option is this data set with vegetation zone.
 
-```{r}
+```{r, eval=F}
 path <- ifelse(dir == "C:", 
@@
 
-```{r}
+```{r, eval=F}
 soner <- sf::read_sf(path)
@@
 
-```{r}
+```{r, eval=F}
 tm_shape(soner)+
@@
 It could make sense to combine some zones, but I will later split the
-zones by region, and then merge sones that are very uncommon in a
-spesific region.
+zones by region, and then merge zones that are very uncommon in a
+specific region.
 
@@
 weighting of the polygons in each region. This is because we don't want
-to assume that the polygons are representative outside of the mapped
-area.
+to assume that the polygons are representative far outside of the mapped
+area. But perhaps we can assume them to be representative inside
+*homogenous ecological areas*.
 
@@
 
--   each *dekningskart* (map of surveyed area)
--   each vegetation zone (e.g. boreonemal zone in Østlandet; excluding
-    zone with too little data points)
--   each vegetation region (aka bioregion, e.g. boreonemal zone and
+1.  each *dekningskart* (map of surveyed area)
+
+2.  each vegetation zone (e.g. boreonemal zone in Østlandet; excluding
+    zones with too little data points)
+
+3.  each vegetation region (aka bioregion, e.g. boreonemal zone and
     continental section in Østlandet)
--   each pixel of a give size, say 1 x 1 km
 
-The last option is problematic because we often will have too few points
-to be able to estimate credible mean. I can try the second option first,
-and see how many data points we get for each zone.
+4.  each region cross tabulated agains infrastructure index classes
+
+5.  each pixel of a give size, say 1 x 1 km
+
+Option 5 is problematic because we often will have too few points to be
+able to estimate credible mean. The second and third option can be good
+for certin biotic variables that depends on local species pools ect.
+However, slitasje is perhaps best explained by human presence, and
+option 4 therefore sounds reasonable.
 

### 2022-12-16 DRAFT_slitasjeOgKjørespor.Rmd
infrastruktur index load

@@
 
+#### Outline of Norway and regions
+
+```{r}
+outline <- sf::read_sf("data/outlineOfNorway_EPSG25833.shp")
+```
+
+```{r}
+regions <- sf::read_sf("data/regions.shp")
+```
+
 #### Infrastructure index
+The infrastructure index is explained [here](https://brage.nina.no/nina-xmlui/handle/11250/2983607).
+I use it as a kind of pressure or human influence index. It is likely to be a good predictor for *slitasje*.
+
+```{r}
+# The original data is here, on the R server, but it is not always working, so I also include a copy on the P server.
+
+#path <- ifelse(dir == "C:", 
+#        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original",
+#        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original")
+#
+
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks")
+
+```
+
+```{r, eval=F}
+infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
+```
+
+It's easuer to see what's happeing if we zoom in a bit. 
+Lets get a boundary box around Trondheim.
+```{r}
+myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
+                ymin = 7032142.5, ymax = 7045245.27),
+                crs = st_crs(infra))
+```
+
+Cropping the raster to speed up processing.
+```{r}
+infra_trd <- sf::st_crop(infra, myBB)
+```
+
+Transform to lat long dye to osm requirements
+```{r, warning=F}
+infra_trd_ll <- sf::st_transform(infra_trd, 4326)
+```
+
+Get the boundary box of the cropped raster
+```{r}
+myBB_ll <- sf::st_bbox(infra_trd_ll)
+```
+
+Download highways using the above bbox.
+```{r, warning=FALSE, message=F, eval=F}
+hw <- 
+  osmplotr::extract_osm_objects(
+    bbox = myBB_ll,
+    key = "highway",
+    sf = T)
+```
+
+Transforming highway data back into utm, although not strictly necessary.
+```{r, eval=F}
+hw_utm <- sf::st_transform(hw, sf::st_crs(infra_trd)) 
+```
+
+This object contains too many roads. I'll take out the unnamed roads. 
+```{r, eval=F}
+hw_utm <- hw_utm[!is.na(hw_utm$name),]
+```
+
+Caching the sf object
+```{r, eval=F}
+saveRDS(hw_utm, "data/cache/highways_trondheim.rds")
+```
+
+Load object
+```{r}
+hw_utm <- readRDS("data/cache/highways_trondheim.rds")
+```
+
+```{r, fig.cap="Infrastrcture indeks over Trondheim. Major roads are in white."}
+tm_shape(infra_trd)+
+  tm_raster(title="Infrastructure Index",
+            style="cont",
+            palette = "viridis")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="white")
+```
+
+The infrastructure index is in EPSG32633. I will need to transform it. It takes a couple of minutes, so I will store the result in P and load it back in afterwards.
+```{r, eval=F}
+
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks")
+
+infra <- sf::st_transform(infra, 25833)
+stars::write_stars(infra, paste0(path, "/infrastrukturindeksEPSG_EPSG35833.tif"))
+```
+
+```{r}
+infra2 <- stars::read_stars(paste0(path, "/infrastrukturindeksEPSG_EPSG35833.tif"))
+```
 

### 2022-12-16 DRAFT_slitasjeOgKjørespor.Rmd
plan

@@
 
+Import a stars proxy (no data imported yet)
 ```{r, eval=F}
@@
 
-It's easuer to see what's happeing if we zoom in a bit. 
+It's easier to see what's happening if we zoom in a bit. 
 Lets get a boundary box around Trondheim.
@@
 ```{r, fig.cap="Infrastrcture indeks over Trondheim. Major roads are in white."}
-tm_shape(infra_trd)+
+map_trd <- tm_shape(infra_trd)+
   tm_raster(title="Infrastructure Index",
@@
   tm_lines(col="white")
+map_trd
 ```
@@
 
-#### Vegetation zones (bioregions)
-
-I want to look at wheter the indicator values can be averages over
-smaller and more meaningful areas (ecological homogeneous areas) than
-the 5 default regions. One option is this data set with vegetation zone.
-
-```{r, eval=F}
-path <- ifelse(dir == "C:", 
-               "R:/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp",
-               "/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp")
+```{r}
+names(infra_trd) <- "infrastructureIndex"
+infra_trd_reclassed <-  infra_trd %>%
+  mutate(infrastructureIndex = case_when(
+    infrastructureIndex < 1 ~ 0,
+    infrastructureIndex < 6 ~ 1,
+    infrastructureIndex < 9 ~ 2,
+    infrastructureIndex >= 9 ~ 3
+  ))
 ```
 
-```{r, eval=F}
-soner <- sf::read_sf(path)
+```{r}
+map_trd_reclassed <- tm_shape(infra_trd_reclassed)+
+  tm_raster(title="Infrastructure Index",
+            #palette = "viridis",
+            style="cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="white")
+
 ```
 
-```{r, eval=F}
-tm_shape(soner)+
-  tm_polygons(col = "NAVN",
-              title = "Vegetation zone")
+```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
+tmap_arrange(map_trd,
+             map_trd_reclassed,
+             ncol=1)
 ```
 
-It could make sense to combine some zones, but I will later split the
-zones by region, and then merge zones that are very uncommon in a
-specific region.
-
-#### Regions
-
-Importing a shape file with the regional delineation.
-
-```{r}
-#reg <- sf::st_read("data/regions.shp")
-#st_crs(reg)
-```
+I tweaked the thresholds so that most of the forest next to Trondheim was i the second lowest class.
 
@@
 *homogenous ecological areas*.
-
-An option is to export one map with just the raw un-scaled slitasje
-variable at the polygon level, and another, scaled indicator map
-aggregated to either
-
-1.  each *dekningskart* (map of surveyed area)
-
-2.  each vegetation zone (e.g. boreonemal zone in Østlandet; excluding
-    zones with too little data points)
-
-3.  each vegetation region (aka bioregion, e.g. boreonemal zone and
-    continental section in Østlandet)
-
-4.  each region cross tabulated agains infrastructure index classes
-
-5.  each pixel of a give size, say 1 x 1 km
-
-Option 5 is problematic because we often will have too few points to be
-able to estimate credible mean. The second and third option can be good
-for certin biotic variables that depends on local species pools ect.
-However, slitasje is perhaps best explained by human presence, and
-option 4 therefore sounds reasonable.
+That's where the infrastructure index comes in.
+I will scale the indicators at the polygon level.
+I then take the simplified infrastructure index and, one class at the time, extract the indicator value that intersect with those cells, and extrapolate the mean of those values to all similar cell within a given region.
 
@@
 <!-- Text here -->
+Bootstrapping of nin-polygons per infrastructure class and region?
 

### 2022-12-19 DRAFT_slitasjeOgKjørespor.Rmd
plan

@@
 #### Infrastructure index
-The infrastructure index is explained [here](https://brage.nina.no/nina-xmlui/handle/11250/2983607).
-I use it as a kind of pressure or human influence index. It is likely to be a good predictor for *slitasje*.
+
+The infrastructure index is explained
+[here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). I use it
+as a kind of pressure or human influence index. It is likely to be a
+good predictor for *slitasje*.
 
@@
 Import a stars proxy (no data imported yet)
+
 ```{r, eval=F}
@@
 
-It's easier to see what's happening if we zoom in a bit. 
-Lets get a boundary box around Trondheim.
+It's easier to see what's happening if we zoom in a bit. Lets get a
+boundary box around Trondheim.
+
 ```{r}
@@
 Cropping the raster to speed up processing.
+
 ```{r}
@@
 Transform to lat long dye to osm requirements
+
 ```{r, warning=F}
@@
 Get the boundary box of the cropped raster
+
 ```{r}
@@
 Download highways using the above bbox.
+
 ```{r, warning=FALSE, message=F, eval=F}
@@
 
-Transforming highway data back into utm, although not strictly necessary.
+Transforming highway data back into utm, although not strictly
+necessary.
+
 ```{r, eval=F}
@@
 This object contains too many roads. I'll take out the unnamed roads.
+
 ```{r, eval=F}
@@
 Caching the sf object
+
 ```{r, eval=F}
@@
 Load object
+
 ```{r}
@@
 
-The infrastructure index is in EPSG32633. I will need to transform it. It takes a couple of minutes, so I will store the result in P and load it back in afterwards.
+The infrastructure index is in EPSG32633. I will need to transform it.
+It takes a couple of minutes, so I will store the result in P and load
+it back in afterwards.
+
 ```{r, eval=F}
@@
 
-I tweaked the thresholds so that most of the forest next to Trondheim was i the second lowest class.
+I tweaked the thresholds so that most of the forest next to Trondheim
+was i the second lowest class.
 
 ### Scaled indicator values
-
-<!-- Text and analyses here -->
 
@@
 area. But perhaps we can assume them to be representative inside
-*homogenous ecological areas*.
-That's where the infrastructure index comes in.
-I will scale the indicators at the polygon level.
-I then take the simplified infrastructure index and, one class at the time, extract the indicator value that intersect with those cells, and extrapolate the mean of those values to all similar cell within a given region.
+*homogenous ecological areas*. That's where the infrastructure index
+comes in. Here's the plan:
+
+1.  scale the indicators at the polygon level.
+
+2.  take the simplified infrastructure index and, one class at the time,
+    extract the indicator value that intersect with those cells, and
+    extrapolate the mean of those values to all similar cell within a
+    given region.
+
+3.  calculate an indicator value for a region by doing a weighted
+    averega based o the relative area of each infrastructure class.
 
@@
 <!-- Text here -->
+
 Bootstrapping of nin-polygons per infrastructure class and region?

### 2022-12-22 index.Rmd
doc: Add reference to eaTools

@@
 
-# About {-}
+# About <img src="images/hexSticker_EC.png" align="right" height="139"/> {.unnumbered}
 
@@
 These indicators are not yet used in real assessments, and may represent unfinished work pending future data availability.
-In addition,  we may chose to also present some analyses documenting indicators that _are_ used in existing assessments of ecosystem condition.
+In addition, we may chose to also present some analyses documenting indicators that *are* used in existing assessments of ecosystem condition.
 
-## Community review {-}
+## eaTools <img src="images/eaTools-logo.png" align="right" height="139"/> {.unnumbered}
+
+`eaTools` is an R package aimed at solving common tasks in GIS oriented ecosystem accounting, especially ecosystem condition accounting.
+We will use functions from this package, such as this function to normalise condition variables seen in Figure \@ref(fig:eaTools-example).
+
+```{r eaTools-example, fig.cap="A normalised condition variable is called an indicator. This indicator had been transformed with a sigmoid function."}
+#devtools::install_github("NINAnor/eaTools")
+library(eaTools)
+
+# Import example data
+data("ex_polygons")
+
+# Use the questionmark to readabout the example data sets or to view the  function documentation 
+#?ex_polygons
+
+eaTools::ea_normalise(data = ex_polygons,
+             vector = "condition_variable_2",
+             upper_reference_level = 7,
+             scaling_function = "sigmoid",
+             plot = T)
+```
+
+See the [package documentation](https://ninanor.github.io/eaTools/index.html), especially under *Articles* for more information and examples of use.
+
+## Community review {.unnumbered}
 
@@
 Therefor we encourage all types of feedback on the analyses of the indicators for ecological condition that you can read about on this web site.
-To submit your comments, please to to [*this github repository*](https://github.com/NINAnor/ecosystemCondition/issues) and open a new _issue_ and we will answer them there.
+To submit your comments, please to to [*this github repository*](https://github.com/NINAnor/ecosystemCondition/issues) and open a new *issue* and we will answer them there.
 
-<img src="images/newIssue.PNG" alt="" width="600"/>
+<img src="images/newIssue.PNG" width="600"/>
 
-## Author guidelines {-}
+## Author guidelines {.unnumbered}
 
@@
 The indicator values, reference values ect should be mapped to the following columns (vector data) or bands (raster data):
-* `v_YYYY` for variable value (i.e. un-scaled or raw indicator values) for year YYYY. 
-* `sd_YYYY` for the standard deviation associated with v_YYYY. Other measures for uncertainty is currently not supported. 
-* `referance_high` for the reference value, i.e. the value of the variable *v* under reference conditions. This value cannot vary between years.
-* `referance_low` for the lower limit reference value, i.e. the worst possible value of the variable *v*. This value cannot vary between years. If missing, this value is assumed to be zero.
-* `thr` for the threshold value defining *good condition* for variable *v*. If missing, this value will default to 0.6.
-* `i_YYYY` for the indicator value at year YYYY.
+
+-   `v_YYYY` for variable value (i.e. un-scaled or raw indicator values) for year YYYY.
+-   `sd_YYYY` for the standard deviation associated with v_YYYY. Other measures for uncertainty is currently not supported.
+-   `referance_high` for the reference value, i.e. the value of the variable *v* under reference conditions. This value cannot vary between years.
+-   `referance_low` for the lower limit reference value, i.e. the worst possible value of the variable *v*. This value cannot vary between years. If missing, this value is assumed to be zero.
+-   `thr` for the threshold value defining *good condition* for variable *v*. If missing, this value will default to 0.6.
+-   `i_YYYY` for the indicator value at year YYYY.
 This value depends on all the above values, but also the scaling function (.e.
@@
 
-* Obtain a _go ahead_ signal from the coordination group to make sure you are free to start your work
-* Read the [`template.Rmd`](template.Rmd) file.
+-   Obtain a *go ahead* signal from the coordination group to make sure you are free to start your work
+-   Read the [`template.Rmd`](template.Rmd) file.
 Make a copy of it and store it, in the same (root) folder as `DRAFT_myIndicator.Rmd`.
@@
 Finish your documentation and analyses there.
-  + You can load data from internal NINA servers, from web hosts, of store smaller data set under `data`.
+    -   You can load data from internal NINA servers, from web hosts, of store smaller data set under `data`.
 The `data` folder also contains some supporting data sets that you may want to use, such as a delineation of the five regions used in the ecosystem assessment (`data/regions.shp`), and an outline of Norway (`outlineOfNorway_EPSG25833.shp`).
-  + To preview you rmarkdown file as html in R, type `rmarkdown::render('DRAFT_myIndicator.Rmd', output_format = "html_document", output_dir = "temp")` in the console (the knitr shortcuts don't work as expected inside bookdown projects).
-  + When filling out the template.Rmd file with your own work, try not to change the headers too much.
-  + Final result from the analyses should be written to the `indicators` folder.
+    -   To preview you rmarkdown file as html in R, type `rmarkdown::render('DRAFT_myIndicator.Rmd', output_format = "html_document", output_dir = "temp")` in the console (the knitr shortcuts don't work as expected inside bookdown projects).
+    -   When filling out the template.Rmd file with your own work, try not to change the headers too much.
+    -   Final result from the analyses should be written to the `indicators` folder.
 The preferred output is georeferenced data (rasters or shape files), and these should be placed under `indators/indicatorMaps`.
 Non-georeferenced data, like data tables, can be stored under `indicators`.
-* Make a pull request to the main [*GitHub repository*](https://github.com/NINAnor/ecosystemCondition) to submit your file.
+-   Make a pull request to the main [*GitHub repository*](https://github.com/NINAnor/ecosystemCondition) to submit your file.
 If you are not comfortable using GitHub, contac Anders to rrange sending the files via email.
-* Anders or someone else in the project wil conduct a rapid code review, making sure the code is reproducible, interpretable and that it renders locally on the R Studio server.
-* When approved, the reviewer will make a copy of the DRAFT_ file, removing the DRAFT_ part of the name, as well as the YML header, and as well as updating the rmd_files part of _bookdown.yml.
+-   Anders or someone else in the project wil conduct a rapid code review, making sure the code is reproducible, interpretable and that it renders locally on the R Studio server.
+-   When approved, the reviewer will make a copy of the DRAFT\_ file, removing the DRAFT\_ part of the name, as well as the YML header, and as well as updating the rmd_files part of \_bookdown.yml.
 (For an example of such a file, see `breareal.Rmd`.)
-* To update the web site (this web site), an admin user needs to pull down the repo to R Studio server, compile the book there (Ctr+Shift+B), commit and push to main.
+-   To update the web site (this web site), an admin user needs to pull down the repo to R Studio server, compile the book there (Ctr+Shift+B), commit and push to main.
 The html version of the book will be in the `docs` folder.

### 2022-12-22 R/functionalPlantIndicators/data handling.R
upload and handling ANO data

@@ -8,12 +8,16 @@ library(tidyverse)
 
 # url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
 
-# download from P-drive
+#### download data from P-drive ####
 
 # ANO
 st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-ANO.dat <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-head(ANO.dat)
+ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+                   layer="ANO_Art")
+ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+                  layer="ANO_SurveyPoint")
+head(ANO.sp)
+head(ANO.geo)
 
 # Tyler indicator data
 ind.Tyler <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
@@ -28,6 +32,7 @@ head(ind.Grime)
 load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
 str(Eco_State)
 
+#### data handling - functional indicator data ####
 # trimming away sub-species & co, and descriptor info
 ind.Grime[,'species.orig'] <- ind.Grime[,'species']
 ind.Grime[,'species'] <- word(ind.Grime[,'species'], 1,2)
@@ -118,3 +123,47 @@ ind.dat$species <- as.factor(ind.dat$species)
 summary(ind.dat$species)
 head(ind.dat)
 
+#### data handling - ANO data ####
+head(ANO.sp)
+head(ANO.geo)
+
+# fix species names
+ANO.sp$Species <- ANO.sp$art_navn
+unique(as.factor(ANO.sp$Species))
+#ANO.sp$Species <- sub(".*?_", "", ANO.sp$Species) # lose the Norwegian name in the front
+ANO.sp[,'Species'] <- word(ANO.sp[,'Species'], 1,2) # lose subspecies
+ANO.sp$Species <- str_to_title(ANO.sp$Species) # make first letter capital
[…]

### 2022-12-22 DRAFT_slitasjeOgKjørespor.Rmd
style: text edits

@@
 
-<!-- Replace 'indicator name' with yor own indicator name  -->
-
 # Slitasje og kjørespor
-
-<br />
 
@@
 ```
-
-<br />
-
-<!-- Load all you dependencies here -->
 
@@
 ```
-
-<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
 
@@
 
-<!-- Don't remove these three html lines -->
-
-<br /> <br />
-
-<hr />
-
-<!-- Document you work below.
-Try not to change  the headers too much.
-Data can be stored on NINA server.
-Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine.
-If it is not too big you may store under /data/ on this repository -->
+<hr/>
 
@@
 ```{r}
-naturetypes <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
-
+path <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
+naturetypes <- readRDS(path)
 ```
@@
 
-Calculate area
+Calculate area for each polygon.
 
@@
 
-The nin variables are combined into a single field which we need to spit
+The NiN variables are combined into a single field which we need to spit
 apart.
@@
 
-Convert the value column to numeric. Ignore NA's
+Convert the value column to numeric. Ignore NA's.
 
@@
 
-Keep only the relevant variables
+Keep only the relevant variables.
 
@@
 
-Colum names starting with a number is problematic, so adding a prefix
+Names starting with a number is problematic, so adding a prefix
 
@@
 
-Remove the old PR variables
+Remove the old PR variables.
 
@@
 
-This dataset only contains data from year 2019 and 2021.
+This dataset only contains data from year 2019 and 2021. We need to
+update this dataset later.
 
@@
 
-```{r}
+```{r, fig.cap="Distribution of the ANO variable andel_hovedoekosystem_punkt for wetland localities."}
 par(mar=c(5,6,4,2))
@@
 Let's now look at the distribution of the variables. First I need to
-sepertae the variable name fro the values themsolves. Now the data looks
+seperate the variable name fro the values themselves. Now the data looks
 like this:
@@
 
-I can reduce the number of columns to keep things a bit more tidy
+I can reduce the number of columns to keep things a bit more tidy.
 
-```{r}
+<!--# this join doesnt work! -->
+
+```{r, eval=F}
 temp <- as.data.frame(ano)

### 2023-01-02 DRAFT_slitasjeOgKjørespor.Rmd
normalise

@@
 library(units)
+library(eaTools)
 library(tidyverse)
@@
 
-![](images/mdirPRscale.PNG){width="400"}
+```{r eight-step, fig.cap="Eight step condition scale"}
+knitr::include_graphics("images/mdirPRscale.PNG")
+```
 
-7TK and 7SE use a 4 step scale ![](images/ninAR4b.PNG)
+7TK and 7SE use a 4 step scale ![Four step condition
+scale](images/ninAR4b.PNG)
 
@@
 
-I will loose the sf class and the crs when joining the datasets, so I
-will save the crs
-
-```{r}
-sf::st_crs(ano) == sf::st_crs(naturetypes) # TRUE
-myCRS <- st_crs(ano)
-```
-
 Fix class
@@
 
-I can reduce the number of columns to keep things a bit more tidy.
+I use `dplyr::select` to can reduce the number of columns to keep things
+a bit more tidy.
 
-<!--# this join doesnt work! -->
-
-```{r, eval=F}
-temp <- as.data.frame(ano)
-temp2 <- as.data.frame(naturetypes)
-slitasje_data <- full_join(select(temp,
+```{r}
+slitasje_data <- dplyr::bind_rows(select(ano,
                                   GlobalID,
@@
                                   SHAPE), 
-                           select(temp2,
+                           select(naturetypes,
                                   identifikasjon_lokalid,
@@
                                   SHAPE))
-# convert back to sf object
-slitasje_data <- st_as_sf(slitasje_data)
-st_crs(slitasje_data) <- myCRS
+
+unique(st_geometry_type(slitasje_data))
+```
+
+The ANO data is point data, and the nature type data is multipolygon.
+
+Grouping and tallying the data for plotting
+
+```{r}
+temp <- as.data.frame(table(slitasje_data$slitasje))
 ```
@@
 ```{r, fig.cap="Slitasje scores."}
-temp <- slitasje_data %>%
-  group_by(slitasje) %>%
-  summarise(sum = n())
-
-ggplot(temp, aes(x = factor(slitasje),
-                 y = sum))+
+ggplot(temp, aes(x = Var1,
+                 y = Freq))+
   geom_bar(stat="identity",
@@
 
-Let's see the propotion of datapoints (not area) origination from each
+Let's see the proportion of data points (not area) origination from each
 data set
@@
 ```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator."}
-barplot(table(slitasje_data$origin),
-        ylab="Number of localities")
+temp <- as.data.frame(table(slitasje_data$origin))
+
+ggplot(temp, aes(x = Var1,
+                 y = Freq))+
+  geom_bar(stat="identity",
+           fill="grey",
+           colour = "black")+
+  theme_bw(base_size = 12)+
+  labs(x = "Data origin",
+       y = "Number of localities")
 ```
 
-So the ANO data is not very important here, bu it can become more
+So the ANO data is not very important here, but it can become more
 important in the future.
@@
 3.  calculate an indicator value for a region by doing a weighted
-    averega based o the relative area of each infrastructure class.
+    average based o the relative area of each infrastructure class.
+
+#### Scaled variable
+
+The variable is ordinal, not continuous. See fig. \@ref(fig:eight-step).
+I could create another variable which was more like original range
+scale. Although not perfect, I will use the median value for each category.
+
+```{r}
+slitasje_data$slitasje_percentage <- slitasje_data$slitasje
+
+slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 1] <- ((1/16)/2)*100
+slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 2] <- (((1/2)/2)+(1/16))*100
+slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 3] <- (3/4)*100
+table(slitasje_data$slitasje_percentage)
+```
+
+```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable"}
+ggplot(slitasje_data, aes(x = slitasje,
+                          y = slitasje_percentage))+
+  geom_point(size=6)+
+  geom_line()+
+  theme_bw(base_size = 16)
+```
+
+I will use the same reference levels/values for all of Norway:
+```{r}
+upper <- 0
+lower <- 100
+threshold <- 10
+```
+
+This implies that it is impossible to detect or measure a state when it is in its most degraded form (100% disturbed).
+This is a problem with the original data resolution.
+But I will compensate somewhat for this by using a non-linear transformation.
+
+I need to do a little trick and reverse the upper and lower reference values for the normalisation to work. This is a bug which can be fixed inside eaTools.
+```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
+eaTools::ea_normalise(data = slitasje_data,
+                      vector = "slitasje_percentage",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      plot=T,
+                      reverse = T
+                      )
+```
+
+This normalisation seems reasonable to me. I can save it as indicator values. I will assign the indicator value to the avergae year of the data.
+```{r}
+mean(slitasje_data$kartleggingsår)
+```
+
+```{r}
+slitasje_data$i_2020 <- eaTools::ea_normalise(data = slitasje_data,
+                      vector = "slitasje_percentage",
+                      upper_reference_level = lower,
+                      lower_reference_level = upper,
+                      break_point = threshold,
+                      reverse = T
+                      )
+```
 

### 2023-01-02 DRAFT_slitasjeOgKjørespor.Rmd
fix: point to polygon

@@
 Fix class
-
 ```{r}
@@
 The ANO data is point data, and the nature type data is multipolygon.
+Because later we want to save this as a shape file, we cannot have a mixed class type.
+I will therefore convert the points to polygons.
+I use the area column to calculate a radii that gives that area.
+```{r}
+slitasje_data_points <- slitasje_data %>%
+  mutate(g_type = st_geometry_type(.)) %>%
+  filter(g_type =="POINT") %>%
+  st_buffer(sqrt(slitasje_data$area/pi))
+```
+
+Checking no that the new polygons have the area corresponding to the proportion of the point that was part of the same main ecosystem:
+```{r new-area, fig.cap="Checking that the area of the new polygons fall in line with the proportion of each point which is part of the main ecosystem."}
+slitasje_data_points$area2 <- st_area(slitasje_data_points)
+plot(slitasje_data_points$area, slitasje_data_points$area2,
+     xlab = "Target area",
+     ylab = "Area of the new polygons")
+```
+
+The area calculation seems to have worked fine.
+
+```{r}
+slitasje_data_polygons <- slitasje_data %>%
+  mutate(g_type = st_geometry_type(.)) %>%
+  filter(g_type !="POINT")
+
+slitasje_data <- bind_rows(slitasje_data_points, slitasje_data_polygons)
+
+unique(st_geometry_type(slitasje_data))
+```
+
+The new dataset only contains polygons.
 
 Grouping and tallying the data for plotting
-
 ```{r}
@@
 ## Prepare export
+According to the workflow spesifications, I will export the indicator map at its highest resolution, but I also need to export a map of the *homogeneous areas* over which I will aggregate the indicator values.
+For the finest resolution indicator map I will not include columns for uncertainty (because this will only be estimated at the aggregated level), or reference levels (because there is no spatial variation).
 
-<!-- Text here -->
+```{r}
+exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
+              origin,
+              area,
+              i_2020,
+              SHAPE
+              )
+
+```
+
+```{r, eval=F}
+st_write(exp_wetland, "indicators/indicatorMaps/wetland_slitasje.shp",
+         append = F)
+```
 
@@
 
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->
+<!-- Export final file. Ideally a georeferenced shape or raster with indicators values (raw and normalised), reference values and errors. -->

### 2023-01-16 DRAFT_slitasjeOgKjørespor.Rmd
doc: Fix headers

--- 2023-01-02 DRAFT_slitasjeOgKjørespor.Rmd
@@
 
-### Combine datasets
+##### Combine Naturtype data and ANO
 
@@
 
+###### Points to polygons
 The ANO data is point data, and the nature type data is multipolygon.
+Because later we want to save this as a shape file, we cannot have a
+mixed class type. I will therefore convert the points to polygons. I use
+the area column to calculate a radii that gives that area.
+
+```{r}
+slitasje_data_points <- slitasje_data %>%
+  mutate(g_type = st_geometry_type(.)) %>%
+  filter(g_type =="POINT") %>%
+  st_buffer(sqrt(slitasje_data$area/pi))
+```
+
+Checking no that the new polygons have the area corresponding to the
+proportion of the point that was part of the same main ecosystem:
+
+```{r new-area, fig.cap="Checking that the area of the new polygons fall in line with the proportion of each point which is part of the main ecosystem."}
+slitasje_data_points$area2 <- st_area(slitasje_data_points)
+plot(slitasje_data_points$area, slitasje_data_points$area2,
+     xlab = "Target area",
+     ylab = "Area of the new polygons")
+```
+
+The area calculation seems to have worked fine.
+
+```{r}
+slitasje_data_polygons <- slitasje_data %>%
+  mutate(g_type = st_geometry_type(.)) %>%
+  filter(g_type !="POINT")
+
+slitasje_data <- bind_rows(slitasje_data_points, slitasje_data_polygons)
+
+unique(st_geometry_type(slitasje_data))
+```
+
+The new dataset only contains polygons.
 
@@
 as a kind of pressure or human influence index. It is likely to be a
-good predictor for *slitasje*.
+good predictor for *slitasje*, and I think to use it to define local reference area,
+or homogeneous (impact) areas.
 
@@
 
-```{r, eval=F}
+```{r}
 infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
@@
 
-Transform to lat long dye to osm requirements
+Transform to lat long due to osm requirements
 
@@
 ```{r, eval=F}
-
 path <- ifelse(dir == "C:", 
@@
 ```
+
+Here I define a simplified categorical typology for the infrastructure
+index using four classes.
 
@@
 I tweaked the thresholds so that most of the forest next to Trondheim
-was i the second lowest class.
+was in the second lowest class. This looks pretty good to me.
 
@@
 area. But perhaps we can assume them to be representative inside
-*homogenous ecological areas*. That's where the infrastructure index
+*homogeneous ecological areas*. That's where the infrastructure index
 comes in. Here's the plan:
@@
 I could create another variable which was more like original range
-scale. Although not perfect, I will use the median value for each category.
+scale. Although not perfect, I will use the median value for each
+category.
 
@@
 I will use the same reference levels/values for all of Norway:
+
 ```{r}
@@
 
-This implies that it is impossible to detect or measure a state when it is in its most degraded form (100% disturbed).
-This is a problem with the original data resolution.
-But I will compensate somewhat for this by using a non-linear transformation.
+This implies that it is impossible to detect or measure a state when it
+is in its most degraded form (100% disturbed). This is a problem with
+the original data resolution. But I will compensate somewhat for this by
+using a non-linear transformation.
 
-I need to do a little trick and reverse the upper and lower reference values for the normalisation to work. This is a bug which can be fixed inside eaTools.
+I need to do a little trick and reverse the upper and lower reference
+values for the normalisation to work. This is a bug which can be fixed
+inside eaTools.
+
 ```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
@@
 
-This normalisation seems reasonable to me. I can save it as indicator values. I will assign the indicator value to the avergae year of the data.
+This normalisation seems reasonable to me. I can save it as indicator
+values. I will assign the indicator value to the average year of the
+data.
+
 ```{r}
@@
 
+### Homogeneous impact areas
+
+```{r, eval=F}
+eaTools::ea_homogeneous_area(infra2)
+```
+
 ### Uncertainty
@@
 
-## Prepare export
+## Eksport files (final product)
 
-<!-- Text here -->
+According to the workflow specifications, I will export the indicator
+map at its highest resolution, for use in various web interfaces etc.
 
-### Eksport file (final product)
+But I also need to export a map of the indicator where we extrapolate the indicator values
+to the homogeneous areas. This map should be interpreted with caution, since they seem
+to provide explicit indicator values for areas where there is no data. This map is mainly for
+aggregation purposes. I will mark this map with the prefix _homogeneous-areas_.
 
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->
+I will also export a map of the *homogeneous areas*, over which I will aggregate the indicator values.
+This map might be useful in later indicator work.
+
+For the finest resolution indicator map I will not include columns for
+uncertainty (because this will only be estimated at the aggregated
+level), or reference levels (because there is no spatial variation).
+
+### Homogeneous areas
+Export a map of the homogeneous areas that we aggregate the indicator
+over:
+
+```{r}
+stars::write_stars(infra_trd_reclassed,
+                   "data/homogeneous_areas/infrastructureIndex_fourClasses.tif")
+```
+
+### Sliatsje in wetlands
+Prepare export
+```{r}
+exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
+              origin,
+              area,
+              i_2020,
+              SHAPE
+              )
+```
+
+Write to file
+```{r, eval=F}
+st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
+         append = F)
+```
+
+### Sliatsje in wetlands, extrapolated to homegeneous impact areas.

### 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
feat: Start homogeneous area draft

---
title: "-"
output:
  html_document
---

# Homogeneous Impact Areas

<br />

_Author and date:_
Anders Kolstad
```{r, echo=FALSE}
Sys.Date()
```

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(tmap)
library(dplyr)
#library(osmplotr)
```

## Introduction
This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
The map is produces by binning values in the infrastructure index into four discrete categories. 
It is likely to be a good predictor for some indicators, such as  *slitasje* and the presence of
*alien species*.

## About the underlying data
The infrastructure index is explained [here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). 
It is a wall-to-wall raster over Norway with 100 m resolution. Each pixel is given a value along a continuous gradient 
from 0 to around 15.4, representing the frequency or surrounding cells within 500 m with human infrastructure (houses, roads, ect).

### Representativity in time and space
The infrastructure index is calculated with data that have slightly different dates, but can be said to represent the current status, the period 2010-2020.
The data covers all of mainland Norway.

## References
Bakkestuen, V., Dervo, B.K., Bærum, K.M. og Erikstad, L. 2022. Prediksjonsmodellering av naturtyper i ferskvann. NINA Rapport 2079. Norsk institutt for naturforskning.

## Analyses
### Import data

The path must be conditional
```{r}
dir <- substr(getwd(), 1,2)
```

```{r}
path <- ifelse(dir == "C:", 
        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif",
        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original/Infrastrukturindeks_UTM33/infra_tiff.tif")
```

Import a stars proxy (no data imported yet)
```{r}
infra <- stars::read_stars(path)
```

```{r}
st_crs(infra)$input
```

### Trondheim example
It's easier to see what's happening if we zoom in a bit. Lets get a
boundary box around Trondheim.
```{r}
myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
                ymin = 7032142.5, ymax = 7045245.27),
                crs = st_crs(infra))
```

Cropping the raster to the bbox
```{r}
infra_trd <- sf::st_crop(infra, myBB)
```

Transform to lat long due to osm requirements
```{r, warning=F}
infra_trd_ll <- sf::st_transform(infra_trd, 4326)
```

Get the boundary box of the cropped raster
```{r}
myBB_ll <- sf::st_bbox(infra_trd_ll)
```

### Get OSM highways
Lets download some base maps to help validate and contextualize the infrastructure index.

Download highways using the above bbox.
```{r, warning=FALSE, message=F, eval=F}
hw <- 
  osmplotr::extract_osm_objects(
    bbox = myBB_ll,
    key = "highway",
    sf = T)
```

Transforming highway data back into utm, although not strictly
necessary.
```{r, eval=F}
hw_utm <- sf::st_transform(hw, sf::st_crs(infra_trd)) 
```

This object contains too many roads (about 30k). I'll take out the unnamed roads.
```{r, eval=F}
hw_utm <- hw_utm[!is.na(hw_utm$name),]
```

```{r, eval=F, echo=F}
# Caching the sf object
saveRDS(hw_utm, "data/cache/highways_trondheim.rds")
```

```{r, echo=F}
#Load object
hw_utm <- readRDS("data/cache/highways_trondheim.rds")
```

### Discretize
Here I define a simplified categorical typology for the infrastructure index using four classes.

```{r}
names(infra_trd) <- "infrastructureIndex"
infra_trd_reclassed <-  infra_trd %>%
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 9 ~ 2,
    infrastructureIndex >= 9 ~ 3
  ))
```

Lets plot these two maps side by side
```{r}
map_trd_reclassed <- tm_shape(infra_trd_reclassed)+
  tm_raster(title="Infrastructure Index",
            #palette = "viridis",
            style="cat")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
```

```{r}
map_trd <- tm_shape(infra_trd)+
  tm_raster(title="Infrastructure Index",
            style="cont",
            palette = "viridis")+
  tm_layout(legend.outside = T)+
  tm_shape(hw_utm)+
  tm_lines(col="white")
```

```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
tmap_arrange(map_trd,
             map_trd_reclassed,
             ncol=1)
```

I tweaked the thresholds for the bins so that most of the forest next to Trondheim (to the left in the map) was in the second lowest class. 
This looks pretty good to me. It depicts a gradient in human presence from high within the build zone, to _no to very low_ in the forests and mountains 
to the west. Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.

## Discretize the entire map

##

## Prepare export
<!-- Text here -->

### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
doc: Why aggregate

@@
 ## Discretize the entire map
+```{r}
+names(infra) <- "infrastructureIndex"
 
-##
+infra_discrete <- infra %>%
+  mutate(infrastructureIndex = case_when(
+    infrastructureIndex < 1 ~ 0,
+    infrastructureIndex < 6 ~ 1,
+    infrastructureIndex < 9 ~ 2,
+    infrastructureIndex >= 9 ~ 3
+  ))
+```
+
+## Aggregate
+The resolution in this map is more than we need, and the size of the data implies that the next step, the vectorization, would take too long.
+We therefore aggregate, ar reduce the resolution.
 

### 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
feat: Vectorize infra

--- 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
@@
 library(dplyr)
+library(stars)
 #library(osmplotr)
@@
 
+### Aggregate
+The resolution in this map is more than we need, and the size of the data implies that the next step, the vectorization, would take too long.
+We therefore aggregate, ar reduce the resolution.
+
+```{r}
+dim <- st_dimensions(infra)
+paste("Resolution is", dim$x$delta, "by", dim$x$delta, "meters")
+```
+
+```{r}
+infra_trd_reclassed_agg <- st_warp(infra_trd_reclassed, cellsize = c(1000, 1000), 
+                                   crs = st_crs(infra_trd_reclassed), 
+                                   use_gdal = TRUE,
+                                   method = "average")
+```
+
+```{r}
+dim <- st_dimensions(infra_trd_reclassed_agg)
+paste("Resolution is", dim$x$delta, "by", dim$x$delta, "meters")
+```
+
+Each cell is no the average of the aggregated cells, and hence the value is continous again. Let's make it discrete.
+```{r}
+names(infra_trd_reclassed_agg) <- "infrastructureIndex"
+infra_trd_reclassed_agg <-  infra_trd_reclassed_agg %>%
+  mutate(infrastructureIndex = case_when(
+    infrastructureIndex < 1 ~ 0,
+    infrastructureIndex < 6 ~ 1,
+    infrastructureIndex < 9 ~ 2,
+    infrastructureIndex >= 9 ~ 3
+  ))
+```
+
+```{r, fig.cap="Infrastrcture indeks over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
+tm_shape(infra_trd_reclassed_agg)+
+  tm_raster(title="Infrastructure Index",
+            style="cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="white")
+```
+
+This resolution is more that good enough for our purpose here.
+
+## Aggregate the entire map to 1x1 km
+```{r}
+# runtime about 30 sec
+infra_agg <- st_warp(infra, cellsize = c(1000, 1000), 
+                                   crs = st_crs(infra), 
+                                   use_gdal = TRUE,
+                                   method = "average")
+```
+
 ## Discretize the entire map
+```{r}
+names(infra_agg) <- "infrastructureIndex"
 
-##
+infra_agg_discrete <- infra_agg %>%
+  mutate(infrastructureIndex = case_when(
+    infrastructureIndex < 1 ~ 0,
+    infrastructureIndex < 6 ~ 1,
+    infrastructureIndex < 9 ~ 2,
+    infrastructureIndex >= 9 ~ 3
+  ))
+```
+
+## Vectorize
+This step might seem rather stupid. We want to vectorize a rather large raster. This makes it a quite big data object.
+The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the
+`{raster}` package. It was not straight forward then either. But calculating intersections between polygons is very fast and easy.
+
+```{r}
+infra_agg_discrete_vect <- eaTools::ea_homogeneous_area(infra_agg_discrete)
+```
+
+Let's plot this now.
+First. lets crop it to reduce computation time.
+```{r}
+infra_agg_discrete_vect_trd <- st_crop(infra_agg_discrete_vect, myBB)
+```
+
+```{r}
+tmap_mode("view")
+tm_shape(infra_agg_discrete_vect_trd)+
+  tm_polygons(col="infrastructureIndex")
+```
 

### 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
feat. Export rds to P server

@@
   html_document
+editor_options: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-_Author and date:_
-Anders Kolstad
+*Author and date:* Anders Kolstad
+
 ```{r, echo=FALSE}
@@
 ## Introduction
+
 This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
 The map is produces by binning values in the infrastructure index into four discrete categories.
-It is likely to be a good predictor for some indicators, such as  *slitasje* and the presence of
-*alien species*.
+It is likely to be a good predictor for some indicators, such as *slitasje* and the presence of *alien species*.
 
 ## About the underlying data
+
 The infrastructure index is explained [here](https://brage.nina.no/nina-xmlui/handle/11250/2983607).
-It is a wall-to-wall raster over Norway with 100 m resolution. Each pixel is given a value along a continuous gradient 
-from 0 to around 15.4, representing the frequency or surrounding cells within 500 m with human infrastructure (houses, roads, ect).
+It is a wall-to-wall raster over Norway with 100 m resolution.
+Each pixel is given a value along a continuous gradient from 0 to around 15.4, representing the frequency or surrounding cells within 500 m with human infrastructure (houses, roads, ect).
 
 ### Representativity in time and space
+
 The infrastructure index is calculated with data that have slightly different dates, but can be said to represent the current status, the period 2010-2020.
@@
 ## References
-Bakkestuen, V., Dervo, B.K., Bærum, K.M. og Erikstad, L. 2022. Prediksjonsmodellering av naturtyper i ferskvann. NINA Rapport 2079. Norsk institutt for naturforskning.
+
+Bakkestuen, V., Dervo, B.K., Bærum, K.M.
+og Erikstad, L.
+2022.
+Prediksjonsmodellering av naturtyper i ferskvann.
+NINA Rapport 2079.
+Norsk institutt for naturforskning.
 
 ## Analyses
+
 ### Import data
@@
 The path must be conditional
+
 ```{r}
@@
 Import a stars proxy (no data imported yet)
+
 ```{r}
@@
 ### Trondheim example
-It's easier to see what's happening if we zoom in a bit. Lets get a
-boundary box around Trondheim.
+
+It's easier to see what's happening if we zoom in a bit.
+Lets get a boundary box around Trondheim.
+
 ```{r}
@@
 Cropping the raster to the bbox
+
 ```{r}
@@
 Transform to lat long due to osm requirements
+
 ```{r, warning=F}
@@
 ### Get OSM highways
+
 Lets download some base maps to help validate and contextualize the infrastructure index.
@@
 
-Transforming highway data back into utm, although not strictly
-necessary.
+Transforming highway data back into utm, although not strictly necessary.
+
 ```{r, eval=F}
@@
 
-This object contains too many roads (about 30k). I'll take out the unnamed roads.
+This object contains too many roads (about 30k).
+I'll take out the unnamed roads.
+
 ```{r, eval=F}
@@
 ### Discretize
+
 Here I define a simplified categorical typology for the infrastructure index using four classes.
@@
 Lets plot these two maps side by side
+
 ```{r}
@@
 I tweaked the thresholds for the bins so that most of the forest next to Trondheim (to the left in the map) was in the second lowest class.
-This looks pretty good to me. It depicts a gradient in human presence from high within the build zone, to _no to very low_ in the forests and mountains 
-to the west. Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.
+This looks pretty good to me.
+It depicts a gradient in human presence from high within the build zone, to *no to very low* in the forests and mountains to the west.
+Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.
 
 ### Aggregate
+
 The resolution in this map is more than we need, and the size of the data implies that the next step, the vectorization, would take too long.
@@
 
-Each cell is no the average of the aggregated cells, and hence the value is continous again. Let's make it discrete.
+Each cell is no the average of the aggregated cells, and hence the value is continous again.
+Let's make it discrete.
+
 ```{r}
@@
 ## Aggregate the entire map to 1x1 km
+
 ```{r}
@@
 ## Discretize the entire map
+
 ```{r}
@@
 ## Vectorize
-This step might seem rather stupid. We want to vectorize a rather large raster. This makes it a quite big data object.
-The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the
-`{raster}` package. It was not straight forward then either. But calculating intersections between polygons is very fast and easy.
+
+This step might seem rather stupid.
+We want to vectorize a rather large raster.
+This makes it a quite big data object.
+The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the `{raster}` package.
+It was not straight forward then either.
+But calculating intersections between polygons is very fast and easy.
 
@@
 Let's plot this now.
-First. lets crop it to reduce computation time.
+First.
+lets crop it to reduce computation time.
+
 ```{r}
@@
 
-```{r}
-tmap_mode("view")
+```{r test-2, fig.cap="A vectorized version of the Infrastructure indeks over Trondheim on a four-step discrete scale."}
 tm_shape(infra_agg_discrete_vect_trd)+
-  tm_polygons(col="infrastructureIndex")
+  tm_polygons(col="infrastructureIndex",
+              style = "cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="white")
 ```
 
-## Prepare export
-<!-- Text here -->
+As can be seen in the figure above, `st_warp` merges grid cells atht have the same value, and the vectorized raster doen's not end up being that big in the end.
 
-### Eksport file (final product)
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->
+## Export
+I will write this data to the NINA P server.
+
+```{r, eval=F}
+saveRDS(infra_agg_discrete_vect_trd, "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
+```

### 2023-01-17 DRAFT_homogeneous_impact_areas.Rmd
doc: Fix eval=F and similar

@@
 
-*Author and date:* Anders Kolstad
+*Author and date:*
+
+Anders Kolstad
 
@@
 
-## References
+### References
 
-Bakkestuen, V., Dervo, B.K., Bærum, K.M.
-og Erikstad, L.
-2022.
+Bakkestuen, V., Dervo, B.K., Bærum, K.M. og Erikstad, L. 2022. 
 Prediksjonsmodellering av naturtyper i ferskvann.
@@
 
-The path must be conditional
+The path must be conditional:
 
@@
 
+Print the coordinate reference system:
 ```{r}
@@
 
+#### Get OSM highways
+
+Lets download some base maps to help validate and contextualize the infrastructure index.
+
 Transform to lat long due to osm requirements
 
-```{r, warning=F}
+```{r, warning=F, eval=F}
 infra_trd_ll <- sf::st_transform(infra_trd, 4326)
@@
 
-```{r}
+```{r, eval=F}
 myBB_ll <- sf::st_bbox(infra_trd_ll)
 ```
-
-### Get OSM highways
-
-Lets download some base maps to help validate and contextualize the infrastructure index.
 
@@
 
-### Discretize
+#### Discretize
 
@@
 
-```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
+```{r, fig.cap="Infrastructure indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
 tmap_arrange(map_trd,
@@
 This looks pretty good to me.
-It depicts a gradient in human presence from high within the build zone, to *no to very low* in the forests and mountains to the west.
+It depicts a gradient in human presence from high within the buildt zone, to *no to very low* in the forests and mountains to the west.
 Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.
 
-### Aggregate
+#### Aggregate
 
@@
 
-```{r}
+```{r, warning=F, message=F}
 infra_trd_reclassed_agg <- st_warp(infra_trd_reclassed, cellsize = c(1000, 1000), 
@@
 
-Each cell is no the average of the aggregated cells, and hence the value is continous again.
+Each cell is now the average of the aggregated cells, and hence the value is continuous again.
 Let's make it discrete.
@@
 
-## Aggregate the entire map to 1x1 km
+### Aggregate the entire map to 1x1 km
 
-```{r}
+```{r, eval=F}
 # runtime about 30 sec
@@
                                    method = "average")
+saveRDS(infra_agg, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
 ```
 
-## Discretize the entire map
+```{r, echo=F}
+infra_agg <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
+```
+
+### Discretize the entire map
 
@@
 
-## Vectorize
+### Vectorize
 
@@
 
-```{r}
+```{r, warning=F}
 infra_agg_discrete_vect <- eaTools::ea_homogeneous_area(infra_agg_discrete)
@@
 Let's plot this now.
-First.
-lets crop it to reduce computation time.
+First, lets crop it to reduce computation time.
 
@@
 ```{r, eval=F}
-saveRDS(infra_agg_discrete_vect_trd, "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
+saveRDS(infra_agg_discrete_vect, "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
 ```

### 2023-01-17 DRAFT_slitasjeOgKjørespor.Rmd
minor: switching branches

@@
 ### Homogeneous impact areas
-
-```{r, eval=F}
-eaTools::ea_homogeneous_area(infra2)
+I want to use the [homogeneous impact areas]() to define smaller regions into which I can extrapolate the indicator values.
+
+```{r}
+HIA <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
+```
+
+```{r}
+tmap_mode("view")
+tm_shape(HIA)+
+  tm_polygons()
 ```

### 2023-01-17 DRAFT_slitasjeOgKjørespor.Rmd
feat: Import HIA

--- 2023-01-16 DRAFT_slitasjeOgKjørespor.Rmd
@@
 
-#### Infrastructure index
-
-The infrastructure index is explained
-[here](https://brage.nina.no/nina-xmlui/handle/11250/2983607). I use it
-as a kind of pressure or human influence index. It is likely to be a
-good predictor for *slitasje*, and I think to use it to define local reference area,
-or homogeneous (impact) areas.
-
-```{r}
-# The original data is here, on the R server, but it is not always working, so I also include a copy on the P server.
-
-#path <- ifelse(dir == "C:", 
-#        "R:/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original",
-#        "/data/R/GeoSpatialData/Utility_governmentalServices/Norway_Infrastructure_Index/Original")
-#
-
-path <- ifelse(dir == "C:", 
-        "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks",
-        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks")
-
-```
-
-Import a stars proxy (no data imported yet)
-
-```{r}
-infra <- stars::read_stars(paste0(path, "/NY_INFRA_IND.tif"))
-```
-
-It's easier to see what's happening if we zoom in a bit. Lets get a
-boundary box around Trondheim.
-
-```{r}
-myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
-                ymin = 7032142.5, ymax = 7045245.27),
-                crs = st_crs(infra))
-```
-
-Cropping the raster to speed up processing.
-
-```{r}
-infra_trd <- sf::st_crop(infra, myBB)
-```
-
-Transform to lat long due to osm requirements
-
-```{r, warning=F}
-infra_trd_ll <- sf::st_transform(infra_trd, 4326)
-```
-
-Get the boundary box of the cropped raster
-
-```{r}
-myBB_ll <- sf::st_bbox(infra_trd_ll)
-```
-
-Download highways using the above bbox.
-
-```{r, warning=FALSE, message=F, eval=F}
-hw <- 
-  osmplotr::extract_osm_objects(
-    bbox = myBB_ll,
-    key = "highway",
-    sf = T)
-```
-
-Transforming highway data back into utm, although not strictly
-necessary.
-
-```{r, eval=F}
-hw_utm <- sf::st_transform(hw, sf::st_crs(infra_trd)) 
-```
-
-This object contains too many roads. I'll take out the unnamed roads.
-
-```{r, eval=F}
-hw_utm <- hw_utm[!is.na(hw_utm$name),]
-```
-
-Caching the sf object
-
-```{r, eval=F}
-saveRDS(hw_utm, "data/cache/highways_trondheim.rds")
-```
-
-Load object
-
-```{r}
-hw_utm <- readRDS("data/cache/highways_trondheim.rds")
-```
-
-```{r, fig.cap="Infrastrcture indeks over Trondheim. Major roads are in white."}
-map_trd <- tm_shape(infra_trd)+
-  tm_raster(title="Infrastructure Index",
-            style="cont",
-            palette = "viridis")+
-  tm_layout(legend.outside = T)+
-  tm_shape(hw_utm)+
-  tm_lines(col="white")
-map_trd
-```
-
-The infrastructure index is in EPSG32633. I will need to transform it.
-It takes a couple of minutes, so I will store the result in P and load
-it back in afterwards.
-
-```{r, eval=F}
-path <- ifelse(dir == "C:", 
-        "P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks",
-        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks")
-
-infra <- sf::st_transform(infra, 25833)
-stars::write_stars(infra, paste0(path, "/infrastrukturindeksEPSG_EPSG35833.tif"))
-```
-
-```{r}
-infra2 <- stars::read_stars(paste0(path, "/infrastrukturindeksEPSG_EPSG35833.tif"))
-```
-
-Here I define a simplified categorical typology for the infrastructure
-index using four classes.
-
-```{r}
-names(infra_trd) <- "infrastructureIndex"
-infra_trd_reclassed <-  infra_trd %>%
-  mutate(infrastructureIndex = case_when(
-    infrastructureIndex < 1 ~ 0,
-    infrastructureIndex < 6 ~ 1,
-    infrastructureIndex < 9 ~ 2,
-    infrastructureIndex >= 9 ~ 3
-  ))
-```
-
-```{r}
-map_trd_reclassed <- tm_shape(infra_trd_reclassed)+
-  tm_raster(title="Infrastructure Index",
-            #palette = "viridis",
-            style="cat")+
-  tm_layout(legend.outside = T)+
-  tm_shape(hw_utm)+
-  tm_lines(col="white")
-
-```
-
-```{r, fig.cap="Infrastrcture indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
-tmap_arrange(map_trd,
-             map_trd_reclassed,
-             ncol=1)
-```
-
-I tweaked the thresholds so that most of the forest next to Trondheim
-was in the second lowest class. This looks pretty good to me.
-
 ### Scaled indicator values
@@
 ### Homogeneous impact areas
+I want to use the [homogeneous impact areas]() (HIA) to define smaller regions into which I can extrapolate the indicator values.
+This data is generated by discretizing the Norwegian Infrastructure Index. I refer to the ordinal values of the
+four impact area classes as their *Human Impact Factor*.
 
-```{r, eval=F}
-eaTools::ea_homogeneous_area(infra2)
+```{r}
+HIA <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
+```
+
+```{r, fig.cap="Counting the number of gridd cells for each Human Impact Factor."}
+barplot(table(HIA$infrastructureIndex),
+        ylab = "Number of 1x1 km grid cells",
+        xlab = "Human impact factor")
+```
+
+I also want to split these four HIA classes based on their region.
+
+<!-- Add function to ea_homogeneous_area() to find intersection with a decond delinimator -->
+
+### Aggregate and spread (extrapolate)
+I now want to find the mean indicator value for each HIA (i.e. to aggregate) and to spread these out
+spatialy to the entire HIA within each region (i.e. to extrapolate).
+
+```{r}
+eaTools::ea_spread()
 ```
@@
 
-### Homogeneous areas
-Export a map of the homogeneous areas that we aggregate the indicator
-over:
-
-```{r}
-stars::write_stars(infra_trd_reclassed,
-                   "data/homogeneous_areas/infrastructureIndex_fourClasses.tif")
-```
-
 ### Sliatsje in wetlands

### 2023-01-18 DRAFT_slitasjeOgKjørespor.Rmd
feat: HIF per region

@@
 ```{r}
-regions <- sf::read_sf("data/regions.shp")
+regions <- sf::read_sf("data/regions.shp", options = "ENCODING=UTF8")
+unique(regions$region)
 ```
@@
 This data is generated by discretizing the Norwegian Infrastructure Index. I refer to the ordinal values of the
-four impact area classes as their *Human Impact Factor*.
+four impact area classes as their *Human Impact Factor* (HIF).
 
@@
 
-I also want to split these four HIA classes based on their region.
+I also want to split these four HIA classes based on their region. To do this I need the two layers to have the same crs
+```{r}
+st_crs(HIA) == st_crs(regions)
+```
 
-<!-- Add function to ea_homogeneous_area() to find intersection with a decond delinimator -->
+```{r}
+HIA <- st_transform(HIA, 25833)
+```
+
+```{r}
+st_crs(HIA) == st_crs(regions)
+```
+
+Since the two layers are completely overlapping, we can get the intersections
+```{r, eval=F}
+HIA_regions <- eaTools::ea_homogeneous_area(HIA,
+                             regions,
+                             keep1 = "infrastructureIndex",
+                             keep2 = "region")
+saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+```
+
+```{r, echo=F}
+# silently read cached file
+HIA_regions <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+```
+
+I now have 20 unique areas that I will, given there is data, extrapolate indicator values over: 
+```{r}
+HIA_regions <- HIA_regions %>%
+  mutate(region_HIF = paste(region, infrastructureIndex))
+length(unique(HIA_regions$region_HIF))
+```
+
+Let's calculate the areas of these polygons and compare the HIF in the five regions. 
+```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
+HIA_regions$area <- sf::st_area(HIA_regions)
+
+temp <- as.data.frame(HIA_regions) %>%
+  group_by(region, infrastructureIndex) %>%
+  summarise(area = mean(area))
+
+ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
+  geom_bar(position = "stack", stat = "identity")
+```
+
+The figure above shows that Nord-Norge for example, has a lot of relatovely untouched areas, and that Østlandet has the highets proportion of impacted areas.
+This is expected.
 

### 2023-01-18 DRAFT_slitasjeOgKjørespor.Rmd
feat: Indicators per region and HIF class

@@
 ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
-  geom_bar(position = "stack", stat = "identity")
+  geom_bar(position = "stack", stat = "identity")+
+  guides(fill = guide_legend("HIF"))
 ```
 
-The figure above shows that Nord-Norge for example, has a lot of relatovely untouched areas, and that Østlandet has the highets proportion of impacted areas.
+The figure above shows that Nord-Norge for example, has a lot of relatively untouched areas, and that Østlandet has the highest proportion of impacted areas.
 This is expected. 
+<!-- It also shows that the second HIF class is perhaps to narrowly defined. This is suggested as a change in another PR: -->
 
@@
 
+```{r, eval=F}
+slitasje_data_extrapolated <- ea_spread(indicator_data = slitasje_data,
+                         indicator = i_2020,
+                         regions = HIA_regions,
+                         groups = region_HIF,
+                         threshold = 20)
+saveRDS(slitasje_data_extrapolated, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
+```
+
+```{r, echo=F}
+# silently read cached file
+slitasje_data_extrapolated <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
+```
+
 ```{r}
-eaTools::ea_spread()
+unique(slitasje_data_extrapolated$meanIndicatorValue)
+```
+<!-- slitasje_data_extrapolated contains 32402 rows, but I expected 5. -->
+
+<!-- Like this: -->
+```{r, eval=F}
+slitasje_data_extrapolated2 <- ea_spread(indicator_data = slitasje_data_extrapolated,
+                         indicator = meanIndicatorValue,
+                         regions = regions,
+                         groups = region,
+                         threshold = 20)
+saveRDS(slitasje_data_extrapolated2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
+```
+```{r, echo=F}
+# silently read cached file
+slitasje_data_extrapolated2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
+```
+
+```{r}
+plot(factor(slitasje_data_extrapolated2$ID), slitasje_data_extrapolated2$meanIndicatorValue,
+     ylab="indicator value", xlab="", ylim=c(0,1))
 ```

### 2023-01-19 DRAFT_homogeneous_impact_areas.Rmd
feat: change threshold between class 2-3

@@
     infrastructureIndex < 6 ~ 1,
-    infrastructureIndex < 9 ~ 2,
-    infrastructureIndex >= 9 ~ 3
+    infrastructureIndex < 12 ~ 2,
+    infrastructureIndex >= 12 ~ 3
   ))
@@
 
-I tweaked the thresholds for the bins so that most of the forest next to Trondheim (to the left in the map) was in the second lowest class.
+I tweaked the thresholds for the bins so that the categories match my knowledge about the land use intensity in Trondheim, for example that most of the forest next to Trondheim (to the left in the map) was in the second lowest class.
 This looks pretty good to me.
@@
     infrastructureIndex < 6 ~ 1,
-    infrastructureIndex < 9 ~ 2,
-    infrastructureIndex >= 9 ~ 3
+    infrastructureIndex < 12 ~ 2,
+    infrastructureIndex >= 1 ~ 3
   ))

### 2023-01-19 DRAFT_homogeneous_impact_areas.Rmd
feat: Change threshold between classes 2 and 3 and look at the distribuiton of the classes thoughout Norway.

@@
 library(stars)
+library(ggplot2)
+library(units)
+library(ggpubr)
+library(DT)
 #library(osmplotr)
@@
 
-As can be seen in the figure above, `st_warp` merges grid cells atht have the same value, and the vectorized raster doen's not end up being that big in the end.
+As can be seen in the figure above, `st_warp` merges grid cells that have the same value, and the vectorized raster doesn't not end up being that big in the end.
+
+## Chech national distribution
+
+```{r}
+regions <- sf::read_sf("data/regions.shp", options = "ENCODING=UTF8")
+unique(regions$region)
+```
+
+```{r}
+st_crs(infra_agg_discrete_vect) == st_crs(regions)
+```
+
+```{r}
+infra_agg_discrete_vect <- st_transform(infra_agg_discrete_vect, 25833)
+```
+
+```{r}
+st_crs(infra_agg_discrete_vect) == st_crs(regions)
+```
+
+Since the two layers are completely overlapping, we can get the intersections
+```{r, eval=F}
+infra_stats <- eaTools::ea_homogeneous_area(infra_agg_discrete_vect,
+                             regions,
+                             keep1 = "infrastructureIndex",
+                             keep2 = "region")
+saveRDS(infra_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_stats.rds")
+```
+
+```{r, echo=F}
+# silently read cached file
+infra_stats <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_stats.rds")
+```
+
+Let's calculate the areas of these polygons and compare the HIF in the five regions. 
+```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
+infra_stats$area <- sf::st_area(infra_stats)
+
+temp <- as.data.frame(infra_stats) %>%
+  group_by(region, infrastructureIndex) %>%
+  summarise(area = mean(area))
+
+ggarrange(
+  ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
+    geom_bar(position = "stack", stat = "identity")+
+    guides(fill = "none")+
+    coord_flip()+
+    xlab("")
+  ,
+  ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
+    geom_bar(position = "fill", stat = "identity")+
+    guides(fill = guide_legend("HIF"))+
+    coord_flip()+
+    xlab("")
+)
+```
+
+This distribution looks reasonable. The relative propotions are similar in the fivre regions, but _Østlandet_ and _Sørlandet_ have more infrastructure in general.
+
+Here are the numbers behind the figure above.
+```{r}
+temp$area <- round(temp$area,0)
+DT::datatable(temp)
+```
 

### 2023-01-19 DRAFT_slitasjeOgKjørespor.Rmd
doc: Some text, switching branches

@@
 This normalisation seems reasonable to me. I can save it as indicator
-values. I will assign the indicator value to the average year of the
+values. There is no point yet making this a time series, and I will 
+assign all the indicator value to the average year of the
 data.
@@
 ### Homogeneous impact areas
-I want to use the [homogeneous impact areas]() (HIA) to define smaller regions into which I can extrapolate the indicator values.
+I want to use the [Homogeneous Impact Areas] (HIA) to define smaller regions into which I can extrapolate the indicator values.
 This data is generated by discretizing the Norwegian Infrastructure Index. I refer to the ordinal values of the
@@
 HIA <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
-```
-
-```{r, fig.cap="Counting the number of gridd cells for each Human Impact Factor."}
-barplot(table(HIA$infrastructureIndex),
-        ylab = "Number of 1x1 km grid cells",
-        xlab = "Human impact factor")
 ```

### 2023-01-19 DRAFT_homogeneous_impact_areas.Rmd
bug: typo

@@
     infrastructureIndex < 12 ~ 2,
-    infrastructureIndex >= 1 ~ 3
+    infrastructureIndex >= 12 ~ 3
   ))
@@
 ```
+
+```{r, eval=F}
+HIA_returned <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
+```

### 2023-01-19 DRAFT_slitasjeOgKjørespor.Rmd
feat: Check correlation between HIF and the indicator

@@
 library(tidyverse)
+library(tidyr)
 knitr::opts_chunk$set(echo = TRUE)
@@
 This is expected.
-<!-- It also shows that the second HIF class is perhaps to narrowly defined. This is suggested as a change in another PR: -->
 
@@
                          threshold = 20)
-saveRDS(slitasje_data_extrapolated, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
+#saveRDS(slitasje_data_extrapolated, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
 ```
 
-```{r, echo=F}
-# silently read cached file
-slitasje_data_extrapolated <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
+Let's split the ID column to get the HIF
+```{r}
+slitasje_data_extrapolated <- slitasje_data_extrapolated %>%
+  tidyr::separate(ID, into = c("region", "HIF"),
+                  sep = " ")
+slitasje_data_extrapolated$HIF <- as.numeric(slitasje_data_extrapolated$HIF)
+```
+
+Now let's check the correlation between the HIF and the indicator values. 
+```{rslitasje-box, fig.cap="The relationship between human impact factor (a categorical version of the Norwegian Infrastructure Index) and the indicator avleus for slitasje."}
+ggplot(slitasje_data_extrapolated, aes(x = factor(HIF), y = meanIndicatorValue))+
+  geom_boxplot(fill = "grey",
+               size=1.5)+
+  theme_bw(base_size = 12)+
+  labs(x = "Human Impact Factor", y = "Slitasje indicator value")
 ```
@@
 ```{r}
-unique(slitasje_data_extrapolated$meanIndicatorValue)
+cor(slitasje_data_extrapolated$HIF, slitasje_data_extrapolated$meanIndicatorValue)
 ```
-<!-- slitasje_data_extrapolated contains 32402 rows, but I expected 5. -->
 
-<!-- Like this: -->
+This relationship is quite good, and I think it justified using HIF as homogeneous areas to extrapolate indicator values onto.
+
+Split one more time, into regions:
 ```{r, eval=F}
@@
 ```
+
 ```{r, echo=F}

### 2023-01-19 DRAFT_slitasjeOgKjørespor.Rmd
feat: bbox for Trondheim

@@
 Let's calculate the areas of these polygons and compare the HIF in the five regions. 
+First we need to remove some marine areas.
+```{r}
+HIA_regions2 <- HIA_regions %>%
+  st_intersection(outline)
+```
+
 ```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
@@
 
+It's easier to see what's happening if we zoom in a bit.
+Lets get a boundary box around Trondheim.
+
+```{r}
+myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
+                ymin = 7032142.5, ymax = 7045245.27),
+                crs = st_crs(slitasje_data_extrapolated))
+```
+
+Cropping the raster to the bbox
+
+```{r}
+slitasje_data_extrapolated_trd <- sf::st_crop(slitasje_data_extrapolated, myBB)
+```
+
+Get map of major roads, for context
+```{r}
+hw_utm <- readRDS("data/cache/highways_trondheim.rds")
+```
+
+```{r}
+tm_shape(slitasje_data_extrapolated_trd)+
+  tm_polygons(col = "meanIndicatorValue",
+    title="Slitasje indicator",
+    palette = "viridis",
+    style="cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="blue")
+```
+
 Let's split the ID column to get the HIF
@@
 Now let's check the correlation between the HIF and the indicator values.
-```{rslitasje-box, fig.cap="The relationship between human impact factor (a categorical version of the Norwegian Infrastructure Index) and the indicator avleus for slitasje."}
+<!-- This is not correct. This count the polygins that have very different sizes -->
+
+```{r slitasje-box, fig.cap="The relationship between human impact factor (a categorical version of the Norwegian Infrastructure Index) and the indicator avleus for slitasje."}
 ggplot(slitasje_data_extrapolated, aes(x = factor(HIF), y = meanIndicatorValue))+

### 2023-01-20 DRAFT_homogeneous_impact_areas.Rmd
feat: Cut out marine areas

@@
 This resolution is more that good enough for our purpose here.
+I see that the map extends a bit into the fjord. We want to cut away these areas. We can use a map of the outline of Norway to do this.
+
+```{r}
+outline <- st_read("data/outlineOfNorway_EPSG25833.shp")
+```
+
+The CRS needs to be identical 
+```{r}
+infra_trd_reclassed_agg <- st_transform(infra_trd_reclassed_agg, 25833)
+```
+
+```{r}
+infra_trd_reclassed_agg_terrestrial <- st_crop(infra_trd_reclassed_agg, outline)
+```
+
+```{r, fig.cap="Infrastrcture indeks over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
+tm_shape(infra_trd_reclassed_agg_terrestrial)+
+  tm_raster(title="Infrastructure Index",
+            style="cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="white")+
+  tm_shape(outline)+
+  tm_borders(lwd =2,
+             col = "black")
+```
+
+I have treated the raster cells as points when cropping, so that cells where the center or the cell is outside he terrestrial delineation are removed.
+I think this makes the most sense for unbiased area statistics, but some coastal indicator data could also be excluded because of this.
+The errors would be smaller with a finer resolution raster, but computation time is a problem.
 
@@
 
+### Cut out marine areas
+I see that the map extends a bit into the fjord. We want to cut away these areas. We can use a map of the outline of Norway to do this.
+
+The CRS needs to be identical 
+```{r}
+infra_agg <- st_transform(infra_agg, 25833)
+```
+
+```{r}
+infra_agg_terrestrial <- st_crop(infra_agg, outline)
+```
+
 ### Discretize the entire map
@@
 ```{r}
-names(infra_agg) <- "infrastructureIndex"
+names(infra_agg_terrestrial) <- "infrastructureIndex"
 
-infra_agg_discrete <- infra_agg %>%
+infra_agg_discrete <- infra_agg_terrestrial %>%
   mutate(infrastructureIndex = case_when(
@@
 
+<!-- Continue from here after fixing ea_spread -->
 ```{r, warning=F}
@@
 
-This distribution looks reasonable. The relative propotions are similar in the fivre regions, but _Østlandet_ and _Sørlandet_ have more infrastructure in general.
+This distribution looks reasonable. The relative proportions are similar in the five regions, but _Østlandet_ and _Sørlandet_ have more infrastructure in general.
 

### 2023-01-20 DRAFT_homogeneous_impact_areas.Rmd
feat: new bbox

@@
 
-<!-- Continue from here after fixing ea_spread -->
+<!-- Continue from here after fixing  ea_homogeneous_area -->
 ```{r, warning=F}
-infra_agg_discrete_vect <- eaTools::ea_homogeneous_area(infra_agg_discrete)
+infra_agg_discrete_vect <- eaTools::ea_homogeneous_area(infra_agg_discrete, groups = infrastructureIndex)
 ```
@@
 ```{r}
+myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
+                ymin = 7032142.5, ymax = 7045245.27),
+                crs = st_crs(infra_agg_discrete_vect))
+
 infra_agg_discrete_vect_trd <- st_crop(infra_agg_discrete_vect, myBB)

### 2023-01-20 R/functionalPlantIndicators/data handling.R
download and unzipping ANO

@@ -1,3 +1,4 @@
+library(downloader)
 library(sf)
 library(tidyr)
 library(dplyr)
@@ -5,11 +6,19 @@ library(plyr)
 library(stringr)
 library(tidyverse)
 
-
+#### download from kartkatalogen to P-drive ####
 # url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
+download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", mode="w") 
+unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", 
+       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
 
-#### download data from P-drive ####
+st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
 
+# SOMETHING NOT WORKING YET
+
+
+
+#### upload data from P-drive ####
 # ANO
 st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
 ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",

### 2023-01-20 DRAFT_slitasjeOgKjørespor.Rmd
doc: Add list of steps

@@
 
-<!-- Text here -->
+1.  Import [Nature type data set](#NTM) set (incl. GRUK) and [ANO data
+    set](#ANO-import)
+
+2.  Identify the [relevant](#naturtype) nature types and [subset](#NTM)
+    the data
+
+3.  Covert [ANO points to polygons](#ANO-points-to-poly)
+
+4.  For each polygon, extract the NiN-variable that has the lowest value
+    (worst-rules rule), e.g. like Fig. \@ref(fig:slitasje-naturetype).
+
+5.  [Combine](#combine-nt-ano) data sets
+
+6.  [Scale](#scale-slitasje-ind) the `slitasje` variable
+
+7.  Define [homogeneous impact areas](#HIA) (HIA)
+
+8.  [Aggregate and spread](#spread-slitasje) indicator value across HIAs
+    and assessment regions
+
+9.  Crop based on ecosystem delineation
+
+10. Confirm relationship between infrastructure index and indicator
+    values to justify the extrapolation
+
+11. Calculate uncertainty?
+
+12. Export indicator maps and extrapolated maps
 
@@
 
-#### Nature type mapping
+#### Nature type mapping {#NTM}
 
@@
 
-```{r, fig.cap="Slitasje scores."}
+```{r slitasje-naturetype, fig.cap="Slitasje scores."}
 temp <- naturetypes_wide %>%
@@
 
-#### ANO
+#### ANO {#ANO-import}
 
@@
 
-##### Combine Naturtype data and ANO
+##### Combine Naturtype data and ANO {#combine-nt-ano}
 
@@
 
-###### Points to polygons
+###### Points to polygons {#ANO-points-to-poly}
+
 The ANO data is point data, and the nature type data is multipolygon.
@@
 
-### Scaled indicator values
+### Scaled indicator values {#scale-slitasje-ind}
 
@@
 values. There is no point yet making this a time series, and I will
-assign all the indicator value to the average year of the
-data.
+assign all the indicator value to the average year of the data.
 
@@
 
-### Homogeneous impact areas
-I want to use the [Homogeneous Impact Areas] (HIA) to define smaller regions into which I can extrapolate the indicator values.
-This data is generated by discretizing the Norwegian Infrastructure Index. I refer to the ordinal values of the
-four impact area classes as their *Human Impact Factor* (HIF).
+### Homogeneous impact areas {#HIA}
+
+I want to use the [Homogeneous Impact Areas](#HIA) (HIA) to define
+smaller regions into which I can extrapolate the indicator values. This
+data is generated by discretizing the Norwegian Infrastructure Index. I
+refer to the ordinal values of the four impact area classes as their
+*Human Impact Factor* (HIF).
 
@@
 
-I also want to split these four HIA classes based on their region. To do this I need the two layers to have the same crs
+Here is a view of the data zooming in on Trondheim
+
+```{r}
+myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
+                ymin = 7032142.5, ymax = 7045245.27),
+                crs = st_crs(HIA))
+```
+
+Cropping the raster to the bbox
+
+```{r}
+HIA <- sf::st_crop(HIA, myBB)
+```
+
+Get map of major roads, for context
+
+```{r}
+hw_utm <- readRDS("data/cache/highways_trondheim.rds")
+```
+
+<!-- ERROR  marine areas still here -->
+```{r}
+tm_shape(HIA)+
+  tm_polygons(col = "infrastructureIndex",
+    title="Infrastructure index\n(modified 4-step scale)",
+    palette = "viridis",
+    style="cat")+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="blue")
+```
+
+I also want to split these four HIA classes based on their region. To do
+this I need the two layers to have the same CRS.
+
 ```{r}
@@
 
-```{r}
-HIA <- st_transform(HIA, 25833)
-```
-
-```{r}
-st_crs(HIA) == st_crs(regions)
-```
-
-Since the two layers are completely overlapping, we can get the intersections
+Since the two layers are completely overlapping, we can get the
+intersections
+
 ```{r, eval=F}
@@
 
-I now have 20 unique areas that I will, given there is data, extrapolate indicator values over: 
+I now have 20 unique areas that I will, given there is data, extrapolate
+indicator values over:
+
 ```{r}
@@
 
-Let's calculate the areas of these polygons and compare the HIF in the five regions. 
-First we need to remove some marine areas.
+Let's calculate the areas of these polygons and compare the HIF in the
+five regions. First we need to remove some marine areas.
+
 ```{r}
@@
 
-The figure above shows that Nord-Norge for example, has a lot of relatively untouched areas, and that Østlandet has the highest proportion of impacted areas.
-This is expected.
-
-### Aggregate and spread (extrapolate)
-I now want to find the mean indicator value for each HIA (i.e. to aggregate) and to spread these out
-spatialy to the entire HIA within each region (i.e. to extrapolate).
+The figure above shows that Nord-Norge for example, has a lot of
+relatively untouched areas, and that Østlandet has the highest
+proportion of impacted areas. This is expected.
+
+### Aggregate and spread (extrapolate) {#spread-slitasje}
+
+I now want to find the mean indicator value for each HIA (i.e. to
+aggregate) and to spread these out spatialy to the entire HIA within
+each region (i.e. to extrapolate).
 
@@
 
-It's easier to see what's happening if we zoom in a bit.
-Lets get a boundary box around Trondheim.
+It's easier to see what's happening if we zoom in a bit. Lets get a
+boundary box around Trondheim.
 
@@
 Get map of major roads, for context
+
 ```{r}
@@
 Let's split the ID column to get the HIF
+
 ```{r}
@@
 
-Now let's check the correlation between the HIF and the indicator values.
+Now let's check the correlation between the HIF and the indicator
+values.
 <!-- This is not correct. This count the polygins that have very different sizes -->
@@
 
-This relationship is quite good, and I think it justified using HIF as homogeneous areas to extrapolate indicator values onto.
+This relationship is quite good, and I think it justified using HIF as
+homogeneous areas to extrapolate indicator values onto.
 
 Split one more time, into regions:
+
 ```{r, eval=F}
@@
 
-But I also need to export a map of the indicator where we extrapolate the indicator values
-to the homogeneous areas. This map should be interpreted with caution, since they seem
-to provide explicit indicator values for areas where there is no data. This map is mainly for
-aggregation purposes. I will mark this map with the prefix _homogeneous-areas_.
-
-I will also export a map of the *homogeneous areas*, over which I will aggregate the indicator values.
-This map might be useful in later indicator work.
+But I also need to export a map of the indicator where we extrapolate
+the indicator values to the homogeneous areas. This map should be
+interpreted with caution, since they seem to provide explicit indicator
+values for areas where there is no data. This map is mainly for
+aggregation purposes. I will mark this map with the prefix
+*homogeneous-areas*.
+
+I will also export a map of the *homogeneous areas*, over which I will
+aggregate the indicator values. This map might be useful in later
+indicator work.
 
@@
 ### Sliatsje in wetlands
+
 Prepare export
+
 ```{r}
@@
 Write to file
+
 ```{r, eval=F}

### 2023-01-20 DRAFT_homogeneous_impact_areas.Rmd
bug: Cache large operation

@@
                                    method = "average")
-saveRDS(infra_agg, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
-```
-
-```{r, echo=F}
-infra_agg <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
+# The CRS needs to be identical for st_crop
+infra_agg <- st_transform(infra_agg, 25833)
 ```
@@
 
-The CRS needs to be identical 
-```{r}
-infra_agg <- st_transform(infra_agg, 25833)
+```{r, eval=F}
+infra_agg_terrestrial <- st_crop(infra_agg, outline)
+saveRDS(infra_agg_terrestrial, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg_terrestrial.rds")
 ```
 
-```{r}
-infra_agg_terrestrial <- st_crop(infra_agg, outline)
+```{r, include=F}
+# silently load cached file
+infra_agg_terrestrial <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
 ```
@@
 
-<!-- Continue from here after fixing  ea_homogeneous_area -->
 ```{r, warning=F}
@@
 unique(regions$region)
-```
-
-```{r}
-st_crs(infra_agg_discrete_vect) == st_crs(regions)
-```
-
-```{r}
-infra_agg_discrete_vect <- st_transform(infra_agg_discrete_vect, 25833)
 ```

### 2023-01-23 DRAFT_slitasjeOgKjørespor.Rmd
bug: Rearrange code blocks

@@
 
+I also want to split these four HIA classes based on their region. To do
+this I need the two layers to have the same CRS.
+
+```{r}
+st_crs(HIA) == st_crs(regions)
+```
+
+Since the two layers are completely overlapping, we can get the
+intersections
+
+```{r, eval=F}
+HIA_regions <- eaTools::ea_homogeneous_area(HIA,
+                             regions,
+                             keep1 = "infrastructureIndex",
+                             keep2 = "region")
+saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+```
+
+```{r, echo=F}
+# silently read cached file
+HIA_regions <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+```
+
+I now have 20 unique areas that I will, given there is data, extrapolate
+indicator values over:
+
+```{r}
+HIA_regions <- HIA_regions %>%
+  mutate(region_HIF = paste(region, infrastructureIndex))
+length(unique(HIA_regions$region_HIF))
+```
+
 Here is a view of the data zooming in on Trondheim
@@
                 ymin = 7032142.5, ymax = 7045245.27),
-                crs = st_crs(HIA))
+                crs = st_crs(HIA_regions))
 ```
@@
 ```{r}
-HIA <- sf::st_crop(HIA, myBB)
+HIA_trd <- sf::st_crop(HIA_regions, myBB)
 ```
@@
 
-<!-- ERROR  marine areas still here -->
-```{r}
-tm_shape(HIA)+
+```{r}
+tm_shape(HIA_trd)+
   tm_polygons(col = "infrastructureIndex",
@@
 
-I also want to split these four HIA classes based on their region. To do
-this I need the two layers to have the same CRS.
-
-```{r}
-st_crs(HIA) == st_crs(regions)
-```
-
-Since the two layers are completely overlapping, we can get the
-intersections
-
-```{r, eval=F}
-HIA_regions <- eaTools::ea_homogeneous_area(HIA,
-                             regions,
-                             keep1 = "infrastructureIndex",
-                             keep2 = "region")
-saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
-```
-
-```{r, echo=F}
-# silently read cached file
-HIA_regions <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
-```
-
-I now have 20 unique areas that I will, given there is data, extrapolate
-indicator values over:
-
-```{r}
-HIA_regions <- HIA_regions %>%
-  mutate(region_HIF = paste(region, infrastructureIndex))
-length(unique(HIA_regions$region_HIF))
-```
-
 Let's calculate the areas of these polygons and compare the HIF in the
-five regions. First we need to remove some marine areas.
-
-```{r}
-HIA_regions2 <- HIA_regions %>%
-  st_intersection(outline)
+five regions. 
+```{r}
+HIA_regions$area <- sf::st_area(HIA_regions)
 ```
@@
 ```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
-HIA_regions$area <- sf::st_area(HIA_regions)
-
 temp <- as.data.frame(HIA_regions) %>%
@@
 each region (i.e. to extrapolate).
+
+<!-- ERROR with summarise = T -->
+```{r}
+summary_temp <- ea_spread(indicator_data = slitasje_data,
+                         indicator = i_2020,
+                         regions = HIA_regions,
+                         groups = region_HIF,
+                         summarise = TRUE)
+```
 
@@
 slitasje_data_extrapolated_trd <- sf::st_crop(slitasje_data_extrapolated, myBB)
-```
-
-Get map of major roads, for context
-
-```{r}
-hw_utm <- readRDS("data/cache/highways_trondheim.rds")
 ```

### 2023-01-23 DRAFT_slitasjeOgKjørespor.Rmd
test: Test 955 with ea_spread_dev()

@@
 
-<!-- ERROR with summarise = T -->
-```{r}
-summary_temp <- ea_spread(indicator_data = slitasje_data,
+<!-- Check with updated ea_spread -->
+```{r}
+mySummary <- ea_spread(indicator_data = slitasje_data,
                          indicator = i_2020,
@@
                          summarise = TRUE)
+
 ```

### 2023-01-23 DRAFT_slitasjeOgKjørespor.Rmd
feat: Subset wetlands and validate

@@
 library(ggplot2)
+library(DT)
 library(units)
@@
 
+I want to check that HIF is in fact a good predictor for _slitasje_.
+
 I also want to split these four HIA classes based on their region. To do
@@
 
+```{r}
+HIA_regions <- HIA_regions %>%
+  mutate(region_HIF = paste(region, infrastructureIndex))
+```
+
+#### Validate
 I now have 20 unique areas that I will, given there is data, extrapolate
 indicator values over:
-
-```{r}
-HIA_regions <- HIA_regions %>%
-  mutate(region_HIF = paste(region, infrastructureIndex))
-length(unique(HIA_regions$region_HIF))
-```
-
+<!-- NEEDS CACHING -->
+
+```{r, eval=F}
+corrCheck <- st_intersection(slitasje_data, HIA_regions)
+
+corrCheck <- as.data.frame(corrCheck) %>%
+  group_by(hovedøkosystem, region) %>%
+  summarise(cor = cor(i_2020, infrastructureIndex))
+```
+
+```{r, fog.cap="Pearson correaltions between the indicator values at the polygons level and the HIF."}
+
+ggplot(corrCheck, aes(x = hovedøkosystem, 
+                      y = cor))+
+  geom_bar(stat = "identity")+
+  facet_wrap(.~region)+
+  coord_flip()
+```
+
+```{r}
+ggarrange(
+ggplot(corrCheck, aes(x = hovedøkosystem, 
+                      y = cor))+
+  geom_bar(stat = "identity")+
+  coord_flip(),
+ggplot(corrCheck, aes(x = region, 
+                      y = cor))+
+  geom_bar(stat = "identity")+
+  coord_flip(),
+ncol=1
+)
+```
+
+From the figures above it appears that the relationship is quite week in general. But the sheer number of data points suggest that all group differences will be very reliable.
+_Naturlig åpne områder_ has the lowest correlation between the indicator and the homogeneous area classes, and of the five regions, _Nord-Norge_ has the lowest correlation.
+
+#### Looking at the HIA
 Here is a view of the data zooming in on Trondheim
@@
 
-```{r}
+```{r, fig.cap="A closer look at the HIA designation over Trondheim"}
 tm_shape(HIA_trd)+
@@
 five regions.
+
 ```{r}
@@
 
-<!-- Check with updated ea_spread -->
-```{r}
-mySummary <- ea_spread(indicator_data = slitasje_data,
+I will focus on wetland from here on out. Only for open wetlands do we have good exosystem delineation maps to base a spatial averaging on.
+
+```{r}
+wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",]
+```
+
+I want to add a threshold so that we don't end up over extrapolating
+based on too few data points. I think we need a minimum of 100 data point to base an estimate on.
+
+<!-- NEEDS CACHING -->
+```{r}
+mySummary <- ea_spread(indicator_data = wetlands,
                          indicator = i_2020,
@@
                          groups = region_HIF,
-                         summarise = TRUE)
-
+                         summarise = TRUE,
+                       threshold = 100)
+```
+
+<!-- Do we need another aggregation step here, to get regional estimates= -->
+
+Let's split the regions column
+
+```{r}
+mySummary <- mySummary %>%
+  tidyr::separate(region_HIF, into = c("region", "HIF"),
+                  sep = " ")
+mySummary$HIF <- as.numeric(mySummary$HIF)
+```
+
+```{r, fig.cap="Indicator values (area weighted averages) for each region and homogeneous area class, as well as the number of data points per groups."}
+ggarrange(
+ggplot(mySummary, aes(x = HIF, y = area_weighted_mean_indicator_value,
+                      fill = area_weighted_mean_indicator_value))+
+  geom_bar(stat = "identity")+
+  theme_bw(base_size = 12)+
+  facet_wrap(.~region, ncol=1)+
+  guides(fill = "none"),
+ggplot(mySummary, aes(x = HIF, y = data_points,
+                      fill = region))+
+  geom_bar(stat = "identity")+
+  theme_bw(base_size = 12)+
+  facet_wrap(.~region, ncol=1)+
+  guides(fill = "none")
+)
+```
+
+The figure above shows that there are much more data on the two middle
+classes (class 1 and 2). We do not have enough data to extrapolate to the 
+highest impacted areas in any of the regions. These are not the areas prioritised for nature type mapping.
+
+We also see that, even though there are small differences, high HIF is assiciated with lower indicator values.
+
+```{r}
+DT::datatable(mySummary)
 ```
@@
 ```{r, eval=F}
-slitasje_data_extrapolated <- ea_spread(indicator_data = slitasje_data,
+wetlands_slitasje_extr <- ea_spread(indicator_data = wetlands,
                          indicator = i_2020,
@@
                          groups = region_HIF,
-                         threshold = 20)
-#saveRDS(slitasje_data_extrapolated, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated.rds")
+                         threshold = 100)
+#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
+```
+```{r, include=F}
+wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
 ```
@@
                 ymin = 7032142.5, ymax = 7045245.27),
-                crs = st_crs(slitasje_data_extrapolated))
+                crs = st_crs(wetlands_slitasje_extr))
 ```
@@
 ```{r}
-slitasje_data_extrapolated_trd <- sf::st_crop(slitasje_data_extrapolated, myBB)
-```
-
-```{r}
-tm_shape(slitasje_data_extrapolated_trd)+
-  tm_polygons(col = "meanIndicatorValue",
+wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
+```
+
+```{r, fig.cap="Sliatsje (wetland) indicator extrapolated over Trondheim"}
+tm_shape(wetlands_slitasje_extr_trd)+
+  tm_polygons(col = "area_weighted_mean_indicator_value",
     title="Slitasje indicator",
     palette = "viridis",
-    style="cat")+
+    style="cont",
+    breaks = c(0,1))+
   tm_layout(legend.outside = T)+
@@
 
-Let's split the ID column to get the HIF
-
-```{r}
-slitasje_data_extrapolated <- slitasje_data_extrapolated %>%
+Let's split the ID column to get the HIF and plot agains indicator values
+
+```{r}
+wetlands_slitasje_extr <- wetlands_slitasje_extr %>%
   tidyr::separate(ID, into = c("region", "HIF"),
                   sep = " ")
-slitasje_data_extrapolated$HIF <- as.numeric(slitasje_data_extrapolated$HIF)
-```
-
-Now let's check the correlation between the HIF and the indicator
-values.
-<!-- This is not correct. This count the polygins that have very different sizes -->
-
-```{r slitasje-box, fig.cap="The relationship between human impact factor (a categorical version of the Norwegian Infrastructure Index) and the indicator avleus for slitasje."}
-ggplot(slitasje_data_extrapolated, aes(x = factor(HIF), y = meanIndicatorValue))+
-  geom_boxplot(fill = "grey",
-               size=1.5)+
-  theme_bw(base_size = 12)+
-  labs(x = "Human Impact Factor", y = "Slitasje indicator value")
-```
-
-```{r}
-cor(slitasje_data_extrapolated$HIF, slitasje_data_extrapolated$meanIndicatorValue)
+wetlands_slitasje_extr$HIF <- as.numeric(wetlands_slitasje_extr$HIF)
+```
+
+```{r}
+summary(wetlands_slitasje_extr)
+ggplot(wetlands_slitasje_extr, aes(x = factor(HIF), y = area_weighted_mean_indicator_value))+
+         geom_boxplot(fill = "grey")+
+  theme_bw(base_size = 12)
 ```
@@
 ```{r, eval=F}
-slitasje_data_extrapolated2 <- ea_spread(indicator_data = slitasje_data_extrapolated,
-                         indicator = meanIndicatorValue,
+wetlands_slitasje_extr2 <- ea_spread_dev(indicator_data = wetlands_slitasje_extr,
+                         indicator = area_weighted_mean_indicator_value,
                          regions = regions,
                          groups = region,
-                         threshold = 20)
-saveRDS(slitasje_data_extrapolated2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
-```
-
-```{r, echo=F}
+                         summarise = F)
+saveRDS(wetlands_slitasje_extr2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
+```
+
+```{r, include=F}
 # silently read cached file
-slitasje_data_extrapolated2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/slitasje_data_extrapolated2.rds")
-```
-
-```{r}
-plot(factor(slitasje_data_extrapolated2$ID), slitasje_data_extrapolated2$meanIndicatorValue,
+wetlands_slitasje_extr2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
+```
+
+```{r}
+plot(factor(wetlands_slitasje_extr2$ID), wetlands_slitasje_extr2$area_weighted_mean_indicator_value,
      ylab="indicator value", xlab="", ylim=c(0,1))

### 2023-01-25 DRAFT_slitasjeOgKjørespor.Rmd
feat: indicator-pressure validation

@@
 library(ggplot2)
+library(ggpubr)
 library(DT)
@@
 #### Validate
-I now have 20 unique areas that I will, given there is data, extrapolate
-indicator values over:
-<!-- NEEDS CACHING -->
+I now have 20 unique areas (for each main ecosystem) that I will, given there is data, extrapolate
+indicator values over
+
+###### Subset ETs
+I will subset the `sliatsje_data` in the three ecosystems. 
+Note that only for open wetlands do we have good ecosystem delineation maps to base a spatial averaging on.
+
+```{r}
+wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",]
+seminat <- slitasje_data[slitasje_data$hovedøkosystem == "Semi-naturlig mark",]
+natOpen <- slitasje_data[slitasje_data$hovedøkosystem == "Naturlig åpne områder under skoggrensa",]
+```
+
+```{r, eval=F}
+wetland_stats <- ea_spread(indicator_data = wetlands,
+          indicator = i_2020,
+          regions = HIA_regions,
+          groups = region_HIF,
+          summarise = TRUE)
+
+seminat_stats <- ea_spread(indicator_data = seminat,
+          indicator = i_2020,
+          regions = HIA_regions,
+          groups = region_HIF,
+          summarise = TRUE)
+
+natOpen_stats <- ea_spread(indicator_data = natOpen,
+          indicator = i_2020,
+          regions = HIA_regions,
+          groups = region_HIF,
+          summarise = TRUE)
+
+wetland_stats <- wetland_stats %>%
+  add_column(eco = "wetland")
+
+seminat_stats <- seminat_stats %>%
+  add_column(eco = "semi-natural")
+
+natOpen_stats <- natOpen_stats %>%
+  add_column(eco = "Naturally-open")
+
+all_stats <- rbind(wetland_stats,
+                   seminat_stats,
+                   natOpen_stats)
+all_stats <- all_stats %>%
+  separate(region_HIF,
+           into = c("region", "HIF"),
+           sep = " ")
+saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
+```
+
+```{r, include=F}
+#silently read cached file
+all_stats <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
+```
+
+```{r dense-stats, fig.height=20, fig.width=12}
+ggarrange(
+ggplot(all_stats, aes(x = region, y = n, group = HIF, fill = HIF))+
+  geom_bar(position = "dodge", stat = "identity")+
+  theme_bw(base_size = 30)+
+  coord_flip()+
+  facet_wrap(.~eco),
+ggplot(all_stats, aes(x = region, y = w_mean, group = HIF, fill = HIF))+
+  geom_bar(position = "dodge", stat = "identity")+
+  theme_bw(base_size = 30)+
+  coord_flip()+
+  facet_wrap(.~eco),
+ggplot(all_stats, aes(x = region, y = sd, group = HIF, fill = HIF))+
+  geom_bar(position = "dodge", stat = "identity")+
+  theme_bw(base_size = 30)+
+  coord_flip()+
+  facet_wrap(.~eco),
+ncol = 1)
+
+```
+
+This rather dense figure needs to be looked at carefully. 
+In the top row we see that the sample size (number of polygons) varies between ecosystems.
+For naturally open ecosystems many polygons are from high infrastructure areas, not surprisingly, and there are few points for the lowest impact class (zero such polygins in _Sørlandet_ for example).
+We there may not be able to extrapolate indicator values in areas in _Sørlandet_ with low levels of infrastructure.
+We can also from the middle row see the association between the indicator values (area weighted means) and the modified infrastructure index (HIF).
+We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row).
+For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to se (middle row).
+The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).
+
+In semi-natural areas and wetlands, we have very little data from HIF class 3, and the sd's are very big.
+The indicator-pressure relationship (middle row) looks the most messy for wetlands, but if ignoring HIF=3, the relationship looks much better.
+
+In addition to looking at the area weighted means and its relationship with the pressure indicator (HIF), we can do the same at the polygon level.
+Because the indicator is pseudo-continuous, boxplots for example become useless, so I will present a relative frequency plot.
 
@@
 corrCheck <- st_intersection(slitasje_data, HIA_regions)
-
-corrCheck <- as.data.frame(corrCheck) %>%
+saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
+```
+```{r, include=F}
+#Silently read cached data
+corrCheck <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
+```
+
+```{r slitasje-precentageplot, fig.height=12, fig.width=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(i_2020)))+
+  geom_bar(position="fill")+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Sliatsje indicator"))+
+  ylab("Fraction of data points")+
+  xlab("HIF")+
+  facet_grid(hovedøkosystem~region)
+```
+
+When down weighting the importance of those groups/categories that have very little The figure above I think supports the indicator-pressure relationship and justifies using the HIA+region intersections as local reference areas (i.e.
+homogeneous impact areas).
+
+We can also calculate Pearsons correlation coefficients.
+
+```{r}
+corrCheck_grouped <- as.data.frame(corrCheck) %>%
   group_by(hovedøkosystem, region) %>%
@@
 
-```{r, fog.cap="Pearson correaltions between the indicator values at the polygons level and the HIF."}
-
-ggplot(corrCheck, aes(x = hovedøkosystem, 
+```{r, fig.cap="Pearson correlations between the indicator values at the polygons level and the HIF."}
+ggplot(corrCheck_grouped, aes(x = hovedøkosystem, 
                       y = cor))+
@@
 
-```{r}
+```{r, fig.cap="Pearson correlations between indicator values and the HIF."}
+corrCheck_eco <- corrCheck_grouped %>%
+  group_by(hovedøkosystem) %>%
+  summarise(cor = mean(cor))
+corrCheck_reg <- corrCheck_grouped %>%
+  group_by(region) %>%
+  summarise(cor = mean(cor))
+
 ggarrange(
-ggplot(corrCheck, aes(x = hovedøkosystem, 
+ggplot(corrCheck_eco, aes(x = hovedøkosystem, 
                       y = cor))+
@@
   coord_flip(),
-ggplot(corrCheck, aes(x = region, 
+ggplot(corrCheck_reg, aes(x = region, 
                       y = cor))+
@@
 
-From the figures above it appears that the relationship is quite week in general. But the sheer number of data points suggest that all group differences will be very reliable.
+From the figures above it appears that the relationship is quite week in general. However, the sheer number of data points suggest that group differences will be very reliable.
 _Naturlig åpne områder_ has the lowest correlation between the indicator and the homogeneous area classes, and of the five regions, _Nord-Norge_ has the lowest correlation.
@@
 I now want to find the mean indicator value for each HIA (i.e. to
-aggregate) and to spread these out spatialy to the entire HIA within
+aggregate) and to spread these out spatially to the entire HIA within
 each region (i.e. to extrapolate).
-
-I will focus on wetland from here on out. Only for open wetlands do we have good exosystem delineation maps to base a spatial averaging on.
-
-```{r}
-wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",]
-```
 

### 2023-01-25 DRAFT_slitasjeOgKjørespor.Rmd
feat: spread to HIAs

@@
 We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row).
-For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to se (middle row).
+For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to see (middle row).
 The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).
@@
 The indicator-pressure relationship (middle row) looks the most messy for wetlands, but if ignoring HIF=3, the relationship looks much better.
+
+```{r, fig.cap="Summary statistics for the slitasje indicator."}
+DT::datatable(all_stats)
+```
 
@@
 
-We can also calculate Pearsons correlation coefficients.
+We can also calculate Kendall tau b coefficients, although I'm not complete sure this is valid when we have ordinal data with only four levels, since there will be a lot of ties.
 
@@
   group_by(hovedøkosystem, region) %>%
-  summarise(cor = cor(i_2020, infrastructureIndex))
+  summarise(cor = cor(i_2020, infrastructureIndex, method = "kendall"))
 ```
@@
 ```{r, fig.cap="A closer look at the HIA designation over Trondheim"}
-tm_shape(HIA_trd)+
+(HIA_trd <- tm_shape(HIA_trd)+
   tm_polygons(col = "infrastructureIndex",
     title="Infrastructure index\n(modified 4-step scale)",
-    palette = "viridis",
+    palette = "-viridis",
     style="cat")+
@@
   tm_shape(hw_utm)+
-  tm_lines(col="blue")
+  tm_lines(col="red")+
+  tm_shape(outline)+
+  tm_borders(col = "black", lwd=2))
 ```
@@
 I want to add a threshold so that we don't end up over extrapolating
-based on too few data points. I think we need a minimum of 100 data point to base an estimate on.
-
-<!-- NEEDS CACHING -->
-```{r}
-mySummary <- ea_spread(indicator_data = wetlands,
-                         indicator = i_2020,
-                         regions = HIA_regions,
-                         groups = region_HIF,
-                         summarise = TRUE,
-                       threshold = 100)
-```
-
-<!-- Do we need another aggregation step here, to get regional estimates= -->
-
-Let's split the regions column
-
-```{r}
-mySummary <- mySummary %>%
-  tidyr::separate(region_HIF, into = c("region", "HIF"),
-                  sep = " ")
-mySummary$HIF <- as.numeric(mySummary$HIF)
-```
-
-```{r, fig.cap="Indicator values (area weighted averages) for each region and homogeneous area class, as well as the number of data points per groups."}
-ggarrange(
-ggplot(mySummary, aes(x = HIF, y = area_weighted_mean_indicator_value,
-                      fill = area_weighted_mean_indicator_value))+
-  geom_bar(stat = "identity")+
-  theme_bw(base_size = 12)+
-  facet_wrap(.~region, ncol=1)+
-  guides(fill = "none"),
-ggplot(mySummary, aes(x = HIF, y = data_points,
-                      fill = region))+
-  geom_bar(stat = "identity")+
-  theme_bw(base_size = 12)+
-  facet_wrap(.~region, ncol=1)+
-  guides(fill = "none")
-)
-```
-
-The figure above shows that there are much more data on the two middle
-classes (class 1 and 2). We do not have enough data to extrapolate to the 
-highest impacted areas in any of the regions. These are not the areas prioritised for nature type mapping.
-
-We also see that, even though there are small differences, high HIF is assiciated with lower indicator values.
-
-```{r}
-DT::datatable(mySummary)
-```
+based on too few data points. From Fig. \@ref(fig:dense-stats) I conclude that we need at least 70 data points for the sd to stabilize. I will conservatively round this up to 100 data points.
 
@@
                          threshold = 100)
+
+seminat_slitasje_extr <- ea_spread(indicator_data = seminat,
+                         indicator = i_2020,
+                         regions = HIA_regions,
+                         groups = region_HIF,
+                         threshold = 100)
+
+natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
+                         indicator = i_2020,
+                         regions = HIA_regions,
+                         groups = region_HIF,
+                         threshold = 100)
 #saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
+#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
+#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
 ```
@@
 wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
+seminat_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
+natOpen_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
 ```
@@
 wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
-```
-
-```{r, fig.cap="Sliatsje (wetland) indicator extrapolated over Trondheim"}
+seminat_slitasje_extr_trd <- sf::st_crop(seminat_slitasje_extr, myBB)
+natOpen_slitasje_extr_trd <- sf::st_crop(natOpen_slitasje_extr, myBB)
+```
+
+```{r, fig.cap="Slitasje indicator extrapolated over Trondheim"}
+tmap_arrange(
 tm_shape(wetlands_slitasje_extr_trd)+
-  tm_polygons(col = "area_weighted_mean_indicator_value",
+  tm_polygons(col = "w_mean",
     title="Slitasje indicator",
@@
   tm_shape(hw_utm)+
-  tm_lines(col="blue")
-```
-
-Let's split the ID column to get the HIF and plot agains indicator values
-
-```{r}
-wetlands_slitasje_extr <- wetlands_slitasje_extr %>%
-  tidyr::separate(ID, into = c("region", "HIF"),
-                  sep = " ")
-wetlands_slitasje_extr$HIF <- as.numeric(wetlands_slitasje_extr$HIF)
-```
-
-```{r}
-summary(wetlands_slitasje_extr)
-ggplot(wetlands_slitasje_extr, aes(x = factor(HIF), y = area_weighted_mean_indicator_value))+
-         geom_boxplot(fill = "grey")+
-  theme_bw(base_size = 12)
-```
-
-This relationship is quite good, and I think it justified using HIF as
-homogeneous areas to extrapolate indicator values onto.
-
-Split one more time, into regions:
-
-```{r, eval=F}
-wetlands_slitasje_extr2 <- ea_spread_dev(indicator_data = wetlands_slitasje_extr,
-                         indicator = area_weighted_mean_indicator_value,
-                         regions = regions,
-                         groups = region,
-                         summarise = F)
-saveRDS(wetlands_slitasje_extr2, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
-```
-
-```{r, include=F}
-# silently read cached file
-wetlands_slitasje_extr2 <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr2.rds")
-```
-
-```{r}
-plot(factor(wetlands_slitasje_extr2$ID), wetlands_slitasje_extr2$area_weighted_mean_indicator_value,
-     ylab="indicator value", xlab="", ylim=c(0,1))
-```
+  tm_lines(col="blue")+
+  tm_layout(title = "Wetland"),
+
+tm_shape(seminat_slitasje_extr_trd)+
+  tm_polygons(col = "w_mean",
+    title="Slitasje indicator",
+    palette = "viridis",
+    style="cont",
+    breaks = c(0,1))+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="blue")+
+  tm_layout(title = "Semi-natural areas"),
+
+tm_shape(natOpen_slitasje_extr_trd)+
+  tm_polygons(col = "w_mean",
+    title="Slitasje indicator",
+    palette = "viridis",
+    style="cont",
+    breaks = c(0,1))+
+  tm_layout(legend.outside = T)+
+  tm_shape(hw_utm)+
+  tm_lines(col="blue")+
+  tm_layout(title = "Naturally open areas"),
+HIA_trd)
+```
+
+There is very little variation both between the ecosystems and the four classes in the modified infrastructure index.
+Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.
+
+The next steps now are to
+
+1. crop the extrapolated indicator maps to ecosystem delineation maps, and then;
+
+2. aggregate these for each region or accounting area.
 

### 2023-01-27 DRAFT_slitasjeOgKjørespor.Rmd
doc: Way forward

@@
 
+Alternatively I can rasterize the indicator map, and mask it using a perfectly aligned ET map. This might be faster.
+
+### Crop to ET
+This is a very heavy operation.
+```{r, eval=F}
+mireET <- sf::st_read("R:/Kladd/Myr/Myrmodell.gdb")
+```
+
+```{r, eval=F}
+wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
+```
+
+### Aggregate to region
+
+```{r, eval=F}
+wetlands_slitasje_extr_cropped_reg <- ea_spread(indicator_data = wetlands_slitasje_extr_cropped,
+                                      indicator = i_2020,
+                                      regions = regions,
+                                      groups = ?)
+```
+
 ### Uncertainty
 
-<!-- Text here -->
-
-Bootstrapping of nin-polygons per infrastructure class and region?
+The calculations of indicator uncertainty is done with bootstrapping, as is described in eaTools.
 

### 2023-01-30 DRAFT_slitasjeOgKjørespor.Rmd
feat: conditional paths; load ET map

@@
 *Author and date:* Anders L. Kolstad
-
-```{r}
-Sys.Date()
-```
 
@@
 knitr::opts_chunk$set(echo = TRUE)
+
+# Set up conditional file paths
+dir <- substr(getwd(), 1,2)
+
+pData <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+
 ```
@@
 
-1.  Import [Nature type data set](#NTM) set (incl. GRUK) and [ANO data
+1.  Import [Nature type data](#NTM) data set (incl. GRUK) and [ANO data
     set](#ANO-import)
@@
 
-6.  [Scale](#scale-slitasje-ind) the `slitasje` variable
+6.  [Scale](#scale-slitasje-ind) the `slitasje` variable based on reference values.
 
@@
 
-9.  Crop based on ecosystem delineation
-
-10. Confirm relationship between infrastructure index and indicator
+9. Confirm relationship between infrastructure index and indicator
     values to justify the extrapolation
 
-11. Calculate uncertainty?
-
-12. Export indicator maps and extrapolated maps
+10.  Sample 10 00 random points from an ecosystem delineation and extract indicator values and sd at those points.
+
+11. Export indicator maps and regional extrapolated maps
 
 ## About the underlying data
+The indicator uses a data set from a standardised field survey of nature types. You can read more about it [here](#naturtyper). I also import a data set called [ANO](#ANO-import)
+
+### Representativity in time and space
+The nature type mapping is not random and cannot be said to be area representative. 
+The data is from 2018 to present. The data from one field season usually becomes available early the following year.
+
+### Original units
+The
+
+### Temporal coverage
 
@@
 
-### Representativity in time and space
+### Aditional comments about the dataset
 
@@
 
-### Original units
+## Ecosystem characteristic
+
+### Norwegain standard
 
@@
 
-### Temporal coverage
-
-<!-- Text here -->
-
-### Aditional comments about the dataset
-
-<!-- Text here -->
-
-## Ecosystem characteristic
-
-### Norwegain standard
-
-<!-- Text here -->
-
-### <!-- Text here -->
+### UN standard
 
@@
 
-```{r}
-dir <- substr(getwd(), 1,2)
-```
-
 ```{r, eval=F}
-path <- ifelse(dir == "C:", 
-               "P:/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
-               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
-
-naturetypes <- sf::st_read(dsn = path)
+naturetypes <- sf::st_read(dsn = paste0(pData, "Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"))
 ```
@@
 ```{r}
-path <- ifelse(dir == "C:", 
-               "P:/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds",
-               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
-naturetypes <- readRDS(path)
+naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
 ```
@@
 
-```{r, fig.cap="7SE scores in the naturetype dataset"}
+```{r, fig.cap="7SE scores in the naturetype dataset", fig.width=3}
 temp <- naturetypes_wide %>%
@@
 
-```{r, fig.cap="7TK scores in the nature type dataset"}
+```{r, fig.cap="7TK scores in the nature type dataset", fig.width=3}
 temp <- naturetypes_wide %>%
@@
 
-```{r slitasje-naturetype, fig.cap="Slitasje scores."}
+```{r slitasje-naturetype, fig.cap="Slitasje scores.", fig.width=3}
 temp <- naturetypes_wide %>%
@@
 
-```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score."}
+```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score.", fig.width=3}
 diff <- naturetypes_wide$var_7SE - 
@@
 ```{r}
-path <- ifelse(dir == "C:", 
-        "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-
-ano <- sf::st_read(path,
+ano <- sf::st_read(paste0(pData, "Naturovervaking_eksport.gdb"),
                    layer = "ANO_SurveyPoint") %>%
@@
 
-```{r, fig.cap="Distribution of slitasje scores in the ANO data"}
-barplot(table(ano$slitasje), xlab="Sliatsje scores")
+```{r, fig.cap="Distribution of slitasje scores in the ANO data", fig.width=6}
+barplot(table(ano$slitasje), xlab="Sliatsje scores", ylab = "Number of localities")
 ```
@@
 
-Checking no that the new polygons have the area corresponding to the
+Checking now that the new polygons have the area corresponding to the
 proportion of the point that was part of the same main ecosystem:
@@
 
-The new dataset only contains polygons.
+The new data set contains **only polygons**.
 
@@
 
-```{r, fig.cap="Slitasje scores."}
+#### Distribution of Slitasje scores
+```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
 ggplot(temp, aes(x = Var1,
@@
 
-```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator."}
+```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator.", fig.width=4}
 temp <- as.data.frame(table(slitasje_data$origin))
@@
 #### Outline of Norway and regions
+
+The GIS layers are used to crop out marine areas, and to define accounting areas, respetively.
 
@@
 
-1.  scale the indicators at the polygon level.
-
-2.  take the simplified infrastructure index and, one class at the time,
-    extract the indicator value that intersect with those cells, and
-    extrapolate the mean of those values to all similar cell within a
-    given region.
-
-3.  calculate an indicator value for a region by doing a weighted
-    average based o the relative area of each infrastructure class.
+1.  normalise the indicators at the polygon level.
+
+2.  take the simplified infrastructure index and
+    extract the indicator value that intersect with a given cell, and
+    extrapolate the area weighted mean of those values to all similar cell within a
+    given region (`ea_spread()`). Errors are calculated from bootstrapping.
+
+3. Take the new area weighted indicator values for the homogeneous impact area (HIA) and match them with ecosystem occurrences of the same HIA. The errors are the same as for the HIAs.
+
+3.  calculate an indicator value for a region (accounting area) by doing a weighted
+    average based of the relative area of ecosystem occurrences. Errors shold be carried somehow, perhaps via weighted resampling
 
@@
 
-```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable"}
+```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable", fig.width=5}
 ggplot(slitasje_data, aes(x = slitasje,
@@
 
-```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable."}
+```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable.", fig.width=5}
 eaTools::ea_normalise(data = slitasje_data,
@@
 ```{r}
-HIA <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/infrastrukturindeks/homogeneous_impact_areas.rds")
+HIA <- readRDS(paste0(pData, "infrastrukturindeks/homogeneous_impact_areas.rds"))
 ```
@@
 # silently read cached file
-HIA_regions <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+
+HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
 ```
@@
 #silently read cached file
-all_stats <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
+all_stats <- readRDS(paste0(pData, "cache/all_stats.rds"))
 ```
@@
 #Silently read cached data
-corrCheck <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
+corrCheck <- readRDS(paste0(pData, "cache/corrCheck.rds"))
 ```
@@
 ```{r, include=F}
-wetlands_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
-seminat_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
-natOpen_slitasje_extr <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
+
+wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
+seminat_slitasje_extr  <- readRDS(paste0(pData, "cache/seminat_slitasje_extr.rds"))
+natOpen_slitasje_extr  <- readRDS(paste0(pData, "cache/natOpen_slitasje_extr.rds"))
 ```
@@
 
-```{r}
+```{r, warning=F}
 wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
@@
 
-There is very little variation both between the ecosystems and the four classes in the modified infrastructure index.
+There is some, but very little variation both between the ecosystems and the four classes in the modified infrastructure index.
 Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.
 
-The next steps now are to
+_These maps are easy to misunderstand, and they should therefor be communicated with caution._
+
+Now, there are multiple way to go proceed. We have indicator maps (vector data), and we have HIA area, and we have the HIA with indicator data (also vector data).
+
+Option 1:
 
@@
 
-Alternatively I can rasterize the indicator map, and mask it using a perfectly aligned ET map. This might be faster.
-
-### Crop to ET
-This is a very heavy operation.
+Step 1 above requires an st_interactions on two very large vactor data sets. This should be avoided.
+
+Option 2:
+
+1. rasterize the extrapolated indicator map, using the ET map as a tamplate, and
+
+2. mask it using the perfectly aligned ET map.
+
+This might be faster.
+
+Option 3:
+
+1. Draw n random point samples from the ET map, and
+
+2. extract indicator values at those points (from the extrapolated indicator map (a stars object)).
+
+This is perhaps even faster.
+I will try the last option, even though it will not provide spatailly explicit indicator values for the entire ecosystem delineation, like option 2 does.
+Perhaps I will revisit option 2 later.
+
+### Ecosystem map
+Importing the wetland map.
+
+This is a quite heavy operation and should be done on a server.
 ```{r, eval=F}
-mireET <- sf::st_read("R:/Kladd/Myr/Myrmodell.gdb")
-```
-
+mireET <- sf::st_read("/data/R/Kladd/Myr/Myrmodell.gdb")
+```
+
+This doesn't take that long.
+```{r, eval=F}
+mireET <- st_transform(mireET, 25833)
+```
+
+This take about 10 min. Not sure I need it.
+```{r, eval=F}
+mireET <- st_make_valid(mireET)
+```
+
+This takes a long!! time, and should be AVOIDED. 
 ```{r, eval=F}
 wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
+```
+
+Instead I will draw some random point samples.
+```{r}
+wetland_points <- st_sample(mireET, 10000)
+```
+
+And intersect this with the extrapolated indicator values.
+```{r}
+wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
+```
+
+Create an empty grid, 50 x 50 m resolution
+```{r}
+grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
+```
+
+Rasterize the indicator
+```{r}
+wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
+```
+
+And extract the values
+```{r}
+wetland_points_extracted <- st_extract(wetlands_slitasje_extr, wetland_points)
 ```

### 2023-01-30 DRAFT_slitasjeOgKjørespor.Rmd
doc: Fill in meta data

@@
 ### Original units
-The
+The variables are recorded on a coarse four-sped ordinal scale (Fig. \@ref(fig:four-step)).
 
 ### Temporal coverage
-
-<!-- Text here -->
+The data goes back to 2018, but is because of data shortage, it is bulked into one time step.
+I then use the mean date for the raw data, and define the variable as belonging to the year 2020 (read more [here](#scaled-slitasje-variable)).
 
 ### Aditional comments about the dataset
-
-<!-- Text here -->
+For a run through of the data set, see [here](#naturtyper)
 
 ## Ecosystem characteristic
-
 ### Norwegain standard
-
-<!-- Text here -->
+Primærproduksjon
+
+The indicator is tagged to the _Økologisk egenskap_ called Primærproduksjon (Primary productivity).
+This is tentative, and perhaps _abiotiske forhold_ is more suited.
+But the thought behind the chiose is that _sliatsje_ affect the potential for primary productivity.
 
 ### UN standard
-
-<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->
+The indicator is tagged as a Physical state characteristics indicator.
+This is quite clear.
+It's a tpe of abiotic characteritic that has to do with the physical structure rather than the chemical composition.
 
@@
 
-7TK and 7SE use a 4 step scale ![Four step condition
-scale](images/ninAR4b.PNG)
+7TK and 7SE use a 4 step scale
+
+```{r four-step, fig.cap="Four step condition scale"}
+knitr::include_graphics("images/ninAR4b.PNG")
+```
 
@@
 
-#### Scaled variable
+#### Scaled variable {#scaled-slitasje-variable}
 

### 2023-01-31 DRAFT_slitasjeOgKjørespor.Rmd
doc: Spell check and formatting

--- 2023-01-30 DRAFT_slitasjeOgKjørespor.Rmd
@@
 
-6.  [Scale](#scale-slitasje-ind) the `slitasje` variable based on reference values.
+6.  [Scale](#scale-slitasje-ind) the `slitasje` variable based on
+    reference values.
 
@@
 
-10.  Sample 10 00 random points from an ecosystem delineation and extract indicator values and sd at those points.
+10. Sample 10 00 random points from an ecosystem delineation and extract
+    indicator values and sd at those points.
 
@@
 ## About the underlying data
-The indicator uses a data set from a standardised field survey of nature types. You can read more about it [here](#naturtyper). I also import a data set called [ANO](#ANO-import)
+
+The indicator uses a data set from a standardised field survey of nature
+types. You can read more about the preliminary analyses
+[here](#naturtype). See also the [official
+site](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
+of the Environmental Agency. I also import a data set called
+[ANO](#ANO-import), which you can read about
+[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)
 
 ### Representativity in time and space
-The nature type mapping is not random and cannot be said to be area representative. 
-The data is from 2018 to present. The data from one field season usually becomes available early the following year.
+
+The nature type mapping is not random and cannot be said to be area
+representative. The ANO dataset however, is area representative. The
+data is from 2018 to present. The data from one field season usually
+becomes available early the following year.
 
 ### Original units
-The
+
+The variables are recorded on a coarse four-sped ordinal scale (Fig.
+\@ref(fig:four-step)).
 
@@
 
-<!-- Text here -->
+The data goes back to 2018, but is because of data shortage, it is
+bulked into one time step. I then use the mean date for the raw data,
+and define the variable as belonging to the year 2020 (read more
+[here](#scaled-slitasje-variable)).
 
@@
 
-<!-- Text here -->
+For a run through of the data set, see [here](#naturtype).
 
@@
 
-<!-- Text here -->
+The indicator is tagged to the *Økologisk egenskap* called
+**Primærproduksjon** (Primary productivity). This is tentative, and
+perhaps *abiotiske forhold* is more suited. But the thought behind the
+chiose is that *sliatsje* affect the potential for primary productivity.
 
@@
 
-<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->
+The indicator is tagged as a **Physical state characteristics**
+indicator. This is quite clear. It's a tpe of abiotic characteritic that
+has to do with the physical structure rather than the chemical
+composition.
 
@@
 
-<!-- Text here -->
-
-## Reference state and values
-
-### Reference state
-
-<!-- Text here -->
+Slitasje is not thought to be very collinear with any other indicator at
+the present.
+
+## Reference condition and values
+
+### Reference condition
+
+The reference condition is one with minimal negative human impact. This
+is also true for semi-natural ecosystems. For example, in the reference
+condition, *slitasje* in semi-natural ecosystems is defined to be zero,
+even though at no point in time would this condition be realized.
 
@@
 
-<!-- Text here -->
+The upper reference value is 0 (no *slitasje*), and the lower reference
+value is 100%. Note that 100% slitasje doesn't mean that all the area
+must be scared or damaged, but that all hypothetical quadrats around a
+point is affected (the variable is frequency-driven, and not
+amount-driven). The threshold for good ecosystem condition is 10%
+damage. This is on hte conservative side. Perhaps is should be set
+somewhat higher, like 20 or 30%.
+
+Read about the normalisation [here](##scaled-slitasje-variable).
 
@@
 
-<!-- Text here -->
+Uncertaintiesare estimated for aggregated indicator values by
+bootstrapping individual indicator values 1000 times and calculating a
+distribution of area weighted means. When aggregating a second time,
+from homogeneous impact areas to accounting areas, we assume a normal
+distribution around the indicator values, with the already mentioned
+errors, and sample n times from these and combine the resamples into an
+a new, area weighted, distribution. The errors for the accounting areas
+thus represents both the spatial variation, and the precision, of the
+indicator values within the accounting areas.
 
@@
 
-<!-- Text here -->
+No additional references.
 
@@
 
-This indicator uses the dataset
+This indicator uses the data set
 `Naturtyper etter Miljødirektoratets Instruks`, which can be found
 [here](https://kartkatalog.geonorge.no/metadata/naturtyper-miljoedirektoratets-instruks/eb48dd19-03da-41e1-afd9-7ebc3079265c).
-See [here](#naturtype) for a detailed description of the dataset.
-
-Importing the full dataset can be done like this, but because it can
-take a few seconds due to its size, I will cache the data on the P
-server.
+See also [here](#naturtype) for a detailed description of the data set.
+
+<!-- Importing the full data set can be done like this, but because it can -->
+
+<!-- take a few seconds due to its size, I will cache the data on the P -->
+
+<!-- server after subsetting. -->
 
@@
 
-We also have a separate summary file where the nature types are manually
-mapped to the correct NiN-main types. We can use this to find the nature
-types associated with the NiN main types that we are interested in.
+We also have a separate [summary file](##exp-natureType-summary) where
+the nature types are manually mapped to the correct NiN-main types. We
+can use this to find the nature types associated with the NiN main types
+that we are interested in.
 
@@
 We are only interested in a few NiN main types and their mapping units
-(see [here](https://ninanor.github.io/ecosystemCondition/naturtype.html)
-for justification)
+(see [here](#naturtype) for justification)
 
@@
 
-Import back in
-
-```{r}
+```{r, include=F}
+# Import back in
 naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
@@
 
-Keep only the relevant variables.
+Keep only the four relevant variables.
 
@@
 
-Names starting with a number is problematic, so adding a prefix
+Column names starting with a number is problematic, so adding a prefix
 
@@
 Now I need to create a single row per locality with a new variable which
-is a product of the four variables "7SE", "7TK", "PRTK" and "PRSL". I
-will create a new data table where I calculate the new variable which I
-then paste back into the sf object.
+is a product/function of the four variables "7SE", "7TK", "PRTK" and
+"PRSL". I will create a new data table where I calculate the new
+variable which I then paste back into the sf object.
 
@@
 highest value (worst condition). This table shows that a locality may
-have been assessed with both a PR variable and a NiN variable
+have been assessed with both a PR variable and a NiN variable:
 
@@
 head(naturetypes_wide)
+```
+
+```{r}
 anyDuplicated(naturetypes_wide$identifikasjon_lokalid)
@@
 
-When both variables are NA, we get -Inf. There were nine cases of this.
-Removing these now.
+When both variables are NA, we get -Inf. There were nine cases of this
+(in 2023). Removing these now.
 
@@
 
-This variable is also recorded in GRUK. The nature type dataset I'm
-working on here includes this data already (presently only 2021
-included). GRUK also records a related variable:
-`% cover in 5m radii circles`. This data is not published. In any case
-it is better to use the harmonized dataset.
+This variable is also recorded in
+[GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
+The nature type data set I'm working on here includes this data already
+(presently only 2021 included). GRUK also records a related variable:
+`% cover in 5m radii circles`, which si much more detailed. This data is
+not published. In any case it is better to use the harmonized data set
+in our case.
 
@@
 
+> *ANO could harmonie better with the Naturetype data set if it included
+> 7SE and 7TK for Naturally open and Semi natural ecosystem.*
+
 ```{r}
@@
 
-This dataset only contains data from year 2019 and 2021. We need to
-update this dataset later.
+This data set only contains data from year 2019 and 2021. We need to
+update this data set later. It is not clear why there is no data from
+2020.
 
@@
 I can remove the NA's, and thus the 2019 data.
+
+> *All of 2019 ANO data is excluded because of missing information*
 
@@
 We want to use area weighting in this indicator, so we can use this
-percentage cover data to calculate the area. Note that both datasets use
-m2 as area units.
+percentage cover data to calculate the area. Note that both data sets
+use m^2^ as area units.
 
@@
 Let's now look at the distribution of the variables. First I need to
-seperate the variable name fro the values themselves. Now the data looks
-like this:
+separate the variable name from the values themselves. Now the data
+looks like this:
 
@@
 No NA's this time.
+
+> *The ANO field app should not allow user to leave blank cells. This
+> resulted in the exclusion of data.*
 
@@
 
-We need to combine the nature type dataset with the ANO dataset. I will
-add a column `origin` to show where the data comes from. I will also and
-add data about the main ecosystem.
+We need to combine the nature type data set with the ANO data set. I
+will add a column `origin` to show where the data comes from. I will
+also add a column with the main ecosystem.
 
@@
 
-I use `dplyr::select` to can reduce the number of columns to keep things
-a bit more tidy.
+I use `dplyr::select` to reduce the number of columns to keep things a
+bit more tidy.
 
@@
 
-The ANO data is point data, and the nature type data is multipolygon.
-Because later we want to save this as a shape file, we cannot have a
-mixed class type. I will therefore convert the points to polygons. I use
-the area column to calculate a radii that gives that area.
+The ANO data is point data, and the nature type data is *multipolygon*.
+Because later we might want to save this as a shape file, we cannot have
+a mixed class type. I will therefore convert the points to polygons. I
+use the area column to calculate a radii that gives that area.
 
@@
 
-The area calculation seems to have worked fine.
+The area calculation seems to have worked fine. Checking that the new
+data set contains **only polygons**.
 
@@
 
-The new data set contains **only polygons**.
-
-Grouping and tallying the data for plotting
-
-```{r}
+Ok.
+
+#### Distribution of Slitasje scores
+
+```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
 temp <- as.data.frame(table(slitasje_data$slitasje))
-```
-
-#### Distribution of Slitasje scores
-```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
 ggplot(temp, aes(x = Var1,
@@
 So the ANO data is not very important here, but it can become more
-important in the future.
+important in the future so good to have them included in the workflow.
 
@@
 
-The GIS layers are used to crop out marine areas, and to define accounting areas, respetively.
+The GIS layers are used to crop out marine areas, and to define
+accounting areas, respectively.
 
@@
 them first. If the scaled value is representative and precise at the
-polygon level, then I could scale at that level. However, the combined
-surveyed area is still so little of the total area of Norway, so that
-only producing indicator values for these leaves the indicator without
-much value for regional assessments. Later when we produce assessments
-and indicator values at the regional level, we cannot simply do an area
-weighting of the polygons in each region. This is because we don't want
-to assume that the polygons are representative far outside of the mapped
-area. But perhaps we can assume them to be representative inside
-*homogeneous ecological areas*. That's where the infrastructure index
-comes in. Here's the plan:
+polygon level, then I could scale at that level. I think they are.
+
+However, the combined surveyed area is a very small fraction of the
+total area of Norway, so that only producing indicator values for the
+mapped areas leaves the indicator without much value for regional
+assessments. When we do regional assessments and calcualte regional
+indicator values, we cannot simply do an area weighting of the polygons
+in each region. This is because we don't want to assume that the
+polygons are representative far outside of the mapped area. But perhaps
+we can assume them to be representative inside *homogeneous ecological
+areas,* or *Homoegeneous Impact Areas* (HIAs). That's where the
+infrastructure index comes in. Here's the plan:
 
@@
 
-2.  take the simplified infrastructure index and
-    extract the indicator value that intersect with a given cell, and
-    extrapolate the area weighted mean of those values to all similar cell within a
-    given region (`ea_spread()`). Errors are calculated from bootstrapping.
-
-3. Take the new area weighted indicator values for the homogeneous impact area (HIA) and match them with ecosystem occurrences of the same HIA. The errors are the same as for the HIAs.
-
-3.  calculate an indicator value for a region (accounting area) by doing a weighted
-    average based of the relative area of ecosystem occurrences. Errors shold be carried somehow, perhaps via weighted resampling
-
-#### Scaled variable
+2.  take a simplified infrastructure index (vector data set with HIAs)
+    and extract the corresponding indicator values that intersect with a
+    given HIA, and extrapolate the area weighted mean of those values to
+    the entire HIA withing a given geographical region. Errors are
+    calculated from bootstrapping.
+
+3.  Take the new area weighted indicator values for the HIAs and match
+    them with ecosystem occurrences of the same HIA (possibly using
+    randomly sampled points from the available ecosystem occurrences).
+    The errors are the same as for the HIAs.
+
+4.  calculate an indicator value for a region (accounting area) by doing
+    a weighted average based of the relative area of ecosystem
+    occurrences. Errors should be carried somehow, perhaps via weighted
+    resampling.
+
+#### Scaled variable {#scaled-slitasje-variable}
 
@@
 scale. Although not perfect, I will use the median value for each
-category.
+category. I.e. when the variable is recorded as 2 (meaning from 1/16th
+to 1/2), this is converted to 31.25%, i.e. in the middle between the two
+end points of the category.
 
@@
 using a non-linear transformation.
+
+Also, the threshold value is set quite conservativly, and should be
+discussed.
 
@@
 
+Assigning the indicator to year 2020.
+
 ```{r}
@@
 data is generated by discretizing the Norwegian Infrastructure Index. I
-refer to the ordinal values of the four impact area classes as their
-*Human Impact Factor* (HIF).
+refer to the ordinal values of the four HIA classes as their *Human
+Impact Factor* (HIF). This is just to keep the approach separate from
+the Norwegian Infrastructure Index.
 
@@
 
-I want to check that HIF is in fact a good predictor for _slitasje_.
+I want to check that HIF is in fact a good predictor for *slitasje*.
 
@@
 
-```{r, echo=F}
+```{r, include=F}
 # silently read cached file
-
 HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
 ```
+
+Create a new column by crossing region and HIF
 
@@
 #### Validate
-I now have 20 unique areas (for each main ecosystem) that I will, given there is data, extrapolate
-indicator values over
+
+I now have 20 unique areas (for each main ecosystem) that I will, given
+there is data, extrapolate indicator values over.
 
 ###### Subset ETs
-I will subset the `sliatsje_data` in the three ecosystems. 
-Note that only for open wetlands do we have good ecosystem delineation maps to base a spatial averaging on.
+
+I will subset the `sliatsje_data` in the three ecosystems. Note that
+only for open wetland do we have good ecosystem delineation maps to base
+a spatial averaging on.
 
@@
 ```
+
+Creating some summary statistics.
 
@@
 
-This rather dense figure needs to be looked at carefully. 
-In the top row we see that the sample size (number of polygons) varies between ecosystems.
-For naturally open ecosystems many polygons are from high infrastructure areas, not surprisingly, and there are few points for the lowest impact class (zero such polygins in _Sørlandet_ for example).
-We there may not be able to extrapolate indicator values in areas in _Sørlandet_ with low levels of infrastructure.
-We can also from the middle row see the association between the indicator values (area weighted means) and the modified infrastructure index (HIF).
-We should then not put too much weight on data were there is little data (low n) and/or high standard error (sd, bottom row).
-For example, with Naturally open ecosystems, Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between the indicator and the pressure that we want to see (middle row).
-The pattern is repeated for all regions except Østlandet, but there we also see that HIF class 3 has very little data, and a high sd, and this could explain the discrepancy (but it doesn't support the general relationship either).
-
-In semi-natural areas and wetlands, we have very little data from HIF class 3, and the sd's are very big.
-The indicator-pressure relationship (middle row) looks the most messy for wetlands, but if ignoring HIF=3, the relationship looks much better.
+This rather dense figure needs to be looked at carefully. In the top row
+we see that the sample size (number of polygons) varies between
+ecosystems. For naturally open ecosystems many polygons are from high
+infrastructure areas, not surprisingly, and there are few points for the
+lowest impact class (zero such polygons in *Sørlandet* for example). We
+may therefore not be able to extrapolate indicator values for Naturally
+open areas in in *Sørlandet* with low levels of infrastructure. We can
+also from the middle row see the association between the indicator
+values (area weighted means) and the modified infrastructure index
+(HIF). We should then not put too much weight on the bars were there is
+little data (low n) and/or high standard error (sd, bottom row). For
+example, with Naturally open ecosystems, Midt-Norge has the overall
+lowest errors, and there we also see the kind of relationship between
+the indicator and the pressure that we want to see (middle row). The
+pattern is repeated for all regions except Østlandet, but there we also
+see that HIF class 3 has very little data, and a high sd, and this could
+explain the discrepancy (but it doesn't support the general relationship
+either).
+
+In semi-natural areas and wetlands, we have very little data from HIF
+class 3, and the sd's are very big. The indicator-pressure relationship
+(middle row) looks the most messy for wetlands, but if ignoring HIF=3,
+the relationship looks much better.
 
@@
 
-In addition to looking at the area weighted means and its relationship with the pressure indicator (HIF), we can do the same at the polygon level.
-Because the indicator is pseudo-continuous, boxplots for example become useless, so I will present a relative frequency plot.
+In addition to looking at the area weighted means and its relationship
+with the pressure indicator (HIF), we can do the same at the polygon
+level. Because the indicator is pseudo-continuous, boxplots for example
+become useless, so I will present a relative frequency plot.
 
@@
 ```
+
 ```{r, include=F}
@@
 
-When down weighting the importance of those groups/categories that have very little The figure above I think supports the indicator-pressure relationship and justifies using the HIA+region intersections as local reference areas (i.e.
-homogeneous impact areas).
-
-We can also calculate Kendall tau b coefficients, although I'm not complete sure this is valid when we have ordinal data with only four levels, since there will be a lot of ties.
+The figure above I think supports the indicator-pressure relationship
+and justifies using the HIA+region intersections as local reference
+areas. Especially when down-weighting the importance of those
+groups/categories that have very little data.
+
+We can also calculate *Kendall tau b* coefficients, although I'm not
+complete sure this is valid when we have ordinal data with only four
+levels, since there will be a lot of ties.
 
@@
 
-From the figures above it appears that the relationship is quite week in general. However, the sheer number of data points suggest that group differences will be very reliable.
-_Naturlig åpne områder_ has the lowest correlation between the indicator and the homogeneous area classes, and of the five regions, _Nord-Norge_ has the lowest correlation.
+From the figures above it appears that the relationship is quite week in
+general. However, the sheer number of data points suggest that group
+differences will be very reliable. *Naturlig åpne områder* has the
+lowest correlation between the indicator and the homogeneous area
+classes, and of the five regions, *Nord-Norge* has the lowest
+correlation.
 
 #### Looking at the HIA
+
 Here is a view of the data zooming in on Trondheim
@@
 I want to add a threshold so that we don't end up over extrapolating
-based on too few data points. From Fig. \@ref(fig:dense-stats) I conclude that we need at least 70 data points for the sd to stabilize. I will conservatively round this up to 100 data points.
+based on too few data points. From Fig. \@ref(fig:dense-stats) I
+conclude that we need at least 70 data points for the sd to stabilize. I
+will conservatively round this up to 100 data points.
 
@@
 ```
+
 ```{r, include=F}
-
 wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
@@
 
-Cropping the raster to the bbox
+Cropping the data to the bbox
 
@@
 
-There is some, but very little variation both between the ecosystems and the four classes in the modified infrastructure index.
-Note however, that for wetland and semi-natural areas we did not extrapolate values for the most urban areas, and that for naturally open areas we did not extrapolate to the least affected areas.
-
-_These maps are easy to misunderstand, and they should therefor be communicated with caution._
-
-Now, there are multiple way to go proceed. We have indicator maps (vector data), and we have HIA area, and we have the HIA with indicator data (also vector data).
+There is some, but very little variation, both between the ecosystems
+and between the four HIF classes in the modified infrastructure index.
+Note however, that for wetland and semi-natural areas we did not
+extrapolate values for the most urban areas, and that for naturally open
+areas we did not extrapolate to the least affected areas.
+
+> *The maps abov are easy to misunderstand, and they should therefor be
+> communicated with caution. They should be considered as temporary
+> files or as by-products.*
+
+Now, there are multiple way to proceed. We have
+
+-   indicator maps (vector data), and we have
+
+-   total HIA areas, and we have the
+
+-   extrapolated indicator maps (HIA maps with aggregated indicator
+    values and errors, also vector data).
+
+-   We also have, for wetlands, an ecosystem type (ET) map.
 
@@
 
-1. crop the extrapolated indicator maps to ecosystem delineation maps, and then;
+1.  crop (i.e. intersect) the extrapolated indicator maps to ecosystem
+    delineation maps, and then;
 
@@
 
-Step 1 above requires an st_interactions on two very large vactor data sets. This should be avoided.
+Step 1 above requires an st_interactions on two very large vector data
+sets with millions of polygons. This should be avoided.
 
@@
 
-1. rasterize the extrapolated indicator map, using the ET map as a tamplate, and
+1.  rasterize the extrapolated indicator map, using the ET map as a
+    template, and
 
@@
 
-1. Draw n random point samples from the ET map, and
-
-2. extract indicator values at those points (from the extrapolated indicator map (a stars object)).
-
-This is perhaps even faster.
-I will try the last option, even though it will not provide spatailly explicit indicator values for the entire ecosystem delineation, like option 2 does.
-Perhaps I will revisit option 2 later.
+1.  Draw *n* random point samples from the ET map, and
+
+2.  extract indicator values at those points (from the extrapolated
+    indicator map (a stars object)).
+
+This is perhaps even faster again. I will try the last option, even
+though it will not provide spatially explicit indicator values for the
+entire ecosystem delineation, like option 2 does. Perhaps I will revisit
+option 2 later.
 
 ### Ecosystem map
+
 Importing the wetland map.
@@
 This is a quite heavy operation and should be done on a server.
+
 ```{r, eval=F}
@@
 This doesn't take that long.
+
 ```{r, eval=F}
@@
 This take about 10 min. Not sure I need it.
+
 ```{r, eval=F}
@@
 This takes a long!! time, and should be AVOIDED.
+
 ```{r, eval=F}
@@
 Instead I will draw some random point samples.
-```{r}
+
+```{r, eval=F}
 wetland_points <- st_sample(mireET, 10000)
@@
 And intersect this with the extrapolated indicator values.
-```{r}
+
+```{r, eval=F}
 wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
@@
 
-Create an empty grid, 50 x 50 m resolution
-```{r}
+And extract the values
+
+```{r, eval=F}
+wetland_points_extracted <- st_intersection(wetlands_slitasje_extr, wetland_points)
+```
+
+If we went for option 2, we could instead do something like this, and
+
+create an empty grid over Norway, 50 x 50 m resolution
+
+<!--# This we should perhaps do once, in a separate misc chapter, and make available for all chapters -->
+
+```{r, eval=F}
 grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
@@
 
-Rasterize the indicator
-```{r}
+and then rasterize the indicator
+
+```{r, eval=F}
 wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
@@
 
-And extract the values
-```{r}
-wetland_points_extracted <- st_extract(wetlands_slitasje_extr, wetland_points)
-```
+To get a perfect alignment with the ET map, we should change `grd` with
+the ET raster map.
+
+Then we can use starts filtering with `[ .. ]` to mask out the areas
+that do not belong to the ET.
 
@@
 
-```{r, eval=F}
-wetlands_slitasje_extr_cropped_reg <- ea_spread(indicator_data = wetlands_slitasje_extr_cropped,
-                                      indicator = i_2020,
-                                      regions = regions,
-                                      groups = ?)
-```
+We need to aggreate the ET occurence and their indicator values, to a
+regional delineation. This functionality will be developed in `eaTools`.
+We need to produce distributions for the indicator valeus for each ET
+occurrence (or groups of them) and resample the distributions wiht an
+area weighting.
 
@@
 
-The calculations of indicator uncertainty is done with bootstrapping, as is described in eaTools.
+The calculations of indicator uncertainty is done with bootstrapping, as
+is described in eaTools.
 
@@
 map at its highest resolution, for use in various web interfaces etc.
-
-But I also need to export a map of the indicator where we extrapolate
-the indicator values to the homogeneous areas. This map should be
+For the finest resolution indicator map I will not include columns for
+uncertainty (because this will only be estimated at the aggregated
+level), or reference levels (because there is no spatial variation).
+
+But I will also export a map of the indicator where we extrapolate the
+indicator values to the homogeneous areas. This map should be
 interpreted with caution, since they seem to provide explicit indicator
@@
 aggregation purposes. I will mark this map with the prefix
-*homogeneous-areas*.
-
-I will also export a map of the *homogeneous areas*, over which I will
-aggregate the indicator values. This map might be useful in later
-indicator work.
-
-For the finest resolution indicator map I will not include columns for
-uncertainty (because this will only be estimated at the aggregated
-level), or reference levels (because there is no spatial variation).
-
-### Sliatsje in wetlands
+*homogeneous-areas*. I can also export this same mappen after masking it
+with the ET map, and this output must come with the same discalimers.
+
+Finally I will export a map of the five accounting areas with
+wall-to-wall indicator value and errors. This map can also contain data
+covering other ETs. Otherwise the maps would not we visually
+interpretative.
+
+### Slitasje in wetlands
 
@@
 
-```{r}
-exp_wetland <- select(slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",],
-              origin,
-              area,
-              i_2020,
-              SHAPE
-              )
+```{r, eval=F}
+# keep col names short and standardized
+exp_wetland <- ...
+```
+
+```{r, eval=F}
+exp_wetland_HIAextrapolated <- …
+```
+
+```{r, eval=F}
+exp_wetland_AAextrapolated <- …
 ```
@@
          append = F)
-```
-
-### Sliatsje in wetlands, extrapolated to homegeneous impact areas.
+
+st_write(exp_wetlandHIAextrapolated, 
+         "indicators/indicatorMaps/extrapolated_to_homogeneous_areas/slitasje/homogeneous_areas_wetland_slitasje.shp",
+         append = F)
+
+st_write(exp_wetland, "indicators/indicatorMaps/extrapolated_to_accounting_areas/slitasje/accounting_areas_wetland_slitasje.shp",
+         append = F)
+```

### 2023-01-31 naturtype.Rmd
doc: Spell check and formatting

@@
 
-# Nature types - Part 1 {#naturtype1}
-
-**On the application of the naturetype dataset - Part 1 - Data exploration and cleaning**
+# Nature types {#naturtype}
+
+**On the application of the nature type dataset**
+
+*Data exploration and an analyses of thematic coverage*
 
@@
 library(tmap)
+library(tidyr)
 library(ggplot2)
@@
 library(stringr)
+library(forcats)
+library(RColorBrewer)
 library(units)
+library(egg)
 library(dplyr)
@@
 ```{r}
+target <- c("Semi-naturlig mark","Våtmark","Naturlig åpne områder under skoggrensa")
 dat <- dat[dat$hovedøkosystem %in% target,]
@@
   ungroup() %>%
-  mutate(NiN_mainType = fct_reorder(NiN_mainType, km2)) %>%
+  mutate(NiN_mainType = forcats::fct_reorder(NiN_mainType, km2)) %>%
   ggplot(aes(x = NiN_mainType,
@@
 
-## Export data
+## Export data {#exp-natureType-summary}
 Exporting the table with the summary statistics and other information that I compled for each nature type in the three main ecosystems.
@@
 
-# Appendix 1
+## Appendix 1
 Here is a deeper look into the causes for why some localities are recorded with seemingly erroneous NiN mapping units (and hence NiN main types)

### 2023-01-31 DRAFT_slitasje.Rmd
bug: Rename and fix color palette

--- 2023-01-31 DRAFT_slitasjeOgKjørespor.Rmd
@@
 
-> *ANO could harmonie better with the Naturetype data set if it included
+> *ANO could harmonise better with the Naturetype data set if it included
 > 7SE and 7TK for Naturally open and Semi natural ecosystem.*
@@
 was not recorded at all in 2019:
+```{r}
+# Hard coding name change
+#unique(ano$hovedtype_250m2)[3]
+ano$hovedtype_250m2[ano$hovedtype_250m2==unique(ano$hovedtype_250m2)[3]] <- "Aapen jordvassmyr"
+```
 
@@
 may therefore not be able to extrapolate indicator values for Naturally
-open areas in in *Sørlandet* with low levels of infrastructure. We can
-also from the middle row see the association between the indicator
+open areas in in *Sørlandet* with low levels of infrastructure.
+
+We can also from the middle row see the association between the indicator
 values (area weighted means) and the modified infrastructure index
@@
 little data (low n) and/or high standard error (sd, bottom row). For
-example, with Naturally open ecosystems, Midt-Norge has the overall
-lowest errors, and there we also see the kind of relationship between
+example, with Naturally open ecosystems,
+
+Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between
 the indicator and the pressure that we want to see (middle row). The
@@
 ```{r, fig.cap="Slitasje indicator extrapolated over Trondheim"}
+myCol <- "RdYlGn"
 tmap_arrange(
@@
     title="Slitasje indicator",
-    palette = "viridis",
+    palette = myCol,
     style="cont",
@@
 
-> *The maps abov are easy to misunderstand, and they should therefor be
+> *The maps above are easy to misunderstand, and they should therefor be
 > communicated with caution. They should be considered as temporary

### 2023-02-01 DRAFT_homogeneous_impact_areas.Rmd
review: Fixed all comments in inital review

@@
 
-<!-- Load all you dependencies here -->
+<!-- Load all your dependencies here -->
 
@@
 library(tmap)
-library(dplyr)
-library(stars)
-library(ggplot2)
+library(stars);packageVersion("stars")
 library(units)
@@
 library(DT)
-#library(osmplotr)
+library(tidyverse)
 ```
@@
 This chapter documents the creation of a wall-to-wall map of homogeneous impact areas in Norway.
-The map is produces by binning values in the infrastructure index into four discrete categories.
+The map is produced by binning values in the infrastructure index into four discrete categories.
 It is likely to be a good predictor for some indicators, such as *slitasje* and the presence of *alien species*.
@@
 It is a wall-to-wall raster over Norway with 100 m resolution.
-Each pixel is given a value along a continuous gradient from 0 to around 15.4, representing the frequency or surrounding cells within 500 m with human infrastructure (houses, roads, ect).
+Each pixel is given a value along a continuous gradient from 0 to around 15.4, representing the frequency of surrounding cells within 500 m with human infrastructure (houses, roads, ect).
 
@@
 
-The infrastructure index is calculated with data that have slightly different dates, but can be said to represent the current status, the period 2010-2020.
+The infrastructure index is calculated with data that have slightly different dates, but can be said to represent year 2015.
 The data covers all of mainland Norway.
@@
 
-```{r, fig.cap="Infrastructure indeks over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
+```{r, fig.cap="Infrastructure index over Trondheim, comparing the continous scale with the ordinal four-step scale. Major roads are in white."}
 tmap_arrange(map_trd,
@@
 This looks pretty good to me.
-It depicts a gradient in human presence from high within the buildt zone, to *no to very low* in the forests and mountains to the west.
+It depicts a gradient in human presence from high within the built zone, to *no to very low* in the forests and mountains to the west.
 Note that there is still considerable human activity also there in the form of outdoor recreation and even forestry.
@@
 The resolution in this map is more than we need, and the size of the data implies that the next step, the vectorization, would take too long.
-We therefore aggregate, ar reduce the resolution.
+We therefore aggregate, or reduce the resolution.
 
@@
 
-```{r, fig.cap="Infrastrcture indeks over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
+```{r, fig.cap="Infrastrcture index over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
 tm_shape(infra_trd_reclassed_agg)+
@@
 
-This resolution is more that good enough for our purpose here.
+This resolution is more than good enough for our purpose here.
 I see that the map extends a bit into the fjord. We want to cut away these areas. We can use a map of the outline of Norway to do this.
@@
 
-```{r, fig.cap="Infrastrcture indeks over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
+```{r, fig.cap="Infrastrcture index over Trondheim on a four-step discrete scale, aggregated to 1x1 km resolution using the mean function."}
 tm_shape(infra_trd_reclassed_agg_terrestrial)+
@@
 
-I have treated the raster cells as points when cropping, so that cells where the center or the cell is outside he terrestrial delineation are removed.
+I have treated the raster cells as points when cropping, so that cells where the center of the cell is outside the terrestrial delineation are removed.
 I think this makes the most sense for unbiased area statistics, but some coastal indicator data could also be excluded because of this.
@@
 # silently load cached file
-infra_agg_terrestrial <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/infra_agg.rds")
+
+infra_agg_terrestrial <- readRDS(path)
 ```

### 2023-02-01 DRAFT_homogeneous_impact_areas.Rmd
bug: Big cache

@@
 Anders Kolstad
-
-```{r, echo=FALSE}
-Sys.Date()
-```
 
@@
 
-```{r, warning=F}
-infra_agg_discrete_vect <- eaTools::ea_homogeneous_area(infra_agg_discrete, groups = infrastructureIndex)
+```{r , eval=F}
+infra_agg_discrete_vect <- 
+  eaTools::ea_homogeneous_area(infra_agg_discrete, 
+                                groups = infrastructureIndex)
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")
+
+saveRDS(infra_agg_discrete_vect, paste0(path, "infra_agg_discrete_vect.rds"))
+```
+
+```{r, include=F}
+infra_agg_discrete_vect <- readRDS(paste0(path, "infra_agg_discrete_vect.rds"))
 ```
@@
 
-```{r test-2, fig.cap="A vectorized version of the Infrastructure indeks over Trondheim on a four-step discrete scale."}
+```{r test-2, fig.cap="A vectorized version of the Infrastructure index over Trondheim on a four-step discrete scale."}
 tm_shape(infra_agg_discrete_vect_trd)+
@@
 
-As can be seen in the figure above, `st_warp` merges grid cells that have the same value, and the vectorized raster doesn't not end up being that big in the end.
+As can be seen in the figure above, `st_warp` merges grid cells that have the same value, and the vectorized raster doesn't end up being that big in the end.
 
-## Chech national distribution
+## Check national distribution
 
@@
 # silently read cached file
-infra_stats <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_stats.rds")
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/infra_stats.rds",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/infra_stats.rds")
+infra_stats <- readRDS(path)
 ```

### 2023-02-01 DRAFT_homogeneous_impact_areas.Rmd
bug: wrong path

@@
 ```{r, include=F}
+path <- ifelse(dir == "C:", 
+        "P:/41201785_okologisk_tilstand_2022_2023/data/cache/",
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/")
+
 infra_agg_discrete_vect <- readRDS(paste0(path, "infra_agg_discrete_vect.rds"))

### 2023-02-02 DRAFT_homogeneous_impact_areas.Rmd

@@
 library(tidyverse)
+library(units)
 ```
@@
 
-```{r, echo=F}
+```{r, include=F}
 # silently read cached file
@@
 ```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
-infra_stats$area <- sf::st_area(infra_stats)
+infra_stats$area_km2 <- units::drop_units(sf::st_area(infra_stats))
+infra_stats$area_km2 <- infra_stats$area_km2/1000
 
@@
   group_by(region, infrastructureIndex) %>%
-  summarise(area = mean(area))
+  summarise(area_km2 = mean(area_km2))
 
 ggarrange(
-  ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
+  ggplot(temp, aes(x = region, y = area_km2, fill = factor(infrastructureIndex)))+
     geom_bar(position = "stack", stat = "identity")+
@@
   ,
-  ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
+  ggplot(temp, aes(x = region, y = area_km2, fill = factor(infrastructureIndex)))+
     geom_bar(position = "fill", stat = "identity")+
@@
     coord_flip()+
+    ylab("Fraction of total area")+
     xlab("")
@@
 ```{r}
-temp$area <- round(temp$area,0)
+temp$area_km2 <- round(temp$area_km2,0)
 DT::datatable(temp)

### 2023-02-02 DRAFT_slitasje.Rmd

@@
 library(tmap)
-library(ggplot2)
 library(ggpubr)
@@
 library(tidyverse)
-library(tidyr)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-Slitasje is not thought to be very collinear with any other indicator at
+Slitasje is not thought to exhibit collinearity with any other indicator at
 the present.
@@
 The upper reference value is 0 (no *slitasje*), and the lower reference
-value is 100%. Note that 100% slitasje doesn't mean that all the area
+value is 100%. Note that 100% slitasje does not mean that all the area
 must be scared or damaged, but that all hypothetical quadrats around a
@@
 amount-driven). The threshold for good ecosystem condition is 10%
-damage. This is on hte conservative side. Perhaps is should be set
+damage. This is on the conservative side. Perhaps is should be set
 somewhat higher, like 20 or 30%.
@@
   theme_bw(base_size = 12)+
-  guides(fill = "none")+
+  theme(legend.position = "top",
+        legend.title = element_blank(),
+        legend.direction = "vertical")+
+  #guides(fill = "none")+
   xlab("")+
@@
 
-I appears most localities are in good condition.
+It appears most localities are in good condition.
 
@@
 (presently only 2021 included). GRUK also records a related variable:
-`% cover in 5m radii circles`, which si much more detailed. This data is
+`% cover in 5m radii circles`, which is much more detailed. This data is
 not published. In any case it is better to use the harmonized data set

### 2023-02-02 slitasje.Rmd

--- 2023-02-02 DRAFT_slitasje.Rmd
@@
----
-title: "Slitasje og kjørespor"
-output:
-  html_document
-editor_options: 
-  markdown: 
-    wrap: 72
----
+# (PART\*) INDICATORS{-}
 
@@
 ## Introduction
+This indicator represent human cause terrain damage from hiking or off-road vehicle traffic.
+The data come from a combination non-random field surveys and area-representative surveys.
+Therefore extra care is taken not to over-extrapolate, and we use local reference area in the form of _homogeneous impact areas_.
 
@@
 
-3.  Covert [ANO points to polygons](#ANO-points-to-poly)
+3.  Convert [ANO points to polygons](#ANO-points-to-poly)
 
 4.  For each polygon, extract the NiN-variable that has the lowest value
-    (worst-rules rule), e.g. like Fig. \@ref(fig:slitasje-naturetype).
+    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype).
 
@@
 
-10. Sample 10 00 random points from an ecosystem delineation and extract
+10. Sample 1000 random points from an ecosystem delineation and extract
     indicator values and sd at those points.
@@
 site](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
-of the Environmental Agency. I also import a data set called
+of the Environment Agency. I also import a data set called
 [ANO](#ANO-import), which you can read about
@@
 The nature type mapping is not random and cannot be said to be area
-representative. The ANO dataset however, is area representative. The
+representative. The ANO data set however, is area representative. The
 data is from 2018 to present. The data from one field season usually
@@
 
-The variables are recorded on a coarse four-sped ordinal scale (Fig.
+The variables are recorded on a coarse four-step ordinal scale (Fig.
 \@ref(fig:four-step)).
@@
 perhaps *abiotiske forhold* is more suited. But the thought behind the
-chiose is that *sliatsje* affect the potential for primary productivity.
+choise is that *sliatsje* affect the potential for primary productivity.
 
@@
 The indicator is tagged as a **Physical state characteristics**
-indicator. This is quite clear. It's a tpe of abiotic characteritic that
+indicator. This is quite clear. It's a type of abiotic characteristic that
 has to do with the physical structure rather than the chemical
@@
 
-Slitasje is not thought to exhibit collinearity with any other indicator at
+_Slitasje_ is not thought to exhibit collinearity with any other indicator at
 the present.
@@
 
+* Upper = 0%
+
+* Threshold = 10%
+
+* Lower = 100%
+
 The upper reference value is 0 (no *slitasje*), and the lower reference
-value is 100%. Note that 100% slitasje does not mean that all the area
-must be scared or damaged, but that all hypothetical quadrats around a
+value is 100%. Note that 100% _slitasje_ does not mean that all the area
+must be scarred or damaged, but that all hypothetical quadrats around a
 point is affected (the variable is frequency-driven, and not
@@
 
-Uncertaintiesare estimated for aggregated indicator values by
+Uncertainties/errors are estimated for aggregated indicator values by
 bootstrapping individual indicator values 1000 times and calculating a
@@
 distribution around the indicator values, with the already mentioned
-errors, and sample n times from these and combine the resamples into an
+errors, and sample _n_ times from these and combine the resamples into an
 a new, area weighted, distribution. The errors for the accounting areas

### 2023-02-17 DRAFT_master_grid.Rmd
create master grid

---
title: ""
output:
  html_document
---

# Master grid

<br />

*Author and date:* Anders L. Kolstad

<br />

<!-- Load all you dependencies here -->

```{r setup, include=FALSE}
library(knitr)
library(sf)
library(stars)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This chapter describes the making of a master raster grid for all of mainland Norway. This grid is used to align all ecosystem and indicator maps so that it becomes possible to do easy resamping, masking and aggregating.

## About the underlying data

I will use the [statistical grid from Norway (5x5km)](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) and resample this to a higher resolutio of 50 x 50 meters.

## Analyses

```{r}
# Set up conditional file paths
dir <- substr(getwd(), 1,2)

pData <- ifelse(dir == "C:", 
               "P:/41201785_okologisk_tilstand_2022_2023/data/",
               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
```

Import data
```{r}
#st_layers(paste0(pData, "Basisdata_0000_Norge_25833_StatistiskRutenett5km_FGDB.gdb"))
masterGrid_50m <- sf::read_sf(paste0(pData, "Basisdata_0000_Norge_25833_StatistiskRutenett5km_FGDB.gdb")) 
```

```{r, eval=F}
tmap_mode("view")
tm_shape(grid_5km$SHAPE)+
  tm_borders()
```

![](images/masterGrid_screenShot.PNG){width="800"}

Confirming CRS is EPSG25833:
```{r}
st_crs(grid_5km)
```

### Make grid

Use the bbox and split into 50 x 50 meter cells

```{r}
(masterGrid_50m <- st_as_stars(st_bbox(grid_5km), dx = 50, dy = 50))
```

This file is 6GB.

Number of cells, in millions, is:
```{r}
(nrow(masterGrid_50m)*ncol(masterGrid_50m))/10^6
```

So, quite a lot. For the area accounts we might consider using 10x10m.

### Eksport file (final product)

```{r}
stars::write_stars(masterGrid_50m, paste0(pData, "masterGrid_50m.tiff"))
```

### Test import
```{r}
temp <- stars::read_stars(paste0(pData, "masterGrid_50m.tiff"))
st_dimensions(temp)
```

### 2023-02-17 DRAFT_master_grid.Rmd
testing rendring on server

@@
 
-*Author and date:* Anders L. Kolstad
+*Author:* Anders L. Kolstad
 
@@
 
-I will use the [statistical grid from Norway (5x5km)](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) and resample this to a higher resolutio of 50 x 50 meters.
+I will use the [statistical grid from Norway (5x5km)](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) and resample this to a higher resolution of 50 x 50 meters.
 
@@
 #st_layers(paste0(pData, "Basisdata_0000_Norge_25833_StatistiskRutenett5km_FGDB.gdb"))
-masterGrid_50m <- sf::read_sf(paste0(pData, "Basisdata_0000_Norge_25833_StatistiskRutenett5km_FGDB.gdb")) 
+grid_5km <- sf::read_sf(paste0(pData, "Basisdata_0000_Norge_25833_StatistiskRutenett5km_FGDB.gdb")) 
 ```

### 2023-02-22 DRAFT_climate_indicators.Rmd
add meta data

---
title: "Climate indicators"
output:
  html_document
---

# Cilmate indicators

<br />

_Author and date:_
<!-- Write your name here -->

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stars)
library(sf)
library(tmap)
library(lubridate)
```

```{r, echo=F}
Ecosystem <- "All" # e.g. "Skog og fjell"
Egenskap  <- "Abiotiske egenskaper" # e.g. "Primærproduksjon"
ECT       <- "Physical state characteristics" # e.g. "Structural state characteristic"
Contact   <- "Anders L. Kolstad" # e.g. "Anders Kolstad"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<br />
<br />
<hr />

## Introduction
This chaters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).

## About the underlying data
The data is in a raster format and extends back to 1957 in the form of multiple interpolated climate variables.
The spatial resolution is 1 x 1 km.

### Representativity in time and space
The data includes the last normal period (1961-1990) which defines the reference condition for climate variables. Therefore the temporal resolution is very good. Also considering tha daily resolution of the data.

Spatially, a 1x1 km resolution is sufficient for most climate variables, esp. in homogeneous terrain, but this needs to be evaluation for each variable and scenario specifically.

### Original units
Varied

### Temporal coverage
1957 - present

### Aditional comments about the dataset
The data format has changes from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.

## Ecosystem characteristic
### Norwegain standard
These variables typically will fall under the _abiotiske egenskaper_ class.

### SEEA EA
In SEEA EA, these variables will typically fall under A1 - Physical state characteristics.

## Collinearities with other indicators
Climate variables are most likely to be correlated with each other (e.g. temperature and snow). Also, some climate variables are better classed as pressure indicators, and these might have a causal association with several condition indicators.

## Reference condition and values
### Reference condition
The reference condition for climate variables is defined as the laste normal period 1961-1990.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values

* Un-scaled indicator value = median value over last 5 year (moving window)

* Upper reference level (best possible condition) = median value from the reference period

* Threshold for good ecosystem condition = 2 standard deviation units for the climate variable during  the reference period.

* Lower reference value = 5 standard deviation units for the climate variable during  the reference period.

## Uncertainties
For the indicator map (1 x 1 km raster) there is no uncertainty associated with the indicator values. For aggregated indicator values (e.g. for regions), the uncertainty in the indicator value is calculated from the spatial variation in the indicator values via bootstrapping.

## References

https://senorge.no/

## Analyses
### Data sets
<!-- Text here -->

#### Data set A
<!-- Text here -->

#### Data set B
<!-- Text here -->

#### Regions
<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
Importing a shape file with the regional delineation. 
```{r}
#reg <- sf::st_read("data/regions.shp")
#st_crs(reg)
```

### Scaled indicator values
<!-- Text and analyses here -->

### Uncertainty
<!-- Text here -->

## Prepare export
<!-- Text here -->

### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-02-22 DRAFT_climate_indicators.Rmd
conseptual workflow

@@
 
-_Author and date:_
-<!-- Write your name here -->
+*Author and date:* <!-- Write your name here -->
 
@@
 library(lubridate)
+
+dir <- substr(getwd(), 1,2)
+
 ```
@@
 
-<br />
-<br />
+<br /> <br />
+
 <hr />
@@
 ## Introduction
+
 This chaters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
@@
 ## About the underlying data
-The data is in a raster format and extends back to 1957 in the form of multiple interpolated climate variables.
-The spatial resolution is 1 x 1 km.
+
+The data is in a raster format and extends back to 1957 in the form of multiple interpolated climate variables. The spatial resolution is 1 x 1 km.
 
 ### Representativity in time and space
+
 The data includes the last normal period (1961-1990) which defines the reference condition for climate variables.
@@
 ### Original units
+
 Varied
@@
 ### Temporal coverage
+
 1957 - present
@@
 ### Aditional comments about the dataset
+
 The data format has changes from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.
@@
 ## Ecosystem characteristic
+
 ### Norwegain standard
-These variables typically will fall under the _abiotiske egenskaper_ class.
+
+These variables typically will fall under the *abiotiske egenskaper* class.
 
 ### SEEA EA
+
 In SEEA EA, these variables will typically fall under A1 - Physical state characteristics.
@@
 ## Collinearities with other indicators
+
 Climate variables are most likely to be correlated with each other (e.g.
@@
 ## Reference condition and values
+
 ### Reference condition
+
 The reference condition for climate variables is defined as the laste normal period 1961-1990.
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
 
-* Un-scaled indicator value = median value over last 5 year (moving window)
+-   Un-scaled indicator value = median value over last 5 year (moving window)
 
-* Upper reference level (best possible condition) = median value from the reference period
+-   Upper reference level (best possible condition) = median value from the reference period
 
-* Threshold for good ecosystem condition = 2 standard deviation units for the climate variable during  the reference period.
+-   Threshold for good ecosystem condition = 2 standard deviation units for the climate variable during the reference period.
 
-* Lower reference value = 5 standard deviation units for the climate variable during  the reference period.
+-   Lower reference value = 5 standard deviation units for the climate variable during the reference period.
 
 ## Uncertainties
+
 For the indicator map (1 x 1 km raster) there is no uncertainty associated with the indicator values.
@@
 
-https://senorge.no/
+<https://senorge.no/>
+
+rr and tm are being download from <https://thredds.met.no/thredds/catalog/senorge/seNorge_2018/Archive/catalog.html>
 
 ## Analyses
-### Data sets
-<!-- Text here -->
 
-#### Data set A
-<!-- Text here -->
+### Data set
 
-#### Data set B
-<!-- Text here -->
+The data is downloaded to a local NINA server, and updated regularly.
+
+```{r}
+path <- ifelse(dir == "C:", 
+      "R:/GeoSpatialData/Meteorology/Norway_SeNorge2018_v22.09/Original",
+      "/data/R/GeoSpatialData/Meteorology/Norway_SeNorge2018_v22.09/Original")
+```
+
+This folder contains folder for the different parameters
+
+```{r}
+(files <- list.files(path))
+```
+
+This table explains them in more detail
+
+```{r}
+senorgelayers <- read_delim("data/senorgelayers.txt", 
+    delim = "\t", escape_double = FALSE, 
+    trim_ws = TRUE)
+DT::datatable(senorgelayers)
+```
+
+Here is the content of one of these folders (first 10 entries)
+
+```{r}
+(files_tm <- list.files(paste0(path, "/rr_tm")))[1:10]
+```
+
+There are `r length(files_tm)` files, i.e. `r length(files_tm)` years of data, one file per year.
+
+Importing (proxy) file:
+
+```{r, warning=F}
+tg_1957 <- stars::read_ncdf(paste0(path, "/rr_tm/seNorge2018_1957.nc"))
+```
+
+I think I know the CRS, so setting it manually.
+
+```{r}
+st_crs(tg_1957) <- 25833
+```
+
+The data has three dimension
+
+```{r}
+dim(tg_1957)
+```
+
+... and four attributes
+
+```{r}
+names(tg_1957)
+```
+
+Precipitation (rr), mean temperature (tg), max temperatur (tx) and min temperature (tn) are included in the same file.
+
+For this example workflow I will only focus on tg (temperature). I could subset the data like this
+
+```{r}
+tg_1957 %>% select(tg)
+```
+
+, but since this is a proxy object, it will not initiate any change untill later, when I write to file and all the lazy operations are done in squeal.
+Therefore, I will import a test data set, which is smaller and which I can import to memory.
+Then I can perform the operations on that data set and we can see the results.
+
+### Dummy data
+
+This test data is included in the {stars} package
+
+```{r}
+tif = system.file("tif/L7_ETMs.tif", package = "stars")
+t1 = as.Date("2018-07-31")
+x = read_stars(c(tif, tif, tif, tif), along = 
+                  list(time = c(t1, t1+1, t1+2, t1+32)), 
+               RasterIO = list(nXOff = c(1), 
+                               nYOff = c(1), 
+                               nXSize = 50, 
+                               nYSize = 50, 
+                               bands = c(1:6)))
+```
+
+A single attribute
+
+```{r}
+names(x)
+```
+
+, and four dimensions
+
+```{r}
+dim(x)
+```
+
+X and y area the coordinates. Band is
+
+```{r}
+st_get_dimension_values(x, "band")
+```
+
+an integer.
+
+And time is
+
+```{r}
+st_get_dimension_values(x, "time")
+```
+
+Four dates covering three months in 2018.
+
+### Workflow
+
+The workflow is as such:
+
+1.  Calculate reference values
+
+    -   Import correct .nc file (loop though year 1961-1990) and subset to the correct attribute (above)
+
+    -   Filter data by dates (optional)
+
+    -   Aggregate across time within a year
+
+    -   Aggregate across years to get median and sd values
+
+    -   Intersect with accounting area polygons
+
+2.  Calculate variable values
+
+    -   Import correct .nc file (loop though the years 1961 to present) and subset to the correct attribute (above)
+
+    -   Repeat the steps from when calculating the reference value, except for calculating the sd.
+
+    -   Add layers to data cube already containing reference values
+
+3.  Normalize climate variable at the individual grid cell level using the three reference values
+
+4.  Write to file (for use with other ecosystem types)
+
+5.  Mask by ecosystem type
+
+6.  Aggregate in space (to accounting areas)
+
+    -   Aggregate across 5 year time steps to smooth out random interannual variation and leav climate signal
+
+    -   Intersect with accounting area polygons
+
+    -   Bootstrap indicator values (cell values) and get median and sd
+
+7.  Make trend figure
+
+### Filter by dates
 
 #### Regions
-<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
+
 Importing a shape file with the regional delineation.
+
 ```{r}
-#reg <- sf::st_read("data/regions.shp")
+reg <- sf::st_read("data/regions.shp")
 #st_crs(reg)
@@
 ### Scaled indicator values
+
 <!-- Text and analyses here -->
@@
 ### Uncertainty
+
 <!-- Text here -->

### 2023-02-22 DRAFT_climate_indicators.Rmd
mean per year

@@
 
+### Additional resources
+
+[Stars package](https://r-spatial.github.io/stars/)
+
+[R as a GIS for economists](https://tmieno2.github.io/R-as-GIS-for-Economists/stars-basics.html) chapter 7
+
 ## Analyses
@@
 tif = system.file("tif/L7_ETMs.tif", package = "stars")
-t1 = as.Date("2018-07-31")
+t1 = as.Date("2018-05-31")
 x = read_stars(c(tif, tif, tif, tif), along = 
-                  list(time = c(t1, t1+1, t1+2, t1+32)), 
+                  list(time = c(t1, t1+1, t1+50, t1+100)), 
                RasterIO = list(nXOff = c(1), 
@@
 
-, and four dimensions
+I can rename it like this
+```{r}
+x <- setNames(x, "Attribute_A")
+```
+
+And I can add another dummy attribute.
+
+```{r}
+x <- x %>%
+  mutate("Attribute_B" = Attribute_A/2)
+```
+
+The dummy data also has four dimensions
 
@@
 
-X and y area the coordinates. Band is
+X and y area the coordinates. Band is an integer:
 
@@
 
-an integer.
-
-And time is
+And time is four dates covering four months in 2018:
 
@@
 
-Four dates covering three months in 2018.
+#### Regions
+
+Importing a shape file with the regional delineation.
+
+```{r}
+reg <- sf::st_read("data/regions.shp")
+#st_crs(reg)
+```
+
+#### Ecosystem map
+
+Coming soon ....
 
@@
 
-The workflow is as such:
+The general workflow is like this:
 
@@
 
-### Filter by dates
+I will try to use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up. Also, masking to ecosystem time earlier in the workflow might prove smart.
 
-#### Regions
+### Calculate reference values
+#### Import
 
-Importing a shape file with the regional delineation.
+I need to code a for loop here for importing the .nc files.
+
+But for then selecting the attribute, simply do like this:
 
 ```{r}
-reg <- sf::st_read("data/regions.shp")
-#st_crs(reg)
+names(x)
 ```
+```{r}
+x_sub <- x %>% 
+  select(Attribute_A)
+names(x_sub)
+```
+
+#### Filter data by dates
+
+Say we want to calculate the mean summer temperature. We then want to exclude the data for the times that is not withing our definition of summer.
+
+Let's try it on the dummy data. Remember this data had three time steps.
+```{r}
+st_get_dimension_values(x, "time")
+```
+
+Our data has 365 time steps for each year.
+
+First I define my start and end dates. I want to keep only the summer data, defined as jun - aug.
+
+```{r}
+start <-  ymd("2018-06-01")
+end <-  ymd("2018-08-31")
+summer <- interval(start, end)
+```
+
+Then I filter
+
+```{r}
+x_aug <- x_sub %>%
+  filter(time %within% summer)
+st_get_dimension_values(x_aug, "time")
+```
+
+#### Aggregate across time within a year
+
+For this example I want to calculate the mean temperature over the summer.
+I therefore need to aggregate over time. Here is how:
+
+```{r}
+x_summerMean <- x_aug %>%
+  filter(band ==1) %>% # for our real data we don't need this filter
+  st_apply(1:2, mean)
+
+dim(x_summerMean)
+```
+
+The time (and band) dimension is now gone as we have aggregated accoss it.
+
+We can also use the aggregate function to produce a data cube with lower temporal resolution, for example per year:
+
+```{r}
+x_summerMean_v2 <- x_aug %>%
+  aggregate(by = "year", mean)
+```
+
+Actually, this second method is better because we don't ned to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name _mean_).
 

### 2023-02-22 DRAFT_climate_indicators.Rmd
plot time aggregation results

@@
 library(lubridate)
-
+library(ggpubr)
 dir <- substr(getwd(), 1,2)
@@
 
+Let me introduce some differences to Attribute_A between these two dates.
+```{r}
+x_aug$Attribute_A[,,,2] <- x_aug$Attribute_A[,,,1]*4
+```
+
 #### Aggregate across time within a year
@@
 
+```{r, fig.cap="Plotting the dummy data showing Attribute_A for the two dates as small maps, and a larger map showing the mean for year 2018"}
+ggarrange(
+ggplot() + 
+  geom_stars(data = x_aug) +
+  coord_equal() +
+  facet_wrap(~time) +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0)),
+
+ggplot() + 
+  geom_stars(data = x_summerMean_v2) +
+  coord_equal() +
+  facet_wrap(~time) +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))
+)
+```
+
 ### Scaled indicator values

### 2023-02-22 DRAFT_climate_indicators.Rmd
trying to normalise

@@
 tif = system.file("tif/L7_ETMs.tif", package = "stars")
-t1 = as.Date("2018-05-31")
+t1 = as.Date("1970-05-31")
 x = read_stars(c(tif, tif, tif, tif), along = 
@@
 I can rename it like this
+
 ```{r}
@@
 
-And time is four dates covering four months in 2018:
+And time is four dates covering four months in 1970:
 
@@
 ### Calculate reference values
+
 #### Import
@@
 ```
+
 ```{r}
@@
 Let's try it on the dummy data. Remember this data had three time steps.
+
 ```{r}
@@
 ```{r}
-start <-  ymd("2018-06-01")
-end <-  ymd("2018-08-31")
+start_month <-  "06"
+end_month <- "08"
+year_temp <- year(st_get_dimension_values(x, "time")[1])
+start <- ym(paste(year_temp, start_month, sep="-"))
+end <-  ym(paste(year_temp, end_month, sep="-"))
 summer <- interval(start, end)
@@
 Let me introduce some differences to Attribute_A between these two dates.
+
 ```{r}
@@
 
-For this example I want to calculate the mean temperature over the summer.
-I therefore need to aggregate over time. Here is how:
+For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. Here is how.
+
+*Method 1:*
 
@@
 
-The time (and band) dimension is now gone as we have aggregated accoss it.
+The time (and band) dimension is now gone as we have aggregated across it.
+
+*Method 2:*
 
@@
 
-Actually, this second method is better because we don't ned to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name _mean_).
+Actually, this second method is better because we don't need to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name *mean*).
 
@@
 
+#### Aggregate across years
+
+For this step we need some more dummy data from different years.
+
+```{r}
+oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
+
+y1 <- x_summerMean_v2
+y2 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3)) 
+y3 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
+y4 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
+y5 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
+
+# We combine the data into one cube:
+y <- c(y1, y2, y3, y4, y5)
+```
+
+```{r}
+ggplot() + 
+  geom_stars(data = y) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~time)
+```
+
+```{r, eval=F, include=F}
+# I would normally do the above code chink using tmap, but I cant get it to work when faceting across dimensions:
+tm_shape(y)+
+  tm_raster()+
+  tm_facets("time")
+```
+
+Now we aggregate across years.
+For the reference period is is probably best to use *Method 1*, and for he indicator variable (the moving window estimation) probably *Method 2* is best.
+So Method 1 now.
+
+```{r}
+median_sd <- function(x) { c(median = median(x), sd = sd(x))}
+```
+
+```{r}
+y_ref <- y %>%
+  filter(band==1) %>%
+  st_apply(c("x", "y"), FUN =  median_sd)
+```
+
+```{r}
+dim(y_ref)
+```
+
+```{r}
+st_get_dimension_values(y_ref, "median_sd")
+```
+
+It's easier if median and sd are unique attributes, rather than levels of a dimension.
+
+```{r}
+y_ref <- y_ref %>% split("median_sd")
+```
+
+The attribute name *mean* we can change this to something more meaningful.
+
+```{r}
+y_ref <- setNames(y_ref, c("reference_value", "sd"))
+```
+
+```{r}
+tmap_arrange(
+tm_shape(y_ref)+
+  tm_raster("reference_value")
+,
+tm_shape(y_ref)+
+  tm_raster("sd",
+            palette = "-viridis")
+)
+```
+
+#### Calculate variable values
+
+The climate variable is the same as y1, y2, y~*n*~, ....
+But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two (pluss some error)
+
+```{r}
+y1v <- y1 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
+y2v <- y2 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
+y3v <- y3 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
+y4v <- y4 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
+y5v <- y5 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop()
+
+#y_v <- c(y1v, y2v, y3v, y4v, y5v, y_ref)
+
+y_v <- c(y1v, select(y_ref, reference_value))
+y_v_merged <- merge(y_v)
+dim(y_v_merged)
+names(y_v_merged)
+y_v_merged <- setNames(y_v_merged, "value")
+
+st_get_dimension_values(y_v_merged, "attributes")
+```
+
+#### Normalise variable
+
+```{r}
+system.time(
+  temp <- st_apply(y_v_merged, c("x", "y"), FUN = function(Attribute_A,reference_value) Attribute_A/reference_value)
+)
+```
+
+```{r}
+tm_shape(temp)+
+  tm_raster("value")
+```
+
+```{r}
+
+y_scaled <- merge(y_v)
+dim(y_scaled)
+names(y_scaled)
+```
+
+```{r}
+tm_shape(y_v_scaled)+
+  tm_raster()
+```
+
 ### Scaled indicator values

### 2023-02-23 DRAFT_climate_indicators.Rmd
normalising linearly

@@
 The climate variable is the same as y1, y2, y~*n*~, ....
-But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two (pluss some error)
+But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two (plus some error).
+Actually, just leaving the reference layer and the variable layer perfectly correlated causes an error when dividing the dimension using st_apply.
 
 ```{r}
-y1v <- y1 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
-y2v <- y2 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
-y3v <- y3 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
-y4v <- y4 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop() 
-y5v <- y5 %>% mutate(Attribute_A = Attribute_A*rnorm(1,.5,.3)) %>% filter(band == 1) %>% adrop()
+
+myRandom <- function(x) x*rnorm(1,1,.1)/2
+myRandom(1)
+
+y1v <- y1 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
+y2v <- y2 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+y3v <- y3 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+y4v <- y4 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+y5v <- y5 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
 
@@
 
+```{r}
+ggplot() + 
+  geom_stars(data = y_v_merged) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~attributes)
+```
+
 #### Normalise variable 
+I now want to to cell-by-cell computations across the _attributes_ dimension, and divide _Attribute_A_ by _reference_value_.
 
 ```{r}
+myFUN <- function(a,b) (a/b)
 system.time(
-  temp <- st_apply(y_v_merged, c("x", "y"), FUN = function(Attribute_A,reference_value) Attribute_A/reference_value)
+  temp <- st_apply(y_v_merged, 
+                   c("x", "y"), 
+                   #1:2,
+                   FUN = myFUN
   )
+)
+
 ```
@@
 tm_shape(temp)+
-  tm_raster("value")
+  tm_raster()+
+  tm_layout(legend.outside = T)
+```
+This looks like perfect noise, and it is., because the variable and reference value are so well correlated.
+If the reference value was static, it would look different.
+
+```{r}
+y_v_merged$value[,,2] <- 200
+```
+
+```{r}
+ggplot() + 
+  geom_stars(data = y_v_merged) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~attributes)
+```
+
+```{r}
+system.time(
+  temp <- st_apply(y_v_merged, 
+                   c("x", "y"), 
+                   #1:2,
+                   FUN = myFUN
+  )
+)
+
+```
+
+```{r}
+tm_shape(temp)+
+  tm_raster()+
+  tm_layout(legend.outside = T)
 ```

### 2023-02-23 DRAFT_climate_indicators.Rmd
add noise earlier in workflow

@@
 
+Firt, let me cut the _band_ dimension.
+```{r}
+temp <- x_summerMean_v2 %>%
+  filter(band==1) %>%
+  adrop(drop=4) %>%
+  aperm(c(2,3,1))
+dim(temp)
+```
+
 ```{r}
@@
 
-y1 <- x_summerMean_v2
-y2 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3)) 
-y3 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
-y4 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
-y5 <- x_summerMean_v2 %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*rnorm(1,1,.3))
+# Function to add random noise
+myRandom <- function(x) x*rnorm(1,1,.1)
+
+y1 <- temp
+y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+
+dim(y1)
+dim(y2)
 
@@
 y_ref <- y %>%
-  filter(band==1) %>%
   st_apply(c("x", "y"), FUN =  median_sd)
@@
 
-It's easier if median and sd are unique attributes, rather than levels of a dimension.
+It's perhaps easier if median and sd are unique attributes, rather than levels of a dimension.
 
@@
 The climate variable is the same as y1, y2, y~*n*~, ....
-But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two (plus some error).
-Actually, just leaving the reference layer and the variable layer perfectly correlated causes an error when dividing the dimension using st_apply.
+But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two.
 
 ```{r}
-
-myRandom <- function(x) x*rnorm(1,1,.1)/2
-myRandom(1)
-
-y1v <- y1 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
-y2v <- y2 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-y3v <- y3 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-y4v <- y4 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-y5v <- y5 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
-
+#y1v <- y1 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
+#y2v <- y2 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+#y3v <- y3 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+#y4v <- y4 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
+#y5v <- y5 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
 #y_v <- c(y1v, y2v, y3v, y4v, y5v, y_ref)
 
+y1v <- y1 %>% adrop() %>% mutate(Attribute_A = Attribute_A/2)
 y_v <- c(y1v, select(y_ref, reference_value))
@@
 ```{r}
+# A function for normalization
+# I need to add a break point and lower reference value here as well.
+# And I need to add truncation.
 myFUN <- function(a,b) (a/b)
+```
+
+```{r}
 system.time(
@@
                    c("x", "y"),
-                   #1:2,
                    FUN = myFUN
@@
 
-```{r}
+```{r, fig.cap"The normalised indicator for year 1970"}
 tm_shape(temp)+
@@
 ```
-This looks like perfect noise, and it is., because the variable and reference value are so well correlated.
+This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
 If the reference value was static, it would look different.
@@
 ```{r}
-system.time(
 temp <- st_apply(y_v_merged, 
@@
 )
-)
-
 ```
@@
   tm_layout(legend.outside = T)
-```
-
-```{r}
-
-y_scaled <- merge(y_v)
-dim(y_scaled)
-names(y_scaled)
-```
-
-```{r}
-tm_shape(y_v_scaled)+
-  tm_raster()
 ```

### 2023-02-23 DRAFT_climate_indicators.Rmd
normalised across all columns

@@
 
+#### Export
+I should export the reference data at this point (write to file) as we don't need to recalculate this every time we want to update the indicator values.
+
 #### Calculate variable values
@@
 But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two.
+I then merge with the reference value and sd, creating one common data cube.
 
 ```{r}
-#y1v <- y1 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom)
-#y2v <- y2 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-#y3v <- y3 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-#y4v <- y4 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-#y5v <- y5 %>% filter(band == 1) %>% adrop() %>% st_apply(1:2, FUN = myRandom) 
-#y_v <- c(y1v, y2v, y3v, y4v, y5v, y_ref)
-
-y1v <- y1 %>% adrop() %>% mutate(Attribute_A = Attribute_A/2)
-y_v <- c(y1v, select(y_ref, reference_value))
-y_v_merged <- merge(y_v)
-dim(y_v_merged)
-names(y_v_merged)
-y_v_merged <- setNames(y_v_merged, "value")
-
-st_get_dimension_values(y_v_merged, "attributes")
+y_var <- y %>% mutate(Attribute_A = Attribute_A/2) %>% 
+  split(3) %>% c(y_ref)
+y_var <- y_var %>%
+  setNames(paste0("v_",substr(names(y_var), 1,4)))%>%
+  rename("reference" = v_refe,
+         "sd" = v_sd)
 ```
@@
 ```{r}
+# select the columns to normalise
+cols <- names(y_var)[!names(y_var) %in% c("reference", "sd") ]
+cols_new <- cols
+names(cols_new) <- gsub("v_", "i_", cols)
+# Mutate 
+y_var_norm <- y_var %>%
+  mutate(across(all_of(cols), ~.x / reference)) %>%
+  rename(all_of(cols_new))
+
+#y_v <- c(y1v, select(y_ref, reference_value))
+#y_v_merged <- merge(y_v)
+#dim(y_v_merged)
+#names(y_v_merged)
+#y_v_merged <- setNames(y_v_merged, "value")
+#st_get_dimension_values(y_v_merged, "attributes")
+```
+
+```{r}
+ggarrange(
 ggplot() + 
-  geom_stars(data = y_v_merged) +
+  geom_stars(data = y_var["v_1970"]) +
   coord_equal() +
@@
   scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~attributes)
+  scale_y_discrete(expand = c(0, 0))
+,
+ggplot() + 
+  geom_stars(data = y_var_norm["reference"]) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))
+,
+ggplot() + 
+  geom_stars(data = y_var_norm["i_1970"]) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))
+, ncol=3
+)
 ```
@@
 #### Normalise variable 
-I now want to to cell-by-cell computations across the _attributes_ dimension, and divide _Attribute_A_ by _reference_value_.
+I now want to do cell-by-cell computations across the _attributes_ dimension, and divide _Attribute_A_ by _reference_value_.
 
@@
 ```
+This operation is quite fast!
 
-```{r, fig.cap"The normalised indicator for year 1970"}
+```{r, fig.cap = "The normalised indicator for year 1970."}
 tm_shape(temp)+
@@
 
-```{r}
+```{r, fig.cap = "The normalised indicator for year 1970 when the reference value is the same all over."}
 tm_shape(temp)+

### 2023-02-23 DRAFT_climate_indicators.Rmd
break points scaling

@@
 # Function to add random noise
-myRandom <- function(x) x*rnorm(1,1,.1)
+myRandom <- function(x) x*rnorm(1,1,.05)
 
@@
 y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-
-dim(y1)
-dim(y2)
 
@@
   facet_wrap(~time)
-```
-
-```{r, eval=F, include=F}
-# I would normally do the above code chink using tmap, but I cant get it to work when faceting across dimensions:
-tm_shape(y)+
-  tm_raster()+
-  tm_facets("time")
 ```
@@
 The climate variable is the same as y1, y2, y~*n*~, ....
-But to create some differences between the variable and the reference values, I will tweak these slightly, more or less divide them by two.
+But to create some differences between the variable and the reference values, I will tweak these slightly, taking them down 2%.
 I then merge with the reference value and sd, creating one common data cube.
@@
 ```{r}
-y_var <- y %>% mutate(Attribute_A = Attribute_A/2) %>% 
+y_var <- y %>% mutate(Attribute_A = Attribute_A * 0.98) %>% 
   split(3) %>% c(y_ref)
@@
 names(cols_new) <- gsub("v_", "i_", cols)
+
 # Mutate
+# can I find a way to keep the v_ columns?
 y_var_norm <- y_var %>%
-  mutate(across(all_of(cols), ~.x / reference)) %>%
+  mutate(lower = reference - 5*sd ) %>%
+  mutate(threshold = reference -2*sd ) %>%
+  mutate(across(all_of(cols), ~ if_else(.x < threshold, 
+                                        (.x - lower) / (threshold - lower),
+                                        (.x - threshold) / (reference - threshold),
+                                        ))) %>%
+  mutate(across(all_of(cols), ~ if_else(.x > 1, 1, .x))) %>%
+  mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
   rename(all_of(cols_new))
-
-#y_v <- c(y1v, select(y_ref, reference_value))
-#y_v_merged <- merge(y_v)
-#dim(y_v_merged)
-#names(y_v_merged)
-#y_v_merged <- setNames(y_v_merged, "value")
-#st_get_dimension_values(y_v_merged, "attributes")
 ```

### 2023-02-24 DRAFT_climate_indicators.Rmd
microbenchmark

@@
 library(ggpubr)
+library(microbenchmark)
 dir <- substr(getwd(), 1,2)
@@
 
-For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. Here is how.
+For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways.
 
@@
 ```{r}
-x_summerMean <- x_aug %>%
+microOut <- microbenchmark(
+st_apply =  x_aug %>%
   filter(band ==1) %>% # for our real data we don't need this filter
-  st_apply(1:2, mean)
-
-dim(x_summerMean)
+  st_apply(1:2, mean, rename=F),
+aggregate = x_aug %>%
+  aggregate(by = "year", mean),
+times=30
+)
 ```
@@
 
-*Method 2:*
+```{r, fig.cap="Comparing computation tome for two spatial aggregation functions."}
+autoplot(microOut)
+```
 
-We can also use the aggregate function to produce a data cube with lower temporal resolution, for example per year:
+`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension. 
+I think I will try to use `aggregate`, and it it takes too long on the real data, then I can come back to this point, and perhaps switch to st_apply.
 
 ```{r}
-x_summerMean_v2 <- x_aug %>%
-  aggregate(by = "year", mean)
+x_summerMean <-   x_aug %>%
+  filter(band ==1) %>% 
+  st_apply(1:2, mean, rename=F)
 ```
-
-Actually, this second method is better because we don't need to filter out additional dimensions, and it retains the attribute name (the first method replaced it with the function name *mean*).
 
@@
 
-Firt, let me cut the _band_ dimension.
+First, let me cut the _band_ dimension.
 ```{r}

### 2023-02-24 DRAFT_climate_indicators.Rmd
switch to st_apply

@@
 ```{r}
+temp_names <- year(st_get_dimension_values(x_aug, "time")[1])
+
 microOut <- microbenchmark(
@@
   filter(band ==1) %>% # for our real data we don't need this filter
-  st_apply(1:2, mean, rename=F),
+  st_apply(1:2, mean) %>%
+  setNames(paste0("v_",temp_names)),
 aggregate = x_aug %>%
@@
 
-`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension. 
-I think I will try to use `aggregate`, and it it takes too long on the real data, then I can come back to this point, and perhaps switch to st_apply.
+`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year.
+I will try to use `st_apply`.
 
 ```{r}
+temp_names <- year(st_get_dimension_values(x_aug, "time")[1])
 x_summerMean <-   x_aug %>%
   filter(band ==1) %>% 
-  st_apply(1:2, mean, rename=F)
+  st_apply(1:2, mean) %>%
+  setNames(paste0("v_",temp_names))
 ```
@@
 ggplot() + 
-  geom_stars(data = x_summerMean_v2) +
+  geom_stars(data = x_summerMean) +
   coord_equal() +
-  facet_wrap(~time) +
+  #facet_wrap(~time) +
   theme_void() +
@@
 ```{r}
-temp <- x_summerMean_v2 %>%
-  filter(band==1) %>%
-  adrop(drop=4) %>%
-  aperm(c(2,3,1))
-dim(temp)
+#temp <- x_summerMean %>%
+#  filter(band==1) %>%
+#  adrop(drop=4) %>%
+#  aperm(c(2,3,1))
+#dim(temp)
 ```
@@
 ```{r}
-oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
+#oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
+#
+## Function to add random noise
+#myRandom <- function(x) x*rnorm(1,1,.05)
+#
+#y1 <- temp
+#y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+#y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+#y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+#y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
+#
+#
+## We combine the data into one cube:
+#y <- c(y1, y2, y3, y4, y5)
+```
 
+```{r}
 # Function to add random noise
 myRandom <- function(x) x*rnorm(1,1,.05)
+```
 
-y1 <- temp
-y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-
+```{r}
+y1 <- x_summerMean
+y2 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-1)) 
+y3 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-2)) 
+y4 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-3)) 
+y5 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-4)) 
 # We combine the data into one cube:
-y <- c(y1, y2, y3, y4, y5)
+y <- c(y1, y2, y3, y4, y5) %>% 
+  merge(name = "Year") %>%
+  setNames("Attribute_A")
 ```
@@
   scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~time)
+  facet_wrap(~Year)
 ```
 
-Now we aggregate across years.
-For the reference period is is probably best to use *Method 1*, and for he indicator variable (the moving window estimation) probably *Method 2* is best.
-So Method 1 now.
+Now we aggregate across years, again using st_apply.
 
@@
 ```{r}
-y_ref <- setNames(y_ref, c("reference_value", "sd"))
+y_ref <- setNames(y_ref, c("reference_upper", "sd"))
 ```
@@
 tm_shape(y_ref)+
-  tm_raster("reference_value")
+  tm_raster("reference_upper")
 ,
@@
   split(3) %>% c(y_ref)
-y_var <- y_var %>%
-  setNames(paste0("v_",substr(names(y_var), 1,4)))%>%
-  rename("reference" = v_refe,
-         "sd" = v_sd)
+```
+
+#### Normalise variable
+
+```{r}
+# select the columns to normalise
+cols <- names(y_var)[!names(y_var) %in% c("reference_upper", "sd") ]
+cols_new <- cols
+names(cols_new) <- gsub("v_", "i_", cols)
 ```
@@
 ```{r}
-# select the columns to normalise
-cols <- names(y_var)[!names(y_var) %in% c("reference", "sd") ]
-cols_new <- cols
-names(cols_new) <- gsub("v_", "i_", cols)
-
 # Mutate
 # can I find a way to keep the v_ columns?
+system.time(
 y_var_norm <- y_var %>%
-  mutate(lower = reference - 5*sd ) %>%
-  mutate(threshold = reference -2*sd ) %>%
+  mutate(lower = reference_upper - 5*sd ) %>%
+  mutate(threshold = reference_upper -2*sd ) %>%
   mutate(across(all_of(cols), ~ if_else(.x < threshold, 
                                         (.x - lower) / (threshold - lower),
-                                        (.x - threshold) / (reference - threshold),
+                                        (.x - threshold) / (reference_upper - threshold),
                                         ))) %>%
@@
   mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
-  rename(all_of(cols_new))
+  rename(all_of(cols_new)) %>%
+  c(select(y_var, all_of(cols)))
+)
 ```
+
+I'm not sure mutate-across is the fastest way to do this. Perhaps st_apply would be better.
 
@@
 ggplot() + 
-  geom_stars(data = y_var["v_1970"]) +
+  geom_stars(data = y_var_norm["v_1970"]) +
   coord_equal() +
@@
 ggplot() + 
-  geom_stars(data = y_var_norm["reference"]) +
+  geom_stars(data = y_var_norm["reference_upper"]) +
   coord_equal() +
@@
   scale_y_discrete(expand = c(0, 0))
-, ncol=3
+, ncol=3, align = "hv"
 )
@@
 
-#### Normalise variable 
-I now want to do cell-by-cell computations across the _attributes_ dimension, and divide _Attribute_A_ by _reference_value_.
+This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
+If the reference value was static, or if there was a pattern in v_1970, the scaled indicator map would look different.
+
+#### Plot time series
 
 ```{r}
-# A function for normalization
-# I need to add a break point and lower reference value here as well.
-# And I need to add truncation.
-myFUN <- function(a,b) (a/b)
-```
-
-```{r}
-system.time(
-  temp <- st_apply(y_v_merged, 
-                   c("x", "y"),
-                   FUN = myFUN
-  )
-)
-
-```
-This operation is quite fast!
-
-```{r, fig.cap = "The normalised indicator for year 1970."}
-tm_shape(temp)+
-  tm_raster()+
-  tm_layout(legend.outside = T)
-```
-This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
-If the reference value was static, it would look different.
-
-```{r}
-y_v_merged$value[,,2] <- 200
-```
-
-```{r}
-ggplot() + 
-  geom_stars(data = y_v_merged) +
-  coord_equal() +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +  
-  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~attributes)
-```
-
-```{r}
-temp <- st_apply(y_v_merged, 
-                   c("x", "y"), 
-                   #1:2,
-                   FUN = myFUN
-)
-```
-
-```{r, fig.cap = "The normalised indicator for year 1970 when the reference value is the same all over."}
-tm_shape(temp)+
-  tm_raster()+
-  tm_layout(legend.outside = T)
+ts <- y_var_norm %>%
+  select(all_of(starts_with("i_"))) %>%
+  st_apply(MARGIN = 3, mean)
 ```

### 2023-02-24 DRAFT_climate_indicators.Rmd
prepare data for plotting

@@
 ```{r}
-ts <- y_var_norm %>%
+  ts1 <- y_var_norm %>%
   select(all_of(starts_with("i_"))) %>%
-  st_apply(MARGIN = 3, mean)
+  st_redimension() %>%
+  st_apply(3, median_sd) %>%
+  as_tibble() %>%
+  setNames(c("metric", "i_year", "value") ) %>%
+  mutate(year = as.numeric(substr(i_year, 3, 7))) %>%
+  pivot_wider(names_from = "metric") %>%
+  select(-i_year) %>%
+  add_column(type = "indicator_value") %>%
+  rename(value = median) %>%
+  add_column(metric = "median")
+```
+
+```{r}
+  ts2 <- y_var_norm %>%
+  select(all_of(starts_with("v_"))) %>%
+  st_redimension() %>%
+  st_apply(3, median_sd) %>%
+  as_tibble() %>%
+  setNames(c("metric", "v_year", "value") ) %>%
+  mutate(year = as.numeric(substr(v_year, 3, 7))) %>%
+  pivot_wider(names_from = "metric") %>%
+  select(-v_year) %>%
+  add_column(type = "variable_value") %>%
+  rename(value = median) %>%
+  add_column(metric = "median")
+```
+
+```{r}
+  ts3 <- y_var_norm %>%
+  select(!all_of(starts_with(c("v_", "i_")))) %>%
+  st_redimension() %>%
+  st_apply(3, median) %>%
+  as_tibble() %>%
+  setNames(c("type", "value")) %>%
+    add_column(metric = "median")
+```
+
+```{r}
+(ts <- bind_rows(ts1, ts2, ts3))
+```
+
+Calculate diference between indicator and reference
+We don't want to include this in ts, but as a temp file.
+```{r}
+tsX <- ts1
+```
+
+```{r}
+ts %>% filter(type == "indicator_value") %>%
+  ggplot(aes(x = year, y = value))+
+  geom_line()
 ```

### 2023-02-27 DRAFT_climate_indicators.Rmd
Fix conceptual workflow

@@
 
-And time is four dates covering four months in 1970:
+I will remove the _band_ dimension.
+```{r}
+x <- x %>% filter(band==1) %>% adrop()
+```
+
+Time is four dates covering four months in 1970:
 
 ```{r}
-st_get_dimension_values(x, "time")
+(oldTime <- st_get_dimension_values(x, "time"))
+```
+
+I'll add some extra seasonal variation between to the mix
+```{r}
+x$Attribute_A[,,2] <- x$Attribute_A[,,1]*1.2
+x$Attribute_A[,,3] <- x$Attribute_A[,,1]*1.4
+x$Attribute_A[,,4] <- x$Attribute_A[,,1]*1.1
+```
+
+Now I create another four copies of this data, adding some random noise and a continuous decreasing trend.
+
+```{r}
+# Function to add random noise
+myRandom <- function(x) x*rnorm(1,1,.05)
+```
+
+```{r}
+y1 <- x
+y2 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*0.95)
+y3 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*0.90)
+y4 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*0.85)
+y5 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*0.80)
+
+```
+
+```{r}
+# We combine the data into one cube for plotting:
+temp <- c(y1, y2, y3, y4, y5)
+
+ggplot() + 
+  geom_stars(data = temp) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~time, ncol=4)
+```
+
+Saving these to file to demonstrate the importing step of the workflow.
+
+```{r, eval=F}
+path <- "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data"
+saveRDS(y1, paste0(path, "/y1.rds"))
+saveRDS(y2, paste0(path, "/y2.rds"))
+saveRDS(y3, paste0(path, "/y3.rds"))
+saveRDS(y4, paste0(path, "/y4.rds"))
+saveRDS(y5, paste0(path, "/y5.rds"))
+```
+
+##### Dummy regions and ET map
+
+Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well.
+
+```{r}
+dummy_aa <- x %>%
+  filter(time == oldTime[1]) %>%
+  adrop() %>%
+  mutate(accuntingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
+  select(accuntingAreas)
+
+tm_shape(dummy_aa) +
+  tm_raster(style="cat")
+```
+
+```{r}
+dummy_et <- x %>%
+  filter(time == oldTime[1]) %>%
+  adrop() %>%
+  mutate(ecosystemType = rep(c(1,NA), length.out = length(Attribute_A))) %>%
+  select(ecosystemType)
+
+tm_shape(dummy_et) +
+  tm_raster(style="cat")
 ```
@@
 
-### Workflow
+### Conseptual workflow
 
-The general workflow is like this:
+The general, conceptual workflow is like this:
+
+1. Collate variable data series
+
+    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute (above)
+
+    -   Filter data by dates (optional) (`dplyr::filter`)
+
+    -   Aggregate across time within a year (`stars::st_apply`)
+
+    -   Join all data into one data cube (´stars:c´)
 
@@
 
-    -   Import correct .nc file (loop though year 1961-1990) and subset to the correct attribute (above)
+    -   Aggregate across reference years to get median and sd values `st_apply`
 
-    -   Filter data by dates (optional)
+    -   Join with existing data cube
 
-    -   Aggregate across time within a year
+2.  Calculate indicator values
 
-    -   Aggregate across years to get median and sd values
+    -   Normalize climate variable at the individual grid cell level using the three reference values `mutate %>% across`
 
-    -   Intersect with accounting area polygons
+4.  Write to file
 
-2.  Calculate variable values
-
-    -   Import correct .nc file (loop though the years 1961 to present) and subset to the correct attribute (above)
-
-    -   Repeat the steps from when calculating the reference value, except for calculating the sd.
-
-    -   Add layers to data cube already containing reference values
-
-3.  Normalize climate variable at the individual grid cell level using the three reference values
-
-4.  Write to file (for use with other ecosystem types)
-
-5.  Mask by ecosystem type
+5.  Mask by ecosystem type (this could also be done in step one to save time)
 
@@
 
-    -   Intersect with accounting area polygons
+    -   Intersect with accounting area polygons `sf:st_intersects`
 
@@
 
-### Calculate reference values
+### Collate variable data series
 
-#### Import
+I need to code a for loop for importing the .nc files.
 
-I need to code a for loop here for importing the .nc files.
-
-But for then selecting the attribute, simply do like this:
+For then selecting the attribute, simply do like this:
 
 ```{r}
+# list attribute names
 names(x)
@@
 
-Let's try it on the dummy data. Remember this data had three time steps.
+Let's try it on the dummy data. Remember this data had four time steps.
 
@@
 
-Our data has 365 time steps for each year.
+Our real data has 365 time steps for each year.
 
@@
 end_month <- "08"
-year_temp <- year(st_get_dimension_values(x, "time")[1])
+```
+
+Then for each iteration I need to get the year as well
+```{r}
+(year_temp <- year(st_get_dimension_values(x, "time")[1]))
+```
+
+Then I can create a time interval object
+```{r}
 start <- ym(paste(year_temp, start_month, sep="-"))
 end <-  ym(paste(year_temp, end_month, sep="-"))
-summer <- interval(start, end)
+(myInterval <- interval(start, end))
 ```
@@
 x_aug <- x_sub %>%
-  filter(time %within% summer)
+  filter(time %within% myInterval)
 st_get_dimension_values(x_aug, "time")
-```
-
-Let me introduce some differences to Attribute_A between these two dates.
-
-```{r}
-x_aug$Attribute_A[,,,2] <- x_aug$Attribute_A[,,,1]*4
 ```
@@
 
-For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways.
-
-*Method 1:*
+For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways, either using `st_apply` or using `aggregate`.
 
@@
 st_apply =  x_aug %>%
-  filter(band ==1) %>% # for our real data we don't need this filter
   st_apply(1:2, mean) %>%
@@
 
-The time (and band) dimension is now gone as we have aggregated across it.
+The time dimension is now gone as we have aggregated across it.
 
@@
 
-`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year.
+`st_apply` is slightly faster, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year.
 I will try to use `st_apply`.
+
+But let me see if I can save time by masking to ET before doing this operation.
+
+```{r, fig.cap="Demonstrating the effecty of maskig the dummy data using a perfectly aligne raster ET map."}
+x_masked <- x_aug
+x_masked[is.na(dummy_et)] <- NA
+plot(x_masked[,,,1])
+```
+
+```{r, fig.cap="There is no performance increas by making before aggregating"}
+autoplot(
+  microbenchmark(
+No_masking =  x_aug %>%
+  st_apply(1:2, mean) %>%
+  setNames(paste0("v_",temp_names)),
+Masking =  x_masked %>%
+  st_apply(1:2, mean) %>%
+  setNames(paste0("v_",temp_names))
+ ), times=30
+)
+
+```
 
@@
 For this step we need some more dummy data from different years.
-
-First, let me cut the _band_ dimension.
-```{r}
-#temp <- x_summerMean %>%
-#  filter(band==1) %>%
-#  adrop(drop=4) %>%
-#  aperm(c(2,3,1))
-#dim(temp)
-```
-
-```{r}
-#oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
-#
-## Function to add random noise
-#myRandom <- function(x) x*rnorm(1,1,.05)
-#
-#y1 <- temp
-#y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-#y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-#y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-#y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
-#
-#
-## We combine the data into one cube:
-#y <- c(y1, y2, y3, y4, y5)
-```
 

### 2023-02-27 DRAFT_climate_indicators.Rmd
Loop through dir

@@
 
+## --------------------------------------------------
+
 ## Analyses
@@
 
-I will remove the _band_ dimension.
+I will remove the *band* dimension.
+
 ```{r}
@@
 I'll add some extra seasonal variation between to the mix
+
 ```{r}
@@
 
+## --------------------------------------------------
+
 ### Conseptual workflow
@@
 
-    -   Join all data into one data cube (´stars:c´)
+    -   Join all data into one data cube (`stars:c`)
 
-1.  Calculate reference values
+2.  Calculate reference values
 
-    -   Aggregate across reference years to get median and sd values `st_apply`
+    -   Aggregate across reference years to get median and sd values (`st_apply)`
 
-    -   Join with existing data cube
+    -   Join with existing data cube (`stars:c`)
 
-2.  Calculate indicator values
+3.  Calculate indicator values
 
-    -   Normalize climate variable at the individual grid cell level using the three reference values `mutate %>% across`
+    -   Normalize climate variable at the individual grid cell level using the three reference values (`mutate(across()))`
 
@@
 
-    -   Aggregate across 5 year time steps to smooth out random interannual variation and leav climate signal
+    -   Aggregate across 5 year time steps to smooth out random inter-annual variation and leave climate signal
 
@@
 
-### Collate variable data series
+### 1. Collate variable data series
 
@@
 Then for each iteration I need to get the year as well
+
 ```{r}
@@
 Then I can create a time interval object
+
 ```{r}
@@
 
-```{r, fig.cap="There is no performance increas by making before aggregating"}
+```{r withMasking, fig.cap="Compareing computation time before and after masking of raster data. There is no performance increase by masking before aggregating."}
 autoplot(
@@
 
+Since there is no increase in performance from masking, as we see in Fig.
+\@ref(fig:withMasking) then it basically means we it is slower to mask beforehand, since the masking was not past of the benchmark assessment.
+
+So, this is my solution:
+
 ```{r}
-temp_names <- year(st_get_dimension_values(x_aug, "time")[1])
 x_summerMean <-   x_aug %>%
-  filter(band ==1) %>% 
   st_apply(1:2, mean) %>%
@@
 ```
+
+The *v\_* follows the naming convention in Norway.
 
@@
 
-#### Aggregate across years
+Here's the whole first step of the conceptual workflow in action, on the dummy data.
 
-For this step we need some more dummy data from different years.
+```{r}
+
+# Setting up the parameters
+dir <- substr(getwd(), 1,2)
+path <- ifelse(dir == "C:",
+       "P:/", 
+       "/data/P-Prosjekter2/")
+path2 <- paste0(path, 
+                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
+start_month <-  "06"
+end_month <- "08"
+myFiles <- list.files(path2, full.names = T)
+summerTemp_fullSeries <- NULL
+
+# Looping though the files in the directory
+system.time({
+for(i in 1:length(myFiles)){
+
+  temp <- readRDS(myFiles[i]) %>%
+    select(Attribute_A)
+
+  year_temp <- year(st_get_dimension_values(temp, "time")[1])
+  start <- ym(paste(year_temp, start_month, sep="-"))
+  end <-  ym(paste(year_temp, end_month, sep="-"))
+  myInterval <- interval(start, end)
+
+  temp <- temp %>%
+    filter(time %within% myInterval) %>%
+    st_apply(1:2, mean) %>%
+    setNames(paste0("v_",year_temp))
+
+  summerTemp_fullSeries <- c(temp, summerTemp_fullSeries)
+  rm(temp) # same computation time with and without this function, but more tidy this way
+}
+
+# Turn the attributes into a dimension and rename the new attribute
+summerTemp_fullSeries <- summerTemp_fullSeries %>%
+  merge(name="Year") %>%
+  setNames("Attribute_A")
+})
+```
+
+And this is the result.
+
+```{r}
+ggplot()+
+  geom_stars(data = summerTemp_fullSeries) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~Year)
+```
+
+### 2. Calculate reference values
+#### Aggregate across reference years
+
+For this step we need some more dummy data from different years. I made data sets y1 to y5 above.
 
@@
 #### Export
+
 I should export the reference data at this point (write to file) as we don't need to recalculate this every time we want to update the indicator values.
@@
 
-Calculate diference between indicator and reference
-We don't want to include this in ts, but as a temp file.
+Calculate diference between indicator and reference We don't want to include this in ts, but as a temp file.
+
 ```{r}

### 2023-02-27 DRAFT_climate_indicators.Rmd
ready to test real data

@@
 
-# Cilmate indicators
+# Climate indicators
 
@@
 library(microbenchmark)
+library(ncmeta)
 dir <- substr(getwd(), 1,2)
@@
 
-This chaters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
+This chapters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
 
@@
 Therefore the temporal resolution is very good.
-Also considering tha daily resolution of the data.
+Also considering the daily resolution of the data.
 
@@
 
-Varied
+Varied. Specified below for each parameter.
 
@@
 ```{r, warning=F}
-tg_1957 <- stars::read_ncdf(paste0(path, "/rr_tm/seNorge2018_1957.nc"))
+tg_1957 <- stars::read_ncdf(paste0(path, "/rr_tm/seNorge2018_1957.nc"),
+                            var="tg")
 ```
@@
 
-... and four attributes
+Initially the data had four attributes, but I subsettet to only include _tg_.
 
@@
 
-Precipitation (rr), mean temperature (tg), max temperatur (tx) and min temperature (tn) are included in the same file.
-
-For this example workflow I will only focus on tg (temperature). I could subset the data like this
-
-```{r}
-tg_1957 %>% select(tg)
-```
-
-, but since this is a proxy object, it will not initiate any change untill later, when I write to file and all the lazy operations are done in squeal.
-Therefore, I will import a test data set, which is smaller and which I can import to memory.
+`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change untill later, typically I write to file and all the lazy operations are done in squeal.
+Therefore, I will prepare a test data set, which is smaller and which I can import to memory.
 Then I can perform the operations on that data set and we can see the results.
@@
 
-```{r}
+```{r, fig.cap="Showing the example account area delineation (raster format)"}
 dummy_aa <- x %>%
@@
 
-```{r}
+```{r, fig.cap="Showing the example Ecosystem type delineation data."}
 dummy_et <- x %>%
@@
 
-I need to code a for loop for importing the .nc files.
-
-For then selecting the attribute, simply do like this:
-
-```{r}
-# list attribute names
-names(x)
-```
-
-```{r}
-x_sub <- x %>% 
-  select(Attribute_A)
-names(x_sub)
-```
+I include a for loop for importing the .nc files at the end.
 
@@
 
-```{r, fig.cap="Demonstrating the effecty of maskig the dummy data using a perfectly aligne raster ET map."}
+```{r, fig.cap="Demonstrating the effect of maskig the dummy data using a perfectly aligne raster ET map."}
 x_masked <- x_aug
@@
 
-```{r withMasking, fig.cap="Compareing computation time before and after masking of raster data. There is no performance increase by masking before aggregating."}
+```{r withMasking, fig.cap="Comparing computation time before and after masking of raster data. There is no performance increase by masking before aggregating."}
 autoplot(
@@
 Since there is no increase in performance from masking, as we see in Fig.
-\@ref(fig:withMasking) then it basically means we it is slower to mask beforehand, since the masking was not past of the benchmark assessment.
+\@ref(fig: withMasking) then it basically means we it is slower to mask beforehand, since the masking was not past of the benchmark assessment.
 
@@
   facet_wrap(~Year)
+```
+
+Now let's try it on the real data
+
+```{r}
+path <- path <- ifelse(dir == "C:", 
+      "R:/",
+      "/data/R/")
+
+path2 <- paste0(path, "GeoSpatialData/Meteorology/Norway_SeNorge2018_v22.09/Original/rr_tm/") #seNorge2018_1957.nc
+
+myFiles <- list.files(path2, full.names = T)[1:2]
+start_month <-  "06"
+end_month <- "08"
+real_temp_summer <- NULL
+
+for(i in 1:length(myFiles)){
+
+  temp <- stars::read_ncdf(paste(path2, var="tg"))
+  year_temp <- year(st_get_dimension_values(temp, "time")[1])
+  start <- ym(paste(year_temp, start_month, sep="-"))
+  end <-  ym(paste(year_temp, end_month, sep="-"))
+  myInterval <- interval(start, end)
+
+  temp <- temp %>%
+    filter(time %within% myInterval) %>%
+    st_apply(1:2, mean) %>%
+    setNames(paste0("v_",year_temp))
+
+  summerTemp_fullSeries <- c(temp, summerTemp_fullSeries)
+  rm(temp)
+}
 ```

### 2023-02-27 DRAFT_climate_indicators.Rmd
tictoc

@@
 library(microbenchmark)
+library(tictoc)
 library(ncmeta)
@@
 
+  tic("init")
   temp <- stars::read_ncdf(paste(path2, var="tg"))
@@
   myInterval <- interval(start, end)
+  toc()
 
+  tic("middle")
   temp <- temp %>%
@@
     setNames(paste0("v_",year_temp))
+  toc()
 
-  summerTemp_fullSeries <- c(temp, summerTemp_fullSeries)
+  tic("c")
+  real_temp_summer <- c(temp, real_temp_summer)
   rm(temp)
+  toc()
 }

### 2023-02-27 DRAFT_climate_indicators.Rmd
failed execution

@@
 
-```{r}
+```{r, warning=F}
 path <- path <- ifelse(dir == "C:", 
@@
   tic("init")
-  temp <- stars::read_ncdf(paste(path2, var="tg"))
+  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg")
   year_temp <- year(st_get_dimension_values(temp, "time")[1])
@@
 }
+#
+#tic("execute")
+#real_temp_summer <- st_as_stars(real_temp_summer)
+#toc()
+
+write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
+```
+
+```{r}
+dim(real_temp_summer)
+st_get_dimension_values(real_temp_summer, "time")
 ```

### 2023-02-28 DRAFT_climate_indicators.Rmd
import julian dates

--- 2023-02-27 DRAFT_climate_indicators.Rmd
@@
 ```{r}
-x_aug <- x_sub %>%
+x_aug <- x %>%
+  select(Attribute_A) %>%
+  adrop() %>%
   filter(time %within% myInterval)
@@
 ```{r}
+# setNams dont work on stars proxy obejct, so I use rename instead
+lookup <- setNames("mean", paste0("v_", temp_names))
+
 x_summerMean <-   x_aug %>%
   st_apply(1:2, mean) %>%
-  setNames(paste0("v_",temp_names))
+  rename(all_of(lookup))
 ```
@@
 end_month <- "08"
-myFiles <- list.files(path2, full.names = T)
+myFiles <- list.files(path2, pattern=".rds", full.names = T)
 summerTemp_fullSeries <- NULL
@@
   myInterval <- interval(start, end)
+  lookup <- setNames("mean", paste0("v_", year_temp))
 
@@
     st_apply(1:2, mean) %>%
-    setNames(paste0("v_",year_temp))
+    rename(all_of(lookup))
 
@@
 
-Now let's try it on the real data
+Now let's try it on the real data.
 
-```{r}
+_I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects.
+Therefore I read the whole file in without proxy.
+To save time I can subset based on Julian dates, but this is a hard coded solution._
+
+```{r, warning=F}
 path <- path <- ifelse(dir == "C:", 
@@
 
-path2 <- paste0(path, "GeoSpatialData/Meteorology/Norway_SeNorge2018_v22.09/Original/rr_tm/") #seNorge2018_1957.nc
+path2 <- paste0(path, "GeoSpatialData/Meteorology/Norway_SeNorge2018_v22.09/Original/rr_tm/")
 
-myFiles <- list.files(path2, full.names = T)[1:2]
+myFiles <- list.files(path2, pattern=".nc$",full.names = T)[1:2]
 start_month <-  "06"
@@
 
+# Get julian days
+temp <- stars::read_ncdf(paste(myFiles[1]))
+start_month_num <-  6
+end_month_num <- 8
+
+julian_start <- yday(st_get_dimension_values(temp, "time")[1] %m+% months(+start_month_num))
+julian_end <- yday(st_get_dimension_values(temp, "time")[1] %m+% months(+end_month_num))
+step <- julian_end-julian_start
+
 for(i in 1:length(myFiles)){
@@
   tic("init")
-  temp <- stars::read_ncdf(paste(path2, var="tg"))
+  temp <- stars::read_ncdf(paste(myFiles[i]), var="tg", proxy=F,
+                           ncsub = cbind(start = c(1, 1, julian_start), 
+                              count = c(NA, NA, step)))
   year_temp <- year(st_get_dimension_values(temp, "time")[1])
@@
   end <-  ym(paste(year_temp, end_month, sep="-"))
-  myInterval <- interval(start, end)
+  #myInterval <- interval(start, end)
+  myInterval <- interval(start, start+1)
+  lookup <- setNames("mean", paste0("v_", year_temp))
+  st_crs(temp) <- 25833
   toc()
@@
     st_apply(1:2, mean) %>%
-    setNames(paste0("v_",year_temp))
+    rename(all_of(lookup)) 
   toc()
@@
   real_temp_summer <- c(temp, real_temp_summer)
-  rm(temp)
+  #rm(temp)
   toc()
 }
+
+```
+
+```{r}
+tif = system.file("tif/L7_ETMs.tif", package = "stars")
+t1 = as.Date("1970-05-31")
+restProxy = read_stars(tif,
+               proxy = T)
+class(restProxy)
+restProxy2 <- st_as_stars(restProxy)
+class(restProxy2)
+dim(restProxy)
+
+restProxy3 <- restProxy %>% filter(band==1)
+dim(restProxy3)
+restProxy4 <- st_as_stars(restProxy3)
+dim(restProxy4)
+
+restProxy5 <- restProxy %>% filter(band==1) %>%
+  st_apply(1:2, mean)
+dim(restProxy5)
+restProxy6 <- st_as_stars(restProxy5)
+dim(restProxy6)
+
+restProxy7 <- restProxy %>% filter(band==1) 
+restProxy7 <- restProxy7 %>%  st_apply(1:2, mean)
+dim(restProxy7)
+restProxy8 <- st_as_stars(restProxy7)
+dim(restProxy8)
+
+```
+
+```{r}
+dim(temp)
+st_get_dimension_values(temp, "time")
+#
+#tic("execute")
+#
+real_temp_summer2 <- st_as_stars(real_temp_summer)
+real_temp_summer <- st_as_stars(real_temp_summer)
+
+#toc()
+temp <- st_as_stars(temp)
+
+write_stars(temp, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_test.tif")
+
+write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
+
+dim(temp)
+
+dim(real_temp_summer)
+names(real_temp_summer)
+
+st_get_dimension_values(real_temp_summer, "time")
 ```

### 2023-02-28 DRAFT_climate_indicators.Rmd
move to server

@@
 myFiles <- list.files(path2, pattern=".nc$",full.names = T)[1:2]
-start_month <-  "06"
-end_month <- "08"
 real_temp_summer <- NULL
@@
 
-julian_start <- yday(st_get_dimension_values(temp, "time")[1] %m+% months(+start_month_num))
-julian_end <- yday(st_get_dimension_values(temp, "time")[1] %m+% months(+end_month_num))
+julian_start <- yday(st_get_dimension_values(temp, "time")[1] %m+%
+                       months(+start_month_num))
+julian_end <- yday(st_get_dimension_values(temp, "time")[1] %m+%
+                     months(+end_month_num))
 step <- julian_end-julian_start
@@
   year_temp <- year(st_get_dimension_values(temp, "time")[1])
-  start <- ym(paste(year_temp, start_month, sep="-"))
-  end <-  ym(paste(year_temp, end_month, sep="-"))
+  #start <- ym(paste(year_temp, start_month, sep="-"))
+  #end <-  ym(paste(year_temp, end_month, sep="-"))
   #myInterval <- interval(start, end)
-  myInterval <- interval(start, start+1)
+  #myInterval <- interval(start, start+1)
   lookup <- setNames("mean", paste0("v_", year_temp))
@@
 
-  tic("middle")
+  tic("filter and st_apply")
   temp <- temp %>%
-    filter(time %within% myInterval) %>%
+    #filter(time %within% myInterval) %>%
     st_apply(1:2, mean) %>%
@@
 
-  tic("c")
+  tic("c()")
   real_temp_summer <- c(temp, real_temp_summer)

### 2023-02-28 DRAFT_climate_indicators.Rmd
parallel cores

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse)
+#library(tidyverse)
+library(dplyr)
+library(ggplot2)
 library(stars)
@@
 library(ncmeta)
+library(parallel)
+library(readr)
 dir <- substr(getwd(), 1,2)
@@
 
-myFiles <- list.files(path2, pattern=".nc$",full.names = T)[1:2]
+myFiles <- list.files(path2, pattern=".nc$",full.names = T)
 real_temp_summer <- NULL
+
+cl <- makeCluster(10L)
 
@@
   year_temp <- year(st_get_dimension_values(temp, "time")[1])
+  print(year_temp)
   #start <- ym(paste(year_temp, start_month, sep="-"))
@@
     #filter(time %within% myInterval) %>%
-    st_apply(1:2, mean) %>%
+    st_apply(1:2, mean, CLUSTER = cl) %>%
     rename(all_of(lookup)) 
@@
 
+tic("Merge")
+real_temp_summer <- real_temp_summer %>%
+  merge(name = "Year") %>%
+  setNames("climate_variable")
+toc()
+
+stopCluster(cl)
+
+```
+
+Here's a test for the effect of splitting over more cores.
+```{r, eval=F}
+cl <- makeCluster(2L)
+cl2 <- makeCluster(6L)
+cl3 <- makeCluster(10L)
+
+tic("No cluster")
+temp %>%
+    #filter(time %within% myInterval) %>%
+    st_apply(1:2, mean) %>%
+    rename(all_of(lookup)) 
+toc()
+
+tic("Two clusters")
+temp %>%
+    #filter(time %within% myInterval) %>%
+    st_apply(1:2, mean, CLUSTER = cl) %>%
+    rename(all_of(lookup)) 
+stopCluster(cl)
+toc()
+
+tic("Six clusters")
+temp %>%
+    #filter(time %within% myInterval) %>%
+    st_apply(1:2, mean, CLUSTER = cl2) %>%
+    rename(all_of(lookup)) 
+stopCluster(cl2)
+toc()
+
+tic("Ten clusters")
+temp %>%
+    #filter(time %within% myInterval) %>%
+    st_apply(1:2, mean, CLUSTER = cl3) %>%
+    rename(all_of(lookup)) 
+stopCluster(cl3)
+toc()
+
+```
+No cluster: 21.401 sec elapsed
+Two clusters: 21.546 sec elapsed
+Six clusters: 14.548 sec elapsed
+Ten clusters: 15.29 sec elapsed
+Ten clusters (second run): 13.617 sec elapsed
+
+The RStudio server has 48 cores. More parallel cores is probably on average faster. The NINA guidelines is to use max 10 cores, and to remember to close parallel cluster when done.
+
+```{r}
+#write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
+```
+
+I might want to merge this
+
+```{r}
+real_temp_summer <- real_temp_summer %>%
+  merge(name = "Year") %>%
+  setNames("climate_variable")
 ```
@@
 ```{r}
-tif = system.file("tif/L7_ETMs.tif", package = "stars")
-t1 = as.Date("1970-05-31")
-restProxy = read_stars(tif,
-               proxy = T)
-class(restProxy)
-restProxy2 <- st_as_stars(restProxy)
-class(restProxy2)
-dim(restProxy)
-
-restProxy3 <- restProxy %>% filter(band==1)
-dim(restProxy3)
-restProxy4 <- st_as_stars(restProxy3)
-dim(restProxy4)
-
-restProxy5 <- restProxy %>% filter(band==1) %>%
-  st_apply(1:2, mean)
-dim(restProxy5)
-restProxy6 <- st_as_stars(restProxy5)
-dim(restProxy6)
-
-restProxy7 <- restProxy %>% filter(band==1) 
-restProxy7 <- restProxy7 %>%  st_apply(1:2, mean)
-dim(restProxy7)
-restProxy8 <- st_as_stars(restProxy7)
-dim(restProxy8)
-
-```
-
-```{r}
-dim(temp)
-st_get_dimension_values(temp, "time")
-#
-#tic("execute")
-#
-real_temp_summer2 <- st_as_stars(real_temp_summer)
-real_temp_summer <- st_as_stars(real_temp_summer)
-
-#toc()
-temp <- st_as_stars(temp)
-
-write_stars(temp, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_test.tif")
-
-write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
-
-dim(temp)
-
-dim(real_temp_summer)
-names(real_temp_summer)
-
-st_get_dimension_values(real_temp_summer, "time")
+ggplot()+
+  geom_stars(data = real_temp_summer[,,,1:2], downsample = c(10, 10, 0)) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D") +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))+
+  facet_wrap(~Year)
 ```

### 2023-02-28 R/functionalPlantIndicators/data handling.R
adding GAD and fixing species names

@@ -132,6 +132,7 @@ ind.dat$species <- as.factor(ind.dat$species)
 summary(ind.dat$species)
 head(ind.dat)
 
+
 #### data handling - ANO data ####
 head(ANO.sp)
 head(ANO.geo)
@@ -147,32 +148,237 @@ ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital
 ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after space to lowercase
 
 ## merge species data with indicators
-ANO.sp <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
+ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                 y= ind.dat[,c("species","CC", "SS", "RR", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
                 by.x="Species", by.y="species", all.x=T)
-summary(ANO.sp)
+summary(ANO.sp.ind)
+
+
+# checking which species didn't find a match
+unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
+#unique(ANO.sp.ind[!is.na(ANO.sp.ind$Moisture),'Species'])
+
+ANO.sp.ind[ANO.sp.ind$Species=="Taraxacum officinale",]
+ind.dat[ind.dat$species=="Picea sitchensis",]
+ind.Grime[ind.Grime$species=="Picea sitchensis",]
+ind.Tyler[ind.Tyler$species=="Picea sitchensis",]
+ANO.sp[ANO.sp$Species=="Picea sitchensis",]
+
+# fix species name issues
+ind.dat <- ind.dat %>% 
+  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
+  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
+  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
+  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
[…]

### 2023-02-28 DRAFT_climate_indicators.Rmd
calculate reference and sd

@@
 
+Checking that this is correct
+```{r}
+p = st_sfc(st_point(c(7039947.63, 270783.69)), crs=25833)
+```
+
 The data has three dimension
@@
 
-`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change untill later, typically I write to file and all the lazy operations are done in squeal.
+`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change until later, typically I write to file and all the lazy operations are done in squeal.
 Therefore, I will prepare a test data set, which is smaller and which I can import to memory.
@@
 
-```{r, warning=F}
+```{r, warning=F, eval=F}
 path <- path <- ifelse(dir == "C:", 
@@
 myFiles <- list.files(path2, pattern=".nc$",full.names = T)
+# The last file, the last year, is incomplete and dont include all juliani dates that we select below
+myFiles <- myFiles[-length(myFiles)]
 real_temp_summer <- NULL
@@
 # Get julian days
-temp <- stars::read_ncdf(paste(myFiles[1]))
+temp <- stars::read_ncdf(paste(myFiles[1]), var="tg")
 start_month_num <-  6
@@
   #myInterval <- interval(start, start+1)
-  lookup <- setNames("mean", paste0("v_", year_temp))
+  lookup <- setNames("mean", paste0("v_", year_temp)) # Perhaps leave out the v_ to get a numeric vector instead, which is easier to subset
   st_crs(temp) <- 25833
@@
 
-Here's a test for the effect of splitting over more cores.
+This takes about 20 sec per file/year, or 22 min on total. Here's a test for the effect of splitting over more cores.
 ```{r, eval=F}
@@
 ```{r}
-#write_stars(real_temp_summer, "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/real_temp_summer.tif")
-```
-
-I might want to merge this
-
-```{r}
-real_temp_summer <- real_temp_summer %>%
-  merge(name = "Year") %>%
-  setNames("climate_variable")
+names(real_temp_summer)
 ```
@@
 ```{r}
+dim(real_temp_summer)
+```
+
+```{r}
+st_get_dimension_values(real_temp_summer, "Year")
+```
+
+```{r, eval=F}
+saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.rds")
+write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff")
+```
+
+These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.
+
+```{r}
+temp_median_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
+                                 proxy=F)
+# This is too slow:
+# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_median_summer.rds")
+```
+
+Note that GTiff automatically renames the third dimension _band_ and also renames the attribute.
+
+```{r}
+st_get_dimension_values(temp_median_summer, "band")
+```
+
+I can rename them. 
+```{r}
+temp_median_summer <- temp_median_summer %>% 
+  st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
+  setNames("temperature")
+dim(temp_median_summer)
+```
+
+```{r, fig.cap="Showing three random slices of the year dimension."}
 ggplot()+
-  geom_stars(data = real_temp_summer[,,,1:2], downsample = c(10, 10, 0)) +
+  geom_stars(data = temp_median_summer[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
   coord_equal() +
@@
   scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~Year)
+  facet_wrap(~v_YEAR)
 ```
@@
 
-For this step we need some more dummy data from different years. I made data sets y1 to y5 above.
+We need to first to define a reference period and then to subset our data `temp_median_summer`. We can practice on the dummy data `y`.
 
 ```{r}
-# Function to add random noise
-myRandom <- function(x) x*rnorm(1,1,.05)
+dir <- substr(getwd(), 1,2)
+path <- ifelse(dir == "C:",
+       "P:/", 
+       "/data/P-Prosjekter2/")
+path2 <- paste0(path, 
+                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
+myFiles <- list.files(path2, pattern=".rds", full.names = T)
+y <- NULL
+
+# Looping though the files in the directory
+for(i in 1:length(myFiles)){
+  temp <- readRDS(myFiles[i])
+  y <- c(temp, y)
+}
 ```
@@
 ```{r}
-y1 <- x_summerMean
-y2 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-1)) 
-y3 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-2)) 
-y4 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-3)) 
-y5 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-4)) 
-# We combine the data into one cube:
-y <- c(y1, y2, y3, y4, y5) %>% 
-  merge(name = "Year") %>%
-  setNames("Attribute_A")
+st_get_dimension_values(y, "time")
 ```
-
 ```{r}
-ggplot() + 
-  geom_stars(data = y) +
-  coord_equal() +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +  
-  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~Year)
+y_filtered <- y %>%
+  select(Attribute_A) %>%
+  filter(time %within% interval("1961-01-01", "1990-12-31"))
 ```
@@
 ```{r}
-y_ref <- y %>%
+y_ref <- y_filtered %>%
   st_apply(c("x", "y"), FUN =  median_sd)
@@
 
-```{r}
+```{r, fig.cap="Showing the upper reference levels and the standard deviation from the dummy data set."}
 tmap_arrange(
@@
 tm_shape(y_ref)+
+  tm_raster("sd",
+            palette = "-viridis")
+)
+```
+
+Let's transfer this over to the real data. First we need to rename our dimension values and turn them back into dates.
+
+```{r}
+new_dims <- as.Date(paste0(substr(st_get_dimension_values(temp_median_summer, "v_YEAR"), 3, 6), "-01-01"))
+temp_median_summer_ref <- temp_median_summer %>%
+  st_set_dimensions("v_YEAR", values = new_dims)
+```
+
+Then I can filter to leave only the reference period.
+
+```{r}
+temp_median_summer_ref <- temp_median_summer_ref %>%
+  filter(v_YEAR %within% interval("1961-01-01", "1990-12-31"))
+st_get_dimension_values(temp_median_summer_ref, "v_YEAR")
+```
+
+And then we calculate the median and sd like above
+
+```{r, eval=FALSE}
+system.time({
+cl <- makeCluster(10L)
+temp_median_summer_ref <- temp_median_summer_ref %>%
+  st_apply(c("x", "y"), 
+           FUN =  median_sd,
+           CLUSTER = cl)
+stopCluster(cl)
+})
+```
+
+  user  system elapsed 
+  9.624   6.069  20.903
+
+```{r}
+temp_median_summer_ref <- temp_median_summer_ref %>% split("median_sd")
+```
+
+The attribute name *mean* we can change this to something more meaningful.
+
+```{r}
+temp_median_summer_ref <- setNames(temp_median_summer_ref, c("reference_upper", "sd"))
+```
+
+```{r, fig.cap="Showing the upper reference levels and the standard deviation from actual data of median summer temperatures."}
+tmap_arrange(
+tm_shape(st_downsample(temp_median_summer_ref, 10))+
+  tm_raster("reference_upper")
+,
+tm_shape(st_downsample(temp_median_summer_ref, 10))+
   tm_raster("sd",

### 2023-03-02 R/functionalPlantIndicators/data handling.R
GRUK & våtmark GAD

@@ -30,35 +30,30 @@ head(ANO.sp)
 head(ANO.geo)
 
 ## GRUK
-# 2020
-excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2020_181021_edited201021.xlsx")
-GRUK2020.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2020_181021_edited201021.xlsx", 
-                                 sheet = 5)
-GRUK2020.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2020_181021_edited201021.xlsx", 
-                               sheet = 2)
-
-# 2021
-excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2021_181021_GJELDENDE.xlsx")
-
-GRUK2021.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2021_181021_GJELDENDE.xlsx", 
+excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx")
+GRUK.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
                                  sheet = 2)
-GRUK2021.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/S123_GRUK2021_181021_GJELDENDE.xlsx", 
+GRUK.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
                                sheet = 3)
 
-
+# condition evaluation for 2021 data
 excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/NNF_GRUK_GJELDENDE.xls")
 
 GRUK2021.condition <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/NNF_GRUK_GJELDENDE.xls", 
                                  sheet = 1)
 
+head(GRUK.variables)
+head(GRUK.species)
+head(GRUK2021.condition)
 
-# 2022
[…]

### 2023-03-02 DRAFT_climate_indicators.Rmd
normalise real data

@@
 
-```{r}
+```{r, eval=F}
 names(real_temp_summer)
@@
 
-```{r}
+```{r, eval=F}
 dim(real_temp_summer)
@@
 
-```{r}
+```{r, eval=F}
 st_get_dimension_values(real_temp_summer, "Year")
@@
 # readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_median_summer.rds")
+# but it could be that RData works better 
 ```
@@
 ```{r}
-temp_median_summer_ref <- temp_median_summer_ref %>% split("median_sd")
+temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
 ```
@@
 ```{r}
-temp_median_summer_ref <- setNames(temp_median_summer_ref, c("reference_upper", "sd"))
+temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
 ```
@@
 tmap_arrange(
-tm_shape(st_downsample(temp_median_summer_ref, 10))+
+tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
   tm_raster("reference_upper")
 ,
-tm_shape(st_downsample(temp_median_summer_ref, 10))+
+tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
   tm_raster("sd",
@@
 
-#### Export
-
-I should export the reference data at this point (write to file) as we don't need to recalculate this every time we want to update the indicator values.
-
-#### Calculate variable values
-
-The climate variable is the same as y1, y2, y~*n*~, ....
-But to create some differences between the variable and the reference values, I will tweak these slightly, taking them down 2%.
-I then merge with the reference value and sd, creating one common data cube.
-
-```{r}
-y_var <- y %>% mutate(Attribute_A = Attribute_A * 0.98) %>% 
-  split(3) %>% c(y_ref)
+This is also something I can export.
+```{r, eval=F}
+# write_stars can only export a single attribute 
+#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")
+
+saveRDS(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+```
+
+```{r, eval=F, include=F}
+temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+```
+
+#### Combine
+
+I need to combine the variables and the ref values in one data cube
+
+```{r}
+y_var <- temp_median_summer %>%
+  split(3) %>%
+  c(temp_median_summer_ref_long)
+
 ```
@@
 # can I find a way to keep the v_ columns?
+
+# The break point scaling is actually not needed here, since 
+# having the lower ref value to be 5 sd implies that the threshold is
+# 2 sd in a linear scaling.
+
 system.time(
 y_var_norm <- y_var %>%
-  mutate(lower = reference_upper - 5*sd ) %>%
-  mutate(threshold = reference_upper -2*sd ) %>%
-  mutate(across(all_of(cols), ~ if_else(.x < threshold, 
-                                        (.x - lower) / (threshold - lower),
-                                        (.x - threshold) / (reference_upper - threshold),
+  mutate(reference_low = reference_upper - 5*sd ) %>%
+  mutate(reference_low2 = reference_upper + 5*sd ) %>%
+  mutate(threshold_low = reference_upper -2*sd ) %>%
+  mutate(threshold_high = reference_upper +2*sd ) %>%
+  mutate(across(all_of(cols), ~ 
+                  if_else(.x < reference_upper,
+                  if_else(.x < threshold_low, 
+                                        (.x - reference_low) / (threshold_low - reference_low),
+                                        (.x - threshold_low) / (reference_upper - threshold_low),
+                                        ),
+                  if_else(.x > threshold_high, 
+                                        (reference_low2 - .x) / (reference_low2 - threshold_high),
+                                        (threshold_high - .x) / (threshold_high - reference_upper),
+                                        )
                 ))) %>%
@@
 ```
-
-I'm not sure mutate-across is the fastest way to do this. Perhaps st_apply would be better.
-
-```{r}
+  user  system elapsed 
+ 14.803   2.717  17.512
+
+```{r, fig.cap="Example showing median summer tempretur in 1970, the upper and lwoer reference temperture, i.e.
+median and 5 SD units of the temperature between  1961-1990, and finally, the scaled indicator values."}
+
+lims <- c(-5, 22)
+
 ggarrange(
 ggplot() + 
-  geom_stars(data = y_var_norm["v_1970"]) +
+  geom_stars(data = st_downsample(y_var_norm["v_1970"],10)) +
   coord_equal() +
   theme_void() +
-  scale_fill_viridis_c(option = "D") +  
+  scale_fill_viridis_c(option = "D",
+                       limits = lims) +  
   scale_x_discrete(expand = c(0, 0)) +
@@
 ggplot() + 
-  geom_stars(data = y_var_norm["reference_upper"]) +
+  geom_stars(data = st_downsample(y_var_norm["reference_upper"], 10)) +
   coord_equal() +
   theme_void() +
-  scale_fill_viridis_c(option = "D") +  
+  scale_fill_viridis_c(option = "D",
+                       limits = lims) +  
   scale_x_discrete(expand = c(0, 0)) +
@@
 ggplot() + 
-  geom_stars(data = y_var_norm["i_1970"]) +
+  geom_stars(data = st_downsample(y_var_norm["reference_low"], 10)) +
   coord_equal() +
   theme_void() +
-  scale_fill_viridis_c(option = "D") +  
+  scale_fill_viridis_c(option = "D",
+                       limits = lims) +  
   scale_x_discrete(expand = c(0, 0)) +
   scale_y_discrete(expand = c(0, 0))
-, ncol=3, align = "hv"
+,
+
+ggplot() + 
+  geom_stars(data = st_downsample(y_var_norm["reference_low2"], 10)) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "D",
+                       limits = lims) +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))
+,
+
+ggplot() + 
+  geom_stars(data = st_downsample(y_var_norm["i_1970"],10)) +
+  coord_equal() +
+  theme_void() +
+  scale_fill_viridis_c(option = "A",
+                       limits = c(0, 1)) +  
+  scale_x_discrete(expand = c(0, 0)) +
+  scale_y_discrete(expand = c(0, 0))
+, ncol=2, nrow=3, align = "hv"
 )
 ```
-
-This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
-If the reference value was static, or if there was a pattern in v_1970, the scaled indicator map would look different.
 

### 2023-03-03 R/functionalPlantIndicators/data handling.R
gruk data handling

@@ -287,7 +287,152 @@ head(ANO.sp.ind)
 
 
 #### data handling - GRUK data ####
+head(GRUK.variables)
+head(GRUK.species)
+head(GRUK2021.condition)
 
+colnames(GRUK.species)[5] <- "art_dekning"
+
+# fix species names
+GRUK.species <- as.data.frame(GRUK.species)
+
+GRUK.species$Species <- GRUK.species$Navn
+unique(as.factor(GRUK.species$Species))
+GRUK.species$Species <- sub(".*?_", "", GRUK.species$Species) # lose the Norwegian name in the front
+GRUK.species$Species <- gsub("_", " ", GRUK.species$Species) # replace underscore with space
+GRUK.species$Species <- str_to_title(GRUK.species$Species) # make first letter capital
+GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after hyphon to lowercase
+GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after space to lowercase
+GRUK.species[,'Species'] <- word(GRUK.species[,'Species'], 1,2) # lose subspecies
+
+
+
+
+## merge species data with indicators
+GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID")], 
+                    y= ind.dat[,c("species","CC", "SS", "RR", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
+                    by.x="Species", by.y="species", all.x=T)
+summary(GRUK.species.ind)
+
+
+# checking which species didn't find a match
+unique(GRUK.species.ind[is.na(GRUK.species.ind$Moisture & 
+                                is.na(GRUK.species.ind$RR)),'Species'])
+
+
+
+# fix species name issues
+ind.dat <- ind.dat %>% 
+#  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
+#  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+#  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
[…]

### 2023-03-03 DRAFT_climate_indicators.Rmd
reg means

@@
 
-### Dummy data
+#### Dummy data
 
@@
 ```{r}
-reg <- sf::st_read("data/regions.shp")
+reg <- sf::st_read("data/regions.shp", options = "ENCODING=UTF8")
 #st_crs(reg)
+```
+
+Outline of Norway
+
+```{r}
+nor <- sf::st_read("data/outlineOfNorway_EPSG25833.shp")
+```
+Remoce marine areas from regions
+
+```{r}
+reg <- st_intersection(reg, nor)
+```
+
+```{r, fig.cap="Five accounting areas (regions)."}
+
+tm_shape(reg) +
+  tm_polygons(col="region")
+
 ```
@@
 
-```{r}
+This is also something I can export.
+```{r, eval=F}
+# write_stars can only export a single attribute 
+#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")
+
+saveRDS(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+```
+
+```{r, eval=T, include=F}
+temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+```
+
+```{r, eval=T}
 temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
@@
 
-```{r}
+```{r, eval=T}
 temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
@@
 
-This is also something I can export.
-```{r, eval=F}
-# write_stars can only export a single attribute 
-#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")
-
-saveRDS(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
-```
-
-```{r, eval=F, include=F}
-temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
-```
-
 #### Combine
@@
 
-```{r}
+```{r, eval=F}
 # Mutate
@@
  14.803   2.717  17.512
+
+```{r, eval=F}
+gc()
+saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
+```
+
+```{r, include=FALSE}
+y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
+```
 
@@
 
-#### Plot time series
+The _real_ indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming.
+Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.
+
+#### Plot time series as line graphs
+
+We could also plot a time series as a series of maps, but then we need to avergae over this 5 time periods firs. We can get to that later.
+
+To plot time series I first need to do a spatial aggregation. One option is to spatially aggregate the stars object directly, using the `reg` data set:
+
+```{r, eval=F}
+temp <- y_var_norm[1,,] %>%
+  aggregate(reg, mean, na.rm=T)
+plot(temp)
+temp2 <- temp %>%
+  st_as_sf() # too slow
+```
+
+However, this does not preserve the region name, and a subsequent st_intersection taks a little whie and gives some boundary problems.
+
+I can also converting att pixels to points and then do the intersection.
+
+```{r, eval=F}
+temp_points <- y_var_norm[1,,] %>%
+  st_as_sf(as_points = TRUE, merge = FALSE)
+temp_points2 <- temp_points %>%
+  st_intersection(reg) " too slow"
+```
+
+But this also took too long.
+
+I can perhaps aggregate, and then turn to points and intersect, but that did not work perfect either.
+
+I can rastereze the regions data set, turn it into a coded dimension and use st_apply, but thats tedious.
+
+Let me try using `exactextract`, as a parallel to `aggregate`. This requires us to turn to the `terra` package for a bit.
+
+```{r, eval=F}
+(
+regional_means <- rast(y_var_norm[,,]) %>%
+  exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
+  setNames(c("region", names(y_var_norm)))
+  )
+```
+
+That was fast!
+
+```{r, eval=F}
+saveRDS(regional_means, "temp/regional_means.rds")
+```
+
+Reshape and plot
+
+```{r}
+(
+regional_means_long <- regional_means %>%
+  as.data.frame() %>%
+  rename(referenceUpper = reference_upper,
+         referenceLow = reference_low,
+         referenceLow2 = reference_low2,
+         thresholdLow = threshold_low,
+         thresholdHigh = threshold_high) %>%
+  pivot_longer(!region) %>%
+  separate(name, into=c("type", "year"), remove=F)
+)
+```
+
+```{r}
+temp3 <- temp2 %>%
+  st_intersection(reg)
+```
 

### 2023-03-03 R/functionalPlantIndicators/data handling.R
gruk tilstand

@@ -406,33 +406,64 @@ unique(GRUK.species.ind[is.na(GRUK.species.ind$Moisture &
                                 is.na(GRUK.species.ind$RR)),'Species'])
 
 
+### merge GRUK.variables and GRUK2021.condition
+GRUK.variables <- as.data.frame(GRUK.variables)
+GRUK2021.condition <- as.data.frame(GRUK2021.condition)
+
+head(GRUK.variables)
+head(GRUK2021.condition)
+
+## fixing site-ID in the condition-data
+GRUK2021.condition$Flate_ID <- GRUK2021.condition$områdenavn
+unique(as.factor(GRUK2021.condition$Flate_ID))
+GRUK2021.condition$Flate_ID <- gsub("_", " ", GRUK2021.condition$Flate_ID) # replace underscore with space
+unique(as.factor(GRUK2021.condition$Flate_ID))
+GRUK2021.condition$Flate_ID <- gsub(" ", "-", GRUK2021.condition$Flate_ID) # replace space with hyphon
+GRUK2021.condition$Flate_ID <- gsub("[^0-9\\-]", "", GRUK2021.condition$Flate_ID) # remove all non-numerics except '-'
+GRUK2021.condition$Flate_ID <- gsub("--", "-", GRUK2021.condition$Flate_ID) # remove double hyphons
+GRUK2021.condition$Flate_ID
+GRUK2021.condition$Flate_ID <- substr(GRUK2021.condition$Flate_ID, 1, nchar(GRUK2021.condition$Flate_ID)-1) # remove the last character ('-' for all)
+GRUK2021.condition$Flate_ID
 
 
-
-
-
-
+colnames(GRUK.variables)
+colnames(GRUK2021.condition)
 #### continue here ####
 
+
+GRUK.variables <- merge(x=GRUK.variables, 
+                          y=GRUK2021.condition[,c("tilstandsvurdering","tilstandsgrunn","artsmangfoldvurdering","lokalitetskvalitet","Flate_ID")], 
+                          by.x="Flate_ID", by.y="Flate_ID", all.x=T)
+
+
+
+
+
+
+
+
+
+
+
 ## adding information on ecosystem and condition variables
-ANO.sp.ind <- merge(x=ANO.sp.ind, 
+GRUK.species.ind <- merge(x=GRUK.species.ind, 
[…]

### 2023-03-03 DRAFT_climate_indicators.Rmd
climate trend figure

@@
 
-*Author and date:* <!-- Write your name here -->
+*Author and date:* 
+Anders L. Kolstad
+March 2023
 
@@
 knitr::opts_chunk$set(echo = TRUE)
-#library(tidyverse)
+#library(tidyverse) # cant update to v2 on server
 library(dplyr)
@@
 library(readr)
+library(exactextractr)
+library(tidyr)
+library(terra)
 dir <- substr(getwd(), 1,2)
@@
 
-This chapters describes a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
+This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
 
@@
 ```{r}
-# setNams dont work on stars proxy obejct, so I use rename instead
+# setNames dont work on stars proxy obejct, so I use rename instead
 lookup <- setNames("mean", paste0("v_", temp_names))
@@
 ```{r, eval=F}
-names(real_temp_summer)
-```
-
-```{r, eval=F}
-dim(real_temp_summer)
-```
-
-```{r, eval=F}
-st_get_dimension_values(real_temp_summer, "Year")
-```
-
-```{r, eval=F}
 saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.rds")
@@
 # but it could be that RData works better 
+```
+
+```{r, eval=T}
+names(real_temp_summer)
+```
+
+```{r, eval=T}
+dim(real_temp_summer)
+```
+
+```{r, eval=T}
+st_get_dimension_values(real_temp_summer, "Year")
 ```
@@
 
-The attribute name *mean* we can change this to something more meaningful.
+The attribute name is *mean*. We can change this to something more meaningful.
 
@@
 (
-regional_means <- rast(y_var_norm[,,]) %>%
+regional_means <- rast(y_var_norm) %>%
   exact_extract(reg, fun = 'mean', append_cols = "region", progress=T) %>%
@@
 
+I should also get the sd:
+
+```{r, eval=F}
+(
+regional_means_sd <- rast(y_var_norm) %>%
+  exact_extract(reg, fun = function(X) c(mean = mean(x), sd = sd(x)), append_cols = "region", progress=T) %>%
+  setNames(c("region", names(y_var_norm)))
+  )
+```
+
 ```{r, eval=F}
@@
 
+```{r}
+regional_means <- readRDS("temp/regional_means.rds")
+```
+
 Reshape and plot
 
-```{r}
+```{r, warning=F}
+div <- c("reference_upper",
+         "reference_low",
+         "reference_low2",
+         "threshold_low",
+         "threshold_high",
+         "sd")
+
+temp <- regional_means %>%
+  as.data.frame() %>%
+  select(region, div)
+
 (
@@
   as.data.frame() %>%
-  rename(referenceUpper = reference_upper,
-         referenceLow = reference_low,
-         referenceLow2 = reference_low2,
-         thresholdLow = threshold_low,
-         thresholdHigh = threshold_high) %>%
+  select(!all_of(div)) %>%
   pivot_longer(!region) %>%
-  separate(name, into=c("type", "year"), remove=F)
+  separate(name, into=c("type", "year")) %>%
+  pivot_wider(#id_cols = region,
+              names_from = type)  %>%
+  left_join(temp, by=join_by(region)) %>% 
+  mutate(diff = v-reference_upper) %>%
+  mutate(threshold_low_centered = threshold_low-reference_upper) %>%
+  mutate(threshold_high_centered = threshold_high-reference_upper)
 )
-```
-
-```{r}
-temp3 <- temp2 %>%
-  st_intersection(reg)
+
+```
+
+```{r}
+regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")
+
+regional_means_long %>%
+  mutate(col = if_else(diff>0, "1", "2")) %>%
+  ggplot(aes(x = as.numeric(year), 
+           y = diff, fill = col))+
+  geom_bar(stat="identity")+
+  geom_hline(aes(yintercept = threshold_low_centered),
+        linetype=2)+
+  geom_hline(aes(yintercept = threshold_high_centered),
+        linetype=2)+
+  geom_segment(x = 1961, xend=1990,
+               y = 0, yend = 0,
+               linewidth=2)+
+  scale_fill_hue(l=70, c=60)+
+  theme_bw(base_size = 12)+
+  ylab("Sommertempratur\navvik fra 1961-1990")+
+  xlab("")+
+  guides(fill="none")+
+  facet_wrap( .~ factor(region, levels = regOrder),
+              ncol=3,
+              scales = "free_y")
+```
+
+Then we can take the mean over the last 5 years and add to a spatial representation.
+
+```{r}
+(
+i_2022 <- regional_means_long %>%
+  filter(year >2017) %>%
+  group_by(region) %>%
+  summarise(i_2022 = mean(i))
+)
+```
+
+```{r}
+(
+sd_2022 <- regional_means_long %>%
+  filter(year >2017) %>%
+  group_by(region) %>%
+  summarise(i_2022 = mean(i))  
+)
 ```

### 2023-03-03 DRAFT_climate_indicators.Rmd
include spatial sd

@@
 library(tidyr)
+library(data.table)
 library(terra)
@@
 ```{r, eval=F}
-(
+system.time(
 regional_means <- rast(y_var_norm) %>%
@@
 
+ user  system elapsed 
+  7.603   3.071  10.697
+
 That was fast! 
+```{r}
+regional_means
+```
 
@@
 ```{r, eval=F}
-(
-regional_means_sd <- rast(y_var_norm) %>%
-  exact_extract(reg, fun = function(X) c(mean = mean(x), sd = sd(x)), append_cols = "region", progress=T) %>%
+system.time(
+regional_sd <- rast(y_var_norm) %>%
+  exact_extract(reg, fun = 'stdev', append_cols = "region", progress=T) %>%
   setNames(c("region", names(y_var_norm)))
@@
 
+  user  system elapsed 
+  7.420   2.921  10.355
+
 ```{r, eval=F}
 saveRDS(regional_means, "temp/regional_means.rds")
-```
-
-```{r}
+saveRDS(regional_sd, "temp/regional_sd.rds")
+```
+
+```{r, include=F}
 regional_means <- readRDS("temp/regional_means.rds")
+regional_sd <- readRDS("temp/regional_sd.rds")
 ```
@@
 
-(
 regional_means_long <- regional_means %>%
@@
   mutate(threshold_high_centered = threshold_high-reference_upper)
+
+#Adding the spatial sd
+(
+regional_means_long <- regional_sd %>%
+  select(!all_of(div)) %>%
+  pivot_longer(!region) %>%
+  separate(name, into=c("type", "year")) %>%
+  pivot_wider(names_from = type) %>%
+  rename(i_sd = i,
+         v_sd = v) %>%
+  left_join(regional_means_long, by=join_by(region, year))
 )
-
 ```
@@
 (
-i_2022 <- regional_means_long %>%
-  filter(year >2017) %>%
-  group_by(region) %>%
-  summarise(i_2022 = mean(i))
+i_aggregatedToPeriods <- regional_means_long %>%
+  mutate(year = as.numeric(year)) %>%
+  mutate(period = case_when(
+    year %between% c(2018, 2022) ~ "2018-2022",
+    year %between% c(2013, 2017) ~ "2013-2017",
+    year %between% c(2008, 2012) ~ "2008-2012",
+    year %between% c(2003, 2007) ~ "2003-2007",
+    .default = "pre 2003"
+  )) %>%
+  mutate(period_rank = case_when(
+   period == "2018-2022" ~ 5,
+   period == "2013-2017" ~ 4,
+   period == "2008-2012" ~ 3,
+   period == "2003-2007" ~ 2,
+    .default = 1
+  )) %>%
+  group_by(region, period, period_rank) %>%
+  summarise(indicator = mean(i),
+            spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
+            #,
+            #spatial_sd_mean = mean(i_sd),
+            #spatial_sd2_max = max(i_sd),
+            #spatial_sd2_length = length(i_sd)
             )
-```
-
-```{r}
-(
-sd_2022 <- regional_means_long %>%
-  filter(year >2017) %>%
-  group_by(region) %>%
-  summarise(i_2022 = mean(i))  
 )
+```
+
+```{r, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent spatial variation within regions and across years."}
+labs <- unique(i_aggregatedToPeriods$period[order(i_aggregatedToPeriods$period_rank)])
+
+i_aggregatedToPeriods %>%
+  ggplot(aes(x = period_rank, 
+             y = indicator,
+             colour=region))+
+  geom_line() +
+  geom_point() +
+  geom_errorbar(aes(ymin=indicator-spatial_sd, 
+                    ymax=indicator+spatial_sd), 
+                    width=.2,
+                 position=position_dodge(0.2)) +
+  theme_bw(base_size = 12)+
+  scale_x_continuous(breaks = 1:5,
+                     labels = labs)+
+  labs(x = "", y = "indikatorverdi")
+```
+
+Finally, I can add these values to the sp object.
+
+```{r}
+reg2 <- reg %>%
+  left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))
+
+```
+
+```{r}
+ggplot(reg2) + 
+  geom_sf(fill = indicator) +
+  geom_sf_label(aes(label = indicator))
+
+```
+
+```{r}
+tm_shape(reg2)+
+  tm_polygons(col=)
 ```

### 2023-03-03 DRAFT_climate_indicators.Rmd
map with regions

@@
 ```{r, eval=T}
-names(real_temp_summer)
+names(temp_median_summer)
 ```
@@
 ```{r, eval=T}
-dim(real_temp_summer)
+dim(temp_median_summer)
 ```
@@
 ```{r, eval=T}
-st_get_dimension_values(real_temp_summer, "Year")
+st_get_dimension_values(temp_median_summer, "Year")
 ```
@@
 That was fast!
-```{r}
-regional_means
-```
 
@@
 regional_sd <- readRDS("temp/regional_sd.rds")
+```
+
+```{r}
+regional_means
 ```
@@
   left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))
-
-```
-
-```{r}
-ggplot(reg2) + 
-  geom_sf(fill = indicator) +
-  geom_sf_label(aes(label = indicator))
-
-```
-
-```{r}
-tm_shape(reg2)+
-  tm_polygons(col=)
+names(reg2)
+```
+
+```{r}
+myCol <- "RdYlGn"
+myCol2 <- "-RdYlGn"
+
+tm_main <- tm_shape(reg2)+
+  tm_polygons(col="indicator",
+              title="Indikator:\nsommertemperatur",
+    palette = myCol,
+    style="fixed",
+    breaks = seq(0,1,.2))
+
+tm_inset <- tm_shape(reg2)+
+  tm_polygons(col="spatial_sd",
+              title="SD",
+              palette = myCol2,
+              style="cont")+
+  tm_layout(legend.format = list(digits=2))
+
+#tm_main <- tmap_grob(tm_main)
+#tm_inset <- tmap_grob(tm_inset)
+
+#ggdraw() +
+#  draw_plot(tm_main) +
+#  draw_plot(tm_inset,
+#    height = 0.4,
+#    x = 0.12, y = .35
+#  )
+
+tmap_arrange(tm_main, 
+             tm_inset)
 ```

### 2023-03-06 R/functionalPlantIndicators/1 data handling.R
set up bootstraps

@@ -460,8 +460,8 @@ summary(GRUK.species.ind)
 head(GRUK.species.ind)
 
 
-#### continue here ####
-#### data handling - ASO data ####
+
+#### data handling - ASO data ... NOT DONE YET ####
 
 
 
@@ -578,11 +578,14 @@ NiN.sp <- NiN.sp %>%
   mutate(sp=str_replace(sp,"Poa alpigena", "Poa pratensis")) %>%
   mutate(sp=str_replace(sp,"Poa angustifolia", "Poa pratensis")) %>%
   mutate(sp=str_replace(sp,"Poa laxa", "Poa flexuosa")) %>%
+  mutate(sp=str_replace(sp,"Poa _herjedalica", "Poa herjedalica")) %>%
+  mutate(sp=str_replace(sp,"Poa _jemtlandica", "Poa jemtlandica")) %>%
   mutate(sp=str_replace(sp,"Poa lindebergii", "Poa arctica")) %>%
   mutate(sp=str_replace(sp,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
   mutate(sp=str_replace(sp,"Rhamnus catharticus", "Rhamnus cathartica")) %>%
   mutate(sp=str_replace(sp,"Rumex alpestris", "Rumex acetosa")) %>%
   mutate(sp=str_replace(sp,"Salix _fragilis", "Salix fragilis")) %>%
+  mutate(sp=str_replace(sp,"Saxifraga _opdalensis", "Saxifraga opdalensis")) %>%
   mutate(sp=str_replace(sp,"Spergularia salina", "Spergularia marina")) %>%
   mutate(sp=str_replace(sp,"Syringa emodi", "Syringa vulgaris")) %>%
   mutate(sp=str_replace(sp,"Taraxacum crocea", "Taraxacum officinale")) %>%
@@ -606,6 +609,7 @@ unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) &
                     NiN.sp.ind$spgr %in% list("a1a","a1b","a1c")
                   ,'sp'])
 # ok now
+
 
 #### matching with NiN ecosystem types - wetlands ####
 # NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "

### 2023-03-06 R/functionalPlantIndicators/2 Reference bootstrap function.R

@@ -1,0 +1,61 @@
+# function to calculate community weighted means of selected indicator values (ind)
+# for species lists (sp) with given abundances (on a scale from 1 to 6) in one or more 'sites' (abun)
+# with a given number of iterations (iter),
+# with species given a certain minimum abundance ocurring in all bootstraps (obl), and
+# with a given resampling ratio of the original species list (rat)
+# in every bootstrap iteration the abundance of the sampled species can be 
+  # randomly changed by +/-1 if wished (var.abun)
+
+
+indBoot <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
+  
+  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
+  colnames(ind.b) <- colnames(abun)
+  ind.b <- as.data.frame(ind.b)  
+  
+  ind <- as.data.frame(ind)
+  ind.list <- as.list(1:length(colnames(ind)))
+  names(ind.list) <- colnames(ind)
+  
+  for (k in 1:length(colnames(ind)) ) 
+    { ind.list[[k]] <- ind.b }
+  
+  for (j in 1:length(colnames(abun)) ) {
+    
+    dat <- cbind(sp,abun[,j],ind)
+    dat <- dat[dat[,2]>0,]            # only species that are present in the ecoystem
+    dat <- dat[!is.na(dat[,3]),]      # only species that have Ellenberg values
+    
+    for (i in 1:iter) {
+      
+      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2] > (obl-1)])) *rat,0), replace=F)  
+      dat.b <- rbind(dat[dat[,2] > (obl-1),],
+                     dat[match(speciesSample,dat$sp),]
+                     )
+      
+      if (var.abun==T) {
+        for (l in 1: nrow(dat.b)) { dat.b[l,2] <- dat.b[l,2] + sample(c(-1,0,1),size=1) }
+        dat.b[!is.na(dat.b[,2]) & dat.b[,2]==(0),2] <- 1
+        dat.b[!is.na(dat.b[,2]) & dat.b[,2]==7,2] <- 6
+      }
+      
+      for (k in 1:length(colnames(ind))) {
+        
+        if ( nrow(dat.b)>2 ) {
+          
[…]

### 2023-03-06 R/functionalPlantIndicators/3 Reference bootstraps.R

@@ -1,0 +1,87 @@
+#### running bootstraps - wetlands ####
+colnames(NiN.wetland)
+mount.ind.list <- indBoot(sp=NiN.wetland[,1],abun=NiN.wetland[,2:46],ind=NiN.wetland[,47:51],
+                          iter=10000,obl=6,rat=1/3,var.abun=T)
+
+mount.ind.list.cov <- indBoot(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:51],
+                          iter=10000,obl=1,rat=1/3,var.abun=T)
+
+### saving backups
+#mount.ind.list.backup <- mount.ind.list
+#mount.ind.list.cov.backup <- mount.ind.list.cov
+
+
+### fixing NaN's
+for (i in 1:length(mount.ind.list) ) {
+  for (j in 1:ncol(mount.ind.list[[i]]) ) {
+    v <- mount.ind.list[[i]][,j]
+    v[is.nan(v)] <- NA
+    mount.ind.list[[i]][,j] <- v
+  }
+}
+
+for (i in 1:length(mount.ind.list.cov) ) {
+  for (j in 1:ncol(mount.ind.list.cov[[i]]) ) {
+    v <- mount.ind.list.cov[[i]][,j]
+    v[is.nan(v)] <- NA
+    mount.ind.list.cov[[i]][,j] <- v
+  }
+}
+
+
+#### checking NiN-types with several species lists ####
+# mountain
+str(mount.ind.list)
+# L, F, R, N, S, HM, T
+#indID <- colnames(ind.dat)[c(3:8)]
+indID <- colnames(ind.dat)[c(3,5,6)]
+NiNID <- colnames(mount.ind.list[[1]])
+
+for (i in indID) {
+  
+  dirOutput <- "C:/Users/joachim.topper/OneDrive - NINA/work/R projects/projects/økol tilst indikatorverdier/output/checks/mountain"
+  setwd(dirOutput)
+  name <- paste('Ellenberg',i,'.jpg',sep='')
+  jpeg(filename=name,width=4000,height=4000,res=300)
+  
+  par(mfrow=c(5,5))
+  for (j in NiNID) {
+    plot(density(mount.ind.list[[i]][,j],na.rm=T),main=j,xlim=c(3,8))
+  }
+  dev.off()
+}
+
+
+
+
+#### checking NiN-types with several species lists ####
+# mountain
[…]

### 2023-03-06 DRAFT_climate_indicators.Rmd
clean text

@@
 *Author and date:*
+
 Anders L. Kolstad
+
 March 2023
@@
 
-### Aditional comments about the dataset
-
-The data format has changes from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.
+### Additional comments about the data set
+
+The data format has recently changed from .BIL to .nc (netcdf) and now a single file contains all the rasters for one year (365 days), and sometimes for multiple variables also.
 
@@
 
-### Norwegain standard
+### Norwegian standard
 
@@
 
-The reference condition for climate variables is defined as the laste normal period 1961-1990.
+The reference condition for climate variables is defined as the last normal period 1961-1990.
 
@@
 
--   Un-scaled indicator value = median value over last 5 year (moving window)
+-   Un-scaled indicator value = median value over 5 year periods
 
@@
 
--   Threshold for good ecosystem condition = 2 standard deviation units for the climate variable during the reference period.
-
--   Lower reference value = 5 standard deviation units for the climate variable during the reference period.
+-   Thresholds for good ecosystem condition = 2 standard deviation units away from the upper reference level for the climate variable during the reference period.
+
+-   Lower reference values (two-directional) = 5 standard deviation units for the climate variable during the reference period (implies linear scaling).
 
@@
 for regions), the uncertainty in the indicator value is calculated from the spatial variation in the indicator values via bootstrapping.
+This might, however, be changed later to the temporal variation between the five years of each period.
 
@@
 
-rr and tm are being download from <https://thredds.met.no/thredds/catalog/senorge/seNorge_2018/Archive/catalog.html>
+*rr and tm are being download from:* <https://thredds.met.no/thredds/catalog/senorge/seNorge_2018/Archive/catalog.html>
 
@@
 
-```{r}
+```{r, fig.cap="Table explaining the available climate parameters in the data set."}
 senorgelayers <- read_delim("data/senorgelayers.txt", 
@@
 
-I think I know the CRS, so setting it manually.
+I know the CRS, so setting it manually. Although I cannot rule out 32633, I don't think it matters.
 
@@
 
-Checking that this is correct
-```{r}
-p = st_sfc(st_point(c(7039947.63, 270783.69)), crs=25833)
-```
-
 The data has three dimension
@@
 
-Initially the data had four attributes, but I subsettet to only include _tg_.
+Initially the data had four attributes, but I subsettet to only include *tg*.
 
@@
 
-Saving these to file to demonstrate the importing step of the workflow.
+Saving these to file.
 
@@
 
-Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well.
+Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well, in case I need them later.
 
@@
 
-```{r}
+```{r, message=FALSE}
 nor <- sf::st_read("data/outlineOfNorway_EPSG25833.shp")
 ```
-Remoce marine areas from regions
+
+Remove marine areas from regions
 
@@
 
-```{r, fig.cap="Five accounting areas (regions)."}
+```{r, fig.cap="Five accounting areas (regions) in Norway."}
 
@@
 
+The climate indicators need to be masked with ecosystem type maps. This step is part of this chapter.
+
 ## --------------------------------------------------
 
-### Conseptual workflow
-
-The general, conceptual workflow is like this:
+### Conceptual workflow
+
+The general, the conceptual workflow is like this:
 
@@
 
-    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute (above)
-
-    -   Filter data by dates (optional) (`dplyr::filter`)
-
-    -   Aggregate across time within a year (`stars::st_apply`)
+    -   Import .nc files (loop though year 1961-1990) and subset to the correct attribute
+
+    -   Filter data by dates (optional) (`dplyr::filter`).
+*This means reading all 365 rasters into memory, and it is much quicker to filter out the correct rasters in the importing step above (see examples later in this chapter)*
+
+    -   Aggregate across time within a year (`stars::st_apply`). *This is the most time consuming operation in the workflow.*
 
@@
 
-4.  Write to file
-
-5.  Mask by ecosystem type (this could also be done in step one to save time)
-
-6.  Aggregate in space (to accounting areas)
+4.  Mask by ecosystem type (*This could also be done in step one, but it doesn't speed things up to set some cells to NA*)
+
+5.  Aggregate in space (to accounting areas) (*zonal statistics*)
 
@@
 
-    -   Intersect with accounting area polygons `sf:st_intersects`
-
-    -   Bootstrap indicator values (cell values) and get median and sd
-
-7.  Make trend figure
-
-I will try to use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up. Also, masking to ecosystem time earlier in the workflow might prove smart.
+    -   Intersect with accounting area polygons `exactextrar::exact_extract` and get mean/median and (spatial) sd. (*Alternatively, get temporal sd at the grid cell level in the step above.*)
+
+6.  Make trend figure and spatially aggregated maps,
 
@@
 
+*Later I discover that \`dplyr::filter\` work after reading the whole object to file, and therefor it is too slow for our use. I therefore filter data a different way further down.*
+
 #### Aggregate across time within a year
 
-For this example I want to calculate the mean temperature over the summer. I therefore need to aggregate over time. There are two ways, either using `st_apply` or using `aggregate`.
+For this example I want to calculate the mean temperature over the summer.
+I therefore need to aggregate over time.
+Many climate indicators will need this functionality.
+There are two ways, either using `st_apply` or using `aggregate`.
 
@@
 
-The *v\_* follows the naming convention in Norway.
+The *v\_* follows the naming convention we have developed for ourselves.
 
@@
 
-```{r}
+```{r, fig.cap="Temporal aggregation on the dummy data set."}
 ggplot()+
@@
 
-_I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects.
+*I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects.
 Therefore I read the whole file in without proxy.
-To save time I can subset based on Julian dates, but this is a hard coded solution._
+To save time I can subset based on Julian dates.*
+
+I will use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up.
 
@@
 myFiles <- list.files(path2, pattern=".nc$",full.names = T)
-# The last file, the last year, is incomplete and dont include all juliani dates that we select below
+# The last file (the last year) is incomplete and don't include all julian dates that we select below, so I will not include it:
 myFiles <- myFiles[-length(myFiles)]
+
 real_temp_summer <- NULL
 
+# set up parallel cluster using 10 cores
 cl <- makeCluster(10L)
 
-# Get julian days
+# Get julian days after defining months
 temp <- stars::read_ncdf(paste(myFiles[1]), var="tg")
@@
   print(year_temp)
-  #start <- ym(paste(year_temp, start_month, sep="-"))
-  #end <-  ym(paste(year_temp, end_month, sep="-"))
-  #myInterval <- interval(start, end)
-  #myInterval <- interval(start, start+1)
-  lookup <- setNames("mean", paste0("v_", year_temp)) # Perhaps leave out the v_ to get a numeric vector instead, which is easier to subset
+  lookup <- setNames("mean", paste0("v_", year_temp)) 
+    # Perhaps leave out the v_ to get a numeric vector instead, 
+    # which is easier to subset
   st_crs(temp) <- 25833
@@
 
-This takes about 20 sec per file/year, or 22 min on total. Here's a test for the effect of splitting over more cores.
+This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 raster are read into memory. Here's a test for the effect of splitting over more cores.
+
 ```{r, eval=F}
@@
 ```
-No cluster: 21.401 sec elapsed
-Two clusters: 21.546 sec elapsed
-Six clusters: 14.548 sec elapsed
-Ten clusters: 15.29 sec elapsed
-Ten clusters (second run): 13.617 sec elapsed
+
+-   No cluster: 21.401 sec elapsed
+
+-   Two clusters: 21.546 sec elapsed
+
+-   Six clusters: 14.548 sec elapsed
+
+-   Ten clusters: 15.29 sec elapsed
+
+-   Ten clusters (second run): 13.617 sec elapsed
 
@@
 These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.
+*(This could perhaps have been solved with a `gc()`,.
+Later I save as RData with no problems).*
 
@@
 
-Note that GTiff automatically renames the third dimension _band_ and also renames the attribute.
+Note that GTiff automatically renames the third dimension *band* and also renames the attribute.
 
@@
 I can rename them.
+
 ```{r}
@@
 ### 2. Calculate reference values
+
 #### Aggregate across reference years
@@
 
-  user  system elapsed 
-  9.624   6.069  20.903
+user system elapsed 9.624 6.069 20.903
 
 This is also something I can export.
+
 ```{r, eval=F}
@@
 ```
-  user  system elapsed 
- 14.803   2.717  17.512
+
+user system elapsed 14.803 2.717 17.512
 
@@
 
-The _real_ indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming.
+The *real* indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming.
 Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.
@@
 
- user  system elapsed 
-  7.603   3.071  10.697
+user system elapsed 7.603 3.071 10.697
 
@@
 
-  user  system elapsed 
-  7.420   2.921  10.355
+user system elapsed 7.420 2.921 10.355
 

### 2023-03-06 DRAFT_climate_indicators.Rmd
finish cleaning

@@
 
-    -   Aggregate across reference years to get median and sd values (`st_apply)`
+    -   Aggregate (`st_apply)` across reference years (`dplyr::filter`) to get median and sd values
 
@@
 ```{r, eval=F}
-saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.rds")
-write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff")
+saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.rds")
+write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff")
 ```
@@
 ```{r}
-temp_median_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
+summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff",
                                  proxy=F)
 # This is too slow:
-# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/temp_median_summer.rds")
+# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/summer_median_temp.rds")
 # but it could be that RData works better 
@@
 ```{r, eval=T}
-names(temp_median_summer)
+names(summer_median_temp)
 ```
@@
 ```{r, eval=T}
-dim(temp_median_summer)
+dim(summer_median_temp)
 ```
@@
 ```{r, eval=T}
-st_get_dimension_values(temp_median_summer, "Year")
+st_get_dimension_values(summer_median_temp, "Year")
 ```
@@
 ```{r}
-st_get_dimension_values(temp_median_summer, "band")
+st_get_dimension_values(summer_median_temp, "band")
 ```
@@
 ```{r}
-temp_median_summer <- temp_median_summer %>% 
+summer_median_temp <- summer_median_temp %>% 
   st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
   setNames("temperature")
-dim(temp_median_summer)
+dim(summer_median_temp)
 ```
@@
 ggplot()+
-  geom_stars(data = temp_median_summer[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
+  geom_stars(data = summer_median_temp[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
   coord_equal() +
@@
 
-We need to first to define a reference period and then to subset our data `temp_median_summer`. We can practice on the dummy data `y`.
+We need to first to define a reference period and then to subset our data `summer_median_temp`. We can practice on the dummy data `y`.
 
@@
 ```{r}
-new_dims <- as.Date(paste0(substr(st_get_dimension_values(temp_median_summer, "v_YEAR"), 3, 6), "-01-01"))
-temp_median_summer_ref <- temp_median_summer %>%
+new_dims <- as.Date(paste0(
+  substr(st_get_dimension_values(summer_median_temp, "v_YEAR"), 3, 6), "-01-01"))
+summer_median_temp_ref <- summer_median_temp %>%
   st_set_dimensions("v_YEAR", values = new_dims)
@@
 ```{r}
-temp_median_summer_ref <- temp_median_summer_ref %>%
+summer_median_temp_ref <- summer_median_temp_ref %>%
   filter(v_YEAR %within% interval("1961-01-01", "1990-12-31"))
-st_get_dimension_values(temp_median_summer_ref, "v_YEAR")
+st_get_dimension_values(summer_median_temp_ref, "v_YEAR")
 ```
@@
 cl <- makeCluster(10L)
-temp_median_summer_ref <- temp_median_summer_ref %>%
+summer_median_temp_ref <- summer_median_temp_ref %>%
   st_apply(c("x", "y"), 
@@
 
-user system elapsed 9.624 6.069 20.903
+| user  | system | elapsed |
+|-------|--------|---------|
+| 9.624 | 6.069  | 20.903  |
 
@@
 # write_stars can only export a single attribute 
-#write_stars(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.tiff")
-
-saveRDS(temp_median_summer_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+#write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
+
+saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
 ```
@@
 ```{r, eval=T, include=F}
-temp_median_summer_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
-```
+summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
+```
+
+Pivot and turn dimension into attributes:
 
 ```{r, eval=T}
-temp_median_summer_ref_long <- temp_median_summer_ref %>% split("median_sd")
+summer_median_temp_ref_long <- summer_median_temp_ref %>% split("median_sd")
 ```
@@
 ```{r, eval=T}
-temp_median_summer_ref_long <- setNames(temp_median_summer_ref_long, c("reference_upper", "sd"))
+summer_median_temp_ref_long <- setNames(summer_median_temp_ref_long, c("reference_upper", "sd"))
 ```
@@
 tmap_arrange(
-tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
+tm_shape(st_downsample(summer_median_temp_ref_long, 10))+
   tm_raster("reference_upper")
 ,
-tm_shape(st_downsample(temp_median_summer_ref_long, 10))+
+tm_shape(st_downsample(summer_median_temp_ref_long, 10))+
   tm_raster("sd",
@@
 ```{r}
-y_var <- temp_median_summer %>%
+y_var <- summer_median_temp %>%
   split(3) %>%
-  c(temp_median_summer_ref_long)
+  c(summer_median_temp_ref_long)
 
@@
 # Mutate
-# can I find a way to keep the v_ columns?
 
@@
 
-user system elapsed 14.803 2.717 17.512
+| user   | system | elapsed |
+|--------|--------|---------|
+| 14.803 | 2.717  | 17.512  |
 
@@
 gc()
-saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
+saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
 ```
@@
 ```{r, include=FALSE}
-y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
+y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
 ```
@@
 
-The *real* indicator values should be running means over 5 years, but calculating a running mean for all time steps is too time consuming.
-Therefore, the time series will be presented with yearly resolution, and the scaled, regional indicator values will be calculated for distinct time steps.
+The *real* indicator values should be means over 5 year periods.
+Calculating a running mean for all time steps is too time consuming.
+Therefore, the scaled, regional indicator values will be calculated for distinct time steps.
+The time series can perhaps best be presented with yearly resolution.
 
@@
 
-We could also plot a time series as a series of maps, but then we need to avergae over this 5 time periods firs. We can get to that later.
+We could also plot a time series as a series of maps, but then we need to average over these 5 year time periods first. We can get to that later.
 
@@
 
-However, this does not preserve the region name, and a subsequent st_intersection taks a little whie and gives some boundary problems.
-
-I can also converting att pixels to points and then do the intersection.
+However, this does no preserve the region name, and a subsequent `st_intersection` taks a little while and gives some boundary problems.
+
+I can also convert pixels to points and then do the intersection.
 
@@
 
-I can rastereze the regions data set, turn it into a coded dimension and use st_apply, but thats tedious.
-
-Let me try using `exactextract`, as a parallel to `aggregate`. This requires us to turn to the `terra` package for a bit.
+I can rasterize the regions data set, turn it into a coded dimension and use st_apply, but that is tedious.
+
+Let me try using `exact_extract`, as an alternative to `aggregate`. This requires us to turn to the `terra` package for a bit.
 
@@
 
-user system elapsed 7.603 3.071 10.697
+user; system; elapsed: 7.603; 3.071; 10.697
 
@@
 
-user system elapsed 7.420 2.921 10.355
+user; system; elapsed: 7.420; 2.921; 10.355
 
@@
 
-```{r}
+```{r, fig.cap="Times series for median summer temperature centered on the median value during the reference period.
+The reference period is indicated with a thick horizontal line.
+Dottet horisontal lines are 2 sd units for the reference period."}
 regOrder = c("Østlandet","Sørlandet","Vestlandet","Midt-Norge","Nord-Norge")
@@
 
-```{r}
+```{r, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
 myCol <- "RdYlGn"
@@
 
-#tm_main <- tmap_grob(tm_main)
-#tm_inset <- tmap_grob(tm_inset)
-
-#ggdraw() +
-#  draw_plot(tm_main) +
-#  draw_plot(tm_inset,
-#    height = 0.4,
-#    x = 0.12, y = .35
-#  )
-
 tmap_arrange(tm_main, 
@@
 ```
-
-```{r}
-  ts1 <- y_var_norm %>%
-  select(all_of(starts_with("i_"))) %>%
-  st_redimension() %>%
-  st_apply(3, median_sd) %>%
-  as_tibble() %>%
-  setNames(c("metric", "i_year", "value") ) %>%
-  mutate(year = as.numeric(substr(i_year, 3, 7))) %>%
-  pivot_wider(names_from = "metric") %>%
-  select(-i_year) %>%
-  add_column(type = "indicator_value") %>%
-  rename(value = median) %>%
-  add_column(metric = "median")
-```
-
-```{r}
-ts2 <- y_var_norm %>%
-  select(all_of(starts_with("v_"))) %>%
-  st_redimension() %>%
-  st_apply(3, median_sd) %>%
-  as_tibble() %>%
-  setNames(c("metric", "v_year", "value") ) %>%
-  mutate(year = as.numeric(substr(v_year, 3, 7))) %>%
-  pivot_wider(names_from = "metric") %>%
-  select(-v_year) %>%
-  add_column(type = "variable_value") %>%
-  rename(value = median) %>%
-  add_column(metric = "median")
-```
-
-```{r}
-  ts3 <- y_var_norm %>%
-  select(!all_of(starts_with(c("v_", "i_")))) %>%
-  st_redimension() %>%
-  st_apply(3, median) %>%
-  as_tibble() %>%
-  setNames(c("type", "value")) %>%
-    add_column(metric = "median")
-```
-
-```{r}
-(ts <- bind_rows(ts1, ts2, ts3))
-```
-
-Calculate diference between indicator and reference We don't want to include this in ts, but as a temp file.
-
-```{r}
-tsX <- ts1
-```
-
-```{r}
-ts %>% filter(type == "indicator_value") %>%
-  ggplot(aes(x = year, y = value))+
-  geom_line()
-```
-
-### Scaled indicator values
-
-<!-- Text and analyses here -->
-
-### Uncertainty
-
-<!-- Text here -->
-
-## Prepare export
-
-<!-- Text here -->
-
-### Eksport file (final product)
-
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-03-07 DRAFT_climate_indicators.Rmd
Add review changes on DRAFT_climate_indicators.Rmd

@@
 
-## --------------------------------------------------
+<hl/>
 
@@
 
-```{r, fig.cap="Table explaining the available climate parameters in the data set."}
+```{r,warning=FALSE, message=FALSE, fig.cap="Table explaining the available climate parameters in the data set."}
 senorgelayers <- read_delim("data/senorgelayers.txt", 
@@
   adrop() %>%
-  mutate(accuntingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
-  select(accuntingAreas)
+  mutate(accountingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
+  select(accountingAreas)
 
@@
 ```{r}
-summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff",
+summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
                                  proxy=F)
@@
 ```{r, eval=T, include=F}
-summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
+summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
 ```
@@
 ```{r, include=FALSE}
-y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
+y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
 ```
@@
 
-```{r}
-regional_means
-```
-
 Reshape and plot

### 2023-03-07 DRAFT_climate_indicators.Rmd

@@
 
-The reference condition for climate variables is defined as the last normal period 1961-1990.
+The reference condition for climate variables is defined as the normal period 1961-1990.
 
@@
 
--   Un-scaled indicator value = median value over 5 year periods
+-   Un-scaled indicator value = median value over 5 year periods (5 years being a pragmatic choice.
+It is long enough to smooth out a lot of inter-annual variation, and it's long enough to enable estimating errors)
 
@@
 -   Lower reference values (two-directional) = 5 standard deviation units for the climate variable during the reference period (implies linear scaling).
+
+The choice to use 2 SD units as the threshold values is a subjective choice in many ways, and not founded in any empirical or known relationship between the indicators and ecosystem integrity.
+The value comes from the common practice of calling something _extreme weather_ when it is outside 2 SD units of the current running average.
+So, if the indicator value today is <0.6 it implies that the mean for that variable over the last year would have been referred to as _extreme_ if it had occurred between 1961 and 1990.
+This is I think a conservative threshold, since one would/could call it _extreme_ if only one year is outside the 2SD range, and having the mean of 5 years being outside this range is _really_ extreme.
 

### 2023-03-07 naturtype.Rmd
render

@@
   ungroup() %>%
+  mutate(count = replace_na(count, 0)) %>%
   mutate(NiN_mainType = fct_reorder(NiN_mainType, count)) %>%

### 2023-03-07 slitasje.Rmd
render

@@
 library(eaTools)
-library(tidyverse)
+#library(tidyverse)
+library(ggplot2)
+library(tidyr)
+library(dplyr)
 knitr::opts_chunk$set(echo = TRUE)

### 2023-03-07 R/functionalPlantIndicators/1 data handling.R
bootstrap, scaling 1st steps, ANO.geo NiN fixing

@@ -165,6 +165,11 @@ head(ind.dat)
 head(ANO.sp)
 head(ANO.geo)
 
+# fix NiN information
+ANO.geo$hovedtype_rute <- substr(ANO.geo$kartleggingsenhet_1m2,1,3) # take the 3 first characters
+ANO.geo$hovedtype_rute <- gsub("-", "", ANO.geo$hovedtype_rute) # remove hyphon
+unique(as.factor(ANO.geo$hovedtype_rute))
+
 # fix species names
 ANO.sp$Species <- ANO.sp$art_navn
 unique(as.factor(ANO.sp$Species))

### 2023-03-07 R/functionalPlantIndicators/3 Reference bootstraps.R

@@ -1,41 +1,45 @@
 #### running bootstraps - wetlands ####
 colnames(NiN.wetland)
-mount.ind.list <- indBoot(sp=NiN.wetland[,1],abun=NiN.wetland[,2:46],ind=NiN.wetland[,47:51],
-                          iter=10000,obl=6,rat=1/3,var.abun=T)
+wetland.ref <- indBoot(sp=NiN.wetland[,1],abun=NiN.wetland[,2:46],ind=NiN.wetland[,47:51],
+                          iter=1000,obl=6,rat=1/3,var.abun=T)
 
-mount.ind.list.cov <- indBoot(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:51],
-                          iter=10000,obl=1,rat=1/3,var.abun=T)
+wetland.ref.cov <- indBoot(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:51],
+                          iter=1000,obl=1,rat=1/3,var.abun=T)
 
 ### saving backups
-#mount.ind.list.backup <- mount.ind.list
-#mount.ind.list.cov.backup <- mount.ind.list.cov
+#wetland.ref.backup <- wetland.ref
+#wetland.ref.cov.backup <- wetland.ref.cov
+
+
+wetland.ref <- mount.ind.list
+wetland.ref.cov <- mount.ind.list.cov
 
 
 ### fixing NaN's
-for (i in 1:length(mount.ind.list) ) {
-  for (j in 1:ncol(mount.ind.list[[i]]) ) {
-    v <- mount.ind.list[[i]][,j]
+for (i in 1:length(wetland.ref) ) {
+  for (j in 1:ncol(wetland.ref[[i]]) ) {
+    v <- wetland.ref[[i]][,j]
     v[is.nan(v)] <- NA
-    mount.ind.list[[i]][,j] <- v
+    wetland.ref[[i]][,j] <- v
   }
 }
 
-for (i in 1:length(mount.ind.list.cov) ) {
-  for (j in 1:ncol(mount.ind.list.cov[[i]]) ) {
-    v <- mount.ind.list.cov[[i]][,j]
+for (i in 1:length(wetland.ref.cov) ) {
+  for (j in 1:ncol(wetland.ref.cov[[i]]) ) {
+    v <- wetland.ref.cov[[i]][,j]
     v[is.nan(v)] <- NA
[…]

### 2023-03-07 R/functionalPlantIndicators/4 scaled value calculations_wetland.R

@@ -1,0 +1,262 @@
+#### scaled values ####
+r.s <- 1    # reference value
+l.s <- 0.6  # limit value
+a.s <- 0    # abscence of indicator, or indicator at maximum
+
+#### function for calculating scaled values for measured value ####
+
+## scaling function including truncation
+scal <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value equal 1
+    if (val >= ref) {x <- 1}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
+    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
+    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
+    # value equals or lower than maxmin
+    if (val <= maxmin) {x <-0}
+  } else {
+    # values <= the reference value equal 1
+    if (val <= ref) {x <- 1}
+    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
+    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
+    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
+    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
+    # value equals or larger than maxmin
+    if (val >= maxmin) {x <-0}
+  }
+  return(x)
+  
+}
+
+## scaling function without truncation
+scal.2 <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
+    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
[…]

### 2023-03-08 slitasje.Rmd
import 2022 data

@@
 
-*Author and date:* Anders L. Kolstad
+*Author and date:*
+
+Anders L. Kolstad
+
+March 2023
 
@@
 
-pData <- ifelse(dir == "C:", 
-               "P:/41201785_okologisk_tilstand_2022_2023/data/",
-               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
 
@@
 ## Introduction
-This indicator represent human cause terrain damage from hiking or off-road vehicle traffic.
+This indicator represent human caused damage or **disturbance to soils and vegetation**, typically from things like recreational activities and off-road vehicle traffic.
+Effects from domestic animal is not included here, and the upper reference value is 0% disturbance.
 The data come from a combination non-random field surveys and area-representative surveys.
-Therefore extra care is taken not to over-extrapolate, and we use local reference area in the form of _homogeneous impact areas_.
+
+Since the data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
+We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region (_five regions in Norway_) where it was recorded.
+We then calculate an aerea weighted mean (and error) indicator value for each HIA and region combination, as long as there is more than 100 data point for a give combination of HIA and region.
+
+Here a general workflow for the calculation of the indicator.
 
@@
 
-The data goes back to 2018, but is because of data shortage, it is
-bulked into one time step. I then use the mean date for the raw data,
+The data goes back to 2018. I therefore bulk all the data from 2018 to 2022 into one time step. I then use the mean date for the raw data,
 and define the variable as belonging to the year 2020 (read more
@@
 perhaps *abiotiske forhold* is more suited. But the thought behind the
-choise is that *sliatsje* affect the potential for primary productivity.
+choice is that *slitasje* affect the potential for primary productivity.
 
@@
 amount-driven). The threshold for good ecosystem condition is 10%
-damage. This is on the conservative side. Perhaps is should be set
-somewhat higher, like 20 or 30%.
+damage. Perhaps is should be set somewhat higher, like 20 or 30%.
+The indicator value could be interpretted differently at the polygon
+vs at the landscape scale. At the polygon scale, 10% damage may seem 
+like not that much, whilst at the landscape scale, 10% seems more serious.
+This final threshold value should be debated and illustrated with examples.
 
@@
 bootstrapping individual indicator values 1000 times and calculating a
-distribution of area weighted means. When aggregating a second time,
+distribution of area weighted means. This uncertainty is different from 
+the spatial variation which we could get more straight forward without 
+bootstrapping. When aggregating a second time,
 from homogeneous impact areas to accounting areas, we assume a normal
@@
 
-<!-- Importing the full data set can be done like this, but because it can -->
-
-<!-- take a few seconds due to its size, I will cache the data on the P -->
-
-<!-- server after subsetting. -->
-
-```{r, eval=F}
-naturetypes <- sf::st_read(dsn = paste0(pData, "Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"))
-```
-
 We also have a separate [summary file](##exp-natureType-summary) where
-the nature types are manually mapped to the correct NiN-main types. We
-can use this to find the nature types associated with the NiN main types
+the nature types are manually mapped the NiN variables and to the correct NiN-main types. We can use this to find the nature types associated with the NiN main types
 that we are interested in.
@@
 ```{r}
-naturetypes_summary <- readRDS("data/naturetypes/natureType_summary.rds")
-```
-
-We are only interested in a few NiN main types and their mapping units
-(see [here](#naturtype) for justification)
-
-```{r}
-main_NiN_types <- c("T12",
-          "T18",
-          "T21",
-          "T31",
-          "T33",
-          "T34",
-          "V1",
-          "V3")
-```
-
-```{r}
-naturetypes_summary <- naturetypes_summary[naturetypes_summary$NiN_mainType %in% main_NiN_types,]
-```
-
-Sub-setting the main data file:
-
-```{r, eval=F}
-naturetypes <- naturetypes[naturetypes$naturtype %in% naturetypes_summary$Nature_type,]
-```
-
-Fix duplicated name for an ecosystem
-
-```{r, eval=F}
-naturetypes$hovedøkosystem[naturetypes$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
-```
-
-Export subsetted data set
-
-```{r, eval=F}
-saveRDS(naturetypes, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/rds/naturtyper.rds")
-```
-
-```{r, include=F}
-# Import back in
-naturetypes <- readRDS(paste0(pData, "rds/naturtyper.rds"))
-```
-
-```{r, fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation."}
+naturetypes_summary_import <- readRDS("data/naturetypes/natureType_summary.rds")
+```
+
+We are only interested in a mapping units that include our target variables.
+
+```{r}
+naturetypes_summary <- naturetypes_summary_import %>%
+  rowwise() %>%
+  mutate(keepers = sum(c_across(
+    c('7TK', '7SE', 'PRTK', 'PRSL'))>0, na.rm=T)) %>%
+  filter(keepers > 0) %>%
+  select(Nature_type, NiN_mainType, Year)
+```
+
+This deleted `r nrow(naturetypes_summary_import) - nrow(naturetypes_summary)` nature types and left us with these:
+
+```{r}
+DT::datatable(naturetypes_summary)
+```
+
+naturligÅpneOmråderILavlandet"] <- "naturligÅpneOmråderUnderSkoggrensa"
+
+Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
+```{r, message=F}
+naturetypes <- sf::st_read(dsn = path) %>%
+  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
+  mutate(hovedøkosystem = recode(hovedøkosystem,
+                                 "naturligÅpneOmråderUnderSkoggrensa" = "naturligÅpneOmråderILavlandet"),
+         area = st_area(.)) %>%
+  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
+  separate(col = ninBeskrivelsesvariable,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F) %>%
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
+  filter(NiN_variable_code %in% c("7SE", "7TK", "PRTK", "PRSL"))
+```
+
+```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation.", fig.height=10, fig.width=4}
 ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
@@
 
-Calculate area for each polygon.
-
-```{r}
-naturetypes$area <- sf::st_area(naturetypes)
-```
-
-The NiN variables are combined into a single field which we need to spit
-apart.
-
-```{r}
-naturetypes <- tidyr::separate_rows(naturetypes,
-                                    ninbeskrivelsesvariabler,
-                                    sep=",") %>%
-  separate(col = ninbeskrivelsesvariabler,
-           into = c("NiN_variable_code", "NiN_variable_value"),
-           sep = "_",
-           remove=F
-           )
-```
-
-Convert the value column to numeric. Ignore NA's.
-
-```{r}
-naturetypes$NiN_variable_value <- as.numeric(naturetypes$NiN_variable_value)
-```
-
-Keep only the four relevant variables.
-
-```{r}
-myVars <- c(
-  #"7FA",     #treated separately in a different indicator
-  "7SE", "7TK", "PRTK", "PRSL")
-naturetypes <- filter(naturetypes,
-                      NiN_variable_code %in% myVars)
-```
-
 Column names starting with a number is problematic, so adding a prefix
@@
                                 values_from = "NiN_variable_value",
-                                id_cols = "identifikasjon_lokalid")
+                                id_cols = "identifikasjon_lokalId")
 naturetypes_wide <- as.data.frame(naturetypes_wide)

### 2023-03-08 R/functionalPlantIndicators/1 data handling.R
scaling and various edits

@@ -170,6 +170,16 @@ ANO.geo$hovedtype_rute <- substr(ANO.geo$kartleggingsenhet_1m2,1,3) # take the 3
 ANO.geo$hovedtype_rute <- gsub("-", "", ANO.geo$hovedtype_rute) # remove hyphon
 unique(as.factor(ANO.geo$hovedtype_rute))
 
+# check that every point is present only once
+length(levels(as.factor(ANO.geo$ano_flate_id)))
+length(levels(as.factor(ANO.geo$ano_punkt_id)))
+summary(as.factor(ANO.geo$ano_punkt_id))
+# there's many double presences, probably some wrong registrations of point numbers,
+# but also double registrations (e.g. ANO0159_55)
+# CHECK THIS when preparing ecosystem-datasets for scaling
+
+
+
 # fix species names
 ANO.sp$Species <- ANO.sp$art_navn
 unique(as.factor(ANO.sp$Species))
@@ -182,7 +192,7 @@ ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital
 
 ## merge species data with indicators
 ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
-                y= ind.dat[,c("species","CC", "SS", "RR", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
+                y= ind.dat[,c("species","CC", "SS", "RR","Continentality", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
                 by.x="Species", by.y="species", all.x=T)
 summary(ANO.sp.ind)
 
@@ -263,7 +273,7 @@ ANO.sp <- ANO.sp %>%
 
 ## merge species data with indicators
 ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
-                    y= ind.dat[,c("species","CC", "SS", "RR", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
+                    y= ind.dat[,c("species","CC", "SS", "RR","Continentality","Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Grazing_mowing")],
                     by.x="Species", by.y="species", all.x=T)
 summary(ANO.sp.ind)
[…]

### 2023-03-08 R/functionalPlantIndicators/3 Reference bootstraps.R

@@ -11,10 +11,6 @@ wetland.ref.cov <- indBoot(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],in
 #wetland.ref.cov.backup <- wetland.ref.cov
 
 
-wetland.ref <- mount.ind.list
-wetland.ref.cov <- mount.ind.list.cov
-
-
 ### fixing NaN's
 for (i in 1:length(wetland.ref) ) {
   for (j in 1:ncol(wetland.ref[[i]]) ) {
@@ -73,7 +69,7 @@ for (i in indID) {
   
   par(mfrow=c(5,9))
   for (j in NiNID) {
-    plot(density(wetland.ref.cov[[i]][,j],na.rm=T),main=j,xlim=c(1,10))
+    plot(density(wetland.ref.cov[[i]][,j],na.rm=T),main=j,xlim=c(-2,10))
   }
   dev.off()
 }

### 2023-03-08 R/functionalPlantIndicators/4 Reference scaling values.R

@@ -1,0 +1,246 @@
+#### defining reference limits, mountain ####
+
+# limit table Ellenberg and aliens, mountain, median and 95% CI
+# NB! working at the 1:5000 mapping unit level !!!
+
+
+## extract from mountain ref-list
+
+# every NiN-type is represented by one 'generalisert artsliste'
+# some NiN-types are represented by two species lists
+# in some cases two NiN-types are represented by the same species list
+head(wetland.ref[[1]])
+wetland.ref[[1]][0,]
+
+# checking the actual NiN-types in the wetland lists
+wetland.NiNtypes <- colnames(wetland.ref[["Continentality"]])
+wetland.NiNtypes <- substr(wetland.NiNtypes,1,5)
+wetland.NiNtypes2 <- wetland.NiNtypes[-x]
+unique(wetland.NiNtypes2)
+
+# 5 indicator-value indicators: Tyler's Continentality, Light, Moisture, Soil_reaction_pH, "Nitrogen"
+indEll.n=5
+# creating a table to hold:
+# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
+# for every nature type (nrows)
+tab <- matrix(ncol=3*indEll.n, nrow=length(unique(wetland.NiNtypes)) ) # 43 basic ecosystem types
+# coercing the values into the table
+# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
+x <- c(26,27,39:43)
+
+for (i in 1:length(x) ) {
+  tab[i,1:3] <- quantile(as.matrix(wetland.ref[["Continentality"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,4:6] <- quantile(as.matrix(wetland.ref[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,7:9] <- quantile(as.matrix(wetland.ref[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,10:12] <- quantile(as.matrix(wetland.ref[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
[…]

### 2023-03-08 R/functionalPlantIndicators/4 scaled value calculations_wetland.R

@@ -60,22 +60,58 @@ scal.2 <- function() {
 }
 
 
+unique(ANO.geo$hovedtype_rute) # NiN types in data
+unique(substr(wetland.ref.val$grunn,1,2)) # NiN types in reference
+#### creating dataframe to holde the results for wetlands ####
+# all ANO points
+nrow(ANO.geo)
+# all wetland ANO points
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),])
+# all wetland ANO points with a NiN-type represented in the reference
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.val$grunn,1,2)),])
+# ok, we'll be losing 70 (our of 1349) that are not covered by the reference
+ANO.wetland <- ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),]
 
-#### creating dataframe to holde the results for wetlandains ####
-ANO.wetland <- ANO.geo[ANO.geo$hovedoekosystem_punkt=='vaatmark',]
+head(ANO.wetland)
+# update row-numbers
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland)
+head(ANO.wetland)
+dim(ANO.wetland)
+colnames(ANO.wetland)
+
+length(levels(as.factor(ANO.wetland$ano_flate_id)))
+length(levels(as.factor(ANO.wetland$ano_punkt_id)))
+summary(as.factor(ANO.wetland$ano_punkt_id))
+# four points that are double
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0159_55",] # double registration, said so in comment. -> choose row 207 over 206
+ANO.wetland <- ANO.wetland[-206,]
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0283_22",] # 2019 & 2021. Lot of NA's in 2019 -> omit 2019
+ANO.wetland <- ANO.wetland[-156,]
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0363_24",]
[…]

### 2023-03-08 slitasje.Rmd
convert to continous

@@
 ```{r}
+myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
+```
+
+```{r}
 naturetypes_summary <- naturetypes_summary_import %>%
@@
   mutate(keepers = sum(c_across(
-    c('7TK', '7SE', 'PRTK', 'PRSL'))>0, na.rm=T)) %>%
+    myVars)>0, na.rm=T)) %>%
   filter(keepers > 0) %>%
@@
 ```
-
-naturligÅpneOmråderILavlandet"] <- "naturligÅpneOmråderUnderSkoggrensa"
 
@@
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
-  filter(NiN_variable_code %in% c("7SE", "7TK", "PRTK", "PRSL"))
+  filter(NiN_variable_code %in% myVars)
 ```
@@
 is a product/function of the four variables "7SE", "7TK", "PRTK" and
-"PRSL". I will create a new data table where I calculate the new
+"PRSL".
+I want to base this indicator on whichever of the variables have the highest value (worst condition).
+Therefore I first convert the ordinal scale to a continuous scale, using the median value of each ordinal step.
+
+The variables use slightly different scales. PRSL and PRTK use this 8
+step scale:
+
+```{r eight-step, fig.cap="Eight step condition scale"}
+knitr::include_graphics("images/mdirPRscale.PNG")
+```
+
+7TK and 7SE use a 4 step scale
+
+```{r four-step, fig.cap="Four step condition scale"}
+knitr::include_graphics("images/ninAR4b.PNG")
+```
+
+```{r}
+naturetypes <- naturetypes %>%
+  mutate(NiN_variable_value = case_when(
+    NiN_variable_code %in% c("var_7TK", "var_7SE") ~ 
+      case_match(NiN_variable_value,
+             0 ~ 0,
+             1 ~ mean(c(0, 1/16))*100,
+             2 ~ mean(c(1/16, 1/2))*100,
+             3 ~ 75),
+    NiN_variable_code %in% c('var_PRTK', 'var_PRSL') ~ 
+      case_match(NiN_variable_value,
+             0 ~ 0,
+             1 ~ 1.5,
+             2 ~ mean(c(3, 6.25)),
+             3 ~ mean(c(6.25, 12.5)),
+             4 ~ mean(c(12.5, 25)),
+             5 ~ mean(c(25, 50)),
+             6 ~ mean(c(50, 75)),
+             7 ~ mean(c(75, 100)))
+    )
+  )
+```
+
+I will create a new data table where I calculate the new
 variable which I then paste back into the sf object.
@@
 naturetypes_wide <- as.data.frame(naturetypes_wide)
-head(naturetypes_wide)
-```
-
-The variables use slightly different scales. PRSL and PRTK use this 8
-step scale:
-
-```{r eight-step, fig.cap="Eight step condition scale"}
-knitr::include_graphics("images/mdirPRscale.PNG")
-```
-
-7TK and 7SE use a 4 step scale
-
-```{r four-step, fig.cap="Four step condition scale"}
-knitr::include_graphics("images/ninAR4b.PNG")
-```
-
-I want to base this indicator on whichever of the variables have the
-highest value (worst condition). This table shows that a locality may
-have been assessed with both a PR variable and a NiN variable:
-
-```{r}
-table(naturetypes_wide$var_7SE, naturetypes_wide$var_PRTK)
-```
-
-Therefore I need to reconcile these scales. The only way I can think of
-is to decrease the resolution of the PR variables to the ar4b scale.
-This is of course a shame to loose this detail.
-
-A locality will never have values for both 7SE and PRSL
-
-```{r}
-table(naturetypes_wide$var_7SE, naturetypes_wide$var_PRSL)
-```
-
-nor 7TK and PRTK
-
-```{r}
-table(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK)
-```
-
-```{r}
-naturetypes_wide$var_7TK <-  ifelse(is.na(naturetypes_wide$var_7TK),
-                              ifelse(naturetypes_wide$var_PRTK == 0, 0,
-                                ifelse(naturetypes_wide$var_PRTK < 3, 1,
-                                  ifelse(naturetypes_wide$var_PRTK < 6, 2, 3))),
-                            naturetypes_wide$var_7TK)
-```
-
-Checking that I successfully combined classes 1+2, 3+4+5 and 6+7.
-
-```{r}
-table(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK,
-      deparse.level = 2)
-```
-
-That looks fine. Now doing the same for the other variable PRSL:
-
-```{r}
-naturetypes_wide$var_7SE <-  ifelse(is.na(naturetypes_wide$var_7SE),
-                              ifelse(naturetypes_wide$var_PRSL == 0, 0,
-                                ifelse(naturetypes_wide$var_PRSL < 3, 1,
-                                  ifelse(naturetypes_wide$var_PRSL < 6, 2, 3))),
-                            naturetypes_wide$var_7SE)
-```
-
-Remove the old PR variables.
-
-```{r}
-naturetypes_wide <- select(naturetypes_wide,
-                           -var_PRTK,
-                           -var_PRSL)
-```
-
-There are some very few cells with X's. These are NA's.
-
-```{r}
-naturetypes_wide$var_7SE[naturetypes_wide$var_7SE == "X"] <- NA
-naturetypes_wide$var_7TK[naturetypes_wide$var_7TK == "X"] <- NA
-
-naturetypes_wide$var_7SE <- as.numeric(naturetypes_wide$var_7SE)
-naturetypes_wide$var_7TK <- as.numeric(naturetypes_wide$var_7TK)
-```
-
-```{r}
-head(naturetypes_wide)
-```
-
-```{r}
-anyDuplicated(naturetypes_wide$identifikasjon_lokalid)
+head(naturetypes_wide, 10)
 ```
@@
 ```{r, warning=F}
-naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
-```
-
-When both variables are NA, we get -Inf. There were nine cases of this
+naturetypes_wide$slitasje <- apply(naturetypes_wide[,-1], 1, max, na.rm=T)
+```
+
+When both variables are NA, we get -Inf. There were 13 cases of this
 (in 2023). Removing these now.
@@
   theme_bw(base_size = 12)+
-  labs(x = "slitasje score",
-       y = "Number of localities")
+  labs(x = "Slitasje score converted to percentage",
+       y = "Number of localities")+
+  coord_flip()
 ```
@@
 slitasje-indicator, and how often it was 7SE. I can do this by taking
-the difference
-
-```{r, fig.cap="Counting the number of cases where 7TK or 7SE ws defining of the slitasje indicator score.", fig.width=3}
-diff <- naturetypes_wide$var_7SE - 
-        naturetypes_wide$var_7TK
+the difference.
+
+```{r, fig.cap="Counting the number of cases where 7TK or 7SE was defining of the slitasje indicator score.", fig.width=3}
+
+diff <- max(naturetypes_wide$var_7SE, naturetypes_wide$var_PRSL, na.rm=T) - 
+        max(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK, na.rm=T)
 

### 2023-03-09 DRAFT_functional plant indicators_wetland.Rmd
markdown start

--- 2022-08-12 AndersTest.Rmd
@@
 ---
-
-This is a test
 
@@
 <!-- Replace 'indicator name' with yor own indicator name  -->
-# Indicator name
+# Functional plant indicators (Moisture, Light, pH, Nitrogen)
 
@@
 _Author and date:_
-<!-- Write your name here -->
+Joachim Töpper
 
@@
 ```{r setup, include=FALSE}
+library(sf)
+library(tidyr)
+library(dplyr)
+library(plyr)
+library(stringr)
+library(tidyverse)
+library(readxl)
 library(knitr)
@@
 ```{r, echo=F}
-Ecosystem <- "" # e.g. "Skog og fjell"
-Egenskap  <- "" # e.g. "Primærproduksjon"
+Ecosystem <- "våtmark"
+Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
 ECT       <- "" # e.g. "Structural state characteristic"
-Contact   <- "" # e.g. "Anders Kolstad"
+Contact   <- "Joachim Töpper"
 ```
@@
 ## Ecosystem characteristic
-### Norwegain standard
+### Norwegian standard
 <!-- Text here -->
@@
 ### Data sets
-<!-- Text here -->
 
-#### Data set A
-<!-- Text here -->
+ANO data
+
+```{r}
+st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
+ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+                   layer="ANO_Art")
+ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+                  layer="ANO_SurveyPoint")
+head(ANO.sp)
+head(ANO.geo)
+```
+
+Plant indicators from Tyler et al. 2021 and Grime
+```{r}
+## Tyler indicator data
+ind.dat <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
+                        sep = '\t', header=T, quote = '')
+head(ind.datr)
+
+```
+
+Generalized species lists NiN
+```{r}
+load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
+str(Eco_State)
+```
+
+#### Plant indicator data handling
+```{r}
+
+ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
+ind.dat <- ind.dat %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
+                                                            "Anemone nemorosa x ranunculoides",
+                                                            "Armeria maritima ssp. elongata",
+                                                            "Asplenium trichomanes ssp. quadrivalens",
+                                                            "Calystegia sepium ssp. spectabilis",
+                                                            "Campanula glomerata 'Superba'",
+                                                            "Dactylorhiza maculata ssp. fuchsii",
+                                                            "Erigeron acris ssp. droebachensis",
+                                                            "Erigeron acris ssp. politus",
+                                                            "Erysimum cheiranthoides L. ssp. alatum",
+                                                            "Euphrasia nemorosa x stricta var. brevipila",
+                                                            "Galium mollugo x verum",
+                                                            "Geum rivale x urbanum",
+                                                            "Hylotelephium telephium (ssp. maximum)",
+                                                            "Juncus alpinoarticulatus ssp. rariflorus",
+                                                            "Lamiastrum galeobdolon ssp. argentatum",
+                                                            "Lathyrus latifolius ssp. heterophyllus",
+                                                            "Medicago sativa ssp. falcata",
+                                                            "Medicago sativa ssp. x varia",
+                                                            "Monotropa hypopitys ssp. hypophegea",
+                                                            "Ononis spinosa ssp. hircina",
+                                                            "Ononis spinosa ssp. procurrens",
+                                                            "Pilosella aurantiaca ssp. decolorans",
+                                                            "Pilosella aurantiaca ssp. dimorpha",
+                                                            "Pilosella cymosa ssp. gotlandica",
+                                                            "Pilosella cymosa ssp. praealta",
+                                                            "Pilosella officinarum ssp. peleteranum",
+                                                            "Poa x jemtlandica (Almq.) K. Richt.",
+                                                            "Poa x herjedalica Harry Sm.",
+                                                            "Ranunculus peltatus ssp. baudotii",
+                                                            "Sagittaria natans x sagittifolia",
+                                                            "Salix repens ssp. rosmarinifolia",
+                                                            "Stellaria nemorum L. ssp. montana",
+                                                            "Trichophorum cespitosum ssp. germanicum")
+) )
+ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
+# getting rid of sect. for Hieracium
+ind.dat <- ind.dat %>% mutate(species=gsub("sect. ","",species.orig))
+ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2)
+
+ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
+# only hybrids left -> get rid of these
+ind.dat <- ind.dat[!duplicated(ind.dat[,'species']),]
+ind.dat[duplicated(ind.dat[,'species']),"species"]
+
+ind.dat$species <- as.factor(ind.dat$species)
+summary(ind.dat$species)
+# no duplicates left
+```
 

### 2023-03-09 DRAFT_functional plant indicators_wetland.Rmd
data handling wetland Rmd continued

@@
 
-Plant indicators from Tyler et al. 2021 and Grime
+Plant indicators from Tyler et al. 2021
 ```{r}
@@
 
-#### Plant indicator data handling
+#### Plant indicator data - data handling
 ```{r}
 
-ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
+names(ind.dat)
+names(ind.dat)[1] <- 'species'
+ind.dat$species <- as.factor(ind.dat$species)
+summary(ind.dat$species)
+ind.dat <- ind.dat[!is.na(ind.dat$species),] # removing species-NAs
+ind.dat[,'species.orig'] <- ind.dat[,'species'] # make a backup of the original species variable
+ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2) # trimming away sub-species & co, and descriptor info
+
+# the trimming above leaves some duplicates that need to be taken care of
+ind.dat[duplicated(ind.dat[,'species']),"species"] ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
 ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
@@
 ) )
+
+# checking duplicates again
 ind.dat[duplicated(ind.dat[,'species']),"species"]
@@
 # no duplicates left
+
+head(ind.dat)
+
 ```
 
-#### Data set B
-<!-- Text here -->
+#### ANO monitoring data - data handling
+```{r}
+head(ANO.sp)
+head(ANO.geo)
+
+## fix NiN information
+ANO.geo$hovedtype_rute <- substr(ANO.geo$kartleggingsenhet_1m2,1,3) # take the 3 first characters
+ANO.geo$hovedtype_rute <- gsub("-", "", ANO.geo$hovedtype_rute) # remove hyphen
+unique(as.factor(ANO.geo$hovedtype_rute))
+
+## fix NiN-variables
+colnames(ANO.geo)
+colnames(ANO.geo)[42:47] <- c("groeftingsintensitet",
+                              "bruksintensitet",
+                              "beitetrykk",
+                              "slatteintensitet",
+                              "tungekjoretoy",
+                              "slitasje")
+head(ANO.geo)
+
+# remove variable code in the data
+ANO.geo$groeftingsintensitet <- gsub("7GR-GI_", "", ANO.geo$groeftingsintensitet) 
+unique(ANO.geo$groeftingsintensitet)
+ANO.geo$groeftingsintensitet <- gsub("X", "NA", ANO.geo$groeftingsintensitet)
+unique(ANO.geo$groeftingsintensitet)
+ANO.geo$groeftingsintensitet <- as.numeric(ANO.geo$groeftingsintensitet)
+unique(ANO.geo$groeftingsintensitet)
+
+ANO.geo$bruksintensitet <- gsub("7JB-BA_", "", ANO.geo$bruksintensitet) 
+unique(ANO.geo$bruksintensitet)
+ANO.geo$bruksintensitet <- gsub("X", "NA", ANO.geo$bruksintensitet)
+unique(ANO.geo$bruksintensitet)
+ANO.geo$bruksintensitet <- as.numeric(ANO.geo$bruksintensitet)
+unique(ANO.geo$bruksintensitet)
+
+ANO.geo$beitetrykk <- gsub("7JB-BT_", "", ANO.geo$beitetrykk) 
+unique(ANO.geo$beitetrykk)
+ANO.geo$beitetrykk <- gsub("X", "NA", ANO.geo$beitetrykk)
+unique(ANO.geo$beitetrykk)
+ANO.geo$beitetrykk <- as.numeric(ANO.geo$beitetrykk)
+unique(ANO.geo$beitetrykk)
+
+ANO.geo$slatteintensitet <- gsub("7JB-SI_", "", ANO.geo$slatteintensitet) 
+unique(ANO.geo$slatteintensitet)
+ANO.geo$slatteintensitet <- gsub("X", "NA", ANO.geo$slatteintensitet)
+unique(ANO.geo$slatteintensitet)
+ANO.geo$slatteintensitet <- as.numeric(ANO.geo$slatteintensitet)
+unique(ANO.geo$slatteintensitet)
+
+ANO.geo$tungekjoretoy <- gsub("7TK_", "", ANO.geo$tungekjoretoy) 
+unique(ANO.geo$tungekjoretoy)
+ANO.geo$tungekjoretoy <- gsub("X", "NA", ANO.geo$tungekjoretoy)
+unique(ANO.geo$tungekjoretoy)
+ANO.geo$tungekjoretoy <- as.numeric(ANO.geo$tungekjoretoy)
+unique(ANO.geo$tungekjoretoy)
+
+ANO.geo$slitasje <- gsub("7SE_", "", ANO.geo$slitasje) 
+unique(ANO.geo$slitasje)
+ANO.geo$slitasje <- gsub("X", "NA", ANO.geo$slitasje)
+unique(ANO.geo$slitasje)
+ANO.geo$slitasje <- as.numeric(ANO.geo$slitasje)
+unique(ANO.geo$slitasje)
+
+## check that every point is present only once
+length(levels(as.factor(ANO.geo$ano_flate_id)))
+length(levels(as.factor(ANO.geo$ano_punkt_id)))
+summary(as.factor(ANO.geo$ano_punkt_id))
+# there's many double presences, probably some wrong registrations of point numbers,
+# but also double registrations (e.g. ANO0159_55)
+# CHECK THIS when preparing ecosystem-datasets for scaling
+
+# fix species names
+ANO.sp$Species <- ANO.sp$art_navn
+unique(as.factor(ANO.sp$Species))
+ANO.sp[,'Species'] <- word(ANO.sp[,'Species'], 1,2) # lose subspecies
+ANO.sp$Species <- str_to_title(ANO.sp$Species) # make first letter capital
+ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after hyphen to lowercase
+ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after space to lowercase
+
+## merge species data with indicators
+ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
+                y= ind.dat[,c("species", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")],
+                by.x="Species", by.y="species", all.x=T)
+summary(ANO.sp.ind)
+
+## checking which species didn't find a match
+unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
+
+# fix species name issues
+ind.dat <- ind.dat %>% 
+  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
+  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
+  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
+  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
+  mutate(species=str_replace(species,"Cherleria biflora", "Minuartia biflora"))
+
+ANO.sp <- ANO.sp %>% 
+  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
+  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
+  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
+  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
+  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
+  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
+  mutate(Species=str_replace(Species,"Carex paupercula", "Carex magellanica")) %>%
+  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+  mutate(Species=str_replace(Species,"Carex viridula", "Carex flava")) %>%
+  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
+  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
+  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
+  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
+  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
+  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
+  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
+  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
+  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
+  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
+  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
+  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
+  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Vulgata")) %>%
+  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
+  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
+  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
+  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
+  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
+  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
+  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
+  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
+  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
+  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
+  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
+  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
+  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
+  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
+  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
+  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
+  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
+  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
+  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
+  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
+  mutate(Species=str_replace(Species,"Poa alpigena", "Poa pratensis")) %>%
+  mutate(Species=str_replace(Species,"Poa angustifolia", "Poa pratensis")) %>%
+  mutate(Species=str_replace(Species,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
+  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa")) %>%
+  mutate(Species=str_replace(Species,"Syringa emodi", "Syringa vulgaris")) %>%
+  mutate(Species=str_replace(Species,"Taraxacum crocea", "Taraxacum officinale")) %>%
+  mutate(Species=str_replace(Species,"Taraxacum croceum", "Taraxacum officinale")) %>%
+  mutate(Species=str_replace(Species,"Trientalis europaea", "Lysimachia europaea")) %>%
+  mutate(Species=str_replace(Species,"Trifolium pallidum", "Trifolium pratense"))
+
+## merge species data with indicators
+ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
+                    y= ind.dat[,c("species", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")],
+                    by.x="Species", by.y="species", all.x=T)
+summary(ANO.sp.ind)
+# checking which species didn't find a match
+unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
+# don't find synonyms for these in the ind lists
+
+## trimming away the points without information on NiN, species or cover
+ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$Species),]
+ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$art_dekning),]
+
+summary(ANO.sp.ind)
+head(ANO.sp.ind)
+rm(ANO.sp)
+
+```
+
+#### NiN reference data - data handling
+
+```{r}
+
+## generalized species lists from NiN
+str(Eco_State)
+
+# species
+Eco_State$Concept_Data$Species$Species_List$species
+# environments
+t(Eco_State$Concept_Data$Env$Env_Data)
+# abundances
+t(Eco_State$Concept_Data$Species$Species_Data)
+
+## transposing abundance data
+NiN.sp <- t(Eco_State$Concept_Data$Species$Species_Data)
+NiN.sp <- as.data.frame(NiN.sp)
+NiN.sp$sp <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$species))
+# only genus and species name
+NiN.sp$sp <- word(NiN.sp$sp, 1,2)
+NiN.sp$spgr <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code))
+# if relevant, trimming to desired species groups (for forests e.g. removing trees)
+#NiN.sp <- NiN.sp[NiN.sp$spgr!="a1a",]
+
+## environment data
+NiN.env <- Eco_State$Concept_Data$Env$Env_Data
+
+## merging with indicator values
+NiN.sp.ind <- merge(NiN.sp,ind.dat, by.x="sp", by.y="species", all.x=T)
+summary(NiN.sp.ind)
+
+NiN.sp.ind[NiN.sp.ind==999] <- NA
+
+## checking which species didn't find a match
+unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) & NiN.sp.ind$spgr %in% list("a1a","a1b","a1c"),'sp'])
+
+## fix species name issues
+ind.dat <- ind.dat %>% 
+  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
+  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
+  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
+  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
+  mutate(species=str_replace(species,"Cherleria biflora", "Minuartia biflora"))
+
+NiN.sp <- NiN.sp %>% 
+  mutate(sp=str_replace(sp,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
+  mutate(sp=str_replace(sp,"Anagallis minima", "Lysimachia minima")) %>% 
+  mutate(sp=str_replace(sp,"Arctous alpinus", "Arctous alpina")) %>%
+  mutate(sp=str_replace(sp,"Betula tortuosa", "Betula pubescens")) %>%
+  mutate(sp=str_replace(sp,"Blysmopsis rufa", "Blysmus rufus")) %>%
+  mutate(sp=str_replace(sp,"Cardamine nymanii", "Cardamine pratensis")) %>%
+  mutate(sp=str_replace(sp,"Carex adelostoma", "Carex buxbaumii")) %>%
+  mutate(sp=str_replace(sp,"Carex leersii", "Carex echinata")) %>%
+  mutate(sp=str_replace(sp,"Carex paupercula", "Carex magellanica")) %>%
+  mutate(sp=str_replace(sp,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
+  mutate(sp=str_replace(sp,"Carex _vacillans", "Carex vacillans")) %>%
+  mutate(sp=str_replace(sp,"Carex viridula", "Carex flava")) %>%
+  mutate(sp=str_replace(sp,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
+  mutate(sp=str_replace(sp,"Cornus suecia", "Cornus suecica")) %>%
+  mutate(sp=str_replace(sp,"Cicerbita alpina", "Lactuca alpina")) %>%
+  mutate(sp=str_replace(sp,"Dactylorhiza sphagnicola", "Dactylorhiza majalis")) %>%
+  mutate(sp=str_replace(sp,"Diphasiastrum complanatum", "Lycopodium complanatum")) %>%
+  mutate(sp=str_replace(sp,"Elymus alaskanus", "Elymus kronokensis")) %>%
+  mutate(sp=str_replace(sp,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
+  mutate(sp=str_replace(sp,"Erigeron eriocephalus", "Erigeron uniflorus")) %>%
+  mutate(sp=str_replace(sp,"Festuca prolifera", "Festuca rubra")) %>%
+  mutate(sp=str_replace(sp,"Galium album", "Galium mollugo")) %>%
+  mutate(sp=str_replace(sp,"Galium elongatum", "Galium palustre")) %>%
+  mutate(sp=str_replace(sp,"Glaux maritima", "Lysimachia maritima")) %>%
+  mutate(sp=str_replace(sp,"Helictotrichon pratense", "Avenula pratensis")) %>%
+  mutate(sp=str_replace(sp,"Helictotrichon pubescens", "Avenula pubescens")) %>%
+  mutate(sp=str_replace(sp,"Hieracium alpina", "Hieracium Alpina")) %>%
+  mutate(sp=str_replace(sp,"Hieracium alpinum", "Hieracium Alpina")) %>%
+  mutate(sp=str_replace(sp,"Hieracium aurantiacum", "Pilosella aurantiaca")) %>%
+  mutate(sp=str_replace(sp,"Hieracium hieracium", "Hieracium Hieracium")) %>%
+  mutate(sp=str_replace(sp,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
+  mutate(sp=str_replace(sp,"Hieracium lactucella", "Pilosella lactucella")) %>%
+  mutate(sp=str_replace(sp,"Hieracium murorum", "Hieracium Vulgata")) %>%
+  mutate(sp=str_replace(sp,"Hieracium oreadea", "Hieracium Oreadea")) %>%
+  mutate(sp=str_replace(sp,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
+  mutate(sp=str_replace(sp,"Hieracium vulgata", "Hieracium Vulgata")) %>%
+  mutate(sp=str_replace(sp,"Hieracium pilosella", "Pilosella officinarum")) %>%
+  mutate(sp=str_replace(sp,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
+  mutate(sp=str_replace(sp,"Hierochloã« alpina", "Hierochloë alpina")) %>%
+  mutate(sp=str_replace(sp,"Hierochloã« hirta", "Hierochloë hirta")) %>%
+  mutate(sp=str_replace(sp,"Hierochloã« odorata", "Hierochloë odorata")) %>%
+  mutate(sp=str_replace(sp,"Huperzia appressa", "Huperzia selago")) %>%
+  mutate(sp=str_replace(sp,"Hylotelephium maximum", "Hylotelephium telephium")) %>%
+  mutate(sp=str_replace(sp,"Lappula myosotis", "Lappula squarrosa")) %>%
+  mutate(sp=str_replace(sp,"Lepidotheca suaveolens", "Matricaria discoidea")) %>%
+  mutate(sp=str_replace(sp,"Listera cordata", "Neottia cordata")) %>%
+  mutate(sp=str_replace(sp,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
+  mutate(sp=str_replace(sp,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
+  mutate(sp=str_replace(sp,"Logfia arvensis", "Filago arvensis")) %>%
+  mutate(sp=str_replace(sp,"Mentha _verticillata", "Mentha verticillata")) %>%
+  mutate(sp=str_replace(sp,"Minuartia rubella", "Sabulina rubella")) %>%
+  mutate(sp=str_replace(sp,"Minuartia stricta", "Sabulina stricta")) %>%
+  mutate(sp=str_replace(sp,"Mycelis muralis", "Lactuca muralis")) %>%
+  mutate(sp=str_replace(sp,"Omalotheca supina", "Gnaphalium supinum")) %>%
+  mutate(sp=str_replace(sp,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
+  mutate(sp=str_replace(sp,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
+  mutate(sp=str_replace(sp,"Ononis arvensis", "Ononis spinosa")) %>%
+  mutate(sp=str_replace(sp,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
+  mutate(sp=str_replace(sp,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
+  mutate(sp=str_replace(sp,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
+  mutate(sp=str_replace(sp,"Phalaris minor", "Phalaris arundinacea")) %>%
+  mutate(sp=str_replace(sp,"Phalaroides arundinacea", "Phalaris arundinacea")) %>%
+  mutate(sp=str_replace(sp,"Pinus unicinata", "Pinus mugo")) %>%
+  mutate(sp=str_replace(sp,"Platanthera montana", "Platanthera chlorantha")) %>%
+  mutate(sp=str_replace(sp,"Poa alpigena", "Poa pratensis")) %>%
+  mutate(sp=str_replace(sp,"Poa angustifolia", "Poa pratensis")) %>%
+  mutate(sp=str_replace(sp,"Poa laxa", "Poa flexuosa")) %>%
+  mutate(sp=str_replace(sp,"Poa _herjedalica", "Poa herjedalica")) %>%
+  mutate(sp=str_replace(sp,"Poa _jemtlandica", "Poa jemtlandica")) %>%
+  mutate(sp=str_replace(sp,"Poa lindebergii", "Poa arctica")) %>%
+  mutate(sp=str_replace(sp,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
+  mutate(sp=str_replace(sp,"Rhamnus catharticus", "Rhamnus cathartica")) %>%
+  mutate(sp=str_replace(sp,"Rumex alpestris", "Rumex acetosa")) %>%
+  mutate(sp=str_replace(sp,"Salix _fragilis", "Salix fragilis")) %>%
+  mutate(sp=str_replace(sp,"Saxifraga _opdalensis", "Saxifraga opdalensis")) %>%
+  mutate(sp=str_replace(sp,"Spergularia salina", "Spergularia marina")) %>%
+  mutate(sp=str_replace(sp,"Syringa emodi", "Syringa vulgaris")) %>%
+  mutate(sp=str_replace(sp,"Taraxacum crocea", "Taraxacum officinale")) %>%
+  mutate(sp=str_replace(sp,"Taraxacum croceum", "Taraxacum officinale")) %>%
+  mutate(sp=str_replace(sp,"Taraxacum erythrospermum", "Taraxacum officinale")) %>%
+  mutate(sp=str_replace(sp,"Taraxacum hamatum", "Taraxacum officinale")) %>%
+  mutate(sp=str_replace(sp,"Trientalis europaea", "Lysimachia europaea")) %>%
+  mutate(sp=str_replace(sp,"Trifolium pallidum", "Trifolium pratense")) %>%
+  mutate(sp=str_replace(sp,"Vicia orobus", "Vicia cassubica"))
+
+## merge species data with indicators
+NiN.sp.ind <- merge(NiN.sp,ind.dat, by.x="sp", by.y="species", all.x=T)
+summary(NiN.sp.ind)
+
+NiN.sp.ind[NiN.sp.ind==999] <- NA
+
+# checking which species didn't find a match
+unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) & NiN.sp.ind$spgr %in% list("a1a","a1b","a1c"),'sp'])
+# ok now
+
+#### CONTINUE HERE ####
+
+## matching with NiN ecosystem types
+# NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
+NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[19:23])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
+NiN.wetland[1,]
+names(NiN.wetland)
+
+#NiN.wetland <- NiN.wetland[,c(1:41,46:48,59:69)]
+cbind(colnames(NiN.wetland),
+      c("",
+        'V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
+        'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
+        'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
+        'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
+        'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
+        'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
+        'V3-C2','V1-C5',
+        'V1-C6a','V1-C6b',
+        'V1-C7a','V1-C7b',
+        'V1-C8a','V1-C8b',
+        'V2-C1a','V2-C1b',
+        'V2-C2a','V2-C2b',
+        'V2-C3a','V2-C3b',
+        "V4-C4","V4-C5",
+        "V4-C2","V4-C3",
+        'V8-C1','V8-C2','V8-C3',
+        rep("",10),
+        rep("",5)
+      )
+)
+
+NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:63)]
+colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
+                              'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
+                              'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
+                              'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
+                              'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
+                              'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
+                              'V3-C2','V1-C5',
+                              'V1-C6a','V1-C6b',
+                              'V1-C7a','V1-C7b',
+                              'V1-C8a','V1-C8b',
+                              'V2-C1a','V2-C1b',
+                              'V2-C2a','V2-C2b',
+                              'V2-C3a','V2-C3b',
+                              'V4-C2','V4-C3',
+                              'V8-C1','V8-C2','V8-C3'
+)
+head(NiN.wetland)
+
+# translating the abundance classes into %-cover
+coverscale <- data.frame(orig=0:6,
+                         cov=c(0,1/32,1/8,3/8,0.6,4/5,1)
+)
+
+NiN.wetland.cov <- NiN.wetland
+colnames(NiN.wetland.cov)
+for (i in 2:46) {
+  NiN.wetland.cov[,i] <- coverscale[,2][ match(NiN.wetland[,i], 0:6 ) ]
+}
+
+summary(NiN.wetland)
+summary(NiN.wetland.cov)
+
+```
 

### 2023-03-10 slitasje.Rmd
add ano data

@@
                "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
+pData <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
 
@@
 ```
+
+I here transfer these ordinal variables over to a continuous scale.
 
@@
 
-7SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7SE),])`
-NA's, and 7TK has
-`r nrow(naturetypes_wide[is.na(naturetypes_wide$var_7TK),])`.
-
-```{r, fig.cap="7SE scores in the naturetype dataset", fig.width=3}
+First I will combine 7TK and PRTK, and also 7SE and PRSL
+
+```{r}
+naturetypes_wide <- naturetypes_wide %>%
+  mutate(TK = if_else(
+    is.na(var_PRTK), var_7TK, var_PRTK),
+    SE = if_else(
+      is.na(var_PRSL), var_7SE, var_PRSL)
+    )
+```
+
+SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$SE),])`
+NA's, and TK has
+`r nrow(naturetypes_wide[is.na(naturetypes_wide$TK),])`.
+
+```{r, fig.cap="7SE scores in the naturetype dataset", fig.width=5}
 temp <- naturetypes_wide %>%
-  group_by(var_7SE) %>%
+  group_by(SE) %>%
   summarise(sum = n())
 
-ggplot(temp, aes(x = factor(var_7SE),
+ggplot(temp, aes(x = factor(SE),
                  y = sum))+
@@
   theme_bw(base_size = 12)+
-  labs(x = "7SE score",
+  labs(x = "7SE or PRSL score",
        y = "Number of localities")
@@
 
-The NA fraction is quite big. These are localities with 7TK recorded,
-but not 7SE.
-
-```{r, fig.cap="7TK scores in the nature type dataset", fig.width=3}
+The NA fraction is quite big. These are localities with 7TK or PRTK recorded,
+but not 7SE or PRSL.
+
+```{r, fig.cap="7TK scores in the nature type dataset", fig.width=5}
 temp <- naturetypes_wide %>%
-  group_by(var_7TK) %>%
+  group_by(TK) %>%
   summarise(sum = n())
 
-ggplot(temp, aes(x = factor(var_7TK),
+ggplot(temp, aes(x = factor(TK),
                  y = sum))+
@@
   theme_bw(base_size = 12)+
-  labs(x = "7TK score",
+  labs(x = "7TK og PRTK score",
        y = "Number of localities")
@@
 ```{r, warning=F}
-naturetypes_wide$slitasje <- apply(naturetypes_wide[,-1], 1, max, na.rm=T)
-```
-
-When both variables are NA, we get -Inf. There were 13 cases of this
-(in 2023). Removing these now.
+naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("TK", "SE")], 1, max, na.rm=T)
+```
+
+When both variables are NA, we get -Inf. There were `r nrow(naturetypes_wide[naturetypes_wide$slitasje<0,])` cases of this. Removing these now.
 
@@
   labs(x = "Slitasje score converted to percentage",
-       y = "Number of localities")+
-  coord_flip()
+       y = "Number of localities") + 
+  theme(axis.text.x = element_text(angle=90, vjust=0.5))
 ```
@@
 
-I would also like to know how often 7TK was defining of the
-slitasje-indicator, and how often it was 7SE. I can do this by taking
-the difference.
-
-```{r, fig.cap="Counting the number of cases where 7TK or 7SE was defining of the slitasje indicator score.", fig.width=3}
-
-diff <- max(naturetypes_wide$var_7SE, naturetypes_wide$var_PRSL, na.rm=T) - 
-        max(naturetypes_wide$var_7TK, naturetypes_wide$var_PRTK, na.rm=T)
-
-diff <- ifelse(diff == 0, "7TK and 7SE",
-               ifelse(diff <0, "7TK", "7SE"))
+I would also like to know, when both tk and SE was recordedm
+how often TK was defining of the slitasje-indicator, and how 
+often it was SE. I can do this by taking the difference.
+
+```{r, fig.cap="Counting the number of cases where TK or SE was defining of the slitasje indicator score.", fig.width=3}
+
+diff <- naturetypes_wide$SE - naturetypes_wide$TK
+
+diff <- ifelse(diff == 0, "TK and SE",
+               ifelse(diff <0, "TK", "SE"))
 
@@
 
-Looks like 7SE is more likely to be in a detrimental state.
+Looks like SE (= 7SE or PRSL) is more likely to be in a detrimental state.
 
@@
 ```{r}
-naturetypes$slitasje <- naturetypes_wide$slitasje[match(naturetypes$identifikasjon_lokalid, naturetypes_wide$identifikasjon_lokalid)]
+naturetypes$slitasje <- naturetypes_wide$slitasje[match(naturetypes$identifikasjon_lokalId, naturetypes_wide$identifikasjon_lokalId)]
 #nrow(naturetypes[is.na(naturetypes$slitasje),])  # 13 cases
@@
 : Table showing which main ecosystems the two NiN variables 7SE and 7TK
-is recorded in withing the ANO dataset.
+is recorded in withing the ANO data set.
 
@@
 #unique(ano$hovedtype_250m2)[3]
-ano$hovedtype_250m2[ano$hovedtype_250m2==unique(ano$hovedtype_250m2)[3]] <- "Aapen jordvassmyr"
+#ano$hovedtype_250m2[ano$hovedtype_250m2==unique(ano$hovedtype_250m2)[3]] <- "Aapen jordvassmyr"
 ```
@@
 
-Combining these two variables into the indicator, same as for the nature
+The ANO sites seem to be in an even better condition than the localities in the naturtype data set.
+
+Combining these two variables 7TJ and 7SE into one, same as for the nature
 type data.
@@
 
+Then I extract the mid values for each ordinal range step, like I did previously for the nature type data set.
+
+```{r}
+ano <- ano %>%
+  mutate(slitasje = case_match(slitasje,
+             0 ~ 0,
+             1 ~ mean(c(0, 1/16))*100,
+             2 ~ mean(c(1/16, 1/2))*100,
+             3 ~ 75))
+```
+
 ```{r, fig.cap="Distribution of slitasje scores in the ANO data", fig.width=6}
-barplot(table(ano$slitasje), xlab="Sliatsje scores", ylab = "Number of localities")
+barplot(table(ano$slitasje), xlab="% Slitasje", ylab = "Number of localities")
 ```
@@
                            select(naturetypes,
-                                  identifikasjon_lokalid,
+                                  identifikasjon_lokalId,
                                   origin,
@@
      ylab = "Area of the new polygons")
+abline(0,1)
 ```
@@
 mapped areas leaves the indicator without much value for regional
-assessments. When we do regional assessments and calcualte regional
+assessments. When we do regional assessments and calculate regional
 indicator values, we cannot simply do an area weighting of the polygons
@@
 
-The variable is ordinal, not continuous. See fig. \@ref(fig:eight-step).
-I could create another variable which was more like original range
-scale. Although not perfect, I will use the median value for each
-category. I.e. when the variable is recorded as 2 (meaning from 1/16th
-to 1/2), this is converted to 31.25%, i.e. in the middle between the two
-end points of the category.
-
-```{r}
-slitasje_data$slitasje_percentage <- slitasje_data$slitasje
-
-slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 1] <- ((1/16)/2)*100
-slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 2] <- (((1/2)/2)+(1/16))*100
-slitasje_data$slitasje_percentage[slitasje_data$slitasje_percentage == 3] <- (3/4)*100
-table(slitasje_data$slitasje_percentage)
-```
-
-```{r continuous-scale, fig.cap="Showing the conversion from ordinal to (pseudo)continous cale for the slitasje variable", fig.width=5}
-ggplot(slitasje_data, aes(x = slitasje,
-                          y = slitasje_percentage))+
-  geom_point(size=6)+
-  geom_line()+
-  theme_bw(base_size = 16)
-```
-
 I will use the same reference levels/values for all of Norway:
@@
 lower <- 100
-threshold <- 10
+threshold <- 5
 ```
@@
 
-Also, the threshold value is set quite conservativly, and should be
+Also, the threshold value is set quite conservatively (I think), and should be
 discussed.
@@
 
-```{r slitasje-normalise, fig.cap="Performing a non-linear break-point type normalisation of the slitasje variable.", fig.width=5}
+<!-- THE BREAKPOINT IS AT 0.5 RATHER THAN 0.6. TEST CODE FROM HERE: https://ninanor.github.io/eaTools/articles/normalise-condition-variable.html -->
+
+```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=5}
 eaTools::ea_normalise(data = slitasje_data,
-                      vector = "slitasje_percentage",
+                      vector = "slitasje",
                       upper_reference_level = lower,

### 2023-03-13 slitasje.Rmd
new figs

@@
 #library(tidyverse)
+library(tibble)
 library(ggplot2)
@@
 slitasje_data$i_2020 <- eaTools::ea_normalise(data = slitasje_data,
-                      vector = "slitasje_percentage",
+                      vector = "slitasje",
                       upper_reference_level = lower,
@@
 
-Since the two layers are completely overlapping, we can get the
+Since the HIA map and the _region_ map are completely overlapping (HIA was masked using the region map), we can get the
 intersections
@@
 
-I now have 20 unique areas (for each main ecosystem) that I will, given
+I now have `r length(unique(HIA_regions$region_HIF))` unique areas (for each main ecosystem) that I will, given
 there is data, extrapolate indicator values over.
@@
 
-I will subset the `sliatsje_data` in the three ecosystems. Note that
+I will subset the `slitasje_data` into the three ecosystems. Note that
 only for open wetland do we have good ecosystem delineation maps to base
@@
 ```{r}
-wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "Våtmark",]
-seminat <- slitasje_data[slitasje_data$hovedøkosystem == "Semi-naturlig mark",]
-natOpen <- slitasje_data[slitasje_data$hovedøkosystem == "Naturlig åpne områder under skoggrensa",]
+#Check spelling
+#unique(slitasje_data$hovedøkosystem)
+
+# Rename one level
+slitasje_data <- slitasje_data %>%
+  mutate(hovedøkosystem = recode(hovedøkosystem, "Våtmark" = "våtmark"))
+
+#make valid
+slitasje_data <- st_make_valid(slitasje_data)
+
+#subset
+wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "våtmark",]
+seminat <- slitasje_data[slitasje_data$hovedøkosystem == "semi-naturligMark",]
+natOpen <- slitasje_data[slitasje_data$hovedøkosystem == "naturligÅpneOmråderILavlandet",]
 ```
@@
            sep = " ")
-saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
+
+saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
 ```
@@
 
-```{r dense-stats, fig.height=20, fig.width=12}
+```{r dense-stats, fig.height= 12, fig.width=6}
+temp_plot <- all_stats %>%
+  mutate(HIF_NA = case_when(
+    n > 150 ~ HIF
+  ))
+
+myColour_list <- unique(temp_plot$myColour)
+names(myColour_list) <-   myColour_list
+
 ggarrange(
-ggplot(all_stats, aes(x = region, y = n, group = HIF, fill = HIF))+
-  geom_bar(position = "dodge", stat = "identity")+
-  theme_bw(base_size = 30)+
+ggplot(temp_plot, aes(x = region, y = n, group = HIF, fill = HIF_NA))+
+  geom_bar(position = "dodge", stat = "identity", col="black")+
+  theme_bw(base_size = 20)+
+  scale_fill_brewer(palette="Oranges")+
+  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
   facet_wrap(.~eco),
-ggplot(all_stats, aes(x = region, y = w_mean, group = HIF, fill = HIF))+
-  geom_bar(position = "dodge", stat = "identity")+
-  theme_bw(base_size = 30)+
+ggplot(temp_plot, aes(x = region, y = w_mean, group = HIF, fill = HIF_NA))+
+  geom_bar(position = "dodge", stat = "identity", col="black")+
+  theme_bw(base_size = 20)+
+  scale_fill_brewer(palette="Oranges")+
+  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
   facet_wrap(.~eco),
-ggplot(all_stats, aes(x = region, y = sd, group = HIF, fill = HIF))+
-  geom_bar(position = "dodge", stat = "identity")+
-  theme_bw(base_size = 30)+
+ggplot(temp_plot, aes(x = region, y = sd, group = HIF, fill = HIF_NA))+
+  geom_bar(position = "dodge", stat = "identity", col="black")+
+  theme_bw(base_size = 20)+
+  scale_fill_brewer(palette="Oranges")+
+  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
@@
 
-This rather dense figure needs to be looked at carefully. In the top row
+In the figure above, colors refer to Human Impact Factor (amount or frequency of infrastructure) 
+and NA means that there are less than 150 data points which we set as a threshold for evaluating the indicator values.
+
+In the top row
 we see that the sample size (number of polygons) varies between
@@
 
+In semi-natural areas and wetlands, we have very little data from HIF
+class 3.
+
 We can also from the middle row see the association between the indicator
 values (area weighted means) and the modified infrastructure index
-(HIF). We should then not put too much weight on the bars were there is
-little data (low n) and/or high standard error (sd, bottom row). For
-example, with Naturally open ecosystems,
-
-Midt-Norge has the overall lowest errors, and there we also see the kind of relationship between
-the indicator and the pressure that we want to see (middle row). The
-pattern is repeated for all regions except Østlandet, but there we also
-see that HIF class 3 has very little data, and a high sd, and this could
-explain the discrepancy (but it doesn't support the general relationship
-either).
-
-In semi-natural areas and wetlands, we have very little data from HIF
-class 3, and the sd's are very big. The indicator-pressure relationship
-(middle row) looks the most messy for wetlands, but if ignoring HIF=3,
-the relationship looks much better.
+(HIF). We should then not put any weight on the bars were there is
+little data (transparent bars). Then we se the general pattern that the indicator 
+value decreases with increasing HIF. But there are exceptions, like for Semi-natural areas
+in Vestlandet.
+
+Let us look at the indicator to pressure relationship across all ecosystems.
+
+```{r, fig.cap="Indicator to pressure relationship across all ecosystems."}
+temp_plot %>%
+  filter(!is.na(HIF_NA)) %>%
+  ggplot(aes(x = HIF_NA, y = w_mean))+
+  geom_point(size=2, position = position_dodge2(.1))+
+  geom_violin(alpha=0)+
+  theme_bw()
+```
+
+The pattern is not perfect, and there is quite some noise. But this is perhaps also expected.
+
+Let us look at the effect of sample size on the indicator uncertainty.
+
+```{r, fig.cap="Smaple size against indicator uncertainty."}
+temp_plot %>%
+  ggplot(aes(x = n, y = sd))+
+  geom_point(size=2, position = position_dodge2(.1))+
+  theme_bw()+
+    facet_wrap(.~eco)
+```
+
+This shows that the uncertainty is inflated with sample sizes less than about 300-500.
 
@@
 corrCheck <- st_intersection(slitasje_data, HIA_regions)
-saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
+saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
 ```
@@
 ```{r slitasje-precentageplot, fig.height=12, fig.width=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
-ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(i_2020)))+
+ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
   geom_bar(position="fill")+
@@
   xlab("HIF")+
+  scale_fill_brewer(palette = "RdYlGn")+
   facet_grid(hovedøkosystem~region)
@@
 groups/categories that have very little data.
-
-We can also calculate *Kendall tau b* coefficients, although I'm not
-complete sure this is valid when we have ordinal data with only four
-levels, since there will be a lot of ties.
-
-```{r}
-corrCheck_grouped <- as.data.frame(corrCheck) %>%
-  group_by(hovedøkosystem, region) %>%
-  summarise(cor = cor(i_2020, infrastructureIndex, method = "kendall"))
-```
-
-```{r, fig.cap="Pearson correlations between the indicator values at the polygons level and the HIF."}
-ggplot(corrCheck_grouped, aes(x = hovedøkosystem, 
-                      y = cor))+
-  geom_bar(stat = "identity")+
-  facet_wrap(.~region)+
-  coord_flip()
-```
-
-```{r, fig.cap="Pearson correlations between indicator values and the HIF."}
-corrCheck_eco <- corrCheck_grouped %>%
-  group_by(hovedøkosystem) %>%
-  summarise(cor = mean(cor))
-corrCheck_reg <- corrCheck_grouped %>%
-  group_by(region) %>%
-  summarise(cor = mean(cor))
-
-ggarrange(
-ggplot(corrCheck_eco, aes(x = hovedøkosystem, 
-                      y = cor))+
-  geom_bar(stat = "identity")+
-  coord_flip(),
-ggplot(corrCheck_reg, aes(x = region, 
-                      y = cor))+
-  geom_bar(stat = "identity")+
-  coord_flip(),
-ncol=1
-)
-```
-
-From the figures above it appears that the relationship is quite week in
-general. However, the sheer number of data points suggest that group
-differences will be very reliable. *Naturlig åpne områder* has the
-lowest correlation between the indicator and the homogeneous area
-classes, and of the five regions, *Nord-Norge* has the lowest
-correlation.
 

### 2023-03-13 slitasje.Rmd
upp threshold

@@
 
-The figure above shows that Nord-Norge for example, has a lot of
-relatively untouched areas, and that Østlandet has the highest
+The figure above shows that _Nord-Norge_ for example, has a lot of
+relatively untouched areas, and that _Østlandet_ has the highest
 proportion of impacted areas. This is expected.
@@
 I want to add a threshold so that we don't end up over extrapolating
-based on too few data points. From Fig. \@ref(fig:dense-stats) I
-conclude that we need at least 70 data points for the sd to stabilize. I
-will conservatively round this up to 100 data points.
+based on too few data points. I have previousl used 150 data pionts, and I think that was about right.
 
@@
                          groups = region_HIF,
-                         threshold = 100)
+                         threshold = 150)
 
@@
                          groups = region_HIF,
-                         threshold = 100)
-#saveRDS(wetlands_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
-#saveRDS(seminat_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
-#saveRDS(natOpen_slitasje_extr, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
+                         threshold = 150)
+#saveRDS(wetlands_slitasje_extr, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
+#saveRDS(seminat_slitasje_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
+#saveRDS(natOpen_slitasje_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
 ```

### 2023-03-13 slitasje.Rmd
try to warp

@@
 
-There is some, but very little variation, both between the ecosystems
-and between the four HIF classes in the modified infrastructure index.
-Note however, that for wetland and semi-natural areas we did not
+Note that for wetland and semi-natural areas we did not
 extrapolate values for the most urban areas, and that for naturally open
 areas we did not extrapolate to the least affected areas.
+For the other regions we generally excluded more HIFs.
 
@@
 
--   indicator maps (vector data), and we have
-
--   total HIA areas, and we have the
+-   indicator maps (vector data, small polygons), and we have
+
+-   total area for each HIA class and region combo, and we have the
 
 -   extrapolated indicator maps (HIA maps with aggregated indicator
-    values and errors, also vector data).
+    values and errors, also vector data, wall-to-wall with some NA areas).
 
@@
 ```{r, eval=F}
-mireET <- sf::st_read("/data/R/Kladd/Myr/Myrmodell.gdb")
-```
-
-This doesn't take that long.
-
-```{r, eval=F}
-mireET <- st_transform(mireET, 25833)
-```
-
-This take about 10 min. Not sure I need it.
-
-```{r, eval=F}
-mireET <- st_make_valid(mireET)
-```
-
-This takes a long!! time, and should be AVOIDED.
-
-```{r, eval=F}
-wetlands_slitasje_extr_cropped <- sf::st_intersection(wetlands_slitasje_extr, mireET)
-```
-
+#takes about one minute
+mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
+                             proxy=F)
+```
+
+This is in the wrong CRS.
+
+```{r}
+mireET <- sf::st_transform(mireET, 25833)
+```
+
+```{r, eval=F}
+system.time(
+  mireET_warped <- stars::st_warp(mireET, grid)
+)
+```
+
+user: 597.308
+
+system: 139.164
+
+elapsed: 735.907
+
+```{r}
+mireET
+```
+
+```{r, eval=F}
+write_stars(mireET_warped, paste0(pData, "wetland_ecosystem_EPSG25833_50m.tif"))
+```
+
+```{r}
+wetlands_trd <- sf::st_crop(mireET, myBB)
+tm_shape(wetlands_trd)+
+  tm_raster()+
+  tm_shape(hw_utm)+
+  tm_lines(col="blue")
+```
+
+## OLD
 Instead I will draw some random point samples.

### 2023-03-13 slitasje.Rmd
trying to mask

@@
 
-This is a quite heavy operation and should be done on a server.
-
 ```{r, eval=F}
-#takes about one minute
+#takes about one minute without the proxy
+mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
+                             proxy=T)
+mireET
+```
+
+This is in the wrong CRS. We want to stay with 25833.
+
+We have a 50x50 m [master grid](#master-grid) with the correct CRS, and I would warp mireET to fit that grid like this:
+```{r, eval=F}
+grid <- stars::read_stars(paste0(pData, "masterGrid_50m.tiff"))
+mireET_warped <- stars::st_warp(mireET, grid)
+```
+
+... but I don't want to down sample the mire data, since we will loose a lot of the smaller ecosystem occurrences.
+
+I could make a 10x10 master grid, but I have not found a way to stop this from being just a massive file.
+
+I can transform the ET map like this, and not worry about the rasters being perfectly aligned, but even this operation is too heavy for the R Studio Server:
+
+```{r, eval=F}
+mireET_trans <- sf::st_transform(mireET, 25833)
+```
+
+So I will transform the extrapolated indicator map ti match the ET instead. Note that the ET map is just for southern Norway up to about the middle of the country.
+
+```{r}
+system.time(
+  wetlands_slitasje_extr <- st_transform(wetlands_slitasje_extr, st_crs(mireET))
+)
+```
+
+Then I import the real ET data and mask
+
+```{r, eval=F}
 mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
@@
 
-This is in the wrong CRS.
-
-```{r}
-mireET <- sf::st_transform(mireET, 25833)
-```
-
-```{r, eval=F}
-system.time(
-  mireET_warped <- stars::st_warp(mireET, grid)
-)
-```
-
-user: 597.308
-
-system: 139.164
-
-elapsed: 735.907
-
-```{r}
-mireET
-```
-
-```{r, eval=F}
-write_stars(mireET_warped, paste0(pData, "wetland_ecosystem_EPSG25833_50m.tif"))
-```
-
-```{r}
-wetlands_trd <- sf::st_crop(mireET, myBB)
-tm_shape(wetlands_trd)+
-  tm_raster()+
-  tm_shape(hw_utm)+
-  tm_lines(col="blue")
-```
-
-## OLD
-Instead I will draw some random point samples.
-
-```{r, eval=F}
-wetland_points <- st_sample(mireET, 10000)
-```
-
-And intersect this with the extrapolated indicator values.
-
-```{r, eval=F}
-wetland_points_match <- st_intersection(wetlands_slitasje_extr, wetland_points)
-```
-
-And extract the values
-
-```{r, eval=F}
-wetland_points_extracted <- st_intersection(wetlands_slitasje_extr, wetland_points)
-```
-
-If we went for option 2, we could instead do something like this, and
-
-create an empty grid over Norway, 50 x 50 m resolution
-
-<!--# This we should perhaps do once, in a separate misc chapter, and make available for all chapters -->
-
-```{r, eval=F}
-grd = st_as_stars(st_bbox(wetlands_slitasje_extr), dx = 50, dy = 50, values = NA_real_)
-```
-
-and then rasterize the indicator
-
-```{r, eval=F}
-wetlands_slitasje_extr_rast <- st_rasterize(wetlands_slitasje_extr, grd)
-```
-
-To get a perfect alignment with the ET map, we should change `grd` with
-the ET raster map.
-
-Then we can use starts filtering with `[ .. ]` to mask out the areas
-that do not belong to the ET.
+<!-- PLAN: rasterize wetland_slitasje_extr and c() with mireET, and mask -->
+<!-- Or use  -->
+<!-- wetlands_slitasje_extr[mireET==0] <- NA -->
+
+```{r}
+wetlands_slitasje_extr_stars <- stars::st_as_stars(wetlands_slitasje_extr)
+temp <- c(wetlands_slitasje_extr_masked, mireET)
+class(wetlands_slitasje_extr)
+dim(temp)
+```
 

### 2023-03-14 R/functionalPlantIndicators/2 Reference bootstrap function.R
wetland markdown

@@ -1,3 +1,4 @@
+#### bootstrap for abundance classes ####
 # function to calculate community weighted means of selected indicator values (ind)
 # for species lists (sp) with given abundances (on a scale from 1 to 6) in one or more 'sites' (abun)
 # with a given number of iterations (iter),
@@ -37,6 +38,78 @@ indBoot <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
         for (l in 1: nrow(dat.b)) { dat.b[l,2] <- dat.b[l,2] + sample(c(-1,0,1),size=1) }
         dat.b[!is.na(dat.b[,2]) & dat.b[,2]==(0),2] <- 1
         dat.b[!is.na(dat.b[,2]) & dat.b[,2]==7,2] <- 6
+      }
+      
+      for (k in 1:length(colnames(ind))) {
+        
+        if ( nrow(dat.b)>2 ) {
+          
+          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
+          ind.list[[k]][i,j] <- ind.b
+          
+        } else {ind.list[[k]][i,j] <- NA}
+        
+      }
+      #      su <- sum(ind.list[["CC"]][i,j],ind.list[["SS"]][i,j],ind.list[["RR"]][i,j])
+      #      print(paste(i,"",j,su))   
+      print(paste(i,"",j)) 
+    }
+    
+  }
+  return(ind.list)
+}
+
+
+
+
+#### bootstrap for frequency abundance ####
+# function to calculate community weighted means of selected indicator values (ind)
+# for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
+# with a given number of iterations (iter),
+# with species given a certain minimum abundance ocurring in all bootstraps (obl), and
+# with a given resampling ratio of the original species list (rat)
+# in every bootstrap iteration the abundance of the sampled species can be 
+# randomly changed by +/-1 if wished (var.abun)
+
+
[…]

### 2023-03-14 R/functionalPlantIndicators/3 Reference bootstraps.R

@@ -3,8 +3,8 @@ colnames(NiN.wetland)
 wetland.ref <- indBoot(sp=NiN.wetland[,1],abun=NiN.wetland[,2:46],ind=NiN.wetland[,47:51],
                           iter=1000,obl=6,rat=2/3,var.abun=T)
 
-# NB! the indboot function does not work with var.abun=T for cover data yet
-wetland.ref.cov <- indBoot(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:51],
+
+wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:51],
                           iter=1000,obl=1,rat=2/3,var.abun=F)
 
 ### saving backups
@@ -28,26 +28,6 @@ for (i in 1:length(wetland.ref.cov) ) {
   }
 }
 
-
-#### checking NiN-types with several species lists ####
-# mountain
-str(wetland.ref)
-indID <- colnames(ind.dat[,c("Continentality", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")])
-NiNID <- colnames(wetland.ref[[1]])
-
-for (i in indID) {
-  
-  dirOutput <- "C:/Users/joachim.topper/OneDrive - NINA/work/R projects/projects/økol tilst indikatorverdier/output/checks/mountain"
-  setwd(dirOutput)
-  name <- paste('wetland',i,'.jpg',sep='')
-  jpeg(filename=name,width=8000,height=4000,res=300)
-  
-  par(mfrow=c(5,9))
-  for (j in NiNID) {
-    plot(density(wetland.ref[[i]][,j],na.rm=T),main=j,xlim=c(-2,10))
-  }
-  dev.off()
-}
 
 
 

### 2023-03-14 R/functionalPlantIndicators/4 Reference scaling values.R

@@ -71,7 +71,6 @@ tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphon after C for NiN-types
 tab
 
 
-
 # restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
 y.Cont <- numeric(length=nrow(tab)*2)
 y.Cont[((1:dim(tab)[1])*2)-1] <- tab$Cont_q2.5 
@@ -131,7 +130,7 @@ summary(wetland.ref.val)
 
 
 
-## doing the same for wetland.ref.cov.cov
+## doing the same for wetland.ref.cov
 # 5 indicator-value indicators: Tyler's Continentality, Light, Moisture, Soil_reaction_pH, "Nitrogen"
 indEll.n=5
 # creating a table to hold:

### 2023-03-14 R/functionalPlantIndicators/5 scaled value calculations_wetland.R

@@ -62,7 +62,7 @@ scal.2 <- function() {
 
 unique(ANO.geo$hovedtype_rute) # NiN types in data
 unique(substr(wetland.ref.val$grunn,1,2)) # NiN types in reference
-#### creating dataframe to holde the results for wetlands ####
+#### creating dataframe to hold the results for wetlands ####
 # all ANO points
 nrow(ANO.geo)
 # all wetland ANO points

### 2023-03-14 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -20,7 +20,7 @@ hist(results.wetland[['original']][results.wetland[['original']]$kartleggingsenh
 
 
 #### regressions vs. ANO variables ####
-library(betareg)
+
 
 head(results.wetland[['original']])
 colnames(results.wetland[['original']])[c(42:47,54:56)]
@@ -54,6 +54,10 @@ summary(lm(Moist1~tresjikt_dekning, data=df ))
 points(0:100,predict(lm(Moist1~tresjikt_dekning, data=df ),
                      newdata=data.frame(tresjikt_dekning=0:100) ),
        type='l', lty=2, lwd=3)
+
+
+library(betareg)
+summary( betareg(Moist1 ~ tresjikt_dekning, data=df[!is.na(df$Moist1),]) )
 
 #with(results.wetland[['scaled']],plot(busker_dekning,Light1))
 #summary(lm(Light1~busker_dekning, data=results.wetland[['scaled']]))

### 2023-03-14 slitasje.Rmd
explain net steps and simplify workflow

@@
 
-10. Sample 1000 random points from an ecosystem delineation and extract
-    indicator values and sd at those points.
-
-11. Export indicator maps and regional extrapolated maps
+10. TO DO: Prepare ecosystem delineation maps and us ethese to mask the extrapolated indicator maps
+
+11. Spatial aggregation of indicator values and uncertainties to accounting areas
+
+12. Export indicator maps and regional extrapolated maps
 
@@
 
-```{r, fig.cap="Slitasje indicator extrapolated over Trondheim"}
+```{r extrapolated-slitasje, fig.cap="Slitasje indicator extrapolated over Trondheim"}
 myCol <- "RdYlGn"
@@
 
-Now, there are multiple way to proceed. We have
-
--   indicator maps (vector data, small polygons), and we have
-
--   total area for each HIA class and region combo, and we have the
-
--   extrapolated indicator maps (HIA maps with aggregated indicator
-    values and errors, also vector data, wall-to-wall with some NA areas).
-
--   We also have, for wetlands, an ecosystem type (ET) map.
-
-Option 1:
-
-1.  crop (i.e. intersect) the extrapolated indicator maps to ecosystem
-    delineation maps, and then;
-
-2.  aggregate these for each region or accounting area.
-
-Step 1 above requires an st_interactions on two very large vector data
-sets with millions of polygons. This should be avoided.
-
-Option 2:
-
-1.  rasterize the extrapolated indicator map, using the ET map as a
-    template, and
-
-2.  mask it using the perfectly aligned ET map.
-
-This might be faster.
-
-Option 3:
+The next step now is to
+
+1. Prepare ecosystem delineation maps in EPSG25833 and perfectly aligned to a master grid
+
+2. Rasterize the extrapolated indicator map, using the ET map as a template
+
+3. and mask it using the perfectly aligned ET map.
+
+A faster way could be to:
 
@@
 
-This is perhaps even faster again. I will try the last option, even
-though it will not provide spatially explicit indicator values for the
-entire ecosystem delineation, like option 2 does. Perhaps I will revisit
-option 2 later.
+... but his will not provide spatially explicit indicator values for the
+entire ecosystem delineation.
 
@@
 
-```{r, eval=F}
+```{r, eval=T}
 #takes about one minute without the proxy
-mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
-                             proxy=T)
-mireET
+(mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
+                             proxy=T))
 ```
@@
 
-So I will transform the extrapolated indicator map ti match the ET instead. Note that the ET map is just for southern Norway up to about the middle of the country.
-
-```{r}
-system.time(
-  wetlands_slitasje_extr <- st_transform(wetlands_slitasje_extr, st_crs(mireET))
-)
-```
-
-Then I import the real ET data and mask
-
-```{r, eval=F}
-mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
-                             proxy=F)
-```
-
-<!-- PLAN: rasterize wetland_slitasje_extr and c() with mireET, and mask -->
-<!-- Or use  -->
-<!-- wetlands_slitasje_extr[mireET==0] <- NA -->
-
-```{r}
-wetlands_slitasje_extr_stars <- stars::st_as_stars(wetlands_slitasje_extr)
-temp <- c(wetlands_slitasje_extr_masked, mireET)
-class(wetlands_slitasje_extr)
-dim(temp)
-```
+I will therefore ask GIS engineers to prepare ecosystem maps to the standard that we chose, so that we can simply combine the stars object with `stars::c()`, or perhaps side-step that operatio and just mask directly like this: `wetlands_slitasje_extr[mireET==0] <- NA`.
 
@@
 We need to aggreate the ET occurence and their indicator values, to a
-regional delineation. This functionality will be developed in `eaTools`.
-We need to produce distributions for the indicator valeus for each ET
-occurrence (or groups of them) and resample the distributions wiht an
+regional delineation. This functionality can be developed in `eaTools`,
+but will in any case build around `exactextratr::exact_extract()`.
+
+To carry the errors, we can for example
+produce distributions for the indicator values for each ET
+occurrence (or groups of them) and resample the distributions with an
 area weighting.
@@
 
-But I will also export a map of the indicator where we extrapolate the
-indicator values to the homogeneous areas. This map should be
-interpreted with caution, since they seem to provide explicit indicator
+I can also export a map of the indicator where we extrapolate the
+indicator values to the homogeneous areas (e.g. Fig. \@ref(fig:extrapolated-slitasje) . This map should be
+interpreted with caution, since they appear as providing explicit indicator
 values for areas where there is no data. This map is mainly for
 aggregation purposes. I will mark this map with the prefix
-*homogeneous-areas*. I can also export this same mappen after masking it
-with the ET map, and this output must come with the same discalimers.
+*homogeneous-areas*. I will also export this same map after masking it
+with the ET map, and this output must come with the same disclaimers.
 
 Finally I will export a map of the five accounting areas with
-wall-to-wall indicator value and errors. This map can also contain data
+wall-to-wall indicator value and errors (e.g. Fig \@ref(fig:wall-to-wall-summer-temp-indicator)). This map can also contain data
 covering other ETs. Otherwise the maps would not we visually
@@
 
-Prepare export
+This is how I plan to export the final product.
 

### 2023-03-14 climate_indicators_explained.Rmd
add chapter medain-summer-temp and rename generic chapter

--- 2023-03-07 DRAFT_climate_indicators.Rmd
@@
----
-title: "Climate indicators"
-output:
-  html_document
----
-
-# Climate indicators
+# Climate indicators excplained
 
@@
 This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
+I describe the process using dummy data and also discuss alternate approaches and dead ends that occured in the development process.
+For a succinct workflow example for a spesific indikcator, see [Median summer temperature](#median-summer-temp).
 

### 2023-03-14 medianSummerTemperature.Rmd
add chapter medain-summer-temp and rename generic chapter

--- 2023-03-07 DRAFT_climate_indicators.Rmd
@@
----
-title: "Climate indicators"
-output:
-  html_document
----
-
-# Climate indicators
+# Median Summer Temperature
 
@@
 
-This chapters describes the development of a workflow for generating or preparing indicators based on interpolated climate data from [SeNorge](https://senorge.no/).
+This chapters describes the workflow for an indicator describing the median summer temperature.
+For a more comprehensive documentaion on the development of the workflow itself, se [here](#climate-indicators-explained).
+The data comes from interpolated climate surfaces from [SeNorge](https://senorge.no/) which contain one 1x1km raster for each day since 1957 to present.
+The reference levels are extracted from the time period 1961-1990 (a temporal reference condition).
+
+**Workflow**
+
+1.  Collate variable data series, ending up with one raster per year after aggregating across days within a year
+
+2.
+Calculate spatially explicit reference values by aggregating across the years 1961-1990.
+The upper reference value is equal to the median value of this period, and the lower reference values (two-sided) is equal to 5 SD units.
+The indicator value is the mean over a five year period.
+
+3.  Calculate indicator values through spatially explicit rescaling based on the reference values
+
+4.  Mask by ecosystem type (*This step is not yet done as we do not have ready available ecosystem maps*)
+
+5.  Aggregate in space (to accounting areas) and take the mean over a five year period to get final indicator values
+
+6.  Make trend figure and spatially aggregated maps
 
@@
 
-This table explains them in more detail
-
-```{r,warning=FALSE, message=FALSE, fig.cap="Table explaining the available climate parameters in the data set."}
-senorgelayers <- read_delim("data/senorgelayers.txt", 
-    delim = "\t", escape_double = FALSE, 
-    trim_ws = TRUE)
-DT::datatable(senorgelayers)
-```
-
-Here is the content of one of these folders (first 10 entries)
-
-```{r}
-(files_tm <- list.files(paste0(path, "/rr_tm")))[1:10]
-```
-
-There are `r length(files_tm)` files, i.e. `r length(files_tm)` years of data, one file per year.
-
-Importing (proxy) file:
-
-```{r, warning=F}
-tg_1957 <- stars::read_ncdf(paste0(path, "/rr_tm/seNorge2018_1957.nc"),
-                            var="tg")
-```
-
-I know the CRS, so setting it manually. Although I cannot rule out 32633, I don't think it matters.
-
-```{r}
-st_crs(tg_1957) <- 25833
-```
-
-The data has three dimension
-
-```{r}
-dim(tg_1957)
-```
-
-Initially the data had four attributes, but I subsettet to only include *tg*.
-
-```{r}
-names(tg_1957)
-```
-
-`tg_1957` is a stars proxy object and [most commands](https://r-spatial.github.io/stars/articles/stars8.html) will not initiate any change until later, typically I write to file and all the lazy operations are done in squeal.
-Therefore, I will prepare a test data set, which is smaller and which I can import to memory.
-Then I can perform the operations on that data set and we can see the results.
-
-#### Dummy data
-
-This test data is included in the {stars} package
-
-```{r}
-tif = system.file("tif/L7_ETMs.tif", package = "stars")
-t1 = as.Date("1970-05-31")
-x = read_stars(c(tif, tif, tif, tif), along = 
-                  list(time = c(t1, t1+1, t1+50, t1+100)), 
-               RasterIO = list(nXOff = c(1), 
-                               nYOff = c(1), 
-                               nXSize = 50, 
-                               nYSize = 50, 
-                               bands = c(1:6)))
-```
-
-A single attribute
-
-```{r}
-names(x)
-```
-
-I can rename it like this
-
-```{r}
-x <- setNames(x, "Attribute_A")
-```
-
-And I can add another dummy attribute.
-
-```{r}
-x <- x %>%
-  mutate("Attribute_B" = Attribute_A/2)
-```
-
-The dummy data also has four dimensions
-
-```{r}
-dim(x)
-```
-
-X and y area the coordinates. Band is an integer:
-
-```{r}
-st_get_dimension_values(x, "band")
-```
-
-I will remove the *band* dimension.
-
-```{r}
-x <- x %>% filter(band==1) %>% adrop()
-```
-
-Time is four dates covering four months in 1970:
-
-```{r}
-(oldTime <- st_get_dimension_values(x, "time"))
-```
-
-I'll add some extra seasonal variation between to the mix
-
-```{r}
-x$Attribute_A[,,2] <- x$Attribute_A[,,1]*1.2
-x$Attribute_A[,,3] <- x$Attribute_A[,,1]*1.4
-x$Attribute_A[,,4] <- x$Attribute_A[,,1]*1.1
-```
-
-Now I create another four copies of this data, adding some random noise and a continuous decreasing trend.
-
-```{r}
-# Function to add random noise
-myRandom <- function(x) x*rnorm(1,1,.05)
-```
-
-```{r}
-y1 <- x
-y2 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% mutate(Attribute_A = Attribute_A*0.95)
-y3 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% mutate(Attribute_A = Attribute_A*0.90)
-y4 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% mutate(Attribute_A = Attribute_A*0.85)
-y5 <- x %>% st_apply(1:3, myRandom) %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% mutate(Attribute_A = Attribute_A*0.80)
-
-```
-
-```{r}
-# We combine the data into one cube for plotting:
-temp <- c(y1, y2, y3, y4, y5)
-
-ggplot() + 
-  geom_stars(data = temp) +
-  coord_equal() +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +  
-  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~time, ncol=4)
-```
-
-Saving these to file.
-
-```{r, eval=F}
-path <- "P:/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data"
-saveRDS(y1, paste0(path, "/y1.rds"))
-saveRDS(y2, paste0(path, "/y2.rds"))
-saveRDS(y3, paste0(path, "/y3.rds"))
-saveRDS(y4, paste0(path, "/y4.rds"))
-saveRDS(y5, paste0(path, "/y5.rds"))
-```
-
-##### Dummy regions and ET map
-
-Here is some dummy data for accounting areas (regions) and ecosystem occurrences as well, in case I need them later.
-
-```{r, fig.cap="Showing the example account area delineation (raster format)"}
-dummy_aa <- x %>%
-  filter(time == oldTime[1]) %>%
-  adrop() %>%
-  mutate(accountingAreas = rep(c(1,2), each=length(Attribute_A)/2)) %>%
-  select(accountingAreas)
-
-tm_shape(dummy_aa) +
-  tm_raster(style="cat")
-```
-
-```{r, fig.cap="Showing the example Ecosystem type delineation data."}
-dummy_et <- x %>%
-  filter(time == oldTime[1]) %>%
-  adrop() %>%
-  mutate(ecosystemType = rep(c(1,NA), length.out = length(Attribute_A))) %>%
-  select(ecosystemType)
-
-tm_shape(dummy_et) +
-  tm_raster(style="cat")
-```
+We are interested in _tm_, which is temperatur in celcius. For some reason the same variable is called _tg_ in the data itself.
 
@@
 
-## --------------------------------------------------
-
 ### Conceptual workflow
@@
 
-6.  Make trend figure and spatially aggregated maps,
+6.  Make trend figure and spatially aggregated maps
 
-### 1. Collate variable data series
-
-I include a for loop for importing the .nc files at the end.
-
-#### Filter data by dates
-
-Say we want to calculate the mean summer temperature. We then want to exclude the data for the times that is not withing our definition of summer.
-
-Let's try it on the dummy data. Remember this data had four time steps.
-
-```{r}
-st_get_dimension_values(x, "time")
-```
-
-Our real data has 365 time steps for each year.
-
-First I define my start and end dates. I want to keep only the summer data, defined as jun - aug.
-
-```{r}
-start_month <-  "06"
-end_month <- "08"
-```
-
-Then for each iteration I need to get the year as well
-
-```{r}
-(year_temp <- year(st_get_dimension_values(x, "time")[1]))
-```
-
-Then I can create a time interval object
-
-```{r}
-start <- ym(paste(year_temp, start_month, sep="-"))
-end <-  ym(paste(year_temp, end_month, sep="-"))
-(myInterval <- interval(start, end))
-```
-
-Then I filter
-
-```{r}
-x_aug <- x %>%
-  select(Attribute_A) %>%
-  adrop() %>%
-  filter(time %within% myInterval)
-st_get_dimension_values(x_aug, "time")
-```
-
-*Later I discover that \`dplyr::filter\` work after reading the whole object to file, and therefor it is too slow for our use. I therefore filter data a different way further down.*
-
-#### Aggregate across time within a year
-
-For this example I want to calculate the mean temperature over the summer.
-I therefore need to aggregate over time.
-Many climate indicators will need this functionality.
-There are two ways, either using `st_apply` or using `aggregate`.
-
-```{r}
-temp_names <- year(st_get_dimension_values(x_aug, "time")[1])
-
-microOut <- microbenchmark(
-st_apply =  x_aug %>%
-  st_apply(1:2, mean) %>%
-  setNames(paste0("v_",temp_names)),
-aggregate = x_aug %>%
-  aggregate(by = "year", mean),
-times=30
-)
-```
-
-The time dimension is now gone as we have aggregated across it.
-
-```{r, fig.cap="Comparing computation tome for two spatial aggregation functions."}
-autoplot(microOut)
-```
-
-`st_apply` is slightly faster, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year.
-I will try to use `st_apply`.
-
-But let me see if I can save time by masking to ET before doing this operation.
-
-```{r, fig.cap="Demonstrating the effect of maskig the dummy data using a perfectly aligne raster ET map."}
-x_masked <- x_aug
-x_masked[is.na(dummy_et)] <- NA
-plot(x_masked[,,,1])
-```
-
-```{r withMasking, fig.cap="Comparing computation time before and after masking of raster data. There is no performance increase by masking before aggregating."}
-autoplot(
-  microbenchmark(
-No_masking =  x_aug %>%
-  st_apply(1:2, mean) %>%
-  setNames(paste0("v_",temp_names)),
-Masking =  x_masked %>%
-  st_apply(1:2, mean) %>%
-  setNames(paste0("v_",temp_names))
- ), times=30
-)
-
-```
-
-Since there is no increase in performance from masking, as we see in Fig.
-\@ref(fig: withMasking) then it basically means we it is slower to mask beforehand, since the masking was not past of the benchmark assessment.
-
-So, this is my solution:
-
-```{r}
-# setNames dont work on stars proxy obejct, so I use rename instead
-lookup <- setNames("mean", paste0("v_", temp_names))
-
-x_summerMean <-   x_aug %>%
-  st_apply(1:2, mean) %>%
-  rename(all_of(lookup))
-```
-
-The *v\_* follows the naming convention we have developed for ourselves.
-
-```{r, fig.cap="Plotting the dummy data showing Attribute_A for the two dates as small maps, and a larger map showing the mean for year 2018"}
-ggarrange(
-ggplot() + 
-  geom_stars(data = x_aug) +
-  coord_equal() +
-  facet_wrap(~time) +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0)),
-
-ggplot() + 
-  geom_stars(data = x_summerMean) +
-  coord_equal() +
-  #facet_wrap(~time) +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +
-  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))
-)
-```
-
-Here's the whole first step of the conceptual workflow in action, on the dummy data.
-
-```{r}
-
-# Setting up the parameters
-dir <- substr(getwd(), 1,2)
-path <- ifelse(dir == "C:",
-       "P:/", 
-       "/data/P-Prosjekter2/")
-path2 <- paste0(path, 
-                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
-start_month <-  "06"
-end_month <- "08"
-myFiles <- list.files(path2, pattern=".rds", full.names = T)
-summerTemp_fullSeries <- NULL
-
-# Looping though the files in the directory
-system.time({
-for(i in 1:length(myFiles)){
-
-  temp <- readRDS(myFiles[i]) %>%
-    select(Attribute_A)
-
-  year_temp <- year(st_get_dimension_values(temp, "time")[1])
-  start <- ym(paste(year_temp, start_month, sep="-"))
-  end <-  ym(paste(year_temp, end_month, sep="-"))
-  myInterval <- interval(start, end)
-  lookup <- setNames("mean", paste0("v_", year_temp))
-
-  temp <- temp %>%
-    filter(time %within% myInterval) %>%
-    st_apply(1:2, mean) %>%
-    rename(all_of(lookup))
-
-  summerTemp_fullSeries <- c(temp, summerTemp_fullSeries)
-  rm(temp) # same computation time with and without this function, but more tidy this way
-}
-
-# Turn the attributes into a dimension and rename the new attribute
-summerTemp_fullSeries <- summerTemp_fullSeries %>%
-  merge(name="Year") %>%
-  setNames("Attribute_A")
-})
-```
-
-And this is the result.
-
-```{r, fig.cap="Temporal aggregation on the dummy data set."}
-ggplot()+
-  geom_stars(data = summerTemp_fullSeries) +
-  coord_equal() +
-  theme_void() +
-  scale_fill_viridis_c(option = "D") +  
-  scale_x_discrete(expand = c(0, 0)) +
-  scale_y_discrete(expand = c(0, 0))+
-  facet_wrap(~Year)
-```
-
-Now let's try it on the real data.
-
-*I initially tried to read stars proxy object and add steps to the call_list and execute these with `st_as_stars`, but this still returned proxy objects.
-Therefore I read the whole file in without proxy.
-To save time I can subset based on Julian dates.*
-
-I will use [parallelization](https://tmieno2.github.io/R-as-GIS-for-Economists/EE.html) to speed things up.
+### Step 1 - temporal aggregation within a year
 
@@
 ```{r, eval=F}
-cl <- makeCluster(2L)
-cl2 <- makeCluster(6L)
-cl3 <- makeCluster(10L)
-
-tic("No cluster")
-temp %>%
-    #filter(time %within% myInterval) %>%
-    st_apply(1:2, mean) %>%
-    rename(all_of(lookup)) 
-toc()
-
-tic("Two clusters")
-temp %>%
-    #filter(time %within% myInterval) %>%
-    st_apply(1:2, mean, CLUSTER = cl) %>%
-    rename(all_of(lookup)) 
-stopCluster(cl)
-toc()
-
-tic("Six clusters")
-temp %>%
-    #filter(time %within% myInterval) %>%
-    st_apply(1:2, mean, CLUSTER = cl2) %>%
-    rename(all_of(lookup)) 
-stopCluster(cl2)
-toc()
-
-tic("Ten clusters")
-temp %>%
-    #filter(time %within% myInterval) %>%
-    st_apply(1:2, mean, CLUSTER = cl3) %>%
-    rename(all_of(lookup)) 
-stopCluster(cl3)
-toc()
-
-```
-
--   No cluster: 21.401 sec elapsed
-
--   Two clusters: 21.546 sec elapsed
-
--   Six clusters: 14.548 sec elapsed
-
--   Ten clusters: 15.29 sec elapsed
-
--   Ten clusters (second run): 13.617 sec elapsed
-
-The RStudio server has 48 cores. More parallel cores is probably on average faster. The NINA guidelines is to use max 10 cores, and to remember to close parallel cluster when done.
-
-```{r, eval=F}
-saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.rds")
 write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff")
@@
 
-These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.
-*(This could perhaps have been solved with a `gc()`,.
-Later I save as RData with no problems).*
-
-```{r}
+```{r, include=F}
 summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
                                  proxy=F)
-# This is too slow:
-# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/summer_median_temp.rds")
-# but it could be that RData works better 
 ```
 
-```{r, eval=T}
-names(summer_median_temp)
-```
-
-```{r, eval=T}
-dim(summer_median_temp)
-```
-
-```{r, eval=T}
-st_get_dimension_values(summer_median_temp, "Year")
-```
-
-Note that GTiff automatically renames the third dimension *band* and also renames the attribute.
-
-```{r}
-st_get_dimension_values(summer_median_temp, "band")
-```
-
-I can rename them.
+Note that GTiff automatically renames the third dimension *band* and also renames the attribute. I can rename them.
 
@@
   setNames("temperature")
-dim(summer_median_temp)
 ```
@@
 
-### 2. Calculate reference values
+### Step 2 - calculate reference values
 
-#### Aggregate across reference years
+We need to first to define a reference period and then to subset our data `summer_median_temp`.
 
-We need to first to define a reference period and then to subset our data `summer_median_temp`. We can practice on the dummy data `y`.
-
-```{r}
-dir <- substr(getwd(), 1,2)
-path <- ifelse(dir == "C:",
-       "P:/", 
-       "/data/P-Prosjekter2/")
-path2 <- paste0(path, 
-                "41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data")
-myFiles <- list.files(path2, pattern=".rds", full.names = T)
-y <- NULL
-
-# Looping though the files in the directory
-for(i in 1:length(myFiles)){
-  temp <- readRDS(myFiles[i])
-  y <- c(temp, y)
-}
-```
-
-```{r}
-st_get_dimension_values(y, "time")
-```
-
-```{r}
-y_filtered <- y %>%
-  select(Attribute_A) %>%
-  filter(time %within% interval("1961-01-01", "1990-12-31"))
-```
-
-Now we aggregate across years, again using st_apply.
-
-```{r}
-median_sd <- function(x) { c(median = median(x), sd = sd(x))}
-```
-
-```{r}
-y_ref <- y_filtered %>%
-  st_apply(c("x", "y"), FUN =  median_sd)
-```
-
-```{r}
-dim(y_ref)
-```
-
-```{r}
-st_get_dimension_values(y_ref, "median_sd")
-```
-
-It's perhaps easier if median and sd are unique attributes, rather than levels of a dimension.
-
-```{r}
-y_ref <- y_ref %>% split("median_sd")
-```
-
-The attribute name *mean* we can change this to something more meaningful.
-
-```{r}
-y_ref <- setNames(y_ref, c("reference_upper", "sd"))
-```
-
-```{r, fig.cap="Showing the upper reference levels and the standard deviation from the dummy data set."}
-tmap_arrange(
-tm_shape(y_ref)+
-  tm_raster("reference_upper")
-,
-tm_shape(y_ref)+
-  tm_raster("sd",
-            palette = "-viridis")
-)
-```
-
-Let's transfer this over to the real data. First we need to rename our dimension values and turn them back into dates.
+First we need to rename our dimension values and turn them back into dates.
 
@@
 
-This is also something I can export.
-
 ```{r, eval=F}
-# write_stars can only export a single attribute 
-#write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
-
 saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
@@
 
-Pivot and turn dimension into attributes:
+Pivot and turn dimension into attributes, and rename attribute:
 
 ```{r, eval=T}
-summer_median_temp_ref_long <- summer_median_temp_ref %>% split("median_sd")
-```
-
-The attribute name is *mean*. We can change this to something more meaningful.
-
-```{r, eval=T}
-summer_median_temp_ref_long <- setNames(summer_median_temp_ref_long, c("reference_upper", "sd"))
+summer_median_temp_ref_long <- summer_median_temp_ref %>% 
+  split("median_sd") %>%
+  setNames(c("reference_upper", "sd"))
 ```
@@
 
-#### Combine
-
 I need to combine the variables and the ref values in one data cube
@@
   c(summer_median_temp_ref_long)
-
 ```
 
-#### Normalise variable
+### Step 3 - normalise variable
 
@@
 ```{r, eval=F}
-# Mutate
-
 # The break point scaling is actually not needed here, since 
@@
 Therefore, the scaled, regional indicator values will be calculated for distinct time steps.
-The time series can perhaps best be presented with yearly resolution.
+The time series can perhaps still be presented with yearly resolution.
 
-#### Plot time series as line graphs
+### Step 4 - Mask with ecosystem delineation map
 
-We could also plot a time series as a series of maps, but then we need to average over these 5 year time periods first. We can get to that later.
+This step we simply ignore for now. It should be relatively easy to do, when we have the maps.
 
-To plot time series I first need to do a spatial aggregation. One option is to spatially aggregate the stars object directly, using the `reg` data set:
+### Step 5 - Make figures
 
-```{r, eval=F}
-temp <- y_var_norm[1,,] %>%
-  aggregate(reg, mean, na.rm=T)
-plot(temp)
-temp2 <- temp %>%
-  st_as_sf() # too slow
-```
-
-However, this does no preserve the region name, and a subsequent `st_intersection` taks a little while and gives some boundary problems.
-
-I can also convert pixels to points and then do the intersection.
-
-```{r, eval=F}
-temp_points <- y_var_norm[1,,] %>%
-  st_as_sf(as_points = TRUE, merge = FALSE)
-temp_points2 <- temp_points %>%
-  st_intersection(reg) " too slow"
-```
-
-But this also took too long.
-
-I can perhaps aggregate, and then turn to points and intersect, but that did not work perfect either.
-
-I can rasterize the regions data set, turn it into a coded dimension and use st_apply, but that is tedious.
-
-Let me try using `exact_extract`, as an alternative to `aggregate`. This requires us to turn to the `terra` package for a bit.
+To plot time series I first need to do a spatial aggregation.
 
@@
 
-That was fast!
-
-I should also get the sd:
+I could also get the sd like this, if I wanted to base the indicator uncertainty on a measure of spatial variation:
 
@@
 regional_means <- readRDS("temp/regional_means.rds")
-regional_sd <- readRDS("temp/regional_sd.rds")
+#regional_sd <- readRDS("temp/regional_sd.rds")
 ```
@@
 #Adding the spatial sd
-(
-regional_means_long <- regional_sd %>%
-  select(!all_of(div)) %>%
-  pivot_longer(!region) %>%
-  separate(name, into=c("type", "year")) %>%
-  pivot_wider(names_from = type) %>%
-  rename(i_sd = i,
-         v_sd = v) %>%
-  left_join(regional_means_long, by=join_by(region, year))
-)
+#(
+#regional_means_long <- regional_sd %>%
+#  select(!all_of(div)) %>%
+#  pivot_longer(!region) %>%
+#  separate(name, into=c("type", "year")) %>%
+#  pivot_wider(names_from = type) %>%
+#  rename(i_sd = i,
+#         v_sd = v) %>%
+#  left_join(regional_means_long, by=join_by(region, year))
+#)
 ```
 
-```{r, fig.cap="Times series for median summer temperature centered on the median value during the reference period.
+```{r summer-temp-time-series, fig.cap="Times series for median summer temperature centered on the median value during the reference period.
 The reference period is indicated with a thick horizontal line.
@@
   theme_bw(base_size = 12)+
-  ylab("Sommertempratur\navvik fra 1961-1990")+
+  ylab("Sommertemperatur\navvik fra 1961-1990")+
   xlab("")+
@@
 
-Then we can take the mean over the last 5 years and add to a spatial representation.
+Then we can take the mean and sd over the last 5 years and add to a spatial representation.
 
@@
   summarise(indicator = mean(i),
-            spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
-            #,
-            #spatial_sd_mean = mean(i_sd),
-            #spatial_sd2_max = max(i_sd),
-            #spatial_sd2_length = length(i_sd)
+            sd = sd(i)
+            # If I inluded a spatial measure for the uncertainty, here is how I would carry the errors:
+            #spatial_sd = sqrt(sum(i_sd^2))/length(i_sd)
             )
@@
 
-```{r, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent spatial variation within regions and across years."}
+```{r summer-temp-time-periods, fig.cap="Scaled indicator values, aggregated over 5 year intervals. Errors represent temporal variation (standard errors) within regions and across 5 years."}
 labs <- unique(i_aggregatedToPeriods$period[order(i_aggregatedToPeriods$period_rank)])
@@
   geom_point() +
-  geom_errorbar(aes(ymin=indicator-spatial_sd, 
-                    ymax=indicator+spatial_sd), 
+  geom_errorbar(aes(ymin=indicator-sd, 
+                    ymax=indicator+sd), 
                     width=.2,
@@
 
-Finally, I can add these values to the sp object.
+Finally, I can add these values to the sp object with the accounting areas.
 
@@
   left_join(i_aggregatedToPeriods[i_aggregatedToPeriods$period_rank==5,], by=join_by(region))
-names(reg2)
 ```
 
-```{r, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
+```{r wall-to-wall-summer-temp-indicator, fig.cap="Summer tempreature indicator values for five accounting areas in Norway. SD is the spatial variation."}
 myCol <- "RdYlGn"
@@
 tm_inset <- tm_shape(reg2)+
-  tm_polygons(col="spatial_sd",
+  tm_polygons(col="sd",
               title="SD",

### 2023-03-14 DRAFT_alien_species.Rmd
write intro

--- 2023-03-14 slitasje.Rmd
@@
-# (PART\*) INDICATORS{-}
-
-# Slitasje og kjørespor
+---
+title: "Alien species"
+output:
+  html_document
+---
+# Alien species {#alien-species}
 
@@
 ## Introduction
-This indicator represent human caused damage or **disturbance to soils and vegetation**, typically from things like recreational activities and off-road vehicle traffic.
-Effects from domestic animal is not included here, and the upper reference value is 0% disturbance.
-The data come from a combination non-random field surveys and area-representative surveys.
+This indicator represent the the relative area with an absence of alien species.
+It is in a way the flip-side of _alien species abundance_.
+The later can perhaps be classed as a pressure variable, whilst the former reacts as direct and linear consequence of changes to this variable.
+
+A related variable has been used previously for [alpine](https://ninanor.github.io/IBECA/faktaark.html#Frav%C3%A6r_av_fremmede_arter) and forest ecosystems in Norway.
+These indicators were based on ANO, a national monitoring program, where alien species cover was estimated to the closest whole %.
+The upper and lower reference levels that was used back then was 0 or 100% of the species composition made up of alien species, respectively.
+The threshold value between good and reduced condition was defined as 5%.
+
+This time we are using the [nature type mapping dataset](#naturtype) which records the _effect_ of alien species,, rather than the % cover.
+The recording scale is called R7 (Fig.
+\@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with %cover recorded in ANO (Fig.
+\@ref(fig:fremmedart-prosentVSinnslag)).
+In order to combine these two data sets, we must convert one of these scales to match the other.
+
+Here I base the indicator on the R7 scale, since this is the bigger data set. I therfore convert %cover into R7 classes based on what I can interpret from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
+
+```{r R7, fig.cap="The R7 scale used when recording the effect of alien species in the nature type monitoring program"}
+knitr::include_graphics("images/R7 scale.PNG")
+```
+
+```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring progra with %cover (y-axis), which is how alien species is recorded in ANO."}
+knitr::include_graphics("images/fremmedart_prosentVSinnslag.PNG")
+```
 

### 2023-03-14 DRAFT_alien_species.Rmd
filter nature type data

@@
 ## Introduction
-This indicator represent the the relative area with an absence of alien species.
-It is in a way the flip-side of _alien species abundance_.
-The later can perhaps be classed as a pressure variable, whilst the former reacts as direct and linear consequence of changes to this variable.
+This indicator represent the the ecological on-site effect from  alien species.
+Whilst the introduction of alien species is best considered a pressure variable, the physical abundance or relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a good condition indicator.
 
@@
 The threshold value between good and reduced condition was defined as 5%.
+In this approach, the indicator represented the relative area with an absence of alien species.
+It is in a way the flip-side of _alien species abundance_.
+The latter was said to be a pressure variable, whilst the former reacts as direct and linear consequence of changes to this variable, and was therefore accepted as a condition indicator.
+We do not make this distinction here.
 
@@
 
-Here I base the indicator on the R7 scale, since this is the bigger data set. I therfore convert %cover into R7 classes based on what I can interpret from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
+Here I base the indicator on the R7 scale, since this is, presumably, the bigger data set.
+I therefore convert %cover into R7 classes based on what I can interpret from Fig.
+\@ref(fig:fremmedart-prosentVSinnslag).
+
+Upper reference value is then 1, and lower reference value is 7, on the R7 scale.
+For the threshold value I will set this to value 4 on the R7 scale. 
+This is based on interpretation of the definition of this value in Fig. \@ref(fig:R7). 
+Step 4 on the R7 scale is the step that, concordantly, best matches the 5% threshold in the old approach, as seen from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
 
@@
 
-```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring progra with %cover (y-axis), which is how alien species is recorded in ANO."}
+```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO.
+Cropped from Evju et al (2022)."}
 knitr::include_graphics("images/fremmedart_prosentVSinnslag.PNG")
@@
 
-Since the data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
-We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region (_five regions in Norway_) where it was recorded.
-We then calculate an aerea weighted mean (and error) indicator value for each HIA and region combination, as long as there is more than 100 data point for a give combination of HIA and region.
+Since the nature type data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
+We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region.
+In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas as regions (5 accounting areas in Norway).
+For this indicator, we expect more gradients also inside accounting areas, and therefore we define areas as municipalities.
+This means that there will be more missing data in our final indicator maps, since not all municipalities will have sufficient data points to estimate indicator values.
+
+We then calculate an area weighted mean (and error) indicator value for each HIA and municipality combination, as long as there is more than _n_ data points for a given combination of HIA and municipality.
 
@@
 
-1.  Import [Nature type data](#NTM) data set (incl. GRUK) and [ANO data
-    set](#ANO-import)
+1.  Import [Nature type data](#NTM2) data set (incl. GRUK) and [ANO data
+    set](#ANO-import2)
 
@@
 
-4.  For each polygon, extract the NiN-variable that has the lowest value
-    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype).
-
 5.  [Combine](#combine-nt-ano) data sets
 
-6.  [Scale](#scale-slitasje-ind) the `slitasje` variable based on
+6.  [Scale](#scale-alien-ind) the `alien species` variable based on
     reference values.
 
-7.  Define [homogeneous impact areas](#HIA) (HIA)
+7.  Define [homogeneous impact areas](#HIA) (HIA) based on an infrastructure index and municipalities
 
 8.  [Aggregate and spread](#spread-slitasje) indicator value across HIAs
-    and assessment regions
+    (and municipalities)
 
@@
 of the Environment Agency. I also import a data set called
-[ANO](#ANO-import), which you can read about
+[ANO](#ANO-import2), which you can read about
 [here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)
@@
 
-The variables are recorded on a coarse four-step ordinal scale (Fig.
-\@ref(fig:four-step)).
+The variables are recorded on a seven-step ordinal scale (Fig.
+\@ref(fig:R7)) or a %cover of alien species.
 
@@
 and define the variable as belonging to the year 2020 (read more
-[here](#scaled-slitasje-variable)).
+[here](#scaled-alien-variable)).
 
@@
 
-For a run through of the data set, see [here](#naturtype).
+For a run through of the nature type data set, see [here](#naturtype).
 
@@
 The indicator is tagged to the *Økologisk egenskap* called
-**Primærproduksjon** (Primary productivity). This is tentative, and
-perhaps *abiotiske forhold* is more suited. But the thought behind the
-choice is that *slitasje* affect the potential for primary productivity.
+**Funksjonelt viktige arter og strukturer** (Functionally important species and structures).
 
@@
 
-The indicator is tagged as a **Physical state characteristics**
-indicator. This is quite clear. It's a type of abiotic characteristic that
-has to do with the physical structure rather than the chemical
-composition.
+The indicator is tagged as a **B1 Compositional state characteristics**
+indicator.
 
@@
 
-_Slitasje_ is not thought to exhibit collinearity with any other indicator at
+_Alien species_ is not thought to exhibit collinearity with any other indicator at
 the present.
@@
 The reference condition is one with minimal negative human impact. This
-is also true for semi-natural ecosystems. For example, in the reference
-condition, *slitasje* in semi-natural ecosystems is defined to be zero,
-even though at no point in time would this condition be realized.
+is also true for semi-natural ecosystems. In a sense, the reference condition is pre 1800s,
+since alien species are defined as species established post 1800.
 
@@
 
-* Upper = 0%
-
-* Threshold = 10%
-
-* Lower = 100%
-
-The upper reference value is 0 (no *slitasje*), and the lower reference
-value is 100%. Note that 100% _slitasje_ does not mean that all the area
-must be scarred or damaged, but that all hypothetical quadrats around a
-point is affected (the variable is frequency-driven, and not
-amount-driven). The threshold for good ecosystem condition is 10%
-damage. Perhaps is should be set somewhat higher, like 20 or 30%.
-The indicator value could be interpretted differently at the polygon
-vs at the landscape scale. At the polygon scale, 10% damage may seem 
-like not that much, whilst at the landscape scale, 10% seems more serious.
-This final threshold value should be debated and illustrated with examples.
-
-Read about the normalisation [here](##scaled-slitasje-variable).
+* Upper = 1 (R7 scale), corresponding to 0% alien species
+
+* Threshold = 4 (R7 scale), corresponding to about 5-30% alien species cover.
+
+* Lower = 7 (R7 scale), corresponding to 100% alien species cover.
+
+Read about the normalisation [here](##scaled-alien-variable).
 
@@
 
-No additional references.
+[Nature type data](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/naturkartlegging/naturtyper/)
+
+[ANO](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)
+
+Evju, M., Skrindo, A.B. & Solstad, H. (red.) 2022. Overvåking av åpen grunnlendt kalkmark 2021‒2024. Årsrapport 2022. NINA Rapport 2195. Norsk institutt for naturforskning.
 
@@
 
-#### Nature type mapping {#NTM}
+#### Nature type mapping {#NTM2}
 

### 2023-03-15 DRAFT_alien_species.Rmd
minor text edits

@@
 For the threshold value I will set this to value 4 on the R7 scale. 
-This is based on interpretation of the definition of this value in Fig. \@ref(fig:R7). 
+This is based on interpretation by the author and based of the definition of this value in Fig. \@ref(fig:R7). 
 Step 4 on the R7 scale is the step that, concordantly, best matches the 5% threshold in the old approach, as seen from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
@@
 We also have a separate [summary file](##exp-natureType-summary) where
-the nature types are manually mapped the NiN variables and to the correct NiN-main types. We can use this to find the nature types associated with the NiN main types
+the nature types are manually mapped the NiN variables and to the correct NiN-main types. 
+We can use this to find the nature types associated with the NiN main types
 that we are interested in.
@@
 
-We are only interested in a mapping units that include our target variables.
-
-```{r}
-myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
+We are only interested in a mapping units that include our target variables. The variable _fremmedartsinnslag_ is coded as 7FA.
+
+```{r}
+myVars <- '7FA'
 ```
@@
   mutate(keepers = sum(c_across(
-    myVars)>0, na.rm=T)) %>%
-  filter(keepers > 0) %>%
+    all_of(myVars))>0, na.rm=T)) %>%
+  filter(keepers >0) %>%
   select(Nature_type, NiN_mainType, Year)
@@
 
-This deleted `r nrow(naturetypes_summary_import) - nrow(naturetypes_summary)` nature types and left us with these:
+This deleted `r nrow(naturetypes_summary_import) - nrow(naturetypes_summary)` nature types and left us with these `r nrow(naturetypes_summary)` naturetypes where 7FA was recorded:
 

### 2023-03-15 DRAFT_alien_species.Rmd
combine with ano
Merge branch 'main' into fremmedarter

@@
 ```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO.
+I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here exualrs to an R7 value of 1.
 Cropped from Evju et al (2022)."}
@@
 
-This deleted `r nrow(naturetypes_summary_import) - nrow(naturetypes_summary)` nature types and left us with these `r nrow(naturetypes_summary)` naturetypes where 7FA was recorded:
+This deleted `r nrow(naturetypes_summary_import) - nrow(naturetypes_summary)` nature types and left us with these `r nrow(naturetypes_summary)` nature types where 7FA was recorded:
 
@@
 
-```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation.", fig.height=10, fig.width=4}
+```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=4}
 ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
@@
   coord_flip()+
-  theme_bw(base_size = 12)+
+  theme_bw(base_size = 10)+
   theme(legend.position = "top",
@@
         legend.direction = "vertical")+
-  #guides(fill = "none")+
+  guides(fill = "none")+
   xlab("")+
-  ylab("Number of localities")
+  ylab("Number of localities")+
+  facet_wrap(.~hovedøkosystem, ncol=1, scale="free")
 ```
@@
 
-Now I need to create a single row per locality with a new variable which
-is a product/function of the four variables "7SE", "7TK", "PRTK" and
-"PRSL".
-I want to base this indicator on whichever of the variables have the highest value (worst condition).
-Therefore I first convert the ordinal scale to a continuous scale, using the median value of each ordinal step.
-
-The variables use slightly different scales. PRSL and PRTK use this 8
-step scale:
-
-```{r eight-step, fig.cap="Eight step condition scale"}
-knitr::include_graphics("images/mdirPRscale.PNG")
-```
-
-7TK and 7SE use a 4 step scale
-
-```{r four-step, fig.cap="Four step condition scale"}
-knitr::include_graphics("images/ninAR4b.PNG")
-```
-
-I here transfer these ordinal variables over to a continuous scale.
-
-```{r}
-naturetypes <- naturetypes %>%
-  mutate(NiN_variable_value = case_when(
-    NiN_variable_code %in% c("var_7TK", "var_7SE") ~ 
-      case_match(NiN_variable_value,
-             0 ~ 0,
-             1 ~ mean(c(0, 1/16))*100,
-             2 ~ mean(c(1/16, 1/2))*100,
-             3 ~ 75),
-    NiN_variable_code %in% c('var_PRTK', 'var_PRSL') ~ 
-      case_match(NiN_variable_value,
-             0 ~ 0,
-             1 ~ 1.5,
-             2 ~ mean(c(3, 6.25)),
-             3 ~ mean(c(6.25, 12.5)),
-             4 ~ mean(c(12.5, 25)),
-             5 ~ mean(c(25, 50)),
-             6 ~ mean(c(50, 75)),
-             7 ~ mean(c(75, 100)))
-    )
-  )
-```
-
-I will create a new data table where I calculate the new
-variable which I then paste back into the sf object.
-
-```{r}
-naturetypes_wide <- pivot_wider(naturetypes,
-                                names_from = "NiN_variable_code",
-                                values_from = "NiN_variable_value",
-                                id_cols = "identifikasjon_lokalId")
-naturetypes_wide <- as.data.frame(naturetypes_wide)
-head(naturetypes_wide, 10)
-```
-
-First I will combine 7TK and PRTK, and also 7SE and PRSL
-
-```{r}
-naturetypes_wide <- naturetypes_wide %>%
-  mutate(TK = if_else(
-    is.na(var_PRTK), var_7TK, var_PRTK),
-    SE = if_else(
-      is.na(var_PRSL), var_7SE, var_PRSL)
-    )
-```
-
-SE has `r nrow(naturetypes_wide[is.na(naturetypes_wide$SE),])`
-NA's, and TK has
-`r nrow(naturetypes_wide[is.na(naturetypes_wide$TK),])`.
-
-```{r, fig.cap="7SE scores in the naturetype dataset", fig.width=5}
-temp <- naturetypes_wide %>%
-  group_by(SE) %>%
-  summarise(sum = n())
-
-ggplot(temp, aes(x = factor(SE),
-                 y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
+Removing NA's
+
+```{r}
+naturetypes <- naturetypes[!is.na(naturetypes$NiN_variable_value),]
+```
+
+```{r alien-naturetype, fig.cap="Alien species variable scores."}
+ggplot(naturetypes, aes(x = NiN_variable_value))+
+  geom_histogram(fill="grey",
+           colour = "black",
+           binwidth = 1)+
   theme_bw(base_size = 12)+
-  labs(x = "7SE or PRSL score",
-       y = "Number of localities")
-```
-
-The NA fraction is quite big. These are localities with 7TK or PRTK recorded,
-but not 7SE or PRSL.
-
-```{r, fig.cap="7TK scores in the nature type dataset", fig.width=5}
-temp <- naturetypes_wide %>%
-  group_by(TK) %>%
-  summarise(sum = n())
-
-ggplot(temp, aes(x = factor(TK),
-                 y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
-  theme_bw(base_size = 12)+
-  labs(x = "7TK og PRTK score",
-       y = "Number of localities")
-```
-
-Then I can combine these two variables into a composite variable called
-`slitasje`
-
-```{r, warning=F}
-naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("TK", "SE")], 1, max, na.rm=T)
-```
-
-When both variables are NA, we get -Inf. There were `r nrow(naturetypes_wide[naturetypes_wide$slitasje<0,])` cases of this. Removing these now.
-
-```{r}
-naturetypes_wide <- naturetypes_wide[naturetypes_wide$slitasje>=0,]
-```
-
-```{r slitasje-naturetype, fig.cap="Slitasje scores.", fig.width=3}
-temp <- naturetypes_wide %>%
-  group_by(slitasje) %>%
-  summarise(sum = n())
-
-ggplot(temp, aes(x = factor(slitasje),
-                 y = sum))+
-  geom_bar(stat="identity",
-           fill="grey",
-           colour = "black")+
-  theme_bw(base_size = 12)+
-  labs(x = "Slitasje score converted to percentage",
-       y = "Number of localities") + 
-  theme(axis.text.x = element_text(angle=90, vjust=0.5))
-```
-
-It appears most localities are in good condition.
-
-I would also like to know, when both tk and SE was recordedm
-how often TK was defining of the slitasje-indicator, and how 
-often it was SE. I can do this by taking the difference.
-
-```{r, fig.cap="Counting the number of cases where TK or SE was defining of the slitasje indicator score.", fig.width=3}
-
-diff <- naturetypes_wide$SE - naturetypes_wide$TK
-
-diff <- ifelse(diff == 0, "TK and SE",
-               ifelse(diff <0, "TK", "SE"))
-
-diff_tbl <- as.data.frame(table(diff))
-
-ggplot(diff_tbl, aes(x = diff, y = Freq))+
-  geom_bar(stat = "identity",
-           colour = "black",
-           fill = "grey")+
-  theme_bw(base_size = 12)+
-  labs(x = "Defining variable",
-       y = "Number of localities")
-```
-
-Looks like SE (= 7SE or PRSL) is more likely to be in a detrimental state.
-
-Now I will copy these slitasje-values into the sf object.
-
-```{r}
-naturetypes$slitasje <- naturetypes_wide$slitasje[match(naturetypes$identifikasjon_lokalId, naturetypes_wide$identifikasjon_lokalId)]
-#nrow(naturetypes[is.na(naturetypes$slitasje),])  # 13 cases
-naturetypes <- naturetypes[!is.na(naturetypes$slitasje),]
-```
+  labs(x = "Alien species variable score") +
+  facet_wrap(.~hovedøkosystem, scales="free")
+```
+
+It appears most localities are in good condition, and most localities have no alien species at all.
 
@@
 
-#### ANO {#ANO-import}
+_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all futute GRUK data, will not be included in this data set._
+
+#### ANO {#ANO-import2}
 
@@
 circle of 250 m2 is visualised, and the main ecosystem is recorded.
-Depending on the main ecosystem, certain NiN variables are also
-recorded. 7SE is recorded for våtmark, but not semi-natural areas or
-naturally open areas. 7TK is recorded in våtmark and naturally open
-areas only.
-
-| Variable | Våtmark | Naturlig åpne områder | Semi-naturlige områder |
-|----------|---------|-----------------------|------------------------|
-| 7SE      | X       | \-                    | \-                     |
-| 7TK      | X       | X                     | \-                     |
-
-: Table showing which main ecosystems the two NiN variables 7SE and 7TK
-is recorded in withing the ANO data set.
-
-It would be very nice to have 7SE recorded for naturally open areas.
-This variable is very relevant here.
-
-I think I will only use våtmark here since my approach will be to the
-'worst value' of the variable 7SE and 7TK (see below). I think not
-having 7SE for naturally open areas will underestimate the degree of
-degredation in these areas.
-
-> *ANO could harmonise better with the Naturetype data set if it included
-> 7SE and 7TK for Naturally open and Semi natural ecosystem.*
-
-```{r}
+
+The %cover of alien vascular plant is estimated, independent on main ecosystem type. 
+In contrast to the nature type data set, ANO only looks at vascular plants, and then also only the three strictest categories
+in the alien plant list (SE, HI, PH). It is therefore not completely unproblematic to combine these to data sets,
+but we will do so anyway. We expect plants to be the by far most common species group to influence the 7FA variable as well.
+And similarly, we expect the three strictest  alien plant categories to contain the most aggressive species which will
+drive the 7FA variable more so than the other categories.
+
+```{r, message=F}
+ano_eco <- c("vaatmark", "naturlig_apne", "semi_naturlig_mark")
 ano <- sf::st_read(paste0(pData, "Naturovervaking_eksport.gdb"),
                    layer = "ANO_SurveyPoint") %>%
-  dplyr::filter(hovedoekosystem_punkt == "vaatmark")
+  dplyr::filter(hovedoekosystem_punkt %in% ano_eco)
 ```
@@
 
-Let's now look at the distribution of the variables. First I need to
-separate the variable name from the values themselves. Now the data
-looks like this:
-
-```{r}
-ano$bv_7se[1:6]
-```
-
-So I create a new variable prefixed var\_:
-
-```{r}
-ano$var_7SE <- as.numeric(sub(pattern = "7SE_",
-                 replacement = "",
-                 x = ano$bv_7se))
-```
-
-The NA's is the case of blank cells. The field app should not allow
-users leaving this field blank. I will need to remove these rows. There
-are fourteen cases:
-
-```{r}
-nrow(ano[is.na(ano$var_7SE),])
-```
-
-```{r}
-ano <- ano[!is.na(ano$var_7SE),]
-```
-
-Same with the other variable 7TK
-
-```{r}
-ano$var_7TK <- as.numeric(sub(pattern = "7TK_",
-                 replacement = "",
-                 x = ano$bv_7tk))
-```
-
-No NA's this time.
-
-> *The ANO field app should not allow user to leave blank cells. This
-> resulted in the exclusion of data.*
-
-```{r, fig.cap="Distribution of 7TK and 7SE scores in the ANO data"}
-par(mfrow=c(1,2))
-barplot(table(ano$var_7TK), xlab="7TK scores")
-barplot(table(ano$var_7SE), xlab="7SE scores")
-```
-
-The ANO sites seem to be in an even better condition than the localities in the naturtype data set.
-
-Combining these two variables 7TJ and 7SE into one, same as for the nature
-type data.
-
-```{r}
-temp <- as.data.frame(ano)
-temp$slitasje <- apply(temp[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
-ano$slitasje <- temp$slitasje
-```
-
-Then I extract the mid values for each ordinal range step, like I did previously for the nature type data set.
+Let's now look at the distribution of the variable. It is coded as `fa_total_dekning`. 
+First, there are `r length(ano[is.na(ano$fa_total_dekning),])` NA's in this column which we can remove.
+
+```{r}
+ano <- ano[!is.na(ano$fa_total_dekning),]
+```
+
+Let us plot the distribution of values.
+
+```{r ano-alien-dist, fig.cap"Distribution of %cover of alien plants variable in the ANO data set"}
+ggplot(ano, aes(x = fa_total_dekning))+
+  geom_histogram(fill="grey",
+           colour = "black",
+           binwidth = 1)+
+  theme_bw(base_size = 12)+
+  labs(x = "% cover alien vascular plants") +
+  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
+```
+
+The ANO localities seemingly have less alien plants in them than the nature type localities. This can be du to sampling biases in the latter.
+
+Now I will convert these %cover values into 7FA classes. This is a critical step, since there is not one-to-one mapping here, 
+but rather there is considerable overlap, as can be seen in Fig. \@ref(fig:fremmedart-prosentVSinnslag).
 
@@
 ano <- ano %>%
-  mutate(slitasje = case_match(slitasje,
-             0 ~ 0,
-             1 ~ mean(c(0, 1/16))*100,
-             2 ~ mean(c(1/16, 1/2))*100,
-             3 ~ 75))
-```
-
-```{r, fig.cap="Distribution of slitasje scores in the ANO data", fig.width=6}
-barplot(table(ano$slitasje), xlab="% Slitasje", ylab = "Number of localities")
-```
-
-##### Combine Naturtype data and ANO {#combine-nt-ano}
+  mutate(var_7FA = case_when(
+             fa_total_dekning>90 ~ 7,
+             fa_total_dekning>25 ~ 6,
+             fa_total_dekning>10 ~ 5,
+             fa_total_dekning>5 ~ 4,
+             fa_total_dekning>2 ~ 3,
+             fa_total_dekning>1 ~ 2,
+             fa_total_dekning<1 ~ 1))
+```
+
+```{r ano-alien-dist-R7, fig.cap"Distribution of R7 values in the ANO data set"}
+ggplot(ano, aes(x = var_7FA))+
+  geom_histogram(fill="grey",
+           colour = "black",
+           binwidth = 1)+
+  theme_bw(base_size = 12)+
+  labs(x = "% cover alien vascular plants\n converted to the R7 scale") +
+  facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
+```
+
+##### Combine Naturtype data and ANO {#combine-nt-ano2}
 
@@
 naturetypes$origin <- "Nature type mapping"
-ano$hovedøkosystem <- "Våtmark"
+ano$hovedøkosystem <- ano$hovedoekosystem_punkt
 ano$kartleggingsår <- ano$aar
@@
 ```{r}
-slitasje_data <- dplyr::bind_rows(select(ano,
+alien_data <- dplyr::bind_rows(select(ano,
                                   GlobalID,
@@
                                   area,
-                                  slitasje,
+                                  var_7FA,
                                   SHAPE), 
@@
                                   area,
-                                  slitasje,
+                                  var_7FA = NiN_variable_value,
                                   SHAPE))
-
-unique(st_geometry_type(slitasje_data))
 ```
@@
 
-The ANO data is point data, and the nature type data is *multipolygon*.
+The ANO data is point data and the nature type data is *multipolygon*:
+```{r}
+unique(st_geometry_type(alien_data))
+```
+
 Because later we might want to save this as a shape file, we cannot have
@@
 ```{r}
-slitasje_data_points <- slitasje_data %>%
+alien_data_points <- alien_data %>%
   mutate(g_type = st_geometry_type(.)) %>%
   filter(g_type =="POINT") %>%
-  st_buffer(sqrt(slitasje_data$area/pi))
+  st_buffer(sqrt(alien_data$area/pi))
 ```
@@
 
-```{r new-area, fig.cap="Checking that the area of the new polygons fall in line with the proportion of each point which is part of the main ecosystem."}
-slitasje_data_points$area2 <- st_area(slitasje_data_points)
-plot(slitasje_data_points$area, slitasje_data_points$area2,
+```{r new-area2, fig.cap="Checking that the area of the new polygons fall in line with the proportion of each point which is part of the main ecosystem."}
+alien_data_points$area2 <- st_area(alien_data_points)
+plot(alien_data_points$area, alien_data_points$area2,
      xlab = "Target area",
@@
 ```{r}
-slitasje_data_polygons <- slitasje_data %>%
+alien_data_polygons <- alien_data %>%
   mutate(g_type = st_geometry_type(.)) %>%
@@
 
-slitasje_data <- bind_rows(slitasje_data_points, slitasje_data_polygons)
-
-unique(st_geometry_type(slitasje_data))
+alien_data <- bind_rows(alien_data_points, alien_data_polygons)
+
+unique(st_geometry_type(alien_data))
 ```
@@
 
-#### Distribution of Slitasje scores
-
-```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
-temp <- as.data.frame(table(slitasje_data$slitasje))
+#### Distribution of 7FA scores
+
+```{r, fig.cap="7FA (alien species) scores (ANO Naturetype (and GRUK) data combined)."}
+temp <- as.data.frame(table(alien_data$var_7FA))
 ggplot(temp, aes(x = Var1,
@@
   theme_bw(base_size = 12)+
-  labs(x = "slitasje score",
+  labs(x = "7FA score",
        y = "Number of localities")
@@
 
-```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator.", fig.width=4}
-temp <- as.data.frame(table(slitasje_data$origin))
+```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the alien species indicator.", fig.width=4}
+temp <- as.data.frame(table(alien_data$origin))
 

### 2023-03-15 DRAFT_alien_species.Rmd
try to normalise

@@
 
+Shift variable to start from 1 rather than from 0.
+
+```{r}
+naturetypes$NiN_variable_value <- naturetypes$NiN_variable_value+1
+```
+
 ```{r alien-naturetype, fig.cap="Alien species variable scores."}
@@
   labs(x = "Alien species variable score") +
-  facet_wrap(.~hovedøkosystem, scales="free")
+  facet_wrap(.~hovedøkosystem, scales="free_y")
 ```
@@
 
-### Scaled indicator values {#scale-slitasje-ind}
+### Scaled indicator values {#scale-alien-ind}
 
@@
     given HIA, and extrapolate the area weighted mean of those values to
-    the entire HIA withing a given geographical region. Errors are
+    the entire HIA withing a given geographical region (municipality). Errors are
     calculated from bootstrapping.
@@
 3.  Take the new area weighted indicator values for the HIAs and match
-    them with ecosystem occurrences of the same HIA (possibly using
-    randomly sampled points from the available ecosystem occurrences).
+    them with ecosystem occurrences of the same HIA (_possibly using
+    randomly sampled points from the available ecosystem occurrences_).
     The errors are the same as for the HIAs.
@@
 ```{r}
-upper <- 0
-lower <- 100
-threshold <- 5
-```
-
-This implies that it is impossible to detect or measure a state when it
-is in its most degraded form (100% disturbed). This is a problem with
-the original data resolution. But I will compensate somewhat for this by
-using a non-linear transformation.
-
-Also, the threshold value is set quite conservatively (I think), and should be
-discussed.
+upper <- 1
+lower <- 7
+threshold <- 4
+```
 
@@
 values for the normalisation to work. This is a bug which can be fixed
-inside eaTools.
-
-<!-- THE BREAKPOINT IS AT 0.5 RATHER THAN 0.6. TEST CODE FROM HERE: https://ninanor.github.io/eaTools/articles/normalise-condition-variable.html -->
-
-```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=5}
-eaTools::ea_normalise(data = slitasje_data,
-                      vector = "slitasje",
-                      upper_reference_level = lower,
-                      lower_reference_level = upper,
+inside eaTools (see [GitHub issue](https://github.com/NINAnor/eaTools/issues/15)).
+
+```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=3}
+eaTools::ea_normalise(data = alien_data,
+                      vector = "var_7FA",
+                      upper_reference_level = upper,
+                      lower_reference_level = lower,
                       break_point = threshold,

### 2023-03-15 DRAFT_alien_species.Rmd
import municipalities

@@
 
+Municipalities
+
+```{r}
+path1 <- ifelse(dir == "C:", 
+               "R:/",
+               "/data/R/")
+
+path <- paste0(path1, "GeoSpatialData/AdministrativeUnits/Norway_AdministrativeUnits/Original/Norway_Municipalities/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")
+
+# find the correct layer
+#st_layers(path)
+muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
+```
+
+There are som multisurfaces in here that I will convert to multoipolygons before plotting.
+
+```{r kommuner, fig.cap="Municipalities in Norway."}
+muni <- st_cast(muni, "MULTIPOLYGON")
+tm_shape(muni)+
+  tm_polygons(col="kommunenummer")+
+  tm_layout(legend.outside = T)
+```
+
 ### Scaled indicator values {#scale-alien-ind}
@@
 ```{r}
-upper <- 1
-lower <- 7
+low <- 1
+high <- 7
 threshold <- 4
@@
 
-```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=3}
+```{r alien-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=4}
 eaTools::ea_normalise(data = alien_data,
                       vector = "var_7FA",
-                      upper_reference_level = upper,
-                      lower_reference_level = lower,
+                      upper_reference_level = high,
+                      lower_reference_level = low,
                       break_point = threshold,
@@
 ```{r}
-mean(slitasje_data$kartleggingsår)
+mean(alien_data$kartleggingsår)
 ```
@@
 ```{r}
-slitasje_data$i_2020 <- eaTools::ea_normalise(data = slitasje_data,
-                      vector = "slitasje",
-                      upper_reference_level = lower,
-                      lower_reference_level = upper,
+alien_data$i_2020 <- eaTools::ea_normalise(data = alien_data,
+                      vector = "var_7FA",
+                      upper_reference_level = high,
+                      lower_reference_level = low,
                       break_point = threshold,
@@
 
-I want to check that HIF is in fact a good predictor for *slitasje*.
-
-I also want to split these four HIA classes based on their region. To do
+I want to check that HIF is in fact a good predictor for *alien species*.
+
+I also want to split these four HIA classes based on municipality. To do
 this I need the two layers to have the same CRS.

### 2023-03-20 DRAFT_alien_species.Rmd
start to validate

@@
   filter(naturtype %in% naturetypes_summary$Nature_type) %>%
-  mutate(hovedøkosystem = recode(hovedøkosystem,
-                                 "naturligÅpneOmråderUnderSkoggrensa" = "naturligÅpneOmråderILavlandet"),
+  mutate(hovedøkosystem = case_match(hovedøkosystem,
+                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
+                                  .default = hovedøkosystem),
          area = st_area(.)) %>%
@@
 
-```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=4}
+```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=6}
 ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
@@
                    layer = "ANO_SurveyPoint") %>%
-  dplyr::filter(hovedoekosystem_punkt %in% ano_eco)
+  dplyr::filter(hovedoekosystem_punkt %in% ano_eco) %>%
+   mutate(hovedoekosystem_punkt = case_match(hovedoekosystem_punkt,
+                                       "vaatmark" ~ "våtmark",
+                                       "naturlig_apne" ~ "naturligÅpneOmråderUnderSkoggrensa",
+                                       "semi_naturlig_mark" ~ "semi-naturligMark",
+                                       .default = hovedoekosystem_punkt))
 ```
@@
 muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
-```
-
-There are som multisurfaces in here that I will convert to multoipolygons before plotting.
+
+```
+
+```{r, include=F, eval=FALSE}
+# If I import the manuicipality names:
+muni_names <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "administrativenhetnavn")
+#The encoding option fail for gdb-files, so I fix the names again here
+muni_names$navn <- stringi::stri_encode(muni_names$navn, from = "UTF-8", to = "UTF-8")
+
+# ... there is no common ID to match these with the municipality polygons!!
+```
+
+There are some multi-surfaces in here that I will convert to multi-polygons before plotting.
 
@@
 
-I need to do a little trick and reverse the upper and lower reference
-values for the normalisation to work. This is a bug which can be fixed
-inside eaTools (see [GitHub issue](https://github.com/NINAnor/eaTools/issues/15)).
-
 ```{r alien-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=4}
@@
 
-This normalisation seems reasonable to me. I can save it as indicator
-values. There is no point yet making this a time series, and I will
+There is no point yet making this a time series, and I will
 assign all the indicator value to the average year of the data.
@@
 ```{r}
-st_crs(HIA) == st_crs(regions)
-```
-
-Since the HIA map and the _region_ map are completely overlapping (HIA was masked using the region map), we can get the
-intersections
+st_crs(HIA) == st_crs(muni)
+```
+
+Then we get the intersections (unique combinations)
 
 ```{r, eval=F}
-HIA_regions <- eaTools::ea_homogeneous_area(HIA,
-                             regions,
+HIA_muni <- eaTools::ea_homogeneous_area(HIA,
+                             muni,
                              keep1 = "infrastructureIndex",
-                             keep2 = "region")
-saveRDS(HIA_regions, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_regions.rds")
+                             keep2 = "kommunenummer")
+saveRDS(HIA_muni, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/HIA_muni.rds")
 ```
@@
 # silently read cached file
-HIA_regions <- readRDS(paste0(pData, "cache/HIA_regions.rds"))
-```
-
-Create a new column by crossing region and HIF
-
-```{r}
-HIA_regions <- HIA_regions %>%
-  mutate(region_HIF = paste(region, infrastructureIndex))
+HIA_muni <- readRDS(paste0(pData, "cache/HIA_muni.rds"))
+```
+
+Create a new column by crossing municipality number and HIF
+
+```{r}
+HIA_muni <- HIA_muni %>%
+  mutate(muni_HIF = paste("ID", kommunenummer, infrastructureIndex, sep="_"))
 ```
@@
 
-I now have `r length(unique(HIA_regions$region_HIF))` unique areas (for each main ecosystem) that I will, given
+I now have `r length(unique(HIA_muni$muni_HIF))` unique areas (for each main ecosystem) that I will, given
 there is data, extrapolate indicator values over.
@@
 
-I will subset the `slitasje_data` into the three ecosystems. Note that
+I will subset the `alien_data` into the three ecosystems. Note that
 only for open wetland do we have good ecosystem delineation maps to base
@@
 ```{r}
-#Check spelling
-#unique(slitasje_data$hovedøkosystem)
-
-# Rename one level
-slitasje_data <- slitasje_data %>%
-  mutate(hovedøkosystem = recode(hovedøkosystem, "Våtmark" = "våtmark"))
-
 #make valid
-slitasje_data <- st_make_valid(slitasje_data)
+alien_data <- st_make_valid(alien_data)
 
 #subset
-wetlands <- slitasje_data[slitasje_data$hovedøkosystem == "våtmark",]
-seminat <- slitasje_data[slitasje_data$hovedøkosystem == "semi-naturligMark",]
-natOpen <- slitasje_data[slitasje_data$hovedøkosystem == "naturligÅpneOmråderILavlandet",]
+wetlands <- alien_data[alien_data$hovedøkosystem == "våtmark",]
+seminat <- alien_data[alien_data$hovedøkosystem == "semi-naturligMark",]
+natOpen <- alien_data[alien_data$hovedøkosystem == "naturligÅpneOmråderUnderSkoggrensa",]
 ```
@@
           indicator = i_2020,
-          regions = HIA_regions,
-          groups = region_HIF,
+          regions = HIA_muni,
+          groups = muni_HIF,
           summarise = TRUE)
@@
 all_stats <- all_stats %>%
-  separate(region_HIF,
-           into = c("region", "HIF"),
-           sep = " ")
-
-saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats.rds")
+  separate(muni_HIF,
+           into = c("tempLink", "municipalityNumber", "HIF"),
+           sep = "_")
+
+#saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
+saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
+
 ```
@@
 #silently read cached file
-all_stats <- readRDS(paste0(pData, "cache/all_stats.rds"))
-```
-
-```{r dense-stats, fig.height= 12, fig.width=6}
-temp_plot <- all_stats %>%
-  mutate(HIF_NA = case_when(
-    n > 150 ~ HIF
-  ))
-
-myColour_list <- unique(temp_plot$myColour)
-names(myColour_list) <-   myColour_list
-
-ggarrange(
-ggplot(temp_plot, aes(x = region, y = n, group = HIF, fill = HIF_NA))+
-  geom_bar(position = "dodge", stat = "identity", col="black")+
-  theme_bw(base_size = 20)+
-  scale_fill_brewer(palette="Oranges")+
-  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
-  coord_flip()+
-  facet_wrap(.~eco),
-ggplot(temp_plot, aes(x = region, y = w_mean, group = HIF, fill = HIF_NA))+
-  geom_bar(position = "dodge", stat = "identity", col="black")+
-  theme_bw(base_size = 20)+
-  scale_fill_brewer(palette="Oranges")+
-  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
-  coord_flip()+
-  facet_wrap(.~eco),
-ggplot(temp_plot, aes(x = region, y = sd, group = HIF, fill = HIF_NA))+
-  geom_bar(position = "dodge", stat = "identity", col="black")+
-  theme_bw(base_size = 20)+
-  scale_fill_brewer(palette="Oranges")+
-  theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
-  coord_flip()+
-  facet_wrap(.~eco),
-ncol = 1)
-
-```
-
-In the figure above, colors refer to Human Impact Factor (amount or frequency of infrastructure) 
-and NA means that there are less than 150 data points which we set as a threshold for evaluating the indicator values.
-
-In the top row
-we see that the sample size (number of polygons) varies between
-ecosystems. For naturally open ecosystems many polygons are from high
-infrastructure areas, not surprisingly, and there are few points for the
-lowest impact class (zero such polygons in *Sørlandet* for example). We
-may therefore not be able to extrapolate indicator values for Naturally
-open areas in in *Sørlandet* with low levels of infrastructure.
-
-In semi-natural areas and wetlands, we have very little data from HIF
-class 3.
-
-We can also from the middle row see the association between the indicator
-values (area weighted means) and the modified infrastructure index
-(HIF). We should then not put any weight on the bars were there is
-little data (transparent bars). Then we se the general pattern that the indicator 
-value decreases with increasing HIF. But there are exceptions, like for Semi-natural areas
-in Vestlandet.
-
-Let us look at the indicator to pressure relationship across all ecosystems.
+all_stats <- readRDS(paste0(pData, "cache/all_stats_alienSpecies.rds"))
+```
+
+We have data from `r length(unique(all_stats$municipalityNumber))` out of `r nrow(muni)` municipalities.
+This figure shows how many municipalities we have at least one data point for, specific for each human impact factor level:
+
+```{r, fig.cap="Number of municipalities we have at least one data point for, specific for each human impact factor level."}
+all_stats %>%
+  group_by(HIF, eco) %>%
+  summarise(n = n()) %>%
+  ggplot()+
+  geom_bar(aes(x = HIF, y = n), stat ="identity")+
+  theme_bw()+
+  ylab("Number of municipalities")+
+  facet_wrap(.~eco)
+```
+
+If we say that we need 20 data points for each HIA and municipality combintaion in order to extrapolate to the entire HIA for that municipality, we get much less to work with:
+
+```{r, fig.cap="Number of municipalities we have at least 20 data points for, specific for each human impact factor level."}
+all_stats %>%
+  filter(n>20) %>%
+  group_by(HIF, eco) %>%
+  summarise(n = n()) %>%
+  ggplot()+
+  geom_bar(aes(x = HIF, y = n), stat ="identity")+
+  theme_bw()+
+  ylab("Number of municipalities")+
+  facet_wrap(.~eco)
+
+```
 
 ```{r, fig.cap="Indicator to pressure relationship across all ecosystems."}
-temp_plot %>%
-  filter(!is.na(HIF_NA)) %>%
-  ggplot(aes(x = HIF_NA, y = w_mean))+
+all_stats %>%
+  filter(n > 20) %>%
+  ggplot(aes(x = HIF, y = w_mean))+
   geom_point(size=2, position = position_dodge2(.1))+
@@
 
-The pattern is not perfect, and there is quite some noise. But this is perhaps also expected.
+There is a trend here, but not very strong.
 
@@
 
-```{r, fig.cap="Smaple size against indicator uncertainty."}
-temp_plot %>%
+```{r, fig.cap="Sample size against indicator uncertainty."}
+all_stats %>%
+  filter(n > 5) %>%
   ggplot(aes(x = n, y = sd))+
@@
   theme_bw()+
-    facet_wrap(.~eco)
-```
-
-This shows that the uncertainty is inflated with sample sizes less than about 300-500.
-
-```{r, fig.cap="Summary statistics for the slitasje indicator."}
-DT::datatable(all_stats)
+    facet_wrap(.~eco, scales="free")
+```
+
+This shows that the uncertainty is really inflated with sample sizes less than about 50-100.
+
+```{r, fig.cap="Summary statistics for the alien species indicator."}
+DT::datatable(all_stats) %>%
+  formatRound(columns=3:8, digits=2)
 ```
@@
 ```{r, eval=F}
-corrCheck <- st_intersection(slitasje_data, HIA_regions)
+corrCheck <- st_intersection(alien_data, HIA_muni)
 saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")

### 2023-03-20 DRAFT_alien_species.Rmd
test over Trd

@@
 ```{r, eval=F}
-corrCheck <- st_intersection(alien_data, HIA_muni)
-saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck.rds")
+corrCheck <- st_intersection(alien_data, HIA)
+#saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
+saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
 ```
@@
 #Silently read cached data
-corrCheck <- readRDS(paste0(pData, "cache/corrCheck.rds"))
-```
-
-```{r slitasje-precentageplot, fig.height=12, fig.width=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+corrCheck <- readRDS(paste0(pData, "cache/corrCheck_alienSpecies.rds"))
+```
+
+```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
 ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
@@
   theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Sliatsje indicator"))+
+  guides(fill = guide_legend("Alien species indicator"))+
   ylab("Fraction of data points")+
@@
   scale_fill_brewer(palette = "RdYlGn")+
-  facet_grid(hovedøkosystem~region)
+  facet_wrap(.~hovedøkosystem)
 ```
@@
 The figure above I think supports the indicator-pressure relationship
-and justifies using the HIA+region intersections as local reference
-areas. Especially when down-weighting the importance of those
-groups/categories that have very little data.
+and justifies using the HIA+municipality intersections as local reference
+areas.
 
@@
                 ymin = 7032142.5, ymax = 7045245.27),
-                crs = st_crs(HIA_regions))
+                crs = st_crs(HIA_muni))
 ```
@@
 ```{r}
-HIA_trd <- sf::st_crop(HIA_regions, myBB)
+HIA_trd <- sf::st_crop(HIA_muni, myBB)
 ```
@@
 
-Let's calculate the areas of these polygons and compare the HIF in the
-five regions.
-
-```{r}
-HIA_regions$area <- sf::st_area(HIA_regions)
-```
-
-```{r HIF-region, fig.cap="Stacked barplot showing the distribution of human impact factor across five regions in Norway."}
-temp <- as.data.frame(HIA_regions) %>%
-  group_by(region, infrastructureIndex) %>%
-  summarise(area = mean(area))
-
-ggplot(temp, aes(x = region, y = area, fill = factor(infrastructureIndex)))+
-  geom_bar(position = "stack", stat = "identity")+
-  guides(fill = guide_legend("HIF"))
-```
-
-The figure above shows that _Nord-Norge_ for example, has a lot of
-relatively untouched areas, and that _Østlandet_ has the highest
-proportion of impacted areas. This is expected.
-
-### Aggregate and spread (extrapolate) {#spread-slitasje}
+See Fig. \@ref(fig.HIF-region) for a closer look at the distribution of HIA classes across Norway.
+
+### Aggregate and spread (extrapolate) {#spread-alien}
 
@@
 aggregate) and to spread these out spatially to the entire HIA within
-each region (i.e. to extrapolate).
+each municipality (i.e. to extrapolate).
+
+<!-- FORTSETT HER -->
 

### 2023-03-21 DRAFT_alien_species.Rmd
ea_spread

@@
 
-Let us look at the effect of sample size on the indicator uncertainty.
-
-```{r, fig.cap="Sample size against indicator uncertainty."}
-all_stats %>%
-  filter(n > 5) %>%
-  ggplot(aes(x = n, y = sd))+
-  geom_point(size=2, position = position_dodge2(.1))+
-  theme_bw()+
-    facet_wrap(.~eco, scales="free")
-```
-
-This shows that the uncertainty is really inflated with sample sizes less than about 50-100.
-
-```{r, fig.cap="Summary statistics for the alien species indicator."}
-DT::datatable(all_stats) %>%
-  formatRound(columns=3:8, digits=2)
-```
-
-In addition to looking at the area weighted means and its relationship
-with the pressure indicator (HIF), we can do the same at the polygon
-level. Because the indicator is pseudo-continuous, boxplots for example
-become useless, so I will present a relative frequency plot.
+Here is a similar figure, looking at the relative frequency of indicator level of polygons and within each HIA.
 
@@
 areas.
+
+Let us look at the effect of sample size on the indicator uncertainty.
+
+```{r, fig.cap="Sample size against indicator uncertainty."}
+all_stats %>%
+  filter(n > 5) %>%
+  ggplot(aes(x = n, y = sd))+
+  geom_point(size=2, position = position_dodge2(.1))+
+  theme_bw()+
+    facet_wrap(.~eco, scales="free")
+```
+
+This shows that the uncertainty is really inflated with sample sizes less than about 50-100.
+
+```{r, fig.cap="Summary statistics for the alien species indicator."}
+DT::datatable(all_stats) %>%
+  formatRound(columns=3:8, digits=2)
+```
 
@@
 
-<!-- FORTSETT HER -->
-
 I want to add a threshold so that we don't end up over extrapolating
-based on too few data points. I have previousl used 150 data pionts, and I think that was about right.
+based on too few data points. I will use 20 data points as a minimum.
 
 ```{r, eval=F}
-wetlands_slitasje_extr <- ea_spread(indicator_data = wetlands,
+wetland_alien_extr <- ea_spread(indicator_data = wetlands,
                          indicator = i_2020,
-                         regions = HIA_regions,
-                         groups = region_HIF,
-                         threshold = 150)
-
-seminat_slitasje_extr <- ea_spread(indicator_data = seminat,
+                         regions = HIA_muni,
+                         groups = muni_HIF,
+                         threshold = 20)
+
+seminat_alien_extr <- ea_spread(indicator_data = seminat,
                          indicator = i_2020,
-                         regions = HIA_regions,
-                         groups = region_HIF,
-                         threshold = 150)
-
-natOpen_slitasje_extr <- ea_spread(indicator_data = natOpen,
+                         regions = HIA_muni,
+                         groups = muni_HIF,
+                         threshold = 20)
+
+natOpen_alien_extr <- ea_spread(indicator_data = natOpen,
                          indicator = i_2020,
-                         regions = HIA_regions,
-                         groups = region_HIF,
-                         threshold = 150)
-#saveRDS(wetlands_slitasje_extr, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_slitasje_extr.rds")
-#saveRDS(seminat_slitasje_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_slitasje_extr.rds")
-#saveRDS(natOpen_slitasje_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_slitasje_extr.rds")
+                         regions = HIA_muni,
+                         groups = muni_HIF,
+                         threshold = 20)
+#saveRDS(wetlands_alien_extr, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_alien_extr.rds")
+#saveRDS( seminat_alien_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_alien_extr.rds")
+#saveRDS( natOpen_alien_extr,  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_alien_extr.rds")
+
+#saveRDS(wetlands_alien_extr,  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_alien_extr.rds")
+#saveRDS( seminat_alien_extr,  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/seminat_alien_extr.rds")
+#saveRDS( natOpen_alien_extr,  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_alien_extr.rds")
+
 ```
@@
 ```{r, include=F}
-wetlands_slitasje_extr <- readRDS(paste0(pData, "cache/wetlands_slitasje_extr.rds"))
-seminat_slitasje_extr  <- readRDS(paste0(pData, "cache/seminat_slitasje_extr.rds"))
-natOpen_slitasje_extr  <- readRDS(paste0(pData, "cache/natOpen_slitasje_extr.rds"))
-```
-
-It's easier to see what's happening if we zoom in a bit. Lets get a
-boundary box around Trondheim.
-
-```{r}
-myBB <- st_bbox(c(xmin=260520.12, xmax = 278587.56,
-                ymin = 7032142.5, ymax = 7045245.27),
-                crs = st_crs(wetlands_slitasje_extr))
-```
-
-Cropping the data to the bbox
-
-```{r, warning=F}
-wetlands_slitasje_extr_trd <- sf::st_crop(wetlands_slitasje_extr, myBB)
-seminat_slitasje_extr_trd <- sf::st_crop(seminat_slitasje_extr, myBB)
-natOpen_slitasje_extr_trd <- sf::st_crop(natOpen_slitasje_extr, myBB)
-```
-
-```{r extrapolated-slitasje, fig.cap="Slitasje indicator extrapolated over Trondheim"}
+wetland_alien_extr <- readRDS(paste0(pData, "cache/wetlands_alien_extr.rds"))
+seminat_alien_extr  <- readRDS(paste0(pData,  "cache/seminat_alien_extr.rds"))
+natOpen_alien_extr  <- readRDS(paste0(pData,  "cache/natOpen_alien_extr.rds"))
+```
+
+It's easier to see what's happening if we zoom in a bit.
+
+Trondheim is municipality nr 5001
+
+3020 is Nordre Follo
+
+5418 is Målselv
+
+```{r}
+wetland_alien_extr <- wetland_alien_extr %>%
+  separate(ID,
+           into = c("ID", "municipalityNumber", "HIF"),
+           sep = "_") %>%
+  add_column(eco = "wetlands")
+
+seminat_alien_extr <- seminat_alien_extr %>%
+  separate(ID,
+           into = c("ID", "municipalityNumber", "HIF"),
+           sep = "_") %>%
+  add_column(eco = "semi_natural")
+
+natOpen_alien_extr <- natOpen_alien_extr %>%
+  separate(ID,
+           into = c("ID", "municipalityNumber", "HIF"),
+           sep = "_") %>%
+  add_column(eco = "naturally_open")
+
+alien_extr <- rbind(
+  wetland_alien_extr,
+  seminat_alien_extr,
+  natOpen_alien_extr
+)
+```
+
+```{r}
+alien_extr_trd_nf <- alien_extr %>%
+  filter(municipalityNumber %in% c("5001", "3020", "5418"))
+```
+
+```{r extrapolated-alien, fig.cap="Alein species indicator extrapolated over Nordre Follo (3020), Trondheim (5001) and Målselv (5418).
+Note that the maps are not masked by ecosystem type, so the three ecosystem specific indicators are overlaping.
+The bottom row shows the location of the four different human impact sones."}
 myCol <- "RdYlGn"
+
 tmap_arrange(
-tm_shape(wetlands_slitasje_extr_trd)+
+tm_shape(alien_extr_trd_nf)+
   tm_polygons(col = "w_mean",
-    title="Slitasje indicator",
+    title="Alien species indicator",
     palette = myCol,
@@
   tm_layout(legend.outside = T)+
-  tm_shape(hw_utm)+
-  tm_lines(col="blue")+
-  tm_layout(title = "Wetland"),
-
-tm_shape(seminat_slitasje_extr_trd)+
-  tm_polygons(col = "w_mean",
-    title="Slitasje indicator",
-    palette = myCol,
+  tm_facets(by=c("eco", "municipalityNumber"))
+,
+tm_shape(alien_extr_trd_nf)+
+  tm_polygons(col = "HIF",
+    title="Human impact factor",
+    palette = "-viridis",
     style="cont",
@@
   tm_layout(legend.outside = T)+
-  tm_shape(hw_utm)+
-  tm_lines(col="blue")+
-  tm_layout(title = "Semi-natural areas"),
-
-tm_shape(natOpen_slitasje_extr_trd)+
-  tm_polygons(col = "w_mean",
-    title="Slitasje indicator",
-    palette = myCol,
-    style="cont",
-    breaks = c(0,1))+
-  tm_layout(legend.outside = T)+
-  tm_shape(hw_utm)+
-  tm_lines(col="blue")+
-  tm_layout(title = "Naturally open areas"),
-HIA_trd)
-```
-
-Note that for wetland and semi-natural areas we did not
-extrapolate values for the most urban areas, and that for naturally open
-areas we did not extrapolate to the least affected areas.
-For the other regions we generally excluded more HIFs.
-
-> *The maps above are easy to misunderstand, and they should therefor be
-> communicated with caution. They should be considered as temporary
-> files or as by-products.*
+  tm_facets(by=c("municipalityNumber"))
+,
+ncol=1,
+heights = c(1,0.3))
+```
+
+Figure \@ref(fig:extrapolated-alien) shows that this indicator will have a lot of missing data.
+For Nordre Follo and Trondheim for example, we do not have any indicator values tied to naturally open or wetland ecosystems, and for semi natural ecosystems we only have data for HIF class 2.
+This is, however, still a lot better than discarding the entire data set due to issues related to area representativity.
+The amount of missing data in the indicator maps is closely ties to the chosen threshold value for how many data points we must have in order to calculate an (area weighted) men indicator value.
+We have used 20 as the threshold here.
+For Målselv (far right column), we ended up having enough data for wetlands in HIF class 0, and since most of the municipality is in this HIF class we get a lot of indicator coverage.
 

### 2023-03-21 DRAFT_alien_species.Rmd
use tidyverse

@@
 library(knitr)
+library(tidyverse)
 library(sf)
@@
 library(eaTools)
-#library(tidyverse)
 library(tibble)
-library(ggplot2)
-library(tidyr)
-library(dplyr)
+#library(ggplot2)
+#library(tidyr)
+#library(dplyr)
 knitr::opts_chunk$set(echo = TRUE)
@@
     palette = myCol,
-    style="cont",
-    breaks = c(0,1))+
+     style="fixed",
+    breaks = seq(0,1,.1))+
   tm_layout(legend.outside = T)+
@@
 
+The following table contains the data values underlying Figure \@ref(fig:extrapolated-alien), including the combination with <20 data points.
+
+```{r, fig.cap="Summary statistics for three example municipalities."}
+all_stats %>%
+  filter(municipalityNumber %in% c("5001", "3020", "5418")) %>%
+  arrange(municipalityNumber, eco, HIF) %>%
+  DT::datatable()%>%
+  formatRound(columns=3:8, digits=2)
+```
+
 The next step now is to
@@
 
-A faster way could be to:
-
-1.  Draw *n* random point samples from the ET map, and
-
-2.  extract indicator values at those points (from the extrapolated
-    indicator map (a stars object)).
-
-... but his will not provide spatially explicit indicator values for the
-entire ecosystem delineation.
-
-### Ecosystem map
-
-Importing the wetland map.
-
-```{r, eval=T}
-#takes about one minute without the proxy
-(mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
-                             proxy=T))
-```
-
-This is in the wrong CRS. We want to stay with 25833.
-
-We have a 50x50 m [master grid](#master-grid) with the correct CRS, and I would warp mireET to fit that grid like this:
-```{r, eval=F}
-grid <- stars::read_stars(paste0(pData, "masterGrid_50m.tiff"))
-mireET_warped <- stars::st_warp(mireET, grid)
-```
-
-... but I don't want to down sample the mire data, since we will loose a lot of the smaller ecosystem occurrences.
-
-I could make a 10x10 master grid, but I have not found a way to stop this from being just a massive file.
-
-I can transform the ET map like this, and not worry about the rasters being perfectly aligned, but even this operation is too heavy for the R Studio Server:
-
-```{r, eval=F}
-mireET_trans <- sf::st_transform(mireET, 25833)
-```
-
-I will therefore ask GIS engineers to prepare ecosystem maps to the standard that we chose, so that we can simply combine the stars object with `stars::c()`, or perhaps side-step that operatio and just mask directly like this: `wetlands_slitasje_extr[mireET==0] <- NA`.
-
-### Aggregate to region
-
-We need to aggreate the ET occurence and their indicator values, to a
-regional delineation. This functionality can be developed in `eaTools`,
-but will in any case build around `exactextratr::exact_extract()`.
-
-To carry the errors, we can for example
-produce distributions for the indicator values for each ET
-occurrence (or groups of them) and resample the distributions with an
-area weighting.
-
-### Uncertainty
-
-The calculations of indicator uncertainty is done with bootstrapping, as
-is described in eaTools.
-
-## Eksport files (final product)
-
-According to the workflow specifications, I will export the indicator
-map at its highest resolution, for use in various web interfaces etc.
-For the finest resolution indicator map I will not include columns for
-uncertainty (because this will only be estimated at the aggregated
-level), or reference levels (because there is no spatial variation).
-
-I can also export a map of the indicator where we extrapolate the
-indicator values to the homogeneous areas (e.g. Fig. \@ref(fig:extrapolated-slitasje) . This map should be
-interpreted with caution, since they appear as providing explicit indicator
-values for areas where there is no data. This map is mainly for
-aggregation purposes. I will mark this map with the prefix
-*homogeneous-areas*. I will also export this same map after masking it
-with the ET map, and this output must come with the same disclaimers.
-
-Finally I will export a map of the five accounting areas with
-wall-to-wall indicator value and errors (e.g. Fig \@ref(fig:wall-to-wall-summer-temp-indicator)). This map can also contain data
-covering other ETs. Otherwise the maps would not we visually
-interpretative.
-
-### Slitasje in wetlands
-
-This is how I plan to export the final product.
-
-```{r, eval=F}
-# keep col names short and standardized
-exp_wetland <- ...
-```
-
-```{r, eval=F}
-exp_wetland_HIAextrapolated <- …
-```
-
-```{r, eval=F}
-exp_wetland_AAextrapolated <- …
-```
-
-Write to file
-
-```{r, eval=F}
-st_write(exp_wetland, "indicators/indicatorMaps/slitasje/wetland_slitasje.shp",
-         append = F)
-
-st_write(exp_wetlandHIAextrapolated, 
-         "indicators/indicatorMaps/extrapolated_to_homogeneous_areas/slitasje/homogeneous_areas_wetland_slitasje.shp",
-         append = F)
-
-st_write(exp_wetland, "indicators/indicatorMaps/extrapolated_to_accounting_areas/slitasje/accounting_areas_wetland_slitasje.shp",
-         append = F)
-```
+This workflow should be syncronized with the [slitasje](#slitasje-og-kjørespor) indicator.

### 2023-03-21 DRAFT_alien_species.Rmd
fix typos and fig captions

@@
 
-```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=6}
+```{r, fig.cap="An overview of the naturetypes for which we will calculate the indicator.", fig.height=8, fig.width=6}
 ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
@@
 
+Note that there is a lot more data for semi-natural ecosystem compared to the other two.
+
 Column names starting with a number is problematic, so adding a prefix
@@
 
-```{r, fig.cap="Distribution of the ANO variable andel_hovedoekosystem_punkt for wetland localities."}
+```{r, fig.cap="Distribution of the ANO variable andel_hovedoekosystem_punkt."}
 par(mar=c(5,6,4,2))
@@
 
-```{r, warning=F, fig.cap="Percentage cover of the Våtmark main ecosystem in the 250m2 circle"}
+```{r, warning=F, fig.cap="Percentage cover of the main ecosystem in the 250m2 circle"}
 ggplot(ano, aes(x = andel_hovedoekosystem_punkt))+
@@
   theme_bw(base_size = 12)+
-  xlab("Percentage cover of the Våtmark main ecosystem\n in the 250m2 circle")+
+  xlab("Percentage cover of the main ecosystem\n in the 250m2 circle")+
   scale_x_continuous(limits = c(0,101),
@@
 
-```{r}
+```{r, warning=F}
 path1 <- ifelse(dir == "C:", 
@@
 
-```{r kommuner, fig.cap="Municipalities in Norway."}
+```{r kommuner, fig.cap="Municipalities in Norway.", warning=F}
 muni <- st_cast(muni, "MULTIPOLYGON")
@@
 
-#### Scaled variable {#scaled-slitasje-variable}
+#### Scaled variable {#scaled-alien-variable}
 
@@
 
-```{r alien-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=4}
+```{r alien-normalise, fig.cap="Performing a linear break-point type normalisation of the alien species variable.", fig.width=4}
 eaTools::ea_normalise(data = alien_data,
@@
 
-If we say that we need 20 data points for each HIA and municipality combintaion in order to extrapolate to the entire HIA for that municipality, we get much less to work with:
+If we say that we need 20 data points for each HIA and municipality combination in order to extrapolate to the entire HIA for that municipality, we get much less to work with:
 
@@
 
-```{r, fig.cap="Indicator to pressure relationship across all ecosystems."}
+```{r, fig.cap="Indicator-pressure relationship across all ecosystems."}
 all_stats %>%
@@
   geom_violin(alpha=0)+
-  theme_bw()
-```
-
-There is a trend here, but not very strong.
+  theme_bw()+
+  labs(x = "Human impact factor",
+       y = "Indicator value (area weighted means)")
+```
+
+There is a trend here, but not very strong perhaps.
 
@@
 The figure above I think supports the indicator-pressure relationship
-and justifies using the HIA+municipality intersections as local reference
+and justifies using the HIA x municipality intersections as local reference
 areas.
@@
                          threshold = 20)
+```
+
+```{r, eval=F, include=F}
 #saveRDS(wetlands_alien_extr, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/wetlands_alien_extr.rds")
@@
 #saveRDS( natOpen_alien_extr,  "P:/41201785_okologisk_tilstand_2022_2023/data/cache/natOpen_alien_extr.rds")
-
 ```
@@
 
-It's easier to see what's happening if we zoom in a bit.
-
-Trondheim is municipality nr 5001
-
-3020 is Nordre Follo
-
-5418 is Målselv
+It's easier to see what's happening if we zoom in a bit. I will look more closely at three example municipalities: Trondheim (municipality nr 5001), Nordre Follo (3020) and Målselv (5418).
 
@@
 
-The next step now is to
+Note that we can also show the uncertainties as maps
+
+```{r sd-alien, fig.cap="Map showing the uncertainties around the indicator values for alien species in three example municipalities for Semi-natural ecosystems."}
+tm_shape(alien_extr_trd_nf[alien_extr_trd_nf$eco=="semi_natural",])+
+  tm_polygons(col = "sd",
+    title="SD(alien\nspecies indicator)",
+    palette = "-viridis"
+     )+
+  tm_layout(legend.outside = T)+
+  tm_facets(by="municipalityNumber")
+```
+
+## Next steps
+
+The next steps now are to
 
@@
 
-This workflow should be syncronized with the [slitasje](#slitasje-og-kjørespor) indicator.
+This workflow should be synchronized with the [slitasje](#slitasje-og-kjørespor) indicator.

### 2023-03-22 DRAFT_alien_species.Rmd
analysi of data area coverage

@@
 
-#### Validate
-
-I now have `r length(unique(HIA_muni$muni_HIF))` unique areas (for each main ecosystem) that I will, given
-there is data, extrapolate indicator values over.
-
-###### Subset ETs
-
-I will subset the `alien_data` into the three ecosystems. Note that
-only for open wetland do we have good ecosystem delineation maps to base
-a spatial averaging on.
-
-```{r}
-#make valid
-alien_data <- st_make_valid(alien_data)
-
-#subset
-wetlands <- alien_data[alien_data$hovedøkosystem == "våtmark",]
-seminat <- alien_data[alien_data$hovedøkosystem == "semi-naturligMark",]
-natOpen <- alien_data[alien_data$hovedøkosystem == "naturligÅpneOmråderUnderSkoggrensa",]
-```
-
-Creating some summary statistics.
-
-```{r, eval=F}
-wetland_stats <- ea_spread(indicator_data = wetlands,
-          indicator = i_2020,
-          regions = HIA_muni,
-          groups = muni_HIF,
-          summarise = TRUE)
-
-seminat_stats <- ea_spread(indicator_data = seminat,
-          indicator = i_2020,
-          regions = HIA_muni,
-          groups = muni_HIF,
-          summarise = TRUE)
-
-natOpen_stats <- ea_spread(indicator_data = natOpen,
-          indicator = i_2020,
-          regions = HIA_muni,
-          groups = muni_HIF,
-          summarise = TRUE)
-
-wetland_stats <- wetland_stats %>%
-  add_column(eco = "wetland")
-
-seminat_stats <- seminat_stats %>%
-  add_column(eco = "semi-natural")
-
-natOpen_stats <- natOpen_stats %>%
-  add_column(eco = "Naturally-open")
-
-all_stats <- rbind(wetland_stats,
-                   seminat_stats,
-                   natOpen_stats)
-all_stats <- all_stats %>%
-  separate(muni_HIF,
-           into = c("tempLink", "municipalityNumber", "HIF"),
-           sep = "_")
-
-#saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
-saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
-
-```
-
-```{r, include=F}
-#silently read cached file
-all_stats <- readRDS(paste0(pData, "cache/all_stats_alienSpecies.rds"))
-```
-
-We have data from `r length(unique(all_stats$municipalityNumber))` out of `r nrow(muni)` municipalities.
-This figure shows how many municipalities we have at least one data point for, specific for each human impact factor level:
-
-```{r, fig.cap="Number of municipalities we have at least one data point for, specific for each human impact factor level."}
-all_stats %>%
-  group_by(HIF, eco) %>%
-  summarise(n = n()) %>%
-  ggplot()+
-  geom_bar(aes(x = HIF, y = n), stat ="identity")+
-  theme_bw()+
-  ylab("Number of municipalities")+
-  facet_wrap(.~eco)
-```
-
-If we say that we need 20 data points for each HIA and municipality combination in order to extrapolate to the entire HIA for that municipality, we get much less to work with:
-
-```{r, fig.cap="Number of municipalities we have at least 20 data points for, specific for each human impact factor level."}
-all_stats %>%
-  filter(n>20) %>%
-  group_by(HIF, eco) %>%
-  summarise(n = n()) %>%
-  ggplot()+
-  geom_bar(aes(x = HIF, y = n), stat ="identity")+
-  theme_bw()+
-  ylab("Number of municipalities")+
-  facet_wrap(.~eco)
-
-```
-
-```{r, fig.cap="Indicator-pressure relationship across all ecosystems."}
-all_stats %>%
-  filter(n > 20) %>%
-  ggplot(aes(x = HIF, y = w_mean))+
-  geom_point(size=2, position = position_dodge2(.1))+
-  geom_violin(alpha=0)+
-  theme_bw()+
-  labs(x = "Human impact factor",
-       y = "Indicator value (area weighted means)")
-```
-
-There is a trend here, but not very strong perhaps.
-
-Here is a similar figure, looking at the relative frequency of indicator level of polygons and within each HIA.
-
-```{r, eval=F}
-corrCheck <- st_intersection(alien_data, HIA)
-#saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
-saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
-```
-
-```{r, include=F}
-#Silently read cached data
-corrCheck <- readRDS(paste0(pData, "cache/corrCheck_alienSpecies.rds"))
-```
-
-```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
-ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
-  geom_bar(position="fill")+
-  theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Alien species indicator"))+
-  ylab("Fraction of data points")+
-  xlab("HIF")+
-  scale_fill_brewer(palette = "RdYlGn")+
-  facet_wrap(.~hovedøkosystem)
-```
-
-The figure above I think supports the indicator-pressure relationship
-and justifies using the HIA x municipality intersections as local reference
-areas.
-
-Let us look at the effect of sample size on the indicator uncertainty.
-
-```{r, fig.cap="Sample size against indicator uncertainty."}
-all_stats %>%
-  filter(n > 5) %>%
-  ggplot(aes(x = n, y = sd))+
-  geom_point(size=2, position = position_dodge2(.1))+
-  theme_bw()+
-    facet_wrap(.~eco, scales="free")
-```
-
-This shows that the uncertainty is really inflated with sample sizes less than about 50-100.
-
-```{r, fig.cap="Summary statistics for the alien species indicator."}
-DT::datatable(all_stats) %>%
-  formatRound(columns=3:8, digits=2)
-```
-
-#### Looking at the HIA
-
 Here is a view of the data zooming in on Trondheim
@@
 
-See Fig. \@ref(fig.HIF-region) for a closer look at the distribution of HIA classes across Norway.
+See Fig. \@ref(fig:HIF-region) for a closer look at the distribution of HIA classes across Norway.
+
+#### Validate
+
+I now have `r length(unique(HIA_muni$muni_HIF))` unique areas (for each main ecosystem) that I will, given
+there is data, extrapolate indicator values over. But first I want to validate the assumption that the HIF 
+explains a considerable portion of the variation i the indicator values.
+
+I will subset the `alien_data` into the three ecosystems.
+
+```{r}
+#make geometries valid 
+alien_data <- st_make_valid(alien_data)
+
+#subset
+wetlands <- alien_data[alien_data$hovedøkosystem == "våtmark",]
+seminat <- alien_data[alien_data$hovedøkosystem == "semi-naturligMark",]
+natOpen <- alien_data[alien_data$hovedøkosystem == "naturligÅpneOmråderUnderSkoggrensa",]
+```
+
+Creating some summary statistics.
+
+```{r, eval=F}
+wetland_stats <- ea_spread(indicator_data = wetlands,
+          indicator = i_2020,
+          regions = HIA_muni,
+          groups = muni_HIF,
+          summarise = TRUE)
+
+seminat_stats <- ea_spread(indicator_data = seminat,
+          indicator = i_2020,
+          regions = HIA_muni,
+          groups = muni_HIF,
+          summarise = TRUE)
+
+natOpen_stats <- ea_spread(indicator_data = natOpen,
+          indicator = i_2020,
+          regions = HIA_muni,
+          groups = muni_HIF,
+          summarise = TRUE)
+
+wetland_stats <- wetland_stats %>%
+  add_column(eco = "wetland")
+
+seminat_stats <- seminat_stats %>%
+  add_column(eco = "semi-natural")
+
+natOpen_stats <- natOpen_stats %>%
+  add_column(eco = "Naturally-open")
+
+all_stats <- rbind(wetland_stats,
+                   seminat_stats,
+                   natOpen_stats)
+all_stats <- all_stats %>%
+  separate(muni_HIF,
+           into = c("tempLink", "municipalityNumber", "HIF"),
+           sep = "_")
+
+#saveRDS(all_stats, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
+saveRDS(all_stats, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/all_stats_alienSpecies.rds")
+
+```
+
+```{r, include=F}
+#silently read cached file
+all_stats <- readRDS(paste0(pData, "cache/all_stats_alienSpecies.rds"))
+```
+
+We have data from `r length(unique(all_stats$municipalityNumber))` out of `r nrow(muni)` municipalities.
+This figure shows how many municipalities we have at least one data point for, specific for each human impact factor level:
+
+```{r, fig.cap="Number of municipalities we have at least one data point for, specific for each human impact factor level."}
+all_stats %>%
+  group_by(HIF, eco) %>%
+  summarise(n = n()) %>%
+  ggplot()+
+  geom_bar(aes(x = HIF, y = n), stat ="identity")+
+  theme_bw()+
+  ylab("Number of municipalities")+
+  facet_wrap(.~eco)
+```
+
+If we say that we need 20 data points for each HIA and municipality combination in order to extrapolate to the entire HIA for that municipality, we get much less to work with:
+
+```{r, fig.cap="Number of municipalities we have at least 20 data points for, specific for each human impact factor level."}
+all_stats %>%
+  filter(n>20) %>%
+  group_by(HIF, eco) %>%
+  summarise(n = n()) %>%
+  ggplot()+
+  geom_bar(aes(x = HIF, y = n), stat ="identity")+
+  theme_bw()+
+  ylab("Number of municipalities")+
+  facet_wrap(.~eco)
+
+```
+
+```{r, fig.cap="Indicator-pressure relationship across all ecosystems."}
+all_stats %>%
+  filter(n > 20) %>%
+  ggplot(aes(x = HIF, y = w_mean))+
+  geom_point(size=2, position = position_dodge2(.1))+
+  geom_violin(alpha=0)+
+  theme_bw()+
+  labs(x = "Human impact factor",
+       y = "Indicator value (area weighted means)")
+```
+
+There is a trend here, but not very strong perhaps.
+
+Here is a similar figure, looking at the relative frequency of indicator level of polygons and within each HIA.
+
+```{r, eval=F}
+corrCheck <- st_intersection(alien_data, HIA)
+#saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
+saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
+```
+
+```{r, include=F}
+#Silently read cached data
+corrCheck <- readRDS(paste0(pData, "cache/corrCheck_alienSpecies.rds"))
+```
+
+```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
+  geom_bar(position="fill")+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Alien species indicator"))+
+  ylab("Fraction of data points")+
+  xlab("HIF")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_wrap(.~hovedøkosystem)
+```
+
+The figure above I think supports the indicator-pressure relationship
+and justifies using the HIA x municipality intersections as local reference
+areas.
+
+Let us look at the effect of sample size on the indicator uncertainty.
+
+```{r, fig.cap="Sample size against indicator uncertainty."}
+all_stats %>%
+  filter(n > 5) %>%
+  ggplot(aes(x = n, y = sd))+
+  geom_point(size=2, position = position_dodge2(.1))+
+  theme_bw()+
+    facet_wrap(.~eco, scales="free")
+```
+
+This shows that the uncertainty is really inflated with sample sizes less than about 50-100.
+
+```{r, fig.cap="Summary statistics for the alien species indicator."}
+DT::datatable(all_stats) %>%
+  formatRound(columns=3:8, digits=2)
+```
 
@@
 
+### Exploring the indicator values
+
+Most HIAs have good condition:
+
+```{r, fig.cap="Number of HIAs with >20 data points (sum of all three ecosystems)"}
+alien_extr %>%
+  filter(!is.na(w_mean)) %>%
+  ggplot()+
+  geom_histogram(aes(x = w_mean))+
+  theme_bw()+
+  labs(x = "Area weighted indicator value",
+       y = "Number of HIAs with >20 data points")
+```
+
+#### Data coverage
+
+Let's see how big proportion of Norway we ended up having data for. Ideally I would do this after masking the indicator maps with ecosystem delineations, 
+but I can also get a good feel for the data coverage like this as well.
+
+The area of Norway is 
+```{r}
+(total_area <- st_area(outline))
+```
+
+And here is the total and relative area of the indicator maps
+```{r relative-area, fig.cap="Relative area cover of the indicator maps for the alien species indicator (i.e.
+value 1 means that all HIAs had > 20 data points, and 0.2 means 20% of them did).
+If assuming the ecosystems are equally common where they are mapped and where hey are not, this measure of data coverage is representative of the real indicator coverage after maksing the maps with ecosystem delineations."}
+alien_extr %>%
+  filter(!is.na(w_mean)) %>%
+  mutate(m2 = st_area(.)) %>%
+  group_by(eco) %>%
+  summarise(sum_area = sum(m2)) %>%
+  add_column(total_area = total_area) %>%
+  mutate(relative_area = sum_area/total_area) %>%
+  ggplot()+
+  geom_bar(aes(x = eco, y = relative_area),
+           stat = "identity")+
+  theme_bw()
+```
+
+Figure \@ref(fig:relative-area) shows us that this indicator will provide indicator values for about 20% of the semi-natural ecosystem, and considerably less for the other two ecosystems.
+The data coverage is very sensitive to the threshold for minimum data points, set to >20 in this case.
+Also, with more nature type data accumulating over time, the data coverage will increase.
+Therefore this indicator is worth while, and is seen as considerably better than the alternative to discard the entire data set.
+
 ## Next steps
@@
 
+4. Then, look at [Median Summer Temperature] for how to aggregate spatially tp accounting areas (regions)
+
 This workflow should be synchronized with the [slitasje](#slitasje-og-kjørespor) indicator.

### 2023-03-22 DRAFT_alien_species.Rmd
latitudinal effect

@@
 
+#### Effect of latitude
+
+Alien species tend to invade from the south. Let us see if there is an effect of latitude.
+
+Extracting x and y coordinates
+
+```{r}
+temp <- alien_extr %>%
+  filter(!is.na(w_mean)) %>%
+  st_centroid() %>%
+  st_coordinates()
+
+temp2 <- alien_extr %>%
+  filter(!is.na(w_mean))  %>%
+  cbind(temp)
+```
+
+```{r lat-effect, fig.cap="Effect of latitude on the alien species indicator values. The blue line is a loess smoother (span=0.75)."}
+temp2 %>%
+  ggplot(aes(x = X, y = w_mean))+
+  geom_point()+
+  theme_bw()+
+  geom_smooth(linewidth=2,
+              method="loess",
+              span=0.75)+
+  labs(x = "Latitude (UTM with offset)",
+       y = "Alien species indicator value\n(area weighted)")+
+  facet_wrap(~eco)
+
+```
+
+```{r, eval=F, include=F}
+# Here is the same analysis, but at the polygon scale (it's a bit more messy)
+temp <- alien_data %>%
+  st_centroid() %>%
+  st_coordinates()
+
+temp2 <- alien_data %>%
+  cbind(temp)
+
+temp2 %>%
+  ggplot(aes(x = X, y = i_2020))+
+  geom_point()+
+  theme_bw()+
+  geom_smooth(linewidth=2,
+              method="loess",
+              span=0.75)+
+  labs(x = "Latitude (EPSG:25833)",
+       y = "Alien species indicator value")+
+  facet_wrap(~hovedøkosystem)
+
+```
+
+There is a quite clear association here between latitude and the on-site effect of alien species.
+
 ## Next steps

### 2023-03-22 DRAFT_alien_species.Rmd
masking example

@@
 
-4. Then, look at [Median Summer Temperature] for how to aggregate spatially tp accounting areas (regions)
-
-This workflow should be synchronized with the [slitasje](#slitasje-og-kjørespor) indicator.
+4. Then, look at [Median Summer Temperature] for how to aggregate spatially to accounting areas (regions)
+
+This workflow should be synchronized with the [slitasje](#slitasje-og-kjørespor) indicator. I will attempt a little proof of concept below.
+
+## Masking with an ecosystem delineation - example
+
+Let's try steps 2 and 3 on a reduced data set. I'll work with wetlands in Gran municipality.
+
+```{r}
+gran <- wetland_alien_extr %>%
+  filter(municipalityNumber == "3446")
+```
+
+```{r, fig.cap="Alien species indicator over wetlands in Gran municipality."}
+tm_shape(gran)+
+  tm_polygons(col = "w_mean",
+    title="Alien species indicator - wetlands",
+    palette = myCol,
+     style="fixed",
+    breaks = seq(0,1,.1))+
+  tm_layout(legend.outside = T,
+            title = "Gran municipality")
+```
+
+Then I can import and crop the ecosystem delineation map.
+
+```{r, eval=F}
+file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
+EDM <- stars::read_stars(file, proxy=F)
+```
+
+The EDM is in UTM32 when we actually want to have it in UTM33. But transforming it takes too long, so I will transform the indicator map instead.
+
+```{r, eval=FALSE}
+gran <- st_transform(gran, st_crs(EDM))
+```
+
+```{r, eval=F}
+mire_gran <- st_crop(EDM, gran)
+saveRDS(mire_gran, paste0(pData, "cache/mire_gran.rds"))
+```
+
+```{r, include=FALSE}
+mire_gran <- readRDS(paste0(pData, "cache/mire_gran.rds"))
+```
+
+```{r, fig.cap="Wetlands in Gran municipality"}
+ggplot()+
+  geom_stars(data = st_downsample(mire_gran, 10))
+```
+
+Then I rasterize the indicator map to match the EDM. This takes a while, even for just one municipality. I'm not sure this would work for the entire country.
+
+```{r, eval=F}
+gran_rast <- st_rasterize(gran, template = mire_gran)  
+saveRDS(gran_rast, paste0(pData, "cache/gran_rast.rds"))
+```
+
+```{r, include=FALSE}
+gran_rast <- readRDS(paste0(pData, "cache/gran_rast.rds"))
+```
+
+```{r}
+gran_outline <- muni %>%
+  filter(kommunenummer == "3446")
+```
+
+```{r, fig.cap="Indictor values (alien species, extrapolated and area weighted values) in Gran municipality"}
+tm_shape(st_downsample(gran_rast, 10))+
+  tm_raster(col="w_mean")+
+  tm_shape(gran_outline)+
+  tm_borders(col="red")
+```
+
+Then i mask
+
+```{r}
+gran_rast_masked <- gran_rast
+gran_rast_masked[mire_gran == 0] <- NA
+```
+
+```{r, fig.cap="Indictor values (alien species, extrapolated and area weighted values) in Gran municipality"}
+tm_shape(st_downsample(gran_rast_masked, 10))+
+  tm_raster(col="w_mean",
+            palette = "red")+
+  tm_shape(gran_outline)+
+  tm_borders(col="red")
+```
+
+From here it is easy to get zonal statistics using `exactextratr`.
+If I have this kind of map for all of Norway I can get the mean value for the regions (defined via a polygon set) and it will of course be area weighted.

### 2023-03-22 DRAFT_alien_species.Rmd
types etc

--- 2023-03-22 DRAFT_alien_species.Rmd
@@
 Ecosystem <- "Våtmark, Naturlig åpne områder under skoggrensa, Semi-naturlig mark" 
-Egenskap  <- "Primærproduksjon (evt. Abiotiske fohold)"
-ECT       <- "Physical state characteristics" 
+Egenskap  <- "Funksjonelt viktige arter og strukturer"
+ECT       <- "B1 Compositional state characteristics" 
 Contact   <- "anders.kolstad@nina.no"
@@
 This indicator represent the the ecological on-site effect from  alien species.
-Whilst the introduction of alien species is best considered a pressure variable, the physical abundance or relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a good condition indicator.
+Whilst the introduction of alien species is best considered a pressure variable, the relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a valid condition indicator.
 
@@
 
-This time we are using the [nature type mapping dataset](#naturtype) which records the _effect_ of alien species,, rather than the % cover.
+This time we are using the [nature type mapping dataset](#naturtype) which records the assumed _effect_ of alien species, rather than a direct measure of abundance such as % cover.
 The recording scale is called R7 (Fig.
@@
 ```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO.
-I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here exualrs to an R7 value of 1.
+I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here equals an R7 value of 1.
 Cropped from Evju et al (2022)."}
@@
 Since the nature type data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
-We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region.
-In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas as regions (5 accounting areas in Norway).
-For this indicator, we expect more gradients also inside accounting areas, and therefore we define areas as municipalities.
+We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure intersected with municipality borders, and we say that if we have more than _n_ data points than this field data is representative inside the entire HIA.
+In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas to define the HIA (5 accounting areas in Norway), in addition to the infrastructure classes.
+For this indicator, we expect more gradients also inside accounting areas, and therefore we define HIAs based on municipalities.
 This means that there will be more missing data in our final indicator maps, since not all municipalities will have sufficient data points to estimate indicator values.
 
-We then calculate an area weighted mean (and error) indicator value for each HIA and municipality combination, as long as there is more than _n_ data points for a given combination of HIA and municipality.
+We then calculate an area weighted mean (and error) indicator value for each HIA, as long as there is more than _n_ data points for a given combination of HIA and municipality.
 
@@
 The indicator uses a data set from a standardised field survey of nature
-types. You can read more about the preliminary analyses
+types. You can read more about this data set from my preliminary analyses
 [here](#naturtype). See also the [official
@@
 [ANO](#ANO-import2), which you can read about
-[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/)
+[here](https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/).
 
@@
 
-The variables are recorded on a seven-step ordinal scale (Fig.
-\@ref(fig:R7)) or a %cover of alien species.
+The variables are recorded on a unitless seven-step ordinal scale (Fig.
+\@ref(fig:R7)) or a as a % cover of alien species.
 
@@
 **Funksjonelt viktige arter og strukturer** (Functionally important species and structures).
+This emphasises the negaivte effect that alien species has on for example supressing or excluding native species.
+In some cases, alien species can also affect nutrient cycling or recruitment.
 
@@
 We also have a separate [summary file](##exp-natureType-summary) where
-the nature types are manually mapped the NiN variables and to the correct NiN-main types. 
-We can use this to find the nature types associated with the NiN main types
-that we are interested in.
+the nature types are manually mapped to the NiN variables and to the correct NiN-main types. 
+We can use this to find the nature types of interest.
 
@@
 Importing and sub-setting the main data file, fix duplicate _hovedøkosystem_, calculate area, split one column in two, make numeric, and select target variables:
-```{r, message=F}
+```{r, message=F, warning=FALSE}
 naturetypes <- sf::st_read(dsn = path) %>%
@@
 
-```{r alien-naturetype, fig.cap="Alien species variable scores."}
+```{r alien-naturetype, fig.cap="Alien species variable scores in the nature type data set."}
 ggplot(naturetypes, aes(x = NiN_variable_value))+
@@
 
-_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all futute GRUK data, will not be included in this data set._
+_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all future GRUK data, will not be included in this data set._
 
@@
 Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
-placed locations each with 18 sample points. In each sample point a
+placed locations in Norway, each with 18 sample points. In each sample point a
 circle of 250 m2 is visualised, and the main ecosystem is recorded.
@@
 was not recorded at all in 2019:
-```{r}
-# Hard coding name change
-#unique(ano$hovedtype_250m2)[3]
-#ano$hovedtype_250m2[ano$hovedtype_250m2==unique(ano$hovedtype_250m2)[3]] <- "Aapen jordvassmyr"
-```
 
@@
 
-The zero in there is an obvious mistake.
+The zero in there is an obvious mistake. Removing it:
 
@@
 
-```{r ano-alien-dist, fig.cap"Distribution of %cover of alien plants variable in the ANO data set"}
+```{r ano-alien-dist, fig.cap="Distribution of %cover of alien plants variable in the ANO data set"}
 ggplot(ano, aes(x = fa_total_dekning))+
@@
 
-Now I will convert these %cover values into 7FA classes. This is a critical step, since there is not one-to-one mapping here, 
+Now I will convert these %cover values into 7FA classes. This is a critical step, since there is no one-to-one mapping here, 
 but rather there is considerable overlap, as can be seen in Fig. \@ref(fig:fremmedart-prosentVSinnslag).
@@
 
-```{r ano-alien-dist-R7, fig.cap"Distribution of R7 values in the ANO data set"}
+```{r ano-alien-dist-R7, fig.cap="Distribution of R7 values in the ANO data set"}
 ggplot(ano, aes(x = var_7FA))+
@@
 
-Because later we might want to save this as a shape file, we cannot have
-a mixed class type. I will therefore convert the points to polygons. I
+Because later we will rasterize these data, we will  convert the points to polygons. I
 use the area column to calculate a radii that gives that area.
@@
 
-Let's see the proportion of data points (not area) origination from each
+Let's see the proportion of data points (not area) originating from each
 data set
 
-```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the alien species indicator.", fig.width=4}
+```{r, fig.cap="Barplot showing the contribution (number of localities) of different data sets to the alien species indicator.", fig.width=4}
 temp <- as.data.frame(table(alien_data$origin))
@@
 
-The GIS layers are used to crop out marine areas, and to define
+These layers are used to crop out marine areas, and to define
 accounting areas, respectively.
@@
 mapped areas leaves the indicator without much value for regional
-assessments. When we do regional assessments and calculate regional
-indicator values, we cannot simply do an area weighting of the polygons
-in each region. This is because we don't want to assume that the
+assessments. But we cannot simply do an area weighting of the polygons
+in each region. This is because we expect considerable sampling bias
+and we can't assume that the
 polygons are representative far outside of the mapped area. But perhaps
@@
 
-2.  take a simplified infrastructure index (vector data set with HIAs)
-    and extract the corresponding indicator values that intersect with a
+2.  take a simplified infrastructure index (vector data) and intersect this with municipality boarders to produce a map of homogeneous impact areas (HIAs).
+
+1.  extract the corresponding indicator values that intersect with a
     given HIA, and extrapolate the area weighted mean of those values to
-    the entire HIA withing a given geographical region (municipality). Errors are
+    the entire HIA. Errors are
     calculated from bootstrapping.
 
-3.  Take the new area weighted indicator values for the HIAs and match
-    them with ecosystem occurrences of the same HIA (_possibly using
-    randomly sampled points from the available ecosystem occurrences_).
-    The errors are the same as for the HIAs.
+3.  Take the new area weighted indicator values for the HIAs, rasterize it, and mask these
+    with ecosystem occurrences. The errors are the same as for the HIAs.
 

### 2023-03-22 DRAFT_alien_species.Rmd
more typos and text edits

@@
 
-2.  take a simplified infrastructure index (vector data) and intersect this with municipality boarders to produce a map of homogeneous impact areas (HIAs).
+2.  take a simplified infrastructure index (vector data) and intersect this with municipality borders to produce a map of homogeneous impact areas (HIAs).
 
@@
 
-I now have `r length(unique(HIA_muni$muni_HIF))` unique areas (for each main ecosystem) that I will, given
+I now have `r length(unique(HIA_muni$muni_HIF))` HIAs (for each main ecosystem) that I will, given
 there is data, extrapolate indicator values over. But first I want to validate the assumption that the HIF 
-explains a considerable portion of the variation i the indicator values.
+explains a considerable portion of the variation in the indicator values.
 
@@
 
-```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot showing the distribution of polygons with different indicator values."}
 ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
@@
 I now want to find the mean indicator value for each HIA (i.e. to
-aggregate) and to spread these out spatially to the entire HIA within
-each municipality (i.e. to extrapolate).
+aggregate) and to _spread_ these out spatially to the entire HIA (i.e. to extrapolate).
 
@@
 
-It's easier to see what's happening if we zoom in a bit. I will look more closely at three example municipalities: Trondheim (municipality nr 5001), Nordre Follo (3020) and Målselv (5418).
+It's easier to see what's happening if we zoom in a bit. I will look more closely at three example municipalities:
+
+Trondheim (municipality nr 5001), Nordre Follo (3020) and Målselv (5418).
 
@@
 
-```{r extrapolated-alien, fig.cap="Alein species indicator extrapolated over Nordre Follo (3020), Trondheim (5001) and Målselv (5418).
+```{r extrapolated-alien, fig.cap="Alien species indicator extrapolated over Nordre Follo (3020), Trondheim (5001) and Målselv (5418).
 Note that the maps are not masked by ecosystem type, so the three ecosystem specific indicators are overlaping.
-The bottom row shows the location of the four different human impact sones."}
+The bottom row shows the location of the four human impact sones."}
 myCol <- "RdYlGn"
@@
 
-Figure \@ref(fig:extrapolated-alien) shows that this indicator will have a lot of missing data.
+Figure \@ref(fig:extrapolated-alien) again indicates that this indicator will have a lot of missing data (see also [here](#alien-data-coverage)).
 For Nordre Follo and Trondheim for example, we do not have any indicator values tied to naturally open or wetland ecosystems, and for semi natural ecosystems we only have data for HIF class 2.
 This is, however, still a lot better than discarding the entire data set due to issues related to area representativity.
-The amount of missing data in the indicator maps is closely ties to the chosen threshold value for how many data points we must have in order to calculate an (area weighted) men indicator value.
+The amount of missing data in the indicator maps is closely tied to the chosen threshold value for how many data points we must have in order to calculate an (area weighted) mean indicator value.
 We have used 20 as the threshold here.
@@
 
-The following table contains the data values underlying Figure \@ref(fig:extrapolated-alien), including the combination with <20 data points.
+The following table contains the data values underlying Figure \@ref(fig:extrapolated-alien), including HIAs with <20 data points.
 
@@
 
-Note that we can also show the uncertainties as maps
-
-```{r sd-alien, fig.cap="Map showing the uncertainties around the indicator values for alien species in three example municipalities for Semi-natural ecosystems."}
+Note that we can also show the uncertainties on maps:
+
+```{r sd-alien, fig.cap="Map showing the uncertainties (spatial variation) around the indicator values for alien species in three example municipalities for Semi-natural ecosystems."}
 tm_shape(alien_extr_trd_nf[alien_extr_trd_nf$eco=="semi_natural",])+
@@
 
-#### Data coverage
+#### Data coverage {#alien-data-coverage}
 
@@
 value 1 means that all HIAs had > 20 data points, and 0.2 means 20% of them did).
-If assuming the ecosystems are equally common where they are mapped and where hey are not, this measure of data coverage is representative of the real indicator coverage after maksing the maps with ecosystem delineations."}
+If assuming the ecosystems are equally common where they are mapped and where they are not, this measure of data coverage is representative of the real indicator coverage after masking the maps with ecosystem delineations."}
 alien_extr %>%
@@
            stat = "identity")+
+  labs(x = "Ecosystem", y = "Relative indicator data coverage")+
   theme_bw()
@@
 Also, with more nature type data accumulating over time, the data coverage will increase.
-Therefore this indicator is worth while, and is seen as considerably better than the alternative to discard the entire data set.
+Therefore I think this indicator is worth while, and that this is considerably better than the alternative to discard the entire data set.
 
@@
 
-```{r}
+```{r, warning=FALSE}
 temp <- alien_extr %>%
@@
 From here it is easy to get zonal statistics using `exactextratr`.
-If I have this kind of map for all of Norway I can get the mean value for the regions (defined via a polygon set) and it will of course be area weighted.
+If I have this kind of map for all of Norway I can get the mean value for the regions (defined via a polygon set) and it will of course be area weighted (but no weight given to ecosystem occurences that don't have indicator values).

### 2023-03-23 DRAFT_alien_species.Rmd
even more types and text edits

@@
 library(tibble)
-#library(ggplot2)
-#library(tidyr)
-#library(dplyr)
+library(stars)
 knitr::opts_chunk$set(echo = TRUE)
@@
 corrCheck <- st_intersection(alien_data, HIA)
-#saveRDS(corrCheck, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
-saveRDS(corrCheck, "P:/41201785_okologisk_tilstand_2022_2023/data/cache/corrCheck_alienSpecies.rds")
+saveRDS(corrCheck, paste0(pData, "cache/corrCheck_alienSpecies.rds")
 ```
@@
   add_column(total_area = total_area) %>%
-  mutate(relative_area = sum_area/total_area) %>%
+  mutate(relative_area = units::drop_units(sum_area/total_area)) %>%
   ggplot()+
@@
 
-```{r lat-effect, fig.cap="Effect of latitude on the alien species indicator values. The blue line is a loess smoother (span=0.75)."}
+```{r lat-effect, fig.cap="Effect of latitude on the alien species indicator values. The blue line is a loess smoother (span=0.75).", message=FALSE}
 temp2 %>%
@@
 
-1. Prepare ecosystem delineation maps in EPSG25833 and perfectly aligned to a master grid
+1. Prepare ecosystem delineation maps in EPSG:25833 and perfectly aligned to a master grid
 
@@
 
-```{r, fig.cap="Indictor values (alien species, extrapolated and area weighted values) in Gran municipality"}
+```{r, fig.cap="Indictor values (alien species, extrapolated and area weighted values) in Gran municipality.
+Note that the pixelation is very small (10 x 10 m), and that the larger squares are remnants from the HIA map which was 1 x 1 km resolution."}
 tm_shape(st_downsample(gran_rast, 10))+
@@
 
-Then i mask
+Then i mask to remove areas that are not actually watlands.
 
@@
 
-```{r, fig.cap="Indictor values (alien species, extrapolated and area weighted values) in Gran municipality"}
+```{r, fig.cap="Stratified, area weighted indictor values (alien species) in Gran municipality masked by a wetland ecosystem delimination map."}
 tm_shape(st_downsample(gran_rast_masked, 10))+

### 2023-03-23 slitasje.Rmd
edit bugs and typos

@@
 This indicator represent human caused damage or **disturbance to soils and vegetation**, typically from things like recreational activities and off-road vehicle traffic.
-Effects from domestic animal is not included here, and the upper reference value is 0% disturbance.
 The data come from a combination non-random field surveys and area-representative surveys.
+Disturbance to soils and vegetation is recorded in the various data sets are NiN variables 7SE and 7TK, or as the similar, but more fine-scaled variable, PRSL and PRTK.
+Effects from domestic animal is not included in the definition of these variables, and the upper reference value is therefore set to 0% disturbance.
 
@@
 We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure, and we say that the field data is representative inside the HIA and region (_five regions in Norway_) where it was recorded.
-We then calculate an aerea weighted mean (and error) indicator value for each HIA and region combination, as long as there is more than 100 data point for a give combination of HIA and region.
-
-Here a general workflow for the calculation of the indicator.
+We then calculate an area weighted mean (and error) indicator value for each HIA and region combination, as long as there is more than 150 data points for a give combination of HIA and region.
+
+Here is a general workflow for the calculation of the indicator.
 
@@
 The variables are recorded on a coarse four-step ordinal scale (Fig.
-\@ref(fig:four-step)).
+\@ref(fig:four-step)) or eigth-step scale.
 
@@
 bootstrapping individual indicator values 1000 times and calculating a
-distribution of area weighted means. This uncertainty is different from 
+distribution of area weighted means. See the documentation for `eaTools::ea_spread` [here](https://ninanor.github.io/eaTools/reference/ea_spread.html). 
+This uncertainty is different from 
 the spatial variation which we could get more straight forward without 
@@
   mutate(keepers = sum(c_across(
-    myVars)>0, na.rm=T)) %>%
+    all_of(myVars))>0, na.rm=T)) %>%
   filter(keepers > 0) %>%
@@
 
-```{r dense-stats, fig.height= 12, fig.width=6}
+```{r dense-stats, fig.height= 12, fig.cap="Barplot showing the number of data points, the area weighted indicator values and the sd for the slitasje indicator.
+Transparrent bars represents categories with less than 150 data points."}
 temp_plot <- all_stats %>%
@@
   geom_bar(position = "dodge", stat = "identity", col="black")+
-  theme_bw(base_size = 20)+
+  theme_bw(base_size = 10)+
   scale_fill_brewer(palette="Oranges")+
@@
 
-2. Rasterize the extrapolated indicator map, using the ET map as a template
+2. Rasterize the extrapolated indicator map, using the ET map as a template (see Fig. \@ref(fig:#masking-example))
 
@@
 
-Importing the wetland map.
-
-```{r, eval=T}
-#takes about one minute without the proxy
-(mireET <- stars::read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif",
-                             proxy=T))
-```
-
-This is in the wrong CRS. We want to stay with 25833.
-
-We have a 50x50 m [master grid](#master-grid) with the correct CRS, and I would warp mireET to fit that grid like this:
-```{r, eval=F}
-grid <- stars::read_stars(paste0(pData, "masterGrid_50m.tiff"))
-mireET_warped <- stars::st_warp(mireET, grid)
-```
-
-... but I don't want to down sample the mire data, since we will loose a lot of the smaller ecosystem occurrences.
-
-I could make a 10x10 master grid, but I have not found a way to stop this from being just a massive file.
-
-I can transform the ET map like this, and not worry about the rasters being perfectly aligned, but even this operation is too heavy for the R Studio Server:
-
-```{r, eval=F}
-mireET_trans <- sf::st_transform(mireET, 25833)
-```
-
-I will therefore ask GIS engineers to prepare ecosystem maps to the standard that we chose, so that we can simply combine the stars object with `stars::c()`, or perhaps side-step that operatio and just mask directly like this: `wetlands_slitasje_extr[mireET==0] <- NA`.
+See [here](https://github.com/NINAnor/ecosystemCondition/issues/34).
 

### 2023-03-23 climate_indicators_explained.Rmd
include chapter in yml
typos
bug: c.stars() cases bugs due to old cached files

@@
-# Climate indicators excplained
+# Climate indicators explained {#climate-indicators-explained}
 
@@
 ```{r, eval=F}
-saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.rds")
-write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff")
+#saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.rds")
+#write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff")
 ```
@@
 ```{r}
-summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
+summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
                                  proxy=F)
@@
 ```{r, eval=T, include=F}
-summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
 ```

### 2023-03-23 medianSummerTemperature.Rmd
typos
bug: c.stars() cases bugs due to old cached files

@@
 knitr::opts_chunk$set(echo = TRUE)
-#library(tidyverse) # cant update to v2 on server
-library(dplyr)
-library(ggplot2)
+library(tidyverse) # cant update to v2 on server
 library(stars)
@@
 library(parallel)
-library(readr)
 library(exactextractr)
-library(tidyr)
 library(data.table)
@@
 dir <- substr(getwd(), 1,2)
-
 ```
@@
 ```{r, echo=F}
-Ecosystem <- "All" # e.g. "Skog og fjell"
-Egenskap  <- "Abiotiske egenskaper" # e.g. "Primærproduksjon"
-ECT       <- "Physical state characteristics" # e.g. "Structural state characteristic"
-Contact   <- "Anders L. Kolstad" # e.g. "Anders Kolstad"
+Ecosystem <- "All" 
+Egenskap  <- "Abiotiske egenskaper"
+ECT       <- "Physical state characteristics" 
+Contact   <- "Anders L. Kolstad" 
 ```
@@
 
-```{r}
+```{r, warning=F}
 reg <- st_intersection(reg, nor)
@@
 ```{r, fig.cap="Five accounting areas (regions) in Norway."}
-
 tm_shape(reg) +
@@
 
-This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 raster are read into memory. Here's a test for the effect of splitting over more cores.
+This takes about 20 sec per file/year, or 22 min on total. That is not too bad. About 6000 rasters are read into memory. Here's a test for the effect of splitting over more cores.
 
 ```{r, eval=F}
-write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp.tiff")
+write_stars(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff")
+```
+
+```{r, eval=F}
+# a backup caching in case tiff causes problems down the line
+saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.rds")
+
 ```
@@
 ```{r, include=F}
-summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer.tiff",
+real_temp_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
                                  proxy=F)
@@
 ```{r}
-summer_median_temp <- summer_median_temp %>% 
+summer_median_temp <- real_temp_summer %>% 
   st_set_dimensions(names = c("x", "y", "v_YEAR")) %>%
@@
 ggplot()+
-  geom_stars(data = summer_median_temp[,,,c(1,11,22)], downsample = c(10, 10, 0)) +
+  geom_stars(data = summer_median_temp[,,,c(1,11,66)], downsample = c(10, 10, 0)) +
   coord_equal() +
@@
 
+```{r}
+median_sd <- function(x) { c(median = median(x), sd = sd(x))}
+```
+
+<!-- RUN AGAIN -->
 ```{r, eval=FALSE}
@@
 ```{r, eval=T, include=F}
-summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_ref.RData")
+summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
 ```
 
-Pivot and turn dimension into attributes, and rename attribute:
+Pivot and turn dimension into attributes, and rename attributes:
 
@@
 y_var <- summer_median_temp %>%
-  split(3) %>%
+  split("v_YEAR") %>%
   c(summer_median_temp_ref_long)

### 2023-04-13 DRAFT_functional plant indicators_wetland.Rmd
fpi wetland - until analysis of results

@@
 ---
-title: "Template for indicator documentation"
+title: "Functional pant indicators - wetland, indicator documentation"
 output:
@@
 Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name te file DRAFT_myIndicatorName.Rmd.
+Name the file DRAFT_myIndicatorName.Rmd.
 
@@
 
-<!-- Replace 'indicator name' with yor own indicator name  -->
+<!-- Replace 'indicator name' with your own indicator name  -->
 # Functional plant indicators (Moisture, Light, pH, Nitrogen)
@@
 library(readxl)
+library(broom)
+library(RColorBrewer)
+library("gridExtra") 
+library(ggridges)
+library(ggplot2)
+library(tmap)
+library(raster)
+library(stars)
+
 library(knitr)
@@
 Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
-ECT       <- "" # e.g. "Structural state characteristic"
+ECT       <- "Functional state characteristic"
 Contact   <- "Joachim Töpper"
@@
 
-```{r}
+```{r, echo = F}
 st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
@@
 Plant indicators from Tyler et al. 2021
-```{r}
+```{r, echo = F}
 ## Tyler indicator data
@@
                         sep = '\t', header=T, quote = '')
-head(ind.datr)
+head(ind.dat)
 
@@
 Generalized species lists NiN
-```{r}
+```{r, echo = F}
 load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
@@
 
-#### Plant indicator data - data handling
-```{r}
+#### Data handling
+- Checking for errors
+- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
+- Merging indicator data with monitoring data and indicator data with reference data
+(not shown here, but documented in the code)
 
+```{r, include = F}
+
+### Plant indicator data
 names(ind.dat)
@@
 # the trimming above leaves some duplicates that need to be taken care of
-ind.dat[duplicated(ind.dat[,'species']),"species"] ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat[duplicated(ind.dat[,'species']),"species"]
+ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
 ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
@@
 
-```
+### ANO monitoring data
 
-#### ANO monitoring data - data handling
-```{r}
 head(ANO.sp)
@@
 
-```
-
-#### NiN reference data - data handling
-
-```{r}
+### NiN reference data - data handling
 
@@
 
-#### CONTINUE HERE ####
-
-## matching with NiN ecosystem types
+## matching with NiN ecosystem types - for wetlands
 # NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
-NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[19:23])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
+NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[15:18])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
 NiN.wetland[1,]
@@
 
-#NiN.wetland <- NiN.wetland[,c(1:41,46:48,59:69)]
 cbind(colnames(NiN.wetland),
@@
         'V2-C3a','V2-C3b',
-        "V4-C4","V4-C5",
         "V4-C2","V4-C3",
+        "","",
         'V8-C1','V8-C2','V8-C3',
         rep("",10),
-        rep("",5)
+        rep("",4)
       )
@@
 
-NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:63)]
+NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:62)]
 colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
@@
 
-summary(NiN.wetland)
-summary(NiN.wetland.cov)
+NiN.wetland.cov$sp <- as.factor(NiN.wetland.cov$sp)
+```
+leaving us with the monitoring data including plant indicators (ANO.sp.ind) and the reference data including plant indicators (NiN.wetland.cov)
+
+```{r}
+#summary(NiN.wetland)
+head(ANO.sp.ind)
+head(NiN.wetland.cov)
+```
+For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
+For wetland ecosystem we are testing "Light", "Moisture", "Soil_reaction_pH", and "Nitrogen".
+In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.
+
+##### bootstrap function for frequency abundance
+- function to calculate community weighted means of selected indicator values (ind)
+- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
+- with a given number of iterations (iter),
+- with species given a certain minimum abundance ocurring in all bootstraps (obl), and
+- with a given resampling ratio of the original species list (rat)
+- in every bootstrap iteration the abundance of the sampled species can be 
+- randomly changed by +/-1 if wished (var.abun)
+
+```{r, echo = F}
+indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
+
+  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
+  colnames(ind.b) <- colnames(abun)
+  ind.b <- as.data.frame(ind.b)
+
+  ind <- as.data.frame(ind)
+  ind.list <- as.list(1:length(colnames(ind)))
+  names(ind.list) <- colnames(ind)
+
+  for (k in 1:length(colnames(ind)) ) {
+    ind.list[[k]] <- ind.b }
+
+  for (j in 1:length(colnames(abun)) ) {
+
+    dat <- cbind(sp,abun[,j],ind)
+    dat <- dat[dat[,2]>0,]            # only species that are present in the ecoystem
+    dat <- dat[!is.na(dat[,3]),]      # only species that have Ellenberg values
+
+    for (i in 1:iter) {
+
+      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
+      dat.b <- rbind(dat[dat[,2] >= obl,],
+                     dat[match(speciesSample,dat$sp),]
+      )
+
+      if (var.abun==T) {
+        for (m in 1:nrow(coverscale[-1,]) ) {
+          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
+          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
+          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
+          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
+          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
+          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
+          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
+        }
+        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
+        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
+      }
+
+      for (k in 1:length(colnames(ind))) {
+
+        if ( nrow(dat.b)>2 ) {
+
+          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
+          ind.list[[k]][i,j] <- ind.b
+
+        } else {ind.list[[k]][i,j] <- NA}
+
+      }
+
+#      print(paste(i,"",j)) 
+    }
+
+  }
+  return(ind.list)
+}
+```
+
+Running the bootstraps
+
+```{r, echo = T, eval = F}
+colnames(NiN.wetland)
+# 1st column is the species
+# 2nd-46th column is the abundances of sp in different ecosystem types
+# 47th-51st column is the indicator values of the respective sp
+# we choose 1000 iterations
+# species with abundance 1 (i.e. 100%, must be included in each sample)
+# each sample resamples 2/3 of the number of species
+# the abundance of the resampled species may vary (see bootstrap function for details)
+wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
+                          iter=1000,obl=1,rat=2/3,var.abun=T)
+
+# fixing NaNs
+for (i in 1:length(wetland.ref.cov) ) {
+  for (j in 1:ncol(wetland.ref.cov[[i]]) ) {
+    v <- wetland.ref.cov[[i]][,j]
+    v[is.nan(v)] <- NA
+    wetland.ref.cov[[i]][,j] <- v
+  }
+}
+```
+
+```{r, include = F}
+load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_wetland.RData")
+```
+
+```{r}
+head(wetland.ref.cov[[1]])
+```
+
+This results in a list with a slot for every selected indicator, and in every slot there's as many columns as there are NiN species lists and as many row as there were iterations in the bootstrap.
+Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here we define things in the following way:
+
+Median = reference values
+0.025 and 0.975 quantiles = lower and upper limit values
+min and max of the respective indicator's scale = min/max values
+
+```{r, echo = F}
+
+# every NiN-type is represented by one 'generalisert artsliste'
+# some NiN-types are represented by two such species lists
+# in some cases two NiN-types are represented by the same species list
+head(wetland.ref.cov[[1]])
+wetland.ref.cov[[1]][0,]
+
+# checking the actual NiN-types in the wetland lists
+wetland.NiNtypes <- colnames(wetland.ref.cov[["Light"]])
+wetland.NiNtypes <- substr(wetland.NiNtypes,1,5)
+unique(wetland.NiNtypes)
+
+# 4 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, "Nitrogen"
+indEll.n=4
+# creating a table to hold:
+# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
+# for every nature type (nrows)
+tab <- matrix(ncol=3*indEll.n, nrow=length(unique(wetland.NiNtypes)) ) # 43 basic ecosystem types
+# coercing the values into the table
+# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
+names(wetland.ref.cov[["Light"]])
+x <- c(27,28,41:45)
+
+for (i in 1:length(x) ) {
+  tab[i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+}
+
+tab <- as.data.frame(tab)
+tab$NiN <- NA
+tab$NiN[1:length(x)] <- names(wetland.ref.cov[[1]])[x]
+
+# NiN-types represented by several species lists
+wetland.NiNtypes2 <- wetland.NiNtypes[-x]
+unique(wetland.NiNtypes2)
+grep(pattern=unique(wetland.NiNtypes2)[1], x=wetland.NiNtypes) # finds columns in e.g. colnames(wetland.ref.cov[["Continentality"]]) that match the first NiN-type
+
+for (i in 1:length(unique(wetland.NiNtypes2)) ) {
+  tab[length(x)+i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[length(x)+i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[length(x)+i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[length(x)+i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab$NiN[length(x)+i] <- unique(wetland.NiNtypes2)[i]
+
+}
+
+tab
+
+# making it a proper data frame
+round(tab[,1:12],digits=2)
+
+colnames(tab) <- c("Light_q2.5","Light_q50","Light_q97.5",
+                   "Moist_q2.5","Moist_q50","Moist_q97.5",
+                   "pH_q2.5","pH_q50","pH_q97.5",
+                   "Nitrogen_q2.5","Nitrogen_q50","Nitrogen_q97.5",
+                   "NiN")
+summary(tab)
+tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphon after C for NiN-types
+tab
+
+# restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
+y.Light <- numeric(length=nrow(tab)*2)
+y.Light[((1:dim(tab)[1])*2)-1] <- tab$Light_q2.5 
+y.Light[((1:dim(tab)[1])*2)] <- tab$Light_q97.5
+
+y.Moist <- numeric(length=nrow(tab)*2)
+y.Moist[((1:dim(tab)[1])*2)-1] <- tab$Moist_q2.5 
+y.Moist[((1:dim(tab)[1])*2)] <- tab$Moist_q97.5
+
+y.pH <- numeric(length=nrow(tab)*2)
+y.pH[((1:dim(tab)[1])*2)-1] <- tab$pH_q2.5 
+y.pH[((1:dim(tab)[1])*2)] <- tab$pH_q97.5
+
+y.Nitrogen <- numeric(length=nrow(tab)*2)
+y.Nitrogen[((1:dim(tab)[1])*2)-1] <- tab$Nitrogen_q2.5 
+y.Nitrogen[((1:dim(tab)[1])*2)] <- tab$Nitrogen_q97.5
+
+# creating final objects holding the reference and limit values for all indicators
+
+# ref object for indicators
+wetland.ref.cov.val <- data.frame(N1=rep('wetland',(nrow(tab)*2*indEll.n)),
+                              hoved=c(rep('NA',(nrow(tab)*2*indEll.n))),
+                              grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
+                              county=rep('all',(nrow(tab)*2*indEll.n)),
+                              region=rep('all',(nrow(tab)*2*indEll.n)),
+                              Ind=c(rep(c('Light1','Light2'),nrow(tab)),
+                                    rep(c('Moist1','Moist2'),nrow(tab)),
+                                    rep(c('pH1','pH2'),nrow(tab)),
+                                    rep(c('Nitrogen1','Nitrogen2'),nrow(tab))
+                              ),
+                              Rv=c(rep(tab$Light_q50,each=2),
+                                   rep(tab$Moist_q50,each=2),
+                                   rep(tab$pH_q50,each=2),
+                                   rep(tab$Nitrogen_q50,each=2)
+                              ),
+                              Gv=c(y.Light,y.Moist,y.pH,y.Nitrogen),
+                              maxmin=c(rep(c(1,7),nrow(tab)), # 7 levels of light
+                                       rep(c(1,12),nrow(tab)), # 12 levels of moisture
+                                       rep(c(1,8),nrow(tab)), # 8 levels of soil reaction pH
+                                       rep(c(1,9),nrow(tab))  # 9 levels of nitrogen
+                              )
+)
+
+wetland.ref.cov.val
+wetland.ref.cov.val$grunn <- as.factor(wetland.ref.cov.val$grunn)
+wetland.ref.cov.val$Ind <- as.factor(wetland.ref.cov.val$Ind)
+summary(wetland.ref.cov.val)
 
 ```
+
+```{r}
+head(wetland.ref.cov.val)
+```
+
+Once testing data (ANO), and reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution.
+
+Here is the scaling function (with and without truncation at 1)
+```{r}
+
+#### scaled values ####
+r.s <- 1    # reference value
+l.s <- 0.6  # limit value
+a.s <- 0    # abscence of indicator, or indicator at maximum
+
+#### function for calculating scaled values for measured value ####
+
+## scaling function including truncation
+scal <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value equal 1
+    if (val >= ref) {x <- 1}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
+    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
+    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
+    # value equals or lower than maxmin
+    if (val <= maxmin) {x <-0}
+  } else {
+    # values <= the reference value equal 1
+    if (val <= ref) {x <- 1}
+    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
+    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
+    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
+    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
+    # value equals or larger than maxmin
+    if (val >= maxmin) {x <-0}
+  }
+  return(x)
+
+}
+
+## scaling function without truncation
+scal.2 <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
+    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
+    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
+    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
+    # value equal or lower than maxmin
+    if (val <= maxmin) {x <-0}
+  } else {
+    # values <= the reference value estimated from the linear relationship for lim < x < ref (line below)
+    if (val <= ref) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
+    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
+    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
+    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
+    if (val > lim & val < maxmin) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
+    # value equal og larger than maxmin
+    if (val >= maxmin) {x <-0}
+  }
+  return(x)
+
+}
+```
+
+We then can prepare a list of dataframes to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
+```{r, echo = T, eval = F}
+
+unique(ANO.geo$hovedtype_rute) # NiN types in data
+unique(substr(wetland.ref.cov.val$grunn,1,2)) # NiN types in reference
+#### creating dataframe to hold the results for wetlands ####
+# all ANO points
+nrow(ANO.geo)
+# all wetland ANO points
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),])
+# all wetland ANO points with a NiN-type represented in the reference
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.cov.val$grunn,1,2)),])
+# ok, we'll be losing 70 (our of 1349) that are not covered by the reference
+ANO.wetland <- ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),]
+
+head(ANO.wetland)
+# update row-numbers
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland)
+head(ANO.wetland)
+dim(ANO.wetland)
+colnames(ANO.wetland)
+
+length(levels(as.factor(ANO.wetland$ano_flate_id)))
+length(levels(as.factor(ANO.wetland$ano_punkt_id)))
+summary(as.factor(ANO.wetland$ano_punkt_id))
+# four points that are double
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0159_55",] # double registration, said so in comment. -> choose row 207 over 206
+ANO.wetland <- ANO.wetland[-206,]
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0283_22",] # 2019 & 2021. Lot of NA's in 2019 -> omit 2019
+ANO.wetland <- ANO.wetland[-156,]
+row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0363_24",]
+ANO.wetland[ANO.wetland$ano_flate_id=="ANO0363","ano_punkt_id"] # point-ID 15 is missing, but 24 is double.
+Likely that registrations are valid, but wrong point-ID.
+-> keep both, call the second obs the one that's missing
+ANO.wetland[311,"ano_punkt_id"] <- "ANO0363_15"
+ANO.wetland[ANO.wetland$ano_punkt_id=="ANO1550_64",] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
+ANO.wetland[ANO.wetland$ano_flate_id=="ANO1550","ano_punkt_id"] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
+ANO.wetland[1273,"ano_punkt_id"] <- "ANO1550_66"
+
+unique(ANO.wetland$hovedoekosystem_punkt)
+unique(ANO.wetland$hovedtype_rute)
+unique(ANO.wetland$kartleggingsenhet_1m2)
+ANO.wetland$hovedtype_rute <- factor(ANO.wetland$hovedtype_rute)
+ANO.wetland$kartleggingsenhet_1m2 <- factor(ANO.wetland$kartleggingsenhet_1m2)
+summary(ANO.wetland$Hovedtype_rute)
+summary(ANO.wetland$Kartleggingsenhet_rute)
+
+results.wetland <- list()
+ind <- unique(wetland.ref.cov.val$Ind)
+# choose columns for site description
+colnames(ANO.wetland)
+results.wetland[['original']] <- ANO.wetland
+# drop geometry
+st_geometry(results.wetland[['original']]) <- NULL
+results.wetland[['original']]
+
+# add columns for indicators
+nvar.site <- ncol(results.wetland[['original']])
+for (i in 1:length(ind) ) {results.wetland[['original']][,i+nvar.site] <- NA}
+colnames(results.wetland[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
+for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.wetland[['original']][,i] <- as.numeric(results.wetland[['original']][,i])}
+summary(results.wetland[['original']])
+#results.wetland[['original']]$Region <- as.factor(results.wetland[['original']]$Region)
+results.wetland[['original']]$GlobalID <- as.factor(results.wetland[['original']]$GlobalID)
+results.wetland[['original']]$ano_flate_id <- as.factor(results.wetland[['original']]$ano_flate_id)
+results.wetland[['original']]$ano_punkt_id <- as.factor(results.wetland[['original']]$ano_punkt_id)
+results.wetland[['original']]$hovedoekosystem_punkt <- as.factor(results.wetland[['original']]$hovedoekosystem_punkt)
+#results.wetland[['original']]$Hovedoekosystem_rute  <- as.factor(results.wetland[['original']]$Hovedoekosystem_rute )
+results.wetland[['original']]$kartleggingsenhet_1m2 <- as.factor(results.wetland[['original']]$kartleggingsenhet_1m2)
+results.wetland[['original']]$hovedtype_rute    <- as.factor(results.wetland[['original']]$hovedtype_rute)
+
+# roll out
+results.wetland[['scaled']] <- results.wetland[['non-truncated']] <- results.wetland[['original']]
+
+#### calculating scaled and non-truncated values for the indicators based on the dataset ####
+for (i in 1:nrow(ANO.wetland) ) {  #
+  tryCatch({
+    print(i)
+    print(paste(ANO.wetland$ano_flate_id[i]))
+    print(paste(ANO.wetland$ano_punkt_id[i]))
+#    ANO.wetland$Hovedoekosystem_sirkel[i]
+#    ANO.wetland$Hovedoekosystem_rute[i]
+
+    # if the ANO.hovedtype exists in the reference
+    if (ANO.wetland$hovedtype_rute[i] %in% unique(substr(wetland.ref.cov.val$grunn,1,2)) ) {
+
+      # if there is any species present in current ANO point  
+      if ( length(ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),'Species']) > 0 ) {
+
+          # Light
+          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Light')]
+          results.wetland[['original']][i,'richness'] <- nrow(dat)
+          dat <- dat[!is.na(dat$Light),]
+
+          if ( nrow(dat)>0 ) {
+
+            val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
+            # lower part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Light1'] <- scal() 
+            results.wetland[['non-truncated']][i,'Light1'] <- scal.2() 
+            results.wetland[['original']][i,'Light1'] <- val
+
+            # upper part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Light2'] <- scal() 
+            results.wetland[['non-truncated']][i,'Light2'] <- scal.2() 
+            results.wetland[['original']][i,'Light2'] <- val
+          }
+
+          # Moisture
+          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Moisture')]
+          results.wetland[['original']][i,'richness'] <- nrow(dat)
+          dat <- dat[!is.na(dat$Moisture),]
+
+          if ( nrow(dat)>0 ) {
+
+            val <- sum(dat[,'art_dekning'] * dat[,'Moisture'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
+            # lower part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Moist1'] <- scal() 
+            results.wetland[['non-truncated']][i,'Moist1'] <- scal.2() 
+            results.wetland[['original']][i,'Moist1'] <- val
+
+            # upper part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Moist2'] <- scal() 
+            results.wetland[['non-truncated']][i,'Moist2'] <- scal.2() 
+            results.wetland[['original']][i,'Moist2'] <- val
+          }
+
+          # Soil_reaction_pH
+          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Soil_reaction_pH')]
+          results.wetland[['original']][i,'richness'] <- nrow(dat)
+          dat <- dat[!is.na(dat$Soil_reaction_pH),]
+
+          if ( nrow(dat)>0 ) {
+
+            val <- sum(dat[,'art_dekning'] * dat[,'Soil_reaction_pH'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
+            # lower part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'pH1'] <- scal() 
+            results.wetland[['non-truncated']][i,'pH1'] <- scal.2() 
+            results.wetland[['original']][i,'pH1'] <- val
+
+            # upper part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'pH2'] <- scal() 
+            results.wetland[['non-truncated']][i,'pH2'] <- scal.2() 
+            results.wetland[['original']][i,'pH2'] <- val
+          }
+
+          # Nitrogen
+          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Nitrogen')]
+          results.wetland[['original']][i,'richness'] <- nrow(dat)
+          dat <- dat[!is.na(dat$Nitrogen),]
+
+          if ( nrow(dat)>0 ) {
+
+            val <- sum(dat[,'art_dekning'] * dat[,'Nitrogen'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
+            # lower part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Nitrogen1'] <- scal() 
+            results.wetland[['non-truncated']][i,'Nitrogen1'] <- scal.2() 
+            results.wetland[['original']][i,'Nitrogen1'] <- val
+
+            # upper part of distribution
+            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
+            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
+            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
+            # coercing x into results.wetland dataframe
+            results.wetland[['scaled']][i,'Nitrogen2'] <- scal() 
+            results.wetland[['non-truncated']][i,'Nitrogen2'] <- scal.2() 
+            results.wetland[['original']][i,'Nitrogen2'] <- val
+          }
+
+        }
+      }
+
+  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+}
+
+# for using both sides of the plant indicators
+results.wetland[['2-sided']] <- results.wetland[['non-truncated']]
+
+# remove values >1 for 2-sided indicators
+results.wetland[['2-sided']]$Light1[results.wetland[['2-sided']]$Light1>1] <- NA
+results.wetland[['2-sided']]$Light2[results.wetland[['2-sided']]$Light2>1] <- NA
+
+results.wetland[['2-sided']]$Moist1[results.wetland[['2-sided']]$Moist1>1] <- NA
+results.wetland[['2-sided']]$Moist2[results.wetland[['2-sided']]$Moist2>1] <- NA
+
+results.wetland[['2-sided']]$pH1[results.wetland[['2-sided']]$pH1>1] <- NA
+results.wetland[['2-sided']]$pH2[results.wetland[['2-sided']]$pH2>1] <- NA
+
+results.wetland[['2-sided']]$Nitrogen1[results.wetland[['2-sided']]$Nitrogen1>1] <- NA
+results.wetland[['2-sided']]$Nitrogen2[results.wetland[['2-sided']]$Nitrogen2>1] <- NA
+
+```
+
+```{r, include = F}
+load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/results.wetland.RData")
+```
+
+```{r}
+head(results.wetland[['2-sided']])
+```
+
+#### CONTINUE HERE ####
 

### 2023-04-13 R/functionalPlantIndicators/1 data handling.R

@@ -682,11 +682,10 @@ unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) &
 
 #### matching with NiN ecosystem types - wetlands ####
 # NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
-NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[19:23])]   # Continentality, Light, Moisture, Soil_reaction_pH, Nitrogen
+NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[15:18])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
 NiN.wetland[1,]
 names(NiN.wetland)
 
-#NiN.wetland <- NiN.wetland[,c(1:41,46:48,59:69)]
 cbind(colnames(NiN.wetland),
       c("",
         'V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
@@ -702,30 +701,30 @@ cbind(colnames(NiN.wetland),
         'V2-C1a','V2-C1b',
         'V2-C2a','V2-C2b',
         'V2-C3a','V2-C3b',
-        "V4-C4","V4-C5",
         "V4-C2","V4-C3",
+        "","",
         'V8-C1','V8-C2','V8-C3',
         rep("",10),
-        rep("",5)
+        rep("",4)
       )
 )
 
-NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:63)]
-colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
-                              'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
-                              'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
-                              'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
-                              'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
-                              'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
-                              'V3-C2','V1-C5',
-                              'V1-C6a','V1-C6b',
-                              'V1-C7a','V1-C7b',
-                              'V1-C8a','V1-C8b',
[…]

### 2023-04-13 R/functionalPlantIndicators/5 scaled value calculations_wetland.R

@@ -61,14 +61,14 @@ scal.2 <- function() {
 
 
 unique(ANO.geo$hovedtype_rute) # NiN types in data
-unique(substr(wetland.ref.val$grunn,1,2)) # NiN types in reference
+unique(substr(wetland.ref.cov.val$grunn,1,2)) # NiN types in reference
 #### creating dataframe to hold the results for wetlands ####
 # all ANO points
 nrow(ANO.geo)
 # all wetland ANO points
 nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),])
 # all wetland ANO points with a NiN-type represented in the reference
-nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.val$grunn,1,2)),])
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.cov.val$grunn,1,2)),])
 # ok, we'll be losing 70 (our of 1349) that are not covered by the reference
 ANO.wetland <- ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),]
 
@@ -145,43 +145,15 @@ for (i in 1:nrow(ANO.wetland) ) {  #
 
 
     # if the ANO.hovedtype exists in the reference
-    if (ANO.wetland$hovedtype_rute[i] %in% unique(substr(wetland.ref.val$grunn,1,2)) ) {
+    if (ANO.wetland$hovedtype_rute[i] %in% unique(substr(wetland.ref.cov.val$grunn,1,2)) ) {
       
       # if there is any species present in current ANO point  
       if ( length(ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),'Species']) > 0 ) {
         
-        
-          # Continentality
-          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Continentality','kartleggingsenhet_1m2')]
-          results.wetland[['original']][i,'richness'] <- nrow(dat)
-          dat <- dat[!is.na(dat$Continentality),]
-          
-          if ( nrow(dat)>0 ) {
-            
[…]

### 2023-04-13 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -1,3 +1,23 @@
+#### scaled values by NiN-hovedtype ####
+res <- results.wetland[['2-sided']]
+levels(res$hovedtype_rute)
+levels(res$hovedtype_rute) <- c("V1",NA,NA,NA,"V2","V3","V4",NA,"V8",NA)
+x11()
+par(mfrow=c(2,4))
+with(res, plot(hovedtype_rute,Light1) )
+with(res, plot(hovedtype_rute,Moist1) )
+with(res, plot(hovedtype_rute,pH1) )
+with(res, plot(hovedtype_rute,Nitrogen1) )
+with(res, plot(hovedtype_rute,Light2) )
+with(res, plot(hovedtype_rute,Moist2) )
+with(res, plot(hovedtype_rute,pH2) )
+with(res, plot(hovedtype_rute,Nitrogen2) )
+
+
+
+
+
+
 #### distribution comparison, reference vs. data ####
 summary(results.wetland[['original']]$kartleggingsenhet_1m2) #V1-C1 & V1-C5 are the most abundant ecosystem types
 

### 2023-04-14 index.Rmd
build_book_test_license

@@
 url: https://ninanor.github.io/ecosystemCondition/index.html
-# cover-image: path to the social sharing image like images/cover.jpg
+cover-image: images/Picture1.png
 description: |
@@
 
-On this site you may view the nitty gritty details for how certain indicators for ecosystem condition for Norway has been calculated.
+On this site you may view the details for how certain indicators for ecosystem condition for Norway have been calculated.
 
@@
 The html version of the book will be in the `docs` folder.
+
+## Citation {.unnumbered}
+
+To cite this book please use the following citation:
+
+Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]
+
+## License
+
+The book is licensed under a [CC By 4.0 licence](https://creativecommons.org/licenses/by/4.0/) which means that you can share and adapt the material
+for any purpose. However, you must give appropriate credit, provide a link to the license, and indicate if changes were made. You can not add additional restrictions on the material even when adapted.

### 2023-04-14 alien_species.Rmd
the edited file with new name (without "DRAFT")

--- 2023-03-23 DRAFT_alien_species.Rmd
@@
 A related variable has been used previously for [alpine](https://ninanor.github.io/IBECA/faktaark.html#Frav%C3%A6r_av_fremmede_arter) and forest ecosystems in Norway.
-These indicators were based on ANO, a national monitoring program, where alien species cover was estimated to the closest whole %.
+These indicators were based on ANO, a national monitoring program, where alien species cover was estimated to the closest whole percent.
 The upper and lower reference levels that was used back then was 0 or 100% of the species composition made up of alien species, respectively.
@@
 
-This time we are using the [nature type mapping dataset](#naturtype) which records the assumed _effect_ of alien species, rather than a direct measure of abundance such as % cover.
+This time we are using the [nature type mapping dataset](#naturtype) which records the assumed _effect_ of alien species, rather than a direct measure of abundance such as percentage cover.
 The recording scale is called R7 (Fig.
-\@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with %cover recorded in ANO (Fig.
+\@ref(fig:R7)) and is an ordinal scale which has a non-linear relationship with percentage cover recorded in ANO (Fig.
 \@ref(fig:fremmedart-prosentVSinnslag)).
@@
 Here I base the indicator on the R7 scale, since this is, presumably, the bigger data set.
-I therefore convert %cover into R7 classes based on what I can interpret from Fig.
+I therefore convert percentage cover into R7 classes based on what I can interpret from Fig.
 \@ref(fig:fremmedart-prosentVSinnslag).
@@
 
-```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with %cover (y-axis), which is how alien species is recorded in ANO.
+```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with percentage cover (y-axis), which is how alien species is recorded in ANO.
 I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here equals an R7 value of 1.
@@
 Since the nature type data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
-We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure intersected with municipality borders, and we say that if we have more than _n_ data points than this field data is representative inside the entire HIA.
+We delineate _homogeneous impact areas_ (HIA) based on four classes of increasing infrastructure intersected with municipality borders, and we say that if we have more than _n_ data points then this field data is representative inside the entire HIA.
 In [a related indicator](#slitasje-og-kjørespor) we used the accounting areas to define the HIA (5 accounting areas in Norway), in addition to the infrastructure classes.
@@
 
-10. TO DO: Prepare ecosystem delineation maps and us ethese to mask the extrapolated indicator maps
+10. TO DO: Prepare ecosystem delineation maps and use these to mask the extrapolated indicator maps
 
@@
 The variables are recorded on a unitless seven-step ordinal scale (Fig.
-\@ref(fig:R7)) or a as a % cover of alien species.
+\@ref(fig:R7)) or a as a percentage cover of alien species.
 
@@
 **Funksjonelt viktige arter og strukturer** (Functionally important species and structures).
-This emphasises the negaivte effect that alien species has on for example supressing or excluding native species.
+This emphasises the negative effect that alien species have on for example supressing or excluding native species.
 In some cases, alien species can also affect nutrient cycling or recruitment.
@@
 The reference condition is one with minimal negative human impact. This
-is also true for semi-natural ecosystems. In a sense, the reference condition is pre 1800s,
-since alien species are defined as species established post 1800.
+is also true for semi-natural ecosystems. In a sense, the reference condition is pre-1800,
+since the Alien Species List of Norway only contains species established after 1800.
 
@@
 distribution of area weighted means. This uncertainty is different from 
-the spatial variation which we could get more straight forward without 
+the spatial variation which we could get more straightforward without 
 bootstrapping. When aggregating a second time,
@@
 errors, and sample _n_ times from these and combine the resamples into an
-a new, area weighted, distribution. The errors for the accounting areas
-thus represents both the spatial variation, and the precision, of the
+a new, area-weighted, distribution. The errors for the accounting areas
+thus represents both the spatial variation and the precision of the
 indicator values within the accounting areas.
@@
 
-We are only interested in a mapping units that include our target variables. The variable _fremmedartsinnslag_ is coded as 7FA.
+We are only interested in mapping units that include our target variables. The variable _fremmedartsinnslag_ (presence of alien species) is coded as 7FA.
 
@@
 
-It appears most localities are in good condition, and most localities have no alien species at all.
+It appears most localities are in good condition; actually, most localities have no alien species at all.
 
@@
 
-This variable is also recorded in
+This variable (7FA) is also recorded in
 [GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
@@
 Arealrepresentativ Naturovervåking (ANO) consist of 1000 systematically
-placed locations in Norway, each with 18 sample points. In each sample point a
-circle of 250 m2 is visualised, and the main ecosystem is recorded.
-
-The %cover of alien vascular plant is estimated, independent on main ecosystem type. 
-In contrast to the nature type data set, ANO only looks at vascular plants, and then also only the three strictest categories
-in the alien plant list (SE, HI, PH). It is therefore not completely unproblematic to combine these to data sets,
+placed locations in Norway, each with 18 sample points. In each sample point, a
+circle of 250 m^2^ is inspected, and the main ecosystem is recorded.
+
+The percentage cover of alien vascular plant is estimated, irrespective of main ecosystem type. 
+In contrast to the nature type data set, ANO only looks at vascular plants, and then also only the three highest impact categories
+of the Alien Species List (SE, HI, PH). It is therefore not completely unproblematic to combine these to data sets,
 but we will do so anyway. We expect plants to be the by far most common species group to influence the 7FA variable as well.
-And similarly, we expect the three strictest  alien plant categories to contain the most aggressive species which will
-drive the 7FA variable more so than the other categories.
+And similarly, we expect the three highest impact categories to contain the most widespread species, which will
+drive the 7FA variable more than the other categories.
 
@@
 
-```{r ano-alien-dist, fig.cap="Distribution of %cover of alien plants variable in the ANO data set"}
+```{r ano-alien-dist, fig.cap="Distribution of percentage cover of alien plants variable in the ANO data set"}
 ggplot(ano, aes(x = fa_total_dekning))+
@@
   theme_bw(base_size = 12)+
-  labs(x = "% cover alien vascular plants") +
+  labs(x = "Percentage cover alien vascular plants") +
   facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
@@
 
-The ANO localities seemingly have less alien plants in them than the nature type localities. This can be du to sampling biases in the latter.
-
-Now I will convert these %cover values into 7FA classes. This is a critical step, since there is no one-to-one mapping here, 
+The ANO localities seemingly have less alien plants in them than the nature type localities. This can be due to sampling biases in the latter.
+
+Now I will convert these percentage cover values into 7FA classes. This is a critical step, since there is no one-to-one mapping here, 
 but rather there is considerable overlap, as can be seen in Fig. \@ref(fig:fremmedart-prosentVSinnslag).
@@
   theme_bw(base_size = 12)+
-  labs(x = "% cover alien vascular plants\n converted to the R7 scale") +
+  labs(x = "Percentage cover alien vascular plants\n converted to the R7 scale") +
   facet_wrap(.~hovedoekosystem_punkt, scales="free_y")
@@
 Because later we will rasterize these data, we will  convert the points to polygons. I
-use the area column to calculate a radii that gives that area.
+use the area column to calculate a radius that gives that area.
 
@@
 So the ANO data is not very important here, but it can become more
-important in the future so good to have them included in the workflow.
+important in the future, so good to have them included in the workflow.
 
@@
 we can assume them to be representative inside *homogeneous ecological
-areas,* or *Homoegeneous Impact Areas* (HIAs). That's where the
+areas,* or *Homogeneous Impact Areas* (HIAs). That's where the
 infrastructure index comes in. Here's the plan:
@@
 4.  calculate an indicator value for a region (accounting area) by doing
-    a weighted average based of the relative area of ecosystem
+    a weighted average based on the relative area of ecosystem
     occurrences. Errors should be carried somehow, perhaps via weighted
@@
 There is no point yet making this a time series, and I will
-assign all the indicator value to the average year of the data.
+assign all the indicator values to the average year of the data.
 
@@
 Also, with more nature type data accumulating over time, the data coverage will increase.
-Therefore I think this indicator is worth while, and that this is considerably better than the alternative to discard the entire data set.
+Therefore I think this indicator is worthwhile, and that this is considerably better than the alternative to discard the entire data set.
 
@@
 
-Alien species tend to invade from the south. Let us see if there is an effect of latitude.
+Alien species can be expected to decrease towards the north. Let us see if there is an effect of latitude.
 
@@
 
-## Masking with an ecosystem delineation - example
+## Masking with an ecosystem delineation - example {#masking-example}
 
@@
 
-Then I can import and crop the ecosystem delineation map.
+Then I can import and crop the ecosystem delineation map (EDM).
 
@@
 
-Then i mask to remove areas that are not actually watlands.
+Then I mask to remove areas that are not actually wetlands.
 

### 2023-04-14 DRAFT_functional plant indicators_wetland.Rmd
fpi wetland result analyses continued

@@
 ```
+#### Scaled value analyses
+
+```{r}
+### scaled values by NiN-hovedtype
+res <- results.wetland[['2-sided']]
+
+res2 <- 
+  res %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = TRUE
+  )
+head(res2)
+
+ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) + 
+  geom_boxplot()
+
+ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+  geom_boxplot(width=0.2, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4)
+
+```
 

### 2023-04-14 R/functionalPlantIndicators/5 scaled value calculations_wetland.R

@@ -329,6 +329,9 @@ hist(results.wetland[['2-sided']]$Nitrogen2,breaks=40)
 
 
 
+
+
+
 write.table(results.wetland[['scaled']], file='output/scaled data/results.wetland_scaled.txt',
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 write.table(results.wetland[['non-truncated']], file='output/scaled data/results.wetland_non-truncated.txt',

### 2023-04-14 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -1,3 +1,36 @@
+#### some data handling again ####
+
+res.wetland <- results.wetland[['2-sided']]
+
+
+# add geometry
+
+
+res.wetland <-
+  res.wetland %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = FALSE
+  )
+
+res.wetland <- 
+  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
+                               pivot_longer(
+                                 cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+                                 names_to = NULL,
+                                 values_to = "original",
+                                 values_drop_na = FALSE
+                               ) %>%
+                               pull(original)
+  )
+
+
+
+
+head(res.wetland[,70:75])
+
 #### scaled values by NiN-hovedtype ####
 res <- results.wetland[['2-sided']]
 
@@ -15,9 +48,17 @@ ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) +
   geom_boxplot()
 
 ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) + 
-  geom_boxplot() +
-  facet_wrap(~fp_ind)
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+  geom_boxplot(width=0.2, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4)
 
+
+geom_point(
+    data = filter(Example_data, Species_loc_n == 1), 
+    aes(color = Location), 
+    show.legend = FALSE
+  ) +
 
 levels(res$hovedtype_rute)
 levels(res$hovedtype_rute) <- c("V1",NA,NA,NA,"V2","V3","V4",NA,"V8",NA)

### 2023-04-14 DRAFT_functional plant indicators_wetland.Rmd
wetland fpi result analyses and maps

--- 2023-04-13 DRAFT_functional plant indicators_wetland.Rmd
@@
 ```
+#### Scaled value analyses
+
+```{r}
+#### plotting scaled values by main ecosystem type ####
+## continuing with 2-sided
+res.wetland <- results.wetland[['2-sided']]
+
+# make long version of the scaled value part
+res.wetland <-
+  res.wetland %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = TRUE
+  )
+
+# summarizing the indicator scores
+res.wetland %>%
+  group_by(fp_ind) %>%
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
+
+# making the plot
+ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+#  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value")
+
+#### getting scaled AND original values ####
+res.wetland <- results.wetland[['2-sided']]
+
+# make long version of the scaled value part
+res.wetland <-
+  res.wetland %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = TRUE
+  )
+
+# add original values as well
+res.wetland <- 
+  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
+                               pivot_longer(
+                                 cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+                                 names_to = NULL,
+                                 values_to = "original",
+                                 values_drop_na = FALSE
+                               ) %>%
+                               pull(original)
+  )
+
+head(res.wetland[,70:76])
+
+#### scaled value maps ####
+# keep wide format and add geometry again
+res.wetland <- results.wetland[['2-sided']]
+st_geometry(res.wetland) <- st_geometry(ANO.wetland)
+
+nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+  st_as_sf()
+
+## scaled value maps
+# Moist1 (lower indicator)
+tm_shape(nor) +
+  tm_fill('GID_0', labels="", title="") + #tm_borders() +
+  tm_shape(res.wetland) +
+  tm_dots('Moist1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
+  tm_layout(main.title = "Moisture index (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
+                labels = c("0.4", "...", "...", 
+                           "...", "1", "NA"),
+                title = "index values")
+
+# Moist2 (upper indicator)
+tm_shape(nor) +
+  tm_fill('GID_0', labels="", title="") + #tm_borders() +
+  tm_shape(res.wetland) +
+  tm_dots('Moist2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
+  tm_layout(main.title = "Moisture index (upper), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
+                labels = c("0.4", "...", "...", 
+                           "...", "1", "NA"),
+                title = "index values")
+
+```
 

### 2023-04-14 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -1,20 +1,59 @@
-#### some data handling again ####
-
+#### plotting scaled values by main ecosystem type ####
+## continuing with 2-sided
 res.wetland <- results.wetland[['2-sided']]
 
-
-# add geometry
-
-
+# make long version of the scaled value part
 res.wetland <-
   res.wetland %>% 
   pivot_longer(
     cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
     names_to = "fp_ind",
     values_to = "scaled_value",
-    values_drop_na = FALSE
+    values_drop_na = TRUE
   )
 
+
+# summarizing the indicator scores
+res.wetland %>%
+  group_by(fp_ind) %>%
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
+
+
+# making the plot
+ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+#  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value") 
+
+
+
+
+
+
+
+
+
+
+
+
+#### getting scaled AND original values ####
+res.wetland <- results.wetland[['2-sided']]
+
+# make long version of the scaled value part
+res.wetland <-
+  res.wetland %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = TRUE
+  )
+
+# add original values as well
 res.wetland <- 
   res.wetland %>% add_column(original = results.wetland[['original']] %>% 
                                pivot_longer(
@@ -26,52 +65,50 @@ res.wetland <-
                                pull(original)
   )
 
+head(res.wetland[,70:76])
+
+
+#### scaled value maps ####
+# keep wide format and add geometry again
[…]

### 2023-04-17 DRAFT_functional plant indicators_wetland.Rmd
tmap wetland analyses

--- 2023-04-13 DRAFT_functional plant indicators_wetland.Rmd
@@
 library(tmap)
+library(tmaptools)
 library(raster)
@@
 ```
+#### Scaled value analyses
+
+```{r}
+#### plotting scaled values by main ecosystem type ####
+## continuing with 2-sided
+res.wetland <- results.wetland[['2-sided']]
+
+# make long version of the scaled value part
+res.wetland <-
+  res.wetland %>% 
+  pivot_longer(
+    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+    names_to = "fp_ind",
+    values_to = "scaled_value",
+    values_drop_na = TRUE
+  )
+
+# summarizing the indicator scores
+res.wetland %>%
+  group_by(fp_ind) %>%
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
+
+# making the plot
+ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+#  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value") 
+```
+
+We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
+```{r}
+#### scaled value maps ####
+# keep wide format and add geometry again
+res.wetland2 <- results.wetland[['2-sided']]
+st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
+
+nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+# change region names to something R-friendly
+reg$region
+reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
+
+regnor <- st_intersection(reg,nor)
+
+## scaled value maps
+# pH1 (lower indicator)
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="") + tm_borders() +
+  tm_shape(res.wetland2) +
+  tm_dots('pH1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
+  tm_layout(main.title = "Moisture index (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
+                labels = c("0.4", "...", "...", 
+                           "...", "1", "NA"),
+                title = "index values")
+
+```
+...or by showing the region-wise means and standard deviations
+```{r}
+regnor <- regnor %>%
+  mutate(Light1.reg.mean = c(mean(res.wetland2$Light1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Light1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Light1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Light1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Light1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Light2.reg.mean = c(mean(res.wetland2$Light2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Light2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Light2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Light2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Light2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Moist1.reg.mean = c(mean(res.wetland2$Moist1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Moist1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Moist1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Moist1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Moist1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Moist2.reg.mean = c(mean(res.wetland2$Moist2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Moist2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Moist2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Moist2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Moist2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         pH1.reg.mean = c(mean(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         pH2.reg.mean = c(mean(res.wetland2$pH2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$pH2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$pH2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$pH2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$pH2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Nitrogen1.reg.mean = c(mean(res.wetland2$Nitrogen1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Nitrogen2.reg.mean = c(mean(res.wetland2$Nitrogen2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$Nitrogen2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+
+         Light1.reg.sd = c(sd(res.wetland2$Light1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             sd(res.wetland2$Light1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             sd(res.wetland2$Light1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             sd(res.wetland2$Light1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             sd(res.wetland2$Light1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Light2.reg.sd = c(sd(res.wetland2$Light2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             sd(res.wetland2$Light2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             sd(res.wetland2$Light2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             sd(res.wetland2$Light2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             sd(res.wetland2$Light2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Moist1.reg.sd = c(sd(res.wetland2$Moist1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             sd(res.wetland2$Moist1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             sd(res.wetland2$Moist1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             sd(res.wetland2$Moist1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             sd(res.wetland2$Moist1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Moist2.reg.sd = c(sd(res.wetland2$Moist2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             sd(res.wetland2$Moist2[res.wetland2$region=="Central Norway"],na.rm=T),
+                             sd(res.wetland2$Moist2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             sd(res.wetland2$Moist2[res.wetland2$region=="Western Norway"],na.rm=T),
+                             sd(res.wetland2$Moist2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         pH1.reg.sd = c(sd(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         pH2.reg.sd = c(sd(res.wetland2$pH2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                          sd(res.wetland2$pH2[res.wetland2$region=="Central Norway"],na.rm=T),
+                          sd(res.wetland2$pH2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                          sd(res.wetland2$pH2[res.wetland2$region=="Western Norway"],na.rm=T),
+                          sd(res.wetland2$pH2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Nitrogen1.reg.sd = c(sd(res.wetland2$Nitrogen1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen1[res.wetland2$region=="Central Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen1[res.wetland2$region=="Western Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+         Nitrogen2.reg.sd = c(sd(res.wetland2$Nitrogen2[res.wetland2$region=="Northern Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen2[res.wetland2$region=="Central Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen2[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen2[res.wetland2$region=="Western Norway"],na.rm=T),
+                                sd(res.wetland2$Nitrogen2[res.wetland2$region=="Southern Norway"],na.rm=T)),
+
+         Light1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Light1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Light1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Light1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Light1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Light1),])),
+         Light2.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Light2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Light2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Light2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Light2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Light2),])),
+         Moist1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Moist1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Moist1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Moist1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Moist1),]),
+                             nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Moist1),])),
+         Moist2.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Moist2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Moist2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Moist2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Moist2),]),
+                             nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Moist2),])),
+         pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),])),
+         pH2.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH2),]),
+                          nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH2),]),
+                          nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH2),]),
+                          nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH2),]),
+                          nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH2),])),
+         Nitrogen1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Nitrogen1),]),
+                                nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Nitrogen1),]),
+                                nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Nitrogen1),]),
+                                nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Nitrogen1),]),
+                                nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Nitrogen1),])),
+         Nitrogen2.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$Nitrogen2),]),
+                                nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$Nitrogen2),]),
+                                nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$Nitrogen2),]),
+                                nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$Nitrogen2),]),
+                                nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$Nitrogen2),]))
+         )
+
+## scaled value maps
+# pH1 (lower indicator)
+tm_shape(regnor) +
+  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
+
+tm_shape(regnor) +
+  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))
+
+```
 

### 2023-04-17 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -70,19 +70,26 @@ head(res.wetland[,70:76])
 
 #### scaled value maps ####
 # keep wide format and add geometry again
-res.wetland <- results.wetland[['2-sided']]
-st_geometry(res.wetland) <- st_geometry(ANO.wetland)
-
+res.wetland2 <- results.wetland[['2-sided']]
+st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
 
 nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
-  st_as_sf()
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+# change region names to something R-friendly
+reg$region
+reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
 
+regnor <- st_intersection(reg,nor)
 
 ## scaled value maps
 # Moist1 (lower indicator)
-tm_shape(nor) +
-  tm_fill('GID_0', labels="", title="") + #tm_borders() +
-  tm_shape(res.wetland) +
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="") + tm_borders() +
+  tm_shape(res.wetland2) +
   tm_dots('Moist1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
   tm_layout(main.title = "Moisture index (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
   tm_add_legend(type = "fill", 
@@ -92,9 +99,9 @@ tm_shape(nor) +
                 title = "index values")
 
 # Moist2 (upper indicator)
-tm_shape(nor) +
-  tm_fill('GID_0', labels="", title="") + #tm_borders() +
-  tm_shape(res.wetland) +
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="") + tm_borders() +
+  tm_shape(res.wetland2) +
   tm_dots('Moist2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
[…]

### 2023-04-20 DRAFT_functional plant indicators_wetland.Rmd
fix wetland maps and start writing explanatory text

--- 2023-04-14 DRAFT_functional plant indicators_wetland.Rmd
@@
 ## Introduction
-<!-- Text here -->
+Functional plant indicators can be used to describe the functional signature of plant communities.
+The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at.
+For instance, using an indicator for moisture on would find a functional signature with higher moisture values for plant communities in mires compared to e.g.
+grasslands or forests.
+Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
+Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO for wetland ecosystems.
+We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen.
+These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity.
 
@@
 ```{r}
-### scaled values by NiN-hovedtype
-res <- results.wetland[['2-sided']]
-
-res2 <- 
-  res %>% 
+#### plotting scaled values by main ecosystem type ####
+## continuing with 2-sided
+res.wetland <- results.wetland[['2-sided']]
+
+# make long version of the scaled value part
+res.wetland <-
+  res.wetland %>% 
   pivot_longer(
@@
   )
-head(res2)
-
-ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) + 
-  geom_boxplot()
-
-ggplot(res2, aes(x=hovedtype_rute, y=scaled_value, fill=fp_ind)) + 
+
+# summarizing the indicator scores
+res.wetland %>%
+  group_by(fp_ind) %>%
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
+
+# making the plot
+ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
   geom_hline(yintercept=0.6, linetype="dashed") + 
   geom_violin() +
-  geom_boxplot(width=0.2, color="grey") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4)
+#  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value") 
+```
+
+We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
+```{r}
+#### scaled value maps ####
+# keep wide format and add geometry again
+res.wetland2 <- results.wetland[['2-sided']]
+st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
+
+nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = crs(ANO.geo))
+# change region names to something R-friendly
+reg$region
+reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
+
+regnor <- st_intersection(reg,nor)
+
+## scaled value maps
+# pH1 (lower indicator)
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(res.wetland2) +
+  tm_dots('pH1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
+  tm_layout(main.title = "pH (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
+                labels = c("0.0 - 0.2", "0.2 - 0.4", "0.4 - 0.6", 
+                           "0.6 - 0.8", "0.8 - 1.0", "NA"),
+                title = "index values")
+
+```
+...but here the colors and values of the data points on the map are hard to make out.
+Alternatively we can calculate and show the region-wise means and standard deviations
+```{r}
+
+res.wetland2 = st_join(res.wetland2, regnor, left = TRUE)
+
+regnor <- regnor %>%
+  mutate(
+    pH1.reg.mean = c(mean(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+    pH1.reg.sd = c(sd(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T))
+    )
+
+## scaled value maps for pH1 (lower indicator)
+# mean
+tm_shape(regnor) +
+  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
+
+# sd
+tm_shape(regnor) +
+  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))
 

### 2023-04-20 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -86,28 +86,46 @@ reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Nor
 regnor <- st_intersection(reg,nor)
 
 ## scaled value maps
+
+# pH1 (lower indicator)
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(res.wetland2) +
+  tm_dots('pH1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
+  tm_layout(main.title = "pH (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
+                labels = c("0.0 - 0.2", "0.2 - 0.4", "0.4 - 0.6", 
+                           "0.6 - 0.8", "0.8 - 1.0", "NA"),
+                title = "index values")
+
+
+
 # Moist1 (lower indicator)
 tm_shape(regnor) +
-  tm_fill('GID_0', labels="", title="") + tm_borders() +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
   tm_shape(res.wetland2) +
   tm_dots('Moist1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
   tm_layout(main.title = "Moisture index (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
   tm_add_legend(type = "fill", 
-                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
-                labels = c("0.4", "...", "...", 
-                           "...", "1", "NA"),
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),'grey'),
+                labels = c("0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
+                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", "NA"),
                 title = "index values")
 
 # Moist2 (upper indicator)
 tm_shape(regnor) +
[…]

### 2023-04-20 DRAFT_functional plant indicators_wetland.Rmd
adding more explanation text to the wetland markdown

@@
 ## Introduction
-Functional plant indicators can be used to describe the functional signature of plant communities.
-The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at.
+Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
+The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf.
+Töpper et al.
+2018).
 For instance, using an indicator for moisture on would find a functional signature with higher moisture values for plant communities in mires compared to e.g.
@@
 Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
-Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO for wetland ecosystems.
+Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
+2019) for wetland ecosystems.
 We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen.
-These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity.
+These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al.
+2020).
+This plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
+2021, Töpper & Jakobsson 2021).
 
 ## About the underlying data
-<!-- Text here -->
+In the 'plant functional indicator' project for wetlands, We use three sets of data for building indicators for ecological condition
+- as testing data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
+- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
+- Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)
+
+(1) ANO monitoring data:
+ANO stands for 'areal-representativ naturovervåking', i.e.
+'area representative nature monitoring'.
+Within a 5-year cycle, 1000 sites are  randomly distributed across mainland Norway.
+Each ANO site spans a 500 x 500 m grid cell and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
+For the vegetation analyses, the cover of each vascular plant species in the plot is recorded.
+Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
+In the wetland-analysis in this document, we only use the plots which were classified as on of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
+In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
+
+(2) NiN reference data:
+The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species.
+In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1 > 1/32, 2 = 1/32 - 1/8 , 3 = 1/8 - 3/8, 4 = 3/8 - 4/5, 5 = 3/8 - 4/5 + dominance, 6 = 4/5 -1 ).
+For the purpose of this project these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
+
+(3) The Swedish plant indicator set published by Tyler et al.
+(2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well.
+From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub-/tree encroachment, drainage, and pollution.
 
 ### Representativity in time and space
-<!-- Text here -->
-
-### Original units
-<!-- Text here -->
+For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country.
+As wetlands occur more often in certain regions of Norway than in others the amount of plots and sites is not equal among the five regions ecological condition is assessed for (Northern Norway, Central Norway, Eastern Norway, Western Norway, Southern Norway).
+The in total 1351 plots are distributed across regions in the following way:
+
+Northern Norway: 416
+Central Norway:  362
+Eastern Norway:  295
+Western Norway:  151
+Southern Norway: 127
 
 ### Temporal coverage
-<!-- Text here -->
-
-### Aditional comments about the dataset
-<!-- Text here -->
-
-## Ecosystem characteristic
-### Norwegian standard
-<!-- Text here -->
-
-### <!-- Text here -->
-<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->
+The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme.
+Thus there is no actual time series to these data and the indicator evaluation does therefore not include any temporal analyses.
 
@@
 ### Reference state
-<!-- Text here -->
+The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al.
+2018).
+By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
+These distributions are calculated for ecosystem sub-types (grunntyper or kartleggingsenheter at a 1:5000 mapping scale) within the main wetland types in NiN.
+A more extensive discussion on the use of reference communities can be found in Jakobsson et al.
+(2020).
 
 ### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-<!-- Text here -->
+In this analysis, we derive scaling values from statistical (non-parametric here) distributions (see Jakobsson et al.
+2010).
+For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution.
+As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al.
+2022), and deviation from the optimal state thus may occur in both direction (e.g.
+indicating too low or too high pH) we need to define two threshold values for good ecological condition as well as both a minimum and maximum value.
+In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e.
+its 0.025 and 0.975 quantiles.
+The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator.
+For details on the scaling principles in IBECA, please see Töpper & Jakobsson 2021.
 
 ## Uncertainties
-<!-- Text here -->
+We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
 
 ## References
-<!-- Text here -->
+Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185
+
+Framstad, E., Kolstad, A.
+L., Nybø, S., Töpper, J.
+& Vandvik, V.
+2022.
+The condition of forest and mountain ecosystems in Norway.
+Assessment by the IBECA method.
+NINA Report 2100.
+Norwegian Institute for Nature Research.
+
+Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E.
+(2020).
+Towards a systematics of ecodiversity: The EcoSyst framework.
+Global Ecology and Biogeography, 29(11), 1887-1906.
+doi:10.1111/geb.13164
+
+Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik.
+V., Velle, L.G., Aarrestad, P.A.
+& Nybø, S.
+2020.
+Setting reference levels and limits for good ecological condition in terrestrial ecosystems.
+Insights from a case study based on the IBECA approach.
+Ecological Indicators 116: 106492.
+
+Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A.
+& Nybø, S.
+2021.
+An index-based assessment of ecological condition and its links to international frameworks.
+Ecological Indicators 124: 107252.
+
+Tingstad, L., Evju, M., Sickel, H., & Töpper, J.
+2019.
+Utvikling av nasjonal arealrepresentativ naturovervåking (ANO).
+Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag.
+NINA Rapport 1642.
+
+Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967. Norwegian Institute for Nature Research.
+
+Töpper, J., Velle, L.G.
+& Vandvik, V.
+2018.
+Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition).
+NINA Report 1529b.
+Norwegian Institute for Nature Research.
+
+Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923
 
@@
 
-Plant indicators from Tyler et al. 2021
+Plant indicators from Tyler et al. (2021)
 ```{r, echo = F}
@@
 - with a given number of iterations (iter),
-- with species given a certain minimum abundance ocurring in all bootstraps (obl), and
-- with a given resampling ratio of the original species list (rat)
-- in every bootstrap iteration the abundance of the sampled species can be 
-- randomly changed by +/-1 if wished (var.abun)
+- with species given a certain minimum abundance occurring in all bootstraps (obl), and
+- with a given re-sampling ratio of the original species list (rat)
+- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)
 
@@
     dat <- cbind(sp,abun[,j],ind)
-    dat <- dat[dat[,2]>0,]            # only species that are present in the ecoystem
-    dat <- dat[!is.na(dat[,3]),]      # only species that have Ellenberg values
+    dat <- dat[dat[,2]>0,]            # only species that are present in the ecosystem
+    dat <- dat[!is.na(dat[,3]),]      # only species that have indicator values
 
@@
 # 2nd-46th column is the abundances of sp in different ecosystem types
-# 47th-51st column is the indicator values of the respective sp
+# 47th-50th column is the indicator values of the respective species
 # we choose 1000 iterations
-# species with abundance 1 (i.e. 100%, must be included in each sample)
-# each sample resamples 2/3 of the number of species
-# the abundance of the resampled species may vary (see bootstrap function for details)
+# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
+# each sample re-samples 2/3 of the number of species
+# the abundance of the re-sampled species may vary (see bootstrap function for details)
 wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
@@
 
-This results in a list with a slot for every selected indicator, and in every slot there's as many columns as there are NiN species lists and as many row as there were iterations in the bootstrap.
-Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here we define things in the following way:
+This results in an R-list with a slot for every selected indicator, and in every slot there's a data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.
+Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:
 
@@
 
-Once testing data (ANO), and reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution.
-
-Here is the scaling function (with and without truncation at 1)
+Once testing data (ANO), and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution.
+Note that we scale each ANO-plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
+Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type.
+An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.
+
+Here is the scaling function
 ```{r}
@@
 
-## scaling function without truncation
-scal.2 <- function() {
-  # place to hold the result
-  x <- numeric()
-  if (maxmin < ref) {
-    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
-    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
-    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
-    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
-    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
-    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
-    # value equal or lower than maxmin
-    if (val <= maxmin) {x <-0}
-  } else {
-    # values <= the reference value estimated from the linear relationship for lim < x < ref (line below)
-    if (val <= ref) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
-    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
-    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
-    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
-    if (val > lim & val < maxmin) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
-    # value equal og larger than maxmin
-    if (val >= maxmin) {x <-0}
-  }
-  return(x)
-
-}
-```
-
-We then can prepare a list of dataframes to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
+```
+
+We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
 ```{r, echo = T, eval = F}
@@
 
-```{r}
+We can calculate simple summary statistics like means, standard deviations, and number of observations (note that there is both a lower and an upper indicator for each of the four plant functional indicators)
+
+```{r, message = FALSE}
 #### plotting scaled values by main ecosystem type ####
@@
   group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value))
+
+```
+
+And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
+```{r}
 
@@
 We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
-```{r}
+```{r, message = FALSE}
 #### scaled value maps ####

### 2023-04-20 R/functionalPlantIndicators/1 data handling.R

@@ -731,7 +731,7 @@ head(NiN.wetland)
 
 # translating the abundance classes into %-cover
 coverscale <- data.frame(orig=0:6,
-                         cov=c(0,1/32,1/8,3/8,0.6,4/5,1)
+                         cov=c(0, 1/32 ,1/8, 3/8, 0.6, 4/5, 1)
 )
 
 NiN.wetland.cov <- NiN.wetland

### 2023-04-20 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -293,21 +293,28 @@ tm_shape(regnor) +
 tm_shape(regnor) +
   tm_polygons(col="Light1.reg.mean", title="Light (upper)", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))
 
-  
-  tm_fill('GID_0', labels="", title="") + tm_borders() +
-  tm_shape(res.wetland) +
-  tm_dots('Moist2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
-  tm_layout(main.title = "Moisture index (upper), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
-  tm_add_legend(type = "fill", 
-                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
-                labels = c("0.4", "...", "...", 
-                           "...", "1", "NA"),
-                title = "index values")
 
 
 
+#### how many observations are there? ####
+# total ANO
+length(unique(ANO.geo$ano_flate_id))
+length(unique(ANO.geo$ano_punkt_id))
+# total wetland
+length(unique(res.wetland2$ano_flate_id))
+length(unique(res.wetland2$ano_punkt_id))
+# by region
+length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Northern Norway"]))
+length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Central Norway"]))
+length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Eastern Norway"]))
+length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Western Norway"]))
+length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Southern Norway"]))
 
-
+length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Northern Norway"]))
+length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Central Norway"]))
+length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Eastern Norway"]))
+length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Western Norway"]))
+length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Southern Norway"]))
 
 
 

### 2023-04-21 DRAFT_functional plant indicators_wetland.Rmd
finishing up first fraft of the wetland file

@@
 library(tidyr)
+library(plyr)
 library(dplyr)
-library(plyr)
 library(stringr)
@@
 We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
-```{r, echo = T, eval = F}
+```{r, include=FALSE}
 
@@
 results.wetland[['scaled']] <- results.wetland[['non-truncated']] <- results.wetland[['original']]
+```
+
+```{r, echo = T, eval = F}
 
@@
 
+### Simple summary statistics
 We can calculate simple summary statistics like means, standard deviations, and number of observations (note that there is both a lower and an upper indicator for each of the four plant functional indicators)
@@
 
+### Ecosystem sub-types
 And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
-```{r}
+```{r, message=FALSE}
 
@@
 
+### Indicator index maps
 We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
@@
 
-nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+#nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+#  st_as_sf() %>%
+#  st_transform(crs = crs(ANO.geo))
+
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
   st_as_sf() %>%
   st_transform(crs = crs(ANO.geo))
-reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+
+#reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+#  st_as_sf() %>%
+#  st_transform(crs = crs(ANO.geo))
+
+reg <- st_read("data/regions.shp")%>%
   st_as_sf() %>%
   st_transform(crs = crs(ANO.geo))
+
 # change region names to something R-friendly
@@
 ...but here the colors and values of the data points on the map are hard to make out.
+
+### Regions
 Alternatively we can calculate and show the region-wise means and standard deviations
@@
 
-#### CONTINUE HERE ####
-
-#### Regions
-<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
-Importing a shape file with the regional delineation. 
-```{r}
-#reg <- sf::st_read("data/regions.shp")
-#st_crs(reg)
-```
-
-### Scaled indicator values
-<!-- Text and analyses here -->
-
-### Uncertainty
-<!-- Text here -->
-
-## Prepare export
-<!-- Text here -->
-
 ### Eksport file (final product)

### 2023-04-21 R/functionalPlantIndicators/6 post scaling analyses.R

@@ -73,12 +73,22 @@ head(res.wetland[,70:76])
 res.wetland2 <- results.wetland[['2-sided']]
 st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
 
-nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+#nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
+#  st_as_sf() %>%
+#  st_transform(crs = crs(ANO.geo))
+
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
   st_as_sf() %>%
   st_transform(crs = crs(ANO.geo))
-reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+
+#reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
+#  st_as_sf() %>%
+#  st_transform(crs = crs(ANO.geo))
+
+reg <- st_read("data/regions.shp")%>%
   st_as_sf() %>%
   st_transform(crs = crs(ANO.geo))
+
 # change region names to something R-friendly
 reg$region
 reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/3 Reference bootstraps_semi-nat.R
seminat reference

@@ -5,7 +5,7 @@ seminat.ref <- indBoot(sp=NiN.seminat[,1],abun=NiN.seminat[,2:35],ind=NiN.semina
 
 
 seminat.ref.cov <- indBoot.freq(sp=NiN.seminat.cov[,1],abun=NiN.seminat.cov[,2:35],ind=NiN.seminat.cov[,36:42],
-                                iter=1000,obl=1,rat=2/3,var.abun=T)
+                                iter=1000,obl=1,rat=1/3,var.abun=T)
 
 ### saving backups
 #seminat.ref.backup <- seminat.ref
@@ -60,8 +60,9 @@ for (i in indID) {
 
 
 #### storing reference lists ####
-#rm(list= ls()[!(ls() %in% c('forest.ind.list','mire.ind.list','semiNat.ind.list','heath.ind.list','seminat.ref','ind.dat'))])
-#rm(list= ls()[!(ls() %in% c('seminat.ref.backup','seminat.ref','seminat.ref.cov.backup','seminat.ref.cov', 'ANO.sp.ind', 'NiN.seminat.cov','ind.dat'))])
-settings <- "iter=1000,obl=1,rat=2/3,var.abun=T"
+#rm(list= ls()[!(ls() %in% c('seminat.ref.cov.backup','seminat.ref.cov', 'ANO.sp.ind', 'ANO.geo', 'ASO.sp.ind', 'ASO.geo', 'NiN.seminat.cov','ind.dat'))])
+settings <- "iter=1000,obl=1,rat=1/3,var.abun=T"
 rm(list= ls()[!(ls() %in% c('seminat.ref.cov', 'seminat.ref.cov.backup','settings'))])
 save.image("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_seminat.RData")
+
+load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_seminat.RData")

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/4 Reference scaling values_semi-nat.R

@@ -1,0 +1,169 @@
+#### defining reference limits, mountain ####
+
+# limit table Ellenberg and aliens, mountain, median and 95% CI
+# NB! working at the 1:5000 mapping unit level !!!
+
+
+## extract from mountain ref-list
+
+# every NiN-type is represented by one 'generalisert artsliste'
+# some NiN-types are represented by two species lists
+# in some cases two NiN-types are represented by the same species list
+head(seminat.ref.cov[[1]])
+seminat.ref.cov[[1]][0,]
+
+# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
+names(seminat.ref.cov[["Light"]])
+x <- c(1,5,16,17,19,23,28,31:34)
+
+# checking the actual NiN-types in the seminat lists
+seminat.NiNtypes <- colnames(seminat.ref.cov[["Light"]])
+seminat.NiNtypes[-x] <- substr(seminat.NiNtypes[-x], 1, nchar(seminat.NiNtypes[-x])-1)
+seminat.NiNtypes
+
+# 7 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, Nitrogen, Phosphorus, Grazing_mowing, Soil_disturbance"
+indEll.n=7
+# creating a table to hold:
+# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
+# for every nature type (nrows)
+tab <- matrix(ncol=3*indEll.n, nrow=length(unique(seminat.NiNtypes)) ) # 34 basic ecosystem types
+# coercing the values into the table
+
+
+for (i in 1:length(x) ) {
+  tab[i,1:3] <- quantile(as.matrix(seminat.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,4:6] <- quantile(as.matrix(seminat.ref.cov[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,7:9] <- quantile(as.matrix(seminat.ref.cov[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
+  tab[i,10:12] <- quantile(as.matrix(seminat.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
[…]

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/4 Reference scaling values_wetland.R

@@ -1,236 +1,0 @@
-#### defining reference limits, mountain ####
-
-# limit table Ellenberg and aliens, mountain, median and 95% CI
-# NB! working at the 1:5000 mapping unit level !!!
-
-
-## extract from mountain ref-list
-
-# every NiN-type is represented by one 'generalisert artsliste'
-# some NiN-types are represented by two species lists
-# in some cases two NiN-types are represented by the same species list
-head(wetland.ref[[1]])
-wetland.ref[[1]][0,]
-
-# checking the actual NiN-types in the wetland lists
-wetland.NiNtypes <- colnames(wetland.ref[["Continentality"]])
-wetland.NiNtypes <- substr(wetland.NiNtypes,1,5)
-
-
-# 4 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, "Nitrogen"
-indEll.n=4
-# creating a table to hold:
-# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
-# for every nature type (nrows)
-tab <- matrix(ncol=3*indEll.n, nrow=length(unique(wetland.NiNtypes)) ) # 43 basic ecosystem types
-# coercing the values into the table
-# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
-x <- c(27,28,41:45)
-
-for (i in 1:length(x) ) {
-  tab[i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-}
-
-tab <- as.data.frame(tab)
-tab$NiN <- NA
-tab$NiN[1:length(x)] <- names(wetland.ref[[1]])[x]
-
-# NiN-types represented by several species lists
[…]

### 2023-05-03 DRAFT_functional plant indicators_wetland.Rmd
Update DRAFT_functional plant indicators_wetland.Rmd
corrected some typos and formatting issues directly in the file

@@
 2018).
-For instance, using an indicator for moisture on would find a functional signature with higher moisture values for plant communities in mires compared to e.g.
+For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g.
 grasslands or forests.
@@
 2020).
-This plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
+These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
 2021, Töpper & Jakobsson 2021).
@@
 ## About the underlying data
-In the 'plant functional indicator' project for wetlands, We use three sets of data for building indicators for ecological condition
-- as testing data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
+In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
+
+- as test data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
 - as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
@@
 'area representative nature monitoring'.
-Within a 5-year cycle, 1000 sites are  randomly distributed across mainland Norway.
-Each ANO site spans a 500 x 500 m grid cell and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
+1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle.
+Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
 For the vegetation analyses, the cover of each vascular plant species in the plot is recorded.
 Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
-In the wetland-analysis in this document, we only use the plots which were classified as on of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
+In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
 In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
@@
 The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species.
-In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1 > 1/32, 2 = 1/32 - 1/8 , 3 = 1/8 - 3/8, 4 = 3/8 - 4/5, 5 = 3/8 - 4/5 + dominance, 6 = 4/5 -1 ).
-For the purpose of this project these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
+In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5).
+For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
 
@@
 (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well.
-From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub-/tree encroachment, drainage, and pollution.
+From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub/tree encroachment, drainage, and pollution.
 
@@
 For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country.
-As wetlands occur more often in certain regions of Norway than in others the amount of plots and sites is not equal among the five regions ecological condition is assessed for (Northern Norway, Central Norway, Eastern Norway, Western Norway, Southern Norway).
-The in total 1351 plots are distributed across regions in the following way:
-
-Northern Norway: 416
-Central Norway:  362
-Eastern Norway:  295
-Western Norway:  151
-Southern Norway: 127
+As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions.
+The 1351 plots are distributed across regions in the following way:
+
+- Northern Norway: 416
+- Central Norway:  362
+- Eastern Norway:  295
+- Western Norway:  151
+- Southern Norway: 127
 
@@
 The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme.
-Thus there is no actual time series to these data and the indicator evaluation does therefore not include any temporal analyses.
+Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.
 
@@
 By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
-These distributions are calculated for ecosystem sub-types (grunntyper or kartleggingsenheter at a 1:5000 mapping scale) within the main wetland types in NiN.
+These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN.
 A more extensive discussion on the use of reference communities can be found in Jakobsson et al.
@@
 ### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-In this analysis, we derive scaling values from statistical (non-parametric here) distributions (see Jakobsson et al.
+In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al.
 2010).
@@
 2022), and deviation from the optimal state thus may occur in both direction (e.g.
-indicating too low or too high pH) we need to define two threshold values for good ecological condition as well as both a minimum and maximum value.
+indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value.
 In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e.
@@
 The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator.
-For details on the scaling principles in IBECA, please see Töpper & Jakobsson 2021.
+For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).
 
@@
 
-Median = reference values
-0.025 and 0.975 quantiles = lower and upper limit values
-min and max of the respective indicator's scale = min/max values
+- Median = reference values
+- 0.025 and 0.975 quantiles = lower and upper limit values
+- min and max of the respective indicator's scale = min/max values
 
@@
 
-Once testing data (ANO), and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO-community data and scale them against the scaling values from the reference distribution.
-Note that we scale each ANO-plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
+Once test data (ANO) and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO community data and scale them against the scaling values from the reference distribution.
+Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
 Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type.

### 2023-05-03 DRAFT_functional plant indicators_wetland.Rmd
Add data to cache to allow book to build from data/ folder

@@
 library(sf)
-library(tidyr)
 library(plyr)
-library(dplyr)
-library(stringr)
 library(tidyverse)
-library(readxl)
-library(broom)
 library(RColorBrewer)
@@
 library(ggridges)
-library(ggplot2)
 library(tmap)
@@
 library(stars)
-
 library(knitr)
@@
 ```{r, echo = F}
-st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                   layer="ANO_Art")
-ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                  layer="ANO_SurveyPoint")
-head(ANO.sp)
-head(ANO.geo)
+# Add data from cache
+ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
+ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))
+
+# st_layers(dsn = "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
+# ANO.sp <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+#                    layer="ANO_Art")
+# ANO.geo <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+#                   layer="ANO_SurveyPoint")
+# head(ANO.sp)
+# head(ANO.geo)
 ```
@@
 ## Tyler indicator data
-ind.dat <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
-                        sep = '\t', header=T, quote = '')
-head(ind.dat)
+## data from cache
+ind.dat <-readRDS(paste0(here::here(),"/data/cache/ind.dat.RDS"))
+# ind.dat <- read.table("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
+#                         sep = '\t', header=T, quote = '')
+# head(ind.dat)
 
@@
 ```{r, echo = F}
-load("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators//reference from NiN/Eco_State.RData")
-str(Eco_State)
+# data from cache
+Eco_State<-readRDS(paste0(here::here(),"/data/cache/Eco_State.RDS"))
+# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
+# str(Eco_State)
 ```
@@
 ```{r, include = F}
-load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_wetland.RData")
+# Data from cache
+wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))
+#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
 ```
@@
 ```{r, include = F}
-load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/results.wetland.RData")
+# Use cache data
+results.wetland<-readRDS(paste0(here::here(),"/data/cache/results.wetland.RDS"))
+
+#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")
 ```

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/3 Reference bootstraps_seminat.R
seminat scaling

@@ -63,6 +63,6 @@ for (i in indID) {
 #rm(list= ls()[!(ls() %in% c('seminat.ref.cov.backup','seminat.ref.cov', 'ANO.sp.ind', 'ANO.geo', 'ASO.sp.ind', 'ASO.geo', 'NiN.seminat.cov','ind.dat'))])
 settings <- "iter=1000,obl=1,rat=1/3,var.abun=T"
 rm(list= ls()[!(ls() %in% c('seminat.ref.cov', 'seminat.ref.cov.backup','settings'))])
-save.image("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_seminat.RData")
+save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
 
-load("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/ref_lists_seminat.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat.R

@@ -1,0 +1,459 @@
+#### scaled values ####
+r.s <- 1    # reference value
+l.s <- 0.6  # limit value
+a.s <- 0    # abscence of indicator, or indicator at maximum
+
+#### function for calculating scaled values for measured value ####
+
+## scaling function including truncation
+scal <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value equal 1
+    if (val >= ref) {x <- 1}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
+    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
+    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
+    # value equals or lower than maxmin
+    if (val <= maxmin) {x <-0}
+  } else {
+    # values <= the reference value equal 1
+    if (val <= ref) {x <- 1}
+    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
+    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
+    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
+    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
+    # value equals or larger than maxmin
+    if (val >= maxmin) {x <-0}
+  }
+  return(x)
+  
+}
+
+## scaling function without truncation
+scal.2 <- function() {
+  # place to hold the result
+  x <- numeric()
+  if (maxmin < ref) {
+    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
+    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
+    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
[…]

### 2023-05-03 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_wetland.R

@@ -1,346 +1,0 @@
-#### scaled values ####
-r.s <- 1    # reference value
-l.s <- 0.6  # limit value
-a.s <- 0    # abscence of indicator, or indicator at maximum
-
-#### function for calculating scaled values for measured value ####
-
-## scaling function including truncation
-scal <- function() {
-  # place to hold the result
-  x <- numeric()
-  if (maxmin < ref) {
-    # values >= the reference value equal 1
-    if (val >= ref) {x <- 1}
-    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
-    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
-    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
-    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
-    # value equals or lower than maxmin
-    if (val <= maxmin) {x <-0}
-  } else {
-    # values <= the reference value equal 1
-    if (val <= ref) {x <- 1}
-    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
-    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
-    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
-    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
-    # value equals or larger than maxmin
-    if (val >= maxmin) {x <-0}
-  }
-  return(x)
-  
-}
-
-## scaling function without truncation
-scal.2 <- function() {
-  # place to hold the result
-  x <- numeric()
-  if (maxmin < ref) {
-    # values >= the reference value estimated from the linear relationship for lim < x < ref (line below)
-    if (val >= ref) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
-    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
[…]

### 2023-05-03 R/NDVI/Landsat.R
Made changes to Landsat.R to allow it to be used within the Bookdown inside the NINA server.

@@ -9,9 +9,22 @@ library(data.table)
 library(tidyverse)
 library(sf)
 
+# Set up conditional file paths
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
+
+pData <- ifelse(dir == "C:", 
+                 "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
+                 "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")
+ 
+ 
+
 #############Import NiN data
 
-nin <- st_read("R:\\GeoSpatialData\\Habitats_biotopes\\Norway_Miljodirektoratet_Naturtyper_nin\\Original\\versjon20221231\\Natur_Naturtyper_nin_norge_med_svalbard_25833\\Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")%>%
+nin <- st_read(path)%>%
   mutate(hovedøkosystem = recode(hovedøkosystem, 
                                  "Våtmark" = 'Våtmark',
                                  "Skog" = "Skog",
@@ -31,10 +44,23 @@ nin <- st_read("R:\\GeoSpatialData\\Habitats_biotopes\\Norway_Miljodirektoratet_
 
 
 ##############Import Landsat NDVI Data
-df <-
-  list.files("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_Landsat\\", pattern = "*.csv") %>%
-  map_df(~fread(.))
-df
+
+## Fread doesn't like the weird path to the server version of the P drive
+## hence this horrendous work around (there must be an easier way than this but I kept hitting dead-ends)
+files=list.files(pData, pattern = "*.csv", full.names = TRUE)
[…]

### 2023-05-05 R/functionalPlantIndicators/code semi-nat/1 data handling_seminat.R
seminat further analysis of results

@@ -333,11 +333,56 @@ ASO.geo <- st_as_sf(x = ASO.geo,
                         crs = "+proj=longlat +datum=WGS84 +ellps=WGS84")
 
 colnames(ASO.geo)
-colnames(ASO.geo)[c(6,7,8,10,16,18,23,24)] <- c("Omradenummer_flatenummer","Eng_ID","ASO_ID","NiN_grunntype",
-                                  "Beitetrykk",
-                                  "Slatteintensitet",
-                                  "tungekjoretoy",
-                                  "slitasje")
+colnames(ASO.geo)[c(6,7,8,10,14,16,18,23,24)] <- c("Omradenummer_flatenummer","Eng_ID","ASO_ID","NiN_grunntype",
+                                                   "bruksintensitet",
+                                                   "beitetrykk",
+                                                   "slatteintensitet",
+                                                   "tungekjoretoy",
+                                                   "slitasje")
+
+
+## fix NiN-variables
+# remove variable code in the data
+ASO.geo$bruksintensitet
+ASO.geo$bruksintensitet <- gsub("7JB-BA_", "", ASO.geo$bruksintensitet) 
+unique(ASO.geo$bruksintensitet)
+ASO.geo$bruksintensitet <- gsub("X", "NA", ASO.geo$bruksintensitet)
+unique(ASO.geo$bruksintensitet)
+ASO.geo$bruksintensitet <- as.numeric(ASO.geo$bruksintensitet)
+unique(ASO.geo$bruksintensitet)
+
+ASO.geo$beitetrykk
+ASO.geo$beitetrykk <- gsub("7JB-BT_", "", ASO.geo$beitetrykk) 
+unique(ASO.geo$beitetrykk)
+ASO.geo$beitetrykk <- gsub("X", "NA", ASO.geo$beitetrykk)
+unique(ASO.geo$beitetrykk)
+ASO.geo$beitetrykk <- as.numeric(ASO.geo$beitetrykk)
+unique(ASO.geo$beitetrykk)
+
+ASO.geo$slatteintensitet
+ASO.geo$slatteintensitet <- gsub("7JB-SI_", "", ASO.geo$slatteintensitet) 
+unique(ASO.geo$slatteintensitet)
[…]

### 2023-05-05 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R

@@ -504,11 +504,13 @@ regnor <- regnor %>%
 ## scaled value maps
 # Light1 (lower indicator)
 tm_shape(regnor) +
-  tm_polygons(col="Light1.reg.mean", title="Light (lower)", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
+  tm_polygons(col="Light1.reg.mean", title="Light (lower)", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("Light1.reg.n",col="black",bg.color="grey")
 
 # Grazing_mowing1 (lower indicator)
 tm_shape(regnor) +
-  tm_polygons(col="Grazing_mowing1.reg.mean", title="Grazing_mowing (lower)", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
+  tm_polygons(col="Grazing_mowing1.reg.mean", title="Grazing_mowing (lower)", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("Grazing_mowing1.reg.n",col="black",bg.color="grey")
 
 
 
@@ -957,3 +959,13 @@ points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Grazing_mowing1" & res.seminat.A
        rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Grazing_mowing1" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original)),
        col="red")
 
+
+
+#### relating scaled values to NiN condition variables ####
+ggplot(res.seminat, aes(x=bruksintensitet, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.seminat.ASO, aes(x="Aktuell bruksintensitet (7JB-BA)", y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")

### 2023-05-05 DRAFT_functional plant indicators_wetland.Rmd

@@
----
-title: "Functional pant indicators - wetland, indicator documentation"
-output:
-  html_document
----
-
-<!-- This is a template for how to document the indicator analyses.
-
-Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name the file DRAFT_myIndicatorName.Rmd.
-
-Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
-you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work.
-
-Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout.
-
-When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
-the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. When completed, Anders will make a copy
-of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
-the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
-
-<!-- Examples:
-
-For an example, see 'indicators/breareal-workingFile.Rmd' 
-and the subsequent 'breareal.Rmd' 
-which again, when rendered, produces this:
-https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
-
-Notice the degree of annotation.
-
--->
-
-<!-- Replace 'indicator name' with your own indicator name  -->
+# Functional plant indicators - wetland {#Functional-plant-indicators-wetland}
 # Functional plant indicators (Moisture, Light, pH, Nitrogen)
@@
 
-```{r}
-Sys.Date()
-```
+May 2023
 
@@
 
-<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
 ```{r, echo=F}
@@
 <hr />
-
-<!-- Document you work below.
-Try not to change  the headers too much.
-Data can be stored on NINA server.
-Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine.
-If it is not too big you may store under /data/ on this repository -->
 
@@
 ```{r}
-#summary(NiN.wetland)
 head(ANO.sp.ind)
@@
 And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
-```{r, message=FALSE}
+```{r,echo=FALSE,message=FALSE, warning=FALSE}
 
@@
 We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
-```{r, message = FALSE}
+```{r, echo=FALSE, warning=FALSE,message = FALSE}
 #### scaled value maps ####

### 2023-05-05 functional_plant_indicators_wetland.Rmd
Add functional plant indicators to the book

--- 2023-05-05 DRAFT_functional plant indicators_wetland.Rmd
@@
-# Functional plant indicators - wetland {#Functional-plant-indicators-wetland}
-# Functional plant indicators (Moisture, Light, pH, Nitrogen)
+# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
 
@@
 
-```{r, echo=F}
+```{r, echo=FALSE}
 Ecosystem <- "våtmark"
@@
 
-```{r, echo=F}
+```{r, echo=FALSE}
 metaData <- data.frame(Ecosystem,

### 2023-05-09 functional_plant_indicators_wetland.Rmd
edits wetland markdown & setup seminat markdown

@@
-# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
+# Functional plant indicators, wetland ecosystems (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
 
@@
 Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
-Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
+Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
 2019) for wetland ecosystems.
@@
 2020).
-These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
+These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
 2021, Töpper & Jakobsson 2021).
+Note that deviations from the reference may occur in both directions, e.g.
+the moisture signature from the testing data may be higher or lower than in the reference.
+Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately.
+Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).
 
 ## About the underlying data
-In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
+In the 'functional plant indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
 
@@
 'area representative nature monitoring'.
-1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle.
+1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle.
 Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
@@
 For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution.
-As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al.
+As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al.
 2022), and deviation from the optimal state thus may occur in both direction (e.g.
@@
 # species with abundance 1 (i.e. a max of 100%, must be included in each sample)
-# each sample re-samples 2/3 of the number of species
+# each sample re-samples 1/3 of the number of species
 # the abundance of the re-sampled species may vary (see bootstrap function for details)
 wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
-                          iter=1000,obl=1,rat=2/3,var.abun=T)
+                          iter=1000,obl=1,rat=1/3,var.abun=T)
 
@@
 Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
-Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type.
+Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each functional plant indicator.
 An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.
@@
     values_to = "scaled_value",
-    values_drop_na = TRUE
+    values_drop_na = FALSE
+  )
+
+# add original values as well
+res.wetland <- 
+  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
+                               pivot_longer(
+                                 cols =  c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+                                 names_to = NULL,
+                                 values_to = "original",
+                                 values_drop_na = FALSE
+                               ) %>%
+                               pull(original)
   )
@@
   group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value))
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value[!is.na(scaled_value)]))
 
@@
 # making the plot
-ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+ggplot(res.wetland[!is.na(res.wetland$scaled_value),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
   geom_hline(yintercept=0.6, linetype="dashed") + 
@@
 ```
+The functional signature from the plant communities in wetland ecosystems looks largely good, except for V2 (swamp forests), which show rather low index scores for the lower pH and Nitrogen indicators, i.e.
+suggesting too acidic conditions and too low nitrogen availability.
+
+We can look at V2 specifically to see if this is the case for all sub-types.
+
+```{r}
+# making the plot for V2 only
+ggplot(res.wetland[res.wetland$hovedtype_rute=="V2",], aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+  #  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value") 
+```
+The deviations towards low pH and low nitrogen are mainly an issue for V2-C1, limestone-poor swamp forests.
 
@@
                           sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
-                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T))
+                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+             pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
+                          nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),]))
     )
@@
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
-
+  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("pH1.reg.n",col="black",bg.color="grey")
+```
+Mean index value by region for the lower pH indicator (i.e. index shows deviations towards a more acidic environment). Numbers in grey fields show the number of observations in the respective region.
+
+```{r}
 # sd
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))
-
-```
+  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
+  tm_text("pH1.reg.n",col="black",bg.color="grey")
+
+```
+Standard deviation to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
+
+### unscaled values vs. reference
+We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
+Since pH and nitrogen show some deviation of the lower pH indicator we exemplify this with these indicators for unscaled values.
+
+```{r}
+summary(res.wetland$kartleggingsenhet_1m2)
+length(unique(res.wetland$kartleggingsenhet_1m2))
+# 16 NiN-types to plot
+colnames(wetland.ref.cov[['Soil_reaction_pH']])
+
+### pH
+par(mfrow=c(4,4))
+## V1s
+# V1-C-1
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C1',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
+       col="red")
+
+# V1-C-2
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C2',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
+       col="red")
+
+# V1-C-3
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C3',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
+       col="red")
+
+# V1-C-4
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C4',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
+       col="red")
+
+# V1-C-5
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C5")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C5',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
+       col="red")
+
+# V1-C-6
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C6',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
+       col="red")
+
+# V1-C-7
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C7',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
+       col="red")
+
+# V1-C-8
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C8',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
+       col="red")
+
+## V2s
+# V2-C-1
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
+       col="red")
+
+# V2-C-2
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C2',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
+       col="red")
+
+# V2-C-3
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C3',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
+       col="red")
+
+## V3s
+# V3-C-1
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V3-C1',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
+       col="red")
+
+# V3-C-2
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V3-C2',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
+       col="red")
+
+## V4s
+# V4-C-2
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V4-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V4-C2',xlab='pH value')
+points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
+       col="red")
+
+## V8s
+# V8-C-1
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C1")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V8-C1',xlab='pH value')
+#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
+#       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
+       col="red")
+
+# V8-C-2
+plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V8-C2',xlab='pH value')
+#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
+#       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
+       col="red")
+legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
+
+### Nitrogen
+par(mfrow=c(4,4))
+## V1s
+# V1-C-1
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C1',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
+       col="red")
+
+# V1-C-2
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C2',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
+       col="red")
+
+# V1-C-3
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C3',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
+       col="red")
+
+# V1-C-4
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C4',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
+       col="red")
+
+# V1-C-5
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C5")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C5',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
+       col="red")
+
+# V1-C-6
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C6',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
+       col="red")
+
+# V1-C-7
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C7',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
+       col="red")
+
+# V1-C-8
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V1-C8',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
+       col="red")
+
+## V2s
+# V2-C-1
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C1',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
+       col="red")
+
+# V2-C-2
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C2',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
+       col="red")
+
+# V2-C-3
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V2-C3',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
+       col="red")
+
+## V3s
+# V3-C-1
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V3-C1',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
+       col="red")
+
+# V3-C-2
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V3-C2',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
+       col="red")
+
+## V4s
+# V4-C-2
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V4-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V4-C2',xlab='Nitrogen value')
+points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
+       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
+       col="red")
+
+## V8s
+# V8-C-1
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C1")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V8-C1',xlab='Nitrogen value')
+#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
+#       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
+       col="red")
+
+# V8-C-2
+plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C2")]) ,na.rm=T),
+     xlim=c(1,7), type="l", main='V8-C2',xlab='Nitrogen value')
+#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
+#       type="l", col="red")
+points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
+       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
+       col="red")
+
+legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
+
+```
+THe figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests.
+According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
 

### 2023-05-09 R/functionalPlantIndicators/code naturally open/1 data handling_natopen.R

@@ -23,17 +23,47 @@ st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_e
 ## ANO
 st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
 ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                   layer="ANO_Art")
+                  layer="ANO_Art")
 ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                  layer="ANO_SurveyPoint")
+                   layer="ANO_SurveyPoint")
 head(ANO.sp)
 head(ANO.geo)
 
+## GRUK
+excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx")
+GRUK.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
+                             sheet = 2)
+GRUK.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
+                           sheet = 3)
+
+# condition evaluation for 2021 data
+excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/NNF_GRUK_GJELDENDE.xls")
+
+GRUK2021.condition <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/NNF_GRUK_GJELDENDE.xls", 
+                                 sheet = 1)
+
+head(GRUK.variables)
+head(GRUK.species)
+head(GRUK2021.condition)
+
+
+
+
+
+## ASO data
+
+
+
+
 
 ## Tyler indicator data
-ind.dat <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
+ind.Tyler <- read.table("P:/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
                         sep = '\t', header=T, quote = '')
-head(ind.dat)
+head(ind.Tyler)
+
+## Grime CSR-values
[…]

### 2023-05-09 R/functionalPlantIndicators/code semi-nat/1 data handling_seminat.R

@@ -540,6 +540,7 @@ NiN.sp$sp <- word(NiN.sp$sp, 1,2)
 NiN.sp$spgr <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code))
 levels(NiN.sp$spgr)
 # trimming to desired species groups (for åpen systems eg. removing trees ("a1a") )
+# here we remove trees as they represent a driver for vegetation change in the field layer
 NiN.sp <- NiN.sp[NiN.sp$spgr!="a1a",]
 
 # environment data
@@ -668,7 +669,7 @@ unique(NiN.sp.ind[is.na(NiN.sp.ind$Light) & is.na(NiN.sp.ind$Nitrogen) &
 #### matching with NiN ecosystem types - semi-natural types ####
 # NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
 levels(NiN.env$Nature_Type)
-NiN.seminat <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type %in% list("Semi_Natural ","Coastal_Heath"),"ID"]),colnames(ind.dat)[c(15:19,21,22)])]   # Light, Moisture, Soil_reaction_pH, Nitrogen, Grazing_mowing, Soil_disturbance
+NiN.seminat <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type %in% list("Semi_Natural ","Coastal_Heath"),"ID"]),colnames(ind.dat)[c(15:19,21,22)])]   # Light, Moisture, Soil_reaction_pH, Nitrogen, Phosphorus, Grazing_mowing, Soil_disturbance
 
 NiN.seminat[1,]
 names(NiN.seminat)

### 2023-05-09 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ANO.R

@@ -61,14 +61,14 @@ scal.2 <- function() {
 
 
 unique(ANO.geo$hovedtype_rute) # NiN types in data
-unique(substr(seminat.ref.cov.val$grunn,1,2)) # NiN types in reference
+unique(substr(seminat.ref.cov.val$grunn,1,3)) # NiN types in reference
 levels(as.factor(ANO.geo$kartleggingsenhet_1m2)) # NiN types in data
 levels(seminat.ref.cov.val$grunn) # NiN types in reference
 #### creating dataframe to hold the results for seminats ####
 # all ANO points
 nrow(ANO.geo)
-# all seminat ANO points (incl. T41 and T45)
-nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T31","T32","T34","T40","T41","T45","V10"),])
+# all seminat ANO points (incl. T40 and T41)
+nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T31","T32","T34","T40","T41","V10"),])
 # all seminat ANO points with a NiN-type represented in the reference
 nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(seminat.ref.cov.val$grunn,1,3)),])
 nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T32"),])
@@ -79,9 +79,7 @@ nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T40"),])
 # ok, we'll be losing 258 T31 and 10 T40, as T31 and T40 are not covered by reference data
 # which leaves us with 329 points, of which...
 nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T41"),])
-nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T45"),])
 # 29 points are T41 (heavily altered meadows appearing as semi-natural after decades of extensive use)
-# 96 points are T45 (intensive and regular soil manipulation)
 # we omit T45
 nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V10"),])
 # 6 points are semi-natural wetlands
@@ -451,13 +449,13 @@ hist(results.seminat[['2-sided']]$Soil_disturbance2,breaks=40)
 #            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 #write.table(results.seminat[['non-truncated']], file='output/scaled data/results.seminat_non-truncated.txt',
[…]

### 2023-05-09 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ASO.R

@@ -87,6 +87,8 @@ ASO.geo$NiN_grunntype <- as.factor(ASO.geo$NiN_grunntype)
 summary(ASO.geo$NiN_grunntype)
 # 230 out of 381 ASO-meadows are not T32
 # for anything that is not T32 no data were registered, these won't be processed further
+# which leaves us with 151 ASO-meadows we can expect a result for
+
 
 results.seminat.ASO <- list()
 ind <- unique(seminat.ref.cov.val$Ind)

### 2023-05-09 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R

@@ -1,3 +1,12 @@
+#### getting all the data from previous scripts in ####
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/data_seminat.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ANO.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")
+
+
+
+
 #### plotting scaled values by main ecosystem type ####
 ## continuing with 2-sided
 res.seminat <- results.seminat[['2-sided']]
@@ -40,26 +49,6 @@ res.seminat <-
 head(res.seminat[,70:76])
 
 
-# summarizing the indicator scores
-res.seminat %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
-
-
-# making the plot
-ggplot(res.seminat, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
-  geom_hline(yintercept=0.6, linetype="dashed") + 
-  geom_violin(color=NA) +
-#  geom_boxplot(width=0.2, color="grey") +
-  geom_point(size=1, shape=16, color="black") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Phosphorus1","Grazing_mowing1","Soil_disturbance1",
-                                     "Light2","Moist2","pH2","Nitrogen2","Phosphorus2","Grazing_mowing2","Soil_disturbance2")), ncol = 7) + 
-  xlab("Main ecosystem type") + 
-  ylab("Scaled indicator value (ANO data)") 
-
-
-
-
 # similarly for ASO
 res.seminat.ASO <- results.seminat.ASO[['2-sided']]
 colnames(res.seminat.ASO)
@@ -83,22 +72,51 @@ res.seminat.ASO <-
 # add original values as well
 res.seminat.ASO <- 
   res.seminat.ASO %>% add_column(original = results.seminat.ASO[['original']] %>% 
-                               pivot_longer(
-                                 cols = c("Light1","Light2",
[…]

### 2023-05-09 R/functionalPlantIndicators/code wetland/3 Reference bootstraps_wetland.R

@@ -5,7 +5,7 @@ wetland.ref <- indBoot(sp=NiN.wetland[,1],abun=NiN.wetland[,2:46],ind=NiN.wetlan
 
 
 wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
-                                iter=1000,obl=1,rat=2/3,var.abun=F)
+                                iter=1000,obl=1,rat=1/3,var.abun=T)
 
 ### saving backups
 #wetland.ref.backup <- wetland.ref
@@ -34,19 +34,22 @@ for (i in 1:length(wetland.ref.cov) ) {
 
 #### checking NiN-types with several species lists ####
 str(wetland.ref.cov)
+# V3-C1a has only NA's, too few species in the NiN-list
 
-indID <- colnames(ind.dat[,c("Continentality", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")])
+indID <- colnames(ind.dat[,c("Light", "Moisture", "Soil_reaction_pH", "Nitrogen")])
 NiNID <- colnames(wetland.ref.cov[[1]])
 
+
+dirOutput <- "C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/testoutput"
+#setwd(dirOutput)
+
 for (i in indID) {
-  
-  dirOutput <- "C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition_PP&FPI/R/functionalPlantIndicators/testoutput"
-  setwd(dirOutput)
-  name <- paste('wetland',i,'.jpg',sep='')
+
+  name <- paste(dirOutput,'/wetland',i,'.jpg',sep='')
   jpeg(filename=name,width=8000,height=4000,res=300)
   
   par(mfrow=c(5,9))
-  for (j in NiNID) {
+  for (j in NiNID[-1]) {
     plot(density(wetland.ref.cov[[i]][,j],na.rm=T),main=j,xlim=c(-2,10))
   }
   dev.off()
@@ -62,6 +65,8 @@ for (i in indID) {
 #### storing reference lists ####
 #rm(list= ls()[!(ls() %in% c('forest.ind.list','mire.ind.list','semiNat.ind.list','heath.ind.list','wetland.ref','ind.dat'))])
[…]

### 2023-05-09 R/functionalPlantIndicators/code wetland/5 scaled value calculations_wetland.R

@@ -278,8 +278,6 @@ summary(results.wetland[['scaled']])
 results.wetland[['2-sided']] <- results.wetland[['non-truncated']]
 
 # check if there are values equalling exactly 1
-results.wetland[['2-sided']]$Cont1[results.wetland[['2-sided']]$Cont1==1]
-results.wetland[['2-sided']]$Cont2[results.wetland[['2-sided']]$Cont2==1]
 results.wetland[['2-sided']]$Light1[results.wetland[['2-sided']]$Light1==1]
 results.wetland[['2-sided']]$Light2[results.wetland[['2-sided']]$Light2==1]
 results.wetland[['2-sided']]$Moist1[results.wetland[['2-sided']]$Moist1==1]
@@ -293,9 +291,6 @@ results.wetland[['2-sided']]$Nitrogen2[results.wetland[['2-sided']]$Nitrogen2==1
 
 
 # remove values >1 for Ellenberg
-results.wetland[['2-sided']]$Cont1[results.wetland[['2-sided']]$Cont1>1] <- NA
-results.wetland[['2-sided']]$Cont2[results.wetland[['2-sided']]$Cont2>1] <- NA
-
 results.wetland[['2-sided']]$Light1[results.wetland[['2-sided']]$Light1>1] <- NA
 results.wetland[['2-sided']]$Light2[results.wetland[['2-sided']]$Light2>1] <- NA
 
@@ -312,9 +307,6 @@ results.wetland[['2-sided']]$Nitrogen2[results.wetland[['2-sided']]$Nitrogen2>1]
 # check distribution
 x11()
 par(mfrow=c(2,5))
-hist(results.wetland[['2-sided']]$Cont1,breaks=40)
-hist(results.wetland[['2-sided']]$Cont2,breaks=40)
-
 hist(results.wetland[['2-sided']]$Light1,breaks=40)
 hist(results.wetland[['2-sided']]$Light2,breaks=40)
 
@@ -342,5 +334,7 @@ write.table(results.wetland[['2-sided']], file='C:/Users/joachim.topper/OneDrive
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
 
-rm(list= ls()[!(ls() %in% c('results.wetland','settings'))])
-save.image("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/results.wetland.RData")
[…]

### 2023-05-09 R/functionalPlantIndicators/code wetland/6 post scaling analyses_wetland.R

@@ -1,3 +1,10 @@
+#### getting all the data from previous scripts in ####
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/data_wetland.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
+load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")
+
+
+
 #### plotting scaled values by main ecosystem type ####
 ## continuing with 2-sided
 res.wetland <- results.wetland[['2-sided']]
@@ -9,48 +16,7 @@ res.wetland <-
     cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
     names_to = "fp_ind",
     values_to = "scaled_value",
-    values_drop_na = TRUE
-  )
-
-
-# summarizing the indicator scores
-res.wetland %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
-
-
-# making the plot
-ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
-  geom_hline(yintercept=0.6, linetype="dashed") + 
-  geom_violin() +
-#  geom_boxplot(width=0.2, color="grey") +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
-  xlab("Main ecosystem type") + 
-  ylab("Scaled indicator value") 
-
-
-
-
-
-
-
-
-
-
-
-
-#### getting scaled AND original values ####
-res.wetland <- results.wetland[['2-sided']]
-
-# make long version of the scaled value part
-res.wetland <-
-  res.wetland %>% 
-  pivot_longer(
-    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
-    names_to = "fp_ind",
[…]

### 2023-05-09 R/functionalPlantIndicators/code semi-nat/6x beta regression.R
introducing beta regression, seminat

@@ -1,0 +1,138 @@
+expit <- function(L) exp(L) / (1+exp(L))
+
+res.seminat3 <- res.seminat2
+st_geometry(res.seminat3) <- NULL
+
+dfx <- res.seminat3[res.seminat3$region=="Northern Norway",]
+
+mean(dfx$Grazing_mowing1,na.rm=T)
+
+nrow(dfx[!is.na(dfx$Grazing_mowing1),])
+
+sd(dfx$Grazing_mowing1,na.rm=T)/sqrt(nrow(dfx[!is.na(dfx$Grazing_mowing1),]))
+
+
+library(betareg)
+summary( betareg(Grazing_mowing1 ~ 1, data=dfx) )
+expit(coefficients(betareg(Grazing_mowing1 ~ 1, data=dfx))[1])
+
+str(summary( betareg(Grazing_mowing1 ~ 1, data=dfx) ))
+
+expit(summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1])
+summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[2]
+
+expit( summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1] + 
+         summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[2] )-
+  expit( summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1] )
+
+library(glmmTMB)
+
+summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )
+
+expit(summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[1])
+summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[2]
+
+expit( summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[1] + 
+         summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[2] )-
[…]

### 2023-05-10 functional_plant_indicators_wetland.Rmd
tidying up the betareg-additions

@@
 
-### Simple summary statistics
-We can calculate simple summary statistics like means, standard deviations, and number of observations (note that there is both a lower and an upper indicator for each of the four plant functional indicators)
+In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).
 
@@
   )
-
-# summarizing the indicator scores
-res.wetland %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value[!is.na(scaled_value)]))
 
@@
 
-### Regions
-Alternatively we can calculate and show the region-wise means and standard deviations
+### Regions - maps and statistics
+Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because:
+(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
+(ii) the ANO dataset has a nested structure
+
+Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
+Here, we apply the following function using either a glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
+```{r}
+library(betareg)
+library(glmmTMB)
+
+expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale
+
+# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
+# if that is not the case, then the function performs a betareg if theres >= 2 observations
+# if that is not the case either, then the function returns the value of the single observation with NA standard error
+
+indmean.beta <- function(df) {
+
+  st_geometry(df) <- NULL
+  colnames(df) <- c("y","ran")
+
+  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {
+
+    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {
+
+      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)
+
+      return(c(
+        expit(summary( mod1 )$coefficients$cond[1]),
+
+        expit( summary( mod1 )$coefficients$cond[1] + 
+                 summary( mod1 )$coefficients$cond[2] )-
+          expit( summary( mod1 )$coefficients$cond[1] ),
+
+        nrow(df[!is.na(df$y),]),
+        summary( mod1 )$coefficients$cond[1],
+        summary( mod1 )$coefficients$cond[2]
+      ))
+
+    } else {
+
+      mod2 <- betareg(y ~ 1, data=df)
+
+      return(c(
+        expit(summary( mod2 )$coefficients$mean[1]),
+        expit( summary( mod2 )$coefficients$mean[1] + 
+                 summary( mod2 )$coefficients$mean[2] )-
+          expit( summary( mod2 )$coefficients$mean[1] ),
+        nrow(df[!is.na(df$y),]),
+        summary( mod2 )$coefficients$mean[1],
+        summary( mod2 )$coefficients$mean[2]
+      ))
+
+    }
+
+  } else {
+
+    return(c(df$y,NA,1,NA,NA))
+
+  }
+
+}
+
+```
+
 ```{r}
@@
   mutate(
-    pH1.reg.mean = c(mean(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
-                             mean(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
-                             mean(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
-                             mean(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
-                             mean(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
-    pH1.reg.sd = c(sd(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
-                          sd(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
-                          sd(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
-                          sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
-                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+    pH1.reg.mean = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[1],
+                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[1],
+                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[1],
+                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[1],
+                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[1]
+    ),
+    pH1.reg.se = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[2],
+                               indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[2],
+                               indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[2],
+                               indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[2],
+                               indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[2]
+    ),
     pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
@@
                               nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
-                          nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),]))
+                              nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),])
+    )
   )
@@
 ```{r}
-# sd
+# se
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
+  tm_polygons(col="pH1.reg.se", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
   tm_text("pH1.reg.n",col="black",bg.color="grey")
@@
 ```
-Standard deviation to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
+Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
 

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ANO.R

@@ -455,7 +455,7 @@ write.table(results.seminat[['2-sided']], file='P:/41201785_okologisk_tilstand_2
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
 
-rm(list= ls()[!(ls() %in% c('ANO.seminat','results.seminat','settings'))])
+rm(list= ls()[!(ls() %in% c('seminat.ref.cov.val','ANO.seminat','results.seminat','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ANO.RData")
 
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ANO.RData")

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ASO.R

@@ -426,7 +426,7 @@ write.table(results.seminat.ASO[['2-sided']], file='P:/41201785_okologisk_tilsta
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
 
-rm(list= ls()[!(ls() %in% c('results.seminat.ASO','settings'))])
+rm(list= ls()[!(ls() %in% c('ASO.geo','seminat.ref.cov.val','results.seminat.ASO','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")
 
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R

@@ -91,19 +91,6 @@ head(res.seminat.ASO[,70:76])
 
 
 
-# summarizing the indicator scores
-
-#ANO
-res.seminat %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value[!is.na(scaled_value)]))
-
-#ASO
-res.seminat.ASO %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value[!is.na(scaled_value)]))
-
-
 # making the plot, ANO
 ggplot(res.seminat, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
   geom_hline(yintercept=0.6, linetype="dashed") + 
@@ -117,14 +104,6 @@ ggplot(res.seminat, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) 
 
 
 
-
-
-# summarizing the indicator scores
-res.seminat.ASO %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
-
-
 # making the plot, ASO
 res.seminat.ASO$NiN_grunntype2 <- substring(res.seminat.ASO$NiN_grunntype,5)
 ggplot(res.seminat.ASO, aes(x=factor(NiN_grunntype2), y=scaled_value, fill=fp_ind)) + 
@@ -136,13 +115,6 @@ ggplot(res.seminat.ASO, aes(x=factor(NiN_grunntype2), y=scaled_value, fill=fp_in
                                      "Light2","Moist2","pH2","Nitrogen2","Phosphorus2","Grazing_mowing2","Soil_disturbance2")), ncol = 7) + 
   xlab("T32 (semi-natural meadow) basic ecosystem type") + 
   ylab("Scaled indicator value (ASO data)") 
-
-
-
-
-
-
-
 
 
 
@@ -252,16 +224,7 @@ tm_shape(regnor) +
 res.seminat2 = st_join(res.seminat2, regnor, left = TRUE)
 colnames(res.seminat2)
 
-res.seminat2 %>% 
-  group_by(as.factor(region)) %>% 
[…]

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/6x beta regression.R

@@ -1,138 +1,0 @@
-expit <- function(L) exp(L) / (1+exp(L))
-
-res.seminat3 <- res.seminat2
-st_geometry(res.seminat3) <- NULL
-
-dfx <- res.seminat3[res.seminat3$region=="Northern Norway",]
-
-mean(dfx$Grazing_mowing1,na.rm=T)
-
-nrow(dfx[!is.na(dfx$Grazing_mowing1),])
-
-sd(dfx$Grazing_mowing1,na.rm=T)/sqrt(nrow(dfx[!is.na(dfx$Grazing_mowing1),]))
-
-
-library(betareg)
-summary( betareg(Grazing_mowing1 ~ 1, data=dfx) )
-expit(coefficients(betareg(Grazing_mowing1 ~ 1, data=dfx))[1])
-
-str(summary( betareg(Grazing_mowing1 ~ 1, data=dfx) ))
-
-expit(summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1])
-summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[2]
-
-expit( summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1] + 
-         summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[2] )-
-  expit( summary(betareg(Grazing_mowing1 ~ 1, data=dfx))$coefficients$mean[1] )
-
-library(glmmTMB)
-
-summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )
-
-expit(summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[1])
-summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[2]
-
-expit( summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[1] + 
-         summary( glmmTMB(Grazing_mowing1 ~ 1 +(1|ano_flate_id), family=beta_family(), data=dfx) )$coefficients$cond[2] )-
[…]

### 2023-05-10 R/functionalPlantIndicators/code wetland/5 scaled value calculations_wetland.R

@@ -334,7 +334,7 @@ write.table(results.wetland[['2-sided']], file='C:/Users/joachim.topper/OneDrive
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
 
-rm(list= ls()[!(ls() %in% c('ANO.wetland','results.wetland','settings'))])
+rm(list= ls()[!(ls() %in% c('wetland.ref.cov.val','ANO.wetland','results.wetland','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")
 
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")

### 2023-05-10 R/functionalPlantIndicators/code wetland/6 post scaling analyses_wetland.R

@@ -33,11 +33,6 @@ res.wetland <-
 
 res.wetland <- res.wetland[!is.na(res.wetland$scaled_value) | !is.na(res.wetland$original),]
 
-# summarizing the indicator scores
-res.wetland %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE))
-
 
 
 # making the plot
@@ -74,7 +69,7 @@ st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
 
 nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
   st_as_sf() %>%
-  st_transform(crs = crs(ANO.geo))
+  st_transform(crs = st_crs(ANO.geo))
 
 #reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
 #  st_as_sf() %>%
@@ -82,7 +77,7 @@ nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
 
 reg <- st_read("data/regions.shp")%>%
   st_as_sf() %>%
-  st_transform(crs = crs(ANO.geo))
+  st_transform(crs = st_crs(ANO.geo))
 
 # change region names to something R-friendly
 reg$region
@@ -139,16 +134,7 @@ tm_shape(regnor) +
 res.wetland2 = st_join(res.wetland2, regnor, left = TRUE)
 colnames(res.wetland2)
 
-res.wetland2 %>% 
-  group_by(as.factor(region)) %>% 
-  dplyr::summarise(Light1.reg.mean = mean(Light1,na.rm=T))
-
-res.wetland2 %>% 
-  group_by(as.factor(region)) %>% 
-  dplyr::summarise(Light2.reg.mean = mean(Light2,na.rm=T))
-  
-
-
+# simple means, not appropriate with 0-1 bound data
 res.wetland2 %>% 
   group_by(as.factor(region)) %>% 
   mutate(Light1.reg.mean = mean(Light1,na.rm=T)) %>%
@@ -161,144 +147,99 @@ res.wetland2 %>%
   mutate(Nitrogen2.reg.mean = mean(Nitrogen2,na.rm=T))
   
 
+# rather use beta-regression
+expit <- function(L) exp(L) / (1+exp(L))
+library(betareg)
+library(glmmTMB)
+
+
[…]

### 2023-05-10 functional_plant_indicators_wetland.Rmd
wrapping up the markdowns for seminat and wetland

@@
 wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))
+
 #load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
@@
 ### pH
-par(mfrow=c(4,4))
+par(mfrow=c(2,4))
 ## V1s
@@
 
+par(mfrow=c(2,4))
 ## V2s
@@
 ### Nitrogen
-par(mfrow=c(4,4))
+par(mfrow=c(2,4))
 ## V1s

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/1 data handling_seminat.R

@@ -11,11 +11,11 @@ library(tmaptools)
 
 #### download ANO data from kartkatalogen to P-drive ####
 # url <- "https://nedlasting.miljodirektoratet.no/naturovervaking/naturovervaking_eksport.gdb.zip"
-download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", mode="w") 
-unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", 
-       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
+#download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", mode="w") 
+#unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb.zip", 
+#       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
 
-st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
+#st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_eksport.gdb2")
 
 # SOMETHING NOT WORKING YET
 
@@ -23,7 +23,7 @@ st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/naturovervaking_e
 
 #### upload data from P-drive ####
 ## ANO
-st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
+#st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
 ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
                    layer="ANO_Art")
 ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
@@ -31,9 +31,11 @@ ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervakin
 head(ANO.sp)
 head(ANO.geo)
 
+saveRDS(ANO.sp, "data/cache/ANO.sp.RDS")
[…]

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/3 Reference bootstraps_seminat.R

@@ -64,6 +64,8 @@ setwd("C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemC
 #### storing reference lists ####
 #rm(list= ls()[!(ls() %in% c('seminat.ref.cov.backup','seminat.ref.cov', 'ANO.sp.ind', 'ANO.geo', 'ASO.sp.ind', 'ASO.geo', 'NiN.seminat.cov','ind.dat'))])
 settings <- "iter=1000,obl=1,rat=1/3,var.abun=T"
+
+saveRDS(seminat.ref.cov, "data/cache/seminat.ref.cov.RDS")
 rm(list= ls()[!(ls() %in% c('seminat.ref.cov', 'seminat.ref.cov.backup','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
 

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ANO.R

@@ -115,33 +115,33 @@ ANO.seminat[ANO.seminat$kartleggingsenhet_1m2=="T32",]
 ANO.seminat[ANO.seminat$kartleggingsenhet_1m2=="T34",]
 # for T32 og T34 without grunntype-registration there's no useful reference, these won't be processed further
 
-results.seminat <- list()
+results.seminat.ANO <- list()
 ind <- unique(seminat.ref.cov.val$Ind)
 # choose columns for site description
 colnames(ANO.seminat)
-results.seminat[['original']] <- ANO.seminat
+results.seminat.ANO[['original']] <- ANO.seminat
 # drop geometry
-st_geometry(results.seminat[['original']]) <- NULL
-results.seminat[['original']]
+st_geometry(results.seminat.ANO[['original']]) <- NULL
+results.seminat.ANO[['original']]
 
 # add columns for indicators
-nvar.site <- ncol(results.seminat[['original']])
-for (i in 1:length(ind) ) {results.seminat[['original']][,i+nvar.site] <- NA}
-colnames(results.seminat[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
-for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.seminat[['original']][,i] <- as.numeric(results.seminat[['original']][,i])}
-summary(results.seminat[['original']])
-#results.seminat[['original']]$Region <- as.factor(results.seminat[['original']]$Region)
-results.seminat[['original']]$GlobalID <- as.factor(results.seminat[['original']]$GlobalID)
-results.seminat[['original']]$ano_flate_id <- as.factor(results.seminat[['original']]$ano_flate_id)
-results.seminat[['original']]$ano_punkt_id <- as.factor(results.seminat[['original']]$ano_punkt_id)
-results.seminat[['original']]$hovedoekosystem_punkt <- as.factor(results.seminat[['original']]$hovedoekosystem_punkt)
-#results.seminat[['original']]$Hovedoekosystem_rute  <- as.factor(results.seminat[['original']]$Hovedoekosystem_rute )
-results.seminat[['original']]$kartleggingsenhet_1m2 <- as.factor(results.seminat[['original']]$kartleggingsenhet_1m2)
-results.seminat[['original']]$hovedtype_rute    <- as.factor(results.seminat[['original']]$hovedtype_rute)
[…]

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/5 scaled value calculations_seminat_ASO.R

@@ -420,12 +420,12 @@ hist(results.seminat.ASO[['2-sided']]$Soil_disturbance2,breaks=40)
 #            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 #write.table(results.seminat.ASO[['non-truncated']], file='output/scaled data/results.seminat.ASO_non-truncated.txt',
 #            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
-write.table(results.seminat.ASO[['original']], file='P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO_original.txt',
-            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
-write.table(results.seminat.ASO[['2-sided']], file='P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO_2-sided.txt',
-            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
+#write.table(results.seminat.ASO[['original']], file='P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO_original.txt',
+#            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
+#write.table(results.seminat.ASO[['2-sided']], file='P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO_2-sided.txt',
+#            quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
-
+saveRDS(results.seminat.ASO, "data/cache/results.seminat.ASO.RDS")
 rm(list= ls()[!(ls() %in% c('ASO.geo','seminat.ref.cov.val','results.seminat.ASO','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.seminat.ASO.RData")
 

### 2023-05-10 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R

@@ -9,12 +9,12 @@ load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for m
 
 #### plotting scaled values by main ecosystem type ####
 ## continuing with 2-sided
-res.seminat <- results.seminat[['2-sided']]
-colnames(res.seminat)
+res.seminat.ANO <- results.seminat.ANO[['2-sided']]
+colnames(res.seminat.ANO)
 
 # make long version of the scaled value part
-res.seminat <-
-  res.seminat %>% 
+res.seminat.ANO <-
+  res.seminat.ANO %>% 
   pivot_longer(
     cols = c("Light1","Light2",
              "Moist1","Moist2",
@@ -29,8 +29,8 @@ res.seminat <-
   )
 
 # add original values as well
-res.seminat <- 
-  res.seminat %>% add_column(original = results.seminat[['original']] %>% 
+res.seminat.ANO <- 
+  res.seminat.ANO %>% add_column(original = results.seminat.ANO[['original']] %>% 
                                pivot_longer(
                                  cols = c("Light1","Light2",
                                           "Moist1","Moist2",
@@ -46,7 +46,7 @@ res.seminat <-
                                pull(original)
   )
 
-head(res.seminat[,70:76])
+head(res.seminat.ANO[,70:76])
 
 
 # similarly for ASO
@@ -92,7 +92,7 @@ head(res.seminat.ASO[,70:76])
 
 
 # making the plot, ANO
-ggplot(res.seminat, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+ggplot(res.seminat.ANO, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
   geom_hline(yintercept=0.6, linetype="dashed") + 
   geom_violin(color=NA) +
 #  geom_boxplot(width=0.2, color="grey") +
@@ -122,14 +122,14 @@ ggplot(res.seminat.ASO, aes(x=factor(NiN_grunntype2), y=scaled_value, fill=fp_in
 
 #### scaled value maps ####
 # keep wide format and add geometry again
-res.seminat2 <- results.seminat[['2-sided']]
-st_geometry(res.seminat2) <- st_geometry(ANO.seminat)
+res.seminat.ANO2 <- results.seminat.ANO[['2-sided']]
[…]

### 2023-05-10 R/functionalPlantIndicators/code wetland/3 Reference bootstraps_wetland.R

@@ -66,6 +66,9 @@ for (i in indID) {
 #rm(list= ls()[!(ls() %in% c('forest.ind.list','mire.ind.list','semiNat.ind.list','heath.ind.list','wetland.ref','ind.dat'))])
 #rm(list= ls()[!(ls() %in% c('wetland.ref.backup','wetland.ref','wetland.ref.cov.backup','wetland.ref.cov', 'ANO.sp.ind', 'NiN.wetland.cov','ind.dat'))])
 settings <- "iter=1000,obl=1,rat=1/3,var.abun=T"
+
+saveRDS(wetland.ref.cov, "data/cache/wetland.ref.cov.RDS")
+
 rm(list= ls()[!(ls() %in% c('wetland.ref.cov', 'wetland.ref.cov.backup','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
 

### 2023-05-10 R/functionalPlantIndicators/code wetland/5 scaled value calculations_wetland.R

@@ -333,7 +333,7 @@ write.table(results.wetland[['original']], file='output/scaled data/results.wetl
 write.table(results.wetland[['2-sided']], file='C:/Users/joachim.topper/OneDrive - NINA/work/R projects/github/ecosystemCondition/R/functionalPlantIndicators/output large files for markdown/results.wetland_2-sided.txt',
             quote=FALSE,sep="\t",col.names=TRUE,row.names=FALSE,dec=".")
 
-
+saveRDS(results.wetland, "data/cache/results.wetland.RDS")
 rm(list= ls()[!(ls() %in% c('wetland.ref.cov.val','ANO.wetland','results.wetland','settings'))])
 save.image("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")
 

### 2023-06-07 DRAFT_trophic_levels-biomass_indicators.qmd

---
title: "TrophicLevelBiomassRatios"
author: "James D. M. Speed"
format: 
  html:
    embed-resources: TRUE
editor: visual
knitr:
  opts_knit:
    root.dir: "S://R projects//TrophicIntIndicators"
execute: 
  cache: true
---

## Introduction

Here we will develop indicators of ecological condition for Norway, based on the distribution of biomass between trophic levels - specifically between vegetation and vertebrate consumers. This includes three trophic levels, plants, herbivores and carnivores. Indicators of trophic interactions are based on deviation between the observed biomass at each consumer trophic level from the expected biomass based on the biomass at the prey trophic level. The expected biomass estimates are derived from global relationships between plants and vertebrate herbivores ([Fløjgaard et al. 2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N))

As the plant biomass level we use Net Primary Productivity. MODIS NPP was selected since it provides the highest temporal extent of available data sources (1999 to present). Spatial resolution is low relative to other remote sensing sources, however, the spatial resolution is still far higher than the consumer data, so the temporal extent was prioritised.

The herbivore data is an extension of the data presented in Austrheim et al. 2008 <https://onlinelibrary.wiley.com/doi/pdf/10.2981/10-038> and Speed et al. 2015 <https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0217166>. For the development of the trophic interaction indicators the same methods were applied to herbivore data going back to 1907 (contra 1949 in the two studies above). This data was extracted by Gunnar Austrheim (livestock) and Erling J. Solberg (wild cervids). In each dataset, the metabolic biomass and the raw biomass of a given species is provided for each year and municipality. The 2007 municipalities were used. The data were rasterized to the same grid as the NPP data.

Metabolic biomass measures were used in the previous studies on Norwegian herbivores cited above. However, for consistency with the global studies used to find predicted biomass, we use raw (unscaled) biomass here.

Herbivore species are wild herbivores: moose/elg, red deer/hjort, roe deer/rådyr, wild reindeer/villrein and livestock: cattle/storfe, sheep/sau, horses/hest, goats/geit and semi-domestic reindeer/tamrein

The carnivore data is taken from back-cast population modelling for the large carnivores in Norway (wolf/ulv, brown bear/bjørn, lynx/gaupe and wolverine/jerv). The data was based on hunting statistics, and was collated and modelled by Anna Sobocinski in a master thesis: <https://ntnuopen.ntnu.no/ntnu-xmlui/handle/11250/3047580> The thesis also presented metabolic biomass, but we use raw biomass to make the indicators. This data was also rasterized.

We calculate indicators (1) across all ecosystems and the total consumer communities (all localities and all herbivores and all carnivores) and (2) specific indicators for each main ecosystem type

## Part 1: Total communities

We start by calculation biomass ratios across all herbivores and carnivores, with no regard to the ecosystem type.

```{r}
#| label: Load packages
#| include: false
#| cache: TRUE

library(terra)
library(tidyterra)
library(ggplot2)
library(sf)
library(scales)
library(RColorBrewer)
library(tidyverse)

```

```{r}
#| label: Data import
#| cache: TRUE

#NPP
#Read in
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")

#Vector of years we work with
allyears_vect<-c(1907,1917,1929,1938,1949,1959,1969,1979,1989,1999,2009,2015)

#Livestock data from 1939 and vilt from 1938
names(livestockrast)<- sub(1939,1938,names(livestockrast))
names(livestockrast)

#Norwegian counties to help plotting
norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
#Simplify by county nr
norcounty<-norcounty_shp %>% 
  group_by(FylkeNr) %>%
  summarise(geometry = st_union(geometry))

```

Now we can plot the raw biomass data.

First NPP

```{r}
#| label: Plotting basic data
#| cache: TRUE

#Select relevant years for NPP
npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]

ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
         geom_spatraster(data=npprast_yrs)+facet_grid(~lyr)+
         scale_fill_continuous(low="lightgreen",high="darkgreen",na.value=NA,expression('kgC km'^-2~'year'^-1))+theme_bw()

```

Next total herbivore biomass

```{r}
#| label: PlotTotalHerbivoreBiomass
#| cache: TRUE
#Total herbivore biomass
#For some reason need to read the raster in again
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
totalherbivorebiomass
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  geom_spatraster(data=totalherbivorebiomass)+facet_wrap(~lyr)+
  scale_fill_continuous(trans='log10',low="lightskyblue",high="darkblue", na.value=NA,expression('Herbivore \nbiomass \nkg km'^-2 ))+theme_bw()

```

And finally total carnivore biomass

```{r}
#| label: Plot carnivore biomass
#| cache: TRUE
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")

totalcarnivores<-carnivorerast[[49:60]]

ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  geom_spatraster(data=totalcarnivores)+facet_wrap(~lyr)+
  scale_fill_continuous(low='white',high='red', na.value=NA, expression('Carnviore\nBiomass \nkg km'^-2 ))+theme_bw()

```

Now we use the NPP to estimate the expected herbivore biomass Fløjgaard et al. 2015 Figure 1/Table 1 Global relationship (mean NPP1km)

```{r}
#| label: Calculating expected herbivore biomass
#| cache: TRUE
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]

expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  ggtitle("Expected herbivore biomass")+
  geom_spatraster(data=expected_herbivore_biomass)+facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
writeRaster(expected_herbivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif",overwrite=TRUE)

```

Now we can calculate the deviation between the observed herbivore biomass and the expected herbivore biomass. We calculate this as a % change.

100 x (ActualBiomass -- ExpectedBiomass)/(ExpectedBiomass +1)

To display this we truncated to 100%\|-100%

A value of 0 indicates that the locality has expected biomass. Value of 100 indicates that the locality has 100% (or more) higher biomass than expected Value of -100 indicates that the locality has 100% (or more) lower biomass than expected

«Good» condition assumed to be 0 to \|40%\|

```{r}
#| label: DifferenceHerbivoreBiomass
#| cache: TRUE
expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
totalherbivorebiomass_npp<-totalherbivorebiomass[[names(totalherbivorebiomass) %in% c("1999","2009","2015")]]

 diffherb<-100*(totalherbivorebiomass_npp-expected_herbivore_biomass)/(expected_herbivore_biomass+1)
 names(diffherb)<-c("2000",'2009','2015')

# #Plot on a diverging colour scale around 0
 ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=diffherb)+
   ggtitle("Difference from expected herbivore biomass")+
   scale_fill_distiller(type="div",limit=c(-100,100),oob=squish, na.value=NA,'% Difference from \nexpected biomass')+
   facet_wrap(~lyr)+theme_bw()

writeRaster(diffherb,"Vertebrates/outputs/HerbivoreBiomassDiffExpected.tif",overwrite=TRUE)

```

Above we see that on the West coast, Trøndelag, Oppland have double herbivore biomass than expected. There is more herbivore biomass than expected in parts of Finnmark, while Nordland and Lierne have lower herbivore biomass than expected based on NPP. Vestfold & Telemark (and parts of Viken) change from having more herbivore biomass than expected in 2000 to less than expected in 2015 (this is consistent with reduction in moose population in these regions)

-- Next we move on to estimate the expected carnivore biomass based on observed herbivore biomass and calculate the deviation as a % difference from expected as for the herbivores. This is based on Hatton et al. 2015 Figure 1 (predator prey biomass relationships)

```{r}
#| label: Expected carnivore biomass
#| cache: TRUE

livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")

totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
names(totalherbivorebiomass)<-allyears_vect
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
totalcarnivores<-carnivorerast[[49:60]]

#Expected carnivore biomass 
#Taken from Hatton et al. Figure 1
expected_carnivore_biomass<-(totalherbivorebiomass^0.73)*0.014#African carnivore and herbivore communities
ggplot()+geom_spatraster(data=expected_carnivore_biomass)+
  ggtitle("Expected carnivore biomass")+
  facet_wrap(~lyr)+theme_bw()+
  scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2),trans='log10')

diffcarn<-100* (totalcarnivores-expected_carnivore_biomass)/(expected_carnivore_biomass+1)
names(diffcarn)<-allyears_vect

ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
  geom_spatraster(data=diffcarn)+
  ggtitle('Difference from expected carnivore biomass')+
  scale_fill_distiller(type="div",limit=c(-100,100),oob=squish,na.value=NA,"%difference from \nexpected biomass")+
  facet_wrap(~lyr)+theme_bw()

writeRaster(expected_carnivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif",overwrite=TRUE)

writeRaster(diffcarn,"Vertebrates/outputs/CarnivoreBiomassDiffExpected.tif",overwrite=TRUE)

```

From the above we see that Norway has had less carnivore biomass than expected based on the availabile herbivore biomass across the time period. Some counties and time periods (e.g. Hedmark today and Finnmark in 1907 had about the same carnivore biomass as predicted.

## Part 2 - Habitat specific biomass distributions

We now create biomass ratios for specific habitat types. To do this we need an ecosystem map and to place each consumer species within each habitat.

Indicators were not developed for mire ecosystems. This decision was taken in conversation with the mire ecosystem group, on the basis that herbivores have minimum use of the mrie ecosystems.

Table of the consumer species associated with each habitat.

| Ecosystem             | Livestock               | Wild herbivores   | Carnivores        |
|------------------|-------------------|------------------|------------------|
| Forest                | Storfe                  | Elg, rådyr, hjort | Ulv, gaupe, bjørn |
| Alpine                | Sau, tamrein, geit      | Villrein          | Jerv              |
| Semi-naturlig mark    | Sau, hest, geit, storfe | Hjort             | Gaupe, ulv        |
| Naturlig åpne områder | Sau, geit, storfe       | Hjort             | Gaupe, ulv        |
| Myr                   | \-\--                   | \-\--             | \-\--             |

Since a suitable ecosystem map is still under development, we use the AR50 land-cover map as a place-holding solution. AR50 does not give a good representation of naturlig åpen områder or semi-naturlig mark, so we demonstrate the habitat specific biomass distributions using the forest class in AR50 and the species outlined above.

```{r}
#| label: Habitat specific rasters
#| cache: TRUE

#Forest
#AR50 type raster
artype50<-rast("T:\\vm\\alle\\GIS_data\\Norge\\Cartography\\N50\\AR50\\AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")
#artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")

#Reclassify
artype50_F<-artype50
arTdf<-data.frame(id=c(10,20,30,50,60,70,81,82,128),label=c("BuiltUp","Pastures/Arable","Forest","Open","Mires","Ice","Freshwater","Sea","Unmapped"))
levels(artype50_F)<-arTdf

arcols<-c("black","orange","darkgreen","wheat","lightblue","white","blue","blue",NA)
ggplot()+geom_spatraster(data=artype50_F)+scale_fill_manual(values=arcols,na.value = NA)

#Load biomass rasters
npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")

#Project the AR50 to the NPP rasters
artype50_FR<-project(resample(artype50_F,npprast,method='near'),npprast)

#Combine herbivores
allherbivores<-c(viltrast,livestockrast)

#Mask out forest from rasters
forest_expected_herbivore_biomass<-mask(expected_herbivore_biomass,artype50_FR,maskvalues=30,inverse=TRUE)
forest_herbivores<-mask(allherbivores,artype50_FR,maskvalues=30,inverse=TRUE)
forest_carnivores<-mask(carnivorerast,artype50_FR,maskvalues=30,inverse=TRUE)
forest_expected_carnivore_biomass<-mask(expected_carnivore_biomass,artype50_FR,maskvalues=30,inverse=TRUE)

#Sum up selected forest herbivore species
#Here we select the relevant herbivore species for the ecosystem type
allyears_vect<-c(1907,1917,1929,1938,1949,1959,1969,1979,1989,1999,2009,2015)

forest_herbivores_selectspp<-rast(forest_herbivores,nlyrs=12)
for(i in 1:length(allyears_vect)){
  print(i)
  forest_herbivores_selectspp[[i]]<-sum(forest_herbivores[[sapply(strsplit(names(forest_herbivores),"_"),'[',3)
                                                         %in% allyears_vect[i] &
                                      sapply(strsplit(names(forest_herbivores),"_"),'[',2)
                                                          %in% c("roe","hjort","elg","storf")]]
)
}
names(forest_herbivores_selectspp)<-allyears_vect

 #Sum up selected forest carnivore species
 #Here we select the relevant carnivore species for the habitat type
 forest_carnivores_selectspp<-rast(forest_carnivores,nlyrs=12)
 for(i in 1:length(allyears_vect)){
   print(i)
   forest_carnivores_selectspp[[i]]<-sum(forest_carnivores[[sapply(strsplit(names(forest_carnivores),"_"),'[',3) 
                                                            %in% allyears_vect[i] &
                                                              sapply(strsplit(names(forest_carnivores),"_"),'[',2)
                                                            %in% c("Wolf","Lynx","Bear")]]
   )
 }
 names(forest_carnivores_selectspp)<-allyears_vect

 #Difference in expected and actual biomass in forest
 forest_herbivores_selectspp_nppyrs<-forest_herbivores_selectspp[[names(forest_herbivores_selectspp)%in% c("1999","2009","2015")]]

forest_diffherb<-100*(forest_herbivores_selectspp_nppyrs - forest_expected_herbivore_biomass)/
 (forest_expected_herbivore_biomass+1)

norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
   group_by(FylkeNr) %>%
   summarise(geometry = st_union(geometry))

ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=forest_diffherb)+
   ggtitle("Difference in forest herbivores from expectation")+
   scale_fill_distiller(type="div",palette='RdYlBu',limit=c(-100,100),oob=squish,na.value=NA, "% difference from \nexpected biomass")+
   facet_wrap(~lyr)+theme_bw()

forest_diffcarn<-100*(forest_carnivores_selectspp-forest_expected_carnivore_biomass)/
   (forest_expected_carnivore_biomass+1)

ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=forest_diffcarn)+
   ggtitle("Difference from expected carnivore biomass")+
   scale_fill_distiller(type="div",palette='RdYlBu',limit=c(-100,100),oob=squish,na.value=NA,"% difference from\nexpected biomass")+
   facet_wrap(~lyr)+theme_bw()

# #Scaling as indicators

 ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=forest_diffherb)+
   ggtitle("Difference from expected herbivore biomass")+
  scale_fill_gradientn(colors= c("blue","yellow",'yellow' ,  "red"),
                       breaks=c(-100,-40,0,40,100),limits=c(-100,100),na.value=NA,"% difference from\nexpected biomass")+
   facet_wrap(~lyr)+theme_bw()

 ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=forest_diffcarn)+
   ggtitle("Difference from expected carnivore biomass")+
   scale_fill_gradientn(colors= c("blue","yellow",'yellow' ,  "red"),
                     breaks=c(-100,-40,0,40,100),limits=c(-100,100),na.value=NA,"% difference from\nexpected biomass")+
   facet_wrap(~lyr)+theme_bw()

 writeRaster(forest_diffcarn,"Vertebrates/outputs/forest_carnivores_diffexpected.tif",overwrite=TRUE)
 writeRaster(forest_diffherb,"Vertebrates/outputs/forest_herbivores_diffexpected.tif",overwrite=TRUE)
```

In the above maps, yellow indicates good condition (within 40% of expected biomass between 60% and 140% of expected biomass). Blue indicates poor condition, with \>40% less biomass than expected (i.e. \<60% of expected biomass), while red indicates poor conditions with \>40% biomass than expected (i.e. \>140% of expected biomass).

## Part 3 County averages over time

Finally we summarise the biomass deviations by county over time. 0% deviation represents reference condition. The lines at 40% and -40% show the limits for good condition.

```{r}
#| label: CountySummaries
#| cache: TRUE

forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")

norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
 #Simplify by county nr
norcounty<-norcounty_shp %>% 
   group_by(FylkeNr) %>%
   summarise(geometry = st_union(geometry))

#County timegraphs
#Make spatVect 
norcounty$CountyName<-norcounty$FylkeNr
#Provide names (some issues with encoding...)
norcounty$CountyName<-c("_stfold","Akershus","Oslo","Hedmark","Oppland","Buskerud","Vestfold","Telemark","Aust-Agder","Vest-Agder","Rogaland","Hordaland","Sogn og Fjordane",
                        "M_re og Romsdal","S_r-Tr_ndelag","Nord-Tr_ndelag","Nordland","Troms","Finnmark")
norcounty_vect<-vect(norcounty)

forest_carn_herb_county<-terra::extract(forest_diffcarn,norcounty_vect,mean,na.rm=T)
forest_carn_herb_county$ID<-norcounty_vect$FylkeNr#Replace IDs with Fylke Nrs
forest_carn_herb_county$CountyName<-norcounty_vect$CountyName#Replace IDs with Fylke Nrs
forest_carn_herb_countyDF<-gather(forest_carn_herb_county,key="Year",value="Difference",-ID,-CountyName)
forest_carn_herb_countyDF$YearN<-as.numeric(substr(forest_carn_herb_countyDF$Year,2,5))
ggplot(data=forest_carn_herb_countyDF,aes(x=YearN,y=Difference,color=CountyName))+geom_line()+scale_color_discrete()+
  ggtitle("Forest: % Difference from expected carnivore biomass")+theme_bw()+xlim(c(1905,2025))+geom_hline(yintercept=c(-40,0),lty=c(2,1))+
  geom_text(data=forest_carn_herb_countyDF[forest_carn_herb_countyDF$YearN==2015,],aes(label = CountyName, x = 2015, y = Difference), hjust = -.1)

forest_herb_veg_county<-terra::extract(forest_diffherb,norcounty_vect,mean,na.rm=T)
forest_herb_veg_county$ID<-norcounty_vect$FylkeNr#Replace IDs with Fylke Nrs
forest_herb_veg_county$CountyName<-norcounty_vect$CountyName#Replace IDs with Fylke Nrs
forest_herb_veg_countyDF<-gather(forest_herb_veg_county,key="Year",value="Difference",-ID,-CountyName)
forest_herb_veg_countyDF$YearN<-as.numeric(substr(forest_herb_veg_countyDF$Year,2,5))
ggplot(data=forest_herb_veg_countyDF,aes(x=YearN,y=Difference,color=CountyName))+geom_line()+scale_color_discrete()+
  ggtitle("Forest: % Difference from expected herbivore biomass")+theme_bw()+xlim(c(1999,2018))+geom_hline(yintercept=c(-40,0,40),lty=c(2,1,2))+
  geom_text(data=forest_herb_veg_countyDF[forest_herb_veg_countyDF$YearN==2015,],aes(label = CountyName, x = 2015, y = Difference), hjust = -.1)

```

### 2023-06-08 alien_species.Rmd
Render book with trophic level biomass

@@
----
-title: "Alien species"
-output:
-  html_document
----
 # Alien species {#alien-species}
@@
 library(knitr)
-library(tidyverse)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
 library(sf)

### 2023-06-08 climate_indicators_explained.Rmd
Render book with trophic level biomass

@@
 knitr::opts_chunk$set(echo = TRUE)
-#library(tidyverse) # cant update to v2 on server
-library(dplyr)
-library(ggplot2)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse) 
 library(stars)
@@
 dir <- substr(getwd(), 1,2)
-
+pData <- ifelse(dir == "C:", 
+  "P:/41201785_okologisk_tilstand_2022_2023/data/",
+  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
 ```
@@
 I describe the process using dummy data and also discuss alternate approaches and dead ends that occured in the development process.
-For a succinct workflow example for a spesific indikcator, see [Median summer temperature](#median-summer-temp).
+For a succinct workflow example for a specific indicator, see [Median summer temperature](#median-summer-temp).
 
@@
 ```{r, fig.cap="Five accounting areas (regions) in Norway."}
-
 tm_shape(reg) +
   tm_polygons(col="region")
-
 ```
@@
 The climate indicators need to be masked with ecosystem type maps. This step is part of this chapter.
-
-## --------------------------------------------------
 
@@
 
-### 1. Collate variable data series
+### Collate variable data series
 
@@
 
-#### Filter data by dates
+### Filter data by dates
 
@@
 
-#### Aggregate across time within a year
+### Aggregate across time within a year
 
@@
 These files are about 500 MB, but the tiff file reads in a lot quicker then the rds file, so from now on I'll just export as tiff.
-*(This could perhaps have been solved with a `gc()`,.
-Later I save as RData with no problems).*
-
-```{r}
-summer_median_temp <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
+*(This could perhaps have been solved with a `gc()`.
+Later I tried saving as RData with no speed issues problems, but then I had a problem with RData files seemingly becoming corrupted over time - stars objects saved as RData files would not bind together using `c.stars()` with other matching stars object and I think this could be related to a time stamp in the RData format.
+So, stick to tiff).*
+
+```{r}
+summer_median_temp <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                  proxy=F)
-# This is too slow:
-# readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/dummy_data/summer_median_temp.rds")
-# but it could be that RData works better 
 ```
@@
 
-### 2. Calculate reference values
+### Calculate reference values
 
@@
 ```{r, eval=F}
-# write_stars can only export a single attribute 
-#write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
-
-saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
+# Note that write_stars can only export a single attribute, but we have put the 'attributes' as dimensions, and these will be read in as 'band' when importing tiff.
+write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
+
+# I initially saved as RData, like this, but ran into problems later when trying to bind stars objects using c.stars().
+It appears RData becomes corrutpt in a sense, over time, perhaps due to some time stamp.
+If I recreated the object in the global environment and saved as RDatam I could read it in and work with it, but the next day it would not work again.
+#saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
 ```
@@
 ```{r, eval=T, include=F}
-summer_median_temp_ref <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
+summer_median_temp_ref <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff"),
+                                     proxy=F)
 ```
@@
 ```{r, eval=T}
-summer_median_temp_ref_long <- summer_median_temp_ref %>% split("median_sd")
+summer_median_temp_ref_long <- summer_median_temp_ref %>% split("band")
 ```
@@
   c(summer_median_temp_ref_long)
-
-```
-
-#### Normalise variable
+```
+
+### Normalise variable
 
@@
 
+Writing to file, this time as RData (could use rds) because tiff don't allow multiple attributes, and because we don't need to use `c.stars()` in this object.
+
 ```{r, eval=F}
@@
 ```{r, include=FALSE}
-y_var_norm <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData")
+y_var_norm <- readRDS(paste0(pData, "climate_indicators/aggregated_climate_time_series/temp_median_summer_normalised.RData"))
 ```
@@
 
-#### Plot time series as line graphs
+### Plot time series as line graphs
 

### 2023-06-08 functional_plant_indicators_wetland.Rmd
Render book with trophic level biomass

@@
 library(plyr)
-library(tidyverse)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
 library(RColorBrewer)

### 2023-06-08 homogeneous_impact_areas.Rmd
Render book with trophic level biomass

--- 2023-02-02 DRAFT_homogeneous_impact_areas.Rmd
@@
----
-title: "-"
-output:
-  html_document
-editor_options: 
-  markdown: 
-    wrap: sentence
----
-
 # Homogeneous Impact Areas
@@
 library(DT)
-library(tidyverse)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
+library(ggplot2)
+library(dplyr)
 library(units)

### 2023-06-08 index.Rmd
Render book with trophic level biomass

@@
 
+```{r workaround-lib-issue, echo=FALSE, warning=FALSE, message=FALSE}
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+```
+
 ```{r eaTools-example, fig.cap="A normalised condition variable is called an indicator. This indicator had been transformed with a sigmoid function."}
 #devtools::install_github("NINAnor/eaTools")
+
 library(eaTools)
@@
 
-Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]
+Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. (2023) Indicators for Ecosystem Condition in Norway. https://doi.org/10.5281/zenodo.7854835
 
-## License
+## License {.unnumbered}
 

### 2023-06-08 medianSummerTemperature.Rmd
Render book with trophic level biomass

@@
-# Median Summer Temperature
+# Median Summer Temperature {#median-summer-temp}
 
@@
 knitr::opts_chunk$set(echo = TRUE)
-library(tidyverse) # cant update to v2 on server
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
 library(stars)
@@
 dir <- substr(getwd(), 1,2)
+pData <- ifelse(dir == "C:", 
+  "P:/41201785_okologisk_tilstand_2022_2023/data/",
+  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
 ```
@@
 This chapters describes the workflow for an indicator describing the median summer temperature.
-For a more comprehensive documentaion on the development of the workflow itself, se [here](#climate-indicators-explained).
+For a more comprehensive documentation on the development of the workflow itself, see [here](#climate-indicators-explained).
 The data comes from interpolated climate surfaces from [SeNorge](https://senorge.no/) which contain one 1x1km raster for each day since 1957 to present.
@@
 
-```{r, eval=F}
-# a backup caching in case tiff causes problems down the line
-saveRDS(real_temp_summer, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.rds")
-
-```
-
 ```{r, include=F}
-real_temp_summer <- read_stars("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff",
+real_temp_summer <- read_stars(paste0(pData, "climate_indicators/aggregated_climate_time_series/real_temp_summer.tiff"),
                                  proxy=F)
@@
 
-<!-- RUN AGAIN -->
 ```{r, eval=FALSE}
@@
 ```{r, eval=F}
-saveRDS(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.RData")
+write_stars(summer_median_temp_ref, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_ref.tiff")
 ```
@@
 summer_median_temp_ref_long <- summer_median_temp_ref %>% 
-  split("median_sd") %>%
+  split("band") %>%
   setNames(c("reference_upper", "sd"))
@@
 saveRDS(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.RData")
+
+# Tiff dont allow for multiple attributes:
+#write_stars(y_var_norm, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/climate_indicators/aggregated_climate_time_series/summer_median_temp_normalised.tiff")
 ```

### 2023-06-08 naturtype.Rmd
Render book with trophic level biomass

@@
 
-**On the application of the nature type dataset**
+**On the application of the nature type data set**
 
@@
 library(data.table)
-library(stringr)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+#library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
 library(forcats)
@@
 
-Author: Anders L. Kolstad
-
-Date: `r Sys.Date()`
+Author:
+
+Anders L. Kolstad
+
+March 2023
 
@@
 ```{r import_nin_data, results="hide"}
+dir <- substr(getwd(), 1,2)
+
 # local path
 #path <- "data/R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
-
-# server path
-path <- "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
-
 # temporary path 
 #path <- "/data/Egenutvikling/41001581_egenutvikling_anders_kolstad/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.g#db"
+
+# server path pre 2022 data
+#path <- "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/Naturtyper_nin_0000_norge_25833_FILEGDB/Naturtyper_nin_0000_norge_25833_FILEGDB.gdb")
+
 dat <- sf::st_read(dsn = path)
@@
 
-The dataset has 95k polygons, each with 36 variables:
+The data set has 117k polygons, each with 37 variables:
 ```{r}
@@
 ```
-The smallest polygons are 20 m2, and the biggest is 8.5 km2.
-
-The largest polygon is a 
-```{r, echo=F}
-dat$naturtype[dat$area == max(dat$area)]
-```
-
-The smallest polygon is a 
-```{r, echo=F}
-dat$naturtype[dat$area == min(dat$area)]
-```
+The smallest polygons are 20 m2, and the biggest is 15.8 km2.
+
+The largest polygon is a `r dat$naturtype[dat$area == max(dat$area)]`, and the smallest polygon is a `r dat$naturtype[dat$area == min(dat$area)]`.
 
@@
 ```
-About 0.5% of Norway has been mapped (note that a bigger area than this has been surveyed, but only a small fraction is delineated). It is therefore essential that these 0.5% are representative.
+About 0.6% of Norway has been mapped (note that a bigger area than this has been surveyed, but only a small fraction is delineated). It is therefore essential that these 0.5% are representative.
 
@@
 
-The rate of nature type mapping is strongly positive last three years.
-There are some differences in the field mapping instructions between the years.
-I will need to decide whether to include all years, or to perhaps exclude the first year.
+There are some differences in the field mapping instructions between the years. I will need to decide whether to include all years, or to perhaps exclude the first year.
 
@@
 ```{r, fig.cap="The overal distribution of condition scores"}
+dat <- dat %>%
+  mutate(tilstand =  recode(tilstand, 
+                            "sværtRedusert" = "1 - Svært redusert",
+                            "dårlig" = "2 - Dårlig",
+                            "moderat" = "3 - Moderat",
+                            "god" = "4 - God"))
+
 barplot(tapply(dat$area/10^6, factor(dat$tilstand), sum))
@@
 
-There are only six nature types (if you exclude those that are not mapped) that don't have a condition score.
+There are only seven nature types (if you exclude those that are not mapped) that don't have a condition score.
 
@@
 `Isinnfrysingsmark` is assessed for condition now, but not in 2018.
+`Snøleieberg` is new in 2022 and is only mapped for area.
 We can therefore exclude these localities from our data set:
@@
 #unique(dat$hovedøkosystem)
-dat$hovedøkosystem[dat$hovedøkosystem=="Naturlig åpne områder i lavlandet"] <- "Naturlig åpne områder under skoggrensa"
+dat$hovedøkosystem[dat$hovedøkosystem=="naturligÅpneOmråderILavlandet"] <- "naturligÅpneOmråderUnderSkoggrensa"
 ```
@@
 par(mar=c(5,20,1,1))
-barplot(table(dat$hovedøkosystem[dat$mosaikk=="Ja"]),
+barplot(table(dat$hovedøkosystem[dat$mosaikk=="ja"]),
         horiz = T, las=2,
@@
 ## NiN types across main ecosystems
-The NiN main types can be extracted from the column `ninkartleggingsenheter`.
-This is the mapping NiN mapping units recorded in the field.
+The NiN main types can be extracted from the column `ninKartleggingsenheter`.
+These are NiN mapping units recorded in the field.
 The NiN main type makes up the first part of this mapping unit code.
@@
 ```{r, fig.cap="The number of NiN main types within one locality should normally be one, except for mosaic localities."}
-#dat$ninkartleggingsenheter[1:10]
-dat$ninkartleggingsenheter2 <- gsub("NA_", dat$ninkartleggingsenheter, replacement = "")
-#dat$ninkartleggingsenheter2[1:30]
-dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, ".[CE].[\\d]{1,}", replacement = "")
-#dat$ninkartleggingsenheter2[1:30]
-dat$ninkartleggingsenheter2 <- str_replace_all(dat$ninkartleggingsenheter2, "-.", replacement = "")
+#dat$ninKartleggingsenheter  [1:10]
+# remove NA prefix:
+dat$ninKartleggingsenheter2 <- gsub("NA_", dat$ninKartleggingsenheter, replacement = "")
+#dat$ninKartleggingsenheter2[1:30]
+
+dat$ninKartleggingsenheter2 <- str_replace_all(dat$ninKartleggingsenheter2, ".[CE].[\\d]{1,}", replacement = "")
+#dat$ninKartleggingsenheter2[1:30]
+dat$ninKartleggingsenheter2 <- str_replace_all(dat$ninKartleggingsenheter2, "-.", replacement = "")
 uni <- function(x){paste(unique(unlist(strsplit(x, ","))), collapse = ",")}
-dat$ninkartleggingsenheter3 <- lapply(dat$ninkartleggingsenheter2, uni)
+dat$ninKartleggingsenheter3 <- lapply(dat$ninKartleggingsenheter2, uni)
 n_uni <- function(x){length(unique(unlist(strsplit(x, ","))))}
-dat$n_ninkartleggingsenheter <- lapply(dat$ninkartleggingsenheter2, n_uni)
-dat$n_ninkartleggingsenheter <- as.numeric(dat$n_ninkartleggingsenheter)
+dat$n_ninKartleggingsenheter <- lapply(dat$ninKartleggingsenheter2, n_uni)
+dat$n_ninKartleggingsenheter <- as.numeric(dat$n_ninKartleggingsenheter)
 
 par(mfrow=c(1,3))
-barplot(table(dat$n_ninkartleggingsenheter),
+barplot(table(dat$n_ninKartleggingsenheter),
         xlab = "Number of NiN main types",
         ylab = "Number of localities")
-barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Nei"]),
+barplot(table(dat$n_ninKartleggingsenheter[dat$mosaikk=="nei"]),
         xlab = "Number of NiN main types\n(Mosaic localities excluded)")
-barplot(table(dat$n_ninkartleggingsenheter[dat$mosaikk=="Ja"]),
+barplot(table(dat$n_ninKartleggingsenheter[dat$mosaikk=="ja"]),
         xlab = "Number of NiN main types\n(Mosaic localities only)")
@@
 
-Mosaic localities (right pane) have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (about 8000 cases).
+Mosaic localities (right pane) have a much higher likelihood of including multiple NiN main types, but it also occurs in non-mosaic localities (over 10 000 cases).
 Most, if not all, nature types are defined within a single NiN main type, so we need to see whats going on there.
@@
 ```{r}
-dat_melt <- tidyr::separate_rows(dat, ninkartleggingsenheter3)
+dat_melt <- tidyr::separate_rows(dat, ninKartleggingsenheter3)
 ```
@@
 ```{r, fig.cap="Figure showing the number of localities for each NiN main type. One locality can have more than one NiN main type. Mosaic localities are excluded in this figure."}
-ggplot(dat_melt[dat_melt$mosaikk!="Ja",])+
-  geom_bar(aes(x = ninkartleggingsenheter3))+
+ggplot(dat_melt[dat_melt$mosaikk!="ja",])+
+  geom_bar(aes(x = ninKartleggingsenheter3))+
   coord_flip()+
@@
 
-There is a lot of miss-match between NiN main types and *hovedøkosystem*. 
-Taking one example: T2 (åpen grunnledt mark) is not found in mountains, but there is one case where this occurs. If I view that case
+This is a messy figure, but the point is, there is a lot of miss-match between NiN main types and *hovedøkosystem*.
+Taking one example: T2 (åpen grunnledt mark) is not found in mountains, but there is one case where this occurs.
+If I view that case
 
 ```{r, eval= FALSE}
-View(dat_melt[dat_melt$mosaikk!="Ja" & dat_melt$hovedøkosystem=="Fjell" & dat_melt$ninkartleggingsenheter3=="T2",])
+View(dat_melt[dat_melt$mosaikk!="Ja" & dat_melt$hovedøkosystem=="Fjell" & dat_melt$ninKartleggingsenheter3=="T2",])
 ```
@@
 Miljødirektoratet could consider adding this information to the downloadable dataset.
-The order of the NiN types in the `ninkartleggingsenheter` column is not reflecting the commonness of the types ether.
+The order of the NiN types in the `ninKartleggingsenheter` column is not reflecting the commonness of the types ether.
 
@@
 Miljødirektoratet may consider adding this as well.
-If we exclude all localities that are not mosaics, but that have multiple NiN main types, we first need to know if there are not any `naturtype` which allow for multiple NiN main types.
-
-As this next figure show, these cases are not restricted to mapping year, and hence to any changes in the field protocol. Rather this phenomenon is increasing in commonness.
+If we exclude all localities that are not mosaics, but that have multiple NiN main types (no, we're not going to do that, here at least), we first need to know if there are not any `naturtype` which allow for multiple NiN main types.
+
+As this next figure show, these cases are not restricted to mapping year, and hence to any changes in the field protocol.
 ```{r, fig.cap="Temporal trend in the recording of non-mosaic localities that are recorded as consisting of multiple NiN mapping units from different NiN main types."}
 par(mar=c(2,6,2,2))
-temp <- dat[dat$n_ninkartleggingsenheter>1 & dat$mosaikk!="Ja",]
+temp <- dat[dat$n_ninKartleggingsenheter>1 & dat$mosaikk!="ja",]
 barplot(table(temp$kartleggingsår), ylab="Number of localities with more than one NiN main type\n(mosaic localitie excluded) ")
@@
 Lets look at the most common nature types that are recorded this way.
-```{r, fig.cap="The ten most common nature types and the number of localities with with multiple NiN main types"}
-temp2 <- as.data.frame(table(temp$naturtype, temp$n_ninkartleggingsenheter))
+```{r multipleNiN, fig.cap="The ten most common nature types and the number of localities with with multiple NiN main types"}
+temp2 <- as.data.frame(table(temp$naturtype, temp$n_ninKartleggingsenheter))
 temp2 <- temp2[base::order(temp2$Freq, decreasing = T),][1:10,]
@@
 'Hule eiker' is a special case because these are recorded as points and not polygons, and they can be found in any NiN main type and any *hovedøkosystem*.
+
 ```{r, ="Table of the first 5 cases of Hule eiker, showing that this nature type can be found in several NiN main types and hovedøkosystem"}
 DT::datatable(
-  dat[dat$naturtype=="Hule eiker",][1:5,c("naturtype", "hovedøkosystem", "ninkartleggingsenheter")])
+  dat[dat$naturtype=="Hule eiker",][1:5,c("naturtype", "hovedøkosystem", "ninKartleggingsenheter")])
 
@@
 ```{r}
-target <- c("Semi-naturlig mark","Våtmark","Naturlig åpne områder under skoggrensa")
+target <- c("semi-naturligMark",
+            "våtmark",
+            "naturligÅpneOmråderUnderSkoggrensa")
 dat <- dat[dat$hovedøkosystem %in% target,]
@@
 
-I will also exclude `Hule eiker`, as this nature type is not easy to tie to a single main ecosystem
+I will also exclude `Hule eiker`, as this nature type is not easy to tie to a single main ecosystem.
+_In some other scenarios, one might wish to keep hule eiker depending on the main ecosystem that it was recorded to in the field._
 
@@
 
-We have 66 nature types to consider. Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *open* wetland.
+We have `r nrow(ntypDF)` nature types to consider.
+
+Some of the wetland types can actually be excluded because of the way we limited this ecosystem to mean *open* wetland:
 
@@
              )
-`%!in%` <- Negate(`%in%`)
-ntypDF <- ntypDF[ntypDF$Nature_type %!in% excl_nt,]
+ntypDF$Nature_type <- trimws(ntypDF$Nature_type)
+ntypDF <- ntypDF[!ntypDF$Nature_type %in% excl_nt,]
 ```
@@
 
+This removed `r nrow(ntypDF) - nrow(ntypDF2)` nature types.
+
 Next I want to add the NiN main types.
@@
 ### Add NiN main type and degree of representativity
+```{r}
+#remove trailing white spaces
+ntypDF2$Nature_type <- trimws(ntypDF2$Nature_type)
+```
+
 ```{r}
@@
 #Exclude non-relevant forest wetland types
-dat2 <- dat[dat$naturtype %!in% excl_nt,]
+dat2 <- dat[!dat$naturtype %in% excl_nt,]
 
 # Melt data
-dat2L <- tidyr::separate_rows(dat2, ninbeskrivelsesvariabler, sep=",")
+dat2L <- tidyr::separate_rows(dat2, ninBeskrivelsesvariable, sep=",")
 
@@
 dat2L <- tidyr::separate(dat2L, 
-                              col = ninbeskrivelsesvariabler,
+                              col = ninBeskrivelsesvariable,
                               into = c("NiN_variable_code", "NiN_variable_value"),
@@
 ```{r}
-temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BA_X",]
+temp <- dat2L[dat2L$ninBeskrivelsesvariable == "7JB-BA_X",]
 table(temp$kartleggingsår, temp$naturtype)
@@
 ```{r}
-temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-BT_X",]
+temp <- dat2L[dat2L$ninBeskrivelsesvariable == "7JB-BT_X",]
 table(temp$kartleggingsår, temp$naturtype)
@@
 ```{r}
-temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7JB-KU-BY_X" |
-                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-DE_X" |
-                dat2L$ninbeskrivelsesvariabler == "7JB-KU-MO_X" |
-                     dat2L$ninbeskrivelsesvariabler == "7JB-KU-PI_X",]
+temp <- dat2L[dat2L$ninBeskrivelsesvariable == "7JB-KU-BY_X" |
+                     dat2L$ninBeskrivelsesvariable == "7JB-KU-DE_X" |
+                dat2L$ninBeskrivelsesvariable == "7JB-KU-MO_X" |
+                     dat2L$ninBeskrivelsesvariable == "7JB-KU-PI_X",]
 table(temp$kartleggingsår, temp$naturtype)
@@
 ```{r}
-temp <- dat2L[dat2L$ninbeskrivelsesvariabler == "7VR-RI_X",]
+temp <- dat2L[dat2L$ninBeskrivelsesvariable == "7VR-RI_X",]
 table(temp$kartleggingsår, temp$naturtype)
@@
 # Exclude selected NiN variables 
-dat2L2 <- dat2L2[dat2L2$NiN_variable_code %!in% exclude,]
+dat2L2 <- dat2L2[!dat2L2$NiN_variable_code %in% exclude,]
 ```
@@
 Mosaics are included.
-Several non-relevant NiN variables are excluded.
+Several non-relevant NiN variables are excluded (forested wetlands).
 The localities are duplicated in as many rows as there are NiN variables for that locality.
@@
 
-Combine datasets
+Combine data sets
 ```{r}
@@
 
-Adding the non-mapped open types
-```{r}
-open[open %!in% ntyp_fill$NiN_mainType]
+Adding these non-mapped open types
+```{r}
+open[!open %in% ntyp_fill$NiN_mainType]
 ```
@@
 ntyp_fill2 <- ntyp_fill %>%
-  add_row(NiN_mainType = "T6", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T11", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T13", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T16", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T23", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T24", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T25", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T26", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
-  add_row(NiN_mainType = "T27", Ecosystem = "Naturlig åpne områder under skoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+  add_row(NiN_mainType = "T6",  Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T11", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T13", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T16", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T23", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T24", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T25", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T26", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0) %>%
+  add_row(NiN_mainType = "T27", Ecosystem = "naturligÅpneOmråderUnderSkoggrensa", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
 
@@
 ```{r}
-semi[semi %!in% ntyp_fill$NiN_mainType]
+semi[!semi %in% ntyp_fill$NiN_mainType]
 ```
@@
 ntyp_fill2 <- ntyp_fill2 %>%
-  add_row(NiN_mainType = "T40", Ecosystem = "Semi-naturlig mark", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
+  add_row(NiN_mainType = "T40", Ecosystem = "semi-naturligMark", NiN_mainTypeCoverage = "Not mapped", km2 = 0)
 ```
@@
 ```{r}
-wetland[wetland %!in% ntyp_fill$NiN_mainType]
+wetland[!wetland %in% ntyp_fill$NiN_mainType]
 ```
@@
 
-From the figure above I take that *Naturlig åpne områder* is very poorly represented in general as there are many NiN main types that are completely missing from the dataset.
+From the figure above I take that *Naturlig åpne områder* is very poorly represented in general as there are many NiN main types that are completely missing from the data set.
 The three most common types by area, however, have complete thematic coverage (if excluding the year 2018 and including all sub-types).
@@
 The other three types probably makes up a considerably smaller area, but note that T33 has some coverage, and is classes as *all (-2018)*.
-Of the three main ecosystems, semi-naturlig mark dominates clearly in ters of area and in terms of number of localities.
-
-For *Våtmark* we have good thematic coverage for V3 nedbørsmyr and V9 semi-naturlig myr, but for V1 åpen jordvassmyr only calcareous localities are mapped and assessed.
+Of the three main ecosystems, semi-naturlig mark dominates clearly in terms of area and in terms of number of localities.
+
+For *Våtmark* we have good thematic coverage for V3 nedbørsmyr and _V9 semi-naturlig myr_, but for _V1 åpen jordvassmyr_ only calcareous localities are mapped and assessed.
 A question is then whether, for a given NiN variable, that calcareous localities can be representative for all mires (poor and rich) or if this will introduce too much bias in one way or another.
@@
 
-I want to see how common the different NiN variables are, and if we can use this to select core variables (kjernevariabler).
-
-First I will get the total area mappe per main ecosystem
+I want to see how common the different NiN variables are, and if we can use this to select core variables \(_kjernevariabler_\).
+
+First I will get the total area mapped per main ecosystem
 ```{r}
@@
 ## Tables
-Here are some output tables with numbers the numbers underlying the above figures.
+Here are some output tables with the numbers underlying the above figures.
 
@@
 ## Export data {#exp-natureType-summary}
-Exporting the table with the summary statistics and other information that I compled for each nature type in the three main ecosystems.
-This can be merged with the main dataset in order to subset according to a specific research question.
+Exporting the table with the summary statistics and other information that I compiled for each nature type in the three main ecosystems.
+This can be merged with the main data set in order to subset according to a specific research question.
 
@@
 
-Note also that the main dataset still contains some localities where there is a mis-match between the nature type definition and the NiN mapping uits recorded in the field.
+Note also that the main data set still contains some localities where there is a mis-match between the nature type definition and the NiN mapping uits recorded in the field.
 Each researcher should decide to include or exclude these.
 
-The main data set also includes mosaic localities, that are less suitable to use as ground trouth data for remote sensing methods.
+The main data set also includes mosaic localities, that are less suitable to use as ground truth data for remote sensing methods.
 
@@
 ```{r}
-dat_red <- dat_melt[dat_melt$n_ninkartleggingsenheter==1 & 
+dat_red <- dat_melt[dat_melt$n_ninKartleggingsenheter==1 & 
                       dat_melt$naturtype != "Hule eiker" &
-                      dat_melt$mosaikk == "Nei",]
+                      dat_melt$mosaikk == "nei",]
 ```
@@
 dat_red_tally <- stats::aggregate(data = dat_red,
-                                  area~hovedøkosystem+ninkartleggingsenheter3, FUN = length)
+                                  area~hovedøkosystem+ninKartleggingsenheter3, FUN = length)
 names(dat_red_tally)[3] <- "count"
-```
-
-```{r}
-target <- c("Semi-naturlig mark",
-            "Våtmark",
-            "Naturlig åpne områder under skoggrensa")
 ```
@@
 ggplot(dat_red_tally[dat_red_tally$hovedøkosystem %in% target,],
-       aes(x = ninkartleggingsenheter3,
+       aes(x = ninKartleggingsenheter3,
                y = count))+
@@
 
-#### Naturlig åpne områder under skoggrensa
-Listing the NiN main types recorded for nature types belonging to *Naturlig åpne områder under skoggrensa*, but which clearly don't belong there:
+#### naturligÅpneOmråderUnderSkoggrensa
+Listing the NiN main types recorded for nature types belonging to *naturligÅpneOmråderUnderSkoggrensa*, but which clearly don't belong there:
 
@@
 dat_red2 <- dat_red[
-  dat_red$hovedøkosystem!="Naturlig åpne områder under skoggrensa" |
-  dat_red$hovedøkosystem=="Naturlig åpne områder under skoggrensa" & 
-    dat_red$ninkartleggingsenheter3 %!in% c("T30", "T32", "T33", "V10"),]
+  dat_red$hovedøkosystem!="naturligÅpneOmråderUnderSkoggrensa" |
+  dat_red$hovedøkosystem=="naturligÅpneOmråderUnderSkoggrensa" & 
+    dat_red$ninKartleggingsenheter3 %!in% c("T30", "T32", "T33", "V10"),]
 ```
@@
 dat_red3 <- dat_red2[
-  dat_red2$hovedøkosystem!="Semi-naturlig mark" |
-  dat_red2$hovedøkosystem=="Semi-naturlig mark" & dat_red2$ninkartleggingsenheter3 %!in% c("V9", "T4", "T35", "T3", "T12", "T1"),]
+  dat_red2$hovedøkosystem!="semi-naturligMark" |
+  dat_red2$hovedøkosystem=="semi-naturligMark" & dat_red2$ninKartleggingsenheter3 %!in% c("V9", "T4", "T35", "T3", "T12", "T1"),]
 ```
@@
 dat_red4 <- dat_red3[
-  dat_red3$hovedøkosystem!="Våtmark" |
-  dat_red3$hovedøkosystem=="Våtmark" & dat_red3$ninkartleggingsenheter3 %!in% c("T4", "T30", "T31", "T32", "V11", "V12", "V13"),]
+  dat_red3$hovedøkosystem!="våtmark" |
+  dat_red3$hovedøkosystem=="våtmark" & dat_red3$ninKartleggingsenheter3 %!in% c("T4", "T30", "T31", "T32", "V11", "V12", "V13"),]
 ```
 This resulted in the deletion of `r nrow(dat_red3) - nrow(dat_red4)` localities.
+
+In total `r nrow(dat_red) - nrow(dat_red4)` localities, or `r round( (nrow(dat_red) - nrow(dat_red4)) / nrow(dat_red) * 100, 1)`% of localities, have a field-recorded NiN main that don't match the definition of the nature type.
+I see this as a symptom of a bigger problem with data validation.

### 2023-06-08 trophic_levels-biomass_indicators.rmd
Render book with trophic level biomass

--- 2023-06-07 DRAFT_trophic_levels-biomass_indicators.qmd
@@
----
-title: "TrophicLevelBiomassRatios"
-author: "James D. M. Speed"
-format: 
-  html:
-    embed-resources: TRUE
-editor: visual
-knitr:
-  opts_knit:
-    root.dir: "S://R projects//TrophicIntIndicators"
-execute: 
-  cache: true
----
+# Trophic level biomass ratios
+
+<br />
+
+_Author and date:_
+James Speed
+
+June 2023
+
+<br />
+
+<!-- Load all you dependencies here -->
+```{r setup, include=FALSE}
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE)
+library(terra)
+library(tidyterra)
+library(ggplot2)
+library(sf)
+library(scales)
+library(RColorBrewer)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
+
+# Set up conditional file paths
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/",
+               "/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/")
+
+pData <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+
+```
+
+<!-- Don't remove these three html lines -->
+<br />
+<br />
+<hr />
+
+<!-- Document you work below.
+Try not to change  the headers too much.
+Data can be stored on NINA server.
+Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine.
+If it is not too big you may store under /data/ on this repository -->
 
@@
 ```{r}
-#| label: Load packages
-#| include: false
-#| cache: TRUE
-
-library(terra)
-library(tidyterra)
-library(ggplot2)
-library(sf)
-library(scales)
-library(RColorBrewer)
-library(tidyverse)
-
-```
-
-```{r}
 #| label: Data import
@@
 #Read in
-npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
-viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
-livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
-carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
+npprast<-rast(paste0(pData,"/Trophic_levels/NPP.tiff"))
+viltrast<-rast(paste0(pData,"/Trophic_levels/Vilt_RawBiomass.tiff"))
+livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))
+carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
 
@@
 #Norwegian counties to help plotting
-norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
+norcounty_shp<-st_read(paste0(pData,"/Trophic_levels/"),"ViltdataCounty")
 #Simplify by county nr
@@
 #For some reason need to read the raster in again
-livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
+livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))
 
@@
 #| cache: TRUE
-carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
+carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
 
@@
 #| cache: TRUE
-npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
-npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]
+# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
+# npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]
 
@@
   scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
-writeRaster(expected_herbivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif",overwrite=TRUE)
+#writeRaster(expected_herbivore_biomass,paste0(pData,"/Trophic_levels/expectedherbivorebiomass.tif"),overwrite=TRUE)
 
@@
 #| cache: TRUE
-expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
-livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
+#expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
+#livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
 
@@
 
-writeRaster(diffherb,"Vertebrates/outputs/HerbivoreBiomassDiffExpected.tif",overwrite=TRUE)
+#writeRaster(diffherb,paste0(pData,"/Trophic_levels/HerbivoreBiomassDiffExpected.tif"),overwrite=TRUE)
 
@@
 
-livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
+#livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))
 
@@
 names(totalherbivorebiomass)<-allyears_vect
-carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
+#carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
 totalcarnivores<-carnivorerast[[49:60]]
@@
 
-writeRaster(expected_carnivore_biomass,"Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif",overwrite=TRUE)
+#writeRaster(expected_carnivore_biomass,paste0(pData,"/Trophic_levels/expectedcarnivorebiomass.tif"),overwrite=TRUE)
 
-writeRaster(diffcarn,"Vertebrates/outputs/CarnivoreBiomassDiffExpected.tif",overwrite=TRUE)
+#writeRaster(diffcarn,paste0(pData,"/Trophic_levels/CarnivoreBiomassDiffExpected.tif"),overwrite=TRUE)
 
@@
 
-From the above we see that Norway has had less carnivore biomass than expected based on the availabile herbivore biomass across the time period.
+From the above we see that Norway has had less carnivore biomass than expected based on the available herbivore biomass across the time period.
 Some counties and time periods (e.g.
@@
 #AR50 type raster
-artype50<-rast("T:\\vm\\alle\\GIS_data\\Norge\\Cartography\\N50\\AR50\\AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")
+artype50<-rast("/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/AR50_Norge.tif")
+
 #artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")
@@
 #Load biomass rasters
-npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
-viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
-livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
-carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
-expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
-expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")
+# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
+# viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
+# livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
+# carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
+# expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
+# expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")
 
 #Project the AR50 to the NPP rasters
+
+### NEED TO CHECK IF THIS IS THE CORRECT LAYER OR GET THE CORRECT ONE FROM JAMES
+
 artype50_FR<-project(resample(artype50_F,npprast,method='near'),npprast)
@@
 
-norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
+#norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
  #Simplify by county nr
@@
 
- writeRaster(forest_diffcarn,"Vertebrates/outputs/forest_carnivores_diffexpected.tif",overwrite=TRUE)
- writeRaster(forest_diffherb,"Vertebrates/outputs/forest_herbivores_diffexpected.tif",overwrite=TRUE)
+ #writeRaster(forest_diffcarn,paste0(pData,"/Trophic_levels/forest_carnivores_diffexpected.tif"),overwrite=TRUE)
+ #writeRaster(forest_diffherb,paste0(pData,"/Trophic_levels/forest_herbivores_diffexpected.tif"),overwrite=TRUE)
 ```
@@
 
-forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
-forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")
+#forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
+#forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")
 
-norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
+#norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
  #Simplify by county nr

### 2023-06-08 R/NDVI/Sentinel.R

@@ -6,7 +6,9 @@
 #To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
 
 library(data.table)
-library(tidyverse)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+#library(tidyverse)
 library(sf)
 
 #############Import NiN data

### 2023-06-08 trophic_levels-biomass_indicators.rmd
Fixed libpath issue

@@
 library(RColorBrewer)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse)
+library(tidyverse)
 

### 2023-06-08 slitasje.Rmd

@@
 library(eaTools)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+ library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
 #library(tidyverse)
 library(tibble)
-library(ggplot2)
-library(tidyr)
-library(dplyr)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation.", fig.height=10, fig.width=4}
+```{r fig.cap="An overview of the naturetypes for which we will calculate the indicator. Colours refer to the main ecosystem affiliation.", fig.height=10}
 ggplot(data = naturetypes, aes(x = naturtype, fill = hovedøkosystem))+
@@
 naturetypes$NiN_variable_code <- paste0("var_", naturetypes$NiN_variable_code)
+```
+
+```{r, fig.cap="Distribution of data points over time."}
+barplot(table(naturetypes$kartleggingsår))
 ```
@@
 
-```{r slitasje-precentageplot, fig.height=12, fig.width=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+```{r slitasje-precentageplot, fig.height=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
 ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+

### 2023-06-13 R/NDVI/2 regression analyses Sentinel.R
further beta distribution tries

@@ -1,5 +1,19 @@
+# NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
+SentinelNDVI.wetland %>%
+  filter(year != '2022') %>%
+  
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+  
+  filter(year == kartleggingsaar) %>%
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  facet_grid( region~hovedtype)
+
+
 library(betareg)
 library(glmmTMB)
+library(StepBeta)
 SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
 
 
@@ -38,6 +52,9 @@ summary( betareg(mean_beta~tilstand + region + hovedtype + tilstand:hovedtype, d
 summary( glmmTMB(mean_beta~tilstand + region + hovedtype + tilstand:hovedtype +(1|id), family=beta_family(),data=SentinelNDVI.wetland.max) )
 
 
+summary( glmmTMB(mean_beta~tilstand*region*hovedtype +(1|id), family=beta_family(),data=SentinelNDVI.wetland.max) )
+
+
 SentinelNDVI.wetland.max %>%
   filter(region=="Northern Norway") %>%
   
@@ -72,11 +89,17 @@ dbeta()
 https://stats.stackexchange.com/questions/12232/calculating-the-parameters-of-a-beta-distribution-using-the-mean-and-variance
 
 
-mod_wetland <- betareg(mean_beta~tilstand + region + hovedtype + tilstand:hovedtype, data=SentinelNDVI.wetland.max)
+mod_wetland.full <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.max)
+mod_wetland.1 <- betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:hovedtype + tilstand_num:region + hovedtype:region, data=SentinelNDVI.wetland.max)
+summary(mod_wetland.1)
[…]

### 2023-06-13 DRAFT_ndvi_wetland.Rmd
starting the ndvi markdown

--- 2023-05-10 functional_plant_indicators_wetland.Rmd
@@
 _Author and date:_
-Joachim Töpper
-
-May 2023
+Joachim Töpper, Tessa Bargmann
+
+June 2023
 
@@
 library(plyr)
-library(tidyverse)
+library(stringr, lib.loc = "/usr/lib/R/site-library")
+library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+library(data.table)
+library(lubridate)
 library(RColorBrewer)
@@
 Ecosystem <- "våtmark"
-Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
-ECT       <- "Functional state characteristic"
+Egenskap  <- c("Primærproduksjon")
+ECT       <- "Structural state characteristics"
 Contact   <- "Joachim Töpper"
@@
 ## Introduction
-Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
+The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and the plant productivity in the given area.
+
+of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
 The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf.

### 2023-06-14 DRAFT_ndvi_wetland.Rmd
markdown text

@@
 ## Introduction
-The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and the plant productivity in the given area.
-
-of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
-The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf.
-Töpper et al.
-2018).
-For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g.
-grasslands or forests.
-Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
-Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
-2019) for wetland ecosystems.
-We calculate the functional signature of plant communities in the monitored sites with respect to light, moisture, pH and nitrogen.
-These functional signatures are then compared to reference distributions of functional signature, separately for each wetland ecosystem type, calculated from 'generalized species lists' underlying the Norwegian categorization system for eco-diversity (Halvorsen et al.
-2020).
-These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
-2021, Töpper & Jakobsson 2021).
-Note that deviations from the reference may occur in both directions, e.g.
-the moisture signature from the testing data may be higher or lower than in the reference.
-Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately.
-Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).
+The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area.
+Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance.
+Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem.
+E.g.
+lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems.
+In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied.
+For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas.
+For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed.
+In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e.
+national maps for the forest and mountain ecosystems).
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
+excluding forest swamps).
+For this task, two initial challenges can be identified early on:
+1) We don't yet have a national wetland map, i.e.
+we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
+2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a nice time series back to year 2000.
+
+In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
+1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
+However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
+Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
+2) We include Sentinel-II imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+However, since Sentinel-II only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
+In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
 
 ## About the underlying data
-In the 'functional plant indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
-
-- as test data we use plant community data from the ANO monitoring scheme (cf. Tingstad et al. 2019)
-- as reference data we use generalized species lists from the Norwegian categorization system for eco-diversity (NiN) (cf. Halvorsen et al. 2020)
-- Swedish plant indicator data for light, moisture, pH, and nitrogen from Tyler et al. (2021)
-
-(1) ANO monitoring data:
-ANO stands for 'areal-representativ naturovervåking', i.e.
-'area representative nature monitoring'.
-1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle.
-Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
-For the vegetation analyses, the cover of each vascular plant species in the plot is recorded.
-Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state.
-In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
-In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
-
-(2) NiN reference data:
-The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species.
-In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5).
-For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
-
-(3) The Swedish plant indicator set published by Tyler et al.
-(2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well.
-From this set, we decided to include indicator data for light, moisture, pH, and nitrogen for wetlands as these are thought to be subject to potential change due to shrub/tree encroachment, drainage, and pollution.
+In the NDVI project for wetlands, we use four sets of data for building indicators for ecological condition:
+
+- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+- Sentinel-II satellite NDVI data
+- MODIS satellite NDVI data
+- Landsat satellite NDVI data
+
+(1) NiN ecosystem data:
+
+(2) Sentinel-II satellite NDVI data:
+
+(3) MODIS satellite NDVI data:
+
+(4) Landsat satellite NDVI data:
+
+have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which reduces the number of available polygons greatly.
+To explore temporal trends even further back in time, we also include Landsat NDVI data, which span a time frame from ???
+to ???, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
 
 ### Representativity in time and space
-For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country.
-As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions.
-The 1351 plots are distributed across regions in the following way:
-
-- Northern Norway: 416
-- Central Norway:  362
-- Eastern Norway:  295
-- Western Norway:  151
-- Southern Norway: 127
+For wetlands, the NiN data contain 6503 wetland polygons of known area size across mainland Norway.
+However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
+The 6503 wetland polygons are distributed across regions and main ecosystem types in the following way:
+
+----build a matrix-table----
+
+The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-II and Landsat.
 
 ### Temporal coverage
-The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme.
-Thus there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.
+The currently available NiN data on wetland ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
+Note that none of the polygons in the NiN-data has been visited more than once during that period.
+
+Sentinel-II satellite NDVI data span the years 2015-2022.
+MODIS satellite NDVI data span the years ????-????.
+Landsat satellite NDVI data span the years ????-????.
 
 ## Collinearities with other indicators
-<!-- Text here -->
+The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in drained wetland areas under shrub and tree encroachment.
 
@@
 ### Reference state
-The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al.
-2018).
-By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
-These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN.
-A more extensive discussion on the use of reference communities can be found in Jakobsson et al.
-(2020).
+This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
+Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for wetland ecosystems.
+In this exploratory work, we will use the condition classification in the NiN-data to explore its applicability for defining a reference state.
+We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
 
 ### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al.
-2010).
-For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution.
-As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al.
-2022), and deviation from the optimal state thus may occur in both direction (e.g.
-indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value.
-In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e.
-its 0.025 and 0.975 quantiles.
-The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator.
-For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).
+We will explore (i) if the condition classes in the NiN data can be used to define scaling values, and (ii) the potential for taking a probabilistic approach to scaling original NDVI values to a condition index based on probability density curves derived from above named models.
 
 ## Uncertainties
-We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
+Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
 
 ## References
-Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185
+
+----not referred to in text yet, but likely will :)----
 
@@
 
-Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A.
-& Nybø, S.
-2021.
-An index-based assessment of ecological condition and its links to international frameworks.
-Ecological Indicators 124: 107252.
-
-Tingstad, L., Evju, M., Sickel, H., & Töpper, J.
-2019.
-Utvikling av nasjonal arealrepresentativ naturovervåking (ANO).
-Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag.
-NINA Rapport 1642.
-
 Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967. Norwegian Institute for Nature Research.
-
-Töpper, J., Velle, L.G.
-& Vandvik, V.
-2018.
-Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition).
-NINA Report 1529b.
-Norwegian Institute for Nature Research.
-
-Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923
 
@@
 
-ANO data
-
+#### NiN data
+We read the NiN data for this document from a cache at the NINA server...
 ```{r, echo = F}
 # Add data from cache
-ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
-ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))
+nin<-readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
+```
+
+...but it can also be directly downloaded from Miiljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+```{r, echo = T, eval = F}
 

### 2023-06-14 R/NDVI/1 data handling.R

@@ -24,7 +24,7 @@
 
 
 
-
+library(downloader)
 library(data.table)
 library(tidyverse)
 library(lubridate)
@@ -33,7 +33,18 @@ library(sf)
 #### data upload ####
 ### Import NiN data
 nin <- st_read("R:\\GeoSpatialData\\Habitats_biotopes\\Norway_Miljodirektoratet_Naturtyper_nin\\Original\\versjon20221231\\Natur_Naturtyper_nin_norge_med_svalbard_25833\\Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
-#nin.andel <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/NiN/naturtyper_nin_20230516.gdb/naturtyper_nin_20230516.gdb")
+
+
+## alternatively download directly from kartkatalogen to P-drive
+url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
+download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="w") 
+unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
+       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB")
+
+## read in from NINA server
+nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/naturtyper_nin_20230516.gdb/naturtyper_nin_20230516.gdb")
+
+
 
 ### Import region- og Norgeskart
 nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
@@ -257,6 +268,7 @@ nin.natopen <- nin.natopen %>%
 
 
 #### save and load processed NiN-data from cache ####
+saveRDS(nin, "data/cache/nin.RDS")
 saveRDS(nin.wetland, "data/cache/nin.wetland.RDS")
 saveRDS(nin.seminat, "data/cache/nin.seminat.RDS")
 saveRDS(nin.natopen, "data/cache/nin.natopen.RDS")

### 2023-06-14 R/NDVI/1 data handling.R
markdown, data import

@@ -37,12 +37,12 @@ nin <- st_read("R:\\GeoSpatialData\\Habitats_biotopes\\Norway_Miljodirektoratet_
 
 ## alternatively download directly from kartkatalogen to P-drive
 url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
-download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="w") 
+download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
 unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
-       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB")
-
-## read in from NINA server
-nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/naturtyper_nin_20230516.gdb/naturtyper_nin_20230516.gdb")
+       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
+nin.2 <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
+                 layer="naturtyper_nin_omr")
 
 
 

### 2023-06-15 DRAFT_ndvi_wetland.Rmd
sdlfksdf

@@
 # Add data from cache
-nin<-readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
-```
-
-...but it can also be directly downloaded from Miiljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
+```
+
+...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
 ```{r, echo = T, eval = F}
 
-# st_layers(dsn = "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-# ANO.sp <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-#                    layer="ANO_Art")
-# ANO.geo <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-#                   layer="ANO_SurveyPoint")
-# head(ANO.sp)
-# head(ANO.geo)
-```
-
+url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
+download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
+unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
+       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
+nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
+                 layer="naturtyper_nin_omr")
+
+```
+
+#### Geometry for Norway and the five regions
+
+```{r}
+# Import region- og Norgeskart, make sure they have the same geometry as the nin data
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = st_crs(nin))
+
+reg <- st_read("data/regions.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = st_crs(nin))
+
+# change region names to something R-friendly
+reg$region
+reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
+
+# combine the Norway and the region maps
+regnor <- st_intersection(reg,nor)
+```
+
+#### NDVI data
 Plant indicators from Tyler et al. (2021)

### 2023-06-15 DRAFT_ndvi_wetland.Rmd
markdown until data handling

--- 2023-06-14 DRAFT_ndvi_wetland.Rmd
@@
 ```{r setup, include=FALSE}
+library(downloader)
+library(data.table)
+library(lubridate)
 library(sf)
@@
 library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-library(data.table)
-library(lubridate)
 library(RColorBrewer)
@@
 library(tmaptools)
-library(raster)
-library(stars)
+library(betareg)
+library(StepBeta)
 library(knitr)
@@
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2) We include Sentinel-II imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
-However, since Sentinel-II only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
+2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
 In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
@@
 - NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
-- Sentinel-II satellite NDVI data
+- Sentinel-2 satellite NDVI data
 - MODIS satellite NDVI data
@@
 
-(2) Sentinel-II satellite NDVI data:
+(2) Sentinel-2 satellite NDVI data:
 
@@
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-II and Landsat.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
 
@@
 
-Sentinel-II satellite NDVI data span the years 2015-2022.
+Sentinel-2 satellite NDVI data span the years 2015-2022.
 MODIS satellite NDVI data span the years ????-????.
@@
 ### Data sets
-
-#### NiN data
+The analyses in this document make use of the following data sets:
+-NiN data (Naturtyper etter Miljødirektoratets innstruks)
+-Sentinel-2 NDVI data
+-MODIS NDVI data
+-Landsat NDVI data
+
+#### NiN data and geometry for Norway and the five regions
 We read the NiN data for this document from a cache at the NINA server...
 ```{r, echo = F}
-# Add data from cache
-nin<-readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
-```
-
-...but it can also be directly downloaded from Miiljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+# Add NiN data from cache
+nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
+```
+
+...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
 ```{r, echo = T, eval = F}
 
-# st_layers(dsn = "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-# ANO.sp <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-#                    layer="ANO_Art")
-# ANO.geo <- st_read("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-#                   layer="ANO_SurveyPoint")
-# head(ANO.sp)
-# head(ANO.geo)
-```
+url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
+download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
+unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
+       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
+nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
+                 layer="naturtyper_nin_omr")
+```
+And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, adn Northern Norway
+```{r}
+# Import region- og Norgeskart, make sure they have the same geometry as the nin data
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = st_crs(nin))
+
+reg <- st_read("data/regions.shp")%>%
+  st_as_sf() %>%
+  st_transform(crs = st_crs(nin))
+
+# change region names to something R-friendly
+reg$region
+reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
+
+# combine the Norway and the region maps
+regnor <- st_intersection(reg,nor)
+```
+
+#### NDVI data
+All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE).
+
+##### Sentinel-2 NDVI data
+The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
+The image collection contains Sentinel imagery from June, July and August 2015-2022
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+```{r}
+# Import Sentinel NDVI Data
+df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
+  map_df(~fread(.))
+df.s
+```
+
+##### MODIS NDVI data
+The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
+The image collection contains MODIS imagery from June, July and August 2000-2022
+
+```{r}
+# Import MODIS NDVI Data
+# MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
+df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
+df.m
+df.m$mean<-df.m$mean/10000
+df.m
+```
+
+##### Landsat NDVI data
+The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
+The image collection contains Landsat imagery from June, July and August 1984-2022
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+```{r}
+# Import Landsat NDVI Data
+# Set up conditional file paths
+dir <- substr(getwd(), 1,2)
+
+path <- ifelse(dir == "C:", 
+               "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
+               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
+
+pData <- ifelse(dir == "C:", 
+                "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
+                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")
+
+# Fread doesn't like the weird path to the server version of the P drive
+# hence this horrendous work around
+files=list.files(pData, pattern = "*.csv", full.names = TRUE)
+
+df_list<-list()# initialise a list of dataframes
+# read in a dataframe in each slot of the df_list
+for (i in files){
+  name <- gsub("-",".",i)
+  name <- gsub(".csv","",name)  
+  i <- paste(i,sep="")
+  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
+}
+
+df.l<-bind_rows(df_list, .id = "column_label")
+df.l
+
+```
+
+### Data handling
 

### 2023-06-16 DRAFT_ndvi_wetland.Rmd
until modelling in ndvi-markdown

@@
 ### Data handling
+#### NiN data
+```{r, echo = T, eval = F}
+# fixing variable- and ecosystem-names with special characters
+colnames(nin)
+colnames(nin)[c(3,8,17,26,31,33,34)] <- c("hovedoekosystem","kartleggingsaar","noyaktighet",
+                                       "omraadenavn","uk_naertruet","uk_sentraloekosystemfunksjon",
+                                       "uk_spesieltdaarligkartlagt")
+unique(nin$hovedoekosystem)
+
+nin <- nin %>% mutate(hovedoekosystem = recode(hovedoekosystem, 
+                                 "Våtmark" = 'Vaatmark',
+                                 "Semi-naturlig mark" = 'Semi_naturlig',
+                                 "Naturlig åpne områder i lavlandet" = 'Naturlig-aapne',
+                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapne')) %>%
+  mutate(validGeo = st_is_valid(SHAPE))
+
+# checking how many polygons have multiple ecosystem types
+unique(nin$ninkartleggingsenheter)
+
+nrow(nin)
+# 95469 polygons altogether
+nrow(nin %>%
+  filter(grepl(',', ninkartleggingsenheter))
+)
+# 21094 polygons have more than 1 ecosystem type (they are separated by commas in the ninkartleggingsenheter-variable)
+
+nrow(nin %>%
+       filter(!grepl(',', ninkartleggingsenheter))
+)
+# 74375 polygons should have only 1 ecosystem type
+
+# there's no information on the proportion of ecosystem types in the polygons, so we got to omit all polygons with multiple ecosystem types :(
+nin <- nin %>%
+  filter(!grepl(',', ninkartleggingsenheter))
+
+# fix the content in the ninkartleggingsenheter-variable
+summary(as.factor(nin$ninkartleggingsenheter))
+# get rid of the NA- in the beginning
+nin <- nin %>% mutate(ninkartleggingsenheter = str_remove(ninkartleggingsenheter, 'NA_'))
+# making a main ecosystem type variable
+nin <- nin %>% mutate(hovedtype = substr(ninkartleggingsenheter, 1, 3),
+               hovedtype = str_remove(hovedtype, '-'))
+# checking mapping unit against main ecosystem type
+nin[,c("hovedoekosystem","hovedtype")]
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Vaatmark"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Semi_naturlig"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Naturlig_aapne"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Skog"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Fjell"]))
+# making a new variable for the overarching ecosystem types based on the main ecosystem types
+nin$hovedoekosystem.orig <- nin$hovedoekosystem
+
+nin <- nin %>%
+  mutate(hovedoekosystem = case_when(hovedtype %in% paste("V",c(1,3:7,9:10),sep="") ~ 'Vaatmark',
+                                     hovedtype %in% paste("V",11:13,sep="") ~ 'Vaatmark_sterkt_endret',
+                                     hovedtype %in% paste("T",c(31:34,40:41),sep="") ~ 'Semi_naturlig',
+                                     hovedtype %in% paste("T",c(2,12,18,20:21),sep="") ~ 'Naturlig_aapent',
+                                     hovedtype %in% c(paste("T",c(4,30,38),sep=""),paste("V",c(2,8),sep="")) ~ 'Skog',
+                                     hovedtype %in% c(paste("T",c(3,7,9,10,14,22,26),sep=""),paste("V",c(6,7),sep="")) ~ 'Fjell',
+                                     TRUE ~ 'NA'))
+
+summary(as.factor(nin$tilstand))
+
+nin <- nin %>% mutate(tilstand = recode(tilstand,
+                                        "Dårlig" = "Redusert",
+                                        "Svært redusert" = "Svaert_redusert"))
+summary(as.factor(nin$tilstand))
+
+# need to split the ninbeskrivelsesvariabler-column into one per variable and these then into two for the variable name and value
+levels(as.factor(nin$ninbeskrivelsesvariabler))
+
+nin2 <- nin %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
+  separate(col = ninbeskrivelsesvariabler,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F) %>%
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
+  mutate(NiN_variable_code = as.factor(NiN_variable_code))
+
+nin2$NiN_variable_code <- as.factor(paste0("var_", nin2$NiN_variable_code, "_end",sep=""))
+
+nin3 <- nin2 %>%
+  pivot_wider(
+    names_from = "NiN_variable_code",
+    values_from = "NiN_variable_value"
+  )
+# this does not do what it is expected to do. nrow(nin3) should go down towards nrow(nin), but it stays at almost nrow(nin2)
+# continue with nin for now, ignoring the beskrivelsesvariabler and only using information from 'tilstand'
+rm(nin2)
+rm(nin3)
+
+## filter out only wetland data
+nin.wetland <- nin %>% 
+  filter(hovedoekosystem %in% c('Vaatmark')) %>%
+  mutate(id = identifikasjon_lokalid) %>%
+  filter(validGeo) %>%
+  drop_na(tilstand) %>%
+  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, kartleggingsaar)
+
+# merge NiN-data with region
+nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
+nin.wetland
+
+colnames(nin.wetland)[c(1,8)] <- c("id","region_id")
+
+nin.wetland <- nin.wetland %>% 
+  mutate(area_meters_nin = st_area(nin.wetland)
+  )
+
+```
+
+```{r, echo = F}
+
+#load processed NiN-data from cache
+nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
+
+```
+
+#### NDVI data
+```{r}
+## Sentinel-2
+# join nin.wetland & Sentinel NDVI data
+SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
+summary(SentinelNDVI.wetland)
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+         hovedtype = as.factor(hovedtype),
+         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
+         lokalitetskvalitet = as.factor(lokalitetskvalitet),
+         tilstand = as.factor(tilstand),
+         area_meters = st_area(SentinelNDVI.wetland))
+summary(SentinelNDVI.wetland)
+# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
+summary(SentinelNDVI.wetland)
+# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
+dim(SentinelNDVI.wetland)
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
+dim(SentinelNDVI.wetland)
+# split date into year, month & day
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  dplyr::mutate(year = lubridate::year(date), 
+                month = lubridate::month(date), 
+                day = lubridate::day(date))
+# add column for sub-ecosystem types
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
+                                                        subtype = str_remove(subtype, '-'))
+# we are using max NDVI per year in every NiN polygon
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+
+summary(SentinelNDVI.wetland)
+
+## MODIS
+# join nin.wetland and Modis NDVI data
+ModisNDVI.wetland <- full_join(nin.wetland, df.m, by="id")
+
+summary(ModisNDVI.wetland)
+ModisNDVI.wetland <- ModisNDVI.wetland %>%
+  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+         hovedtype = as.factor(hovedtype),
+         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
+         lokalitetskvalitet = as.factor(lokalitetskvalitet),
+         tilstand = as.factor(tilstand),
+         area_meters = st_area(ModisNDVI.wetland))
+summary(ModisNDVI.wetland)
+# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
+ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(hovedtype))
+ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
+summary(ModisNDVI.wetland)
+# get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
+dim(ModisNDVI.wetland)
+ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
+dim(ModisNDVI.wetland)
+# split date into year, month & day
+ModisNDVI.wetland <- ModisNDVI.wetland %>%
+  dplyr::mutate(year = lubridate::year(date), 
+                month = lubridate::month(date), 
+                day = lubridate::day(date))
+# we are using max NDVI per year in every NiN polygon
+ModisNDVI.wetland <- ModisNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+summary(ModisNDVI.wetland)
+
+## Landsat
+# join nin.wetland & Landsat NDVI data
+LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by="id")
+
+summary(LandsatNDVI.wetland)
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+         hovedtype = as.factor(hovedtype),
+         ninkartleggingsenheter = as.factor(ninkartleggingsenheter), 
+         lokalitetskvalitet = as.factor(lokalitetskvalitet),
+         tilstand = as.factor(tilstand),
+         area_meters = st_area(LandsatNDVI.wetland))
+summary(LandsatNDVI.wetland)
+# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(hovedtype))
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
+summary(LandsatNDVI.wetland)
+# get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
+dim(LandsatNDVI.wetland)
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
+dim(LandsatNDVI.wetland)
+# split date into year, month & day
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+  dplyr::mutate(year = lubridate::year(date), 
+                month = lubridate::month(date), 
+                day = lubridate::day(date))
+# we are using max NDVI per year in every NiN polygon
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+
+summary(LandsatNDVI.wetland)
+```
+
+### Exploratory analyses, Sentinel-2
+
+```{r}
+## how are polygon sizes distributed?
+summary(SentinelNDVI.wetland$area_meters)
+
+hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
+abline(v=100,lty=2)
+
+## how does NDVI vary over the years (all years)
+SentinelNDVI.wetland %>%
+  ggplot( aes(x=year, y=mean )) + 
+  geom_point() +
+  facet_grid( tilstand~hovedtype)
+# 2022 stands out with the highest NDVI values missing
+# the pattern is strange, suggest to omit 2022 for this analysis
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+
+## NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
+SentinelNDVI.wetland %>%
+  filter(year == kartleggingsaar) %>%
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  facet_wrap( ~hovedtype)
+# NDVI largely varies between hovedtyper: higher in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
+# @Tilstand: in seminat types, kaldkilde, and bogs ndvi seems to increase as condition deteriorates, in fens not much of a pattern
+
+## looking at subtypes for polygons in good condition only (and only for NDVI data matching NiN-mapping)
+SentinelNDVI.wetland %>%
+  filter(tilstand == 'God') %>%
+  filter(year == kartleggingsaar) %>%
+  ggplot(aes(x=subtype, y=mean )) + 
+  geom_violin() +
+  facet_wrap( ~hovedtype)
+# NDVI varies between ecosystem subtypes
+# -at least equally much as between main ecosystem types
+# -stronger than between condition classes in previous plots
+
+```
+
+### Regression analyses, NDVI as a function of condition, Sentinel-2
+```{r}
+# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that handles bounded data
+# we can transform the variable to be bounded between 0 and 1 and use beta-regression models
+SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
+# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
+SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
+  filter(year == kartleggingsaar)
+
+summary( betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train) )
+# full model does not work, remove the 3-way interaction
+summary( betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train) )
+# this works! Let's run a stepwise-function to fin the most parsimonious model
+model.wetland.train <- betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train)
+model.wetland.train <- StepBeta(model.wetland.train)
+summary(model.wetland.train)
+```
+
+NDVI values do vary between:
+- Regions
+- Ecosystem types
+- Condition classes
+
+Since the 3-way interaction term was not possible to be evaluated by the model, but it is evident that all three variables matter, we run the model separately for every region
+```{r}
+model.wetland.train.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
+model.wetland.train.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
+model.wetland.train.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
+model.wetland.train.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
+model.wetland.train.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
+
+```
+
+### Regression analyses, NDVI across time
+
+### Scaled value calculations
 

### 2023-06-16 R/NDVI/2 exploratory analyses Sentinel.R

@@ -1,9 +1,14 @@
 ############ working with max NDVI per year in every NiN polygon
-SentinelNDVI.wetland %>%
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  ggplot(aes(x=as.numeric(area_meters))) + 
-  geom_histogram()
+  filter(mean == max(mean, na.rm=TRUE))
+
+## distribution of polygon sizes
+summary(SentinelNDVI.wetland$area_meters)
+
+hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
+abline(v=100,lty=2)
+
 
 SentinelNDVI.seminat %>%
   group_by(id, year) %>%

### 2023-06-16 R/NDVI/2 regression analyses Sentinel.R

@@ -10,59 +10,53 @@ SentinelNDVI.wetland %>%
   geom_violin() +
   facet_grid( region~hovedtype)
 
-
+#library(glmmTMB)
 library(betareg)
-library(glmmTMB)
 library(StepBeta)
 SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
 
-
-SentinelNDVI.wetland.max <- SentinelNDVI.wetland %>%
-  filter(year != '2022') %>%
-  
-  filter(year == kartleggingsaar) %>%
-  
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
+SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
+  filter(year == kartleggingsaar)
 
 # make an extra numeric condition variable
-unique(SentinelNDVI.wetland.max$tilstand)
-SentinelNDVI.wetland.max <- SentinelNDVI.wetland.max %>% mutate(tilstand_num = recode(tilstand, 
+unique(SentinelNDVI.wetland.train$tilstand)
+SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
                                             "God" = '1',
                                             "Moderat" = '2',
                                             "Redusert" = '3',
                                             "Svært redusert" = '4'))
-SentinelNDVI.wetland.max$tilstand_num <- as.numeric(SentinelNDVI.wetland.max$tilstand_num)
+SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
 
-summary(SentinelNDVI.wetland.max$tilstand_num)
+summary(SentinelNDVI.wetland.train$tilstand_num)
   
   
 
-summary( betareg(mean_beta~tilstand_num*region,data=SentinelNDVI.wetland.max) )
+summary( betareg(mean_beta~tilstand_num*region,data=SentinelNDVI.wetland.train) )
 
-summary( betareg(mean_beta~tilstand*region,data=SentinelNDVI.wetland.max) )
-
-
[…]

### 2023-06-22 DRAFT_ndvi_wetland.Rmd
ndvi data handling and regression fixes

@@
 
-# need to split the ninbeskrivelsesvariabler-column into one per variable and these then into two for the variable name and value
-levels(as.factor(nin$ninbeskrivelsesvariabler))
-
-nin2 <- nin %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
-  separate(col = ninbeskrivelsesvariabler,
-           into = c("NiN_variable_code", "NiN_variable_value"),
-           sep = "_",
-           remove=F) %>%
-  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
-  mutate(NiN_variable_code = as.factor(NiN_variable_code))
-
-nin2$NiN_variable_code <- as.factor(paste0("var_", nin2$NiN_variable_code, "_end",sep=""))
-
-nin3 <- nin2 %>%
-  pivot_wider(
-    names_from = "NiN_variable_code",
-    values_from = "NiN_variable_value"
-  )
-# this does not do what it is expected to do. nrow(nin3) should go down towards nrow(nin), but it stays at almost nrow(nin2)
-# continue with nin for now, ignoring the beskrivelsesvariabler and only using information from 'tilstand'
-rm(nin2)
-rm(nin3)
+##############################
+##############################
+# code on ninbeskrivelsesvariabler here if relevant
+##############################
+##############################
 
@@
 
+## looking at ndvi in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
+SentinelNDVI.wetland %>%
+  filter(year != '2022') %>%
+
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+
+  filter(year == kartleggingsaar) %>%
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  facet_grid( region~hovedtype)
+# V4 has lacking data for quite some condition classes in most regions
+
 ```
@@
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
+# we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
+
 SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
-  filter(year == kartleggingsaar)
-
-summary( betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train) )
-# full model does not work, remove the 3-way interaction
-summary( betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train) )
-# this works! Let's run a stepwise-function to fin the most parsimonious model
-model.wetland.train <- betareg(mean_beta~tilstand_num + region + hovedtype + tilstand_num:region + tilstand_num:hovedtype + region:hovedtype, data=SentinelNDVI.wetland.train)
+  filter(year == kartleggingsaar) %>%
+  filter(hovedtype != 'V4')
+
+# We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
+model.wetland.train <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
 model.wetland.train <- StepBeta(model.wetland.train)
@@
 
-Since the 3-way interaction term was not possible to be evaluated by the model, but it is evident that all three variables matter, we run the model separately for every region
+It is evident that all three variables matter, for ease of interpretation we run the model separately for every region
 ```{r}

### 2023-06-22 R/NDVI/1 data handling.R

@@ -178,22 +178,37 @@ summary(as.factor(nin$tilstand))
 # need to split the ninbeskrivelsesvariabler-column into one per variable and these then into two for the variable name and value
 levels(as.factor(nin$ninbeskrivelsesvariabler))
 
+# two attempts of doing that
 nin2 <- nin %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
   separate(col = ninbeskrivelsesvariabler,
            into = c("NiN_variable_code", "NiN_variable_value"),
            sep = "_",
            remove=F) %>%
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
-  mutate(NiN_variable_code = as.factor(NiN_variable_code))
+  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
+  subset(select = -ninbeskrivelsesvariabler)
 
-nin2$NiN_variable_code <- as.factor(paste0("var_", nin2$NiN_variable_code, "_end",sep=""))
+nin2 <- as.data.frame(nin2)
+
+# or
+nin2<- nin %>% 
+  mutate(ninbeskrivelsesvariabler = strsplit(as.character(ninbeskrivelsesvariabler), ",")) %>% 
+  unnest(ninbeskrivelsesvariabler) %>%
+  separate(col=ninbeskrivelsesvariabler, into=c('NiN_variable_code', 'NiN_variable_value'), sep='_')
+
+nin2 <- as.data.frame(nin2)
+
+#nin2$NiN_variable_code <- as.factor(paste0("var_", nin2$NiN_variable_code, "_end",sep=""))
 
 nin3 <- nin2 %>%
   pivot_wider(
     names_from = "NiN_variable_code",
     values_from = "NiN_variable_value"
   )
-# this does not do what it is expected to do. nrow(nin3) should go down towards nrow(nin), but it stays at almost nrow(nin2)
+
+nin3 <- as.data.frame(nin3)
+
+
 # continue with nin for now, ignoring the beskrivelsesvariabler and only using information from 'tilstand'
 rm(nin2)
 rm(nin3)
@@ -552,3 +567,4 @@ summary(LandsatNDVI.natopen)
 
 
 
+#rm(list= ls()[!(ls() %in% c('SentinelNDVI.wetland','ModisNDVI.wetland','LandsatNDVI.wetland'))])

### 2023-06-22 R/NDVI/2 regression analyses Sentinel.R

@@ -9,6 +9,7 @@ SentinelNDVI.wetland %>%
   ggplot( aes(x=tilstand, y=mean )) + 
   geom_violin() +
   facet_grid( region~hovedtype)
+# V4 has lacking data for quite some condition classes in most regions
 
 #library(glmmTMB)
 library(betareg)
@@ -16,8 +17,10 @@ library(StepBeta)
 SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
 
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
+# we drop V4 for the analysis as it lacks data for most combinations of condition and region
 SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
-  filter(year == kartleggingsaar)
+  filter(year == kartleggingsaar) %>%
+  filter(hovedtype != 'V4')
 
 # make an extra numeric condition variable
 unique(SentinelNDVI.wetland.train$tilstand)

### 2023-06-23 DRAFT_ndvi_wetland.Rmd
wetland ndvi continued

@@
   drop_na(tilstand) %>%
-  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, kartleggingsaar)
+  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, ninbeskrivelsesvariabler, kartleggingsaar)
 
@@
   )
+# check and edit the order of regions
+levels(nin.wetland$region)
+nin.wetland$region <- as.factor(nin.wetland$region)
+levels(nin.wetland$region)
+nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+levels(nin.wetland$region)
 
@@
 
+# make an extra numeric condition variable
+unique(SentinelNDVI.wetland.train$tilstand)
+SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
+                                            "God" = '0',
+                                            "Moderat" = '1',
+                                            "Redusert" = '2',
+                                            "Svært redusert" = '3'))
+SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
+
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.wetland.train <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
-model.wetland.train <- StepBeta(model.wetland.train)
-summary(model.wetland.train)
-```
-
+model.wetland.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
+model.wetland.cond.Sent <- StepBeta(model.wetland.cond.Sent)
+summary(model.wetland.cond.Sent)
+```
 NDVI values do vary between:
@@
 
-It is evident that all three variables matter, for ease of interpretation we run the model separately for every region
+It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
+```{r, include = F}
+rm(model.wetland.cond.Sent)
+```
+
 ```{r}
-model.wetland.train.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
-model.wetland.train.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
-model.wetland.train.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
-model.wetland.train.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
-model.wetland.train.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
+model.wetland.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
+model.wetland.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
+model.wetland.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
+model.wetland.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
+model.wetland.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
 

### 2023-06-23 R/NDVI/1 data handling.R

@@ -26,9 +26,18 @@
 
 library(downloader)
 library(data.table)
-library(tidyverse)
 library(lubridate)
 library(sf)
+library(plyr)
+library(stringr)
+library(tidyverse)
+library(RColorBrewer)
+library("gridExtra") 
+library(ggridges)
+library(tmap)
+library(tmaptools)
+library(betareg)
+library(StepBeta)
 
 #### data upload ####
 ### Import NiN data
@@ -175,6 +184,7 @@ summary(as.factor(nin$tilstand))
 
 
 
+  
 # need to split the ninbeskrivelsesvariabler-column into one per variable and these then into two for the variable name and value
 levels(as.factor(nin$ninbeskrivelsesvariabler))
 
@@ -222,19 +232,24 @@ nin.wetland <- nin %>%
   mutate(id = identifikasjon_lokalid) %>%
   filter(validGeo) %>%
   drop_na(tilstand) %>%
-  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, kartleggingsaar)
+  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, ninbeskrivelsesvariabler, kartleggingsaar)
 
 
 # merge NiN-data with region
 nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
 nin.wetland
 
-colnames(nin.wetland)[c(1,8)] <- c("id","region_id")
+colnames(nin.wetland)[c(1,9)] <- c("id","region_id")
 
 nin.wetland <- nin.wetland %>% 
   mutate(area_meters_nin = st_area(nin.wetland)
   )
-
+# check and edit the order of regions
+levels(nin.wetland$region)
+nin.wetland$region <- as.factor(nin.wetland$region)
+levels(nin.wetland$region)
+nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+levels(nin.wetland$region)
 
 
 
@@ -245,17 +260,23 @@ nin.seminat <- nin %>%
   mutate(id = identifikasjon_lokalid) %>%
   filter(validGeo) %>%
   drop_na(tilstand) %>%
[…]

### 2023-06-23 R/NDVI/2 regression analyses Sentinel.R

@@ -1,3 +1,10 @@
+# make sure to run these if not running script '2 exploratory analyses Sentinel'
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+
+
 # NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
 SentinelNDVI.wetland %>%
   filter(year != '2022') %>%
@@ -25,50 +32,62 @@ SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
 # make an extra numeric condition variable
 unique(SentinelNDVI.wetland.train$tilstand)
 SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
-                                            "God" = '1',
-                                            "Moderat" = '2',
-                                            "Redusert" = '3',
-                                            "Svært redusert" = '4'))
+                                            "God" = '0',
+                                            "Moderat" = '1',
+                                            "Redusert" = '2',
+                                            "Svært redusert" = '3'))
 SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
 
 summary(SentinelNDVI.wetland.train$tilstand_num)
   
   
 
-summary( betareg(mean_beta~tilstand_num*region,data=SentinelNDVI.wetland.train) )
+model.wetland.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
+model.wetland.cond.Sent <- StepBeta(model.wetland.cond.Sent)
+summary(model.wetland.cond.Sent)
 
-summary( betareg(mean_beta~tilstand*region,data=SentinelNDVI.wetland.train) )
-
-
-summary( glmmTMB(mean_beta~tilstand_num*region +(1|hovedtype/id), family=beta_family(),data=SentinelNDVI.wetland.train) )
-
[…]

### 2023-06-30 DRAFT_insectIndicators.Rmd
first draft, not done

---
title: "Template for indicator documentation"
output:
  html_document
---

<!-- This is a template for how to document the indicator analyses.

Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name te file DRAFT_myIndicatorName.Rmd.

Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
you need to type rmarkdown::render("template.Rmd") in the console, 
as the the usual Ctrl + Alt + k will not work.

Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
easier to read the web page as a whole of the indicators are presented using the same layout.

When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
the analysis is repeatable by someone else. Be therefore very careful to annotate 
and comment on every step in the analysis. When completed, Anders will make a copy
of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->

<!-- Examples:

For an example, see 'indicators/breareal-workingFile.Rmd' 
and the subsequent 'breareal.Rmd' 
which again, when rendered, produces this:
https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html

Notice the degree of annotation.

-->

<!-- Replace 'indicator name' with yor own indicator name  -->
# Insect indicators

<br />

_Author and date:_
Jens Åström

```{r}
Sys.Date()
```

<br />

#Notes on this draft
This is a draft. The plan to finish it up i the fall of 2023.

Incomprehensive list of things that will change:

* Fill in missing sections.
* Choices affecting the individual indicators, such as what taxa to select for each indicator'
* Upgrade of genetic diversity calculation. This is just a placeholder, awaiting some new data.
* Bumblebee and butterfly indicators will be added. The routines are done but should be streamlined.
* Packaging of R-package versions for reproducibility? Or just point to a version on Github..
* Make bombLepiSurv public.
* Fix any comments in this draft inside hard brackets.
* Write something about the choice of each indicator.

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

require(Norimon)
require(bombLepiSurv)
require(tmap)
require(sf)
suppressPackageStartupMessages({
  require(tidyverse)
}

)

```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- rep("Semi-naturlig mark", 3) # e.g. "Skog og fjell"
Egenskap  <- c("Biologisk mangfold, Funksjonelle grupper innen trofiske nivåer, Funksjonelt viktige arter og biofysiske strukturer") # e.g. "Primærproduksjon"
ECT       <- rep("Functional state characteristics", 3) # e.g. "Structural state characteristic"
Contact   <- rep("Jens Åström", 3) # e.g. "Anders Kolstad"
```

```{r, echo=F}
metaData <- tibble(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
We here show how to calculate a set of indicators for terrestrial insects in Norway. Two of them are adapted from earlier work on the Nature Index for Norway and use data from the Norwegian monitoring program on bumblebees and butterflies (**NMBB**). The others are developed during 2023, and use data from the Norwegian insect monitoring program (**NorIns**).

This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github. The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators (**Not public yet**), and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.

For the indicators from the NMBB program, community reference values has been elicitated from experts, and is explained further below. For the indicators from the NorIns program, this is significantly harder to do, because of the size of the communities, lack of historical timeseries and changes in sampling and identification techniques. We therefore currently lack reference values for many of these indicators.

Below is a list of the indicators calculated in this document, and their state of development. Some details is further expanded on below.

```{r, echo=F}
indDataset <- c(rep("NMBB", 2), 
             rep("NorIns", 6))

Indicator  <- c("Bumblebees in semi-natural grasslands", 
                "Butterflies in semi-natural grasslands",
                "Biomass of flying insects",
                "Species richness of flying insects",
                "Species richness of pollinators",
                "Species richness of X",
                "Species richness Apocrita / Symphyta",
                "Intraspesific genetic variation")

Geo_scope       <- c(rep("Sørlandet, Østfold-Vestfold, Vestlandet, Trøndelag", 2), 
                     rep("Sørlandet, Østlandet, Trøndelag, Nord-Norge", 6))

Ind_data <- c(rep("Present", 2),
              rep("Present", 6))

Ref_val   <- c(rep("Present", 2), 
               rep("Lacking", 6))
```

```{r, echo=F}
indList <- tibble("Dataset" = indDataset,
                  "Indicator" = Indicator,
                  "Current geographical scope" = Geo_scope,
                  "Indicator data" = Ind_data,
                  "Reference values" = Ref_val)

knitr::kable(indList)
```

## About the underlying data
<!-- Text here -->

### Representativity in time and space
<!-- Text here -->

### Original units
<!-- Text here -->

### Temporal coverage
<!-- Text here -->

### Aditional comments about the dataset
<!-- Text here -->

## Ecosystem characteristic
### Norwegian standard
<!-- Text here -->

### <!-- Text here -->
<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->

## Collinearities with other indicators
None measured so far, but it is reasonable to assume that some of these indicators will share some causal relationships with eachother, and therefore correlate to some degree. This could potentially happen with some other indicators not described here.

## Reference state and values
Assessing the state of insect communities can be a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability. Still, the most complicated part for an indicator is to establishing reference values, when the reference states are lacking or prohibitively difficult to measure empirically. We currently lack reference values for all indicators from the **NorIns** program. It is unlikely that this can be solved "simply" by surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely don't exist. The question of how to handle these reference values is currently unresolved, and we present these indicator value calculations with made up reference values as placeholders in the code.

### Reference state
<!-- Text here -->

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
<!-- Text here -->

## Uncertainties
The uncertainties for the indicator values is calculated by bootstrapping the average values for each locality within a year. This takes the variability between the localities into account.

## References
* Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
Insektovervåking på Østlandet, Sørlandet og i Trøndelag. Rapport fra feltsesong 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978‐82‐426‐5037‐5) 99 s. NINA rapport(2241)

* Åström, Sandra Charlotte Helene; Åström, Jens; Bøhn, Kristoffer; Gjershaug, Jan Ove; Staverløkk, Arnstein; Dahle, Sondre; Ødegaard, Frode. Nasjonal overvåking av dagsommerfugler og humler i Norge. Oppsummering av aktiviteten i 2022. Trondheim: Norsk institutt for naturforskning (NINA) 2023 (ISBN 978-82-426-5009-2) 54 s. NINA rapport(2214)

* Connor M French, Laura D Bertola, Ana C Carnaval, Evan P Economo, Jamie M Kass, David J Lohman, Katharine A Marske, Rudolf Meier, Isaac Overcast, Andrew J. Rominger, Phillip Staniczenko, Michael J Hickerson. 2023. Global determinants of insect mitochondrial genetic diversity. https://doi.org/10.1101/2022.02.09.479762

## Analyses

### Calculation principles for NorIns indicators
The indicators from the Norwegian insect monitoring are calculated with functions in the `Norimon` package. This method is meant to capture a broad variety of insect indicators using data from **NorIns**.

**Main steps of workflow:**

1. Fetch data of biomass or community diversity from a centralized database (or GBIF). Diversity data can be subsetted on several taxonomic levels, but biomass is only available for whole samples.
2. Bootstrap observations and calculate mean and uncertainty of point estimates.
3. Compare observations to reference points.
4. Display and plot the results

```{r flow_chart, out.width = '400pt', dev='png', echo=F}
knitr::include_graphics("images/NorIns_flow_chart.png")

```

Most of the ecology comes into step 1, in choosing what data to use to describe a quality. This can be a custom selection of species, or a set of higher taxa such as specific genuses, families or an order. The goal is to choose a set of taxa that represent specific qualities of the community that can indicate the ecological status of the ecosystem. In addition to the selection of taxa, we have to decide on the level of aggregation, e.g. if we should aggregate the data on the region scale, locality scale, or even the individual sample occasion scale. For ecological indicators, we will aggregate the values to a year locality scale, where the variables summarises all catches in a locality in a year.

The second step is to get an estimate of the uncertainty of the data, were we use the bootstrap method. This method is flexible, as we could bootstrap the samples on different sampling levels (samples within localities, regions within country). For the insect indicators, we will bootstrap and summarize the variables to regional values, for example expressing the mean values of insect biomass in Trøndelag in 2021, with uncertainty.

Ecological knowledge also comes into step 3, comparing the values to a reference state. Here we have several options. We could for example use a single defined value as the reference state. But we can also use a point in time as a reference point (e.g. the start of a time series).

Taken together, this framework is meant to facilitate the calculation of an arbitrary set of insect indicators, based on the combination of choices in data to fetch, aggregation level, and reference comparison. We can work through an example with pollinators, to showcase the framework.

In the example, we start with connecting to the database in order to fetch the data.

```{r}
connect_to_insect_db()
```

For convenience, the pollinator families can be retreived by the `get_pollinators()` function.

```{r}
pollinators <- get_pollinators()

pollinators_fam <- pollinators %>%
  select(family_latin) %>%
  pull()
```

```{r, echo = FALSE}
kable(pollinators)
```
We then fetch the community data for these families through the `get_observations` function. We additionally subset the habitat type that we are interested in, here semi-natural land. The result is a tibble of the aggregated number of species, shannon diversity, and the mean number of genetic variants per species. The default aggregation level is "year_locality", meaning the total observations for a locality within a year. [The output currently contains some experimental values. Shannon diversity doesn't really make sense without counts/amounts, etc...]

```{r}

poll_loc_year <- get_observations(subset_families = pollinators_fam,
                                  subset_habitat = "Semi-nat")

```
```{r, echo = FALSE}
poll_loc_year %>% 
  slice(1:5) %>% 
kableExtra::kbl()
```

## Bootstrap observations
The localities in the NorIns program are (semi-)randomly selected and can be viewed as samples of a larger population. Individually, they represent a single measurement of an insect community in a habitat type in a region. To get the sampling uncertainty for this representation, we can bootstrap the values by this simple prosess: we choose a random set of localities within a year and region (with replacement) and calculate the average values. We then repeat this process a large number of times to get a bootstrap sample of values, which can be used to express the uncertainty in the dataset.

We typically have 10 localities for a given habitat type and region each year, which is not very much to base our uncertainty estimates on. But bootstrapping works fairly well on small samples and has the advantage that is doesn't make assumptions of the statistical distribution of the errors. Note that once we get a larger dataset with repeating surveys every 5 years, it probably makes sense to do some kind of parametric (hierarcichal) modelling of the values instead of this simple bootstrap.

When **NorIns** is fully scaled up, it will have 50 localities for each covered habitat type in every country region, with 10 localities every year, in a rotating survey scheme over 5 years. Since the program is rolled out sequentially starting in 2020, for semi-natural land it currently covers 4/5 of the country, with varying amounts of localities within each region. The map below displays the localities surveyed so far.

```{r}
semi_loc <- get_localities(dataset = "NasIns",
                           habitat_type = "Semi-nat")

norway_regions <- get_map()
```

```{r location_map, echo = F}
tm_shape(norway_regions) +
  tm_polygons("region") +
  tm_shape(semi_loc) +
  tm_symbols(col = "red",
             shape = 22,
             size = 0.8)

```

The bootstrap routine is implemented in the function `bootstrap_value()`, which takes a community (from `get_observation`) or weight dataset (from `get_biomass`) as its first input. It also needs to know what measurement in the dataset to bootstrap, and what, if any, grouping structure to aggregate the results on. In this example, we bootstrap the number of pollinator species, and aggregate the results on the year and regional scale.

```{r}

poll_richness_boot <- bootstrap_value(poll_loc_year,
                                        value = no_species,
                                        groups = c("year",
                                                   "region_name")
                                        )

```

This creates an object of type `boot_stat`. Calling it prints a simple summary of the bootstrap values.
```{r}
poll_richness_boot
```

But the `boot_stat` object also stores the individual bootstrap values for later computation. [Why is is only 5 994 rows? Fix/figure out]

```{r}
poll_richness_boot[2]
```

## Comparing bootstrap values to a reference point
Setting aside the difficulties in establishing robust reference values, the next step is to compare the observed values (with bootstrapped uncertainty) to the chosen reference value. This can be made in several ways. Most simply, if we have a single numeric value as a reference value, we can simply subtract that from the observed values. For example, if we observe 13 species of pollinators in 2022 at a location, and the reference point is 10, the 2022 has increased by 13 - 10 = 3 species. Such subtractions should be made on the set of bootstrapped values, and new summary statistics calculated, preserving the uncertainty from the bootstrap. The `boot_stat` has its own subtraction method do to just that. In this example, we set the reference value arbitrarily to 30 species.

```{r}
poll_richness_boot
```

```{r}
diff_poll_richness_boot <- poll_richness_boot - 30
```

```{r}
diff_poll_richness_boot
```

Alternatively, we could use a reference point in the time series itself. Say for example that we want to use the values for species richness of pollinators in semi-natural land i Østlandet 2020 as a reference point. We can then calculate the difference (the contrast) between this level and all the other levels. We do this by the function `boot_contrast()`

**NB! This functionality is in development. It currently works for single rows as reference points, but needs updating to allow for referencing several values, e.g using the start values for all regions and habitat types as their own reference points.**

```{r}

diff_poll_richness_boot2 <- poll_richness_boot %>%
  boot_contrast(year == 2020 & region_name == 'Østlandet')
```
```{r}
diff_poll_richness_boot2
```

##Normalizing the values
After the comparison to a reference value, we need to normalize the indicator values to lie between 0 and 1. This can be done in several ways (insert link to scaling chapter). The simplest case is to use a linear scaling, with a natural zero. This can be done by dividing the indicator values by the highest value state. A `boot_stat` object has a `/` function that divides each bootstap value by a given value, truncates the highest values to 1, and recalculates the summary values.

[Show this, but need to think on how to do this with difference values]

## Display and plot bootstrap values
The boot_stat class also has its own plot function. It tries to plot a comparison of the bootstrap distributions over years, for each group. For example, if we plot the object `diff_poll_richness_boot`, we can look at the yearly differences in beetles species richness in the two geographic regions: (Note that we have only 1 year of data from Sørlandet so far)

```{r beetle_richness_boot}
plot(diff_poll_richness_boot)
```

In the cases where we have used a single row as a reference point, this shows up as a sharp spike at 0.

```{r diff_beetle_richness_boot2}
plot(diff_poll_richness_boot2)
```

More plotting options are available, see the `Norimon` indicator workflow vignette.

## Map plots
In addition to the density plots above, we can also display the values geographically, in a map. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.

```{r map_plot_cutout}
map_plot(poll_richness_boot)
```

By default, it only shows the regions with data, but this can be overridden manually:
```{r map_plot_whole}
map_plot(poll_richness_boot,
         whole_country = TRUE)
```

We can also choose a different palette, for example from the NinaR package, and visualize the uncertainty by setting the transparency of the colors from the bootstrap standard deviations. By the way, most of these functions can be piped as well:

```{r diff_beetle_richness_boot_map}
diff_poll_richness_boot %>%
map_plot(palette = "orange-green",
         whole_country = FALSE,
         alpha_from_sd = TRUE)
```

This concludes the tour of the `Norimon` functionality.

### Data sets
There are a few different ways to access the data required for these indicators. Both the NMBB and the NorIns project stores its data in an internal database at NINA. They both also export most of their data to GBIF, but those exports need to be restructured before they can be processed in the following scripts.

For the time being, we will here use some internal functions in the bombLepiSurv and `Norimon` packages to fetch the required data. [We may add a separate data fetching script later on, describing how to get this data externally in a stable way.]

#### NorIns data
The `Norimon` package has convenience functions to fetch data from the database, housed internally at NINA. This database is currently not available outside NINA, but we will implement a solution for this. Either we will make the database externally available, or we will provide an alternative route to the data from the GBIF export. For now, we will fetch the data through the `Norimon` functions.

#### NMBB data
The data for the bumblebee and butterfly indicators can be accessed through the R-package `bombLepiSurv`. [Replace this with the GBIF version?. At least make the package public.]

```{r}
require(bombLepiSurv)
bombLepiSurv::humlesommerfConnect(connectionName = "bombLepiCon")
bombus_obs <- getAllData(dataConnection = "bombLepiCon",
                         type = "bumblebees",
                         year = 2009:2022)
lepi_obs <- getAllData(dataConnection = "bombLepiCon",
                         type = "butterflies",
                         year = 2009:2022)

```

#### Regions
<!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
Importing a shape file with the regional delineation. The indicators associated with the NorIns project can be attributed to the 5 country regions of Norway. Currently however, the data program only covers 4 of the 5 regions.

```{r}
norway_regions <- Norimon::get_map()
norway_reg_NorIns <- norway_regions %>% 
  filter(region != "Vestlandet") %>% 
  select(region) %>% 
  group_by(region) %>% 
  summarize(geom = st_union(geom))
```

```{r}
tm_shape(norway_reg_NorIns) + 
  tm_polygons(col = "region") 
```

The indicators from the NMBB program is similary connected to regions, with the exception that the south-east region only covers the old fylke's of Vestfold and Østfold, and that Nord-Norge isn't covered yet. [Find the old fylke geometries and add this map to bombLepiSurv]

### Indicator values for NorIns indicators
#### Biomass of flying insects in semi-natural land

Fetch the data.
```{r}
biomass_sn <- get_biomass(subset_year = 2020:2022,
                          subset_region = NULL)
```
Calculate the indicator values for each region and year.
```{r}
biomass_sn_boot <- bootstrap_value(df = biomass_sn,
                                   value = avg_wet_weight,
                                   groups = c("year",
                                              "region_name")
                                   )
```

Compare values to a reference state. The reference state is here set uniformly for all regions, with a made up value just to be able to continue the calculations. Alternatively, we could calculate the indicator values for each region separately, compare to individual reference states, and then put it all together again.

We here use a single value as a reference state and normalize the values at the same time, using the `/` function.

```{r}
biomass_ref <- 50

biomass_sn_diff <- biomass_sn_boot / biomass_ref
```

Plot the results and display uncertainty.
```{r}
biomass_sn_diff
```

```{r}
plot(biomass_sn_diff)
```

Prepare export format.

```{r}
biomass_sn_to_exp <- biomass_sn_diff$bootstrap_summary

biomass_sn_to_exp <- biomass_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

biomass_sn_to_exp <- norway_reg_NorIns  %>% 
  inner_join(biomass_sn_to_exp,
            by = c("region" = "region"),
            multiple = "all")

biomass_sn_to_exp
```

#### Species richness of flying insects in semi-natural land

Fetch the data.
```{r}
richness_sn <- get_observations(subset_year = 2020:2022,
                                subset_region = NULL)
```
Calculate the indicator values for each region and year.
```{r}
richness_sn_boot <- bootstrap_value(df = richness_sn,
                                   value = no_species,
                                   groups = c("year",
                                              "region_name")
                                   )
```

Compare to arbitrary reference value and normalize.
```{r}
richness_ref <- 4000

richness_sn_diff <- richness_sn_boot / richness_ref
```

Plot the results and display uncertainty.
```{r}
richness_sn_diff
```

```{r}
plot(richness_sn_diff)
```

Prepare export format.

```{r}
richness_sn_to_exp <- richness_sn_diff$bootstrap_summary

richness_sn_to_exp <- richness_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

richness_sn_to_exp <- norway_reg_NorIns  %>% 
  inner_join(richness_sn_to_exp,
            by = c("region" = "region"),
            multiple = "all")

richness_sn_to_exp
```

#### Species richness of pollinators in semi-natural land

Fetch the data.
```{r}
pollinator_fam <- get_pollinators() %>% 
  select(family_latin) %>% 
  pull()

pollinators_sn <- get_observations(subset_year = 2020:2022,
                                   subset_families = pollinator_fam,
                                   subset_region = NULL)
```
Calculate the indicator values for each region and year.
```{r}
pollinators_sn_boot <- bootstrap_value(df = pollinators_sn,
                                       value = no_species,
                                       groups = c("year",
                                                  "region_name")
                                       )
```
Compare to arbitrary reference value and normalize.
```{r}
pollinators_ref <- 50

pollinators_sn_diff <- pollinators_sn_boot / pollinators_ref
```

Plot the results and display uncertainty.
```{r}
pollinators_sn_diff
```

```{r}
plot(pollinators_sn_diff)
```

Prepare export format.

```{r}
pollinators_sn_to_exp <- pollinators_sn_diff$bootstrap_summary

pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

pollinators_sn_to_exp <- norway_reg_NorIns  %>% 
  inner_join(pollinators_sn_to_exp,
            by = c("region" = "region"),
            multiple = "all")

pollinators_sn_to_exp
```

#### Relationship between Symphyta and Parasitica in semi-natural land

```{r}
symphyta_fam <- c("Argidae", "Cephidae", "Cimbicidae", 
                  "Diprionidae", "Orussidae", "Pamphiliidae", 
                  "Pergidae", "Siricidae", "Anaxyelidae",
                  "Tenthredinidae", "Xiphydriidae", "Xyelidae")

symphyta_sn <- get_observations(subset_year = 2020:2022,
                                subset_families = symphyta_fam,
                                subset_region = NULL)

symphyta_sn
```

[Note, this list needs to be checked over by e.g. Arnstein.]
```{r}
parasitica_fam <- c("Braconidae", "Ichneumonidae", "Chalcididae", 
                    "Eulophidae", "Pteromalidae", "Aphelinidae", 
                    "Scelionidae", "Eupelmidae", "Encyrtidae", 
                    "Mymaridae", "Diapriidae", "Bethylidae", 
                    "Evaniidae", "Ceraphronidae", "Torymidae", 
                    "Dryinidae", "Eucharitidae", "Mymarommatidae", 
                    "Orussidae", "Megaspilidae", "Stephanidae", 
                    "Trigonalidae", "Platygastridae", "Aulacidae", 
                    "Gasteruptiidae", "Rhopalosomatidae", "Larridae", 
                    "Agaonidae", "Pompilidae", "Bradynobaenidae"
                    )

parasitica_sn <- get_observations(subset_year = 2020:2022,
                                subset_families = parasitica_fam,
                                subset_region = NULL)

parasitica_sn
```

Divide the richness of parasitica by symphyta. [Could make a function for this later on.]

```{r}
par_sym <- symphyta_sn %>% 
  full_join(parasitica_sn,
       by = c("year" ="year",
              "locality" = "locality",
              "habitat_type" = "habitat_type",
              "region_name" = "region_name"),
       suffix = c("_sym", "_par")
       ) %>% 
  mutate(par_per_sym_richn = no_species_par / no_species_sym)

```

Bootstrap the fraction of parasitica to symphyta.

```{r}
par_sym_sn_boot <- bootstrap_value(par_sym,
                                value = par_per_sym_richn,
                                groups = c("year", 
                                           "region_name"))
```

Compare to arbitrary reference value and normalize.

```{r}
par_sym_ref <- 40

par_sym_sn_diff <- par_sym_sn_boot / par_sym_ref
```

Plot the results and display uncertainty.
```{r}
par_sym_sn_diff
```

```{r}
plot(par_sym_sn_diff)
```

Prepare export format.

```{r}
par_sym_sn_exp <- par_sym_sn_diff$bootstrap_summary

par_sym_sn_exp <- par_sym_sn_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

par_sym_sn_exp <- norway_reg_NorIns  %>% 
  inner_join(par_sym_sn_exp,
            by = c("region" = "region"),
            multiple = "all")

par_sym_sn_exp
```

#### Genetic diversity evenness of flying insects in semi-natural land

[This is not fully implemented yet. Need full COI sequence to calculate the true GDE]. 
This could also be done for a subset of species.
(doi: https://doi.org/10.1101/2022.02.09.479762)

Fetch the data.
```{r}
richness_sn <- get_observations(subset_year = 2020:2022,
                                subset_region = NULL)
```
Calculate the indicator values for each region and year.
```{r}
gde_sn_boot <- bootstrap_value(df = richness_sn,
                                   value = GDE_by_asv,
                                   groups = c("year",
                                              "region_name")
                                   )
```

Compare to arbitrary reference value and normalize.
```{r}
gde_sn_ref <- 0.7

gde_sn_diff <- gde_sn_boot / gde_sn_ref
```

Plot the results and display uncertainty.
```{r}
gde_sn_diff
```

```{r}
plot(gde_sn_diff)
```

Prepare export format.

```{r}
gde_sn_to_exp <- gde_sn_diff$bootstrap_summary

gde_sn_to_exp <- gde_sn_to_exp %>% 
  select(year,
         region = region_name,
         ind_value = boot_value,
         ind_sd = boot_sd,
         ind_lower2.5 = boot_lower2.5,
         ind_upper97.5 = boot_upper97.5)

gde_sn_to_exp <- norway_reg_NorIns  %>% 
  inner_join(gde_sn_to_exp,
            by = c("region" = "region"),
            multiple = "all")

gde_sn_to_exp
```

## Prepare export
<!-- Text here -->

### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster with indicators values (raw and normalised), reference values and errors. -->

### 2023-06-30 indicators/insectindicators/tutorials/tutorial.Rmd
first draft, not done

@@
 
-Bla bla
+Intro
+======
+Not much here yet, only some quick commands.
 
@@
 ```
+
+```{r}
+require(DBI)
+ require(dplyr)
+locations <- tbl(con,
+                 Id(schema = "locations",
+                    table = "localities"))
+
+```
+```{r}
+locations
+```
+```{r}
+require(sf)
+
+locations_sf <- read_sf(con,
+                 Id(schema = "locations",
+                    table = "localities"))
+
+```
+```{r}
+require(ggplot2)
+
+ggplot(locations_sf) +
+  geom_sf(aes(fill = habitat_type))
+```
+
+New locality function
+---------
+
+```{r}
+insect_localities_sf <- get_localities()
+```
+
+```{r}
+insect_localities <- get_localities(as_sf = FALSE)
+```

### 2023-07-03 DRAFT_ndvi_wetland.Rmd
Made some changes to the text (Intro, references, GEE scripts)

@@
 library(plyr)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
+library(stringr)
+library(tidyverse)
 library(RColorBrewer)
@@
 library(StepBeta)
+library(glmmTMB)
 library(knitr)
@@
 For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed.
-In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e.
-national maps for the forest and mountain ecosystems).
+In both cases, MODIS-derived NDVI values (MOD13Q1 V6-product) were used.
+National maps for forest and mountain ecosystems were used to define the areas of interest.
 Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
@@
 we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a nice time series back to year 2000.
+2) The spatial extent of the single wetland ecosystem occurrences ('polygons' hereafter) includes many wetland areas which are smaller than the resolution of the MODIS imagery (250 x 250 m), which was used for forests and mountains and has a time series going back to the year 2000.
 
 In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
-1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
+1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the correct ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m, which is smaller than almost all wetland polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
+Thus, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
 
@@
 
-- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
-- Sentinel-2 satellite NDVI data
-- MODIS satellite NDVI data
-- Landsat satellite NDVI data
-
-(1) NiN ecosystem data:
-
-(2) Sentinel-2 satellite NDVI data:
-
-(3) MODIS satellite NDVI data:
-
-(4) Landsat satellite NDVI data:
-
-have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which reduces the number of available polygons greatly.
-To explore temporal trends even further back in time, we also include Landsat NDVI data, which span a time frame from ???
-to ???, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
+(1) NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the correct locations and to provide information on the field-assessed ecological condition of mapped polygons.
+
+(2) Sentinel-2 satellite-derived NDVI data (10 x 10 m resolution) - See https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4 for the Google Earth Engine code used to process Sentinel imagery and export NDVI data.
+
+(3) Landsat satellite-derived NDVI data (30 x 30 m resolution).
+See https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7 for the Google Earth Engine code used to process Landsat imagery and export NDVI data.
+
+(4) MODIS satellite-derived NDVI data (250 x 250 m resolution).
+The MOD13Q1 data product was used for NDVI values.
+The product provides a vegetation index value (NDVI) per pixel.
+MODIS NDVI is calculated from atmospherically-corrected bi-directional surface reflectances that have been masked for water, clouds, heavy aerosols and cloud shadows.
+The algorithm chooses the best available pixel value from all the acquisitions from the 16 day period.
+See https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84 for the Google Earth Engine code used to export NDVI data
+
+We have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which greatly reduces the number of available polygons.
+Thus, to explore temporal trends further back in time, we also include Landsat-derived NDVI data, which span a time frame from 1984 to 2021, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
 
@@
 
-The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+The NDVI data from either satellite follows the spatial distribution and bias status of the NiN-data.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than is the case for the NDVI data from Sentinel-2 and Landsat satellites.
 
@@
 
-Sentinel-2 satellite NDVI data span the years 2015-2022.
-MODIS satellite NDVI data span the years ????-????.
-Landsat satellite NDVI data span the years ????-????.
+Available Sentinel-2 satellite NDVI data span the years 2015-2022.
+Available MODIS satellite NDVI data span the years 2000-2022.
+Available Landsat satellite NDVI data span the years 1984-2021.
 
@@
 
+NDVI Literature
+
+Cunliffe, A.M., Assmann, J.J., Daskalova, G.N., Kerby, J.T.
+& Myers-Smith, I.H.
+2020.
+Aboveground biomass corresponds strongly with drone-derived canopy height but weakly with greenness (NDVI) in a shrub tundra landscape.
+Environmental Research Letters 15(12): 125004.
+
+Karlsen, S.R., Elvebakk, A., Høgda, K.A.
+& Grydeland, T.
+2014.
+Spatial and temporal variability in the onset of the growing season on Svalbard, Arctic Norway—measured by MODIS-NDVI satellite data.
+Remote Sensing 6(9): 8088-8106.
+
+Taddeo, S., Dronova, I.
+& Depsky, N.
+2019.
+Spectral vegetation indices of wetland greenness: Responses to vegetation structure, composition, and spatial distribution.
+Remote sensing of Environment 234: 111467.
+
+Tshabalala, N.N., Mutanga, O. & Sibanda, M. 2021. The Utility of Sentinel-2 MSI Data to Estimate Wetland Vegetation Leaf Area Index in Natural and Rehabilitated Wetlands. Geographies 1(3): 178-191.
+
+Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. NTNU.
+
 ## Analyses
@@
 -Sentinel-2 NDVI data
+-Landsat NDVI data
 -MODIS NDVI data
--Landsat NDVI data
 
@@
 ```
-And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, adn Northern Norway
+And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
 ```{r}
@@
 The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
-The image collection contains Sentinel imagery from June, July and August 2015-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+The image collection contains Sentinel imagery from June, July and August 2015-2022.
+The GEE script creates a folder in Google Drive and places the exports there.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files.
 ```{r}
@@
 The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
-The image collection contains MODIS imagery from June, July and August 2000-2022
+The image collection contains MODIS imagery from June, July and August 2000-2022. The GEE script creates a folder in Google Drive and places the exported file there.
 
@@
 df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
-df.m
 df.m$mean<-df.m$mean/10000
@@
 The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
-The image collection contains Landsat imagery from June, July and August 1984-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+The image collection contains Landsat imagery from June, July and August 1984-2022.
+The GEE script creates a folder in Google Drive and places the exports there.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files.
+This script merges them and then merges the dataframe to the NiN data.
 ```{r}
@@
                                  "Semi-naturlig mark" = 'Semi_naturlig',
-                                 "Naturlig åpne områder i lavlandet" = 'Naturlig-aapne',
-                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapne')) %>%
+                                 "Naturlig åpne områder i lavlandet" = 'Naturlig-aapent',
+                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapent')) %>%
   mutate(validGeo = st_is_valid(SHAPE))
@@
 
-colnames(nin.wetland)[c(1,8)] <- c("id","region_id")
+colnames(nin.wetland)[c(1,9)] <- c("id","region_id")
+
+# check that every nin-polygon still occurs only once
+summary(as.factor(nin.wetland$id))
+nin.wetland[nin.wetland$id=="NINFP2110008060",]
+# this one was assigned to both Western and Central Norway, we drop the latter
+nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% '1144.1',]
 
@@
 SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
-summary(SentinelNDVI.wetland)
+#summary(SentinelNDVI.wetland)
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
@@
          area_meters = st_area(SentinelNDVI.wetland))
-summary(SentinelNDVI.wetland)
+#summary(SentinelNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
-summary(SentinelNDVI.wetland)
+#summary(SentinelNDVI.wetland)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
@@
 
-summary(ModisNDVI.wetland)
+#summary(ModisNDVI.wetland)
 ModisNDVI.wetland <- ModisNDVI.wetland %>%
@@
          area_meters = st_area(ModisNDVI.wetland))
-summary(ModisNDVI.wetland)
+#summary(ModisNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
-summary(ModisNDVI.wetland)
+#summary(ModisNDVI.wetland)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
@@
   filter(mean == max(mean, na.rm=TRUE))
-summary(ModisNDVI.wetland)
+#summary(ModisNDVI.wetland)
 
@@
 
-summary(LandsatNDVI.wetland)
+#summary(LandsatNDVI.wetland)
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
@@
          area_meters = st_area(LandsatNDVI.wetland))
-summary(LandsatNDVI.wetland)
+#summary(LandsatNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
-summary(LandsatNDVI.wetland)
+#summary(LandsatNDVI.wetland)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
@@
 
-summary(LandsatNDVI.wetland)
+#summary(LandsatNDVI.wetland)
 ```
@@
 # the pattern is strange, suggest to omit 2022 for this analysis
+
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
@@
 SentinelNDVI.wetland %>%
-  filter(year != '2022') %>%
-
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
 
@@
 
-### Regression analyses, NDVI as a function of condition, Sentinel-2
+### Regression analyses Sentinel-2, NDVI as a function of condition
 ```{r}
@@
 
+# Northern Norway
+```{r}
+SentinelNDVI.wetland.train %>%
+  filter(region=="Northern Norway") %>%
+
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~hovedtype)
+
+summary(model.wetland.cond.Sent.N)$coefficients
+```
+NDVI in good condition: V10 > V1 > V9 > V3
+
+V1: NDVI decreases as condition deteriorates
+V3: NDVI does not change as condition deteriorates
+V9 & V10: NDVI increases as condition deteriorates
+
+# Central Norway
+```{r}
+SentinelNDVI.wetland.train %>%
+  filter(region=="Central Norway") %>%
+
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~hovedtype)
+
+summary(model.wetland.cond.Sent.C)$coefficients
+```
+NDVI in good condition: V10 > V1 & V9 > V3
+
+V1: NDVI decreases as condition deteriorates
+V3, V9 & V10: NDVI increases as condition deteriorates
+
+# Western Norway
+```{r}
+SentinelNDVI.wetland.train %>%
+  filter(region=="Western Norway") %>%
+
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~hovedtype)
+
+summary(model.wetland.cond.Sent.W)$coefficients
+```
+NDVI in good condition: V10 > V9 & V3 > V1
+
+V1 & V10: NDVI increases as condition deteriorates
+V3: NDVI decreases as condition deteriorates (but really insufficient data)
+V9: NDVI does not change as condition deteriorates
+
+# Eastern Norway
+```{r}
+SentinelNDVI.wetland.train %>%
+  filter(region=="Eastern Norway") %>%
+
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~hovedtype)
+
+summary(model.wetland.cond.Sent.E)$coefficients
+```
+NDVI in good condition: V10 > V1 > V3 & V9
+
+V1 & V3: NDVI does not change as condition deteriorates
+V9 & V10: NDVI increases as condition deteriorates
+
+# Southern Norway
+```{r}
+SentinelNDVI.wetland.train %>%
+  filter(region=="Southern Norway") %>%
+
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~hovedtype)
+
+summary(model.wetland.cond.Sent.S)$coefficients
+```
+NDVI in good condition: V10 > V9 > V1 & V3
+
+V1, V9 & V10: NDVI increases as condition deteriorates
+V3: NDVI does not change as condition deteriorates (but also insuffcient data)
+
+Overall take home messages:
+
+???
+
 ### Regression analyses, NDVI across time
-
-### Scaled value calculations
-
-Plant indicators from Tyler et al. (2021)
-```{r, echo = F}
-## Tyler indicator data
-## data from cache
-ind.dat <-readRDS(paste0(here::here(),"/data/cache/ind.dat.RDS"))
-# ind.dat <- read.table("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
-#                         sep = '\t', header=T, quote = '')
-# head(ind.dat)
-
-```
-
-Generalized species lists NiN
-```{r, echo = F}
-# data from cache
-Eco_State<-readRDS(paste0(here::here(),"/data/cache/Eco_State.RDS"))
-# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
-# str(Eco_State)
-```
-
-#### Data handling
-- Checking for errors
-- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
-- Merging indicator data with monitoring data and indicator data with reference data
-(not shown here, but documented in the code)
-
-```{r, include = F}
-
-### Plant indicator data
-names(ind.dat)
-names(ind.dat)[1] <- 'species'
-ind.dat$species <- as.factor(ind.dat$species)
-summary(ind.dat$species)
-ind.dat <- ind.dat[!is.na(ind.dat$species),] # removing species-NAs
-ind.dat[,'species.orig'] <- ind.dat[,'species'] # make a backup of the original species variable
-ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2) # trimming away sub-species & co, and descriptor info
-
-# the trimming above leaves some duplicates that need to be taken care of
-ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
-ind.dat <- ind.dat %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
-                                                            "Anemone nemorosa x ranunculoides",
-                                                            "Armeria maritima ssp. elongata",
-                                                            "Asplenium trichomanes ssp. quadrivalens",
-                                                            "Calystegia sepium ssp. spectabilis",
-                                                            "Campanula glomerata 'Superba'",
-                                                            "Dactylorhiza maculata ssp. fuchsii",
-                                                            "Erigeron acris ssp. droebachensis",
-                                                            "Erigeron acris ssp. politus",
-                                                            "Erysimum cheiranthoides L. ssp. alatum",
-                                                            "Euphrasia nemorosa x stricta var. brevipila",
-                                                            "Galium mollugo x verum",
-                                                            "Geum rivale x urbanum",
-                                                            "Hylotelephium telephium (ssp. maximum)",
-                                                            "Juncus alpinoarticulatus ssp. rariflorus",
-                                                            "Lamiastrum galeobdolon ssp. argentatum",
-                                                            "Lathyrus latifolius ssp. heterophyllus",
-                                                            "Medicago sativa ssp. falcata",
-                                                            "Medicago sativa ssp. x varia",
-                                                            "Monotropa hypopitys ssp. hypophegea",
-                                                            "Ononis spinosa ssp. hircina",
-                                                            "Ononis spinosa ssp. procurrens",
-                                                            "Pilosella aurantiaca ssp. decolorans",
-                                                            "Pilosella aurantiaca ssp. dimorpha",
-                                                            "Pilosella cymosa ssp. gotlandica",
-                                                            "Pilosella cymosa ssp. praealta",
-                                                            "Pilosella officinarum ssp. peleteranum",
-                                                            "Poa x jemtlandica (Almq.) K. Richt.",
-                                                            "Poa x herjedalica Harry Sm.",
-                                                            "Ranunculus peltatus ssp. baudotii",
-                                                            "Sagittaria natans x sagittifolia",
-                                                            "Salix repens ssp. rosmarinifolia",
-                                                            "Stellaria nemorum L. ssp. montana",
-                                                            "Trichophorum cespitosum ssp. germanicum")
-) )
-
-# checking duplicates again
-ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
-# getting rid of sect. for Hieracium
-ind.dat <- ind.dat %>% mutate(species=gsub("sect. ","",species.orig))
-ind.dat[,'species'] <- word(ind.dat[,'species'], 1,2)
-
-ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat.dup <- ind.dat[duplicated(ind.dat[,'species']),"species"]
-ind.dat[ind.dat$species %in% ind.dat.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
-# only hybrids left -> get rid of these
-ind.dat <- ind.dat[!duplicated(ind.dat[,'species']),]
-ind.dat[duplicated(ind.dat[,'species']),"species"]
-
-ind.dat$species <- as.factor(ind.dat$species)
-summary(ind.dat$species)
-# no duplicates left
-
-head(ind.dat)
-
-### ANO monitoring data
-
-head(ANO.sp)
-head(ANO.geo)
-
-## fix NiN information
-ANO.geo$hovedtype_rute <- substr(ANO.geo$kartleggingsenhet_1m2,1,3) # take the 3 first characters
-ANO.geo$hovedtype_rute <- gsub("-", "", ANO.geo$hovedtype_rute) # remove hyphen
-unique(as.factor(ANO.geo$hovedtype_rute))
-
-## fix NiN-variables
-colnames(ANO.geo)
-colnames(ANO.geo)[42:47] <- c("groeftingsintensitet",
-                              "bruksintensitet",
-                              "beitetrykk",
-                              "slatteintensitet",
-                              "tungekjoretoy",
-                              "slitasje")
-head(ANO.geo)
-
-# remove variable code in the data
-ANO.geo$groeftingsintensitet <- gsub("7GR-GI_", "", ANO.geo$groeftingsintensitet) 
-unique(ANO.geo$groeftingsintensitet)
-ANO.geo$groeftingsintensitet <- gsub("X", "NA", ANO.geo$groeftingsintensitet)
-unique(ANO.geo$groeftingsintensitet)
-ANO.geo$groeftingsintensitet <- as.numeric(ANO.geo$groeftingsintensitet)
-unique(ANO.geo$groeftingsintensitet)
-
-ANO.geo$bruksintensitet <- gsub("7JB-BA_", "", ANO.geo$bruksintensitet) 
-unique(ANO.geo$bruksintensitet)
-ANO.geo$bruksintensitet <- gsub("X", "NA", ANO.geo$bruksintensitet)
-unique(ANO.geo$bruksintensitet)
-ANO.geo$bruksintensitet <- as.numeric(ANO.geo$bruksintensitet)
-unique(ANO.geo$bruksintensitet)
-
-ANO.geo$beitetrykk <- gsub("7JB-BT_", "", ANO.geo$beitetrykk) 
-unique(ANO.geo$beitetrykk)
-ANO.geo$beitetrykk <- gsub("X", "NA", ANO.geo$beitetrykk)
-unique(ANO.geo$beitetrykk)
-ANO.geo$beitetrykk <- as.numeric(ANO.geo$beitetrykk)
-unique(ANO.geo$beitetrykk)
-
-ANO.geo$slatteintensitet <- gsub("7JB-SI_", "", ANO.geo$slatteintensitet) 
-unique(ANO.geo$slatteintensitet)
-ANO.geo$slatteintensitet <- gsub("X", "NA", ANO.geo$slatteintensitet)
-unique(ANO.geo$slatteintensitet)
-ANO.geo$slatteintensitet <- as.numeric(ANO.geo$slatteintensitet)
-unique(ANO.geo$slatteintensitet)
-
-ANO.geo$tungekjoretoy <- gsub("7TK_", "", ANO.geo$tungekjoretoy) 
-unique(ANO.geo$tungekjoretoy)
-ANO.geo$tungekjoretoy <- gsub("X", "NA", ANO.geo$tungekjoretoy)
-unique(ANO.geo$tungekjoretoy)
-ANO.geo$tungekjoretoy <- as.numeric(ANO.geo$tungekjoretoy)
-unique(ANO.geo$tungekjoretoy)
-
-ANO.geo$slitasje <- gsub("7SE_", "", ANO.geo$slitasje) 
-unique(ANO.geo$slitasje)
-ANO.geo$slitasje <- gsub("X", "NA", ANO.geo$slitasje)
-unique(ANO.geo$slitasje)
-ANO.geo$slitasje <- as.numeric(ANO.geo$slitasje)
-unique(ANO.geo$slitasje)
-
-## check that every point is present only once
-length(levels(as.factor(ANO.geo$ano_flate_id)))
-length(levels(as.factor(ANO.geo$ano_punkt_id)))
-summary(as.factor(ANO.geo$ano_punkt_id))
-# there's many double presences, probably some wrong registrations of point numbers,
-# but also double registrations (e.g. ANO0159_55)
-# CHECK THIS when preparing ecosystem-datasets for scaling
-
-# fix species names
-ANO.sp$Species <- ANO.sp$art_navn
-unique(as.factor(ANO.sp$Species))
-ANO.sp[,'Species'] <- word(ANO.sp[,'Species'], 1,2) # lose subspecies
-ANO.sp$Species <- str_to_title(ANO.sp$Species) # make first letter capital
-ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after hyphen to lowercase
-ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after space to lowercase
-
-## merge species data with indicators
-ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
-                y= ind.dat[,c("species", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")],
-                by.x="Species", by.y="species", all.x=T)
-summary(ANO.sp.ind)
-
-## checking which species didn't find a match
-unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
-
-# fix species name issues
-ind.dat <- ind.dat %>% 
-  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
-  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
-  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
-  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
-  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
-  mutate(species=str_replace(species,"Cherleria biflora", "Minuartia biflora"))
-
-ANO.sp <- ANO.sp %>% 
-  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
-  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
-  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
-  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
-  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
-  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
-  mutate(Species=str_replace(Species,"Carex paupercula", "Carex magellanica")) %>%
-  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
-  mutate(Species=str_replace(Species,"Carex viridula", "Carex flava")) %>%
-  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
-  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
-  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
-  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
-  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
-  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
-  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
-  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
-  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
-  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
-  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
-  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
-  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Vulgata")) %>%
-  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
-  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
-  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
-  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
-  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
-  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
-  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
-  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
-  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
-  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
-  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
-  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
-  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
-  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
-  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
-  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
-  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
-  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
-  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
-  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
-  mutate(Species=str_replace(Species,"Poa alpigena", "Poa pratensis")) %>%
-  mutate(Species=str_replace(Species,"Poa angustifolia", "Poa pratensis")) %>%
-  mutate(Species=str_replace(Species,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
-  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa")) %>%
-  mutate(Species=str_replace(Species,"Syringa emodi", "Syringa vulgaris")) %>%
-  mutate(Species=str_replace(Species,"Taraxacum crocea", "Taraxacum officinale")) %>%
-  mutate(Species=str_replace(Species,"Taraxacum croceum", "Taraxacum officinale")) %>%
-  mutate(Species=str_replace(Species,"Trientalis europaea", "Lysimachia europaea")) %>%
-  mutate(Species=str_replace(Species,"Trifolium pallidum", "Trifolium pratense"))
-
-## merge species data with indicators
-ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
-                    y= ind.dat[,c("species", "Light", "Moisture", "Soil_reaction_pH", "Nitrogen")],
-                    by.x="Species", by.y="species", all.x=T)
-summary(ANO.sp.ind)
-# checking which species didn't find a match
-unique(ANO.sp.ind[is.na(ANO.sp.ind$Moisture),'Species'])
-# don't find synonyms for these in the ind lists
-
-## trimming away the points without information on NiN, species or cover
-ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$Species),]
-ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$art_dekning),]
-
-summary(ANO.sp.ind)
-head(ANO.sp.ind)
-rm(ANO.sp)
-
-### NiN reference data - data handling
-
-## generalized species lists from NiN
-str(Eco_State)
-
-# species
-Eco_State$Concept_Data$Species$Species_List$species
-# environments
-t(Eco_State$Concept_Data$Env$Env_Data)
-# abundances
-t(Eco_State$Concept_Data$Species$Species_Data)
-
-## transposing abundance data
-NiN.sp <- t(Eco_State$Concept_Data$Species$Species_Data)
-NiN.sp <- as.data.frame(NiN.sp)
-NiN.sp$sp <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$species))
-# only genus and species name
-NiN.sp$sp <- word(NiN.sp$sp, 1,2)
-NiN.sp$spgr <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code))
-# if relevant, trimming to desired species groups (for forests e.g. removing trees)
-#NiN.sp <- NiN.sp[NiN.sp$spgr!="a1a",]
-
-## environment data
-NiN.env <- Eco_State$Concept_Data$Env$Env_Data
-
-## merging with indicator values
-NiN.sp.ind <- merge(NiN.sp,ind.dat, by.x="sp", by.y="species", all.x=T)
-summary(NiN.sp.ind)
-
-NiN.sp.ind[NiN.sp.ind==999] <- NA
-
-## checking which species didn't find a match
-unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) & NiN.sp.ind$spgr %in% list("a1a","a1b","a1c"),'sp'])
-
-## fix species name issues
-ind.dat <- ind.dat %>% 
-  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
-  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
-  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
-  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
-  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
-  mutate(species=str_replace(species,"Cherleria biflora", "Minuartia biflora"))
-
-NiN.sp <- NiN.sp %>% 
-  mutate(sp=str_replace(sp,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
-  mutate(sp=str_replace(sp,"Anagallis minima", "Lysimachia minima")) %>% 
-  mutate(sp=str_replace(sp,"Arctous alpinus", "Arctous alpina")) %>%
-  mutate(sp=str_replace(sp,"Betula tortuosa", "Betula pubescens")) %>%
-  mutate(sp=str_replace(sp,"Blysmopsis rufa", "Blysmus rufus")) %>%
-  mutate(sp=str_replace(sp,"Cardamine nymanii", "Cardamine pratensis")) %>%
-  mutate(sp=str_replace(sp,"Carex adelostoma", "Carex buxbaumii")) %>%
-  mutate(sp=str_replace(sp,"Carex leersii", "Carex echinata")) %>%
-  mutate(sp=str_replace(sp,"Carex paupercula", "Carex magellanica")) %>%
-  mutate(sp=str_replace(sp,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
-  mutate(sp=str_replace(sp,"Carex _vacillans", "Carex vacillans")) %>%
-  mutate(sp=str_replace(sp,"Carex viridula", "Carex flava")) %>%
-  mutate(sp=str_replace(sp,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
-  mutate(sp=str_replace(sp,"Cornus suecia", "Cornus suecica")) %>%
-  mutate(sp=str_replace(sp,"Cicerbita alpina", "Lactuca alpina")) %>%
-  mutate(sp=str_replace(sp,"Dactylorhiza sphagnicola", "Dactylorhiza majalis")) %>%
-  mutate(sp=str_replace(sp,"Diphasiastrum complanatum", "Lycopodium complanatum")) %>%
-  mutate(sp=str_replace(sp,"Elymus alaskanus", "Elymus kronokensis")) %>%
-  mutate(sp=str_replace(sp,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
-  mutate(sp=str_replace(sp,"Erigeron eriocephalus", "Erigeron uniflorus")) %>%
-  mutate(sp=str_replace(sp,"Festuca prolifera", "Festuca rubra")) %>%
-  mutate(sp=str_replace(sp,"Galium album", "Galium mollugo")) %>%
-  mutate(sp=str_replace(sp,"Galium elongatum", "Galium palustre")) %>%
-  mutate(sp=str_replace(sp,"Glaux maritima", "Lysimachia maritima")) %>%
-  mutate(sp=str_replace(sp,"Helictotrichon pratense", "Avenula pratensis")) %>%
-  mutate(sp=str_replace(sp,"Helictotrichon pubescens", "Avenula pubescens")) %>%
-  mutate(sp=str_replace(sp,"Hieracium alpina", "Hieracium Alpina")) %>%
-  mutate(sp=str_replace(sp,"Hieracium alpinum", "Hieracium Alpina")) %>%
-  mutate(sp=str_replace(sp,"Hieracium aurantiacum", "Pilosella aurantiaca")) %>%
-  mutate(sp=str_replace(sp,"Hieracium hieracium", "Hieracium Hieracium")) %>%
-  mutate(sp=str_replace(sp,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
-  mutate(sp=str_replace(sp,"Hieracium lactucella", "Pilosella lactucella")) %>%
-  mutate(sp=str_replace(sp,"Hieracium murorum", "Hieracium Vulgata")) %>%
-  mutate(sp=str_replace(sp,"Hieracium oreadea", "Hieracium Oreadea")) %>%
-  mutate(sp=str_replace(sp,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
-  mutate(sp=str_replace(sp,"Hieracium vulgata", "Hieracium Vulgata")) %>%
-  mutate(sp=str_replace(sp,"Hieracium pilosella", "Pilosella officinarum")) %>%
-  mutate(sp=str_replace(sp,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
-  mutate(sp=str_replace(sp,"Hierochloã« alpina", "Hierochloë alpina")) %>%
-  mutate(sp=str_replace(sp,"Hierochloã« hirta", "Hierochloë hirta")) %>%
-  mutate(sp=str_replace(sp,"Hierochloã« odorata", "Hierochloë odorata")) %>%
-  mutate(sp=str_replace(sp,"Huperzia appressa", "Huperzia selago")) %>%
-  mutate(sp=str_replace(sp,"Hylotelephium maximum", "Hylotelephium telephium")) %>%
-  mutate(sp=str_replace(sp,"Lappula myosotis", "Lappula squarrosa")) %>%
-  mutate(sp=str_replace(sp,"Lepidotheca suaveolens", "Matricaria discoidea")) %>%
-  mutate(sp=str_replace(sp,"Listera cordata", "Neottia cordata")) %>%
-  mutate(sp=str_replace(sp,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
-  mutate(sp=str_replace(sp,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
-  mutate(sp=str_replace(sp,"Logfia arvensis", "Filago arvensis")) %>%
-  mutate(sp=str_replace(sp,"Mentha _verticillata", "Mentha verticillata")) %>%
-  mutate(sp=str_replace(sp,"Minuartia rubella", "Sabulina rubella")) %>%
-  mutate(sp=str_replace(sp,"Minuartia stricta", "Sabulina stricta")) %>%
-  mutate(sp=str_replace(sp,"Mycelis muralis", "Lactuca muralis")) %>%
-  mutate(sp=str_replace(sp,"Omalotheca supina", "Gnaphalium supinum")) %>%
-  mutate(sp=str_replace(sp,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
-  mutate(sp=str_replace(sp,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
-  mutate(sp=str_replace(sp,"Ononis arvensis", "Ononis spinosa")) %>%
-  mutate(sp=str_replace(sp,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
-  mutate(sp=str_replace(sp,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
-  mutate(sp=str_replace(sp,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
-  mutate(sp=str_replace(sp,"Phalaris minor", "Phalaris arundinacea")) %>%
-  mutate(sp=str_replace(sp,"Phalaroides arundinacea", "Phalaris arundinacea")) %>%
-  mutate(sp=str_replace(sp,"Pinus unicinata", "Pinus mugo")) %>%
-  mutate(sp=str_replace(sp,"Platanthera montana", "Platanthera chlorantha")) %>%
-  mutate(sp=str_replace(sp,"Poa alpigena", "Poa pratensis")) %>%
-  mutate(sp=str_replace(sp,"Poa angustifolia", "Poa pratensis")) %>%
-  mutate(sp=str_replace(sp,"Poa laxa", "Poa flexuosa")) %>%
-  mutate(sp=str_replace(sp,"Poa _herjedalica", "Poa herjedalica")) %>%
-  mutate(sp=str_replace(sp,"Poa _jemtlandica", "Poa jemtlandica")) %>%
-  mutate(sp=str_replace(sp,"Poa lindebergii", "Poa arctica")) %>%
-  mutate(sp=str_replace(sp,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
-  mutate(sp=str_replace(sp,"Rhamnus catharticus", "Rhamnus cathartica")) %>%
-  mutate(sp=str_replace(sp,"Rumex alpestris", "Rumex acetosa")) %>%
-  mutate(sp=str_replace(sp,"Salix _fragilis", "Salix fragilis")) %>%
-  mutate(sp=str_replace(sp,"Saxifraga _opdalensis", "Saxifraga opdalensis")) %>%
-  mutate(sp=str_replace(sp,"Spergularia salina", "Spergularia marina")) %>%
-  mutate(sp=str_replace(sp,"Syringa emodi", "Syringa vulgaris")) %>%
-  mutate(sp=str_replace(sp,"Taraxacum crocea", "Taraxacum officinale")) %>%
-  mutate(sp=str_replace(sp,"Taraxacum croceum", "Taraxacum officinale")) %>%
-  mutate(sp=str_replace(sp,"Taraxacum erythrospermum", "Taraxacum officinale")) %>%
-  mutate(sp=str_replace(sp,"Taraxacum hamatum", "Taraxacum officinale")) %>%
-  mutate(sp=str_replace(sp,"Trientalis europaea", "Lysimachia europaea")) %>%
-  mutate(sp=str_replace(sp,"Trifolium pallidum", "Trifolium pratense")) %>%
-  mutate(sp=str_replace(sp,"Vicia orobus", "Vicia cassubica"))
-
-## merge species data with indicators
-NiN.sp.ind <- merge(NiN.sp,ind.dat, by.x="sp", by.y="species", all.x=T)
-summary(NiN.sp.ind)
-
-NiN.sp.ind[NiN.sp.ind==999] <- NA
-
-# checking which species didn't find a match
-unique(NiN.sp.ind[is.na(NiN.sp.ind$Moisture) & NiN.sp.ind$spgr %in% list("a1a","a1b","a1c"),'sp'])
-# ok now
-
-## matching with NiN ecosystem types - for wetlands
-# NB! beware of rogue spaces in the 'Nature_type' & 'Sub_Type' variables, e.g. "Spring_Forest "
-NiN.wetland <- NiN.sp.ind[,c("sp",paste(NiN.env[NiN.env$Nature_Type=="Mire","ID"]),colnames(ind.dat)[15:18])]   # Light, Moisture, Soil_reaction_pH, Nitrogen
-NiN.wetland[1,]
-names(NiN.wetland)
-
-cbind(colnames(NiN.wetland),
-      c("",
-        'V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
-        'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
-        'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
-        'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
-        'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
-        'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
-        'V3-C2','V1-C5',
-        'V1-C6a','V1-C6b',
-        'V1-C7a','V1-C7b',
-        'V1-C8a','V1-C8b',
-        'V2-C1a','V2-C1b',
-        'V2-C2a','V2-C2b',
-        'V2-C3a','V2-C3b',
-        "V4-C2","V4-C3",
-        "","",
-        'V8-C1','V8-C2','V8-C3',
-        rep("",10),
-        rep("",4)
+First, there's again some data handling to do. We merge the Sentinel, Modis, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
+```{r}
+## data handling for time series analysis
+# Sentinel time series checked in exploratory analysis script
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+
+# checking time series for MODIS
+ModisNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+  ggplot( aes(x=year, y=mean )) + 
+  geom_point() +
+  facet_wrap(~hovedtype)
+# 2022 does not stand out as in the Sentinel data, so we keep it
+ModisNDVI.wetland <- ModisNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+
+# checking time series for Landsat
+LandsatNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+  ggplot( aes(x=year, y=mean )) + 
+  geom_point() +
+  facet_wrap(~hovedtype)
+# nothing worrying to see here either
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE))
+
+# transformation of NDVI scale to beta scale
+#SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
+ModisNDVI.wetland$mean_beta <- (ModisNDVI.wetland$mean + 1) / 2
+LandsatNDVI.wetland$mean_beta <- (LandsatNDVI.wetland$mean + 1) / 2
+
+# check if there's any 0s or 1s (which beta cannot handle)
+summary(SentinelNDVI.wetland$mean_beta)
+summary(ModisNDVI.wetland$mean_beta)
+summary(LandsatNDVI.wetland$mean_beta)
+# replace 1s in Landsat data with 0.9999
+LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+  mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
+
+# check if the three Sattelite objects have the same structure (for concatenating them)
+names(SentinelNDVI.wetland)
+names(ModisNDVI.wetland)
+names(LandsatNDVI.wetland)
+# Sentinel and Landsat have each an extra column -> omit them when concatenating further below
+# one column is named slightly differently in the Sentinel data -> rename it
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+  rename('system.index' = 'system:index')
+
+# check if they have the same geometry
+st_crs(SentinelNDVI.wetland)
+st_crs(ModisNDVI.wetland)
+st_crs(LandsatNDVI.wetland)
+# all good
+
+# add an increment to the year variable to avoid overlapping data being hidden in figures
+SentinelNDVI.wetland$year_jit <- SentinelNDVI.wetland$year + 0.3
+ModisNDVI.wetland$year_jit <- ModisNDVI.wetland$year - 0.3
+LandsatNDVI.wetland$year_jit <- LandsatNDVI.wetland$year
+
+# concatenate the three Satellite objects
+allSatNDVI.wetland <- rbind(
+SentinelNDVI.wetland[,!names(SentinelNDVI.wetland) %in% "subtype"],
+ModisNDVI.wetland,
+LandsatNDVI.wetland[,!names(LandsatNDVI.wetland) %in% "column_label"]
 )
+# add variable for Satellite indentity
+allSatNDVI.wetland$Sat <- c(
+  rep("Sentinel",nrow(SentinelNDVI.wetland)),
+  rep("Modis",nrow(ModisNDVI.wetland)),
+  rep("Landsat",nrow(LandsatNDVI.wetland))
 )
-
-NiN.wetland <- NiN.wetland[,c(1:43,46:48,59:62)]
-colnames(NiN.wetland)[2:46] <- c('V3-C1a','V3-C1b','V3-C1c','V3-C1d','V3-C1e',
-                              'V1-C1a','V1-C1b','V1-C1c','V1-C1d','V1-C1e',
-                              'V1-C2a','V1-C2b','V1-C2c','V1-C2d',
-                              'V1-C3a','V1-C3b','V1-C3c','V1-C3d',
-                              'V1-C4a','V1-C4b','V1-C4c','V1-C4d',
-                              'V1-C4e','V1-C4f','V1-C4g','V1-C4h',
-                              'V3-C2','V1-C5',
-                              'V1-C6a','V1-C6b',
-                              'V1-C7a','V1-C7b',
-                              'V1-C8a','V1-C8b',
-                              'V2-C1a','V2-C1b',
-                              'V2-C2a','V2-C2b',
-                              'V2-C3a','V2-C3b',
-                              'V4-C2','V4-C3',
-                              'V8-C1','V8-C2','V8-C3'
-)
-head(NiN.wetland)
-
-# translating the abundance classes into %-cover
-coverscale <- data.frame(orig=0:6,
-                         cov=c(0,1/32,1/8,3/8,0.6,4/5,1)
-)
-
-NiN.wetland.cov <- NiN.wetland
-colnames(NiN.wetland.cov)
-for (i in 2:46) {
-  NiN.wetland.cov[,i] <- coverscale[,2][ match(NiN.wetland[,i], 0:6 ) ]
-}
-
-NiN.wetland.cov$sp <- as.factor(NiN.wetland.cov$sp)
-```
-leaving us with the monitoring data including plant indicators (ANO.sp.ind) and the reference data including plant indicators (NiN.wetland.cov)
-
-```{r}
-head(ANO.sp.ind)
-head(NiN.wetland.cov)
-```
-For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
-For wetland ecosystem we are testing "Light", "Moisture", "Soil_reaction_pH", and "Nitrogen".
-In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.
-
-##### bootstrap function for frequency abundance
-- function to calculate community weighted means of selected indicator values (ind)
-- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
-- with a given number of iterations (iter),
-- with species given a certain minimum abundance occurring in all bootstraps (obl), and
-- with a given re-sampling ratio of the original species list (rat)
-- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)
-
-```{r, echo = F}
-indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
-
-  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
-  colnames(ind.b) <- colnames(abun)
-  ind.b <- as.data.frame(ind.b)
-
-  ind <- as.data.frame(ind)
-  ind.list <- as.list(1:length(colnames(ind)))
-  names(ind.list) <- colnames(ind)
-
-  for (k in 1:length(colnames(ind)) ) {
-    ind.list[[k]] <- ind.b }
-
-  for (j in 1:length(colnames(abun)) ) {
-
-    dat <- cbind(sp,abun[,j],ind)
-    dat <- dat[dat[,2]>0,]            # only species that are present in the ecosystem
-    dat <- dat[!is.na(dat[,3]),]      # only species that have indicator values
-
-    for (i in 1:iter) {
-
-      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
-      dat.b <- rbind(dat[dat[,2] >= obl,],
-                     dat[match(speciesSample,dat$sp),]
-      )
-
-      if (var.abun==T) {
-        for (m in 1:nrow(coverscale[-1,]) ) {
-          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
-          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
-          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
-          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
-          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
-          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
-          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
-        }
-        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
-        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
-      }
-
-      for (k in 1:length(colnames(ind))) {
-
-        if ( nrow(dat.b)>2 ) {
-
-          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
-          ind.list[[k]][i,j] <- ind.b
-
-        } else {ind.list[[k]][i,j] <- NA}
-
-      }
-
-#      print(paste(i,"",j)) 
-    }
-
-  }
-  return(ind.list)
-}
-```
-
-Running the bootstraps
-
-```{r, echo = T, eval = F}
-colnames(NiN.wetland)
-# 1st column is the species
-# 2nd-46th column is the abundances of sp in different ecosystem types
-# 47th-50th column is the indicator values of the respective species
-# we choose 1000 iterations
-# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
-# each sample re-samples 1/3 of the number of species
-# the abundance of the re-sampled species may vary (see bootstrap function for details)
-wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
-                          iter=1000,obl=1,rat=1/3,var.abun=T)
-
-# fixing NaNs
-for (i in 1:length(wetland.ref.cov) ) {
-  for (j in 1:ncol(wetland.ref.cov[[i]]) ) {
-    v <- wetland.ref.cov[[i]][,j]
-    v[is.nan(v)] <- NA
-    wetland.ref.cov[[i]][,j] <- v
-  }
-}
-```
-
-```{r, include = F}
-# Data from cache
-wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))
-
-#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
-```
-
-```{r}
-head(wetland.ref.cov[[1]])
-```
-
-This results in an R-list with a slot for every selected indicator, and in every slot there's a data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.
-Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:
-
-- Median = reference values
-- 0.025 and 0.975 quantiles = lower and upper limit values
-- min and max of the respective indicator's scale = min/max values
-
-```{r, echo = F}
-
-# every NiN-type is represented by one 'generalisert artsliste'
-# some NiN-types are represented by two such species lists
-# in some cases two NiN-types are represented by the same species list
-head(wetland.ref.cov[[1]])
-wetland.ref.cov[[1]][0,]
-
-# checking the actual NiN-types in the wetland lists
-wetland.NiNtypes <- colnames(wetland.ref.cov[["Light"]])
-wetland.NiNtypes <- substr(wetland.NiNtypes,1,5)
-unique(wetland.NiNtypes)
-
-# 4 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, "Nitrogen"
-indEll.n=4
-# creating a table to hold:
-# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
-# for every nature type (nrows)
-tab <- matrix(ncol=3*indEll.n, nrow=length(unique(wetland.NiNtypes)) ) # 43 basic ecosystem types
-# coercing the values into the table
-# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
-names(wetland.ref.cov[["Light"]])
-x <- c(27,28,41:45)
-
-for (i in 1:length(x) ) {
-  tab[i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
-}
-
-tab <- as.data.frame(tab)
-tab$NiN <- NA
-tab$NiN[1:length(x)] <- names(wetland.ref.cov[[1]])[x]
-
-# NiN-types represented by several species lists
-wetland.NiNtypes2 <- wetland.NiNtypes[-x]
-unique(wetland.NiNtypes2)
-grep(pattern=unique(wetland.NiNtypes2)[1], x=wetland.NiNtypes) # finds columns in e.g. colnames(wetland.ref.cov[["Continentality"]]) that match the first NiN-type
-
-for (i in 1:length(unique(wetland.NiNtypes2)) ) {
-  tab[length(x)+i,1:3] <- quantile(as.matrix(wetland.ref.cov[["Light"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[length(x)+i,4:6] <- quantile(as.matrix(wetland.ref.cov[["Moisture"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[length(x)+i,7:9] <- quantile(as.matrix(wetland.ref.cov[["Soil_reaction_pH"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab[length(x)+i,10:12] <- quantile(as.matrix(wetland.ref.cov[["Nitrogen"]][,grep(pattern=unique(wetland.NiNtypes2)[i], x=wetland.NiNtypes)]),probs=c(0.025,0.5,0.975),na.rm=T)
-  tab$NiN[length(x)+i] <- unique(wetland.NiNtypes2)[i]
-
-}
-
-tab
-
-# making it a proper data frame
-round(tab[,1:12],digits=2)
-
-colnames(tab) <- c("Light_q2.5","Light_q50","Light_q97.5",
-                   "Moist_q2.5","Moist_q50","Moist_q97.5",
-                   "pH_q2.5","pH_q50","pH_q97.5",
-                   "Nitrogen_q2.5","Nitrogen_q50","Nitrogen_q97.5",
-                   "NiN")
-summary(tab)
-tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphon after C for NiN-types
-tab
-
-# restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
-y.Light <- numeric(length=nrow(tab)*2)
-y.Light[((1:dim(tab)[1])*2)-1] <- tab$Light_q2.5 
-y.Light[((1:dim(tab)[1])*2)] <- tab$Light_q97.5
-
-y.Moist <- numeric(length=nrow(tab)*2)
-y.Moist[((1:dim(tab)[1])*2)-1] <- tab$Moist_q2.5 
-y.Moist[((1:dim(tab)[1])*2)] <- tab$Moist_q97.5
-
-y.pH <- numeric(length=nrow(tab)*2)
-y.pH[((1:dim(tab)[1])*2)-1] <- tab$pH_q2.5 
-y.pH[((1:dim(tab)[1])*2)] <- tab$pH_q97.5
-
-y.Nitrogen <- numeric(length=nrow(tab)*2)
-y.Nitrogen[((1:dim(tab)[1])*2)-1] <- tab$Nitrogen_q2.5 
-y.Nitrogen[((1:dim(tab)[1])*2)] <- tab$Nitrogen_q97.5
-
-# creating final objects holding the reference and limit values for all indicators
-
-# ref object for indicators
-wetland.ref.cov.val <- data.frame(N1=rep('wetland',(nrow(tab)*2*indEll.n)),
-                              hoved=c(rep('NA',(nrow(tab)*2*indEll.n))),
-                              grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
-                              county=rep('all',(nrow(tab)*2*indEll.n)),
-                              region=rep('all',(nrow(tab)*2*indEll.n)),
-                              Ind=c(rep(c('Light1','Light2'),nrow(tab)),
-                                    rep(c('Moist1','Moist2'),nrow(tab)),
-                                    rep(c('pH1','pH2'),nrow(tab)),
-                                    rep(c('Nitrogen1','Nitrogen2'),nrow(tab))
-                              ),
-                              Rv=c(rep(tab$Light_q50,each=2),
-                                   rep(tab$Moist_q50,each=2),
-                                   rep(tab$pH_q50,each=2),
-                                   rep(tab$Nitrogen_q50,each=2)
-                              ),
-                              Gv=c(y.Light,y.Moist,y.pH,y.Nitrogen),
-                              maxmin=c(rep(c(1,7),nrow(tab)), # 7 levels of light
-                                       rep(c(1,12),nrow(tab)), # 12 levels of moisture
-                                       rep(c(1,8),nrow(tab)), # 8 levels of soil reaction pH
-                                       rep(c(1,9),nrow(tab))  # 9 levels of nitrogen
-                              )
-)
-
-wetland.ref.cov.val
-wetland.ref.cov.val$grunn <- as.factor(wetland.ref.cov.val$grunn)
-wetland.ref.cov.val$Ind <- as.factor(wetland.ref.cov.val$Ind)
-summary(wetland.ref.cov.val)
-
-```
-
-```{r}
-head(wetland.ref.cov.val)
-```
-
-Once test data (ANO) and the scaling values from the reference data (NiN) are in place, we can calculate CWMs of the selected indicators for the ANO community data and scale them against the scaling values from the reference distribution.
-Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
-Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each functional plant indicator.
-An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.
-
-Here is the scaling function
-```{r}
-
-#### scaled values ####
-r.s <- 1    # reference value
-l.s <- 0.6  # limit value
-a.s <- 0    # abscence of indicator, or indicator at maximum
-
-#### function for calculating scaled values for measured value ####
-
-## scaling function including truncation
-scal <- function() {
-  # place to hold the result
-   x <- numeric()
-  if (maxmin < ref) {
-    # values >= the reference value equal 1
-    if (val >= ref) {x <- 1}
-    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
-    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
-    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
-    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
-    # value equals or lower than maxmin
-    if (val <= maxmin) {x <-0}
-  } else {
-    # values <= the reference value equal 1
-    if (val <= ref) {x <- 1}
-    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
-    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
-    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
-    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
-    # value equals or larger than maxmin
-    if (val >= maxmin) {x <-0}
-  }
-  return(x)
-
-}
-
-```
-
-We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
-```{r, include=FALSE}
-
-unique(ANO.geo$hovedtype_rute) # NiN types in data
-unique(substr(wetland.ref.cov.val$grunn,1,2)) # NiN types in reference
-#### creating dataframe to hold the results for wetlands ####
-# all ANO points
-nrow(ANO.geo)
-# all wetland ANO points
-nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),])
-# all wetland ANO points with a NiN-type represented in the reference
-nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(substr(wetland.ref.cov.val$grunn,1,2)),])
-# ok, we'll be losing 70 (our of 1349) that are not covered by the reference
-ANO.wetland <- ANO.geo[ANO.geo$hovedtype_rute %in% list("V1","V2","V3","V4","V5","V6","V7","V8","V9","V10","V11","V12","V13"),]
-
-head(ANO.wetland)
-# update row-numbers
-row.names(ANO.wetland) <- 1:nrow(ANO.wetland)
-head(ANO.wetland)
-dim(ANO.wetland)
-colnames(ANO.wetland)
-
-length(levels(as.factor(ANO.wetland$ano_flate_id)))
-length(levels(as.factor(ANO.wetland$ano_punkt_id)))
-summary(as.factor(ANO.wetland$ano_punkt_id))
-# four points that are double
-ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0159_55",] # double registration, said so in comment. -> choose row 207 over 206
-ANO.wetland <- ANO.wetland[-206,]
-row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
-ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0283_22",] # 2019 & 2021. Lot of NA's in 2019 -> omit 2019
-ANO.wetland <- ANO.wetland[-156,]
-row.names(ANO.wetland) <- 1:nrow(ANO.wetland) # update row-numbers
-ANO.wetland[ANO.wetland$ano_punkt_id=="ANO0363_24",]
-ANO.wetland[ANO.wetland$ano_flate_id=="ANO0363","ano_punkt_id"] # point-ID 15 is missing, but 24 is double.
-Likely that registrations are valid, but wrong point-ID.
--> keep both, call the second obs the one that's missing
-ANO.wetland[311,"ano_punkt_id"] <- "ANO0363_15"
-ANO.wetland[ANO.wetland$ano_punkt_id=="ANO1550_64",] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
-ANO.wetland[ANO.wetland$ano_flate_id=="ANO1550","ano_punkt_id"] # point-ID 66 is missing, but 64 is double. Likely that registrations are valid, but wrong point-ID.  -> keep both
-ANO.wetland[1273,"ano_punkt_id"] <- "ANO1550_66"
-
-unique(ANO.wetland$hovedoekosystem_punkt)
-unique(ANO.wetland$hovedtype_rute)
-unique(ANO.wetland$kartleggingsenhet_1m2)
-ANO.wetland$hovedtype_rute <- factor(ANO.wetland$hovedtype_rute)
-ANO.wetland$kartleggingsenhet_1m2 <- factor(ANO.wetland$kartleggingsenhet_1m2)
-summary(ANO.wetland$Hovedtype_rute)
-summary(ANO.wetland$Kartleggingsenhet_rute)
-
-results.wetland <- list()
-ind <- unique(wetland.ref.cov.val$Ind)
-# choose columns for site description
-colnames(ANO.wetland)
-results.wetland[['original']] <- ANO.wetland
-# drop geometry
-st_geometry(results.wetland[['original']]) <- NULL
-results.wetland[['original']]
-
-# add columns for indicators
-nvar.site <- ncol(results.wetland[['original']])
-for (i in 1:length(ind) ) {results.wetland[['original']][,i+nvar.site] <- NA}
-colnames(results.wetland[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
-for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.wetland[['original']][,i] <- as.numeric(results.wetland[['original']][,i])}
-summary(results.wetland[['original']])
-#results.wetland[['original']]$Region <- as.factor(results.wetland[['original']]$Region)
-results.wetland[['original']]$GlobalID <- as.factor(results.wetland[['original']]$GlobalID)
-results.wetland[['original']]$ano_flate_id <- as.factor(results.wetland[['original']]$ano_flate_id)
-results.wetland[['original']]$ano_punkt_id <- as.factor(results.wetland[['original']]$ano_punkt_id)
-results.wetland[['original']]$hovedoekosystem_punkt <- as.factor(results.wetland[['original']]$hovedoekosystem_punkt)
-#results.wetland[['original']]$Hovedoekosystem_rute  <- as.factor(results.wetland[['original']]$Hovedoekosystem_rute )
-results.wetland[['original']]$kartleggingsenhet_1m2 <- as.factor(results.wetland[['original']]$kartleggingsenhet_1m2)
-results.wetland[['original']]$hovedtype_rute    <- as.factor(results.wetland[['original']]$hovedtype_rute)
-
-# roll out
-results.wetland[['scaled']] <- results.wetland[['non-truncated']] <- results.wetland[['original']]
-```
-
-```{r, echo = T, eval = F}
-
-#### calculating scaled and non-truncated values for the indicators based on the dataset ####
-for (i in 1:nrow(ANO.wetland) ) {  #
-  tryCatch({
-    print(i)
-    print(paste(ANO.wetland$ano_flate_id[i]))
-    print(paste(ANO.wetland$ano_punkt_id[i]))
-#    ANO.wetland$Hovedoekosystem_sirkel[i]
-#    ANO.wetland$Hovedoekosystem_rute[i]
-
-    # if the ANO.hovedtype exists in the reference
-    if (ANO.wetland$hovedtype_rute[i] %in% unique(substr(wetland.ref.cov.val$grunn,1,2)) ) {
-
-      # if there is any species present in current ANO point  
-      if ( length(ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),'Species']) > 0 ) {
-
-          # Light
-          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Light')]
-          results.wetland[['original']][i,'richness'] <- nrow(dat)
-          dat <- dat[!is.na(dat$Light),]
-
-          if ( nrow(dat)>0 ) {
-
-            val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
-            # lower part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Light1'] <- scal() 
-            results.wetland[['non-truncated']][i,'Light1'] <- scal.2() 
-            results.wetland[['original']][i,'Light1'] <- val
-
-            # upper part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Light2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Light2'] <- scal() 
-            results.wetland[['non-truncated']][i,'Light2'] <- scal.2() 
-            results.wetland[['original']][i,'Light2'] <- val
-          }
-
-          # Moisture
-          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Moisture')]
-          results.wetland[['original']][i,'richness'] <- nrow(dat)
-          dat <- dat[!is.na(dat$Moisture),]
-
-          if ( nrow(dat)>0 ) {
-
-            val <- sum(dat[,'art_dekning'] * dat[,'Moisture'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
-            # lower part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Moist1'] <- scal() 
-            results.wetland[['non-truncated']][i,'Moist1'] <- scal.2() 
-            results.wetland[['original']][i,'Moist1'] <- val
-
-            # upper part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Moist2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Moist2'] <- scal() 
-            results.wetland[['non-truncated']][i,'Moist2'] <- scal.2() 
-            results.wetland[['original']][i,'Moist2'] <- val
-          }
-
-          # Soil_reaction_pH
-          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Soil_reaction_pH')]
-          results.wetland[['original']][i,'richness'] <- nrow(dat)
-          dat <- dat[!is.na(dat$Soil_reaction_pH),]
-
-          if ( nrow(dat)>0 ) {
-
-            val <- sum(dat[,'art_dekning'] * dat[,'Soil_reaction_pH'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
-            # lower part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'pH1'] <- scal() 
-            results.wetland[['non-truncated']][i,'pH1'] <- scal.2() 
-            results.wetland[['original']][i,'pH1'] <- val
-
-            # upper part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='pH2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'pH2'] <- scal() 
-            results.wetland[['non-truncated']][i,'pH2'] <- scal.2() 
-            results.wetland[['original']][i,'pH2'] <- val
-          }
-
-          # Nitrogen
-          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.wetland$GlobalID[i]),c('art_dekning','Nitrogen')]
-          results.wetland[['original']][i,'richness'] <- nrow(dat)
-          dat <- dat[!is.na(dat$Nitrogen),]
-
-          if ( nrow(dat)>0 ) {
-
-            val <- sum(dat[,'art_dekning'] * dat[,'Nitrogen'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
-            # lower part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen1' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Nitrogen1'] <- scal() 
-            results.wetland[['non-truncated']][i,'Nitrogen1'] <- scal.2() 
-            results.wetland[['original']][i,'Nitrogen1'] <- val
-
-            # upper part of distribution
-            ref <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
-            lim <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
-            maxmin <- wetland.ref.cov.val[wetland.ref.cov.val$Ind=='Nitrogen2' & wetland.ref.cov.val$grunn==as.character(results.wetland[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
-            # coercing x into results.wetland dataframe
-            results.wetland[['scaled']][i,'Nitrogen2'] <- scal() 
-            results.wetland[['non-truncated']][i,'Nitrogen2'] <- scal.2() 
-            results.wetland[['original']][i,'Nitrogen2'] <- val
-          }
-
-        }
-      }
-
-  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
-}
-
-# for using both sides of the plant indicators
-results.wetland[['2-sided']] <- results.wetland[['non-truncated']]
-
-# remove values >1 for 2-sided indicators
-results.wetland[['2-sided']]$Light1[results.wetland[['2-sided']]$Light1>1] <- NA
-results.wetland[['2-sided']]$Light2[results.wetland[['2-sided']]$Light2>1] <- NA
-
-results.wetland[['2-sided']]$Moist1[results.wetland[['2-sided']]$Moist1>1] <- NA
-results.wetland[['2-sided']]$Moist2[results.wetland[['2-sided']]$Moist2>1] <- NA
-
-results.wetland[['2-sided']]$pH1[results.wetland[['2-sided']]$pH1>1] <- NA
-results.wetland[['2-sided']]$pH2[results.wetland[['2-sided']]$pH2>1] <- NA
-
-results.wetland[['2-sided']]$Nitrogen1[results.wetland[['2-sided']]$Nitrogen1>1] <- NA
-results.wetland[['2-sided']]$Nitrogen2[results.wetland[['2-sided']]$Nitrogen2>1] <- NA
-
-```
-
-```{r, include = F}
-# Use cache data
-results.wetland<-readRDS(paste0(here::here(),"/data/cache/results.wetland.RDS"))
-
-#load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/results.wetland.RData")
-```
-
-```{r}
-head(results.wetland[['2-sided']])
-```
-#### Scaled value analyses
-
-In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).
-
-```{r, message = FALSE}
-#### plotting scaled values by main ecosystem type ####
-## continuing with 2-sided
-res.wetland <- results.wetland[['2-sided']]
-
-# make long version of the scaled value part
-res.wetland <-
-  res.wetland %>% 
-  pivot_longer(
-    cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
-    names_to = "fp_ind",
-    values_to = "scaled_value",
-    values_drop_na = FALSE
-  )
-
-# add original values as well
-res.wetland <- 
-  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
-                               pivot_longer(
-                                 cols =  c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
-                                 names_to = NULL,
-                                 values_to = "original",
-                                 values_drop_na = FALSE
-                               ) %>%
-                               pull(original)
-  )
-
-```
-
-### Ecosystem sub-types
-And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
-```{r,echo=FALSE,message=FALSE, warning=FALSE}
-
-# making the plot
-ggplot(res.wetland[!is.na(res.wetland$scaled_value),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
-  geom_hline(yintercept=0.6, linetype="dashed") + 
-  geom_violin() +
-#  geom_boxplot(width=0.2, color="grey") +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
-  xlab("Main ecosystem type") + 
-  ylab("Scaled indicator value") 
-```
-The functional signature from the plant communities in wetland ecosystems looks largely good, except for V2 (swamp forests), which show rather low index scores for the lower pH and Nitrogen indicators, i.e.
-suggesting too acidic conditions and too low nitrogen availability.
-
-We can look at V2 specifically to see if this is the case for all sub-types.
-
-```{r}
-# making the plot for V2 only
-ggplot(res.wetland[res.wetland$hovedtype_rute=="V2",], aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
-  geom_hline(yintercept=0.6, linetype="dashed") + 
-  geom_violin() +
-  #  geom_boxplot(width=0.2, color="grey") +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
-  xlab("Main ecosystem type") + 
-  ylab("Scaled indicator value") 
-```
-The deviations towards low pH and low nitrogen are mainly an issue for V2-C1, limestone-poor swamp forests.
-
-### Indicator index maps
-We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
-```{r, echo=FALSE, warning=FALSE,message = FALSE}
-#### scaled value maps ####
-# keep wide format and add geometry again
-res.wetland2 <- results.wetland[['2-sided']]
-st_geometry(res.wetland2) <- st_geometry(ANO.wetland)
-
-#nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
-#  st_as_sf() %>%
-#  st_transform(crs = crs(ANO.geo))
-
-nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
-  st_as_sf() %>%
-  st_transform(crs = crs(ANO.geo))
-
-#reg <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/regioner/regNorway_wgs84 - MERGED.shp")%>%
-#  st_as_sf() %>%
-#  st_transform(crs = crs(ANO.geo))
-
-reg <- st_read("data/regions.shp")%>%
-  st_as_sf() %>%
-  st_transform(crs = crs(ANO.geo))
-
-# change region names to something R-friendly
-reg$region
-reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
-
-regnor <- st_intersection(reg,nor)
-
-## scaled value maps
-# pH1 (lower indicator)
-tm_shape(regnor) +
-  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
-  tm_borders() +
-  tm_shape(res.wetland2) +
-  tm_dots('pH1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=1, legend.show = FALSE) + # 
-  tm_layout(main.title = "pH (lower), wetland",legend.position = c("right", "bottom"), main.title.size=1.2) + 
-  tm_add_legend(type = "fill", 
-                col = c(tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),'grey'),
-                labels = c("0.0 - 0.2", "0.2 - 0.4", "0.4 - 0.6", 
-                           "0.6 - 0.8", "0.8 - 1.0", "NA"),
-                title = "index values")
-
-```
-...but here the colors and values of the data points on the map are hard to make out.
-
-### Regions - maps and statistics
-Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because:
-(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
-(ii) the ANO dataset has a nested structure
-
-Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
-Here, we apply the following function using either a glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
-```{r}
-library(betareg)
-library(glmmTMB)
-
-expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale
-
-# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
-# if that is not the case, then the function performs a betareg if theres >= 2 observations
-# if that is not the case either, then the function returns the value of the single observation with NA standard error
-
-indmean.beta <- function(df) {
-
-  st_geometry(df) <- NULL
-  colnames(df) <- c("y","ran")
-
-  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {
-
-    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {
-
-      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)
-
-      return(c(
-        expit(summary( mod1 )$coefficients$cond[1]),
-
-        expit( summary( mod1 )$coefficients$cond[1] + 
-                 summary( mod1 )$coefficients$cond[2] )-
-          expit( summary( mod1 )$coefficients$cond[1] ),
-
-        nrow(df[!is.na(df$y),]),
-        summary( mod1 )$coefficients$cond[1],
-        summary( mod1 )$coefficients$cond[2]
-      ))
-
-    } else {
-
-      mod2 <- betareg(y ~ 1, data=df)
-
-      return(c(
-        expit(summary( mod2 )$coefficients$mean[1]),
-        expit( summary( mod2 )$coefficients$mean[1] + 
-                 summary( mod2 )$coefficients$mean[2] )-
-          expit( summary( mod2 )$coefficients$mean[1] ),
-        nrow(df[!is.na(df$y),]),
-        summary( mod2 )$coefficients$mean[1],
-        summary( mod2 )$coefficients$mean[2]
-      ))
-
-    }
-
-  } else {
-
-    return(c(df$y,NA,1,NA,NA))
-
-  }
-
-}
-
-```
-
-```{r}
-
-res.wetland2 = st_join(res.wetland2, regnor, left = TRUE)
-
-regnor <- regnor %>%
-  mutate(
-    pH1.reg.mean = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[1]
-    ),
-    pH1.reg.se = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[2]
-    ),
-    pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),])
-    )
-  )
-
-## scaled value maps for pH1 (lower indicator)
-# mean
-tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
-  tm_text("pH1.reg.n",col="black",bg.color="grey")
-```
-Mean index value by region for the lower pH indicator (i.e. index shows deviations towards a more acidic environment). Numbers in grey fields show the number of observations in the respective region.
-
-```{r}
-# se
-tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.se", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
-  tm_text("pH1.reg.n",col="black",bg.color="grey")
-
-```
-Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
-
-### unscaled values vs. reference
-We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
-Since pH and nitrogen show some deviation of the lower pH indicator we exemplify this with these indicators for unscaled values.
-
-```{r}
-summary(res.wetland$kartleggingsenhet_1m2)
-length(unique(res.wetland$kartleggingsenhet_1m2))
-# 16 NiN-types to plot
-colnames(wetland.ref.cov[['Soil_reaction_pH']])
-
-### pH
-par(mfrow=c(2,4))
-## V1s
-# V1-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
-       col="red")
-
-# V1-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
-       col="red")
-
-# V1-C-3
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C3',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
-       col="red")
-
-# V1-C-4
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C4',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
-       col="red")
-
-# V1-C-5
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C5")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C5',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
-       col="red")
-
-# V1-C-6
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C6',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
-       col="red")
-
-# V1-C-7
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C7',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
-       col="red")
-
-# V1-C-8
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C8',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
-       col="red")
-
-par(mfrow=c(2,4))
-## V2s
-# V2-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
-       col="red")
-
-# V2-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
-       col="red")
-
-# V2-C-3
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C3',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
-       col="red")
-
-## V3s
-# V3-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
-       col="red")
-
-# V3-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
-       col="red")
-
-## V4s
-# V4-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V4-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V4-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
-       col="red")
-
-## V8s
-# V8-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C1")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C1',xlab='pH value')
-#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
-       col="red")
-
-# V8-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C2',xlab='pH value')
-#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
-       col="red")
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-### Nitrogen
-par(mfrow=c(2,4))
-## V1s
-# V1-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
-       col="red")
-
-# V1-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
-       col="red")
-
-# V1-C-3
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C3',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
-       col="red")
-
-# V1-C-4
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C4',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
-       col="red")
-
-# V1-C-5
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C5")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C5',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
-       col="red")
-
-# V1-C-6
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C6',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
-       col="red")
-
-# V1-C-7
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C7',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
-       col="red")
-
-# V1-C-8
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C8',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
-       col="red")
-
-par(mfrow=c(2,4))
-## V2s
-# V2-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
-       col="red")
-
-# V2-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
-       col="red")
-
-# V2-C-3
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C3',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
-       col="red")
-
-## V3s
-# V3-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
-       col="red")
-
-# V3-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
-       col="red")
-
-## V4s
-# V4-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V4-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V4-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
-       col="red")
-
-## V8s
-# V8-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C1")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C1',xlab='Nitrogen value')
-#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
-       col="red")
-
-# V8-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C2',xlab='Nitrogen value')
-#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
-       col="red")
-
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-```
-THe figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests.
-According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
+allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat,levels=c("Sentinel","Modis","Landsat"))
+levels(allSatNDVI.wetland$Sat)
+```
+Now we can plot the time series for each main ecosystem type, and for each satellite separately  
+```{r}
+# plot 
+allSatNDVI.wetland %>%
+  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
+  geom_point() +
+  facet_wrap(~hovedtype, ncol = 1)
+```
+It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable.
+They vary both in their placement along the y-axis and in their variance.
+Modis data are absent from the V4-ecosystem type because of a too large pixel size compared to the NiN-polygon sizes.
+In V10, Modis is poorly represented for the same reason.
+
+As a final step, we model NDVI as a function of year and main ecosystem type separately for each Satellite
+```{r}
+model.wetland.time.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland)
+model.wetland.time.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland)
+model.wetland.time.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland)
+
+summary(model.wetland.time.Sent)$coefficients$cond
+summary(model.wetland.time.Modi)$coefficients$cond
+summary(model.wetland.time.Land)$coefficients$cond
+```
 

### 2023-07-04 DRAFT_insectIndicators.Rmd
added skeleton of NBBM indicators

@@
 #Notes on this draft
-This is a draft. The plan to finish it up i the fall of 2023.
+Most things are done but it's still a rough draft. I plan to finish it in august. But you could see where it's headed so raise the alarm if I'm off course.
 
-Incomprehensive list of things that will change:
+Incomplete list of things that will change:
 
-* Fill in missing sections.
-* Choices affecting the individual indicators, such as what taxa to select for each indicator'
-* Upgrade of genetic diversity calculation. This is just a placeholder, awaiting some new data.
-* Bumblebee and butterfly indicators will be added. The routines are done but should be streamlined.
+* Fill in sections that lack text.
+* Upgrade genetic diversity calculation. The current calculation is just a placeholder, awaiting some new data from Marie.
 * Packaging of R-package versions for reproducibility? Or just point to a version on Github..
-* Make bombLepiSurv public.
+* Make bombLepiSurv public. After touch-up and fixes of github issues.
 * Fix any comments in this draft inside hard brackets.
@@
 
+[Add short motivation for each indicator]
+
 ## About the underlying data
@@
 <!-- Text here -->
-
-### <!-- Text here -->
-<!-- Text here. Note that there are two parallell systems for defining ecosystem properties - one Norwegian and one international (SEEA EA). See man/SEEA EA examples.pdf for help -->
 
@@
 
-## Bootstrap observations
+#### Bootstrap observations
 The localities in the NorIns program are (semi-)randomly selected and can be viewed as samples of a larger population.
@@
 
-## Comparing bootstrap values to a reference point
+#### Comparing bootstrap values to a reference point
 Setting aside the difficulties in establishing robust reference values, the next step is to compare the observed values (with bootstrapped uncertainty) to the chosen reference value.
@@
 
-##Normalizing the values
+#### Normalizing the values
 After the comparison to a reference value, we need to normalize the indicator values to lie between 0 and 1.
@@
 
-## Display and plot bootstrap values
+#### Display and plot bootstrap values
 The boot_stat class also has its own plot function.
@@
 
-## Map plots
+#### Map plots
 In addition to the density plots above, we can also display the values geographically, in a map.
@@
 
+### Calculation principles for NBBM indicators
+These indicators were developed for the Nature Index of Norway, back in 2010 by Ola Diserud and Sandra Åström [ref], and is calculated routinely on a yearly basis [see xx for the most current report].
+
+These indicators summarize the state of a bumblebee or butterfly community by estimating the difference from a reference community.
+Formally, the community indicator (CI) is expressed as the relative change of the community from a state of reference (SR), were the change is calculated by a state of change (ST).
+
+$$CI = \frac{SR-SC}{SR}$$
+The state of reference (SR) represents a community of species that can be expected to be observed in a given habitat type and region.
+It is calculated by assigning each species to class of expected commonality; common (C), sporadic (S) and rare (R) species.
+This classification is done by expert opinion, informed by known present and past species distributions.
+This state of reference doesn't contain all current possible species, only a subset of species, making up a historical reference community.
+Potential observations of "new" species therefore doesn't inform the indicator value at all.
+Common species is expected to be observed in at least 5 % of the surveyed transects in a habitat type and region.
+Sporadic species similarly is defined as having a presence below 5 % but above 1 %.
+Rare species is seen in no more than 1 % of the transects, and lastly, species not seen in any transect is assumed as lost (L) for the purpose of this calculation.
+
+Each commonality class gets it's own weight, so that common species inform the reference state more than sporadic species, followed by rare species.
+The state of reference (SR) is thus a value, where each species adds to the state according to its commonality.
+Formally, it is defined as:
+
+$$SR = n_C * w_{C, SR} + n_S * w_{S, SR} + n_R * w_{R, SR} = \sum_{i = (C, S, R)} n_i * w_{i, SR}, $$
+where $n_i$ is the number of species in a commonality class (C, S, R), and the weights $[w_{C, SR}, w_{S, SR}, w_{R, SR}]$ specifies their respective contribution to the reference state (SR).
+The weights used is $[w_{C, SR}, w_{S, SR}, w_{R, SR}] = [1.0, 0.75, 0.50]$, i.e.
+a sporadic species has 75 % the weight of a common species, and a rare species has 50 % the weight of a common species.
+
+The state of change (SC) is calculated as
+ $$SC = n_{CS} * w_{CS} + n_{CR} * w_{CR} + n_{CL} * w_{CL} + n_{SR} * w_{SR} + n_{SL} * w_{SL} + n_{RL} * w_{RL}, $$ 
+where $n_{CS}$ is the number of expected common species (C) that is observed sporadically (S), and $w_{CS}$ is the weight of this change.
+Similarly, $n_{CR}$ is the number of expected species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on.
+Potential increases in observation rates does not inform the indicator.
+Changes for common species is weighted more heavily than changes for less common species, and larger decreases is weighted more heavily than smalle decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.
+
+The indicator values are estimated with uncertainty, by bootstrapping the set of localities within a region, and calculating the indicator for each of 999 such samples.
+
+#### Calculation example
+These calculations with bootstrap sampling is implemented in the R-package `bombLepiSurv` (Bombus and Lepidoptera Survey).
+We briefly show this functionality with bumblebees in semi-natural lands as an example.
+
+We connect to the internal database with a convenience function `humlesommerfConnect`
+```{r}
+bombLepiSurv::humlesommerfConnect()
+```
+
+We can fetch the all observations of bumblebees in semi-natural land with the `getAllData` function. Species names is here shown in Norwegian.
+
+```{r, fig.cap = "Bumblebees in grasslands 2021"}
+allBombusGrassland2022 <- getAllData(type = "bumblebees",
+                                     habitat = "gressmark",
+                                     year = 2022,
+                                     language = "norsk")
+
+```
+
+A summary of the records is most easily plotted by `surveyBarPlot`
+
+```{r}
+surveyBarplot(allBombusGrassland2022)
+```
+
+But for calculating the indicators, we only consider the transects that had were surveyed for the standard 3 times per year, to cover the phenology of the entire season.
+Complete data sets can be fetched by the function `getComplData`, which fetches data from one habitat type and region at a time.
+
+```{r}
+bombus_trond_2022 <- getComplData(type = "Humler",
+                                  region_short = "Trond",
+                                  habitat = "Gressmark",
+                                  year = 2022)
+```
+
+The observed relative to the expected occurrences can be visualized through the `plotArt` function. This requires a reference community, which can be fetched through the `getExpValues` function.
+
+```{r}
+exp_bombus_trond <- getExpValues(type = "Humler",
+                                 region_short = "Trond",
+                                 habitat = "Gressmark"
+                                 ) %>% 
+  select(species = species_latin,
+         everything())
+```
+
+Species in green bars are species that are expected to be common, and should reach the green areas in the plot.
+Similarly with the blue (sporadic species), and red (rare species).
+We see in this case that Bombus soroeensis is an exptected sporadic species that only were observed rarely.
+
+```{r}
+plotArt(bombus_trond_2022,
+        exp_bombus_trond)
+```
+[add all species in exp to plot]
+
+The actual calculation and bootstrapping is done in the function `calcInd`.
+This function calls the `getComplData` and `getExpectedValues` functions internally.
+It also requires the weights for the commonality classes in the reference community (SR), and the weights for the changes in state (CS).
+This is fetched internally via the functions `getAmountWeights` and `getDiffWeights`, respectively.
+Here the classes is in norwegian (v = common, m = sporadic, s = rare).
+
+```{r}
+getDiffWeights()
+
+```
+```{r}
+getAmountWeights()
+```
+Lastly, we specify the number of samples for the bootstrap
+
+```{r}
+nIter = 999
+```
+
+```{r}
+hInd2022TrondGress <- calcInd(type = "Humler",
+                              region_short = "Trond",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+The result [is going to be] an object of class [xxx]. Add print and plot methods...
+
+Plotting functions for a series of indicator values is shown below in the indicator calculations.
+
 ### Data sets
@@
 
-```{r}
-require(bombLepiSurv)
-bombLepiSurv::humlesommerfConnect(connectionName = "bombLepiCon")
-bombus_obs <- getAllData(dataConnection = "bombLepiCon",
-                         type = "bumblebees",
-                         year = 2009:2022)
-lepi_obs <- getAllData(dataConnection = "bombLepiCon",
-                         type = "butterflies",
-                         year = 2009:2022)
-
-```
-
 #### Regions
@@
 ```{r}
+connect_to_insect_db()
 norway_regions <- Norimon::get_map()
@@
 
-### Indicator values for NorIns indicators
+### Calculation of NorIns indicators
 #### Biomass of flying insects in semi-natural land
@@
 
+### Calculation of NBBM indicators
+
+### Bumblebees in open landscapes
+```{r}
+require(bombLepiSurv)
+humlesommerfConnect(host = "ninradardata01.nina.no")
+
+```
+
+```{r bombusGressBarplot, fig.cap = "Bumblebees in grasslands 2021"}
+allBombusGressmark2022 <- getAllData(type = "bumblebees",
+                        habitat = "gressmark",
+                        year = 2022,
+                        language = "norsk")
+
+surveyBarplot(allBombusGressmark2022) #+
+#  ggtitle("Humler i gressmark 2021")
+
+```
+
+```{r}
+nIter = 999
+
+hInd2022TrondGress <- calcInd(type = "Humler",
+                              region_short = "Trond",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+```{r}
+hInd2022SorGress <- calcInd(type = "Humler",
+                              region_short = "Sor",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+```{r}
+hInd2022OstGress <- calcInd(type = "Humler",
+                              region_short = "Ost",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+[Implement this]
+```{r, eval = F}
+hInd2022VestGress <- calcInd(type = "Humler",
+                              region_short = "Vest",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+### Butterflies in open landscapes
+
+```{r}
+nIter = 999
+
+sInd2022TrondGress <- calcInd(type = "Sommerfugl",
+                              region_short = "Trond",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+```{r}
+sInd2022SorGress <- calcInd(type = "Sommerfugl",
+                              region_short = "Sor",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+```{r}
+sInd2022OstGress <- calcInd(type = "Sommerfugl",
+                              region_short = "Ost",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+[Implement this]
+```{r, eval = F}
+sInd2022VestGress <- calcInd(type = "Sommerfugl",
+                              region_short = "Vest",
+                              habitat = "Gressmark",
+                              year = 2022,
+                              nIter = nIter,
+                              save.draws = T)
+```
+
+The calculations for earlier years is already present as datasets in the R-package and can be fetched individually.
+Once this is done, we can plot the timeseries of values using the function `indPlot`.
+
+Todo, put these in the new version of the package.
+```{r}
+#load("out/IndicatorValues2020.Rdata")
+```
+
+```{r}
+
+data("sInd2019TrondGress")
+data("sInd2019TrondSkog")
+data("sInd2019OstGress")
+data("sInd2019OstSkog")
+data("sInd2019SorGress")
+data("sInd2019SorSkog")
+data("sInd2018TrondGress")
+data("sInd2018TrondSkog")
+data("sInd2018OstGress")
+data("sInd2018OstSkog")
+data("sInd2018SorGress")
+data("sInd2018SorSkog")
+data("sInd2017TrondGress")
+data("sInd2017TrondSkog")
+data("sInd2017OstGress")
+data("sInd2017OstSkog")
+data("sInd2017SorGress")
+data("sInd2017SorSkog")
+data("sInd2016TrondGress")
+data("sInd2016TrondSkog")
+data("sInd2016OstGress")
+data("sInd2016OstSkog")
+data("sInd2016SorGress")
+data("sInd2016SorSkog")
+data("sInd2015TrondGress")
+data("sInd2015TrondSkog")
+data("sInd2015OstGress")
+data("sInd2015OstSkog")
+data("sInd2015SorGress")
+data("sInd2015SorSkog")
+data("sInd2014TrondGress")
+data("sInd2014TrondSkog")
+data("sInd2014OstGress")
+data("sInd2014OstSkog")
+data("sInd2014SorGress")
+data("sInd2014SorSkog")
+data("sInd2013TrondGress")
+data("sInd2013TrondSkog")
+data("sInd2013OstGress")
+data("sInd2013OstSkog")
+data("sInd2013SorGress")
+data("sInd2013SorSkog")
+data("sInd2012TrondGress")
+data("sInd2012TrondSkog")
+data("sInd2012OstGress")
+data("sInd2012OstSkog")
+data("sInd2011OstGress")
+data("sInd2011OstSkog")
+data("sInd2010OstGress")
+data("sInd2010OstSkog")
+data("sInd2009OstGress")
+data("sInd2009OstSkog")
+data("hInd2019TrondGress")
+data("hInd2019TrondSkog")
+data("hInd2019OstGress")
+data("hInd2019OstSkog")
+data("hInd2019SorGress")
+data("hInd2019SorSkog")
+data("hInd2018TrondGress")
+data("hInd2018TrondSkog")
+data("hInd2018OstGress")
+data("hInd2018OstSkog")
+data("hInd2018SorGress")
+data("hInd2018SorSkog")
+data("hInd2017TrondGress")
+data("hInd2017TrondSkog")
+data("hInd2017OstGress")
+data("hInd2017OstSkog")
+data("hInd2017SorGress")
+data("hInd2017SorSkog")
+data("hInd2016TrondGress")
+data("hInd2016TrondSkog")
+data("hInd2016OstGress")
+data("hInd2016OstSkog")
+data("hInd2016SorGress")
+data("hInd2016SorSkog")
+data("hInd2015TrondGress")
+data("hInd2015TrondSkog")
+data("hInd2015OstGress")
+data("hInd2015OstSkog")
+data("hInd2015SorGress")
+data("hInd2015SorSkog")
+data("hInd2014TrondGress")
+data("hInd2014TrondSkog")
+data("hInd2014OstGress")
+data("hInd2014OstSkog")
+data("hInd2014SorGress")
+data("hInd2014SorSkog")
+data("hInd2013TrondGress")
+data("hInd2013TrondSkog")
+data("hInd2013OstGress")
+data("hInd2013OstSkog")
+data("hInd2013SorGress")
+data("hInd2013SorSkog")
+data("hInd2012TrondGress")
+data("hInd2012TrondSkog")
+data("hInd2012OstGress")
+data("hInd2012OstSkog")
+data("hInd2011OstGress")
+data("hInd2011OstSkog")
+data("hInd2010OstGress")
+data("hInd2010OstSkog")
+data("hInd2009OstGress")
+data("hInd2009OstSkog")
+
+```
+
+```{r bombusGressIndPlot}
+hGressIndVals <- collectIndData(taxa = "bumblebees", 
+                                habitat = "gressmark",
+                                years = 2009:2022)
+
+indPlot(input = hGressIndVals,
+        confLevel = "95%",
+        lineWidth = 1,
+        dodge = 0.5,
+        pointSize = 4) +
+  ggtitle("Indicator values for bumblebees in semi-natural landscapes") +
+  scale_x_continuous(breaks = 2009:2022)
+
+```
+
+```{r lepiGressIndPlot}
+sGressIndVals <- collectIndData(taxa = "butterflies", 
+                                habitat = "gressmark",
+                                years = 2009:2022)
+
+indPlot(input = sGressIndVals,
+        confLevel = "95%",
+        lineWidth = 1,
+        dodge = 0.5,
+        pointSize = 4) +
+  ggtitle("Indicator values for butterflies in semi-natural landscapes") +
+  scale_x_continuous(breaks = 2009:2022)
+
+```
+
 ## Prepare export

### 2023-07-05 DRAFT_ndvi_wetland.Rmd
More changes in text

@@
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
+df.m<- read.csv("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv", )
 df.m$mean<-df.m$mean/10000
@@
 
-# there's no information on the proportion of ecosystem types in the polygons, so we got to omit all polygons with multiple ecosystem types :(
+# there's no information on the proportion of ecosystem types in the polygons, so we have to omit all polygons with multiple ecosystem types :(
 nin <- nin %>%
@@
   facet_wrap( ~hovedtype)
-# NDVI largely varies between hovedtyper: higher in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
-# @Tilstand: in seminat types, kaldkilde, and bogs ndvi seems to increase as condition deteriorates, in fens not much of a pattern
+# NDVI largely varies between hovedtyper: NDVI is higher in semi-natural wet-meadows (V10), and lowest in ombrotrophic bogs (V3)
+# @Tilstand: in semi-natural types, NDVI in kaldkilde, and bogs seems to increase as condition deteriorates, in fens there is not much of a pattern
 
@@
 
-## looking at ndvi in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
+## looking at NDVI in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
 SentinelNDVI.wetland %>%
@@
   facet_grid( region~hovedtype)
-# V4 has lacking data for quite some condition classes in most regions
+# V4 kaldkilde has lacking data for quite some condition classes in most regions
 
@@
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
-# we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
+# we drop V4 Kaldkilde for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
 
@@
 
-It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
+It is evident that there are differences between regions.
+But since we are not interested in the statistics for regional differences, we run separate models for every region for ease of interpretation.
 ```{r, include = F}
@@
 
-V1: NDVI decreases as condition deteriorates
-V3: NDVI does not change as condition deteriorates
-V9 & V10: NDVI increases as condition deteriorates
+V1 Åpen jordvannsmyr: NDVI decreases as condition deteriorates
+V3 Nedbørsmyr: NDVI does not change as condition deteriorates
+V9 Semi-naturlig myr & V10 Semi-naturlig våteng: NDVI increases as condition deteriorates
 
@@
 
-V1: NDVI decreases as condition deteriorates
-V3, V9 & V10: NDVI increases as condition deteriorates
+V1 Åpen jordvannsmyr: NDVI decreases as condition deteriorates
+V3 Nedbørsmyr, V9 Semi-naturlig myr & V10 Semi-naturlig våteng: NDVI increases as condition deteriorates
 
@@
 
-V1 & V10: NDVI increases as condition deteriorates
-V3: NDVI decreases as condition deteriorates (but really insufficient data)
-V9: NDVI does not change as condition deteriorates
+V1 Åpen jordvannsmyr & V10 Semi-naturlig våteng: NDVI increases as condition deteriorates
+V3 Nedbørsmyr: NDVI decreases as condition deteriorates (but really insufficient data)
+V9 Semi-naturlig myr: NDVI does not change as condition deteriorates
 
@@
 
-V1 & V3: NDVI does not change as condition deteriorates
-V9 & V10: NDVI increases as condition deteriorates
+V1 Åpen jordvannsmyr & V3 Nedbørsmyr: NDVI does not change as condition deteriorates
+V9 Semi-naturlig myr & V10 Semi-naturlig våteng: NDVI increases as condition deteriorates
 
@@
 
-V1, V9 & V10: NDVI increases as condition deteriorates
-V3: NDVI does not change as condition deteriorates (but also insuffcient data)
+V1 Åpen jordvannsmyr, V9 Semi-naturlig myr & V10 Semi-nautrlig våteng: NDVI increases as condition deteriorates
+V3 Nedbørsmyr: NDVI does not change as condition deteriorates (but also insuffcient data)
 
 Overall take home messages: 
+NDVI in semi-natural wetlands (V9) and semi-natural wet meadows (V10) increases, or is unaffected by reduced condition in all geographical regions.
+This is consistent with one of the main drivers of reduced condition in semi-natural landscapes in general, namely overgrowth/ encroachment by shrubs as a result of a reduction in traditional management practices.
 
-???
+In minerotrophic mires (V1), the relationship between NDVI and condition differs in the northern versus the southern regions of Norway.
+In both northern and central Norway NDVI decreases as condition decreases, while NDVI increases with lower condition in western, eastern and southern Norway.
+This may indicate different drivers of poor condition along a latitudinal gradient.
+
+There is no consistent pattern in NDVI and condition in ombrotrophic mires (V3).
+Neither northern, eastern or southern Norway's ombrotrophic mires exhibit changes in NDVI as condition decreases.
+In central Norway NDVI increases as condition decreases, while NDVI decreases as condition decreases in western Norway.
+However, it should be noted that in the latter case the data available is very limited.
 
 ### Regression analyses, NDVI across time
-First, there's again some data handling to do. We merge the Sentinel, Modis, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
+First, there's some data handling to do. We merge the Sentinel, Modis, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
 ```{r}

### 2023-07-05 DRAFT_insectIndicators.Rmd
added Vest region, and 2020:2022 data to NBBM

@@
 ```{r}
+data("sInd2022VestGress")
+data("sInd2022VestSkog")
+data("sInd2022TrondGress")
+data("sInd2022TrondSkog")
+data("sInd2022OstGress")
+data("sInd2022OstSkog")
+data("sInd2022SorGress")
+data("sInd2022SorSkog")
+
+data("sInd2021TrondGress")
+data("sInd2021TrondSkog")
+data("sInd2021OstGress")
+data("sInd2021OstSkog")
+data("sInd2021SorGress")
+data("sInd2021SorSkog")
+
+data("sInd2020TrondGress")
+data("sInd2020TrondSkog")
+data("sInd2020OstGress")
+data("sInd2020OstSkog")
+data("sInd2020SorGress")
+data("sInd2020SorSkog")
 
@@
 data("sInd2009OstSkog")
+
+data("hInd2022VestGress")
+data("hInd2022VestSkog")
+data("hInd2022TrondGress")
+data("hInd2022TrondSkog")
+data("hInd2022OstGress")
+data("hInd2022OstSkog")
+data("hInd2022SorGress")
+data("hInd2022SorSkog")
+data("hInd2021TrondGress")
+data("hInd2021TrondSkog")
+data("hInd2021OstGress")
+data("hInd2021OstSkog")
+data("hInd2021SorGress")
+data("hInd2021SorSkog")
+data("hInd2020TrondGress")
+data("hInd2020TrondSkog")
+data("hInd2020OstGress")
+data("hInd2020OstSkog")
+data("hInd2020SorGress")
+data("hInd2020SorSkog")
 data("hInd2019TrondGress")

### 2023-08-02 DRAFT_ndvi_wetland.Rmd

@@
 
-# check if the three Sattelite objects have the same structure (for concatenating them)
+# check if the three Satellite objects have the same structure (for concatenating them)
 names(SentinelNDVI.wetland)
@@
 
-### Eksport file (final product)
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->
+### Export file (final product)
+<!-- Export final file. Ideally a georeferenced shape or raster with indicators values (raw and normalised), reference values and errors. -->

### 2023-08-10 DRAFT_gjengroing.Rmd
Add files via upload
added Zander's code for "gjengroing"

---
title: "Vegetation regrowth/encroachment (gjengroing)"
output:
  pdf_document: default
  html_document: default
---

<br />

_Author and date:_
Zander Venter

```{r}
Sys.Date()
```

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
library(sf)
library(tidyverse)
library(gridExtra)
library(RColorBrewer)

# Set global variable defining whether you want to run everything from scratch (very long runtime) or use pre-exported data (short runtime)
runFromScratch <- FALSE

knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "Våtmark og Semi-/Naturlig-åpne" 
Egenskap  <- "Gjengroing"
ECT       <- "Structural state characteristic" 
Contact   <- "Zander Venter"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
The Norwegian word "gjengroing" is directly translated to "regrowing" in English. The gjengroing indicator describes the regrowth of woody vegetation (trees and bushes) in open ecosystems (wetland, semi- and naturally open areas) across Norway. We will use a spatial reference approach where reference areas define good or optimal vegetation regrowth heights.

The workflow spans two platforms including RStudio and Google Earth Engine (GEE). To reproduce this workflow you will need a GEE account and access to the NINA RStudio and R/GeoSpatialData servers.

## About the underlying data
We rely on the following datasets:

* [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
* NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type. The population sample are the areas used to determine ecological condition relative to the reference areas. We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (ie. poor condition),
* LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/). This includes both a terrain (DTM) and surface mode (DSM). We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground. From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
* Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df). We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band. This is done in combination with bioclimatic zones - see next point.
* Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner). We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
* [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
* A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
* The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
* The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

### Representativity in time and space
The index will cover the mainland of Norway. The analysis will be stratified by (1) åpne and (2) våtmark ecosystems. The former includes both semi-naturlig mark and naturlig åpne områder under skoggrensa (due to the fact that AR5 does not differentiate these), and the latter includes våtmark only. The LiDAR data cover a range of years and therefore the indicator represents conditions for circa 2010 to 2021.

### Original units
The original units for ecological condition are meters. This is the height of the vegetation within referecne, polygon and forest areas.

### Temporal coverage
Circa 2010 to 2021. This is a single snapshot and not a change analysis. In the future, when LiDAR data has been repeated across the country, it may be possible to do a change assessment. In addition, the use of optical or radar satellite imagery may serve as a proxy for vegetation height using machine learning. This will allow for annual updates, depending on the uncertainty in the satellite-based maps of vegetation height.

### Aditional comments about the dataset
None.

## Ecosystem characteristic
### Norwegain standard
Not sure what this means??

## Collinearities with other indicators
There is possibly a collinearity with the primary production indicator (primærproduksjon). The primary production indicator uses the normalized difference vegetation index (NDVI) as a proxy for vegetation production. NDVI can be correlated with vegetation height and consequently yield similar results to the LiDAR-based gjengroing indicator.

## Reference state and values
The methodology used to calculate the gjengroing indicator is outlined in the schematic below. The workflow in the schematic is conducted for all reference and population polygons in Norway and repeated for each ecosystem type (åpne and våtmark), respectively. The indicator values are aggregated to a 50km grid and regional level at the end. The individual steps are discussed in turn in the following subsections.

```{r}
include_graphics('./DATA/images/gjengroing_schematic.jpg')
```

### Reference state
The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state. We use the all-encompassing "Tilstand" variable assigned to each NiN polygon. The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie. 1) of the scaled indicator value. We cannot define local reference values based on proximity, because the NiN polygons are spatially biased and not close to all AR5 population polygons. Therefore we calculate regional reference values using elevation (300m bands) and bioclimatic zones as stratification. We calculate the mean reference value for each unique combination of elevation-bioclimatic zone. When calculating the index for each population polygon, the reference value is inhereted from the elecation-bioclimatic zone it falls within.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
Once we have the reference vegetation height for a given ecosystem type and elevation-bioclimatic zone, we need to define the minimum (or worst/bad) condition. We use the 50th percentile of LiDAR-derived vegetation heights within AR5 skog polygons to define a climax vegetation successional stage where gjengroing is at its most extreme. In order not to include forest patches that have recently been harvested, we mask out any forest which has been clear-cut since 1986. Here, 1986 is a hard limit defined by the clear-cut dataset which was based on Landsat imagery. Therefore we can be assured that we are measuring forest that is at least 35 years old.

We use both local and regional "reference" approaches to defining poor ecological condition. To do this we calculate the forest heights in AR5 forest patches that fall within 200m of a population polygon (local reference). For population polygons that do not have forest within their proximity, we use the mean forest height within the unique elevation-bioblimatic zone the population falls within (regional reference).

To define the ecological condition of the population polygon, we measure 50th percentile of LiDAR-derived vegetation height. As a reminder, we use AR5 polygons that have types that are approximately compatible with the NiN ecosystem types. For våtmark we use the AR5 polygons defined as "Myr". For åpne ecosystems we use the AR5 polygons defined as either "Åpent fastmark" or "Innmarksbeite".

The vegetation height percentiles are then scaled to between 0 and 1 using a sigmoid transformation [Oliver at al. (2021)](https://www.sciencedirect.com/science/article/pii/S1470160X21000066).

## Uncertainties
For the indicator map at the polygon level there is no uncertainty associated with the indicator values. For aggregated indicator values (e.g. for 50km grid and regions), the uncertainty in the indicator value is calculated from the spatial variation in the polygon-level reference height values. We could use the alternative of calculating the spatial variation in the indicator values via bootstrapping using the EAtools package. However, given the extreme number of population polygons, the uncertainty values are extremely small and therefore we choose to use the standard deviation in reference vegetation heights, converted to the indicator scale.

## References
Links to data and resources are provided with hyperlinks in-line.

## Analyses
### Data sets
There are several datasets which are used in GEE which will not be imported into the R session here. These datasets have been obtained from the source and ingested into GEE by Zander Venter or Vegard Bakkestuen with the help of Miljædata section at NINA. They include

* [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6)
* [LiDAR-derived digital elevation model from høydedata](https://hoydedata.no/LaserInnsyn/).
* [Kartverket's national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df)
* [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) 
* [European forest clear-cut map](https://www.nature.com/articles/s41893-020-00609-y)

The remaining datasets will be imported into the R session.

#### Regions
The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.

```{r}
regions <- st_read('./DATA/regions.shp') %>%
  mutate(region = factor(region) )
# Rename regions with correct Norwegian characters
levels(regions$region) <- c("Østlandet", "Midt-Norge","Nord-Norge","Sørlandet","Vestlandet")
```

#### SSB 10 and 50km grids
The [SSB 10km and 50km grids](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) are used for visualizing the distribution of available data and for aggregating gjengroing index scores.

```{r}
ssb10km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_10km/ssb10km.shp')%>%
  # mutate the id value so that it aligns with the data coming from GEE
  mutate(SSBID = as.numeric(SSBID)/1000)
# Transform to the correct projection
ssb10km <- st_transform(ssb10km, st_crs(regions))

ssb50km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_50km/ssb50km.shp')%>%
  # mutate the id value so that it aligns with the data coming from GEE
  mutate(SSBID = as.numeric(SSBID)/1000)
# Transform to the correct projection
ssb50km <- st_transform(ssb50km, st_crs(regions))
```

#### Bioclimatic regions
The Moen's bioclimatic regions are imported from NINA R/GeoSpatialData/ server.

```{r}
bioclim <- st_read('/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp') %>%
  mutate(NAVN = ifelse(NAVN == 'S°rboreal', 'Sørboreal', NAVN))
bioclim <- st_transform(bioclim, st_crs(regions))
```

#### NiN polygons
[NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are imported here from the NINA servers:
```{r}
nin <- st_read('/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_NiN_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb')%>%
  # Simplify ecosystem names
  mutate(hovedøkosystem = recode(hovedøkosystem, 
                                 "Våtmark" = 'Våtmark',
                                 "Skog" = "Skog",
                                 "Ingen" = "Ingen",
                                 "Fjell" = "Fjell",
                                 "Semi-naturlig mark" = 'Semi-naturlig',
                                 "Naturlig åpne områder i lavlandet" = 'Naturlig åpne',
                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig åpne')) %>% 
  # Fix any invalid geometries
  mutate(validGeo = st_is_valid(SHAPE)) %>%
  filter(validGeo) %>%
  # Filter out the ecosystems we are not interested in
  filter(!hovedøkosystem %in% c('Skog','Ingen','Fjell')) %>%
  # Get the main ecosystem codes in case we need them later
  mutate(hovedtype = substr(ninkartleggingsenheter, 4, 6),
         hovedtype = str_remove(hovedtype, '-'),
         id = identifikasjon_lokalid) %>%
  # Drop polygons with no condition score
  drop_na(tilstand) %>%
  dplyr::select(id, hovedøkosystem, hovedtype, naturtypekode, tilstand)
```

* **Export the cleaned NiN data and upload to GEE before proceeding**

```{r}
# Only run if you are repeating this workflow for the first time
# nin %>%
#   dplyr::select(id, hovedøkosystem, tilstand) %>%
#   st_transform(st_crs(fylker_wgs84)) %>%
#   st_write('nin_cleaned.shp', append=FALSE)
```

* **Now run the following GEE JavaScript code by clicking on the hyperlink:** [GEE script](https://code.earthengine.google.com/b2d06ad72c515fbbb3024f157987eedc) This script is also found in the SCR repository as "GEE_script.js". When you run the script, then click the Tasks tab in the GEE code editor and run the export tasks. The export files will appear in your Google Drive. Download the files from your Google Drive to the /data/gjengroing/From_GEE directory. This script extracts three CSV files:
    + ‘area_cover_grid.csv’ - this gives the area coverage of NiN, AR5 and LiDAR data for each 10x10km SSB grid cell.
    + 'elevation_stratified.shp' - this is a 300m elevation band stratification that is simply used for visualizing the strata here in R.
    + ‘vegHeights.zip’ - this is a folder with hundreds of CSV files. These files contain the vegetation heights for reference, population, and forest polygons over Norway. The exports were split by 50km grid to prevent hitting user memory limits in GEE. Unzip the folder before you upload to ./DATA/From_GEE/

#### Data from Google Earth Engine
Once you have run the GEE script above, and downloaded the data to the /data/gjengroing/From_GEE directory, you can proceed with this R workflow.

Create functions to import vegetation height files and stratify data by elevation band and vegetation/climate zone

```{r}
# Function to read in multiple files resulting from GEE exports
readVegHeightFiles <- function(uniqueString){
  dir <- './DATA/From_GEE/vegHeights/'
  files <- list.files(dir)[str_detect(list.files(dir), uniqueString)]
  #print(files)

  dat <- tibble()
  for (i in files){
    dat <- dat %>% 
      bind_rows(read_csv(paste0(dir, i)) %>%
                  mutate(ssbid = substr(str_split(i, '_')[[1]][3], 1, 14)))
  }
  return (dat)
}

# Vegetation type lookup
vegLookup <- tibble(
  vegClimZone = c(1,2,3,4,5,6), 
  vegClimZoneLab = c('Nemoral','Boreonemoral','Sørboreal','Mellomboreal','Nordboreal','Alpin')
  )

# Elevation stratification and vegetation type cleaning function
cleanElevVegType <- function(data){

  dataOut <- data %>%
    mutate(e = elevation) %>%
    mutate(elevation = ifelse(e < 300, '0-300',
                              ifelse(e > 300 & e < 600, '300-600',
                                     ifelse(e > 600 & e < 900, '600-900',
                                     ifelse(e > 900 & e < 1200, '900-1200', '>1200'))))) %>%
    mutate(vegClimZone = round(vegClimZone)) %>%
    left_join(vegLookup, by = 'vegClimZone') %>%
    dplyr::select(-e, -vegClimZone) %>%
    drop_na(vegClimZoneLab, elevation)

  return (dataOut)

}

```

Import the CSV files generated in GEE related to area covers and elevation bands:

```{r}
# Import data coverages per SSB 10km grid cell
areaCovers <- read_csv('./DATA/From_GEE/area_cover_grid.csv') %>%
  dplyr::select(-'.geo', -'system:index')

# Calculate percentage cover relative to the land area in each cell
areaCoversPerc <- areaCovers %>%
  gather(key, val, aapne_ar5, vaatmark_ar5,aapne_nin, vaatmark_nin, lidarCover, skog) %>%
  mutate(type = ifelse(str_detect(key, 'ar5'), 'ar5', 
                       ifelse(str_detect(key, 'nin'), 'nin', 
                              ifelse(key == 'lidarCover', 'lidar', NA)))) %>%
  mutate(ecosystem = ifelse(str_detect(key, 'aapne'), 'Semi-/Naturlig åpne', 
                       ifelse(str_detect(key, 'vaatmark'), 'Våtmark', NA))) %>%
  mutate(areaPerc = val/land*100,
         areaPerc = ifelse(is.na(areaPerc), 0, areaPerc)) %>%
  dplyr::select(SSBID, key, type,ecosystem, areaPerc)

elevStrata <- st_read('./DATA/From_GEE/elevation_stratified.shp') %>%
  mutate(elevBand = ifelse(mode == 1, '0-300',
                           ifelse(mode == 2, '300-600',
                                  ifelse(mode == 3, '600-900',
                                         ifelse(mode == 4, '900-1200', '>1200')))),
         elevBand = factor(elevBand, levels=c('0-300','300-600','600-900','900-1200','>1200')))
elevStrata <- st_transform(elevStrata, st_crs(regions))

```

Import the CSV files generated in GEE of vegetation heights (circa 15 min runtime):

```{r}

if (runFromScratch){

  # Import median LiDAR heights for NiN polygons
  refVaatmarkRaw <- readVegHeightFiles('vaatmark_ref') %>%
    mutate(ref = chm) %>%
    dplyr::select(-'.geo', -'system:index', -chm)
  refAapneRaw <- readVegHeightFiles('aapne_ref') %>%
    mutate(ref = chm) %>%
    dplyr::select(-'.geo', -'system:index', -chm)

  # Add elevation and veg/clim zones and summarise per strata
    # this will form the upper limit of the ecological condition score (ie. good condition)
  refVaatmark <- cleanElevVegType(refVaatmarkRaw) %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(ref = mean(ref, na.rm=T))
  refAapne <- cleanElevVegType(refAapneRaw) %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(ref = mean(ref, na.rm=T))

  # Export to file for reference
  refVaatmark %>%
    st_write('./DATA/refVaatmark.csv')
  refAapne %>%
    st_write('./DATA/refAapne.csv')

  # Import median LiDAR heights for AR5 polygons for population measurement
  popVaatmarkHt <- readVegHeightFiles('vaatmark_pop')  %>%
    mutate(pop = chm)%>%
    dplyr::select(-chm)
  popAapneHt <- readVegHeightFiles('aapne_pop')  %>%
    mutate(pop = chm)%>%
    dplyr::select(-chm)

  # Turn into spatial objects using the .geo column in the CSV
  popVaatmarkHt <- st_as_sf(bind_cols(popVaatmarkHt,geojsonsf::geojson_sf(popVaatmarkHt$.geo)))
  popAapneHt <- st_as_sf(bind_cols(popAapneHt,geojsonsf::geojson_sf(popAapneHt$.geo)))

  # Import median LiDAR heights for forest (skog) within 200m of population polygons
    # this will form the lower limit of the ecological condition score (ie. poor condition)
  popVaatmarkSkog <- readVegHeightFiles('vaatmark_skog') %>%
    mutate(skog = chm)%>%
    dplyr::select(-chm, -'.geo', -ssbid)
  popAapneSkog <- readVegHeightFiles('aapne_skog') %>%
    mutate(skog = chm)%>%
    dplyr::select(-chm, -'.geo', -ssbid)

  # Combine the population heights with the forest heights
  popVaatmark <- popVaatmarkHt %>%
    left_join(popVaatmarkSkog, by = 'system:index')%>%
    dplyr::select(-'.geo')
  popAapne <- popAapneHt %>%
    left_join(popAapneSkog, by = 'system:index')%>%
    dplyr::select(-'.geo')

  # Add elevation and veg/clim zones to the combined population and forest heights
  popVaatmark <- cleanElevVegType(popVaatmark)
  popAapne <- cleanElevVegType(popAapne)

  # Get regional reference values for forest height
    # this is to fill in for population polygons which did not have any forest within 200m
  vaatmarkSkogRegionalRef <- popVaatmark %>%
    as_tibble() %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
    drop_na(skogFill)
  aapneSkogRegionalRef <- popAapne %>%
    as_tibble() %>%
    group_by(elevation, vegClimZoneLab) %>%
    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
    drop_na(skogFill)

  # Combine all the above into one data frame
  #set.seed(123)
  vaatmarkHts <- popVaatmark %>%
    #sample_n(100000) %>%
    as_tibble() %>% dplyr::select(-geometry) %>%
    left_join(refVaatmark, by=c('vegClimZoneLab', 'elevation')) %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
    summarise(height = mean(height)) %>%
    pivot_wider(values_from = height, names_from = type) %>%
    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled value will be 1
    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
    drop_na(pop, ref) %>%
    left_join(vaatmarkSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
    dplyr::select(-skogFill) %>%
    filter(skog < 30)
  #colSums(is.na(vaatmarkHts))

  aapneHts <- popAapne %>%
    #sample_n(100000) %>%
    as_tibble() %>% dplyr::select(-geometry) %>%
    left_join(refAapne, by=c('vegClimZoneLab', 'elevation')) %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
    summarise(height = mean(height)) %>%
    pivot_wider(values_from = height, names_from = type) %>%
    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled index will be 1
    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
    drop_na(pop, ref) %>%
    left_join(aapneSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
    dplyr::select(-skogFill) %>%
    filter(skog < 30)
  #colSums(is.na(aapneHts))

} else {

  refVaatmark  <- read_csv('./DATA/refVaatmark.csv')
  refAapne <- read_csv('./DATA/refAapne.csv')

}

```

### Explore data coverages

Here we will make maps of NiN, AR5 and LiDAR coverage over Norway to get a feel for the data distributions. These areas were calculated in GEE and exported to CSV in the GEE script provided above. All we need to do now is join them with the SSB 10km grid and visualize.

```{r fig.height=5, fig.width=15}

c1 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'nin') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('#e6bf6b', '#91b5a2'),
                       name = '% cover',
                       limits=c(0, 2),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('A) NiN polygons with good condition')
#c1
c2 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'ar5') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('#e6bf6b', '#91b5a2'),
                       name = '% cover',
                       limits=c(0, 100),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('B) AR5 polygons')
#c2
c3 <- ssb10km %>% 
  left_join(areaCoversPerc %>%
              filter(type == 'lidar') %>%
              group_by(SSBID) %>%
              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
  ggplot() +
  geom_sf(aes(fill=areaPerc), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.3) +
  scale_fill_gradientn(colours = c('#e6bf6b', '#91b5a2'),
                       name = '% cover',
                       limits=c(0, 100),
                       oob=scales::squish) +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('C) LiDAR canopy height model')
#c3

coverageFig <- grid.arrange(c1, c2, c3, ncol=3, widths=c(1,1,1), padding = unit(0, "line"), newpage = T)

```

### Bioclimatic zone and elevation strata for regional reference values

We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition. For each unique spatial combination of these strata (n = 21), we calcualte a mean reference value later on in the analysis. Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.

```{r}

bioclimPlot <- bioclim %>%
  ggplot() + 
  geom_sf(aes(fill = NAVN), color=NA)+
    scale_fill_manual(values=brewer.pal(10, 'Paired'), name = 'Zone')+
    ggtitle('A) Bioclimatic zones') +
    theme_void()

elevPlot <- elevStrata %>%
  ggplot() +
  geom_sf(aes(fill = elevBand), color=NA, size=0.001)+
    scale_fill_manual(values=brewer.pal(5, 'Pastel1'), name = 'HASL (M)')+
    ggtitle('B) Elevation bands')+
    theme_void()

strataPlot <- grid.arrange(bioclimPlot, elevPlot, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)

```

We can see that while AR5 polygons are well distributed across the country, the NiN polygons are clustered and biased toward low elevation regions of the country. The LiDAR data covers nearly the entire country with some exceptions in the north. Future iterations of this work will need to attempt to quantify the biases introduced by the relative distributions of reference and population polygons and by the datasets used to quantify gjengroing (in this case LiDAR).

### Scaled indicator values at polygon level (circa 120 min runtime)

We will calculate the gjengroing index for each population polygon over Norway using a Sigmoid scaling function.

```{r}

# Define Sigmoid scaling function
scaleSigmoid <- function(variable){
  refLow <- min(variable)
  refHigh <- max(variable)
  thr <- (refHigh-refLow)/2
  indicator_LowHigh <- (variable - refLow)/(refHigh - refLow)
  indicator_LowHigh[indicator_LowHigh[] < 0] <- 0
  indicator_LowHigh[indicator_LowHigh[] > 1] <- 1
  indicator_sigmoid <- 100.68*(1-exp(-5*(indicator_LowHigh)^2.5))/100
  return (round(indicator_sigmoid, 4))
}

if (runFromScratch){

  # Calcualte gjengroing index values
  vaatmarkIndexPoly <- vaatmarkHts  %>%
    # Gather into long format so taht we can mutate the scaleSigmoid() function
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`) %>%
    mutate(index = scaleSigmoid(height)) %>%
    # Because short vegetation is good condition, we must invert the scaled indicator
    mutate(index = 1-index) %>%
    # Get the indicator values for the population polygons only (ref and skog will be 1 and 0)
    filter(type == 'pop') %>%
    dplyr::select(-type, -height) %>%
    # Join with original data to get the actual median vegetaiton heights for each indicator score
    left_join(vaatmarkHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
    left_join(popVaatmark %>% dplyr::select("system:index", ssbid))

  # do the same for åpne polygons - will not comment this because same as above
  aapneIndexPoly <- aapneHts  %>%
    gather(type, height, ref, pop, skog) %>%
    group_by(`system:index`) %>%
    mutate(index = scaleSigmoid(height)) %>%
    mutate(index = 1-index) %>%
    filter(type == 'pop') %>%
    dplyr::select(-type, -height) %>%
    left_join(aapneHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
    left_join(popAapne %>% dplyr::select("system:index", ssbid))

  # Make into spatial objects
  vaatmarkIndexPolySpat <- popVaatmark %>%
    dplyr::select("system:index") %>%
    left_join(vaatmarkIndexPoly) %>%
    filter(!is.na(index))
  aapneIndexPolySpat <- popAapne %>%
    dplyr::select("system:index") %>%
    left_join(aapneIndexPoly) %>%
    filter(!is.na(index))

  # Transform to correct projection
  vaatmarkIndexPolySpat <- st_transform(vaatmarkIndexPolySpat, st_crs(regions))
  aapneIndexPolySpat <- st_transform(aapneIndexPolySpat, st_crs(regions))

} else {

  vaatmarkIndexPolySpat <- st_read('./OUTPUT/vaatmark_index.shp')%>%
    mutate(`system:index` = id,
           elevation = elev, 
           vegClimZoneLab = vegZn) %>%
    dplyr::select(-id, -elev, -vegZn)
  aapneIndexPolySpat <- st_read('./OUTPUT/aapne_index.shp')%>%
    mutate(`system:index` = id,
           elevation = elev, 
           vegClimZoneLab = vegZn) %>%
    dplyr::select(-id, -elev, -vegZn)

}

```

Here we will visualize the spatial variation in scaled gjengroind indicator values, as well as the vegetation heights for reference (good condition), indicator, and mature forest (poor condition) over the country. We could plot all the polygons, but it is computationally intensive and very difficult to visualize spatial variations over the entire country. Therefore we will calculate the area-weighted mean in values for each bioclimatic-elevation strata (n = 21) over the country.

```{r, fig.width=10}

# Create intersection of the two strata layers
bioClimElev <- bioclim %>%
  st_intersection(elevStrata) %>%
  mutate(elevation = elevBand,
         vegClimZoneLab = NAVN)%>%
  dplyr::select(elevation, vegClimZoneLab) %>%
  rowid_to_column("ID")

if (runFromScratch){

  # Get area-weighted mean for reference, forest and population vegetation heights 
  # for each unique bioclimatic-elevation zone combination
  vaatmarkIndexStrata <- bioClimElev %>% 
    #sample_n(10000) %>%
    st_join(vaatmarkIndexPolySpat %>% 
                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
    group_by(ID) %>%
    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
                as_tibble() %>% 
                dplyr::select(-geometry)

  aapneIndexStrata <- bioClimElev %>% 
    #sample_n(10000) %>%
    st_join(aapneIndexPolySpat %>% 
                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
    group_by(ID) %>%
    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
                as_tibble() %>% 
                dplyr::select(-geometry)

  # Write out for import later
  vaatmarkIndexStrata %>% write_csv('./DATA/vaatmarkIndexStrata.csv')
  aapneIndexStrata %>% write_csv('./DATA/aapneIndexStrata.csv')

} else {

  vaatmarkIndexStrata <- read_csv('./DATA/vaatmarkIndexStrata.csv')
  aapneIndexStrata <- read_csv('./DATA/aapneIndexStrata.csv')

}

# Make plots of indicator heights and scaled condition values for each strata polygon

theme_legend_1 <- theme(legend.key.size = unit(0.25, 'cm'), #change legend key size
                        legend.key.height = unit(0.25, 'cm'), #change legend key height
                        legend.key.width = unit(0.25, 'cm'), #change legend key width
                        legend.title = element_text(size=6), #change legend title font size
                        legend.text = element_text(size=6)) #change legend text font size

makeStrataHeightMap <- function(data, var, title, limits){
  p <- bioClimElev %>%
    left_join(data, by = c('ID'))  %>%
    mutate(response = .[[var]]) %>%
    ggplot() + 
    geom_sf(aes(fill = response), color=NA)+
      scale_fill_gradientn(colors=viridis::viridis(10),
                           limits = limits,
                           oob = scales::squish,
                           name = 'Veg ht (m)')+
      ggtitle(title) +
      theme_void() +
    theme(plot.title = element_text(size=9)) +
    theme_legend_1

  return (p)
}

makeStrataIndicatorMap <- function(data, title){
  p <- bioClimElev %>%
    left_join(data, by = c('ID'))  %>%
    ggplot() + 
    geom_sf(aes(fill = index), color=NA)+
      scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
                           limits = c(0,1),
                           name = 'Index')+
      ggtitle(title) +
      theme_void()+
    theme(plot.title = element_text(size=9)) +
    theme_legend_1

  return (p)
}

a1 <- makeStrataIndicatorMap(aapneIndexStrata,  'A) Semi-/Naturlig-åpne scaled indicator')
a2 <- makeStrataHeightMap(aapneIndexStrata, "ref", 'B) Semi-/Naturlig-åpne reference height', c(0,5))
a3 <- makeStrataHeightMap(aapneIndexStrata, "pop", 'C) Semi-/Naturlig-åpne indicator height', c(0,5))
a4 <- makeStrataHeightMap(aapneIndexStrata, "skog", 'D) Semi-/Naturlig-åpne forest height', c(0,5))

v1 <- makeStrataIndicatorMap(vaatmarkIndexStrata,  'E) Våtmark scaled indicator')
v2 <- makeStrataHeightMap(vaatmarkIndexStrata, "ref", 'F) Våtmark reference height', c(0,5))
v3 <- makeStrataHeightMap(vaatmarkIndexStrata, "pop", 'G) Våtmark indicator height', c(0,5))
v4 <- makeStrataHeightMap(vaatmarkIndexStrata, "skog", 'H) Mature forest height', c(0,5))

strataHeightPlot <- grid.arrange(a1,a2,a3,a4, v1,v2,v3,v4, ncol=4, widths=c(1,1,1,1), padding = unit(0, "line"), newpage = T)

```

We can see that overall, the condition scores for both ecosystem types are poor along the coastal areas and lower-altitude zones of the country (A and E). At higher altitudes the gjengroing condition scores are higher.

When comparing våtmark to semi-/naturlig-åpne ecosystems, we can see that the reference heights for våtmark are generally higher in terms of good (B and F) and poor (D and H) ecological condition. This makes ecological sense because wetlands can and do naturally occur under forest and therefore våtmark in good condition can have tree cover without necessarily being gjengroing.

### Scaled indicator values at 50km grid level (circa 45 min runtime)

Using the polygon-level indicator scores, we will aggregate to 50km grid level to explore the spatial variation over Norway. We are doing this as an alternative way of visualizing the broad-scale patterns as presented in the section above where we used bioclimatic-elevation strata.

Here we will use a weighted average, but test out the eaTools::ea_spread function! It takes some time to run so best to do this overnight, or to parallelize in some way.

```{r}

if (runFromScratch){

  # Aggregate polygon index scores to grid level using eaTools ea_spread function
  vaatmarkIndexGrid <- eaTools::ea_spread(indicator_data = vaatmarkIndexPolySpat,
                                            indicator = index,
                                            regions = ssb50km,
                                            groups = SSBID,
                                            threshold = 1) %>%
    mutate(index = w_mean, ssbid = ID) %>%
    dplyr::select(ssbid, index, sd)

  aapneIndexGrid <- eaTools::ea_spread(indicator_data = aapneIndexPolySpat,
                                            indicator = index,
                                            regions = ssb50km,
                                            groups = SSBID,
                                            threshold = 1) %>%
    mutate(index = w_mean, ssbid = ID) %>%
    dplyr::select(ssbid, index, sd)

} else {

  print('Will import pre-exported data')

  vaatmarkIndexGrid <- st_read('./OUTPUT/vaatmark_index_grid.shp')
  aapneIndexGrid <- st_read('./OUTPUT/aapne_index_grid.shp')

}

```

Visualize the spatial distribution of gjengroing indicator scores at grid level.

```{r}

vGrid1 <- aapneIndexGrid %>%
  ggplot() +
  geom_sf(aes(fill=index), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.5) +
  scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
                       limits = c(0,1),
                       name = 'Index') +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('A) Semi-/Naturlig gjengroing condition')
#vGrid1

vGrid2 <- vaatmarkIndexGrid %>%
  ggplot() +
  geom_sf(aes(fill=index), color=NA) +
  geom_sf(data=regions, fill=NA, size=0.5) +
  scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
                       limits = c(0,1),
                       name = 'Index') +
  theme_void() +
  theme(legend.position = c(0.7, 0.4)) +
  ggtitle('B) Våtmark gjengroing condition')
#vGrid2

indexGridFig <- grid.arrange(vGrid1, vGrid2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)

```

These patterns mirror those in the previous section with higher condition scores at higher elevations and lower scores along the coastal zone and low-lying parts of the country.

We see some very low condition scores for a few coastal 50km grid cells - this may be an artifact of creating an average with very few population polygons and therefore vulnerable to skewed average indicator scores.

### Scaled indicator values at regional level (circa 120 min runtime)

Now we will aggregate up to the regional level for final reporting. We will use the same method as in the section above, except here we will use a different method to calculate the uncertainty value around the indicator values. The ea_spread function in eaTools uses a bootstrapping method to calculate the standard error in indicator values when aggregating from the polygon to regional level. Because we have hundreds of thousands of polygons per region, the resulting standard errors are extremely small. Therefore we will use the standard deviation in reference values per region. The standard deviation, as a percentage of the average reference value is used to calcualte the uncertainty in the scaled indicator value.

```{r}

if (runFromScratch){

  # Get the mean population, reference and forest heights per region for interpreting index values
  # The standard deviation is used later to define uncertainty around the indicator score
  vaatmarkHeightsRegion <- regions %>%
    st_join(vaatmarkIndexPolySpat) %>%
    as_tibble() %>%
    drop_na(index) %>%
    group_by(region) %>%
    summarise(pop_mean = mean(pop),
              skog_mean = mean(skog),
              ref_mean = mean(ref),
              sd_ref = sd(ref))
  aapneHeightsRegion <- regions %>%
    st_join(aapneIndexPolySpat) %>%
    as_tibble() %>%
    drop_na(index) %>%
    group_by(region)  %>%
    summarise(pop_mean = mean(pop),
              skog_mean = mean(skog),
              ref_mean = mean(ref),
              sd_ref = sd(ref))

  # Aggregate polygon index scores to regional level using eaTools
  vaatmarkIndexRegion <- eaTools::ea_spread(indicator_data = vaatmarkIndexPolySpat,
                                            indicator = index,
                                            regions = regions,
                                            groups = region,
                                            threshold = 1) %>%
    mutate(index = w_mean, region = ID) %>%
    dplyr::select(region, index, sd) %>%
    left_join(vaatmarkHeightsRegion)

  # Only run on first iteration - takes a long time
  aapneIndexRegion <- eaTools::ea_spread(indicator_data = aapneIndexPolySpat,
                                            indicator = index,
                                            regions = regions,
                                            groups = region,
                                            threshold = 1) %>%
    mutate(index = w_mean, region = ID) %>%
    dplyr::select(region, index, sd) %>%
    left_join(aapneHeightsRegion)

} else {

  print('Will import pre-exported data')

  vaatmarkIndexRegion <- st_read('./OUTPUT/vaatmark_index_region.shp')
  aapneIndexRegion <- st_read('./OUTPUT/aapne_index_region.shp')

}

```

Present the indicator values as table, plots, and maps:

```{r}

# Join into a table to present values
aapentTable <- aapneIndexRegion %>% as_tibble() 
vaatmarkTable <- vaatmarkIndexRegion %>% as_tibble()

# Join ecosystem types and calculate uncertainty
indicatorTable <- aapentTable %>% mutate(Ecosystem = 'Semi-/Naturlig åpne') %>%
  bind_rows(vaatmarkTable %>% mutate(Ecosystem = 'Våtmark')) %>%
  # Calculate the error based on the standard deviation in reference polygon heights
  # as a percentage of the reference height, and multiply by index score
  mutate(sd = (sd_ref/ref_mean)*index)

# Format the table for presentation
indicatorTable_formatted <- regions %>%
  left_join(indicatorTable) %>% as_tibble() %>%
  dplyr::select(-geometry)  %>%
  dplyr::select( region, Ecosystem, index, sd, pop_mean, ref_mean, skog_mean) %>%
  mutate_if(is.numeric, round, 2)
names(indicatorTable_formatted) <- c( "Region", "Ecosystem",
                                      "Scaled indicator",  "Scaled Std.dev",
                                      "Indicator height","Reference height",
                                      "Mature forest height")

# Use knitr to visualize
knitr::kable(indicatorTable_formatted)

# Make plots of regional index values
getIndicatorPlot <- function(label){
  regions %>%
    left_join(indicatorTable %>%
                filter(Ecosystem == label)) %>%
    mutate(lowError = index-sd, 
           upError = index+sd,
           upError = ifelse(upError > 1, 1, upError)) %>%
    ggplot(aes(y=region, x=index)) +
    geom_point() +
    geom_segment(aes(yend=region, x=lowError, xend=upError)) +
    coord_cartesian(xlim=c(0,1)) +
    geom_vline(xintercept = 1) +
    geom_vline(xintercept = 0) +
    labs(x = 'Tilstandsverdi') +
    theme(axis.title.y=element_blank()) +
    ggtitle(label) 
}
i1 <- getIndicatorPlot('Semi-/Naturlig åpne')
i2 <- getIndicatorPlot('Våtmark')
indicatorGraphFig <- grid.arrange(i1, i2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)

# Make maps of regional index values
getIndicatorMap <- function(data, label){
  regions %>%
    left_join(data) %>%
    ggplot() +
    geom_sf(aes(fill=index)) +
    scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
                         limits = c(0,1))+
    ggtitle(label)+
    theme_void()
}
m1 <- getIndicatorMap(aapentTable, 'A) Semi-/Naturlig åpne')
m2 <- getIndicatorMap(vaatmarkTable, 'B) Våtmark')
indicatorGMapFig <- grid.arrange(m1, m2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T) 
```

From these tables, figures and maps we can see that the gjengroing condition indicator scores are on average above 0.8 and therefore indicate a good overall condition in Norway. Østlandet comes out best for våtmark and semi-naturlig åpne ecosystems. Nord-Norge has poorest condition for semi-naturlig åpne ecosystems, while vestlandet has the poorest scores for våtmark.

### Uncertainty and validation

The calculation of the indicator involves many small decisions which can have significant effects on the resulting conditions cores. Therefore in the future it would be good to perform sensitivity analyses with each decision to test the effect on the outcome. The decisions which would benefit from further exploration include:

* **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise. However, both NiN and AR5 are spatially and thematically biased in some ways. Therefore, it may be worth exploring alternative datasets in the future. NiN polygons are biased towards low lying areas which are generally warmer and more productive. For a given bioclimatic-elevation strata, this could lead to higher reference values for vegetation height than one would expect with a representative reference sample. This could consequently result in high elevation population polygons appearing to have very good condition ebcause they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata. This is a possible explanation for why the high elevation parts of norway appear to be in very good ecological condition.

* **Definition of good condition states in NiN data.** Here we used the overall "tilstand" score for each NiN reference polygon. However, the overall score is a combination of several sub-scores that may or may not be relevant for ecological condition. For instance, ditches or wheel marks in wetlands is a common condition score in NiN, however it is debatable as to whether this should be considered a part of ecological condition. Therefore it may be beneficial to attempt repeating the indicator calculation using sub-categories of "tilstand" in the NiN data.

* **Definition of climax successional stage for tree regrowth.** Here we use the 50th percentile of LiDAR-derived tree heights in AR5 forest polygons that have not been harvested since 1986. This defines the zero on the scaled indicator score. The resulting indicator scores will vary widely depending on how you define this zero-point. Therefore, testing different definitions (i.e. tree height percentiles) may be important.

* **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data. However, for poor condition we use climax successional stage for trees (as explained above) within 200m of each population polygon. For polygons with no surrounding forest, we impute with regional reference values (mean of forest height in elevation-bioclamic zone). The choice of 200m buffer zone was largely based on computational limitations. Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex plygons takes time, even in GEE. It may make more ecological sense to define a larger buffer zone to use for defining reference for poor condition. However, longer processing times need to be expected.

* **Quantification of uncertainty around indicator scores.** Here we used the standard deviation in the vegetation heights for reference (NiN) polygons within each region to quantify the uncertainty around scaled indicator scores. We did this because the method used in eaTools::ea_spread() function is a bootstrapping approach which resulted in extremely small uncertainty estimates due to the large number of polygons included in our analysis. In the future, a better method for quantifying uncertainty is needed. Perhaps by running sensitivity analyses that test several variations of the points mentioned above can form the basis for quantifying an error margin around indicator values.

#### Validation of random polygons

After exporting the final polygon-level indicator scores, we imported to QGIS and explored how they look at the landscape scale. We chose random locations spread across the 5 regions in Norway. The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosysystem polygons from AR5. The polygons have jaggered edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space. Each polygon is colored based on the final scaled gjengroing indicator score (0 to 1). The polygons are also labelled with the median LiDAR-derived forest height. An orthophoto from Norgeibilder is displayed in the background for reference.

In the first example from midt-norge, we see a våtmark polygon colored red (ie. in poor condition) due to the growth of a forest in the northern section. It looks like a plantation forest and possibly occured after the wetland was drained. The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m. Therefore it gets a scaped indicator score of 0.1. In contrast the wetland plygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.

```{r}
include_graphics('./DATA/images/val_1.jpg')
```

The second example from Sørlandet shows three AR5 semi-naturlig åpne polygons with different indicator values. The lower polygon has poor ecological condition because its median vegetation height is 6.2m which is close to the local reference value for mature forest (7.9m). Although the centeral polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.

```{r}
include_graphics('./DATA/images/val_2.jpg')
```

The third example from Østlandet shows a range of indicator values for våtmark polygons. Although many of the polygons are partly covered by trees, their indicator values show good condition because the local reference for mature forest (zero on the indicator scale) is very high in this landscape of productive forest.

```{r}
include_graphics('./DATA/images/val_3.jpg')
```

The fourth example from Nord-norge shows AR5 semi-naturlig åpne polygons with low and high condition scores. The large polygon with poor condition score has a median vegetation height of 2.7 which is nearly as high as the local reference for mature forest (3.8m).

```{r}
include_graphics('./DATA/images/val_4.jpg')
```

The fifth and final example from further north in Nord-norge shows two våtmark polygons with contrasting condition scores, yet exactly the same median vegetation heights of 0.6m. The difference in the scaled iindicator value is due to different local reference values for surrounding forest. The polygon on the left has surrounding forest (within 200m buffer zone) with median height 0.5m, while the polygon on the right has surrounding forest of 0.7m. This creates a contrast in the final indicator valye and illustrates the effect that a local reference value can have on defining good and poor ecological condition.

```{r}
include_graphics('./DATA/images/val_5.jpg')
```

### Eksport file (final product)

We will export gjengroing indicator values at three levels:
- polygon level
- 50km grid level
- regional level

```{r}

if (runFromScratch){
  # Polygon level
  vaatmarkIndexPolySpat  %>%
    #mutate(area = as.numeric(st_area(geometry)))%>% 
    #filter(area > 7500) %>%
    mutate(id = `system:index`,
           elev = elevation, 
           vegZn = vegClimZoneLab) %>%
    dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
    st_write('./OUTPUT/vaatmark_index.shp', append=FALSE)

  aapneIndexPolySpat %>% 
    #mutate(area = as.numeric(st_area(geometry)))%>% 
    #filter(area > 7500) %>%
    mutate(id = `system:index`,
           elev = elevation, 
           vegZn = vegClimZoneLab) %>%
    dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
    st_write('./OUTPUT/aapne_index.shp', append=FALSE)

  # Grid level
  vaatmarkIndexGrid %>%
    st_write('./OUTPUT/vaatmark_index_grid.shp', append = FALSE)
  aapneIndexGrid %>%
    st_write('./OUTPUT/aapne_index_grid.shp', append = FALSE)

  # Region level
  vaatmarkIndexRegion %>%
    st_write('./OUTPUT/vaatmark_index_region.shp', append = FALSE)
  aapneIndexRegion %>%
    st_write('./OUTPUT/aapne_index_region.shp', append = FALSE)
}

```

### 2023-08-10 java/GEE_Script.js
Delete GEE_Script.js
deleting the first version, which is replaced by the newrer upload

@@ -1,233 +1,0 @@
-// Author: Zander Venter - zander.venter@nina.no
-
-// Workflow in this script:
-  // 1. Setup global environment: Import ecossytem reference polygons generated in R,
-    // SSB grid, FKB buildings, LiDAR, AR5, clear-cut data
-  // 2. Export data coverage CSV for visualization in R
-  // 3. Extract sample of vegetationheights for NiN reference and AR5 polygons
-
-
-/*
-  // 1. Setup global environment ///////////////////////////////////////////////////////////////////////////////
-*/
-
-// Geometries for debugging
-var geometry = 
-    /* color: #d63000 */
-    /* shown: false */
-    ee.Geometry.Point([10.913510914527835, 60.03670256866523]),
-    geometry2 = 
-    /* color: #98ff00 */
-    /* shown: false */
-    ee.Geometry.MultiPoint(
-        [[11.022080131331995, 59.94877591779931],
-         [11.00148076609762, 59.997223922250754],
-         [11.080444999496057, 60.00374695882402],
-         [11.201294608871057, 59.99688056890364]]);
-
-// Dictionary to inform the AR5 area code types
-var artypeDict = {
- 'Bebygd': 11,
- 'Ferskvann': 81,
- 'Fulldyrka jord': 21,
- 'Hav': 82,
- 'Ikke kartlag': 99,
- 'Innmarksbeite': 23,
- 'Myr': 60,
- 'Overflatedyrka jord': 22,
- 'Samferdsel': 12,
- 'Skog': 30,
- 'Snøisbre': 70,
- 'Åpen fastmark': 50
-}
-
-// Set the map to satellite image and add a background layer for viz purposes
-Map.setOptions('HYBRID')
-Map.addLayer(ee.Image(1), {palette:['white']}, 'white background', 0)
-
-// Color pallete for veg heights
-var virdis = '440154,472878,3E4A89,31688E,25838E,1E9E89,35B779,6CCE59,B5DE2C,FDE725';
-
-// Import 5 regions shapefile
-var regions = ee.FeatureCollection('users/zandersamuel/NINA/Vector/Norway_administrative_regions_2022');
-regions = regions.select(['id'])
-
-// Import SSB grid 
-var grid = ee.FeatureCollection('users/zandersamuel/NINA/Vector/Norway_SSB_grid_10km')
-grid = grid.select(['ssbid'], ['SSBID']);
[…]

### 2023-08-10 DRAFT_gjengroing.Rmd
Update DRAFT_gjengroing.Rmd
changes file paths to P:/...

@@
 ```{r}
-include_graphics('./DATA/images/gjengroing_schematic.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/gjengroing_schematic.jpg')
 ```
@@
 ```{r}
-regions <- st_read('./DATA/regions.shp') %>%
+regions <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/regions.shp') %>%
   mutate(region = factor(region) )
@@
 readVegHeightFiles <- function(uniqueString){
-  dir <- './DATA/From_GEE/vegHeights/'
+  dir <- 'P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/vegHeights/'
   files <- list.files(dir)[str_detect(list.files(dir), uniqueString)]

### 2023-08-10 DRAFT_gjengroing.Rmd
Update DRAFT_gjengroing.Rmd
exchanged path files to P: and R:

@@
 ```{r}
-ssb10km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_10km/ssb10km.shp')%>%
+ssb10km <- st_read('R:/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_10km/ssb10km.shp')%>%
   # mutate the id value so that it aligns with the data coming from GEE
@@
 
-ssb50km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_50km/ssb50km.shp')%>%
+ssb50km <- st_read('R:/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_50km/ssb50km.shp')%>%
   # mutate the id value so that it aligns with the data coming from GEE
@@
 ```{r}
-bioclim <- st_read('/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp') %>%
+bioclim <- st_read('R:/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp') %>%
   mutate(NAVN = ifelse(NAVN == 'S°rboreal', 'Sørboreal', NAVN))
@@
 ```{r}
-nin <- st_read('/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_NiN_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb')%>%
+nin <- st_read('R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_NiN_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb')%>%
   # Simplify ecosystem names
@@
 # Import data coverages per SSB 10km grid cell
-areaCovers <- read_csv('./DATA/From_GEE/area_cover_grid.csv') %>%
+areaCovers <- read_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/area_cover_grid.csv') %>%
   dplyr::select(-'.geo', -'system:index')
@@
 
-elevStrata <- st_read('./DATA/From_GEE/elevation_stratified.shp') %>%
+elevStrata <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/elevation_stratified.shp') %>%
   mutate(elevBand = ifelse(mode == 1, '0-300',
@@
   refVaatmark %>%
-    st_write('./DATA/refVaatmark.csv')
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
   refAapne %>%
-    st_write('./DATA/refAapne.csv')
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
 
@@
 
-  refVaatmark  <- read_csv('./DATA/refVaatmark.csv')
-  refAapne <- read_csv('./DATA/refAapne.csv')
+  refVaatmark  <- read_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
+  refAapne <- read_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
 
@@
 
-  vaatmarkIndexPolySpat <- st_read('./OUTPUT/vaatmark_index.shp')%>%
+  vaatmarkIndexPolySpat <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp')%>%
     mutate(`system:index` = id,
@@
     dplyr::select(-id, -elev, -vegZn)
-  aapneIndexPolySpat <- st_read('./OUTPUT/aapne_index.shp')%>%
+  aapneIndexPolySpat <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp')%>%
     mutate(`system:index` = id,
@@
   # Write out for import later
-  vaatmarkIndexStrata %>% write_csv('./DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata %>% write_csv('./DATA/aapneIndexStrata.csv')
+  vaatmarkIndexStrata %>% write_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
+  aapneIndexStrata %>% write_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
 
@@
 
-  vaatmarkIndexStrata <- read_csv('./DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata <- read_csv('./DATA/aapneIndexStrata.csv')
+  vaatmarkIndexStrata <- read_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
+  aapneIndexStrata <- read_csv('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
 
@@
 
-  vaatmarkIndexGrid <- st_read('./OUTPUT/vaatmark_index_grid.shp')
-  aapneIndexGrid <- st_read('./OUTPUT/aapne_index_grid.shp')
+  vaatmarkIndexGrid <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp')
+  aapneIndexGrid <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp')
 
@@
 
-  vaatmarkIndexRegion <- st_read('./OUTPUT/vaatmark_index_region.shp')
-  aapneIndexRegion <- st_read('./OUTPUT/aapne_index_region.shp')
+  vaatmarkIndexRegion <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp')
+  aapneIndexRegion <- st_read('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp')
 
@@
 ```{r}
-include_graphics('./DATA/images/val_1.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_1.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_2.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_2.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_3.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_3.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_4.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_4.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_5.jpg')
+include_graphics('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_5.jpg')
 ```
@@
     dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('./OUTPUT/vaatmark_index.shp', append=FALSE)
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp', append=FALSE)
 
@@
     dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('./OUTPUT/aapne_index.shp', append=FALSE)
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp', append=FALSE)
 
@@
   vaatmarkIndexGrid %>%
-    st_write('./OUTPUT/vaatmark_index_grid.shp', append = FALSE)
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp', append = FALSE)
   aapneIndexGrid %>%
-    st_write('./OUTPUT/aapne_index_grid.shp', append = FALSE)
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp', append = FALSE)
 
@@
   vaatmarkIndexRegion %>%
-    st_write('./OUTPUT/vaatmark_index_region.shp', append = FALSE)
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp', append = FALSE)
   aapneIndexRegion %>%
-    st_write('./OUTPUT/aapne_index_region.shp', append = FALSE)
-}
-
-```
+    st_write('P:/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp', append = FALSE)
+}
+
+```

### 2023-08-24 R/functionalPlantIndicators/code naturally open/6 post scaling analyses_natopen.R
starting the natopen markdown

@@ -58,7 +58,7 @@ res.natopen.ANO <-
 summary(res.natopen.ANO$scaled_value)
 
 
-summary(res.natopen.ANO[,79:81])
+#summary(res.natopen.ANO[,79:81])
 
 
 # similarly for GRUK
@@ -97,7 +97,7 @@ res.natopen.GRUK <-
                                    pull(original)
   )
 
-summary(res.natopen.GRUK[,52:54])
+#summary(res.natopen.GRUK[,52:54])
 
 # making the plot, ANO, all indicators
 ggplot(data=subset(res.natopen.ANO,!is.na(scaled_value)), aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
@@ -372,6 +372,9 @@ indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Northern Norway",c(
 res.natopen.ANO2$RR1[res.natopen.ANO2$RR1==1] <- 0.999
 res.natopen.ANO2$RR1[res.natopen.ANO2$RR1==0] <- 0.001
 
+res.natopen.ANO2$CC1[res.natopen.ANO2$CC1==1] <- 0.999
+res.natopen.ANO2$CC1[res.natopen.ANO2$CC1==0] <- 0.001
+
 regnor <- regnor %>%
   mutate(
     RR1.ANO.reg.mean = c(indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Northern Norway",c("RR1","ano_flate_id")])[1],
@@ -380,23 +383,44 @@ regnor <- regnor %>%
                                  indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Western Norway",c("RR1","ano_flate_id")])[1],
                                  indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Southern Norway",c("RR1","ano_flate_id")])[1]
     ),
-    RR1.ANO.reg.se = c(indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Northern Norway",c("RR1","ano_flate_id")])[2],
-                               indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Central Norway",c("RR1","ano_flate_id")])[2],
-                               indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Eastern Norway",c("RR1","ano_flate_id")])[2],
[…]

### 2023-08-25 functional_plant_indicators_wetland.Rmd
continuing on the markdown for all three systems

--- 2023-05-10 functional_plant_indicators_wetland.Rmd
@@
 In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
-In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
+In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which as of today comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
 

### 2023-08-28 DRAFT_functional_plant_indicators_natopen.Rmd
analysis bit of natopen markdown

# Functional plant indicators, naturally open ecosystems below the tree line (Grime's CSR values, Light, Nitrogen, Soil disturbance) {#Functional-plant-indicators-seminat}

<br />

_Author and date:_
Joachim Töpper

August 2023

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(readxl)
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra") 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
Ecosystem <- "naturally open below tree line"
Egenskap  <- c("Primærproduksjon","Abiotiske forhold")
ECT       <- "Functional state characteristic"
Contact   <- "Joachim Töpper"
```

```{r, echo=FALSE}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

## Introduction
Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003). The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf. Töpper et al. 2018). For instance, using an indicator for moisture one would find a functional signature with higher moisture values for plant communities in mires compared to e.g. grasslands or forests. Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition. Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring programs ANO (Tingstad et al. 2019) and GRUK (Evju et al. 2020) for naturally open ecosystems below tree line (abbreviated as 'nat-open' henceforth). We calculate the functional signature of plant communities in the monitored sites with respect to Grime's CSR values, light, nitrogen, and soil disturbance. These functional signatures are then compared to reference distributions of functional signature, separately for each nat-open ecosystem type, calculated from 'generalized species lists' developed for ecosystem types in the Norwegian categorization system for eco-diversity (Halvorsen et al. 2020). These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al. 2021, Töpper & Jakobsson 2021). Note that deviations from the reference may occur in both directions, e.g. the nitrogen signature from the testing data may be higher or lower than in the reference. Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately. Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).

## About the underlying data
In the 'functional plant indicator' project for nat-open ecosystems, we use five sets of data for building indicators for ecological condition:

- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020).
- as reference data we use (3) generalized species lists developed by Evju et al. 2023 for nat-open ecosystem types.
- (4) Swedish plant indicator data for light, nitrogen, ans soil disturbance from Tyler et al. (2021), and (5) Grime's CSR values for plant species' strategies (towards competition, stress, and ruderal conditions)

(1) ANO monitoring data:
ANO stands for 'areal-representativ naturovervåking', i.e. 'area representative nature monitoring'. 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle. Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible). For the vegetation analyses, the cover of each vascular plant species in the plot is recorded. Every vegetation analysis is accompanied by an assessment of the ecosystem the plot lies in, including ecosystem type and some additional variables related to ecosystem-specific drivers of state. In the analysis in this document, we only use the plots which were classified as one of the nat-open ecosystem types in the Norwegian categorization system for eco-diversity (NiN) and which are not registered as lying in mountain areas above the tree line.
In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.

(2) GRUK monitoring data
GRUK stands for 'Grunnlendt åpen kalkmark', and represents a monitoring of open, limestone rich ecosystems with soils too shallow for forest establishment. The limestone rich focus of the scheme is motivated from a high biodiversity value and places all monitoring sites in an area around the Oslofjord. As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites, all of which are mapped to the ecosystem type "åpen grunnlendt mark" (T2) and therein to the limestone-rich subtypes T2-C-7 and T2-C-8 in the Norwegian classification system for ecosystem types.
The vegetation analysis in GRUK record percent cover for every species of vascular plants in 1 sqm plots. In addition, a range of site variables related to ecosystem-specific drivers of state are recorded.
In the analysis in this document, we use all GRUK monitoring data data, which span the years 2020-22. The 2022-data also include an assessment of site condition, which is included in the analysis of results in this document. In the future, these data data may become available in Miljødirektoratets kartkatalog.

(3) NiN reference data:
The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species. In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5). For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.

(4) The Swedish plant indicator set published by Tyler et al. (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well. From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.

(5) Grime's system of plant strategy scores (Grime 1974) comprises relative (too one another) scores for the competition-, stress-, and disturbance("ruderality")-related life strategy of plant species. In the analysis, in this document we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g. alien species for competition, climate change for stress, land use change for ruderality).

### Representativity in time and space
For nat-open ecosystems, the ANO data in this analysis contain 143 plots in 52 sites, in principle distributed randomly across the country. As nat-open ecosystems occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions. The 143 plots are distributed across regions in the following way:

- Northern Norway: 29
- Central Norway:  39
- Eastern Norway:  34
- Western Norway:  24
- Southern Norway: 17

For GRUK, this analysis covers 1103 plots in 146 sites.

The 1103 plots are distributed across regions in the following way:

- Northern Norway: 0
- Central Norway:  0
- Eastern Norway:  1042
- Western Norway:  0
- Southern Norway: 61

### Temporal coverage
The ANO evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme. GRUK covers 2020-2022. Thus, there is no actual time series to these data, and the indicator evaluation does therefore not include any temporal analyses.

## Collinearities with other indicators
<!-- Text here -->

## Reference state and values
### Reference state
The reference state is defined via the functional signature of the generalized species lists for NiN ecosystem types (see also Töpper et al. 2018). For the nat-open ecosystem types these lists have been newly prepared by Evju et al. 2023. By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator. These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN. A more extensive discussion on the use of reference communities can be found in Jakobsson et al. (2020).

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al. 2010). For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution. As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al. 2022), and deviation from the optimal state thus may occur in both direction (e.g. indicating too low or too high pH), we need to define two threshold values for good ecological condition as well as both a minimum and maximum value. In line with previous assessments of ecological condition for Norwegian forests and mountains, we define a lower and an upper threshold value via the 95% confidence interval of the reference distribution, i.e. its 0.025 and 0.975 quantiles. The minimum and maximum values are given by the minimum and maximum of the possible indicator values for each respective plant functional indicator. For details on the scaling principles in IBECA, please see Töpper & Jakobsson (2021).

## Uncertainties
We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.

## References
Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185

Evju, M., Stabbetorp, O.E., Olsen, S.L., Bratli, H., Often, A. & Bakkestuen, V. 2020. Dry calcareous grasslands in the Oslofjord region. A test of monitoring protocols and results for 2020. NINA Report 1910. Norwegian Institute for Nature Research.

Evju, M., Stabbetorp, O.E., Olsen, S.L., Bratli, ... . 2023. Generalized species lists for naturally open ecosystem types below the tree line in Norway. in prep.

Framstad, E., Kolstad, A. L., Nybø, S., Töpper, J. & Vandvik, V. 2022. The condition of forest and mountain ecosystems in Norway. Assessment by the IBECA method. NINA Report 2100. Norwegian Institute for Nature Research.

Grime J.P. 1974. Vegetation classification by reference to strategies. Nature 250(5461):26-31.

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E. (2020). Towards a systematics of ecodiversity: The EcoSyst framework. Global Ecology and Biogeography, 29(11), 1887-1906. doi:10.1111/geb.13164

Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik. V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2020. Setting reference levels and limits for good ecological condition in terrestrial ecosystems. Insights from a case study based on the IBECA approach. Ecological Indicators 116: 106492.

Jakobsson, S., Evju, M., Framstad, E., Imbert, A., Lyngstad, A., Sickel, H., Sverdrup-Thygeson, A., Töpper, J., Vandvik, V., Velle, L.G., Aarrestad, P.A. & Nybø, S. 2021. An index-based assessment of ecological condition and its links to international frameworks. Ecological Indicators 124: 107252.

Tingstad, L., Evju, M., Sickel, H., & Töpper, J. 2019. Utvikling av nasjonal arealrepresentativ naturovervåking (ANO). Forslag til gjennomføring, protokoller og kostnadsvurderinger med utgangspunkt i erfaringer fra uttesting i Trøndelag. NINA Rapport 1642.

Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967. Norwegian Institute for Nature Research.

Töpper, J., Velle, L.G. & Vandvik, V. 2018. Developing a method for assessment of ecological state based on indicator values after Ellenberg and Grime (revised edition). NINA Report 1529b. Norwegian Institute for Nature Research.

Tyler, T., Herbertsson, L., Olofsson, J., & Olsson, P. A. 2021. Ecological indicator and traits values for Swedish vascular plants. Ecological In-dicators, 120. doi:10.1016/j.ecolind.2020.106923

## Analyses
### Data sets

ANO data: ANO.sp contains the species data, ANO.geo contains site data
```{r, echo = F}
# Add data from cache
ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))
```

GRUK data: GRUK.species contains the species data, GRUK.variable contains site data, GRUK2021.condition contains a field-based condition assessment from the 2021 season
```{r, echo = F}
# Add data from cache
GRUK.variables<-readRDS(paste0(here::here(),"/data/cache/GRUK.variables.RDS"))
GRUK.species<-readRDS(paste0(here::here(),"/data/cache/GRUK.species.RDS"))
GRUK2021.condition<-readRDS(paste0(here::here(),"/data/cache/GRUK2021.condition.RDS"))
```

Plant indicators from Tyler et al. (2021) and Grime (1974)
```{r, echo = F}
## data from cache
ind.Tyler<-readRDS(paste0(here::here(),"/data/cache/ind.Tyler.RDS"))
ind.Grime<-readRDS(paste0(here::here(),"/data/cache/ind.Grime.RDS"))
```

Generalized species lists (reference communities): natopen_NiN_ref contains the reference species lists, natopen_NiN_ref_spInfo contains additional taxonomic information for each species
```{r, echo = F}
# data from cache
natopen_NiN_ref<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref.RDS"))
natopen_NiN_ref_spInfo<-readRDS(paste0(here::here(),"/data/cache/natopen_NiN_ref_spInfo.RDS"))
```

#### Data handling
- Checking for errors
- Checking species nomenclature in the different species lists to make species and indicator data possible to merge
- Merging indicator data with monitoring data and indicator data with reference data
(not shown here, but documented in the code)

```{r, include = F}

### Plant indicator data
# trimming away sub-species & co, and descriptor info
ind.Grime[,'species.orig'] <- ind.Grime[,'species']
ind.Grime[,'species'] <- word(ind.Grime[,'species'], 1,2)
names(ind.Tyler)[1] <- 'species'
ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
ind.Tyler <- ind.Tyler[!is.na(ind.Tyler$species),]
ind.Tyler[,'species.orig'] <- ind.Tyler[,'species']
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)

# dealing with 'duplicates'
ind.Grime[duplicated(ind.Grime[,'species']),"species"]
ind.Grime.dup <- ind.Grime[duplicated(ind.Grime[,'species']),c("species")]
ind.Grime[ind.Grime$species %in% ind.Grime.dup,]
# getting rid of the duplicates
ind.Grime <- ind.Grime %>% filter( !(species.orig %in% list("Carex viridula brachyrrhyncha",
                                                            "Dactylorhiza fuchsii praetermissa",
                                                            "Medicago sativa varia",
                                                            "Montia fontana chondrosperma",
                                                            "Papaver dubium lecoqii",
                                                            "Sanguisorba minor muricata")
) )
#ind.Grime[duplicated(ind.Grime[,'species']),"species"]

names(ind.Tyler)[1] <- 'species'
ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
ind.Tyler <- ind.Tyler[!is.na(ind.Tyler$species),]
ind.Tyler[,'species.orig'] <- ind.Tyler[,'species']
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)
#ind.Tyler2 <- ind.Tyler
#ind.Tyler <- ind.Tyler2
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
ind.Tyler <- ind.Tyler %>% filter( !(species.orig %in% list("Ammophila arenaria x Calamagrostis epigejos",
                                                            "Anemone nemorosa x ranunculoides",
                                                            "Armeria maritima ssp. elongata",
                                                            "Asplenium trichomanes ssp. quadrivalens",
                                                            "Calystegia sepium ssp. spectabilis",
                                                            "Campanula glomerata 'Superba'",
                                                            "Dactylorhiza maculata ssp. fuchsii",
                                                            "Erigeron acris ssp. droebachensis",
                                                            "Erigeron acris ssp. politus",
                                                            "Erysimum cheiranthoides L. ssp. alatum",
                                                            "Euphrasia nemorosa x stricta var. brevipila",
                                                            "Galium mollugo x verum",
                                                            "Geum rivale x urbanum",
                                                            "Hylotelephium telephium (ssp. maximum)",
                                                            "Juncus alpinoarticulatus ssp. rariflorus",
                                                            "Lamiastrum galeobdolon ssp. argentatum",
                                                            "Lathyrus latifolius ssp. heterophyllus",
                                                            "Medicago sativa ssp. falcata",
                                                            "Medicago sativa ssp. x varia",
                                                            "Monotropa hypopitys ssp. hypophegea",
                                                            "Ononis spinosa ssp. hircina",
                                                            "Ononis spinosa ssp. procurrens",
                                                            "Pilosella aurantiaca ssp. decolorans",
                                                            "Pilosella aurantiaca ssp. dimorpha",
                                                            "Pilosella cymosa ssp. gotlandica",
                                                            "Pilosella cymosa ssp. praealta",
                                                            "Pilosella officinarum ssp. peleteranum",
                                                            "Poa x jemtlandica (Almq.) K. Richt.",
                                                            "Poa x herjedalica Harry Sm.",
                                                            "Ranunculus peltatus ssp. baudotii",
                                                            "Sagittaria natans x sagittifolia",
                                                            "Salix repens ssp. rosmarinifolia",
                                                            "Stellaria nemorum L. ssp. montana",
                                                            "Trichophorum cespitosum ssp. germanicum")
) )
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
# getting rid of sect. for Hieracium
ind.Tyler <- ind.Tyler %>% mutate(species=gsub("sect. ","",species.orig))
ind.Tyler[,'species'] <- word(ind.Tyler[,'species'], 1,2)

#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
ind.Tyler.dup <- ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]
#ind.Tyler[ind.Tyler$species %in% ind.Tyler.dup,c("Light","Moisture","Soil_reaction_pH","Nitrogen","species.orig","species")]
# only hybrids left -> get rid of these
ind.Tyler <- ind.Tyler[!duplicated(ind.Tyler[,'species']),]
#ind.Tyler[duplicated(ind.Tyler[,'species']),"species"]

ind.Tyler$species <- as.factor(ind.Tyler$species)
#summary(ind.Tyler$species)
# no duplicates left

# merge indicator data
ind.dat <- merge(ind.Grime,ind.Tyler, by="species", all=T)
#summary(ind.dat)
#ind.dat[duplicated(ind.dat[,'species']),"species"]
ind.dat$species <- as.factor(ind.dat$species)
#summary(ind.dat$species)
head(ind.dat)

### ANO monitoring data

head(ANO.sp)
head(ANO.geo)

## fix NiN information
ANO.geo$hovedtype_rute <- substr(ANO.geo$kartleggingsenhet_1m2,1,3) # take the 3 first characters
ANO.geo$hovedtype_rute <- gsub("-", "", ANO.geo$hovedtype_rute) # remove hyphen
unique(as.factor(ANO.geo$hovedtype_rute))

## fix NiN-variables
colnames(ANO.geo)
colnames(ANO.geo)[42:47] <- c("groeftingsintensitet",
                              "bruksintensitet",
                              "beitetrykk",
                              "slatteintensitet",
                              "tungekjoretoy",
                              "slitasje")
head(ANO.geo)

# remove variable code in the data
ANO.geo$groeftingsintensitet <- gsub("7GR-GI_", "", ANO.geo$groeftingsintensitet) 
unique(ANO.geo$groeftingsintensitet)
ANO.geo$groeftingsintensitet <- gsub("X", "NA", ANO.geo$groeftingsintensitet)
unique(ANO.geo$groeftingsintensitet)
ANO.geo$groeftingsintensitet <- as.numeric(ANO.geo$groeftingsintensitet)
unique(ANO.geo$groeftingsintensitet)

ANO.geo$bruksintensitet <- gsub("7JB-BA_", "", ANO.geo$bruksintensitet) 
unique(ANO.geo$bruksintensitet)
ANO.geo$bruksintensitet <- gsub("X", "NA", ANO.geo$bruksintensitet)
unique(ANO.geo$bruksintensitet)
ANO.geo$bruksintensitet <- as.numeric(ANO.geo$bruksintensitet)
unique(ANO.geo$bruksintensitet)

ANO.geo$beitetrykk <- gsub("7JB-BT_", "", ANO.geo$beitetrykk) 
unique(ANO.geo$beitetrykk)
ANO.geo$beitetrykk <- gsub("X", "NA", ANO.geo$beitetrykk)
unique(ANO.geo$beitetrykk)
ANO.geo$beitetrykk <- as.numeric(ANO.geo$beitetrykk)
unique(ANO.geo$beitetrykk)

ANO.geo$slatteintensitet <- gsub("7JB-SI_", "", ANO.geo$slatteintensitet) 
unique(ANO.geo$slatteintensitet)
ANO.geo$slatteintensitet <- gsub("X", "NA", ANO.geo$slatteintensitet)
unique(ANO.geo$slatteintensitet)
ANO.geo$slatteintensitet <- as.numeric(ANO.geo$slatteintensitet)
unique(ANO.geo$slatteintensitet)

ANO.geo$tungekjoretoy <- gsub("7TK_", "", ANO.geo$tungekjoretoy) 
unique(ANO.geo$tungekjoretoy)
ANO.geo$tungekjoretoy <- gsub("X", "NA", ANO.geo$tungekjoretoy)
unique(ANO.geo$tungekjoretoy)
ANO.geo$tungekjoretoy <- as.numeric(ANO.geo$tungekjoretoy)
unique(ANO.geo$tungekjoretoy)

ANO.geo$slitasje <- gsub("7SE_", "", ANO.geo$slitasje) 
unique(ANO.geo$slitasje)
ANO.geo$slitasje <- gsub("X", "NA", ANO.geo$slitasje)
unique(ANO.geo$slitasje)
ANO.geo$slitasje <- as.numeric(ANO.geo$slitasje)
unique(ANO.geo$slitasje)

## check that every point is present only once
length(levels(as.factor(ANO.geo$ano_flate_id)))
length(levels(as.factor(ANO.geo$ano_punkt_id)))
summary(as.factor(ANO.geo$ano_punkt_id))
# there's many double presences, probably some wrong registrations of point numbers,
# CHECK THIS when preparing ecosystem-datasets for scaling

# fix species names
ANO.sp$Species <- ANO.sp$art_navn
unique(as.factor(ANO.sp$Species))
#ANO.sp$Species <- sub(".*?_", "", ANO.sp$Species) # lose the Norwegian name in the front
ANO.sp[,'Species'] <- word(ANO.sp[,'Species'], 1,2) # lose subspecies
ANO.sp$Species <- str_to_title(ANO.sp$Species) # make first letter capital
#ANO.sp$Species <- gsub("_", " ", ANO.sp$Species) # replace underscore with space
ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after hyphon to lowercase
ANO.sp$Species <- gsub("( .*)","\\L\\1",ANO.sp$Species,perl=TRUE) # make capital letters after space to lowercase

## merge species data with indicators
ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                    y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                    by.x="Species", by.y="species", all.x=T)
summary(ANO.sp.ind)

## checking which species didn't find a match
unique(ANO.sp.ind[is.na(ANO.sp.ind$Light),'Species'])

# fix species name issues
ind.dat <- ind.dat %>% 
  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
  mutate(species=str_replace(species,"Cherleria biflora", "Minuartia biflora"))

ANO.sp <- ANO.sp %>% 
  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
  mutate(Species=str_replace(Species,"Carex paupercula", "Carex magellanica")) %>%
  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  mutate(Species=str_replace(Species,"Carex viridula", "Carex flava")) %>%
  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Vulgata")) %>%
  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
  mutate(Species=str_replace(Species,"Poa alpigena", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Poa angustifolia", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa")) %>%
  mutate(Species=str_replace(Species,"Syringa emodi", "Syringa vulgaris")) %>%
  mutate(Species=str_replace(Species,"Taraxacum crocea", "Taraxacum officinale")) %>%
  mutate(Species=str_replace(Species,"Taraxacum croceum", "Taraxacum officinale")) %>%
  mutate(Species=str_replace(Species,"Trientalis europaea", "Lysimachia europaea")) %>%
  mutate(Species=str_replace(Species,"Trifolium pallidum", "Trifolium pratense"))

## merge species data with indicators
ANO.sp.ind <- merge(x=ANO.sp[,c("Species", "art_dekning", "ParentGlobalID")], 
                    y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                    by.x="Species", by.y="species", all.x=T)
summary(ANO.sp.ind)
# checking which species didn't find a match
unique(ANO.sp.ind[is.na(ANO.sp.ind$Light),'Species'])
# don't find synonyms for these in the ind lists

# trimming away the points without information on NiN, species or cover
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$Species),]
ANO.sp.ind <- ANO.sp.ind[!is.na(ANO.sp.ind$art_dekning),]

summary(ANO.sp.ind)
head(ANO.sp.ind)
rm(ANO.sp)

### GRUK data handling
names(GRUK.variables)
names(GRUK.species)
names(GRUK2021.condition)

colnames(GRUK.species)[5] <- "art_dekning"

# fix species names
GRUK.species <- as.data.frame(GRUK.species)

GRUK.species$Species <- GRUK.species$Navn
unique(as.factor(GRUK.species$Species))
GRUK.species$Species <- sub(".*?_", "", GRUK.species$Species) # lose the Norwegian name in the front
GRUK.species$Species <- gsub("_", " ", GRUK.species$Species) # replace underscore with space
GRUK.species$Species <- str_to_title(GRUK.species$Species) # make first letter capital
GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after hyphon to lowercase
GRUK.species$Species <- gsub("( .*)","\\L\\1",GRUK.species$Species,perl=TRUE) # make capital letters after space to lowercase
GRUK.species[,'Species'] <- word(GRUK.species[,'Species'], 1,2) # lose subspecies

## merge species data with indicators
GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID")], 
                          y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(GRUK.species.ind)

# checking which species didn't find a match
unique(GRUK.species.ind[is.na(GRUK.species.ind$Light & 
                                is.na(GRUK.species.ind$RR)),'Species'])

# fix species name issues
ind.dat <- ind.dat %>% 
  #  mutate(species=str_replace(species,"Aconitum lycoctonum", "Aconitum septentrionale")) %>% 
  #  mutate(species=str_replace(species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  #  mutate(species=str_replace(species,"Carex myosuroides", "Kobresia myosuroides")) %>%
  #  mutate(species=str_replace(species,"Clinopodium acinos", "Acinos arvensis")) %>%
  #  mutate(species=str_replace(species,"Artemisia rupestris", "Artemisia norvegica")) %>%
  mutate(species=str_replace(species,"Rosa vosagica", "Rosa vosagiaca"))

GRUK.species <- GRUK.species %>% 
  mutate(Species=str_replace(Species,"Arabis wahlenbergii", "Arabis hirsuta")) %>%
  #  mutate(Species=str_replace(Species,"Arctous alpinus", "Arctous alpina")) %>%
  #  mutate(Species=str_replace(Species,"Betula tortuosa", "Betula pubescens")) %>%
  #  mutate(Species=str_replace(Species,"Blysmopsis rufa", "Blysmus rufus")) %>%
  #  mutate(Species=str_replace(Species,"Cardamine nymanii", "Cardamine pratensis")) %>%
  #  mutate(Species=str_replace(Species,"Carex adelostoma", "Carex buxbaumii")) %>%
  #  mutate(Species=str_replace(Species,"Carex leersii", "Carex echinata")) %>%
  mutate(Species=str_replace(Species,"Carex paupercula", "Carex magellanica")) %>%
  #  mutate(Species=str_replace(Species,"Carex simpliciuscula", "Kobresia simpliciuscula")) %>%
  mutate(Species=str_replace(Species,"Carex viridula", "Carex flava")) %>%
  #  mutate(Species=str_replace(Species,"Chamaepericlymenum suecicum", "Cornus suecia")) %>%
  #  mutate(Species=str_replace(Species,"Cicerbita alpina", "Lactuca alpina")) %>%
  mutate(Species=str_replace(Species,"Cotoneaster scandinavicus", "Cotoneaster integerrimus")) %>%
  mutate(Species=str_replace(Species,"Cotoneaster symondsii", "Cotoneaster integrifolius")) %>%
  mutate(Species=str_replace(Species,"Cyanus montanus", "Centaurea montana")) %>%
  #  mutate(Species=str_replace(Species,"Empetrum hermaphroditum", "Empetrum nigrum")) %>%
  mutate(Species=str_replace(Species,"Erysimum virgatum", "Erysimum strictum")) %>%
  #  mutate(Species=str_replace(Species,"Festuca prolifera", "Festuca rubra")) %>%
  mutate(Species=str_replace(Species,"Festuca trachyphylla", "Festuca brevipila")) %>%
  mutate(Species=str_replace(Species,"Galium album", "Galium mollugo")) %>%
  #  mutate(Species=str_replace(Species,"Galium elongatum", "Galium palustre")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pratense", "Avenula pratensis")) %>%
  mutate(Species=str_replace(Species,"Helictotrichon pubescens", "Avenula pubescens")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpina", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium alpinum", "Hieracium Alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracium", "Hieracium Hieracium")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium hieracioides", "Hieracium umbellatum")) %>%
  mutate(Species=str_replace(Species,"Hieracium murorum", "Hieracium Vulgata")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium oreadea", "Hieracium Oreadea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium prenanthoidea", "Hieracium Prenanthoidea")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgata", "Hieracium Vulgata")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium pilosella", "Pilosella officinarum")) %>%
  #  mutate(Species=str_replace(Species,"Hieracium vulgatum", "Hieracium umbellatum")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« alpina", "Hierochloë alpina")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« hirta", "Hierochloë hirta")) %>%
  #  mutate(Species=str_replace(Species,"Hierochloã« odorata", "Hierochloë odorata")) %>%
  mutate(Species=str_replace(Species,"Hylotelephium maximum", "Sedum telephium")) %>%
  #  mutate(Species=str_replace(Species,"Listera cordata", "Neottia cordata")) %>%
  #  mutate(Species=str_replace(Species,"Leontodon autumnalis", "Scorzoneroides autumnalis")) %>%
  mutate(Species=str_replace(Species,"Lepidotheca suaveolens", "Matricaria discoidea")) %>%
  #  mutate(Species=str_replace(Species,"Loiseleuria procumbens", "Kalmia procumbens")) %>%
  mutate(Species=str_replace(Species,"Malus ×domestica", "Malus domestica")) %>%
  #  mutate(Species=str_replace(Species,"Mycelis muralis", "Lactuca muralis")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca supina", "Gnaphalium supinum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca norvegica", "Gnaphalium norvegicum")) %>%
  #  mutate(Species=str_replace(Species,"Omalotheca sylvatica", "Gnaphalium sylvaticum")) %>%
  #  mutate(Species=str_replace(Species,"Oreopteris limbosperma", "Thelypteris limbosperma")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus microcarpus", "Vaccinium microcarpum")) %>%
  #  mutate(Species=str_replace(Species,"Oxycoccus palustris", "Vaccinium oxycoccos")) %>%
  #  mutate(Species=str_replace(Species,"Phalaris minor", "Phalaris arundinacea")) %>%
  #  mutate(Species=str_replace(Species,"Pinus unicinata", "Pinus mugo")) %>%
  #  mutate(Species=str_replace(Species,"Poa alpigena", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Poa angustifolia", "Poa pratensis")) %>%
  mutate(Species=str_replace(Species,"Poa humilis", "Poa pratensis")) %>%
  #  mutate(Species=str_replace(Species,"Pyrola grandiflora", "Pyrola rotundifolia")) %>%
  mutate(Species=str_replace(Species,"Rosa dumalis", "Rosa vosagiaca")) %>%
  #  mutate(Species=str_replace(Species,"Rumex alpestris", "Rumex acetosa")) %>%
  #  mutate(Species=str_replace(Species,"Syringa emodi", "Syringa vulgaris")) %>%
  #  mutate(Species=str_replace(Species,"Taraxacum crocea", "Taraxacum officinale")) %>%
  #  mutate(Species=str_replace(Species,"Taraxacum croceum", "Taraxacum officinale")) %>%
  #  mutate(Species=str_replace(Species,"Trientalis europaea", "Lysimachia europaea")) %>%
  mutate(Species=str_replace(Species,"Trifolium pallidum", "Trifolium pratense"))

## merge species data with indicators
GRUK.species.ind <- merge(x=GRUK.species[,c("Species", "art_dekning", "ParentGlobalID")], 
                          y= ind.dat[,c("species","CC", "SS", "RR","Light", "Nitrogen", "Soil_disturbance")],
                          by.x="Species", by.y="species", all.x=T)
summary(GRUK.species.ind)
# checking which species didn't find a match
unique(GRUK.species.ind[is.na(GRUK.species.ind$Light & 
                                is.na(GRUK.species.ind$RR)),'Species'])

### make GRUK.variables into spatial object
names(GRUK.variables)
GRUK.variables <- st_as_sf(GRUK.variables, coords = c("x","y"),remove=F)
# add CRS
GRUK.variables <- st_set_crs(GRUK.variables,4326)
# transform CRS to match ANO
GRUK.variables <- GRUK.variables %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(ANO.geo))

### merge GRUK.variables and GRUK2021.condition
#GRUK.variables <- as.data.frame(GRUK.variables)
GRUK2021.condition <- as.data.frame(GRUK2021.condition)

head(GRUK.variables)
head(GRUK2021.condition)

## fixing site-ID in the condition-data
GRUK2021.condition$Flate_ID <- GRUK2021.condition$områdenavn
unique(as.factor(GRUK2021.condition$Flate_ID))
GRUK2021.condition$Flate_ID <- gsub("_", " ", GRUK2021.condition$Flate_ID) # replace underscore with space
unique(as.factor(GRUK2021.condition$Flate_ID))
GRUK2021.condition$Flate_ID <- sub(" .*", "", GRUK2021.condition$Flate_ID) # remove everything after space
unique(as.factor(GRUK2021.condition$Flate_ID))

#GRUK2021.condition$Flate_ID <- gsub(" ", "-", GRUK2021.condition$Flate_ID) # replace space with hyphon
#GRUK2021.condition$Flate_ID <- gsub("[^0-9\\-]", "", GRUK2021.condition$Flate_ID) # remove all non-numerics except '-'
#GRUK2021.condition$Flate_ID <- gsub("--", "-", GRUK2021.condition$Flate_ID) # remove double hyphons
#GRUK2021.condition$Flate_ID <- substr(GRUK2021.condition$Flate_ID, 1, nchar(GRUK2021.condition$Flate_ID)-1) # remove the last character ('-' for all)

colnames(GRUK.variables)
colnames(GRUK2021.condition)

GRUK.variables <- merge(x=GRUK.variables, 
                        y=GRUK2021.condition[,c("tilstandsvurdering","tilstandsgrunn","artsmangfoldvurdering","lokalitetskvalitet","Flate_ID")], 
                        by.x="Flate_ID", by.y="Flate_ID", all.x=T)

## adding information on ecosystem and condition variables to species data
GRUK.species.ind <- merge(x=GRUK.species.ind, 
                          y=GRUK.variables[,c("GlobalID","year","Flate_ID","Punkt_ID",
                                              "Total dekning % av karplanter registert","Dekning % av karplanter i feltsjikt","Dekning % av moser",
                                              "Dekning % av lav","Dekning % av strø","Dekning % av bar jord/grus/stein/berg",
                                              "Kartleggingsenhet",
                                              "Spor etter ferdsel med tunge kjøretøy (%)","Spor etter slitasje og slitasjebetinget erosjon (%)","Dekning % av nakent berg","Menneskeskapte objekter i sirkelen?",
                                              "Total dekning % av vedplanter i feltsjikt","Dekning % av busker i busksjikt","Dekning % av tresjikt","Dekning % av problemarter","Total dekning % av fremmede arter",
                                              "x","y")], 
                          by.x="ParentGlobalID", by.y="GlobalID", all.x=T)
# trimming away the points without information on NiN, species or cover
GRUK.species.ind <- GRUK.species.ind[!is.na(GRUK.species.ind$Species),]
GRUK.species.ind <- GRUK.species.ind[!is.na(GRUK.species.ind$art_dekning),]

#rm(GRUK.species)
#rm(GRUK.variables)

summary(GRUK.species.ind)
head(GRUK.species.ind)

### reference data - data handling

head(natopen_NiN_ref)
head(natopen_NiN_ref_spInfo)

colnames(natopen_NiN_ref)[1] <- "sp"
head(natopen_NiN_ref)

natopen_NiN_ref <- merge(natopen_NiN_ref,natopen_NiN_ref_spInfo[,c(1,4)], by.x="sp", by.y="ScientificName", all.x=T)
unique(natopen_NiN_ref[is.na(natopen_NiN_ref$Phylum),'sp']) # Pucinella does not exist in ind.dat, so we don't care
# we're only interested in vascular plants and ferns, which we have indicators on
unique(natopen_NiN_ref$Phylum)
natopen_NiN_ref <- natopen_NiN_ref %>%
  filter(Phylum %in% c("Magnoliophyta","Pteridophyta"))
unique(natopen_NiN_ref$Phylum)

natopen_NiN_ref$sp
# only genus and species name
natopen_NiN_ref$sp.orig <- natopen_NiN_ref$sp
natopen_NiN_ref$sp <- word(natopen_NiN_ref$sp, 1,2)
natopen_NiN_ref <- natopen_NiN_ref[!is.na(natopen_NiN_ref$sp),]
# merging with indicator values
NiN.natopen <- merge(natopen_NiN_ref,ind.dat[,c(1,3:5,20,23,27)], by.x="sp", by.y="species", all.x=T)
head(NiN.natopen)
summary(NiN.natopen)

NiN.natopen
unique(NiN.natopen$sp)
#NiN.sp$spgr <- as.factor(as.vector(Eco_State$Concept_Data$Species$Species_List$art.code))

# checking which species didn't find a match
unique(NiN.natopen[is.na(NiN.natopen$Light) & is.na(NiN.natopen$Nitrogen) & is.na(NiN.natopen$RR),'sp'])

# fix species name issues
natopen_NiN_ref <- natopen_NiN_ref %>% 
  mutate(sp=str_replace(sp,"Cirsium acaulon", "Cirsium acaule")) %>%
  mutate(sp=str_replace(sp,"Hylotelephium maximum", "Hylotelephium telephium"))

ind.dat[2556,'species'] <- "Saxifraga osloënsis"
ind.dat[17,'species'] <- "Hierochloë odorata"
ind.dat[9,'species'] <- "Hippophaë rhamnoides"

# merging with indicator values
NiN.natopen <- merge(natopen_NiN_ref,ind.dat[,c(1,3:5,20,23,27)], by.x="sp", by.y="species", all.x=T)
# checking which species didn't find a match
unique(NiN.natopen[is.na(NiN.natopen$Light) & is.na(NiN.natopen$Nitrogen) & is.na(NiN.natopen$RR),'sp'])
# ok now

# translating the abundance classes into %-cover
coverscale <- data.frame(orig=0:6,
                         cov=c(0, 1/32 ,1/8, 3/8, 0.6, 4/5, 1)
)

NiN.natopen.cov <- NiN.natopen
colnames(NiN.natopen.cov)
for (i in 2:71) {
  NiN.natopen.cov[,i] <- coverscale[,2][ match(NiN.natopen[,i], 0:6 ) ]
}

summary(NiN.natopen)
summary(NiN.natopen.cov)

```
leaving us with the monitoring data including plant indicators (ANO.sp.ind, GRUK.species.ind) and the reference data including plant indicators (NiN.natopen.cov)

```{r}
head(ANO.sp.ind)
head(GRUK.species.ind)
head(NiN.natopen.cov)
```
For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
For semi-natural ecosystems, we are testing "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Phosphorus", "Grazing_mowing", and "Soil_disturbance". In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.

##### bootstrap function for frequency abundance
- function to calculate community weighted means of selected indicator values (ind)
- for species lists (sp) with given abundances in percent (or on a scale from 0 to 1) in one or more 'sites' (abun)
- with a given number of iterations (iter),
- with species given a certain minimum abundance occurring in all bootstraps (obl), and
- with a given re-sampling ratio of the original species list (rat)
- in every bootstrap iteration the abundance of the sampled species can be randomly changed by a limited amount if wished by introducing a re-sampling of abundance values from adjacent abundance steps with a certain probability (var.abun)

```{r, echo = F}
indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {

  ind.b <- matrix(nrow=iter,ncol=length(colnames(abun)))
  colnames(ind.b) <- colnames(abun)
  ind.b <- as.data.frame(ind.b)

  ind <- as.data.frame(ind)
  ind.list <- as.list(1:length(colnames(ind)))
  names(ind.list) <- colnames(ind)

  for (k in 1:length(colnames(ind)) ) {
    ind.list[[k]] <- ind.b }

  for (j in 1:length(colnames(abun)) ) {

    dat <- cbind(sp,abun[,j],ind)
    dat <- dat[dat[,2]>0,]            # only species that are present in the ecosystem
    dat <- dat[!is.na(dat[,3]),]      # only species that have indicator values

    for (i in 1:iter) {

      speciesSample <- sample(dat$sp[dat[,2] < obl], size=round( (length(dat$sp)-length(dat$sp[dat[,2]>=obl])) *rat,0), replace=F)  
      dat.b <- rbind(dat[dat[,2] >= obl,],
                     dat[match(speciesSample,dat$sp),]
      )

      if (var.abun==T) {
        for (m in 1:nrow(coverscale[-1,]) ) {
          xxx <- dat.b[dat.b[,2]==coverscale[-1,][m,2],2]
          if ( m==1 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.5, 0.5, 0.0, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==2 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.2, 0.3, 0.5, 0.0, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==3 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.2, 0.3, 0.5, 0.0, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==4 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.2, 0.3, 0.5, 0.0, 0.0) ,size=length(xxx),replace=T) }
          if ( m==5 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.2, 0.3, 0.5, 0.0) ,size=length(xxx),replace=T) }
          if ( m==6 ) { dat.b[dat.b[,2]==coverscale[-1,][m,2],2] <- sample( c(0.01,coverscale[2:7,2]), prob = c(0.0, 0.0, 0.0, 0.0, 0.2, 0.3, 0.5) ,size=length(xxx),replace=T) }
        }
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]<=(0),2] <- 0.01
        dat.b[!is.na(dat.b[,2]) & dat.b[,2]>1,2] <- 1
      }

      for (k in 1:length(colnames(ind))) {

        if ( nrow(dat.b)>2 ) {

          ind.b <- sum(dat.b[!is.na(dat.b[,2+k]),2] * dat.b[!is.na(dat.b[,2+k]),2+k] , na.rm=T) / sum(dat.b[!is.na(dat.b[,2+k]),2],na.rm=T)
          ind.list[[k]][i,j] <- ind.b

        } else {ind.list[[k]][i,j] <- NA}

      }

#      print(paste(i,"",j)) 
    }

  }
  return(ind.list)
}
```

Running the bootstraps

```{r, echo = T, eval = F}
colnames(NiN.natopen)
# 1st column is the species
# 6th-71st column is the abundances of sp in different ecosystem types
# 74th-79th column is the indicator values of the respective species
# we choose 1000 iterations
# species with abundance 1 (i.e. a max of 100%, must be included in each sample)
# each sample re-samples 1/3 of the number of species
# the abundance of the re-sampled species may vary (see bootstrap function for details)
natopen.ref.cov <- indBoot.freq(sp=NiN.natopen.cov[,1],abun=NiN.natopen.cov[,6:71],ind=NiN.natopen.cov[,74:79],iter=1000,obl=1,rat=1/3,var.abun=T)

### fixing NaN's
for (i in 1:length(natopen.ref.cov) ) {
  for (j in 1:ncol(natopen.ref.cov[[i]]) ) {
    v <- natopen.ref.cov[[i]][,j]
    v[is.nan(v)] <- NA
    natopen.ref.cov[[i]][,j] <- v
  }
}
```

```{r, include = F}
# Data from cache
natopen.ref.cov<-readRDS(paste0(here::here(), "/data/cache/natopen.ref.cov.RDS"))
```

```{r}
head(natopen.ref.cov[[1]])
```

This results in an R-list with a slot for every selected indicator, and in every slot there's a data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.
Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:

- Median = reference values
- 0.025 and 0.975 quantiles = lower and upper limit values
- min and max of the respective indicator's scale = min/max values

```{r, echo = F}

# NiN-types where each type is represented by one species list (including when one species list represents two NiN-types)
names(natopen.ref.cov[["Light"]])
x <- 1:66

# checking the actual NiN-types in the natopen lists
natopen.NiNtypes <- colnames(natopen.ref.cov[["Light"]])
natopen.NiNtypes[-x] <- substr(natopen.NiNtypes[-x], 1, nchar(natopen.NiNtypes[-x])-1)
natopen.NiNtypes

# 7 indicator-value indicators: Grimes C, S and R, Tyler's Light, Nitrogen, and Soil_disturbance
indEll.n=6
# creating a table to hold:
# Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type
# for every nature type (nrows)
tab <- matrix(ncol=3*indEll.n, nrow=length(unique(natopen.NiNtypes)) ) # 66 basic ecosystem types
# coercing the values into the table

for (i in 1:length(x) ) {
  tab[i,1:3] <- quantile(as.matrix(natopen.ref.cov[["CC"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,4:6] <- quantile(as.matrix(natopen.ref.cov[["SS"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,7:9] <- quantile(as.matrix(natopen.ref.cov[["RR"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,10:12] <- quantile(as.matrix(natopen.ref.cov[["Light"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,13:15] <- quantile(as.matrix(natopen.ref.cov[["Nitrogen"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)
  tab[i,16:18] <- quantile(as.matrix(natopen.ref.cov[["Soil_disturbance"]][,x[i]]),probs=c(0.025,0.5,0.975),na.rm=T)

}

tab <- as.data.frame(tab)
tab$NiN <- NA
tab$NiN[1:length(x)] <- names(natopen.ref.cov[[1]])[x]
tab

# making it a proper data frame
dim(tab)
round(tab[,1:18],digits=2)

colnames(tab) <- c("CC_q2.5","CC_q50","CC_q97.5",
                   "SS_q2.5","SS_q50","SS_q97.5",
                   "RR_q2.5","RR_q50","RR_q97.5",
                   "Light_q2.5","Light_q50","Light_q97.5",
                   "Nitrogen_q2.5","Nitrogen_q50","Nitrogen_q97.5",
                   "Soil_disturbance_q2.5","Soil_disturbance_q50","Soil_disturbance_q97.5",
                   "NiN")
summary(tab)
tab$NiN
#tab$NiN <- gsub("C", "C-", tab$NiN) # add extra hyphen after C for NiN-types
tab

# restructuring into separate indicators for lower (q2.5) and higher (q97.5) than reference value (=median, q50)
y.Light <- numeric(length=nrow(tab)*2)
y.Light[((1:dim(tab)[1])*2)-1] <- tab$Light_q2.5 
y.Light[((1:dim(tab)[1])*2)] <- tab$Light_q97.5

y.CC <- numeric(length=nrow(tab)*2)
y.CC[((1:dim(tab)[1])*2)-1] <- tab$CC_q2.5 
y.CC[((1:dim(tab)[1])*2)] <- tab$CC_q97.5

y.SS <- numeric(length=nrow(tab)*2)
y.SS[((1:dim(tab)[1])*2)-1] <- tab$SS_q2.5 
y.SS[((1:dim(tab)[1])*2)] <- tab$SS_q97.5

y.Nitrogen <- numeric(length=nrow(tab)*2)
y.Nitrogen[((1:dim(tab)[1])*2)-1] <- tab$Nitrogen_q2.5 
y.Nitrogen[((1:dim(tab)[1])*2)] <- tab$Nitrogen_q97.5

y.RR <- numeric(length=nrow(tab)*2)
y.RR[((1:dim(tab)[1])*2)-1] <- tab$RR_q2.5 
y.RR[((1:dim(tab)[1])*2)] <- tab$RR_q97.5

y.Soil_disturbance <- numeric(length=nrow(tab)*2)
y.Soil_disturbance[((1:dim(tab)[1])*2)-1] <- tab$Soil_disturbance_q2.5 
y.Soil_disturbance[((1:dim(tab)[1])*2)] <- tab$Soil_disturbance_q97.5

# creating final objects holding the reference and limit values for all indicators

# ref object for indicators
natopen.ref.cov.val <- data.frame(N1=rep('natopen',(nrow(tab)*2*indEll.n)),
                              hoved=c(rep('NA',(nrow(tab)*2*indEll.n))),
                              grunn=c(rep(rep(tab$NiN,each=2),indEll.n)),
                              county=rep('all',(nrow(tab)*2*indEll.n)),
                              region=rep('all',(nrow(tab)*2*indEll.n)),
                              Ind=c(rep(c('CC1','CC2'),nrow(tab)),
                                    rep(c('SS1','SS2'),nrow(tab)),
                                    rep(c('RR1','RR2'),nrow(tab)),
                                    rep(c('Light1','Light2'),nrow(tab)),
                                    rep(c('Nitrogen1','Nitrogen2'),nrow(tab)),
                                    rep(c('Soil_disturbance1','Soil_disturbance2'),nrow(tab))
                              ),
                              Rv=c(rep(tab$CC_q50,each=2),
                                   rep(tab$SS_q50,each=2),
                                   rep(tab$RR_q50,each=2),
                                   rep(tab$Light_q50,each=2),
                                   rep(tab$Nitrogen_q50,each=2),
                                   rep(tab$Soil_disturbance_q50,each=2)
                              ),
                              Gv=c(y.CC,y.SS,y.RR,y.Light,y.Nitrogen,y.Soil_disturbance),
                              maxmin=c(rep(c(0,1),nrow(tab)), # CC
                                       rep(c(0,1),nrow(tab)), # SS
                                       rep(c(0,1),nrow(tab)), # RR
                                       rep(c(1,7),nrow(tab)),  # 7 levels of Light
                                       rep(c(1,9),nrow(tab)),  # 9 levels of Nitrogen
                                       rep(c(1,9),nrow(tab))  # 9 levels of Soil_disturbance
                              )
)

natopen.ref.cov.val
natopen.ref.cov.val$grunn <- as.factor(natopen.ref.cov.val$grunn)
natopen.ref.cov.val$Ind <- as.factor(natopen.ref.cov.val$Ind)
summary(natopen.ref.cov.val)

```

```{r}
head(natopen.ref.cov.val)
```

Once test data (ANO, GRUK) and the scaling values from the reference data are in place, we can calculate community-weighted means (CWM) of the selected indicators for the ANO and GRUK community data and scale them against the scaling values from the reference distribution. Note that we scale each ANO/GRUK plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value. Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each  functional plant indicator. An ANO/GRUK plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.

Here is the scaling function
```{r}

#### scaled values ####
r.s <- 1    # reference value
l.s <- 0.6  # limit value
a.s <- 0    # abscence of indicator, or indicator at maximum

#### function for calculating scaled values for measured value ####

## scaling function including truncation
scal <- function() {
  # place to hold the result
   x <- numeric()
  if (maxmin < ref) {
    # values >= the reference value equal 1
    if (val >= ref) {x <- 1}
    # values < the reference value and >= the limit value can be deducted from the linear relationship between these two
    if (val < ref & val >= lim) {x <- (l.s + (val-lim) * ( (r.s-l.s) / (ref-lim) ) )}
    # values < the limit value and > maxmin can be deducted from the linear relationship between these two
    if (val < lim & val > maxmin) {x <- (a.s + (val-maxmin) * ( (l.s-a.s) / (lim-maxmin) ) )}
    # value equals or lower than maxmin
    if (val <= maxmin) {x <-0}
  } else {
    # values <= the reference value equal 1
    if (val <= ref) {x <- 1}
    # values > the reference value and <= the limit value can be deducted from the linear relationship between these two
    if (val > ref & val <= lim) {x <- ( r.s - ( (r.s - l.s) * (val - ref) / (lim - ref) ) )}
    # values > the limit value and < maxmin can be deducted from the linear relationship between these two
    if (val > lim) {x <- ( l.s - (l.s * (val - lim) / (maxmin - lim) ) )}
    # value equals or larger than maxmin
    if (val >= maxmin) {x <-0}
  }
  return(x)

}

```

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
This is done separately for ANO and ASO. First for ANO:
```{r, include=FALSE}

unique(ANO.geo$hovedtype_rute) # NiN types in data
unique(substr(natopen.ref.cov.val$grunn,1,3)) # NiN types in reference
#### creating dataframe to hold the results for natopens ####
# all ANO points
nrow(ANO.geo)
# all natopen ANO points
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T2","T8","T11","T12","T13","T15","T16","T18","T21","T24","T29") & ANO.geo$hovedoekosystem_punkt!='fjell',])
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T2") & ANO.geo$hovedoekosystem_punkt!='fjell',])
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T13") & ANO.geo$hovedoekosystem_punkt!='fjell',])
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% list("T16") & ANO.geo$hovedoekosystem_punkt!='fjell',])
# all natopen ANO points with a NiN-type represented in the reference
nrow(ANO.geo[ANO.geo$hovedtype_rute %in% unique(gsub("-", "",substr(natopen.ref.cov.val$grunn,1,3))) & ANO.geo$hovedoekosystem_punkt!='fjell',])
# leaving us with a total of 272 ANO-points with naturally open ecosystems that are not in the mountains
ANO.natopen <- ANO.geo[ANO.geo$hovedtype_rute %in% unique(gsub("-", "",substr(natopen.ref.cov.val$grunn,1,3))) & ANO.geo$hovedoekosystem_punkt!='fjell',]
ANO.natopen <- ANO.natopen[!is.na(ANO.natopen$kartleggingsenhet_1m2),]
head(ANO.natopen)
# update row-numbers
row.names(ANO.natopen) <- 1:nrow(ANO.natopen)
head(ANO.natopen)
dim(ANO.natopen)
colnames(ANO.natopen)

length(levels(as.factor(ANO.natopen$ano_flate_id)))
length(levels(as.factor(ANO.natopen$ano_punkt_id)))
summary(as.factor(ANO.natopen$ano_punkt_id))
# one point is double
ANO.natopen[ANO.natopen$ano_punkt_id=="ANO0084_51",] # double registration of # 51
ANO.geo[ANO.geo$ano_punkt_id=="ANO0084_51",] # double registration of # 51

ANO.geo[ANO.geo$ano_flate_id=="ANO0084","ano_punkt_id"] # also # 62 is registered double
ANO.geo[ANO.geo$ano_punkt_id=="ANO0084_62",] # but this is not a natopen-type
# seems to be a legit registration of a naturally open type on both ANO-points # 51.
# Both points # 22, 35, 44, and 53 are missing, so we simply assign the 2nd # 51 to # 53 for this analysis
ANO.natopen[ANO.natopen$ano_punkt_id=="ANO0084_51" & !is.na(ANO.natopen$ano_punkt_id),][2,"ano_punkt_id"] <- "ANO0084_53"
ANO.natopen[ANO.natopen$ano_flate_id=="ANO0084","ano_punkt_id"]
summary(as.factor(ANO.natopen$ano_punkt_id))

#ANO.natopen <- ANO.natopen[-206,]
#row.names(ANO.natopen) <- 1:nrow(ANO.natopen) # update row-numbers

unique(ANO.natopen$hovedoekosystem_punkt)
unique(ANO.natopen$hovedtype_rute)
unique(ANO.natopen$kartleggingsenhet_1m2)
ANO.natopen$hovedtype_rute <- as.factor(ANO.natopen$hovedtype_rute)
ANO.natopen$kartleggingsenhet_1m2 <- as.factor(ANO.natopen$kartleggingsenhet_1m2)
summary(ANO.natopen$hovedtype_rute)
summary(ANO.natopen$kartleggingsenhet_1m2)
# we have no reference for overall T2 registrations, these 24 observations won't be evaluated
ANO.natopen[ANO.natopen$kartleggingsenhet_1m2=="T2",] # were largely inaccessible and thus mapped from arial pictures

results.natopen.ANO <- list()
ind <- unique(natopen.ref.cov.val$Ind)
# choose columns for site description
colnames(ANO.natopen)
results.natopen.ANO[['original']] <- ANO.natopen
# drop geometry
st_geometry(results.natopen.ANO[['original']]) <- NULL
results.natopen.ANO[['original']]

# add columns for indicators
nvar.site <- ncol(results.natopen.ANO[['original']])
for (i in 1:length(ind) ) {results.natopen.ANO[['original']][,i+nvar.site] <- NA}
colnames(results.natopen.ANO[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.natopen.ANO[['original']][,i] <- as.numeric(results.natopen.ANO[['original']][,i])}
summary(results.natopen.ANO[['original']])
#results.natopen.ANO[['original']]$Region <- as.factor(results.natopen.ANO[['original']]$Region)
results.natopen.ANO[['original']]$GlobalID <- as.factor(results.natopen.ANO[['original']]$GlobalID)
results.natopen.ANO[['original']]$ano_flate_id <- as.factor(results.natopen.ANO[['original']]$ano_flate_id)
results.natopen.ANO[['original']]$ano_punkt_id <- as.factor(results.natopen.ANO[['original']]$ano_punkt_id)
results.natopen.ANO[['original']]$hovedoekosystem_punkt <- as.factor(results.natopen.ANO[['original']]$hovedoekosystem_punkt)
#results.natopen.ANO[['original']]$Hovedoekosystem_rute  <- as.factor(results.natopen.ANO[['original']]$Hovedoekosystem_rute )
results.natopen.ANO[['original']]$kartleggingsenhet_1m2 <- as.factor(results.natopen.ANO[['original']]$kartleggingsenhet_1m2)
results.natopen.ANO[['original']]$hovedtype_rute    <- as.factor(results.natopen.ANO[['original']]$hovedtype_rute)

# roll out
results.natopen.ANO[['scaled']] <- results.natopen.ANO[['non-truncated']] <- results.natopen.ANO[['original']]

```

```{r, echo = T, eval = F}

#### calculating scaled and non-truncated values for the indicators based on the dataset ####
for (i in 1:nrow(ANO.natopen) ) {  #
  tryCatch({
    print(i)
    print(paste(ANO.natopen$ano_flate_id[i]))
    print(paste(ANO.natopen$ano_punkt_id[i]))
#    ANO.natopen$Hovedoekosystem_sirkel[i]
#    ANO.natopen$Hovedoekosystem_rute[i]

    # if the ANO.hovedtype exists in the reference
    if (ANO.natopen$hovedtype_rute[i] %in% gsub("-","",unique(substr(natopen.ref.cov.val$grunn,1,3))) ) {

      # if there is any species present in current ANO point  
      if ( length(ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),'Species']) > 0 ) {

        # Grime's C

        dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','CC')]
        results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$CC),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'CC'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'CC1'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'CC1'] <- scal.2() 
          results.natopen.ANO[['original']][i,'CC1'] <- val

          # upper part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }          
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'CC2'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'CC2'] <- scal.2() 
          results.natopen.ANO[['original']][i,'CC2'] <- val
        }

        # Grime's S
        dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','SS')]
        results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$SS),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'SS'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }          
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'SS1'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'SS1'] <- scal.2() 
          results.natopen.ANO[['original']][i,'SS1'] <- val

          # upper part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }          
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'SS2'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'SS2'] <- scal.2() 
          results.natopen.ANO[['original']][i,'SS2'] <- val
        }

        # Grime's R
        dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','RR')]
        results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$RR),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'RR'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }          
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'RR1'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'RR1'] <- scal.2() 
          results.natopen.ANO[['original']][i,'RR1'] <- val

          # upper part of distribution
          if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
          } else {
            ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
            lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
          }          
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
          # coercing x into results.natopen.ANO dataframe
          results.natopen.ANO[['scaled']][i,'RR2'] <- scal() 
          results.natopen.ANO[['non-truncated']][i,'RR2'] <- scal.2() 
          results.natopen.ANO[['original']][i,'RR2'] <- val
        }

          # Light
          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','Light')]
          results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
          dat <- dat[!is.na(dat$Light),]

          if ( nrow(dat)>0 ) {

            val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
            # lower part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }            
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Light1'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Light1'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Light1'] <- val

            # upper part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }            
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Light2'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Light2'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Light2'] <- val
          }

          # Nitrogen
          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','Nitrogen')]
          results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
          dat <- dat[!is.na(dat$Nitrogen),]

          if ( nrow(dat)>0 ) {

            val <- sum(dat[,'art_dekning'] * dat[,'Nitrogen'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
            # lower part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }            
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Nitrogen1'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Nitrogen1'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Nitrogen1'] <- val

            # upper part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }            
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Nitrogen2'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Nitrogen2'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Nitrogen2'] <- val
          }

          # Soil_disturbance
          dat <- ANO.sp.ind[ANO.sp.ind$ParentGlobalID==as.character(ANO.natopen$GlobalID[i]),c('art_dekning','Soil_disturbance')]
          results.natopen.ANO[['original']][i,'richness'] <- nrow(dat)
          dat <- dat[!is.na(dat$Soil_disturbance),]

          if ( nrow(dat)>0 ) {

            val <- sum(dat[,'art_dekning'] * dat[,'Soil_disturbance'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
            # lower part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }           
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Soil_disturbance1'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Soil_disturbance1'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Soil_disturbance1'] <- val

            # upper part of distribution
            if( ANO.natopen$kartleggingsenhet_1m2[i] %in% c("T2-C-7","T2-C-8") ) {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),"_BN",sep=""),'Gv']
            } else {
              ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Rv']
              lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'Gv']
            }
            maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==as.character(results.natopen.ANO[['original']][i,"kartleggingsenhet_1m2"]),'maxmin']
            # coercing x into results.natopen.ANO dataframe
            results.natopen.ANO[['scaled']][i,'Soil_disturbance2'] <- scal() 
            results.natopen.ANO[['non-truncated']][i,'Soil_disturbance2'] <- scal.2() 
            results.natopen.ANO[['original']][i,'Soil_disturbance2'] <- val
          }

        }
      }

  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# for using both sides of the plant indicators
results.natopen.ANO[['2-sided']] <- results.natopen.ANO[['non-truncated']]

# remove values >1 for 2-sided indicators
results.natopen.ANO[['2-sided']]$CC1[results.natopen.ANO[['2-sided']]$CC1>1] <- NA
results.natopen.ANO[['2-sided']]$CC2[results.natopen.ANO[['2-sided']]$CC2>1] <- NA

results.natopen.ANO[['2-sided']]$SS1[results.natopen.ANO[['2-sided']]$SS1>1] <- NA
results.natopen.ANO[['2-sided']]$SS2[results.natopen.ANO[['2-sided']]$SS2>1] <- NA

results.natopen.ANO[['2-sided']]$RR1[results.natopen.ANO[['2-sided']]$RR1>1] <- NA
results.natopen.ANO[['2-sided']]$RR2[results.natopen.ANO[['2-sided']]$RR2>1] <- NA

results.natopen.ANO[['2-sided']]$Light1[results.natopen.ANO[['2-sided']]$Light1>1] <- NA
results.natopen.ANO[['2-sided']]$Light2[results.natopen.ANO[['2-sided']]$Light2>1] <- NA

results.natopen.ANO[['2-sided']]$Nitrogen1[results.natopen.ANO[['2-sided']]$Nitrogen1>1] <- NA
results.natopen.ANO[['2-sided']]$Nitrogen2[results.natopen.ANO[['2-sided']]$Nitrogen2>1] <- NA

results.natopen.ANO[['2-sided']]$Soil_disturbance1[results.natopen.ANO[['2-sided']]$Soil_disturbance1>1] <- NA
results.natopen.ANO[['2-sided']]$Soil_disturbance2[results.natopen.ANO[['2-sided']]$Soil_disturbance2>1] <- NA

```

and for GRUK:
```{r, include=FALSE}

colnames(GRUK.variables)
levels(as.factor(GRUK.variables$Kartleggingsenhet)) # NiN types in data
levels(natopen.ref.cov.val$grunn) # NiN types in reference
#### creating dataframe to hold the results for natopens ####
# all GRUK points
nrow(GRUK.variables)
GRUK.natopen <- GRUK.variables

head(GRUK.natopen)
# update row-numbers
row.names(GRUK.natopen) <- 1:nrow(GRUK.natopen)
head(GRUK.natopen)
dim(GRUK.natopen)
colnames(GRUK.natopen)

length(levels(as.factor(GRUK.natopen$Flate_ID)))
length(levels(as.factor(GRUK.natopen$uPunkt_ID)))
summary(as.factor(GRUK.natopen$uPunkt_ID))
# none are double

unique(GRUK.natopen$Kartleggingsenhet)
GRUK.natopen$Kartleggingsenhet <- as.factor(GRUK.natopen$Kartleggingsenhet)
summary(GRUK.natopen$Kartleggingsenhet)

results.natopen.GRUK <- list()
ind <- unique(natopen.ref.cov.val$Ind)
# choose columns for site description
colnames(GRUK.natopen)
results.natopen.GRUK[['original']] <- GRUK.natopen
# drop geometry
st_geometry(results.natopen.GRUK[['original']]) <- NULL
results.natopen.GRUK[['original']] <- as.data.frame(results.natopen.GRUK[['original']])

# add columns for indicators
nvar.site <- ncol(results.natopen.GRUK[['original']])
for (i in 1:length(ind) ) {results.natopen.GRUK[['original']][,i+nvar.site] <- NA}
colnames(results.natopen.GRUK[['original']])[(nvar.site+1):(length(ind)+nvar.site)] <- paste(ind)
for (i in (nvar.site+1):(length(ind)+nvar.site) ) {results.natopen.GRUK[['original']][,i] <- as.numeric(results.natopen.GRUK[['original']][,i])}
summary(results.natopen.GRUK[['original']])
#results.natopen.GRUK[['original']]$Region <- as.factor(results.natopen.GRUK[['original']]$Region)
results.natopen.GRUK[['original']]$GlobalID <- as.factor(results.natopen.GRUK[['original']]$GlobalID)
results.natopen.GRUK[['original']]$Flate_ID <- as.factor(results.natopen.GRUK[['original']]$Flate_ID)
results.natopen.GRUK[['original']]$uPunkt_ID <- as.factor(results.natopen.GRUK[['original']]$uPunkt_ID)
results.natopen.GRUK[['original']]$Kartleggingsenhet <- as.factor(results.natopen.GRUK[['original']]$Kartleggingsenhet)

# roll out
results.natopen.GRUK[['scaled']] <- results.natopen.GRUK[['non-truncated']] <- results.natopen.GRUK[['original']]

```

```{r, echo = T, eval = F}

#### calculating scaled and non-truncated values for the indicators based on the dataset ####
for (i in 1:nrow(GRUK.natopen) ) {  #
  tryCatch({
    print(i)
    print(paste(GRUK.natopen$Flate_ID[i]))
    print(paste(GRUK.natopen$uPunkt_ID[i]))
    #    GRUK.natopen$Hovedoekosystem_sirkel[i]
    #    GRUK.natopen$Hovedoekosystem_rute[i]

    # if the GRUK.hovedtype exists in the reference
#    if (GRUK.natopen$hovedtype_rute[i] %in% unique(substr(natopen.ref.cov.val$grunn,1,2)) ) {

      # if there is any species present in current GRUK point  
      if ( length(GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),'Species']) > 0 ) {

        # Grime's C

        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','CC')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$CC),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'CC'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'CC1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'CC1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'CC1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='CC2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'CC2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'CC2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'CC2'] <- val

        }

        # Grime's S
        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','SS')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$SS),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'SS'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'SS1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'SS1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'SS1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='SS2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'SS2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'SS2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'SS2'] <- val

        }

        # Grime's R
        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','RR')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$RR),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'RR'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'RR1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'RR1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'RR1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='RR2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'RR2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'RR2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'RR2'] <- val

        }

        # Light
        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Light')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$Light),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'Light'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Light1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Light1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Light1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Light2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Light2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Light2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Light2'] <- val

        }

        # Nitrogen
        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Nitrogen')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$Nitrogen),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'Nitrogen'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Nitrogen1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Nitrogen1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Nitrogen1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Nitrogen2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Nitrogen2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Nitrogen2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Nitrogen2'] <- val

        }

        # Soil_disturbance
        dat <- GRUK.species.ind[GRUK.species.ind$ParentGlobalID==as.character(GRUK.natopen$GlobalID[i]),c('art_dekning','Soil_disturbance')]
        results.natopen.GRUK[['original']][i,'richness'] <- nrow(dat)
        dat <- dat[!is.na(dat$Soil_disturbance),]

        if ( nrow(dat)>0 ) {

          val <- sum(dat[,'art_dekning'] * dat[,'Soil_disturbance'],na.rm=T) / sum(dat[,'art_dekning'],na.rm=T)
          # lower part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance1' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Soil_disturbance1'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Soil_disturbance1'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Soil_disturbance1'] <- val

          # upper part of distribution
          ref <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Rv']
          lim <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==paste(as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),"_BN",sep=""),'Gv']
          maxmin <- natopen.ref.cov.val[natopen.ref.cov.val$Ind=='Soil_disturbance2' & natopen.ref.cov.val$grunn==as.character(results.natopen.GRUK[['original']][i,"Kartleggingsenhet"]),'maxmin']
          # coercing x into results.natopen.GRUK dataframe
          results.natopen.GRUK[['scaled']][i,'Soil_disturbance2'] <- scal() 
          results.natopen.GRUK[['non-truncated']][i,'Soil_disturbance2'] <- scal.2() 
          results.natopen.GRUK[['original']][i,'Soil_disturbance2'] <- val

        }

      }
#    }

  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
}

# for using both sides of the plant indicators
results.natopen.GRUK[['2-sided']] <- results.natopen.GRUK[['non-truncated']]

# remove values >1 for 2-sided indicators
results.natopen.GRUK[['2-sided']]$CC1[results.natopen.GRUK[['2-sided']]$CC1>1] <- NA
results.natopen.GRUK[['2-sided']]$CC2[results.natopen.GRUK[['2-sided']]$CC2>1] <- NA

results.natopen.GRUK[['2-sided']]$SS1[results.natopen.GRUK[['2-sided']]$SS1>1] <- NA
results.natopen.GRUK[['2-sided']]$SS2[results.natopen.GRUK[['2-sided']]$SS2>1] <- NA

results.natopen.GRUK[['2-sided']]$RR1[results.natopen.GRUK[['2-sided']]$RR1>1] <- NA
results.natopen.GRUK[['2-sided']]$RR2[results.natopen.GRUK[['2-sided']]$RR2>1] <- NA

results.natopen.GRUK[['2-sided']]$Light1[results.natopen.GRUK[['2-sided']]$Light1>1] <- NA
results.natopen.GRUK[['2-sided']]$Light2[results.natopen.GRUK[['2-sided']]$Light2>1] <- NA

results.natopen.GRUK[['2-sided']]$Nitrogen1[results.natopen.GRUK[['2-sided']]$Nitrogen1>1] <- NA
results.natopen.GRUK[['2-sided']]$Nitrogen2[results.natopen.GRUK[['2-sided']]$Nitrogen2>1] <- NA

results.natopen.GRUK[['2-sided']]$Soil_disturbance1[results.natopen.GRUK[['2-sided']]$Soil_disturbance1>1] <- NA
results.natopen.GRUK[['2-sided']]$Soil_disturbance2[results.natopen.GRUK[['2-sided']]$Soil_disturbance2>1] <- NA

```

```{r, include = F}

# Use cache data
results.natopen.ANO<-readRDS(paste0(here::here(),"/data/cache/results.natopen.ANO.RDS"))
results.natopen.GRUK<-readRDS(paste0(here::here(),"/data/cache/results.natopen.GRUK.RDS"))
```

```{r}
head(results.natopen.ANO[['2-sided']])
head(results.natopen.GRUK[['2-sided']])
```
#### Scaled value analyses

In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).

```{r, message = FALSE}
#### plotting scaled values by main ecosystem type ####
## continuing with 2-sided
res.natopen.ANO <- results.natopen.ANO[['2-sided']]

# make long version of the scaled value part
res.natopen.ANO <-
  res.natopen.ANO %>% 
  pivot_longer(
    cols = c("CC1","CC2",
             "SS1","SS2",
             "RR1","RR2",
             "Light1","Light2",
             "Nitrogen1","Nitrogen2",
             "Soil_disturbance1","Soil_disturbance2"),
    names_to = "fp_ind",
    values_to = "scaled_value",
    values_drop_na = FALSE
  )

# add original values as well
res.natopen.ANO <- 
  res.natopen.ANO %>% add_column(original = results.natopen.ANO[['original']] %>% 
                               pivot_longer(
                                 cols = c("CC1","CC2",
                                          "SS1","SS2",
                                          "RR1","RR2",
                                          "Light1","Light2",
                                          "Nitrogen1","Nitrogen2",
                                          "Soil_disturbance1","Soil_disturbance2"),
                                 names_to = NULL,
                                 values_to = "original",
                                 values_drop_na = FALSE
                               ) %>%
                               pull(original)
  )

# similarly for GRUK
res.natopen.GRUK <- results.natopen.GRUK[['2-sided']]

# make long version of the scaled value part
res.natopen.GRUK <-
  res.natopen.GRUK %>% 
  pivot_longer(
    cols = c("CC1","CC2",
             "SS1","SS2",
             "RR1","RR2",
             "Light1","Light2",
             "Nitrogen1","Nitrogen2",
             "Soil_disturbance1","Soil_disturbance2"),
    names_to = "fp_ind",
    values_to = "scaled_value",
    values_drop_na = FALSE
  )

# add original values as well
res.natopen.GRUK <- 
  res.natopen.GRUK %>% add_column(original = results.natopen.GRUK[['original']] %>% 
                                   pivot_longer(
                                     cols = c("CC1","CC2",
                                              "SS1","SS2",
                                              "RR1","RR2",
                                              "Light1","Light2",
                                              "Nitrogen1","Nitrogen2",
                                              "Soil_disturbance1","Soil_disturbance2"),
                                     names_to = NULL,
                                     values_to = "original",
                                     values_drop_na = FALSE
                                   ) %>%
                                   pull(original)
  )

```

### Ecosystem sub-types
And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
```{r,echo=FALSE,message=FALSE, warning=FALSE}

# making the plot, ANO, CSR-indicators
ggplot(data=subset(res.natopen.ANO,!is.na(scaled_value) & fp_ind %in% c("CC1","SS1","RR1",
                                                                        "CC2","SS2","RR2")), 
       aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin(color=NA) +
  #  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=1, shape=16, color="black") +
  facet_wrap(~factor(fp_ind,levels=c("CC1","SS1","RR1",
                                     "CC2","SS2","RR2"),
                     labels=c("CSR-C1","CSR-S1","CSR-R1",
                                     "CSR-C2","CSR-S2","CSR-R2")
                     ), ncol = 3) + 
  xlab("Basic ecosystem type") + 
  ylab("Scaled indicator value (ANO data, CSR indicators)") +
  scale_fill_discrete(labels=c("CSR-C1","CSR-S1","CSR-R1",
                                     "CSR-C2","CSR-S2","CSR-R2")) + 
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))

# making the plot, ANO, Tyler-indicators
ggplot(data=subset(res.natopen.ANO,!is.na(scaled_value) & fp_ind %in% c("Light1","Nitrogen1","Soil_disturbance1",
                                                                        "Light2","Nitrogen2","Soil_disturbance2")), 
       aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin(color=NA) +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=1, shape=16, color="black") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Nitrogen1","Soil_disturbance1",
                                     "Light2","Nitrogen2","Soil_disturbance2")), ncol = 3) + 
  xlab("Basic ecosystem type") + 
  ylab("Scaled indicator value (ANO data, Tyler indicators)") + 
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))

# making the plot, GRUK
ggplot(res.natopen.GRUK, aes(x=factor(Kartleggingsenhet), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin(color = NA) +
  #  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=1, shape=16, color="black") +
  facet_wrap(~factor(fp_ind,levels=c("CC1","SS1","RR1","Light1","Nitrogen1","Soil_disturbance1",
                                     "CC2","SS2","RR2","Light2","Nitrogen2","Soil_disturbance2"),
                     labels=c("CSR-C1","CSR-S1","CSR-R1","Light1","Nitrogen1","Soil_disturbance1",
                                     "CSR-C2","CSR-S2","CSR-R2","Light2","Nitrogen2","Soil_disturbance2")), ncol = 6) + 
  xlab("Basic ecosystem type") + 
  ylab("Scaled indicator value (GRUK data, all indicators)") + 
  theme(legend.position="none") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))

```
The ANO results show a high frequency of too low CSR-R values (CSR-R1) and Soil disturbance while also showing signs of higher than expected competition (CSR-C2), which  indicates that a number of naturally open areas are changing in their dominance structure and physical stability. The GRUK data show a generally large spread of scaled values, but the violin plots (the shapes indicate the relative amount of observations across the y-axis) suggest that there is a number of sites with too high nitrogen (Nitrogen2), and some sites with too much disturbance (CSR-R2) and light (Light2). Many GRUK sites are popular outdoor locations with the local population, which could increase disturbance and thus cause deviations towards high CSR-R values (CSR-R2).
Increased nitrogen may be an effect of alien species, which the GRUK data records as a major pressure and problem for many monitoring sites.

```{r}
ggplot(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",], aes(x=fremmedartsdekning, y=scaled_value)) +
  geom_point() +
  xlab("Alien species cover (%)") + ylab("Scaled Nitrogen2 value (GRUK data)")
```
And indeed, we can see that very high alien species cover predicts poor scores in the Nitrogen2 indicator.

### Indicator index maps
We can also show the results as a map, for instance for CSR-R1 (the lower ruderal/disturbance indicator) for ANO and Nitrogen2 (the upper nitrogen indicator) for GRUK, either by directly plotting the data onto the map...
```{r, echo=FALSE, warning=FALSE,message = FALSE}
#### scaled value maps ####
# keep wide format and add geometry again
res.natopen.ANO2 <- results.natopen.ANO[['2-sided']]
st_geometry(res.natopen.ANO2) <- st_geometry(ANO.natopen)

res.natopen.GRUK2 <- results.natopen.GRUK[['2-sided']]
st_geometry(res.natopen.GRUK2) <- st_geometry(GRUK.natopen)

nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(ANO.geo))

reg <- st_read("data/regions.shp")%>%
  st_as_sf() %>%
  st_transform(crs = st_crs(ANO.geo))

# change region names to something R-friendly
#reg$region
reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")

regnor <- st_intersection(reg,nor)

## scaled value maps
# ANO, RR1 (lower indicator)
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(res.natopen.ANO2) +
  tm_dots('RR1',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=2, legend.show = FALSE) + # 
  tm_layout(main.title = "Ruderal index (lower), natopen ANO",legend.position = c("right", "bottom"), main.title.size=1.2) + 
  tm_add_legend(type = "fill", 
                col = c(tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),'grey'),
                labels = c("0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", "NA"),
                title = "index values")

# GRUK, Nitrogen2 (upper indicator)
tm_shape(regnor) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Nitrogen2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=2, legend.show = FALSE) + # 
  tm_layout(main.title = "Nitrogen index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2) + 
  tm_add_legend(type = "fill", 
                col = c(tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE),'grey'),
                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", "NA"),
                title = "index values")

```
For GRUK we can zoom in on the area around the Oslofjord...

```{r}
boks <- st_bbox(c(xmin = 10.3, xmax = 10.8, ymax = 59.95, ymin = 59.4), crs = st_crs(4326))

tm_shape(regnor,bbox=boks) +
  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
  tm_borders() +
  tm_shape(res.natopen.GRUK2) +
  tm_dots('Nitrogen2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=3, legend.show = FALSE) + # 
  tm_layout(main.title = "Nitrogen index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2) + 
  tm_add_legend(type = "fill", 
                col = c(tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE),'grey'),
                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", "NA"),
                title = "index values")
```
...but for ANO the colors and values of the data points on the map are hard to make out.

#### continue here ####

### Regions - maps and statistics
Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because:
(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
(ii) both the ANO and the ASO datasets have a nested structure

Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
Here, we apply the following function using either a glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
```{r}
library(betareg)
library(glmmTMB)

expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale

# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
# if that is not the case, then the function performs a betareg if theres >= 2 observations
# if that is not the case either, then the function returns the value of the single observation with NA standard error

indmean.beta <- function(df) {

  st_geometry(df) <- NULL
  colnames(df) <- c("y","ran")

  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {

    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {

      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)

      return(c(
        expit(summary( mod1 )$coefficients$cond[1]),

        expit( summary( mod1 )$coefficients$cond[1] + 
                 summary( mod1 )$coefficients$cond[2] )-
          expit( summary( mod1 )$coefficients$cond[1] ),

        nrow(df[!is.na(df$y),]),
        summary( mod1 )$coefficients$cond[1],
        summary( mod1 )$coefficients$cond[2]
      ))

    } else {

      mod2 <- betareg(y ~ 1, data=df)

      return(c(
        expit(summary( mod2 )$coefficients$mean[1]),
        expit( summary( mod2 )$coefficients$mean[1] + 
                 summary( mod2 )$coefficients$mean[2] )-
          expit( summary( mod2 )$coefficients$mean[1] ),
        nrow(df[!is.na(df$y),]),
        summary( mod2 )$coefficients$mean[1],
        summary( mod2 )$coefficients$mean[2]
      ))

    }

  } else {

    return(c(df$y,NA,1,NA,NA))

  }

}

```

```{r}

res.seminat.ANO2 = st_join(res.seminat.ANO2, regnor, left = TRUE)
res.seminat.ASO2 = st_join(res.seminat.ASO2, regnor, left = TRUE)

regnor <- regnor %>%
  mutate(
    Grazing_mowing1.ANO.reg.mean = c(indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[1],
                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[1]
                                 ),
    Grazing_mowing1.ANO.reg.se = c(indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[2],
                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[2]
                               ),
    Grazing_mowing1.ANO.reg.n = c(nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),])
                              )
  )

# and adding the values for ASO
regnor <- regnor %>%
  mutate(
    Nitrogen2.ASO.reg.mean = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1]
                               ),
    Nitrogen2.ASO.reg.se = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2]
                             ),
    Nitrogen2.ASO.reg.n = c(nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]))
  )

## scaled value maps for Grazing_mowing1 (lower indicator), ANO
# mean
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.ANO.reg.mean", title="Grazing_mowing (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
  tm_text("Grazing_mowing1.ANO.reg.n",col="black",bg.color="grey")
```
Mean index value by region for the upper Grazing_mowing indicator (i.e. index shows deviations towards less grazing or mowing) from the ANO monitoring data. Numbers in grey fields show the number of observations in the respective region.

```{r}
# se
tm_shape(regnor) +
  tm_polygons(col="Grazing_mowing1.ANO.reg.se", title="Grazing_mowing (lower), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Grazing_mowing1.ANO.reg.n",col="black",bg.color="grey")

```
Standard error to the mean index value by region for the lower Grazing_mowing indicator from the ANO monitoring data. Numbers in grey fields show the number of observations in the respective region.

And here are the corresponding maps for the upper Nitrogen indicator in the ASO data:
```{r}
## scaled value maps for Nitrogen2 (upper indicator), ASO
# mean
tm_shape(regnor) +
  tm_polygons(col="Nitrogen2.ASO.reg.mean", title="Nitrogen (upper), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")
```
Mean index value by region for the upper Nitrogen indicator (i.e. index shows deviations towards increased Nitrogen availability) from the ASO monitoring data. Numbers in grey fields show the number of observations in the respective region.

```{r}
# se
tm_shape(regnor) +
  tm_polygons(col="Nitrogen2.ASO.reg.se", title="Nitrogen (upper), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")

```
Standard error to the mean index value by region for the upper Nitrogen indicator from the ASO monitoring data. Numbers in grey fields show the number of observations in the respective region.

#### Unscaled values vs. reference
We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation. Since Grazing_mowing for ANO and nitrogen for ASO show some deviations  we exemplify this with these indicators for unscaled values.

Grazing_mowing, ANO:
```{r}
#summary(res.seminat$kartleggingsenhet_1m2)
#length(unique(res.seminat$kartleggingsenhet_1m2))
# 19 NiN-types, of which T41-C-1 has 0 observations, and T32 & T34 as main types don't have a reference -> so, 16 NiN-Types to plot
#colnames(seminat.ref.cov[['Light']])

### Grazing_mowing
par(mfrow=c(2,4))
## T32s
# T32-C-1
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C1',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original)),
       col="red")

# T32-C-2
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C2',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original)),
       col="red")

# T32-C-3
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C3C4"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C3',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original)),
       col="red")

# T32-C-4
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C3C4"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C4',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original)),
       col="red")

# T32-C-5
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C5',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original)),
       col="red")

# T32-C-6
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C21C6a","T32-C21C6b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C6',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original)),
       col="red")

# T32-C-9
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C9a","T32-C9b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C9',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original)),
       col="red")

# T32-C-10
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C10a","T32-C10b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C10',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original)),
       col="red")

par(mfrow=c(2,4))
# T32-C-15
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C15")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C15',xlab='Grazing_mowing value')
#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original)),
       col="red")

# T32-C-20
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C20',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original)),
       col="red")

## V10s
# V10-C-1
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"V10-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='V10-C1',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original)),
       col="red")

# V10-C-3
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"V10-C3"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='V10-C3',xlab='Grazing_mowing value')
#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original)),
       col="red")

## T34s
# T34-C-1
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T34-C1"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T34-C1',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original)),
       col="red")

# T34-C-2
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T34-C2a","T34-C2b","T34-C2c")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T34-C2',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original)),
       col="red")

# T34-C-4
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T34-C4a","T34-C4b","T34-C4c","T34-C4d")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T34-C4',xlab='Grazing_mowing value')
#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original)),
       col="red")

## T41
plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T41a","T41b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T41',xlab='Grazing_mowing value')
points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original,
       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original)),
       col="red")
legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))

```

Nitrogen, ASO
```{r}

#summary(res.seminat.ASO$NiN_grunntype)
#length(levels(res.seminat.ASO$NiN_grunntype))
# 12 NiN-types, of which C-13 and C-2 have 0 observations -> so, 10 NiN-Types to plot
#colnames(seminat.ref.cov[['Nitrogen']])

### Nitrogen
par(mfrow=c(3,4))
# T32-C-1
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C1C2"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C1',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original)),
       col="red")

# T32-C-3
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C3C4"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C3',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original)),
       col="red")

# T32-C-4
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C3C4"]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C4',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original)),
       col="red")

# T32-C-5
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C5',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original)),
       col="red")

# T32-C-6
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C21C6a","T32-C21C6b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C6',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original)),
       col="red")

# T32-C-7
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C7C8")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C7',xlab='Nitrogen value')
#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original)),
       col="red")

# T32-C-8
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C7C8")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C8',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original)),
       col="red")

# T32-C-9
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C9a","T32-C9b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C9',xlab='Nitrogen value')
#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original)),
       col="red")

# T32-C-10
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C10a","T32-C10b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C10',xlab='Nitrogen value')
#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original,na.rm=T),
#       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original)),
       col="red")

# T32-C-20
plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
     xlim=c(1,8), type="l", main='T32-C20',xlab='Nitrogen value')
points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original,na.rm=T),
       type="l", col="red")
points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original,
       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original)),
       col="red")

legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))

```
The figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests. According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.

### Eksport file (final product)
<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-08-28 functional_plant_indicators_wetland.Rmd
analysis bit of natopen markdown

@@
 ## Uncertainties
-We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
+We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.
 

### 2023-08-28 R/functionalPlantIndicators/code naturally open/1 data handling_natopen.R

@@ -24,54 +24,81 @@ library(tmaptools)
 #### upload data from P-drive ####
 ## ANO
 #st_layers(dsn = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb")
-ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                  layer="ANO_Art")
-ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
-                   layer="ANO_SurveyPoint")
+#ANO.sp <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+#                  layer="ANO_Art")
+#ANO.geo <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/Naturovervaking_eksport.gdb",
+#                   layer="ANO_SurveyPoint")
+
+# load from cache
+ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
+ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))
+
 head(ANO.sp)
 head(ANO.geo)
 
 ## GRUK
 #excel_sheets("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx")
-GRUK.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
-                             sheet = 2)
-GRUK.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
-                           sheet = 3)
+#GRUK.variables <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
+#                             sheet = 2)
+#GRUK.species <- read_excel("P:/41201785_okologisk_tilstand_2022_2023/data/GRUK/GRUKdata_2020-2022_GJELDENDE.xlsx", 
+#                           sheet = 3)
 
 # condition evaluation for 2021 data
[…]

### 2023-08-28 R/functionalPlantIndicators/code naturally open/4 Reference scaling values_natopen.R

@@ -21,7 +21,7 @@ natopen.NiNtypes <- colnames(natopen.ref.cov[["Light"]])
 natopen.NiNtypes[-x] <- substr(natopen.NiNtypes[-x], 1, nchar(natopen.NiNtypes[-x])-1)
 natopen.NiNtypes
 
-# 7 indicator-value indicators: Tyler's Light, Moisture, Soil_reaction_pH, Nitrogen, Phosphorus, Grazing_mowing, Soil_disturbance"
+# 7 indicator-value indicators: Grimes C, S and R, Tyler's Light, Nitrogen, and Soil_disturbance
 indEll.n=6
 # creating a table to hold:
 # Tyler: the 0.5 quantile (median), 0.05 quantile and  0.95 quantile for each NiN-type

### 2023-08-28 R/functionalPlantIndicators/code naturally open/5 scaled value calculations_natopen_GRUK.R

@@ -90,6 +90,9 @@ ind <- unique(natopen.ref.cov.val$Ind)
 # choose columns for site description
 colnames(GRUK.natopen)
 results.natopen.GRUK[['original']] <- GRUK.natopen
+# drop geometry
+st_geometry(results.natopen.GRUK[['original']]) <- NULL
+results.natopen.GRUK[['original']] <- as.data.frame(results.natopen.GRUK[['original']])
 
 
 # add columns for indicators

### 2023-08-28 R/functionalPlantIndicators/code naturally open/6 post scaling analyses_natopen.R

@@ -150,6 +150,79 @@ ggplot(res.natopen.GRUK, aes(x=factor(Kartleggingsenhet), y=scaled_value, fill=f
   ylab("Scaled indicator value (GRUK data)") 
 
 
+#### relating scaled values to NiN condition variables ####
+
+# ANO
+ggplot(res.natopen.ANO, aes(x=fa_total_dekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.ANO, aes(x=vedplanter_total_dekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.ANO, aes(x=busker_dekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.ANO, aes(x=tresjikt_dekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+# GRUK
+colnames(res.natopen.GRUK)[c(27,28,34,35,38)] <- c("driving","erosion","busksjiktsdekning","tresjiktsdekning","fremmedartsdekning")
+ggplot(res.natopen.GRUK, aes(x=fremmedartsdekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.GRUK, aes(x=tresjiktsdekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.GRUK, aes(x=busksjiktsdekning, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.GRUK, aes(x=erosion, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+ggplot(res.natopen.GRUK, aes(x=driving, y=scaled_value)) +
+  geom_point() +
+  facet_wrap(~fp_ind, scale="fixed")
+
+
+plot.erosion.1 <- ggplot(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="RR2",], aes(x=erosion, y=scaled_value)) +
+  geom_point() +
[…]

### 2023-08-28 R/functionalPlantIndicators/n and regions.R

@@ -1,0 +1,6 @@
+nrow(res.natopen.GRUK2[!is.na(res.natopen.GRUK2$region) & !is.na(res.natopen.GRUK2$uPunkt_ID) & res.natopen.GRUK2$region=="Northern Norway",])
+nrow(res.natopen.GRUK2[!is.na(res.natopen.GRUK2$region) & !is.na(res.natopen.GRUK2$uPunkt_ID) & res.natopen.GRUK2$region=="Central Norway",])
+nrow(res.natopen.GRUK2[!is.na(res.natopen.GRUK2$region) & !is.na(res.natopen.GRUK2$uPunkt_ID) & res.natopen.GRUK2$region=="Western Norway",])
+nrow(res.natopen.GRUK2[!is.na(res.natopen.GRUK2$region) & !is.na(res.natopen.GRUK2$uPunkt_ID) & res.natopen.GRUK2$region=="Eastern Norway",])
+nrow(res.natopen.GRUK2[!is.na(res.natopen.GRUK2$region) & !is.na(res.natopen.GRUK2$uPunkt_ID) & res.natopen.GRUK2$region=="Southern Norway",])
+

### 2023-08-28 DRAFT_functional_plant_indicators_natopen.Rmd
markdown analysis continued

@@
 library(stars)
+library(betareg)
+library(glmmTMB)
 library(knitr)
@@
 
-#### continue here ####
-
 ### Regions - maps and statistics
@@
 (i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
-(ii) both the ANO and the ASO datasets have a nested structure
+(ii) both the ANO and the GRUK datasets have a nested structure
 
@@
 ```{r}
-library(betareg)
-library(glmmTMB)
 
@@
 ```{r}
-
-res.seminat.ANO2 = st_join(res.seminat.ANO2, regnor, left = TRUE)
-res.seminat.ASO2 = st_join(res.seminat.ASO2, regnor, left = TRUE)
+# we have to join the ANO and GRUK results spatial objects with the Norway and region mask
+res.natopen.ANO2 = st_join(res.natopen.ANO2, regnor, left = TRUE)
+res.natopen.GRUK2 = st_join(res.natopen.GRUK2, regnor, left = TRUE)
+# we check if all there's any sitethat did not get a region assigned
+nrow(res.natopen.ANO2[is.na(res.natopen.ANO2$region),]) # no NA's for ANO
+nrow(res.natopen.GRUK2[is.na(res.natopen.GRUK2$region),]) # some points didn't get assigned to a region. Why?
+
+tm_shape(regnor, bbox = boks) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(res.natopen.GRUK2[is.na(res.natopen.GRUK2$region),]) +
+  tm_dots('Nitrogen2',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE), scale=2, legend.show = FALSE) + # 
+  tm_layout(main.title = "Nitrogen index (upper), natopen GRUK",legend.position = c("right", "bottom"), main.title.size=1.2) + 
+  tm_add_legend(type = "fill", 
+                col = c(tmaptools::get_brewer_pal("YlOrRd", 7, plot = FALSE),'grey'),
+                labels = c("0.3 - 0.4", "0.4 - 0.5", "0.5 - 0.6", "0.6 - 0.7", 
+                           "0.7 - 0.8", "0.8 - 0.9", "0.9 - 1.0", "NA"),
+                title = "index values")
+# they seem to lie in water
+# all sites but the southernmost one are in Eastern Norway, the remaining one in Southern Norway
+summary(res.natopen.GRUK2[is.na(res.natopen.GRUK2$region),"y"])
+res.natopen.GRUK2[is.na(res.natopen.GRUK2$region) & res.natopen.GRUK2$y<59.83,c("y","Flate_ID")] # site 123-4 is in Southern Norway
+res.natopen.GRUK2[res.natopen.GRUK2$Flate_ID=="123-4","region"] <- "Southern Norway"
+# and all the other region=NA observations are in Eastern Norway
+res.natopen.GRUK2[is.na(res.natopen.GRUK2$region),"region"] <- "Eastern Norway"
+nrow(res.natopen.GRUK2[is.na(res.natopen.GRUK2$region),]) # no NA's left
+
+# now we can calculate regionwise means and standard errors with beta-regression null-models
+# note that beta-models cannot handle observations that are exactly 0 or 1
+res.natopen.ANO2$RR1[res.natopen.ANO2$RR1==1] <- 0.999
+res.natopen.ANO2$RR1[res.natopen.ANO2$RR1==0] <- 0.001
+res.natopen.GRUK2$Nitrogen2[res.natopen.GRUK2$Nitrogen2==1] <- 0.999
+res.natopen.GRUK2$Nitrogen2[res.natopen.GRUK2$Nitrogen2==0] <- 0.001
 
@@
   mutate(
-    Grazing_mowing1.ANO.reg.mean = c(indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[1]
+    RR1.ANO.reg.mean = c(indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Northern Norway",c("RR1","ano_flate_id")])[1],
+                         indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Central Norway",c("RR1","ano_flate_id")])[1],
+                         indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Eastern Norway",c("RR1","ano_flate_id")])[1],
+                         indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Western Norway",c("RR1","ano_flate_id")])[1],
+                         indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Southern Norway",c("RR1","ano_flate_id")])[1]
     ),
-    Grazing_mowing1.ANO.reg.se = c(indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
-                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway",c("Grazing_mowing1","ano_flate_id")])[2],
-                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway",c("Grazing_mowing1","ano_flate_id")])[2],
-                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway",c("Grazing_mowing1","ano_flate_id")])[2],
-                               indmean.beta(df=res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway",c("Grazing_mowing1","ano_flate_id")])[2]
+    RR1.ANO.reg.se = c(indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Northern Norway",c("RR1","ano_flate_id")])[2]*2,
+                       indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Central Norway",c("RR1","ano_flate_id")])[2]*2,
+                       indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Eastern Norway",c("RR1","ano_flate_id")])[2]*2,
+                       indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Western Norway",c("RR1","ano_flate_id")])[2]*2,
+                       indmean.beta(df=res.natopen.ANO2[res.natopen.ANO2$region=="Southern Norway",c("RR1","ano_flate_id")])[2]*2
     ),
-    Grazing_mowing1.ANO.reg.n = c(nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Northern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
-                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Central Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
-                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Eastern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
-                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Western Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),]),
-                              nrow(res.seminat.ANO2[res.seminat.ANO2$region=="Southern Norway" & !is.na(res.seminat.ANO2$Grazing_mowing1),])
+    RR1.ANO.reg.n = c(nrow(res.natopen.ANO2[res.natopen.ANO2$region=="Northern Norway" & !is.na(res.natopen.ANO2$RR1),]),
+                      nrow(res.natopen.ANO2[res.natopen.ANO2$region=="Central Norway" & !is.na(res.natopen.ANO2$RR1),]),
+                      nrow(res.natopen.ANO2[res.natopen.ANO2$region=="Eastern Norway" & !is.na(res.natopen.ANO2$RR1),]),
+                      nrow(res.natopen.ANO2[res.natopen.ANO2$region=="Western Norway" & !is.na(res.natopen.ANO2$RR1),]),
+                      nrow(res.natopen.ANO2[res.natopen.ANO2$region=="Southern Norway" & !is.na(res.natopen.ANO2$RR1),])
+    ),
+    Nitrogen2.GRUK.reg.mean = c(indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Northern Norway",c("Nitrogen2","Flate_ID")])[1],
+                                indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Central Norway",c("Nitrogen2","Flate_ID")])[1],
+                                indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Eastern Norway",c("Nitrogen2","Flate_ID")])[1],
+                                indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Western Norway",c("Nitrogen2","Flate_ID")])[1],
+                                indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Southern Norway",c("Nitrogen2","Flate_ID")])[1]
+    ),
+    Nitrogen2.GRUK.reg.se = c(indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Northern Norway",c("Nitrogen2","Flate_ID")])[2]*2,
+                              indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Central Norway",c("Nitrogen2","Flate_ID")])[2]*2,
+                              indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Eastern Norway",c("Nitrogen2","Flate_ID")])[2]*2,
+                              indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Western Norway",c("Nitrogen2","Flate_ID")])[2]*2,
+                              indmean.beta(df=res.natopen.GRUK2[res.natopen.GRUK2$region=="Southern Norway",c("Nitrogen2","Flate_ID")])[2]*2
+    ),
+    Nitrogen2.GRUK.reg.n = c(0,
+                             0,
+                             nrow(res.natopen.GRUK2[res.natopen.GRUK2$region=="Eastern Norway" & !is.na(res.natopen.GRUK2$Nitrogen2),]),
+                             0,
+                             nrow(res.natopen.GRUK2[res.natopen.GRUK2$region=="Southern Norway" & !is.na(res.natopen.GRUK2$Nitrogen2),])
     )
@@
 
-# and adding the values for ASO
-regnor <- regnor %>%
-  mutate(
-    Nitrogen2.ASO.reg.mean = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
-                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
-                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
-                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1],
-                               indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[1]
-                               ),
-    Nitrogen2.ASO.reg.se = c(indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
-                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
-                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
-                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2],
-                             indmean.beta(df=res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway",c("Nitrogen2","Omradenummer_flatenummer")])[2]
-                             ),
-    Nitrogen2.ASO.reg.n = c(nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Northern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
-                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Central Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
-                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Eastern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
-                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Western Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]),
-                            nrow(res.seminat.ASO2[res.seminat.ASO2$region=="Southern Norway" & !is.na(res.seminat.ASO2$Nitrogen2),]))
-  )
-
-## scaled value maps for Grazing_mowing1 (lower indicator), ANO
+## scaled value maps for CSR-R1 (lower indicator), ANO
 # mean
 tm_shape(regnor) +
-  tm_polygons(col="Grazing_mowing1.ANO.reg.mean", title="Grazing_mowing (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
-  tm_text("Grazing_mowing1.ANO.reg.n",col="black",bg.color="grey")
-```
-Mean index value by region for the upper Grazing_mowing indicator (i.e.
-index shows deviations towards less grazing or mowing) from the ANO monitoring data.
+  tm_polygons(col="RR1.ANO.reg.mean", title="CSR-R (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("RR1.ANO.reg.n",col="black",bg.color="grey")
+```
+Mean index value by region for the lower CSR-R indicator (i.e.
+index shows deviations towards less ruderal species) from the ANO monitoring data.
 Numbers in grey fields show the number of observations in the respective region.
@@
 tm_shape(regnor) +
-  tm_polygons(col="Grazing_mowing1.ANO.reg.se", title="Grazing_mowing (lower), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
-  tm_text("Grazing_mowing1.ANO.reg.n",col="black",bg.color="grey")
-
-```
-Standard error to the mean index value by region for the lower Grazing_mowing indicator from the ANO monitoring data. Numbers in grey fields show the number of observations in the respective region.
-
-And here are the corresponding maps for the upper Nitrogen indicator in the ASO data:
+  tm_polygons(col="RR1.ANO.reg.se", title="CSR-R (lower)", style="quantile", palette=(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("RR1.ANO.reg.n",col="black",bg.color="grey")
+
+```
+Standard error to the mean index value by region for the lower CSR-R indicator from the ANO monitoring data. Numbers in grey fields show the number of observations in the respective region.
+
+And here are the corresponding maps for the upper Nitrogen indicator in the GRUK data:
 ```{r}
@@
 tm_shape(regnor) +
-  tm_polygons(col="Nitrogen2.ASO.reg.mean", title="Nitrogen (upper), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
-  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")
+  tm_polygons(col="Nitrogen2.GRUK.reg.mean", title="Nitrogen (upper), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("Nitrogen2.GRUK.reg.n",col="black",bg.color="grey")
 ```
 Mean index value by region for the upper Nitrogen indicator (i.e.
-index shows deviations towards increased Nitrogen availability) from the ASO monitoring data.
+index shows deviations towards increased Nitrogen affinity in the plant community) from the GRUK monitoring data.
 Numbers in grey fields show the number of observations in the respective region.
@@
 tm_shape(regnor) +
-  tm_polygons(col="Nitrogen2.ASO.reg.se", title="Nitrogen (upper), SE", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
-  tm_text("Nitrogen2.ASO.reg.n",col="black",bg.color="grey")
-
-```
-Standard error to the mean index value by region for the upper Nitrogen indicator from the ASO monitoring data. Numbers in grey fields show the number of observations in the respective region.
+  tm_polygons(col="Nitrogen2.GRUK.reg.se", title="Nitrogen (upper), 2 SE", style="quantile", palette=(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_text("Nitrogen2.GRUK.reg.n",col="black",bg.color="grey")
+
+```
+Standard error to the mean index value by region for the upper Nitrogen indicator from the GRUK monitoring data. Numbers in grey fields show the number of observations in the respective region.
+
+#### continue here ####
 
@@
 We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
-Since Grazing_mowing for ANO and nitrogen for ASO show some deviations  we exemplify this with these indicators for unscaled values.
+Since CSR-R for ANO and nitrogen for GRUK show some deviations, we exemplify this with these indicators for unscaled values.
 

### 2023-08-28 R/functionalPlantIndicators/code naturally open/6 post scaling analyses_natopen.R

@@ -8,6 +8,8 @@ library(tidyverse)
 library(readxl)
 library(tmap)
 library(tmaptools)
+library(betareg)
+library(glmmTMB)
 
 #### getting all the data from previous scripts in ####
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/data_natopen.RData")
@@ -204,8 +206,7 @@ plot.alien.1 <- ggplot(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",], 
 
 # use beta-regression for analysis of response between 0 and 1
 expit <- function(L) exp(L) / (1+exp(L))
-library(betareg)
-library(glmmTMB)
+
 
 mod.GRUK.R2.erosion <- glmmTMB(scaled_value ~ erosion +(1|Flate_ID), family=beta_family(), data=res.natopen.GRUK[res.natopen.GRUK$fp_ind=="RR2",])
 summary(mod.GRUK.R2.erosion)
@@ -232,18 +233,6 @@ st_geometry(res.natopen.ANO2) <- st_geometry(ANO.natopen)
 
 res.natopen.GRUK2 <- results.natopen.GRUK[['2-sided']]
 st_geometry(res.natopen.GRUK2) <- st_geometry(GRUK.natopen)
-
-## similarly for GRUK
-# keep wide format and create geometry from coords in the dataframe
-res.natopen.GRUK2 <- results.natopen.GRUK[['2-sided']]
-# make spatial object
-res.natopen.GRUK2 <- st_as_sf(res.natopen.GRUK2, coords = c("x","y"),remove=F)
-# add CRS
-res.natopen.GRUK2 <- st_set_crs(res.natopen.GRUK2,4326)
-# transform CRS to match ANO
-res.natopen.GRUK2 <- res.natopen.GRUK2 %>%
-  st_as_sf() %>%
-  st_transform(crs = st_crs(ANO.geo))
 
 
 #nor <- readRDS('P:/41201785_okologisk_tilstand_2022_2023/data/rds/norway_outline.RDS')%>%
@@ -356,9 +345,9 @@ nrow(res.natopen.ANO2[is.na(res.natopen.ANO2$region),]) # no NA's for regions
 res.natopen.GRUK2 = st_join(res.natopen.GRUK2, regnor, left = TRUE)
[…]

### 2023-08-29 DRAFT_functional_plant_indicators_natopen.Rmd
natopen markdown continued

@@
 
-Grazing_mowing, ANO:
+CSR-R, ANO:
 ```{r}
-#summary(res.seminat$kartleggingsenhet_1m2)
-#length(unique(res.seminat$kartleggingsenhet_1m2))
-# 19 NiN-types, of which T41-C-1 has 0 observations, and T32 & T34 as main types don't have a reference -> so, 16 NiN-Types to plot
-#colnames(seminat.ref.cov[['Light']])
-
-### Grazing_mowing
-par(mfrow=c(2,4))
-## T32s
-# T32-C-1
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C1',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original,na.rm=T),
+# summary(res.natopen.ANO$kartleggingsenhet_1m2)
+# 15 NiN-types, of which T2, T13 & T18 as main types don't have a reference
+# many T13-types (rasmark) have too few vascular plants in the reference lists for getting a reference distribution
+# C1, C2 and C10 in the ANO data have none
+#-> so, 9 NiN-Types to plot
+
+x11()
+par(mfrow=c(3,3))
+
+for ( i in sort(unique(res.natopen.ANO$kartleggingsenhet_1m2)[-c(2,8,9,12,13,15)]) ) {
+
+  tryCatch({
+
+    plot(density( as.matrix(natopen.ref.cov[['RR']][,i]) ,na.rm=T),
+         xlim=c(0,1), ylim=c(0,10), type="l", main=i,xlab='CSR-R value')
+    points(density(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,na.rm=T),
            type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-1",]$original)),
+    points(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,
+           rep(0,length(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original)),
            col="red")
 
-# T32-C-2
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C1C2"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C2',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original,na.rm=T),
+  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+
+}
+legend("topright", legend=c("reference","field data"), pch=c(NULL,1), lty=1, col=c("black","red"))
+
+```
+
+Nitrogen, GRUK
+```{r}
+
+# only two NiN types in GRUK, T2-C-7 and T2-C-8
+par(mfrow=c(1,2))
+
+for ( i in unique(res.natopen.GRUK$Kartleggingsenhet) ) {
+
+  tryCatch({
+
+    plot(density( as.matrix(natopen.ref.cov[['Nitrogen']][,i]) ,na.rm=T),
+         xlim=c(1,9), ylim=c(0,2), type="l", main=i,xlab='Nitrogen value')
+    points(density(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original,na.rm=T),
            type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-2",]$original)),
+    points(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original,
+           rep(0,length(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original)),
            col="red")
 
-# T32-C-3
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C3C4"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C3',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-3",]$original)),
-       col="red")
-
-# T32-C-4
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T32-C3C4"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C4',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-4",]$original)),
-       col="red")
-
-# T32-C-5
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C5',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-5",]$original)),
-       col="red")
-
-# T32-C-6
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C21C6a","T32-C21C6b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C6',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-6",]$original)),
-       col="red")
-
-# T32-C-9
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C9a","T32-C9b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C9',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-9",]$original)),
-       col="red")
-
-# T32-C-10
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C10a","T32-C10b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C10',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-10",]$original)),
-       col="red")
-
-par(mfrow=c(2,4))
-# T32-C-15
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C15")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C15',xlab='Grazing_mowing value')
-#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-15",]$original)),
-       col="red")
-
-# T32-C-20
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C20',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T32-C-20",]$original)),
-       col="red")
-
-## V10s
-# V10-C-1
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"V10-C1C2"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='V10-C1',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-1",]$original)),
-       col="red")
-
-# V10-C-3
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"V10-C3"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='V10-C3',xlab='Grazing_mowing value')
-#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="V10-C-3",]$original)),
-       col="red")
-
-## T34s
-# T34-C-1
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,"T34-C1"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T34-C1',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-1",]$original)),
-       col="red")
-
-# T34-C-2
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T34-C2a","T34-C2b","T34-C2c")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T34-C2',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-2",]$original)),
-       col="red")
-
-# T34-C-4
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T34-C4a","T34-C4b","T34-C4c","T34-C4d")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T34-C4',xlab='Grazing_mowing value')
-#points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T34-C-4",]$original)),
-       col="red")
-
-## T41
-plot(density( as.matrix(seminat.ref.cov[['Grazing_mowing']][,c("T41a","T41b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T41',xlab='Grazing_mowing value')
-points(density(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original,
-       rep(0,length(res.seminat.ANO[res.seminat.ANO$fp_ind=="Grazing_mowing1" & res.seminat.ANO$kartleggingsenhet_1m2=="T41",]$original)),
-       col="red")
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-```
-
-Nitrogen, ASO
-```{r}
-
-#summary(res.seminat.ASO$NiN_grunntype)
-#length(levels(res.seminat.ASO$NiN_grunntype))
-# 12 NiN-types, of which C-13 and C-2 have 0 observations -> so, 10 NiN-Types to plot
-#colnames(seminat.ref.cov[['Nitrogen']])
-
-### Nitrogen
-par(mfrow=c(3,4))
-# T32-C-1
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C1C2"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C1',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-1",]$original)),
-       col="red")
-
-# T32-C-3
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C3C4"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C3',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-3",]$original)),
-       col="red")
-
-# T32-C-4
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,"T32-C3C4"]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C4',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-4",]$original)),
-       col="red")
-
-# T32-C-5
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C5',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-5",]$original)),
-       col="red")
-
-# T32-C-6
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C21C6a","T32-C21C6b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C6',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-6",]$original)),
-       col="red")
-
-# T32-C-7
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C7C8")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C7',xlab='Nitrogen value')
-#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-7",]$original)),
-       col="red")
-
-# T32-C-8
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C7C8")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C8',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-8",]$original)),
-       col="red")
-
-# T32-C-9
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C9a","T32-C9b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C9',xlab='Nitrogen value')
-#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-9",]$original)),
-       col="red")
-
-# T32-C-10
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C10a","T32-C10b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C10',xlab='Nitrogen value')
-#points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-10",]$original)),
-       col="red")
-
-# T32-C-20
-plot(density( as.matrix(seminat.ref.cov[['Nitrogen']][,c("T32-C5C20a","T32-C5C20b")]) ,na.rm=T),
-     xlim=c(1,8), type="l", main='T32-C20',xlab='Nitrogen value')
-points(density(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original,na.rm=T),
-       type="l", col="red")
-points(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original,
-       rep(0,length(res.seminat.ASO[res.seminat.ASO$fp_ind=="Nitrogen2" & res.seminat.ASO$NiN_grunntype=="T32-C-20",]$original)),
-       col="red")
-
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
+  }, error=function(e){cat("ERROR :",conditionMessage(e), "\n")})
+
+}
+legend("topright", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
 

### 2023-08-29 R/functionalPlantIndicators/code naturally open/6 post scaling analyses_natopen.R

@@ -601,7 +601,7 @@ tm_shape(regnor) +
 
 
 
-#### continue here ####
+
 
 #### how many observations are there? ####
 # total ANO
@@ -626,427 +626,59 @@ length(unique(res.natopen.ANO2$ano_punkt_id[res.natopen.ANO2$region=="Southern N
 
 
 
-
+#### continue here ####
 
 #### distribution comparison, reference vs. field data ####
 summary(res.natopen.ANO$kartleggingsenhet_1m2)
 length(unique(res.natopen.ANO$kartleggingsenhet_1m2))
-# 15 NiN-types, of which T2, T13 & T18 as main types don't have a reference -> so, 12 NiN-Types to plot
-colnames(natopen.ref.cov[['RR']])
+### ANO, CSR-R
+# 15 NiN-types, of which T2, T13 & T18 as main types don't have a reference
+# many T13-types (rasmark) have too few vascular plants in the reference lists for getting a reference distribution
+# C1, C2 and C10 in the ANO data have none
+#-> so, 9 NiN-Types to plot
 
-### Light
 x11()
-par(mfrow=c(3,4))
-## T2s
-# T2-C-1
-plot(density( as.matrix(natopen.ref.cov[['RR']][,"T2-C-1"]) ,na.rm=T),
-     xlim=c(0,1), ylim=c(0,10), type="l", main='T2-C-1',xlab='CSR-R value')
-points(density(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2=="T2-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2=="T2-C-1",]$original,
-       rep(0,length(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2=="T2-C-1",]$original)),
-       col="red")
+par(mfrow=c(3,3))
 
-# T2-C-2
[…]

### 2023-08-29 DRAFT_functional_plant_indicators_natopen.Rmd
natopen markdown finished

@@
 # we have no reference for overall T2 registrations, these 24 observations won't be evaluated
-ANO.natopen[ANO.natopen$kartleggingsenhet_1m2=="T2",] # were largely inaccessible and thus mapped from arial pictures
+ANO.natopen[ANO.natopen$kartleggingsenhet_1m2=="T2",] # were largely inaccessible and thus mapped from aerial pictures
 
@@
 ```{r}
+colnames(res.natopen.GRUK)[38] <- "fremmedartsdekning"
 ggplot(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",], aes(x=fremmedartsdekning, y=scaled_value)) +
@@
 ```
-And indeed, we can see that very high alien species cover predicts poor scores in the Nitrogen2 indicator.
+And indeed, we can see that very high alien species cover predicts almost only poor scores in the Nitrogen2 indicator.
 
@@
 ```{r}
-# summary(res.natopen.ANO$kartleggingsenhet_1m2)
-# 15 NiN-types, of which T2, T13 & T18 as main types don't have a reference
-# many T13-types (rasmark) have too few vascular plants in the reference lists for getting a reference distribution
-# C1, C2 and C10 in the ANO data have none
-#-> so, 9 NiN-Types to plot
-
-x11()
-par(mfrow=c(3,3))
-
-for ( i in sort(unique(res.natopen.ANO$kartleggingsenhet_1m2)[-c(2,8,9,12,13,15)]) ) {
+#summary(res.natopen.ANO[!is.na(res.natopen.ANO$original),]$kartleggingsenhet_1m2)
+#sort(unique(res.natopen.ANO$kartleggingsenhet_1m2))
+# 8 NiN-types with data to plot
+
+par(mfrow=c(2,4))
+for ( i in sort(unique(res.natopen.ANO$kartleggingsenhet_1m2))[c(1,6,7,10,12,13,14,15)] ) {
 
@@
          xlim=c(0,1), ylim=c(0,10), type="l", main=i,xlab='CSR-R value')
-    points(density(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,na.rm=T),
-           type="l", col="red")
     points(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,
@@
            col="red")
+    points(density(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,na.rm=T),
+           type="l", col="red")
 
@@
 }
-legend("topright", legend=c("reference","field data"), pch=c(NULL,1), lty=1, col=c("black","red"))
+legend("topright", legend=c("reference","field data"), pch=c(NULL,1), lty=1, col=c("black","red"),bty="n", cex=1)
 
@@
 par(mfrow=c(1,2))
-
 for ( i in unique(res.natopen.GRUK$Kartleggingsenhet) ) {
@@
          xlim=c(1,9), ylim=c(0,2), type="l", main=i,xlab='Nitrogen value')
-    points(density(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original,na.rm=T),
-           type="l", col="red")
     points(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original,
@@
            col="red")
+    points(density(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen1" & res.natopen.GRUK$Kartleggingsenhet==i,]$original,na.rm=T),
+           type="l", col="red")
 
@@
 }
-legend("topright", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-```
-The figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests.
-According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
+legend("topright", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"), cex=1)
+
+```
+The GRUK figure shows that the distributions for the plant communities' nitrogen affinity in the limestone rich T2-areas (åpen grunnlendt mark) around Oslofjord are shifted towards higher nitrogen affinity.
+The ANO figure shows shifts towards less ruderal strategy in plant communities limestone poor T2-areas as well as T12 (strandeng) and T16 (rasmarkhei, -eng) types.
 

### 2023-08-29 R/functionalPlantIndicators/code naturally open/5 scaled value calculations_natopen_ANO.R

@@ -110,7 +110,7 @@ ANO.natopen$kartleggingsenhet_1m2 <- as.factor(ANO.natopen$kartleggingsenhet_1m2
 summary(ANO.natopen$hovedtype_rute)
 summary(ANO.natopen$kartleggingsenhet_1m2)
 # we have no reference for overall T2 registrations, these 24 observations won't be evaluated
-ANO.natopen[ANO.natopen$kartleggingsenhet_1m2=="T2",] # were largely inaccessible and thus mapped from arial pictures
+ANO.natopen[ANO.natopen$kartleggingsenhet_1m2=="T2",] # were largely inaccessible and thus mapped from aerial pictures
 
 results.natopen.ANO <- list()
 ind <- unique(natopen.ref.cov.val$Ind)

### 2023-08-29 R/functionalPlantIndicators/code naturally open/6 post scaling analyses_natopen.R

@@ -204,6 +204,8 @@ plot.alien.1 <- ggplot(res.natopen.GRUK[res.natopen.GRUK$fp_ind=="Nitrogen2",], 
   xlab("Alien species cover (%)") + ylab("Scaled Nitrogen2 value (GRUK data)")
 
 
+
+
 # use beta-regression for analysis of response between 0 and 1
 expit <- function(L) exp(L) / (1+exp(L))
 
@@ -629,33 +631,34 @@ length(unique(res.natopen.ANO2$ano_punkt_id[res.natopen.ANO2$region=="Southern N
 #### continue here ####
 
 #### distribution comparison, reference vs. field data ####
-summary(res.natopen.ANO$kartleggingsenhet_1m2)
-length(unique(res.natopen.ANO$kartleggingsenhet_1m2))
+
 ### ANO, CSR-R
-# 15 NiN-types, of which T2, T13 & T18 as main types don't have a reference
-# many T13-types (rasmark) have too few vascular plants in the reference lists for getting a reference distribution
-# C1, C2 and C10 in the ANO data have none
-#-> so, 9 NiN-Types to plot
+summary(res.natopen.ANO[!is.na(res.natopen.ANO$original),]$kartleggingsenhet_1m2)
+sort(unique(res.natopen.ANO$kartleggingsenhet_1m2))
+# 8 NiN-types with data to plot
 
 x11()
 par(mfrow=c(3,3))
 
-for ( i in sort(unique(res.natopen.ANO$kartleggingsenhet_1m2)[-c(2,8,9,12,13,15)]) ) {
+for ( i in sort(unique(res.natopen.ANO$kartleggingsenhet_1m2))[c(1,6,7,10,12,13,14,15)] ) {
   
   tryCatch({
     
     plot(density( as.matrix(natopen.ref.cov[['RR']][,i]) ,na.rm=T),
          xlim=c(0,1), ylim=c(0,10), type="l", main=i,xlab='CSR-R value')
-    points(density(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,na.rm=T),
-           type="l", col="red")
     points(res.natopen.ANO[res.natopen.ANO$fp_ind=="RR1" & res.natopen.ANO$kartleggingsenhet_1m2==i,]$original,
[…]

### 2023-08-30 DRAFT_gjengroing.Rmd
Change file paths

--- 2023-08-10 DRAFT_gjengroing.Rmd
@@
 
-_Author and date:_
-Zander Venter
+*Author and date:* Zander Venter
 
@@
 ## Introduction
+
 The Norwegian word "gjengroing" is directly translated to "regrowing" in English.
@@
 ## About the underlying data
+
 We rely on the following datasets:
 
-* [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
-* NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type.
+-   [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
+-   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type.
 The population sample are the areas used to determine ecological condition relative to the reference areas.
@@
 poor condition),
-* LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/).
+-   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/).
 This includes both a terrain (DTM) and surface mode (DSM).
@@
 From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
-* Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df).
+-   Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df).
 We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band.
 This is done in combination with bioclimatic zones - see next point.
-* Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner).
+-   Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner).
 We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
-* [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
-* A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
-* The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
-* The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.
+-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
+-   A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
+-   The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
+-   The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.
 
 ### Representativity in time and space
+
 The index will cover the mainland of Norway.
@@
 ### Original units
+
 The original units for ecological condition are meters. This is the height of the vegetation within referecne, polygon and forest areas.
@@
 ### Temporal coverage
+
 Circa 2010 to 2021.
@@
 ### Aditional comments about the dataset
+
 None.
@@
 ## Ecosystem characteristic
+
 ### Norwegain standard
+
 Not sure what this means??
@@
 ## Collinearities with other indicators
+
 There is possibly a collinearity with the primary production indicator (primærproduksjon).
@@
 ## Reference state and values
+
 The methodology used to calculate the gjengroing indicator is outlined in the schematic below.
@@
 
-```{r}
-include_graphics('./DATA/images/gjengroing_schematic.jpg')
-```
+![](images/gjengroing_schematic.jpg)
 
 ### Reference state
+
 The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state.
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+
 Once we have the reference vegetation height for a given ecosystem type and elevation-bioclimatic zone, we need to define the minimum (or worst/bad) condition.
@@
 ## Uncertainties
+
 For the indicator map at the polygon level there is no uncertainty associated with the indicator values.
@@
 ## References
+
 Links to data and resources are provided with hyperlinks in-line.
@@
 ## Analyses
+
 ### Data sets
+
 There are several datasets which are used in GEE which will not be imported into the R session here.
@@
 
-* [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6)
-* [LiDAR-derived digital elevation model from høydedata](https://hoydedata.no/LaserInnsyn/).
-* [Kartverket's national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df)
-* [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) 
-* [European forest clear-cut map](https://www.nature.com/articles/s41893-020-00609-y)
+-   [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6)
+-   [LiDAR-derived digital elevation model from høydedata](https://hoydedata.no/LaserInnsyn/).
+-   [Kartverket's national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df)
+-   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97)
+-   [European forest clear-cut map](https://www.nature.com/articles/s41893-020-00609-y)
 
@@
 #### Regions
+
 The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.
@@
 ```{r}
-regions <- st_read('./DATA/regions.shp') %>%
+regions <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/regions.shp') %>%
   mutate(region = factor(region) )
@@
 #### SSB 10 and 50km grids
+
 The [SSB 10km and 50km grids](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) are used for visualizing the distribution of available data and for aggregating gjengroing index scores.
@@
 #### Bioclimatic regions
+
 The Moen's bioclimatic regions are imported from NINA R/GeoSpatialData/ server.
@@
 #### NiN polygons
+
 [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are imported here from the NINA servers:
+
 ```{r}
@@
 
-* **Export the cleaned NiN data and upload to GEE before proceeding**
+-   **Export the cleaned NiN data and upload to GEE before proceeding**
 
@@
 
-* **Now run the following GEE JavaScript code by clicking on the hyperlink:** [GEE script](https://code.earthengine.google.com/b2d06ad72c515fbbb3024f157987eedc) This script is also found in the SCR repository as "GEE_script.js".
+-   **Now run the following GEE JavaScript code by clicking on the hyperlink:** [GEE script](https://code.earthengine.google.com/b2d06ad72c515fbbb3024f157987eedc) This script is also found in the SCR repository as "GEE_script.js".
 When you run the script, then click the Tasks tab in the GEE code editor and run the export tasks.
@@
 This script extracts three CSV files:
-    + ‘area_cover_grid.csv’ - this gives the area coverage of NiN, AR5 and LiDAR data for each 10x10km SSB grid cell.
-    + 'elevation_stratified.shp' - this is a 300m elevation band stratification that is simply used for visualizing the strata here in R.
-    + ‘vegHeights.zip’ - this is a folder with hundreds of CSV files.
+    -   'area_cover_grid.csv' - this gives the area coverage of NiN, AR5 and LiDAR data for each 10x10km SSB grid cell.
+    -   'elevation_stratified.shp' - this is a 300m elevation band stratification that is simply used for visualizing the strata here in R.
+    -   'vegHeights.zip' - this is a folder with hundreds of CSV files.
 These files contain the vegetation heights for reference, population, and forest polygons over Norway.
@@
 #### Data from Google Earth Engine
+
 Once you have run the GEE script above, and downloaded the data to the /data/gjengroing/From_GEE directory, you can proceed with this R workflow.
@@
 readVegHeightFiles <- function(uniqueString){
-  dir <- './DATA/From_GEE/vegHeights/'
+  dir <- '/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/vegHeights/'
   files <- list.files(dir)[str_detect(list.files(dir), uniqueString)]
@@
 # Import data coverages per SSB 10km grid cell
-areaCovers <- read_csv('./DATA/From_GEE/area_cover_grid.csv') %>%
+areaCovers <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/area_cover_grid.csv') %>%
   dplyr::select(-'.geo', -'system:index')
@@
 
-elevStrata <- st_read('./DATA/From_GEE/elevation_stratified.shp') %>%
+elevStrata <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/elevation_stratified.shp') %>%
   mutate(elevBand = ifelse(mode == 1, '0-300',
@@
   refVaatmark %>%
-    st_write('./DATA/refVaatmark.csv')
+    st_write('data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
   refAapne %>%
-    st_write('./DATA/refAapne.csv')
+    st_write('data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
 
@@
 
-  refVaatmark  <- read_csv('./DATA/refVaatmark.csv')
-  refAapne <- read_csv('./DATA/refAapne.csv')
+  refVaatmark  <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
+  refAapne <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
 
@@
 
-  vaatmarkIndexPolySpat <- st_read('./OUTPUT/vaatmark_index.shp')%>%
+  vaatmarkIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp')%>%
     mutate(`system:index` = id,
@@
     dplyr::select(-id, -elev, -vegZn)
-  aapneIndexPolySpat <- st_read('./OUTPUT/aapne_index.shp')%>%
+  aapneIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp')%>%
     mutate(`system:index` = id,
@@
 # Create intersection of the two strata layers
+#elevStrata <- st_transform(elevStrata, st_crs(regions))
 bioClimElev <- bioclim %>%
@@
   # Write out for import later
-  vaatmarkIndexStrata %>% write_csv('./DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata %>% write_csv('./DATA/aapneIndexStrata.csv')
+  vaatmarkIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
+  aapneIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
 
@@
 
-  vaatmarkIndexStrata <- read_csv('./DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata <- read_csv('./DATA/aapneIndexStrata.csv')
+  vaatmarkIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
+  aapneIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
 
@@
 
-  vaatmarkIndexGrid <- st_read('./OUTPUT/vaatmark_index_grid.shp')
-  aapneIndexGrid <- st_read('./OUTPUT/aapne_index_grid.shp')
+  vaatmarkIndexGrid <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp')
+  aapneIndexGrid <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp')
 
@@
 
-  vaatmarkIndexRegion <- st_read('./OUTPUT/vaatmark_index_region.shp')
-  aapneIndexRegion <- st_read('./OUTPUT/aapne_index_region.shp')
+  vaatmarkIndexRegion <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp')
+  aapneIndexRegion <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp')
 
@@
 
-* **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise.
+-   **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise.
 However, both NiN and AR5 are spatially and thematically biased in some ways.
@@
 
-* **Definition of good condition states in NiN data.** Here we used the overall "tilstand" score for each NiN reference polygon.
+-   **Definition of good condition states in NiN data.** Here we used the overall "tilstand" score for each NiN reference polygon.
 However, the overall score is a combination of several sub-scores that may or may not be relevant for ecological condition.
@@
 
-* **Definition of climax successional stage for tree regrowth.** Here we use the 50th percentile of LiDAR-derived tree heights in AR5 forest polygons that have not been harvested since 1986.
+-   **Definition of climax successional stage for tree regrowth.** Here we use the 50th percentile of LiDAR-derived tree heights in AR5 forest polygons that have not been harvested since 1986.
 This defines the zero on the scaled indicator score.
@@
 
-* **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data.
+-   **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data.
 However, for poor condition we use climax successional stage for trees (as explained above) within 200m of each population polygon.
@@
 
-* **Quantification of uncertainty around indicator scores.** Here we used the standard deviation in the vegetation heights for reference (NiN) polygons within each region to quantify the uncertainty around scaled indicator scores.
+-   **Quantification of uncertainty around indicator scores.** Here we used the standard deviation in the vegetation heights for reference (NiN) polygons within each region to quantify the uncertainty around scaled indicator scores.
 We did this because the method used in eaTools::ea_spread() function is a bootstrapping approach which resulted in extremely small uncertainty estimates due to the large number of polygons included in our analysis.
@@
 ```{r}
-include_graphics('./DATA/images/val_1.jpg')
+include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_1.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_2.jpg')
+include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_2.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_3.jpg')
+include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_3.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_4.jpg')
+include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_4.jpg')
 ```
@@
 ```{r}
-include_graphics('./DATA/images/val_5.jpg')
+include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_5.jpg')
 ```
@@
 
-We will export gjengroing indicator values at three levels:
-- polygon level
-- 50km grid level
-- regional level
+We will export gjengroing indicator values at three levels: - polygon level - 50km grid level - regional level
 
@@
     dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('./OUTPUT/vaatmark_index.shp', append=FALSE)
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp', append=FALSE)
 
@@
     dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('./OUTPUT/aapne_index.shp', append=FALSE)
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp', append=FALSE)
 
@@
   vaatmarkIndexGrid %>%
-    st_write('./OUTPUT/vaatmark_index_grid.shp', append = FALSE)
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp', append = FALSE)
   aapneIndexGrid %>%
-    st_write('./OUTPUT/aapne_index_grid.shp', append = FALSE)
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp', append = FALSE)
 
@@
   vaatmarkIndexRegion %>%
-    st_write('./OUTPUT/vaatmark_index_region.shp', append = FALSE)
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp', append = FALSE)
   aapneIndexRegion %>%
-    st_write('./OUTPUT/aapne_index_region.shp', append = FALSE)
-}
-
-```
+    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp', append = FALSE)
+}
+
+```

### 2023-08-30 gjengroing.Rmd
Alter the gjengroing chapter to display images correctly AND Re-render

--- 2023-08-30 DRAFT_gjengroing.Rmd
@@
----
-title: "Vegetation regrowth/encroachment (gjengroing)"
-output:
-  pdf_document: default
-  html_document: default
----
+# Vegetation regrowth/encroachment (gjengroing) {#gjengroing}
 
@@
 
-```{r}
-include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_1.jpg')
-```
+![](images/val_1.jpg)
 
@@
 
-```{r}
-include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_2.jpg')
-```
+![](images/val_2.jpg)
 
@@
 
-```{r}
-include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_3.jpg')
-```
+![](images/val_3.jpg)
 
@@
 
-```{r}
-include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_4.jpg')
-```
+![](images/val_4.jpg)
 
@@
 
-```{r}
-include_graphics('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/images/val_5.jpg')
-```
+![](images/val_5.jpg)
 

### 2023-08-31 DRAFT_functional_plant_indicators_natopen.Rmd
updates all fpi markdowns

@@
 1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle.
-Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
+Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 18 evenly distributed vegetation analyses in 1 x 1 m plots (up to 18, because some of these evenly distributed points may be in water or otherwise inaccessible).
 For the vegetation analyses, the cover of each vascular plant species in the plot is recorded.
@@
 By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
-These distributions are calculated for ecosystem sub-types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the main wetland types in NiN.
+These distributions are calculated for minor ecosystem types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN.
 A more extensive discussion on the use of reference communities can be found in Jakobsson et al.

### 2023-08-31 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R

@@ -1,3 +1,16 @@
+library(downloader)
+library(sf)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+library(tidyverse)
+library(readxl)
+library(tmap)
+library(tmaptools)
+library(betareg)
+library(glmmTMB)
+
 #### getting all the data from previous scripts in ####
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/data_seminat.RData")
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_seminat.RData")
@@ -398,19 +411,31 @@ length(unique(ANO.geo$ano_punkt_id))
 length(unique(res.seminat.ANO2$ano_flate_id))
 length(unique(res.seminat.ANO2$ano_punkt_id))
 # by region
-length(unique(res.seminat.ANO2$ano_flate_id[res.seminat.ANO2$region=="Northern Norway"]))
-length(unique(res.seminat.ANO2$ano_flate_id[res.seminat.ANO2$region=="Central Norway"]))
-length(unique(res.seminat.ANO2$ano_flate_id[res.seminat.ANO2$region=="Eastern Norway"]))
-length(unique(res.seminat.ANO2$ano_flate_id[res.seminat.ANO2$region=="Western Norway"]))
-length(unique(res.seminat.ANO2$ano_flate_id[res.seminat.ANO2$region=="Southern Norway"]))
+length(unique(res.seminat.ANO2$ano_flate_id[!is.na(res.seminat.ANO2$region) & res.seminat.ANO2$region=="Northern Norway"]))
+length(unique(res.seminat.ANO2$ano_flate_id[!is.na(res.seminat.ANO2$region) & res.seminat.ANO2$region=="Central Norway"]))
+length(unique(res.seminat.ANO2$ano_flate_id[!is.na(res.seminat.ANO2$region) & res.seminat.ANO2$region=="Eastern Norway"]))
+length(unique(res.seminat.ANO2$ano_flate_id[!is.na(res.seminat.ANO2$region) & res.seminat.ANO2$region=="Western Norway"]))
+length(unique(res.seminat.ANO2$ano_flate_id[!is.na(res.seminat.ANO2$region) & res.seminat.ANO2$region=="Southern Norway"]))
 
[…]

### 2023-08-31 R/functionalPlantIndicators/code wetland/6 post scaling analyses_wetland.R

@@ -620,6 +620,10 @@ points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrog
 
 
 
+# ?
+#The figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests. According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
+
+
 
 #### regressions vs. ANO variables ####
 

### 2023-08-31 functional_plant_indicators_wetland.Rmd
finalizing the fpi markdown for wetlands

@@
 - Northern Norway: 416
-- Central Norway:  362
-- Eastern Norway:  295
-- Western Norway:  151
-- Southern Norway: 127
+- Central Norway:  361
+- Eastern Norway:  294
+- Western Norway:  150
+- Southern Norway: 126
 
@@
 We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
-Since pH and nitrogen show some deviation of the lower pH indicator we exemplify this with these indicators for unscaled values.
+Since pH and nitrogen show some deviation in the lower indicator we exemplify this with these two indicators for unscaled values.
 
@@
 ```
-THe figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests.
+Deviations from the reference occur mainly in V2-C1, which represents limestone-poor swamp forests.
 According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.

### 2023-08-31 R/functionalPlantIndicators/code wetland/6 post scaling analyses_wetland.R

@@ -1,3 +1,16 @@
+library(downloader)
+library(sf)
+library(tidyr)
+library(plyr)
+library(dplyr)
+library(stringr)
+library(tidyverse)
+library(readxl)
+library(tmap)
+library(tmaptools)
+library(betareg)
+library(glmmTMB)
+
 #### getting all the data from previous scripts in ####
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/data_wetland.RData")
 load("P:/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
@@ -258,11 +271,21 @@ length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Eastern Norway"]))
 length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Western Norway"]))
 length(unique(res.wetland2$ano_flate_id[res.wetland2$region=="Southern Norway"]))
 
-length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Northern Norway"]))
-length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Central Norway"]))
-length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Eastern Norway"]))
-length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Western Norway"]))
-length(unique(res.wetland2$ano_punkt_id[res.wetland2$region=="Southern Norway"]))
+length(unique(res.wetland2$ano_punkt_id[!is.na(res.wetland2$region) & res.wetland2$region=="Northern Norway"]))
+length(unique(res.wetland2$ano_punkt_id[!is.na(res.wetland2$region) & res.wetland2$region=="Central Norway"]))
+length(unique(res.wetland2$ano_punkt_id[!is.na(res.wetland2$region) & res.wetland2$region=="Eastern Norway"]))
+length(unique(res.wetland2$ano_punkt_id[!is.na(res.wetland2$region) & res.wetland2$region=="Western Norway"]))
+length(unique(res.wetland2$ano_punkt_id[!is.na(res.wetland2$region) & res.wetland2$region=="Southern Norway"]))
+
+length(unique(res.wetland2$ano_punkt_id[is.na(res.wetland2$region)])) # not assigned a region, in water?
+tm_shape(regnor) +
[…]

### 2023-08-31 R/functionalPlantIndicators/code semi-nat/6 post scaling analyses_seminat.R
finalizing fpi markdown for seminat

@@ -116,9 +116,39 @@ ggplot(res.seminat.ANO, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_in
   ylab("Scaled indicator value (ANO data)") 
 
 
+# making the plot, ANO Light, Moist & pH
+ggplot(data=subset(res.seminat.ANO,!is.na(scaled_value) & fp_ind %in% c("Light1","Moist1","pH1",
+                                                                        "Light2","Moist2","pH2")),
+       aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin(color=NA) +
+  #  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=1, shape=16, color="black") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1",
+                                     "Light2","Moist2","pH2")), ncol = 3) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value (ANO data)") +
+  theme(legend.position="none")
+
+# making the plot, ANO Nitrogen, Phosphorus, Grazing_mowing, Soil_disturbance
+ggplot(data=subset(res.seminat.ANO,!is.na(scaled_value) & fp_ind %in% c("Nitrogen1","Phosphorus1","Grazing_mowing1","Soil_disturbance1",
+                                                                        "Nitrogen2","Phosphorus2","Grazing_mowing2","Soil_disturbance2")),
+       aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin(color=NA) +
+  #  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=1, shape=16, color="black") +
+  facet_wrap(~factor(fp_ind,levels=c("Nitrogen1","Phosphorus1","Grazing_mowing1","Soil_disturbance1",
+                                     "Nitrogen2","Phosphorus2","Grazing_mowing2","Soil_disturbance2")), ncol = 4) + 
+  xlab("Main ecosystem type") + 
+  ylab("Scaled indicator value (ANO data)") +
+  theme(legend.position="none")
+
 
 # making the plot, ASO
[…]

### 2023-09-06 DRAFT_insectIndicators.Rmd
mostly complete version

@@
 <!-- Load all you dependencies here -->
+
+```{r}
+#Check if Norimon and bombLepiSurv is installed
+
+if (!"Norimon" %in% installed.packages()[,"Package"]){
+  devtools::install_github("jenast/Norimon")
+}
+
+if (!"bombLepiSurv" %in% installed.packages()[,"Package"]){
+  devtools::install_github("jenast/bombLepiSurv")
+}
+
+#Alternatively, with pacman 
+#pacman::p_load_gh("NINAnor/Norimon")
+#pacman::p_load_gh("jenast/bombLepiSurv")
+
+```
+
 ```{r setup, include=FALSE}
@@
 
-For the indicators from the NMBB program, community reference values has been elicitated from experts, and is explained further below.
+For the indicators from the NMBB program, community reference values has been elicited from experts, and is explained further below.
 For the indicators from the NorIns program, this is significantly harder to do, because of the size of the communities, lack of historical timeseries and changes in sampling and identification techniques.
@@
                 "Species richness of flying insects",
-                "Species richness of pollinators",
-                "Species richness of X",
-                "Species richness Apocrita / Symphyta",
+                "Species richness of pollinating insects",
+                "Species richness of dung associated insects",
+                "Quotent of Apocrita vs Symphyta species richness",
                 "Intraspesific genetic variation")
 
+nor_cat <- c("Funksjonelt viktige arter og biofysiske stukturer",
+             "Funksjonelt viktige arter og biofysiske stukturer",
+             "Funksjonelt viktige arter og biofysiske stukturer",
+             "Biologisk mangfold",
+             "Biologisk mangfold",
+             "Biologisk mangfold",
+             "Funksjonelt viktige arter og biofysiske strukturer",
+             "Biologisk mangfold"
+)
+
+ect_cat <- c("B1",
+             "B1",
+             "B2",
+             "B1",
+             "B1",
+             "B1",
+             "B2",
+             "B1"
+)
 Geo_scope <- c(rep("Sørlandet, Østfold-Vestfold, Vestlandet, Trøndelag", 2), 
@@
                   "Indicator" = Indicator,
+                  "Norwegian type" = nor_cat,
+                  "ECT" = ect_cat,
                   "Current geographical scope" = Geo_scope,
@@
 ```
-
-[Add short motivation for each indicator]
+Many of these indicators are self-explanatory and don't require much justification.
+**Bumblebees in semi-natural grasslands** and **Butterflies in semi-natural grasslands** are functionally important pollinators, and are as a community dependent on a varied herbaceous flora.
+They therefore both represent key functionality, as well as indicators of habitat quality.
+**Biomass of flying insects** represent a key functional characteristic in ecosystem functioning as food for insectivores, and as a response to habitat quality and management.
+**Species richness of flying insects** represents the total breadth of the insect community, and **Species richness of pollinating insects** complements the butterfly and bumblebee indicators in taxonomic breadth and spatial resolution.
+**Species richness of dung associated insects** represent a key functional group in decomposition, and function as an indicator of grazing intensity and diversity.
+**Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community.
+Apocrita represents predatory and parasitic wasps, while Symphyta is plant-feeding wasps.
+A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control.
+Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it.
+The need to conserve genetic diversity is identified in Norwegian law, and decreased intraspecific diversity could signal population decreases, and decreased potential to adapt to environmental changes, before whole species are lost to an area.
 
 ## About the underlying data
-<!-- Text here -->
+The data used here comes from the ongoing monitoring programs "Norwegian insect monitoring" ([NorIns](https://www.nina.no/Naturmangfold/Insekter/Overv%C3%A5king-av-insekter)) and "Norwegian monitoring of bumblebees and butterflies" ([NMBB](https://www.nina.no/V%C3%A5re-fagomr%C3%A5der/Milj%C3%B8overv%C3%A5king-p%C3%A5-land/Humler-og-dagsommerfugler)).
+Both programs are initiated and financed by the Environmental agency, with the aim to produce continuous, areal representative, time-series of insect community data.
+The **NorIns** program use malaise and window traps to sample a broad community of insect species, which are identified through metabarcoding of DNA.
+The **NMBB** program use sweep netting along transect walks where the species are identified manually in the field.
+More information is available in the links above.
 
 ### Representativity in time and space
-<!-- Text here -->
+The **NorIns** program started in 2020 with monitoring in semi-natural land and forests in the south-eastern region of Norway (Østlandet), with the long-term ambition to cover the entire country.
+In 2021, the program was expanded to semi-natural land in central Norway (Trøndelag), in 2022 to semi-natural land in south-western Norway (Sørlandet), and in 2023 to semi-natural land in northern Norway (Nord-Norge), with a possible expansion to western Norway (Vestlandet) in 2024.
+So far, forest habitats are only monitored in south-eastern Norway.
+
+These 5 regions represents the planned spatial precision of the survey.
+Finer resolution could be evaluated in the future, although it is unlikely to be feasible without more intense sampling.
+This program uses a staggered sampling scheme where we visit different locations each year, and return to each location every 5 years.
+This sampling scheme, coupled with the strong yearly variation in insect abundances makes it reasonable to use a temporal resolution of 5 years.
+[How to deal with this now?
+average over 5 years?
+Add 5 year aggregation in bootstrap_value()]
+
+The **NMBB* program started in 2009 in a subset of south-eastern Norway (counties of Vestfold and Østfold), and was expanded to central Norway (Trøndelag) in 2010, south-western Norway in 2013, and to western Norway (Vestland and Møre og Romsdal fylke) in 2022.
+The spatial resolution aligns with these (as of now) 4 regions, and it is not recommended to downscale the predictions due to the limited number of sampling locations in each region.
+Since the same locations are visited each year, yearly estimates of the condition could be calculated, but should be interpreted by caution due to natural yearly variability.
+
+Both programs are designed to be area representative, within their respective habitat types and regions.
+Both use semi-randomized selection schemes of localities, the **NorIns** selects by random draws, and the **NMBB** by a predefined continuous network of survey squares.
+In both programs however, some subjective decisions are made to define habitat criteria and to discard unpractical survey locations.
 
 ### Original units
-<!-- Text here -->
+The original units of **NorIns** include number of species, total biomass as wet weight, and inter-specific genetic diversity (unit to be determined).
+For the **NMBB** program, original units are number of individuals of each species within a transect, although only presences in transects are used for the indicator calculations.
 
 ### Temporal coverage
-<!-- Text here -->
-
-### Aditional comments about the dataset
-<!-- Text here -->
+2009-present for the **NMBB** data, and 2020-present for the **NorIns** data.
 
@@
 ### Norwegian standard
-<!-- Text here -->
+Four of the indicators are tied to the *Økologisk egenskap* (Ecological characteristic) of **Funksjonelt viktige arter og biofysiske stukturer** (Funktionally important species and structures), and the other four to **Biologisk mangfold** (Biodiversity).
+
+### UN standard
+Six of the indicators are tied to the UN standard (ECT-classes) **B1 Compositional state characteristics**, and the two remaining to **B2 Structural state characteristics**.
 
@@
 
-## Reference state and values
-Assessing the state of insect communities can be a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability.
-Still, the most complicated part for an indicator is to establishing reference values, when the reference states are lacking or prohibitively difficult to measure empirically.
+## Reference condition and values
+The reference condition is one with minimal negative human impact, but this has a special interpretation for semi-natural land.
+These habitats are formed and maintained through human activities.
+Thus, "minimal negative human impact" is replaced by a state of "good maintenance", here understood as resembling the traditional agricultural maintenance regime, existing for several hundred years up until the late 1800s.
+This is characterized by extensive grazing, meadows and lays, crop rotation with legumes to bind nitrogen in addition to fertilization from manure, relatively small field sizes and abundant field margins, and a lack of artificial fertilization and mechanized tilling.
+
+### Reference condition
+Assessing the current state of insect communities may be a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability.
+Still, even more challenging is to establishing reference values, when the reference states are extinct or prohibitively difficult to measure empirically.
 We currently lack reference values for all indicators from the **NorIns** program.
-It is unlikely that this can be solved "simply" by surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely don't exist.
+It is unlikely that this can be solved "simply" by surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely no longer exist.
 The question of how to handle these reference values is currently unresolved, and we present these indicator value calculations with made up reference values as placeholders in the code.
 
-### Reference state
-<!-- Text here -->
+For the **NMBB** indicators, the communities are small enough and well known enough to identify reference communities by expert opinion.
+These communities specify the expected rate of observing each species.
+The current rates of observation are then compared to the expected rates.
+The calculations are further explained below.
 
@@
 <!-- Text here -->
+The specific reference communities for the **NMBB** indicators won't be spelled out here, see instead the scripts below.
+We have not specified a value for _good ecological condition_ as of yet, since there lack a straightforward empirical basis to do so.
 
@@
 
+* Öberg, S., Gjershaug, J. O., Diserud, O. & Ødegaard, F. 2011. Videreutvikling av metodikk for 
+arealrepresentativ overvåking av dagsommerfugler og humler. Naturindeks for Norge. – NINA 
+Rapport 663.
+
 ## Analyses
@@
 ### Calculation principles for NorIns indicators
-The indicators from the Norwegian insect monitoring are calculated with functions in the `Norimon` package.
-This method is meant to capture a broad variety of insect indicators using data from **NorIns**.
+The indicators from the Norwegian insect monitoring are calculated with functions in the `Norimon` R-package [link](https://github.com/NINAnor/Norimon).
+This methodology is meant to facilitate the calculation of a broad variety of insect indicators using data from **NorIns**.
 
@@
 
-1. Fetch data of biomass or community diversity from a centralized database (or GBIF). Diversity data can be subsetted on several taxonomic levels, but biomass is only available for whole samples.
+1. Fetch data of biomass or community diversity from a centralized database. Diversity data can be filtered on several taxonomic levels, but biomass is only available for whole samples.
 2. Bootstrap observations and calculate mean and uncertainty of point estimates.
@@
 Most of the ecology comes into step 1, in choosing what data to use to describe a quality.
-This can be a custom selection of species, or a set of higher taxa such as specific genuses, families or an order.
+This can be a custom selection of species, or a set of higher taxa such as specific genuses, families or even order.
 The goal is to choose a set of taxa that represent specific qualities of the community that can indicate the ecological status of the ecosystem.
@@
 if we should aggregate the data on the region scale, locality scale, or even the individual sample occasion scale.
-For ecological indicators, we will aggregate the values to a year locality scale, where the variables summarises all catches in a locality in a year.
+For ecological indicators, we will aggregate the values to a year locality scale, where the variables summarises all catches in a locality in a year.[NOTE, implement 5 year/region aggregation]
 
@@
 
-For convenience, the pollinator families can be retreived by the `get_pollinators()` function.
+For convenience, the pollinator families can be retrieved by the `get_pollinators()` function.
 
@@
 The default aggregation level is "year_locality", meaning the total observations for a locality within a year.
-[The output currently contains some experimental values.
-Shannon diversity doesn't really make sense without counts/amounts, etc...]
+
+Note that the output currently contains some experimental values, awaiting a more robust methodology.
+"Shannon diversity" doesn't really make sense without counts or amounts, etc.
+This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (Connor et al.
+2023), in the (hopefully near) future.
 
@@
 Individually, they represent a single measurement of an insect community in a habitat type in a region.
-To get the sampling uncertainty for this representation, we can bootstrap the values by this simple prosess: we choose a random set of localities within a year and region (with replacement) and calculate the average values.
+To get the sampling uncertainty for this representation, we can bootstrap the values by this simple process: we choose a random set of localities within a year and region (with replacement) and calculate the average values.
 We then repeat this process a large number of times to get a bootstrap sample of values, which can be used to express the uncertainty in the dataset.
@@
 But bootstrapping works fairly well on small samples and has the advantage that is doesn't make assumptions of the statistical distribution of the errors.
-Note that once we get a larger dataset with repeating surveys every 5 years, it probably makes sense to do some kind of parametric (hierarcichal) modelling of the values instead of this simple bootstrap.
+Note that once we get a larger dataset with repeating surveys every 5 years, aggregating and bootstrapping over a whole 5 year period.
 
@@
 
-But the `boot_stat` object also stores the individual bootstrap values for later computation. [Why is is only 5 994 rows? Fix/figure out]
+But the `boot_stat` object also stores the individual bootstrap values for later computation. By default, we use 999 bootstrap samples, which here results in 999 * 6 = 5994 rows of bootstrap values.
 
@@
 #### Comparing bootstrap values to a reference point
-Setting aside the difficulties in establishing robust reference values, the next step is to compare the observed values (with bootstrapped uncertainty) to the chosen reference value.
+Setting aside the practical difficulties in establishing robust reference values, the next step in the methodology is to compare the observed values (with bootstrapped uncertainty) to the chosen reference value.
 This can be made in several ways.
@@
 After the comparison to a reference value, we need to normalize the indicator values to lie between 0 and 1.
-This can be done in several ways (insert link to scaling chapter).
+This can be done in several ways (se e.g.
+[eaTools](https://github.com/NINAnor/eaTools)).
 The simplest case is to use a linear scaling, with a natural zero.
 This can be done by dividing the indicator values by the highest value state.
-A `boot_stat` object has a `/` function that divides each bootstap value by a given value, truncates the highest values to 1, and recalculates the summary values.
-
-[Show this, but need to think on how to do this with difference values]
+A `boot_stat` object has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values.
+We can also simply divide by the reference state, to compare the values to the reference state, and normalize them in one go.
+Here, we exemplify the method by dividing the value by 30.
+
+```{r poll_richness_scaling}
+diff_poll_richness_boot3 <- poll_richness_boot / 30
+```
+
+```{r}
+diff_poll_richness_boot3
+```
 
@@
 
-```{r beetle_richness_boot}
+```{r poll_richness_boot}
 plot(diff_poll_richness_boot)
@@
 ### Calculation principles for NBBM indicators
-These indicators were developed for the Nature Index of Norway, back in 2010 by Ola Diserud and Sandra Åström [ref], and is calculated routinely on a yearly basis [see xx for the most current report].
+These indicators were developed for the Nature Index of Norway, back in 2010 by Ola Diserud and Sandra Öberg (Öberg et.
+al 2011), and is calculated routinely on a yearly basis (see Åström et.
+al 2022 for the latest report).
 
 These indicators summarize the state of a bumblebee or butterfly community by estimating the difference from a reference community.
-Formally, the community indicator (CI) is expressed as the relative change of the community from a state of reference (SR), were the change is calculated by a state of change (ST).
+Formally, the community indicator (CI) is expressed as the relative change of the community from a state of reference (SR), were the change is calculated by a state of change (SC).
 
@@
 This classification is done by expert opinion, informed by known present and past species distributions.
-This state of reference doesn't contain all current possible species, only a subset of species, making up a historical reference community.
+Note that this state of reference only contain a subset of highlighted species, making up a historical reference community.
 Potential observations of "new" species therefore doesn't inform the indicator value at all.
@@
 Potential increases in observation rates does not inform the indicator.
-Changes for common species is weighted more heavily than changes for less common species, and larger decreases is weighted more heavily than smalle decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.
-
-The indicator values are estimated with uncertainty, by bootstrapping the set of localities within a region, and calculating the indicator for each of 999 such samples.
+Changes for common species is weighted more heavily than changes for less common species, and larger decreases is weighted more heavily than smaller decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.
+
+The indicator values are estimated with uncertainty, by bootstrapping the set of localities within a region, and calculating the indicator for each of 9999 such samples.
 
 #### Calculation example
-These calculations with bootstrap sampling is implemented in the R-package `bombLepiSurv` (Bombus and Lepidoptera Survey).
+These calculations with bootstrap sampling is implemented in the R-package `bombLepiSurv` (Bombus and Lepidoptera Survey) [link](https://github.com/jenast/bombLepiSurv).
 We briefly show this functionality with bumblebees in semi-natural lands as an example.
@@
 
-But for calculating the indicators, we only consider the transects that had were surveyed for the standard 3 times per year, to cover the phenology of the entire season.
-Complete data sets can be fetched by the function `getComplData`, which fetches data from one habitat type and region at a time.
+Note that for calculating the indicator values, we only consider the transects that had were surveyed for the standard 3 times per year, to cover the phenology of the entire season.
+Data of these complete survey rounds can be fetched from the database by the function `getComplData`, which fetches data from one habitat type and region at a time.
 
@@
                                  habitat = "Gressmark"
-                                 ) %>% 
-  select(species = species_latin,
-         everything())
+                                 )
 ```
@@
 ```
-[add all species in exp to plot]
-
-The actual calculation and bootstrapping is done in the function `calcInd`.
+
+The actual calculation and bootstrapping of the indicator is done in the function `calcInd`.
 This function calls the `getComplData` and `getExpectedValues` functions internally.
@@
 This is fetched internally via the functions `getAmountWeights` and `getDiffWeights`, respectively.
-Here the classes is in norwegian (v = common, m = sporadic, s = rare).
+Here the classes is coded in norwegian (v = common, m = sporadic, s = rare).
 
@@
 ```
-Lastly, we specify the number of samples for the bootstrap
+Lastly, we specify the number of samples for the bootstrap. To speed up, only 999 samples is shown here.
 
@@
 
-The result [is going to be] an object of class [xxx]. Add print and plot methods...
+The result is an object of class "comm_index" (community index).
+It comes with some (still rudimentary) print and plotting functions, showing the point estimate and the limits of a 95% confidence interval.
+
+```{r}
+hInd2022TrondGress
+```
+
+```{r}
+plot(hInd2022TrondGress)
+```
 
@@
 They both also export most of their data to GBIF, but those exports need to be restructured before they can be processed in the following scripts.
-
-For the time being, we will here use some internal functions in the bombLepiSurv and `Norimon` packages to fetch the required data.
-[We may add a separate data fetching script later on, describing how to get this data externally in a stable way.]
 
@@
 #### NMBB data
-The data for the bumblebee and butterfly indicators can be accessed through the R-package `bombLepiSurv`. [Replace this with the GBIF version?. At least make the package public.]
+The data for the bumblebee and butterfly indicators can be accessed through the R-package `bombLepiSurv`.
 
@@
 <!-- In case you need to map the indicator value to regions, you can do that here. Remove this chapter if not relevant. -->
-Importing a shape file with the regional delineation.
+We here show how to import a shape file with the regional delineation.
 The indicators associated with the NorIns project can be attributed to the 5 country regions of Norway.
@@
 The indicators from the NMBB program is similary connected to regions, with the exception that the south-east region only covers the old fylke's of Vestfold and Østfold, and that Nord-Norge isn't covered yet.
-[Find the old fylke geometries and add this map to bombLepiSurv]
+
+```{r}
+nbbm_norway_regions <- bombLepiSurv::get_map()
+norway_reg_NBBM <- nbbm_norway_regions %>% 
+  filter(region %in% c("Ost", "Sørlandet", "Trøndelag", "Vestlandet")) %>% 
+  select(region) %>% 
+  group_by(region) %>% 
+  summarize(geom = st_union(geom))
+```
+
+```{r}
+tm_shape(norway_reg_NBBM) + 
+  tm_polygons(col = "region") 
+```
 
@@
 #### Genetic diversity evenness of flying insects in semi-natural land
-
-[This is not fully implemented yet. Need full COI sequence to calculate the true GDE]. 
-This could also be done for a subset of species.
-(doi: https://doi.org/10.1101/2022.02.09.479762)
+This is not fully implemented yet, and the method shown here functions as a placeholder, awaiting a proper calculation.
+Here, the intraspecific variation is based simply on the number of unique genetic variants (ASVs), and their relative abundance measured as number of DNA-reads in the samples.
+But to do this properly, we need to use the full COI sequence to calculate the true GDE (see Connor et al.
+2023).
+This temporarily postponed due to lack of software, but will be updated (and amended here) as soon as possible.
 
@@
 
-[Implement this]
-```{r, eval = F}
+```{r}
 hInd2022VestGress <- calcInd(type = "Humler",
@@
 
-[Implement this]
-```{r, eval = F}
+```{r}
 sInd2022VestGress <- calcInd(type = "Sommerfugl",
@@
 Once this is done, we can plot the timeseries of values using the function `indPlot`.
-
-Todo, put these in the new version of the package.
-```{r}
-#load("out/IndicatorValues2020.Rdata")
-```
 
@@
 
+Once loaded, the yearly datasets can be organized through the convenience function `collectIndData`, and plotted with `indPlot`.
+
 ```{r bombusGressIndPlot}
@@
 ## Prepare export
-<!-- Text here -->
+We here fetch the indicator regions as an sf and join the calculated indicator values.
+
+```{r}
+
+norway_reg_NBBM <- bombLepiSurv::get_map() %>% 
+  filter(region %in% c("Sør", "Vestlandet", "Øst", "Trøndelag")) %>% 
+  select(region) %>% 
+  group_by(region) %>% 
+  summarize(geom = st_union(geom))
+
+hGressIndVals_to_exp <- hGressIndVals %>% 
+  select(year = Year,
+         region = Region,
+         ind_value = Indicator_value,
+         ind_sd = NULL,
+         ind_lower2.5 = lower95,
+         ind_upper97.5 = upper95)
+
+hGressIndVals_to_exp <- norway_reg_NBBM  %>% 
+  inner_join(hGressIndVals_to_exp,
+            by = c("region" = "region"),
+            multiple = "all")
+
+hGressIndVals_to_exp
+
+```
+
+```{r}
+
+sGressIndVals_to_exp <- sGressIndVals %>% 
+  select(year = Year,
+         region = Region,
+         ind_value = Indicator_value,
+         ind_sd = NULL,
+         ind_lower2.5 = lower95,
+         ind_upper97.5 = upper95)
+
+sGressIndVals_to_exp <- norway_reg_NBBM  %>% 
+  inner_join(sGressIndVals_to_exp,
+            by = c("region" = "region"),
+            multiple = "all")
+
+sGressIndVals_to_exp
+
+```
 

### 2023-09-06 DRAFT_insectIndicators.Rmd
a few touch-ups

@@
 _Author and date:_
-Jens Åström
-
-```{r}
-Sys.Date()
-```
+Jens Åström `r Sys.Date()`
+
+# ```{r}
+# Sys.Date()
+# ```
 
@@
 
-#Notes on this draft
-Most things are done but it's still a rough draft. I plan to finish it in august. But you could see where it's headed so raise the alarm if I'm off course.
+# Notes on this draft
+Most things are done but some things needs updates or touching up.
 
@@
 
-* Fill in sections that lack text.
-* Upgrade genetic diversity calculation. The current calculation is just a placeholder, awaiting some new data from Marie.
-* Packaging of R-package versions for reproducibility? Or just point to a version on Github..
-* Make bombLepiSurv public. After touch-up and fixes of github issues.
-* Fix any comments in this draft inside hard brackets.
-* Write something about the choice of each indicator.
+* Upgrade genetic diversity calculation. The current calculation is just a placeholder, awaiting some new data from Marie. Waiting for software installation by IT. Unsure of timeline.
+* Calculate NorIns indicators on 5 year interval. (Moving window?)
+* Fix remaining comments in this draft inside hard brackets.
 
@@
 
-```{r}
+```{r pkg_check, include=FALSE}
 #Check if Norimon and bombLepiSurv is installed
@@
 
-#Alternatively, with pacman 
+#Alternatively, load below with pacman, requires pacman :) 
 #pacman::p_load_gh("NINAnor/Norimon")
@@
 
+suppressPackageStartupMessages({
 require(Norimon)
@@
 require(sf)
-suppressPackageStartupMessages({
 require(tidyverse)
-}
-
-)
+})
 
@@
 **Quotent of Apocrita vs Symphyta species richness** tries to go beyond simple amounts, and capture compositional changes in the insect community.
-Apocrita represents predatory and parasitic wasps, while Symphyta is plant-feeding wasps.
+Apocrita covers predatory and parasitic wasps, while Symphyta is plant-feeding wasps.
 A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control.
@@
 
-The **NMBB* program started in 2009 in a subset of south-eastern Norway (counties of Vestfold and Østfold), and was expanded to central Norway (Trøndelag) in 2010, south-western Norway in 2013, and to western Norway (Vestland and Møre og Romsdal fylke) in 2022.
+The **NMBB** program started in 2009 in a subset of south-eastern Norway (counties of Vestfold and Østfold), and was expanded to central Norway (Trøndelag) in 2010, south-western Norway in 2013, and to western Norway (Vestland and Møre og Romsdal fylke) in 2022.
 The spatial resolution aligns with these (as of now) 4 regions, and it is not recommended to downscale the predictions due to the limited number of sampling locations in each region.
@@
 ### Original units
-The original units of **NorIns** include number of species, total biomass as wet weight, and inter-specific genetic diversity (unit to be determined).
-For the **NMBB** program, original units are number of individuals of each species within a transect, although only presences in transects are used for the indicator calculations.
+The original units from **NorIns** include number of species, total biomass as wet weight, and inter-specific genetic diversity (unit to be determined).
+For the **NMBB** program, original units are number of individuals of each species within a transect, although only presences in transects (as presence frequencies) are used for the indicator calculations.
 
@@
 ## Collinearities with other indicators
-None measured so far, but it is reasonable to assume that some of these indicators will share some causal relationships with eachother, and therefore correlate to some degree.
-This could potentially happen with some other indicators not described here.
+None measured so far, but it is reasonable to assume that some of these indicators will share some causal relationships with each other, and therefore correlate to some degree.
+They might also correlate to other indicators, not described here, as well.
 
 ## Reference condition and values
-The reference condition is one with minimal negative human impact, but this has a special interpretation for semi-natural land.
+The reference condition is meant to be one with minimal negative human impact, but this has a special interpretation for semi-natural land.
 These habitats are formed and maintained through human activities.
@@
 ## Uncertainties
-The uncertainties for the indicator values is calculated by bootstrapping the average values for each locality within a year. This takes the variability between the localities into account.
+The uncertainties for the indicator values is calculated by bootstrapping the average values for each locality within a year.
+This takes the variability between the localities into account.
+For the **NorIns** indicators, this uncertainties are given as standard deviation, as well as confidence intervals.The uncertainties for the **NMBB** indicators are constituted by a distribution of discrete values, and therefore we only provide confidence intervals and not a standard deviation.
 
@@
 ```
+
 We then fetch the community data for these families through the `get_observations` function.
@@
 
-```{r location_map, echo = F}
+```{r location_map}
 tm_shape(norway_regions) +
@@
 
-But the `boot_stat` object also stores the individual bootstrap values for later computation. By default, we use 999 bootstrap samples, which here results in 999 * 6 = 5994 rows of bootstrap values.
+But the `boot_stat` object also stores the individual bootstrap values for later computation.
+By default, we use 999 bootstrap samples, which here results in 999 samples * 6 groups = 5994 rows of bootstrap values.
 
@@
 #### Comparing bootstrap values to a reference point
-Setting aside the practical difficulties in establishing robust reference values, the next step in the methodology is to compare the observed values (with bootstrapped uncertainty) to the chosen reference value.
+Setting aside the practical difficulties in establishing robust reference values, the next step in the methodology is to compare the observed values (with bootstrapped uncertainty) to a chosen reference value.
 This can be made in several ways.
 Most simply, if we have a single numeric value as a reference value, we can simply subtract that from the observed values.
-For example, if we observe 13 species of pollinators in 2022 at a location, and the reference point is 10, the 2022 has increased by 13 - 10 = 3 species.
-Such subtractions should be made on the set of bootstrapped values, and new summary statistics calculated, preserving the uncertainty from the bootstrap.
-The `boot_stat` has its own subtraction method do to just that.
+For example, if we observe 13 species of pollinators in 2022 at a location, and the reference point is 10, the 2022 value has increased by 13 - 10 = 3 species.
+Such subtractions should be made on the set of bootstrapped values, followed by new summary statistics being calculated, preserving the uncertainty from the bootstrap.
+The `boot_stat` has its own subtraction method `-` do to just that.
 In this example, we set the reference value arbitrarily to 30 species.
@@
 This functionality is in development.
-It currently works for single rows as reference points, but needs updating to allow for referencing several values, e.g using the start values for all regions and habitat types as their own reference points.**
+It currently works for single rows as reference points, but needs updating to allow for referencing several values simultaneously, e.g using the start values for all regions and habitat types as their own reference points.**
 
@@
 #### Normalizing the values
-After the comparison to a reference value, we need to normalize the indicator values to lie between 0 and 1.
+After the comparison to a reference value, we need to normalize the indicator values so that they lie between 0 and 1.
 This can be done in several ways (se e.g.
 [eaTools](https://github.com/NINAnor/eaTools)).
-The simplest case is to use a linear scaling, with a natural zero.
-This can be done by dividing the indicator values by the highest value state.
-A `boot_stat` object has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values.
+The simplest case is to use a linear scaling, with a natural zero, which e.g.
+can be done by dividing the indicator values by the highest value state.
+A `boot_stat` class has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values.
 We can also simply divide by the reference state, to compare the values to the reference state, and normalize them in one go.
@@
 #### Map plots
-In addition to the density plots above, we can also display the values geographically, in a map.
-The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.
+In addition to the density plots above, we can also display the values geographically. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.
 
@@
 
-We connect to the internal database with a convenience function `humlesommerfConnect`
+We connect to the internal database with a convenience function.
 ```{r}
@@
 
-Note that for calculating the indicator values, we only consider the transects that had were surveyed for the standard 3 times per year, to cover the phenology of the entire season.
+Note that for calculating the indicator values, we only consider the transects that were surveyed for the standard 3 times per year, to cover the phenology of the entire season without bias.
 Data of these complete survey rounds can be fetched from the database by the function `getComplData`, which fetches data from one habitat type and region at a time.
@@
                                  region_short = "Trond",
-                                 habitat = "Gressmark"
-                                 )
-```
-
-Species in green bars are species that are expected to be common, and should reach the green areas in the plot.
+                                 habitat = "Gressmark")
+```
+
+Species in green bars are species that are expected to be common, who need to reach the green areas in the plot not to decrease the indicator value (from 1).
 Similarly with the blue (sporadic species), and red (rare species).
-We see in this case that Bombus soroeensis is an exptected sporadic species that only were observed rarely.
+We see in this case that Bombus soroeensis is an expected to be a sporadic species, but that it were only observed rarely.
 
@@
 It comes with some (still rudimentary) print and plotting functions, showing the point estimate and the limits of a 95% confidence interval.
+Due to the limited number of species and the fixed "weights" of each species and state of change, the indicator calculation returns a distribution of discrete values.
+Therefore it can happen that the 95% and the 90% confidence intervals are the same, as in this case.
 
@@
 norway_reg_NBBM <- nbbm_norway_regions %>% 
-  filter(region %in% c("Ost", "Sørlandet", "Trøndelag", "Vestlandet")) %>% 
+  filter(region %in% c("Øst", "Sørlandet", "Trøndelag", "Vestlandet")) %>% 
   select(region) %>% 
@@
 ### Calculation of NorIns indicators
+Here we calculate the indicators in abbreviated form, following the general framework outlined above.
+
 #### Biomass of flying insects in semi-natural land
@@
 pollinators_sn_to_exp
+```
+
+#### Species richness of dung associated insects
+Note that the species list will probably be expanded, after a taxonomic review.
+
+Fetch the data.
+
+```{r}
+dung_fam <- c("Muscidae")
+
+dung_sn <- get_observations(subset_year = 2020:2022,
+                                   subset_families = dung_fam,
+                                   subset_region = NULL)
+```
+
+[This will be amended with data based on genus and species lists. Need to use the lowest taxonomic resolution. Will implement a species lookup list in the database]
+
+Calculate the indicator values for each region and year.
+```{r}
+dung_sn_boot <- bootstrap_value(df = dung_sn,
+                                       value = no_species,
+                                       groups = c("year",
+                                                  "region_name")
+                                       )
+```
+Compare to arbitrary reference value and normalize.
+```{r}
+dung_ref <- 90
+
+dung_sn_diff <- dung_sn_boot / dung_ref
+```
+
+Plot the results and display uncertainty.
+```{r}
+dung_sn_diff
+```
+
+```{r}
+plot(dung_sn_diff)
+```
+
+Prepare export format.
+
+```{r}
+dung_sn_to_exp <- dung_sn_diff$bootstrap_summary
+
+dung_sn_to_exp <- dung_sn_to_exp %>% 
+  select(year,
+         region = region_name,
+         ind_value = boot_value,
+         ind_sd = boot_sd,
+         ind_lower2.5 = boot_lower2.5,
+         ind_upper97.5 = boot_upper97.5)
+
+dung_sn_to_exp <- norway_reg_NorIns  %>% 
+  inner_join(dung_sn_to_exp,
+            by = c("region" = "region"),
+            multiple = "all")
+
+dung_sn_to_exp
 ```
@@
 
-surveyBarplot(allBombusGressmark2022) #+
-#  ggtitle("Humler i gressmark 2021")
+surveyBarplot(allBombusGressmark2022)
 
@@
 <!-- Export final file. Ideally a georeferenced shape or raster with indicators values (raw and normalised), reference values and errors. -->
+
+TBD. Need to consult with others.

### 2023-09-07 trophic_levels-biomass_indicators.rmd
Change location of data

@@
 _Author and date:_
-James Speed
+James D. M. Speed, Gunnar Austrheim, Erling J. Solberg
 
-June 2023
+August 2023
 
@@
 
+```{r, echo=F}
+Ecosystem <- "All" # e.g. "Skog og fjell"
+Egenskap  <- "Biomass distribution among trophic levels" # e.g. "Primærproduksjon"
+ECT       <- "Structural state characteristic"
+Contact   <- "James Speed" # e.g. "Anders Kolstad"
+```
+
+```{r, echo=F}
+metaData <- data.frame(Ecosystem,
+                       "Økologisk egenskap" = Egenskap,
+                       "ECT class" = ECT)
+knitr::kable(metaData)
+```
+
 <!-- Don't remove these three html lines -->
@@
 2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N))
+
+## About the underlying data
 
@@
 
+### Representativity in time and space
+
+The data spans 1907 to 2015, with data points roughly every 10 years.
+The whole of Norway is covered.
+Herbivore data is collated at the municipality level (with the exception of semi-domestic reindeer in Finnmark at county level) and carnivore data at the county level.
+Municipalities and counties valid in 2007 were used.
+NPP is at a pixel resolution.
+
+### Original units
+
+The biomass of the trophic levels is as kg km^-2.
+
+### Temporal coverage
+
+The data spans 1907 to 2015, with data points roughly every 10 years.
+
+## Ecosystem characteristic
+
+### Norwegain standard
+
+This indicator is within the ecosystem characterstic "biomass distribution among trophic levels"
+
+### SEEA EA (UN standard)
+
+This indicator alines most closely with B2 Structural state characteristics (although arguments could be made also for B1 and B3).
+
+## Collinearities with other indicators
+
+Colinearlities are likely to exist with climatic variables and disturbance.
+
+## Reference condition and values
+
+### Reference condition
+
+Reference values are taken from published global relationships between plants and vertebrate herbivores ([Fløjgaard et al.
+2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N)).
+
+### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+
+Deviations from the reference (expected) biomass of consumer based on consumed trophic level biomass are expressed as a two-sided percentage.
+Good condition was defined as within +/- 40% of the reference value (i.e.
+that expected from global relationships).
+Poor condition as over +/- 40% of the reference level.
+The choice of 40% is arbitary.
+
+## Uncertainties
+
+Uncertainty around the estimates have not been estimated.
+However, there are potential uncertainties around the reported livestock and hunting statistics and the demographic models used to estimate metabolic biomass at the population level.
+See Austrheim et al.
+2011 and Speed et al.
+2015 for further details.
+
+## References
+
+Austrheim, G., Solberg, E.
+J., & Mysterud, A.
+(2011).
+Spatio‐temporal variation in large herbivore pressure in Norway during 1949‐1999: has decreased grazing by livestock been countered by increased browsing by cervids?.
+Wildlife Biology, 17(3), 286-298.
+Fløjgaard, C., Pedersen, P.
+B.
+M., Sandom, C.
+J., Svenning, J.
+C., & Ejrnæs, R.
+(2022).
+Exploring a natural baseline for large‐herbivore biomass in ecological restoration.
+Journal of Applied Ecology, 59(1), 18-24.
+Hatton, I.
+A., McCann, K.
+S., Fryxell, J.
+M., Davies, T.
+J., Smerlak, M., Sinclair, A.
+R., & Loreau, M.
+(2015).
+The predator-prey power law: Biomass scaling across terrestrial and aquatic biomes.
+Science, 349(6252), aac6284.
+Sobocinski, A. (2022). Mapping large carnivores in Norway across 175 years of changing policy (Master's thesis, NTNU). https://hdl.handle.net/11250/3047580 
+Speed, J.
+D.
+M., Austrheim, G., Kolstad, A.
+L., & Solberg, E.
+J.
+(2019).
+Long-term changes in northern large-herbivore communities reveal differential rewilding rates in space and time.
+PLoS One, 14(5), e0217166.
+
+## Additional information about the data
+
+The raw data is available [here](10.6084/m9.figshare.24015072)
+
+## Analyses
+
 ## Part 1: Total communities
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 #| label: Data import
@@
 names(livestockrast)<- sub(1939,1938,names(livestockrast))
-names(livestockrast)
+#names(livestockrast)
 
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 #| label: Plotting basic data
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 #| label: PlotTotalHerbivoreBiomass
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 #| label: Plot carnivore biomass
@@
 
-```{r}
+```{r, message=FALSE, warning=FALSE}
 #| label: Calculating expected herbivore biomass
 #| cache: TRUE
-# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
-# npprast_yrs<-npprast[[names(npprast)%in%c("NPP_2000","NPP_2009","NPP_2015")]]
-
 expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
@@
   scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2))
-#writeRaster(expected_herbivore_biomass,paste0(pData,"/Trophic_levels/expectedherbivorebiomass.tif"),overwrite=TRUE)
 
@@
 #| cache: TRUE
-#expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
-#livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
-
 totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
@@
 
-#writeRaster(diffherb,paste0(pData,"/Trophic_levels/HerbivoreBiomassDiffExpected.tif"),overwrite=TRUE)
-
 ```
@@
 #| cache: TRUE
-
-#livestockrast<-rast(paste0(pData,"/Trophic_levels/Livestock_RawBiomass.tiff"))
 
@@
   facet_wrap(~lyr)+theme_bw()
-
-#writeRaster(expected_carnivore_biomass,paste0(pData,"/Trophic_levels/expectedcarnivorebiomass.tif"),overwrite=TRUE)
-
-#writeRaster(diffcarn,paste0(pData,"/Trophic_levels/CarnivoreBiomassDiffExpected.tif"),overwrite=TRUE)
 
@@
 #AR50 type raster
-artype50<-rast("/data/R/GeoSpatialData/LandCover/Norway_LandResource_AR50/Processed/Raster/AR50_Norge.tif")
+# options(timeout = max(1000, getOption("timeout")))
+# filedownload<-"downloadAR50.zip"
+# download.file("https://ntnu.box.com/shared/static/mfrqmuyx14gtf8u9b8xlzb391m0f9fhy.zip","downloadAR50.zip")
+# unzip('downloadAR50.zip',exdir="AR50")
 
-#artype50<-rast("Vertebrates/AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif")
+artype50<-rast(paste0(pData, "AR50/AR50_artype_25_ETRS_1989_UTM_Zone_33N.tif"))
 
@@
 
-#Load biomass rasters
-# npprast<-rast("Vertebrates/data/TrophicBiomassData/NPP.tiff")
-# viltrast<-rast("Vertebrates/data/TrophicBiomassData/Vilt_RawBiomass.tiff")
-# livestockrast<-rast("Vertebrates/data/TrophicBiomassData/Livestock_RawBiomass.tiff")
-# carnivorerast<-rast("Vertebrates/data/TrophicBiomassData/Carnivores_RawBiomass.tiff")
-# expected_herbivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedherbivorebiomass.tif")
-# expected_carnivore_biomass<-rast("Vertebrates/outputs/expectedbiomass/expectedcarnivorebiomass.tif")
-
 #Project the AR50 to the NPP rasters
-
-### NEED TO CHECK IF THIS IS THE CORRECT LAYER OR GET THE CORRECT ONE FROM JAMES
 
@@
 
- #writeRaster(forest_diffcarn,paste0(pData,"/Trophic_levels/forest_carnivores_diffexpected.tif"),overwrite=TRUE)
- #writeRaster(forest_diffherb,paste0(pData,"/Trophic_levels/forest_herbivores_diffexpected.tif"),overwrite=TRUE)
 ```
@@
 
-#forest_diffcarn<-rast("Vertebrates/outputs/forest_carnivores_diffexpected.tif")
-#forest_diffherb<-rast("Vertebrates/outputs/forest_herbivores_diffexpected.tif")
-
-#norcounty_shp<-st_read("Vertebrates/data/Processed/","ViltdataCounty")
  #Simplify by county nr

### 2023-09-07 trophic_levels-biomass_indicators.rmd
corrections made in the previous version of the draft moved to the current pull request (minor editing/formatting issues)

@@
 
-Here we will develop indicators of ecological condition for Norway, based on the distribution of biomass between trophic levels - specifically between vegetation and vertebrate consumers.
+Here we will develop indicators of ecological condition for Norway, based on the distribution of biomass between trophic levels – specifically between vegetation and vertebrate consumers.
 This includes three trophic levels, plants, herbivores and carnivores.
@@
 The expected biomass estimates are derived from global relationships between plants and vertebrate herbivores ([Fløjgaard et al.
-2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N))
+2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://doi.org/10.1126/science.aac6284)).
 
@@
 
-The herbivore data is an extension of the data presented in Austrheim et al.
-2008 <https://onlinelibrary.wiley.com/doi/pdf/10.2981/10-038> and Speed et al.
-2015 <https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0217166>.
+The herbivore data is an extension of the data presented in [Austrheim et al.
+(2008)](https://doi.org/10.2981/10-038) and [Speed et al.
+(2015)](https://doi.org/10.1371/journal.pone.0217166).
 For the development of the trophic interaction indicators the same methods were applied to herbivore data going back to 1907 (contra 1949 in the two studies above).
@@
 
-Herbivore species are wild herbivores: moose/elg, red deer/hjort, roe deer/rådyr, wild reindeer/villrein and livestock: cattle/storfe, sheep/sau, horses/hest, goats/geit and semi-domestic reindeer/tamrein
+Herbivore species are wild herbivores: moose/elg, red deer/hjort, roe deer/rådyr, wild reindeer/villrein; and livestock: cattle/storfe, sheep/sau, horses/hest, goats/geit and semi-domestic reindeer/tamrein.
 
 The carnivore data is taken from back-cast population modelling for the large carnivores in Norway (wolf/ulv, brown bear/bjørn, lynx/gaupe and wolverine/jerv).
-The data was based on hunting statistics, and was collated and modelled by Anna Sobocinski in a master thesis: <https://ntnuopen.ntnu.no/ntnu-xmlui/handle/11250/3047580> The thesis also presented metabolic biomass, but we use raw biomass to make the indicators.
+The data was based on hunting statistics, and was collated and modelled by Anna Sobocinski in a [master thesis](https://ntnuopen.ntnu.no/ntnu-xmlui/handle/11250/3047580).
+The thesis also presented metabolic biomass, but we use raw biomass to make the indicators.
 This data was also rasterized.
 
-We calculate indicators (1) across all ecosystems and the total consumer communities (all localities and all herbivores and all carnivores) and (2) specific indicators for each main ecosystem type
+We calculate indicators (1) across all ecosystems and the total consumer communities (all localities and all herbivores and all carnivores) and (2) specific indicators for each main ecosystem type.
 
@@
 
-Colinearlities are likely to exist with climatic variables and disturbance.
+Colinearities are likely to exist with climatic variables and disturbance.
 
@@
 Reference values are taken from published global relationships between plants and vertebrate herbivores ([Fløjgaard et al.
-2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://www.science.org/doi/pdf/10.1126/science.aac6284?casa_token=mM3vaCIMkpAAAAAA:bo4aODtkJORM0kIeFfezII4T_0wo7Y7X8ktWxI6vHN64sGmMw9Xi9xCc_g33c437swJaeTeiEE_N)).
+2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://doi.org/10.1126/science.aac6284)).
 
@@
 Deviations from the reference (expected) biomass of consumer based on consumed trophic level biomass are expressed as a two-sided percentage.
-Good condition was defined as within +/- 40% of the reference value (i.e.
+Good condition was defined as within ±40% of the reference value (i.e.
 that expected from global relationships).
-Poor condition as over +/- 40% of the reference level.
+Poor condition as over ±40% of the reference level.
 The choice of 40% is arbitary.
@@
 
-Uncertainty around the estimates have not been estimated.
+Uncertainties around the estimates have not been estimated.
 However, there are potential uncertainties around the reported livestock and hunting statistics and the demographic models used to estimate metabolic biomass at the population level.
 See Austrheim et al.
-2011 and Speed et al.
-2015 for further details.
+(2011) and Speed et al.
+(2015) for further details.
 
@@
 (2011).
-Spatio‐temporal variation in large herbivore pressure in Norway during 1949‐1999: has decreased grazing by livestock been countered by increased browsing by cervids?.
+Spatio‐temporal variation in large herbivore pressure in Norway during 1949‐1999: has decreased grazing by livestock been countered by increased browsing by cervids?
 Wildlife Biology, 17(3), 286-298.
@@
 
-The raw data is available [here](10.6084/m9.figshare.24015072)
+The raw data is available [here](https://doi.org/10.6084/m9.figshare.24015072).
 
@@
 
-First NPP
+First NPP:
 
@@
          geom_spatraster(data=npprast_yrs)+facet_grid(~lyr)+
-         scale_fill_continuous(low="lightgreen",high="darkgreen",na.value=NA,expression('kgC km'^-2~'year'^-1))+theme_bw()
+         scale_fill_continuous(low="lightgreen",high="darkgreen",na.value=NA,expression('kg C km'^-2~'year'^-1))+theme_bw()
 
@@
 
-Next total herbivore biomass
+Next total herbivore biomass:
 
@@
   geom_spatraster(data=totalherbivorebiomass)+facet_wrap(~lyr)+
-  scale_fill_continuous(trans='log10',low="lightskyblue",high="darkblue", na.value=NA,expression('Herbivore \nbiomass \nkg km'^-2 ))+theme_bw()
+  scale_fill_continuous(trans='log10',low="lightskyblue",high="darkblue", na.value=NA,expression(atop('Herbivore \nbiomass', 'kg km'^-2)))+theme_bw()
 
@@
 
-And finally total carnivore biomass
+And finally total carnivore biomass:
 
@@
   geom_spatraster(data=totalcarnivores)+facet_wrap(~lyr)+
-  scale_fill_continuous(low='white',high='red', na.value=NA, expression('Carnviore\nBiomass \nkg km'^-2 ))+theme_bw()
+  scale_fill_continuous(low='white',high='red', na.value=NA, expression(atop('Carnviore\nbiomass', 'kg km'^-2)))+theme_bw()
 
@@
 
-Now we use the NPP to estimate the expected herbivore biomass Fløjgaard et al. 2015 Figure 1/Table 1 Global relationship (mean NPP1km)
+Now we use the NPP to estimate the expected herbivore biomass (Fløjgaard et al. 2015, Figure 1/Table 1, Global relationship "mean NPP1km"):
 
@@
 
-Now we can calculate the deviation between the observed herbivore biomass and the expected herbivore biomass. We calculate this as a % change.
+Now we can calculate the deviation between the observed herbivore biomass and the expected herbivore biomass. We calculate this as a % change:
 
-100 x (ActualBiomass -- ExpectedBiomass)/(ExpectedBiomass +1)
+100 x (ActualBiomass -- ExpectedBiomass) / (ExpectedBiomass + 1)
 
@@
 A value of 0 indicates that the locality has expected biomass.
-Value of 100 indicates that the locality has 100% (or more) higher biomass than expected Value of -100 indicates that the locality has 100% (or more) lower biomass than expected
+Value of 100 indicates that the locality has 100% (or more) higher biomass than expected.
+Value of -100 indicates that the locality has 100% (or more) lower biomass than expected.
 
@@
    ggtitle("Difference from expected herbivore biomass")+
-   scale_fill_distiller(type="div",limit=c(-100,100),oob=squish, na.value=NA,'% Difference from \nexpected biomass')+
+   scale_fill_distiller(type="div",limit=c(-100,100),oob=squish, na.value=NA,'% difference from \nexpected biomass')+
    facet_wrap(~lyr)+theme_bw()
@@
 This is based on Hatton et al.
-2015 Figure 1 (predator prey biomass relationships)
+(2015, Figure 1, predator prey biomass relationships).
 
@@
   ggtitle('Difference from expected carnivore biomass')+
-  scale_fill_distiller(type="div",limit=c(-100,100),oob=squish,na.value=NA,"%difference from \nexpected biomass")+
+  scale_fill_distiller(type="div",limit=c(-100,100),oob=squish,na.value=NA,"% difference from \nexpected biomass")+
   facet_wrap(~lyr)+theme_bw()
@@
 
-Indicators were not developed for mire ecosystems. This decision was taken in conversation with the mire ecosystem group, on the basis that herbivores have minimum use of the mrie ecosystems.
+Indicators were not developed for mire ecosystems. This decision was taken in conversation with the mire ecosystem group, on the basis that herbivores have minimum use of the mire ecosystems.
 
@@
 Since a suitable ecosystem map is still under development, we use the AR50 land-cover map as a place-holding solution.
-AR50 does not give a good representation of naturlig åpen områder or semi-naturlig mark, so we demonstrate the habitat specific biomass distributions using the forest class in AR50 and the species outlined above.
+AR50 does not give a good representation of naturlig åpne områder or semi-naturlig mark, so we demonstrate the habitat specific biomass distributions using the forest class in AR50 and the species outlined above.
 

### 2023-09-07 DRAFT_functional_plant_indicators_natopen.Rmd
corrections of some types and formatting issues. Nat-open is not yet finished (OK until heading "Analyses").

@@
 
-- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020).
-- as reference data we use (3) generalized species lists developed by Evju et al. 2023 for nat-open ecosystem types.
-- (4) Swedish plant indicator data for light, nitrogen, ans soil disturbance from Tyler et al.
+- as test data we use plant community data from (1) the ANO monitoring scheme (cf. Tingstad et al. 2019) and (2) from the GRUK monitoring scheme (cf. Evju et al. 2020)
+- as reference data we use (3) generalized species lists developed by Evju et al. (2023) for nat-open ecosystem types
+- (4) Swedish plant indicator data for light, nitrogen, and soil disturbance from Tyler et al.
 (2021), and (5) Grime's CSR values for plant species' strategies (towards competition, stress, and ruderal conditions)
@@
 As of spring 2023, the GRUK data comprise 596 vegetation plots in 146 sites, all of which are mapped to the ecosystem type "åpen grunnlendt mark" (T2) and therein to the limestone-rich subtypes T2-C-7 and T2-C-8 in the Norwegian classification system for ecosystem types.
-The vegetation analysis in GRUK record percent cover for every species of vascular plants in 1 sqm plots.
+The vegetation analysis in GRUK records percent cover for every species of vascular plants in 1 m^2 plots.
 In addition, a range of site variables related to ecosystem-specific drivers of state are recorded.
-In the analysis in this document, we use all GRUK monitoring data data, which span the years 2020-22.
-The 2022-data also include an assessment of site condition, which is included in the analysis of results in this document.
+In the analysis in this document, we use all GRUK monitoring data, which span the years 2020-22.
+The 2022 data also include an assessment of site condition, which is included in the analysis of results in this document.
 In the future, these data data may become available in Miljødirektoratets kartkatalog.
@@
 (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well.
-From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.
+From this set, we decided to include indicator data for light, moisture, pH, nitrogen, phosphorus, grazing_mowing, and soil disturbance for semi-natural ecosystems, as these are thought to be subject to potential change due to abandonment, drainage/flooding, pollution, and erosion.
 
 (5) Grime's system of plant strategy scores (Grime 1974) comprises relative (too one another) scores for the competition-, stress-, and disturbance("ruderality")-related life strategy of plant species.
-In the analysis, in this document we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g.
+In the analysis in this document, we use all three variables, C, S and R, as different pressures acting on the ecosystem might change every one of the strategies (e.g.
 alien species for competition, climate change for stress, land use change for ruderality).
@@
 For the nat-open ecosystem types these lists have been newly prepared by Evju et al.
-2023.
+(2023).
 By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
@@
 For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution.
-As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al.
+As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Framstad et al.
 2022), and deviation from the optimal state thus may occur in both direction (e.g.

### 2023-09-07 alien_species.Rmd
Fixed libpath issue
Render the re-ordered book 07092023

@@
 library(knitr)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse)
+library(tidyverse)
 library(sf)
@@
 library(stars)
+library(gdalUtilities)
 knitr::opts_chunk$set(echo = TRUE)
@@
 # If I import the manuicipality names:
-muni_names <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "administrativenhetnavn")
+muni_names <- sf::read_sf(path, layer = "administrativenhetnavn")
 #The encoding option fail for gdb-files, so I fix the names again here
@@
 ```{r kommuner, fig.cap="Municipalities in Norway.", warning=F}
-muni <- st_cast(muni, "MULTIPOLYGON")
+# function to make sure that mutipolygons are returned 
+ensure_multipolygons <- function(X) {
+    tmp1 <- tempfile(fileext = ".gpkg")
+    tmp2 <- tempfile(fileext = ".gpkg")
+    st_write(X, tmp1)
+    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
+    Y <- st_read(tmp2)
+    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
+}
+
+muni <- ensure_multipolygons(muni)
+
 tm_shape(muni)+

### 2023-09-08 trophic_levels-biomass_indicators.rmd

@@
 totalcarnivores<-carnivorerast[[49:60]]
-
+# alter col names to show the year in the facet strip text
+names(totalcarnivores)<-gsub("TotalRawBiomassDensity_", "", names(totalcarnivores))
 ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
   geom_spatraster(data=totalcarnivores)+facet_wrap(~lyr)+
-  scale_fill_continuous(low='white',high='red', na.value=NA, expression(atop('Carnviore\nbiomass', 'kg km'^-2)))+theme_bw()
+  scale_fill_continuous(low='white',high='red', na.value=NA, expression(atop('Carnviore\nbiomass', 'kg km'^-2)))+theme_bw()+
+  theme(strip.text.x = element_text(size = 10))
 
@@
 forest_carn_herb_countyDF<-gather(forest_carn_herb_county,key="Year",value="Difference",-ID,-CountyName)
-forest_carn_herb_countyDF$YearN<-as.numeric(substr(forest_carn_herb_countyDF$Year,2,5))
+#forest_carn_herb_countyDF$YearN<-as.numeric(substr(forest_carn_herb_countyDF$Year,2,5))
+forest_carn_herb_countyDF$YearN<-as.numeric(forest_carn_herb_countyDF$Year)
+
 ggplot(data=forest_carn_herb_countyDF,aes(x=YearN,y=Difference,color=CountyName))+geom_line()+scale_color_discrete()+
@@
 forest_herb_veg_countyDF<-gather(forest_herb_veg_county,key="Year",value="Difference",-ID,-CountyName)
-forest_herb_veg_countyDF$YearN<-as.numeric(substr(forest_herb_veg_countyDF$Year,2,5))
+forest_herb_veg_countyDF$YearN<-as.numeric(forest_herb_veg_countyDF$Year)
 ggplot(data=forest_herb_veg_countyDF,aes(x=YearN,y=Difference,color=CountyName))+geom_line()+scale_color_discrete()+

### 2023-09-08 trophic_levels-biomass_indicators.rmd

@@
    scale_fill_distiller(type="div",palette='RdYlBu',limit=c(-100,100),oob=squish,na.value=NA, "% difference from \nexpected biomass")+
-   facet_wrap(~lyr)+theme_bw()
+   facet_wrap(~lyr)+theme_bw()+
+  labs(x="Year", y="Difference")
 
@@
    scale_fill_distiller(type="div",palette='RdYlBu',limit=c(-100,100),oob=squish,na.value=NA,"% difference from\nexpected biomass")+
-   facet_wrap(~lyr)+theme_bw()
+   facet_wrap(~lyr)+theme_bw()+
+  labs(x="Year", y="Difference")
 

### 2023-09-11 functional_plant_indicators_wetland.Rmd
RenderBook and remove sysDate. Fixed some typos in gjengoing and image widths

--- 2023-08-28 functional_plant_indicators_wetland.Rmd
@@
-# Functional plant indicators, wetland ecosystems (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
+# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
 
@@
 _Author and date:_
+
 Joachim Töpper
@@
 Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
-Here, we combine functional plant indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
+Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
 2019) for wetland ecosystems.
@@
 2020).
-These plant functional condition indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
+These plant functional type indicators are developed following the principles and technical protocol of the IBECA framework (Jakobsson et al.
 2021, Töpper & Jakobsson 2021).
-Note that deviations from the reference may occur in both directions, e.g.
-the moisture signature from the testing data may be higher or lower than in the reference.
-Deviations in these two directions indicate very different environmental phenomena and thus have to be treated separately.
-Therefore, we develop two condition indicators for each functional plant indicator, a lower one and an upper one (see further down for more details).
 
 ## About the underlying data
-In the 'functional plant indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
+In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
 
@@
 'area representative nature monitoring'.
-1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle.
+1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle.
 Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
@@
 In the wetland-analysis in this document, we only use the plots which were classified as one of the wetland ecosystem types in the Norwegian categorization system for eco-diversity (NiN).
-In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which as of today comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
+In the analysis in this document, we use the data available on Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset/Details/2054), which comprises data from the first three ANO-years, 2019-2021, and a total of 8887 plots in 498 sites.
 
@@
 For each ecosystem sub-type and plant functional indicator, the reference value is defined as the median value of the indicator value distribution.
-As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided functional plant indicator, Fremstad et al.
+As in most cases the distributions naturally are two-sided (but see the Heat-requirement indicator in the mountain assessment for an example of a one-sided plant functional indicator, Fremstad et al.
 2022), and deviation from the optimal state thus may occur in both direction (e.g.
@@
 ## Uncertainties
-We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.
+We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
 
@@
 # species with abundance 1 (i.e. a max of 100%, must be included in each sample)
-# each sample re-samples 1/3 of the number of species
+# each sample re-samples 2/3 of the number of species
 # the abundance of the re-sampled species may vary (see bootstrap function for details)
 wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
-                          iter=1000,obl=1,rat=1/3,var.abun=T)
+                          iter=1000,obl=1,rat=2/3,var.abun=T)
 
@@
 wetland.ref.cov<-readRDS(paste0(here::here(), "/data/cache/wetland.ref.cov.RDS"))
-
 #load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/FPI_output large files for markdown/ref_lists_wetland.RData")
@@
 Note that we scale each ANO plot's CWM against either the lower threshold value and the min value OR the upper threshold value and the max value based on whether the CWM is smaller or higher than the reference value.
-Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper condition indicators for each functional plant indicator.
+Since the scaled values for both sides range between 0 and 1, we generate separate lower and upper indicators for each plant functional indicator type.
 An ANO plot can only have a scaled value in either the lower or the upper indicator (the other one will be 'NA'), except for the unlikely event that the CWM exactly matches the reference value, in which case both lower and upper indicator will receive a scaled indicator value of 1.
@@
 
-In order to visualize the results we need to rearrange the results-objects from wide to long format (note that there is both a lower and an upper condition indicator for each of the functional plant indicators).
+### Simple summary statistics
+We can calculate simple summary statistics like means, standard deviations, and number of observations (note that there is both a lower and an upper indicator for each of the four plant functional indicators)
 
@@
     values_to = "scaled_value",
-    values_drop_na = FALSE
+    values_drop_na = TRUE
   )
 
-# add original values as well
-res.wetland <- 
-  res.wetland %>% add_column(original = results.wetland[['original']] %>% 
-                               pivot_longer(
-                                 cols =  c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
-                                 names_to = NULL,
-                                 values_to = "original",
-                                 values_drop_na = FALSE
-                               ) %>%
-                               pull(original)
-  )
+# summarizing the indicator scores
+res.wetland %>%
+  group_by(fp_ind) %>%
+  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value))
 
@@
 # making the plot
-ggplot(res.wetland[!is.na(res.wetland$scaled_value),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
   geom_hline(yintercept=0.6, linetype="dashed") + 
@@
 ```
-The functional signature from the plant communities in wetland ecosystems looks largely good, except for V2 (swamp forests), which show rather low index scores for the lower pH and Nitrogen indicators, i.e.
-suggesting too acidic conditions and too low nitrogen availability.
-
-We can look at V2 specifically to see if this is the case for all sub-types.
-
-```{r}
-# making the plot for V2 only
-ggplot(res.wetland[res.wetland$hovedtype_rute=="V2",], aes(x=factor(kartleggingsenhet_1m2), y=scaled_value, fill=fp_ind)) + 
-  geom_hline(yintercept=0.6, linetype="dashed") + 
-  geom_violin() +
-  #  geom_boxplot(width=0.2, color="grey") +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
-  xlab("Main ecosystem type") + 
-  ylab("Scaled indicator value") 
-```
-The deviations towards low pH and low nitrogen are mainly an issue for V2-C1, limestone-poor swamp forests.
 
@@
 
-### Regions - maps and statistics
-Alternatively we can calculate and show the region-wise means and their related standard errors. But note that calculating a simple mean would be inappropriate for these data. This is because:
-(i) the scaled data are bound between 0 and 1, and thus follow a beta-distribution rather than a Gaussian one
-(ii) the ANO dataset has a nested structure
-
-Therefore, we need to (i) use a beta-model, that (ii) can account for the nested structure of the data.
-Here, we apply the following function using either a glmmTMB null-model with a beta-distribution, logit link, and the nesting as a random intercept, or a simple betareg null-model with logit link if the nesting is not extensive enough for a mixed model.
-```{r}
-library(betareg)
-library(glmmTMB)
-
-expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale
-
-# the function performs a glmmTMB if there's >= 5 random levels in the nesting structure
-# if that is not the case, then the function performs a betareg if theres >= 2 observations
-# if that is not the case either, then the function returns the value of the single observation with NA standard error
-
-indmean.beta <- function(df) {
-
-  st_geometry(df) <- NULL
-  colnames(df) <- c("y","ran")
-
-  if ( nrow(df[!is.na(df[,1]),]) >= 2 ) {
-
-    if ( length(unique(df[!is.na(df[,1]),2])) >=5 ) {
-
-      mod1 <- glmmTMB(y ~ 1 +(1|ran), family=beta_family(), data=df)
-
-      return(c(
-        expit(summary( mod1 )$coefficients$cond[1]),
-
-        expit( summary( mod1 )$coefficients$cond[1] + 
-                 summary( mod1 )$coefficients$cond[2] )-
-          expit( summary( mod1 )$coefficients$cond[1] ),
-
-        nrow(df[!is.na(df$y),]),
-        summary( mod1 )$coefficients$cond[1],
-        summary( mod1 )$coefficients$cond[2]
-      ))
-
-    } else {
-
-      mod2 <- betareg(y ~ 1, data=df)
-
-      return(c(
-        expit(summary( mod2 )$coefficients$mean[1]),
-        expit( summary( mod2 )$coefficients$mean[1] + 
-                 summary( mod2 )$coefficients$mean[2] )-
-          expit( summary( mod2 )$coefficients$mean[1] ),
-        nrow(df[!is.na(df$y),]),
-        summary( mod2 )$coefficients$mean[1],
-        summary( mod2 )$coefficients$mean[2]
-      ))
-
-    }
-
-  } else {
-
-    return(c(df$y,NA,1,NA,NA))
-
-  }
-
-}
-
-```
-
+### Regions
+Alternatively we can calculate and show the region-wise means and standard deviations
 ```{r}
@@
   mutate(
-    pH1.reg.mean = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[1],
-                                 indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[1]
-    ),
-    pH1.reg.se = c(indmean.beta(df=res.wetland2[res.wetland2$region=="Northern Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Central Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Eastern Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Western Norway",c("pH1","ano_flate_id")])[2],
-                               indmean.beta(df=res.wetland2[res.wetland2$region=="Southern Norway",c("pH1","ano_flate_id")])[2]
-    ),
-    pH1.reg.n = c(nrow(res.wetland2[res.wetland2$region=="Northern Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Central Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Eastern Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Western Norway" & !is.na(res.wetland2$pH1),]),
-                              nrow(res.wetland2[res.wetland2$region=="Southern Norway" & !is.na(res.wetland2$pH1),])
-    )
+    pH1.reg.mean = c(mean(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                             mean(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T)),
+    pH1.reg.sd = c(sd(res.wetland2$pH1[res.wetland2$region=="Northern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Central Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Eastern Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Western Norway"],na.rm=T),
+                          sd(res.wetland2$pH1[res.wetland2$region=="Southern Norway"],na.rm=T))
     )
@@
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
-  tm_text("pH1.reg.n",col="black",bg.color="grey")
-```
-Mean index value by region for the lower pH indicator (i.e. index shows deviations towards a more acidic environment). Numbers in grey fields show the number of observations in the respective region.
-
-```{r}
-# se
+  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE)))
+
+# sd
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.se", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE)) +
-  tm_text("pH1.reg.n",col="black",bg.color="grey")
-
-```
-Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
-
-### unscaled values vs. reference
-We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
-Since pH and nitrogen show some deviation of the lower pH indicator we exemplify this with these indicators for unscaled values.
-
-```{r}
-summary(res.wetland$kartleggingsenhet_1m2)
-length(unique(res.wetland$kartleggingsenhet_1m2))
-# 16 NiN-types to plot
-colnames(wetland.ref.cov[['Soil_reaction_pH']])
-
-### pH
-par(mfrow=c(2,4))
-## V1s
-# V1-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
-       col="red")
-
-# V1-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
-       col="red")
-
-# V1-C-3
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C3',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
-       col="red")
-
-# V1-C-4
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C4',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
-       col="red")
-
-# V1-C-5
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C5")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C5',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
-       col="red")
-
-# V1-C-6
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C6',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
-       col="red")
-
-# V1-C-7
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C7',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
-       col="red")
-
-# V1-C-8
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C8',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
-       col="red")
-
-par(mfrow=c(2,4))
-## V2s
-# V2-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
-       col="red")
-
-# V2-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
-       col="red")
-
-# V2-C-3
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C3',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
-       col="red")
-
-## V3s
-# V3-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C1',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
-       col="red")
-
-# V3-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V3-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
-       col="red")
-
-## V4s
-# V4-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V4-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V4-C2',xlab='pH value')
-points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
-       col="red")
-
-## V8s
-# V8-C-1
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C1")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C1',xlab='pH value')
-#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
-       col="red")
-
-# V8-C-2
-plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V8-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C2',xlab='pH value')
-#points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
-       col="red")
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-### Nitrogen
-par(mfrow=c(2,4))
-## V1s
-# V1-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
-       col="red")
-
-# V1-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C2a","V1-C2b","V1-C2c","V1-C2d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-2",]$original)),
-       col="red")
-
-# V1-C-3
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C3a","V1-C3b","V1-C3c","V1-C3d")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C3',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-3",]$original)),
-       col="red")
-
-# V1-C-4
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C4a","V1-C4b","V1-C4c","V1-C4d","V1-C4e","V1-C4f","V1-C4g","V1-C4h")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C4',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-4",]$original)),
-       col="red")
-
-# V1-C-5
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C5")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C5',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-5",]$original)),
-       col="red")
-
-# V1-C-6
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C6a","V1-C6b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C6',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-6",]$original)),
-       col="red")
-
-# V1-C-7
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C7a","V1-C7b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C7',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-7",]$original)),
-       col="red")
-
-# V1-C-8
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C8a","V1-C8b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V1-C8',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-8",]$original)),
-       col="red")
-
-par(mfrow=c(2,4))
-## V2s
-# V2-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
-       col="red")
-
-# V2-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C2a","V2-C2b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-2",]$original)),
-       col="red")
-
-# V2-C-3
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C3a","V2-C3b")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V2-C3',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-3",]$original)),
-       col="red")
-
-## V3s
-# V3-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C1a","V3-C1b","V3-C1c","V3-C1d","V3-C1e")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C1',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-1",]$original)),
-       col="red")
-
-# V3-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V3-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V3-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V3-C-2",]$original)),
-       col="red")
-
-## V4s
-# V4-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V4-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V4-C2',xlab='Nitrogen value')
-points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,na.rm=T),
-       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V4-C-2",]$original)),
-       col="red")
-
-## V8s
-# V8-C-1
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C1")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C1',xlab='Nitrogen value')
-#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-1",]$original)),
-       col="red")
-
-# V8-C-2
-plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V8-C2")]) ,na.rm=T),
-     xlim=c(1,7), type="l", main='V8-C2',xlab='Nitrogen value')
-#points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,na.rm=T),
-#       type="l", col="red")
-points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original,
-       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V8-C-2",]$original)),
-       col="red")
-
-legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-```
-THe figure shows that deviations mainly occur in V2-C1, which represents limestone-poor swamp forests.
-According to the functional signature from the plant community composition many of the occurrences of this ecosystem type are too acidic and may have too low availability of nitrogen.
+  tm_polygons(col="pH1.reg.sd", title="pH (lower), sd", style="quantile", palette=get_brewer_pal(palette="OrRd", n=5, plot=FALSE))
+
+```
 

### 2023-09-11 scalingFunctions.Rmd
RenderBook and remove sysDate. Fixed some typos in gjengoing and image widths

@@
 
-Date: `r Sys.Date()`
+March 2023
+
+**Superseeded :** *The functionality explained here is moved over to the [eaTools](https://github.com/NINAnor/eaTools) package. Please see the documentation there for updated examples.*
 
@@
 Define plotting aesthetics
+
 ```{r}
@@
 high <- "green"
-scale_fill_gradient(low = low, high = high)
 ridge <- geom_density_ridges_gradient(scale = 3, size = .3)
@@
 Create random raw data for two variables
+
 ```{r}
@@
 
-Lets say that the reference condition for variable1 to equal 16, and for variable 2 its 35
+Lets say that the reference condition for `variable1` to equal 16, and for variable 2 its 35
 
@@
 To reverse what is good and what is bad condition: change the sign and add one.
+
 ```{r}
@@
 At least this is true around the reference value.
-But perhaps a variable value around 25 for variabel2 is actually quite bad, and in the above example I have assumed that the value zero is the worst possible value.
-That may not always be the case.
+But perhaps a variable value around 25 for `variable2` is actually quite bad.
+In the above example I have assumed that the value zero is the worst possible value, but that may not always be the case.
 
 ## Linear, with defined lower limit
-We can also say that the worst possible condition for these variables is not zero, but for example say that it is 5 for variable1 and 28 for variable2.
+
+We can also say that the worst possible condition for these variables is not zero, but for example say that it is 5 for `variable1` and 28 for `variable2`.
+
 ```{r}
@@
 We can then normalise the data between these two reference values.
+
 ```{r}
@@
 ## Break point scaling
-We can also define a threshold for good ecological condition. Lets for example say that for variable1, ecosystem function really declines when the variable shows values below 9.
+
+We can also define a threshold for good ecological condition. Lets for example say that for `variable1`, ecosystem function really declines when the variable shows values below 9.
 
@@
 
-Then we can normalise again, using what I refer to as break point normalisation. For the maths, see [here](https://stats.stackexchange.com/questions/281162/scale-a-number-between-a-range)
+Then we can normalise again, using what I refer to as break point normalisation. For the maths, see [here](https://stats.stackexchange.com/questions/281162/scale-a-number-between-a-range).
+
 ```{r}
-dat$indicator1_breakPoint <- ifelse(dat$variable1 < dat$thr1,
+dat$indicator1_breakPoint <- ifelse(
+  dat$variable1 < dat$thr1,
     ((dat$variable1-dat$referenceLow1)/(dat$thr1 - dat$referenceLow1))*(0.6-0)+0,
@@
   geom_segment(aes(x = thr1, y = 0, xend = thr1, yend = 0.6),
-               linetype="dashed",
+               linetype="solid",
                size=1.5)+
   geom_segment(aes(x = 0, y = 0.6, xend = thr1, yend = 0.6),
-               linetype="dashed",
+               linetype="solid",
                size=1.5)+
@@
 
+The solid lines point to the indicator value where the variable equals the threshold value.
+
 ## Two-sided indicator
 
-An indicator may have an optimum value somewhere in the middle of its range, rather than a max or min value. You can then define a two-sided indicator.
-Lets say that variable2 is optimal when its at a level of 30, and values either lower or high are both bad.
+An indicator may have an optimum value somewhere in the middle of its range, rather than a max or min value.
+You can then define a two-sided indicator.
+Lets say that `variable2` is optimal when its at a level of 30, and values either lower or high are both bad.
 
@@
 ```{r}
-dat$indicator2_twoSided <- ifelse(dat$variable2<dat$referenceMid2,
+dat$indicator2_twoSided <- ifelse(
+  dat$variable2<dat$referenceMid2,
     (dat$variable2-dat$referenceLow2)/(dat$referenceMid2-dat$referenceLow2),
@@
 ## Sigmoid transformation
+
 There are many other, non-linear ways to normalise an indicator.
@@
 It is less sensitive to changes around the min and max reference values.
-A slight deviation from the reference condition can often not be described as a true reduction in condition.
-This is true for example when reference communities (ecosystemns defines as having or representing good condition) themselves show variation in the variable values, yet the reference value is fixed.
+A slight deviation from the reference condition should perhaps not often not be described as a true reduction in condition.
+This is true for example when reference communities (ecosystems defined as having, or representing, good condition) themselves show variation in the variable values, yet the reference value is fixed.
 A linear scaling will in these cases lead to what is referred to as underestimation biaz (making thing look worse than they really are).
@@
 Finding the transformed indicator value when the variable equals the threshold
+
 ```{r}
@@
 Also plotted against the linearly normalised indicator values:
+
 ```{r}
@@
 
+The dotted helper lines selve as visual guides to judge how different a scaled indicator value of 0.6 is for the two approaches.
+
 ### Accounting for the threshold value
@@
 We can account for the threshold for good ecological condition when applying sigmoid transformation like this
+
 ```{r}
-dat$indicator1_sigmoid_thr <- ifelse(dat$variable1<dat$thr1,
+dat$indicator1_sigmoid_thr <- ifelse(
+  dat$variable1<dat$thr1,
     ((dat$indicator1_sigmoid-0)/(sigmoid_thr-0))*(0.6-0)+0,
@@
 
+The grey circles and horizonal line correspons to the sigmoid transformation without any adjustment for the threshold value.
+
 ## Exponential functoins
-Here are two example of exponential function:
+
+Here are two examples of exponential transformation that can also be used:
 

### 2023-09-11 gjengroing.Rmd

@@
 
-*Author and date:* Zander Venter
-
-```{r}
-Sys.Date()
-```
+*Author and date:*
+
+Zander Venter
+
+September 2023
 
@@
 The population sample are the areas used to determine ecological condition relative to the reference areas.
-We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (ie.
+We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (i.e..
 poor condition),
@@
 
-### Aditional comments about the dataset
+### Additional comments about the dataset
 
@@
 
-![](images/gjengroing_schematic.jpg)
+![](images/gjengroing_schematic.jpg){width="70%"}
 
@@
 We calculate the mean reference value for each unique combination of elevation-bioclimatic zone.
-When calculating the index for each population polygon, the reference value is inhereted from the elecation-bioclimatic zone it falls within.
+When calculating the index for each population polygon, the reference value is inherited from the elevation-bioclimatic zone it falls within.
 
@@
 We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition.
-For each unique spatial combination of these strata (n = 21), we calcualte a mean reference value later on in the analysis.
+For each unique spatial combination of these strata (n = 21), we calculate a mean reference value later on in the analysis.
 Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.
@@
 Therefore we will use the standard deviation in reference values per region.
-The standard deviation, as a percentage of the average reference value is used to calcualte the uncertainty in the scaled indicator value.
+The standard deviation, as a percentage of the average reference value is used to calculate the uncertainty in the scaled indicator value.
 
@@
 For a given bioclimatic-elevation strata, this could lead to higher reference values for vegetation height than one would expect with a representative reference sample.
-This could consequently result in high elevation population polygons appearing to have very good condition ebcause they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata.
-This is a possible explanation for why the high elevation parts of norway appear to be in very good ecological condition.
+This could consequently result in high elevation population polygons appearing to have very good condition because they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata.
+This is a possible explanation for why the high elevation parts of Norway appear to be in very good ecological condition.
 
@@
 The choice of 200m buffer zone was largely based on computational limitations.
-Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex plygons takes time, even in GEE.
+Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex polygons takes time, even in GEE.
 It may make more ecological sense to define a larger buffer zone to use for defining reference for poor condition.
@@
 We chose random locations spread across the 5 regions in Norway.
-The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosysystem polygons from AR5.
-The polygons have jaggered edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space.
+The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosystem polygons from AR5.
+The polygons have jagged edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space.
 Each polygon is colored based on the final scaled gjengroing indicator score (0 to 1).
@@
 in poor condition) due to the growth of a forest in the northern section.
-It looks like a plantation forest and possibly occured after the wetland was drained.
+It looks like a plantation forest and possibly occurred after the wetland was drained.
 The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m.
 Therefore it gets a scaped indicator score of 0.1.
-In contrast the wetland plygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.
-
-![](images/val_1.jpg)
+In contrast the wetland polygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.
+
+![](images/val_1.jpg){width="50%"}
 
@@
 The lower polygon has poor ecological condition because its median vegetation height is 6.2m which is close to the local reference value for mature forest (7.9m).
-Although the centeral polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.
-
-![](images/val_2.jpg)
+Although the central polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.
+
+![](images/val_2.jpg){width="50%"}
 
@@
 
-![](images/val_3.jpg)
+![](images/val_3.jpg){width="50%"}
 
@@
 
-![](images/val_4.jpg)
+![](images/val_4.jpg){width="50%"}
 
 The fifth and final example from further north in Nord-norge shows two våtmark polygons with contrasting condition scores, yet exactly the same median vegetation heights of 0.6m.
-The difference in the scaled iindicator value is due to different local reference values for surrounding forest.
+The difference in the scaled indicator value is due to different local reference values for surrounding forest.
 The polygon on the left has surrounding forest (within 200m buffer zone) with median height 0.5m, while the polygon on the right has surrounding forest of 0.7m.
-This creates a contrast in the final indicator valye and illustrates the effect that a local reference value can have on defining good and poor ecological condition.
-
-![](images/val_5.jpg)
+This creates a contrast in the final indicator value and illustrates the effect that a local reference value can have on defining good and poor ecological condition.
+
+![](images/val_5.jpg){width="50%"}
 

### 2023-09-11 DRAFT_functional_plant_indicators_natopen.Rmd
Update DRAFT_functional_plant_indicators_natopen.Rmd
backing up the last changes (small layout changes)

@@
 
-ANO data: ANO.sp contains the species data, ANO.geo contains site data
+ANO data: ANO.sp contains the species data, ANO.geo contains site data.
 ```{r, echo = F}
@@
 
-GRUK data: GRUK.species contains the species data, GRUK.variable contains site data, GRUK2021.condition contains a field-based condition assessment from the 2021 season
+GRUK data: GRUK.species contains the species data, GRUK.variable contains site data, GRUK2021.condition contains a field-based condition assessment from the 2021 season.
 ```{r, echo = F}
@@
 
-Plant indicators from Tyler et al. (2021) and Grime (1974)
+Plant indicators from Tyler et al. (2021) and Grime (1974) are saved as ind.Tyler and ind.Grime.
 ```{r, echo = F}
@@
 
-Generalized species lists (reference communities): natopen_NiN_ref contains the reference species lists, natopen_NiN_ref_spInfo contains additional taxonomic information for each species
+Generalized species lists (reference communities): natopen_NiN_ref contains the reference species lists, natopen_NiN_ref_spInfo contains additional taxonomic information for each species.
 ```{r, echo = F}

### 2023-09-11 DRAFT_functional_plant_indicators_natopen.Rmd
Update DRAFT_functional_plant_indicators_natopen.Rmd
some more minor changes

@@
 For semi-natural ecosystems, we are testing "Light", "Moisture", "Soil_reaction_pH", "Nitrogen", "Phosphorus", "Grazing_mowing", and "Soil_disturbance".
-In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data) the NiN lists can be bootstrapped.
+In order to get distributions of CWMs rather than one single value (for comparison with the empirical testing data), the NiN lists can be bootstrapped.
 

### 2023-09-11 functional_plant_indicators_wetland.Rmd

--- 2023-08-31 functional_plant_indicators_wetland.Rmd
@@
 'area representative nature monitoring'.
-1000 sites are randomly distributed across mainland Norway and visited within a 5-year cycle.
-Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 18 evenly distributed vegetation analyses in 1 x 1 m plots (up to 18, because some of these evenly distributed points may be in water or otherwise inaccessible).
+1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle.
+Each ANO site spans a 500 x 500 m grid cell, and the data collection at each ANO site includes up to 12 evenly distributed vegetation analyses in 1 x 1 m plots (up to 12, because some of these evenly distributed points may be in water or otherwise inaccessible).
 For the vegetation analyses, the cover of each vascular plant species in the plot is recorded.

### 2023-09-11 rosslyng.Rmd

--- 2022-11-15 DRAFT_rosslyng.Rmd
@@
----
-title: "DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei"
-output:
-  html_document
----
+# (PART\*) DEPRECATED{-}
 
-# DEPRECATED: Areal uten død eller skadet røsslyng i kystlynghei
+# Areal uten død eller skadet røsslyng i kystlynghei
 
@@
 _Author and date:_
+
 Anders L. Kolstad
 
-```{r}
-Sys.Date()
-```
+March 2023
 
@@
 ## Introduction
-Here I quickly go through the ANO data set for the years 2019-2021 to investigate the frequeny of poins that fall within coastal heathands (kystlynghei).
-This will decide wether we can use this data to inform an indikcator about the proportion of dead Calluna vulgaris (røsslyng).
+Here I quickly go through the ANO data set for the years 2019-2021 to investigate the frequeny of poins that fall within coastal heathlands (kystlynghei).
+This will decide whether we can use this data to inform an indicator about the proportion of dead _Calluna vulgaris_ (røsslyng).
 
@@
 
-Imnport and list terms: 
+Import and list terms: 
 ```{r}
@@
 ```
-Only 26 of the 250m2 circles are kystlynghei.
+Only 26 of the 250m2 circles are kystlynghei. The variable of interest is called _Dekning av død/skadet røsslyng_ and is only recorded on these 26 circles.
 
 ## Conclusion
-Because of the random allocation of sampling points, not enough points fall in coastal heathlands to allow us to create a trustworth indicator for this nature type.
-With a few more years of data we might end up with about 100 points, which is borderline what we need in order to calculate condition values at a regional level with some level of precition.
+Because of the random allocation of sampling points, not enough points fall in coastal heathlands to allow us to create a trustworthy indicator for this nature type.
+With a few more years of data we might end up with about 100 points, which is borderline what we need in order to calculate condition values at a regional level with some level of precision.

### 2023-09-12 functional_plant_indicators_wetland.Rmd
fix FPI_wetland bug

@@
 
-# summarizing the indicator scores
-res.wetland %>%
-  group_by(fp_ind) %>%
-  dplyr::summarize(Mean = mean(scaled_value, na.rm=TRUE), SD = sd(scaled_value, na.rm=TRUE), N = length(scaled_value))
+# add original values as well
+res.wetland <-
+res.wetland %>% add_column(original = results.wetland[['original']] %>%
+pivot_longer(
+cols = c("Light1","Light2","Moist1","Moist2","pH1","pH2","Nitrogen1","Nitrogen2"),
+names_to = NULL,
+values_to = "original",
+values_drop_na = FALSE
+) %>%
+pull(original)
+)
 

### 2023-09-12 DRAFT_NDVI_wetland.Rmd
jglu

--- 2023-07-03 DRAFT_ndvi_wetland.Rmd
@@
 For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed.
-In both cases, MODIS-derived NDVI values (MOD13Q1 V6-product) were used.
-National maps for forest and mountain ecosystems were used to define the areas of interest.
+In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e.
+national maps for the forest and mountain ecosystems).
 Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
@@
 we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single wetland ecosystem occurrences ('polygons' hereafter) includes many wetland areas which are smaller than the resolution of the MODIS imagery (250 x 250 m), which was used for forests and mountains and has a time series going back to the year 2000.
+2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
 
 In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
-1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the correct ecosystem.
+1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m, which is smaller than almost all wetland polygons in the NiN-data.
+2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-Thus, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
+In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
 
@@
 
-(1) NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the correct locations and to provide information on the field-assessed ecological condition of mapped polygons.
+- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+- Sentinel-2 satellite NDVI data
+- MODIS satellite NDVI data
+- Landsat satellite NDVI data
 
-(2) Sentinel-2 satellite-derived NDVI data (10 x 10 m resolution) - See https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4 for the Google Earth Engine code used to process Sentinel imagery and export NDVI data.
+(1) NiN ecosystem data:
 
-(3) Landsat satellite-derived NDVI data (30 x 30 m resolution).
-See https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7 for the Google Earth Engine code used to process Landsat imagery and export NDVI data.
+(2) Sentinel-2 satellite NDVI data:
 
-(4) MODIS satellite-derived NDVI data (250 x 250 m resolution).
-The MOD13Q1 data product was used for NDVI values.
-The product provides a vegetation index value (NDVI) per pixel.
-MODIS NDVI is calculated from atmospherically-corrected bi-directional surface reflectances that have been masked for water, clouds, heavy aerosols and cloud shadows.
-The algorithm chooses the best available pixel value from all the acquisitions from the 16 day period.
-See https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84 for the Google Earth Engine code used to export NDVI data
+(3) MODIS satellite NDVI data:
 
-We have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which greatly reduces the number of available polygons.
-Thus, to explore temporal trends further back in time, we also include Landsat-derived NDVI data, which span a time frame from 1984 to 2021, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
+(4) Landsat satellite NDVI data:
+
+have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which reduces the number of available polygons greatly.
+To explore temporal trends even further back in time, we also include Landsat NDVI data, which span a time frame from 1984 to 2021, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
 
@@
 
-The NDVI data from either satellite follows the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than is the case for the NDVI data from Sentinel-2 and Landsat satellites.
+The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
 
@@
 
-NDVI Literature
-
-Cunliffe, A.M., Assmann, J.J., Daskalova, G.N., Kerby, J.T.
-& Myers-Smith, I.H.
-2020.
-Aboveground biomass corresponds strongly with drone-derived canopy height but weakly with greenness (NDVI) in a shrub tundra landscape.
-Environmental Research Letters 15(12): 125004.
-
-Karlsen, S.R., Elvebakk, A., Høgda, K.A.
-& Grydeland, T.
-2014.
-Spatial and temporal variability in the onset of the growing season on Svalbard, Arctic Norway—measured by MODIS-NDVI satellite data.
-Remote Sensing 6(9): 8088-8106.
-
-Taddeo, S., Dronova, I.
-& Depsky, N.
-2019.
-Spectral vegetation indices of wetland greenness: Responses to vegetation structure, composition, and spatial distribution.
-Remote sensing of Environment 234: 111467.
-
-Tshabalala, N.N., Mutanga, O. & Sibanda, M. 2021. The Utility of Sentinel-2 MSI Data to Estimate Wetland Vegetation Leaf Area Index in Natural and Rehabilitated Wetlands. Geographies 1(3): 178-191.
-
-Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. NTNU.
-
 ## Analyses
@@
 -Sentinel-2 NDVI data
+-MODIS NDVI data
 -Landsat NDVI data
--MODIS NDVI data
 
@@
 The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
-The image collection contains Sentinel imagery from June, July and August 2015-2022.
-The GEE script creates a folder in Google Drive and places the exports there.
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files.
+The image collection contains Sentinel imagery from June, July and August 2015-2022
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
 ```{r}
@@
 The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
-The image collection contains MODIS imagery from June, July and August 2000-2022. The GEE script creates a folder in Google Drive and places the exported file there.
+The image collection contains MODIS imagery from June, July and August 2000-2022
 
@@
 The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
-The image collection contains Landsat imagery from June, July and August 1984-2022.
-The GEE script creates a folder in Google Drive and places the exports there.
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files.
-This script merges them and then merges the dataframe to the NiN data.
+The image collection contains Landsat imagery from June, July and August 1984-2022
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
 ```{r}

### 2023-09-12 functional_plant_indicators_natopen.Rmd

--- 2023-09-11 DRAFT_functional_plant_indicators_natopen.Rmd
@@
-# Functional plant indicators, naturally open ecosystems below the tree line (Grime's CSR values, Light, Nitrogen, Soil disturbance) {#Functional-plant-indicators-seminat}
+# Functional plant indicators, naturally open ecosystems below the tree line (Grime's CSR values, Light, Nitrogen, Soil disturbance) {#Functional-plant-indicators-open}
 

### 2023-09-12 functional_plant_indicators_wetland.Rmd

@@
 values_to = "scaled_value",
-values_drop_na = TRUE
+values_drop_na = FALSE
 )

### 2023-09-12 slitasje.Rmd

@@
-# (PART\*) INDICATORS{-}
-
-# Slitasje og kjørespor
+# Slitasje og kjørespor {#slitasje}
 
@@
 library(eaTools)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
- library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse)
+library(tidyverse)
 library(tibble)

### 2023-09-12 DRAFT_NDVI_wetland.Rmd
fsøgjs
ndvi_wetland_continued

@@
-# Functional plant indicators, wetland ecosystems (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
+# NDVI (greenness) indicator, wetland ecosystems  {#NDVI-indicator-wetland}
 
@@
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
+                                                        subtype = str_remove(subtype, '-'),
                                                         subtype = str_remove(subtype, '-'))
@@
 summary(SentinelNDVI.wetland$area_meters)
-
 hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
 abline(v=100,lty=2)
+# wetland polygons stretch a large number of sizes from just above 100 sqm to mor ethan 4 million sqm. 75% of all polygons are below 14264 sqm.
 
@@
 # the pattern is strange, suggest to omit 2022 for this analysis
-
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
 
-## NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
+## looking at major types for polygons in good condition only (and only for NDVI data matching NiN-mapping)
 SentinelNDVI.wetland %>%
+  filter(tilstand == 'God') %>%
   filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
+  ggplot(aes(x=hovedtype, y=mean )) + 
   geom_violin() +
-  facet_wrap( ~hovedtype)
-# NDVI largely varies between hovedtyper: higher in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
-# @Tilstand: in seminat types, kaldkilde, and bogs ndvi seems to increase as condition deteriorates, in fens not much of a pattern
+  geom_point(size=0.7, shape=16, color="grey")
+# NDVI varies between hovedtyper: highest in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
 
@@
   geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
   facet_wrap( ~hovedtype)
-# NDVI varies between ecosystem subtypes
-# -at least equally much as between main ecosystem types
-# -stronger than between condition classes in previous plots
+# NDVI varies between ecosystem subtypes in V1. For the other types things seem more constant, but there's also less data and less minor ecosystem types present in the data.
 
-## looking at ndvi in different condition classes, separately for main types and regions  (only for NDVI data matching NiN-mapping)
+## NDVI across regions for polygons in good condition only (for major types and NDVI years data matching NiN-mapping years)
 SentinelNDVI.wetland %>%
-  group_by(id, year) %>%
+  filter(tilstand == 'God') %>%
+  filter(year == kartleggingsaar) %>%
+  filter(!is.na(region)) %>%
+  ggplot(aes(x=region, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap( ~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
+# no obvious trends in geography
 
+## NDVI across condition classes (for major types and NDVI years data matching NiN-mapping years)
+SentinelNDVI.wetland %>%
   filter(year == kartleggingsaar) %>%
@@
   geom_violin() +
-  facet_grid( region~hovedtype)
-# V4 has lacking data for quite some condition classes in most regions
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap( ~hovedtype)
+# in seminat types (V9,V10), kaldkilde (V4), and bogs (V3) ndvi seems to increase as condition deteriorates, in fens (V1) not much of a pattern
 
@@
 
-???
+In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
+In natural fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates. In bogs (V3) there is no consistent pattern.
 
@@
 )
-# add variable for Satellite indentity
+# add variable for Satellite identity
 allSatNDVI.wetland$Sat <- c(

### 2023-09-12 DRAFT_insectIndicators.Rmd
fixed date on top
Update DRAFT_insectIndicators.Rmd
corrected some typos

@@
 Jens Åström `r Sys.Date()`
-
-# ```{r}
-# Sys.Date()
-# ```
 
@@
 These indicators were developed for the Nature Index of Norway, back in 2010 by Ola Diserud and Sandra Öberg (Öberg et.
-al 2011), and is calculated routinely on a yearly basis (see Åström et.
+al 2011), and are calculated routinely on a yearly basis (see Åström et.
 al 2022 for the latest report).
@@
 The state of reference (SR) represents a community of species that can be expected to be observed in a given habitat type and region.
-It is calculated by assigning each species to class of expected commonality; common (C), sporadic (S) and rare (R) species.
+It is calculated by assigning each species to class of expected commonality: common (C), sporadic (S) and rare (R) species.
 This classification is done by expert opinion, informed by known present and past species distributions.
-Note that this state of reference only contain a subset of highlighted species, making up a historical reference community.
-Potential observations of "new" species therefore doesn't inform the indicator value at all.
-Common species is expected to be observed in at least 5 % of the surveyed transects in a habitat type and region.
-Sporadic species similarly is defined as having a presence below 5 % but above 1 %.
-Rare species is seen in no more than 1 % of the transects, and lastly, species not seen in any transect is assumed as lost (L) for the purpose of this calculation.
-
-Each commonality class gets it's own weight, so that common species inform the reference state more than sporadic species, followed by rare species.
-The state of reference (SR) is thus a value, where each species adds to the state according to its commonality.
+Note that this state of reference only contains a subset of highlighted species, making up a historical reference community.
+Potential observations of "new" species therefore don't inform the indicator value at all.
+Common species are expected to be observed in at least 5 % of the surveyed transects in a habitat type and region.
+Sporadic species similarly are defined as having a presence below 5 % but above 1 %.
+Rare species are seen in no more than 1 % of the transects, and lastly, species not seen in any transect are assumed as lost (L) for the purpose of this calculation.
+
+Each commonality class gets its own weight, so that common species inform the reference state more than sporadic species, followed by rare species.
+The state of reference (SR) is thus a value where each species adds to the state according to its commonality.
 Formally, it is defined as:
@@
 $$SR = n_C * w_{C, SR} + n_S * w_{S, SR} + n_R * w_{R, SR} = \sum_{i = (C, S, R)} n_i * w_{i, SR}, $$
-where $n_i$ is the number of species in a commonality class (C, S, R), and the weights $[w_{C, SR}, w_{S, SR}, w_{R, SR}]$ specifies their respective contribution to the reference state (SR).
-The weights used is $[w_{C, SR}, w_{S, SR}, w_{R, SR}] = [1.0, 0.75, 0.50]$, i.e.
+where $n_i$ is the number of species in a commonality class (C, S, R), and the weights $[w_{C, SR}, w_{S, SR}, w_{R, SR}]$ specify their respective contribution to the reference state (SR).
+The weights used are $[w_{C, SR}, w_{S, SR}, w_{R, SR}] = [1.0, 0.75, 0.50]$, i.e.
 a sporadic species has 75 % the weight of a common species, and a rare species has 50 % the weight of a common species.
@@
 Similarly, $n_{CR}$ is the number of expected species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on.
-Potential increases in observation rates does not inform the indicator.
-Changes for common species is weighted more heavily than changes for less common species, and larger decreases is weighted more heavily than smaller decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.
+Potential increases in observation rates do not inform the indicator.
+Changes for common species are weighted more heavily than changes for less common species, and larger decreases are weighted more heavily than smaller decreases, such that $[w_{CS}, w_{CR}, w_{CL}, w_{SR}, w_{SL}, w_{RL}]= [0.50, 0.75, 1.00, 0.50, 0.75, 0.50]$.
 
@@
 #### Calculation example
-These calculations with bootstrap sampling is implemented in the R-package `bombLepiSurv` (Bombus and Lepidoptera Survey) [link](https://github.com/jenast/bombLepiSurv).
+These calculations with bootstrap sampling are implemented in the R-package `bombLepiSurv` ([Bombus and Lepidoptera Survey](https://github.com/jenast/bombLepiSurv)).
 We briefly show this functionality with bumblebees in semi-natural lands as an example.
@@
 
-We can fetch the all observations of bumblebees in semi-natural land with the `getAllData` function. Species names is here shown in Norwegian.
+We can fetch all observations of bumblebees in semi-natural land with the `getAllData` function. Species names are here shown in Norwegian.
 
@@
 
-A summary of the records is most easily plotted by `surveyBarPlot`
+A summary of the records is most easily plotted by `surveyBarPlot`:
 
@@
 Similarly with the blue (sporadic species), and red (rare species).
-We see in this case that Bombus soroeensis is an expected to be a sporadic species, but that it were only observed rarely.
+We see in this case that _Bombus soroeensis_ is expected to be a sporadic species, but that it was only observed rarely.
 
@@
 This is fetched internally via the functions `getAmountWeights` and `getDiffWeights`, respectively.
-Here the classes is coded in norwegian (v = common, m = sporadic, s = rare).
+Here the classes are coded in Norwegian (v = common, m = sporadic, s = rare).
 
@@
 ```
-Lastly, we specify the number of samples for the bootstrap. To speed up, only 999 samples is shown here.
+Lastly, we specify the number of samples for the bootstrap. To speed up, only 999 samples are used here.
 

### 2023-09-12 DRAFT_insectIndicators.Rmd
cache data

--- 2023-09-06 DRAFT_insectIndicators.Rmd
@@
 
-```{r}
+```{r, eval=FALSE}
 connect_to_insect_db()
@@
 
-```{r}
+```{r, eval=FALSE}
 
@@
 ```
+
 ```{r, echo = FALSE}
+poll_loc_year <- readRDS("data/cache/poll_loc_year.RDS")
 poll_loc_year %>% 
@@
 
-```{r}
+```{r, eval=FALSE}
 semi_loc <- get_localities(dataset = "NasIns",
@@
 
+```
+
+```{r, echo=FALSE}
+semi_loc<-readRDS("data/cache/semi_loc.RDS")
+```
+
+```{r}
 norway_regions <- get_map()
@@
 We connect to the internal database with a convenience function.
-```{r}
+```{r, eval=FALSE}
 bombLepiSurv::humlesommerfConnect()
@@
 
-```{r, fig.cap = "Bumblebees in grasslands 2021"}
+```{r,eval=FALSE, fig.cap = "Bumblebees in grasslands 2021"}
 allBombusGrassland2022 <- getAllData(type = "bumblebees",
@@
 
+```{r, echo=FALSE}
+allBombusGrassland2022<-readRDS("data/cache/allBombusGrassland2022.RDS")
+```
+
 A summary of the records is most easily plotted by `surveyBarPlot`
@@
 
-```{r}
+```{r, eval=FALSE}
 bombus_trond_2022 <- getComplData(type = "Humler",
@@
 
+```{r, echo=FALSE}
+bombus_trond_2022<-readRDS("data/cache/bombus_trond_2022.RDS")
+```
+
 The observed relative to the expected occurrences can be visualized through the `plotArt` function. This requires a reference community, which can be fetched through the `getExpValues` function.
 
-```{r}
+```{r, eval=FALSE}
 exp_bombus_trond <- getExpValues(type = "Humler",
@@
                                  habitat = "Gressmark")
+```
+
+```{r}
+exp_bombus_trond<-readRDS("data/cache/exp_bombus_trond.RDS")
 ```

### 2023-09-12 DRAFT_insectIndicators.Rmd
Update DRAFT_insectIndicators.Rmd
correction of some more typos

--- 2023-09-12 DRAFT_insectIndicators.Rmd
@@
 There are a few different ways to access the data required for these indicators.
-Both the NMBB and the NorIns project stores its data in an internal database at NINA.
+Both the NMBB and the NorIns project store their data in an internal database at NINA.
 They both also export most of their data to GBIF, but those exports need to be restructured before they can be processed in the following scripts.
@@
 
-The indicators from the NMBB program is similary connected to regions, with the exception that the south-east region only covers the old fylke's of Vestfold and Østfold, and that Nord-Norge isn't covered yet.
+The indicators from the NMBB program are similary connected to regions, with the exception that the south-east region only covers the old counties of Vestfold and Østfold, and that Nord-Norge isn't covered yet.
 
@@
 Compare values to a reference state.
-The reference state is here set uniformly for all regions, with a made up value just to be able to continue the calculations.
+The reference state is here set uniformly for all regions, with a made-up value just to be able to continue the calculations.
 Alternatively, we could calculate the indicator values for each region separately, compare to individual reference states, and then put it all together again.
@@
 
-Divide the richness of parasitica by symphyta. [Could make a function for this later on.]
+Divide the richness of Parasitica by Symphyta. [Could make a function for this later on.]
 
@@
 
-Bootstrap the fraction of parasitica to symphyta.
+Bootstrap the fraction of Parasitica to Symphyta.
 
@@
 
-The calculations for earlier years is already present as datasets in the R-package and can be fetched individually.
+The calculations for earlier years are already present as datasets in the R-package and can be fetched individually.
 Once this is done, we can plot the timeseries of values using the function `indPlot`.

### 2023-09-12 DRAFT_insectIndicators.Rmd

--- 2023-09-12 DRAFT_insectIndicators.Rmd
@@
 
-# Notes on this draft
+<!--# Notes on this draft
 Most things are done but some things needs updates or touching up.
@@
 * Calculate NorIns indicators on 5 year interval. (Moving window?)
-* Fix remaining comments in this draft inside hard brackets.
+* Fix remaining comments in this draft inside hard brackets.-->
 
@@
 
+```{r, echo=FALSE}
+allBombusGrassland2022<-readRDS("data/cache/allBombusGrassland2022.RDS")
+```
+
 A summary of the records is most easily plotted by `surveyBarPlot`:
@@
 
-```{r}
+```{r, eval=FALSE}
 getDiffWeights()
@@
 ```
-```{r}
+
+```{r, echo=FALSE}
+diffWeights<-readRDS("data/cache/diffWeights.RDS")
+diffWeights
+```
+
+```{r, eval=FALSE}
 getAmountWeights()
 ```
+
+```{r, echo=FALSE}
+amountWeights<-readRDS("data/cache/amountWeights.RDS")
+amountWeights
+```
+
 Lastly, we specify the number of samples for the bootstrap. To speed up, only 999 samples are used here.
@@
 
-```{r}
+```{r, eval=FALSE}
 hInd2022TrondGress <- calcInd(type = "Humler",
@@
 
+```{r, echo=FALSE}
+hInd2022TrondGress<-readRDS("data/cache/hInd2022TrondGress.RDS")
+```
+
 The result is an object of class "comm_index" (community index).
@@
 ```{r}
-connect_to_insect_db()
+#connect_to_insect_db()
 norway_regions <- Norimon::get_map()
@@
 Fetch the data.
-```{r}
+```{r, eval=FALSE}
 biomass_sn <- get_biomass(subset_year = 2020:2022,
@@
 ```
+
+```{r, echo=FALSE}
+biomass_sn<-readRDS("data/cache/biomass_sn.RDS")
+```
+
 Calculate the indicator values for each region and year.
@@
 Fetch the data.
-```{r}
+```{r, eval=FALSE}
 richness_sn <- get_observations(subset_year = 2020:2022,
@@
 ```
+
+```{r, echo=FALSE}
+richness_sn<-readRDS("data/cache/richness_sn.RDS")
+```
+
 Calculate the indicator values for each region and year.
@@
 
+```
+
+```{r, eval=FALSE}
 pollinators_sn <- get_observations(subset_year = 2020:2022,
@@
 ```
+
+```{r, echo=FALSE}
+pollinators_sn<-readRDS("data/cache(pollinators_sn.RDS")
+```
+
 Calculate the indicator values for each region and year.
@@
 ### Bumblebees in open landscapes
-```{r}
+```{r, eval=FALSE}
 require(bombLepiSurv)
@@
 
-```{r bombusGressBarplot, fig.cap = "Bumblebees in grasslands 2021"}
+```{r bombusGressBarplot,eval=FALSE, fig.cap = "Bumblebees in grasslands 2021"}
 allBombusGressmark2022 <- getAllData(type = "bumblebees",
@@
 
+```
+
+```{r, echo=FALSE}
+allBombusGressmark2022<-readRDS("data/cache/allBombusGressmark2022.RDS")
+
 surveyBarplot(allBombusGressmark2022)
-
-```
-
-```{r}
+```
+
+```{r, eval=FALSE}
 nIter = 999
@@
 
-```{r}
+```{r, eval=FALSE}
 hInd2022SorGress <- calcInd(type = "Humler",
@@
 
-```{r}
+```{r, eval=FALSE}
 hInd2022OstGress <- calcInd(type = "Humler",
@@
 
-```{r}
+```{r, eval=FALSE}
 hInd2022VestGress <- calcInd(type = "Humler",
@@
 
-```{r}
+```{r, eval=FALSE}
 nIter = 999
@@
 
-```{r}
+```{r, eval=FALSE}
 sInd2022SorGress <- calcInd(type = "Sommerfugl",
@@
 
-```{r}
+```{r, eval=FALSE}
 sInd2022OstGress <- calcInd(type = "Sommerfugl",
@@
 
-```{r}
+```{r, eval=FALSE}
 sInd2022VestGress <- calcInd(type = "Sommerfugl",

### 2023-09-13 DRAFT_NDVI_wetland.Rmd
ndvi wetland, condition variables

@@
                                         "Svært redusert" = "Svaert_redusert"))
+nin <- nin %>% mutate(tilstand_num = recode(tilstand,
+                                        "God" = 1, "Moderat" = 2, 
+                                        "Redusert" = 3, "Svaert_redusert" = 4))
+
 summary(as.factor(nin$tilstand))
-
-##############################
-##############################
-# code on ninbeskrivelsesvariabler here if relevant
-##############################
-##############################
+summary(as.numeric(nin$tilstand_num))
 
@@
   drop_na(tilstand) %>%
-  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, ninbeskrivelsesvariabler, kartleggingsaar)
+  dplyr::select(id, hovedoekosystem, hovedtype, ninkartleggingsenheter, lokalitetskvalitet, tilstand, tilstand_num, ninbeskrivelsesvariabler, kartleggingsaar)
 
@@
 
+# splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
+nin.wetland2 <- nin.wetland %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
+  separate(col = ninbeskrivelsesvariabler,
+           into = c("NiN_variable_code", "NiN_variable_value"),
+           sep = "_",
+           remove=F) %>%
+  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
+  mutate(NiN_variable_code = as.factor(NiN_variable_code)) %>%
+  subset(select = -ninbeskrivelsesvariabler)
+nin.wetland2 <- as.data.frame(nin.wetland2)
+
 ```
@@
 nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
+nin.wetland2 <- readRDS(paste0(here::here(),"/data/cache/nin.wetland2.RDS"))
 
@@
   geom_point(size=0.7, shape=16, color="grey")
-# NDVI varies between hovedtyper: highest in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
+# NDVI varies between hovedtyper: highest in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3). The overlaps are considerable.
 
@@
   facet_wrap( ~hovedtype)
-# NDVI varies between ecosystem subtypes in V1. For the other types things seem more constant, but there's also less data and less minor ecosystem types present in the data.
+# NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), but the overlap is large.
 
@@
   facet_wrap( ~hovedtype)
-# in seminat types (V9,V10), kaldkilde (V4), and bogs (V3) ndvi seems to increase as condition deteriorates, in fens (V1) not much of a pattern
+# in seminat types (V9,V10), kaldkilde (V4), and bogs (V3) ndvi seems to increase as condition deteriorates, in fens (V1) not much of a pattern.
+Overlaps between condition classes are generally very large.
+```
+
+Which ecological relevant variable(s) is the condition measure in the NiN data related to?
+```{r}
+nin.wetland2 %>%
+  filter(hovedtype=='V9') %>%
+  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "1AG-B"="Shrub cover",
+                                 "7GR-GI"="Abundance of ditches",
+                                 "1AG-A-G"="Tree cover",
+                                 "1AR-C-L"="Woody plants in field layer",
+                                 "7TK"="Heavy vehicles",
+                                 "PRSL"="Wear/tear")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  theme(legend.position="none")
+
+nin.wetland2 %>%
+  filter(hovedtype=='V10') %>%
+  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7JB-BA"="Land use intensity",
+                                 "7RA-SJ"="Succession",
+                                 "7FA"="Alien species",
+                                 "7JB-GJ"="Fertilization")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
+  theme(legend.position="none")
+
+nin.wetland2 %>%
+  filter(hovedtype=='V1') %>%
+  filter(NiN_variable_code %in% c("7GR-GI","PRTK","7FA","7TK","7SE","PRSL")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7GR-GI"="Abundance of ditches",
+                                 "PRTK"="Heavy vehicles",
+                                 "7FA"="Alien species",
+                                 "7TK"="Heavy vehicles2",
+                                 "7SE"="Wear/tear",
+                                 "PRSL"="Wear/tear2")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  theme(legend.position="none")
 
 ```
+In these plots, condition=1 is best and 4 is worst (degraded). The panels for V1, V9 and V10 differ because Mdir-instruks uses different variables in these major ecosystem types to define condition.
+
+In V9, semi-natural mires, reduced condition is mainly related to increasing cover of shrubs and trees.
+In V10, semi-natural wet-meadows, reduced condition is related to insufficient land-use intensity and succession.
+In both cases, we'd expect an increase in the greenness signal.
+Thus, NDVI can, in principle, be indicative of ecological condition in semi-natural wetlands due to a direct connection to the main pressures on the ecosystems.
+
+In V1 (minerotrophic mires), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
+Under these drivers, greenness would not directly be affected.
+although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run.
+Currently, NDVI seems poorly linked to ecological condition in and the major pressures on minerotrophic mires.
 
 ### Regression analyses Sentinel-2, NDVI as a function of condition
+We can investigate further how NDVI and condition are connected using regression analyses.
 ```{r}
-# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that handles bounded data
+# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data
 # we can transform the variable to be bounded between 0 and 1 and use beta-regression models
@@
 In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
-In natural fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates. In bogs (V3) there is no consistent pattern.
+This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural wetland ecosystems.
+In natural fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates.
+A factor with potential for affecting vegetation composition and productivity, and thus NDVI, is Nitrogen deposition.
+Nitrogen deposition is higher in Southern parts of Norway than further north, and could thus be responsible for increased NDVI with reduced condition in the South, but it is not clear how Nitrogen deposition and the variables underlying the condition assessment in V1 (Ditches, Alien species, Heavy vehicles, Wear/tear) are connected.
+In bogs (V3) there is no consistent pattern.
 
 ### Regression analyses, NDVI across time
+As a last step, we can investigate how NDVI has changed over time.
+
 First, there's again some data handling to do. We merge the Sentinel, Modis, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
@@
 
-# check if the three Sattelite objects have the same structure (for concatenating them)
+# check if the three Satelite objects have the same structure (for concatenating them)
 names(SentinelNDVI.wetland)

### 2023-09-13 R/NDVI/1_data_handling.R
dfksl

@@ -319,6 +319,7 @@ levels(nin.wetland$region)
 #### save and load processed NiN-data from cache ####
 saveRDS(nin, "data/cache/nin.RDS")
 saveRDS(nin.wetland, "data/cache/nin.wetland.RDS")
+saveRDS(nin.wetland2, "data/cache/nin.wetland2.RDS")
 saveRDS(nin.seminat, "data/cache/nin.seminat.RDS")
 saveRDS(nin.natopen, "data/cache/nin.natopen.RDS")
 nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))

### 2023-09-13 DRAFT_NDVI_wetland.Rmd
ndvi wetland year regressions for good condition

@@
 
-### Regression analyses, NDVI across time
-As a last step, we can investigate how NDVI has changed over time.
+### NDVI across time - Sentinel, MODIS & LandSat
+As a last step, we can investigate how NDVI has changed over time. For this we include data from the MODIS and LandSat satellites in addition to Sentinel.
 
@@
 ```
-Now we can plot the time series for each main ecosystem type, and for each satellite separately  
+Now we can plot the time series for each main ecosystem type, showing each satellite separately  
 ```{r}
@@
 
-As a final step, we model NDVI as a function of year and main ecosystem type separately for each Satellite
+We can test if there is consistent tempral changes in NDVI by running regressions including year as a covariate.
+
+We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
+```{r}
+model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])
+
+summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
+summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
+summary(model.wetland.time.GodTilst.Land)$coefficients$cond
+```
+
 ```{r}

### 2023-09-13 R/NDVI/2_regression_analyses_across_time.R

@@ -86,6 +86,16 @@ allSatNDVI.wetland %>%
 
 
 # modeling NDVI as a function of year and main ecosystem type separately for each Satellite
+# only good condition
+model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])
+
+summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
+summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
+summary(model.wetland.time.GodTilst.Land)$coefficients$cond
+
+# all condition
 model.wetland.time.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland)
 model.wetland.time.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland)
 model.wetland.time.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland)

### 2023-09-13 insectIndicators.Rmd
Render book with insect chapter

@@
----
-title: "Template for indicator documentation"
-output:
-  html_document
----
-
 <!-- This is a template for how to document the indicator analyses.
@@
 
+```{r, echo=FALSE}
+connect_to_insect_db(user=Sys.getenv("INSECT_USER"), password=Sys.getenv("INSECT_PASSWORD"))
+```
+
 For convenience, the pollinator families can be retrieved by the `get_pollinators()` function.
@@
 
-```{r, eval=FALSE}
+```{r}
 
@@
 
-```{r, echo = FALSE}
-poll_loc_year <- readRDS("data/cache/poll_loc_year.RDS")
+```{r}
 poll_loc_year %>% 
@@
 
-```{r, eval=FALSE}
+```{r}
 semi_loc <- get_localities(dataset = "NasIns",
@@
 
-```
-
-```{r, echo=FALSE}
-semi_loc<-readRDS("data/cache/semi_loc.RDS")
 ```
@@
 
+```{r, echo=FALSE}
+bombLepiSurv::humlesommerfConnect(username = Sys.getenv("BOMB_USER"), Sys.getenv("BOMB_PASSWORD"))
+```
+
 We can fetch all observations of bumblebees in semi-natural land with the `getAllData` function. Species names are here shown in Norwegian.
 
-```{r,eval=FALSE, fig.cap = "Bumblebees in grasslands 2021"}
+```{r, fig.cap = "Bumblebees in grasslands 2021"}
 allBombusGrassland2022 <- getAllData(type = "bumblebees",
@@
 
-```{r, echo=FALSE}
-allBombusGrassland2022<-readRDS("data/cache/allBombusGrassland2022.RDS")
-```
-
 A summary of the records is most easily plotted by `surveyBarPlot`:
@@
 
-```{r, eval=FALSE}
+```{r}
 bombus_trond_2022 <- getComplData(type = "Humler",
@@
 
-```{r, echo=FALSE}
-bombus_trond_2022<-readRDS("data/cache/bombus_trond_2022.RDS")
-```
-
 The observed relative to the expected occurrences can be visualized through the `plotArt` function. This requires a reference community, which can be fetched through the `getExpValues` function.
 
-```{r, eval=FALSE}
+```{r}
 exp_bombus_trond <- getExpValues(type = "Humler",
@@
                                  habitat = "Gressmark")
-```
-
-```{r}
-exp_bombus_trond<-readRDS("data/cache/exp_bombus_trond.RDS")
 ```
@@
 
-```{r, eval=FALSE}
+```{r}
 getDiffWeights()
@@
 
-```{r, echo=FALSE}
-diffWeights<-readRDS("data/cache/diffWeights.RDS")
-diffWeights
-```
-
-```{r, eval=FALSE}
+```{r}
 getAmountWeights()
@@
 
-```{r, echo=FALSE}
-amountWeights<-readRDS("data/cache/amountWeights.RDS")
-amountWeights
-```
-
 Lastly, we specify the number of samples for the bootstrap. To speed up, only 999 samples are used here.
@@
 
-```{r, eval=FALSE}
+```{r}
 hInd2022TrondGress <- calcInd(type = "Humler",
@@
 
-```{r, echo=FALSE}
-hInd2022TrondGress<-readRDS("data/cache/hInd2022TrondGress.RDS")
-```
-
 The result is an object of class "comm_index" (community index).
@@
 
-```{r}
-#connect_to_insect_db()
+```{r, eval=FALSE}
+connect_to_insect_db()
 norway_regions <- Norimon::get_map()
+
+```
+
+```{r, echo=FALSE}
+connect_to_insect_db(user=Sys.getenv("INSECT_USER"), password=Sys.getenv("INSECT_PASSWORD"))
+norway_regions <- Norimon::get_map()
+
+```
+
+```{r}
 norway_reg_NorIns <- norway_regions %>% 
@@
 nbbm_norway_regions <- bombLepiSurv::get_map()
+```
+
+```{r}
+
 norway_reg_NBBM <- nbbm_norway_regions %>% 
@@
 Fetch the data.
-```{r, eval=FALSE}
+```{r}
 biomass_sn <- get_biomass(subset_year = 2020:2022,
                           subset_region = NULL)
-```
-
-```{r, echo=FALSE}
-biomass_sn<-readRDS("data/cache/biomass_sn.RDS")
 ```
@@
 Fetch the data.
-```{r, eval=FALSE}
+
+```{r}
 richness_sn <- get_observations(subset_year = 2020:2022,
                                 subset_region = NULL)
-```
-
-```{r, echo=FALSE}
-richness_sn<-readRDS("data/cache/richness_sn.RDS")
 ```
@@
 
-```{r, eval=FALSE}
+```{r}
 pollinators_sn <- get_observations(subset_year = 2020:2022,
@@
                                    subset_region = NULL)
-```
-
-```{r, echo=FALSE}
-pollinators_sn<-readRDS("data/cache(pollinators_sn.RDS")
 ```
@@
 Compare to arbitrary reference value and normalize.
+
 ```{r}
@@
 ```
+
 Calculate the indicator values for each region and year.
+
 ```{r}
@@
 
-```{r bombusGressBarplot,eval=FALSE, fig.cap = "Bumblebees in grasslands 2021"}
+```{r, echo=FALSE}
+require(bombLepiSurv)
+humlesommerfConnect(username = Sys.getenv("BOMB_USER"), password = Sys.getenv("BOMB_PASSWORD"), host = "ninradardata01.nina.no")
+
+```
+
+```{r bombusGressBarplot, fig.cap = "Bumblebees in grasslands 2021"}
 allBombusGressmark2022 <- getAllData(type = "bumblebees",
@@
 
-```
-
-```{r, echo=FALSE}
-allBombusGressmark2022<-readRDS("data/cache/allBombusGressmark2022.RDS")
-
 surveyBarplot(allBombusGressmark2022)
+
 ```
@@
 ```{r}
-
 norway_reg_NBBM <- bombLepiSurv::get_map() %>% 
@@
   summarize(geom = st_union(geom))
+
+```
+
+```{r}
 

### 2023-09-13 Nature Index product.Rmd
Add Natureindexproduct

# Nature index product

<br />

_Author and date:_

Chloé R. Nater

September 2023

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(NIflex)
library(magrittr)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "Våtmark (også fjell og skog)" 
Egenskap  <- "Biologisk mangfold" 
ECT       <- "Compositional state characteristic"
Contact   <- "Chloé R. Nater" 
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
The Norwegian Nature Index (NI) is a composite metric for quantifying the state of biodiversity in an abundance-context. It has been established and published for the first time in 2010 and has since been updated for years 2011, 2012, 2013,2014, and 2019. It is owned and presented by the Norwegian Environmental Agency and the latest iteration, including a lot of information, can be found on its webpage: https://www.naturindeks.no/

NI is a composite index made up of single- and multi-species (abundance) indicators that are scaled to fall between 0 and 1 (where 0 = absent/extinct and 1 = abundance corresponding to expected abundance in a natural environment with minimum human disturbance). During the aggregation process, indicators are  weighed on three levels: functional group (trophic weights), spatial coverage (area weights)and key function (key weights).

The methodology for calculating NI is well established and has been described in  detail in NINA reports 347 [@nybo2008a], 425 [@nybo2008b], 797 [@pedersen2012], 1130 [@pedersen2015], and 1886 [@jakobsson2020] and several international scientific articles [e.g. @certain2011] and will not be elaborated on further here.

It is important to note that the NI products that are used indicators in Ecosystem Condition Assessment and the Norwegian Nature Index that has been previously published, and will be published again in 2025, as a national biodiversity assessment are not identical. 
We use modified versions of the NI in Ecosystem Condition Assessments, the details of which will be described below. Moving forward, it will be important to stress this distinction, and we will therefore start talking about "NI products" when referring to composite indices built based on data in the Nature Index Database (see below) that are not identical to the regularly published national assessments (i.e. what is published on https://www.naturindeks.no/).

## About the underlying data
The data underlying NI products is stored in the Nature Index Database at https://naturindeks.nina.no/. The database needs login credentials for access.

In the database, data on a large number of indicators are stored separately for different areas and different years. The areas and years available for each indicator vary but most indicators have data for years 2000, 2010, 2014, and 2019. The database currently also holds data for 1990, but for many indicators, no data is available for this year. New data will be added in 2024, and we are currently working on extending the database such that it may also hold data from years in between official NI updates.

The data held in the NI database are a mixture of observational data, modelled estimates, and quantitative expert assessments.

Here, we focus primarily on an NI product for wetlands, and the following table gives an overview of the indicators included in the official NI for wetlands in 2020. It also provides information on each indicators trophic level, underlying data, whether we expect updated indicator data towards NI 2025, and additional notes.

```{r, echo = FALSE}

indicatorTable <- readRDS(paste0(here::here(),"/data/indicatorTable.RDS"))

indicatorTable %>%
  kableExtra::kbl(caption = "Overview of indicators used in wetlands NI 2020:") %>%
  kableExtra::kable_styling(fixed_thead = TRUE) 
```

### Representativity in time and space
NI products for inclusion in Ecological Condition Assessments are can currently be calculated for years 1990, 2000, 2010, 2014, and 2019, which correspond to the years in which the official NI updates took place. 
Ecosystem level NI products are calculated for 5 major regions in Norway: "Nord-Norge" (northern Norway), "Sør-Norge" (southern Norway), "Midt-Norge" (central Norway), "Vestlandet" (western Norway), and "Østlandet". 
In theory, it would be possible to calculate NI products also for other, including smaller, spatial units, but this would require additional coding. It is also important to note that the spatial resolution of the indicators that make up a specific NI product can vary substantially. As many indicators have rather crude spatial resolution, scaling NI products down to smaller spatial units may not make a lot of sense. 
Consequently, it is important to keep in mind that NI products, even plotted to a map, quantify trends in biodiversity at relatively large spatial scales, and are not suitable for quantifying and comparing biodivseristy in small and very localised areas.

### Original units
The NI is a composite index made up of scaled indicators. NI products are therefore numbers between 0 and 1 (more specifically a set of simulated numbers resulting from a bootstrap, and derived summary metrics including mean, median, and 95 \% confidence intervals) that have no unit.

### Temporal coverage
As mentioned above, NI products can currently be calculated for years 1990, 2000, 2010, 2014, and 2019. From 2025, calculations based on new data from 2024 will become available.

### Aditional comments about the dataset
There are missing values in the NI database. The proportion of missing data values depends on the individual indicators that are to be included in an NI product. 
As part of the NI analysis workflow, missing data values can be imputed using the MICE (Multivariate Imputation by Chained Equations) algorithm. This is an imputation procedure that uses (temporal) correlations among included indicators to make inference about missing values. Uncertainty due to NA imputation is propagated into the final NI product. 
Since the imputation procedure is based on correlations, it will be influenced by the composition of the dataset, i.e. the choice of indicators, years, and locations will affect the parameters in the imputation routine. 
For now, we recommend using data for all official NI years when calculating products with NA imputation as this will be the approach that most closely resembles the routine used for the official regular NI analyses.

## Ecosystem characteristic
### Norwegain standard
NI products are regarded as indicators of biodiversity ("biologisk mangfold"). 
It is important to note though that NI products represent species abundance in this context and are **not** indicators of species richness or species distribution.

### SEEA EA (UN standard)
NI products are tagged as "Compositional state characteristic". 
This seems like the obvious choice for biodiversity indicators that focus on species richness, but may need revisiting for NI products, which are indicators of species abundance.

## Collinearities with other indicators
There is a substantial potential for collinearity between NI products and other indicators, primarily due to the fact that a range of indicators that may be part of an NI product for a given ecosystem may also enter the Ecological Condition Assessment as indicators in their own right or through other composite indicators. 
For the wetlands ecosystem, this is the case, for example, for the bird indicators based on data from "Hekkefuglovervåkingen". These indicators form an important proportion of wetlands NI products (at least how they are defined now) but also enter into the Ecological Condition Assessment as a composite bird indicator. Collinearity between the two indicators/indices should be checked.

## Reference condition and values
### Reference condition
The reference condition is defined along the same lines for NI and for other indicators in Ecological Condition Assessments.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
As NI products are already scaled to range from 0 to 1 and reference conditions are set according to the same approach, the reference value is 1. 
For wetlands ecosystems in particular, we have not yet discussed the definition of "good ecological condition", but it seems that in the Ecological Condition Assessments for forest and mountain ecosystems, 0.6 was used as "good ecological condition" consistent with the approach from "Vanndirektivet". 
To what degree this is appropriate is difficult to judge. It may well be best to use 0.6 due to its consistency with other frameworks (e.g. Vanndirektivet) as it would be challenging to otherwise determine what is "good enough" condition for a highly derived variable such as NI products.

## Uncertainties
Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package NIcalc. These draws can be summarised with metrics such as mean, median, confidence intervals. etc. 
The degree of uncertainty is determined by a) the uncertainty associated with each indicator's data as deposited in the NI database and b) the uncertainty resulting from imputation of missing values.

Besides these "quantitative" sources of uncertainty, there is also "qualitative" uncertainty linked to using NI products as indicators in Ecological Condition Assessments. This is linked to methodological questions regarding how (and perhaps if) highly processed composite indices should be included as indicators to make other composite indices. 
While some of these questions are of a more philosohpical nature, some more pragmatic ones include decisions about which indicators to include/exclude (especially considering indicators that may get used directly and via NI products and indicators that do not really quantify species abundance), how to weigh indicators, and whether or not to impute NA values as part of generating the NI product.

We present some thoughts and suggestions regarding these latter questions in the next section. 
## Suggestions for using NI products in Ecological Condition Assessments
[This section will hold some insights and suggestions based on a discussion to be held once Anders is back from leave. Content will be added later.]

## References
<div id="refs"></div>

## Analyses
The following analysis workflow is almost entirely arranged in functions. The functions for the different steps are from the NIflex R package (https://github.com/NINAnor/NIflex).Many of the NIflex functions call on functions from NIcalc (https://github.com/NINAnor/NIcalc), the R package written for entering and extracting data from the NI database and calculating NI products from it. 
Before proceeding any further, we therefore need to make sure that both of these packages and their dependencies are installed:

```{r}
if(system.file(package = 'NIcalc') == ""){
  devtools::install_github("NINAnor/NIcalc", build_vignettes = TRUE)
}

if(system.file(package = 'NIflex') == ""){
  devtools::install_github("NINAnor/NIflex", build_vignettes = TRUE)
}
```

NIflex was written to provide a flexible workflow for calculating a range of NI products including the original NI for all seven main ecosystems, a range of previously used thematic indices, fully custom indices and -- most important here -- NI products aimed for use in Ecological Condition Assessments for forest, mountain, and wetland ecosystems.

Below I walk through the workflow for an NI product for the Ecosystem Condition Assessment for wetlands. I will not provide a lot of additional details on the functions and their arguments here as all necessary information is included in the function documentation.

### Data sets

#### Nature Index database
All data required for making the NI product are contained in the NI database (https://naturindeks.nina.no/). Accessing the database requires a username and a password. 
At present, this setup works with the username and password saved as personal environmental variables `NIDB_USERNAME` and `NIDB_PASSWORD`.

```{r, eval=FALSE}
NIdb_username <- Sys.getenv("NIDB_USRNAME")
NIdb_password <- Sys.getenv("NIDB_PASSWORD")
```

User credentials can be requested from NINA.

#### Regions
As mentioned above, ecosystem-level NI products are calculated for five major regions in Norway: "Nord-Norge" (northern Norway), "Sør-Norge" (southern Norway), "Midt-Norge" (central Norway), "Vestlandet" (western Norway), and "Østlandet".
For georeferencing values and plotting them onto a map, shapefiles for these regions are required. 
The relevant files can be downloaded in JSON format from the NI API (https://ninweb08.nina.no/NaturindeksAPI/index.html) and subsequently need to be transformed into shapefiles.

In this workflow, we access the shapefiles that are deposited on NINA's internal file system:

```{r, eval=FALSE}
shapeLibraryPath <- "/data/P-Prosjekter/41201611_naturindeks_2021_2023_vitenskapelig_ansvar/Shapefiles"
```

### Workflow options
NIflex offers a workflow for calculating NI products that can be customized in terms of a range of options. 
So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include. Since we may want to use NA imputation, we select all official NI years:
```{r}
years <- c("1990", "2000", "2010", "2014", "2019")
```

Next, we decide what type of output we want and whether or not to impute missing values
```{r}
OutputType <- "EcologicalCondition"
EC_ecosystem <- c("Wetlands", "Våtmark")

NAImputation <- TRUE 
```

and continue to set how we want indicators to be weighed. At present, we will not use key indicator weights, but we will still weigh according to area and trophic group:
```{r}
KeyIndicators <- FALSE
KeyWeight <- 0.5

AreaWeights <- TRUE

TrophicWeights <- TRUE
```

We also need to state that we want to use Norwegian indicator names
```{r}
norwegianNames <- TRUE
```

and set the way in which we want to specify which indicators are dropped. 
There are two options for this: "pre-defined" drops the hard-coded set of indicators that should be dropped in any case. So far, there is only one indicator that gets dropped for wetlands and that is "lavhei" (due to low representativeness for the ecosystem). It's also possible to set it to "custom" and specify which indicators should be dropped, but we are not doing that here. 
```{r}
DropIndMode <- "pre-defined"
if(DropIndMode == "pre-defined"){DropIndList <- NULL}
```

With the final options, we choose whether we want dataset diagnostics to be calculated and also whether we want to run a test run or a full run:
```{r}
Diagnostics <- TRUE # Yes
TestRun <- TRUE # Yes
```
The difference for the latter is that in a test run, we only run NA imputation and bootstrap for 10 iterations instead of 1000, which makes the runtime of the workflow a lot more "practical" (but 1000 or even more iterations should be run for the final results).

### Scaled indicator values
Before beginning index calculation, there are a few more conditional variables that need to be set, namely the Norwegian name of the ecosystem and the lists of indicators to use and indicators to drop. 
```{r}
ecosystem_use <- EC_ecosystem[2]
indicators_use <- NULL

dropInd <- selectDropIndicators(DropIndMode = DropIndMode,
                                OutputType = OutputType,
                                ecosystem = ifelse(OutputType %in% c("NatureIndex", "EcologicalCondition"), ecosystem_use, NULL),
                                customList = DropIndList)
```
Note that we set `indicators_use` to `NULL` because we are working with a pre-defined set of indicators (i.e. the set linked to NI for wetlands).

We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectivel (these are used for controlling workflow elements linked to thematic indices):
```{r,eval=FALSE}
theme <- "none"
funArguments <- NULL
```

With this, we are ready to move ahead with the calculation of our wetlands NI product. For this, we use `NIflex::calculateCustomNI()`, a wrapper function that downloads the data, runs dataset diagnostics (if requested), imputes NA values (if requested), and calculates the NI product using functions from the NIcalc package: 
```{r, results = FALSE, eval=FALSE}
CustomNI <- calculateCustomNI(ecosystem = ecosystem_use,
                              indicators = indicators_use,
                              theme = theme,
                              dropInd = dropInd,
                              username = NIdb_username,
                              password = NIdb_password,
                              KeyIndicators = KeyIndicators,
                              KeyWeight = KeyWeight,
                              AreaWeights = AreaWeights,
                              TrophicWeights = TrophicWeights,
                              NAImputation = NAImputation,
                              years = years,
                              OutputType = OutputType,
                              funArguments = funArguments,
                              Diagnostics = Diagnostics,
                              TestRun = TestRun,
                              norwegianNames = norwegianNames,
                              saveSteps = FALSE)
```
```{r, echo=FALSE, message=FALSE, warning=FALSE}
CustomNI<-readRDS("data/cache/CustomNI.RDS")
```

After calculating the index values, we can extract summary metrics for a 
specified year and match them to geospatial data for further analysis and/or 
inclusion in the Ecosystem Condition framework:

```{r}
CustomNI_map <- geomapNI(Index = CustomNI$CustomIndex, 
                         year = 2019, 
                         OutputType = OutputType, 
                         ecosystem = ecosystem_use,
                         shapeLibraryPath = shapeLibraryPath)
```

Next we can have a look at some visualizations of our NI product. 
First, we can plot the "traditional" time-series plot of the index averaged across regions and calculate the corresponding summary metrics:

```{r}
Index <- CustomNI$CustomIndex
plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
summary(Index$wholeArea)
```

The summary metrics (here median) for any given year can then be plotted onto a map by matching them up to the shapefiles:

```{r, message = FALSE, warning = FALSE}
plotNI_Map(shp = CustomNI_map,
           year = 2019,
           OutputType = OutputType,
           ecosystem = ecosystem_use,
           plotMedian = TRUE, plotCI = FALSE, plotDisplacement = FALSE,
           interactiveMap = TRUE)
```

### Uncertainty
Based on recent work done as part of a pilot study on visualizing NI products (https://ninanor.github.io/NIviz/index.html) I have also implemented some more advanced plotting functions. The first one plots a time-series as well, but including the entire uncertainty distribution, not just summary statistics:
```{r, results = FALSE, message = FALSE, warning=FALSE}
Index <- Index <- CustomNI$CustomIndex

plotNI_DensityRidgeTS(Index = Index, 
                      OutputType = OutputType, 
                      ecosystem = ecosystem_use, 
                      theme = theme,
                      allAreas = FALSE, 
                      selectedArea  = "wholeArea")
```

This works for plotting values averaged across the entire Norway as above, but also by area: 
```{r, results = FALSE, message = FALSE, warning=FALSE}
plotNI_DensityRidgeTS(Index = Index, 
                      OutputType = OutputType, 
                      ecosystem = ecosystem_use, 
                      theme = theme,
                      allAreas = TRUE)

```

Uncertainty can also be visualized on a map alongside the median: 
```{r, message = FALSE, warning = FALSE}
plotNI_Map(shp = CustomNI_map,
           year = 2019,
           OutputType = OutputType,
           ecosystem = ecosystem_use,
           plotMedian = TRUE, plotCI = TRUE, plotDisplacement = FALSE,
           interactiveMap = TRUE)
```

Finally, the `plotNI_Map()` function also supports the option of plotting "displacement", which is a statistical artifact generated by the NI calculation routine. In short, it's a negative number that quantifies the degree to which the mean of the index is shifted away from 1 when scaling indicators. The higher the displacement value, the more cautious one may want to be when making inference based on summary metrics alone: 
```{r, message = FALSE, warning = FALSE}
plotNI_Map(shp = CustomNI_map,
           year = 2019,
           OutputType = OutputType,
           ecosystem = ecosystem_use,
           plotMedian = TRUE, plotCI = FALSE, plotDisplacement = TRUE,
           interactiveMap = TRUE)
```

## Prepare export
Preparing the export is only a matter of renaming columns at this stage. Using 
`geomapNI()` we have already extracted summaries and matched them to spatial
data. We can take the resulting object, `CustomNI_map`, rename the required
columns to match the naming conventions outlined in the main document, add
information on reference state, and remove columns that are not needed:

```{r}
CustomNI_map <- CustomNI_map %>%
  dplyr::rename(v_2019 = meanValue,
                sd_2019 = standardDev) %>%
  dplyr::mutate(i_2019 = v_2019,
                reference_high = 1,
                reference_low = 0,
                thr = 0.6) %>%
  dplyr::select(-medianValue, -widthCI, -displacement)
```

Note that Nature Index products only exist as scaled variables. For that reason, 
there is no difference between `i_yyyy` and `v_yyyy`.

### Eksport file (final product)
Lastly, we can save the results in the designated folder:

```{r}
saveRDS(CustomNI, "indicators/naturindeks.rds")
saveRDS(CustomNI_map, "indicators/indicatorMaps/naturindeks_map.rds")
```

### 2023-09-14 slitasje.Rmd
Fix #116

@@
 
-```{r extrapolated-slitasje, fig.cap="Slitasje indicator extrapolated over Trondheim"}
+```{r extrapolated-slitasje, fig.cap="Slitasje indicator extrapolated over Trondheim, with one map for each ecosystem.
+The map is not masked against ecosystem delineation maps, which make them easy to missinterpret."}
 myCol <- "RdYlGn"

### 2023-09-14 slitasje.Rmd
#105

@@
 
-```{r slitasje-precentageplot, fig.height=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different indicator values."}
+```{r slitasje-precentageplot, fig.height=12, fig.cap="Relative frequency plot (conditioned on region and main ecosystem) showing the distribution of polygons with different scaled indicator values for the indicator named Anthropogenic disturbance to soils and vegetation."}
 ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
@@
   theme_bw(base_size = 12)+
-  guides(fill = guide_legend("Sliatsje indicator"))+
+  guides(fill = guide_legend("Scaled indicator values"))+
   ylab("Fraction of data points")+

### 2023-09-14 slitasje.Rmd
#104

@@
 
-```{r dense-stats, fig.height= 12, fig.cap="Barplot showing the number of data points, the area weighted indicator values and the sd for the slitasje indicator.
-Transparrent bars represents categories with less than 150 data points."}
+```{r dense-stats, fig.height= 12, fig.cap="Barplot showing the number of data points, the area weighted indicator values and the sd for the ADSV indicator.
+Transparent bars represents categories with less than 150 data points."}
 temp_plot <- all_stats %>%
@@
   theme_bw(base_size = 10)+
-  scale_fill_brewer(palette="Oranges")+
+  scale_fill_brewer(palette="Oranges", na.translate=F)+
   theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
+  labs(fill = "HIF")+
   facet_wrap(.~eco),
@@
   theme_bw(base_size = 10)+
-  scale_fill_brewer(palette="Oranges")+
+  scale_fill_brewer(palette="Oranges",  na.translate=F)+
   theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
+  labs(fill = "HIF")+
   facet_wrap(.~eco),

### 2023-09-14 slitasje.Rmd
Change indicator name and add vertical line in figure

--- 2023-09-14 slitasje.Rmd
@@
-# Slitasje og kjørespor {#slitasje}
+# Anthropogenic disturbance to soils and vegetation ADSV {#slitasje}
 
@@
 
-March 2023
+September 2023
+
+Norwegian indicator name: Slitasje
 
@@
 
-```{r dense-stats, fig.height= 12, fig.cap="Barplot showing the number of data points, the area weighted indicator values and the sd for the slitasje indicator.
-Transparrent bars represents categories with less than 150 data points."}
+```{r dense-stats, fig.height= 12, fig.cap="Barplot showing the number of data points, the area weighted scaled indicator values and the sd for the ADSV indicator.
+Transparent bars represents categories with less than 150 data points, indicated by the vertical line in the top row of panes."}
 temp_plot <- all_stats %>%
@@
   theme_bw(base_size = 10)+
-  scale_fill_brewer(palette="Oranges")+
+  scale_fill_brewer(palette="Oranges", na.translate=F)+
   theme(axis.text.x = element_text(angle = 90, vjust=0.5))+
   coord_flip()+
+  labs(fill = "HIF")+
+  geom_hline(yintercept=150)+
   facet_wrap(.~eco),
@@
 
-```{r extrapolated-slitasje, fig.cap="Slitasje indicator extrapolated over Trondheim, with one map for each ecosystem.
+```{r extrapolated-slitasje, fig.cap="ADSV indicator extrapolated over Trondheim, with one map for each ecosystem.
 The map is not masked against ecosystem delineation maps, which make them easy to missinterpret."}
@@
   tm_polygons(col = "w_mean",
-    title="Slitasje indicator",
+    title="Indicator value",
     palette = myCol,
@@
 
-### Slitasje in wetlands
+### ADSV in wetlands
 

### 2023-09-14 slitasje.Rmd
bug: fix crossrefs

@@
 
-7.  Define [homogeneous impact areas](#HIA) (HIA)
+7.  Define [homogeneous impact areas](#HIA-slitasje) (HIA)
 
@@
 
-We also have a separate [summary file](##exp-natureType-summary) where
-the nature types are manually mapped the NiN variables and to the correct NiN-main types. We can use this to find the nature types associated with the NiN main types
+We also have a separate [summary file](#exp-natureType-summary) where
+the nature types are manually mapped to the correct NiN-main typesand to the NiN variables  that are recorded in each nature type.
+We can use this to find the nature types associated with the NiN main types
 that we are interested in.
@@
 
-### Homogeneous impact areas {#HIA}
-
-I want to use the [Homogeneous Impact Areas](#HIA) (HIA) to define
+### Homogeneous impact areas {#HIA-slitasje}
+
+I want to use the [Homogeneous Impact Areas](#homogeneous-impact-areas) (HIA) to define
 smaller regions into which I can extrapolate the indicator values. This

### 2023-09-14 Nature_Index_product.Rmd
Make interactive plots non-interactive and Render the Book

--- 2023-09-13 Nature Index product.Rmd
@@
+---
+editor_options: 
+  markdown: 
+    wrap: sentence
+---
+
 # Nature index product
@@
 
-_Author and date:_
+*Author and date:*
 
@@
 ## Introduction
+
 The Norwegian Nature Index (NI) is a composite metric for quantifying the state of biodiversity in an abundance-context.
 It has been established and published for the first time in 2010 and has since been updated for years 2011, 2012, 2013,2014, and 2019.
-It is owned and presented by the Norwegian Environmental Agency and the latest iteration, including a lot of information, can be found on its webpage: https://www.naturindeks.no/
+It is owned and presented by the Norwegian Environmental Agency and the latest iteration, including a lot of information, can be found on its webpage: <https://www.naturindeks.no/>
 
@@
 Moving forward, it will be important to stress this distinction, and we will therefore start talking about "NI products" when referring to composite indices built based on data in the Nature Index Database (see below) that are not identical to the regularly published national assessments (i.e.
-what is published on https://www.naturindeks.no/).
+what is published on <https://www.naturindeks.no/>).
 
 ## About the underlying data
-The data underlying NI products is stored in the Nature Index Database at https://naturindeks.nina.no/. The database needs login credentials for access.
+
+The data underlying NI products is stored in the Nature Index Database at <https://naturindeks.nina.no/>.
+The database needs login credentials for access.
 
@@
 ### Representativity in time and space
+
 NI products for inclusion in Ecological Condition Assessments are can currently be calculated for years 1990, 2000, 2010, 2014, and 2019, which correspond to the years in which the official NI updates took place.
@@
 ### Original units
+
 The NI is a composite index made up of scaled indicators.
-NI products are therefore numbers between 0 and 1 (more specifically a set of simulated numbers resulting from a bootstrap, and derived summary metrics including mean, median, and 95 \% confidence intervals) that have no unit.
+NI products are therefore numbers between 0 and 1 (more specifically a set of simulated numbers resulting from a bootstrap, and derived summary metrics including mean, median, and 95 % confidence intervals) that have no unit.
 
 ### Temporal coverage
-As mentioned above, NI products can currently be calculated for years 1990, 2000, 2010, 2014, and 2019. From 2025, calculations based on new data from 2024 will become available.
+
+As mentioned above, NI products can currently be calculated for years 1990, 2000, 2010, 2014, and 2019.
+From 2025, calculations based on new data from 2024 will become available.
 
 ### Aditional comments about the dataset
-There are missing values in the NI database. The proportion of missing data values depends on the individual indicators that are to be included in an NI product. 
+
+There are missing values in the NI database.
+The proportion of missing data values depends on the individual indicators that are to be included in an NI product.
 As part of the NI analysis workflow, missing data values can be imputed using the MICE (Multivariate Imputation by Chained Equations) algorithm.
@@
 ## Ecosystem characteristic
+
 ### Norwegain standard
+
 NI products are regarded as indicators of biodiversity ("biologisk mangfold").
@@
 ### SEEA EA (UN standard)
+
 NI products are tagged as "Compositional state characteristic".
@@
 ## Collinearities with other indicators
+
 There is a substantial potential for collinearity between NI products and other indicators, primarily due to the fact that a range of indicators that may be part of an NI product for a given ecosystem may also enter the Ecological Condition Assessment as indicators in their own right or through other composite indicators.
@@
 ## Reference condition and values
+
 ### Reference condition
+
 The reference condition is defined along the same lines for NI and for other indicators in Ecological Condition Assessments.
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+
 As NI products are already scaled to range from 0 to 1 and reference conditions are set according to the same approach, the reference value is 1.
@@
 ## Uncertainties
+
 Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package NIcalc.
@@
 We present some thoughts and suggestions regarding these latter questions in the next section.
-## Suggestions for using NI products in Ecological Condition Assessments
-[This section will hold some insights and suggestions based on a discussion to be held once Anders is back from leave. Content will be added later.]
+## Suggestions for using NI products in Ecological Condition Assessments [This section will hold some insights and suggestions based on a discussion to be held once Anders is back from leave.
+Content will be added later.]
 
 ## References
-<div id="refs"></div>
+
+::: {#refs}
+:::
 
 ## Analyses
+
 The following analysis workflow is almost entirely arranged in functions.
-The functions for the different steps are from the NIflex R package (https://github.com/NINAnor/NIflex).Many of the NIflex functions call on functions from NIcalc (https://github.com/NINAnor/NIcalc), the R package written for entering and extracting data from the NI database and calculating NI products from it.
+The functions for the different steps are from the NIflex R package (<https://github.com/NINAnor/NIflex>).Many of the NIflex functions call on functions from NIcalc (<https://github.com/NINAnor/NIcalc>), the R package written for entering and extracting data from the NI database and calculating NI products from it.
 Before proceeding any further, we therefore need to make sure that both of these packages and their dependencies are installed:
@@
 #### Nature Index database
-All data required for making the NI product are contained in the NI database (https://naturindeks.nina.no/). Accessing the database requires a username and a password. 
+
+All data required for making the NI product are contained in the NI database (<https://naturindeks.nina.no/>).
+Accessing the database requires a username and a password.
 At present, this setup works with the username and password saved as personal environmental variables `NIDB_USERNAME` and `NIDB_PASSWORD`.
@@
 #### Regions
+
 As mentioned above, ecosystem-level NI products are calculated for five major regions in Norway: "Nord-Norge" (northern Norway), "Sør-Norge" (southern Norway), "Midt-Norge" (central Norway), "Vestlandet" (western Norway), and "Østlandet".
 For georeferencing values and plotting them onto a map, shapefiles for these regions are required.
-The relevant files can be downloaded in JSON format from the NI API (https://ninweb08.nina.no/NaturindeksAPI/index.html) and subsequently need to be transformed into shapefiles.
+The relevant files can be downloaded in JSON format from the NI API (<https://ninweb08.nina.no/NaturindeksAPI/index.html>) and subsequently need to be transformed into shapefiles.
 
@@
 
-```{r, eval=FALSE}
+```{r}
 shapeLibraryPath <- "/data/P-Prosjekter/41201611_naturindeks_2021_2023_vitenskapelig_ansvar/Shapefiles"
@@
 ### Workflow options
+
 NIflex offers a workflow for calculating NI products that can be customized in terms of a range of options.
-So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include. Since we may want to use NA imputation, we select all official NI years:
+So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include.
+Since we may want to use NA imputation, we select all official NI years:
+
 ```{r}
@@
 Next, we decide what type of output we want and whether or not to impute missing values
+
 ```{r}
@@
 
-and continue to set how we want indicators to be weighed. At present, we will not use key indicator weights, but we will still weigh according to area and trophic group:
+and continue to set how we want indicators to be weighed.
+At present, we will not use key indicator weights, but we will still weigh according to area and trophic group:
+
 ```{r}
@@
 We also need to state that we want to use Norwegian indicator names
+
 ```{r}
@@
 It's also possible to set it to "custom" and specify which indicators should be dropped, but we are not doing that here.
+
 ```{r}
@@
 With the final options, we choose whether we want dataset diagnostics to be calculated and also whether we want to run a test run or a full run:
+
 ```{r}
@@
 ```
+
 The difference for the latter is that in a test run, we only run NA imputation and bootstrap for 10 iterations instead of 1000, which makes the runtime of the workflow a lot more "practical" (but 1000 or even more iterations should be run for the final results).
@@
 ### Scaled indicator values
+
 Before beginning index calculation, there are a few more conditional variables that need to be set, namely the Norwegian name of the ecosystem and the lists of indicators to use and indicators to drop.
+
 ```{r}
@@
 ```
+
 Note that we set `indicators_use` to `NULL` because we are working with a pre-defined set of indicators (i.e. the set linked to NI for wetlands).
@@
 We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectivel (these are used for controlling workflow elements linked to thematic indices):
+
 ```{r,eval=FALSE}
@@
 For this, we use `NIflex::calculateCustomNI()`, a wrapper function that downloads the data, runs dataset diagnostics (if requested), imputes NA values (if requested), and calculates the NI product using functions from the NIcalc package:
+
 ```{r, results = FALSE, eval=FALSE}
@@
 ```
+
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-After calculating the index values, we can extract summary metrics for a 
-specified year and match them to geospatial data for further analysis and/or 
-inclusion in the Ecosystem Condition framework:
+After calculating the index values, we can extract summary metrics for a specified year and match them to geospatial data for further analysis and/or inclusion in the Ecosystem Condition framework:
 
-```{r}
+```{r, eval=FALSE}
 CustomNI_map <- geomapNI(Index = CustomNI$CustomIndex, 
@@
 
+```{r, echo=FALSE}
+CustomNI_map<-readRDS("data/cache/CustomNI_map.RDS")
+```
+
 Next we can have a look at some visualizations of our NI product.
@@
 
-```{r}
+```{r, eval=FALSE}
 Index <- CustomNI$CustomIndex
@@
 summary(Index$wholeArea)
+```
+
+![](images/custPlot.png)
+
+```{r, echo=FALSE}
+Index <- CustomNI$CustomIndex
+#plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
+summary(Index$wholeArea)
+
 ```
@@
            plotMedian = TRUE, plotCI = FALSE, plotDisplacement = FALSE,
-           interactiveMap = TRUE)
+           interactiveMap = FALSE)
+
 ```
@@
 ### Uncertainty
-Based on recent work done as part of a pilot study on visualizing NI products (https://ninanor.github.io/NIviz/index.html) I have also implemented some more advanced plotting functions.
+
+Based on recent work done as part of a pilot study on visualizing NI products (<https://ninanor.github.io/NIviz/index.html>) I have also implemented some more advanced plotting functions.
 The first one plots a time-series as well, but including the entire uncertainty distribution, not just summary statistics:
+
 ```{r, results = FALSE, message = FALSE, warning=FALSE}
@@
 This works for plotting values averaged across the entire Norway as above, but also by area:
+
 ```{r, results = FALSE, message = FALSE, warning=FALSE}
@@
 Uncertainty can also be visualized on a map alongside the median:
+
 ```{r, message = FALSE, warning = FALSE}
@@
            plotMedian = TRUE, plotCI = TRUE, plotDisplacement = FALSE,
-           interactiveMap = TRUE)
+           interactiveMap = FALSE)
 ```
@@
 The higher the displacement value, the more cautious one may want to be when making inference based on summary metrics alone:
+
 ```{r, message = FALSE, warning = FALSE}
@@
            plotMedian = TRUE, plotCI = FALSE, plotDisplacement = TRUE,
-           interactiveMap = TRUE)
+           interactiveMap = FALSE)
 ```
+## Prepare export
 
-## Prepare export
-Preparing the export is only a matter of renaming columns at this stage. Using 
-`geomapNI()` we have already extracted summaries and matched them to spatial
-data. We can take the resulting object, `CustomNI_map`, rename the required
-columns to match the naming conventions outlined in the main document, add
-information on reference state, and remove columns that are not needed:
+Preparing the export is only a matter of renaming columns at this stage.
+Using `geomapNI()` we have already extracted summaries and matched them to spatial data.
+We can take the resulting object, `CustomNI_map`, rename the required columns to match the naming conventions outlined in the main document, add information on reference state, and remove columns that are not needed:
 
@@
 
-Note that Nature Index products only exist as scaled variables. For that reason, 
-there is no difference between `i_yyyy` and `v_yyyy`.
+Note that Nature Index products only exist as scaled variables.
+For that reason, there is no difference between `i_yyyy` and `v_yyyy`.
 
 ### Eksport file (final product)
+
 Lastly, we can save the results in the designated folder:

### 2023-09-14 slitasje.Rmd
Add some refs and a new image (A8 scale)

@@
 
-No additional references.
+Halvorsen, R.
+& Bratli, H.
+2019.
+Dokumentasjon av NiN versjon 2.2 tilrettelagt for praktisk naturkartlegging: utvalgte variabler fra beskrivelsessystemet.
+– Natur i Norge, Artikkel 11 (versjon 2.2.0): 1–218 (Artsdatabanken, Trondheim; http://www.artsdatabanken.no.)
+
+Miljødirektoratet 2022.
+Kartleggingsinstruks - kartlegging av terrestriske Naturtyper etter NiN2.
+URL: https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/
+
+Miljødirektoratet 2023.
+Arealrepresnetativ Naturovervåking.
+Webpage.
+URL: https://www.miljodirektoratet.no/ansvarsomrader/overvaking-arealplanlegging/miljoovervaking/overvakingsprogrammer/natur-pa-land/arealrepresentativ-naturovervakning-ano/
 
@@
 
-```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the slitasje variable.", fig.width=5}
+```{r slitasje-normalise, fig.cap="Performing a linear break-point type normalisation of the ADSV variable.", fig.width=5}
 eaTools::ea_normalise(data = slitasje_data,

### 2023-09-14 slitasje.Rmd
Add some cross refs and hyperlinks to GH issues

@@
-# Anthropogenic disturbance to soils and vegetation ADSV {#slitasje}
+# Anthropogenic disturbance to soils and vegetation (ADSV) {#slitasje}
 
@@
 
-1.  Import [Nature type data](#NTM) data set (incl. GRUK) and [ANO data
+1.  Import [Nature type data](#NTM) data set and [ANO data
     set](#ANO-import)
@@
 
-4.  For each polygon, extract the NiN-variable that has the lowest value
-    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype).
-
-5.  [Combine](#combine-nt-ano) data sets
+4.  For each polygon, extract the variable that has the lowest value
+    (one-out all-out principle), e.g. Fig. \@ref(fig:slitasje-naturetype). The four variables are 7TK, 7SE (both NiN2-variables) and PRTK and PRSL (both defined by the Norwegain Environmental Agency)
+
+5.  [Combine](#combine-nt-ano) data sets (nature types and ANO) but keep ecosystems separate.
 
@@
 9.  Confirm relationship between infrastructure index and indicator
-    values to justify the extrapolation
+    values to [justify the extrapolation](#validate-slitasje)
 
@@
 
-The data goes back to 2018. I therefore bulk all the data from 2018 to 2022 into one time step. I then use the mean date for the raw data,
+The data goes back to 2018.
+I therefore [bulk all the data from 2018 to 2022](https://github.com/NINAnor/ecosystemCondition/issues/101#issuecomment-1719170208) into one time step.
+I then use the mean date for the raw data,
 and define the variable as belonging to the year 2020 (read more
@@
 The indicator is tagged to the *Økologisk egenskap* called
-**Primærproduksjon** (Primary productivity). This is tentative, and
+**Primærproduksjon** (Primary productivity). This is [somewhat tentative](https://github.com/NINAnor/ecosystemCondition/issues/96), and
 perhaps *abiotiske forhold* is more suited. But the thought behind the
-choice is that *slitasje* affect the potential for primary productivity.
+choice is that *ADSV* affect the potential for primary productivity.
 
@@
 
-_Slitasje_ is not thought to exhibit collinearity with any other indicator at
+_ADSV_ is not thought to exhibit collinearity with any other indicator at
 the present.
@@
 is also true for semi-natural ecosystems. For example, in the reference
-condition, *slitasje* in semi-natural ecosystems is defined to be zero,
+condition, *ADSV* in semi-natural ecosystems is defined to be zero,
 even though at no point in time would this condition be realized.
@@
 ### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+For a further discussion on this topic, see [GitHub issue 88](https://github.com/NINAnor/ecosystemCondition/issues/88).
 
@@
 
-The upper reference value is 0 (no *slitasje*), and the lower reference
-value is 100%. Note that 100% _slitasje_ does not mean that all the area
+The upper reference value is 0% (no *damage*), and the lower reference
+value is 100%. Note that 100% _damage_ does not mean that all the area
 must be scarred or damaged, but that all hypothetical quadrats around a
 point is affected (the variable is frequency-driven, and not
-amount-driven). The threshold for good ecosystem condition is 10%
-damage. Perhaps is should be set somewhat higher, like 20 or 30%.
-The indicator value could be interpretted differently at the polygon
-vs at the landscape scale. At the polygon scale, 10% damage may seem 
+amount-driven). The threshold for good ecosystem condition is set to 10%
+damage.
+
+The indicator value could be interpreted differently at the polygon
+_vs_ landscape scale. At the polygon scale, 10% damage may seem 
 like not that much, whilst at the landscape scale, 10% seems more serious.
-This final threshold value should be debated and illustrated with examples.
-
-Read about the normalisation [here](##scaled-slitasje-variable).
+
+Read about the normalisation [here](#scaled-slitasje-variable).
 
@@
 ```{r eight-step, fig.cap="Eight step condition scale"}
-knitr::include_graphics("images/mdirPRscale.PNG")
+knitr::include_graphics("images/ninAR8.PNG")
 ```
@@
 
-I here transfer these ordinal variables over to a continuous scale.
+I here transfer these ordinal variables over to a continuous scale based on the mean value of each step, but we acknowledge that there are important assumptions being made here, for example that the mean _real_ value for 7TK or 7SE 3 is the mid-point in the range from 0.5 to 1 (i.e.
+0.75).
+This is probably not correct.
+The large range of the steps, particularly for the 7SE and 7TK variables, poses a problem for assessing ecological condition at the polygon scale, but less so at aggregated scales.
+Hence, the indicator should not be presented and interpreted at the polygon scale.
 
@@
 
-First I will combine 7TK and PRTK, and also 7SE and PRSL
+First I will combine 7TK and PRTK, and also 7SE and PRSL using the one-out-all-out principle.
+This is a more conservative approach than for example simply adding these together, but we do it this way because the two variables are likely to overlap considerably at small spatial scales.
 
@@
 [GRUK](https://www.nina.no/Naturmangfold/Trua-natur/%C3%85pen-grunnlendt-kalkmark).
-The nature type data set I'm working on here includes this data already
-(presently only 2021 included). GRUK also records a related variable:
-`% cover in 5m radii circles`, which is much more detailed. This data is
-not published. In any case it is better to use the harmonized data set
+This is a stratified and intensive monitoring program for calcareous naturally open areas around the Oslo fjord. 
+The nature type data set I'm working on here includes some data from GRUK already
+(presently only 2021 included). The full GRUK data set also contains a related variable:
+`% cover in 5m radii circles`, which is much more detailed than the four variable we have looked at so far. This data is
+not published. In any case it might be better to use the harmonized data set
 in our case.
@@
 Depending on the main ecosystem, certain NiN variables are also
-recorded. 7SE is recorded for våtmark, but not semi-natural areas or
-naturally open areas. 7TK is recorded in våtmark and naturally open
+recorded. 7SE is recorded for våtmark (wetlands), but not semi-natural areas or
+naturally open areas. 7TK is recorded in wetlands and naturally open
 areas only.
@@
 update this data set later. It is not clear why there is no data from
-2020.
+2020. This topic is maintained as a [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/101).
 
@@
 
-Combining these two variables 7TJ and 7SE into one, same as for the nature
+Combining these two variables 7TK and 7SE into one, same as for the nature
 type data.
@@
 
-#### Validate
+#### Validate {#validate-slitasje}
 

### 2023-09-14 slitasje.Rmd
Working on additive approach and fixing break point value

@@
 "PRSL".
-I want to base this indicator on whichever of the variables have the highest value (worst condition).
-Therefore I first convert the ordinal scale to a continuous scale, using the median value of each ordinal step.
+
+I initially baseed this indicator on whichever of the variables have the highest value (worst condition), but after some thought and discussion I now use an additive approach (adding the variable values togther).
+I first convert the ordinal scale to a continuous scale, using the median value of each ordinal step.
 
@@
 
-First I will combine 7TK and PRTK, and also 7SE and PRSL using the one-out-all-out principle.
-This is a more conservative approach than for example simply adding these together, but we do it this way because the two variables are likely to overlap considerably at small spatial scales.
+First I will combine 7TK and PRTK, and also 7SE and PRSL.
 
@@
 
-Then I can combine these two variables into a composite variable called
+Then I can add these two variables together into a composite variable called
 `slitasje`
@@
 ```{r, warning=F}
-naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("TK", "SE")], 1, max, na.rm=T)
-```
-
-When both variables are NA, we get -Inf. There were `r nrow(naturetypes_wide[naturetypes_wide$slitasje<0,])` cases of this. Removing these now.
-
-```{r}
-naturetypes_wide <- naturetypes_wide[naturetypes_wide$slitasje>=0,]
-```
-
-```{r slitasje-naturetype, fig.cap="Slitasje scores.", fig.width=3}
+naturetypes_wide$slitasje <- apply(naturetypes_wide[,c("TK", "SE")], 1, sum, na.rm=T)
+```
+
+Chekhing for -Inf
+```{r}
+nrow(naturetypes_wide[naturetypes_wide$slitasje<0,])
+```
+No issues with that.
+
+```{r slitasje-naturetype, fig.cap="ADSV scores.", fig.width=3}
 temp <- naturetypes_wide %>%
@@
 
-ggplot(temp, aes(x = factor(slitasje),
+ggplot(temp, aes(x = slitasje,
                  y = sum))+
@@
   theme_bw(base_size = 12)+
-  labs(x = "Slitasje score converted to percentage",
+  labs(x = "ADSV score converted to percentage",
        y = "Number of localities") + 
-  theme(axis.text.x = element_text(angle=90, vjust=0.5))
-```
-
-It appears most localities are in good condition.
-
-I would also like to know, when both tk and SE was recordedm
-how often TK was defining of the slitasje-indicator, and how 
-often it was SE. I can do this by taking the difference.
-
-```{r, fig.cap="Counting the number of cases where TK or SE was defining of the slitasje indicator score.", fig.width=3}
+  theme(axis.text.x = element_text(angle=90, vjust=0.5))+
+  scale_x_continuous(
+    labels = scales::label_number(accuracy = 1))
+```
+
+It appears most localities are in good condition. Note also that we can get >100% for the unscaled variable.
+
+I would also like to know, when both TK and SE was recorded,
+which one is likely to be in the worst (or best) condition. 
+I can do this by taking the difference.
+
+```{r, fig.cap="Counting the number of cases where TK or SE was in the worst condition for a given locality.", fig.width=3}
 
@@
   theme_bw(base_size = 12)+
-  labs(x = "Defining variable",
+  labs(x = "Variable indicating the worst state",
        y = "Number of localities")
@@
 
-Now I will copy these slitasje-values into the sf object.
+Now I will copy these ADVS-values into the sf object.
 
@@
 
-I think I will only use våtmark here since my approach will be to the
+I think I will only use wetlands here since my approach will be to the
 'worst value' of the variable 7SE and 7TK (see below). I think not
 having 7SE for naturally open areas will underestimate the degree of
-degredation in these areas.
+degradation in these areas.
 
@@
 
-```{r}
-temp <- as.data.frame(ano)
-temp$slitasje <- apply(temp[,c("var_7TK", "var_7SE")], 1, max, na.rm=T)
-ano$slitasje <- temp$slitasje
-```
-
-Then I extract the mid values for each ordinal range step, like I did previously for the nature type data set.
+First I extract the mid values for each ordinal range step, like I did previously for the nature type data set.
 
@@
 ano <- ano %>%
-  mutate(slitasje = case_match(slitasje,
+  mutate(var_7TK = case_match(var_7TK,
+             0 ~ 0,
+             1 ~ mean(c(0, 1/16))*100,
+             2 ~ mean(c(1/16, 1/2))*100,
+             3 ~ 75)) %>%
+  mutate(var_7SE = case_match(var_7SE,
              0 ~ 0,
@@
 
+```{r}
+temp <- as.data.frame(ano)
+temp$slitasje <- apply(temp[,c("var_7TK", "var_7SE")], 1, sum, na.rm=T)
+ano$slitasje <- temp$slitasje
+```
+
 ```{r, fig.cap="Distribution of slitasje scores in the ANO data", fig.width=6}
-barplot(table(ano$slitasje), xlab="% Slitasje", ylab = "Number of localities")
+barplot(table(ano$slitasje), xlab="ADVS variable score", ylab = "Number of localities", las=2)
 ```
@@
 
-#### Distribution of Slitasje scores
-
-```{r, fig.cap="Slitasje scores (ANO Naturetype (and GRUK) data combined). The score for a locality equals the highest (worst) score of the related variables 7TK, 7SE, PRSL and PRTK."}
+#### Distribution of ADVS scores
+
+```{r, fig.cap="ADVS variable scores (ANO Naturetype data combined). The score for a locality equals the sum of the related variables 7TK, 7SE, PRSL and PRTK."}
 temp <- as.data.frame(table(slitasje_data$slitasje))
-ggplot(temp, aes(x = Var1,
+ggplot(temp, aes(x = as.numeric(Var1),
                  y = Freq))+
@@
   theme_bw(base_size = 12)+
-  labs(x = "slitasje score",
-       y = "Number of localities")
+  labs(x = "ADVS score",
+       y = "Number of localities")+
+  scale_x_continuous(
+    labels = scales::label_number(accuracy = 1))
 ```
@@
 
-```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the slitasje indicator.", fig.width=4}
+```{r, fig.cap="Barplot show the contribution (number of localities) of different data sets to the ADVS indicator.", fig.width=4}
 temp <- as.data.frame(table(slitasje_data$origin))
@@
 lower <- 100
-threshold <- 5
+threshold <- 10
 ```

### 2023-09-14 DRAFT_NDVI_wetland.Rmd
almost done with the ndvi wetland markdown

--- 2023-09-13 DRAFT_NDVI_wetland.Rmd
@@
 In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied.
-For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas.
-For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed.
+For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al.
+2020).
+For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021).
 In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e.
@@
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-In addition, we still explore NDVI data from the MODIS and Landsat satellites (see more details below).
+In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurments.
+We also still explore NDVI data from MODIS (see more details below).
 
 ## About the underlying data
-In the NDVI project for wetlands, we use four sets of data for building indicators for ecological condition:
+In the NDVI project for wetlands, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
 
 - NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+- remote sensed NDVI data
   - Sentinel-2 satellite NDVI data
@@
 (1) NiN ecosystem data:
+The NiN data in this study contain data about the ecosystem categories (cf.
+Halvorsen et al.
+2020) present at mapped sites as wall as their spatial delimitation (in the form of polygons).
+Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
+Polygon sizes in these NiN data may vary from <100 sqm to 1 sqkm.
 
-(2) Sentinel-2 satellite NDVI data:
+(2) NDVI data:
+The NDVI data used in the work at hand are pixel means for polygons in the NiN-data, i.e.
+each NiN polygon is represented by one NDVI value per measurement time (see below for more details on the temporal resolution of NDVI data).
+With respect to that, it does only make sense to use NDVI data for polygons that are at least of the same size as the resolution of the respective satellite, and thus we excluded any polygons smaller than each satellite's resolution.
 
-(3) MODIS satellite NDVI data:
+(2a) Sentinel-2 satellite NDVI data:
+The Sentinel data have a resolution of 10 x 10 m, and we thus excluded all NiN-polygons smaller than 100 sqm.
+Sentinel-2 has a revisit frequency of 10 days and should thus provide enough measurements during peak growing season to provide a robust measure of maximum greenness.
 
-(4) Landsat satellite NDVI data:
+(2b) MODIS satellite NDVI data:
+The MODIS data have a resolution of 250 x 250 m, and we thus excluded all NiN-polygons smaller than 62500 sqm.
+MODIS has a revisit frequency of 1-2 days and thus provides an extensive sample to provide a robust measure of maximum greenness.
 
-have to exclude all polygons smaller than the MODIS pixel size of 250 x 250 m, which reduces the number of available polygons greatly.
-To explore temporal trends even further back in time, we also include Landsat NDVI data, which span a time frame from 1984 to 2021, and have a resolution of 30 x 30 m, thus excluding only few polygons of insufficient size.
+(2c) Landsat satellite NDVI data:
+The Landsat data have a resolution of 30 x 30 m, and we thus excluded all NiN-polygons smaller than 900 sqm.
+Landsat has a revisit frequency of 16 days and thus provides the poorest growing season sample among the available NDVI data for a robust measure of maximum greenness.
 
 ### Representativity in time and space
-For wetlands, the NiN data contain 6503 wetland polygons of known area size across mainland Norway.
+For wetlands, the NiN data contain 6502 wetland polygons of known area size across mainland Norway.
 However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 6503 wetland polygons are distributed across regions and main ecosystem types in the following way:
+The 6502 wetland polygons are distributed across regions and major ecosystem types in the following way:
 
-----build a matrix-table----
+                  V1    V3    V4    V6    V9    V10
+Northern Norway   638   275   55    102   360   315
+Central Norway    702   1250  34    0     302   116
+Western Norway    296   111   30    0     460   158
+Eastern Norway    251   89    103   0     251   223
+Southern Norway   58    44    5     0     155   102
+
+In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region.
 
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data likely are even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
 
@@
 ## Reference state and values
-### Reference state
+### Reference state and scaling values
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
-Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for wetland ecosystems.
-In this exploratory work, we will use the condition classification in the NiN-data to explore its applicability for defining a reference state.
+Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for wetland ecosystems (cf.
+Jakobsson et al.
+2020, Töpper & Jakobsson 2021).
+In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to wetland ecology and condition and (2) its applicability for defining a reference state.
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-We will explore (i) if the condition classes in the NiN data can be used to define scaling values, and (ii) the potential for taking a probabilistic approach to scaling original NDVI values to a condition index based on probability density curves derived from above named models.
-
 ## Uncertainties
-Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area.
+Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
 
 ## References
-
-----not referred to in text yet, but likely will :)----
 
@@
 Ecological Indicators 116: 106492.
+
+Miljødiretoratets mapping manual: https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/
 
@@
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
-```{r}
+```{r message=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
   map_df(~fread(.))
-df.s
 ```
@@
 df.m$mean<-df.m$mean/10000
-df.m
 ```
@@
 df.l<-bind_rows(df_list, .id = "column_label")
-df.l
-
 ```
@@
 #### NDVI data
-```{r}
+```{r, results='hide'}
 ## Sentinel-2
@@
 
+# How are polygon sizes distributed?
 ```{r}
-## how are polygon sizes distributed?
 summary(SentinelNDVI.wetland$area_meters)
@@
 abline(v=100,lty=2)
-# wetland polygons stretch a large number of sizes from just above 100 sqm to mor ethan 4 million sqm. 75% of all polygons are below 14264 sqm.
+```
+Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
 
-## how does NDVI vary over the years (all years)
+# How does NDVI vary over the years
+```{r}
 SentinelNDVI.wetland %>%
@@
   facet_grid( tilstand~hovedtype)
-# 2022 stands out with the highest NDVI values missing
-# the pattern is strange, suggest to omit 2022 for this analysis
+```
+2022 stands out with the highest NDVI values missing.
+WE omit 2022 for this analysis
+```{r}
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+```
 
-## looking at major types for polygons in good condition only (and only for NDVI data matching NiN-mapping)
+# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
+```{r}
 SentinelNDVI.wetland %>%
@@
   geom_point(size=0.7, shape=16, color="grey")
-# NDVI varies between hovedtyper: highest in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3). The overlaps are considerable.
 
-## looking at subtypes for polygons in good condition only (and only for NDVI data matching NiN-mapping)
 SentinelNDVI.wetland %>%
@@
   facet_wrap( ~hovedtype)
-# NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), but the overlap is large.
+```
+NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3).
+The overlaps between the ecosystem types are considerable.
+With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.
 
-## NDVI across regions for polygons in good condition only (for major types and NDVI years data matching NiN-mapping years)
+# How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)
+```{r}
 SentinelNDVI.wetland %>%
@@
   theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
-# no obvious trends in geography
+```
+There is no obvious trends in geography.
 
-## NDVI across condition classes (for major types and NDVI years data matching NiN-mapping years)
+# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
+```{r}
 SentinelNDVI.wetland %>%
@@
   facet_wrap( ~hovedtype)
-# in seminat types (V9,V10), kaldkilde (V4), and bogs (V3) ndvi seems to increase as condition deteriorates, in fens (V1) not much of a pattern.
+```
+In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern.
 Overlaps between condition classes are generally very large.
-```
 
-Which ecological relevant variable(s) is the condition measure in the NiN data related to?
+# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
 ```{r}
@@
   facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none")
+  theme(legend.position="none") +
+  ggtitle("Condition and condition variables in semi-natural mires (V9)") +
+  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
 
@@
   facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
-  theme(legend.position="none")
+  theme(legend.position="none") +
+  ggtitle("Condition and condition variables in semi-natural wet-meadows (V10)") +
+  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
 
@@
   facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none")
+  theme(legend.position="none") +
+  ggtitle("Condition and condition variables in minerotrophic fens (V1)") +
+  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
 
 ```
-In these plots, condition=1 is best and 4 is worst (degraded). The panels for V1, V9 and V10 differ because Mdir-instruks uses different variables in these major ecosystem types to define condition.
+In these plots, the panels for V1, V9 and V10 differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
 
@@
 Thus, NDVI can, in principle, be indicative of ecological condition in semi-natural wetlands due to a direct connection to the main pressures on the ecosystems.
-
-In V1 (minerotrophic mires), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
+In V1 (minerotrophic fens), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
 Under these drivers, greenness would not directly be affected.
 although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run.
-Currently, NDVI seems poorly linked to ecological condition in and the major pressures on minerotrophic mires.
+Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.
 
@@
 We can investigate further how NDVI and condition are connected using regression analyses.
+
+NDVI data are bounded between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data.
+We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
 ```{r}
-# NDVI data are bound between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data
-# we can transform the variable to be bounded between 0 and 1 and use beta-regression models
 SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
@@
 # we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
-
 SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
@@
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 model.wetland.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
@@
 # Northern Norway
-```{r}
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
 # Central Norway
-```{r}
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
 # Western Norway
-```{r}
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
 # Eastern Norway
-```{r}
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
 # Southern Norway
-```{r}
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
 This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural wetland ecosystems.
-In natural fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates.
+In natural minerotrophic fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates.
 A factor with potential for affecting vegetation composition and productivity, and thus NDVI, is Nitrogen deposition.
 Nitrogen deposition is higher in Southern parts of Norway than further north, and could thus be responsible for increased NDVI with reduced condition in the South, but it is not clear how Nitrogen deposition and the variables underlying the condition assessment in V1 (Ditches, Alien species, Heavy vehicles, Wear/tear) are connected.
+Furthermore we are to date lacking documented empirical ground-truthed information about the connection between NDVI and Nitrogen effects in wetlands.
 In bogs (V3) there is no consistent pattern.
 
-### Regression analyses, NDVI across time
-As a last step, we can investigate how NDVI has changed over time.
+### NDVI across time - Sentinel, MODIS & LandSat
+As a last step, we can investigate how NDVI has changed over time. For this we include data from both MODIS and Landsat in addition to Sentinel-2.
 
-First, there's again some data handling to do. We merge the Sentinel, Modis, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
-```{r}
+First, there's again some data handling to do. We merge the Sentinel, MODIS, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
+```{r results='hide'}
 ## data handling for time series analysis
@@
 
-# check if the three Satelite objects have the same structure (for concatenating them)
+# check if the three Satellite objects have the same structure (for concatenating them)
 names(SentinelNDVI.wetland)
@@
 ```
-Now we can plot the time series for each main ecosystem type, and for each satellite separately  
+Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
 ```{r}
@@
 
-As a final step, we model NDVI as a function of year and main ecosystem type separately for each Satellite
-```{r}
-model.wetland.time.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland)
-model.wetland.time.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland)
-model.wetland.time.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland)
+We can test if there are consistent temporal changes in NDVI by running regressions including year as a covariate.
 
-summary(model.wetland.time.Sent)$coefficients$cond
-summary(model.wetland.time.Modi)$coefficients$cond
-summary(model.wetland.time.Land)$coefficients$cond
+We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
+```{r message=FALSE, warning=FALSE}
+model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
+model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])
+
+summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
+summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
+summary(model.wetland.time.GodTilst.Land)$coefficients$cond
 ```
+There's an increase in NDVI over time in all three satellite time series in essentially all major wetland ecosystem types in good condition.
+It is worth pointing out though that the model on the MODIS data shows 1/10 of the effect size for the time slope and has way less statistical power compared to Sentinel and Landsat.
+
+Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger in semi-natural wetlands?
+```{r message=FALSE, warning=FALSE}
+model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
+model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9","V10"),])
+model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9","V10"),])
+summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
+summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
+summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
+```
+Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat.
+For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.
+
+# Overall conclusion
+NDVI in wetland ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
+Especially the semi-natural wetland ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
+The large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
+Since ecological condition also relates to the temporal development of NDVI values in the analyses above, an alternative approach can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
+2022, see references above).
+
+Beyond these considerations, there are two general issues which limit a more comprehensive and robust application of satellite based NDVI-data in an ecological condition context:
+
+# 1) We are lacking ecosystem maps for wetland ecosystems
+The study at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within wetlands separately.
+# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
+Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands.
+Here, we need field studies investigating:
+- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition
+
+- How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.
+The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
+Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
+At the same time, wetlands are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+
+In addition, further investigations should test how NDVI-condition pattern vary across biogeographic regions in addition to administrative regions (which are ecologically irrelevant), in order to better understand the results for the administrative regions.
 

### 2023-09-14 R/NDVI/1_data_handling.R

@@ -312,7 +312,8 @@ levels(nin.wetland$region)
 nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
 levels(nin.wetland$region)
 
-
+# check # of wetland types in different regions
+nin.wetland %>% group_by(hovedtype, region) %>% summarise(n = n()) %>% print(n=30)
 
 
 

### 2023-09-14 DRAFT_NDVI_seminat.Rmd
starting the seminat markdown

--- 2023-09-14 DRAFT_NDVI_wetland.Rmd
@@
-# NDVI (greenness) indicator, wetland ecosystems  {#NDVI-indicator-wetland}
+# NDVI (greenness) indicator, semi-natural ecosystems  {#NDVI-indicator-seminat}
 
@@
 
-June 2023
+September 2023
 
@@
 national maps for the forest and mountain ecosystems).
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
-excluding forest swamps).
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems.
 For this task, two initial challenges can be identified early on:
-1) We don't yet have a national wetland map, i.e.
-we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+1) We don't yet have a national map for semi-natural ecosystems, i.e.
+we don't know the exact locations of all semi-natural areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
+2) The spatial extent of the single semi-natural ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
 
-In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
+In this draft development for a NDVI indicator for ecological condition in semi-natural  systems we approach these challenges in the following ways:
 1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
-Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of semi-natural ecosystems are available.
+2) We include Sentinel-2 imagery for semi-natural ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all semi-natural polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurments.
+In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurements.
 We also still explore NDVI data from MODIS (see more details below).
@@
 ## About the underlying data
-In the NDVI project for wetlands, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
+In the NDVI project for semi-natural systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
 
-- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets instruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
 - remote sensed NDVI data
@@
 ### Representativity in time and space
-For wetlands, the NiN data contain 6502 wetland polygons of known area size across mainland Norway.
+For semi-natural systems, the NiN data contain 6502 semi-natural polygons of known area size across mainland Norway.
 However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 6502 wetland polygons are distributed across regions and major ecosystem types in the following way:
+The 6502 semi-natural polygons are distributed across regions and major ecosystem types in the following way:
 
@@
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the semi-natural occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
 
 ### Temporal coverage
-The currently available NiN data on wetland ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
+The currently available NiN data on semi-natural ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
 Note that none of the polygons in the NiN-data has been visited more than once during that period.
@@
 ## Collinearities with other indicators
-The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in drained wetland areas under shrub and tree encroachment.
+The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in abandoned semi-natural areas under shrub and tree encroachment.
 
@@
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
-Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for wetland ecosystems (cf.
+Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for semi-natural ecosystems (cf.
 Jakobsson et al.
 2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to wetland ecology and condition and (2) its applicability for defining a reference state.
+In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to semi-natural ecology and condition and (2) its applicability for defining a reference state.
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
@@
 The analyses in this document make use of the following data sets:
--NiN data (Naturtyper etter Miljødirektoratets innstruks)
+-NiN data (Naturtyper etter Miljødirektoratets instruks)
 -Sentinel-2 NDVI data
@@
 
-## filter out only wetland data
-nin.wetland <- nin %>% 
+## filter out only semi-natural data
+nin.semi-natural <- nin %>% 
   filter(hovedoekosystem %in% c('Vaatmark')) %>%
@@
 # merge NiN-data with region
-nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
-nin.wetland
+nin.semi-natural = st_join(nin.semi-natural, regnor, left = TRUE)
+nin.semi-natural
 
-colnames(nin.wetland)[c(1,9)] <- c("id","region_id")
+colnames(nin.semi-natural)[c(1,9)] <- c("id","region_id")
 
 # check that every nin-polygon still occurs only once
-summary(as.factor(nin.wetland$id))
-nin.wetland[nin.wetland$id=="NINFP2110008060",]
+summary(as.factor(nin.semi-natural$id))
+nin.semi-natural[nin.semi-natural$id=="NINFP2110008060",]
 # this one was assigned to both Western and Central Norway, we drop the latter
-nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% '1144.1',]
+nin.semi-natural <- nin.semi-natural[!row.names(nin.semi-natural) %in% '1144.1',]
 
-nin.wetland <- nin.wetland %>% 
-  mutate(area_meters_nin = st_area(nin.wetland)
+nin.semi-natural <- nin.semi-natural %>% 
+  mutate(area_meters_nin = st_area(nin.semi-natural)
   )
 # check and edit the order of regions
-levels(nin.wetland$region)
-nin.wetland$region <- as.factor(nin.wetland$region)
-levels(nin.wetland$region)
-nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
-levels(nin.wetland$region)
+levels(nin.semi-natural$region)
+nin.semi-natural$region <- as.factor(nin.semi-natural$region)
+levels(nin.semi-natural$region)
+nin.semi-natural$region <- factor(nin.semi-natural$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+levels(nin.semi-natural$region)
 
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.wetland2 <- nin.wetland %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
+nin.semi-natural2 <- nin.semi-natural %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
   separate(col = ninbeskrivelsesvariabler,
@@
   subset(select = -ninbeskrivelsesvariabler)
-nin.wetland2 <- as.data.frame(nin.wetland2)
+nin.semi-natural2 <- as.data.frame(nin.semi-natural2)
 
@@
 #load processed NiN-data from cache
-nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
-nin.wetland2 <- readRDS(paste0(here::here(),"/data/cache/nin.wetland2.RDS"))
+nin.semi-natural <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural.RDS"))
+nin.semi-natural2 <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural2.RDS"))
 
@@
 ## Sentinel-2
-# join nin.wetland & Sentinel NDVI data
-SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
-#summary(SentinelNDVI.wetland)
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+# join nin.semi-natural & Sentinel NDVI data
+SentinelNDVI.semi-natural <- full_join(nin.semi-natural, df.s, by="id")
+#summary(SentinelNDVI.semi-natural)
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.wetland))
-#summary(SentinelNDVI.wetland)
-# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(hovedtype))
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
-#summary(SentinelNDVI.wetland)
+         area_meters = st_area(SentinelNDVI.semi-natural))
+#summary(SentinelNDVI.semi-natural)
+# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(hovedtype))
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(mean))
+#summary(SentinelNDVI.semi-natural)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-dim(SentinelNDVI.wetland)
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
-dim(SentinelNDVI.wetland)
+dim(SentinelNDVI.semi-natural)
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 100)
+dim(SentinelNDVI.semi-natural)
 # split date into year, month & day
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # add column for sub-ecosystem types
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                         subtype = str_remove(subtype, '-'),
@@
 # we are using max NDVI per year in every NiN polygon
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 
-summary(SentinelNDVI.wetland)
+summary(SentinelNDVI.semi-natural)
 
 ## MODIS
-# join nin.wetland and Modis NDVI data
-ModisNDVI.wetland <- full_join(nin.wetland, df.m, by="id")
+# join nin.semi-natural and Modis NDVI data
+ModisNDVI.semi-natural <- full_join(nin.semi-natural, df.m, by="id")
 
-#summary(ModisNDVI.wetland)
-ModisNDVI.wetland <- ModisNDVI.wetland %>%
+#summary(ModisNDVI.semi-natural)
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.wetland))
-#summary(ModisNDVI.wetland)
-# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
-ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(hovedtype))
-ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
-#summary(ModisNDVI.wetland)
+         area_meters = st_area(ModisNDVI.semi-natural))
+#summary(ModisNDVI.semi-natural)
+# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(hovedtype))
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(mean))
+#summary(ModisNDVI.semi-natural)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-dim(ModisNDVI.wetland)
-ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
-dim(ModisNDVI.wetland)
+dim(ModisNDVI.semi-natural)
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 62500)
+dim(ModisNDVI.semi-natural)
 # split date into year, month & day
-ModisNDVI.wetland <- ModisNDVI.wetland %>%
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-ModisNDVI.wetland <- ModisNDVI.wetland %>%
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
   group_by(id, year) %>%
   filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.wetland)
+#summary(ModisNDVI.semi-natural)
 
 ## Landsat
-# join nin.wetland & Landsat NDVI data
-LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by="id")
+# join nin.semi-natural & Landsat NDVI data
+LandsatNDVI.semi-natural <- full_join(nin.semi-natural, df.l, by="id")
 
-#summary(LandsatNDVI.wetland)
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+#summary(LandsatNDVI.semi-natural)
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.wetland))
-#summary(LandsatNDVI.wetland)
-# get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(hovedtype))
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
-#summary(LandsatNDVI.wetland)
+         area_meters = st_area(LandsatNDVI.semi-natural))
+#summary(LandsatNDVI.semi-natural)
+# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(hovedtype))
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(mean))
+#summary(LandsatNDVI.semi-natural)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-dim(LandsatNDVI.wetland)
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
-dim(LandsatNDVI.wetland)
+dim(LandsatNDVI.semi-natural)
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 900)
+dim(LandsatNDVI.semi-natural)
 # split date into year, month & day
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 
-#summary(LandsatNDVI.wetland)
+#summary(LandsatNDVI.semi-natural)
 ```
@@
 ```{r}
-summary(SentinelNDVI.wetland$area_meters)
-hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
+summary(SentinelNDVI.semi-natural$area_meters)
+hist(SentinelNDVI.semi-natural$area_meters,xlim=c(0,20000),breaks=50000)
 abline(v=100,lty=2)
 ```
-Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
+semi-natural polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
 
@@
 ```{r}
-SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural %>%
   ggplot( aes(x=year, y=mean )) + 
@@
 ```{r}
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(year != '2022')
 ```
@@
 ```{r}
-SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural %>%
   filter(tilstand == 'God') %>%
@@
 
-SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural %>%
   filter(tilstand == 'God') %>%
@@
 
-# How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)
+# How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)
 ```{r}
-SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural %>%
   filter(tilstand == 'God') %>%
@@
 ```{r}
-SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural %>%
   filter(year == kartleggingsaar) %>%
@@
 ```{r}
-nin.wetland2 %>%
+nin.semi-natural2 %>%
   filter(hovedtype=='V9') %>%
@@
 
-nin.wetland2 %>%
+nin.semi-natural2 %>%
   filter(hovedtype=='V10') %>%
@@
 
-nin.wetland2 %>%
+nin.semi-natural2 %>%
   filter(hovedtype=='V1') %>%
@@
 In both cases, we'd expect an increase in the greenness signal.
-Thus, NDVI can, in principle, be indicative of ecological condition in semi-natural wetlands due to a direct connection to the main pressures on the ecosystems.
+Thus, NDVI can, in principle, be indicative of ecological condition in semi-natural semi-naturals due to a direct connection to the main pressures on the ecosystems.
 In V1 (minerotrophic fens), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
@@
 ```{r}
-SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
+SentinelNDVI.semi-natural$mean_beta <- (SentinelNDVI.semi-natural$mean + 1) / 2
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
 # we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
-SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural.train <- SentinelNDVI.semi-natural %>%
   filter(year == kartleggingsaar) %>%
@@
 # make an extra numeric condition variable
-unique(SentinelNDVI.wetland.train$tilstand)
-SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
+unique(SentinelNDVI.semi-natural.train$tilstand)
+SentinelNDVI.semi-natural.train <- SentinelNDVI.semi-natural.train %>% mutate(tilstand_num = recode(tilstand, 
                                             "God" = '0',
@@
                                             "Svært redusert" = '3'))
-SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
+SentinelNDVI.semi-natural.train$tilstand_num <- as.numeric(SentinelNDVI.semi-natural.train$tilstand_num)
 
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.wetland.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
-model.wetland.cond.Sent <- StepBeta(model.wetland.cond.Sent)
-summary(model.wetland.cond.Sent)
+model.semi-natural.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.semi-natural.train)
+model.semi-natural.cond.Sent <- StepBeta(model.semi-natural.cond.Sent)
+summary(model.semi-natural.cond.Sent)
 ```
@@
 ```{r, include = F}
-rm(model.wetland.cond.Sent)
+rm(model.semi-natural.cond.Sent)
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-model.wetland.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
-model.wetland.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
-model.wetland.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
-model.wetland.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
-model.wetland.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
+model.semi-natural.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Northern Norway",])
+model.semi-natural.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Central Norway",])
+model.semi-natural.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Western Norway",])
+model.semi-natural.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Eastern Norway",])
+model.semi-natural.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Southern Norway",])
 
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.wetland.train %>%
+SentinelNDVI.semi-natural.train %>%
   filter(region=="Northern Norway") %>%
@@
 
-summary(model.wetland.cond.Sent.N)$coefficients
+summary(model.semi-natural.cond.Sent.N)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.wetland.train %>%
+SentinelNDVI.semi-natural.train %>%
   filter(region=="Central Norway") %>%
@@
 
-summary(model.wetland.cond.Sent.C)$coefficients
+summary(model.semi-natural.cond.Sent.C)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.wetland.train %>%
+SentinelNDVI.semi-natural.train %>%
   filter(region=="Western Norway") %>%
@@
 
-summary(model.wetland.cond.Sent.W)$coefficients
+summary(model.semi-natural.cond.Sent.W)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.wetland.train %>%
+SentinelNDVI.semi-natural.train %>%
   filter(region=="Eastern Norway") %>%
@@
 
-summary(model.wetland.cond.Sent.E)$coefficients
+summary(model.semi-natural.cond.Sent.E)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.wetland.train %>%
+SentinelNDVI.semi-natural.train %>%
   filter(region=="Southern Norway") %>%
@@
 
-summary(model.wetland.cond.Sent.S)$coefficients
+summary(model.semi-natural.cond.Sent.S)$coefficients
 ```
@@
 
-In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
-This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural wetland ecosystems.
+In semi-natural semi-natural ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
+This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural semi-natural ecosystems.
 In natural minerotrophic fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates.
@@
 Nitrogen deposition is higher in Southern parts of Norway than further north, and could thus be responsible for increased NDVI with reduced condition in the South, but it is not clear how Nitrogen deposition and the variables underlying the condition assessment in V1 (Ditches, Alien species, Heavy vehicles, Wear/tear) are connected.
-Furthermore we are to date lacking documented empirical ground-truthed information about the connection between NDVI and Nitrogen effects in wetlands.
+Furthermore we are to date lacking documented empirical ground-truthed information about the connection between NDVI and Nitrogen effects in semi-naturals.
 In bogs (V3) there is no consistent pattern.
@@
 # Sentinel time series checked in exploratory analysis script
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
   group_by(id, year) %>%
   filter(mean == max(mean, na.rm=TRUE))
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(year != '2022')
 
 # checking time series for MODIS
-ModisNDVI.wetland %>%
+ModisNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 # 2022 does not stand out as in the Sentinel data, so we keep it
-ModisNDVI.wetland <- ModisNDVI.wetland %>%
+ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 # checking time series for Landsat
-LandsatNDVI.wetland %>%
+LandsatNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 # nothing worrying to see here either
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
   group_by(id, year) %>%
@@
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
-ModisNDVI.wetland$mean_beta <- (ModisNDVI.wetland$mean + 1) / 2
-LandsatNDVI.wetland$mean_beta <- (LandsatNDVI.wetland$mean + 1) / 2
+#SentinelNDVI.semi-natural$mean_beta <- (SentinelNDVI.semi-natural$mean + 1) / 2
+ModisNDVI.semi-natural$mean_beta <- (ModisNDVI.semi-natural$mean + 1) / 2
+LandsatNDVI.semi-natural$mean_beta <- (LandsatNDVI.semi-natural$mean + 1) / 2
 
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.wetland$mean_beta)
-summary(ModisNDVI.wetland$mean_beta)
-summary(LandsatNDVI.wetland$mean_beta)
+summary(SentinelNDVI.semi-natural$mean_beta)
+summary(ModisNDVI.semi-natural$mean_beta)
+summary(LandsatNDVI.semi-natural$mean_beta)
 # replace 1s in Landsat data with 0.9999
-LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
+LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
   mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-names(SentinelNDVI.wetland)
-names(ModisNDVI.wetland)
-names(LandsatNDVI.wetland)
+names(SentinelNDVI.semi-natural)
+names(ModisNDVI.semi-natural)
+names(LandsatNDVI.semi-natural)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
 # one column is named slightly differently in the Sentinel data -> rename it
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
+SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
   rename('system.index' = 'system:index')
@@
 # check if they have the same geometry
-st_crs(SentinelNDVI.wetland)
-st_crs(ModisNDVI.wetland)
-st_crs(LandsatNDVI.wetland)
+st_crs(SentinelNDVI.semi-natural)
+st_crs(ModisNDVI.semi-natural)
+st_crs(LandsatNDVI.semi-natural)
 # all good
@@
 # add an increment to the year variable to avoid overlapping data being hidden in figures
-SentinelNDVI.wetland$year_jit <- SentinelNDVI.wetland$year + 0.3
-ModisNDVI.wetland$year_jit <- ModisNDVI.wetland$year - 0.3
-LandsatNDVI.wetland$year_jit <- LandsatNDVI.wetland$year
+SentinelNDVI.semi-natural$year_jit <- SentinelNDVI.semi-natural$year + 0.3
+ModisNDVI.semi-natural$year_jit <- ModisNDVI.semi-natural$year - 0.3
+LandsatNDVI.semi-natural$year_jit <- LandsatNDVI.semi-natural$year
 
 # concatenate the three Satellite objects
-allSatNDVI.wetland <- rbind(
-SentinelNDVI.wetland[,!names(SentinelNDVI.wetland) %in% "subtype"],
-ModisNDVI.wetland,
-LandsatNDVI.wetland[,!names(LandsatNDVI.wetland) %in% "column_label"]
+allSatNDVI.semi-natural <- rbind(
+SentinelNDVI.semi-natural[,!names(SentinelNDVI.semi-natural) %in% "subtype"],
+ModisNDVI.semi-natural,
+LandsatNDVI.semi-natural[,!names(LandsatNDVI.semi-natural) %in% "column_label"]
 )
 # add variable for Satellite identity
-allSatNDVI.wetland$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.wetland)),
-  rep("Modis",nrow(ModisNDVI.wetland)),
-  rep("Landsat",nrow(LandsatNDVI.wetland))
+allSatNDVI.semi-natural$Sat <- c(
+  rep("Sentinel",nrow(SentinelNDVI.semi-natural)),
+  rep("Modis",nrow(ModisNDVI.semi-natural)),
+  rep("Landsat",nrow(LandsatNDVI.semi-natural))
 )
-allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat,levels=c("Sentinel","Modis","Landsat"))
-levels(allSatNDVI.wetland$Sat)
+allSatNDVI.semi-natural$Sat <- factor(allSatNDVI.semi-natural$Sat,levels=c("Sentinel","Modis","Landsat"))
+levels(allSatNDVI.semi-natural$Sat)
 ```
@@
 # plot 
-allSatNDVI.wetland %>%
+allSatNDVI.semi-natural %>%
   ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
@@
 ```{r message=FALSE, warning=FALSE}
-model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
-model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
-model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])
+model.semi-natural.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.semi-natural[SentinelNDVI.semi-natural$tilstand=="God",])
+model.semi-natural.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.semi-natural[ModisNDVI.semi-natural$tilstand=="God",])
+model.semi-natural.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.semi-natural[LandsatNDVI.semi-natural$tilstand=="God",])
 
-summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
-summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
-summary(model.wetland.time.GodTilst.Land)$coefficients$cond
+summary(model.semi-natural.time.GodTilst.Sent)$coefficients$cond
+summary(model.semi-natural.time.GodTilst.Modi)$coefficients$cond
+summary(model.semi-natural.time.GodTilst.Land)$coefficients$cond
 ```
-There's an increase in NDVI over time in all three satellite time series in essentially all major wetland ecosystem types in good condition.
+There's an increase in NDVI over time in all three satellite time series in essentially all major semi-natural ecosystem types in good condition.
 It is worth pointing out though that the model on the MODIS data shows 1/10 of the effect size for the time slope and has way less statistical power compared to Sentinel and Landsat.
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger in semi-natural wetlands?
+Given the systematically higher NDVI-values with decreasing condition in semi-natural semi-naturals, We may ask the question if the temporal NDVI increase might be stronger in semi-natural semi-naturals?
 ```{r message=FALSE, warning=FALSE}
-model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
-model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9","V10"),])
-model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9","V10"),])
-summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
-summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
-summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
+model.semi-natural.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.semi-natural[SentinelNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
+model.semi-natural.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.semi-natural[ModisNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
+model.semi-natural.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.semi-natural[LandsatNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
+summary(model.semi-natural.time.tilst.V9V10.Sent)$coefficients$cond
+summary(model.semi-natural.time.tilst.V9V10.Modi)$coefficients$cond
+summary(model.semi-natural.time.tilst.V9V10.Land)$coefficients$cond
 ```
-Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat.
+Indeed, the temporal NDVI increase has been stronger in semi-natural semi-naturals with reduced ecological condition, in both Sentinel and Landsat.
 For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.
@@
 # Overall conclusion
-NDVI in wetland ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
-Especially the semi-natural wetland ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
+NDVI in semi-natural ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
+Especially the semi-natural semi-natural ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
 The large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
@@
 
-# 1) We are lacking ecosystem maps for wetland ecosystems
-The study at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
-Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within wetlands separately.
-# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands.
+# 1) We are lacking ecosystem maps for semi-natural ecosystems
+The study at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within semi-naturals separately.
+# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
+Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-naturals.
 Here, we need field studies investigating:
@@
 
-- How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.
+- How reflectance in different semi-natural types is affected by how moist the moss layer is at the time of measuring.
 The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
-Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, wetlands are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+Minerotrophic semi-naturals (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
+At the same time, semi-naturals are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
 

### 2023-09-15 DRAFT_NDVI_seminat.Rmd
ndvi seminat conntinued (data hanndling)

@@
 ## filter out only semi-natural data
-nin.semi-natural <- nin %>% 
+nin.seminat <- nin %>% 
   filter(hovedoekosystem %in% c('Vaatmark')) %>%
@@
 # merge NiN-data with region
-nin.semi-natural = st_join(nin.semi-natural, regnor, left = TRUE)
-nin.semi-natural
+nin.seminat = st_join(nin.seminat, regnor, left = TRUE)
+nin.seminat
 
-colnames(nin.semi-natural)[c(1,9)] <- c("id","region_id")
+colnames(nin.seminat)[c(1,9)] <- c("id","region_id")
 
 # check that every nin-polygon still occurs only once
-summary(as.factor(nin.semi-natural$id))
-nin.semi-natural[nin.semi-natural$id=="NINFP2110008060",]
+summary(as.factor(nin.seminat$id))
+nin.seminat[nin.seminat$id=="NINFP2110008060",]
 # this one was assigned to both Western and Central Norway, we drop the latter
-nin.semi-natural <- nin.semi-natural[!row.names(nin.semi-natural) %in% '1144.1',]
+nin.seminat <- nin.seminat[!row.names(nin.seminat) %in% '1144.1',]
 
-nin.semi-natural <- nin.semi-natural %>% 
-  mutate(area_meters_nin = st_area(nin.semi-natural)
+nin.seminat <- nin.seminat %>% 
+  mutate(area_meters_nin = st_area(nin.seminat)
   )
 # check and edit the order of regions
-levels(nin.semi-natural$region)
-nin.semi-natural$region <- as.factor(nin.semi-natural$region)
-levels(nin.semi-natural$region)
-nin.semi-natural$region <- factor(nin.semi-natural$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
-levels(nin.semi-natural$region)
+levels(nin.seminat$region)
+nin.seminat$region <- as.factor(nin.seminat$region)
+levels(nin.seminat$region)
+nin.seminat$region <- factor(nin.seminat$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+levels(nin.seminat$region)
 
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.semi-natural2 <- nin.semi-natural %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
+nin.seminat2 <- nin.seminat %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
   separate(col = ninbeskrivelsesvariabler,
@@
   subset(select = -ninbeskrivelsesvariabler)
-nin.semi-natural2 <- as.data.frame(nin.semi-natural2)
+nin.seminat2 <- as.data.frame(nin.seminat2)
 
@@
 #load processed NiN-data from cache
-nin.semi-natural <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural.RDS"))
-nin.semi-natural2 <- readRDS(paste0(here::here(),"/data/cache/nin.semi-natural2.RDS"))
+nin.seminat <- readRDS(paste0(here::here(),"/data/cache/nin.seminat.RDS"))
+nin.seminat2 <- readRDS(paste0(here::here(),"/data/cache/nin.seminat2.RDS"))
 
@@
 ## Sentinel-2
-# join nin.semi-natural & Sentinel NDVI data
-SentinelNDVI.semi-natural <- full_join(nin.semi-natural, df.s, by="id")
-#summary(SentinelNDVI.semi-natural)
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
+# join nin.seminat & Sentinel NDVI data
+SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by="id")
+#summary(SentinelNDVI.seminat)
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.semi-natural))
-#summary(SentinelNDVI.semi-natural)
+         area_meters = st_area(SentinelNDVI.seminat))
+#summary(SentinelNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(hovedtype))
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(!is.na(mean))
-#summary(SentinelNDVI.semi-natural)
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(hovedtype))
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(mean))
+#summary(SentinelNDVI.seminat)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-dim(SentinelNDVI.semi-natural)
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 100)
-dim(SentinelNDVI.semi-natural)
+dim(SentinelNDVI.seminat)
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(as.numeric(area_meters) >= 100)
+dim(SentinelNDVI.seminat)
 # split date into year, month & day
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # add column for sub-ecosystem types
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                         subtype = str_remove(subtype, '-'),
@@
 # we are using max NDVI per year in every NiN polygon
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
   group_by(id, year) %>%
@@
 
-summary(SentinelNDVI.semi-natural)
+summary(SentinelNDVI.seminat)
 
 ## MODIS
-# join nin.semi-natural and Modis NDVI data
-ModisNDVI.semi-natural <- full_join(nin.semi-natural, df.m, by="id")
+# join nin.seminat and Modis NDVI data
+ModisNDVI.seminat <- full_join(nin.seminat, df.m, by="id")
 
-#summary(ModisNDVI.semi-natural)
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
+#summary(ModisNDVI.seminat)
+ModisNDVI.seminat <- ModisNDVI.seminat %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.semi-natural))
-#summary(ModisNDVI.semi-natural)
+         area_meters = st_area(ModisNDVI.seminat))
+#summary(ModisNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(hovedtype))
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(!is.na(mean))
-#summary(ModisNDVI.semi-natural)
+ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(!is.na(hovedtype))
+ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(!is.na(mean))
+#summary(ModisNDVI.seminat)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-dim(ModisNDVI.semi-natural)
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 62500)
-dim(ModisNDVI.semi-natural)
+dim(ModisNDVI.seminat)
+ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(as.numeric(area_meters) >= 62500)
+dim(ModisNDVI.seminat)
 # split date into year, month & day
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
+ModisNDVI.seminat <- ModisNDVI.seminat %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
+ModisNDVI.seminat <- ModisNDVI.seminat %>%
   group_by(id, year) %>%
   filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.semi-natural)
+#summary(ModisNDVI.seminat)
 
 ## Landsat
-# join nin.semi-natural & Landsat NDVI data
-LandsatNDVI.semi-natural <- full_join(nin.semi-natural, df.l, by="id")
+# join nin.seminat & Landsat NDVI data
+LandsatNDVI.seminat <- full_join(nin.seminat, df.l, by="id")
 
-#summary(LandsatNDVI.semi-natural)
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
+#summary(LandsatNDVI.seminat)
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.semi-natural))
-#summary(LandsatNDVI.semi-natural)
+         area_meters = st_area(LandsatNDVI.seminat))
+#summary(LandsatNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(hovedtype))
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(!is.na(mean))
-#summary(LandsatNDVI.semi-natural)
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(!is.na(hovedtype))
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(!is.na(mean))
+#summary(LandsatNDVI.seminat)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-dim(LandsatNDVI.semi-natural)
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>% filter(as.numeric(area_meters) >= 900)
-dim(LandsatNDVI.semi-natural)
+dim(LandsatNDVI.seminat)
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(as.numeric(area_meters) >= 900)
+dim(LandsatNDVI.seminat)
 # split date into year, month & day
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
   group_by(id, year) %>%
@@
 
-#summary(LandsatNDVI.semi-natural)
+#summary(LandsatNDVI.seminat)
 ```
@@
 ```{r}
-summary(SentinelNDVI.semi-natural$area_meters)
-hist(SentinelNDVI.semi-natural$area_meters,xlim=c(0,20000),breaks=50000)
+summary(SentinelNDVI.seminat$area_meters)
+hist(SentinelNDVI.seminat$area_meters,xlim=c(0,20000),breaks=50000)
 abline(v=100,lty=2)
@@
 ```{r}
-SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat %>%
   ggplot( aes(x=year, y=mean )) + 
@@
 ```{r}
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(year != '2022')
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != '2022')
 ```
@@
 ```{r}
-SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat %>%
   filter(tilstand == 'God') %>%
@@
 
-SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat %>%
   filter(tilstand == 'God') %>%
@@
 ```{r}
-SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat %>%
   filter(year == kartleggingsaar) %>%
@@
 ```{r}
-nin.semi-natural2 %>%
+nin.seminat2 %>%
   filter(hovedtype=='V9') %>%
@@
 
-nin.semi-natural2 %>%
+nin.seminat2 %>%
   filter(hovedtype=='V10') %>%
@@
 
-nin.semi-natural2 %>%
+nin.seminat2 %>%
   filter(hovedtype=='V1') %>%
@@
 ```{r}
-SentinelNDVI.semi-natural$mean_beta <- (SentinelNDVI.semi-natural$mean + 1) / 2
+SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
 # we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
-SentinelNDVI.semi-natural.train <- SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
   filter(year == kartleggingsaar) %>%
@@
 # make an extra numeric condition variable
-unique(SentinelNDVI.semi-natural.train$tilstand)
-SentinelNDVI.semi-natural.train <- SentinelNDVI.semi-natural.train %>% mutate(tilstand_num = recode(tilstand, 
+unique(SentinelNDVI.seminat.train$tilstand)
+SentinelNDVI.seminat.train <- SentinelNDVI.seminat.train %>% mutate(tilstand_num = recode(tilstand, 
                                             "God" = '0',
@@
                                             "Svært redusert" = '3'))
-SentinelNDVI.semi-natural.train$tilstand_num <- as.numeric(SentinelNDVI.semi-natural.train$tilstand_num)
+SentinelNDVI.seminat.train$tilstand_num <- as.numeric(SentinelNDVI.seminat.train$tilstand_num)
 
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.semi-natural.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.semi-natural.train)
+model.semi-natural.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.seminat.train)
 model.semi-natural.cond.Sent <- StepBeta(model.semi-natural.cond.Sent)
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Northern Norway",])
-model.semi-natural.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Central Norway",])
-model.semi-natural.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Western Norway",])
-model.semi-natural.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Eastern Norway",])
-model.semi-natural.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.semi-natural.train[SentinelNDVI.semi-natural.train$region=="Southern Norway",])
+model.semi-natural.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Northern Norway",])
+model.semi-natural.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Central Norway",])
+model.semi-natural.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Western Norway",])
+model.semi-natural.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Eastern Norway",])
+model.semi-natural.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Southern Norway",])
 
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.semi-natural.train %>%
+SentinelNDVI.seminat.train %>%
   filter(region=="Northern Norway") %>%
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.semi-natural.train %>%
+SentinelNDVI.seminat.train %>%
   filter(region=="Central Norway") %>%
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.semi-natural.train %>%
+SentinelNDVI.seminat.train %>%
   filter(region=="Western Norway") %>%
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.semi-natural.train %>%
+SentinelNDVI.seminat.train %>%
   filter(region=="Eastern Norway") %>%
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.semi-natural.train %>%
+SentinelNDVI.seminat.train %>%
   filter(region=="Southern Norway") %>%
@@
 # Sentinel time series checked in exploratory analysis script
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
   group_by(id, year) %>%
   filter(mean == max(mean, na.rm=TRUE))
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>% filter(year != '2022')
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != '2022')
 
 # checking time series for MODIS
-ModisNDVI.semi-natural %>%
+ModisNDVI.seminat %>%
   group_by(id, year) %>%
@@
 # 2022 does not stand out as in the Sentinel data, so we keep it
-ModisNDVI.semi-natural <- ModisNDVI.semi-natural %>%
+ModisNDVI.seminat <- ModisNDVI.seminat %>%
   group_by(id, year) %>%
@@
 # checking time series for Landsat
-LandsatNDVI.semi-natural %>%
+LandsatNDVI.seminat %>%
   group_by(id, year) %>%
@@
 # nothing worrying to see here either
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
   group_by(id, year) %>%
@@
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.semi-natural$mean_beta <- (SentinelNDVI.semi-natural$mean + 1) / 2
-ModisNDVI.semi-natural$mean_beta <- (ModisNDVI.semi-natural$mean + 1) / 2
-LandsatNDVI.semi-natural$mean_beta <- (LandsatNDVI.semi-natural$mean + 1) / 2
+#SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
+ModisNDVI.seminat$mean_beta <- (ModisNDVI.seminat$mean + 1) / 2
+LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
 
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.semi-natural$mean_beta)
-summary(ModisNDVI.semi-natural$mean_beta)
-summary(LandsatNDVI.semi-natural$mean_beta)
+summary(SentinelNDVI.seminat$mean_beta)
+summary(ModisNDVI.seminat$mean_beta)
+summary(LandsatNDVI.seminat$mean_beta)
 # replace 1s in Landsat data with 0.9999
-LandsatNDVI.semi-natural <- LandsatNDVI.semi-natural %>%
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
   mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-names(SentinelNDVI.semi-natural)
-names(ModisNDVI.semi-natural)
-names(LandsatNDVI.semi-natural)
+names(SentinelNDVI.seminat)
+names(ModisNDVI.seminat)
+names(LandsatNDVI.seminat)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
 # one column is named slightly differently in the Sentinel data -> rename it
-SentinelNDVI.semi-natural <- SentinelNDVI.semi-natural %>%
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
   rename('system.index' = 'system:index')
@@
 # check if they have the same geometry
-st_crs(SentinelNDVI.semi-natural)
-st_crs(ModisNDVI.semi-natural)
-st_crs(LandsatNDVI.semi-natural)
+st_crs(SentinelNDVI.seminat)
+st_crs(ModisNDVI.seminat)
+st_crs(LandsatNDVI.seminat)
 # all good
@@
 # add an increment to the year variable to avoid overlapping data being hidden in figures
-SentinelNDVI.semi-natural$year_jit <- SentinelNDVI.semi-natural$year + 0.3
-ModisNDVI.semi-natural$year_jit <- ModisNDVI.semi-natural$year - 0.3
-LandsatNDVI.semi-natural$year_jit <- LandsatNDVI.semi-natural$year
+SentinelNDVI.seminat$year_jit <- SentinelNDVI.seminat$year + 0.3
+ModisNDVI.seminat$year_jit <- ModisNDVI.seminat$year - 0.3
+LandsatNDVI.seminat$year_jit <- LandsatNDVI.seminat$year
 
 # concatenate the three Satellite objects
-allSatNDVI.semi-natural <- rbind(
-SentinelNDVI.semi-natural[,!names(SentinelNDVI.semi-natural) %in% "subtype"],
-ModisNDVI.semi-natural,
-LandsatNDVI.semi-natural[,!names(LandsatNDVI.semi-natural) %in% "column_label"]
+allSatNDVI.seminat <- rbind(
+SentinelNDVI.seminat[,!names(SentinelNDVI.seminat) %in% "subtype"],
+ModisNDVI.seminat,
+LandsatNDVI.seminat[,!names(LandsatNDVI.seminat) %in% "column_label"]
 )
 # add variable for Satellite identity
-allSatNDVI.semi-natural$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.semi-natural)),
-  rep("Modis",nrow(ModisNDVI.semi-natural)),
-  rep("Landsat",nrow(LandsatNDVI.semi-natural))
+allSatNDVI.seminat$Sat <- c(
+  rep("Sentinel",nrow(SentinelNDVI.seminat)),
+  rep("Modis",nrow(ModisNDVI.seminat)),
+  rep("Landsat",nrow(LandsatNDVI.seminat))
 )
-allSatNDVI.semi-natural$Sat <- factor(allSatNDVI.semi-natural$Sat,levels=c("Sentinel","Modis","Landsat"))
-levels(allSatNDVI.semi-natural$Sat)
+allSatNDVI.seminat$Sat <- factor(allSatNDVI.seminat$Sat,levels=c("Sentinel","Modis","Landsat"))
+levels(allSatNDVI.seminat$Sat)
 ```
@@
 # plot 
-allSatNDVI.semi-natural %>%
+allSatNDVI.seminat %>%
   ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.semi-natural[SentinelNDVI.semi-natural$tilstand=="God",])
-model.semi-natural.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.semi-natural[ModisNDVI.semi-natural$tilstand=="God",])
-model.semi-natural.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.semi-natural[LandsatNDVI.semi-natural$tilstand=="God",])
+model.semi-natural.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
+model.semi-natural.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
+model.semi-natural.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])
 
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.semi-natural[SentinelNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
-model.semi-natural.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.semi-natural[ModisNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
-model.semi-natural.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.semi-natural[LandsatNDVI.semi-natural$hovedtype %in% c("V9","V10"),])
+model.semi-natural.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("V9","V10"),])
+model.semi-natural.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("V9","V10"),])
+model.semi-natural.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("V9","V10"),])
 summary(model.semi-natural.time.tilst.V9V10.Sent)$coefficients$cond

### 2023-09-15 R/NDVI/1_data_handling.R

@@ -322,7 +322,9 @@ saveRDS(nin, "data/cache/nin.RDS")
 saveRDS(nin.wetland, "data/cache/nin.wetland.RDS")
 saveRDS(nin.wetland2, "data/cache/nin.wetland2.RDS")
 saveRDS(nin.seminat, "data/cache/nin.seminat.RDS")
+saveRDS(nin.seminat2, "data/cache/nin.seminat.RDS")
 saveRDS(nin.natopen, "data/cache/nin.natopen.RDS")
+saveRDS(nin.natopen2, "data/cache/nin.natopen.RDS")
 nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
 nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
 nin.seminat <- readRDS(paste0(here::here(),"/data/cache/nin.seminat.RDS"))

### 2023-09-15 trophic_levels-biomass_indicators.rmd
Update trophic_levels-biomass_indicators.rmd
just a minor formatting issue
Minor changes to fix issues #117, #115, #114, #112 and #111

@@
 
-_Author and date:_
+*Author and date:*
+
 James D. M. Speed, Gunnar Austrheim, Erling J. Solberg
@@
 
-Here we will develop indicators of ecological condition for Norway, based on the distribution of biomass between trophic levels – specifically between vegetation and vertebrate consumers.
+Here we will develop indicators of ecological condition for Norway, based on the distribution of biomass between trophic levels -- specifically between vegetation and vertebrate consumers.
 This includes three trophic levels, plants, herbivores and carnivores.
@@
 2022](https://doi.org/10.1111/1365-2664.14047)) and between vertebrate prey and predators ([Hatton et al 2015](https://doi.org/10.1126/science.aac6284)).
+
+Indicators developed are:
+
+**Deviation from expected vertebrate herbivore biomass** (in total, and for each ecosystem)
+
+**Deviation from expected vertebrate carnivore biomass** (in total, and for each ecosystem)
 
@@
 Municipalities and counties valid in 2007 were used.
-NPP is at a pixel resolution.
+NPP is at a 500 m resolution.
 
@@
 
-The biomass of the trophic levels is as kg km^-2.
+The biomass of the herbivore and carnivore trophic levels is as kg km^-2^. For NPP the original units is kg C m^-2^ year^-1^.
 
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+For herbivores the expected biomass is calculated as:
+
+NPP^0.47^ \* 0.643
+
+For carnivores the expected biomass is calculated as:
+
+Herbivore biomass^0.73^ \* 0.014
+
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
 
@@
 Poor condition as over ±40% of the reference level.
-The choice of 40% is arbitary.
+The choice of 40% is arbitary, and is based on the expert opinion of the authors.
 
@@
 (2015) for further details.
+There are further uncertainties related to the deviance around the equations used to estimate consumer biomass from consumed biomass, as well as the applicability of the global relationships to the Norwegian context.
 
@@
 Wildlife Biology, 17(3), 286-298.
+
 Fløjgaard, C., Pedersen, P.
@@
 Journal of Applied Ecology, 59(1), 18-24.
+
 Hatton, I.
@@
 Science, 349(6252), aac6284.
-Sobocinski, A. (2022). Mapping large carnivores in Norway across 175 years of changing policy (Master's thesis, NTNU). https://hdl.handle.net/11250/3047580 
+
+Sobocinski, A. (2022). Mapping large carnivores in Norway across 175 years of changing policy (Master's thesis, NTNU). <https://hdl.handle.net/11250/3047580>
+
 Speed, J.

### 2023-09-15 DRAFT_NDVI_seminat.Rmd
ndvi seminat analyses, almost done

@@
 nin <- nin %>% mutate(tilstand_num = recode(tilstand,
-                                        "God" = 1, "Moderat" = 2, 
-                                        "Redusert" = 3, "Svaert_redusert" = 4))
+                                        "God" = 0, "Moderat" = 1, 
+                                        "Redusert" = 2, "Svaert_redusert" = 3))
 
@@
 nin.seminat <- nin %>% 
-  filter(hovedoekosystem %in% c('Vaatmark')) %>%
+  filter(hovedoekosystem %in% c('Semi_naturlig')) %>%
   mutate(id = identifikasjon_lokalid) %>%
@@
 
-colnames(nin.seminat)[c(1,9)] <- c("id","region_id")
+colnames(nin.seminat)[c(1,10)] <- c("id","region_id")
 
 # check that every nin-polygon still occurs only once
-summary(as.factor(nin.seminat$id))
-nin.seminat[nin.seminat$id=="NINFP2110008060",]
-# this one was assigned to both Western and Central Norway, we drop the latter
-nin.seminat <- nin.seminat[!row.names(nin.seminat) %in% '1144.1',]
+summary(as.factor(nin.seminat$id)) # ok
 
@@
 #### NDVI data
-```{r, results='hide'}
+```{r, results='hide', message=FALSE}
 ## Sentinel-2
@@
 ```
-semi-natural polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
+semi-natural polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.
 
@@
 
+SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","C9","C10","C11","C12","C13","C14","C15","C16","C17","C18","C19","C20","C21","E1","E2","E3","E4","E5","E6","E7"))
 SentinelNDVI.seminat %>%
@@
   geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype)
+  facet_wrap( ~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2, size=6))
 ```
-NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3).
+NDVI varies between major ecosystem types with the highest values in semi-natural grasslands (T32) and in previously intensively used and altered grasslands that today resemble  a semi-natural grassland (T40, T41), lowest values in coastal heathlands (T34).
+Other ecossytem types: Boreal heath (T31), semi-natural coastal grasslands (T33).
 The overlaps between the ecosystem types are considerable.
-With respect to basic ecosystem types within the major types, NDVI appears to increase with increasing limestone level (C1-C4, C5-C8, and E1-E3 in V1, C1-C3 in the other major types), and also here the overlap is large.
+With respect to basic ecosystem types within the major types, NDVI appears to cover very similar ranges across the subtypes, and those with deviating violin plots have a smaller sample size.
+There is no obvious trend related to the structuring environmental gradients withing the major types, and also here the overlap is large.
 
@@
 ```
-There is no obvious trends in geography.
+For the meadow systems (T32, T33, T40, T41) there is no obvious trends in geography.
+In the heath systems (T31, T34) the distribution of NDVI values in Central Norway is somewhat lower than in the other regions.
 
@@
 ```
-In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern.
+In the coastal types (T33, T34, coastal meadows and heathlands, respectively) NDVI seems to increase as condition deteriorates, but in the remaining types such a pattern is absent.
 Overlaps between condition classes are generally very large.
@@
 nin.seminat2 %>%
-  filter(hovedtype=='V9') %>%
-  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
+  filter(hovedtype=='T31') %>%
+  filter(NiN_variable_code %in% c("7JB-BT","7RA-BA","7FA","7TK")) %>%
   mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "1AG-B"="Shrub cover",
-                                 "7GR-GI"="Abundance of ditches",
-                                 "1AG-A-G"="Tree cover",
-                                 "1AR-C-L"="Woody plants in field layer",
-                                 "7TK"="Heavy vehicles",
-                                 "PRSL"="Wear/tear")) %>%
+                                 "7JB-BT"="Grazing pressure",
+                                 "7RA-BA"="Succession",
+                                 "7FA"="Alien species",
+                                 "7TK"="Heavy vehicles")) %>%
   ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
@@
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in semi-natural mires (V9)") +
+  ggtitle("Condition variables in boreal heath (T31)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
 nin.seminat2 %>%
-  filter(hovedtype=='V10') %>%
+  filter(hovedtype=='T32') %>%
+  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7JB-BA"="Land-use intensity",
+                                 "7RA-SJ"="Succession",
+                                 "7FA"="Alien species",
+                                 "7SE"="Wear/tear",
+                                 "7JB-GJ"="Fertilization")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  theme(legend.position="none") +
+  ggtitle("Condition variables in semi-natural grasslands (T32)") +
+  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+
+nin.seminat2 %>%
+  filter(hovedtype=='T33') %>%
   filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
   mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land use intensity",
+                                 "7JB-BA"="Land-use intensity",
                                  "7RA-SJ"="Succession",
@@
   geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in semi-natural wet-meadows (V10)") +
+  ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
 nin.seminat2 %>%
-  filter(hovedtype=='V1') %>%
-  filter(NiN_variable_code %in% c("7GR-GI","PRTK","7FA","7TK","7SE","PRSL")) %>%
+  filter(hovedtype=='T34') %>%
+  filter(NiN_variable_code %in% c("7JB-BT","7RA-SJ","7FA","7TK")) %>%
   mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7GR-GI"="Abundance of ditches",
-                                 "PRTK"="Heavy vehicles",
+                                 "7JB-BT"="Grazing pressure",
+                                 "7RA-SJ"="Succession",
                                  "7FA"="Alien species",
-                                 "7TK"="Heavy vehicles2",
-                                 "7SE"="Wear/tear",
-                                 "PRSL"="Wear/tear2")) %>%
+                                 "7TK"="Heavy vehicles")) %>%
   ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
@@
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in minerotrophic fens (V1)") +
+  ggtitle("Condition variables in coastal heathlands (T34)") +
+  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+
+nin.seminat2 %>%
+  filter(hovedtype %in% c('T40','T41')) %>%
+  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7JB-BA"="Land-use intensity",
+                                 "7RA-SJ"="Succession",
+                                 "7FA"="Alien species",
+                                 "7SE"="Wear/tear",
+                                 "7JB-GJ"="Fertilization")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  theme(legend.position="none") +
+  ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
 ```
-In these plots, the panels for V1, V9 and V10 differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
+In these plots, the panels for the different major ecosystem types differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
 
-In V9, semi-natural mires, reduced condition is mainly related to increasing cover of shrubs and trees.
-In V10, semi-natural wet-meadows, reduced condition is related to insufficient land-use intensity and succession.
-In both cases, we'd expect an increase in the greenness signal.
-Thus, NDVI can, in principle, be indicative of ecological condition in semi-natural semi-naturals due to a direct connection to the main pressures on the ecosystems.
-In V1 (minerotrophic fens), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
-Under these drivers, greenness would not directly be affected.
-although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run.
-Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.
+In the heath systems - T31 and T34 - reduced condition is mainly related to low levels of grazing and ongoing succession (but also too intensive grazing).
+In the meadow systems - T32, T33, T40, T41 - reduced condition is related to low levels of land-use intensity and ongoing succession  (but also too intensive land use).
+In addition, reduced condition is related to high abundance of alien species in all major semi-natural ecosystem types.
 
@@
 We can investigate further how NDVI and condition are connected using regression analyses.
+
+First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
+```{r warning=FALSE, message=FALSE}
+# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
+SentinelNDVI.seminat %>%
+
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+
+  filter(year == kartleggingsaar) %>%
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_grid( region~hovedtype)
+```
+T41 is has only very few data and is present in 2 regions only, we omit it.
+Northern Norway does not have any T34 occurrences and very few T33 and T40, we won't be able to model it together with the other regions, and might need to model T31 and T32 there only.
 
@@
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
-# we drop V4 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
+# we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
 SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
   filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'V4')
+  filter(hovedtype != 'T41')
 
@@
 
+# check if there's any 0s or 1s (which beta cannot handle)
+summary(SentinelNDVI.seminat.train$mean_beta)
+# replace 1s in Landsat data with 0.9999
+LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+  mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
+
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.semi-natural.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.seminat.train)
-model.semi-natural.cond.Sent <- StepBeta(model.semi-natural.cond.Sent)
-summary(model.semi-natural.cond.Sent)
+model.seminat.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region!='Northern Norway',])
+model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
+summary(model.seminat.cond.Sent)
 ```
@@
 ```{r, include = F}
-rm(model.semi-natural.cond.Sent)
+rm(model.seminat.cond.Sent)
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Northern Norway",])
-model.semi-natural.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Central Norway",])
-model.semi-natural.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Western Norway",])
-model.semi-natural.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Eastern Norway",])
-model.semi-natural.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Southern Norway",])
+model.seminat.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Northern Norway" & SentinelNDVI.seminat.train$hovedtype!='T34',])
+model.seminat.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Central Norway",])
+model.seminat.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Western Norway",])
+model.seminat.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Eastern Norway",])
+model.seminat.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Southern Norway",])
 
@@
 
-summary(model.semi-natural.cond.Sent.N)$coefficients
+summary(model.seminat.cond.Sent.N)$coefficients
 ```
-NDVI in good condition: V10 > V1 > V9 > V3
+NDVI in good condition: T32 > T31/T33 > T40 (last not sig. but very few data)
 
-V1: NDVI decreases as condition deteriorates
-V3: NDVI does not change as condition deteriorates
-V9 & V10: NDVI increases as condition deteriorates
+T31,T33,T40: NDVI increases as condition deteriorates
+T32: NDVI does not change with condition
 
@@
 
-summary(model.semi-natural.cond.Sent.C)$coefficients
+summary(model.seminat.cond.Sent.C)$coefficients
 ```
-NDVI in good condition: V10 > V1 & V9 > V3
+NDVI in good condition: T32 > T33/T34/T40 > T31
 
-V1: NDVI decreases as condition deteriorates
-V3, V9 & V10: NDVI increases as condition deteriorates
+T31: NDVI increases as condition deteriorates
+T32/T33/T34/T40: NDVI increases sig. less than in T31 as condition deteriorates, but it still increases
 
@@
 
-summary(model.semi-natural.cond.Sent.W)$coefficients
+summary(model.seminat.cond.Sent.W)$coefficients
 ```
-NDVI in good condition: V10 > V9 & V3 > V1
+NDVI in good condition: T31/T32/T40 > T33/T34
 
-V1 & V10: NDVI increases as condition deteriorates
-V3: NDVI decreases as condition deteriorates (but really insufficient data)
-V9: NDVI does not change as condition deteriorates
+all types: NDVI increases slightly as condition deteriorates, but just about sig.
 
@@
 
-summary(model.semi-natural.cond.Sent.E)$coefficients
+summary(model.seminat.cond.Sent.E)$coefficients
 ```
-NDVI in good condition: V10 > V1 > V3 & V9
+NDVI in good condition: T32 > T31/T40 > T33 > T34
 
-V1 & V3: NDVI does not change as condition deteriorates
-V9 & V10: NDVI increases as condition deteriorates
+T31/T32/T34/T40: NDVI increases slightly as condition deteriorates
+T33: NDVI increases as condition deteriorates, more strongly than in the other types
 
@@
 
-summary(model.semi-natural.cond.Sent.S)$coefficients
+summary(model.seminat.cond.Sent.S)$coefficients
 ```
-NDVI in good condition: V10 > V9 > V1 & V3
+NDVI in good condition: T32 > T31/T40 > T33/T34
 
-V1, V9 & V10: NDVI increases as condition deteriorates
-V3: NDVI does not change as condition deteriorates (but also insuffcient data)
+T31/T32: borderline NDVI increases as condition deteriorates
+T33/T44: NDVI increases as condition deteriorates
+T40: insufficient power
 
@@
 
-In semi-natural semi-natural ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
-This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural semi-natural ecosystems.
-In natural minerotrophic fens (V1) NDVI increases in the South but decreases in the North as ecosystem condition deteriorates.
-A factor with potential for affecting vegetation composition and productivity, and thus NDVI, is Nitrogen deposition.
-Nitrogen deposition is higher in Southern parts of Norway than further north, and could thus be responsible for increased NDVI with reduced condition in the South, but it is not clear how Nitrogen deposition and the variables underlying the condition assessment in V1 (Ditches, Alien species, Heavy vehicles, Wear/tear) are connected.
-Furthermore we are to date lacking documented empirical ground-truthed information about the connection between NDVI and Nitrogen effects in semi-naturals.
-In bogs (V3) there is no consistent pattern.
+In semi-natural ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
+Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country.
+Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.
 
@@
 They vary both in their placement along the y-axis and in their variance.
-Modis data are absent from the V4-ecosystem type because of a too large pixel size compared to the NiN-polygon sizes.
-In V10, Modis is poorly represented for the same reason.
+Modis data are absent from the T40/T41-ecosystem types because of a too large pixel size compared to the NiN-polygon sizes.
+In the other types, Modis is poorly represented for the same reason.
 
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
-model.semi-natural.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
-model.semi-natural.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])
+model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
+model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
+model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])
 
-summary(model.semi-natural.time.GodTilst.Sent)$coefficients$cond
-summary(model.semi-natural.time.GodTilst.Modi)$coefficients$cond
-summary(model.semi-natural.time.GodTilst.Land)$coefficients$cond
+summary(model.seminat.time.GodTilst.Sent)$coefficients$cond
+summary(model.seminat.time.GodTilst.Modi)$coefficients$cond
+summary(model.seminat.time.GodTilst.Land)$coefficients$cond
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-model.semi-natural.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.semi-natural.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.semi-natural.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("V9","V10"),])
-summary(model.semi-natural.time.tilst.V9V10.Sent)$coefficients$cond
-summary(model.semi-natural.time.tilst.V9V10.Modi)$coefficients$cond
-summary(model.semi-natural.time.tilst.V9V10.Land)$coefficients$cond
+model.seminat.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("V9","V10"),])
+model.seminat.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("V9","V10"),])
+model.seminat.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("V9","V10"),])
+summary(model.seminat.time.tilst.V9V10.Sent)$coefficients$cond
+summary(model.seminat.time.tilst.V9V10.Modi)$coefficients$cond
+summary(model.seminat.time.tilst.V9V10.Land)$coefficients$cond
 ```

### 2023-09-15 DRAFT_NDVI_wetland.Rmd
ndvi seminat analyses, almost done

@@
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurments.
+In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurements.
 We also still explore NDVI data from MODIS (see more details below).
@@
 
-- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets innstruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets instruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
 - remote sensed NDVI data
@@
 
-colnames(nin.wetland)[c(1,9)] <- c("id","region_id")
+colnames(nin.wetland)[c(1,10)] <- c("id","region_id")
 
@@
   geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype)
+  facet_wrap( ~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
 ```
@@
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in semi-natural mires (V9)") +
+  ggtitle("Condition variables in semi-natural mires (V9)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in semi-natural wet-meadows (V10)") +
+  ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
   theme(legend.position="none") +
-  ggtitle("Condition and condition variables in minerotrophic fens (V1)") +
+  ggtitle("Condition variables in minerotrophic fens (V1)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
@@
 We can investigate further how NDVI and condition are connected using regression analyses.
+
+First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
+```{r warning=FALSE, message=FALSE}
+# NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
+SentinelNDVI.wetland %>%
+
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm=TRUE)) %>%
+
+  filter(year == kartleggingsaar) %>%
+  ggplot( aes(x=tilstand, y=mean )) + 
+  geom_violin() +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_grid( region~hovedtype)
+```
+V4 has sparse data and is lacking in some combinations of region and condition level, so we drop it.
 

### 2023-09-15 R/NDVI/1_data_handling.R
cfhf

@@ -322,7 +322,7 @@ saveRDS(nin, "data/cache/nin.RDS")
 saveRDS(nin.wetland, "data/cache/nin.wetland.RDS")
 saveRDS(nin.wetland2, "data/cache/nin.wetland2.RDS")
 saveRDS(nin.seminat, "data/cache/nin.seminat.RDS")
-saveRDS(nin.seminat2, "data/cache/nin.seminat.RDS")
+saveRDS(nin.seminat2, "data/cache/nin.seminat2.RDS")
 saveRDS(nin.natopen, "data/cache/nin.natopen.RDS")
 saveRDS(nin.natopen2, "data/cache/nin.natopen.RDS")
 nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))

### 2023-09-16 DRAFT_NDVI_seminat.Rmd
fixing errors in ndvi wetland and seminat

@@
 regnor <- st_intersection(reg,nor)
+rm(reg)
+rm(nor)
 ```
@@
 # How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
-```{r}
+```{r warning=FALSE}
 SentinelNDVI.seminat %>%
@@
 # How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)
-```{r}
+```{r warning=FALSE}
 SentinelNDVI.seminat %>%
@@
 # How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
-```{r}
+```{r warning=FALSE}
 SentinelNDVI.seminat %>%
@@
 # Which ecological relevant variable(s) is the condition measure in the NiN data related to?
-```{r}
+```{r warning=FALSE}
 nin.seminat2 %>%
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
 
@@
 
-# make an extra numeric condition variable
-unique(SentinelNDVI.seminat.train$tilstand)
-SentinelNDVI.seminat.train <- SentinelNDVI.seminat.train %>% mutate(tilstand_num = recode(tilstand, 
-                                            "God" = '0',
-                                            "Moderat" = '1',
-                                            "Redusert" = '2',
-                                            "Svært redusert" = '3'))
-SentinelNDVI.seminat.train$tilstand_num <- as.numeric(SentinelNDVI.seminat.train$tilstand_num)
-
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.seminat.train$mean_beta)
+summary(SentinelNDVI.seminat.train$mean_beta) #ok
 # replace 1s in Landsat data with 0.9999
@@
 T31/T32: borderline NDVI increases as condition deteriorates
-T33/T44: NDVI increases as condition deteriorates
+T33/T34: NDVI increases as condition deteriorates
 T40: insufficient power
@@
 # Sentinel time series checked in exploratory analysis script
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != '2022')
 
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.seminat$mean_beta)
-summary(ModisNDVI.seminat$mean_beta)
-summary(LandsatNDVI.seminat$mean_beta)
+summary(SentinelNDVI.seminat$mean_beta) # ok
+summary(ModisNDVI.seminat$mean_beta) # ok
+summary(LandsatNDVI.seminat$mean_beta) # has 1's
 # replace 1s in Landsat data with 0.9999

### 2023-09-16 DRAFT_NDVI_wetland.Rmd
fixing errors in ndvi wetland and seminat

@@
   filter(hovedtype != 'V4')
-
-# make an extra numeric condition variable
-unique(SentinelNDVI.wetland.train$tilstand)
-SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
-                                            "God" = '0',
-                                            "Moderat" = '1',
-                                            "Redusert" = '2',
-                                            "Svært redusert" = '3'))
-SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
 
@@
 # Sentinel time series checked in exploratory analysis script
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
 

### 2023-09-17 DRAFT_NDVI_wetland.Rmd
wetland ndvi done

@@
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger in semi-natural wetlands?
+Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
 ```{r message=FALSE, warning=FALSE}
@@
 Since ecological condition also relates to the temporal development of NDVI values in the analyses above, an alternative approach can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
-2022, see references above).
+2022 in references above, see coded example at https://ninanor.github.io/IBECA/ndvi-trend_fjell.html).
 

### 2023-09-18 DRAFT_NDVI_natopen.Rmd
corrections ndvi seminat & wetland, starting ndvi natopen

--- 2023-09-16 DRAFT_NDVI_seminat.Rmd
@@
-# NDVI (greenness) indicator, semi-natural ecosystems  {#NDVI-indicator-seminat}
+# NDVI (greenness) indicator, naturally open ecosystems below the treeline  {#NDVI-indicator-natopen}
 
@@
 ```{r, echo=FALSE}
-Ecosystem <- "våtmark"
+Ecosystem <- "naturlig åpen"
 Egenskap  <- c("Primærproduksjon")
@@
 national maps for the forest and mountain ecosystems).
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems.
-For this task, two initial challenges can be identified early on:
-1) We don't yet have a national map for semi-natural ecosystems, i.e.
-we don't know the exact locations of all semi-natural areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single semi-natural ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
 
-In this draft development for a NDVI indicator for ecological condition in semi-natural  systems we approach these challenges in the following ways:
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing naturally open ecosystems below the treeline (called 'natopen' hereafter).
+For this task, three initial challenges can be identified early on:
+1) We don't yet have a national map for natopen ecosystems, i.e.
+we don't know the exact locations of all natopen areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
+2) The spatial extent of the single natopen ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+3) Natopen ecosystems in Norway are widely different in their physiognomy, ecology, and thus vegetation, spanning from highly disturbed hill slopes with sparse ruderal vegetation to nutrient rich shallow soils with resource demanding plant species.
+Thus, large variation in NDVI signals from different natopen ecosystem types has to be expected.
+
+In this draft development for a NDVI indicator for ecological condition in natopen  systems we approach these challenges in the following ways:
 1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
-Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of semi-natural ecosystems are available.
-2) We include Sentinel-2 imagery for semi-natural ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all semi-natural polygons in the NiN-data.
+Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of natopen ecosystems are available.
+2) We include Sentinel-2 imagery for natopen ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all natopen polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
@@
 ## About the underlying data
-In the NDVI project for semi-natural systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
+In the NDVI project for natopen systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
 
@@
 Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
+#### check here #### 
 Polygon sizes in these NiN data may vary from <100 sqm to 1 sqkm.
@@
 ### Representativity in time and space
-For semi-natural systems, the NiN data contain 6502 semi-natural polygons of known area size across mainland Norway.
+#### check here ####
+For natopen systems, the NiN data contain 6502 natopen polygons of known area size across mainland Norway.
 However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 6502 semi-natural polygons are distributed across regions and major ecosystem types in the following way:
+The 6502 natopen polygons are distributed across regions and major ecosystem types in the following way:
 
@@
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the semi-natural occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the natopen occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
 
 ### Temporal coverage
-The currently available NiN data on semi-natural ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
+The currently available NiN data on natopen ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
 Note that none of the polygons in the NiN-data has been visited more than once during that period.
@@
 ## Collinearities with other indicators
-The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in abandoned semi-natural areas under shrub and tree encroachment.
+Unlike wetlands and semi.natural systems, natopen areas are not subject to shrub and tree encroachment due to changes in land-use, Thus, we do not expect any co-linearity of the NDVI signal with e.g.
+the re-growth indicator based on LiDAr derived canopy height.
 
@@
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
-Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for semi-natural ecosystems (cf.
+Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for natopen ecosystems (cf.
 Jakobsson et al.
 2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to semi-natural ecology and condition and (2) its applicability for defining a reference state.
+In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to natopen ecology and condition and (2) its applicability for defining a reference state.
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
@@
 
-## filter out only semi-natural data
-nin.seminat <- nin %>% 
-  filter(hovedoekosystem %in% c('Semi_naturlig')) %>%
+## filter out only natopen data
+nin.natopen <- nin %>% 
+  filter(hovedoekosystem %in% c('Naturlig_aapent')) %>%
   mutate(id = identifikasjon_lokalid) %>%
@@
 # merge NiN-data with region
-nin.seminat = st_join(nin.seminat, regnor, left = TRUE)
-nin.seminat
+nin.natopen = st_join(nin.natopen, regnor, left = TRUE)
+nin.natopen
 
-colnames(nin.seminat)[c(1,10)] <- c("id","region_id")
+colnames(nin.natopen)[c(1,10)] <- c("id","region_id")
 
 # check that every nin-polygon still occurs only once
-summary(as.factor(nin.seminat$id)) # ok
+summary(as.factor(nin.natopen$id)) # ok
 
-nin.seminat <- nin.seminat %>% 
-  mutate(area_meters_nin = st_area(nin.seminat)
+nin.natopen <- nin.natopen %>% 
+  mutate(area_meters_nin = st_area(nin.natopen)
   )
 # check and edit the order of regions
-levels(nin.seminat$region)
-nin.seminat$region <- as.factor(nin.seminat$region)
-levels(nin.seminat$region)
-nin.seminat$region <- factor(nin.seminat$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
-levels(nin.seminat$region)
+levels(nin.natopen$region)
+nin.natopen$region <- as.factor(nin.natopen$region)
+levels(nin.natopen$region)
+nin.natopen$region <- factor(nin.natopen$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+levels(nin.natopen$region)
 
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.seminat2 <- nin.seminat %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
+nin.natopen2 <- nin.natopen %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
   separate(col = ninbeskrivelsesvariabler,
@@
   subset(select = -ninbeskrivelsesvariabler)
-nin.seminat2 <- as.data.frame(nin.seminat2)
+nin.natopen2 <- as.data.frame(nin.natopen2)
 
@@
 #load processed NiN-data from cache
-nin.seminat <- readRDS(paste0(here::here(),"/data/cache/nin.seminat.RDS"))
-nin.seminat2 <- readRDS(paste0(here::here(),"/data/cache/nin.seminat2.RDS"))
+nin.natopen <- readRDS(paste0(here::here(),"/data/cache/nin.natopen.RDS"))
+nin.natopen2 <- readRDS(paste0(here::here(),"/data/cache/nin.natopen2.RDS"))
 
@@
 ## Sentinel-2
-# join nin.seminat & Sentinel NDVI data
-SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by="id")
-#summary(SentinelNDVI.seminat)
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
+# join nin.natopen & Sentinel NDVI data
+SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by="id")
+#summary(SentinelNDVI.natopen)
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.seminat))
-#summary(SentinelNDVI.seminat)
-# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(hovedtype))
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(mean))
-#summary(SentinelNDVI.seminat)
+         area_meters = st_area(SentinelNDVI.natopen))
+#summary(SentinelNDVI.natopen)
+# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(hovedtype))
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
+#summary(SentinelNDVI.natopen)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-dim(SentinelNDVI.seminat)
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(as.numeric(area_meters) >= 100)
-dim(SentinelNDVI.seminat)
+dim(SentinelNDVI.natopen)
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
+dim(SentinelNDVI.natopen)
 # split date into year, month & day
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # add column for sub-ecosystem types
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
                                                         subtype = str_remove(subtype, '-'),
@@
 # we are using max NDVI per year in every NiN polygon
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
   group_by(id, year) %>%
@@
 
-summary(SentinelNDVI.seminat)
+summary(SentinelNDVI.natopen)
 
 ## MODIS
-# join nin.seminat and Modis NDVI data
-ModisNDVI.seminat <- full_join(nin.seminat, df.m, by="id")
+# join nin.natopen and Modis NDVI data
+ModisNDVI.natopen <- full_join(nin.natopen, df.m, by="id")
 
-#summary(ModisNDVI.seminat)
-ModisNDVI.seminat <- ModisNDVI.seminat %>%
+#summary(ModisNDVI.natopen)
+ModisNDVI.natopen <- ModisNDVI.natopen %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.seminat))
-#summary(ModisNDVI.seminat)
-# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(!is.na(hovedtype))
-ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(!is.na(mean))
-#summary(ModisNDVI.seminat)
+         area_meters = st_area(ModisNDVI.natopen))
+#summary(ModisNDVI.natopen)
+# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
+ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(hovedtype))
+ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
+#summary(ModisNDVI.natopen)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-dim(ModisNDVI.seminat)
-ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(as.numeric(area_meters) >= 62500)
-dim(ModisNDVI.seminat)
+dim(ModisNDVI.natopen)
+ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
+dim(ModisNDVI.natopen)
 # split date into year, month & day
-ModisNDVI.seminat <- ModisNDVI.seminat %>%
+ModisNDVI.natopen <- ModisNDVI.natopen %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-ModisNDVI.seminat <- ModisNDVI.seminat %>%
+ModisNDVI.natopen <- ModisNDVI.natopen %>%
   group_by(id, year) %>%
   filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.seminat)
+#summary(ModisNDVI.natopen)
 
 ## Landsat
-# join nin.seminat & Landsat NDVI data
-LandsatNDVI.seminat <- full_join(nin.seminat, df.l, by="id")
+# join nin.natopen & Landsat NDVI data
+LandsatNDVI.natopen <- full_join(nin.natopen, df.l, by="id")
 
-#summary(LandsatNDVI.seminat)
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+#summary(LandsatNDVI.natopen)
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   mutate(hovedoekosystem = as.factor(hovedoekosystem),
@@
          tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.seminat))
-#summary(LandsatNDVI.seminat)
-# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(!is.na(hovedtype))
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(!is.na(mean))
-#summary(LandsatNDVI.seminat)
+         area_meters = st_area(LandsatNDVI.natopen))
+#summary(LandsatNDVI.natopen)
+# get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(hovedtype))
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
+#summary(LandsatNDVI.natopen)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-dim(LandsatNDVI.seminat)
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(as.numeric(area_meters) >= 900)
-dim(LandsatNDVI.seminat)
+dim(LandsatNDVI.natopen)
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
+dim(LandsatNDVI.natopen)
 # split date into year, month & day
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   dplyr::mutate(year = lubridate::year(date), 
@@
 # we are using max NDVI per year in every NiN polygon
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   group_by(id, year) %>%
@@
 
-#summary(LandsatNDVI.seminat)
+#summary(LandsatNDVI.natopen)
 ```
@@
 ```{r}
-summary(SentinelNDVI.seminat$area_meters)
-hist(SentinelNDVI.seminat$area_meters,xlim=c(0,20000),breaks=50000)
+summary(SentinelNDVI.natopen$area_meters)
+hist(SentinelNDVI.natopen$area_meters,xlim=c(0,20000),breaks=50000)
 abline(v=100,lty=2)
 ```
-semi-natural polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.
+natopen polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.
 
@@
 ```{r}
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen %>%
   ggplot( aes(x=year, y=mean )) + 
@@
 ```{r}
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != '2022')
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(year != '2022')
 ```
@@
 ```{r warning=FALSE}
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen %>%
   filter(tilstand == 'God') %>%
@@
 
-SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","C9","C10","C11","C12","C13","C14","C15","C16","C17","C18","C19","C20","C21","E1","E2","E3","E4","E5","E6","E7"))
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","C9","C10","C11","C12","C13","C14","C15","C16","C17","C18","C19","C20","C21","E1","E2","E3","E4","E5","E6","E7"))
+SentinelNDVI.natopen %>%
   filter(tilstand == 'God') %>%
@@
 ```
-NDVI varies between major ecosystem types with the highest values in semi-natural grasslands (T32) and in previously intensively used and altered grasslands that today resemble  a semi-natural grassland (T40, T41), lowest values in coastal heathlands (T34).
-Other ecossytem types: Boreal heath (T31), semi-natural coastal grasslands (T33).
+NDVI varies between major ecosystem types with the highest values in natopen grasslands (T32) and in previously intensively used and altered grasslands that today resemble  a natopen grassland (T40, T41), lowest values in coastal heathlands (T34).
+Other ecossytem types: Boreal heath (T31), natopen coastal grasslands (T33).
 The overlaps between the ecosystem types are considerable.
@@
 
-# How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)
+# How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)
 ```{r warning=FALSE}
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen %>%
   filter(tilstand == 'God') %>%
@@
 ```{r warning=FALSE}
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen %>%
   filter(year == kartleggingsaar) %>%
@@
 ```{r warning=FALSE}
-nin.seminat2 %>%
+nin.natopen2 %>%
   filter(hovedtype=='T31') %>%
@@
 
-nin.seminat2 %>%
+nin.natopen2 %>%
   filter(hovedtype=='T32') %>%
@@
   theme(legend.position="none") +
-  ggtitle("Condition variables in semi-natural grasslands (T32)") +
+  ggtitle("Condition variables in natopen grasslands (T32)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
 
-nin.seminat2 %>%
+nin.natopen2 %>%
   filter(hovedtype=='T33') %>%
@@
   theme(legend.position="none") +
-  ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
+  ggtitle("Condition variables in natopen coastal grasslands (T33)") +
   labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
 
-nin.seminat2 %>%
+nin.natopen2 %>%
   filter(hovedtype=='T34') %>%
@@
 
-nin.seminat2 %>%
+nin.natopen2 %>%
   filter(hovedtype %in% c('T40','T41')) %>%
@@
 In the meadow systems - T32, T33, T40, T41 - reduced condition is related to low levels of land-use intensity and ongoing succession  (but also too intensive land use).
-In addition, reduced condition is related to high abundance of alien species in all major semi-natural ecosystem types.
+In addition, reduced condition is related to high abundance of alien species in all major natopen ecosystem types.
 
@@
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
-SentinelNDVI.seminat %>%
+SentinelNDVI.natopen %>%
 
@@
 ```{r}
-SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
+SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
 # NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
 # we drop T41 for the analysis as it lacks data for most combinations of condition and region, and thus would cause convergence issues
-SentinelNDVI.seminat.train <- SentinelNDVI.seminat %>%
+SentinelNDVI.natopen.train <- SentinelNDVI.natopen %>%
   filter(year == kartleggingsaar) %>%
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.seminat.train$mean_beta) #ok
+summary(SentinelNDVI.natopen.train$mean_beta) #ok
 # replace 1s in Landsat data with 0.9999
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
@@
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.seminat.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region!='Northern Norway',])
-model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
-summary(model.seminat.cond.Sent)
+model.natopen.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region!='Northern Norway',])
+model.natopen.cond.Sent <- StepBeta(model.natopen.cond.Sent)
+summary(model.natopen.cond.Sent)
 ```
@@
 ```{r, include = F}
-rm(model.seminat.cond.Sent)
+rm(model.natopen.cond.Sent)
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-model.seminat.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Northern Norway" & SentinelNDVI.seminat.train$hovedtype!='T34',])
-model.seminat.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Central Norway",])
-model.seminat.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Western Norway",])
-model.seminat.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Eastern Norway",])
-model.seminat.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Southern Norway",])
+model.natopen.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Northern Norway" & SentinelNDVI.natopen.train$hovedtype!='T34',])
+model.natopen.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Central Norway",])
+model.natopen.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Western Norway",])
+model.natopen.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Eastern Norway",])
+model.natopen.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Southern Norway",])
 
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.seminat.train %>%
+SentinelNDVI.natopen.train %>%
   filter(region=="Northern Norway") %>%
@@
 
-summary(model.seminat.cond.Sent.N)$coefficients
+summary(model.natopen.cond.Sent.N)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.seminat.train %>%
+SentinelNDVI.natopen.train %>%
   filter(region=="Central Norway") %>%
@@
 
-summary(model.seminat.cond.Sent.C)$coefficients
+summary(model.natopen.cond.Sent.C)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.seminat.train %>%
+SentinelNDVI.natopen.train %>%
   filter(region=="Western Norway") %>%
@@
 
-summary(model.seminat.cond.Sent.W)$coefficients
+summary(model.natopen.cond.Sent.W)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.seminat.train %>%
+SentinelNDVI.natopen.train %>%
   filter(region=="Eastern Norway") %>%
@@
 
-summary(model.seminat.cond.Sent.E)$coefficients
+summary(model.natopen.cond.Sent.E)$coefficients
 ```
@@
 ```{r message=FALSE, warning=FALSE}
-SentinelNDVI.seminat.train %>%
+SentinelNDVI.natopen.train %>%
   filter(region=="Southern Norway") %>%
@@
 
-summary(model.seminat.cond.Sent.S)$coefficients
+summary(model.natopen.cond.Sent.S)$coefficients
 ```
@@
 
-In semi-natural ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
+In natopen ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
 Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country.
-Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.
+natopen and natopen-looking meadows show very tiny increases if any at all.
 
@@
 # checking time series for MODIS
-ModisNDVI.seminat %>%
+ModisNDVI.natopen %>%
   group_by(id, year) %>%
@@
 # 2022 does not stand out as in the Sentinel data, so we keep it
-ModisNDVI.seminat <- ModisNDVI.seminat %>%
+ModisNDVI.natopen <- ModisNDVI.natopen %>%
   group_by(id, year) %>%
@@
 # checking time series for Landsat
-LandsatNDVI.seminat %>%
+LandsatNDVI.natopen %>%
   group_by(id, year) %>%
@@
 # nothing worrying to see here either
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   group_by(id, year) %>%
@@
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
-ModisNDVI.seminat$mean_beta <- (ModisNDVI.seminat$mean + 1) / 2
-LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
+#SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
+ModisNDVI.natopen$mean_beta <- (ModisNDVI.natopen$mean + 1) / 2
+LandsatNDVI.natopen$mean_beta <- (LandsatNDVI.natopen$mean + 1) / 2
 
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.seminat$mean_beta) # ok
-summary(ModisNDVI.seminat$mean_beta) # ok
-summary(LandsatNDVI.seminat$mean_beta) # has 1's
+summary(SentinelNDVI.natopen$mean_beta) # ok
+summary(ModisNDVI.natopen$mean_beta) # ok
+summary(LandsatNDVI.natopen$mean_beta) # has 1's
 # replace 1s in Landsat data with 0.9999
-LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
+LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
   mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-names(SentinelNDVI.seminat)
-names(ModisNDVI.seminat)
-names(LandsatNDVI.seminat)
+names(SentinelNDVI.natopen)
+names(ModisNDVI.natopen)
+names(LandsatNDVI.natopen)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
 # one column is named slightly differently in the Sentinel data -> rename it
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
   rename('system.index' = 'system:index')
@@
 # check if they have the same geometry
-st_crs(SentinelNDVI.seminat)
-st_crs(ModisNDVI.seminat)
-st_crs(LandsatNDVI.seminat)
+st_crs(SentinelNDVI.natopen)
+st_crs(ModisNDVI.natopen)
+st_crs(LandsatNDVI.natopen)
 # all good
@@
 # add an increment to the year variable to avoid overlapping data being hidden in figures
-SentinelNDVI.seminat$year_jit <- SentinelNDVI.seminat$year + 0.3
-ModisNDVI.seminat$year_jit <- ModisNDVI.seminat$year - 0.3
-LandsatNDVI.seminat$year_jit <- LandsatNDVI.seminat$year
+SentinelNDVI.natopen$year_jit <- SentinelNDVI.natopen$year + 0.3
+ModisNDVI.natopen$year_jit <- ModisNDVI.natopen$year - 0.3
+LandsatNDVI.natopen$year_jit <- LandsatNDVI.natopen$year
 
 # concatenate the three Satellite objects
-allSatNDVI.seminat <- rbind(
-SentinelNDVI.seminat[,!names(SentinelNDVI.seminat) %in% "subtype"],
-ModisNDVI.seminat,
-LandsatNDVI.seminat[,!names(LandsatNDVI.seminat) %in% "column_label"]
+allSatNDVI.natopen <- rbind(
+SentinelNDVI.natopen[,!names(SentinelNDVI.natopen) %in% "subtype"],
+ModisNDVI.natopen,
+LandsatNDVI.natopen[,!names(LandsatNDVI.natopen) %in% "column_label"]
 )
 # add variable for Satellite identity
-allSatNDVI.seminat$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.seminat)),
-  rep("Modis",nrow(ModisNDVI.seminat)),
-  rep("Landsat",nrow(LandsatNDVI.seminat))
+allSatNDVI.natopen$Sat <- c(
+  rep("Sentinel",nrow(SentinelNDVI.natopen)),
+  rep("Modis",nrow(ModisNDVI.natopen)),
+  rep("Landsat",nrow(LandsatNDVI.natopen))
 )
-allSatNDVI.seminat$Sat <- factor(allSatNDVI.seminat$Sat,levels=c("Sentinel","Modis","Landsat"))
-levels(allSatNDVI.seminat$Sat)
+allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat,levels=c("Sentinel","Modis","Landsat"))
+levels(allSatNDVI.natopen$Sat)
 ```
@@
 # plot 
-allSatNDVI.seminat %>%
+allSatNDVI.natopen %>%
   ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
@@
 ```{r message=FALSE, warning=FALSE}
-model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
-model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
-model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])
+model.natopen.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$tilstand=="God",])
+model.natopen.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.natopen[ModisNDVI.natopen$tilstand=="God",])
+model.natopen.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$tilstand=="God",])
 
-summary(model.seminat.time.GodTilst.Sent)$coefficients$cond
-summary(model.seminat.time.GodTilst.Modi)$coefficients$cond
-summary(model.seminat.time.GodTilst.Land)$coefficients$cond
+summary(model.natopen.time.GodTilst.Sent)$coefficients$cond
+summary(model.natopen.time.GodTilst.Modi)$coefficients$cond
+summary(model.natopen.time.GodTilst.Land)$coefficients$cond
 ```
-There's an increase in NDVI over time in all three satellite time series in essentially all major semi-natural ecosystem types in good condition.
+There's an increase in NDVI over time in all three satellite time series in essentially all major natopen ecosystem types in good condition.
 It is worth pointing out though that the model on the MODIS data shows 1/10 of the effect size for the time slope and has way less statistical power compared to Sentinel and Landsat.
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural semi-naturals, We may ask the question if the temporal NDVI increase might be stronger in semi-natural semi-naturals?
+Given the systematically higher NDVI-values with decreasing condition in natopen natopens, We may ask the question if the temporal NDVI increase might be stronger in natopen natopens?
 ```{r message=FALSE, warning=FALSE}
-model.seminat.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.seminat.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.seminat.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("V9","V10"),])
-summary(model.seminat.time.tilst.V9V10.Sent)$coefficients$cond
-summary(model.seminat.time.tilst.V9V10.Modi)$coefficients$cond
-summary(model.seminat.time.tilst.V9V10.Land)$coefficients$cond
+model.natopen.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("V9","V10"),])
+model.natopen.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.natopen[ModisNDVI.natopen$hovedtype %in% c("V9","V10"),])
+model.natopen.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("V9","V10"),])
+summary(model.natopen.time.tilst.V9V10.Sent)$coefficients$cond
+summary(model.natopen.time.tilst.V9V10.Modi)$coefficients$cond
+summary(model.natopen.time.tilst.V9V10.Land)$coefficients$cond
 ```
-Indeed, the temporal NDVI increase has been stronger in semi-natural semi-naturals with reduced ecological condition, in both Sentinel and Landsat.
+Indeed, the temporal NDVI increase has been stronger in natopen natopens with reduced ecological condition, in both Sentinel and Landsat.
 For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.
@@
 # Overall conclusion
-NDVI in semi-natural ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
-Especially the semi-natural semi-natural ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
+NDVI in natopen ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
+Especially the natopen natopen ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
 The large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
@@
 
-# 1) We are lacking ecosystem maps for semi-natural ecosystems
-The study at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
-Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within semi-naturals separately.
-# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-naturals.
+# 1) We are lacking ecosystem maps for natopen ecosystems
+The study at hand clearly shows that different natopen ecosystem types display different greenness levels and thus need to be evaluated separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within natopens separately.
+# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in natopen ecosystems
+Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in natopens.
 Here, we need field studies investigating:
@@
 
-- How reflectance in different semi-natural types is affected by how moist the moss layer is at the time of measuring.
+- How reflectance in different natopen types is affected by how moist the moss layer is at the time of measuring.
 The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
-Minerotrophic semi-naturals (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, semi-naturals are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+Minerotrophic natopens (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
+At the same time, natopens are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
 

### 2023-09-18 DRAFT_NDVI_wetland.Rmd
corrections ndvi seminat & wetland, starting ndvi natopen

@@
 national maps for the forest and mountain ecosystems).
+
 Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
 excluding forest swamps).
-For this task, two initial challenges can be identified early on:
+For this task, three initial challenges can be identified early on:
 1) We don't yet have a national wetland map, i.e.
@@
 2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+3) NDVI in wetland systems lacks the thorough ground-truthing known from globally large ecosystems as forests, tundra or Savannah.
+Normal variation, like stochastic rain events and moss moisture status may influence NDVI signals greatly (Valøen, 2019).
+Our current understanding of what NDVI and its changes in different wetland ecosystems mean is thus limited.
 
@@
 Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
-Polygon sizes in these NiN data may vary from <100 sqm to 1 sqkm.
+Polygon sizes in these NiN data may vary from <100 sqm to the dimension of sqkm's.
 
@@
 Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967. Norwegian Institute for Nature Research.
+
+Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. Master Thesis NTNU.
 

### 2023-09-18 DRAFT_NDVI_seminat.Rmd
ndvi wetland & seminat done

@@
 ```{r, echo=FALSE}
-Ecosystem <- "våtmark"
+Ecosystem <- "semi-naturlig"
 Egenskap  <- c("Primærproduksjon")
@@
 national maps for the forest and mountain ecosystems).
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems.
-For this task, two initial challenges can be identified early on:
+
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems. For this task, two initial challenges can be identified early on:
 1) We don't yet have a national map for semi-natural ecosystems, i.e.
@@
 
-                  V1    V3    V4    V6    V9    V10
-Northern Norway   638   275   55    102   360   315
-Central Norway    702   1250  34    0     302   116
-Western Norway    296   111   30    0     460   158
-Eastern Norway    251   89    103   0     251   223
-Southern Norway   58    44    5     0     155   102
+                  T31   T32   T33   T34   T40   T41
+Northern Norway   1194  2081  14    1     15    0
+Central Norway    641   2518  210   2338  97    0
+Western Norway    132   2922  94    2200  66    54
+Eastern Norway    1064  4488  10    10    371   54
+Southern Norway   234   2317  112   749   111   140
 
-In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region.
+In addition, there is 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40 that are not assigned to any region.
 
@@
 ```
-There's an increase in NDVI over time in all three satellite time series in essentially all major semi-natural ecosystem types in good condition.
-It is worth pointing out though that the model on the MODIS data shows 1/10 of the effect size for the time slope and has way less statistical power compared to Sentinel and Landsat.
+There's an increase in NDVI over time in mainly the Sentinel and Landsat time series in boreal heath in good condition, but this pattern is reduced or absent in the other semi-natural ecosystem types.
+MODIS does not show any trends in time but also has way less statistical power compared to Sentinel and Landsat.
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural semi-naturals, We may ask the question if the temporal NDVI increase might be stronger in semi-natural semi-naturals?
+#### check here ####
+Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
 ```{r message=FALSE, warning=FALSE}
-model.seminat.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.seminat.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("V9","V10"),])
-model.seminat.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("V9","V10"),])
-summary(model.seminat.time.tilst.V9V10.Sent)$coefficients$cond
-summary(model.seminat.time.tilst.V9V10.Modi)$coefficients$cond
-summary(model.seminat.time.tilst.V9V10.Land)$coefficients$cond
+model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"),])
+model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"),])
+model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"),])
+summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
+summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
+summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
 ```
-Indeed, the temporal NDVI increase has been stronger in semi-natural semi-naturals with reduced ecological condition, in both Sentinel and Landsat.
-For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.
+The temporal NDVI increase in boreal heath has NOT been stronger with reduced ecological condition. So a temporal NDVI approach does not seem like a promising way to go.
+We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
+```{r message=FALSE, warning=FALSE}
+model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"),])
+model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"),])
+model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"),])
+summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
+summary(model.seminat.time.tilst.T32.Modi)$coefficients$cond
+summary(model.seminat.time.tilst.T32.Land)$coefficients$cond
+
+model.seminat.time.tilst.T34.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T34"),])
+model.seminat.time.tilst.T34.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T34"),])
+model.seminat.time.tilst.T34.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T34"),])
+summary(model.seminat.time.tilst.T34.Sent)$coefficients$cond
+summary(model.seminat.time.tilst.T34.Modi)$coefficients$cond
+summary(model.seminat.time.tilst.T34.Land)$coefficients$cond
+```
+Again, patterns are only identifiable with Sentinel and Landsat data, but not with Modis.
+Semi-natural grasslands (T32) do indeed show a stronger increase in NDVI as condition deteriorates, so here a temporal approach may be useful under a reduced land-management/shrub-/tree-encroachment scenario.
+Coastal heathlands (T34), on the other hand, do not show stronger temporal increases with reduced condition.
+Coastal heathlands are subject to a management cycle, which includes developmental and successional changes over a course of 15-25 years followed by prescribed burning which resets that process.
+So, we have to expect increases in greenness in coastal heathlands over time also in well-managed areas, and thus temporal changes in NDVI are largely irrelevant for assessing ecological condition.
 
 # Overall conclusion
-NDVI in semi-natural ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
-Especially the semi-natural semi-natural ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
-The large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
-Since ecological condition also relates to the temporal development of NDVI values in the analyses above, an alternative approach can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
-2022, see references above).
+NDVI in semi-natural ecosystem does vary with condition and also with ecosystem type, but the variance and thus overlap between condition classes and ecosystem types is large.
+Especially in boreal heath and semi-natural grasslands greenness and its change over time appaer to be indicative of the ecological condition of local ecoystems, linking well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
+However, the large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
+Since ecological condition also relates to the temporal development of NDVI values in semi-natural grasslands, a potential way way to approach this can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
+2022 in references above, see coded example at https://ninanor.github.io/IBECA/ndvi-trend_fjell.html).
 
@@
 The study at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
-Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within semi-naturals separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within semi-natural systems separately.
 # 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-naturals.
-Here, we need field studies investigating:
-- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition
-
-- How reflectance in different semi-natural types is affected by how moist the moss layer is at the time of measuring.
-The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
-Minerotrophic semi-naturals (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, semi-naturals are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems.
+Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like for example shrub-/tree encroachment, are connected.
 

### 2023-09-18 DRAFT_NDVI_wetland.Rmd
ndvi wetland & seminat done

@@
 The study at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
-Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within wetlands separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
 # 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems

### 2023-09-18 R/NDVI/1_data_handling.R

@@ -314,6 +314,8 @@ levels(nin.wetland$region)
 
 # check # of wetland types in different regions
 nin.wetland %>% group_by(hovedtype, region) %>% summarise(n = n()) %>% print(n=30)
+nin.seminat %>% group_by(hovedtype, region) %>% summarise(n = n()) %>% print(n=30)
+nin.natopen %>% group_by(hovedtype, region) %>% summarise(n = n()) %>% print(n=30)
 
 
 

### 2023-09-18 DRAFT_NDVI_natopen.Rmd
exploratory analyses ndvi natopen, correction ndvi seminat

@@
 Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
-#### check here #### 
-Polygon sizes in these NiN data may vary from <100 sqm to 1 sqkm.
+Polygon sizes in these NiN data may vary from <100 sqm to 240 000 sqm.
 
@@
 #### check here ####
-For natopen systems, the NiN data contain 6502 natopen polygons of known area size across mainland Norway.
+For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway.
 However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 6502 natopen polygons are distributed across regions and major ecosystem types in the following way:
+The 2881 natopen polygons are distributed across regions and major ecosystem types in the following way:
 
-                  V1    V3    V4    V6    V9    V10
-Northern Norway   638   275   55    102   360   315
-Central Norway    702   1250  34    0     302   116
-Western Norway    296   111   30    0     460   158
-Eastern Norway    251   89    103   0     251   223
-Southern Norway   58    44    5     0     155   102
+                  T2    T12   T18   T20   T21
+Northern Norway   0     230   648   10    31
+Central Norway    41    282   201   0     0
+Western Norway    49    44    79    0     9
+Eastern Norway    217   84    483   2     51
+Southern Norway   52    155   21    0     73
 
-In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region.
+In addition, there is 54 T2s, 57 T12s, 3 T18s, and 5 T21s that are not assigned to any region.
 
@@
 All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
-
+```{r, echo = F}
+# downloading the unprocessed NDVI data from a data-cache et the NINA server
+df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
+df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
+df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
+```
 ##### Sentinel-2 NDVI data
@@
 To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
-```{r}
+```{r, echo = T, eval = F}
 # Import Sentinel NDVI Data
@@
 
-```{r}
+```{r, echo = T, eval = F}
 # Import MODIS NDVI Data
@@
 To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
-```{r}
+```{r, echo = T, eval = F}
 # Import Landsat NDVI Data
@@
 ```
-natopen polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.
+natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.
 
@@
 
-SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","C9","C10","C11","C12","C13","C14","C15","C16","C17","C18","C19","C20","C21","E1","E2","E3","E4","E5","E6","E7"))
+SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","E1"))
 SentinelNDVI.natopen %>%
@@
 ```
-NDVI varies between major ecosystem types with the highest values in natopen grasslands (T32) and in previously intensively used and altered grasslands that today resemble  a natopen grassland (T40, T41), lowest values in coastal heathlands (T34).
-Other ecossytem types: Boreal heath (T31), natopen coastal grasslands (T33).
+NDVI varies between major ecosystem types with the highest values in open flood plains (T18) and in natural coastal meadows (T12), followed by naturally open areas on shallow soils (T2) and vegetated sand dunes (T21).
+There's very little data on T20 (isinnfrysingsmark) and all mapped areas of that type are in good condition.
+Therefore we will omit this ecosystem type from the statistical analyses further below.
 The overlaps between the ecosystem types are considerable.
-With respect to basic ecosystem types within the major types, NDVI appears to cover very similar ranges across the subtypes, and those with deviating violin plots have a smaller sample size.
+With respect to basic ecosystem types within the major types, NDVI appears to cover very similar ranges across the subtypes, but sample size are also rather small.
 There is no obvious trend related to the structuring environmental gradients withing the major types, and also here the overlap is large.
@@
 ```
-For the meadow systems (T32, T33, T40, T41) there is no obvious trends in geography.
-In the heath systems (T31, T34) the distribution of NDVI values in Central Norway is somewhat lower than in the other regions.
+There is no obvious trends in geography.
 
@@
 ```
-In the coastal types (T33, T34, coastal meadows and heathlands, respectively) NDVI seems to increase as condition deteriorates, but in the remaining types such a pattern is absent.
-Overlaps between condition classes are generally very large.
-
+The plots suggest a slight decrease in NDVI as condition deteriorates. Overlaps between condition classes are generally very large.
+#### continue here ####
 # Which ecological relevant variable(s) is the condition measure in the NiN data related to?
@@
 Northern Norway does not have any T34 occurrences and very few T33 and T40, we won't be able to model it together with the other regions, and might need to model T31 and T32 there only.
+
+We remove T20 from the statistical analyses because this ecosystem type is only ever registerd as in good condition, and it appears only in two regions.
+```{r}
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(hovetype != 'T20')
+```
 

### 2023-09-18 DRAFT_NDVI_seminat.Rmd
exploratory analyses ndvi natopen, correction ndvi seminat

@@
 ### Representativity in time and space
-For semi-natural systems, the NiN data contain 6502 semi-natural polygons of known area size across mainland Norway.
+For semi-natural systems, the NiN data contain 24494 semi-natural polygons of known area size across mainland Norway.
 However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 6502 semi-natural polygons are distributed across regions and major ecosystem types in the following way:
+The 24494 semi-natural polygons are distributed across regions and major ecosystem types in the following way:
 
@@
 #### NiN data and geometry for Norway and the five regions
-We read the NiN data for this document from a cache at the NINA server...
-```{r, echo = F}
+We read the NiN data for this document from a cache at the GitHub repository...
+```{r}
 # Add NiN data from cache
@@
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
-```{r message=FALSE}
+```{r, message=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
 
-#### check here ####
 Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?

### 2023-09-18 R/NDVI/1_data_handling.R

@@ -326,7 +326,7 @@ saveRDS(nin.wetland2, "data/cache/nin.wetland2.RDS")
 saveRDS(nin.seminat, "data/cache/nin.seminat.RDS")
 saveRDS(nin.seminat2, "data/cache/nin.seminat2.RDS")
 saveRDS(nin.natopen, "data/cache/nin.natopen.RDS")
-saveRDS(nin.natopen2, "data/cache/nin.natopen.RDS")
+saveRDS(nin.natopen2, "data/cache/nin.natopen2.RDS")
 nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
 nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
 nin.seminat <- readRDS(paste0(here::here(),"/data/cache/nin.seminat.RDS"))

### 2023-09-19 DRAFT_NDVI_natopen.Rmd
ndvi wetland, seminat & natopen finished

@@
 
+# Where are the natopen sites?
+```{r}
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(SentinelNDVI.natopen) +
+  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale=2.5, legend.show = FALSE) + 
+  tm_add_legend(type = "symbol", 
+                col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
+                labels = c("T12 - coastal meadows",
+                           "T18 - flood plains",
+                           "T2 - shallow soils",
+                           "T20 - innfrysningsmark",
+                           "T21 - sand dunes"),
+                title = "Major ecosystem type",
+                size=3) +
+  tm_layout(main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
+            main.title.size=1.2, legend.position = c("right", "bottom"), 
+            legend.text.size = 1.3, legend.title.size = 1.4)
+```
+
 # How are polygon sizes distributed?
@@
 nin.natopen2 %>%
-  filter(hovedtype=='T31') %>%
-  filter(NiN_variable_code %in% c("7JB-BT","7RA-BA","7FA","7TK")) %>%
+  filter(hovedtype=='T2') %>%
+  filter(NiN_variable_code %in% c("7FA","7SE","7TK","1AG-B")) %>%
   mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BT"="Grazing pressure",
-                                 "7RA-BA"="Succession",
                                  "7FA"="Alien species",
+                                 "7SE"="Wear/tear",
+                                 "7TK"="Heavy vehicles",
+                                 "1AG-B"="Shrub cover")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des, ncol=2) +
+  theme(legend.position="none") +
+  ggtitle("Condition variables in open systems on shallow soils (T2)") +
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
+
+nin.natopen2 %>%
+  filter(hovedtype=='T12') %>%
+  filter(NiN_variable_code %in% c("7FA","7SE","7TK")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7FA"="Alien species",
+                                 "7SE"="Wear/tear",
                                  "7TK"="Heavy vehicles")) %>%
@@
   geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  facet_wrap(~NiN_variable_des, ncol=2) +
   theme(legend.position="none") +
-  ggtitle("Condition variables in boreal heath (T31)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  ggtitle("Condition variables in natural coastal meadows (T12)") +
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.natopen2 %>%
-  filter(hovedtype=='T32') %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
+  filter(hovedtype=='T18') %>%
+  filter(NiN_variable_code %in% c("7FA","7JB-BT","7SE","7VR-RI")) %>%
   mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
                                  "7FA"="Alien species",
+                                 "7JB-BT"="Grazing pressure",
                                  "7SE"="Wear/tear",
+                                 "7VR-RI"="Water regulation")) %>%
+  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
+  geom_point(size=2) +  
+  geom_jitter(width = 0.25, height=0.25) +
+  facet_wrap(~NiN_variable_des, ncol=2) +
+  theme(legend.position="none") +
+  ggtitle("Condition variables in flood plains (T18)") +
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
+
+nin.natopen2 %>%
+  filter(hovedtype %in% c('T21')) %>%
+  filter(NiN_variable_code %in% c("7FA","7JB-BT","7SE","7TK","7JB-GJ")) %>%
+  mutate(NiN_variable_des=recode(NiN_variable_code,
+                                 "7FA"="Alien species",
+                                 "7JB-BT"="Grazing pressure",
+                                 "7SE"="Wear/tear",
+                                 "7TK"="Heavy vehicles",
                                  "7JB-GJ"="Fertilization")) %>%
@@
   geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
+  facet_wrap(~NiN_variable_des, ncol=2) +
   theme(legend.position="none") +
-  ggtitle("Condition variables in natopen grasslands (T32)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
-
-nin.natopen2 %>%
-  filter(hovedtype=='T33') %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
-  ggtitle("Condition variables in natopen coastal grasslands (T33)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
-
-nin.natopen2 %>%
-  filter(hovedtype=='T34') %>%
-  filter(NiN_variable_code %in% c("7JB-BT","7RA-SJ","7FA","7TK")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BT"="Grazing pressure",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7TK"="Heavy vehicles")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
-  ggtitle("Condition variables in coastal heathlands (T34)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
-
-nin.natopen2 %>%
-  filter(hovedtype %in% c('T40','T41')) %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7SE"="Wear/tear",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
-  ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  ggtitle("Condition variables in sand dunes (T21)") +
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
 
-In the heath systems - T31 and T34 - reduced condition is mainly related to low levels of grazing and ongoing succession (but also too intensive grazing).
-In the meadow systems - T32, T33, T40, T41 - reduced condition is related to low levels of land-use intensity and ongoing succession  (but also too intensive land use).
-In addition, reduced condition is related to high abundance of alien species in all major natopen ecosystem types.
+In open systems on shallow soils - T2 - reduced condition is mainly related to high levels of alien species, and also to damage from heavy vehicles.
+In the coastal meadows - T12 - reduced condition is related to high levels of alien species, but also to wear/tear and damage from heavy vehicles.
+I flood plains - T18 - reduced condition is mainly related to effects of water regulation.
+In sand dunes - T21 - reduced condition is mainly related to high levels of alien species and wear/tear.
+Wear/tear and damage related issues as well as water regulation might alter the amount of vegetation and thus the NDVI signal.
 
@@
 ```
-T41 is has only very few data and is present in 2 regions only, we omit it.
-Northern Norway does not have any T34 occurrences and very few T33 and T40, we won't be able to model it together with the other regions, and might need to model T31 and T32 there only.
-
-We remove T20 from the statistical analyses because this ecosystem type is only ever registerd as in good condition, and it appears only in two regions.
-```{r}
-SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(hovetype != 'T20')
-```
+T20 is has only very few data and is present in only 1 conditino class and 2 regions, we omit it.
+Northern Norway does not have any T2 occurrences, central Norway doesn't have any T21 occurrences, we won't be able to model these types across all regions.
 
@@
   filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'T41')
+  filter(hovedtype != 'T20')
 
@@
 summary(SentinelNDVI.natopen.train$mean_beta) #ok
-# replace 1s in Landsat data with 0.9999
-LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
-  mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
 
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.natopen.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region!='Northern Norway',])
+# we exclude Northern Norway and T21 for this first exploratory model
+model.natopen.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region!='Northern Norway' & SentinelNDVI.natopen.train$hovedtype!='T21',])
 model.natopen.cond.Sent <- StepBeta(model.natopen.cond.Sent)
@@
 ```{r message=FALSE, warning=FALSE}
-model.natopen.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Northern Norway" & SentinelNDVI.natopen.train$hovedtype!='T34',])
+model.natopen.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Northern Norway",])
 model.natopen.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Central Norway",])
@@
 ```
-NDVI in good condition: T32 > T31/T33 > T40 (last not sig. but very few data)
-
-T31,T33,T40: NDVI increases as condition deteriorates
-T32: NDVI does not change with condition
+Not much of a pattern in either model or figure.
 
@@
 ```
-NDVI in good condition: T32 > T33/T34/T40 > T31
+NDVI in good condition: T12/T18 > T2
 
-T31: NDVI increases as condition deteriorates
-T32/T33/T34/T40: NDVI increases sig. less than in T31 as condition deteriorates, but it still increases
+No changes with condition.
 
@@
 ```
-NDVI in good condition: T31/T32/T40 > T33/T34
+NDVI in good condition: T18 marginally > T2/T12/T21
 
-all types: NDVI increases slightly as condition deteriorates, but just about sig.
+No changes with condition.
 
@@
 ```
-NDVI in good condition: T32 > T31/T40 > T33 > T34
+NDVI in good condition: T18 > T2/T12 > T21
 
-T31/T32/T34/T40: NDVI increases slightly as condition deteriorates
-T33: NDVI increases as condition deteriorates, more strongly than in the other types
+No changes with condition.
 
@@
 ```
-NDVI in good condition: T32 > T31/T40 > T33/T34
+NDVI in good condition: T12/T18 > T2/T21
 
-T31/T32: borderline NDVI increases as condition deteriorates
-T33/T34: NDVI increases as condition deteriorates
-T40: insufficient power
+Reduced NDVI with reduced condition in sand dune vegetation types in Southern Norway, Otherwise no changes with condition.
 
@@
 
-In natopen ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
-Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country.
-natopen and natopen-looking meadows show very tiny increases if any at all.
+In naturally open ecosystems below the tree line, NDVI seems largely independent of ecosystem condition.
+This is probably because many of the regarding ecosystems are structured by natural disturbances anyways.
+The only system showing a clear and convincing trend is the sand dunes in Southern Norway with a reduction in NDVI as condition deteriorates (due to Wear/tear and alien species).
 
@@
 summary(ModisNDVI.natopen$mean_beta) # ok
-summary(LandsatNDVI.natopen$mean_beta) # has 1's
+summary(LandsatNDVI.natopen$mean_beta) 
 # replace 1s in Landsat data with 0.9999
@@
 They vary both in their placement along the y-axis and in their variance.
-Modis data are absent from the T40/T41-ecosystem types because of a too large pixel size compared to the NiN-polygon sizes.
-In the other types, Modis is poorly represented for the same reason.
+Modis data are are basically only available for flood plains (T18), as the other ecosystems types' polygons are too small for the Modis pixel size.
 
@@
 
-We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
+We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for Sentinel and Landsat.
 ```{r message=FALSE, warning=FALSE}
 model.natopen.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$tilstand=="God",])
-model.natopen.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.natopen[ModisNDVI.natopen$tilstand=="God",])
 model.natopen.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$tilstand=="God",])
@@
 summary(model.natopen.time.GodTilst.Sent)$coefficients$cond
-summary(model.natopen.time.GodTilst.Modi)$coefficients$cond
 summary(model.natopen.time.GodTilst.Land)$coefficients$cond
 ```
-There's an increase in NDVI over time in all three satellite time series in essentially all major natopen ecosystem types in good condition.
-It is worth pointing out though that the model on the MODIS data shows 1/10 of the effect size for the time slope and has way less statistical power compared to Sentinel and Landsat.
+There's an increase in NDVI over time in both satellite time series in T12 (coastal meadows) and T18 (flood plains) in good condition.
+In the Sentinel data this pattern is absent for T2 and T21, in the Landsat data the increase is reduced.
 
-Given the systematically higher NDVI-values with decreasing condition in natopen natopens, We may ask the question if the temporal NDVI increase might be stronger in natopen natopens?
-```{r message=FALSE, warning=FALSE}
-model.natopen.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("V9","V10"),])
-model.natopen.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.natopen[ModisNDVI.natopen$hovedtype %in% c("V9","V10"),])
-model.natopen.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("V9","V10"),])
-summary(model.natopen.time.tilst.V9V10.Sent)$coefficients$cond
-summary(model.natopen.time.tilst.V9V10.Modi)$coefficients$cond
-summary(model.natopen.time.tilst.V9V10.Land)$coefficients$cond
+We can check if these temporal changes in T12 and T18, as well as for T21 in Southern Norway, vary between condition classes?
+```{r, warning=FALSE}
+model.natopen.time.tilst.T12.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T12"),])
+model.natopen.time.tilst.T12.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T12"),])
+model.natopen.time.tilst.T18.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T18"),])
+model.natopen.time.tilst.T18.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T18"),])
+model.natopen.time.tilst.T21.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T21") & SentinelNDVI.natopen$region=="Southern Norway",])
+model.natopen.time.tilst.T21.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T21") & LandsatNDVI.natopen$region=="Southern Norway",])
+summary(model.natopen.time.tilst.T12.Sent)$coefficients$cond
+summary(model.natopen.time.tilst.T18.Sent)$coefficients$cond
+summary(model.natopen.time.tilst.T21.Sent)$coefficients$cond
+
+summary(model.natopen.time.tilst.T12.Land)$coefficients$cond
+summary(model.natopen.time.tilst.T18.Land)$coefficients$cond
+summary(model.natopen.time.tilst.T21.Land)$coefficients$cond
 ```
-Indeed, the temporal NDVI increase has been stronger in natopen natopens with reduced ecological condition, in both Sentinel and Landsat.
-For MODIS, the time slope is again 1/10 of the slopes in the other two satellite time series, and it is not significant.
+The temporal NDVI increase has been stronger in flood plains that are in reduced ecological condition, but only in the Landsat time series. The other ecosystem types do not show any obvious patterns.
 
 # Overall conclusion
-NDVI in natopen ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
-Especially the natopen natopen ecosystems show a very consistent pattern of higher NDVI and larger temporal NDVI increase in sites in reduced ecological condition, which also links well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
-The large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
-Since ecological condition also relates to the temporal development of NDVI values in the analyses above, an alternative approach can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
-2022, see references above).
+NDVI in natopen ecosystem varies very little with condition.
+The variance and thus overlap between condition classes and ecosystem types is large.
+Exceptions are sand dunes in Southern Norway, which show reduced NDVI under deteriorating condition in the Sentinel data, and flood plains nationally, where areas in deteriorating condition show a stronger increase in NDVI over time.
+Other than for these two phenomena, NDVI does not appear as a sensitive measure of ecological condition in natopen systems.
 
@@
 # 1) We are lacking ecosystem maps for natopen ecosystems
-The study at hand clearly shows that different natopen ecosystem types display different greenness levels and thus need to be evaluated separately.
+The analysis at hand clearly shows that different natopen ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within natopens separately.
 # 2) We are lacking a good ecological understanding and ground-truthing for NDVI in natopen ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in natopens.
-Here, we need field studies investigating:
-- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition
-
-- How reflectance in different natopen types is affected by how moist the moss layer is at the time of measuring.
-The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
-Minerotrophic natopens (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, natopens are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
-
-In addition, further investigations should test how NDVI-condition pattern vary across biogeographic regions in addition to administrative regions (which are ecologically irrelevant), in order to better understand the results for the administrative regions.
+Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems.
+Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like different land-use schemes in the different ecosystem types, are connected.
+This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
 

### 2023-09-19 DRAFT_NDVI_seminat.Rmd
ndvi wetland, seminat & natopen finished

@@
 
+# Where are the seminat sites?
+```{r}
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(SentinelNDVI.seminat) +
+  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale=2.5, legend.show = FALSE) + 
+  tm_add_legend(type = "symbol", 
+                col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
+                labels = c("T31 - boreal heath",
+                           "T32 - semi-natural grasslands",
+                           "T33 - semi-natural coastal grasslands",
+                           "T34 - coastal heathlands",
+                           "T40/41 - previously heavily altered grassland"),
+                title = "Major ecosystem type",
+                size=3) +
+  tm_layout(main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
+            main.title.size=1.2, legend.position = c("right", "bottom"), 
+            legend.text.size = 1.3, legend.title.size = 1.4)
+```
+
 # How are polygon sizes distributed?
@@
   ggtitle("Condition variables in boreal heath (T31)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in semi-natural grasslands (T32)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in coastal heathlands (T34)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-```{r message=FALSE, warning=FALSE}
+```{r, warning=FALSE}
 model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
@@
 Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
-```{r message=FALSE, warning=FALSE}
+```{r, warning=FALSE}
 model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"),])
@@
 We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
-```{r message=FALSE, warning=FALSE}
+```{r, warning=FALSE}
 model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"),])
@@
 # 1) We are lacking ecosystem maps for semi-natural ecosystems
-The study at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
+The analysis at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within semi-natural systems separately.
@@
 Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like for example shrub-/tree encroachment, are connected.
+This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
 

### 2023-09-19 DRAFT_NDVI_wetland.Rmd
ndvi wetland, seminat & natopen finished

@@
 
+# Where are the wetland sites?
+```{r}
+tm_shape(regnor) +
+  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_borders() +
+  tm_shape(SentinelNDVI.wetland) +
+  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale=2.5, legend.show = FALSE) + 
+  tm_add_legend(type = "symbol", 
+                col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
+                labels = c("V1 - minerotrophic mires",
+                           "V3 - ombrotrophic bogs",
+                           "V4 - cold springs",
+                           "V6 - wet snowbeds",
+                           "V9 - semi-natural mires",
+                           "V10 - semi-natural wet meadows"),
+                title = "Major ecosystem type",
+                size=3) +
+  tm_layout(main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
+            main.title.size=1.2, legend.position = c("right", "bottom"), 
+            legend.text.size = 1.3, legend.title.size = 1.4)
+```
+
 # How are polygon sizes distributed?
@@
   ggtitle("Condition variables in semi-natural mires (V9)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
   ggtitle("Condition variables in minerotrophic fens (V1)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r warning=FALSE, message=FALSE}
+```{r warning=FALSE}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-```{r message=FALSE, warning=FALSE}
+```{r warning=FALSE}
 model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
@@
 Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
-```{r message=FALSE, warning=FALSE}
+```{r warning=FALSE}
 model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
@@
 # 1) We are lacking ecosystem maps for wetland ecosystems
-The study at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
+The analysis at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
@@
 Here, we need field studies investigating:
-- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition
+- The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition.
+This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
 

### 2023-09-20 NDVI_natopen.Rmd
Small changes in the code to make it run in the book

--- 2023-09-19 DRAFT_NDVI_natopen.Rmd
@@
 library(RColorBrewer)
-library("gridExtra") 
+library(gridExtra)
 library(ggridges)
@@
   "Økologisk egenskap" = Egenskap,
-                       "ECT class" = ECT)
+  "ECT class" = ECT
+)
 knitr::kable(metaData)
@@
 # Add NiN data from cache
-nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
+nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
 ```
@@
 ```{r, echo = T, eval = F}
-
 url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
-download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
-unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
-       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
+unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
+  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
+)
 st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
 nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
-                 layer="naturtyper_nin_omr")
+  layer = "naturtyper_nin_omr"
+)
 ```
+
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
@@
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
-nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
   st_as_sf() %>%
@@
 
-reg <- st_read("data/regions.shp")%>%
+reg <- st_read("data/regions.shp") %>%
   st_as_sf() %>%
@@
 reg$region
-reg$region <- c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway")
+reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")
 
 # combine the Norway and the region maps
-regnor <- st_intersection(reg,nor)
+regnor <- st_intersection(reg, nor)
 rm(reg)
@@
 # downloading the unprocessed NDVI data from a data-cache et the NINA server
-df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
-df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
-df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
+df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
+df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
+df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
 ```
@@
 # Import Sentinel NDVI Data
-df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
-  map_df(~fread(.))
+df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
+  map_df(~ fread(.))
 ```
@@
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
-df.m$mean<-df.m$mean/10000
+df.m <- read.csv("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv")
+df.m$mean <- df.m$mean / 10000
 ```
@@
 # Set up conditional file paths
-dir <- substr(getwd(), 1,2)
+dir <- substr(getwd(), 1, 2)
 
@@
   "R:/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb",
-               "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")
+  "/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_nin_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb"
+)
 
@@
   "P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat",
-                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat")
+  "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Landsat"
+)
 
@@
 # hence this horrendous work around
-files=list.files(pData, pattern = "*.csv", full.names = TRUE)
+files <- list.files(pData, pattern = "*.csv", full.names = TRUE)
 
-df_list<-list()# initialise a list of dataframes
+df_list <- list() # initialise a list of dataframes
 # read in a dataframe in each slot of the df_list
-for (i in files){
-  name <- gsub("-",".",i)
-  name <- gsub(".csv","",name)  
-  i <- paste(i,sep="")
-  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
+for (i in files) {
+  name <- gsub("-", ".", i)
+  name <- gsub(".csv", "", name)
+  i <- paste(i, sep = "")
+  df_list[[i]] <- assign(name, read.csv(i, header = TRUE))
 }
 
-df.l<-bind_rows(df_list, .id = "column_label")
+df.l <- bind_rows(df_list, .id = "column_label")
 ```
@@
 colnames(nin)
-colnames(nin)[c(3,8,17,26,31,33,34)] <- c("hovedoekosystem","kartleggingsaar","noyaktighet",
-                                          "omraadenavn","uk_naertruet","uk_sentraloekosystemfunksjon",
-                                          "uk_spesieltdaarligkartlagt")
+colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
+  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
+  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
+  "uk_spesieltdaarligkartlagt"
+)
 unique(nin$hovedoekosystem)
 
-nin <- nin %>% mutate(hovedoekosystem = recode(hovedoekosystem, 
-                                               "Våtmark" = 'Vaatmark',
-                                               "Semi-naturlig mark" = 'Semi_naturlig',
-                                               "Naturlig åpne områder i lavlandet" = 'Naturlig-aapent',
-                                               "Naturlig åpne områder under skoggrensa" = 'Naturlig_aapent')) %>%
+nin <- nin %>%
+  mutate(hovedoekosystem = recode(hovedoekosystem,
+    "Våtmark" = "Vaatmark",
+    "Semi-naturlig mark" = "Semi_naturlig",
+    "Naturlig åpne områder i lavlandet" = "Naturlig-aapent",
+    "Naturlig åpne områder under skoggrensa" = "Naturlig_aapent"
+  )) %>%
   mutate(validGeo = st_is_valid(SHAPE))
@@
 nrow(nin %>%
-       filter(grepl(',', ninkartleggingsenheter))
-)
+  filter(grepl(",", ninkartleggingsenheter)))
 # 21094 polygons have more than 1 ecosystem type (they are separated by commas in the ninkartleggingsenheter-variable)
@@
 nrow(nin %>%
-       filter(!grepl(',', ninkartleggingsenheter))
-)
+  filter(!grepl(",", ninkartleggingsenheter)))
 # 74375 polygons should have only 1 ecosystem type
@@
 nin <- nin %>%
-  filter(!grepl(',', ninkartleggingsenheter))
+  filter(!grepl(",", ninkartleggingsenheter))
 
@@
 # get rid of the NA- in the beginning
-nin <- nin %>% mutate(ninkartleggingsenheter = str_remove(ninkartleggingsenheter, 'NA_'))
+nin <- nin %>% mutate(ninkartleggingsenheter = str_remove(ninkartleggingsenheter, "NA_"))
 # making a main ecosystem type variable
-nin <- nin %>% mutate(hovedtype = substr(ninkartleggingsenheter, 1, 3),
-                      hovedtype = str_remove(hovedtype, '-'))
+nin <- nin %>% mutate(
+  hovedtype = substr(ninkartleggingsenheter, 1, 3),
+  hovedtype = str_remove(hovedtype, "-")
+)
 # checking mapping unit against main ecosystem type
-nin[,c("hovedoekosystem","hovedtype")]
-summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Vaatmark"]))
-summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Semi_naturlig"]))
-summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Naturlig_aapne"]))
-summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Skog"]))
-summary(as.factor(nin$hovedtype[nin$hovedoekosystem=="Fjell"]))
+nin[, c("hovedoekosystem", "hovedtype")]
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Vaatmark"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Semi_naturlig"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Naturlig_aapne"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Skog"]))
+summary(as.factor(nin$hovedtype[nin$hovedoekosystem == "Fjell"]))
 # making a new variable for the overarching ecosystem types based on the main ecosystem types
@@
 nin <- nin %>%
-  mutate(hovedoekosystem = case_when(hovedtype %in% paste("V",c(1,3:7,9:10),sep="") ~ 'Vaatmark',
-                                     hovedtype %in% paste("V",11:13,sep="") ~ 'Vaatmark_sterkt_endret',
-                                     hovedtype %in% paste("T",c(31:34,40:41),sep="") ~ 'Semi_naturlig',
-                                     hovedtype %in% paste("T",c(2,12,18,20:21),sep="") ~ 'Naturlig_aapent',
-                                     hovedtype %in% c(paste("T",c(4,30,38),sep=""),paste("V",c(2,8),sep="")) ~ 'Skog',
-                                     hovedtype %in% c(paste("T",c(3,7,9,10,14,22,26),sep=""),paste("V",c(6,7),sep="")) ~ 'Fjell',
-                                     TRUE ~ 'NA'))
+  mutate(hovedoekosystem = case_when(
+    hovedtype %in% paste("V", c(1, 3:7, 9:10), sep = "") ~ "Vaatmark",
+    hovedtype %in% paste("V", 11:13, sep = "") ~ "Vaatmark_sterkt_endret",
+    hovedtype %in% paste("T", c(31:34, 40:41), sep = "") ~ "Semi_naturlig",
+    hovedtype %in% paste("T", c(2, 12, 18, 20:21), sep = "") ~ "Naturlig_aapent",
+    hovedtype %in% c(paste("T", c(4, 30, 38), sep = ""), paste("V", c(2, 8), sep = "")) ~ "Skog",
+    hovedtype %in% c(paste("T", c(3, 7, 9, 10, 14, 22, 26), sep = ""), paste("V", c(6, 7), sep = "")) ~ "Fjell",
+    TRUE ~ "NA"
+  ))
 
@@
   "Dårlig" = "Redusert",
-                                        "Svært redusert" = "Svaert_redusert"))
+  "Svært redusert" = "Svaert_redusert"
+))
 nin <- nin %>% mutate(tilstand_num = recode(tilstand,
   "God" = 0, "Moderat" = 1,
-                                        "Redusert" = 2, "Svaert_redusert" = 3))
+  "Redusert" = 2, "Svaert_redusert" = 3
+))
 
@@
 nin.natopen <- nin %>%
-  filter(hovedoekosystem %in% c('Naturlig_aapent')) %>%
+  filter(hovedoekosystem %in% c("Naturlig_aapent")) %>%
   mutate(id = identifikasjon_lokalid) %>%
@@
 # merge NiN-data with region
-nin.natopen = st_join(nin.natopen, regnor, left = TRUE)
+nin.natopen <- st_join(nin.natopen, regnor, left = TRUE)
 nin.natopen
 
-colnames(nin.natopen)[c(1,10)] <- c("id","region_id")
+colnames(nin.natopen)[c(1, 10)] <- c("id", "region_id")
 
@@
 nin.natopen <- nin.natopen %>%
-  mutate(area_meters_nin = st_area(nin.natopen)
-  )
+  mutate(area_meters_nin = st_area(nin.natopen))
 # check and edit the order of regions
@@
 levels(nin.natopen$region)
-nin.natopen$region <- factor(nin.natopen$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+nin.natopen$region <- factor(nin.natopen$region, levels = c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway"))
 levels(nin.natopen$region)
@@
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.natopen2 <- nin.natopen %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
-  separate(col = ninbeskrivelsesvariabler,
+nin.natopen2 <- nin.natopen %>%
+  separate_rows(ninbeskrivelsesvariabler, sep = ",") %>%
+  separate(
+    col = ninbeskrivelsesvariabler,
     into = c("NiN_variable_code", "NiN_variable_value"),
     sep = "_",
-           remove=F) %>%
+    remove = F
+  ) %>%
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
@@
 nin.natopen2 <- as.data.frame(nin.natopen2)
-
 ```
@@
 ```{r, echo = F}
-
-#load processed NiN-data from cache
-nin.natopen <- readRDS(paste0(here::here(),"/data/cache/nin.natopen.RDS"))
-nin.natopen2 <- readRDS(paste0(here::here(),"/data/cache/nin.natopen2.RDS"))
-
+# load processed NiN-data from cache
+nin.natopen <- readRDS(paste0(here::here(), "/data/cache/nin.natopen.RDS"))
+nin.natopen2 <- readRDS(paste0(here::here(), "/data/cache/nin.natopen2.RDS"))
 ```
@@
 # join nin.natopen & Sentinel NDVI data
-SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by="id")
-#summary(SentinelNDVI.natopen)
+SentinelNDVI.natopen <- full_join(nin.natopen, df.s, by = "id")
+# summary(SentinelNDVI.natopen)
 SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.natopen))
-#summary(SentinelNDVI.natopen)
+    area_meters = st_area(SentinelNDVI.natopen)
+  )
+# summary(SentinelNDVI.natopen)
 # get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
@@
 SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(!is.na(mean))
-#summary(SentinelNDVI.natopen)
+# summary(SentinelNDVI.natopen)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
@@
 SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # add column for sub-ecosystem types
-SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
-                                                        subtype = str_remove(subtype, '-'),
-                                                        subtype = str_remove(subtype, '-'))
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% mutate(
+  subtype = substring(ninkartleggingsenheter, 4),
+  subtype = str_remove(subtype, "-"),
+  subtype = str_remove(subtype, "-")
+)
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
@@
 # join nin.natopen and Modis NDVI data
-ModisNDVI.natopen <- full_join(nin.natopen, df.m, by="id")
+ModisNDVI.natopen <- full_join(nin.natopen, df.m, by = "id")
 
-#summary(ModisNDVI.natopen)
+# summary(ModisNDVI.natopen)
 ModisNDVI.natopen <- ModisNDVI.natopen %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.natopen))
-#summary(ModisNDVI.natopen)
+    area_meters = st_area(ModisNDVI.natopen)
+  )
+# summary(ModisNDVI.natopen)
 # get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
@@
 ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(!is.na(mean))
-#summary(ModisNDVI.natopen)
+# summary(ModisNDVI.natopen)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
@@
 ModisNDVI.natopen <- ModisNDVI.natopen %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.natopen)
+  filter(mean == max(mean, na.rm = TRUE))
+# summary(ModisNDVI.natopen)
 
@@
 # join nin.natopen & Landsat NDVI data
-LandsatNDVI.natopen <- full_join(nin.natopen, df.l, by="id")
+LandsatNDVI.natopen <- full_join(nin.natopen, df.l, by = "id")
 
-#summary(LandsatNDVI.natopen)
+# summary(LandsatNDVI.natopen)
 LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.natopen))
-#summary(LandsatNDVI.natopen)
+    area_meters = st_area(LandsatNDVI.natopen)
+  )
+# summary(LandsatNDVI.natopen)
 # get rid of NAs (i.e. NDVI cells that were not in natopen polygons)
@@
 LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(!is.na(mean))
-#summary(LandsatNDVI.natopen)
+# summary(LandsatNDVI.natopen)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
@@
 LandsatNDVI.natopen <- LandsatNDVI.natopen %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
-#summary(LandsatNDVI.natopen)
+# summary(LandsatNDVI.natopen)
 ```
@@
 tm_shape(regnor) +
-  tm_fill('GID_0', labels="", title="", legend.show = FALSE) + 
+  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
   tm_borders() +
   tm_shape(SentinelNDVI.natopen) +
-  tm_dots('hovedtype',midpoint=NA, palette=tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale=2.5, legend.show = FALSE) + 
-  tm_add_legend(type = "symbol", 
+  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
+  tm_add_legend(
+    type = "symbol",
     col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
-                labels = c("T12 - coastal meadows",
+    labels = c(
+      "T12 - coastal meadows",
       "T18 - flood plains",
@@
       "T20 - innfrysningsmark",
-                           "T21 - sand dunes"),
+      "T21 - sand dunes"
+    ),
     title = "Major ecosystem type",
-                size=3) +
-  tm_layout(main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
-            main.title.size=1.2, legend.position = c("right", "bottom"), 
-            legend.text.size = 1.3, legend.title.size = 1.4)
+    size = 3
+  ) +
+  tm_layout(
+    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
+    main.title.size = 1.2, legend.position = c("right", "bottom"),
+    legend.text.size = 1.3, legend.title.size = 1.4
+  )
 ```
@@
 summary(SentinelNDVI.natopen$area_meters)
-hist(SentinelNDVI.natopen$area_meters,xlim=c(0,20000),breaks=50000)
-abline(v=100,lty=2)
+hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
+abline(v = 100, lty = 2)
 ```
@@
 SentinelNDVI.natopen %>%
-  ggplot( aes(x=year, y=mean )) + 
+  ggplot(aes(x = year, y = mean)) +
   geom_point() +
-  facet_grid( tilstand~hovedtype)
+  facet_grid(tilstand ~ hovedtype)
 ```
@@
 ```{r}
-SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(year != '2022')
+SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(year != "2022")
 ```
@@
 SentinelNDVI.natopen %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=hovedtype, y=mean )) + 
+  ggplot(aes(x = hovedtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey")
+  geom_point(size = 0.7, shape = 16, color = "grey")
 
-SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","E1"))
+SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
 SentinelNDVI.natopen %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=subtype, y=mean )) + 
+  ggplot(aes(x = subtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2, size=6))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
 ```
@@
 SentinelNDVI.natopen %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
   filter(!is.na(region)) %>%
-  ggplot(aes(x=region, y=mean )) + 
+  ggplot(aes(x = region, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
 ```
@@
   filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype)
 ```
@@
 nin.natopen2 %>%
-  filter(hovedtype=='T2') %>%
-  filter(NiN_variable_code %in% c("7FA","7SE","7TK","1AG-B")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7FA"="Alien species",
-                                 "7SE"="Wear/tear",
-                                 "7TK"="Heavy vehicles",
-                                 "1AG-B"="Shrub cover")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des, ncol=2) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T2") %>%
+  filter(NiN_variable_code %in% c("7FA", "7SE", "7TK", "1AG-B")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7FA" = "Alien species",
+    "7SE" = "Wear/tear",
+    "7TK" = "Heavy vehicles",
+    "1AG-B" = "Shrub cover"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, ncol = 2) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in open systems on shallow soils (T2)") +
-  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.natopen2 %>%
-  filter(hovedtype=='T12') %>%
-  filter(NiN_variable_code %in% c("7FA","7SE","7TK")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7FA"="Alien species",
-                                 "7SE"="Wear/tear",
-                                 "7TK"="Heavy vehicles")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des, ncol=2) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T12") %>%
+  filter(NiN_variable_code %in% c("7FA", "7SE", "7TK")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7FA" = "Alien species",
+    "7SE" = "Wear/tear",
+    "7TK" = "Heavy vehicles"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, ncol = 2) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in natural coastal meadows (T12)") +
-  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.natopen2 %>%
-  filter(hovedtype=='T18') %>%
-  filter(NiN_variable_code %in% c("7FA","7JB-BT","7SE","7VR-RI")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7FA"="Alien species",
-                                 "7JB-BT"="Grazing pressure",
-                                 "7SE"="Wear/tear",
-                                 "7VR-RI"="Water regulation")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des, ncol=2) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T18") %>%
+  filter(NiN_variable_code %in% c("7FA", "7JB-BT", "7SE", "7VR-RI")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7FA" = "Alien species",
+    "7JB-BT" = "Grazing pressure",
+    "7SE" = "Wear/tear",
+    "7VR-RI" = "Water regulation"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, ncol = 2) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in flood plains (T18)") +
-  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.natopen2 %>%
-  filter(hovedtype %in% c('T21')) %>%
-  filter(NiN_variable_code %in% c("7FA","7JB-BT","7SE","7TK","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7FA"="Alien species",
-                                 "7JB-BT"="Grazing pressure",
-                                 "7SE"="Wear/tear",
-                                 "7TK"="Heavy vehicles",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des, ncol=2) +
-  theme(legend.position="none") +
+  filter(hovedtype %in% c("T21")) %>%
+  filter(NiN_variable_code %in% c("7FA", "7JB-BT", "7SE", "7TK", "7JB-GJ")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7FA" = "Alien species",
+    "7JB-BT" = "Grazing pressure",
+    "7SE" = "Wear/tear",
+    "7TK" = "Heavy vehicles",
+    "7JB-GJ" = "Fertilization"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, ncol = 2) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in sand dunes (T21)") +
-  labs(y= "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
-
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 ```
@@
 SentinelNDVI.natopen %>%
-
   group_by(id, year) %>%
-
   filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_grid( region~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_grid(region ~ hovedtype)
 ```
@@
   filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'T20')
+  filter(hovedtype != "T20")
 
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.natopen.train$mean_beta) #ok
+summary(SentinelNDVI.natopen.train$mean_beta) # ok
 
@@
 # we exclude Northern Norway and T21 for this first exploratory model
-model.natopen.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region!='Northern Norway' & SentinelNDVI.natopen.train$hovedtype!='T21',])
+model.natopen.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region != "Northern Norway" & SentinelNDVI.natopen.train$hovedtype != "T21", ])
 model.natopen.cond.Sent <- StepBeta(model.natopen.cond.Sent)
@@
 ```{r message=FALSE, warning=FALSE}
-model.natopen.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Northern Norway",])
-model.natopen.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Central Norway",])
-model.natopen.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Western Norway",])
-model.natopen.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Eastern Norway",])
-model.natopen.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region=="Southern Norway",])
-
+model.natopen.cond.Sent.N <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region == "Northern Norway", ])
+model.natopen.cond.Sent.C <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region == "Central Norway", ])
+model.natopen.cond.Sent.W <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region == "Western Norway", ])
+model.natopen.cond.Sent.E <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region == "Eastern Norway", ])
+model.natopen.cond.Sent.S <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.natopen.train[SentinelNDVI.natopen.train$region == "Southern Norway", ])
 ```
@@
 SentinelNDVI.natopen.train %>%
-  filter(region=="Northern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Northern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.natopen.train %>%
-  filter(region=="Central Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Central Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.natopen.train %>%
-  filter(region=="Western Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Western Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.natopen.train %>%
-  filter(region=="Eastern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Eastern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.natopen.train %>%
-  filter(region=="Southern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Southern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  ggplot( aes(x=year, y=mean )) + 
+  filter(mean == max(mean, na.rm = TRUE)) %>%
+  ggplot(aes(x = year, y = mean)) +
   geom_point() +
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
+# SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
 ModisNDVI.natopen$mean_beta <- (ModisNDVI.natopen$mean + 1) / 2
@@
 SentinelNDVI.natopen <- SentinelNDVI.natopen %>%
-  rename('system.index' = 'system:index')
+  dplyr::rename("system.index" = "system:index")
 
@@
 allSatNDVI.natopen <- rbind(
-SentinelNDVI.natopen[,!names(SentinelNDVI.natopen) %in% "subtype"],
+  SentinelNDVI.natopen[, !names(SentinelNDVI.natopen) %in% "subtype"],
   ModisNDVI.natopen,
-LandsatNDVI.natopen[,!names(LandsatNDVI.natopen) %in% "column_label"]
+  LandsatNDVI.natopen[, !names(LandsatNDVI.natopen) %in% "column_label"]
 )
@@
 allSatNDVI.natopen$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.natopen)),
-  rep("Modis",nrow(ModisNDVI.natopen)),
-  rep("Landsat",nrow(LandsatNDVI.natopen))
+  rep("Sentinel", nrow(SentinelNDVI.natopen)),
+  rep("Modis", nrow(ModisNDVI.natopen)),
+  rep("Landsat", nrow(LandsatNDVI.natopen))
 )
-allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat,levels=c("Sentinel","Modis","Landsat"))
+allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat, levels = c("Sentinel", "Modis", "Landsat"))
 levels(allSatNDVI.natopen$Sat)
@@
 allSatNDVI.natopen %>%
-  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
+  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
   geom_point() +
@@
 ```{r message=FALSE, warning=FALSE}
-model.natopen.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$tilstand=="God",])
-model.natopen.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$tilstand=="God",])
+model.natopen.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$tilstand == "God", ])
+model.natopen.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$tilstand == "God", ])
 
@@
 ```{r, warning=FALSE}
-model.natopen.time.tilst.T12.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T12"),])
-model.natopen.time.tilst.T12.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T12"),])
-model.natopen.time.tilst.T18.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T18"),])
-model.natopen.time.tilst.T18.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T18"),])
-model.natopen.time.tilst.T21.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T21") & SentinelNDVI.natopen$region=="Southern Norway",])
-model.natopen.time.tilst.T21.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T21") & LandsatNDVI.natopen$region=="Southern Norway",])
+model.natopen.time.tilst.T12.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T12"), ])
+model.natopen.time.tilst.T12.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T12"), ])
+model.natopen.time.tilst.T18.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T18"), ])
+model.natopen.time.tilst.T18.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T18"), ])
+model.natopen.time.tilst.T21.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T21") & SentinelNDVI.natopen$region == "Southern Norway", ])
+model.natopen.time.tilst.T21.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T21") & LandsatNDVI.natopen$region == "Southern Norway", ])
 summary(model.natopen.time.tilst.T12.Sent)$coefficients$cond

### 2023-09-20 NDVI_seminat.Rmd
Small changes in the code to make it run in the book

--- 2023-09-18 DRAFT_NDVI_seminat.Rmd
@@
 ```{r, echo = T, eval = F}
-
 url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
-download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
-unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
-       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
+unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
+  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
+)
 st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
 nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
-                 layer="naturtyper_nin_omr")
+  layer = "naturtyper_nin_omr"
+)
 ```
@@
 # Import Sentinel NDVI Data
-df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
-  map_df(~fread(.))
+df.s <- list.files("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
+  map_df(~ fread(.))
 ```
@@
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
-df.m$mean<-df.m$mean/10000
+df.m <- read.csv("/data/P-Prosjekter241201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv", )
+df.m$mean <- df.m$mean / 10000
 ```
@@
 # hence this horrendous work around
-files=list.files(pData, pattern = "*.csv", full.names = TRUE)
-
-df_list<-list()# initialise a list of dataframes
+files <- list.files(pData, pattern = "*.csv", full.names = TRUE)
+
+df_list <- list() # initialise a list of dataframes
 # read in a dataframe in each slot of the df_list
-for (i in files){
-  name <- gsub("-",".",i)
-  name <- gsub(".csv","",name)  
-  i <- paste(i,sep="")
-  df_list[[i]]=assign(name,read.csv(i, header=TRUE))
+for (i in files) {
+  name <- gsub("-", ".", i)
+  name <- gsub(".csv", "", name)
+  i <- paste(i, sep = "")
+  df_list[[i]] <- assign(name, read.csv(i, header = TRUE))
 }
 
-df.l<-bind_rows(df_list, .id = "column_label")
+df.l <- bind_rows(df_list, .id = "column_label")
 ```
@@
 nin.seminat <- nin %>%
-  filter(hovedoekosystem %in% c('Semi_naturlig')) %>%
+  filter(hovedoekosystem %in% c("Semi_naturlig")) %>%
   mutate(id = identifikasjon_lokalid) %>%
@@
 # merge NiN-data with region
-nin.seminat = st_join(nin.seminat, regnor, left = TRUE)
+nin.seminat <- st_join(nin.seminat, regnor, left = TRUE)
 nin.seminat
 
-colnames(nin.seminat)[c(1,10)] <- c("id","region_id")
+colnames(nin.seminat)[c(1, 10)] <- c("id", "region_id")
 
@@
 nin.seminat <- nin.seminat %>%
-  mutate(area_meters_nin = st_area(nin.seminat)
-  )
+  mutate(area_meters_nin = st_area(nin.seminat))
 # check and edit the order of regions
@@
 levels(nin.seminat$region)
-nin.seminat$region <- factor(nin.seminat$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+nin.seminat$region <- factor(nin.seminat$region, levels = c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway"))
 levels(nin.seminat$region)
@@
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.seminat2 <- nin.seminat %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
-  separate(col = ninbeskrivelsesvariabler,
+nin.seminat2 <- nin.seminat %>%
+  separate_rows(ninbeskrivelsesvariabler, sep = ",") %>%
+  separate(
+    col = ninbeskrivelsesvariabler,
     into = c("NiN_variable_code", "NiN_variable_value"),
     sep = "_",
-           remove=F) %>%
+    remove = F
+  ) %>%
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
@@
 nin.seminat2 <- as.data.frame(nin.seminat2)
-
 ```
@@
 ```{r, echo = F}
-
-#load processed NiN-data from cache
-nin.seminat <- readRDS(paste0(here::here(),"/data/cache/nin.seminat.RDS"))
-nin.seminat2 <- readRDS(paste0(here::here(),"/data/cache/nin.seminat2.RDS"))
-
+# load processed NiN-data from cache
+nin.seminat <- readRDS(paste0(here::here(), "/data/cache/nin.seminat.RDS"))
+nin.seminat2 <- readRDS(paste0(here::here(), "/data/cache/nin.seminat2.RDS"))
 ```
@@
 # join nin.seminat & Sentinel NDVI data
-SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by="id")
-#summary(SentinelNDVI.seminat)
+SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by = "id")
+# summary(SentinelNDVI.seminat)
 SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.seminat))
-#summary(SentinelNDVI.seminat)
+    area_meters = st_area(SentinelNDVI.seminat)
+  )
+# summary(SentinelNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
@@
 SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(mean))
-#summary(SentinelNDVI.seminat)
+# summary(SentinelNDVI.seminat)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
@@
 SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # add column for sub-ecosystem types
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
-                                                        subtype = str_remove(subtype, '-'),
-                                                        subtype = str_remove(subtype, '-'))
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(
+  subtype = substring(ninkartleggingsenheter, 4),
+  subtype = str_remove(subtype, "-"),
+  subtype = str_remove(subtype, "-")
+)
 # we are using max NDVI per year in every NiN polygon
@@
 # join nin.seminat and Modis NDVI data
-ModisNDVI.seminat <- full_join(nin.seminat, df.m, by="id")
-
-#summary(ModisNDVI.seminat)
+ModisNDVI.seminat <- full_join(nin.seminat, df.m, by = "id")
+
+# summary(ModisNDVI.seminat)
 ModisNDVI.seminat <- ModisNDVI.seminat %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.seminat))
-#summary(ModisNDVI.seminat)
+    area_meters = st_area(ModisNDVI.seminat)
+  )
+# summary(ModisNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
@@
 ModisNDVI.seminat <- ModisNDVI.seminat %>% filter(!is.na(mean))
-#summary(ModisNDVI.seminat)
+# summary(ModisNDVI.seminat)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
@@
 ModisNDVI.seminat <- ModisNDVI.seminat %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.seminat)
+  filter(mean == max(mean, na.rm = TRUE))
+# summary(ModisNDVI.seminat)
 
@@
 # join nin.seminat & Landsat NDVI data
-LandsatNDVI.seminat <- full_join(nin.seminat, df.l, by="id")
-
-#summary(LandsatNDVI.seminat)
+LandsatNDVI.seminat <- full_join(nin.seminat, df.l, by = "id")
+
+# summary(LandsatNDVI.seminat)
 LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.seminat))
-#summary(LandsatNDVI.seminat)
+    area_meters = st_area(LandsatNDVI.seminat)
+  )
+# summary(LandsatNDVI.seminat)
 # get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
@@
 LandsatNDVI.seminat <- LandsatNDVI.seminat %>% filter(!is.na(mean))
-#summary(LandsatNDVI.seminat)
+# summary(LandsatNDVI.seminat)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
@@
 LandsatNDVI.seminat <- LandsatNDVI.seminat %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-
-#summary(LandsatNDVI.seminat)
+  filter(mean == max(mean, na.rm = TRUE))
+
+# summary(LandsatNDVI.seminat)
 ```
@@
 ### Exploratory analyses, Sentinel-2
+
+# Where are the seminat sites?
+```{r}
+tm_shape(regnor) +
+  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
+  tm_borders() +
+  tm_shape(SentinelNDVI.seminat) +
+  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
+  tm_add_legend(
+    type = "symbol",
+    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
+    labels = c(
+      "T31 - boreal heath",
+      "T32 - semi-natural grasslands",
+      "T33 - semi-natural coastal grasslands",
+      "T34 - coastal heathlands",
+      "T40/41 - previously heavily altered grassland"
+    ),
+    title = "Major ecosystem type",
+    size = 3
+  ) +
+  tm_layout(
+    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
+    main.title.size = 1.2, legend.position = c("right", "bottom"),
+    legend.text.size = 1.3, legend.title.size = 1.4
+  )
+```
 
@@
 summary(SentinelNDVI.seminat$area_meters)
-hist(SentinelNDVI.seminat$area_meters,xlim=c(0,20000),breaks=50000)
-abline(v=100,lty=2)
+hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
+abline(v = 100, lty = 2)
 ```
@@
 SentinelNDVI.seminat %>%
-  ggplot( aes(x=year, y=mean )) + 
+  ggplot(aes(x = year, y = mean)) +
   geom_point() +
-  facet_grid( tilstand~hovedtype)
+  facet_grid(tilstand ~ hovedtype)
 ```
@@
 ```{r}
-SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != '2022')
+SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != "2022")
 ```
@@
 SentinelNDVI.seminat %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=hovedtype, y=mean )) + 
+  ggplot(aes(x = hovedtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey")
-
-SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype,levels=c("C1","C2","C3","C4","C5","C6","C7","C8","C9","C10","C11","C12","C13","C14","C15","C16","C17","C18","C19","C20","C21","E1","E2","E3","E4","E5","E6","E7"))
+  geom_point(size = 0.7, shape = 16, color = "grey")
+
+SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "C13", "C14", "C15", "C16", "C17", "C18", "C19", "C20", "C21", "E1", "E2", "E3", "E4", "E5", "E6", "E7"))
 SentinelNDVI.seminat %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=subtype, y=mean )) + 
+  ggplot(aes(x = subtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2, size=6))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
 ```
@@
 SentinelNDVI.seminat %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
   filter(!is.na(region)) %>%
-  ggplot(aes(x=region, y=mean )) + 
+  ggplot(aes(x = region, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
 ```
@@
 nin.seminat2 %>%
-  filter(hovedtype=='T31') %>%
-  filter(NiN_variable_code %in% c("7JB-BT","7RA-BA","7FA","7TK")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BT"="Grazing pressure",
-                                 "7RA-BA"="Succession",
-                                 "7FA"="Alien species",
-                                 "7TK"="Heavy vehicles")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T31") %>%
+  filter(NiN_variable_code %in% c("7JB-BT", "7RA-BA", "7FA", "7TK")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BT" = "Grazing pressure",
+    "7RA-BA" = "Succession",
+    "7FA" = "Alien species",
+    "7TK" = "Heavy vehicles"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in boreal heath (T31)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.seminat2 %>%
-  filter(hovedtype=='T32') %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7SE"="Wear/tear",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T32") %>%
+  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7SE", "7JB-GJ")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BA" = "Land-use intensity",
+    "7RA-SJ" = "Succession",
+    "7FA" = "Alien species",
+    "7SE" = "Wear/tear",
+    "7JB-GJ" = "Fertilization"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in semi-natural grasslands (T32)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.seminat2 %>%
-  filter(hovedtype=='T33') %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T33") %>%
+  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7JB-GJ")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BA" = "Land-use intensity",
+    "7RA-SJ" = "Succession",
+    "7FA" = "Alien species",
+    "7JB-GJ" = "Fertilization"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in semi-natural coastal grasslands (T33)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.seminat2 %>%
-  filter(hovedtype=='T34') %>%
-  filter(NiN_variable_code %in% c("7JB-BT","7RA-SJ","7FA","7TK")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BT"="Grazing pressure",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7TK"="Heavy vehicles")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "T34") %>%
+  filter(NiN_variable_code %in% c("7JB-BT", "7RA-SJ", "7FA", "7TK")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BT" = "Grazing pressure",
+    "7RA-SJ" = "Succession",
+    "7FA" = "Alien species",
+    "7TK" = "Heavy vehicles"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in coastal heathlands (T34)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.seminat2 %>%
-  filter(hovedtype %in% c('T40','T41')) %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7SE","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land-use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7SE"="Wear/tear",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype %in% c("T40", "T41")) %>%
+  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7SE", "7JB-GJ")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BA" = "Land-use intensity",
+    "7RA-SJ" = "Succession",
+    "7FA" = "Alien species",
+    "7SE" = "Wear/tear",
+    "7JB-GJ" = "Fertilization"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in previously heavily altered grasslands (T40/T41)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
-
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 ```
@@
 SentinelNDVI.seminat %>%
-
   group_by(id, year) %>%
-
   filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_grid( region~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_grid(region ~ hovedtype)
 ```
@@
   filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'T41')
+  filter(hovedtype != "T41")
 
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.seminat.train$mean_beta) #ok
+summary(SentinelNDVI.seminat.train$mean_beta) # ok
+LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
 # replace 1s in Landsat data with 0.9999
@@
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.seminat.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region!='Northern Norway',])
+model.seminat.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region != "Northern Norway", ])
 model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
@@
 ```{r message=FALSE, warning=FALSE}
-model.seminat.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Northern Norway" & SentinelNDVI.seminat.train$hovedtype!='T34',])
-model.seminat.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Central Norway",])
-model.seminat.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Western Norway",])
-model.seminat.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Eastern Norway",])
-model.seminat.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region=="Southern Norway",])
-
+model.seminat.cond.Sent.N <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Northern Norway" & SentinelNDVI.seminat.train$hovedtype != "T34", ])
+model.seminat.cond.Sent.C <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Central Norway", ])
+model.seminat.cond.Sent.W <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Western Norway", ])
+model.seminat.cond.Sent.E <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Eastern Norway", ])
+model.seminat.cond.Sent.S <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.seminat.train[SentinelNDVI.seminat.train$region == "Southern Norway", ])
 ```
@@
 SentinelNDVI.seminat.train %>%
-  filter(region=="Northern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Northern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.seminat.train %>%
-  filter(region=="Central Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Central Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.seminat.train %>%
-  filter(region=="Western Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Western Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.seminat.train %>%
-  filter(region=="Eastern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Eastern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.seminat.train %>%
-  filter(region=="Southern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Southern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
+# SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
 ModisNDVI.seminat$mean_beta <- (ModisNDVI.seminat$mean + 1) / 2
@@
 SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
-  rename('system.index' = 'system:index')
+  dplyr::rename("system.index" = "system:index")
 
@@
 allSatNDVI.seminat <- rbind(
-SentinelNDVI.seminat[,!names(SentinelNDVI.seminat) %in% "subtype"],
+  SentinelNDVI.seminat[, !names(SentinelNDVI.seminat) %in% "subtype"],
   ModisNDVI.seminat,
-LandsatNDVI.seminat[,!names(LandsatNDVI.seminat) %in% "column_label"]
+  LandsatNDVI.seminat[, !names(LandsatNDVI.seminat) %in% "column_label"]
 )
@@
 allSatNDVI.seminat$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.seminat)),
-  rep("Modis",nrow(ModisNDVI.seminat)),
-  rep("Landsat",nrow(LandsatNDVI.seminat))
-)
-allSatNDVI.seminat$Sat <- factor(allSatNDVI.seminat$Sat,levels=c("Sentinel","Modis","Landsat"))
+  rep("Sentinel", nrow(SentinelNDVI.seminat)),
+  rep("Modis", nrow(ModisNDVI.seminat)),
+  rep("Landsat", nrow(LandsatNDVI.seminat))
+)
+allSatNDVI.seminat$Sat <- factor(allSatNDVI.seminat$Sat, levels = c("Sentinel", "Modis", "Landsat"))
 levels(allSatNDVI.seminat$Sat)
@@
 allSatNDVI.seminat %>%
-  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
+  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
   geom_point() +
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-```{r message=FALSE, warning=FALSE}
-model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand=="God",])
-model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$tilstand=="God",])
-model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand=="God",])
+```{r, warning=FALSE}
+model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand == "God", ])
+model.seminat.time.GodTilst.Modi <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$tilstand == "God", ])
+model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand == "God", ])
 
@@
 Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
-```{r message=FALSE, warning=FALSE}
-model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"),])
-model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"),])
-model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"),])
+```{r, warning=FALSE}
+model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"), ])
+model.seminat.time.tilst.T31.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T31"), ])
+model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"), ])
 summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
@@
 We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
-```{r message=FALSE, warning=FALSE}
-model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"),])
-model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"),])
-model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"),])
+```{r, warning=FALSE}
+model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"), ])
+model.seminat.time.tilst.T32.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T32"), ])
+model.seminat.time.tilst.T32.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T32"), ])
 summary(model.seminat.time.tilst.T32.Sent)$coefficients$cond
@@
 
-model.seminat.time.tilst.T34.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T34"),])
-model.seminat.time.tilst.T34.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T34"),])
-model.seminat.time.tilst.T34.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T34"),])
+model.seminat.time.tilst.T34.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T34"), ])
+model.seminat.time.tilst.T34.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.seminat[ModisNDVI.seminat$hovedtype %in% c("T34"), ])
+model.seminat.time.tilst.T34.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T34"), ])
 summary(model.seminat.time.tilst.T34.Sent)$coefficients$cond

### 2023-09-20 NDVI_wetland.Rmd
Small changes in the code to make it run in the book

--- 2023-09-18 DRAFT_NDVI_wetland.Rmd
@@
 #### NiN data and geometry for Norway and the five regions
-We read the NiN data for this document from a cache at the NINA server...
+
 ```{r, echo = F}
 # Add NiN data from cache
-nin <- readRDS(paste0(here::here(),"/data/cache/nin.RDS"))
+nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
 ```
 
-...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+The NiN data can be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+
 ```{r, echo = T, eval = F}
-
 url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
-download(url, dest="P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode="wb") 
-unzip ("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", 
-       exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data")
+download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
+unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
+  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
+)
 st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
 nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
-                 layer="naturtyper_nin_omr")
+  layer = "naturtyper_nin_omr"
+)
 ```
+
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
+
 ```{r message=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
-nor <- st_read("data/outlineOfNorway_EPSG25833.shp")%>%
+nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
   st_as_sf() %>%
@@
 
+```{r, echo = F}
+# downloading the unprocessed NDVI data from a data-cache et the NINA server
+df.s <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
+df.m <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
+df.l <- readRDS("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
+```
+
 ##### Sentinel-2 NDVI data
@@
 To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
-```{r}
+
+```{r, echo = T, eval = F}
 # Import Sentinel NDVI Data
-df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
-  map_df(~fread(.))
+df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
+  map_df(~ fread(.))
 ```
@@
 
-```{r}
+```{r, echo = T, eval = F}
 # Import MODIS NDVI Data
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
-df.m$mean<-df.m$mean/10000
+df.m <- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
+df.m$mean <- df.m$mean / 10000
 ```
@@
 To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
-```{r}
+
+```{r, echo = T, eval = F}
 # Import Landsat NDVI Data
 # Set up conditional file paths
-dir <- substr(getwd(), 1,2)
+dir <- substr(getwd(), 1, 2)
 
@@
 nin.wetland <- nin %>%
-  filter(hovedoekosystem %in% c('Vaatmark')) %>%
+  filter(hovedoekosystem %in% c("Vaatmark")) %>%
   mutate(id = identifikasjon_lokalid) %>%
@@
 # merge NiN-data with region
-nin.wetland = st_join(nin.wetland, regnor, left = TRUE)
+nin.wetland <- st_join(nin.wetland, regnor, left = TRUE)
 nin.wetland
 
-colnames(nin.wetland)[c(1,10)] <- c("id","region_id")
+colnames(nin.wetland)[c(1, 10)] <- c("id", "region_id")
 
@@
 summary(as.factor(nin.wetland$id))
-nin.wetland[nin.wetland$id=="NINFP2110008060",]
+nin.wetland[nin.wetland$id == "NINFP2110008060", ]
 # this one was assigned to both Western and Central Norway, we drop the latter
-nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% '1144.1',]
+nin.wetland <- nin.wetland[!row.names(nin.wetland) %in% "1144.1", ]
 
 nin.wetland <- nin.wetland %>%
-  mutate(area_meters_nin = st_area(nin.wetland)
-  )
+  mutate(area_meters_nin = st_area(nin.wetland))
 # check and edit the order of regions
@@
 levels(nin.wetland$region)
-nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway","Central Norway","Eastern Norway","Western Norway","Southern Norway"))
+nin.wetland$region <- factor(nin.wetland$region, levels = c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway"))
 levels(nin.wetland$region)
@@
 # splitting content in the ninbeskrivelsesvariabler-column into separate rows per variable and then separating that column into two for the variable name and value
-nin.wetland2 <- nin.wetland %>% separate_rows(ninbeskrivelsesvariabler, sep=",") %>%
-  separate(col = ninbeskrivelsesvariabler,
+nin.wetland2 <- nin.wetland %>%
+  separate_rows(ninbeskrivelsesvariabler, sep = ",") %>%
+  separate(
+    col = ninbeskrivelsesvariabler,
     into = c("NiN_variable_code", "NiN_variable_value"),
     sep = "_",
-           remove=F) %>%
+    remove = F
+  ) %>%
   mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
@@
 nin.wetland2 <- as.data.frame(nin.wetland2)
-
 ```
@@
 ```{r, echo = F}
-
-#load processed NiN-data from cache
-nin.wetland <- readRDS(paste0(here::here(),"/data/cache/nin.wetland.RDS"))
-nin.wetland2 <- readRDS(paste0(here::here(),"/data/cache/nin.wetland2.RDS"))
-
+# load processed NiN-data from cache
+nin.wetland <- readRDS(paste0(here::here(), "/data/cache/nin.wetland.RDS"))
+nin.wetland2 <- readRDS(paste0(here::here(), "/data/cache/nin.wetland2.RDS"))
 ```
@@
 # join nin.wetland & Sentinel NDVI data
-SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by="id")
-#summary(SentinelNDVI.wetland)
+SentinelNDVI.wetland <- full_join(nin.wetland, df.s, by = "id")
+# summary(SentinelNDVI.wetland)
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(SentinelNDVI.wetland))
-#summary(SentinelNDVI.wetland)
+    area_meters = st_area(SentinelNDVI.wetland)
+  )
+# summary(SentinelNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(!is.na(mean))
-#summary(SentinelNDVI.wetland)
+# summary(SentinelNDVI.wetland)
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
@@
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # add column for sub-ecosystem types
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(subtype = substring(ninkartleggingsenheter, 4),
-                                                        subtype = str_remove(subtype, '-'),
-                                                        subtype = str_remove(subtype, '-'))
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% mutate(
+  subtype = substring(ninkartleggingsenheter, 4),
+  subtype = str_remove(subtype, "-"),
+  subtype = str_remove(subtype, "-")
+)
 # we are using max NDVI per year in every NiN polygon
@@
 # join nin.wetland and Modis NDVI data
-ModisNDVI.wetland <- full_join(nin.wetland, df.m, by="id")
+ModisNDVI.wetland <- full_join(nin.wetland, df.m, by = "id")
 
-#summary(ModisNDVI.wetland)
+# summary(ModisNDVI.wetland)
 ModisNDVI.wetland <- ModisNDVI.wetland %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(ModisNDVI.wetland))
-#summary(ModisNDVI.wetland)
+    area_meters = st_area(ModisNDVI.wetland)
+  )
+# summary(ModisNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(!is.na(mean))
-#summary(ModisNDVI.wetland)
+# summary(ModisNDVI.wetland)
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
@@
 ModisNDVI.wetland <- ModisNDVI.wetland %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-#summary(ModisNDVI.wetland)
+  filter(mean == max(mean, na.rm = TRUE))
+# summary(ModisNDVI.wetland)
 
@@
 # join nin.wetland & Landsat NDVI data
-LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by="id")
+LandsatNDVI.wetland <- full_join(nin.wetland, df.l, by = "id")
 
-#summary(LandsatNDVI.wetland)
+# summary(LandsatNDVI.wetland)
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
-  mutate(hovedoekosystem = as.factor(hovedoekosystem),
+  mutate(
+    hovedoekosystem = as.factor(hovedoekosystem),
     hovedtype = as.factor(hovedtype),
@@
     tilstand = as.factor(tilstand),
-         area_meters = st_area(LandsatNDVI.wetland))
-#summary(LandsatNDVI.wetland)
+    area_meters = st_area(LandsatNDVI.wetland)
+  )
+# summary(LandsatNDVI.wetland)
 # get rid of NAs (i.e. NDVI cells that were not in wetland polygons)
@@
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(!is.na(mean))
-#summary(LandsatNDVI.wetland)
+# summary(LandsatNDVI.wetland)
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
@@
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>%
-  dplyr::mutate(year = lubridate::year(date), 
+  dplyr::mutate(
+    year = lubridate::year(date),
     month = lubridate::month(date),
-                day = lubridate::day(date))
+    day = lubridate::day(date)
+  )
 # we are using max NDVI per year in every NiN polygon
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
-#summary(LandsatNDVI.wetland)
+# summary(LandsatNDVI.wetland)
 ```
@@
 
+# Where are the wetland sites?
+```{r}
+tm_shape(regnor) +
+  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
+  tm_borders() +
+  tm_shape(SentinelNDVI.wetland) +
+  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE), scale = 2.5, legend.show = FALSE) +
+  tm_add_legend(
+    type = "symbol",
+    col = tmaptools::get_brewer_pal("YlOrRd", 6, plot = FALSE),
+    labels = c(
+      "V1 - minerotrophic mires",
+      "V3 - ombrotrophic bogs",
+      "V4 - cold springs",
+      "V6 - wet snowbeds",
+      "V9 - semi-natural mires",
+      "V10 - semi-natural wet meadows"
+    ),
+    title = "Major ecosystem type",
+    size = 3
+  ) +
+  tm_layout(
+    main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
+    main.title.size = 1.2, legend.position = c("right", "bottom"),
+    legend.text.size = 1.3, legend.title.size = 1.4
+  )
+```
+
 # How are polygon sizes distributed?
@@
 summary(SentinelNDVI.wetland$area_meters)
-hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
-abline(v=100,lty=2)
+hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000)
+abline(v = 100, lty = 2)
 ```
@@
 SentinelNDVI.wetland %>%
-  ggplot( aes(x=year, y=mean )) + 
+  ggplot(aes(x = year, y = mean)) +
   geom_point() +
-  facet_grid( tilstand~hovedtype)
+  facet_grid(tilstand ~ hovedtype)
 ```
@@
 ```{r}
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
+SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != "2022")
 ```
@@
 SentinelNDVI.wetland %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=hovedtype, y=mean )) + 
+  ggplot(aes(x = hovedtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey")
+  geom_point(size = 0.7, shape = 16, color = "grey")
 
 SentinelNDVI.wetland %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
-  ggplot(aes(x=subtype, y=mean )) + 
+  ggplot(aes(x = subtype, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
 ```
@@
 SentinelNDVI.wetland %>%
-  filter(tilstand == 'God') %>%
+  filter(tilstand == "God") %>%
   filter(year == kartleggingsaar) %>%
   filter(!is.na(region)) %>%
-  ggplot(aes(x=region, y=mean )) + 
+  ggplot(aes(x = region, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_wrap( ~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust=0.2))
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
 ```
@@
 nin.wetland2 %>%
-  filter(hovedtype=='V9') %>%
-  filter(NiN_variable_code %in% c("1AG-B","7GR-GI","1AG-A-G","1AR-C-L","7TK","PRSL")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "1AG-B"="Shrub cover",
-                                 "7GR-GI"="Abundance of ditches",
-                                 "1AG-A-G"="Tree cover",
-                                 "1AR-C-L"="Woody plants in field layer",
-                                 "7TK"="Heavy vehicles",
-                                 "PRSL"="Wear/tear")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "V9") %>%
+  filter(NiN_variable_code %in% c("1AG-B", "7GR-GI", "1AG-A-G", "1AR-C-L", "7TK", "PRSL")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "1AG-B" = "Shrub cover",
+    "7GR-GI" = "Abundance of ditches",
+    "1AG-A-G" = "Tree cover",
+    "1AR-C-L" = "Woody plants in field layer",
+    "7TK" = "Heavy vehicles",
+    "PRSL" = "Wear/tear"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in semi-natural mires (V9)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.wetland2 %>%
-  filter(hovedtype=='V10') %>%
-  filter(NiN_variable_code %in% c("7JB-BA","7RA-SJ","7FA","7JB-GJ")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7JB-BA"="Land use intensity",
-                                 "7RA-SJ"="Succession",
-                                 "7FA"="Alien species",
-                                 "7JB-GJ"="Fertilization")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=2) +
-  theme(legend.position="none") +
+  filter(hovedtype == "V10") %>%
+  filter(NiN_variable_code %in% c("7JB-BA", "7RA-SJ", "7FA", "7JB-GJ")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7JB-BA" = "Land use intensity",
+    "7RA-SJ" = "Succession",
+    "7FA" = "Alien species",
+    "7JB-GJ" = "Fertilization"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 2) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in semi-natural wet-meadows (V10)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 
 nin.wetland2 %>%
-  filter(hovedtype=='V1') %>%
-  filter(NiN_variable_code %in% c("7GR-GI","PRTK","7FA","7TK","7SE","PRSL")) %>%
-  mutate(NiN_variable_des=recode(NiN_variable_code,
-                                 "7GR-GI"="Abundance of ditches",
-                                 "PRTK"="Heavy vehicles",
-                                 "7FA"="Alien species",
-                                 "7TK"="Heavy vehicles2",
-                                 "7SE"="Wear/tear",
-                                 "PRSL"="Wear/tear2")) %>%
-  ggplot(aes(x=NiN_variable_value, y=tilstand_num )) +
-  geom_point(size=2) +  
-  geom_jitter(width = 0.25, height=0.25) +
-  facet_wrap(~NiN_variable_des,nrow=2, ncol=3) +
-  theme(legend.position="none") +
+  filter(hovedtype == "V1") %>%
+  filter(NiN_variable_code %in% c("7GR-GI", "PRTK", "7FA", "7TK", "7SE", "PRSL")) %>%
+  mutate(NiN_variable_des = recode(NiN_variable_code,
+    "7GR-GI" = "Abundance of ditches",
+    "PRTK" = "Heavy vehicles",
+    "7FA" = "Alien species",
+    "7TK" = "Heavy vehicles2",
+    "7SE" = "Wear/tear",
+    "PRSL" = "Wear/tear2"
+  )) %>%
+  ggplot(aes(x = NiN_variable_value, y = tilstand_num)) +
+  geom_point(size = 2) +
+  geom_jitter(width = 0.25, height = 0.25) +
+  facet_wrap(~NiN_variable_des, nrow = 2, ncol = 3) +
+  theme(legend.position = "none") +
   ggtitle("Condition variables in minerotrophic fens (V1)") +
-  labs(y= "Condition level (1-4='good' to 'very reduced')", x = "Condition variable score")
-
+  labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
 ```
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r warning=FALSE, message=FALSE}
+```{r warning=FALSE}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
 SentinelNDVI.wetland %>%
-
   group_by(id, year) %>%
-
   filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
-  facet_grid( region~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey") +
+  facet_grid(region ~ hovedtype)
 ```
@@
   filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'V4')
+  filter(hovedtype != "V4")
 
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
-model.wetland.cond.Sent <- betareg(mean_beta~tilstand_num*region*hovedtype, data=SentinelNDVI.wetland.train)
+model.wetland.cond.Sent <- betareg(mean_beta ~ tilstand_num * region * hovedtype, data = SentinelNDVI.wetland.train)
 model.wetland.cond.Sent <- StepBeta(model.wetland.cond.Sent)
@@
 ```{r message=FALSE, warning=FALSE}
-model.wetland.cond.Sent.N <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Northern Norway",])
-model.wetland.cond.Sent.C <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Central Norway",])
-model.wetland.cond.Sent.W <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Western Norway",])
-model.wetland.cond.Sent.E <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Eastern Norway",])
-model.wetland.cond.Sent.S <- betareg(mean_beta~tilstand_num*hovedtype, data=SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region=="Southern Norway",])
-
+model.wetland.cond.Sent.N <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region == "Northern Norway", ])
+model.wetland.cond.Sent.C <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region == "Central Norway", ])
+model.wetland.cond.Sent.W <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region == "Western Norway", ])
+model.wetland.cond.Sent.E <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region == "Eastern Norway", ])
+model.wetland.cond.Sent.S <- betareg(mean_beta ~ tilstand_num * hovedtype, data = SentinelNDVI.wetland.train[SentinelNDVI.wetland.train$region == "Southern Norway", ])
 ```
@@
 SentinelNDVI.wetland.train %>%
-  filter(region=="Northern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Northern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.wetland.train %>%
-  filter(region=="Central Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Central Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.wetland.train %>%
-  filter(region=="Western Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Western Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.wetland.train %>%
-  filter(region=="Eastern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Eastern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
 SentinelNDVI.wetland.train %>%
-  filter(region=="Southern Norway") %>%
-
-  ggplot( aes(x=tilstand, y=mean )) + 
+  filter(region == "Southern Norway") %>%
+  ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size=0.7, shape=16, color="grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey") +
   facet_wrap(~hovedtype)
@@
   group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
+  filter(mean == max(mean, na.rm = TRUE))
 
 # transformation of NDVI scale to beta scale
-#SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
+# SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
 ModisNDVI.wetland$mean_beta <- (ModisNDVI.wetland$mean + 1) / 2
@@
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  rename('system.index' = 'system:index')
+  dplyr::rename("system.index" = "system:index")
 
@@
 allSatNDVI.wetland <- rbind(
-SentinelNDVI.wetland[,!names(SentinelNDVI.wetland) %in% "subtype"],
+  SentinelNDVI.wetland[, !names(SentinelNDVI.wetland) %in% "subtype"],
   ModisNDVI.wetland,
-LandsatNDVI.wetland[,!names(LandsatNDVI.wetland) %in% "column_label"]
+  LandsatNDVI.wetland[, !names(LandsatNDVI.wetland) %in% "column_label"]
 )
@@
 allSatNDVI.wetland$Sat <- c(
-  rep("Sentinel",nrow(SentinelNDVI.wetland)),
-  rep("Modis",nrow(ModisNDVI.wetland)),
-  rep("Landsat",nrow(LandsatNDVI.wetland))
+  rep("Sentinel", nrow(SentinelNDVI.wetland)),
+  rep("Modis", nrow(ModisNDVI.wetland)),
+  rep("Landsat", nrow(LandsatNDVI.wetland))
 )
-allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat,levels=c("Sentinel","Modis","Landsat"))
+allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat, levels = c("Sentinel", "Modis", "Landsat"))
 levels(allSatNDVI.wetland$Sat)
@@
 allSatNDVI.wetland %>%
-  ggplot( aes(x=year_jit, y=mean, color=Sat )) + 
+  ggplot(aes(x = year_jit, y = mean, color = Sat)) +
   geom_point() +
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-```{r message=FALSE, warning=FALSE}
-model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand=="God",])
-model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$tilstand=="God",])
-model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta~year*hovedtype + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand=="God",])
+```{r warning=FALSE}
+model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand == "God", ])
+model.wetland.time.GodTilst.Modi <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = ModisNDVI.wetland[ModisNDVI.wetland$tilstand == "God", ])
+model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand == "God", ])
 
@@
 Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
-```{r message=FALSE, warning=FALSE}
-model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9","V10"),])
-model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9","V10"),])
-model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta~year*tilstand + (1|id), family=beta_family(),data=LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9","V10"),])
+```{r warning=FALSE}
+model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
+model.wetland.time.tilst.V9V10.Modi <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = ModisNDVI.wetland[ModisNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
+model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
 summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond

### 2023-09-20 NDVI_wetland.Rmd
Change paths for bookdown

@@
 # Import Sentinel NDVI Data
-df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
+df.s <- list.files("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
   map_df(~ fread(.))
@@
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m <- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
+df.m <- read.csv("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv")
 df.m$mean <- df.m$mean / 10000

### 2023-09-26 NDVI_natopen.Rmd
Update NDVI_natopen.Rmd
Fixed typos and more simple rendering errors

@@
-# NDVI (greenness) indicator, naturally open ecosystems below the treeline  {#NDVI-indicator-natopen}
+# NDVI - Naturally Open Ecosystems {#NDVI-indicator-natopen}
+
+NDVI (greenness) indicator, naturally open ecosystems below the treeline
 
@@
 _Author and date:_
+
 Joachim Töpper, Tessa Bargmann
@@
 
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing naturally open ecosystems below the treeline (called 'natopen' hereafter).
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing naturally open ecosystems below the treeline (called _natopen_ hereafter).
 For this task, three initial challenges can be identified early on:
-1) We don't yet have a national map for natopen ecosystems, i.e.
+
+1.
+We don't yet have a national map for natopen ecosystems, i.e.
 we don't know the exact locations of all natopen areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single natopen ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
-3) Natopen ecosystems in Norway are widely different in their physiognomy, ecology, and thus vegetation, spanning from highly disturbed hill slopes with sparse ruderal vegetation to nutrient rich shallow soils with resource demanding plant species.
+2.
+The spatial extent of the single natopen ecosystem occurrences (hereafter _polygons_) includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+3.
+Natopen ecosystems in Norway are widely different in their physiognomy, ecology, and thus vegetation, spanning from highly disturbed hill slopes with sparse ruderal vegetation to nutrient rich shallow soils with resource demanding plant species.
 Thus, large variation in NDVI signals from different natopen ecosystem types has to be expected.
@@
 In this draft development for a NDVI indicator for ecological condition in natopen  systems we approach these challenges in the following ways:
-1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
+
+1.
+We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of natopen ecosystems are available.
-2) We include Sentinel-2 imagery for natopen ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all natopen polygons in the NiN-data.
+2.
+We include Sentinel-2 imagery for natopen ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all natopen polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
@@
 2020) present at mapped sites as wall as their spatial delimitation (in the form of polygons).
-Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
+Since we use NiN data that were collected following ['Miljødirektoratets instruks'](https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/), these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
 Polygon sizes in these NiN data may vary from <100 sqm to 240 000 sqm.
@@
 
-                  T2    T12   T18   T20   T21
-Northern Norway   0     230   648   10    31
-Central Norway    41    282   201   0     0
-Western Norway    49    44    79    0     9
-Eastern Norway    217   84    483   2     51
-Southern Norway   52    155   21    0     73
+|Region|T2|T12|T18|T20|T21|
+Northern Norway |  0  |   230 |  648 |  10 |   31|
+Central Norway  |  41 |   282 |  201 |  0  |   0|
+Western Norway  |  49 |   44  |  79  |  0  |   9|
+Eastern Norway  |  217 |  84  |  483 |  2   |  51|
+Southern Norway |  52  |  155  | 21  |  0   |  73|
 
@@
 The analyses in this document make use of the following data sets:
--NiN data (Naturtyper etter Miljødirektoratets instruks)
--Sentinel-2 NDVI data
--MODIS NDVI data
--Landsat NDVI data
+
+- NiN data (Naturtyper etter Miljødirektoratets instruks)
+- Sentinel-2 NDVI data
+- MODIS NDVI data
+- Landsat NDVI data
 
@@
 ```
+
 ##### Sentinel-2 NDVI data
@@
 
-# Where are the natopen sites?
+**Where are the natopen sites?**
 ```{r}
@@
 
-# How are polygon sizes distributed?
+**How are polygon sizes distributed?**
 ```{r}
@@
 ```
+
 natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.
 
-# How does NDVI vary over the years
+**How does NDVI vary over the years?**
 ```{r}
@@
 ```
+
 2022 stands out with the highest NDVI values missing.
@@
 
-# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
+**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
 ```{r warning=FALSE}
@@
 ```
+
 NDVI varies between major ecosystem types with the highest values in open flood plains (T18) and in natural coastal meadows (T12), followed by naturally open areas on shallow soils (T2) and vegetated sand dunes (T21).
@@
 
-# How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)
+**How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
 ```{r warning=FALSE}
@@
 ```
+
 There is no obvious trends in geography.
 
-# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
+**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
 ```{r warning=FALSE}
@@
 ```
+
 The plots suggest a slight decrease in NDVI as condition deteriorates. Overlaps between condition classes are generally very large.
 #### continue here ####
-# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
+
+**Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
 ```{r warning=FALSE}
@@
 ```
+
 In these plots, the panels for the different major ecosystem types differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
@@
 ```
+
 T20 is has only very few data and is present in only 1 conditino class and 2 regions, we omit it.
@@
 
-Overall take home messages:
+**Overall take home messages:**
 
@@
 ```
+
 It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable.
@@
 
-# 1) We are lacking ecosystem maps for natopen ecosystems
+1.  We are lacking ecosystem maps for natopen ecosystems
+
     The analysis at hand clearly shows that different natopen ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within natopens separately.
-# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in natopen ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems.
-Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like different land-use schemes in the different ecosystem types, are connected.
-This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
+
+2.  We are lacking ecosystem maps for natopen ecosystems
+
+    The analysis at hand clearly shows that different natopen ecosystem types display different greenness levels and thus need to be evaluated separately.
+Thus, we need ecosystem maps, that allow us to treat NDVI values for at least the different major ecosystem types within natopens separately.
 
 ### Eksport file (final product)
+Nothing to export.
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-09-27 slitasje.Rmd
New way to convert range categories into percentages

@@
 Disturbance to soils and vegetation is recorded in the various data sets are NiN variables 7SE and 7TK, or as the similar, but more fine-scaled variable, PRSL and PRTK.
-Effects from domestic animal is not included in the definition of these variables, and the upper reference value is therefore set to 0% disturbance.
+Effects from domestic animals are not included in the definition of these variables, and the upper reference value is therefore set to 0% disturbance.
 
@@
 12. Export indicator maps and regional extrapolated maps
+
+Note that besides the wetland map, we do not have ecosystem delineation maps to complete step 10 in the workflow.
+
+| Variable    | Name |Measurement scale | Description |
+|------       |------|------|-----|
+|7TK          |Spor etter ferdsel med tunge kjøretøy      |A4b      | Det som skal registreres er andelen (tenkte) småruter á 10 × 10 m (100 m2) innenfor en kartfigur som inneholder kjøretøyspor. |
+|7SE          |Spor etter slitasje og slitasjebetinget erosjon      |A4b  | Det som skal registreres er antall (tenkte) småruter á 2 × 2 m (4 m2) som inneholder spor etter slitasje og slitasjebetinget erosjon (slitasjespor).
+|
+|PRTK         |Kjørespor      |A8 | Same as 7TK, but with a finer measurement scale |
+|PRSL         |Slitasje | A8 | Same as 7SE, but with a finer measurement scale |
+
+: Variables used in this composite ADSV indicator
 
@@
 
-I here transfer these ordinal variables over to a continuous scale based on the mean value of each step, but we acknowledge that there are important assumptions being made here, for example that the mean _real_ value for 7TK or 7SE 3 is the mid-point in the range from 0.5 to 1 (i.e.
-0.75).
-This is probably not correct.
-The large range of the steps, particularly for the 7SE and 7TK variables, poses a problem for assessing ecological condition at the polygon scale, but less so at aggregated scales.
-Hence, the indicator should not be presented and interpreted at the polygon scale.
+I here transfer these ordinal variables over to a continuous scale. The data is strongly right skewed, so simply taking the center value of each value range will not work.
+
+```{r naturetypes-histogram-vars, fig.cap="Histogram showing the distribution of the four variables used in this indicator."}
+naturetypes %>% ggplot()+
+  theme_bw() +
+  geom_histogram(aes(x = NiN_variable_value),
+                 binwidth = 1) +
+  facet_wrap(.~NiN_variable_code)
+```
+
+```{r}
+naturetypes %>%
+  as.tibble() %>%
+  group_by(NiN_variable_code)%>%
+  summarise(mean = mean(NiN_variable_value, na.rm=T),
+            median = median(NiN_variable_value, na.rm=T))
+```
+
+The table above shows that the median value for all the variables is zero.
+The most common % value for each range category should then be the lower limit of that category.
+For example, when the category is `1/2 - 1` then the % value is most likely 50%, and not 75% which is the center value for that value range.
+One  issue is when the variable value is 1, because there the lower limit is the same as when the variable value is 0.
+Here I think I must manually assign it to something a bit higher than 0.
+
+You can read more about this issue [here](https://github.com/NINAnor/ecosystemCondition/issues/100).
 
@@
              1 ~ mean(c(0, 1/16))*100,
-             2 ~ mean(c(1/16, 1/2))*100,
-             3 ~ 75),
+             2 ~ 1/16*100,
+             3 ~ 50),
     NiN_variable_code %in% c('var_PRTK', 'var_PRSL') ~ 
@@
              1 ~ 1.5,
-             2 ~ mean(c(3, 6.25)),
-             3 ~ mean(c(6.25, 12.5)),
-             4 ~ mean(c(12.5, 25)),
-             5 ~ mean(c(25, 50)),
-             6 ~ mean(c(50, 75)),
-             7 ~ mean(c(75, 100)))
+             2 ~ 3,
+             3 ~ 6.25,
+             4 ~ 12.5,
+             5 ~ 25,
+             6 ~ 50,
+             7 ~ 75)
     )
@@
 
-```{r slitasje-naturetype, fig.cap="ADSV scores.", fig.width=3}
+```{r slitasje-naturetype, fig.cap="ADSV scores.", fig.width=5}
 temp <- naturetypes_wide %>%
@@
 
-First I extract the mid values for each ordinal range step, like I did previously for the nature type data set.
+First I convert the range categories into percentages, like I did previously for the nature type data set.
 
@@
              1 ~ mean(c(0, 1/16))*100,
-             2 ~ mean(c(1/16, 1/2))*100,
-             3 ~ 75)) %>%
+             2 ~ 1/16*100,
+             3 ~ 50)) %>%
   mutate(var_7SE = case_match(var_7SE,
@@
              1 ~ mean(c(0, 1/16))*100,
-             2 ~ mean(c(1/16, 1/2))*100,
-             3 ~ 75))
+             2 ~ 1/16*100,
+             3 ~ 50))
 ```
@@
 temp <- as.data.frame(table(slitasje_data$slitasje))
-ggplot(temp, aes(x = as.numeric(Var1),
+temp$Var1 <- as.numeric(as.character(temp$Var1))
+ggplot(temp, aes(x = Var1,
                  y = Freq))+

### 2023-09-27 slitasje.Rmd

@@
 
+The indicator is calculated separately for the three ecosystems, but all nature types associated with the same ecosystem are pooled and treated equally.
+This means we assume the nature types that are mapped and which we have data from are, when pooled, representative for the main ecosystem.
+This is less true for Naturally open ecosystems than it is for the other two.
+You can read more about that [here](#naturtype).
+
 Since the data is not sampled in a systematic or random way, we must take extra care not to over-extrapolate in space.
@@
 
+Regarding update frequency, the Nature type data set is not replicated, nor will it be in the future.
+We must therefore trust that the selection criteria for which areas to survey will not change drastically and systematically over time.
+For example, we are now surveying a lot in pressure areas, meaning lowland areas close to human population centers.
+If we in the future allocate more time to surveying wilderness areas, then the indicator becomes biased.
+On the other hand, ANO is replicated every 5 years.
+
 ### Aditional comments about the dataset
@@
 point is affected (the variable is frequency-driven, and not
-amount-driven). The threshold for good ecosystem condition is set to 10%
-damage.
+amount-driven).
+
+The threshold for good ecosystem condition is set to 10%
+damage. This is an expert judgement, and a rather weak one at that. 
+In the future we should try to empirically validate the dose-response relationship,
+for example by comparing different variable scores with erosion rates or the potential for plant growth.
+This would need to be done specifically for each nature type.
 
@@
 ```{r naturetypes-histogram-vars, fig.cap="Histogram showing the distribution of the four variables used in this indicator."}
-naturetypes %>% ggplot()+
+naturetypes %>% 
+  ggplot()+
   theme_bw() +
@@
 
-```{r}
-naturetypes %>%
-  as.tibble() %>%
-  group_by(NiN_variable_code)%>%
-  summarise(mean = mean(NiN_variable_value, na.rm=T),
-            median = median(NiN_variable_value, na.rm=T))
-```
-
-The table above shows that the median value for all the variables is zero.
+The table below shows that, if I assume there is some ADSV and therefore exlude case with variable value <1, the median value for all the variables is 1.
 The most common % value for each range category should then be the lower limit of that category.
@@
 Here I think I must manually assign it to something a bit higher than 0.
+```{r}
+naturetypes %>%
+  as.tibble() %>%
+  filter(NiN_variable_value >0) %>%
+  group_by(NiN_variable_code)%>%
+  summarise(mean = mean(NiN_variable_value, na.rm=T),
+            median = median(NiN_variable_value, na.rm=T))
+```
 
@@
 
+This scaling of the variable implies that it very difficult to measure a location when it
+is in its most degraded form (100% disturbed), since the highest % value one can use is 75% for PRSL and PRTK and just 50% for the other two. 
+This is a problem with the original data resolution. 
+But I will compensate somewhat for this by
+using a non-linear transformation.
+
 I will create a new data table where I calculate the new
@@
 
-It appears most localities are in good condition. Note also that we can get >100% for the unscaled variable.
+It appears most localities are in good condition. Note also that we theoretically can get >100% for the unscaled variable.
 
@@
 ```
-
-This implies that it is impossible to detect or measure a state when it
-is in its most degraded form (100% disturbed). This is a problem with
-the original data resolution. But I will compensate somewhat for this by
-using a non-linear transformation.
 

### 2023-09-28 functional_plant_indicators_natopen.Rmd
Shorten chapter name and list indicator names #77

@@
-# Functional plant indicators, naturally open ecosystems below the tree line (Grime's CSR values, Light, Nitrogen, Soil disturbance) {#Functional-plant-indicators-open}
-
+# Functional Plant Indicators - Naturally Open Ecosystems {#Functional-plant-indicators-open}
+
+<br />
+
+Norwegian name: **Planteindikatorer**
 <br />
@@
 ```
+
+Indicators described in this chapter:
+
+- Grime's CSR values
+- Light
+- Nitrogen
+- Soil disturbance
 

### 2023-09-28 climate_indicators_explained.Rmd
Using a template for the indicator chapters causes repeating headers and duplicate anchors #39

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse) 
+library(tidyverse) 
 library(stars)
@@
 
-## Introduction
+## Introduction {#intro-cli}
 
@@
 
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-vals-cli}
 
@@
 
-## Analyses
+## Analyses {#analyses-cli}
 
@@
 
-#### Regions
+#### Regions {#regions-cli}
 

### 2023-09-28 alien_species.Rmd

@@
 # Alien species {#alien-species}
+
+Norwegian name: **Fravær av fremmede arter**
 
@@
 
-## Introduction
+## Introduction {#intro-ali}
 This indicator represent the the ecological on-site effect from  alien species.
@@
 
-## About the underlying data
+## About the underlying data {#about-data-ali}
 
@@
 
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-vals-ali}
 
@@
 
-## Analyses
+## Analyses {analyses-ali}
 
@@
 
-#### Validate
+#### Validate {#validate-ali}
 

### 2023-09-28 functional_plant_indicators_natopen.Rmd

@@
 
-## Introduction
+## Introduction {#intro-fpin}
 Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-vals-fpin}
 In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al.
@@
 
-## Analyses
+## Analyses {#analyses-fpin}
 ### Data sets
@@
 
-### Eksport file (final product)
+### Eksport file (final product) {#exp-fpin}
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-09-28 functional_plant_indicators_wetland.Rmd
Shorten chapter name and list indicator names #77

@@
-# Functional plant indicators (Moisture, Light, pH, Nitrogen) {#Functional-plant-indicators-wetland}
+# Functional Plant Indicators - Wetlands {#Functional-plant-indicators-wetland}
 
 <br />
+
+Norwegian name: **Planteindikatorer**
 
@@
 
+Indicators described in this chaper:
+
+- Moisture
+- Light
+- pH
+- Nitrogen
+
 <!-- Don't remove these three html lines -->
@@
 
-## Introduction
+## Introduction {#intro-fpiw}
 Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-values-fpiw}
 In this analysis, we derive scaling values from statistical (here, non-parametric) distributions (see Jakobsson et al.
@@
 
-## Analyses
+## Analyses {#analyses-fpiw}
 ### Data sets
@@
 
-### Eksport file (final product)
+### Eksport file (final product) {#exp-fpiw}
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-09-28 gjengroing.Rmd
RenderBook12092023
#75 shorten indicator name

@@
-# Vegetation regrowth/encroachment (gjengroing) {#gjengroing}
-
+# (PART\*) INDICATORS{-}
+
+# Enchroachment {#gjengroing}
 <br />
+
+Norwegian name: **Gjengroing**
+
+_Vegetation regrowth/encroachment_
 
@@
 
-## Introduction
+## Introduction {#intro-enc}
 
@@
 
-## About the underlying data
+## About the underlying data {#about-data-enc}
 
@@
 
-### Representativity in time and space
+### Representativity in time and space {#rep-enc}
 
@@
 
-### Original units
+### Original units {#units-enc}
 
@@
 
-### Temporal coverage
+### Temporal coverage {#temp-enc}
 
@@
 
-### Additional comments about the dataset
+### Additional comments about the dataset {#comm-enc}
 
@@
 
-### Reference state
+### Reference state {#state-enc}
 
@@
 
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}
 
@@
 
-## Uncertainties
+## Uncertainties {#err-enc}
 
@@
 
-## Analyses
-
-### Data sets
+## Analyses {#analyses-enc}
+
+### Data sets {#data-enc}
 
@@
 
-#### Regions
+#### Regions {#reg-enc}
 
@@
 
-### Eksport file (final product)
+### Eksport file (final product) {#exp-enc}
 

### 2023-09-28 homogeneous_impact_areas.Rmd

--- 2023-02-02 DRAFT_homogeneous_impact_areas.Rmd
@@
 library(DT)
-library(tidyverse)
+#library(tidyverse)
+library(ggplot2)
+library(dplyr)
 library(units)
@@
 
-## Introduction
+## Introduction {#intro-hia}
 
@@
 The map is produced by binning values in the infrastructure index into four discrete categories.
-It is likely to be a good predictor for some indicators, such as *slitasje* and the presence of *alien species*.
+It is likely to be a good predictor for some indicators, such as [*slitasje*](#slitasje) and the presence of [*alien species*](#alien-species).
 
@@
 
-## Export
+## Export {#exp-hia}
 I will write this data to the NINA P server.

### 2023-09-28 index.Rmd
Fixed libpath issue

--- 2023-04-14 index.Rmd
@@
 
-Kolstad, A.L., Topper, J. Grainger, M.J., Sandvik, H. Indicators for Ecosystem Condition in Norway. https://ninanor.github.io/ecosystemCondition/index.html.doi [TBA]
+Kolstad, A.L., Topper, J.
+Grainger, M.J., Sandvik, H., Venter, Z., Töpper, J., Speed, J.D.M, Nater, C.R., Evju, M., Åström, J.
+(2023) Indicators for Ecosystem Condition in Norway.
+https://doi.org/10.5281/zenodo.7854835
 
-## License
+**Contributions**
+The list of authors include code developers and reviewers.
+Not all authors interacted with GitHub through their own user, but for those that did, theiy contributions can be tracked, for example [here](https://github.com/NINAnor/ecosystemCondition/graphs/contributors).
+
+## License {.unnumbered}
 

### 2023-09-28 alien_species.Rmd
Update threhold value #38
Change _Økologisk egenskap_ assignment #86

@@
 
-March 2023
+September 2023
 
@@
 Ecosystem <- "Våtmark, Naturlig åpne områder under skoggrensa, Semi-naturlig mark" 
-Egenskap  <- "Funksjonelt viktige arter og strukturer"
+Egenskap  <- "Biologisk mangfold"
 ECT       <- "B1 Compositional state characteristics" 
@@
 Upper reference value is then 1, and lower reference value is 7, on the R7 scale.
-For the threshold value I will set this to value 4 on the R7 scale. 
+For the threshold value I will set this to value 3 on the R7 scale. 
 This is based on interpretation by the author and based of the definition of this value in Fig. \@ref(fig:R7). 
-Step 4 on the R7 scale is the step that, concordantly, best matches the 5% threshold in the old approach, as seen from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
+Step 3 on the R7 scale is the step that, concordantly, best matches the 5% threshold in the old approach, as seen from Fig. \@ref(fig:fremmedart-prosentVSinnslag).
 
@@
 The indicator is tagged to the *Økologisk egenskap* called
-**Funksjonelt viktige arter og strukturer** (Functionally important species and structures).
+**Biologisk mangfold**. 
 This emphasises the negative effect that alien species have on for example supressing or excluding native species. 
-In some cases, alien species can also affect nutrient cycling or recruitment.
+In some cases, alien species can also affect nutrient cycling or recruitment and could then also be appropriate under _funksjonelt viktige arter og strukturer_.
 
@@
 
-* Threshold = 4 (R7 scale), corresponding to about 5-30% alien species cover.
+* Threshold = 3 (R7 scale), corresponding to about 2-5% alien species cover.
 
 * Lower = 7 (R7 scale), corresponding to 100% alien species cover.
+
+For a discussion around the reference values, see [here](https://github.com/NINAnor/ecosystemCondition/issues/38)
 
@@
 high <- 7
-threshold <- 4
+threshold <- 3
 ```

### 2023-09-28 alien_species.Rmd
Specify that indicators are unique also to each ecosystem #83

@@
 
+The indicator is calculated separately for the three ecosystems, but all nature types associated with the same ecosystem are pooled and treated equally.
+This means we assume the nature types that are mapped and which we have data from are, when pooled, representative for the main ecosystem.
+This is less true for Naturally open ecosystems than it is for the other two.
+You can read more about that [here](#naturtype).
+
 ```{r R7, fig.cap="The R7 scale used when recording the effect of alien species in the nature type monitoring program"}
@@
 The nature type mapping is not random and cannot be said to be area
-representative. The ANO data set however, is area representative. The
+representative. 
+This will not improve over time either.
+
+The ANO data set however, is area representative. The
 data is from 2018 to present. The data from one field season usually
 becomes available early the following year.
+
+Regarding update frequency, the Nature type data set is not replicated, nor will it be in the future.
+We must therefore trust that the selection criteria for which areas to survey will not change drastically and systematically over time.
+For example, we are now surveying a lot in pressure areas, meaning lowland areas close to human population centers.
+If we in the future allocate more time to surveying wilderness areas, then the indicator becomes biased.
+On the other hand, ANO is replicated every 5 years.
 

### 2023-09-28 alien_species.Rmd
Reasoning for calculating HIA-values at municipality level #82
Investiagte how the R7 scale is used and fix figure text #81

@@
 
-```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with percentage cover (y-axis), which is how alien species is recorded in ANO.
+```{r fremmedart-prosentVSinnslag, fig.cap="Comparing the R7 scale (x axis) which is used in the nature type monitoring program with percentage cover (y-axis), which is how alien species is recorded in GRUK.
 I am assuming the x axis is shifted downwards compared to the original R7 scale which range from 1 to 7, so that a value of 0 here equals an R7 value of 1.
+Note also that this relationship is based on limited data from just one nature type and in smaller diameter circles than the other data we use in this indicator.
 Cropped from Evju et al (2022)."}
@@
 
-I also want to split these four HIA classes based on municipality. To do
-this I need the two layers to have the same CRS.
+I also want to split these four HIA classes based on municipality. 
+For a [related indicator](#HIA-slitasje) I used combinations of HIA and region, but for alien species I think
+that there is too much variation also inside each region (esp north-south) so that we need a finer division of strata.
+To do this I need the two layers to have the same CRS.
 

### 2023-09-28 alien_species.Rmd
Highlight the focus on vascular plants #80
Use ANO and GRUK data sets #79

@@
 This indicator represent the the ecological on-site effect from  alien species.
+In practice, the indicator is restricted to alien vascular plants.
 Whilst the introduction of alien species is best considered a pressure variable, the relative abundance of alien species on a given location is seen here as a manifestation of this negative anthropogenic impact, and is therefore a valid condition indicator.
@@
 
-_NOTE: I'm not sure, but I wonder if the 2021 GRUK data was a pilot project and that most GRUK data, and all future GRUK data, will not be included in this data set._
+See also [this GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/79).
 

### 2023-09-29 NDVI_seminat.Rmd
cut short friday

@@
-# NDVI (greenness) indicator, semi-natural ecosystems  {#NDVI-indicator-seminat}
+---
+editor_options: 
+  markdown: 
+    wrap: 72
+---
+
+# NDVI- Semi-natural Ecosystems {#NDVI-indicator-seminat}
+
+<br /> Norwegian name: **Primærproduksjon** <br />
+
+*Author and date:*
+
+Joachim Töpper, Tessa Bargmann
+
+September 2023
 
@@
 
-_Author and date:_
-Joachim Töpper, Tessa Bargmann
-
-September 2023
-
-<br />
-
 <!-- Load all you dependencies here -->
+
 ```{r setup, include=FALSE}
@@
 <!-- Don't remove these three html lines -->
-<br />
-<br />
+
+<br /> <br />
+
 <hr />
 
-## Introduction
-The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area.
-Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance.
-Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem.
-E.g.
-lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems.
-In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied.
-For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al.
-2020).
-For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021).
-In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e.
-national maps for the forest and mountain ecosystems).
-
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems. For this task, two initial challenges can be identified early on:
-1) We don't yet have a national map for semi-natural ecosystems, i.e.
-we don't know the exact locations of all semi-natural areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single semi-natural ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
-
-In this draft development for a NDVI indicator for ecological condition in semi-natural  systems we approach these challenges in the following ways:
-1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
-However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
-Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of semi-natural ecosystems are available.
-2) We include Sentinel-2 imagery for semi-natural ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all semi-natural polygons in the NiN-data.
-However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
-In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurements.
-We also still explore NDVI data from MODIS (see more details below).
-
-## About the underlying data
-In the NDVI project for semi-natural systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
-
-- NiN ecosystem data from 'Naturtypekartlegging etter Miljødirektoratets instruks' to guide extraction of NDVI data for the right locations and to provide information on the field-assessed ecological condition of mapped polygons
+## Introduction {#intro-ndvis}
+
+The Normalized difference vegetation index (NDVI) can be used to
+describe the greenness of an area, and may thus represent a proxy for
+the amount of chlorophyll and plant productivity in the given area.
+Different ecosystem types naturally have greenness signals of different
+intensity but also with a certain variance. Deviations beyond the
+natural variation in either direction may indicate a reduction in
+ecological condition of a given ecosystem. E.g. lower than normal NDVI
+may indicate excessive disturbance, while higher than normal NDVI may
+indicate regrowth and transitions towards more densely vegetated
+systems. In recent assessments of ecological condition for forest and
+mountain ecosystems, two different NDVI-based indicators were developed
+and applied. For forests, NDVI-values across the country were compared
+to a spatial reference NDVI-signal from protected areas (Framstad et al.
+2020). For mountains, the extent and consistence of the temporal change
+of NDVI-values over the last twenty years was assessed (Framstad et al
+2021). In both cases, the NDVI-product from the MODIS satellite
+reflectance imagery was used based on available maps of spatial
+ecosystem extent in Norway (i.e. national maps for the forest and
+mountain ecosystems).
+
+Here, we explore the potential for developing an NDVI-indicator for
+ecological condition representing open semi-natural ecosystems. For this
+task, two initial challenges can be identified early on:
+
+1.  We don't yet have a national map for semi-natural ecosystems, i.e.
+    we don't know the exact locations of all semi-natural areas in
+    Norway, which is a prerequisite to extract satellite data for the
+    right locations and ecosystems, and thus achieve a good and unbiased
+    spatial representation for the assessment area.
+2.  The spatial extent of the single semi-natural ecosystem occurrences
+    (hereafter 'polygons') includes many areas which are way smaller
+    than the pixel size in the MODIS imagery (250 x 250 m) which was
+    used for forests and mountains and has a time series back to year
+    2000.
+
+In this draft development for a NDVI indicator for ecological condition
+in semi-natural systems we approach these challenges in the following
+ways:
+
+1.  We use ecosystem occurrence data from NiN-mapping to guide the
+    extraction of satellite NDVI data for pixels representing the right
+    ecosystem. However, the ecosystem occurrences in the NiN data are
+    highly spatially biased and thus cannot provide a good and unbiased
+    spatial representation for the assessment area. Our work therefore
+    represents a concept and method development which may be applied in
+    an assessment of ecological condition once spatially unbiased
+    occurrence data or a national map of semi-natural ecosystems are
+    available.
+2.  We include Sentinel-2 imagery for semi-natural ecosystems as the
+    resolution for the bands underlying the NDVI product here is 10 x 10
+    m which is smaller than almost all semi-natural polygons in the
+    NiN-data. However, since Sentinel-2 only goes back to 2015, we lose
+    much of the time series aspect as compared to the MODIS data. In
+    addition, we include Landsat imagery which goes back to 1984 and has
+    a resolution of 30 x 30 m, but also has more variable data due to
+    less frequent seasonal measurements. We also still explore NDVI data
+    from MODIS (see more details below).
+
+## About the underlying data {#underlying-data-ndvis}
+
+In the NDVI project for semi-natural systems, we use two kinds of data
+(coming as four datasets) for building indicators for ecological
+condition:
+
+-   NiN ecosystem data from 'Naturtypekartlegging etter
+    Miljødirektoratets instruks' to guide extraction of NDVI data for
+    the right locations and to provide information on the field-assessed
+    ecological condition of mapped polygons
 -   remote sensed NDVI data
@@
 
-(1) NiN ecosystem data:
-The NiN data in this study contain data about the ecosystem categories (cf.
-Halvorsen et al.
-2020) present at mapped sites as wall as their spatial delimitation (in the form of polygons).
-Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
-Polygon sizes in these NiN data may vary from <100 sqm to the dimension of sqkm's.
-
-(2) NDVI data:
-The NDVI data used in the work at hand are pixel means for polygons in the NiN-data, i.e.
-each NiN polygon is represented by one NDVI value per measurement time (see below for more details on the temporal resolution of NDVI data).
-With respect to that, it does only make sense to use NDVI data for polygons that are at least of the same size as the resolution of the respective satellite, and thus we excluded any polygons smaller than each satellite's resolution.
-
-(2a) Sentinel-2 satellite NDVI data:
-The Sentinel data have a resolution of 10 x 10 m, and we thus excluded all NiN-polygons smaller than 100 sqm.
-Sentinel-2 has a revisit frequency of 10 days and should thus provide enough measurements during peak growing season to provide a robust measure of maximum greenness.
-
-(2b) MODIS satellite NDVI data:
-The MODIS data have a resolution of 250 x 250 m, and we thus excluded all NiN-polygons smaller than 62500 sqm.
-MODIS has a revisit frequency of 1-2 days and thus provides an extensive sample to provide a robust measure of maximum greenness.
-
-(2c) Landsat satellite NDVI data:
-The Landsat data have a resolution of 30 x 30 m, and we thus excluded all NiN-polygons smaller than 900 sqm.
-Landsat has a revisit frequency of 16 days and thus provides the poorest growing season sample among the available NDVI data for a robust measure of maximum greenness.
-
-### Representativity in time and space
-For semi-natural systems, the NiN data contain 24494 semi-natural polygons of known area size across mainland Norway.
-However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased.
-The 24494 semi-natural polygons are distributed across regions and major ecosystem types in the following way:
-
-                  T31   T32   T33   T34   T40   T41
-Northern Norway   1194  2081  14    1     15    0
-Central Norway    641   2518  210   2338  97    0
-Western Norway    132   2922  94    2200  66    54
-Eastern Norway    1064  4488  10    10    371   54
-Southern Norway   234   2317  112   749   111   140
-
-In addition, there is 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40 that are not assigned to any region.
-
-The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the semi-natural occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
-
-### Temporal coverage
-The currently available NiN data on semi-natural ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
-Note that none of the polygons in the NiN-data has been visited more than once during that period.
+(1) NiN ecosystem data: The NiN data in this study contain data about
+    the ecosystem categories (cf. Halvorsen et al. 2020) present at
+    mapped sites as wall as their spatial delimitation (in the form of
+    polygons). Since we use NiN data that were collected following
+    'Miljødirektoratets instruks', these data also include information
+    on the ecological condition and underlying condition variabels of
+    the mapped sites (see the Miljødirektoratets mapping manual for
+    details). Polygon sizes in these NiN data may vary from \<100 sqm to
+    the dimension of sqkm's.
+
+(2) NDVI data: The NDVI data used in the work at hand are pixel means
+    for polygons in the NiN-data, i.e. each NiN polygon is represented
+    by one NDVI value per measurement time (see below for more details
+    on the temporal resolution of NDVI data). With respect to that, it
+    does only make sense to use NDVI data for polygons that are at least
+    of the same size as the resolution of the respective satellite, and
+    thus we excluded any polygons smaller than each satellite's
+    resolution.
+
+(2a) Sentinel-2 satellite NDVI data: The Sentinel data have a resolution
+of 10 x 10 m, and we thus excluded all NiN-polygons smaller than 100
+sqm. Sentinel-2 has a revisit frequency of 10 days and should thus
+provide enough measurements during peak growing season to provide a
+robust measure of maximum greenness.
+
+(2b) MODIS satellite NDVI data: The MODIS data have a resolution of 250
+x 250 m, and we thus excluded all NiN-polygons smaller than 62500 sqm.
+MODIS has a revisit frequency of 1-2 days and thus provides an extensive
+sample to provide a robust measure of maximum greenness.
+
+(2c) Landsat satellite NDVI data: The Landsat data have a resolution of
+30 x 30 m, and we thus excluded all NiN-polygons smaller than 900 sqm.
+Landsat has a revisit frequency of 16 days and thus provides the poorest
+growing season sample among the available NDVI data for a robust measure
+of maximum greenness.
+
+### Representativity in time and space {#representativity-ndvis}
+
+For semi-natural systems, the NiN data contain 24494 semi-natural
+polygons of known area size across mainland Norway. However, NiN-mapping
+is not planned and performed in a spatially representative way and thus
+both the ecosystem occurrences and their spatial extent have to be
+treated as spatially biased. The 24494 semi-natural polygons are
+distributed across regions and major ecosystem types in the following
+way:
+
+| Region          | T31  | T32  | T33 | T34  | T40 |
+|-----------------|------|------|-----|------|-----|
+| Northern Norway | 1194 | 2081 | 14  | 1    | 15  |
+| Central Norway  | 641  | 2518 | 210 | 2338 | 97  |
+| Western Norway  | 132  | 2922 | 94  | 2200 | 66  |
+| Eastern Norway  | 1064 | 4488 | 10  | 10   | 371 |
+| Southern Norway | 234  | 2317 | 112 | 749  | 111 |
+
+In addition, there are 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40
+that are not assigned to any region.
+
+The NDVI data from either satellite follow the spatial distribution and
+bias status of the NiN-data. Due to the larger number of excluded
+polygons in the NDVI data from MODIS (because of the large pixel size in
+MODIS, see above), the MODIS NDVI data for the semi-natural occurrences
+in the NiN data are likely even more spatially biased than it is the
+case for the NDVI data from Sentinel-2 and Landsat.
+
+### Temporal coverage {#temp-cov-ndvis}
+
+The currently available NiN data on semi-natural ecosystems span a time
+period from 2018-2021 and thus represent a contemporary sample of
+ecosystem occurrences. Note that none of the polygons in the NiN-data
+has been visited more than once during that period.
 
 Available Sentinel-2 satellite NDVI data span the years 2015-2022.
+
 Available MODIS satellite NDVI data span the years 2000-2022.
+
 Available Landsat satellite NDVI data span the years 1984-2021.
 
-## Collinearities with other indicators
-The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in abandoned semi-natural areas under shrub and tree encroachment.
-
-## Reference state and values
-### Reference state and scaling values
-This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
-Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for semi-natural ecosystems (cf.
-Jakobsson et al.
-2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to semi-natural ecology and condition and (2) its applicability for defining a reference state.
-We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
-
-## Uncertainties
-Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
+## Collinearities with other indicators {#coll-ndvis}
+
+The NDVI signal may likely be co-linear with [the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html)
+based on LiDAR derived canopy height in abandoned semi-natural areas
+under shrub and tree encroachment.
+
+## Reference state and values {#ref-state-ndvis}
+
+### Reference state and scaling values {#ref-state2-ndvis}
+
+This indicator project is mainly exploratory, mapping the possibilities
+of and lacking prerequisites for satellite derived NDVI data to inform
+ecological condition assessments. Therefore, we lack a concise strategy
+for how to define the reference state, especially prior to having access
+to an ecosystem map for semi-natural ecosystems (cf. Jakobsson et al.
+2020, Töpper & Jakobsson 2021). In this exploratory work, we will use
+the condition classification in the NiN-data to explore (1) how NDVI
+connects to semi-natural ecology and condition and (2) its applicability
+for defining a reference state. We will use statistical models to test
+whether or not NDVI varies systematically with condition and across
+time.
+
+## Uncertainties {#uncertainties-ndvis}
+
+Given a condition index can be achieved, we can calculate a mean
+indicator value (after scaling) for every region (or any other delimited
+area of interest) as well as its corresponding standard error and
+standard deviation as a measure of spatial uncertainty for a
+geographical area (see Töpper & Jakobsson 2021).
 
@@
 
-Framstad, E., Kolstad, A.
-L., Nybø, S., Töpper, J.
-& Vandvik, V.
-2022.
-The condition of forest and mountain ecosystems in Norway.
-Assessment by the IBECA method.
-NINA Report 2100.
+Framstad, E., Kolstad, A. L., Nybø, S., Töpper, J. & Vandvik, V. 2022.
+The condition of forest and mountain ecosystems in Norway. Assessment by
+the IBECA method. NINA Report 2100. Norwegian Institute for Nature
+Research.
+
+Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L.,
+Simensen, T., & Lieungh, E. (2020). Towards a systematics of
+ecodiversity: The EcoSyst framework. Global Ecology and Biogeography,
+29(11), 1887-1906. <doi:10.1111/geb.13164>
+
+Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A.,
+Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik. V., Velle,
+L.G., Aarrestad, P.A. & Nybø, S. 2020. Setting reference levels and
+limits for good ecological condition in terrestrial ecosystems. Insights
+from a case study based on the IBECA approach. Ecological Indicators
+116: 106492.
+
+Miljødiretoratets mapping manual:
+<https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/>
+
+Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition
+Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967.
 Norwegian Institute for Nature Research.
 
-Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E.
-(2020).
-Towards a systematics of ecodiversity: The EcoSyst framework.
-Global Ecology and Biogeography, 29(11), 1887-1906.
-doi:10.1111/geb.13164
-
-Jakobsson, S., Töpper, J.P., Evju, M., Framstad, E., Lyngstad, A., Pedersen, B., Sickel, H., Sverdrup-Thygeson, A., Vandvik.
-V., Velle, L.G., Aarrestad, P.A.
-& Nybø, S.
-2020.
-Setting reference levels and limits for good ecological condition in terrestrial ecosystems.
-Insights from a case study based on the IBECA approach.
-Ecological Indicators 116: 106492.
-
-Miljødiretoratets mapping manual: https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/
-
-Töpper, J. & Jakobsson, S. 2021. The Index-Based Ecological Condition Assessment (IBECA) - Technical protocol, version 1.0. NINA Report 1967. Norwegian Institute for Nature Research.
-
-## Analyses
-### Data sets
+## Analyses {#analyses-ndvis}
+
+### Data sets {#datasets-ndvis}
+
 The analyses in this document make use of the following data sets:
--NiN data (Naturtyper etter Miljødirektoratets instruks)
--Sentinel-2 NDVI data
--MODIS NDVI data
--Landsat NDVI data
+
+- NiN data (Naturtyper etter Miljødirektoratets instruks) 
+- Sentinel-2 NDVI data 
+- MODIS NDVI data 
+- Landsat NDVI data
 
 #### NiN data and geometry for Norway and the five regions
-We read the NiN data for this document from a cache at the GitHub repository...
+
+We read the NiN data for this document from a cache at the GitHub
+repository...
+
 ```{r}
@@
 
-...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
+...but it can also be directly downloaded from Miljødirektoratets
+kartkatalog (<https://kartkatalog.miljodirektoratet.no/Dataset>)
+
 ```{r, echo = T, eval = F}
@@
 ```
-And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
+
+And we upload maps for Norway and the five regions: Southern, Western,
+Eastern, Central, and Northern Norway
+
 ```{r, message=FALSE}
@@
 #### NDVI data
-All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
+
+All NDVI data are calculated mean NDVI values for each respective NiN
+polygon from each available Sentinel-2, MODIS, or Landsat image in
+Google Earth Engine (GEE).
+
+<!-- RUNTIME 30 sec -->
 ```{r, echo = F}
@@
 ```
+
 ##### Sentinel-2 NDVI data
-The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
-The image collection contains Sentinel imagery from June, July and August 2015-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+
+The GEE code can be seen here:
+<https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4>
+The image collection contains Sentinel imagery from June, July and
+August 2015-2022 To not exceed GEE memory limits, the exported files had
+to be iterated over a grid which resulted in 42 separate csv files. This
+script merges them and then merges the dataframe to the NiN data.
+
 ```{r, echo = T, eval = F}
@@
 ##### MODIS NDVI data
-The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
-The image collection contains MODIS imagery from June, July and August 2000-2022
+
+The GEE code can be seen here:
+<https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84>
+The image collection contains MODIS imagery from June, July and August
+2000-2022
 
@@
 ##### Landsat NDVI data
-The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
-The image collection contains Landsat imagery from June, July and August 1984-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+
+The GEE code can be seen here:
+<https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7>
+The image collection contains Landsat imagery from June, July and August
+1984-2022 To not exceed GEE memory limits, the exported files had to be
+iterated over a grid which resulted in 42 separate csv files. This
+script merges them and then merges the dataframe to the NiN data.
+
 ```{r, echo = T, eval = F}
@@
 
-### Data handling
-#### NiN data
+### Data handling {#data-handling-ndvis}
+
+#### NiN data {#nin-data-ndvis}
+
 ```{r, echo = T, eval = F}
@@
 
+<!-- RUNTIME 2 sec -->
 ```{r, echo = F}
@@
 
-#### NDVI data
-```{r, results='hide', message=FALSE}
+#### NDVI data {#ndvi-data-ndvis}
+
+<!-- RUNTIME 4 min. Caching it. -->
+```{r, results='hide', message=FALSE, eval=F}
 ## Sentinel-2
@@
 
+```{r cahing data ndvis, eval=FALSE, include=F}
+saveRDS(LandsatNDVI.seminat, 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/LandsatNDVI.seminat.rds")
+
+saveRDS(ModisNDVI.seminat, 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/ModisNDVI.seminat.rds")
+
+saveRDS(SentinelNDVI.seminat, 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/SentinelNDVI.seminat.rds")
+```
+
+```{r, echo=F}
+LandsatNDVI.seminat <- readRDS(
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/LandsatNDVI.seminat.rds")
+
+ModisNDVI.seminat <- readRDS( 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/ModisNDVI.seminat.rds")
+
+SentinelNDVI.seminat <- readRDS( 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/SentinelNDVI.seminat.rds")
+```
+
 ### Exploratory analyses, Sentinel-2
 
-# Where are the seminat sites?
-```{r}
+**Where are the seminat sites?**
+
+```{r where-are-seminat-sites, }
 tm_shape(regnor) +
@@
 
-# How are polygon sizes distributed?
+**How are polygon sizes distributed?**
+
 ```{r}
@@
 ```
-semi-natural polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.
-
-# How does NDVI vary over the years
+
+semi-natural polygons stretch a large number of sizes from just above
+100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
+sqm.
+
+**How does NDVI vary over the years**
+
 ```{r}
@@
 ```
-2022 stands out with the highest NDVI values missing.
-WE omit 2022 for this analysis
+
+2022 stands out with the highest NDVI values missing. WE omit 2022 for
+this analysis
+
 ```{r}
@@
 
-# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
+**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
+
 ```{r warning=FALSE}
@@
 ```
-NDVI varies between major ecosystem types with the highest values in semi-natural grasslands (T32) and in previously intensively used and altered grasslands that today resemble  a semi-natural grassland (T40, T41), lowest values in coastal heathlands (T34).
-Other ecossytem types: Boreal heath (T31), semi-natural coastal grasslands (T33).
-The overlaps between the ecosystem types are considerable.
-With respect to basic ecosystem types within the major types, NDVI appears to cover very similar ranges across the subtypes, and those with deviating violin plots have a smaller sample size.
-There is no obvious trend related to the structuring environmental gradients withing the major types, and also here the overlap is large.
-
-# How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)
+
+NDVI varies between major ecosystem types with the highest values in
+semi-natural grasslands (T32) and in previously intensively used and
+altered grasslands that today resemble a semi-natural grassland (T40,
+T41), lowest values in coastal heathlands (T34). Other ecosystem types:
+Boreal heath (T31), semi-natural coastal grasslands (T33). The overlaps
+between the ecosystem types are considerable. With respect to basic
+ecosystem types within the major types, NDVI appears to cover very
+similar ranges across the subtypes, and those with deviating violin
+plots have a smaller sample size. There is no obvious trend related to
+the structuring environmental gradients withing the major types, and
+also here the overlap is large.
+
+**How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
+
 ```{r warning=FALSE}
@@
 ```
-For the meadow systems (T32, T33, T40, T41) there is no obvious trends in geography.
-In the heath systems (T31, T34) the distribution of NDVI values in Central Norway is somewhat lower than in the other regions.
-
-# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
+
+For the meadow systems (T32, T33, T40, T41) there is no obvious trends
+in geography. In the heath systems (T31, T34) the distribution of NDVI
+values in Central Norway is somewhat lower than in the other regions.
+
+**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)**
+
 ```{r warning=FALSE}
@@
 ```
-In the coastal types (T33, T34, coastal meadows and heathlands, respectively) NDVI seems to increase as condition deteriorates, but in the remaining types such a pattern is absent.
-Overlaps between condition classes are generally very large.
-
-# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
+
+In the coastal types (T33, T34, coastal meadows and heathlands,
+respectively) NDVI seems to increase as condition deteriorates, but in
+the remaining types such a pattern is absent. Overlaps between condition
+classes are generally very large.
+
+**Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
+
 ```{r warning=FALSE}
@@
 ```
-In these plots, the panels for the different major ecosystem types differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
-
-In the heath systems - T31 and T34 - reduced condition is mainly related to low levels of grazing and ongoing succession (but also too intensive grazing).
-In the meadow systems - T32, T33, T40, T41 - reduced condition is related to low levels of land-use intensity and ongoing succession  (but also too intensive land use).
-In addition, reduced condition is related to high abundance of alien species in all major semi-natural ecosystem types.
+
+In these plots, the panels for the different major ecosystem types
+differ because Miljødirektoratets mapping manual (see reference list
+above) uses different variables in these major ecosystem types to define
+condition.
+
+In the heath systems - T31 and T34 - reduced condition is mainly related
+to low levels of grazing and ongoing succession (but also too intensive
+grazing). In the meadow systems - T32, T33, T40, T41 - reduced condition
+is related to low levels of land-use intensity and ongoing succession
+(but also too intensive land use). In addition, reduced condition is
+related to high abundance of alien species in all major semi-natural
+ecosystem types.
 
 ### Regression analyses Sentinel-2, NDVI as a function of condition
-We can investigate further how NDVI and condition are connected using regression analyses.
-
-First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
+
+We can investigate further how NDVI and condition are connected using
+regression analyses.
+
+First, we take a look at how balanced the data are across major
+ecosystem types, regions and condition classes
+
 ```{r warning=FALSE, message=FALSE}
@@
 ```
-T41 is has only very few data and is present in 2 regions only, we omit it.
-Northern Norway does not have any T34 occurrences and very few T33 and T40, we won't be able to model it together with the other regions, and might need to model T31 and T32 there only.
-
-NDVI data are bounded between -1 and 1, and thus require modelling with an appropriate method that can handle bounded data.
-We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
+
+T41 is has only very few data and is present in 2 regions only, we omit
+it. Northern Norway does not have any T34 occurrences and very few T33
+and T40, we won't be able to model it together with the other regions,
+and might need to model T31 and T32 there only.
+
+NDVI data are bounded between -1 and 1, and thus require modelling with
+an appropriate method that can handle bounded data. We can transform the
+variable to be bounded between 0 and 1 and use beta-regression models:
+
 ```{r}
@@
 ```
-NDVI values do vary between:
-- Regions
-- Ecosystem types
-- Condition classes
-
-It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
+
+NDVI values do vary between: - Regions - Ecosystem types - Condition
+classes
+
+It is evident that there are differences between regions. But since we
+are not interested in the statistics for region differences, we run
+separate models for every region for ease of interpretation.
+
 ```{r, include = F}
@@
 # Northern Norway
+
 ```{r message=FALSE, warning=FALSE}
@@
 ```
-NDVI in good condition: T32 > T31/T33 > T40 (last not sig. but very few data)
-
-T31,T33,T40: NDVI increases as condition deteriorates
-T32: NDVI does not change with condition
+
+NDVI in good condition: T32 \> T31/T33 \> T40 (last not sig. but very
+few data)
+
+T31,T33,T40: NDVI increases as condition deteriorates T32: NDVI does not
+change with condition
 
 # Central Norway
+
 ```{r message=FALSE, warning=FALSE}
@@
 ```
-NDVI in good condition: T32 > T33/T34/T40 > T31
-
-T31: NDVI increases as condition deteriorates
-T32/T33/T34/T40: NDVI increases sig. less than in T31 as condition deteriorates, but it still increases
+
+NDVI in good condition: T32 \> T33/T34/T40 \> T31
+
+T31: NDVI increases as condition deteriorates T32/T33/T34/T40: NDVI
+increases sig. less than in T31 as condition deteriorates, but it still
+increases
 
 # Western Norway
+
 ```{r message=FALSE, warning=FALSE}
@@
 ```
-NDVI in good condition: T31/T32/T40 > T33/T34
-
-all types: NDVI increases slightly as condition deteriorates, but just about sig.
+
+NDVI in good condition: T31/T32/T40 \> T33/T34
+
+all types: NDVI increases slightly as condition deteriorates, but just
+about sig.
 
 # Eastern Norway
+
 ```{r message=FALSE, warning=FALSE}
@@
 ```
-NDVI in good condition: T32 > T31/T40 > T33 > T34
-
-T31/T32/T34/T40: NDVI increases slightly as condition deteriorates
-T33: NDVI increases as condition deteriorates, more strongly than in the other types
+
+NDVI in good condition: T32 \> T31/T40 \> T33 \> T34
+
+T31/T32/T34/T40: NDVI increases slightly as condition deteriorates T33:
+NDVI increases as condition deteriorates, more strongly than in the
+other types
 
 # Southern Norway
+
 ```{r message=FALSE, warning=FALSE}
@@
 ```
-NDVI in good condition: T32 > T31/T40 > T33/T34
-
-T31/T32: borderline NDVI increases as condition deteriorates
-T33/T34: NDVI increases as condition deteriorates
-T40: insufficient power
+
+NDVI in good condition: T32 \> T31/T40 \> T33/T34
+
+T31/T32: borderline NDVI increases as condition deteriorates T33/T34:
+NDVI increases as condition deteriorates T40: insufficient power
 
@@
 
-In semi-natural ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
-Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country.
-Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.
+In semi-natural ecosystems NDVI increases as ecosystem condition
+deteriorates, but the effect sizes vary with major ecosystem types.
+Boreal heath shows the most consistent increases, and also coastal
+meadows and coastal heathlands show increases in parts of the country.
+Semi-natural and semi-natural-looking meadows show very tiny increases
+if any at all.
 
 ### NDVI across time - Sentinel, MODIS & LandSat
-As a last step, we can investigate how NDVI has changed over time. For this we include data from both MODIS and Landsat in addition to Sentinel-2.
-
-First, there's again some data handling to do. We merge the Sentinel, MODIS, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
+
+As a last step, we can investigate how NDVI has changed over time. For
+this we include data from both MODIS and Landsat in addition to
+Sentinel-2.
+
+First, there's again some data handling to do. We merge the Sentinel,
+MODIS, and Landsat data to show the full picture across time. Then we
+model the time series for each Satellite separately.
+
 ```{r results='hide'}
@@
 ```
-Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
+
+Now we can plot the time series for each main ecosystem type, showing
+each satellite time series separately:
+
 ```{r}
@@
 ```
-It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable.
-They vary both in their placement along the y-axis and in their variance.
-Modis data are absent from the T40/T41-ecosystem types because of a too large pixel size compared to the NiN-polygon sizes.
-In the other types, Modis is poorly represented for the same reason.
-
-We can test if there are consistent temporal changes in NDVI by running regressions including year as a covariate.
-
-We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
+
+It is quite obvious that the NDVI values from the three satellites are
+not quantitatively comparable. They vary both in their placement along
+the y-axis and in their variance. Modis data are absent from the
+T40/T41-ecosystem types because of a too large pixel size compared to
+the NiN-polygon sizes. In the other types, Modis is poorly represented
+for the same reason.
+
+We can test if there are consistent temporal changes in NDVI by running
+regressions including year as a covariate.
+
+We first investigate a model testing NDVI as a function of year and main
+ecosystem type for polygons in good condition, and we do so separately
+for each satellite.
+
 ```{r, warning=FALSE}
@@
 ```
-There's an increase in NDVI over time in mainly the Sentinel and Landsat time series in boreal heath in good condition, but this pattern is reduced or absent in the other semi-natural ecosystem types.
-MODIS does not show any trends in time but also has way less statistical power compared to Sentinel and Landsat.
-
-Given the consistently higher NDVI-values with decreasing condition in boreal heath, We may ask the question if the temporal NDVI increase might be stronger under reduced condition?
+
+There's an increase in NDVI over time in mainly the Sentinel and Landsat
+time series in boreal heath in good condition, but this pattern is
+reduced or absent in the other semi-natural ecosystem types. MODIS does
+not show any trends in time but also has way less statistical power
+compared to Sentinel and Landsat.
+
+Given the consistently higher NDVI-values with decreasing condition in
+boreal heath, We may ask the question if the temporal NDVI increase
+might be stronger under reduced condition?
+
 ```{r, warning=FALSE}
@@
 ```
-The temporal NDVI increase in boreal heath has NOT been stronger with reduced ecological condition. So a temporal NDVI approach does not seem like a promising way to go.
-We check for the same pattern in T32 and T34, jsut because they are the other semi-natural major ecosystem types with a substantial number of occurrences in the the NiN data:
+
+The temporal NDVI increase in boreal heath has NOT been stronger with
+reduced ecological condition. So a temporal NDVI approach does not seem
+like a promising way to go. We check for the same pattern in T32 and
+T34, just because they are the other semi-natural major ecosystem types
+with a substantial number of occurrences in the the NiN data:
+
 ```{r, warning=FALSE}
@@
 ```
-Again, patterns are only identifiable with Sentinel and Landsat data, but not with Modis.
-Semi-natural grasslands (T32) do indeed show a stronger increase in NDVI as condition deteriorates, so here a temporal approach may be useful under a reduced land-management/shrub-/tree-encroachment scenario.
-Coastal heathlands (T34), on the other hand, do not show stronger temporal increases with reduced condition.
-Coastal heathlands are subject to a management cycle, which includes developmental and successional changes over a course of 15-25 years followed by prescribed burning which resets that process.
-So, we have to expect increases in greenness in coastal heathlands over time also in well-managed areas, and thus temporal changes in NDVI are largely irrelevant for assessing ecological condition.
+
+Again, patterns are only identifiable with Sentinel and Landsat data,
+but not with Modis. Semi-natural grasslands (T32) do indeed show a
+stronger increase in NDVI as condition deteriorates, so here a temporal
+approach may be useful under a reduced
+land-management/shrub-/tree-encroachment scenario. Coastal heathlands
+(T34), on the other hand, do not show stronger temporal increases with
+reduced condition. Coastal heathlands are subject to a management cycle,
+which includes developmental and successional changes over a course of
+15-25 years followed by prescribed burning which resets that process.
+So, we have to expect increases in greenness in coastal heathlands over
+time also in well-managed areas, and thus temporal changes in NDVI are
+largely irrelevant for assessing ecological condition.
 
 # Overall conclusion
-NDVI in semi-natural ecosystem does vary with condition and also with ecosystem type, but the variance and thus overlap between condition classes and ecosystem types is large.
-Especially in boreal heath and semi-natural grasslands greenness and its change over time appaer to be indicative of the ecological condition of local ecoystems, linking well to the registered condition variables behind the lower condition scores (shrub/tree encroachment, low land-use intensity, succession) and thus reduced land-use as the main pressure identified for these systems.
-However, the large variance and thus overlap between condition classes and ecosystem types makes it generally difficult to define useful scaling values for absolute NDVI values.
-Since ecological condition also relates to the temporal development of NDVI values in semi-natural grasslands, a potential way way to approach this can be to define scaling values for time slopes as for instance has been done for the NDVI indicator in the ecological condition assessment for Norwegian mountains (Framstad et al.
-2022 in references above, see coded example at https://ninanor.github.io/IBECA/ndvi-trend_fjell.html).
-
-Beyond these considerations, there are two general issues which limit a more comprehensive and robust application of satellite based NDVI-data in an ecological condition context:
-
-# 1) We are lacking ecosystem maps for semi-natural ecosystems
-The analysis at hand clearly shows that different semi-natural ecosystem types display different greenness levels and thus need to be evaluated separately.
-Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within semi-natural systems separately.
-# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in semi-natural ecosystems
-Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in semi-natural systems.
-Here, we need field studies investigating more thoroughly how NDVI and ecological drivers, like for example shrub-/tree encroachment, are connected.
-This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
-
-In addition, further investigations should test how NDVI-condition pattern vary across biogeographic regions in addition to administrative regions (which are ecologically irrelevant), in order to better understand the results for the administrative regions.
+
+NDVI in semi-natural ecosystem does vary with condition and also with
+ecosystem type, but the variance and thus overlap between condition
+classes and ecosystem types is large. Especially in boreal heath and
+semi-natural grasslands greenness and its change over time appaer to be
+indicative of the ecological condition of local ecoystems, linking well
+to the registered condition variables behind the lower condition scores
+(shrub/tree encroachment, low land-use intensity, succession) and thus
+reduced land-use as the main pressure identified for these systems.
+However, the large variance and thus overlap between condition classes
+and ecosystem types makes it generally difficult to define useful
+scaling values for absolute NDVI values. Since ecological condition also
+relates to the temporal development of NDVI values in semi-natural
+grasslands, a potential way way to approach this can be to define
+scaling values for time slopes as for instance has been done for the
+NDVI indicator in the ecological condition assessment for Norwegian
+mountains (Framstad et al. 2022 in references above, see coded example
+at <https://ninanor.github.io/IBECA/ndvi-trend_fjell.html>).
+
+Beyond these considerations, there are two general issues which limit a
+more comprehensive and robust application of satellite based NDVI-data
+in an ecological condition context:
+
+1. We are lacking ecosystem maps for semi-natural ecosystems
+The analysis at hand clearly shows that different semi-natural ecosystem
+types display different greenness levels and thus need to be evaluated
+separately. Thus, we need ecosystem maps, that allow us to treat NDVI
+values, and temporal changes of these, for at least the different major
+ecosystem types within semi-natural systems separately. 
+2. We are lacking a good ecological understanding and ground-truthing for NDVI in
+semi-natural ecosystems Unlike for ecosystems like forests, tundra,
+savannah and similar, there is little empirical work and ground-truthing
+data documented that link NDVI values to ecological phenomena in
+semi-natural systems. Here, we need field studies investigating more
+thoroughly how NDVI and ecological drivers, like for example shrub-/tree
+encroachment, are connected. This includes a need for understanding
+temporal increases in NDVI in ecosystems in good condition.
+
+In addition, further investigations should test how NDVI-condition
+pattern vary across biogeographic regions in addition to administrative
+regions (which are ecologically irrelevant), in order to better
+understand the results for the administrative regions.
 
 ### Eksport file (final product)
+
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-10-04 NDVI_seminat.Rmd
adding fig captions and caching some data

@@
 
-```{r where-are-seminat-sites, }
+```{r where-are-seminat-sites, fig.cap="A map of Norway showing the location of the semi-natural ecosystem sites."}
 tm_shape(regnor) +
@@
 
-```{r}
+```{r hist-sentinel-ndvis, fig.cap="A histogram showing the distribution of polygon sizes."}
 summary(SentinelNDVI.seminat$area_meters)
@@
 
-semi-natural polygons stretch a large number of sizes from just above
+Semi-natural polygons stretch a large number of sizes from just above
 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
@@
 
-```{r}
+```{r NDVI-values-over-time-ndvis, fig.cap="A figure showing NDVI values for semi-natural ecosystem over time for six different main nature types in varying degrees of condition."}
 SentinelNDVI.seminat %>%
@@
 
-2022 stands out with the highest NDVI values missing. WE omit 2022 for
-this analysis
+2022 stands out with the highest NDVI values missing. 
+We omit 2022 for this analysis.
 
@@
 
-```{r warning=FALSE}
+```{r NDVI-major-basic-ndvis, warning=FALSE, fig.cap="NDVI values in major and basic semi natural ecosystem types in good condition."}
 SentinelNDVI.seminat %>%
@@
 
-```{r warning=FALSE}
+```{r ndvi-regions-ndvis, warning=FALSE, fig.cap="DVI variation across regions in semi-natural ecosystems (major types) in good condition"}
 SentinelNDVI.seminat %>%
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
+  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1))+
   facet_wrap(~hovedtype)
@@
 
-```{r warning=FALSE}
+```{r important-variables-ndvis, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
 nin.seminat2 %>%
@@
 
-```{r warning=FALSE, message=FALSE}
+```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradien, faceted by six semi natural ecosystem types in five regions in Norway. "}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
 
-```{r}
+<!-- RUNTIME approx. 2 min -->
+```{r, eval=F}
 SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
@@
 model.seminat.cond.Sent <- StepBeta(model.seminat.cond.Sent)
+summary(model.seminat.cond.Sent)
+
+# Caching
+saveRDS(model.seminat.cond.Sent, 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
+```
+
+```{r, include=}
+model.seminat.cond.Sent <- readRDS( 
+        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.cond.Sent.rds")
 summary(model.seminat.cond.Sent)

### 2023-10-04 NDVI_seminat.Rmd

@@
 
-<!-- RUNTIME approx. 2 min -->
-```{r, eval=F}
+```{r}
 SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
@@
   mutate(mean_beta = replace(mean_beta, mean_beta == 1, 0.9999))
-
+```
+
+<!-- RUNTIME approx. 2 min -->
+```{r, eval=F}
 # We run a stepwise-function on the full model including condition, ecosystem type, and region to find the most parsimonious model
@@
 
-```{r, include=}
+```{r, include=F}
 model.seminat.cond.Sent <- readRDS( 
@@
 
-It is evident that there are differences between regions. But since we
-are not interested in the statistics for region differences, we run
-separate models for every region for ease of interpretation.
+It is evident that there are differences between regions. 
+But since we are not interested in the statistics for region differences, 
+we run **separate models for every region** for ease of interpretation.
 
@@
 
+<!-- RUNTIME 5 sec -->
 ```{r message=FALSE, warning=FALSE}
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r north-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Northern Norway."}
 SentinelNDVI.seminat.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.seminat.cond.Sent.N)$coefficients
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Central Norway."}
 SentinelNDVI.seminat.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.seminat.cond.Sent.C)$coefficients
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r west-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in W-Norway."}
 SentinelNDVI.seminat.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.seminat.cond.Sent.W)$coefficients
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in E-Norway."}
 SentinelNDVI.seminat.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.seminat.cond.Sent.E)$coefficients
@@
 
-```{r message=FALSE, warning=FALSE}
+```{r  south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in S-Norway."}
 SentinelNDVI.seminat.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.seminat.cond.Sent.S)$coefficients
@@
 
-Overall take home messages:
-
-In semi-natural ecosystems NDVI increases as ecosystem condition
-deteriorates, but the effect sizes vary with major ecosystem types.
-Boreal heath shows the most consistent increases, and also coastal
-meadows and coastal heathlands show increases in parts of the country.
-Semi-natural and semi-natural-looking meadows show very tiny increases
-if any at all.
+> Overall take home messages:
+>
+> In semi-natural ecosystems NDVI increases as ecosystem condition deteriorates, but the effect sizes vary with major ecosystem types.
+Boreal heath shows the most consistent increases, and also coastal meadows and coastal heathlands show increases in parts of the country.
+Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.
 
@@
 
-```{r results='hide'}
+```{r, results='hide'}
 ## data handling for time series analysis
@@
 # checking time series for MODIS
-ModisNDVI.seminat %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   #ModisNDVI.seminat %>%
+   #  group_by(id, year) %>%
+   #  filter(mean == max(mean, na.rm = TRUE)) %>%
+   #  ggplot(aes(x = year, y = mean)) +
+   #  geom_point() +
+   #  facet_wrap(~hovedtype)
 # 2022 does not stand out as in the Sentinel data, so we keep it
@@
 # checking time series for Landsat
-LandsatNDVI.seminat %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   #LandsatNDVI.seminat %>%
+   #  group_by(id, year) %>%
+   #  filter(mean == max(mean, na.rm = TRUE)) %>%
+   #  ggplot(aes(x = year, y = mean)) +
+   #  geom_point() +
+   #  facet_wrap(~hovedtype)
 # nothing worrying to see here either
@@
 
-```{r}
+```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets."}
 # plot
@@
 
-```{r, warning=FALSE}
+<!-- RUNTIME 10 min -->
+```{r, eval=F}
 model.seminat.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$tilstand == "God", ])
@@
 model.seminat.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$tilstand == "God", ])
-
+```
+
+```{r, include=FALSE, eval=F}
+saveRDS(model.seminat.time.GodTilst.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Sent.rds")
+saveRDS(model.seminat.time.GodTilst.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Modi.rds")
+saveRDS(model.seminat.time.GodTilst.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Land.rds")
+```
+
+```{r, include=F}
+model.seminat.time.GodTilst.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Sent.rds")
+model.seminat.time.GodTilst.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Modi.rds")
+model.seminat.time.GodTilst.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.GodTilst.Land.rds")
+```
+
+```{r}
 summary(model.seminat.time.GodTilst.Sent)$coefficients$cond
+```
+
+```{r}
 summary(model.seminat.time.GodTilst.Modi)$coefficients$cond
+```
+
+```{r}
 summary(model.seminat.time.GodTilst.Land)$coefficients$cond
@@
 Given the consistently higher NDVI-values with decreasing condition in
-boreal heath, We may ask the question if the temporal NDVI increase
+**boreal heath**, We may ask the question if the temporal NDVI increase
 might be stronger under reduced condition?
 
-```{r, warning=FALSE}
+<!-- RUNTIME 10 min -->
+```{r, warning=FALSE, eval=F}
 model.seminat.time.tilst.T31.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T31"), ])
@@
 model.seminat.time.tilst.T31.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.seminat[LandsatNDVI.seminat$hovedtype %in% c("T31"), ])
+```
+
+```{r, include=FALSE, eval=F}
+saveRDS(model.seminat.time.tilst.T31.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Sent.rds")
+saveRDS(model.seminat.time.tilst.T31.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Modi.rds")
+saveRDS(model.seminat.time.tilst.T31.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Land.rds")
+```
+
+```{r, include=F}
+model.seminat.time.tilst.T31.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Sent.rds")
+model.seminat.time.tilst.T31.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Modi.rds")
+model.seminat.time.tilst.T31.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.seminat.time.tilst.T31.Land.rds")
+```
+
+```{r}
 summary(model.seminat.time.tilst.T31.Sent)$coefficients$cond
+```
+
+```{r}
 summary(model.seminat.time.tilst.T31.Modi)$coefficients$cond
+```
+
+```{r}
 summary(model.seminat.time.tilst.T31.Land)$coefficients$cond
@@
 reduced ecological condition. So a temporal NDVI approach does not seem
-like a promising way to go. We check for the same pattern in T32 and
+like a promising way to go.
+
+We could check for the same pattern in T32 and
 T34, just because they are the other semi-natural major ecosystem types
-with a substantial number of occurrences in the the NiN data:
-
-```{r, warning=FALSE}
+with a substantial number of occurrences in the the NiN data. Here's the code for that, stored for later.
+
+```{r, eval=FALSE}
 model.seminat.time.tilst.T32.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.seminat[SentinelNDVI.seminat$hovedtype %in% c("T32"), ])
@@
 
-Again, patterns are only identifiable with Sentinel and Landsat data,
+But again, patterns are only identifiable with Sentinel and Landsat data,
 but not with Modis. Semi-natural grasslands (T32) do indeed show a
@@
 
-# Overall conclusion
+## Overall conclusion
 
@@
 classes and ecosystem types is large. Especially in boreal heath and
-semi-natural grasslands greenness and its change over time appaer to be
-indicative of the ecological condition of local ecoystems, linking well
+semi-natural grasslands greenness and its change over time appear to be
+indicative of the ecological condition of local ecosystems, linking well
 to the registered condition variables behind the lower condition scores

### 2023-10-04 DRAFT_NDVI_seminat.Rmd

@@
 # Import Sentinel NDVI Data
-df.s <- list.files("P:/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
+df.s <- list.files("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names=TRUE) %>%
   map_df(~fread(.))
@@
 # MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df.m<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
+df.m<- read.csv("/data/P-Prosjekter241201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_MODIS/modis_ndvi_ts_2000_2022.csv", )
 df.m$mean<-df.m$mean/10000
@@
 summary(SentinelNDVI.seminat.train$mean_beta) #ok
+LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
 # replace 1s in Landsat data with 0.9999
@@
 SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
-  rename('system.index' = 'system:index')
+  dplyr::rename('system.index' = 'system:index')
 

### 2023-10-04 NDVI_natopen.Rmd

@@
 
-|Region|T2|T12|T18|T20|T21|
+Region|T2|T12|T18|T20|T21|
+-|-|-|-|-|-|
 Northern Norway |  0  |   230 |  648 |  10 |   31|

### 2023-10-04 NDVI_wetland.Rmd
Fix headers, anchors, typos, line breaks, tables, fig caps, ect. Minor things.

@@
-# NDVI (greenness) indicator, wetland ecosystems  {#NDVI-indicator-wetland}
-
-<br />
+# NDVI - Wetland ecosystems  {#NDVI-indicator-wetland}
+
+<br /> Norwegian name: **Primærproduksjon** <br />
 
 _Author and date:_
+
 Joachim Töpper, Tessa Bargmann
@@
 ```{r, echo=FALSE}
-Ecosystem <- "våtmark"
+Ecosystem <- "Våtmark"
 Egenskap <- c("Primærproduksjon")
@@
 
-## Introduction
-The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area.
+## Introduction {#intro-ndviw}
+The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, 
+and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. 
 Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. 
 Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. 
-E.g.
-lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems.
+E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. 
 In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. 
-For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al.
-2020).
+For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). 
 For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). 
@@
 
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
-excluding forest swamps).
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e. excluding forest swamps). 
 For this task, three initial challenges can be identified early on:
-1) We don't yet have a national wetland map, i.e.
+
+1.
+We don't yet have a national wetland map, i.e.
 we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
-3) NDVI in wetland systems lacks the thorough ground-truthing known from globally large ecosystems as forests, tundra or Savannah.
+2.
+The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+3.
+NDVI in wetland systems lacks the thorough ground-truthing known from globally large ecosystems as forests, tundra or Savannah.
 Normal variation, like stochastic rain events and moss moisture status may influence NDVI signals greatly (Valøen, 2019).
@@
 In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
-1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
+
+1.
+We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+2.
+We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
@@
 
-## About the underlying data
+## About the underlying data {#underlying-data-ndviw}
 In the NDVI project for wetlands, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
@@
 
-(1) NiN ecosystem data:
+1. NiN ecosystem data:
 The NiN data in this study contain data about the ecosystem categories (cf.
@@
 
-(2) NDVI data:
+2. NDVI data:
 The NDVI data used in the work at hand are pixel means for polygons in the NiN-data, i.e.
@@
 
-### Representativity in time and space
+### Representativity in time and space {#representativity-ndviw}
 For wetlands, the NiN data contain 6502 wetland polygons of known area size across mainland Norway.
@@
 
-                  V1    V3    V4    V6    V9    V10
-Northern Norway   638   275   55    102   360   315
-Central Norway    702   1250  34    0     302   116
-Western Norway    296   111   30    0     460   158
-Eastern Norway    251   89    103   0     251   223
-Southern Norway   58    44    5     0     155   102
+|Region            |V1    |V3    |V4   |V6   | V9   |V10|
+|------------------|------|------|-----|-----|------|---|
+|Northern Norway   |638   |275   |55   |102  | 360  |315|
+|Central Norway    |702   |1250  |34   |0    | 302  |116|
+|Western Norway    |296   |111   |30   |0    | 460  |158|
+|Eastern Norway    |251   |89    |103  |0    | 251  |223|
+|Southern Norway   |58    |44    |5    |0    | 155  |102|
 
@@
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data. 
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
-
-### Temporal coverage
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), 
+the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+
+### Temporal coverage {#temp-cov-ndvis
 The currently available NiN data on wetland ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. 
@@
 
-Available Sentinel-2 satellite NDVI data span the years 2015-2022.
-Available MODIS satellite NDVI data span the years 2000-2022.
-Available Landsat satellite NDVI data span the years 1984-2021.
-
-## Collinearities with other indicators
-The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in drained wetland areas under shrub and tree encroachment.
-
-## Reference state and values
-### Reference state and scaling values
+- Available Sentinel-2 satellite NDVI data span the years 2015-2022.
+- Available MODIS satellite NDVI data span the years 2000-2022.
+- Available Landsat satellite NDVI data span the years 1984-2021.
+
+## Collinearities with other indicators {#coll-ndvis}
+The NDVI signal may likely be co-linear with [the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html) based on LiDAR derived canopy height in drained wetland areas under shrub and tree encroachment.
+
+## Reference state and values {#ref-state-ndviw}
+
+### Reference state and scaling values {#ref-state2-ndviw}
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. 
@@
 2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to wetland ecology and condition and (2) its applicability for defining a reference state.
+In this exploratory work, we will use the condition classification in the NiN-data to explore
+
+1. how NDVI connects to wetland ecology and condition and 
+2. its applicability for defining a reference state.
+
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
 
-## Uncertainties
-Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
+## Uncertainties {#uncertainties-ndviw}
+Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region 
+(or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a 
+measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
 
@@
 
-## Analyses
-### Data sets
+## Analyses {#analyses-ndviw}
+
+### Data sets {#datasets-ndviw}
 The analyses in this document make use of the following data sets:
--NiN data (Naturtyper etter Miljødirektoratets instruks)
--Sentinel-2 NDVI data
--MODIS NDVI data
--Landsat NDVI data
-
-#### NiN data and geometry for Norway and the five regions
+
+- NiN data (Naturtyper etter Miljødirektoratets instruks) 
+- Sentinel-2 NDVI data 
+- MODIS NDVI data 
+- Landsat NDVI data
+
+#### NiN data and geometry for Norway and the five regions {#nindata-ndviw}
 
@@
 
-```{r message=FALSE}
+```{r, message=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
 # change region names to something R-friendly
-reg$region
+#reg$region
 reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")
@@
 
-#### NDVI data
+#### NDVI data {#ndvidata-ndviw}
 All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE).
@@
 
-##### Sentinel-2 NDVI data
+##### Sentinel-2 NDVI data {#sentinelldata-ndviw}
 The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
 The image collection contains Sentinel imagery from June, July and August 2015-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. 
+This script merges them and then merges the dataframe to the NiN data.
 
@@
 
-##### MODIS NDVI data
+##### MODIS NDVI data {#modisdata-ndviw}
 The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
@@
 
-##### Landsat NDVI data
+##### Landsat NDVI data {#landsatdata-ndviw}
 The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
 The image collection contains Landsat imagery from June, July and August 1984-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. 
+This script merges them and then merges the dataframe to the NiN data.
 
@@
 
-### Data handling
-#### NiN data
+### Data handling {#data-handling-ndviw}
+#### NiN data {#nin-data-handling-ndviw}
 ```{r, echo = T, eval = F}
@@
 
+```{r, include=F, eval=F}
+saveRDS(nin.wetland, paste0(here::here(), "/data/cache/nin.wetland.RDS"))
+saveRDS(nin.wetland2, paste0(here::here(), "/data/cache/nin.wetland2.RDS"))
+```
+
 ```{r, echo = F}
@@
 
-#### NDVI data
+#### NDVI data {#ndvi-data-ndviw}
 ```{r, results='hide'}
@@
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-dim(SentinelNDVI.wetland)
+  # dim(SentinelNDVI.wetland)
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
-dim(SentinelNDVI.wetland)
+  # dim(SentinelNDVI.wetland)
 # split date into year, month & day
@@
 
-summary(SentinelNDVI.wetland)
+  # summary(SentinelNDVI.wetland)
 
@@
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-dim(ModisNDVI.wetland)
+  # dim(ModisNDVI.wetland)
 ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
-dim(ModisNDVI.wetland)
+  # dim(ModisNDVI.wetland)
 # split date into year, month & day
@@
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-dim(LandsatNDVI.wetland)
+  # dim(LandsatNDVI.wetland)
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
-dim(LandsatNDVI.wetland)
+  # dim(LandsatNDVI.wetland)
 # split date into year, month & day
@@
 
-### Exploratory analyses, Sentinel-2
-
-# Where are the wetland sites?
-```{r}
+### Exploratory analyses, Sentinel-2 {#exploratory-ndviw}
+
+**Where are the wetland sites?**
+```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows.}
 tm_shape(regnor) +
@@
     labels = c(
-      "V1 - minerotrophic mires",
-      "V3 - ombrotrophic bogs",
-      "V4 - cold springs",
-      "V6 - wet snowbeds",
-      "V9 - semi-natural mires",
-      "V10 - semi-natural wet meadows"
+      "V1",
+      "V3",
+      "V4",
+      "V6",
+      "V9",
+      "V10"
     ),
@@
 # How are polygon sizes distributed?
-```{r}
+```{r histogram-sentinall-ndviw, fig.cap="Histogram of wetland ecosystem polygon sizes in Norway."}
 summary(SentinelNDVI.wetland$area_meters)
@@
 ```
+
 Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
 
-# How does NDVI vary over the years
-```{r}
+**How does NDVI vary over the years**
+```{r NDVI-values-over-time-ndviw, fig.cap="A figure showing NDVI values for wetland ecosystem over time for six different main nature types in varying degrees of condition.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
 SentinelNDVI.wetland %>%
@@
 
-# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
-```{r}
+**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
+```{r NDVI-major-ndviw, warning=FALSE, fig.cap="NDVI values in major wetland ecosystem types in good condition.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows"}
 SentinelNDVI.wetland %>%
@@
   geom_point(size = 0.7, shape = 16, color = "grey")
-
+```
+```{r NDVI-basic-ndviw, warning=FALSE, fig.cap="NDVI values in basic wetland ecosystem types in good condition.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
 SentinelNDVI.wetland %>%
@@
 ```
+
 NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3).
@@
 
-# How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)
-```{r}
+**How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
+```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="DVI variation across regions in wetland ecosystems (major types) in good condition.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
 SentinelNDVI.wetland %>%
@@
 ```
+
 There is no obvious trends in geography.
 
-# How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)
-```{r}
+**How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)**
+```{r ndvi-condition-ndviw, warning=FALSE, fig.cap="NDVI values across wetland sites in different ecological condition.
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
 SentinelNDVI.wetland %>%
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
   facet_wrap(~hovedtype)
 ```
-In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern.
+
+In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, 
+but in minertrophic fens (V1) there is not much of a pattern. 
 Overlaps between condition classes are generally very large.
 
-# Which ecological relevant variable(s) is the condition measure in the NiN data related to?
-```{r}
+**Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
+```{r important-variables-ndviwv9, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
 nin.wetland2 %>%
@@
   labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
-
+```
+```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
 nin.wetland2 %>%
@@
   labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
-
+```
+
+```{r important-variables-ndviwv1, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
 nin.wetland2 %>%
@@
 ```
+
 In these plots, the panels for V1, V9 and V10 differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
@@
 In V1 (minerotrophic fens), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
-Under these drivers, greenness would not directly be affected.
-although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run.
-Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.
-
-### Regression analyses Sentinel-2, NDVI as a function of condition
+Under these drivers, greenness would not directly be affected, although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run. 
+**Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.**
+
+### Regression analyses Sentinel-2, NDVI as a function of condition {#regression-ndviw}
 We can investigate further how NDVI and condition are connected using regression analyses.
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r warning=FALSE}
+```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. "}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
 ```
+
 V4 has sparse data and is lacking in some combinations of region and condition level, so we drop it.
@@
 ```
+
 NDVI values do vary between:
+
 - Regions
@@
 
-# Northern Norway
-```{r message=FALSE, warning=FALSE}
+**Northern Norway**
+```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for four wetland ecosystems in Northern Norway."}
 SentinelNDVI.wetland.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.wetland.cond.Sent.N)$coefficients
 ```
+
 NDVI in good condition: V10 > V1 > V9 > V3
@@
 
-# Central Norway
-```{r message=FALSE, warning=FALSE}
+**Central Norway**
+```{r central-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Central Norway."}
 SentinelNDVI.wetland.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.wetland.cond.Sent.C)$coefficients
 ```
+
 NDVI in good condition: V10 > V1 & V9 > V3
@@
 
-# Western Norway
-```{r message=FALSE, warning=FALSE}
+**Western Norway**
+```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for fou wetland ecosystems in W-Norway."}
 SentinelNDVI.wetland.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.wetland.cond.Sent.W)$coefficients
-```
+
+```
+
 NDVI in good condition: V10 > V9 & V3 > V1
@@
 
-# Eastern Norway
-```{r message=FALSE, warning=FALSE}
+**Eastern Norway**
+```{r east-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in E-Norway."}
 SentinelNDVI.wetland.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.wetland.cond.Sent.E)$coefficients
 ```
+
 NDVI in good condition: V10 > V1 > V3 & V9
@@
 
-# Southern Norway
-```{r message=FALSE, warning=FALSE}
+**Southern Norway**
+```{r south-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in S-Norway."}
 SentinelNDVI.wetland.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.wetland.cond.Sent.S)$coefficients
 ```
+
 NDVI in good condition: V10 > V9 > V1 & V3
@@
 
-Overall take home messages:
-
-In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
+>Overall take home messages:
+>
+>In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
 This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural wetland ecosystems.
@@
 
-### NDVI across time - Sentinel, MODIS & LandSat
+### NDVI across time - Sentinel, MODIS & LandSat {#across-time-ndviw}
 As a last step, we can investigate how NDVI has changed over time. For this we include data from both MODIS and Landsat in addition to Sentinel-2.
@@
 # checking time series for MODIS
-ModisNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   # ModisNDVI.wetland %>%
+   #   group_by(id, year) %>%
+   #   filter(mean == max(mean, na.rm = TRUE)) %>%
+   #   ggplot(aes(x = year, y = mean)) +
+   #   geom_point() +
+   #   facet_wrap(~hovedtype)
 # 2022 does not stand out as in the Sentinel data, so we keep it
@@
 # checking time series for Landsat
-LandsatNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   # LandsatNDVI.wetland %>%
+   #   group_by(id, year) %>%
+   #   filter(mean == max(mean, na.rm = TRUE)) %>%
+   #   ggplot(aes(x = year, y = mean)) +
+   #   geom_point() +
+   #   facet_wrap(~hovedtype)
 # nothing worrying to see here either
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.wetland$mean_beta)
-summary(ModisNDVI.wetland$mean_beta)
-summary(LandsatNDVI.wetland$mean_beta)
+   # summary(SentinelNDVI.wetland$mean_beta)
+   # summary(ModisNDVI.wetland$mean_beta)
+   # summary(LandsatNDVI.wetland$mean_beta)
 # replace 1s in Landsat data with 0.9999
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-names(SentinelNDVI.wetland)
-names(ModisNDVI.wetland)
-names(LandsatNDVI.wetland)
+   # names(SentinelNDVI.wetland)
+   # names(ModisNDVI.wetland)
+   # names(LandsatNDVI.wetland)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
@@
 # check if they have the same geometry
-st_crs(SentinelNDVI.wetland)
-st_crs(ModisNDVI.wetland)
-st_crs(LandsatNDVI.wetland)
+   # st_crs(SentinelNDVI.wetland)
+   # st_crs(ModisNDVI.wetland)
+   # st_crs(LandsatNDVI.wetland)
 # all good
@@
 allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat, levels = c("Sentinel", "Modis", "Landsat"))
-levels(allSatNDVI.wetland$Sat)
-```
+  # levels(allSatNDVI.wetland$Sat)
+```
+
 Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
-```{r}
+```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets."}
 # plot
@@
 ```
+
 It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable. 
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-```{r warning=FALSE}
+
+<!-- RUNTIME approx. 15 min -->
+```{r, warning=FALSE, eval=F}
 model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand == "God", ])
@@
 model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand == "God", ])
-
+```
+
+```{r, include=FALSE, eval=F}
+saveRDS(model.wetland.time.GodTilst.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Sent.rds")
+saveRDS(model.wetland.time.GodTilst.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Modi.rds")
+saveRDS(model.wetland.time.GodTilst.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Land.rds")
+```
+
+```{r, include=F}
+model.wetland.time.GodTilst.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Sent.rds")
+model.wetland.time.GodTilst.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Modi.rds")
+model.wetland.time.GodTilst.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Land.rds")
+```
+
+```{r}
 summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
+```
+
+```{r}
 summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
+```
+
+```{r}
 summary(model.wetland.time.GodTilst.Land)$coefficients$cond
 ```
+
 There's an increase in NDVI over time in all three satellite time series in essentially all major wetland ecosystem types in good condition. 
@@
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
-```{r warning=FALSE}
+Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, we may ask the question if the temporal NDVI increase might be stronger with reduced condition?
+```{r, warning=FALSE, eval=F}
 model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
@@
 model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
+```
+
+```{r, include=FALSE, eval=F}
+saveRDS(model.wetland.time.tilst.V9V10.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Sent.rds")
+saveRDS(model.wetland.time.tilst.V9V10.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Modi.rds")
+saveRDS(model.wetland.time.tilst.V9V10.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Land.rds")
+```
+
+```{r, include=F}
+model.wetland.time.tilst.V9V10.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Sent.rds")
+model.wetland.time.tilst.V9V10.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Modi.rds")
+model.wetland.time.tilst.V9V10.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Land.rds")
+```
+
+```{r}
 summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
+```
+```{r}
 summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
+```
+```{r}
 summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
 ```
+
 Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat. 
@@
 
-# Overall conclusion
+## Overall conclusion
+
 NDVI in wetland ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
@@
 
-# 1) We are lacking ecosystem maps for wetland ecosystems
+1. We are lacking ecosystem maps for wetland ecosystems
 The analysis at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
-# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
+2. We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
 Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands.
+
 Here, we need field studies investigating:
+
 - The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition.
 This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
-
 - How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.
+
 The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
@@
 
-### Eksport file (final product)
+### Eksport {#export-ndviw}
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-10-04 NDVI_seminat.Rmd
Some erroneos headers

@@
 
-```{r warning=FALSE}
+```{r ndvi-condition-ndvis, warning=FALSE, fig.cap="NDVI values across semi-natural ecosystem sites in different ecological condition."}
 SentinelNDVI.seminat %>%
@@
 
-```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradien, faceted by six semi natural ecosystem types in five regions in Norway. "}
+```{r balanced-data-ndvis, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by six semi natural ecosystem types in five regions in Norway. "}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
 
-# Northern Norway
+**Northern Norway**
 
@@
 
-# Central Norway
-
-```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in Central Norway."}
+**Central Norway**
+
+```{r central-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for five semi-natural ecosystems in Central Norway."}
 SentinelNDVI.seminat.train %>%
@@
 
-# Western Norway
+**Western Norway**
 
@@
 
-# Eastern Norway
-
-```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in E-Norway."}
+**Eastern Norway**
+
+```{r east-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for five semi-natural ecosystems in E-Norway."}
 SentinelNDVI.seminat.train %>%
@@
 
-# Southern Norway
-
-```{r  south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for five semi-natural ecosystems in S-Norway."}
+**Southern Norway**
+
+```{r south-nvdi-vs-tilstand-ndvis, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for five semi-natural ecosystems in S-Norway."}
 SentinelNDVI.seminat.train %>%

### 2023-10-04 NDVI_wetland.Rmd
Unquoted fig caps

@@
 ```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows.}
+V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
 tm_shape(regnor) +
@@
 ```
-```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
+```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
 nin.wetland2 %>%
@@
 
-```{r important-variables-ndviwv1, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
+```{r important-variables-ndviwv1, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
 nin.wetland2 %>%
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. "}
+```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. ", fig.height=12}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
   facet_grid(region ~ hovedtype)
@@
 We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
-```{r}
+```{r, message=F}
 SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
@@
 **Northern Norway**
-```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for four wetland ecosystems in Northern Norway."}
+```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Northern Norway."}
 SentinelNDVI.wetland.train %>%
@@
 V1: NDVI decreases as condition deteriorates
+
 V3: NDVI does not change as condition deteriorates
+
 V9 & V10: NDVI increases as condition deteriorates
@@
 V1: NDVI decreases as condition deteriorates
+
 V3, V9 & V10: NDVI increases as condition deteriorates
@@
 V1 & V10: NDVI increases as condition deteriorates
+
 V3: NDVI decreases as condition deteriorates (but really insufficient data)
+
 V9: NDVI does not change as condition deteriorates
@@
 V1 & V3: NDVI does not change as condition deteriorates
+
 V9 & V10: NDVI increases as condition deteriorates
@@
 V1, V9 & V10: NDVI increases as condition deteriorates
+
 V3: NDVI does not change as condition deteriorates (but also insuffcient data)
+
+<br />
+<br />
+<hr />
 
@@
 In bogs (V3) there is no consistent pattern.
+
+<hr />
 
@@
 Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
-```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets."}
+```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
 # plot

### 2023-10-04 NDVI_seminat.Rmd
Increase figure height

@@
 
+<br />
+<br />
+<hr />
+
 > Overall take home messages:
@@
 Semi-natural and semi-natural-looking meadows show very tiny increases if any at all.
+
+<hr />
 
@@
 
-```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets."}
+```{r ndvi-time-series-comparison-ndvis, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
 # plot

### 2023-10-04 NDVI_wetland.Rmd
typo

@@
 **Western Norway**
-```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for fou wetland ecosystems in W-Norway."}
+```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in W-Norway."}
 SentinelNDVI.wetland.train %>%

### 2023-10-04 NDVI_natopen.Rmd

@@
 
-## Introduction
+## Introduction {#intro-ndvin}
 The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area.
@@
 
-## About the underlying data
+## About the underlying data {#about-the-data-ndvin}
 In the NDVI project for natopen systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
@@
 
-### Representativity in time and space
-#### check here ####
+### Representativity in time and space {#rep-in-space-and-time-ndvin}
+<!-- #### check here #### -->
 For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway.
@@
 
-### Temporal coverage
+### Temporal coverage {#temp-coverage-ndvin}
 The currently available NiN data on natopen ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
@@
 
-Available Sentinel-2 satellite NDVI data span the years 2015-2022.
-Available MODIS satellite NDVI data span the years 2000-2022.
-Available Landsat satellite NDVI data span the years 1984-2021.
-
-## Collinearities with other indicators
+- Available Sentinel-2 satellite NDVI data span the years 2015-2022.
+- Available MODIS satellite NDVI data span the years 2000-2022.
+- Available Landsat satellite NDVI data span the years 1984-2021.
+
+## Collinearities with other indicators {#collinearity-ndvin}
 Unlike wetlands and semi.natural systems, natopen areas are not subject to shrub and tree encroachment due to changes in land-use, Thus, we do not expect any co-linearity of the NDVI signal with e.g.
-the re-growth indicator based on LiDAr derived canopy height.
-
-## Reference state and values
-### Reference state and scaling values
+[the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html) based on LiDAr derived canopy height.
+
+## Reference state and values {#ref-ndvin}
+### Reference state and scaling values {#ref2-ndvin}
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
@@
 2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to natopen ecology and condition and (2) its applicability for defining a reference state.
+In this exploratory work, we will use the condition classification in the NiN-data to explore
+1. how NDVI connects to natopen ecology and condition and 
+2. its applicability for defining a reference state.
+
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
 
-## Uncertainties
+## Uncertainties {#uncertainties-ndvin}
 Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
 
-## References
+## References {#references-ndvin}
 
@@
 
-## Analyses
-### Data sets
+## Analyses {#analyses-ndvin}
+### Data sets {#datasets-ndvin}
 The analyses in this document make use of the following data sets:
@@
 # change region names to something R-friendly
-reg$region
+  # reg$region
 reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")
@@
 
-#### NDVI data
+#### NDVI data {#ndvi-data-ndvin}
 All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
@@
 
-### Data handling
+### Data handling {#data-handling-ndvin}
 #### NiN data
@@
 
+```{r, include=F, eval=F}
+saveRDS(nin.natopen, paste0(here::here(), "/data/cache/nin.natopen.RDS"))
+saveRDS(nin.natopen2, paste0(here::here(), "/data/cache/nin.natopen2.RDS"))
+```
+
 ```{r, echo = F}
@@
 
-#### NDVI data
+#### NDVI data {#ndvi-data-ndvin}
 ```{r, results='hide', message=FALSE}
@@
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-dim(SentinelNDVI.natopen)
+  # dim(SentinelNDVI.natopen)
 SentinelNDVI.natopen <- SentinelNDVI.natopen %>% filter(as.numeric(area_meters) >= 100)
-dim(SentinelNDVI.natopen)
+  # dim(SentinelNDVI.natopen)
 # split date into year, month & day
@@
 
-summary(SentinelNDVI.natopen)
+  # summary(SentinelNDVI.natopen)
 
@@
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-dim(ModisNDVI.natopen)
+  # dim(ModisNDVI.natopen)
 ModisNDVI.natopen <- ModisNDVI.natopen %>% filter(as.numeric(area_meters) >= 62500)
-dim(ModisNDVI.natopen)
+  # dim(ModisNDVI.natopen)
 # split date into year, month & day
@@
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-dim(LandsatNDVI.natopen)
+  # dim(LandsatNDVI.natopen)
 LandsatNDVI.natopen <- LandsatNDVI.natopen %>% filter(as.numeric(area_meters) >= 900)
-dim(LandsatNDVI.natopen)
+  # dim(LandsatNDVI.natopen)
 # split date into year, month & day
@@
 
-### Exploratory analyses, Sentinel-2
+### Exploratory analyses, Sentinel-2 {#exporatory-analyses-ndvin}
 
 **Where are the natopen sites?**
-```{r}
+```{r where-are-natopen-sites, fig.cap="A map of Norway showing the location of the naturally open ecosystem sites.
+T12 = coastal meadows, T18 = flood plains, T2 = shallow soils, T20 = innfrysningsmark, T21 = sand dunes."}
 tm_shape(regnor) +
@@
     labels = c(
-      "T12 - coastal meadows",
-      "T18 - flood plains",
-      "T2 - shallow soils",
-      "T20 - innfrysningsmark",
-      "T21 - sand dunes"
+      "T12",
+      "T18",
+      "T2",
+      "T20",
+      "T21"
     ),
@@
 **How are polygon sizes distributed?**
-```{r}
-summary(SentinelNDVI.natopen$area_meters)
+```{r histogram-sentinall-ndvin, fig.cap="Histogram of naturally open ecosystem polygon sizes in Norway."}
+#summary(SentinelNDVI.natopen$area_meters)
 hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
@@
 **How does NDVI vary over the years?**
-```{r}
+```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition.}
 SentinelNDVI.natopen %>%
@@
 **How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r warning=FALSE}
+```{r NDVI-major-ndvin, warning=FALSE, fig.cap="NDVI values in major natopen ecosystem types in good condition."}
 SentinelNDVI.natopen %>%
@@
   geom_point(size = 0.7, shape = 16, color = "grey")
-
+```
+```{r}
 SentinelNDVI.natopen$subtype <- factor(SentinelNDVI.natopen$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "E1"))
+```
+
+```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition.}
 SentinelNDVI.natopen %>%
@@
 **How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r warning=FALSE}
+```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="DVI variation across regions in natopen ecosystems (major types) in good condition.}
 SentinelNDVI.natopen %>%
@@
 **How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
-```{r warning=FALSE}
+```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition.}
 SentinelNDVI.natopen %>%
@@
 
-The plots suggest a slight decrease in NDVI as condition deteriorates. Overlaps between condition classes are generally very large.
-#### continue here ####
+The plots suggest a slight decrease in NDVI as condition deteriorates. 
+Overlaps between condition classes are generally very large.
+
+<!-- #### continue here #### -->
 
@@
 
-### Regression analyses Sentinel-2, NDVI as a function of condition
+### Regression analyses Sentinel-2, NDVI as a function of condition {#regressions-ndvin}
 We can investigate further how NDVI and condition are connected using regression analyses.
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r warning=FALSE, message=FALSE}
+```{r balanced-data-ndvin, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by natopen ecosystem types in five regions in Norway. ", fig.height=12}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
 
-T20 is has only very few data and is present in only 1 conditino class and 2 regions, we omit it.
-Northern Norway does not have any T2 occurrences, central Norway doesn't have any T21 occurrences, we won't be able to model these types across all regions.
+T20 is has only very few data and is present in only 1 condition class and 2 regions, we omit it.
+Northern Norway does not have any T2 occurrences, 
+central Norway doesn't have any T21 occurrences, 
+we won't be able to model these types across all regions.
 
@@
 We can transform the variable to be bounded between 0 and 1 and use beta-regression models:
-```{r}
+```{r, message=F}
 SentinelNDVI.natopen$mean_beta <- (SentinelNDVI.natopen$mean + 1) / 2
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.natopen.train$mean_beta) # ok
+  # summary(SentinelNDVI.natopen.train$mean_beta) # ok
 
@@
 
-It is evident that there are differences between regions. But since we are not interested in the statistics for region differences, we run separate models for every region for ease of interpretation.
+It is evident that there are differences between regions. 
+But since we are not interested in the statistics for region differences, we run **separate models for every region** for ease of interpretation.
 ```{r, include = F}
@@
 
-# Northern Norway
-```{r message=FALSE, warning=FALSE}
+**Northern Norway**
+```{r north-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Northern Norway."}
 SentinelNDVI.natopen.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.natopen.cond.Sent.N)$coefficients
 ```
+
 Not much of a pattern in either model or figure.
 
-# Central Norway
-```{r message=FALSE, warning=FALSE}
+**Central Norway**
+```{r central-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in Central Norway."}
 SentinelNDVI.natopen.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.natopen.cond.Sent.C)$coefficients
 ```
+
 NDVI in good condition: T12/T18 > T2
@@
 
-# Western Norway
-```{r message=FALSE, warning=FALSE}
+**Western Norway**
+```{r west-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for natopen ecosystems in W-Norway."}
 SentinelNDVI.natopen.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+
+```{r}
 summary(model.natopen.cond.Sent.W)$coefficients
 ```
+
 NDVI in good condition: T18 marginally > T2/T12/T21
@@
 
-# Eastern Norway
-```{r message=FALSE, warning=FALSE}
+**Eastern Norway**
+```{r east-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in E-Norway."}
 SentinelNDVI.natopen.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.natopen.cond.Sent.E)$coefficients
 ```
+
 NDVI in good condition: T18 > T2/T12 > T21
@@
 
-# Southern Norway
-```{r message=FALSE, warning=FALSE}
+**Southern Norway**
+```{r south-nvdi-vs-tilstand-ndvin, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for natopen ecosystems in S-Norway."}
 SentinelNDVI.natopen.train %>%
@@
   facet_wrap(~hovedtype)
-
+```
+```{r}
 summary(model.natopen.cond.Sent.S)$coefficients
 ```
+
 NDVI in good condition: T12/T18 > T2/T21
 
-Reduced NDVI with reduced condition in sand dune vegetation types in Southern Norway, Otherwise no changes with condition.
-
-**Overall take home messages:**
-
-In naturally open ecosystems below the tree line, NDVI seems largely independent of ecosystem condition.
+Reduced NDVI with reduced condition in sand dune vegetation types in Southern Norway. Otherwise no changes with condition.
+
+<br />
+<br />
+<hr />
+
+>Overall take home messages:
+>
+>In naturally open ecosystems below the tree line, NDVI seems largely independent of ecosystem condition.
 This is probably because many of the regarding ecosystems are structured by natural disturbances anyways.
@@
 
-### NDVI across time - Sentinel, MODIS & LandSat
-As a last step, we can investigate how NDVI has changed over time. For this we include data from both MODIS and Landsat in addition to Sentinel-2.
-
-First, there's again some data handling to do. We merge the Sentinel, MODIS, and Landsat data to show the full picture across time. Then we model the time series for each Satellite separately.
+<hr />
+
+### NDVI across time - Sentinel, MODIS & LandSat {#ndvi-across-time-ndvin}
+As a last step, we can investigate how NDVI has changed over time. 
+For this we include data from both MODIS and Landsat in addition to Sentinel-2.
+
+First, there's again some data handling to do. We merge the Sentinel, MODIS, and Landsat data to show the full picture across time. 
+Then we model the time series for each Satellite separately.
 ```{r results='hide'}
@@
 # checking time series for MODIS
-ModisNDVI.natopen %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   # ModisNDVI.natopen %>%
+   #   group_by(id, year) %>%
+   #   filter(mean == max(mean, na.rm = TRUE)) %>%
+   #   ggplot(aes(x = year, y = mean)) +
+   #   geom_point() +
+   #   facet_wrap(~hovedtype)
 # 2022 does not stand out as in the Sentinel data, so we keep it
@@
 # checking time series for Landsat
-LandsatNDVI.natopen %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm = TRUE)) %>%
-  ggplot(aes(x = year, y = mean)) +
-  geom_point() +
-  facet_wrap(~hovedtype)
+   # LandsatNDVI.natopen %>%
+   #   group_by(id, year) %>%
+   #   filter(mean == max(mean, na.rm = TRUE)) %>%
+   #   ggplot(aes(x = year, y = mean)) +
+   #   geom_point() +
+   #   facet_wrap(~hovedtype)
 # nothing worrying to see here either
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-summary(SentinelNDVI.natopen$mean_beta) # ok
-summary(ModisNDVI.natopen$mean_beta) # ok
-summary(LandsatNDVI.natopen$mean_beta)
+   # summary(SentinelNDVI.natopen$mean_beta) # ok
+   # summary(ModisNDVI.natopen$mean_beta) # ok
+   # summary(LandsatNDVI.natopen$mean_beta)
 # replace 1s in Landsat data with 0.9999
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-names(SentinelNDVI.natopen)
-names(ModisNDVI.natopen)
-names(LandsatNDVI.natopen)
+   # names(SentinelNDVI.natopen)
+   # names(ModisNDVI.natopen)
+   # names(LandsatNDVI.natopen)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
@@
 # check if they have the same geometry
-st_crs(SentinelNDVI.natopen)
-st_crs(ModisNDVI.natopen)
-st_crs(LandsatNDVI.natopen)
+   # st_crs(SentinelNDVI.natopen)
+   # st_crs(ModisNDVI.natopen)
+   # st_crs(LandsatNDVI.natopen)
 # all good
@@
 allSatNDVI.natopen$Sat <- factor(allSatNDVI.natopen$Sat, levels = c("Sentinel", "Modis", "Landsat"))
-levels(allSatNDVI.natopen$Sat)
+   # levels(allSatNDVI.natopen$Sat)
 ```
 Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
-```{r}
+```{r ndvi-time-series-comparison-ndvin, fig.cap="Comparing the NDVI time series in the three data sets.", fig.height=10}
 # plot
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for Sentinel and Landsat.
-```{r message=FALSE, warning=FALSE}
+```{r message=FALSE, warning=FALSE, eval=F}
 model.natopen.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$tilstand == "God", ])
 model.natopen.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$tilstand == "God", ])
-
+```
+
+```{r, include=F, eval=F}
+saveRDS(model.natopen.time.GodTilst.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.GodTilst.Sent.rds")
+saveRDS(model.natopen.time.GodTilst.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.GodTilst.Land.rds")
+```
+
+```{r, include=F}
+model.natopen.time.GodTilst.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.GodTilst.Sent.rds")
+model.natopen.time.GodTilst.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.GodTilst.Land.rds")
+```
+
+```{r}
 summary(model.natopen.time.GodTilst.Sent)$coefficients$cond
+```
+```{r}
 summary(model.natopen.time.GodTilst.Land)$coefficients$cond
 ```
+
 There's an increase in NDVI over time in both satellite time series in T12 (coastal meadows) and T18 (flood plains) in good condition. 
@@
 We can check if these temporal changes in T12 and T18, as well as for T21 in Southern Norway, vary between condition classes?
-```{r, warning=FALSE}
+
+<!-- RUNTIME >10 min -->
+```{r, warning=FALSE, eval=F}
 model.natopen.time.tilst.T12.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.natopen[SentinelNDVI.natopen$hovedtype %in% c("T12"), ])
@@
 model.natopen.time.tilst.T21.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.natopen[LandsatNDVI.natopen$hovedtype %in% c("T21") & LandsatNDVI.natopen$region == "Southern Norway", ])
+```
+
+```{r, eval=F, include=F}
+saveRDS(model.natopen.time.tilst.T12.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T12.Sent.rds")
+saveRDS(model.natopen.time.tilst.T12.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T12.Land.rds")
+saveRDS(model.natopen.time.tilst.T18.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T18.Sent.rds")
+saveRDS(model.natopen.time.tilst.T18.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T18.Land.rds")
+saveRDS(model.natopen.time.tilst.T21.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T21.Sent.rds")
+saveRDS(model.natopen.time.tilst.T21.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T21.Land.rds")
+```
+
+```{r, include=F}
+model.natopen.time.tilst.T12.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T12.Sent.rds")
+model.natopen.time.tilst.T12.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T12.Land.rds")
+model.natopen.time.tilst.T18.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T18.Sent.rds")
+model.natopen.time.tilst.T18.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T18.Land.rds")
+model.natopen.time.tilst.T21.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T21.Sent.rds")
+model.natopen.time.tilst.T21.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.natopen.time.tilst.T21.Land.rds")
+```
+
+```{r}
 summary(model.natopen.time.tilst.T12.Sent)$coefficients$cond
+```
+```{r}
 summary(model.natopen.time.tilst.T18.Sent)$coefficients$cond
+```
+```{r}
 summary(model.natopen.time.tilst.T21.Sent)$coefficients$cond
-
+```
+```{r}
 summary(model.natopen.time.tilst.T12.Land)$coefficients$cond
+```
+```{r}
 summary(model.natopen.time.tilst.T18.Land)$coefficients$cond
+```
+```{r}
 summary(model.natopen.time.tilst.T21.Land)$coefficients$cond
 ```
-The temporal NDVI increase has been stronger in flood plains that are in reduced ecological condition, but only in the Landsat time series. The other ecosystem types do not show any obvious patterns.
-
-# Overall conclusion
+
+The temporal NDVI increase has been stronger in flood plains that are in reduced ecological condition, but only in the Landsat time series. 
+The other ecosystem types do not show any obvious patterns.
+
+## Overall conclusion
 NDVI in natopen ecosystem varies very little with condition.
@@
 
-### Eksport file (final product)
+### Eksport file {#export-ndvin}
 Nothing to export.

### 2023-10-04 NDVI_natopen.Rmd
revision of ndvi wetland, seminat & natopen bookdown-files

--- 2023-10-04 NDVI_natopen.Rmd
@@
 ### Representativity in time and space
-#### check here ####
+
 For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway.
@@
 In addition, there is 54 T2s, 57 T12s, 3 T18s, and 5 T21s that are not assigned to any region.
+(These polygons likely lie in water and thus outside of the regional polygons due to imprecise coordinates from the mapping device).
 
@@
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
-```{r message=FALSE}
+```{r message=FALSE, warning=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
 regnor <- st_intersection(reg, nor)
+```
+
+```{r, echo = F}
 rm(reg)
@@
 tm_shape(regnor) +
-  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
+  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE, alpha=0) +
   tm_borders() +
@@
     labels = c(
-      "T12 - coastal meadows",
-      "T18 - flood plains",
-      "T2 - shallow soils",
-      "T20 - innfrysningsmark",
-      "T21 - sand dunes"
+      "T12",
+      "T18",
+      "T2",
+      "T20",
+      "T21"
     ),
     title = "Major ecosystem type",
-    size = 3
+    size = 1.5
   ) +
   tm_layout(
-    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
+    main.title = "Naturally open ecosystems below \nthe tree line \n(NiN, Miljødirektoratets instruks)",
     main.title.size = 1.2, legend.position = c("right", "bottom"),
@@
 ```
+T12...Coastal meadows, T18...flood plains, T2...shallow soils, T20...innfrysningsmark, T21...sand dunes
 
@@
 summary(SentinelNDVI.natopen$area_meters)
-hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
+hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
 abline(v = 100, lty = 2)
@@
 
-natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.
+Natopen polygons stretch a large number of sizes from just above 100 sqm to almost 250 000 sqm. 75% of all polygons are below 4213 sqm.
+The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.
 
@@
 SentinelNDVI.natopen %>%
+  filter(tilstand!="") %>%
   ggplot(aes(x = year, y = mean)) +
   geom_point() +
-  facet_grid(tilstand ~ hovedtype)
+  facet_grid(tilstand ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
+  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Year")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey")
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  ggtitle("NDVI for different ecosystem types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
+  ggtitle("NDVI for different ecosystem sub-types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
+
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("NDVI for different ecosystems in good condition across regions") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Region")
 ```
@@
   filter(year == kartleggingsaar) %>%
+  filter(tilstand != "") %>%
   ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
+  ggtitle("NDVI for different ecosystems in different condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   filter(year == kartleggingsaar) %>%
+  filter(tilstand != "") %>%
+  filter(!is.na(region)) %>%
   ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_grid(region ~ hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey",) +
+  facet_grid(region ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Northern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Central Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Western Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Eastern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Southern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
 
@@
   geom_point() +
+  ggtitle("NDVI across time") +
+  labs(y = "NDVI values", x = "Year") +
   facet_wrap(~hovedtype, ncol = 1)

### 2023-10-04 NDVI_seminat.Rmd
revision of ndvi wetland, seminat & natopen bookdown-files

--- 2023-10-04 NDVI_seminat.Rmd
@@
 
-# NDVI- Semi-natural Ecosystems {#NDVI-indicator-seminat}
+# NDVI - Semi-natural Ecosystems {#NDVI-indicator-seminat}
+
+NDVI (greenness) indicator, semi-natural ecosystems
 
@@
     labels = c(
-      "T31 - boreal heath",
-      "T32 - semi-natural grasslands",
-      "T33 - semi-natural coastal grasslands",
-      "T34 - coastal heathlands",
-      "T40/41 - previously heavily altered grassland"
+      "T31",
+      "T32",
+      "T33",
+      "T34",
+      "T40/41"
     ),
     title = "Major ecosystem type",
-    size = 3
+    size = 1.5
   ) +
   tm_layout(
-    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
+    main.title = "Semi-natural ecosystems \n(NiN, Miljødirektoratets instruks)",
     main.title.size = 1.2, legend.position = c("right", "bottom"),
@@
 ```
+T31...Boreal heath, T32...semi-natural grasslands, T33...semi-natural coastal grasslands, T34...Coastal heathlands, T40/T41...previously heavily managed grasslands (now similar to T32)
 
@@
 summary(SentinelNDVI.seminat$area_meters)
-hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
+hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
 abline(v = 100, lty = 2)
@@
 sqm.
+The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.
 
@@
 SentinelNDVI.seminat %>%
+  filter(tilstand!="") %>%
   ggplot(aes(x = year, y = mean)) +
   geom_point() +
-  facet_grid(tilstand ~ hovedtype)
+  facet_grid(tilstand ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
+  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Year")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey" ) +
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 7)) +
+  ggtitle("NDVI for different ecosystem sub-types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
 ```
@@
 
-**How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
+**How does NDVI vary across regions in semi-natural ecosystems (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  theme(axis.text.x = element_text(angle = 35, vjust = 1, hjust=1))+
-  facet_wrap(~hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
+  ggtitle("NDVI for different ecosystems in different condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   group_by(id, year) %>%
+  filter(!is.na(region)) %>%
   filter(year == kartleggingsaar) %>%
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_grid(region ~ hovedtype)
+  facet_grid(region ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Central Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Western Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Eastern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Southern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)

### 2023-10-04 NDVI_wetland.Rmd
revision of ndvi wetland, seminat & natopen bookdown-files

--- 2023-10-04 NDVI_wetland.Rmd
@@
-# NDVI - Wetland ecosystems  {#NDVI-indicator-wetland}
-
-<br /> Norwegian name: **Primærproduksjon** <br />
+# NDVI - wetland ecosystems  {#NDVI-indicator-wetland}
+
+NDVI (greenness) indicator, wetland ecosystems
+
+<br />
 
 _Author and date:_
-
 Joachim Töpper, Tessa Bargmann
@@
 ```{r, echo=FALSE}
-Ecosystem <- "Våtmark"
+Ecosystem <- "våtmark"
 Egenskap <- c("Primærproduksjon")
@@
 
-## Introduction {#intro-ndviw}
-The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, 
-and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. 
+## Introduction
+The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area.
 Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance.
 Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem.
-E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. 
+E.g.
+lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems.
 In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied.
-For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). 
+For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al.
+2020).
 For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021).
@@
 
-Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e. excluding forest swamps). 
+Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open wetland ecosystems (i.e.
+excluding forest swamps).
 For this task, three initial challenges can be identified early on:
-
-1.
-We don't yet have a national wetland map, i.e.
+1) We don't yet have a national wetland map, i.e.
 we don't know the exact locations of all wetlands in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
-2.
-The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
-3.
-NDVI in wetland systems lacks the thorough ground-truthing known from globally large ecosystems as forests, tundra or Savannah.
+2) The spatial extent of the single wetland ecosystem occurrences (hereafter 'polygons') includes many wetland areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.
+3) NDVI in wetland systems lacks the thorough ground-truthing known from globally large ecosystems as forests, tundra or Savannah.
 Normal variation, like stochastic rain events and moss moisture status may influence NDVI signals greatly (Valøen, 2019).
@@
 In this draft development for a wetland NDVI indicator for ecological condition we approach these challenges in the following ways:
-
-1.
-We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
+1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem.
 However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area.
 Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of wetland ecosystems are available.
-2.
-We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
+2) We include Sentinel-2 imagery for wetland ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all wetland polygons in the NiN-data.
 However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data.
@@
 
-## About the underlying data {#underlying-data-ndviw}
+## About the underlying data
 In the NDVI project for wetlands, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:
@@
 
-1. NiN ecosystem data:
+(1) NiN ecosystem data:
 The NiN data in this study contain data about the ecosystem categories (cf.
@@
 
-2. NDVI data:
+(2) NDVI data:
 The NDVI data used in the work at hand are pixel means for polygons in the NiN-data, i.e.
@@
 
-### Representativity in time and space {#representativity-ndviw}
+### Representativity in time and space
 For wetlands, the NiN data contain 6502 wetland polygons of known area size across mainland Norway.
@@
 
-|Region            |V1    |V3    |V4   |V6   | V9   |V10|
-|------------------|------|------|-----|-----|------|---|
-|Northern Norway   |638   |275   |55   |102  | 360  |315|
-|Central Norway    |702   |1250  |34   |0    | 302  |116|
-|Western Norway    |296   |111   |30   |0    | 460  |158|
-|Eastern Norway    |251   |89    |103  |0    | 251  |223|
-|Southern Norway   |58    |44    |5    |0    | 155  |102|
+| Region          | V1  | V3   | V4  | V6  | V9  | V10 |
+|-----------------|-----|------|-----|-----|-----|-----|
+| Northern Norway | 638 | 275  | 55  | 102 | 360 | 315 |
+| Central Norway  | 702 | 1250 | 34  | 0   | 302 | 116 |
+| Western Norway  | 296 | 111  | 30  | 0   | 460 | 158 |
+| Eastern Norway  | 251 | 89   | 103 | 0   | 251 | 223 |
+| Southern Norway | 58  | 44   | 5   | 0   | 155 | 102 |
 
@@
 The NDVI data from either satellite follow the spatial distribution and bias status of the NiN-data.
-Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), 
-the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
-
-### Temporal coverage {#temp-cov-ndvis
+Due to the larger number of excluded polygons in the NDVI data from MODIS (because of the large pixel size in MODIS, see above), the MODIS NDVI data for the wetland occurrences in the NiN data are likely even more spatially biased than it is the case for the NDVI data from Sentinel-2 and Landsat.
+
+### Temporal coverage
 The currently available NiN data on wetland ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences.
@@
 
-- Available Sentinel-2 satellite NDVI data span the years 2015-2022.
-- Available MODIS satellite NDVI data span the years 2000-2022.
-- Available Landsat satellite NDVI data span the years 1984-2021.
-
-## Collinearities with other indicators {#coll-ndvis}
-The NDVI signal may likely be co-linear with [the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html) based on LiDAR derived canopy height in drained wetland areas under shrub and tree encroachment.
-
-## Reference state and values {#ref-state-ndviw}
-
-### Reference state and scaling values {#ref-state2-ndviw}
+Available Sentinel-2 satellite NDVI data span the years 2015-2022.
+Available MODIS satellite NDVI data span the years 2000-2022.
+Available Landsat satellite NDVI data span the years 1984-2021.
+
+## Collinearities with other indicators
+The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in drained wetland areas under shrub and tree encroachment.
+
+## Reference state and values
+### Reference state and scaling values
 This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments.
@@
 2020, Töpper & Jakobsson 2021).
-In this exploratory work, we will use the condition classification in the NiN-data to explore
-
-1. how NDVI connects to wetland ecology and condition and 
-2. its applicability for defining a reference state.
-
+In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to wetland ecology and condition and (2) its applicability for defining a reference state.
 We will use statistical models to test whether or not NDVI varies systematically with condition and across time.
 
-## Uncertainties {#uncertainties-ndviw}
-Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region 
-(or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a 
-measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
+## Uncertainties
+Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).
 
@@
 
-## Analyses {#analyses-ndviw}
-
-### Data sets {#datasets-ndviw}
+## Analyses
+### Data sets
 The analyses in this document make use of the following data sets:
-
-- NiN data (Naturtyper etter Miljødirektoratets instruks) 
-- Sentinel-2 NDVI data 
-- MODIS NDVI data 
-- Landsat NDVI data
-
-#### NiN data and geometry for Norway and the five regions {#nindata-ndviw}
+-NiN data (Naturtyper etter Miljødirektoratets instruks)
+-Sentinel-2 NDVI data
+-MODIS NDVI data
+-Landsat NDVI data
+
+#### NiN data and geometry for Norway and the five regions
 
@@
 
-```{r, message=FALSE}
+```{r message=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
 # change region names to something R-friendly
-#reg$region
+reg$region
 reg$region <- c("Northern Norway", "Central Norway", "Eastern Norway", "Western Norway", "Southern Norway")
@@
 
-#### NDVI data {#ndvidata-ndviw}
+#### NDVI data
 All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE).
@@
 
-##### Sentinel-2 NDVI data {#sentinelldata-ndviw}
+##### Sentinel-2 NDVI data
 The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
 The image collection contains Sentinel imagery from June, July and August 2015-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. 
-This script merges them and then merges the dataframe to the NiN data.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
 
@@
 
-##### MODIS NDVI data {#modisdata-ndviw}
+##### MODIS NDVI data
 The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
@@
 
-##### Landsat NDVI data {#landsatdata-ndviw}
+##### Landsat NDVI data
 The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
 The image collection contains Landsat imagery from June, July and August 1984-2022
-To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. 
-This script merges them and then merges the dataframe to the NiN data.
+To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
 
@@
 
-### Data handling {#data-handling-ndviw}
-#### NiN data {#nin-data-handling-ndviw}
+### Data handling
+#### NiN data
 ```{r, echo = T, eval = F}
@@
 
-```{r, include=F, eval=F}
-saveRDS(nin.wetland, paste0(here::here(), "/data/cache/nin.wetland.RDS"))
-saveRDS(nin.wetland2, paste0(here::here(), "/data/cache/nin.wetland2.RDS"))
-```
-
 ```{r, echo = F}
@@
 
-#### NDVI data {#ndvi-data-ndviw}
+#### NDVI data
 ```{r, results='hide'}
@@
 # get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
-  # dim(SentinelNDVI.wetland)
+dim(SentinelNDVI.wetland)
 SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(as.numeric(area_meters) >= 100)
-  # dim(SentinelNDVI.wetland)
+dim(SentinelNDVI.wetland)
 # split date into year, month & day
@@
 
-  # summary(SentinelNDVI.wetland)
+summary(SentinelNDVI.wetland)
 
@@
 # get rid of any nin-polygons smaller than the Modis grid cell size (62500 sqm)
-  # dim(ModisNDVI.wetland)
+dim(ModisNDVI.wetland)
 ModisNDVI.wetland <- ModisNDVI.wetland %>% filter(as.numeric(area_meters) >= 62500)
-  # dim(ModisNDVI.wetland)
+dim(ModisNDVI.wetland)
 # split date into year, month & day
@@
 # get rid of any nin-polygons smaller than the Landsat grid cell size (900 sqm)
-  # dim(LandsatNDVI.wetland)
+dim(LandsatNDVI.wetland)
 LandsatNDVI.wetland <- LandsatNDVI.wetland %>% filter(as.numeric(area_meters) >= 900)
-  # dim(LandsatNDVI.wetland)
+dim(LandsatNDVI.wetland)
 # split date into year, month & day
@@
 
-### Exploratory analyses, Sentinel-2 {#exploratory-ndviw}
+### Exploratory analyses, Sentinel-2
 
 **Where are the wetland sites?**
-```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V6 = wet snowbeds, V9 = semi-natural mires, V10 = semi-natural wet meadows.}
+```{r}
 tm_shape(regnor) +
-  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
+  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE, alpha=0) +
   tm_borders() +
@@
     title = "Major ecosystem type",
-    size = 3
+    size = 1.5
   ) +
   tm_layout(
-    main.title = "Wetland ecosystems (NiN, Miljødirektoratets instruks)",
+    main.title = "Wetland ecosystems \n(NiN, Miljødirektoratets instruks)",
     main.title.size = 1.2, legend.position = c("right", "bottom"),
@@
 ```
-
-# How are polygon sizes distributed?
-```{r histogram-sentinall-ndviw, fig.cap="Histogram of wetland ecosystem polygon sizes in Norway."}
+V1...Minerotrophic fens, V3...ombrotrophic bogs, V4...cold springs, V6...wet snowbeds, V9...semi-natural mires, V10...semi-natural wet meadows
+
+**How are polygon sizes distributed?**
+```{r}
 summary(SentinelNDVI.wetland$area_meters)
-hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000)
+hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
 abline(v = 100, lty = 2)
 ```
-
 Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
-
-**How does NDVI vary over the years**
-```{r NDVI-values-over-time-ndviw, fig.cap="A figure showing NDVI values for wetland ecosystem over time for six different main nature types in varying degrees of condition.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
+The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.
+
+**How does NDVI vary over the years?**
+```{r}
 SentinelNDVI.wetland %>%
@@
   geom_point() +
-  facet_grid(tilstand ~ hovedtype)
-```
-
+  facet_grid(tilstand ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
+  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Year")
+```
 2022 stands out with the highest NDVI values missing.
@@
 **How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r NDVI-major-ndviw, warning=FALSE, fig.cap="NDVI values in major wetland ecosystem types in good condition.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows"}
+```{r}
 SentinelNDVI.wetland %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey")
-```
-```{r NDVI-basic-ndviw, warning=FALSE, fig.cap="NDVI values in basic wetland ecosystem types in good condition.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  ggtitle("NDVI for different ecosystem types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")
+
 SentinelNDVI.wetland %>%
@@
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
-```
-
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("NDVI for different ecosystem sub-types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
+```
 NDVI varies between major ecosystem types with the highest values in semi-natural wet-meadows (V10), and lowest values in ombrotrophic bogs (V3).
@@
 **How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="DVI variation across regions in wetland ecosystems (major types) in good condition.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
+```{r}
 SentinelNDVI.wetland %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
-```
-
-There is no obvious trends in geography.
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("NDVI for different ecosystems in good condition across regions") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Region")
+```
+There is no obvious trends in geography, with the exception of V3 which shows lower values in Central Norway.
 
 **How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)**
-```{r ndvi-condition-ndviw, warning=FALSE, fig.cap="NDVI values across wetland sites in different ecological condition.
-V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}
+```{r}
 SentinelNDVI.wetland %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
-  facet_wrap(~hovedtype)
-```
-
-In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, 
-but in minertrophic fens (V1) there is not much of a pattern. 
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
+  ggtitle("NDVI for different ecosystems in different condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
+```
+In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, but in minertrophic fens (V1) there is not much of a pattern.
 Overlaps between condition classes are generally very large.
@@
 **Which ecological relevant variable(s) is the condition measure in the NiN data related to?**
-```{r important-variables-ndviwv9, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
+```{r}
 nin.wetland2 %>%
@@
   labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
-```
-```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
+
 nin.wetland2 %>%
@@
   labs(y = "Condition level (0-3='good' to 'very reduced')", x = "Condition variable score")
-```
-
-```{r important-variables-ndviwv1, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?}
+
 nin.wetland2 %>%
@@
 ```
-
 In these plots, the panels for V1, V9 and V10 differ because Miljødirektoratets mapping manual (see reference list above) uses different variables in these major ecosystem types to define condition.
@@
 In V1 (minerotrophic fens), reduced condition is mainly related to the abundance of ditches, but also increased use of heavy vehicles and high alien species abundance are related to very reduce condition.
-Under these drivers, greenness would not directly be affected, although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run. 
-**Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.**
-
-### Regression analyses Sentinel-2, NDVI as a function of condition {#regression-ndviw}
+Under these drivers, greenness would not directly be affected.
+although drainage due to ditches may lead to succession and woody plant encroachment, and thus greening, in the longer run.
+Currently, NDVI seems poorly linked to ecological condition in, and the major pressures on, minerotrophic fens.
+
+### Regression analyses Sentinel-2, NDVI as a function of condition
 We can investigate further how NDVI and condition are connected using regression analyses.
@@
 First, we take a look at how balanced the data are across major ecosystem types, regions and condition classes
-```{r balanced-data-ndviw, warning=FALSE, message=FALSE, fig.cap="NDVI values across a condition gradient, faceted by five wetland ecosystem types in five regions in Norway. "}
+```{r warning=FALSE}
 # NDVI across hovedtyper, regions and condition classes (only for NDVI years data matching NiN-mapping years)
@@
   filter(year == kartleggingsaar) %>%
+  filter(!is.na(region)) %>%
   ggplot(aes(x = tilstand, y = mean)) +
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_grid(region ~ hovedtype)
-```
-
+  facet_grid(region ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
+```
 V4 has sparse data and is lacking in some combinations of region and condition level, so we drop it.
@@
 ```
-
 NDVI values do vary between:
-
 - Regions
@@
 
-**Northern Norway**
-```{r north-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for four wetland ecosystems in Northern Norway."}
+# Northern Norway
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Northern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
-```
-```{r}
+
 summary(model.wetland.cond.Sent.N)$coefficients
 ```
-
 NDVI in good condition: V10 > V1 > V9 > V3
@@
 
-**Central Norway**
-```{r central-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in Central Norway."}
+# Central Norway
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Central Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
-```
-
-```{r}
+
 summary(model.wetland.cond.Sent.C)$coefficients
 ```
-
 NDVI in good condition: V10 > V1 & V9 > V3
@@
 
-**Western Norway**
-```{r west-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI agains ecosystm condition for fou wetland ecosystems in W-Norway."}
+# Western Norway
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Western Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
-```
-```{r}
+
 summary(model.wetland.cond.Sent.W)$coefficients
-
-```
-
+```
 NDVI in good condition: V10 > V9 & V3 > V1
@@
 
-**Eastern Norway**
-```{r east-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in E-Norway."}
+# Eastern Norway
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Eastern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
-```
-```{r}
+
 summary(model.wetland.cond.Sent.E)$coefficients
 ```
-
 NDVI in good condition: V10 > V1 > V3 & V9
@@
 
-**Southern Norway**
-```{r south-nvdi-vs-tilstand-ndviw, message=FALSE, warning=FALSE, fig.cap="NDVI against ecosystem condition for four wetland ecosystems in S-Norway."}
+# Southern Norway
+```{r message=FALSE, warning=FALSE}
 SentinelNDVI.wetland.train %>%
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 1, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Southern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
-```
-```{r}
+
 summary(model.wetland.cond.Sent.S)$coefficients
 ```
-
 NDVI in good condition: V10 > V9 > V1 & V3
@@
 
->Overall take home messages:
->
->In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
+Overall take home messages:
+
+In semi-natural wetland ecosystems (V9, V10) NDVI increases as ecosystem condition deteriorates, and this is consistent for the entire country.
 This supports the conclusion from the exploratory analyses that NDVI is indicative of ecological condition in semi-natural wetland ecosystems.
@@
 
-### NDVI across time - Sentinel, MODIS & LandSat {#across-time-ndviw}
+### NDVI across time - Sentinel, MODIS & LandSat
 As a last step, we can investigate how NDVI has changed over time. For this we include data from both MODIS and Landsat in addition to Sentinel-2.
@@
 # checking time series for MODIS
-   # ModisNDVI.wetland %>%
-   #   group_by(id, year) %>%
-   #   filter(mean == max(mean, na.rm = TRUE)) %>%
-   #   ggplot(aes(x = year, y = mean)) +
-   #   geom_point() +
-   #   facet_wrap(~hovedtype)
+ModisNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm = TRUE)) %>%
+  ggplot(aes(x = year, y = mean)) +
+  geom_point() +
+  facet_wrap(~hovedtype)
 # 2022 does not stand out as in the Sentinel data, so we keep it
@@
 # checking time series for Landsat
-   # LandsatNDVI.wetland %>%
-   #   group_by(id, year) %>%
-   #   filter(mean == max(mean, na.rm = TRUE)) %>%
-   #   ggplot(aes(x = year, y = mean)) +
-   #   geom_point() +
-   #   facet_wrap(~hovedtype)
+LandsatNDVI.wetland %>%
+  group_by(id, year) %>%
+  filter(mean == max(mean, na.rm = TRUE)) %>%
+  ggplot(aes(x = year, y = mean)) +
+  geom_point() +
+  facet_wrap(~hovedtype)
 # nothing worrying to see here either
@@
 # check if there's any 0s or 1s (which beta cannot handle)
-   # summary(SentinelNDVI.wetland$mean_beta)
-   # summary(ModisNDVI.wetland$mean_beta)
-   # summary(LandsatNDVI.wetland$mean_beta)
+summary(SentinelNDVI.wetland$mean_beta)
+summary(ModisNDVI.wetland$mean_beta)
+summary(LandsatNDVI.wetland$mean_beta)
 # replace 1s in Landsat data with 0.9999
@@
 # check if the three Satellite objects have the same structure (for concatenating them)
-   # names(SentinelNDVI.wetland)
-   # names(ModisNDVI.wetland)
-   # names(LandsatNDVI.wetland)
+names(SentinelNDVI.wetland)
+names(ModisNDVI.wetland)
+names(LandsatNDVI.wetland)
 # Sentinel and Landsat have each an extra column -> omit them when concatenating further below
@@
 # check if they have the same geometry
-   # st_crs(SentinelNDVI.wetland)
-   # st_crs(ModisNDVI.wetland)
-   # st_crs(LandsatNDVI.wetland)
+st_crs(SentinelNDVI.wetland)
+st_crs(ModisNDVI.wetland)
+st_crs(LandsatNDVI.wetland)
 # all good
@@
 allSatNDVI.wetland$Sat <- factor(allSatNDVI.wetland$Sat, levels = c("Sentinel", "Modis", "Landsat"))
-  # levels(allSatNDVI.wetland$Sat)
-```
-
+levels(allSatNDVI.wetland$Sat)
+```
 Now we can plot the time series for each main ecosystem type, showing each satellite time series separately:
-```{r ndvi-time-series-comparison-ndviw, fig.cap="Comparing the NDVI time series in the three data sets."}
+```{r}
 # plot
@@
   geom_point() +
+  ggtitle("NDVI across time") +
+  labs(y = "NDVI values", x = "Year") +
   facet_wrap(~hovedtype, ncol = 1)
 ```
-
 It is quite obvious that the NDVI values from the three satellites are not quantitatively comparable.
@@
 We first investigate a model testing NDVI as a function of year and main ecosystem type for polygons in good condition, and we do so separately for each satellite.
-
-<!-- RUNTIME approx. 15 min -->
-```{r, warning=FALSE, eval=F}
+```{r warning=FALSE}
 model.wetland.time.GodTilst.Sent <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$tilstand == "God", ])
@@
 model.wetland.time.GodTilst.Land <- glmmTMB(mean_beta ~ year * hovedtype + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$tilstand == "God", ])
-```
-
-```{r, include=FALSE, eval=F}
-saveRDS(model.wetland.time.GodTilst.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Sent.rds")
-saveRDS(model.wetland.time.GodTilst.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Modi.rds")
-saveRDS(model.wetland.time.GodTilst.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Land.rds")
-```
-
-```{r, include=F}
-model.wetland.time.GodTilst.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Sent.rds")
-model.wetland.time.GodTilst.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Modi.rds")
-model.wetland.time.GodTilst.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.GodTilst.Land.rds")
-```
-
-```{r}
+
 summary(model.wetland.time.GodTilst.Sent)$coefficients$cond
-```
-
-```{r}
 summary(model.wetland.time.GodTilst.Modi)$coefficients$cond
-```
-
-```{r}
 summary(model.wetland.time.GodTilst.Land)$coefficients$cond
 ```
-
 There's an increase in NDVI over time in all three satellite time series in essentially all major wetland ecosystem types in good condition.
@@
 
-Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, we may ask the question if the temporal NDVI increase might be stronger with reduced condition?
-```{r, warning=FALSE, eval=F}
+Given the systematically higher NDVI-values with decreasing condition in semi-natural wetlands, We may ask the question if the temporal NDVI increase might be stronger with reduced condition?
+```{r warning=FALSE}
 model.wetland.time.tilst.V9V10.Sent <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = SentinelNDVI.wetland[SentinelNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
@@
 model.wetland.time.tilst.V9V10.Land <- glmmTMB(mean_beta ~ year * tilstand + (1 | id), family = beta_family(), data = LandsatNDVI.wetland[LandsatNDVI.wetland$hovedtype %in% c("V9", "V10"), ])
-```
-
-```{r, include=FALSE, eval=F}
-saveRDS(model.wetland.time.tilst.V9V10.Sent, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Sent.rds")
-saveRDS(model.wetland.time.tilst.V9V10.Modi, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Modi.rds")
-saveRDS(model.wetland.time.tilst.V9V10.Land, "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Land.rds")
-```
-
-```{r, include=F}
-model.wetland.time.tilst.V9V10.Sent <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Sent.rds")
-model.wetland.time.tilst.V9V10.Modi <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Modi.rds")
-model.wetland.time.tilst.V9V10.Land <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/model.wetland.time.tilst.V9V10.Land.rds")
-```
-
-```{r}
 summary(model.wetland.time.tilst.V9V10.Sent)$coefficients$cond
-```
-```{r}
 summary(model.wetland.time.tilst.V9V10.Modi)$coefficients$cond
-```
-```{r}
 summary(model.wetland.time.tilst.V9V10.Land)$coefficients$cond
 ```
-
 Indeed, the temporal NDVI increase has been stronger in semi-natural wetlands with reduced ecological condition, in both Sentinel and Landsat.
@@
 
-## Overall conclusion
-
+# Overall conclusion
 NDVI in wetland ecosystem does vary with condition and also with ecosystem type but the variance and thus overlap between condition classes and ecosystem types is large.
@@
 
-1. We are lacking ecosystem maps for wetland ecosystems
+# 1) We are lacking ecosystem maps for wetland ecosystems
 The analysis at hand clearly shows that different wetland ecosystem types display different greenness levels and thus need to be evaluated separately.
 Thus, we need ecosystem maps, that allow us to treat NDVI values, and temporal changes of these, for at least the different major ecosystem types within wetlands separately.
-2. We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
+# 2) We are lacking a good ecological understanding and ground-truthing for NDVI in wetland ecosystems
 Unlike for ecosystems like forests, tundra, savannah and similar, there is little empirical work and ground-truthing data documented that link NDVI values to ecological phenomena in wetlands.
-
 Here, we need field studies investigating:
-
 - The connection between NDVI and ecological drivers like for example potential effects of Nitrogen deposition.
 This includes a need for understanding temporal increases in NDVI in ecosystems in good condition.
+
 - How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.
-
-The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
+The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (Valøen, 2019).
 Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, wetlands are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+At the same time, wetlands are generally drier in the western parts of Norway than in the East (Moen et al.
+2011), so NDVI patterns may, at least in part, be determined by moisture.
 
@@
 
-### Eksport {#export-ndviw}
+**References introduced in the conclusion:**
+
+Moen, A., Lyngstad, A. & Øien, D.-I. 2011. Kunnskapsstatus og innspill til faggrunnlag for oseanisk nedbørmyr som utvalgt naturtype. – NTNU Vitensk.mus. Rapp. bot. Ser. 2011-7: 1-72
+
+Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. NTNU
+
+### Eksport file (final product)
 <!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->

### 2023-10-04 NDVI_natopen.Rmd
more prallel-comit-issues fixed

--- 2023-10-04 NDVI_natopen.Rmd
@@
 ### Representativity in time and space {#rep-in-space-and-time-ndvin}
-<!-- #### check here #### -->
 For natopen systems, the NiN data contain 2881 natopen polygons of known area size across mainland Norway.
@@
 And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
-```{r, message=FALSE}
+```{r message=FALSE, warning=FALSE}
 # Import region- og Norgeskart, make sure they have the same geometry as the nin data
@@
     title = "Major ecosystem type",
-    size = 3
+    size = 1.5
   ) +
   tm_layout(
-    main.title = "Naturally open ecosystems below the tree line (NiN, Miljødirektoratets instruks)",
+    main.title = "Naturally open ecosystems below \nthe tree line \n(NiN, Miljødirektoratets instruks)",
     main.title.size = 1.2, legend.position = c("right", "bottom"),
@@
 ```{r histogram-sentinall-ndvin, fig.cap="Histogram of naturally open ecosystem polygon sizes in Norway."}
-#summary(SentinelNDVI.natopen$area_meters)
-hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000)
+summary(SentinelNDVI.natopen$area_meters)
+hist(SentinelNDVI.natopen$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
 abline(v = 100, lty = 2)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey")
-```
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  ggtitle("NDVI for different ecosystem types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")
+```
+
 ```{r}
@@
   filter(year == kartleggingsaar) %>%
+  filter(tilstand != "") %>%
   ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  facet_wrap(~hovedtype)
-```
-
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
+  ggtitle("NDVI for different ecosystems in different condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
+```
 The plots suggest a slight decrease in NDVI as condition deteriorates. 
 Overlaps between condition classes are generally very large.
-
-<!-- #### continue here #### -->
 
@@
   filter(year == kartleggingsaar) %>%
+  filter(tilstand != "") %>%
+  filter(!is.na(region)) %>%
   ggplot(aes(x = tilstand, y = mean)) +
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
-  facet_grid(region ~ hovedtype)
+  geom_point(size = 0.7, shape = 16, color = "grey",) +
+  facet_grid(region ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Central Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype) +
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Western Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Eastern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
   facet_wrap(~hovedtype)
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Southern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +  facet_wrap(~hovedtype)
   facet_wrap(~hovedtype)

### 2023-10-04 NDVI_wetland.Rmd
more prallel-comit-issues fixed

--- 2023-10-04 NDVI_wetland.Rmd
@@
-# NDVI - Wetland ecosystems  {#NDVI-indicator-wetland}
+# NDVI - wetland ecosystems  {#NDVI-indicator-wetland}
+
+NDVI (greenness) indicator, wetland ecosystems
 
@@
 
-|Region            |V1    |V3    |V4   |V6   | V9   |V10|
-|------------------|------|------|-----|-----|------|---|
-|Northern Norway   |638   |275   |55   |102  | 360  |315|
-|Central Norway    |702   |1250  |34   |0    | 302  |116|
-|Western Norway    |296   |111   |30   |0    | 460  |158|
-|Eastern Norway    |251   |89    |103  |0    | 251  |223|
-|Southern Norway   |58    |44    |5    |0    | 155  |102|
+| Region          | V1  | V3   | V4  | V6  | V9  | V10 |
+|-----------------|-----|------|-----|-----|-----|-----|
+| Northern Norway | 638 | 275  | 55  | 102 | 360 | 315 |
+| Central Norway  | 702 | 1250 | 34  | 0   | 302 | 116 |
+| Western Norway  | 296 | 111  | 30  | 0   | 460 | 158 |
+| Eastern Norway  | 251 | 89   | 103 | 0   | 251 | 223 |
+| Southern Norway | 58  | 44   | 5   | 0   | 155 | 102 |
 
 In addition, there is 3 V3s, 1 V4, 4 V9s, and 9 V10s that are not assigned to any region.
+(These polygons likely lie in water and thus outside of the regional polygons due to imprecise coordinates from the mapping device).
 
@@
 **Where are the wetland sites?**
+
 ```{r  where-are-wetland-sites, fig.cap="A map of Norway showing the location of the wetland ecosystem sites.
@@
 
-# How are polygon sizes distributed?
+**How are polygon sizes distributed?**
 ```{r histogram-sentinall-ndviw, fig.cap="Histogram of wetland ecosystem polygon sizes in Norway."}
 summary(SentinelNDVI.wetland$area_meters)
-hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000)
+hist(SentinelNDVI.wetland$area_meters, xlim = c(0, 20000), breaks = 50000,xlab="Area",main="")
 abline(v = 100, lty = 2)
@@
 Wetland polygons stretch a large number of sizes from just above 100 sqm to more than 4 million sqm. 75% of all polygons are below 14264 sqm.
-
-**How does NDVI vary over the years**
+The dashed line indicates the minimum size of the Sentinel-2 pixels for NDVI, all NiN-polygons smaller than that have been removed form the data.
+
+**How does NDVI vary over the years?**
 ```{r NDVI-values-over-time-ndviw, fig.cap="A figure showing NDVI values for wetland ecosystem over time for six different main nature types in varying degrees of condition.
@@
   geom_point() +
-  facet_grid(tilstand ~ hovedtype)
+  facet_grid(tilstand ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12)) +
+  ggtitle("NDVI across years for different ecosystem types in all condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Year")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey")
-```
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  ggtitle("NDVI for different ecosystem types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Major ecosystem type")
+
+```
+
 ```{r NDVI-basic-ndviw, warning=FALSE, fig.cap="NDVI values in basic wetland ecosystem types in good condition.
@@
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("NDVI for different ecosystem sub-types in good condition") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Ecosystem sub-type")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
   facet_wrap(~hovedtype) +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
-```
-
-There is no obvious trends in geography.
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("NDVI for different ecosystems in good condition across regions") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Region")
+```
+There is no obvious trends in geography, with the exception of V3 which shows lower values in Central Norway.
 
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
-  facet_wrap(~hovedtype)
-```
-
+  geom_point(size = 0.7, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  facet_wrap(~hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2)) +
+  ggtitle("NDVI for different ecosystems in different condition classes") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
+```
 In semi-natural types (V9,V10), kaldkilde (V4), and ombrotrophic bogs (V3) NDVI seems to increase as condition deteriorates, 
@@
 ```
+
 ```{r important-variables-ndviwv10, warning=FALSE, fig.cap="Which ecological relevant variable(s) is the condition measure in the NiN data related to?"}
@@
   geom_point(size = 0.7, shape = 16, color = "grey") +
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))+
-  facet_grid(region ~ hovedtype)
+  facet_grid(region ~ hovedtype) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition")
 ```
@@
   geom_violin() +
-  geom_point(size = 0.7, shape = 16, color = "grey") +
+  geom_point(size = 1, shape = 16, color = "grey", position = position_jitter(.1) ) +
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 9)) +
+  ggtitle("Southern Norway") +
+  labs(y = "NDVI values (Sentinel-2)", x = "Condition") +
   facet_wrap(~hovedtype)
@@
 - How reflectance in different wetland types is affected by how moist the moss layer is at the time of measuring.
-
-The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (REF).
+The literature suggests a high variability in NDVI measurements from mosses in particular, in part due to moisture content (Valøen, 2019).
 Minerotrophic wetlands (V1) are more dominated by graminoids than V3, and the moss-layer is generally more visible from above.
-At the same time, wetlands are generally drier in the western parts of Norway than in the East (REF), so NDVI patterns may, at least in part, be determined by moisture.
+At the same time, wetlands are generally drier in the western parts of Norway than in the East (Moen et al.
+2011), so NDVI patterns may, at least in part, be determined by moisture.
 
 In addition, further investigations should test how NDVI-condition pattern vary across biogeographic regions in addition to administrative regions (which are ecologically irrelevant), in order to better understand the results for the administrative regions.
+
+**References introduced in the conclusion:**
+
+Moen, A., Lyngstad, A. & Øien, D.-I. 2011. Kunnskapsstatus og innspill til faggrunnlag for oseanisk nedbørmyr som utvalgt naturtype. – NTNU Vitensk.mus. Rapp. bot. Ser. 2011-7: 1-72
+
+Valøen, K. 2019. Stochastic rain events increase NDVI through moss water content: a High-Arctic field experiment. NTNU
 

### 2023-10-05 R/NDVI/2_exploratory_analyses_Sentinel.R
renoame folder again

@@ -1,232 +1,0 @@
-############ working with max NDVI per year in every NiN polygon
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-
-## distribution of polygon sizes
-summary(SentinelNDVI.wetland$area_meters)
-
-hist(SentinelNDVI.wetland$area_meters,xlim=c(0,20000),breaks=50000)
-abline(v=100,lty=2)
-
-
-SentinelNDVI.seminat %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  ggplot(aes(x=as.numeric(area_meters))) + 
-  geom_histogram()
-
-SentinelNDVI.natopen %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  ggplot(aes(x=as.numeric(area_meters))) + 
-  geom_histogram()
-
-
-#### Wetland analyses, Sentinel data ####
-# how does NDVI vary over the years (all years)
-SentinelNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  
-  ggplot( aes(x=year, y=mean )) + 
-  geom_point() +
-  facet_grid( tilstand~hovedtype)
-# 2022 stands out with the highest NDVI values missing
-# the pattern is strange, suggest to omit 2022 for this analysis
-#SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
-
-# NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
-SentinelNDVI.wetland %>%
-  filter(year != '2022') %>%
-  
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  
-  filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
-  geom_violin() +
-  facet_wrap( ~hovedtype)
-# NDVI largely varies between hovedtyper: higher in seminat wet-meadows (V10), lowest in ombrotrophic bogs (V3)
-# @Tilstand: in seminat types, kaldkilde, and bogs ndvi seems to increase as condition deteriorates, in fens not much of a pattern
-
-# add column for sub-ecosystem types
[…]

### 2023-10-05 R/NDVI/2_exploratory_analyses_Sentinel_Modis_Landsat.R

@@ -1,65 +1,0 @@
-#### running a regression for the temporal trend of NDVI for each region, separately for the ecosystems and for the satellites ####
-library(betareg)
-library(glmmTMB)
-library(tidyverse)
-library(broom)
-library("gridExtra")
-
-## make NDVI variable fit for regression analyses
-# NDVI is bound between -1 and +1
-# transforming it by (x+1)/2 to be bound between 0 and 1
-# then we can use a beta regression for analysis
-SentinelNDVI.seminat$mean_beta <- (SentinelNDVI.seminat$mean + 1) / 2
-ModisNDVI.seminat$mean_beta <- (ModisNDVI.seminat$mean + 1) / 2
-LandsatNDVI.seminat$mean_beta <- (LandsatNDVI.seminat$mean + 1) / 2
-
-
-
-Sentinel.seminat.ndviTrends <- SentinelNDVI.seminat  %>% 
-  group_by(id) %>% 
-  nest()%>% 
-  mutate(model = map(data, ~betareg(mean_beta ~ year, data = .x))) %>%
-  mutate(tidy = map(model, tidy),
-         glance = map(model, glance),
-         augment = map(model, augment),
-         rsq = glance %>% map_dbl('r.squared'),
-         pvalue = glance %>% map_dbl('p.value'),
-         intercept = tidy %>% map_dbl(~ filter(.x, term == "(Intercept)") %>% pull(estimate)),
-         slope = tidy %>% map_dbl(~ filter(.x, term == "year") %>% pull(estimate))) %>%
-  dplyr::select(id, intercept, slope, rsq, pvalue)
-
-Sentinel.seminat.ndviTrends
-
-
-
-Sentinel.seminat.ndviTrends <- SentinelNDVI.seminat  %>% 
-  group_by(region_id) %>% 
-  nest()%>% 
-  mutate(model = map(data, ~betareg(mean_beta ~ year, data = .x))) %>%
-  mutate(tidy = map(model, tidy),
-         glance = map(model, glance),
-         augment = map(model, augment),
-         rsq = glance %>% map_dbl('r.squared'),
-         pvalue = glance %>% map_dbl('p.value'),
-         intercept = tidy %>% map_dbl(~ filter(.x, term == "(Intercept)") %>% pull(estimate)),
-         slope = tidy %>% map_dbl(~ filter(.x, term == "year") %>% pull(estimate))) %>%
[…]

### 2023-10-05 R/NDVI/2_regression_analyses_Sentinel.R

@@ -1,181 +1,0 @@
-# make sure to run these if not running script '2 exploratory analyses Sentinel'
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>%
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE))
-SentinelNDVI.wetland <- SentinelNDVI.wetland %>% filter(year != '2022')
-
-
-# NDVI across hovedtyper (only for NDVI years data matching NiN-mapping years)
-SentinelNDVI.wetland %>%
-  filter(year != '2022') %>%
-  
-  group_by(id, year) %>%
-  filter(mean == max(mean, na.rm=TRUE)) %>%
-  
-  filter(year == kartleggingsaar) %>%
-  ggplot( aes(x=tilstand, y=mean )) + 
-  geom_violin() +
-  facet_grid( region~hovedtype)
-# V4 has lacking data for quite some condition classes in most regions
-
-#library(glmmTMB)
-library(betareg)
-library(StepBeta)
-SentinelNDVI.wetland$mean_beta <- (SentinelNDVI.wetland$mean + 1) / 2
-
-# NDVI data from the year of NiN-mapping (and thus with condition assessment) to train the condition models
-# we drop V4 for the analysis as it lacks data for most combinations of condition and region
-SentinelNDVI.wetland.train <- SentinelNDVI.wetland %>%
-  filter(year == kartleggingsaar) %>%
-  filter(hovedtype != 'V4')
-
-# make an extra numeric condition variable
-unique(SentinelNDVI.wetland.train$tilstand)
-SentinelNDVI.wetland.train <- SentinelNDVI.wetland.train %>% mutate(tilstand_num = recode(tilstand, 
-                                            "God" = '0',
-                                            "Moderat" = '1',
-                                            "Redusert" = '2',
-                                            "Svært redusert" = '3'))
-SentinelNDVI.wetland.train$tilstand_num <- as.numeric(SentinelNDVI.wetland.train$tilstand_num)
-
[…]

### 2023-10-05 R/NDVI/Modis.R

@@ -1,43 +1,0 @@
-#Åpen Lavland - NDVI from MODIS
-
-#mean NDVI for each åpen lavland NiN polygon has been calculated from each available MODIS image in Google Earth Engine. 
-#The code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
-#The image collection contains MODIS imagery from june, july and august 2000-2022
-
-
-library(data.table)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse)
-library(sf)
-
-#############Import NiN data
-
-nin <- st_read("R:\\GeoSpatialData\\Habitats_biotopes\\Norway_Miljodirektoratet_Naturtyper_nin\\Original\\versjon20221231\\Natur_Naturtyper_nin_norge_med_svalbard_25833\\Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb")%>%
-  mutate(hovedøkosystem = recode(hovedøkosystem, 
-                                 "Våtmark" = 'Våtmark',
-                                 "Skog" = "Skog",
-                                 "Ingen" = "Ingen",
-                                 "Fjell" = "Fjell",
-                                 "Semi-naturlig mark" = 'Semi-naturlig',
-                                 "Naturlig åpne områder i lavlandet" = 'Naturlig åpne',
-                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig åpne')) %>% 
-  mutate(validGeo = st_is_valid(SHAPE)) %>%
-  filter(!hovedøkosystem %in% c('Skog','Ingen','Fjell')) %>%
-  mutate(hovedtype = substr(ninkartleggingsenheter, 4, 6),
-         hovedtype = str_remove(hovedtype, '-'),
-         id = identifikasjon_lokalid) %>%
-  filter(validGeo) %>%
-  drop_na(tilstand) %>%
-  dplyr::select(id, hovedøkosystem, hovedtype, naturtypekode, tilstand)
-
-nin
-##############Import MODIS NDVI Data
-#MODIS NDVI is scaled by 0.0001. Mean must be divided by 10000.
-df<- read.csv("P:\\41201785_okologisk_tilstand_2022_2023\\data\\NDVI_åpenlavland\\NDVI_data_MODIS\\modis_ndvi_ts_2000_2022.csv", )
-df
[…]

### 2023-10-05 NDVI_natopen.Rmd

@@
+# (PART\*) R & D{-}
+
 # NDVI - Naturally Open Ecosystems {#NDVI-indicator-natopen}
@@
 **How does NDVI vary over the years?**
-```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition.}
+```{r NDVI-values-over-time-ndvin, fig.cap="A figure showing NDVI values for naturally open ecosystem over time for six different main nature types in varying degrees of condition."}
 SentinelNDVI.natopen %>%
@@
 
-```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition.}
+```{r NDVI-basic-ndvin, warning=FALSE, fig.cap="NDVI values in basic natopen ecosystem types in good condition."}
 SentinelNDVI.natopen %>%
@@
 **How does NDVI vary across regions in natopens (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="DVI variation across regions in natopen ecosystems (major types) in good condition.}
+```{r ndvi-regions-ndvin, warning=FALSE, fig.cap="NDVI variation across regions in natopen ecosystems (major types) in good condition."}
 SentinelNDVI.natopen %>%
@@
 **How does NDVI vary across condition classes (for major types and NDVI data matching NiN-mapping years)?**
-```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition.}
+```{r ndvi-condition-ndvin, warning=FALSE, fig.cap="NDVI values across natopen sites in different ecological condition."}
 SentinelNDVI.natopen %>%

### 2023-10-05 NDVI_wetland.Rmd

@@
 **How does NDVI vary across regions in wetlands (major types) in good condition? (using only NDVI data matching NiN-mapping years)**
-```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="DVI variation across regions in wetland ecosystems (major types) in good condition.
+```{r ndvi-regions-ndviw, warning=FALSE, fig.cap="NDVI variation across regions in wetland ecosystems (major types) in good condition.
 V1 = minerotrophic mires, V3 = ombrotrophic bogs, V4 = cold springs, V9 = semi-natural mires, V10 = semi-natural wet meadows."}

### 2023-10-09 functional_plant_indicators_wetland.Rmd
text formatting

@@
 
-Norwegian name: **Planteindikatorer**
+Norwegian name: **Vegetasjon og fuktighet/lys/pH/nitrogen**
 
@@
 
+<br />
+
 Indicators described in this chaper:
@@
 ## Introduction {#intro-fpiw}
-Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of plant indicator values for plant communities (Diekmann 2003).
+Functional plant indicators can be used to describe the functional signature of plant communities by calculating community-weighted means of functional trait values for plant communities (Diekmann 2003).
 The functional signature of plant communities may be indicative of ecosystem identity, depending on which functional plant indicator we look at (cf.
@@
 Deviations in the functional signature of such an indicator beyond a certain range of indicator values (as there of course is natural variation of functional signatures within an ecosystem type) may be related to a reduction in ecological condition.
+
 Here, we combine functional indicator data with field sampled plant community data from the Norwegian nature monitoring program ANO (Tingstad et al.
@@
 
+Parallel indicator sets are explored and developed for [semi-natural ecosystems](#functional-plant-indicators-seminat) and naturally [open ecosystems](#functional-plant-indicators-open).
+
 ## About the underlying data
-In the 'plant functional indicator' project for wetlands, we use three sets of data for building indicators for ecological condition:
+We use three sets of data for building indicators for ecological condition:
 
@@
 (1) ANO monitoring data:
-ANO stands for 'areal-representativ naturovervåking', i.e.
-'area representative nature monitoring'.
+ANO stands for _areal-representativ naturovervåking_, i.e.
+_area representative nature monitoring_.
 1000 sites are randomly distributed across mainland Norway and visitied within a 5-year cycle.

### 2023-10-09 functional_plant_indicators_wetland.Rmd
Print more of the code, add fig captions, and produce scrollable text output

@@
 library(knitr)
+library(kableExtra)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-## About the underlying data
+## About the underlying data {#underlying-data-fpiw}
 We use three sets of data for building indicators for ecological condition:
@@
 
-(1) ANO monitoring data:
+(1) **ANO monitoring data**:
 ANO stands for _areal-representativ naturovervåking_, i.e.
@@
 
-(2) NiN reference data:
-The generalized species lists underlying the ecosystem categorization in NiN represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species.
-In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence (1: < 1/32, 2: 1/32 - 1/8 , 3: 1/8 - 3/8, 4: 3/8 - 4/5, 5: 3/8 - 4/5 + dominance, 6: > 4/5).
-For the purpose of this project, these steps are simplified to maximum 'expected combination of frequency and cover', whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
-
-(3) The Swedish plant indicator set published by Tyler et al.
+(2) **NiN reference data**:
+The generalized species lists underlying the ecosystem categorization in [NiN](https://artsdatabanken.no/NiN) represent expert-compiled species lists based on empirical evidence from the literature and expert knowledge of the systems and their species.
+In these lists, every species is assigned an abundance value on a 6-step scale, with each step representing a range for the 'expected combination of frequency and cover' of occurrence
+
+    -   1 = \< 1/32
+
+    -   2 = 1/32 - 1/8
+
+    -   3 = 1/8 - 3/8
+
+    -   4 = 3/8 - 4/5
+
+    -   5 = 3/8 - 4/5 + dominance
+
+    -   6 = \> 4/5
+
+For the purpose of this project, these steps are simplified to maximum _expected combination of frequency and cover_, whereby steps 4 & 5 are assigned 0.6 and 0.8, respectively, in order to distinguish between them.
+
+(3) **The Swedish plant indicator set** published by Tyler et al.
 (2021) contains a large collection of plant indicators based on the Swedish flora, which is well representative of the Norwegian flora as well.
@@
 
-### Representativity in time and space
+### Representativity in time and space {#representativity-fpiw}
 For wetlands, the ANO data contain 1351 plots in 330 sites, in principle distributed randomly across the country.
 As wetlands occur more often in certain regions of Norway than in others, the amount of plots and sites is not equal among Norway's five regions.
-The 1351 plots are distributed across regions in the following way:
+The 1351
+<!-- I cannot recreate this number. Use in-line R code, here and elsewhere in similar situations -->
+plots are distributed across regions in the following way:
 
@@
 
-### Temporal coverage
+### Temporal coverage {#representativity-temporal-fpiw}
 The evaluation data cover the first three years, 2019-2021, of the first 5-year-cycle in the ANO monitoring scheme.
@@
 ## Collinearities with other indicators
-<!-- Text here -->
-
-## Reference state and values
-### Reference state
+Collinearity could occur with other indicators that are responding to the same drivers/pressures, but no really problematic issues have been discussed.
+
+## Reference state and values {#ref-state-fpiw}
+### Reference state {#ref-state2-fpiw}
 The reference state is defined via the functional signature of the generalized species lists in NiN (see also Töpper et al.
@@
 By bootstrapping the species lists (see details further below) and calculating community-weighted means of functional plant indicators for every re-sampled community, we describe the reference state as a distribution of indicator values for each respective plant functional indicator.
-These distributions are calculated for minor ecosystem types ("grunntyper" or "kartleggingsenheter" at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN.
+These distributions are calculated for minor ecosystem types ( _grunntyper_ or _kartleggingsenheter_ at a 1:5000 mapping scale) within the major ecosystem types (hovedtyper) in NiN.
 A more extensive discussion on the use of reference communities can be found in Jakobsson et al.
@@
 
-## Uncertainties
+## Uncertainties {#uncertainties-fpiw}
 We can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error as a measure of spatial uncertainty for a geographical area.
 
-## References
+## References {#references-fpiw}
 Diekmann, M. 2003. Species indicator values as an important tool in applied plant ecology - a review. Basic and Applied Ecology 4: 493-506, doi:10.1078/1439-1791-00185
@@
 ### Data sets
-
+<!-- Please print the entire workflow with echo=T.
+Also, simply reading cached data creates loose ends - I cant find out where this data comes from.
+You may also chose to read tin this data (or a cached version of it) way at the top of the file, so that you can add some in-line R code for the summary stats in the introduction part (see previous comments)-->
 ANO data
@@
 (not shown here, but documented in the code)
-
-```{r, include = F}
+<!-- Please include everything like this, but you can chose to have it take up less space by adding the attr.output option, like this: -->
+
+```{r, include = T, attr.output='style="max-height: 300px;"'}
 
@@
 
+<!-- Print head of data set with horizontal scrolling -->
 ```{r}
-head(ANO.sp.ind)
-head(NiN.wetland.cov)
-```
+head(ANO.sp.ind) %>%
+  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+```
+
+```{r}
+head(NiN.wetland.cov) %>% 
+  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+```
+
 For each ecosystem type with a NiN species list, we can calculate a community weighted mean (CWM) for the relevant functional plant indicators.
@@
 
-```{r, echo = F}
+```{r, echo = T}
 indBoot.freq <- function(sp,abun,ind,iter,obl,rat=2/3,var.abun=F) {
@@
 
-```{r, echo = T, eval = F}
+```{r}
 colnames(NiN.wetland)
@@
 # the abundance of the re-sampled species may vary (see bootstrap function for details)
+```
+
+```{r, eval = F}
 wetland.ref.cov <- indBoot.freq(sp=NiN.wetland.cov[,1],abun=NiN.wetland.cov[,2:46],ind=NiN.wetland.cov[,47:50],
@@
 
+<!-- Please insert all the code that produces the cached data. Probably just needs to add the saveRDS() part to the code chunk above. -->
+
 ```{r, include = F}
@@
 ```{r}
-head(wetland.ref.cov[[1]])
+head(wetland.ref.cov[[1]]) %>%
+    kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+
 ```
@@
 
-```{r, echo = F}
+```{r, attr.output='style="max-height: 300px;"'}
 
@@
 ```{r}
-head(wetland.ref.cov.val)
+head(wetland.ref.cov.val) %>%
+      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+
 ```
@@
 We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
-```{r, include=FALSE}
+```{r}
 
@@
 ```{r}
-head(results.wetland[['2-sided']])
+head(results.wetland[['2-sided']]) %>%
+      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
 ```
@@
 
-### Ecosystem sub-types
+### Ecosystem sub-types {#ecosystem-sub-types-fpiw}
 And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
-```{r,echo=FALSE,message=FALSE, warning=FALSE}
+```{r fpiw-plot-maintype,echo=FALSE,message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands.
+1 = cases where variable is lower than the reference value- 2 = cases where variablei si higher than the reference value."}
 
@@
 We can also show the results as a map, for instance for pH1 (the lower pH indicator), either by directly plotting the data onto the map...
-```{r, echo=FALSE, warning=FALSE,message = FALSE}
+
+```{r ph-lower-map-fpiw, echo=FALSE, warning=FALSE,message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands."}
 #### scaled value maps ####
@@
 ```
+
 ...but here the colors and values of the data points on the map are hard to make out.

### 2023-10-12 functional_plant_indicators_wetland.Rmd

@@
+---
+output:
+  html_document:
+    code_folding: show
+---
+
 # Functional Plant Indicators - Wetlands {#Functional-plant-indicators-wetland}
@@
 <!-- Load all you dependencies here -->
-```{r setup, include=FALSE}
+```{r setup, class.source = 'fold-hide'}
 library(sf)
@@
 library(RColorBrewer)
-library("gridExtra") 
+library(gridExtra) 
 library(ggridges)
@@
 library(kableExtra)
+library(betareg)
+library(glmmTMB)
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+<!-- Limit code block height and force scrolling -->
+```{css, echo=FALSE}
+pre {
+  max-height: 300px;
+  overflow-y: auto;
+}
+
+pre[class] {
+  max-height: 200px;
+}
 ```
@@
 
-```{r, echo = F}
+```{r, class.source = 'fold-hide'}
 # Add data from cache
@@
 Plant indicators from Tyler et al. (2021)
-```{r, echo = F}
+```{r, class.source = 'fold-hide'}
 ## Tyler indicator data
@@
 Generalized species lists NiN
-```{r, echo = F}
+```{r, class.source = 'fold-hide'}
 # data from cache
@@
 
-```{r, include = T, attr.output='style="max-height: 300px;"'}
+```{r, include = T, results='hide', warning=F, message=F, class.source = 'fold-hide'}
 
@@
 ```
-leaving us with the monitoring data including plant indicators (ANO.sp.ind) and the reference data including plant indicators (NiN.wetland.cov)
+
+This leaves us with the monitoring data including plant indicators (ANO.sp.ind) and the reference data including plant indicators (NiN.wetland.cov):
 
 <!-- Print head of data set with horizontal scrolling -->
-```{r}
+```{r ano-sp-tab, class.source = 'fold-hide'}
 head(ANO.sp.ind) %>%
-  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
-```
-
-```{r}
+  kable("html",caption = "ANO species occurence data set with attached plant trait data.") %>% 
+  kable_styling("striped") %>% scroll_box(width = "100%")
+```
+
+```{r wetland-ref-data-tab, class.source = 'fold-hide'}
 head(NiN.wetland.cov) %>% 
-  kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
-```
+  kable("html", caption = "Reference data set") %>% 
+  kable_styling("striped") %>% scroll_box(width = "100%")
+```
+
+<br />
 
@@
 
-##### bootstrap function for frequency abundance
+<br />
+
+##### Bootstrap function for frequency abundance
 - function to calculate community weighted means of selected indicator values (ind)
@@
 
-Running the bootstraps
-
 ```{r}
@@
 
+Running the bootstraps:
 ```{r, eval = F}
@@
 head(wetland.ref.cov[[1]]) %>%
-    kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+    kable("html", caption = "Table showing the first 6 rows of the boostrapped data set.") %>% kable_styling("striped") %>% scroll_box(width = "100%")
 
@@
 This results in an R-list with a slot for every selected indicator, and in every slot there's a data frame with as many columns as there are NiN species lists and as many rows as there were iterations in the bootstrap.
+
 Next, we need to derive scaling values from these bootstrap-lists (the columns) for every mapping unit in NiN. Here, we define things in the following way:
@@
 
-```{r, attr.output='style="max-height: 300px;"'}
+```{r, attr.output='style="max-height: 300px;"', results='hide', class.source = 'fold-hide'}
 
@@
 ```{r}
-head(wetland.ref.cov.val) %>%
-      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+wetland.ref.cov.val %>%
+      kable("html", caption = "Reference values (Rv and threshold values (Gv) for each indicator and nature type combination") %>% kable_styling("striped") %>% scroll_box(width = "100%", height = "300px")
 
@@
 
-Here is the scaling function
+Here is the scaling function:
 ```{r}
@@
 
-We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021)
-```{r}
+We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021).
+```{r, results='hide', class.source = 'fold-hide'}
 
@@
 
-```{r, echo = T, eval = F}
+```{r, eval = F, class.source = 'fold-hide'}
 
@@
 head(results.wetland[['2-sided']]) %>%
-      kable("html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
+      kable("html", caption = "ANO plots for wetland ecosystems with scaled indicator values in the far-right columns.") %>% 
+  kable_styling("striped") %>% scroll_box(width = "100%")
 ```
@@
 And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
-```{r fpiw-plot-maintype,echo=FALSE,message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands.
-1 = cases where variable is lower than the reference value- 2 = cases where variablei si higher than the reference value."}
+```{r fpiw-plot-maintype, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands.
+1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}
 
@@
   facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
-  xlab("Main ecosystem type") + 
+  xlab("Wetland ecosystem sub-type") + 
   ylab("Scaled indicator value") 
+```
+
+This figure is a bit easier to read if we remove the sub-types with very few data points:
+```{r fpiw-plot-maintype-v2, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for the four most common sub-types of wetland.
+1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}
+ggplot(res.wetland[res.wetland$hovedtype_rute %in% c("V1", "V2", "V3", "V4"),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
+  geom_hline(yintercept=0.6, linetype="dashed") + 
+  geom_violin() +
+#  geom_boxplot(width=0.2, color="grey") +
+  geom_point(size=0.7, shape=16, color="grey") +
+  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
+  xlab("Wetland ecosystem sub-type") + 
+  ylab("Scaled indicator value") +
+  guides(fill='none')
 ```
@@
 
-```{r ph-lower-map-fpiw, echo=FALSE, warning=FALSE,message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands."}
+```{r ph-lower-map-fpiw, warning=FALSE, message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands.", class.source = 'fold-hide'}
 #### scaled value maps ####
@@
 ```{r}
-library(betareg)
-library(glmmTMB)
-
 expit <- function(L) exp(L) / (1+exp(L)) # since the beta-models use a logit link, we need to calculate the estimates back to the identity scale
@@
 
-```{r}
+```{r ph-map-fpiw, fig.cap="Mean index value by region for the lower pH indicator (i.e.
+index shows deviations towards a more acidic environment).
+Numbers in grey fields show the number of observations in the respective region."}
 
@@
 
+# Colour palettes
+myCol <- "RdYlGn"
+
 ## scaled value maps for pH1 (lower indicator)
@@
 tm_shape(regnor) +
-  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="quantile", palette=rev(get_brewer_pal(palette="OrRd", n=5, plot=FALSE))) +
+  tm_polygons(col="pH1.reg.mean", title="pH (lower), mean", style="fixed", 
+              palette=myCol,
+              breaks = seq(0,1,.2)) +
   tm_text("pH1.reg.n",col="black",bg.color="grey")
 ```
-Mean index value by region for the lower pH indicator (i.e. index shows deviations towards a more acidic environment). Numbers in grey fields show the number of observations in the respective region.
-
-```{r}
-# se
+
+We can also plot the uncertainty, here represented as the spatial variation:
+```{r ph-se-fpiw, fig.cap="Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region."}
+
 tm_shape(regnor) +
@@
 ```
-Standard error to the mean index value by region for the lower pH indicator. Numbers in grey fields show the number of observations in the respective region.
-
-### unscaled values vs. reference
+
+### Unscaled values vs. reference
 We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation.
@@
 
-```{r}
-summary(res.wetland$kartleggingsenhet_1m2)
-length(unique(res.wetland$kartleggingsenhet_1m2))
+```{r pH-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-type V1."}
+    #summary(res.wetland$kartleggingsenhet_1m2)
+    #length(unique(res.wetland$kartleggingsenhet_1m2))
 # 16 NiN-types to plot
-colnames(wetland.ref.cov[['Soil_reaction_pH']])
+    #colnames(wetland.ref.cov[['Soil_reaction_pH']])
 
@@
        col="red")
-
+```
+
+```{r ph-distcomp2-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-types other than V1."}
 par(mfrow=c(2,4))
@@
 legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
+```
+
+```{r n-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-type V1."}
 ### Nitrogen
@@
        col="red")
-
+```
+
+```{r n-distcomp2-fpiw,class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-types other than V1."}
 par(mfrow=c(2,4))
@@
 legend("topleft", legend=c("reference","field data"), pch=c(NA,1), lty=1, col=c("black","red"))
-
-```
+```
+
 Deviations from the reference occur mainly in V2-C1, which represents limestone-poor swamp forests.
@@
 ### Eksport file (final product) {#exp-fpiw}
-<!-- Export final file. Ideally a georeferenced shape or raster wit indicators values (raw and normalised), reference values and errors. -->
+Nothing exported yet.

### 2023-10-13 functional_plant_indicators_wetland.Rmd

@@
----
-output:
-  html_document:
-    code_folding: show
----
-
 # Functional Plant Indicators - Wetlands {#Functional-plant-indicators-wetland}
@@
 <!-- Load all you dependencies here -->
-```{r setup, class.source = 'fold-hide'}
+```{r setup, include=FALSE, warning=F, message=F}
 library(sf)
@@
 
-Indicators described in this chaper:
+Indicators described in this chapter:
 
@@
 You may also chose to read tin this data (or a cached version of it) way at the top of the file, so that you can add some in-line R code for the summary stats in the introduction part (see previous comments)-->
-ANO data
-
-```{r, class.source = 'fold-hide'}
+
+Import ANO data:
+```{r}
 # Add data from cache
@@
 
-Plant indicators from Tyler et al. (2021)
-```{r, class.source = 'fold-hide'}
+Import Plant indicators from Tyler et al. (2021):
+```{r}
 ## Tyler indicator data
@@
 
-Generalized species lists NiN
-```{r, class.source = 'fold-hide'}
+Import Generalized species lists NiN
+```{r}
 # data from cache
@@
 
-```{r, include = T, results='hide', warning=F, message=F, class.source = 'fold-hide'}
+```{r, include = T, results='hide', warning=F, message=F}
 
@@
 <!-- Print head of data set with horizontal scrolling -->
-```{r ano-sp-tab, class.source = 'fold-hide'}
+```{r ano-sp-tab, echo=F}
 head(ANO.sp.ind) %>%
@@
 
-```{r wetland-ref-data-tab, class.source = 'fold-hide'}
+```{r wetland-ref-data-tab, echo=FALSE}
 head(NiN.wetland.cov) %>% 
@@
 
-```{r, attr.output='style="max-height: 300px;"', results='hide', class.source = 'fold-hide'}
+```{r, attr.output='style="max-height: 300px;"', results='hide'}
 
@@
 We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021).
-```{r, results='hide', class.source = 'fold-hide'}
+```{r, results='hide'}
 
@@
 
-```{r, eval = F, class.source = 'fold-hide'}
+```{r, eval = F}
 
@@
 ```{r fpiw-plot-maintype, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands.
-1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}
-
+1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', echo=F}
 # making the plot
@@
 ```{r fpiw-plot-maintype-v2, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for the four most common sub-types of wetland.
-1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}
+1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%'}
 ggplot(res.wetland[res.wetland$hovedtype_rute %in% c("V1", "V2", "V3", "V4"),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
@@
 
-```{r ph-lower-map-fpiw, warning=FALSE, message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands.", class.source = 'fold-hide'}
+```{r ph-lower-map-fpiw, warning=FALSE, message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands."}
 #### scaled value maps ####
@@
 Since pH and nitrogen show some deviation in the lower indicator we exemplify this with these two indicators for unscaled values.
-
-```{r pH-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-type V1."}
+(Only the code for the first figure is shown).
+
+```{r pH-distcomp-fpiw, fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-type V1."}
     #summary(res.wetland$kartleggingsenhet_1m2)
@@
 
-```{r ph-distcomp2-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-types other than V1."}
+```{r ph-distcomp2-fpiw, fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-types other than V1.", echo=FALSE}
 par(mfrow=c(2,4))
@@
 
-```{r n-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-type V1."}
+```{r n-distcomp-fpiw, fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-type V1.", echo=F}
 ### Nitrogen
@@
 
-```{r n-distcomp2-fpiw,class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-types other than V1."}
+```{r n-distcomp2-fpiw, fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-types other than V1.", echo=F}
 par(mfrow=c(2,4))

### 2023-10-13 Konnetivitet.Rmd
Add files via upload

Konnektivitet

Kode
Koden er utviklet i JAVA for Google Earth Engine:
// Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå
Map.setCenter(11.0534047, 62.14137, 12);

// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine (GEE). Datasettene er delt og fitt tilgjengelig for alle fra GEE

Data 
var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND');
var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
var myrpredNN = ee.ImageCollection('users/vegarbakkestuen/Myr168NN');

var myrpredNN = myrpredNN.mean()
var myrpred = myrpred2.mean()//.rename('DTM');

// Legg til myrdata som et lag på kartet med tilpassede visningsparametere
Map.addLayer(myr, {min: 0, max: 1, palette: ['white', 'green']}, 'myr');
Map.addLayer(infrastruktur, {min: 0, max: 13, palette: ['blue', 'yellow', 'red']}, 'infrastruktur');

// Definer en bufferavstand (i meter) som bestemmer området som blir analysert rundt hvert myrpolygon
var bufferDistance = 500;

// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur
var calculateMeanDistance = function(feature) {
  // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon
  var distance = infrastruktur.rename('infrastruktur').reduceRegion({
    reducer: ee.Reducer.mean(),
    geometry: feature.geometry().buffer(bufferDistance),
    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
    maxPixels: 1e13
  }).get('infrastruktur');
  return feature.set('mean_distance', distance);
};

// Velg områder med myr (myrinnhold > 80%) og bruk dette som geometri for analysen
var omraderMedMyr = myr.gt(0.8).selfMask();

// Hent geometriene for hvert område med myr
var omraderGeometries = omraderMedMyr.reduceToVectors({
  geometry: Map.getBounds(true),
  scale: 10,  // Oppløsningen til datasettet (meters per piksel)
  maxPixels: 1e13
});

// Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr

// For scenariet "Konnektivitet med infrastruktur" 
var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance);

// Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet
Map.addLayer(omraderMedMyr, {min: 0, max: 1, palette: ['white', 'green']}, 'Myrpolygoner');

// Vis scenariet "Konnektivitet med infrastruktur" på kartet
var visParams = {
  min: 0,
  max: 1, // Juster dette basert på din infrastrukturindeksens min og maks
  palette: ['blue', 'yellow', 'red'],
};
Map.addLayer(omraderMedGjennomsnittligAvstand, visParams, 'Gjennomsnittlig avstand til infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstand,
  description:'Konnektivitet_med_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet med infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

// Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);

// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

# Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);

// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

### 2023-10-15 DRAFT_Connectivity.Rmd
Add files via upload

---
title: "Template for indicator documentation"
output:
  html_document
---

<!-- This is a template for how to document the indicator analyses.

Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name the file DRAFT_myIndicatorName.Rmd.

Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
you need to type rmarkdown::render("template.Rmd") in the console, 
as the the usual Ctrl + Alt + k will not work.

Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
easier to read the web page as a whole of the indicators are presented using the same layout.

When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
the analysis is repeatable by someone else. Be therefore very careful to annotate 
and comment on every step in the analysis. When completed, Anders will make a copy
of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->

<!-- Examples:

For an example, see 'indicators/breareal-workingFile.Rmd' 
and the subsequent 'breareal.Rmd' 
which again, when rendered, produces this:
https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html

Notice the degree of annotation.

-->

<!-- Replace 'indicator name' with yor own indicator name  -->
# Connectivity

<br />

_Author and date:_
Vegar Bakkestuen, October 2023_

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "Våtmark"
Egenskap  <- "Konnektivitet"
ECT       <- " Strukturell konnektivitet"
Contact   <- "Vegar Bakkestuen"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
Loss of natural habitat and changes in landscape patterns are central issues in biogeography and conservation biology. The loss of connectivity (connections or linkages) between suitable habitats affects individuals, populations, and communities through various ecosystem interactions. In landscape ecology, connectivity is essentially defined as "the degree to which the landscape facilitates or impedes movement between patches of resources or suitable habitats." The concept of connectivity encompasses both structural connectivity, which involves the physical locations of suitable and unsuitable habitats or patches being analyzed, and functional connectivity, which focuses on the actual movement of individuals across suitable and unsuitable habitats and disturbance barriers. We have assessed that there is currently insufficient data available to start calculating functional connectivity on a national scale in relation to the assessment of ecological condition, as there is almost a total lack of relevant data for almost all species. Therefore, we have developed an index for structural connectivity based on physical interventions calculated using an infrastructure index (Erikstad et al. 2023) and a new wetland map for Norway (Bakkestuen et al. 2023, Bakkestuen et al. in prep).
Here a general workflow for the calculation of the indicator.
1.	You must load the necessary libraries for geospatial analysis.
2.	You must load the infrastructure index and wetland data.
3.	You define the buffer distance (500 meters).
4.	You must create a function (calculateMeanDistance) to calculate the mean distance to infrastructure within a buffer around each wetland polygon.
5.	Then you filter the wetlands with more than 80% coverage.
6.	You have to apply the calculateMeanDistance function to these selected wetlands, both with and without infrastructure.
7.	You export the results as shapefiles for both scenarios.
8.	The reference values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure. The reference value is calculated as the ratio of the mean distances between myrpolygon areas in these two scenarios.

## About the underlying data
The infrastructure index is one part of the connectivity index for wetlands. The infrastructure index was calculated for three time points: 2003, 2013, and 2023. This provides a temporal perspective for assessing changes over time. The data used included information on various types of infrastructure, such as roads and buildings. Additionally, a new wetland map for Norway has been developed. This map is based on Sentinel-2 and LiDAR data from the year 2020. The U-Net model, implemented through Google Earth Engine and TensorFlow, has been used for mapping wetlands in southern Norway. The model achieved a balanced accuracy rate of 90.9% when validated against ground-truth samples, significantly improving upon the accuracy rates achieved with manually digitized maps in Norway.

### Representativity in time and space
The data for the infrastructure index covers the entire country and was collected for three time points over a 20-year period. This provides a broad and representative dataset for assessing changes in connectivity in Norwegian landscapes. The wetland mapping was conducted using data from the year 2020, making it representative for that specific time frame. However, the model used in the mapping process can be applied to other time frames if necessary. 
### Original units
The infrastructure index has no specific units as it is a measure of landscape fragmentation and disruption. It is expressed as a dimensionless index.

### Temporal coverage
The data covers three time points: 2003, 2013, and 2023. This provides a 20-year temporal perspective for assessing changes in connectivity over time. The wetland mapping provides information in the form of pixel values representing wetland presence or absence.

### Aditional comments about the dataset
It is important to note that infrastructure data are based on available map and geographical data sources. Sometimes, there may be variations and uncertainties in this data that can affect the accuracy of the infrastructure index.

## Ecosystem characteristic
In the context of the Norwegian standard, the "Connectivity" indicator aligns with the ecosystem characteristic known as "Biologisk mangfold" or "Biological diversity." This classification underscores the significance of the indicator in assessing its impact on biological diversity within ecosystems. It emphasizes how connectivity can influence biological diversity by facilitating or impeding the movement of species across habitats. Additionally, the indicator could also be relevant in the context of "funksjonelt viktige arter og strukturer," or "functionally important species and structures," particularly when assessing its effects on nutrient cycling, species recruitment, or other key ecosystem functions.
UN Standard
### SEEA EA (UN standard)
In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Connectivity" indicator can be classified as a B1 "Compositional state characteristics" indicator. This classification within the SEEA EA framework highlights the indicator's role in assessing the compositional state of ecosystems. By examining how connectivity influences the arrangement and presence of species and structures within an ecosystem, it provides valuable insights into the overall composition and structure of ecosystems.

## Collinearities with other indicators
Connectivity is not thought to exhibit collinearity with any other indicator at the present.

## Reference condition and values
### Reference condition
The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems. In this reference condition, the landscape exhibits natural and undisturbed patterns of connectivity between suitable habitats. This state represents the optimal ecological condition for connectivity assessment.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
The reference values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure. The reference value is calculated as the ratio of the mean distances between myrpolygon areas in these two scenarios. This ratio provides a value between 0 and 1, where 0 indicates a severe disruption of connectivity due to infrastructure, and 1 represents an undisturbed natural state of connectivity.
Thresholds for defining "good ecological condition" can be established based on this reference value. For instance, a reference value close to 1 would indicate that connectivity is minimally affected by infrastructure, suggesting a state of good ecological condition. In contrast, a reference value closer to 0 would signify a significant negative impact on connectivity, indicating a less desirable ecological condition. The specific thresholds may vary based on the desired ecological quality standards and policy objectives.
In the report «Vurdering av økologisk tilstand for fjell i Norge i 2021» (Framstad et al. 2021) the value of 0,8 was used as the threshold for good ecological condition.

## Uncertainties
Our wetland mapping model is a powerful tool for large-scale mapping but does come with certain uncertainties. While our U-Net model achieved an impressive balanced accuracy rate of 90.9% when validated against an independent ground-truth sample, it's essential to recognize that some level of uncertainty is inherent. The model's accuracy may vary across different wetland types and landscape conditions. The model's performance might vary in regions with distinct wetland characteristics that differ from our training data. It may not fully capture variations in wetland types or subtle differences in wetland boundaries. The accuracy of the wetland mapping model is influenced by the quality and representativeness of the training data. We made significant efforts to ensure the highest quality, but the inherent variability in ground-truth data may introduce some level of uncertainty.

The infrastructure data sources used for this analysis might contain inaccuracies, errors, or omissions. These can arise from factors like inconsistent reporting and data collection methods. While infrastructure data is a valuable proxy for landscape barriers, it may not capture all potential barriers accurately. For instance, small-scale features or unreported structures could introduce uncertainties in connectivity assessments. The resolution of the infrastructure data might not capture all relevant features at a fine scale. This can affect the precision of connectivity assessments in areas with small, localized infrastructure.

## References
Bakkestuen, V.; Venter, Z.; Ganerød, A.J.; Framstad, E. Delineation of Wetland Areas in South Norway from Sentinel-2 Imagery and LiDAR Using TensorFlow, U-Net, and Google Earth Engine. Remote Sens. 2023, 15, 1203. https://doi.org/10.3390/rs15051203
Bakkestuen, V. et al. in prep. Comprehensive Mapping of Wetland Areas in Norway: An Integration of Sentinel-2 Imagery, LiDAR, and Deep Learning Using TensorFlow and Google Earth Engine
Erikstad, L.; Simensen, T.; Bakkestuen, V.; Halvorsen, R. Index Measuring Land Use Intensity—A Gradient-Based Approach. Geomatics 2023, 3, 188-204. https://doi.org/10.3390/geomatics3010010
Framstad, E., Eide, N.E., Eide, W., Klanderud, K., Kolstad, A., Töpper, J. & Vandvik, V. 2022. Vurdering av økologisk tilstand for fjell i Norge i 2021. NINA Rapport 2050. Norsk institutt for naturforskning.

## Analyses
### Data sets
Two main data sets were included. (A) The infrastructure index (Erikstad et al. 2023) and the wetland model (Bakkestuen et al. 2023) and Bakkestuen et al. In prep.

#### Data set A
The infrastructure index utilized in this analysis was derived from data sources available through Google Earth Engine (GEE). The index quantifies the presence and distribution of key infrastructure elements such as roads, buildings, and related human-made structures across Norway. The data were available for the years 2003, 2013, and 2023, facilitating an assessment of how infrastructure has evolved in terms of both magnitude and distribution over time. It is important to note that this data forms a fundamental component in the calculation of the connectivity indicator.

#### Data set B
The wetland model used in this study was developed based on the integration of Sentinel-2 satellite imagery and LiDAR data, utilizing deep learning techniques. The model provides high-resolution information regarding the distribution and extent of wetland areas within southern Norway, specifically for the year 2020. By fusing spectral and elevation data, this model offers significant improvements over conventional, manually-digitized land cover maps. The wetland model is a critical component of the connectivity analysis, as it aids in the identification of ecologically relevant areas. The data set B for South-Norway is already accessible here: Åpen våtmark i Sør-Norge basert på Natur i Norge (NiN) typologi - Kartkatalogen (geonorge.no)

### Scaled indicator values
The connectivity indicator was created by examining the interplay between infrastructure and wetland areas. We initially calculated the average distance between wetland patches using the wetland model, representing a baseline measure of natural connectivity in the absence of infrastructure (i.e., the reference condition). Subsequently, the infrastructure index was introduced, portraying the extent of human-made structural elements within the landscape.
To assess the impact of infrastructure on connectivity, we conducted two scenarios: one considering connectivity with infrastructure and another without it. In both scenarios, we measured the average distance between wetland patches. This comparison enabled the quantification of changes in connectivity attributed to the presence of infrastructure. The results provide a clear depiction of how infrastructure influences connectivity between wetland areas, with reference values set according to the reference condition.

### Uncertainty
Wetland Model Uncertainties: The wetland model's accuracy is dependent on various factors, including the quality of Sentinel-2 imagery and the accuracy of LiDAR data. While we achieved a balanced accuracy rate of 90.9% when validating the model against an independent ground-truth sample, it's important to acknowledge that uncertainties may arise due to classification errors, misalignment between satellite imagery and LiDAR data, and variations in wetland characteristics. The accuracy rate of this model significantly surpasses traditional mapping methods but should be interpreted with awareness of these potential sources of uncertainty.
Infrastructure Index Uncertainties: The infrastructure index is based on the availability and accuracy of infrastructure data sources. Variations in the quality of data and the ability to correctly identify and classify different infrastructure elements may introduce uncertainties in the index. While efforts have been made to minimize such uncertainties, it is important to recognize that the index's accuracy may vary depending on data sources and their completeness.

## Prepare export
You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.

### Eksport file (final product)
Kode
Koden er utviklet i JAVA for Google Earth Engine:
// Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå
Map.setCenter(11.0534047, 62.14137, 12);

// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine (GEE). Datasettene er delt og fitt tilgjengelig for alle fra GEE

Data 
var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND');
var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
var myrpredNN = ee.ImageCollection('users/vegarbakkestuen/Myr168NN');

var myrpredNN = myrpredNN.mean()
var myrpred = myrpred2.mean()//.rename('DTM');

// Legg til myrdata som et lag på kartet med tilpassede visningsparametere
Map.addLayer(myr, {min: 0, max: 1, palette: ['white', 'green']}, 'myr');
Map.addLayer(infrastruktur, {min: 0, max: 13, palette: ['blue', 'yellow', 'red']}, 'infrastruktur');

// Definer en bufferavstand (i meter) som bestemmer området som blir analysert rundt hvert myrpolygon
var bufferDistance = 500;

// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur
var calculateMeanDistance = function(feature) {
  // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon
  var distance = infrastruktur.rename('infrastruktur').reduceRegion({
    reducer: ee.Reducer.mean(),
    geometry: feature.geometry().buffer(bufferDistance),
    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
    maxPixels: 1e13
  }).get('infrastruktur');
  return feature.set('mean_distance', distance);
};

// Velg områder med myr (myrinnhold > 80%) og bruk dette som geometri for analysen
var omraderMedMyr = myr.gt(0.8).selfMask();

// Hent geometriene for hvert område med myr
var omraderGeometries = omraderMedMyr.reduceToVectors({
  geometry: Map.getBounds(true),
  scale: 10,  // Oppløsningen til datasettet (meters per piksel)
  maxPixels: 1e13
});

// Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr

// For scenariet "Konnektivitet med infrastruktur" 
var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance);

// Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet
Map.addLayer(omraderMedMyr, {min: 0, max: 1, palette: ['white', 'green']}, 'Myrpolygoner');

// Vis scenariet "Konnektivitet med infrastruktur" på kartet
var visParams = {
  min: 0,
  max: 1, // Juster dette basert på din infrastrukturindeksens min og maks
  palette: ['blue', 'yellow', 'red'],
};
Map.addLayer(omraderMedGjennomsnittligAvstand, visParams, 'Gjennomsnittlig avstand til infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstand,
  description:'Konnektivitet_med_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet med infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

// Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);

// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

# Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);

// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');

// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
Export.table.toDrive({
  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
});

// The "Connectivity" reference values is created by dividing the result without infrastructure by the result that includes infrastructure. This ratio will give you a value between 0 and 1, where 0 represents complete disruption of connectivity due to infrastructure, and 1 represents undisturbed natural connectivity.

# Calculate the reference value for connectivity
reference_value <- selected_wetlands_without_infrastructure$mean_distance / selected_wetlands_with_infrastructure$mean_distance

### 2023-10-16 Nature_Index_product.Rmd
Added text to NI chapter & changed workflow parameters

@@
 
-September 2023
+October 2023
 
@@
 
-Here, we focus primarily on an NI product for wetlands, and the following table gives an overview of the indicators included in the official NI for wetlands in 2020.
+Here, we focus primarily on showcasing thw workflow for creating an NI product for wetlands, and the following table gives an overview of all the indicators included in the official NI for wetlands in 2020.
 It also provides information on each indicators trophic level, underlying data, whether we expect updated indicator data towards NI 2025, and additional notes.
@@
 For now, we recommend using data for all official NI years when calculating products with NA imputation as this will be the approach that most closely resembles the routine used for the official regular NI analyses.
+NI products can also be calculated without NA imputation.
+In that case, indicators with NA values in a given year get dropped from the NI product for that year, meaning that the same NI product for different years may contain a sightly different indicator set.
+This is no an issue unless the focus is on the time series itself, i.e.
+on the change in the characteristic that an NI product represents over time.
 
@@
 
-There is a substantial potential for collinearity between NI products and other indicators, primarily due to the fact that a range of indicators that may be part of an NI product for a given ecosystem may also enter the Ecological Condition Assessment as indicators in their own right or through other composite indicators.
+There is a substantial potential for collinearity between NI products and other indicators, primarily due to the fact that a range of indicators that may be part of an NI product for a given ecosystem have been suggested to enter the Ecological Condition Assessment also as indicators in their own right or through other composite indicators.
 For the wetlands ecosystem, this is the case, for example, for the bird indicators based on data from "Hekkefuglovervåkingen".
 These indicators form an important proportion of wetlands NI products (at least how they are defined now) but also enter into the Ecological Condition Assessment as a composite bird indicator.
-Collinearity between the two indicators/indices should be checked.
+Collinearity and double-use of data are issues here, and we discuss these at the end of the introduction of this chapter below.
 
@@
 It may well be best to use 0.6 due to its consistency with other frameworks (e.g.
-Vanndirektivet) as it would be challenging to otherwise determine what is "good enough" condition for a highly derived variable such as NI products.
+Vanndirektivet) as it would be challenging to otherwise determine what is "good enough" condition for a highly derived variables such as NI products.
 
@@
 We present some thoughts and suggestions regarding these latter questions in the next section.
-## Suggestions for using NI products in Ecological Condition Assessments [This section will hold some insights and suggestions based on a discussion to be held once Anders is back from leave.
-Content will be added later.]
+
+## Suggestions for using NI products in Ecological Condition Assessments
+
+NI products used for ecological condition assessments may deviate from the regularly published official version of NI in terms of 1) indicator selection, 2) weighting, and 3) treatment of missing data.
+
+The general rule is that NI products used for ecosystem condition assessments should follow a definition of NI as an indicator of the abundance of species.
+This means that the indicators that are part of such NI products should be indicators representing abundance and/or population status of specific species and species groups.
+Indicators that describe properties linked to the condition and/or occurrence of habitat types, on the other hand, should be dropped.
+For wetlands, for example, this includes the NI indicators "Atlantisk høgmyr areal" and "Palsmyr areal".
+The second group of indicators that should not be included in NI products are indicators that are also used as additional independent indicators in the ecosystem assessment, or that use the same underlying data as other indicators in the ecosystem assessment.
+This includes, e.g., indicators such as breeding birds, wolves, small rodents etc.
+Including the same data/information through different channel is problematic for several reasons including artificially increased indicator weight, inflated precision of the resulting aggregate metrics, high correlations among indicators, and increased risk for mistakes due to paralell workflows.
+The last group of indicators from the NI database that needs discussing are indicators based on expert knowledge.
+It has been decided that ecosystem condition assessments using the IBECA method should only include indicators based on data and model predictions.
+To what degree NI products that contain a portion of indicators based on expert knowledge should be allowed will require discussion.
+Considering all the groups of indicators that should not be included in NI products for ecosystem condition assessments opens up a discussion on whether, instead of using a composite index across a small and very unbalanced set of species (groups), a better approach might be to design NI products that represent specific taxonomic and/or functional species groups.
+Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains is eight species of vasular plants and wild reindeer.
+Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer as a separate indicator.
+
+Weighting of indicators during aggregation happens at three levels in the NI framework: area weighting, trophic weighting, and key indicator weighting. 
+Area and trophic weighting aim to smooth over imbalances in spatial and taxonomic coverage of indicators while increased weights for a priori defined key indicators are supposed to increase the influence of indicators that are thought to be functionally important species and habitats.
+Area weighting is central to the method and should always be done when calculating NI products, while the need for trophic weighting depends on the indicator selection of the specific NI product: weighting by trophic group can be useful for indicator sets that are taxonomically diverse while it is largely moot for sets of indicators representing functionally and/or taxonomically closely related species.
+For the majority of NI products for inclusion in ecosystem condition assessments key indicator weighting will not be relevant because most NI key indicators fall into the categories of either habitat indicators and/or indicators that are used independently in the assessments and are therefore removed from NI products.
+If any key indicators remain in the relevant indicator set after that, there is a good chance it will be only one or two, and allocating literally half the weight in a composite index to a single indicator is not reasonable.
+
+The last adjustment that can be made for NI products for use in ecosystem condition assessments deals with how gaps (NA values) in the data are treated.
+The official NI has always had a time series perspective, i.e.
+the focus lies on changes in the index over time.
+To be able to compare NI across different years it is necessary to impute missing indicator values as this ensures that the same (complete) set of indicators is used every year.
+The MICE algorithm that is used for the NA imputation is a practical solution that works well enough when there are many indicators and as long as the data gaps are randomly distributed across years.
+That being said, it has not been tested how well the imputation performs with smaller indicator sets, which is going to be the case with the NI products for use in the ecosystem condition assessments.
+The tidiest and most robust approach as of now is therefore to not impute missing data and calculate NI products based on indicators without missing data in the relevant time period.
+This is likely to be the best way to go as long as the focus of ecosystem condition assessments does not shift over to a primary focus on time series.
 
@@
 So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include.
-Since we may want to use NA imputation, we select all official NI years:
+Since we will not use NA imputation here, we drop the first year with available data for in the NI database (1990) as many indicators do not actually have values for this year.
 
 ```{r}
-years <- c("1990", "2000", "2010", "2014", "2019")
+years <- c("2000", "2010", "2014", "2019")
 ```
@@
 
-NAImputation <- TRUE 
+NAImputation <- FALSE 
 ```
@@
 There are two options for this: "pre-defined" drops the hard-coded set of indicators that should be dropped in any case.
-So far, there is only one indicator that gets dropped for wetlands and that is "lavhei" (due to low representativeness for the ecosystem).
-It's also possible to set it to "custom" and specify which indicators should be dropped, but we are not doing that here.
+So far, there is only one indicator that gets dropped for wetlands by default and that is "lavhei" (due to low representativeness for the ecosystem).
+The alternative is to set the mode to "custom" and specify which indicators should be dropped.
+We will demonstrate this here as we would also like to also drop the two indicators that do not represent species abundance but habitat condition, "Atlantisk høgmyr areal" and "Palsmyr areal", as well as the indicator "Tamrein" which is not relevant when considering natural biodiversity.
 
 ```{r}
-DropIndMode <- "pre-defined"
+DropIndMode <- "custom"
+DropIndList <- c(7, 142, 155, 436)
+
 if(DropIndMode == "pre-defined"){DropIndList <- NULL}
@@
 Diagnostics <- TRUE # Yes
-TestRun <- TRUE # Yes
+TestRun <- FALSE # Yes
 ```
@@
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-CustomNI<-readRDS("data/cache/CustomNI.RDS")
+CustomNI <- readRDS("data/cache/CustomNI.RDS")
 ```
@@
 ```{r, echo=FALSE}
-CustomNI_map<-readRDS("data/cache/CustomNI_map.RDS")
+CustomNI_map <- readRDS("data/cache/CustomNI_map.RDS")
 ```
@@
 
-```{r, results = FALSE, message = FALSE, warning=FALSE}
-Index <- Index <- CustomNI$CustomIndex
+```{r, results = FALSE, message = FALSE, warning = FALSE}
+Index <- CustomNI$CustomIndex
 

### 2023-10-16 DRAFT_birdIndicators.Rmd
Set up draft chapter for bird indicators

---
title: "Breeding bird indicators"
output:
  html_document
---

# Breeding bird indicators

<br />

_Author and date:_
Chloé R. Nater, Diego Pavón-Jordán

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F}
Ecosystem <- "Skog, fjell, våtmark, seminaturlig mark" 
Egenskap  <- "Biologisk mangfold" 
ECT       <- "Compositional state characteristic" 
Contact   <- "Chloé R. Nater" 
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<br />
<br />
<hr />

## Introduction
<!-- @Diego: please add a few short paragraphs on the monitoring programme (data details will be in the next section), on its role in national and international reporting, and on the linkage with PECBMS (incl. what PECBMS is and why its good/important). Don't worry too much about flow, I'll add to this section as well and try to make it into a cohesive description. -->

## About the underlying data
<!-- @Diego: please add content here and in the following four subsections -->

### Representativity in time and space
<!-- Text here -->

### Original units
<!-- Text here -->

### Temporal coverage
<!-- Text here -->

### Aditional comments about the dataset
<!-- Text here -->

## Ecosystem characteristic

### Norwegain standard

Breeding bird indicators are regarded as indicators of biodiversity ("biologisk mangfold").
It is important to note though that the indicators represent species abundance (trends) in this context and are **not** indicators of species richness or species distribution.

### SEEA EA (UN standard)
Breeding bird indicators are tagged as "Compositional state characteristic".
We develop the indicators here to be as consistent as possible with the internationally reported and well-established farmland bird index (https://agridata.ec.europa.eu/Qlik_Downloads/InfoSheetEnvironmental/infoC35.html) and common bird indices (https://www.eea.europa.eu/ims/abundance-and-distribution-of-selected).

## Collinearities with other indicators
Several species that are part of the breeding bird indicators are also part of the Norwegian Nature Index (NI). For some ecosystems, e.g. wetlands, NI is dominated by bird indicators, potentially leading to substantial collinearity with the corresponding breeding bird index. In the chapter on NI products, we suggest to remove bird indicators from NI products when using them in ecosystem condition assessments that also make use of composite breeding bird indicators.

## Reference condition and values
### Reference condition
In general, the reference condition for breeding bird indicators should be defined along the same lines as for the other indicators in ecological condition assessments.

The internationally established composite bird indicators (farmland bird index, common bird index, etc.), however, are trend indicators can be presented with a few different baseline years. It is the availability of monitoring data of sufficient quality that determines which years may be used as baselines, and the "Hekkefuglovervåkingen", the monitoring programme that supplies the data for the Norwegian breeding bird indicators, has been running since 2007/2008. For some species, a predecessor monitoring scheme has been providing data since 1996. The possible baselines for the breeding bird indicators can therefore be no earlier than 1996, and using baselines consistent with the other nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/; https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx) and internationally reported indices (https://pecbms.info/trends-and-indicators/) seems sensible.

We note that composite breeding bird indicators could also be built as Nature Index (NI) product (see separate chapters). For the species included in NI, reference values corresponding to "ecosystems with minimal human impacts" were determined previously, and an NI product for a selection of breeding birds would therefore be an indicator taking values between 0 and 1, where 1 = reference condition.
We refer to the chapter on NI products for more information (and implementation code) for this alternative approach.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
The breeding bird indicators represent average trends relative to a reference year and are therefore scaled with respect to population size in the reference year. Therefore, the indicator value in the reference year is 1. In theory, the indicators can take values between 0 (representing complete extinction of all species) and infinity (infinite population increase). 
The reference value 1 represents unchanged abundance relative to the reference year. Determining a threshold for what level of decrease in bird abundance is acceptable under "good ecological condition", however, is a challenging task. One approach to setting a threshold may be to consider thresholds of population decline used in other frameworks. The IUCN Red List of threatened species, for example, considers single species as "endangered" when a population decrease of 30% or more has been observed for the most recent decade [@iucn2001]. While it is unclear to what degree that can (or should) be extended to assemblages of multiple species, one may use this as a guideline to stipulate that >70% of the original population (indicator value 0.7) should be present (i.e. the species assemblage should not be endangered) when¨the ecosystem is in good condition. Another approach to determining a threshold would be to assume that bird populations should not be declining at all if an ecosystem is in good condition, i.e. that 1 is the threshold for good ecological condition instead of the reference value. In that case, however, one may need to come up with a reference value that is greater than 1, and determining what level of population increase would be ideal is a nearly impossible task.
We want to highlight here that these are very preliminary thoughts, and that the reference and threshold values have not been discussed in any more details yet. These discussions will nonetheless be needed, and possibly augmented with simulations, before putting the breeding bird indicators to use for ecosystem condition assessments.

## Uncertainties
<!-- Text here -->

## References

::: {#refs}
:::

## Analyses
### Data sets
<!-- Text here -->

#### Monitoring data from "Hekkefuglovervåkingen"
<!-- Text here -->

#### PECBMS Trim results from predecessor monitoring programme
<!-- Text here -->

#### Regions

As per now, the breeding bird indicators are calculated at the national scale only and the data is therefore not linked to any geofiles here.

We note that some species may be sufficiently common and have enough data to allow for calculating indicator values for two major regions, Northern Norway and Southern Norway, but this requires careful evaluation that needs to be planned and organized as part of a future project.

### PECBMS Trim analyses

### Combining data series using SWAN

### Scaled indicator values

### Uncertainty

## Prepare export

### Eksport file (final product)

### 2023-10-17 DRAFT_Connectivity.Rmd
Review with added suggestions

@@
 <br />
+Norwegian name: **Konnektivitet**
+<br />
 
 _Author and date:_
-Vegar Bakkestuen, October 2023_
+
+Vegar Bakkestuen
+
+October 2023
 
@@
 Egenskap  <- "Konnektivitet"
-ECT       <- " Strukturell konnektivitet"
+ECT       <- "Landscape and seascape characteristics"
 Contact   <- "Vegar Bakkestuen"
@@
 
-## Introduction
+## Introduction {#intro-con}
 Loss of natural habitat and changes in landscape patterns are central issues in biogeography and conservation biology.
@@
 in prep).
+
 Here a general workflow for the calculation of the indicator.
+
 1.	You must load the necessary libraries for geospatial analysis.
@@
 3.	You define the buffer distance (500 meters).
-4.	You must create a function (calculateMeanDistance) to calculate the mean distance to infrastructure within a buffer around each wetland polygon.
-5.	Then you filter the wetlands with more than 80% coverage.
-6.	You have to apply the calculateMeanDistance function to these selected wetlands, both with and without infrastructure.
+4.	You must create a function (``calculateMeanDistance``) to calculate the mean distance to infrastructure within a buffer around each wetland polygon.
+5.	Then you filter and keep only the homogeneous wetlands with more than 80% mire coverage.
+6.	You have to apply the ``calculateMeanDistance`` function to these selected wetlands, both with and without infrastructure.
 7.	You export the results as shapefiles for both scenarios.
-8.	The reference values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure.
+8.	The scaled values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure.
 The reference value is calculated as the ratio of the mean distances between myrpolygon areas in these two scenarios.
 
-## About the underlying data
-The infrastructure index is one part of the connectivity index for wetlands.
+Currently step 8, and a subsequent step 9 (to aggregate scaled indicators to regions and simultaneously calculate errors) are not complete.
+Se [this issue]()https://github.com/NINAnor/ecosystemCondition/issues/144.
+
+## About the underlying data {#underlying-data-con}
+The **infrastructure index** is one part of the connectivity index for wetlands.
 The infrastructure index was calculated for three time points: 2003, 2013, and 2023.
@@
 The data used included information on various types of infrastructure, such as roads and buildings.
-Additionally, a new wetland map for Norway has been developed.
+Additionally, a **new wetland** map for Norway has been developed.
 This map is based on Sentinel-2 and LiDAR data from the year 2020.
@@
 
-### Representativity in time and space
+### Representativity in time and space {#representativity-con}
 The data for the infrastructure index covers the entire country and was collected for three time points over a 20-year period.
@@
 However, the model used in the mapping process can be applied to other time frames if necessary.
-### Original units
+
+### Original units {#units-con}
 The infrastructure index has no specific units as it is a measure of landscape fragmentation and disruption. It is expressed as a dimensionless index.
 
-### Temporal coverage
+### Temporal coverage {#temporal-coverage-con}
 The data covers three time points: 2003, 2013, and 2023.
@@
 
-### Aditional comments about the dataset
+### Aditional comments about the dataset {#additional-comments-con}
 It is important to note that infrastructure data are based on available map and geographical data sources.
@@
 
-## Ecosystem characteristic
-In the context of the Norwegian standard, the "Connectivity" indicator aligns with the ecosystem characteristic known as "Biologisk mangfold" or "Biological diversity." This classification underscores the significance of the indicator in assessing its impact on biological diversity within ecosystems.
-It emphasizes how connectivity can influence biological diversity by facilitating or impeding the movement of species across habitats.
-Additionally, the indicator could also be relevant in the context of "funksjonelt viktige arter og strukturer," or "functionally important species and structures," particularly when assessing its effects on nutrient cycling, species recruitment, or other key ecosystem functions.
-UN Standard
-### SEEA EA (UN standard)
-In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Connectivity" indicator can be classified as a B1 "Compositional state characteristics" indicator.
-This classification within the SEEA EA framework highlights the indicator's role in assessing the compositional state of ecosystems.
-By examining how connectivity influences the arrangement and presence of species and structures within an ecosystem, it provides valuable insights into the overall composition and structure of ecosystems.
+## Ecosystem characteristic {#egenskap-con}
+In the context of the Norwegian standard, the "Connectivity" indicator aligns with the ecosystem characteristic known as **Landskapsøkologiske mønstre**.
 
-## Collinearities with other indicators
+### SEEA EA (UN standard) {#seea-con}
+In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Connectivity" indicator can be classified as a C1 "Landscape and seascape characteristics" indicator.
+
+## Collinearities with other indicators {#collinear-con}
 Connectivity is not thought to exhibit collinearity with any other indicator at the present.
 
-## Reference condition and values
-### Reference condition
+## Reference condition and values {#ref-conditions-and-values-con}
+### Reference condition {#ref-condition-con}
 The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems.
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-The reference values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure.
-The reference value is calculated as the ratio of the mean distances between myrpolygon areas in these two scenarios.
-This ratio provides a value between 0 and 1, where 0 indicates a severe disruption of connectivity due to infrastructure, and 1 represents an undisturbed natural state of connectivity.
-Thresholds for defining "good ecological condition" can be established based on this reference value.
-For instance, a reference value close to 1 would indicate that connectivity is minimally affected by infrastructure, suggesting a state of good ecological condition.
-In contrast, a reference value closer to 0 would signify a significant negative impact on connectivity, indicating a less desirable ecological condition.
-The specific thresholds may vary based on the desired ecological quality standards and policy objectives.
-In the report «Vurdering av økologisk tilstand for fjell i Norge i 2021» (Framstad et al. 2021) the value of 0,8 was used as the threshold for good ecological condition.
+### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-values-con}
+The reference values for the "Connectivity" indicator are determined by calculating the connectivity or mean distance to mire polygons, when excluding infrastructure from the analysis.
 
-## Uncertainties
+In the report «Vurdering av økologisk tilstand for fjell i Norge i 2021» (Framstad et al.
+2021) the value of 0,6 was used as the threshold for good ecological condition.
+We will tentatively use this threshold value again here.
+A perpetual discussion about the reference (and threshold values) for the connectivity indicator is maintained on [GitHub.](https://github.com/NINAnor/ecosystemCondition/issues/143)
+
+## Uncertainties {#uncertainties-con}
 Our wetland mapping model is a powerful tool for large-scale mapping but does come with certain uncertainties.
@@
 https://doi.org/10.3390/rs15051203
+
 Bakkestuen, V. et al. in prep. Comprehensive Mapping of Wetland Areas in Norway: An Integration of Sentinel-2 Imagery, LiDAR, and Deep Learning Using TensorFlow and Google Earth Engine
+
 Erikstad, L.; Simensen, T.; Bakkestuen, V.; Halvorsen, R. Index Measuring Land Use Intensity—A Gradient-Based Approach. Geomatics 2023, 3, 188-204. https://doi.org/10.3390/geomatics3010010
+
 Framstad, E., Eide, N.E., Eide, W., Klanderud, K., Kolstad, A., Töpper, J.
@@
 
-## Analyses
-### Data sets
-Two main data sets were included. (A) The infrastructure index (Erikstad et al. 2023) and the wetland model (Bakkestuen et al. 2023) and Bakkestuen et al. In prep.
+## Analyses {#analyses-con}
+### Data sets {#datasets-con}
+Two main data sets were included.
 
-#### Data set A
+A. The infrastructure index (Erikstad et al. 2023) and;
+
+B. the wetland model (Bakkestuen et al. 2023, Bakkestuen et al. _In prep._)
+
+#### Data set A - Infrastructure index {#data-a-con}
 The infrastructure index utilized in this analysis was derived from data sources available through Google Earth Engine (GEE).
@@
 
-#### Data set B
+#### Data set B - Wetland model {#data-b-con}
 The wetland model used in this study was developed based on the integration of Sentinel-2 satellite imagery and LiDAR data, utilizing deep learning techniques.
@@
 The wetland model is a critical component of the connectivity analysis, as it aids in the identification of ecologically relevant areas.
-The data set B for South-Norway is already accessible here: Åpen våtmark i Sør-Norge basert på Natur i Norge (NiN) typologi - Kartkatalogen (geonorge.no)
+The data set B for South-Norway is already accessible [here](https://kartkatalog.geonorge.no/metadata/aapen-vtmark-i-soer-norge-basert-p-natur-i-norge-nin-typologi/6b5a34f9-c23d-4cca-a51b-01251c217cb1).
 
-### Scaled indicator values
+### Scaled indicator values {#scaled-values-con}
 The connectivity indicator was created by examining the interplay between infrastructure and wetland areas.
@@
 Subsequently, the infrastructure index was introduced, portraying the extent of human-made structural elements within the landscape.
-To assess the impact of infrastructure on connectivity, we conducted two scenarios: one considering connectivity with infrastructure and another without it.
-In both scenarios, we measured the average distance between wetland patches.
-This comparison enabled the quantification of changes in connectivity attributed to the presence of infrastructure.
-The results provide a clear depiction of how infrastructure influences connectivity between wetland areas, with reference values set according to the reference condition.
+The scaled indicator is the quotient of the values.
 
-### Uncertainty
-Wetland Model Uncertainties: The wetland model's accuracy is dependent on various factors, including the quality of Sentinel-2 imagery and the accuracy of LiDAR data.
-While we achieved a balanced accuracy rate of 90.9% when validating the model against an independent ground-truth sample, it's important to acknowledge that uncertainties may arise due to classification errors, misalignment between satellite imagery and LiDAR data, and variations in wetland characteristics.
-The accuracy rate of this model significantly surpasses traditional mapping methods but should be interpreted with awareness of these potential sources of uncertainty.
-Infrastructure Index Uncertainties: The infrastructure index is based on the availability and accuracy of infrastructure data sources.
-Variations in the quality of data and the ability to correctly identify and classify different infrastructure elements may introduce uncertainties in the index.
-While efforts have been made to minimize such uncertainties, it is important to recognize that the index's accuracy may vary depending on data sources and their completeness.
+### Uncertainty {#uncertainty-con}
+The current state of the indicator is as a geospatial data set with one scaled indicator value per mire polygon, and no errors around this value.
+Future improvement must include aggregating these indicator values to regions.
+The uncertainties around these aggregated indicator values could then be represented by the spatial variation in the indicator values.
 
-## Prepare export
+## Prepare export {#pre-export-con}
 You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.
 
-### Eksport file (final product)
-Kode
+### JAVA code {#java-con}
+
 Koden er utviklet i JAVA for Google Earth Engine:
+
+```{}
 // Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå
@@
 
-// The "Connectivity" reference values is created by dividing the result without infrastructure by the result that includes infrastructure.
+```
+
+The sclaed "Connectivity" values is created by dividing the result without infrastructure by the result that includes infrastructure.
 This ratio will give you a value between 0 and 1, where 0 represents complete disruption of connectivity due to infrastructure, and 1 represents undisturbed natural connectivity.
 
-# Calculate the reference value for connectivity
+This step is not yet done, but in R this is how one would do it.
+Calculate the reference value for connectivity
+```{r, eval=FALSE}
 reference_value <- selected_wetlands_without_infrastructure$mean_distance / selected_wetlands_with_infrastructure$mean_distance
+```

### 2023-10-17 DRAFT_Connectivity.Rmd
rm duplicate java code in final code block

@@
 Ecosystem <- "Våtmark"
-Egenskap  <- "Konnektivitet"
+Egenskap  <- "Landskapsøkologiske mønstre"
 ECT       <- "Landscape and seascape characteristics"
@@
 
+It is also possible to create a time series for this indicator, but this is not yet done. See [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/145).
+
 ## About the underlying data {#underlying-data-con}
@@
 ### SEEA EA (UN standard) {#seea-con}
-In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Connectivity" indicator can be classified as a C1 "Landscape and seascape characteristics" indicator.
+In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the _Connectivity_ indicator can be classified as a **C1 Landscape and seascape characteristics** indicator.
 
@@
 
-# Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
-infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
-var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);
-
-// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
-Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');
-
-// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
-Export.table.toDrive({
-  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
-  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
-  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
-});
-
 ```
 
-The sclaed "Connectivity" values is created by dividing the result without infrastructure by the result that includes infrastructure.
+The scaled "Connectivity" values are created by dividing the result without infrastructure by the result that includes infrastructure.
 This ratio will give you a value between 0 and 1, where 0 represents complete disruption of connectivity due to infrastructure, and 1 represents undisturbed natural connectivity.

### 2023-10-18 Nature_Index_product.Rmd
Suggestions for text; added anchors

@@
 library(knitr)
+library(flextable)
 knitr::opts_chunk$set(echo = TRUE)
@@
                        "Økologisk egenskap" = Egenskap,
-                       "ECT class" = ECT)
+                       "ECT class" = ECT,
+                       check.names = F)
 knitr::kable(metaData)
@@
 
-## Introduction
+## Introduction {#intro-ni}
 
@@
 
-## About the underlying data
+## About the underlying data {#underlying-data-ni}
 
@@
 
-### Representativity in time and space
+### Representativity in time and space {#representativity-ni}
 
@@
 
-### Original units
+### Original units {#units-ni}
 
@@
 
-### Temporal coverage
+### Temporal coverage {#temporal-coverage-ni}
 
@@
 
-### Aditional comments about the dataset
+### Aditional comments about the dataset {#comments-ni}
 
@@
 
-## Ecosystem characteristic
+## Ecosystem characteristic {#ecological-char-ni}
 
-### Norwegain standard
+### Norwegain standard {#norwegian-standard-ni}
 
@@
 
-### SEEA EA (UN standard)
+### SEEA EA (UN standard) {#seea-ni}
 
@@
 
-## Collinearities with other indicators
+## Collinearities with other indicators {#collinear-ni}
 
@@
 
-## Reference condition and values
+## Reference condition and values {#ref-con-ni}
 
-### Reference condition
+### Reference condition {#ref-con2-ni}
 
@@
 
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-values-ni}
 
 As NI products are already scaled to range from 0 to 1 and reference conditions are set according to the same approach, the reference value is 1.
-For wetlands ecosystems in particular, we have not yet discussed the definition of "good ecological condition", but it seems that in the Ecological Condition Assessments for forest and mountain ecosystems, 0.6 was used as "good ecological condition" consistent with the approach from "Vanndirektivet".
+For wetlands ecosystems in particular, we have not yet discussed the definition of "good ecological condition", but in the Ecological Condition Assessments for forest and mountain ecosystems, the default 0.6 was used as "good ecological condition".
 To what degree this is appropriate is difficult to judge.
-It may well be best to use 0.6 due to its consistency with other frameworks (e.g.
-Vanndirektivet) as it would be challenging to otherwise determine what is "good enough" condition for a highly derived variables such as NI products.
+It may well be best to use 0.6 due to its consistency previous work as it would be challenging to at least empirically determine what is "good enough" condition for a highly derived variables such as NI products.
 
-## Uncertainties
+## Uncertainties {#uncertainties-ni}
 
-Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package NIcalc.
+Calculated NI products are actually composed of a pre-defined number of random samples from a probability distribution drawn during an internal bootstrap implemented in the R package ``NIcalc``.
 These draws can be summarised with metrics such as mean, median, confidence intervals.
@@
 This is linked to methodological questions regarding how (and perhaps if) highly processed composite indices should be included as indicators to make other composite indices.
-While some of these questions are of a more philosohpical nature, some more pragmatic ones include decisions about which indicators to include/exclude (especially considering indicators that may get used directly and via NI products and indicators that do not really quantify species abundance), how to weigh indicators, and whether or not to impute NA values as part of generating the NI product.
+While some of these questions are of a more philosophical nature, some more pragmatic ones include decisions about which indicators to include/exclude (especially considering indicators that may get used directly and via NI products and indicators that do not really quantify species abundance), how to weigh indicators, and whether or not to impute NA values as part of generating the NI product.
 
@@
 
-## Suggestions for using NI products in Ecological Condition Assessments
+## Recommendations for using NI products in Ecological Condition Assessments {#recomendations-ni}
 
@@
 This includes, e.g., indicators such as breeding birds, wolves, small rodents etc.
-Including the same data/information through different channel is problematic for several reasons including artificially increased indicator weight, inflated precision of the resulting aggregate metrics, high correlations among indicators, and increased risk for mistakes due to paralell workflows.
+Including the same data/information through different channels is problematic for several reasons including artificially increased indicator weight, inflated precision of the resulting aggregate metrics, high correlations among indicators, and increased risk for mistakes due to parallel workflows.
 The last group of indicators from the NI database that needs discussing are indicators based on expert knowledge.
@@
 To what degree NI products that contain a portion of indicators based on expert knowledge should be allowed will require discussion.
+It is clear, however, that expert knowledge can also be considered data.
+There is also large variation in how expert elicitation is conducted which will influence the quality and transparency of the this kind of data.
 Considering all the groups of indicators that should not be included in NI products for ecosystem condition assessments opens up a discussion on whether, instead of using a composite index across a small and very unbalanced set of species (groups), a better approach might be to design NI products that represent specific taxonomic and/or functional species groups.
-Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains is eight species of vasular plants and wild reindeer.
+Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains are eight species of vascular plants and wild reindeer.
 Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer as a separate indicator.
@@
 Weighting of indicators during aggregation happens at three levels in the NI framework: area weighting, trophic weighting, and key indicator weighting. 
-Area and trophic weighting aim to smooth over imbalances in spatial and taxonomic coverage of indicators while increased weights for a priori defined key indicators are supposed to increase the influence of indicators that are thought to be functionally important species and habitats.
+Area and trophic weighting aim to smooth over imbalances in spatial and taxonomic coverage of indicators while increased weights for _a priori_ defined key indicators are supposed to increase the influence of indicators that are thought to be functionally important species or habitats.
 Area weighting is central to the method and should always be done when calculating NI products, while the need for trophic weighting depends on the indicator selection of the specific NI product: weighting by trophic group can be useful for indicator sets that are taxonomically diverse while it is largely moot for sets of indicators representing functionally and/or taxonomically closely related species.
@@
 If any key indicators remain in the relevant indicator set after that, there is a good chance it will be only one or two, and allocating literally half the weight in a composite index to a single indicator is not reasonable.
+
+NI indicators are normalised (i.e. scaled relative to their reference values) at what is referred to as the BSU level, or basic spatial units. 
+This is the spatial resolution of the raw observation data, and the reference values have the same spatial resolution as the observation data.
+Quite often the BSU it is a municipal scale, but not always. 
+The EAA (ecosystem assessment area), however, is typically much larger, with only five regions in Norway.
+Therefore, scaled indicator valeus are aggregated to produce regional estimates, with the effect that the regional estimated now represents the (area weighted) condition of the BSUs, and not the staus of the population.
+An alternative to this, for indicators where the raw data is in the form of counts or mass (densities are not suited), one might sum the variable and the reference values for each region and then scale the variable based on these sums (e.g.
+number of individuals).
+This approach would produce an indicator which reflect the _population_ status rather than the mean status for any given area.
 
@@
 
-## References
+## References {#refs.ni}
 
@@
 
-## Analyses
+## Analyses {#analyses-ni}
 
 The following analysis workflow is almost entirely arranged in functions.
-The functions for the different steps are from the NIflex R package (<https://github.com/NINAnor/NIflex>).Many of the NIflex functions call on functions from NIcalc (<https://github.com/NINAnor/NIcalc>), the R package written for entering and extracting data from the NI database and calculating NI products from it.
+The functions for the different steps are from the ``NIflex`` R package (<https://github.com/NINAnor/NIflex>).Many of the NIflex functions call on functions from ``NIcalc`` (<https://github.com/NINAnor/NIcalc>), the R package written for entering and extracting data from the NI database and calculating NI products from it.
 Before proceeding any further, we therefore need to make sure that both of these packages and their dependencies are installed:
@@
 
-NIflex was written to provide a flexible workflow for calculating a range of NI products including the original NI for all seven main ecosystems, a range of previously used thematic indices, fully custom indices and -- most important here -- NI products aimed for use in Ecological Condition Assessments for forest, mountain, and wetland ecosystems.
+`NIflex`` was written to provide a flexible workflow for calculating a range of NI products including the original NI for all seven main ecosystems, a range of previously used thematic indices, fully custom indices and -- most important here -- NI products aimed for use in Ecological Condition Assessments for forest, mountain, and wetland ecosystems.
 
@@
 
-### Data sets
+### Data sets {#data-sets-ni}
 
-#### Nature Index database
+#### Nature Index database {#ni-database-ni}
 
@@
 
-#### Regions
+#### Regions {#regaions.ni}
 
@@
 
-### Workflow options
+### Workflow options {#workflow-options-ni}
 
-NIflex offers a workflow for calculating NI products that can be customized in terms of a range of options.
+`NIflex`` offers a workflow for calculating NI products that can be customized in terms of a range of options.
 So next, we set the global options for calculating our wetlands NI product, starting with the years of data we want to include.
@@
 Diagnostics <- TRUE # Yes
-TestRun <- FALSE # Yes
+TestRun <- FALSE # No
 ```
@@
 
-### Scaled indicator values
+### Scaled indicator values {#scaled-values-ni}
 
@@
 
-We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectivel (these are used for controlling workflow elements linked to thematic indices):
+We also need to set variables "theme" and "funArguments" to "none" and `NULL`, respectively (these are used for controlling workflow elements linked to thematic indices):
 
@@
 ```
+```{r, include=F}
+dir <- substr(getwd(), 1,2)
+pData <- ifelse(dir == "C:", 
+               "P:/41201785_okologisk_tilstand_2022_2023/data/",
+               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+saveRDS(CustomNI, paste0(pData, "cache/CustomNI.RDS"))
+```
 
 ```{r, echo=FALSE, message=FALSE, warning=FALSE}
-CustomNI <- readRDS("data/cache/CustomNI.RDS")
+CustomNI <- readRDS(paste0(pData, "cache/CustomNI.RDS"))
 ```
@@
 
-```{r, eval=FALSE}
+```{r geomapNI, eval=FALSE}
 CustomNI_map <- geomapNI(Index = CustomNI$CustomIndex, 
@@
 
-```{r, eval=FALSE}
+```{r}
 Index <- CustomNI$CustomIndex
 plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
-summary(Index$wholeArea)
 ```
-
-![](images/custPlot.png)
+```{r}
+data.frame("Region"= rownames(summary(Index$wholeArea)), summary(Index$wholeArea), check.names = F) %>%
+  flextable() %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
+```
 

### 2023-10-18 Nature_Index_product.Rmd
rm duplicate chunck

@@
 
-```{r, echo=FALSE}
-Index <- CustomNI$CustomIndex
-#plot(Index$wholeArea, main = "Custom index", cex = 1, lwd = 2, shade = TRUE)
-summary(Index$wholeArea)
-
-```
-
 The summary metrics (here median) for any given year can then be plotted onto a map by matching them up to the shapefiles:
@@
 
-### Uncertainty
+### Uncertainty {#uncertainty-ni}
 
@@
 ```
-## Prepare export
+## Prepare export {#prepare-export-ni}
 

### 2023-10-18 indipri.Rmd
add chapter to embed indipri

# IndiPri

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, out.width='100%', echo=FALSE}
knitr::include_app("https://anderskolstad.shinyapps.io/indiPri/",
                   height = "800px")
```

### 2023-10-19 DRAFT_Connectivity.Rmd
Update DRAFT_Connectivity.Rmd
Replacing java code as provided by @VegarBakkestuen

@@
 
-// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine (GEE). Datasettene er delt og fitt tilgjengelig for alle fra GEE
-
-Data 
+// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine Data
 var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND');
 var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
-var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
-var myrpredNN = ee.ImageCollection('users/vegarbakkestuen/Myr168NN');
-
-var myrpredNN = myrpredNN.mean()
-var myrpred = myrpred2.mean()//.rename('DTM');
+var myr = myrpred2.mean();
 
 // Legg til myrdata som et lag på kartet med tilpassede visningsparametere
-Map.addLayer(myr, {min: 0, max: 1, palette: ['white', 'green']}, 'myr');
-Map.addLayer(infrastruktur, {min: 0, max: 13, palette: ['blue', 'yellow', 'red']}, 'infrastruktur');
+Map.addLayer(myr, { min: 0, max: 1, palette: ['white', 'green'] }, 'myr');
+Map.addLayer(infrastruktur, { min: 0, max: 13, palette: ['blue', 'yellow', 'red'] }, 'infrastruktur');
 
@@
 
-// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur
-var calculateMeanDistance = function(feature) {
+// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur basert på grenseverdien
+var calculateMeanDistance = function (feature) {
+  var threshold = 3;  // Definer din grenseverdi for infrastruktur
+
+  // Filtrer infrastrukturbildet basert på grenseverdien
+  var filteredInfrastructure = infrastruktur.gt(threshold);
+
+  // Hent myrpolygonets sentrum
+  var myrCenter = feature.geometry().centroid(100);
+
   // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon
-  var distance = infrastruktur.rename('infrastruktur').reduceRegion({
+  var distanceWithInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
     reducer: ee.Reducer.mean(),
@@
     maxPixels: 1e13
-  }).get('infrastruktur');
-  return feature.set('mean_distance', distance);
+  }).get('filtered_infra');
+
+  var distanceWithoutInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
+    reducer: ee.Reducer.mean(),
+    geometry: myrCenter.buffer(bufferDistance),
+    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
+    maxPixels: 1e13
+  }).get('filtered_infra');
+
+  // Skalér brøkverdien til meter ved å multiplisere med bufferavstanden
+  var scaledDistanceWithInfrastructure = ee.Number(distanceWithInfrastructure).multiply(bufferDistance);
+  var scaledDistanceWithoutInfrastructure = ee.Number(distanceWithoutInfrastructure).multiply(bufferDistance);
+
+  // Beregn kvotienten (forholdet mellom avstand uten infrastruktur og avstand med infrastruktur)
+  var kvotient = scaledDistanceWithoutInfrastructure.divide(scaledDistanceWithInfrastructure);
+
+  var properties = {
+    'distinfra_meters': scaledDistanceWithInfrastructure,
+    'distnoinfra_meters': scaledDistanceWithoutInfrastructure,
+    'kvotient': kvotient
+  };
+
+  return feature.set(properties);
 };
@@
 var omraderGeometries = omraderMedMyr.reduceToVectors({
-  geometry: Map.getBounds(true),
+  geometry: geometry,
   scale: 10,  // Oppløsningen til datasettet (meters per piksel)
@@
 // Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr
-
-// For scenariet "Konnektivitet med infrastruktur" 
 var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance);
@@
 // Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet
-Map.addLayer(omraderMedMyr, {min: 0, max: 1, palette: ['white', 'green']}, 'Myrpolygoner');
+Map.addLayer(omraderMedMyr, { min: 0, max: 1, palette: ['white', 'green'] }, 'Myrpolygoner');
 
@@
   collection: omraderMedGjennomsnittligAvstand,
-  description:'Konnektivitet_med_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet med infrastruktur"
-  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
-});
-
-// Skift til scenariet "Konnektivitet uten infrastruktur" ved å fjerne infrastrukturdataene
-infrastruktur = ee.Image(0); // Nullstiller infrastruktur til et tomt bilde
-var omraderMedGjennomsnittligAvstandUtenInfrastruktur = omraderGeometries.map(calculateMeanDistance);
-
-// Vis scenariet "Konnektivitet uten infrastruktur" på kartet
-Map.addLayer(omraderMedGjennomsnittligAvstandUtenInfrastruktur, visParams, 'Gjennomsnittlig avstand uten infrastruktur');
-
-// Eksporter resultatene som en Shapefile til Google Drive for scenariet "Konnektivitet uten infrastruktur"
-Export.table.toDrive({
-  collection: omraderMedGjennomsnittligAvstandUtenInfrastruktur,
-  description:'Konnektivitet_uten_infrastruktur', // Navnet på eksportfilen for scenariet "Konnektivitet uten infrastruktur"
-  fileFormat: 'SHP' // Filformat for eksport (Shapefile)
+  description: 'Konnektivitet_med_infrastruktur',
+  fileFormat: 'SHP',
 });

### 2023-10-19 Nature_Index_product.Rmd
Update Nature_Index_product.Rmd
Set `eval = F` for saveRDS

@@
 
-```{r, include=F}
+```{r, include=F, eval=F}
 dir <- substr(getwd(), 1,2)

### 2023-10-19 DRAFT_birdIndicators.Rmd
Test for Diego
intro info

@@
 ## Introduction
-<!-- @Diego: please add a few short paragraphs on the monitoring programme (data details will be in the next section), on its role in national and international reporting, and on the linkage with PECBMS (incl.
-what PECBMS is and why its good/important).
-Don't worry too much about flow, I'll add to this section as well and try to make it into a cohesive description.
--->
+The Norwegian common breeding bird monitoring scheme (hekkefuglovervåking in Norwegian; formerly called TOV-E) was established between 2005 and 2010.
+The main goal of the programme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (available at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway.
+These data are the basis to update the Red List of Birds in Norway following the IUCN protocol, and are also used to generate indicators of mountain, farmland and forest habitats.
 
+Furthermore, these data are shared with the Pan-European Common Bird Monitoring Scheme (PECBMS; https://pecbms.info/), which, by combining data from different long-term and large-scale bird monitoring programmes in all European countries, aims at producing indicators of the general state of nature at continental scale.
+
+Data from hekkefuglovervåking are also use in scientific research both at a national and supra-national scale.
+
+The hekkefuglovervåking, coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), replaced the previous scheme, which ran between 1995 and 2008.
 ## About the underlying data
@@
 ### Representativity in time and space
-<!-- Text here -->
+Diego is on it!
 

### 2023-10-19 DRAFT_birdIndicators.Rmd
About the data

@@
 The hekkefuglovervåking, coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), replaced the previous scheme, which ran between 1995 and 2008.
+
 ## About the underlying data
-<!-- @Diego: please add content here and in the following four subsections -->
+These data are gathered mostly by volunteer experienced bird-watchers following a standardized sampling protocol (https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf).
+Observers will visit each route once a year between 20.05 and 10.06 (depending on latitude) and carry out the census in good weather conditions.
+The observer will record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route.
+A total of 229 species are typically recorded during this census.
+After 5 minutes, the observer walks to the next point in a straight line (when possible).
+During this 'transect' the observer is requested to record also a subset of species that are considered rare or difficult to capture during the 5-minute point counts (n = 121).
 
 ### Representativity in time and space
-Diego is on it!
+Nation-wide monitoring scheme consisting in 492 fix routes across all terrestrial habitats present in Norway (MAP HERE??
+I can provide this).
+Each route has between 12 and 20 counting points (ca.
+9000 across Norway), which typically are 300 m apart, describing a square with side 1.5 km long (the observer walks on a straight line between two consecutive points).
+Deviations of this typical route shape are normally due to the landscape configuration (e.g.
+rough terrain, presence of a lake or river, sea, glacier...).
+Due to lack on resources, only a random sample of 370 routes are surveyed every year.
+The aim is to survey at least 80% of the routes yearly.
+
+This monitoring scheme is designed to gather representative sample in all (common) habitats in both the latitudinal and altitudinal gradients.
+
+Although this monitoring scheme started in 2005, full coverage at national scale was achieved in 2010.
 
 ### Original units
-<!-- Text here -->
+The observers will record 'number of individuals' seen or heard during the census.
+Then, those values are converted into 'number of pairs' by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only).
+Information on bird abundance is then processed in the software TRIM to estimate abundance indices relative to a base year (in Norway this is typically 2008).
+Such abundance indices are used for reporting population trends across Norway, generate multi-species indicators in Norway and shared with PECBMS for pan-European assessment of bird populations and their habitats.
 
 ### Temporal coverage
-<!-- Text here -->
+The Norwegian common breeding bird monitoring scheme was established in 2005 in central Norway and but reached full development (nation-wide) in 2010.
+These scheme provides high quality data for many species already since 2008.
+This monitoring programme is still ongoing as of 2023.
+A previous monitoring scheme existed, which was established in 1995 and was discontinued in 2008. 
+Although these two monitoring programmes are not the same, tools available (e.g.
+RSWAN) can combine the abundance indices calculated with TRIM from the two different schemes and then combine them for a single trend estimate.
+This is possible because both monitoring schemes overlapped for one year (2008).
 
 ### Aditional comments about the dataset
-<!-- Text here -->
+From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), only 71 species are requested by PECBMS (most widely distribute with continental-wide relevance).
+Longer (combined) time series can be calculated for 55 species of which 39 species are used for the multi-species indicators (farmland = 7 species; forest = 24 species; mountain = 8 species).
 

### 2023-10-19 indipri.Rmd
add some text

@@
-# IndiPri
+# Oversikt over indikatorforslag
+
+Dette kapittelet gir en oversikt over gamle og nye forslag på indikatorer på økologisk tilstand i de tre hovedøkosystemene semi-naturlig mark, naturlig åpne områder under skoggrensa og våtmark. 
+Under vises en interaktiv applikasjon med navn IndiPri hvor man kan bla i indikatortabellen og foreta ulike filtreringer og/eller uthevinger.
+Appen vedlikeholdes gjennom en separat [GitHub repo](https://github.com/anders-kolstad/indiPri).
+
+## IndiPri
 

### 2023-10-19 gjengroing.Rmd
#59 update encroachment chapter following review

@@
 Norwegian name: **Gjengroing**
-
-_Vegetation regrowth/encroachment_
 
@@
 library(RColorBrewer)
-
+library(flextable)
 # Set global variable defining whether you want to run everything from scratch (very long runtime) or use pre-exported data (short runtime)
@@
 Ecosystem <- "Våtmark og Semi-/Naturlig-åpne" 
-Egenskap  <- "Gjengroing"
+Egenskap  <- "Funksjonell sammensetning innen trofiske nivåer"
 ECT       <- "Structural state characteristic" 
@@
                        "Økologisk egenskap" = Egenskap,
-                       "ECT class" = ECT)
-knitr::kable(metaData)
+                       "ECT class" = ECT,
+                       check.names = F)
+flextable(metaData) %>%
+  bg(bg = "lightblue", part="header") %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
 ```
@@
 
--   [NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are used to identify reference areas with good ecological condition.
+-   [Nature type polygons](https://kartkatalog.miljodirektoratet.no/Dataset/Details/2031) are used to identify reference areas with good ecological condition.
 -   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type.
@@
 -   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/).
-This includes both a terrain (DTM) and surface mode (DSM).
+This includes both a terrain (DTM) and surface model (DSM).
 We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground.
@@
 
-Not sure what this means??
+Funksjonell sammensetning innen trofiske nivåer, reflecting the change in dominant growth form.
 
@@
 
-There is possibly a collinearity with the primary production indicator (primærproduksjon).
+There is possibly a collinearity with the [primary production indicator](#NDVI-indicator-natopen) (primærproduksjon).
 The primary production indicator uses the normalized difference vegetation index (NDVI) as a proxy for vegetation production.
@@
 
-The methodology used to calculate the gjengroing indicator is outlined in the schematic below.
+The methodology used to calculate the _gjengroing_ indicator is outlined in the schematic below.
 The workflow in the schematic is conducted for all reference and population polygons in Norway and repeated for each ecosystem type (åpne and våtmark), respectively.
@@
 
-![](images/gjengroing_schematic.jpg){width="70%"}
+```{r enc-workflow, fig.cap="Schematic illustration of how the encroachment indicator is calculated.
+The lower bound for poor condition is set by LiDAR heights from mature forest surrounding the population polygon.
+The upper bound for good condition is set by regional median LiDAR heights within NiN polygons in the same bioclimatic-elevation zone as the population polygon.", out.width='70%', echo=FALSE}
+knitr::include_graphics("images/gjengroing_schematic.jpg")
+```
 
@@
 
+The reference state is defined as one where åpne ecosystems are indeed open, with little to no woody encroachment. 
+For natural ecosystems this reflects a state where climate warming, wetland drainage, and alien species etc have not led to any net loss of these nature types and their internal functioning. 
+For semi natural ecosystems it reflects a state where traditional husbandry is keeping he ecosystems open due to grazing and hay making.
+
+### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}
 The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state.
-We use the all-encompassing "Tilstand" variable assigned to each NiN polygon.
+We use the aggregated "Tilstand" variable assigned to each NiN polygon.
 The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie.
@@
 
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}
-
 Once we have the reference vegetation height for a given ecosystem type and elevation-bioclimatic zone, we need to define the minimum (or worst/bad) condition.
@@
 
+We acknowledge that the AR5 data set has several shortcomings in terms of lags in update frequency and with precision in the maps themselves. 
+This indicator will therefore benefit strongly from new ecosystem delineation maps, that update more frequently, and preferably once that can differentiate between semi-natural and naturally open ecosystems.
+
 The vegetation height percentiles are then scaled to between 0 and 1 using a sigmoid transformation [Oliver at al. (2021)](https://www.sciencedirect.com/science/article/pii/S1470160X21000066).
@@
 There are several datasets which are used in GEE which will not be imported into the R session here.
-These datasets have been obtained from the source and ingested into GEE by Zander Venter or Vegard Bakkestuen with the help of Miljædata section at NINA.
+These datasets have been obtained from the source and ingested into GEE by Zander Venter or Vegard Bakkestuen with the help of Miljødata section at NINA.
 They include
@@
 ```{r}
-regions <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/regions.shp') %>%
-  mutate(region = factor(region) )
-# Rename regions with correct Norwegian characters
-levels(regions$region) <- c("Østlandet", "Midt-Norge","Nord-Norge","Sørlandet","Vestlandet")
+regions <- sf::st_read("data/regions.shp", options = "ENCODING=UTF8") %>%
+  mutate(region = factor(region))
 ```
@@
 
-The Moen's bioclimatic regions are imported from NINA R/GeoSpatialData/ server.
+The [Moen's bioclimatic regions](https://data.artsdatabanken.no//Natur_i_Norge/Natursystem/Beskrivelsessystem/Regional_naturvariasjon/Bioklimatisk_sone) are imported from NINA R/GeoSpatialData/ server.
 
@@
 
-[NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2050) are imported here from the NINA servers:
+[NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2031) are imported here from the NINA servers:
 
@@
 
-Once you have run the GEE script above, and downloaded the data to the /data/gjengroing/From_GEE directory, you can proceed with this R workflow.
+Once you have run the GEE script above, and downloaded the data to the `/gjengroing/R_project/DATA/From_GEE/` directory, you can proceed with this R workflow.
 
@@
   refVaatmark %>%
-    st_write('data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
+    st_write('data/encroachment/refVaatmark.csv')
   refAapne %>%
-    st_write('data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
+    st_write('data/encroachment/refAapne.csv')
 
@@
 
-  refVaatmark  <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refVaatmark.csv')
-  refAapne <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/refAapne.csv')
-
-}
-
+  refVaatmark  <- read_csv('data/encroachment/refVaatmark.csv')
+  refAapne <- read_csv('data/encroachment/refAapne.csv')
+
+}
+
+```
+
+```{r ref-open-enc, fig.cap="Refenerence vegetation height in semi-natural and naturally open ecosystems under the forest line in Norway."}
+refAapne %>%
+  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
+         vegClimZoneLab = factor(vegClimZoneLab)) %>%
+  ggplot(aes(x = elevation, y = ref))+
+  geom_bar(stat="identity")+
+  theme_bw(base_size = 12)+
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
+  facet_wrap(.~vegClimZoneLab)+
+  labs(y = "Reference vegetation height (m)",
+       x = "Elevation (meters above sea level)")
+```
+
+```{r ref-open-enc, fig.cap="Refenerence vegetation height in wetland ecosystems under the forest line in Norway."}
+refVaatmark %>%
+  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
+         vegClimZoneLab = factor(vegClimZoneLab)) %>%
+  ggplot(aes(x = elevation, y = ref))+
+  geom_bar(stat="identity")+
+  theme_bw(base_size = 12)+
+  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
+  facet_wrap(.~vegClimZoneLab)+
+  labs(y = "Reference vegetation height (m)",
+       x = "Elevation (meters above sea level)")
 ```
@@
 
-```{r fig.height=5, fig.width=15}
+```{r coverage-map-enc, fig.height=5, fig.width=15, fig.cap="Map of Norway showing coverage of different data sets."}
 
@@
   geom_sf(data=regions, fill=NA, size=0.3) +
-  scale_fill_gradientn(colours = c('#e6bf6b', '#91b5a2'),
+  scale_fill_gradientn(colours = c('black', 'orange'),
                        name = '% cover',
@@
 
-```{r}
+```{r sones-and-bands, fig.cap="Bioclimatic sones and elevation bands used to define strata in the enchroachment indicator."}
 
@@
 
-Here we will visualize the spatial variation in scaled gjengroind indicator values, as well as the vegetation heights for reference (good condition), indicator, and mature forest (poor condition) over the country.
+Here we will visualize the spatial variation in scaled gjengroings indicator values, as well as the vegetation heights for reference (good condition), indicator, and mature forest (poor condition) over the country.
 We could plot all the polygons, but it is computationally intensive and very difficult to visualize spatial variations over the entire country.
@@
 
-```{r, fig.width=10}
+```{r multiple-maps-enc, fig.width=10, fig.cap="Maps of scaled and unscaled indicator values, as well as reference values.
+For the unscaled variables the legend is truncated at 5 m to ensure the same colour gradient for all maps."}
 
@@
 
-```{r}
+```{r grid-level-enc, fig.cap="Maps displaying scaled encroachment indicator values (A, E) along with the raw vegetation heights defining upper limits for good (B, F), and bad (D, H) condition, and observed population vegetation heights (C, G).
+Values are aggregaed to polygons defining unique bioclimatic-elevation strata over the country."}
 
@@
 -   **Quantification of uncertainty around indicator scores.** Here we used the standard deviation in the vegetation heights for reference (NiN) polygons within each region to quantify the uncertainty around scaled indicator scores.
-We did this because the method used in eaTools::ea_spread() function is a bootstrapping approach which resulted in extremely small uncertainty estimates due to the large number of polygons included in our analysis.
-In the future, a better method for quantifying uncertainty is needed.
-Perhaps by running sensitivity analyses that test several variations of the points mentioned above can form the basis for quantifying an error margin around indicator values.
+We did this because the method used in `eaTools::ea_spread()` function is a bootstrapping approach which resulted in extremely small uncertainty estimates due to the large number of polygons included in our analysis.
+[In the future, a better method for quantifying uncertainty is needed.](https://github.com/NINAnor/eaTools/issues/17) Perhaps by running sensitivity analyses that test several variations of the points mentioned above can form the basis for quantifying an error margin around indicator values.
 
@@
 The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m.
-Therefore it gets a scaped indicator score of 0.1.
+Therefore it gets a scaled indicator score of 0.1.
 In contrast the wetland polygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.

### 2023-10-19 gjengroing.Rmd

@@
 <br />
+
+<!-- Limit code block height and force scrolling -->
+```{css, echo=FALSE}
+pre {
+  max-height: 300px;
+  overflow-y: auto;
+}
+
+pre[class] {
+  max-height: 200px;
+}
+```
 
@@
 
-```{r ref-open-enc, fig.cap="Refenerence vegetation height in wetland ecosystems under the forest line in Norway."}
+```{r ref-wet-enc, fig.cap="Refenerence vegetation height in wetland ecosystems under the forest line in Norway."}
 refVaatmark %>%

### 2023-10-19 Nature_Index_product.Rmd
#151

--- 2023-10-18 Nature_Index_product.Rmd
@@
                        check.names = F)
-knitr::kable(metaData)
+flextable(metaData) %>%
+  bg(bg = "lightblue", part="header") %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
 ```

### 2023-10-19 slitasje.Rmd
Specify that indicators are unique also to each ecosystem #83
New table. Fig for report.

@@
 
-Norwegian indicator name: Slitasje
+Norwegian indicator name: **Slitasje**
 
@@
 library(tibble)
+library(flextable)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-### Temporal coverage
+### Temporal coverage {#temoporal-sli}
 
@@
 
+```{r, include=FALSE, eval=FALSE}
+# Figure for the report 2023
+
+corrCheck %>%
+  filter(hovedøkosystem == "naturligÅpneOmråderILavlandet") %>%
+ggplot(aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
+  geom_bar(position="fill")+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Skalerte indikatorverdier"))+
+  ylab("Andel av datapunkter")+
+  xlab("Infrastrukturindeksen (kategorisk versjon)")+
+  scale_fill_brewer(palette = "RdYlGn")
+```
 #### Looking at the HIA

### 2023-10-19 DRAFT_birdIndicators.Rmd
A bit on data analysis
to be continued...

@@
 ### Data sets
-<!-- Text here -->
+All data used are gathered under the Norwegian common breeding bird monitoring scheme (Hekkeguglovervåking in Norwegian; formely named TOV-E).
+For the calculation of multi-species indicators (MSI) of farmland and forest habitats, however, a longer dataset is used by combininng data from the 'predecessor monitoring programme', whic ran from 1995 to 2008.
 
 #### Monitoring data from "Hekkefuglovervåkingen"
-<!-- Text here -->
+The route-based data are compiled after each breeding season and stored at NINA's database.
+The route-based data are then used to calculate the long-term trends (since 2008) in abundance for XX species, which have sufficient data to properly estimate the yearly abundance indices at a National scale.
 
 #### PECBMS Trim results from predecessor monitoring programme
-<!-- Text here -->
+The abundance indices generated by TRIM using data from the predecessor scheme are stored in different repositories.
+These remain unchanged as this programme was discontinued in 2008 and no new data are being collected.
 

### 2023-10-20 NDVI_wetland.Rmd
updating tables and minor edits

@@
                        "Økologisk egenskap" = Egenskap,
-  "ECT class" = ECT
-)
-knitr::kable(metaData)
+                       "ECT class" = ECT,
+                       check.names = F)
+flextable(metaData) %>%
+  bg(bg = "lightblue", part="header") %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
 ```

### 2023-10-20 functional_plant_indicators_natopen.Rmd
updating tables and minor edits

@@
-# Functional Plant Indicators - Naturally Open Ecosystems {#Functional-plant-indicators-open}
+# Functional Plant Indicators - Naturally Open Ecosystems {#functional-plant-indicators-open}
 

### 2023-10-20 insectIndicators.Rmd
updating tables and minor edits

@@
-<!-- This is a template for how to document the indicator analyses.
-
-Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name te file DRAFT_myIndicatorName.Rmd.
-
-Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
-you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work.
-
-Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout.
-
-When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
-the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. When completed, Anders will make a copy
-of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
-the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
-
-<!-- Examples:
-
-For an example, see 'indicators/breareal-workingFile.Rmd' 
-and the subsequent 'breareal.Rmd' 
-which again, when rendered, produces this:
-https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
-
-Notice the degree of annotation.
-
--->
-
-<!-- Replace 'indicator name' with yor own indicator name  -->
 # Insect indicators
@@
 _Author and date:_
+
 Jens Åström `r Sys.Date()`
@@
 require(tidyverse)
+require(flextable)
 })
@@
 ```{r, echo=F}
-Ecosystem <- rep("Semi-naturlig mark", 3) # e.g. "Skog og fjell"
-Egenskap  <- c("Biologisk mangfold, Funksjonelle grupper innen trofiske nivåer, Funksjonelt viktige arter og biofysiske strukturer") # e.g. "Primærproduksjon"
-ECT       <- rep("Functional state characteristics", 3) # e.g. "Structural state characteristic"
-Contact   <- rep("Jens Åström", 3) # e.g. "Anders Kolstad"
+Ecosystem <- rep("Semi-naturlig mark", 3) 
+Egenskap  <- c("Biologisk mangfold", 
+               "Funksjonelle grupper innen trofiske nivåer", 
+                "Funksjonelt viktige arter og biofysiske strukturer") 
+ECT       <- rep("Functional state characteristics", 3) 
+Contact   <- rep("Jens Åström", 3)
 ```
@@
 ```{r, echo=F}
-metaData <- tibble(Ecosystem,
+metaData <- data.frame(Ecosystem,
                        "Økologisk egenskap" = Egenskap,
-                       "ECT class" = ECT)
-knitr::kable(metaData)
+                       "ECT class" = ECT,
+                       check.names = F)
+flextable(metaData) %>%
+  bg(bg = "lightblue", part="header") %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
 ```

### 2023-10-20 rosslyng.Rmd
updating tables and minor edits

@@
 library(sf)
+library(flextable)
 knitr::opts_chunk$set(echo = TRUE)

### 2023-10-20 Nature_Index_product.Rmd
Fixed typo

@@
 The EAA (ecosystem assessment area), however, is typically much larger, with only five regions in Norway.
-Therefore, scaled indicator valeus are aggregated to produce regional estimates, with the effect that the regional estimated now represents the (area weighted) condition of the BSUs, and not the staus of the population.
+Therefore, scaled indicator valeus are aggregated to produce regional estimates, with the effect that the regional estimated now represents the (area weighted) condition of the BSUs, and not the status of the population.
 An alternative to this, for indicators where the raw data is in the form of counts or mass (densities are not suited), one might sum the variable and the reference values for each region and then scale the variable based on these sums (e.g.

### 2023-10-20 DRAFT_Flood_frequency.Rmd
First comitt for Flood Frequency
Here is the first version

---
title: "Template for indicator documentation"
output:
  html_document
---

<!-- This is a template for how to document the indicator analyses.

Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
Name the file DRAFT_myIndicatorName.Rmd.

Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
you need to type rmarkdown::render("template.Rmd") in the console, 
as the the usual Ctrl + Alt + k will not work.

Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
easier to read the web page as a whole of the indicators are presented using the same layout.

When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
the analysis is repeatable by someone else. Be therefore very careful to annotate 
and comment on every step in the analysis. When completed, Anders will make a copy
of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->

<!-- Examples:

For an example, see 'indicators/breareal-workingFile.Rmd' 
and the subsequent 'breareal.Rmd' 
which again, when rendered, produces this:
https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html

Notice the degree of annotation.

-->

<!-- Replace 'indicator name' with yor own indicator name  -->
# Flood frequency

<br />

_Author and date:_
Vegar Bakkestuen, October 2023_

<br />

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Fill in which ecosystem the indicator belongs to, as well as the ecosystem characteristic it should be linked to. It's OK to use some Norwegian here -->
```{r, echo=F}
Ecosystem <- "Våtmark"
Egenskap  <- "Oversvømmelse"
ECT       <- " Oversvømmelseshyppighet"
Contact   <- "Vegar Bakkestuen"
```

```{r, echo=F}
metaData <- data.frame(Ecosystem,
                       "Økologisk egenskap" = Egenskap,
                       "ECT class" = ECT)
knitr::kable(metaData)
```

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

<!-- Document you work below. Try not to change  the headers too much. Data can be stored on NINA server. Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine. If it is not too big you may store under /data/ on this repository -->

## Introduction
Water and inundated areas are a land cover class that can be relatively easily distinguished from other land cover classes with high accuracy using remote sensing data, particularly Copernicus Sentinel-1 data. Sentinel-1 employs Synthetic Aperture Radar (SAR) instruments that can "see through" clouds. In Norway, Sentinel-1 data is available at a minimum frequency of once a week.
Utilizing Google Earth Engine, all available Sentinel-1 data can be accessed, and there are pre-established scripts for distinguishing water-covered areas from other land cover classes. For example, you can refer to "Flood Mapping and Area Calculation of Flood Extent Using Sentinel-I SAR Data in Google Earth Engine: the case of Super Typhoon Odette (Rai)" [^1^] and "Applying Google Earth Engine for flood mapping and monitoring in the downstream provinces of the Mekong river" [^2^].
By aggregating Sentinel-1 images over an entire growing season, it becomes possible to model when areas are covered by water. This approach has been successfully used in various countries to map deltas and floodplains that are periodically submerged. We aim to apply this methodology to the deltas and floodplains in Norway to assess the accuracy of this method in Norwegian conditions.
The code is developed in Java for Google Earth Engine and calculates flood frequency over a five-year period, which can be extended to ten years. Reference states need to be defined both before and after flood events, which is accomplished by specifying pre- and post-flood dates. You can determine these dates using hydrology scripts.
This script uses Sentinel-1 imagery in a change detection approach, comparing images before and after a flood event. The key preprocessing steps for Sentinel-1 Ground Range Detected (GRD) data are thermal noise removal, radiometric calibration, and terrain correction. Therefore, the primary preprocessing step is applying a speckle filter.
In this script, you can define your study area, select your desired time frame, and set SAR parameters. Make sure to adjust these parameters according to your study area's specific conditions. 
Workflow: Calculate Flood Indicator Using Google Earth Engine Script
1.	Initialization:
•	Set up your Google Earth Engine environment.
•	Import the necessary geometries or AOI.
2.	Define Functions:
•	Define a function for calculating flood extent and area based on the provided script. The function should take parameters like the year, polarization, and AOI.
3.	Define Constants:
•	Set constant values such as polarization, pass direction, smoothing radius, and difference threshold. Ensure these values are appropriate for your study area.
4.	Define Years and Date Ranges:
•	Create an array or list of years for which you want to calculate the flood indicator.
•	Define the "before" and "after" date ranges for each year based on your specific study periods.
5.	Loop Through Years:
•	Use a loop to iterate through each year in your defined list.
•	For each year, call the flood extent calculation function you defined, passing the appropriate parameters (e.g., SAR collections for "before" and "after," polarization, AOI).
6.	Calculate Flood Frequency:
•	After calculating flood extent for each year, calculate the flood frequency. You can create a frequency count image by accumulating flood extent layers from each year.
7.	Data Visualization:
•	Define a color palette for displaying flood frequency on the map.
•	Add the flood frequency layer to the map for visualization.
•	Optionally, you can also display individual flood extents for each year.
8.	Print Results:
•	In the loop, after calculating flood extent for each year, print the calculated flood extent areas to the console for each year.
9.	Data Export (Optional):
•	If needed, you can export the results, such as the frequency map, for further analysis or reporting.
10.	Fine-Tuning:
•	As needed, adjust parameters in the script, such as smoothing radius, difference threshold, or additional data sources, to optimize the flood indicator calculation.
11.	Interpretation:
•	Analyze the results to understand the flood extent areas and frequencies for different years.

## About the underlying data
The flood frequency indicator is derived from a time series of Sentinel-1 SAR data covering multiple years. The analysis involves the following steps:
1.	Data Collection: Sentinel-1 GRD imagery, including the VH polarization, was used to create a comprehensive dataset. The data covered several years, enabling a temporal perspective on flood frequency.
2.	Preprocessing: The data underwent preprocessing, including thermal-noise removal, radiometric calibration, and terrain correction, resulting in high-quality radar backscatter information. A speckle filter was applied for data refinement.
3.	Flood Extent Calculation: Flood extents for each year were determined through a change detection approach, comparing "before" and "after" flood event images. The selected polarization, VH, was utilized to differentiate flooded areas.
4.	Flood Frequency Mapping: The accumulated flood extents for different years were used to create a flood frequency map. This map illustrates the frequency of flooding events in the study area over the analyzed period.

### Representativity in time and space
The analysis spans multiple years, offering a long-term view of flooding events in the study area. This extended time frame enables the identification of recurrent patterns and trends, which is vital for understanding wetland dynamics. 
### Original units
The original units of the calculated flood extent area are in square meters.
### Temporal coverage
The data covers the last five years. It is possible to go back to 2015.

### Aditional comments about the dataset
Sentinel-1 data provides valuable information; however, its spatial and temporal resolution can affect the accuracy of flood extent mapping. In this analysis, a 10-meter spatial resolution was used, which means that smaller-scale features may not be fully captured. Additionally, Sentinel-1 data has a temporal revisit time, which may result in missed flood events or difficulties in distinguishing between different flood events that occurred close in time.

## Ecosystem characteristic
In the context of the Norwegian standard, the 'Flood frequency' indicator aligns with the ecosystem characteristic known as 'Biologisk mangfold' or 'Biological diversity.' This classification underscores the significance of the indicator in assessing its impact on biological diversity within ecosystems. Flood frequency is a crucial factor in understanding how changes in hydrology, including flood events, influence the composition and distribution of species within ecosystems. It can also be instrumental in evaluating the resilience of ecosystems to such disturbances.
Furthermore, the indicator could also be relevant in the context of 'funksjonelt viktige arter og strukturer,' or 'functionally important species and structures.' Flood frequency plays a pivotal role in shaping the habitat availability and accessibility for species critical to ecosystem functioning. This includes species that are essential for nutrient cycling, species recruitment, and other key ecosystem functions. Frequent or infrequent flood events can have far-reaching consequences on the viability of these functionally important components of ecosystems.
Therefore, understanding and monitoring flood frequency not only serve as a hydrological metric but also have profound implications for ecosystem dynamics, which are vital for the overall health and sustainability of ecosystems. This integrated perspective highlights the interconnectedness of hydrological patterns and ecosystem characteristics, emphasizing the indicator's significance in ecological and conservation assessments.
UN Standard
### SEEA EA (UN standard)
In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Flood frequency" indicator can be classified as a B1 "Compositional state characteristics" indicator. This classification within the SEEA EA framework highlights the indicator's role in assessing the compositional state of ecosystems. By examining how connectivity influences the arrangement and presence of species and structures within an ecosystem, it provides valuable insights into the overall composition and structure of ecosystems.

## Collinearities with other indicators
Flood frequency is not thought to exhibit collinearity with any other indicator at the present.

## Reference condition and values
### Reference condition
The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems. In this reference condition, the landscape exhibits natural and undisturbed patterns of connectivity between suitable habitats. This state represents the optimal ecological condition for connectivity assessment.

### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
Reference values for hydrological parameters, such as water level, water flow, and temperature, are crucial for assessing the ecological condition of aquatic ecosystems. In the context of this analysis, we are planning to derive reference values from extensive datasets collected by the Norwegian Water Resources and Energy Directorate (NVE) through its hydrological API (HydAPI). NVE has been instrumental in gathering long-term hydrological data, with records dating back to the 1800s. These historical measurements serve as a valuable baseline to evaluate contemporary conditions. W 
Data Retrieval Solution - Python Code (get-observations-NINA.py): This Python script (get-observations-NINA.py) serves as a data retrieval solution to access hydrological data from the Norwegian Water Resources and Energy Directorate's (NVE) Hydrological API (HydAPI). It is designed to fetch observational data and metadata for hydrological stations across Norway. The script is versatile and allows users to specify parameters such as the station ID, observed parameters, resolution time, reference time, and a mandatory API key.
•	Station ID (-s): Users can input one or multiple station IDs separated by a comma, allowing access to specific hydrological data from various locations.
•	Parameters (-p): This parameter allows users to specify the type of observation data they need. Multiple parameters can be specified, enhancing flexibility.
•	Resolution Time (-r): Users can select the resolution time, choosing between instantaneous (0), hourly (60), or daily (1440) data.
•	API Key (-a): The API key is a mandatory component to authenticate and access the HydAPI data. Users need to register and obtain their API key from the HydAPI platform.
The script leverages Python's request library to send HTTP requests to the HydAPI, and it retrieves data in JSON format. It also provides the ability to define a reference time for data retrieval. The retrieved data is written to a user-defined output file, facilitating further analysis and utilization.
Data Retrieval Procedure:
To retrieve data using the code:
1.	Set Working Directory: Define the desired folder path, typically using the os.chdir() function, to ensure that the script operates in the intended directory.
2.	Define the Command: Create a command as a list of strings, including the Python interpreter (python), the script name (get-observations-NINA.py), and the various parameters required for data retrieval (e.g., station, parameters, resolution time, and API key).
3.	Execute the Command: Use subprocess.call() to execute the command, which triggers the Python script to fetch the requested data from the HydAPI.
4.	Data Output: The retrieved data is written to a JSON file named according to the station, e.g., output_file = r'C:\Konnektivitet\Stasjon_xxx.json'.

Script (in python)

Koden er skrevet i python og henter hydrologiske data fra NVE sin API. Skriptet heter get-observations-NINA.py. Den fungerer hvis man oppretter en mappe på C:/ kalt Konnektivetet/python og legger scriptet der.

#!/usr/bin/python

import getopt
import json
import sys

try:
    from urllib.request import Request, urlopen  # Python 3
except ImportError:
    from urllib2 import Request, urlopen  # Python 2

def usage():
    print()
    print("Get observations from the NVE Hydrological API (HydAPI)")
    print("Parameters:")
    print("   -a: ApiKey (mandatory). ")
    print("   -s: StationId (mandatory). Several stations can be given separated by a comma. Example \"6.10.0,12.209.0\"")
    print("   -p: Parameter (mandatory). Several Parameters can be given separated by a comma.")
    print("   -r: Resolution time. 0 (instantaneous), 60 (hourly), or 1440 (daily). (mandatory)")
    print("   -t: Reference time. See documentation for reference time. Example \"P1D/\", gives one day back in time. If none given, the last observed value will be returned")
    print("   -h: This help")
    print()
    print("Example:")
    print("    python get-observations.py -a \"INSERT_APIKEY_HERE\" -s \"6.10.0,12.209.0\" -p \"1000,1001\" -r 60 -t \"P1D/\"")
    print()

def main(argv):
    try:
        opts, args = getopt.getopt(argv, "a:s:p:r:ht:")
    except getopt.GetoptError as err:
        print(str(err))  # will print something like "option -a not recognized"
        usage()
        sys.exit(2)

    station = None
    parameter = None
    resolution_time = None
    api_key = None
    reference_time = None

    for opt, arg in opts:
        if opt == "-s":
            station = arg
        elif opt == "-p":
            parameter = arg
        elif opt == "-r":
            resolution_time = arg
        elif opt == "-a":
            api_key = arg
        elif opt == "-t":
            reference_time = arg
        elif opt == "-h":
            usage()
            sys.exit()
        else:
            assert False, "unhandled option"

    if api_key is None:
        print("Error: You must supply the API key with your request (-a)")
        usage()
        sys.exit(2)

    if station is None or parameter is None or resolution_time is None:
        print("Error: You must supply the parameters station (-s), parameter (-p), and resolution time (-r)")
        usage()
        sys.exit(2)

    baseurl = "https://hydapi.nve.no/api/v1/Observations?StationId={station}&Parameter={parameter}&ResolutionTime={resolution_time}"

    url = baseurl.format(station=station, parameter=parameter, resolution_time=resolution_time)

    if reference_time is not None:
        url = "{url}&ReferenceTime={reference_time}".format(url=url, reference_time=reference_time)

    print(url)

    request_headers = {
        "Accept": "application/json",
        "X-API-Key": api_key
    }

    request = Request(url, headers=request_headers)

    response = urlopen(request)
    content = response.read().decode('utf-8')

    output_file = r'C:\Konnektivitet\Stasjon_xxx.json'  # Definerer stien til utdatafilen

    with open(output_file, 'w') as outfile:
        outfile.write(content)

    print(f"Data written to {output_file}")

if __name__ == "__main__":
    main(sys.argv[1:])

Dataene hentes ut ved å skrive inn følgende kommandoer i python vinduet:

import os
#Angi banen til den ønskede mappen
ny_mappe = r'C:\Konnektivitet\python'
Endre den aktive mappen til den nye mappen
os.chdir(ny_mappe)

import subprocess
# Definer kommandoen som en liste av strenger
kommando = [
     'python',                # Dette er kommandoen du vil kjøre (Python)
     'get-observations-NINA.py',   # Dette er navnet på Python-skriptet du vil kjøre
     '-a', '4B2BQZYxbEiommYw8YCB8w==', # Mulig alle må lage sin egen API key: hydapi.nve.no/Users
     '-s', '12.209.0', # Dette er stasjonsnummer som kan hentes fra NINA innsynsløsning (se under).
     '-p', '1000', # spesifikk type observasjon – se hydapi.nve.no/UserDocumentation/
     '-r', '60', # oppløsningstid. 60 indikerer en times oppløsning
     '-t', '2018-01-01T10:00:00/2018-01-14T20:00:00'  # Merk at timestamp er uten ekstra anførselstegn
 ]
subprocess.call(kommando)
Innsynsløsning: https://nina.maps.arcgis.com/apps/instant/basic/index.html?appid=4344dd7f769048ed9dedf1803638ff48

Sentinel-1 radar data can be affected by speckle noise, which appears as random variations in pixel values. While speckle filtering was applied to reduce this noise, some level of uncertainty remains, particularly in areas with complex terrain.

## References
## Analyses
### Data sets

### Scaled indicator values
### Uncertainty
## Prepare export
You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.

### Eksport file (final product)
/*===========================================================================================
                       SAR-FLOOD MAPPING USING A CHANGE DETECTION APPROACH
  ===========================================================================================
  This script uses Sentinel-1 SAR data to create a flood extent map. It employs a change 
  detection method, comparing before- and after-flood images. The script is designed to work 
  with Ground Range Detected (GRD) imagery, which undergoes preprocessing steps, including 
  Thermal-Noise Removal, Radiometric Calibration, and Terrain Correction. Therefore, only a 
  Speckle filter needs to be applied during preprocessing.

  ===========================================================================================

                                  SELECT YOUR OWN STUDY AREA

   Use the polygon tool in the top left corner of the map pane to outline your study area. 
   Single clicks add vertices, and double-clicking completes the polygon.
   **CAUTION**: Under 'Geometry Imports' (top left in the map panel), uncheck the 
               geometry box to avoid obscuring the view of the imagery later.

  *******************************************************************************************
                                       SET TIME FRAME

   Specify the start and end dates for a period BEFORE the flood event. Ensure that the 
   duration is sufficient for Sentinel-1 to acquire images (repetition rate = 6 days). 
   Adjust these parameters if the ImageCollections (see Console) are empty.

/********************************************************************************************
                           SET SAR PARAMETERS (can be left default)*/

var polarization = "VH"; /* or 'VV' --> VH is commonly preferred for flood mapping, but 
                           the choice may depend on your study area. You can choose 'VV' as 
                           well. */

var pass_direction = "DESCENDING"; /* or 'ASCENDING'. When comparing images, use only one 
                           pass direction. Consider changing this parameter if your image 
                           collection is empty. In some areas, more Ascending images exist 
                           than Descending, or vice versa. */

var difference_threshold = 1.25; /* Threshold applied to the difference image (after flood
                           - before flood). The value is chosen through trial and error. 
                           If your flood extent result contains many false positives or 
                           negatives, consider adjusting it. */
//var relative_orbit = 79; 
                          /* If you know the relative orbit for your study area, you can 
                           filter your image collection by it here to avoid errors due to 
                           comparisons across different relative orbits. */

/********************************************************************************************
  ---->>> DO NOT EDIT THE SCRIPT PAST THIS POINT! (unless you know what you are doing) <<<---
  ------------------>>> now hit the 'RUN' at the top of the script! <<<-----------------------
  -----> The final flood product will be ready for download on the right (under tasks) <-----
  ******************************************************************************************/

//---------------------------------- Translating User Inputs ------------------------------//
Map.setCenter(9.98383,  60.95329, 14); // Sets the map center.

var before_start = '2023-04-01';
var before_end = '2023-07-01'; 
// Define the start and end dates for the period before the flood.

var after_start = '2023-08-01'; 
var after_end = '2023-08-25'; 
// Define the start and end dates for the period after the flood.

//------------------------------- DATA SELECTION & PREPROCESSING --------------------------//

// Rename the selected geometry feature 
var aoi = ee.FeatureCollection(geometry);

// Load and filter Sentinel-1 GRD data based on predefined parameters 
var collection = ee.ImageCollection('COPERNICUS/S1_GRD')
  .filter(ee.Filter.eq('instrumentMode', 'IW'))
  .filter(ee.Filter.listContains('transmitterReceiverPolarisation', polarization))
  .filter(ee.Filter.eq('orbitProperties_pass', pass_direction)) 
  .filter(ee.Filter.eq('resolution_meters', 10))
  //.filter(ee.Filter.eq('relativeOrbitNumber_start', relative_orbit ))
  .filterBounds(aoi)
  .select(polarization);

// Select images based on predefined dates
var before_collection = collection.filterDate(before_start, before_end);
var after_collection = collection.filterDate(after_start, after_end);

// Print selected tiles to the console
function dates(imgcol) {
    var range = imgcol.reduceColumns(ee.Reducer.minMax(), ["system:time_start"]);
    var printed = ee.String('from ')
      .cat(ee.Date(range.get('min')).format('YYYY-MM-dd'))
      .cat(' to ')
      .cat(ee.Date(range.get('max')).format('YYYY-MM-dd'));
    return printed;
}

var before_count = before_collection.size();
print(ee.String('Tiles selected: Before Flood ').cat('(').cat(before_count).cat(')'),
    dates(before_collection), before_collection);

var after_count = before_collection.size();
print(ee.String('Tiles selected: After Flood ').cat('(').cat(after_count).cat(')'),
    dates(after_collection), after_collection);

// Create a mosaic of selected tiles and clip to the study area
var before = before_collection.mosaic().clip(aoi);
var after = after_collection.mosaic().clip(aoi);

// Apply radar speckle reduction by smoothing  
var smoothing_radius = 50;
var before_filtered = before.focal_mean(smoothing_radius, 'circle', 'meters');
var after_filtered = after.focal_mean(smoothing_radius, 'circle', 'meters');

//------------------------------- FLOOD EXTENT CALCULATION -------------------------------//
// Calculate the difference between the before and after images
var difference = after_filtered.divide(before_filtered);

//Map.addLayer(difference, {}, 'Difference 2023', 1);

// Apply the predefined difference threshold to create the flood extent mask 
var threshold = difference_threshold;
var difference_binary = difference.gt(threshold);

// Refine the flood result using additional datasets
// Include the JRC layer on surface water seasonality to mask flood pixels from areas
// with "permanent" water (water > 10 months of the year)
var swater = ee.Image('JRC/GSW1_0/GlobalSurfaceWater').select('seasonality');
var swater_mask = swater.gte(10).updateMask(swater.gte(10));

// Create a flooded mask where perennial water bodies (water > 10 mo/yr) are assigned a 0 value
var flooded_mask = difference_binary.where(swater_mask, 0);
// Create the final flooded area without pixels in perennial water bodies
var flooded = flooded_mask.updateMask(flooded_mask);

// Compute the connectivity of pixels to eliminate those connected to 8 or fewer neighbors
// This operation reduces noise in the flood extent product
var connections = flooded.connectedPixelCount();
var flooded = flooded.updateMask(connections.gte(8));

// Mask out areas with more than 5 percent slope using the Norwegian Digital Elevation Model
//var DEM = ee.Image('WWF/HydroSHEDS/03VFDEM');
var DEM = ee.Image('users/vegar/dem_10m_kartverket');
var terrain = ee.Algorithms.Terrain(DEM);
var slope = terrain.select('slope');
var flooded = flooded.updateMask(slope.lt(5));

// Calculate flood extent area
// Create a raster layer containing the area information of each pixel 
var flood_pixelarea = flooded.select(polarization)
  .multiply(ee.Image.pixelArea());

// Sum the areas of flooded pixels
// The default 'bestEffort' is set to true to reduce computation time; for a more 
// accurate result, set 'bestEffort' to false and increase 'maxPixels'. 
var flood_stats = flood_pixelarea.reduceRegion({
  reducer: ee.Reducer.sum(),              
  geometry: aoi,
  scale: 10, // native resolution 
  //maxPixels: 1e9,
  bestEffort: true
});

// Convert the flood extent to hectares (area calculations are originally given in meters)  
var flood_area_ha = flood_stats
  .getNumber(polarization)
  .divide(10000)
  .round();

// Set visualization parameters
var frequencyVis = {
  min: 0,
  max: 4,
  palette: ['white', 'lightblue', 'mediumblue', 'darkblue', 'black']
};

// Display flooded areas
Map.addLayer(flooded, { palette: "0000FF" }, 'Flooded areas Hans 2023', 0);

//----------------------------- Compare with the previous year 2022------------------------

var before_start2 = '2022-04-01';
var before_end2 = '2022-09-01';

var after_start2 = '2023-08-01'; // Should match the reference state 
var after_end2 = '2023-08-25'; // Should match the reference state

var before_collection2 = collection.filterDate(before_start2, before_end2);
var after_collection2 = collection.filterDate(after_start2, after_end2);

//----------------------------- Compare with the previous year 2022------------------------

var before_start2 = '2022-04-01';
var before_end2 = '2022-09-01';

var after_start2 = '2023-08-01'; // Should match the reference state 
var after_end2 = '2023-08-25'; // Should match the reference state

var before_collection2 = collection.filterDate(before_start2, before_end2);
var after_collection2 = collection.filterDate(after_start2, after_end2);

// Create a mosaic of selected tiles and clip to the study area
var before2 = before_collection2.mosaic().clip(aoi);
var after2 = after_collection2.mosaic().clip(aoi);

// Apply radar speckle reduction by smoothing  
var smoothing_radius = 50;
var before_filtered2 = before2.focal_mean(smoothing_radius, 'circle', 'meters');
var after_filtered2 = after2.focal_mean(smoothing_radius, 'circle', 'meters');

// Calculate the difference between the before and after images
var difference2 = after_filtered2.divide(before_filtered2);

// Apply the predefined difference-threshold and create the flood extent mask 
var threshold2 = difference_threshold;
var difference_binary2 = difference2.gt(threshold2);

// Refine the flood result using additional datasets
// Include the JRC layer on surface water seasonality to mask flood pixels from areas
// with "permanent" water (water > 10 months of the year)
var swater = ee.Image('JRC/GSW1_0/GlobalSurfaceWater').select('seasonality');
var swater_mask = swater.gte(10).updateMask(swater.gte(10));

// Create a flooded mask where perennial water bodies (water > 10 mo/yr) are assigned a 0 value
var flooded_mask2 = difference_binary2.where(swater_mask, 0);
// Create the final flooded area without pixels in perennial water bodies
var flooded2 = flooded_mask2.updateMask(flooded_mask2);

// Compute the connectivity of pixels to eliminate those connected to 8 or fewer neighbors
// This operation reduces noise in the flood extent product
var connections2 = flooded2.connectedPixelCount();
var flooded2 = flooded2.updateMask(connections2.gte(8));

// Mask out areas with more than 5 percent slope using a Digital Elevation Model
var DEM = ee.Image('users/vegar/dem_10m_kartverket');
var terrain = ee.Algorithms.Terrain(DEM);
var slope = terrain.select('slope');
flooded2 = flooded2.updateMask(slope.lt(5));

// Calculate flood extent area
// Create a raster layer containing the area information of each pixel 
var flood_pixelarea2 = flooded2.select(polarization)
  .multiply(ee.Image.pixelArea());

// Sum the areas of flooded pixels
// The default 'bestEffort' is set to true to reduce computation time; for a more 
// accurate result, set 'bestEffort' to false and increase 'maxPixels'. 
var flood_stats2 = flood_pixelarea2.reduceRegion({
  reducer: ee.Reducer.sum(),              
  geometry: aoi,
  scale: 10, // native resolution 
  //maxPixels: 1e9,
  bestEffort: true
});

// Convert the flood extent to hectares (area calculations are originally given in meters)  
var flood_area_ha2 = flood_stats2
  .getNumber(polarization)
  .divide(10000)
  .round();

// Flooded areas
//Map.addLayer(flooded2, {palette: "0FFFFF"}, 'Flooded areas 2022', 1);

//----------------------------- Compare with the previous year 2021------------------------

var before_start3 = '2021-04-01';
var before_end3 = '2021-09-01';

var after_start3 = '2023-08-01'; // Should match the reference state 
var after_end3 = '2023-08-25'; // Should match the reference state

var before_collection3 = collection.filterDate(before_start3, before_end3);
var after_collection3 = collection.filterDate(after_start3, after_end3);

// Create a mosaic of selected tiles and clip to the study area
var before3 = before_collection3.mosaic().clip(aoi);
var after3 = after_collection3.mosaic().clip(aoi);

// Apply radar speckle reduction by smoothing  
var before_filtered3 = before3.focal_mean(smoothing_radius, 'circle', 'meters');
var after_filtered3 = after3.focal_mean(smoothing_radius, 'circle', 'meters');

// Calculate the difference between the before and after images
var difference3 = after_filtered3.divide(before_filtered3);

// Apply the predefined difference-threshold and create the flood extent mask 
var threshold3 = difference_threshold;
var difference_binary3 = difference3.gt(threshold3);

// Refine the flood result using additional datasets
// Include the JRC layer on surface water seasonality to mask flood pixels from areas
// with "permanent" water (water > 10 months of the year)
swater = ee.Image('JRC/GSW1_0/GlobalSurfaceWater').select('seasonality');
swater_mask = swater.gte(10).updateMask(swater.gte(10));

// Create a flooded mask where perennial water bodies (water > 10 mo/yr) are assigned a 0 value
var flooded_mask3 = difference_binary3.where(swater_mask, 0);
// Create the final flooded area without pixels in perennial water bodies
var flooded3 = flooded_mask3.updateMask(flooded_mask3);

// Compute the connectivity of pixels to eliminate those connected to 8 or fewer neighbors
// This operation reduces noise in the flood extent product
var connections3 = flooded3.connectedPixelCount();
flooded3 = flooded3.updateMask(connections3.gte(8));

// Mask out areas with more than 5 percent slope using a Digital Elevation Model
flooded3 = flooded3.updateMask(slope.lt(5));

// Calculate flood extent area
// Create a raster layer containing the area information of each pixel 
var flood_pixelarea3 = flooded3.select(polarization)
  .multiply(ee.Image.pixelArea());

// Sum the areas of flooded pixels
// The default 'bestEffort' is set to true to reduce computation time; for a more 
// accurate result, set 'bestEffort' to false and increase 'maxPixels'. 
var flood_stats3 = flood_pixelarea3.reduceRegion({
  reducer: ee.Reducer.sum(),              
  geometry: aoi,
  scale: 10, // native resolution 
  //maxPixels: 1e9,
  bestEffort: true
});

// Convert the flood extent to hectares (area calculations are originally given in meters)  
var flood_area_ha3 = flood_stats3
  .getNumber(polarization)
  .divide(10000)
  .round();

// Flooded areas
//Map.addLayer(flooded3, {palette: "0FFFFF"}, 'Flooded areas 2021', 1);

//----------------------------- Compare with the previous year 2020------------------------

var before_start4 = '2020-04-01'; 
var before_end4 = '2020-09-01';

var after_start4 = '2023-08-01'; // Should match the reference state 
var after_end4 = '2023-08-25'; // Should match the reference state

var before_collection4 = collection.filterDate(before_start4, before_end4);
var after_collection4 = collection.filterDate(after_start4, after_end4);

// Create a mosaic of selected tiles and clip to the study area
var before4 = before_collection4.mosaic().clip(aoi);
var after4 = after_collection4.mosaic().clip(aoi);

// Apply radar speckle reduction by smoothing  
var before_filtered4 = before4.focal_mean(smoothing_radius, 'circle', 'meters');
var after_filtered4 = after4.focal_mean(smoothing_radius, 'circle', 'meters');

// Calculate the difference between the before and after images
var difference4 = after_filtered4.divide(before_filtered4);

// Apply the predefined difference-threshold and create the flood extent mask 
var threshold4 = difference_threshold;
var difference_binary4 = difference4.gt(threshold4);

// Refine the flood result using additional datasets
// Include the JRC layer on surface water seasonality to mask flood pixels from areas
// with "permanent" water (water > 10 months of the year)
swater = ee.Image('JRC/GSW1_0/GlobalSurfaceWater').select('seasonality');
swater_mask = swater.gte(10).updateMask(swater.gte(10));

// Create a flooded mask where perennial water bodies (water > 10 mo/yr) are assigned a 0 value
var flooded_mask4 = difference_binary4.where(swater_mask, 0);
// Create the final flooded area without pixels in perennial water bodies
var flooded4 = flooded_mask4.updateMask(flooded_mask4);

// Compute the connectivity of pixels to eliminate those connected to 8 or fewer neighbors
// This operation reduces noise in the flood extent product
var connections4 = flooded4.connectedPixelCount();
flooded4 = flooded4.updateMask(connections4.gte(8));

// Mask out areas with more than 5 percent slope using a Digital Elevation Model
flooded4 = flooded4.updateMask(slope.lt(5));

// Calculate flood extent area
// Create a raster layer containing the area information of each pixel 
var flood_pixelarea4 = flooded4.select(polarization)
  .multiply(ee.Image.pixelArea());

// Sum the areas of flooded pixels
// The default 'bestEffort' is set to true to reduce computation time; for a more 
// accurate result, set 'bestEffort' to false and increase 'maxPixels'. 
var flood_stats4 = flood_pixelarea4.reduceRegion({
  reducer: ee.Reducer.sum(),              
  geometry: aoi,
  scale: 10, // native resolution 
  //maxPixels: 1e9,
  bestEffort: true
});

// Convert the flood extent to hectares (area calculations are originally given in meters)  
var flood_area_ha4 = flood_stats4
  .getNumber(polarization)
  .divide(10000)
  .round();

// Flooded areas
//Map.addLayer(flooded4, {palette: "0FFFFF"}, 'Flooded areas 2020', 1);

//----------------------------- Compare with the previous year 2019------------------------

var before_start5 = '2019-04-01';
var before_end5 = '2019-09-01';

var after_start5 = '2019-08-01'; // Should match the reference state 
var after_end5 = '2019-08-25'; // Should match the reference state

var before_collection5 = collection.filterDate(before_start5, before_end5);
var after_collection5 = collection.filterDate(after_start5, after_end5);

// Create a mosaic of selected tiles and clip to the study area
var before5 = before_collection5.mosaic().clip(aoi);
var after5 = after_collection5.mosaic().clip(aoi);

// Apply radar speckle reduction by smoothing  
var before_filtered5 = before5.focal_mean(smoothing_radius, 'circle', 'meters');
var after_filtered5 = after5.focal_mean(smoothing_radius, 'circle', 'meters');

// Calculate the difference between the before and after images
var difference5 = after_filtered5.divide(before_filtered5);

// Apply the predefined difference-threshold and create the flood extent mask 
var threshold5 = difference_threshold;
var difference_binary5 = difference5.gt(threshold5);

// Refine the flood result using additional datasets
// Include the JRC layer on surface water seasonality to mask flood pixels from areas
// with "permanent" water (water > 10 months of the year)
swater = ee.Image('JRC/GSW1_0/GlobalSurfaceWater').select('seasonality');
swater_mask = swater.gte(10).updateMask(swater.gte(10));

// Create a flooded mask where perennial water bodies (water > 10 mo/yr) are assigned a 0 value
var flooded_mask5 = difference_binary5.where(swater_mask, 0);
// Create the final flooded area without pixels in perennial water bodies
var flooded5 = flooded_mask5.updateMask(flooded_mask5);

// Compute the connectivity of pixels to eliminate those connected to 8 or fewer neighbors
// This operation reduces noise in the flood extent product
var connections5 = flooded5.connectedPixelCount();
flooded5 = flooded5.updateMask(connections5.gte(8));

// Mask out areas with more than 5 percent slope using a Digital Elevation Model
flooded5 = flooded5.updateMask(slope.lt(5));

// Calculate flood extent area
// Create a raster layer containing the area information of each pixel 
var flood_pixelarea5 = flooded5.select(polarization)
  .multiply(ee.Image.pixelArea());

// Sum the areas of flooded pixels
// The default 'bestEffort' is set to true to reduce computation time; for a more 
// accurate result, set 'bestEffort' to false and increase 'maxPixels'. 
var flood_stats5 = flood_pixelarea5.reduceRegion({
  reducer: ee.Reducer.sum(),              
  geometry: aoi,
  scale: 10, // native resolution 
  //maxPixels: 1e9,
  bestEffort: true
});

// Convert the flood extent to hectares (area calculations are originally given in meters)  
var flood_area_ha5 = flood_stats5
  .getNumber(polarization)
  .divide(10000)
  .round();

// Flooded areas
//Map.addLayer(flooded5, {palette: "0FFFFF"}, 'Flooded areas 2019', 1);

//----------------------------- Frequency ----------------------------//

// Calculate the frequency

// Unmask the two images and add them together
var flooded_unmasked = flooded.unmask(0);
var flooded2_unmasked = flooded2.unmask(0);
var flooded3_unmasked = flooded3.unmask(0);
var flooded4_unmasked = flooded4.unmask(0);
var flooded5_unmasked = flooded5.unmask(0);

var floodFrequency = flooded_unmasked.add(flooded2_unmasked)
  .add(flooded3_unmasked)
  .add(flooded4_unmasked)
  .add(flooded5_unmasked);

// Define the flood frequency image with a mask for zero values
var maskedFloodFrequency = floodFrequency.updateMask(floodFrequency.gt(0));

// Define a color palette for flood frequency
var colorPalette = ['lightblue', 'blue', 'purple', 'pink', 'red'];

// Add the flood frequency image to the map with the updated color palette
Map.addLayer(maskedFloodFrequency, {min: 1, max: 5, palette: colorPalette}, 'Flood Frequency', 0);

//----------------------------- Affected wetland land ----------------------------//

// Load the wetland map
var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
var myr = myrpred2.mean().select('impervious'); // Select the band 'b1' from the wetland map

// Calculate affected wetland areas using the resampled flood layer and wetland map

// Use areas from the myr dataset as geometry
var wetlandAreas = myr.gt(0.8).selfMask(); // Select areas with wetlands and use geometry as a mask
var wetlandAffected = maskedFloodFrequency.updateMask(wetlandAreas);

// Get pixel area of the affected wetland layer
var wetlandPixelArea = wetlandAffected.multiply(ee.Image.pixelArea()); // Calculate the area of each pixel

// Sum the pixels of the affected wetland layer
var wetlandStats = wetlandPixelArea.reduceRegion({
  reducer: ee.Reducer.sum(), // Sum all pixels with area information                
  geometry: aoi,
  scale: 500,
  maxPixels: 1e9
});

Map.addLayer(wetlandAffected, {min: 1, max: 5, palette: colorPalette}, 'Flood Frequency in Wetlands');

### 2023-10-20 DRAFT_birdIndicators.Rmd
Edited / added to Diego's text

--- 2023-10-19 DRAFT_birdIndicators.Rmd
@@
 ## Introduction
-The Norwegian common breeding bird monitoring scheme (hekkefuglovervåking in Norwegian; formerly called TOV-E) was established between 2005 and 2010.
+The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called TOV-E) was established between 2005 and 2010.
 The main goal of the programme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (available at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway.
-These data are the basis to update the Red List of Birds in Norway following the IUCN protocol, and are also used to generate indicators of mountain, farmland and forest habitats.
+The "hekkefuglovervåking" is coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), and has replaced predecessor monitoring schemes that ran from 1996 to 2008.
 
-Furthermore, these data are shared with the Pan-European Common Bird Monitoring Scheme (PECBMS; https://pecbms.info/), which, by combining data from different long-term and large-scale bird monitoring programmes in all European countries, aims at producing indicators of the general state of nature at continental scale.
+These data collected through "hekkefuglovervåkingen" form the basis for a variety of national and international assessments including nationally reported species population trends (https://hekkefuglovervakingen.nina.no/hekkefugl), updates to the Red List of Birds in Norway following the IUCN protocol (https://artsdatabanken.no/lister/rodlisteforarter), indicators for the Norwegian Nature Index (https://www.naturindeks.no/), and single- and multi-species indicators that are reported nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/) and internationally to e.g.
+the Pan European Bird Monitoring Scheme (PECBMS, https://pecbms.info/) and the OECD.
 
-Data from hekkefuglovervåking are also use in scientific research both at a national and supra-national scale.
+Workflows generating outputs and numbers for several of these reporting routes are already integrated thanks to coordination and standardization efforts by PECMBS, and the calculation of breeding bird indicators for use in ecosystem condition assessments should be incorporated into the same workflow.
+We therefore suggest to use the same multi-species indicators in the framework for ecosystem condition assessment as are reported nationally and internationally: the established "Farmland bird index", "Mountain bird index", and "Forest bird index", as well as a corresponding index for wetland birds.
 
-The hekkefuglovervåking, coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), replaced the previous scheme, which ran between 1995 and 2008.
 ## About the underlying data
-<!-- @Diego: please add content here and in the following four subsections -->
+The data from "hekkefuglovervåkingen" are gathered mostly by experienced volunteer bird-watchers following a standardized sampling protocol (https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf).
+Observers will visit each route once a year between 20.
+May and 10.
+June (depending on latitude) and carry out the census in good weather conditions.
+The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route.
+A total of 229 species are typically recorded during this census.
+After 5 minutes, the observer walks to the next point in a straight line (when possible).
+During this 'transect' the observer is requested to record also a subset of species that are considered rare or difficult to capture during the 5-minute point counts (n = 121).
+
+The data are subsequently entered into a database (https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx) and quality-checked by NINA.
+Subsequently, the raw data are processed using a combination of country-specific wrangling routines and internationally standardized scripts distributed by PECBMS, which together make up the now integrated workflow below:
+
+<img src="Hekkefugl_Workflow.png" width="600"/>
 
 ### Representativity in time and space
-Diego is on it!
+"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway (MAP HERE??
+I can provide this).
+Each route has between 12 and 20 counting points (ca.
+9000 across Norway), which typically are 300 m apart, describing a square with side 1.5 km long (the observer walks on a straight line between two consecutive points).
+Deviations of this typical route shape are normally due to the landscape configuration (e.g.
+rough terrain, presence of a lake or river, sea, glacier, etc.).
+Due to limited resources, only a random sample of 370 routes are surveyed every year with an aim to survey at least 80% of the routes yearly.
+
+This monitoring scheme is therefore designed to gather representative samples in all (common) habitats along both latitudinal and altitudinal gradients.
+
+Even though implementation of the monitoring scheme started in 2005, full coverage at the national scale was not achieved until 2010.
 
 ### Original units
-<!-- Text here -->
+The observers in the field record "number of individuals" seen or heard during the census.
+Those values are then converted into "number of pairs" by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only).
+In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of the population in a given baseline year (in Norway this is typically 2008).
+The abundance indices then form the basis of further analyses and aggregation routines.
 
 ### Temporal coverage
-<!-- Text here -->
+The "hekkefuglovervåkingen" was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010.
+For many species, data is already of good coverage and high quality starting in 2008.
+The monitoring programme is still ongoing as of 2023.
+
+Predecessor monitoring schemes existed prior to the establishment of "hekkefuglovervåkingen" and led to the collection of decent data in the period from 1996 up to and including 2008. 
+Although the predecessor monitoring programmes and "hekkefuglovervåkingen" are not the same, the overlap for the years 2007 and 2008 makes it possible to combine abundance indices from across time using tools like "RSWAN", which is provided by PECBMS.
+For a subset of 55 species, longer time-series of abundance indices (1996-2022, as of now) are therefore available.
 
 ### Aditional comments about the dataset
-<!-- Text here -->
+From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses.
+Abundance indices of 71 of those species, representing those with wide distributions and high continental-scale relevance, are reported annually to PECBMS.
+Longer (combined) time series can be calculated for 55 species, and 54 of those serve as indicators in the Norwegian Nature Index.
+The established breeding bird indicators contain varying numbers of species: the farmland bird index has 7 species, the mountain bird index has 8 species, and the forest bird index has 24 species.
+For the new index for wetland birds, we used 7 out of the 8 species suggested for this purpose by @husby2011.
+The one species we dropped was the Common Crane _(Grus grus)_ as it has seen strong recent population increases that are linked to anthropogenic impacts unrelated to the status of their wetland breeding habitats (Bård Stokke, pers.
+communication).
 
@@
 
-The internationally established composite bird indicators (farmland bird index, common bird index, etc.), however, are trend indicators can be presented with a few different baseline years.
-It is the availability of monitoring data of sufficient quality that determines which years may be used as baselines, and the "Hekkefuglovervåkingen", the monitoring programme that supplies the data for the Norwegian breeding bird indicators, has been running since 2007/2008.
-For some species, a predecessor monitoring scheme has been providing data since 1996.
-The possible baselines for the breeding bird indicators can therefore be no earlier than 1996, and using baselines consistent with the other nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/; https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx) and internationally reported indices (https://pecbms.info/trends-and-indicators/) seems sensible.
+The internationally established composite bird indicators (farmland bird index, common bird index, etc.), however, are trend indicators that can be presented with a few different baseline years.
+It is the availability of monitoring data of sufficient quality that determines which years may be used as baselines, and the "hekkefuglovervåkingen", the monitoring programme that supplies the data for the Norwegian breeding bird indicators, has been running since 2007/2008.
+For some species, predecessor monitoring schemes have been providing data since 1996.
+The possible baselines for the breeding bird indicators can therefore be no earlier than 1996 or 2008 depending on the species (assemblage), and using baselines consistent with the other nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/; https://hekkefuglovervakingen.nina.no/Fugl/Default.aspx) and internationally reported indices (https://pecbms.info/trends-and-indicators/) seems sensible.
 
@@
 The breeding bird indicators represent average trends relative to a reference year and are therefore scaled with respect to population size in the reference year.
-Therefore, the indicator value in the reference year is 1.
+Therefore, the indicator value in the reference year is 100 (percent).
 In theory, the indicators can take values between 0 (representing complete extinction of all species) and infinity (infinite population increase).
-The reference value 1 represents unchanged abundance relative to the reference year.
+The reference value 100 represents unchanged abundance relative to the reference year.
 Determining a threshold for what level of decrease in bird abundance is acceptable under "good ecological condition", however, is a challenging task.
 One approach to setting a threshold may be to consider thresholds of population decline used in other frameworks.
-The IUCN Red List of threatened species, for example, considers single species as "endangered" when a population decrease of 30% or more has been observed for the most recent decade [@iucn2001].
-While it is unclear to what degree that can (or should) be extended to assemblages of multiple species, one may use this as a guideline to stipulate that >70% of the original population (indicator value 0.7) should be present (i.e.
-the species assemblage should not be endangered) when¨the ecosystem is in good condition.
+The IUCN Red List of threatened species, for example, considers single species as "endangered" when a population decrease of 30% or more has been observed for the most recent decade or over the most recent 3 generations [@iucn2001].
+While it is unclear to what degree that can (or should) be extended to assemblages of multiple species, one may use this as a guideline to stipulate that >70% of the original population (indicator value 70 when considering a recent trend over 10 years or 3 average generations) should be present (i.e.
+the species assemblage should not be endangered) when the ecosystem is in good condition.
 Another approach to determining a threshold would be to assume that bird populations should not be declining at all if an ecosystem is in good condition, i.e.
-that 1 is the threshold for good ecological condition instead of the reference value.
-In that case, however, one may need to come up with a reference value that is greater than 1, and determining what level of population increase would be ideal is a nearly impossible task.
+that 100 is the threshold for good ecological condition instead of the reference value.
+In that case, however, one may need to come up with a reference value that is greater than 100, and determining what level of population increase would be ideal is a nearly impossible task.
 We want to highlight here that these are very preliminary thoughts, and that the reference and threshold values have not been discussed in any more details yet.
-These discussions will nonetheless be needed, and possibly augmented with simulations, before putting the breeding bird indicators to use for ecosystem condition assessments.
+These discussions will nonetheless be needed, and possibly augmented with simulation studies, before putting the breeding bird indicators to use for ecosystem condition assessments.
 
 ## Uncertainties
-<!-- Text here -->
+There is a substantial amount of uncertainty related to which year should be considered as a baseline and how the reference and threshold values should be set (see preceeding section).
+
+Another source of uncertainty comes from the assumption that multi-species abundance indices use a representative selection of species sampled in a representative set of habitats.
+However, due to resource- and logistic constraints, that is often not the case and non-random sampling along, for example, rural-urban gradients may bias breeding bird indices derived from data from "hekkefuglovervåkingen" [@husby2021].
 
@@
 ### Data sets
-<!-- Text here -->
+All data used are gathered under the Norwegian common breeding bird monitoring scheme (Hekkeguglovervåking in Norwegian; formely named TOV-E).
+For the calculation of multi-species indicators (MSI) of farmland and forest habitats, however, a longer dataset is used by combininng data from the 'predecessor monitoring programme', whic ran from 1995 to 2008.
 
 #### Monitoring data from "Hekkefuglovervåkingen"
-<!-- Text here -->
+The route-based data are compiled after each breeding season and stored at NINA's database.
+The route-based data are then used to calculate the long-term trends (since 2008) in abundance for XX species, which have sufficient data to properly estimate the yearly abundance indices at a National scale.
 
 #### PECBMS Trim results from predecessor monitoring programme
-<!-- Text here -->
+The abundance indices generated by the PECBMS RTRIM workflow using data from the predecessor scheme are stored in NINAs local file repositories.
+Unlike the Trim analyses for the data from "hekkefuglovervåkingen", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being added.
 
@@
 
-### PECBMS Trim analyses
+### PECBMS RTRIM analyses
 
-### Combining data series using SWAN
+### Combining data series using RSWAN
 
-### Scaled indicator values
+### Aggregation to multi-species indices (MSIs)
 

### 2023-10-21 DRAFT_birdIndicators.Rmd
Completed draft of bird indicators chapter

@@
 library(knitr)
+library(tidyverse)
+library(rtrim)
+library(data.table)
+library(RPostgres)
+library(odbc)
+library(RPostgreSQL)
+library(DBI)
+library(rpostgis)
+library(sf)
+library(lubridate)
+library(xtable)
+library(rtrim)
+library(readxl)
+library(data.table)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-<img src="Hekkefugl_Workflow.png" width="600"/>
+<img src="images/Hekkefugl_Workflow.png" width="600"/>
 
@@
 There is a substantial amount of uncertainty related to which year should be considered as a baseline and how the reference and threshold values should be set (see preceeding section).
+This is also closely related to a second question that requires discussion: over what time period should breeding bird indicators be considered in ecosystem condition assessments?
+One option would be to use an average over the entire period with available data (but note that for mountain and wetland indicators, the starting year may have to be set to 2008 or even 2009 or 2010 to ensure sufficient data coverage).
+Alternatively, averages over shorter periods leading up to the year for which ecosystem condition should be assessed.
+This will require some careful evaluation.
 
@@
 ## Analyses
+
+### Workflow setup
+
+The code for the following analyses has been developed in a separate GitHub repository: https://github.com/NINAnor/birdIndicators
+As a first step, we therefore need to read in all the functions from that repository.
+The repository has not been converted into a package yet, so we need to download the entire repository and source all function files in the "R" folder manually:
+
+```{r}
+download.file(url = "https://github.com/NINAnor/birdIndicators/archive/refs/heads/main.zip",
+              destfile = "temp/birdIndicators_Repo.zip")
+unzip(zipfile = "temp/birdIndicators_Repo.zip",
+      exdir = "temp")
+
+sourceDir <- function(path, trace = TRUE, ...) {
+  for (nm in list.files(path, pattern = "[.][RrSsQq]$")) {
+    if(trace) cat(nm,":")
+    source(file.path(path, nm), ...)
+    if(trace) cat("\n")
+  }
+}
+sourceDir("temp/birdIndicators-main/R")
+```
+
+Next, we need to define a range of relative and absolute directories / file paths.
+The main reason for that is that many of the scripts delivered by PECBMS function by iteratively writing and reading files in a specific folder structure.
+Whenever possible, I have written wrapper functions around the PECBMS code that work with relative paths, but absolute paths are still necessary for the scripts involved in the RSWAN routine as that workflow toggles between different working directories using `setwd()` repeatedly.
+Additionally, the RSWAN workflow also uses some helper functions that include changing working directories that are not provided as function arguments but instead need to be globally defined.
+
+While all code would ideally be rewritten to work with relative paths and without toggling working directories, PECBMS tends to distribute new versions of their scripts periodically and with no formal version control in place.
+This means that it is imperative to leave their core code as untouched as possible to facilitate including new versions (by copy paste) as they come.
+
+Hence we set up both relative and absolute file paths (note that I construct the latter relative to the project working directory to ensure a minimum of portability):
+
+```{r, eval = FALSE}
+## Relative directory/file paths
+
+# Storage of raw PECBMS Trim outputs
+folder <- "PECBMS_Files"
+
+# Storage of processed PECBMS Trim outputs
+subFolderName <- "Species_files"
+
+# Relative general and working folders for RSWAN
+general_folder_rel <- "data"
+working_folder_rel <- paste0(folder, "/", subFolderName)
+
+# Storage of MSI results
+MSI_results_folder <- "MSI_Results"
+
+## Absolute directory/file paths
+
+# NINAs local storage of legacy Trim output files
+legacyFile_folder <- "P:/41201612_naturindeks_2021_2023_database_og_innsynslosning/Hekkefugl_Dataflyt/LegacyFiles_PECBMS_Trim"
+
+# Absolute data and output paths for RSWAN 
+general_folder <- paste0(getwd(), "/", general_folder_rel)
+working_folder <- paste0(getwd(), "/", working_folder_rel)
+
+output_folder <- paste0(working_folder, "/output/") 
+output_folder2 <- paste0(working_folder, "/output")
+
+## Create directories if they are not available
+dir.list <- c(folder, general_folder_rel, working_folder_rel, output_folder2)
+for(i in 1:length(dir.list)){
+  if(!file.exists(dir.list[i])){ 
+  dir.create(dir.list[i])
+  }
+}
+
+```
+
 ### Data sets
-All data used are gathered under the Norwegian common breeding bird monitoring scheme (Hekkeguglovervåking in Norwegian; formely named TOV-E).
-For the calculation of multi-species indicators (MSI) of farmland and forest habitats, however, a longer dataset is used by combininng data from the 'predecessor monitoring programme', whic ran from 1995 to 2008.
+The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåkingen" and the calculated abundance indices from the predecessor monitoring schemes.
+Additionally, we also a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow.
 
 #### Monitoring data from "Hekkefuglovervåkingen"
-The route-based data are compiled after each breeding season and stored at NINA's database.
-The route-based data are then used to calculate the long-term trends (since 2008) in abundance for XX species, which have sufficient data to properly estimate the yearly abundance indices at a National scale.
+The route-based raw data from "hekkefuglovervåkingen" are compiled after each breeding season and stored in NINA's database.
+From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2022).
+
+```{r, eval = FALSE}
+minYear <- 2006
+maxYear <- 2022
+
+Trim_data <- downloadData_TRIM(minYear = minYear, maxYear = maxYear,
+                               drop_negativeSpp = TRUE, DriverName = "FreeTDS")
+```
 
@@
 Unlike the Trim analyses for the data from "hekkefuglovervåkingen", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being added.
+These data are needed for the RSWAN workflow, and they need to be present in the respective "working_folder".
+So we copy them from their location in NINA's local file storage into the correct folder for analyses.
+
+```{r, eval = FALSE}
+collectSpeciesFiles_Legacy(origin_folder = legacyFile_folder,
+                           target_folder = paste0(folder, "/", subFolderName))
+```
+
+#### Lists of species information
+
+The species list files should have been downloaded together with the rest of the `birdIndicators` repository and we can use `makeSpeciesLists()` to assemble the information.
+
+```{r}
+sppLists <- makeSpeciesLists(Trim_data = Trim_data, spp_DataPath = "temp/birdIndicators-main/data")
+Spp_selection <- sppLists$sppData
+```
+
+The table ``Spp_selection`` contains information for each species on which reporting routes is should be included in.
+The multispecies indices are grouped into a single category ("MSI") here, but we can extract the species that go into the MSI for each ecosystem via the additionally provided lists.
+
+Farmland MSI species:
+```{r, echo = F}
+print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_farmland)])
+``` 
+Forest MSI species:
+```{r, echo = F}
+print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_forest)])
+```
+
+Mountain MSI species:
+```{r, echo = F}
+print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_mountain)])
+``` 
+Wetlands MSI species:
+```{r, echo = F}
+print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_wetlands)])
+```
 
@@
 
+In the next step, the downloaded raw data get analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data", @R-rtrim2020).
+The RTRIM analyses are run for a subset of 85 species which have sufficient data for it and using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/).
+For the application here, we have written wrapper function around the PECBMS code for each step of the RTRIM workflow.
+
+```{r, eval = FALSE}
+## Write PECBMS arguments input files for each species
+argument_file <- setupInputFiles_PECBMS_trimShell(Spp_selection = Spp_selection,
+                                                  folderPath = folder)
+
+## Subset data to contain only relevant species
+PECBMS_data <- makeInputData_PECBMS(Trim_data = Trim_data,
+                                    Spp_selection = Spp_selection,
+                                    convertNA = TRUE, 
+                                    save_allSppData = TRUE, returnData = TRUE)
+
+## Run analyses using the PECBMS Rtrim shell
+runRtrimShell_PECBMS(folder = folder)
+
+## Process results
+trimResults_PECBMS <- processRtrimOutput_PECBMS(folder = folder)
+```
+
+The results of the the RTRIM workflow are a series of .csv and .txt files.
+These are delivered directly to PECBMS each year after the analyses have been run.
+Prior to delivery, they get copied into a designated folder:
+
+```{r, eval = FALSE}
+## Collect (and rename) species and summary files
+collectSpeciesFiles_PECBMS(folder = folder, 
+                           subFolderName = subFolderName)
+```
+
 ### Combining data series using RSWAN
+
+For 55 of the 85 species, RTRIM analyses have also been run using data from the predecessor monitoring schemes.
+The RTRIM output files from both old and new analyses should now be present in the same folder.
+From there, they enter into the next step of the workflow: PECBMS' RSWAN routine.
+This is based on another set of scripts distributed through PECBMS which combine RTRIM outputs from different monitoring schemes / time periods.
+
+Again, we have written wrapper functions around the scripts from PECBMS to make for a more streamlined and automated workflow:
+
+```{r, eval = FALSE}
+## Write/load schedule table
+writeSchedule_SWAN(working_folder = working_folder_rel, 
+                   general_folder = general_folder_rel,
+                   MSI_speciesList = sppLists$sppLists$MSI,
+                   loadSchedule = FALSE)
+
+## Truncate first survey year where neccessary
+correctFirstSurveyYear_SWAN(general_folder = general_folder_rel, 
+                            working_folder = working_folder_rel,
+                            maxYear = 2022)
+
+## Run RSWAN
+combineTimeSeries_SWAN(general_folder_abs = general_folder,
+                       working_folder_abs = working_folder)
+
+```
+
+Running RSWAN requires some additional files that are provided by PECBMS, currently: "Swan_schedules.csv", "Swan_schedules_comb.csv", "Species_Countries.csv", and "New_notorics_2022.xlsx".
+
+(Once we turn ``birdIndicators`` into a package, we may add the content of these as integrated data.
+Although we would then need a function to write them to files as the core scripts from PECBMS need them to be files).
+
+THe RSWAN routine produces output files for the selected 55 species that are equivalent to those produced by the RTRIM workflow, but have a prefix "COMB" to indicate that they are based on combined time series.
 
@@
 
+The breeding bird indicators for ecosystem condition assessments are mutli-species indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.
+
+The function for calculating MSIs requires specifying whether or not to make use of combined time series generated by RSWAN (argument ``useCombTS``), which year to use as a base/reference (argument ``baseYear``), and which year to use as a changepoint (argument ``changepointYear``).
+The choice of change point year does not affect index values, but specifies until and from which year additional, shorter term trends should be calculated (i.e.
+if ``changepointYear = 2008``, the procedure will additionally provide trend estimates from starting year to 2008, and from 2008 to the last year in the data).
+
+```{r, eval = FALSE}
+useCombTS <- TRUE
+baseYear <- 1996
+changepointYear <- 2008
+```
+
+The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument ). We list these for all four ecosystem indices:
+
+```{r, eval = FALSE}
+IndexNames <- c("FarmlandBirds", "ForestBirds", "MountainBirds", "WetlandBirds")
+Index_sppLists <- list(sppLists$sppLists$MSI_farmland,
+                       sppLists$sppLists$MSI_forest,
+                       sppLists$sppLists$MSI_mountain,
+                       sppLists$sppLists$MSI_wetlands)
+```
+
+First, we will calculate the farmland and forest bird indices starting from 1996 and using that year as the reference.
+
+```{r, eval = FALSE}
+MSI_longTerm <- data.frame()
+for(i in 1:2){
+  message(paste0(crayon::bold("Calculating ", IndexNames[i], " Index...")))
+  results <- calculateIndex_MultiSpecies(working_folder = working_folder_rel, 
+                                         Spp_subset = Index_sppLists[[i]], 
+                                         IndexName = IndexNames[i], 
+                                         results_folder = MSI_results_folder,
+                                         useCombTS = useCombTS,
+                                         baseYear = baseYear,
+                                         changepointYear = changepointYear)
+  MSI_longTerm <- rbind(MSI_longTerm, results)
+  message("")
+}
+```
+
+```{r, echo = FALSE}
+MSI_results <- readRDS("data/MSI_results.rds")
+MSI_longTerm <- MSI_results$MSI_baseline1996
+MSI_all <- MSI_results$MSI_baseline2008
+```
+
+The MSI is provided both as point estimates (``MSI``) with associated standard error (``sd_MSI``) and confidence interval (``lower_CL_MSI`` and ``upper_CL_MSI``), and as a smoothed trend model fit (characterised by an average trend, ``Trend``, and an associated confidence interval, ``lower_CL_trend`` and ``upper_CL_trend``):
+
+```{r}
+head(MSI_longTerm)
+```
+
+Additionally, the results are saved as a standardized set of files used for reporting to PECBMS.
+
+We can then visualize the results both as point MSI only and as point MSI + trend fit:
+
+```{r}
+plotTimeSeries_MSI(MSI = MSI_longTerm,
+                   results_folder = MSI_results_folder,
+                   plot_name = "MSI_base1996_AllEcosystems", 
+                   displayPlots = TRUE,
+                   savePDF = FALSE)
+```
+
+Now we'll calculate indices for all ecosystems using the same setup, but using 2008 as a reference year for all of them to make values comparable:
+
+```{r, eval = FALSE}
+## Setting function arguments
+useCombTS <- c(TRUE, TRUE, FALSE, FALSE)
+baseYear <- 2008
+
+## Calculate MSI for each ecosystem
+MSI_all <- data.frame()
+for(i in 1:length(IndexNames)){
+  message(paste0(crayon::bold("Calculating ", IndexNames[i], " Index...")))
+  results <- calculateIndex_MultiSpecies(working_folder = working_folder_rel, 
+                                         Spp_subset = Index_sppLists[[i]], 
+                                         IndexName = IndexNames[i], 
+                                         results_folder = MSI_results_folder,
+                                         useCombTS = useCombTS[i],
+                                         baseYear = baseYear,
+                                         changepointYear = changepointYear)
+  MSI_all <- rbind(MSI_all, results)
+  message("")
+}
+```
+```{r}
+## Plot results
+plotTimeSeries_MSI(MSI = MSI_all,
+        results_folder = MSI_results_folder,
+        plot_name = "MSI_base2008_AllEcosystems",
+        displayPlots = TRUE,
+        savePDF = FALSE)
+```
+
 ### Uncertainty
+
+Uncertainty in the point estimate MSI is provided via an associated standard error and confidence interval.
+Uncertainty in the smoothed trend model fit is provided via an associated confidence interval.
 
@@
 
+Below we are preparing the output in the requested format.
+We are reporting the point estimate MSI here but note that we have not discussed to what degree the smoothed trend may be suitable.
+The breeding bird indicators exist only at the national scale as of now, so there are no spatial units associated with them.
+
+```{r}
+MSI_export <- list(NA, NA, NA, NA)
+names(MSI_export) <- unique(MSI_all$IndexName)
+
+for(i in 1:length(MSI_export)){
+
+  # Get mean indicator values in wide format
+  mean_wide <- MSI_all %>%
+    dplyr::filter(IndexName == names(MSI_export)[i]) %>%
+    dplyr::mutate(toColNames = paste0("v_", year)) %>%
+    reshape2::dcast(IndexName ~ toColNames, value.var = "MSI") %>%
+    dplyr::select(-IndexName)
+
+  # Get sd of indicator values in wide format
+  sd_wide <- MSI_all %>%
+    dplyr::filter(IndexName == names(MSI_export)[i]) %>%
+    dplyr::mutate(toColNames = paste0("sd_", year)) %>%
+    reshape2::dcast(IndexName ~ toColNames, value.var = "sd_MSI") %>%
+    dplyr::select(-IndexName)
+
+  # Combine and add additional columns
+  MSI_export[[i]] <- cbind(mean_wide, sd_wide) %>%
+    dplyr::mutate(reference_high = NA, 
+                  reference_low = 0, 
+                  thr = NA)
+}
+```
+Notably, I have not added the ``i_YYYY` columns yet as these can only be calculated once we have discussed and come to an agreement on reference value (year), threshold value, and scaling function.
+
 ### Eksport file (final product)
+
+```{r}
+saveRDS(MSI_export, file = "indicators/BreedingBirdIndicators.rds")
+```
+
+### Tidy-up
+At the very end we want to remove the downloaded "birdIndicators" repository from the temp directory:
+```{r}
+unlink("temp/birdIndicators_Repo.zip", recursive = TRUE)
+unlink("temp/birdIndicators-main", recursive = TRUE)
+```

### 2023-10-21 DRAFT_birdIndicators.Rmd
Updated files after dropping 2007 from mountain/wetlands MSI

@@
 This is also closely related to a second question that requires discussion: over what time period should breeding bird indicators be considered in ecosystem condition assessments?
-One option would be to use an average over the entire period with available data (but note that for mountain and wetland indicators, the starting year may have to be set to 2008 or even 2009 or 2010 to ensure sufficient data coverage).
+One option would be to use an average over the entire period with available data).
 Alternatively, averages over shorter periods leading up to the year for which ecosystem condition should be assessed.
@@
 
-Now we'll calculate indices for all ecosystems using the same setup, but using 2008 as a reference year for all of them to make values comparable:
+Now we'll calculate indices for all ecosystems using the same setup, but using 2008 as a reference year.
+This makes the farmland and forest indices which can be calculated from 1996, and the mountain and wetland indices which can only be calculated from 2008 (due to insufficient data before) comparable:
 

### 2023-10-23 trophic_levels-biomass_indicators.rmd
add captions, exuations, and cache big chunks

@@
 
+<br />
+Norwegian names of indicators:
+
+- **Store beitedyr**
+- **Store rovdyr**
 <br />
@@
 library(tidyverse)
-
+library(flextable)
 # Set up conditional file paths
@@
 
-## Introduction
+## Introduction {#intro-tlb}
 
@@
 
-## About the underlying data
+## About the underlying data {#underlying-data-tlb}
 
@@
 
-### Representativity in time and space
+### Representativity in time and space {#rep-time-and-space-tlb}
 
@@
 
-NPP^0.47^ \* 0.643
+$$
+NPP^{0.47}\times0.643
+$$
 
@@
 
-Herbivore biomass^0.73^ \* 0.014
+$$
+Herbivore\ biomass^{0.73}\times 0.014
+$$
 
@@
 The choice of 40% is arbitary, and is based on the expert opinion of the authors.
+
+Reference levels for this indicator are discussed [here](https://github.com/NINAnor/ecosystemCondition/issues/113).
 
@@
 
-## Analyses
+## Analyses {#analyses-tlb}
 
@@
 ```{r, message=FALSE, warning=FALSE}
-#| label: Plotting basic data
+#| label: Plotting-basic-data
 #| cache: TRUE
+#| fig.cap: "Net primary production over Norway"
 
@@
 #| cache: TRUE
+#| fig.cap: "Herbivore biomass over Norway"
 #Total herbivore biomass
@@
 names(totalherbivorebiomass)<-allyears_vect
-totalherbivorebiomass
+#totalherbivorebiomass
 ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
@@
 ```{r, message=FALSE, warning=FALSE}
-#| label: Plot carnivore biomass
+#| label: Plot-carnivore-biomass
 #| cache: TRUE
+#| fig.cap: "Carnivore biomass over Norway
 carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
@@
 ```{r, message=FALSE, warning=FALSE}
-#| label: Calculating expected herbivore biomass
+#| label: Calculating-expected-herbivore-biomass
 #| cache: TRUE
+#| fig.cap:"Expected herbivore biomass"
 expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
@@
 
-100 x (ActualBiomass -- ExpectedBiomass) / (ExpectedBiomass + 1)
+$$
+100 \times (ActualBiomass - ExpectedBiomass) / (ExpectedBiomass + 1)
+$$
 
@@
 #| cache: TRUE
+#| fig.cap:"Difference from expected herbivore biomass"
 totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
@@
 
--- Next we move on to estimate the expected carnivore biomass based on observed herbivore biomass and calculate the deviation as a % difference from expected as for the herbivores.
+Next we move on to estimate the expected carnivore biomass based on observed herbivore biomass and calculate the deviation as a % difference from expected as for the herbivores.
 This is based on Hatton et al.
@@
 ```{r}
-#| label: Expected carnivore biomass
+#| label: Expected-carnivore-biomass
 #| cache: TRUE 
+#| fig.cap:"Expected carnivore biomass"
+#| warning: FALSE
 
@@
   scale_fill_continuous(low='lightskyblue',high="darkblue",na.value=NA,expression('kg km'^-2),trans='log10')
+```
 
+```{r diff-carnivores-tlb, fig.cap="Difference from expected carnivore biomass"}
 diffcarn<-100* (totalcarnivores-expected_carnivore_biomass)/(expected_carnivore_biomass+1)
@@
 
+<!-- Limit code block height and force scrolling -->
+```{css, echo=FALSE}
+pre {
+  max-height: 300px;
+  overflow-y: auto;
+}
+
+pre[class] {
+  max-height: 200px;
+}
+```
+
 ```{r}
@@
 #| cache: TRUE
+#| eval: FALSE
 
@@
 ggplot()+geom_spatraster(data=artype50_F)+scale_fill_manual(values=arcols,na.value = NA)
+ggsave("data/cache/gg1_tlb.png")
 
@@
   labs(x="Year", y="Difference")
+ ggsave("data/cache/gg2_tlb.png")
 
@@
   labs(x="Year", y="Difference")
+  ggsave("data/cache/gg3_tlb.png")
 
@@
    facet_wrap(~lyr)+theme_bw()
+  ggsave("data/cache/gg4_tlb.png")
 
@@
    facet_wrap(~lyr)+theme_bw()
+  ggsave("data/cache/gg5_tlb.png")
 
+```
+
+```{r ar50, fig.cap="A view of the AR50 data used as a placeholder for an ecosystem map.", out.width='50%'}
+knitr::include_graphics("data/cache/gg1_tlb.png")
+```
+
+```{r forestdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='50%'}
+knitr::include_graphics("data/cache/gg2_tlb.png")
+```
+
+```{r forestcarn-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='50%'}
+knitr::include_graphics("data/cache/gg3_tlb.png")
+```
+
+```{r forestherbdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='50%'}
+knitr::include_graphics("data/cache/gg4_tlb.png")
+```
+
+```{r forestcarn2-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='50%'}
+knitr::include_graphics("data/cache/gg5_tlb.png")
 ```
@@
 
-```{r}
+```{r, eval=F}
 #| label: CountySummaries
@@
 #forest_carn_herb_countyDF$YearN<-as.numeric(substr(forest_carn_herb_countyDF$Year,2,5))
-forest_carn_herb_countyDF$YearN<-as.numeric(forest_carn_herb_countyDF$Year)
+forest_carn_herb_countyDF$Year2 <- substr(forest_carn_herb_countyDF$Year, 2, nchar(forest_carn_herb_countyDF$Year))
+forest_carn_herb_countyDF$YearN<-as.numeric(forest_carn_herb_countyDF$Year2)
 
@@
   geom_text(data=forest_carn_herb_countyDF[forest_carn_herb_countyDF$YearN==2015,],aes(label = CountyName, x = 2015, y = Difference), hjust = -.1) 
+ggsave("data/cache/gg6_tlb.png")
 
@@
 forest_herb_veg_countyDF<-gather(forest_herb_veg_county,key="Year",value="Difference",-ID,-CountyName)
-forest_herb_veg_countyDF$YearN<-as.numeric(forest_herb_veg_countyDF$Year)
+forest_herb_veg_countyDF$Year2 <- substr(forest_herb_veg_countyDF$Year, 2, nchar(forest_herb_veg_countyDF$Year))
+forest_herb_veg_countyDF$YearN<-as.numeric(forest_herb_veg_countyDF$Year2)
 ggplot(data=forest_herb_veg_countyDF,aes(x=YearN,y=Difference,color=CountyName))+geom_line()+scale_color_discrete()+
@@
   geom_text(data=forest_herb_veg_countyDF[forest_herb_veg_countyDF$YearN==2015,],aes(label = CountyName, x = 2015, y = Difference), hjust = -.1) 
+ggsave("data/cache/gg7_tlb.png")
 
 ```
+
+```{r herbtime, fig.cap="Difference in herbivore biomass from expectations"}
+knitr::include_graphics("data/cache/gg6_tlb.png")
+```
+
+```{r carntime, fig.cap="Difference in herbivore biomass from expectations"}
+knitr::include_graphics("data/cache/gg7_tlb.png")
+```

### 2023-10-23 trophic_levels-biomass_indicators.rmd
yml errors and set some echo false

@@
 ```{r, message=FALSE, warning=FALSE}
-#| label: Data import
+#| label: Data-import
 #| cache: TRUE
@@
 
-```{r, message=FALSE, warning=FALSE}
+```{r, message=FALSE, warning=FALSE, fig.cap= "Net primary production over Norway"}
 #| label: Plotting-basic-data
 #| cache: TRUE
-#| fig.cap: "Net primary production over Norway"
 
@@
 #| fig.cap: "Herbivore biomass over Norway"
+
 #Total herbivore biomass
@@
 #| cache: TRUE
-#| fig.cap: "Carnivore biomass over Norway
+#| fig.cap: "Carnivore biomass over Norway"
 carnivorerast<-rast(paste0(pData,"/Trophic_levels/Carnivores_RawBiomass.tiff"))
@@
 #| cache: TRUE
-#| fig.cap:"Expected herbivore biomass"
+#| fig.cap: "Expected herbivore biomass"
 expected_herbivore_biomass<-(npprast_yrs^0.47 * 0.643)#Global 
@@
 #| cache: TRUE
-#| fig.cap:"Difference from expected herbivore biomass"
+#| fig.cap: "Difference from expected herbivore biomass"
 totalherbivorebiomass<-livestockrast[[sapply(strsplit(names(livestockrast),"_"),'[',1) %in% "TotalHerbivoreBiomass"]]
@@
 #| cache: TRUE 
-#| fig.cap:"Expected carnivore biomass"
+#| fig.cap: "Expected carnivore biomass"
 #| warning: FALSE
@@
 
-```{r ar50, fig.cap="A view of the AR50 data used as a placeholder for an ecosystem map.", out.width='50%'}
+```{r ar50, fig.cap="A view of the AR50 data used as a placeholder for an ecosystem map.", out.width='50%', echo=FALSE}
 knitr::include_graphics("data/cache/gg1_tlb.png")
@@
 
-```{r forestdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='50%'}
+```{r forestdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='100%', echo=FALSE}
 knitr::include_graphics("data/cache/gg2_tlb.png")
@@
 
-```{r forestcarn-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='50%'}
+```{r forestcarn-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='100%', echo=FALSE}
 knitr::include_graphics("data/cache/gg3_tlb.png")
@@
 
-```{r forestherbdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='50%'}
+```{r forestherbdiff-tlb, fig.cap="Difference in forest herbivore biomass from expectation", out.width='100%', echo=FALSE}
 knitr::include_graphics("data/cache/gg4_tlb.png")
@@
 
-```{r forestcarn2-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='50%'}
+```{r forestcarn2-tlb, fig.cap="Difference in forest carnivore biomass from expectation", out.width='100%', echo=FALSE}
 knitr::include_graphics("data/cache/gg5_tlb.png")
@@
 
-```{r herbtime, fig.cap="Difference in herbivore biomass from expectations"}
+```{r herbtime, fig.cap="Difference in herbivore biomass from expectations", echo=FALSE}
 knitr::include_graphics("data/cache/gg6_tlb.png")
@@
 
-```{r carntime, fig.cap="Difference in herbivore biomass from expectations"}
+```{r carntime, fig.cap="Difference in herbivore biomass from expectations", echo=FALSE}
 knitr::include_graphics("data/cache/gg7_tlb.png")

### 2023-10-23 DRAFT_Flood_frequency.Rmd
Format text and clarify around ref values

@@
----
-title: "Template for indicator documentation"
-output:
-  html_document
----
-
-<!-- This is a template for how to document the indicator analyses.
-
-Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name the file DRAFT_myIndicatorName.Rmd.
-
-Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
-you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work.
-
-Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout.
-
-When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
-the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. When completed, Anders will make a copy
-of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
-the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
-
-<!-- Examples:
-
-For an example, see 'indicators/breareal-workingFile.Rmd' 
-and the subsequent 'breareal.Rmd' 
-which again, when rendered, produces this:
-https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
-
-Notice the degree of annotation.
-
--->
-
-<!-- Replace 'indicator name' with yor own indicator name  -->
 # Flood frequency
@@
 <br />
+Norwegian name: **Oversvømmelseshyppighet**
+<br />
 
 _Author and date:_
-Vegar Bakkestuen, October 2023_
+
+Vegar Bakkestuen
+
+October 2023
 
@@
 Ecosystem <- "Våtmark"
-Egenskap  <- "Oversvømmelse"
-ECT       <- " Oversvømmelseshyppighet"
+Egenskap  <- "Abiotiske forhold"
+ECT       <- "A1 Physical state characteristics"
 Contact   <- "Vegar Bakkestuen"
@@
 
-<!-- Document you work below.
-Try not to change  the headers too much.
-Data can be stored on NINA server.
-Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine.
-If it is not too big you may store under /data/ on this repository -->
+```{r ff-example, fig.cap="Flood frequency last 5 years, somewhere in Norway. Light blue areas are flooded once, and dark blue areas are flooded five times.", out.width='100%', echo=FALSE}
+knitr::include_graphics("images/floodFreqExample.png")
+```
 
-## Introduction
-Water and inundated areas are a land cover class that can be relatively easily distinguished from other land cover classes with high accuracy using remote sensing data, particularly Copernicus Sentinel-1 data.
-Sentinel-1 employs Synthetic Aperture Radar (SAR) instruments that can "see through" clouds.
+## Introduction {#intro-ff}
+Water and inundated areas is a land cover class that can be relatively easily distinguished from other land cover classes with high accuracy using remote sensing data, particularly Copernicus Sentinel-1 data.
+Sentinel-1 employs Synthetic Aperture Radar (SAR) instruments that can _see through_ clouds. 
 In Norway, Sentinel-1 data is available at a minimum frequency of once a week.
 Utilizing Google Earth Engine, all available Sentinel-1 data can be accessed, and there are pre-established scripts for distinguishing water-covered areas from other land cover classes. 
-For example, you can refer to "Flood Mapping and Area Calculation of Flood Extent Using Sentinel-I SAR Data in Google Earth Engine: the case of Super Typhoon Odette (Rai)" [^1^] and "Applying Google Earth Engine for flood mapping and monitoring in the downstream provinces of the Mekong river" [^2^].
+For example, you can refer to "Flood Mapping and Area Calculation of Flood Extent Using Sentinel-I SAR Data in Google Earth Engine:
+
+"the case of Super Typhoon Odette (Rai)", and;
+
+"Applying Google Earth Engine for flood mapping and monitoring in the downstream provinces of the Mekong river".
+
 By aggregating Sentinel-1 images over an entire growing season, it becomes possible to model when areas are covered by water. 
@@
 This script uses Sentinel-1 imagery in a change detection approach, comparing images before and after a flood event. 
-The key preprocessing steps for Sentinel-1 Ground Range Detected (GRD) data are thermal noise removal, radiometric calibration, and terrain correction.
-Therefore, the primary preprocessing step is applying a speckle filter.
-In this script, you can define your study area, select your desired time frame, and set SAR parameters. Make sure to adjust these parameters according to your study area's specific conditions. 
-Workflow: Calculate Flood Indicator Using Google Earth Engine Script
+The key pre-processing steps for Sentinel-1 Ground Range Detected (GRD) data are thermal noise removal, radiometric calibration, and terrain correction. 
+Therefore, the primary pre-processing step is applying a speckle filter.
+In this script, you can define your study area, select your desired time frame, and set SAR parameters. 
+Make sure to adjust these parameters according to your study area's specific conditions.
+
+**Workflow**:
+
+Calculate Flood Indicator Using Google Earth Engine Script
+
 1.	Initialization:
+
     •	Set up your Google Earth Engine environment.
+
     •	Import the necessary geometries or AOI.
+
 2.	Define Functions:
+
     •	Define a function for calculating flood extent and area based on the provided script. The function should take parameters like the year, polarization, and AOI.
+
 3.	Define Constants:
+
     •	Set constant values such as polarization, pass direction, smoothing radius, and difference threshold. Ensure these values are appropriate for your study area.
+
 4.	Define Years and Date Ranges:
+
     •	Create an array or list of years for which you want to calculate the flood indicator.
+
     •	Define the "before" and "after" date ranges for each year based on your specific study periods.
+
 5.	Loop Through Years:
+
     •	Use a loop to iterate through each year in your defined list.
+
     •	For each year, call the flood extent calculation function you defined, passing the appropriate parameters (e.g., SAR collections for "before" and "after," polarization, AOI).
+
 6.	Calculate Flood Frequency:
+
     •	After calculating flood extent for each year, calculate the flood frequency. You can create a frequency count image by accumulating flood extent layers from each year.
+
 7.	Data Visualization:
+
     •	Define a color palette for displaying flood frequency on the map.
+
     •	Add the flood frequency layer to the map for visualization.
+
     •	Optionally, you can also display individual flood extents for each year.
+
 8.	Print Results:
+
     •	In the loop, after calculating flood extent for each year, print the calculated flood extent areas to the console for each year.
+
 9.	Data Export (Optional):
+
     •	If needed, you can export the results, such as the frequency map, for further analysis or reporting.
+
 10.	Fine-Tuning:
+
     •	As needed, adjust parameters in the script, such as smoothing radius, difference threshold, or additional data sources, to optimize the flood indicator calculation.
+
 11.	Interpretation:
+
     •	Analyze the results to understand the flood extent areas and frequencies for different years.
 
-## About the underlying data
+## About the underlying data {#underlying-data-ff}
+
 The flood frequency indicator is derived from a time series of Sentinel-1 SAR data covering multiple years. The analysis involves the following steps:
+
 1.	Data Collection: Sentinel-1 GRD imagery, including the VH polarization, was used to create a comprehensive dataset.
@@
 
-### Representativity in time and space
+### Representativity in time and space {#rep-time-space-ff}
 The analysis spans multiple years, offering a long-term view of flooding events in the study area.
 This extended time frame enables the identification of recurrent patterns and trends, which is vital for understanding wetland dynamics.
+
 ### Original units
 The original units of the calculated flood extent area are in square meters.
+
 ### Temporal coverage
@@
 ## Ecosystem characteristic
-In the context of the Norwegian standard, the 'Flood frequency' indicator aligns with the ecosystem characteristic known as 'Biologisk mangfold' or 'Biological diversity.' This classification underscores the significance of the indicator in assessing its impact on biological diversity within ecosystems.
+In the context of the Norwegian standard, the 'Flood frequency' indicator aligns with the ecosystem characteristic known as _Abiotiske forhold_. 
 Flood frequency is a crucial factor in understanding how changes in hydrology, including flood events, influence the composition and distribution of species within ecosystems. 
 It can also be instrumental in evaluating the resilience of ecosystems to such disturbances.
+
 Furthermore, the indicator could also be relevant in the context of 'funksjonelt viktige arter og strukturer,' or 'functionally important species and structures.' Flood frequency plays a pivotal role in shaping the habitat availability and accessibility for species critical to ecosystem functioning.
@@
 Frequent or infrequent flood events can have far-reaching consequences on the viability of these functionally important components of ecosystems.
+
 Therefore, understanding and monitoring flood frequency not only serve as a hydrological metric but also have profound implications for ecosystem dynamics, which are vital for the overall health and sustainability of ecosystems.
 This integrated perspective highlights the interconnectedness of hydrological patterns and ecosystem characteristics, emphasizing the indicator's significance in ecological and conservation assessments.
-UN Standard
+
 ### SEEA EA (UN standard)
-In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Flood frequency" indicator can be classified as a B1 "Compositional state characteristics" indicator.
-This classification within the SEEA EA framework highlights the indicator's role in assessing the compositional state of ecosystems.
-By examining how connectivity influences the arrangement and presence of species and structures within an ecosystem, it provides valuable insights into the overall composition and structure of ecosystems.
+In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the "Flood frequency" indicator can be classified as a _Physical state characteristics_.
 
@@
 ### Reference condition
-The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems.
-In this reference condition, the landscape exhibits natural and undisturbed patterns of connectivity between suitable habitats.
-This state represents the optimal ecological condition for connectivity assessment.
+The reference condition is one where human activities that impact large scale hydrological patterns are minimal.
+The most important drivers of anthropogenic change are climate change, water regulation, draning and flood prevention and river channelization.
 
 ### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
-Reference values for hydrological parameters, such as water level, water flow, and temperature, are crucial for assessing the ecological condition of aquatic ecosystems.
-In the context of this analysis, we are planning to derive reference values from extensive datasets collected by the Norwegian Water Resources and Energy Directorate (NVE) through its hydrological API (HydAPI).
+Reference values are not yet developed for this indicator. 
+Note that the term reference level when used for this indicator sometime pertains to baseline hydrological conditions, and not a reference for the level of flood frequency under the reference condition.
+
+## References
+## Analyses
+### NVE data
+
+Innsynsløsning: https://nina.maps.arcgis.com/apps/instant/basic/index.html?appid=4344dd7f769048ed9dedf1803638ff48
+
+Baseline data for hydrological parameters, such as water level, water flow, and temperature, are crucial for assessing the ecological condition of aquatic ecosystems. 
+In the context of this analysis, we are planning to derive baseline values from extensive datasets collected by the Norwegian Water Resources and Energy Directorate (NVE) through its hydrological API (HydAPI).
 NVE has been instrumental in gathering long-term hydrological data, with records dating back to the 1800s. 
 These historical measurements serve as a valuable baseline to evaluate contemporary conditions.
-W
-Data Retrieval Solution - Python Code (get-observations-NINA.py): This Python script (get-observations-NINA.py) serves as a data retrieval solution to access hydrological data from the Norwegian Water Resources and Energy Directorate's (NVE) Hydrological API (HydAPI).
+
+W Data Retrieval Solution - Python Code (get-observations-NINA.py):
+
+This Python script (get-observations-NINA.py) serves as a data retrieval solution to access hydrological data from the Norwegian Water Resources and Energy Directorate's (NVE) Hydrological API (HydAPI).
 It is designed to fetch observational data and metadata for hydrological stations across Norway. 
 The script is versatile and allows users to specify parameters such as the station ID, observed parameters, resolution time, reference time, and a mandatory API key.
+
 •	Station ID (-s): Users can input one or multiple station IDs separated by a comma, allowing access to specific hydrological data from various locations.
+
 •	Parameters (-p): This parameter allows users to specify the type of observation data they need. Multiple parameters can be specified, enhancing flexibility.
+
 •	Resolution Time (-r): Users can select the resolution time, choosing between instantaneous (0), hourly (60), or daily (1440) data.
+
 •	API Key (-a): The API key is a mandatory component to authenticate and access the HydAPI data. Users need to register and obtain their API key from the HydAPI platform.
+
 The script leverages Python's request library to send HTTP requests to the HydAPI, and it retrieves data in JSON format. 
@@
 The retrieved data is written to a user-defined output file, facilitating further analysis and utilization.
+
 Data Retrieval Procedure:
+
 To retrieve data using the code:
+
 1.	Set Working Directory: Define the desired folder path, typically using the os.chdir() function, to ensure that the script operates in the intended directory.
@@
 Skriptet heter get-observations-NINA.py.
-Den fungerer hvis man oppretter en mappe på C:/ kalt Konnektivetet/python og legger scriptet der.
+Den fungerer hvis man oppretter en mappe på `C:/` kalt `Konnektivetet/python` og legger scriptet der.
 
+```{js}
 #!/usr/bin/python
@@
 
+```
+
 Dataene hentes ut ved å skrive inn følgende kommandoer i python vinduet:
 
+```{js}
 import os
@@
 subprocess.call(kommando)
-Innsynsløsning: https://nina.maps.arcgis.com/apps/instant/basic/index.html?appid=4344dd7f769048ed9dedf1803638ff48
+```
 
@@
 
-## References
-## Analyses
-### Data sets
+### Scaled indicator values
+The indicator is not scaled.
 
-### Scaled indicator values
 ### Uncertainty
-## Prepare export
+We have no measure for the uncertainty or errors assosiated with the estimated frequency of floods.
+
+## JAVA code
 You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.
 
-### Eksport file (final product)
+```{js}
+
 /*===========================================================================================
@@
 Map.addLayer(wetlandAffected, {min: 1, max: 5, palette: colorPalette}, 'Flood Frequency in Wetlands');
+
+```

### 2023-10-23 NDVI_natopen.Rmd

@@
 # NDVI - Naturally Open Ecosystems {#NDVI-indicator-natopen}
+
+<br /> Norwegian name: **Primærproduksjon** <br />
 

### 2023-10-23 NDVI_seminat.Rmd
more prallel-comit-issues fixed

@@
 In addition, there are 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40
-that are not assigned to any region.
+that are not assigned to any region. (These polygons likely lie in water and thus outside of the regional polygons due to imprecise coordinates from the mapping device).
 

### 2023-10-23 alien_species.Rmd

@@
 library(gdalUtilities)
+library(flextable)
 knitr::opts_chunk$set(echo = TRUE)
@@
 
-```{r alien-precentageplot, fig.height=5, fig.width=8, fig.cap="Relative frequency plot showing the distribution of polygons with different indicator values."}
+```{r alien-precentageplot, fig.cap="Relative frequency plot showing the distribution of polygons with different indicator values."}
 ggplot(corrCheck, aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
@@
 areas.
+
+```{r, include=FALSE, eval=FALSE}
+# Figure for the report
+corrCheck  %>%
+  mutate(hovedøkosystem = forcats::fct_recode(factor(hovedøkosystem), "Naturlig åpne" = "naturligÅpneOmråderUnderSkoggrensa",
+                      "Semi-naturlig mark" = "semi-naturligMark",
+                      "Våtmark" = "våtmark")) %>%
+ggplot(aes(x = factor(infrastructureIndex), fill = factor(round(i_2020,2))))+
+  geom_bar(position="fill")+
+  theme_bw(base_size = 12)+
+  guides(fill = guide_legend("Indikatorverdi"))+
+  ylab("Andel av datapunkter")+
+  xlab("Infrastrukturindeksen (kategorisk versjon)")+
+  scale_fill_brewer(palette = "RdYlGn")+
+  facet_wrap(.~hovedøkosystem)
+
+```
 

### 2023-10-23 functional_plant_indicators_wetland.Rmd

@@
-# Functional Plant Indicators - Wetlands {#Functional-plant-indicators-wetland}
+# Functional Plant Indicators - Wetlands {#functional-plant-indicators-wetland}
 
@@
 library(glmmTMB)
+library(flextable)
+
 knitr::opts_chunk$set(echo = TRUE)

### 2023-10-23 medianSummerTemperature.Rmd

@@
 knitr::opts_chunk$set(echo = TRUE)
-library(stringr, lib.loc = "/usr/lib/R/site-library")
-library(tidyverse, lib.loc = "/usr/local/lib/R/site-library")
-#library(tidyverse)
+library(tidyverse)
 library(stars)
@@
 library(terra)
+library(flextable)
 dir <- substr(getwd(), 1,2)
@@
                        "Økologisk egenskap" = Egenskap,
-                       "ECT class" = ECT)
-knitr::kable(metaData)
+                       "ECT class" = ECT,
+                       check.names = F)
+flextable(metaData) %>%
+  bg(bg = "lightblue", part="header") %>%
+  theme_vanilla() %>%
+  set_table_properties(layout = "autofit")
+
 ```

### 2023-10-23 Nature_Index_product.Rmd
delete html and md in root. Rm cache and temp files. Render

@@
 ```{r setup, include=FALSE}
-library(NIflex)
 library(magrittr)
@@
 library(flextable)
+library(NIflex)
+library(NIcalc)
 knitr::opts_chunk$set(echo = TRUE)
@@
 This approach would produce an indicator which reflect the _population_ status rather than the mean status for any given area.
+We note, however, that this alternative approach is not (yet) implemented as a routine in the ``NIcalc`` R package, and code to implement and execute it would have to be developed first.
 
@@
 
-```{r, include=F, eval=F}
+```{r, include=F}
 dir <- substr(getwd(), 1,2)
@@
                "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
+
+```
+
+```{r, include=F, eval=F}
 saveRDS(CustomNI, paste0(pData, "cache/CustomNI.RDS"))

### 2023-10-23 trophic_levels-biomass_indicators.rmd

@@
 ```{r, message=FALSE, warning=FALSE}
-#| label: Data-import
-#| cache: TRUE
+#| label: data-import
 
@@
 #| label: Plotting-basic-data
-#| cache: TRUE
 
@@
 #| label: PlotTotalHerbivoreBiomass
-#| cache: TRUE
 #| fig.cap: "Herbivore biomass over Norway"
@@
 #| label: Plot-carnivore-biomass
-#| cache: TRUE
 #| fig.cap: "Carnivore biomass over Norway"
@@
 #| label: Calculating-expected-herbivore-biomass
-#| cache: TRUE
 #| fig.cap: "Expected herbivore biomass"
@@
 #| label: DifferenceHerbivoreBiomass
-#| cache: TRUE
 #| fig.cap: "Difference from expected herbivore biomass"
@@
 #| label: Expected-carnivore-biomass
-#| cache: TRUE 
 #| fig.cap: "Expected carnivore biomass"
@@
 ```{r diff-carnivores-tlb, fig.cap="Difference from expected carnivore biomass"}
-diffcarn<-100* (totalcarnivores-expected_carnivore_biomass)/(expected_carnivore_biomass+1)
-names(diffcarn)<-allyears_vect
+diffcarn <-100*(totalcarnivores-expected_carnivore_biomass)/(expected_carnivore_biomass+1)
+names(diffcarn) <- allyears_vect
 

### 2023-10-24 DRAFT_birdIndicators.Rmd
Removed one eval = FALSE that prevented Trim_data download
Added sentence on specification of database driver

@@
 From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2022).
+The function ``downloadData_TRIM`` does that for a specified year range (from ``minYear`` up to and including ``maxYear``), and requires specifying which driver is used for accessing the database (via argument ``DriverName``).
+When running from the NINA RStudio server, we need to use "FreeTDS".
+When running locally from NINA computers, we need to select "SQL Server".
 
-```{r, eval = FALSE}
+```{r}
 minYear <- 2006

### 2023-10-30 DRAFT_birdIndicators.Rmd
Added spp list as data object to run chapter without data download
Downloading data from Hekkefuglovervåkingen requires user-specific access privileges. To make the chapter run-able by everyone, I have added the object that relies on the downloaded data as an .rds file instead. Reading that in means we do not need to download the data within the chapter.

@@
 When running locally from NINA computers, we need to select "SQL Server".
+Note that downloading this data requires user-specific access privileges which can be requested from NINA.
 
-```{r}
+```{r, eval = FALSE}
 minYear <- 2006
@@
 
-```{r}
+```{r, echo = FALSE}
+sppLists <- readRDS("data/birdIndicators_sppLists.rds")
+Spp_selection <- sppLists$sppData
+```
+
+```{r, eval = FALSE}
 sppLists <- makeSpeciesLists(Trim_data = Trim_data, spp_DataPath = "temp/birdIndicators-main/data")

### 2023-10-30 DRAFT_birdIndicators.Rmd
Formatting and test render

@@
 <br />
+Norwegian name: **Hekkefulgindikatorer**
+<br />
 
 _Author and date:_
-Chloé R. Nater, Diego Pavón-Jordán
+
+Chloé R. Nater,
+
+Diego Pavón-Jordán
+
+Oktober 2023
 
@@
 library(readxl)
+library(flextable)
 library(data.table)
@@
 ```{r, echo=F}
-Ecosystem <- "Skog, fjell, våtmark, seminaturlig mark" 
-Egenskap  <- "Biologisk mangfold" 
-ECT       <- "Compositional state characteristic" 
-Contact   <- "Chloé R. Nater" 
+Ecosystem <- c("Skog", "fjell", "våtmark", "seminaturlig mark")
+Egenskap  <- rep("Biologisk mangfold" , 4)
+ECT       <- rep("Compositional state characteristic" , 4)
+Contact   <- rep("Chloé R. Nater" , 4)
 ```
@@
 ### Representativity in time and space
-"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway (MAP HERE??
-I can provide this).
+"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway.
 Each route has between 12 and 20 counting points (ca.
@@
 
-We note that composite breeding bird indicators could also be built as Nature Index (NI) product (see separate chapters).
+We note that composite breeding bird indicators could also be built as [Nature Index (NI) product]("nature-index-product) (see separate chapters).
 For the species included in NI, reference values corresponding to "ecosystems with minimal human impacts" were determined previously, and an NI product for a selection of breeding birds would therefore be an indicator taking values between 0 and 1, where 1 = reference condition.
@@
 The reference value 100 represents unchanged abundance relative to the reference year.
+
 Determining a threshold for what level of decrease in bird abundance is acceptable under "good ecological condition", however, is a challenging task. 
@@
 
-```{r}
+```{r, results='hide'}
 download.file(url = "https://github.com/NINAnor/birdIndicators/archive/refs/heads/main.zip",
@@
 The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåkingen" and the calculated abundance indices from the predecessor monitoring schemes.
-Additionally, we also a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow.
+Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow.
 
@@
 Farmland MSI species:
-```{r, echo = F}
-print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_farmland)])
+
+```{r, echo = F, results='asis'}
+temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_farmland)])
+cat(paste0("- ", temp), sep = "\n")
 ```
+
 Forest MSI species:
-```{r, echo = F}
-print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_forest)])
+
+```{r, echo = F, results='asis'}
+temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_forest)])
+cat(paste0("- ", temp), sep = "\n")
 ```
@@
 Mountain MSI species:
-```{r, echo = F}
-print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_mountain)])
+
+```{r, echo = F, results='asis'}
+temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_mountain)])
+cat(paste0("- ", temp), sep = "\n")
 ```
+
 Wetlands MSI species:
-```{r, echo = F}
-print(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_wetlands)])
+
+```{r, echo = F, results='asis'}
+temp <- sort(Spp_selection$Species[which(Spp_selection$EURINGCode %in% sppLists$sppLists$MSI_wetlands)])
+cat(paste0("- ", temp), sep = "\n")
 ```
@@
 
-```{r}
+```{r, eval=F}
 MSI_export <- list(NA, NA, NA, NA)
@@
 
-```{r}
+```{r, eval=F}
 saveRDS(MSI_export, file = "indicators/BreedingBirdIndicators.rds")

### 2023-10-30 birdIndicators.Rmd
rename and add some anchors

--- 2023-10-30 DRAFT_birdIndicators.Rmd
@@
----
-title: "Breeding bird indicators"
-output:
-  html_document
----
-
-# Breeding bird indicators
+# Breeding bird indicators {#breeding-birds}
 
@@
 
-## Introduction
+## Introduction {#intro-bb}
 The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called TOV-E) was established between 2005 and 2010.
@@
 
-## About the underlying data
+## About the underlying data {#underlying-bb}
 The data from "hekkefuglovervåkingen" are gathered mostly by experienced volunteer bird-watchers following a standardized sampling protocol (https://hekkefuglovervakingen.nina.no/Fugl/public/papirskjema/MethodologyEng.pdf).
@@
 
-### Representativity in time and space
+### Representativity in time and space {#rep-in-time-and-space-bb}
 "Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway.
@@
 
-### Original units
+### Original units {#original-units-bb}
 The observers in the field record "number of individuals" seen or heard during the census.
@@
 
-### Temporal coverage
+### Temporal coverage {#temp-coverage-bb}
 The "hekkefuglovervåkingen" was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010.
@@
 
-### Aditional comments about the dataset
+### Aditional comments about the dataset {#additional-comments-bb}
 From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses.
@@
 
-## Ecosystem characteristic
+## Ecosystem characteristic {#ecosystem-characteristics-bb}
 
@@
 
-## Reference condition and values
-### Reference condition
+## Reference condition and values {#ref-con-and-vals-bb}
+### Reference condition {#ref-con-bb}
 In general, the reference condition for breeding bird indicators should be defined along the same lines as for the other indicators in ecological condition assessments.
@@
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values
+### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-vals-bb}
 The breeding bird indicators represent average trends relative to a reference year and are therefore scaled with respect to population size in the reference year.
@@
 
-## Uncertainties
+## Uncertainties {#errors-bb}
 There is a substantial amount of uncertainty related to which year should be considered as a baseline and how the reference and threshold values should be set (see preceeding section).
@@
 
-## References
+## References {#refs-bb}
 
@@
 
-## Analyses
+## Analyses {#analyses-bb}
 
@@
 
-### Data sets
+### Data sets {#datasets-bb}
 The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåkingen" and the calculated abundance indices from the predecessor monitoring schemes.
@@
 
-#### Regions
+#### Regions {#regions-bb}
 
@@
 
-## Prepare export
+## Prepare export {#prepare-export-bb}
 
@@
 
-### Eksport file (final product)
+### Eksport file (final product) {#export-bb}
 

### 2023-11-13 Nature_Index_product.Rmd
Included indicator "tamrein"

@@
 
-Here, we focus primarily on showcasing thw workflow for creating an NI product for wetlands, and the following table gives an overview of all the indicators included in the official NI for wetlands in 2020.
+Here, we focus primarily on showcasing the workflow for creating an NI product for wetlands, and the following table gives an overview of all the indicators included in the official NI for wetlands in 2020.
 It also provides information on each indicators trophic level, underlying data, whether we expect updated indicator data towards NI 2025, and additional notes.
@@
 Considering all the groups of indicators that should not be included in NI products for ecosystem condition assessments opens up a discussion on whether, instead of using a composite index across a small and very unbalanced set of species (groups), a better approach might be to design NI products that represent specific taxonomic and/or functional species groups.
-Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains are eight species of vascular plants and wild reindeer.
-Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer as a separate indicator.
+Take, for example, the wetlands NI product presented below: if the bird indicators are removed as they are a separate index, and the four indicators representing expert knowledge on rare mosses and insects are discontinued, all that remains are eight species of vascular plants, wild reindeer, and semi-domestic reindeer.
+Instead of averaging across plants and reindeer here, it may be more sensible to make an NI product for vascular plants, and potentially include reindeer (either joint or divided into wild and semi-domestic) as a separate indicator.
 
@@
 The alternative is to set the mode to "custom" and specify which indicators should be dropped.
-We will demonstrate this here as we would also like to also drop the two indicators that do not represent species abundance but habitat condition, "Atlantisk høgmyr areal" and "Palsmyr areal", as well as the indicator "Tamrein" which is not relevant when considering natural biodiversity.
+We will demonstrate this here as we would also like to also drop the two indicators that do not represent species abundance but habitat condition, "Atlantisk høgmyr areal" and "Palsmyr areal".
 
@@
 DropIndMode <- "custom"
-DropIndList <- c(7, 142, 155, 436)
+DropIndList <- c(7, 142, 155)
 

### 2023-11-14 trophic_levels-biomass_indicators.rmd
add figure for repport and update methods

@@
 The data were rasterized to the same grid as the NPP data.
+The livestock biomass is adjusted to account for only the time of year that the animals is outside in nature.
 
@@
 
+```{r figure-for-rapport, eval=FALSE, include=F}
+
+ggplot()+geom_sf(data=norcounty,fill="white",lwd=0.1)+
+  geom_spatraster(data=diffcarn[[c(1,12)]])+
+  ggtitle('Difference from expected carnivore biomass')+
+  scale_fill_distiller(type="div",limit=c(-100,100),oob=squish,na.value=NA,"% difference from \nexpected biomass")+
+  facet_wrap(~lyr)+
+  theme_bw()
+ggsave("images/expCarnivoreBiomass.jpg")
+```
+
 ## Part 2 - Habitat specific biomass distributions

### 2023-11-14 Connectivity.Rmd

--- 2023-10-19 DRAFT_Connectivity.Rmd
@@
----
-title: "Template for indicator documentation"
-output:
-  html_document
----
-
-<!-- This is a template for how to document the indicator analyses.
-
-Make a copy of this file and store it in the root folder (same as where this template is stored) folder. 
-Name the file DRAFT_myIndicatorName.Rmd.
-
-Because this is a stand-alone Rmd file inside a bookdown project, in order to knit the file for preview purposes,
-you need to type rmarkdown::render("template.Rmd") in the console, 
-as the the usual Ctrl + Alt + k will not work.
-
-Make sure also to not change the order, or modify the headers, unless you really need to. This is because it
-easier to read the web page as a whole of the indicators are presented using the same layout.
-
-When the analyses is complete, Anders Kolstad, or someone else, will perform a rapid code review, making sure 
-the analysis is repeatable by someone else. Be therefore very careful to annotate 
-and comment on every step in the analysis. When completed, Anders will make a copy
-of the working file and put it in the same project root folder, but removing the 'DRAFT_' part of the name. 
-After modifying it slightly (mainly removing the top YAML header and updating the rmd_files argument in '_bookdown.yml') 
-the output will become available on the website (https://ninanor.github.io/ecosystemCondition/). -->
-
-<!-- Examples:
-
-For an example, see 'indicators/breareal-workingFile.Rmd' 
-and the subsequent 'breareal.Rmd' 
-which again, when rendered, produces this:
-https://ninanor.github.io/ecosystemCondition/areal-av-isbreer.html
-
-Notice the degree of annotation.
-
--->
-
-<!-- Replace 'indicator name' with yor own indicator name  -->
 # Connectivity
@@
 
-October 2023
+November 2023
 
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(flextable)
 ```
@@
 The loss of connectivity (connections or linkages) between suitable habitats affects individuals, populations, and communities through various ecosystem interactions. 
-In landscape ecology, connectivity is essentially defined as "the degree to which the landscape facilitates or impedes movement between patches of resources or suitable habitats." The concept of connectivity encompasses both structural connectivity, which involves the physical locations of suitable and unsuitable habitats or patches being analyzed, and functional connectivity, which focuses on the actual movement of individuals across suitable and unsuitable habitats and disturbance barriers.
-We have assessed that there is currently insufficient data available to start calculating functional connectivity on a national scale in relation to the assessment of ecological condition, as there is almost a total lack of relevant data for almost all species.
+In landscape ecology, connectivity is essentially defined as "the degree to which the landscape facilitates or impedes movement between patches of resources or suitable habitats." 
+The concept of connectivity encompasses both structural connectivity, which involves the physical locations of suitable and unsuitable habitats or patches being analyzed, and functional connectivity, which focuses on the actual movement of individuals across suitable and unsuitable habitats and disturbance barriers.
+We have assessed that there is currently insufficient data available to start calculating functional connectivity on a national scale in relation to the assessment of ecological condition, 
+as there is almost a total lack of relevant data for almost all species. 
 Therefore, we have developed an index for structural connectivity based on physical interventions calculated using an infrastructure index (Erikstad et al.
@@
 
-Here a general workflow for the calculation of the indicator.
+> MORE INFO COMING SOON!
 
-1.	You must load the necessary libraries for geospatial analysis.
-2.	You must load the infrastructure index and wetland data.
-3.	You define the buffer distance (500 meters).
-4.	You must create a function (``calculateMeanDistance``) to calculate the mean distance to infrastructure within a buffer around each wetland polygon.
-5.	Then you filter and keep only the homogeneous wetlands with more than 80% mire coverage.
-6.	You have to apply the ``calculateMeanDistance`` function to these selected wetlands, both with and without infrastructure.
-7.	You export the results as shapefiles for both scenarios.
-8.	The scaled values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure.
+<!-- Here a general workflow for the calculation of the indicator. -->
+
+<!-- 1.	You must load the necessary libraries for geospatial analysis. -->
+<!-- 2.	You must load the infrastructure index and wetland data. -->
+<!-- 3.	You define the buffer distance (500 meters). -->
+<!-- 4.	You must create a function (``calculateMeanDistance``) to calculate the mean distance to infrastructure within a buffer around each wetland polygon. -->
+<!-- 5.	Then you filter and keep only the homogeneous wetlands with more than 80% mire coverage. -->
+<!-- 6.	You have to apply the ``calculateMeanDistance`` function to these selected wetlands, both with and without infrastructure. -->
+<!-- 7.	You export the results as shapefiles for both scenarios. -->
+<!-- 8.	The scaled values for the "Connectivity" indicator are determined by comparing two scenarios: one that includes the presence of infrastructure and one that excludes infrastructure.
 The reference value is calculated as the ratio of the mean distances between myrpolygon areas in these two scenarios.
+-->
 
-Currently step 8, and a subsequent step 9 (to aggregate scaled indicators to regions and simultaneously calculate errors) are not complete.
+<!-- Currently step 8, and a subsequent step 9 (to aggregate scaled indicators to regions and simultaneously calculate errors) are not complete.
 Se [this issue]()https://github.com/NINAnor/ecosystemCondition/issues/144.
+-->
 
-It is also possible to create a time series for this indicator, but this is not yet done. See [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/145).
+<!-- It is also possible to create a time series for this indicator, but this is not yet done. See [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/145). -->
 
-## About the underlying data {#underlying-data-con}
-The **infrastructure index** is one part of the connectivity index for wetlands.
+<!-- ## About the underlying data {#underlying-data-con} -->
+<!-- The **infrastructure index** is one part of the connectivity index for wetlands.
 The infrastructure index was calculated for three time points: 2003, 2013, and 2023.
@@
 The model achieved a balanced accuracy rate of 90.9% when validated against ground-truth samples, significantly improving upon the accuracy rates achieved with manually digitized maps in Norway.
+-->
 
-### Representativity in time and space {#representativity-con}
-The data for the infrastructure index covers the entire country and was collected for three time points over a 20-year period.
+<!-- ### Representativity in time and space {#representativity-con} -->
+<!-- The data for the infrastructure index covers the entire country and was collected for three time points over a 20-year period.
 This provides a broad and representative dataset for assessing changes in connectivity in Norwegian landscapes.
@@
 However, the model used in the mapping process can be applied to other time frames if necessary.
+-->
 
-### Original units {#units-con}
-The infrastructure index has no specific units as it is a measure of landscape fragmentation and disruption. It is expressed as a dimensionless index.
+<!-- ### Original units {#units-con} -->
+<!-- The infrastructure index has no specific units as it is a measure of landscape fragmentation and disruption. It is expressed as a dimensionless index. -->
 
-### Temporal coverage {#temporal-coverage-con}
-The data covers three time points: 2003, 2013, and 2023.
+<!-- ### Temporal coverage {#temporal-coverage-con} -->
+<!-- The data covers three time points: 2003, 2013, and 2023.
 This provides a 20-year temporal perspective for assessing changes in connectivity over time.
 The wetland mapping provides information in the form of pixel values representing wetland presence or absence.
+-->
 
-### Aditional comments about the dataset {#additional-comments-con}
-It is important to note that infrastructure data are based on available map and geographical data sources.
+<!-- ### Aditional comments about the dataset {#additional-comments-con} -->
+<!-- It is important to note that infrastructure data are based on available map and geographical data sources.
 Sometimes, there may be variations and uncertainties in this data that can affect the accuracy of the infrastructure index.
+-->
 
-## Ecosystem characteristic {#egenskap-con}
-In the context of the Norwegian standard, the "Connectivity" indicator aligns with the ecosystem characteristic known as **Landskapsøkologiske mønstre**.
+<!-- ## Ecosystem characteristic {#egenskap-con} -->
+<!-- In the context of the Norwegian standard, the "Connectivity" indicator aligns with the ecosystem characteristic known as **Landskapsøkologiske mønstre**. -->
 
-### SEEA EA (UN standard) {#seea-con}
-In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the _Connectivity_ indicator can be classified as a **C1 Landscape and seascape characteristics** indicator.
+<!-- ### SEEA EA (UN standard) {#seea-con} -->
+<!-- In the framework of the United Nations System of Environmental-Economic Accounting (SEEA EA), the _Connectivity_ indicator can be classified as a **C1 Landscape and seascape characteristics** indicator.
+-->
 
-## Collinearities with other indicators {#collinear-con}
-Connectivity is not thought to exhibit collinearity with any other indicator at the present.
+<!-- ## Collinearities with other indicators {#collinear-con} -->
+<!-- Connectivity is not thought to exhibit collinearity with any other indicator at the present. -->
 
-## Reference condition and values {#ref-conditions-and-values-con}
-### Reference condition {#ref-condition-con}
-The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems.
+<!-- ## Reference condition and values {#ref-conditions-and-values-con} -->
+<!-- ### Reference condition {#ref-condition-con} -->
+<!-- The reference condition for the "Connectivity" indicator is characterized as one with minimal negative human impact on the connectivity of ecosystems.
 In this reference condition, the landscape exhibits natural and undisturbed patterns of connectivity between suitable habitats.
 This state represents the optimal ecological condition for connectivity assessment.
+-->
 
-### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-values-con}
-The reference values for the "Connectivity" indicator are determined by calculating the connectivity or mean distance to mire polygons, when excluding infrastructure from the analysis.
+<!-- ### Reference values, thresholds for defining _good ecological condition_, minimum and/or maximum values {#ref-values-con} -->
+<!-- The reference values for the "Connectivity" indicator are determined by calculating the connectivity or mean distance to mire polygons, when excluding infrastructure from the analysis. -->
 
-In the report «Vurdering av økologisk tilstand for fjell i Norge i 2021» (Framstad et al.
+<!-- In the report «Vurdering av økologisk tilstand for fjell i Norge i 2021» (Framstad et al.
 2021) the value of 0,6 was used as the threshold for good ecological condition.
 We will tentatively use this threshold value again here.
-A perpetual discussion about the reference (and threshold values) for the connectivity indicator is maintained on [GitHub.](https://github.com/NINAnor/ecosystemCondition/issues/143)
+A perpetual discussion about the reference (and threshold values) for the connectivity indicator is maintained on [GitHub.](https://github.com/NINAnor/ecosystemCondition/issues/143) -->
 
-## Uncertainties {#uncertainties-con}
-Our wetland mapping model is a powerful tool for large-scale mapping but does come with certain uncertainties.
+<!-- ## Uncertainties {#uncertainties-con} -->
+<!-- Our wetland mapping model is a powerful tool for large-scale mapping but does come with certain uncertainties.
 While our U-Net model achieved an impressive balanced accuracy rate of 90.9% when validated against an independent ground-truth sample, it's essential to recognize that some level of uncertainty is inherent.
@@
 We made significant efforts to ensure the highest quality, but the inherent variability in ground-truth data may introduce some level of uncertainty.
+-->
 
-The infrastructure data sources used for this analysis might contain inaccuracies, errors, or omissions.
+<!-- The infrastructure data sources used for this analysis might contain inaccuracies, errors, or omissions.
 These can arise from factors like inconsistent reporting and data collection methods.
@@
 This can affect the precision of connectivity assessments in areas with small, localized infrastructure.
+-->
 
-## References
-Bakkestuen, V.; Venter, Z.; Ganerød, A.J.; Framstad, E.
+<!-- ## References -->
+<!-- Bakkestuen, V.; Venter, Z.; Ganerød, A.J.; Framstad, E.
 Delineation of Wetland Areas in South Norway from Sentinel-2 Imagery and LiDAR Using TensorFlow, U-Net, and Google Earth Engine.
@@
 2023, 15, 1203.
-https://doi.org/10.3390/rs15051203
+https://doi.org/10.3390/rs15051203 -->
 
-Bakkestuen, V. et al. in prep. Comprehensive Mapping of Wetland Areas in Norway: An Integration of Sentinel-2 Imagery, LiDAR, and Deep Learning Using TensorFlow and Google Earth Engine
+<!-- Bakkestuen, V. et al. in prep. Comprehensive Mapping of Wetland Areas in Norway: An Integration of Sentinel-2 Imagery, LiDAR, and Deep Learning Using TensorFlow and Google Earth Engine -->
 
-Erikstad, L.; Simensen, T.; Bakkestuen, V.; Halvorsen, R. Index Measuring Land Use Intensity—A Gradient-Based Approach. Geomatics 2023, 3, 188-204. https://doi.org/10.3390/geomatics3010010
+<!-- Erikstad, L.; Simensen, T.; Bakkestuen, V.; Halvorsen, R. Index Measuring Land Use Intensity—A Gradient-Based Approach. Geomatics 2023, 3, 188-204. https://doi.org/10.3390/geomatics3010010 -->
 
-Framstad, E., Eide, N.E., Eide, W., Klanderud, K., Kolstad, A., Töpper, J.
+<!-- Framstad, E., Eide, N.E., Eide, W., Klanderud, K., Kolstad, A., Töpper, J.
 & Vandvik, V.
@@
 Norsk institutt for naturforskning.
+-->
 
-## Analyses {#analyses-con}
-### Data sets {#datasets-con}
-Two main data sets were included.
+<!-- Husk en refERANSE TIL KKLG -->
 
-A. The infrastructure index (Erikstad et al. 2023) and;
+<!-- ## Analyses {#analyses-con} -->
+<!-- ### Data sets {#datasets-con} -->
+<!-- Two main data sets were included. -->
 
-B. the wetland model (Bakkestuen et al. 2023, Bakkestuen et al. _In prep._)
+<!-- A. The infrastructure index (Erikstad et al. 2023) and; -->
 
-#### Data set A - Infrastructure index {#data-a-con}
-The infrastructure index utilized in this analysis was derived from data sources available through Google Earth Engine (GEE).
+<!-- B. the wetland model (Bakkestuen et al. 2023, Bakkestuen et al. _In prep._) -->
+
+<!-- #### Data set A - Infrastructure index {#data-a-con} -->
+<!-- The infrastructure index utilized in this analysis was derived from data sources available through Google Earth Engine (GEE).
 The index quantifies the presence and distribution of key infrastructure elements such as roads, buildings, and related human-made structures across Norway.
@@
 It is important to note that this data forms a fundamental component in the calculation of the connectivity indicator.
+-->
 
-#### Data set B - Wetland model {#data-b-con}
-The wetland model used in this study was developed based on the integration of Sentinel-2 satellite imagery and LiDAR data, utilizing deep learning techniques.
+<!-- #### Data set B - Wetland model {#data-b-con} -->
+<!-- The wetland model used in this study was developed based on the integration of Sentinel-2 satellite imagery and LiDAR data, utilizing deep learning techniques.
 The model provides high-resolution information regarding the distribution and extent of wetland areas within southern Norway, specifically for the year 2020.
@@
 The data set B for South-Norway is already accessible [here](https://kartkatalog.geonorge.no/metadata/aapen-vtmark-i-soer-norge-basert-p-natur-i-norge-nin-typologi/6b5a34f9-c23d-4cca-a51b-01251c217cb1).
+-->
 
-### Scaled indicator values {#scaled-values-con}
-The connectivity indicator was created by examining the interplay between infrastructure and wetland areas.
+<!-- ### Scaled indicator values {#scaled-values-con} -->
+<!-- The connectivity indicator was created by examining the interplay between infrastructure and wetland areas.
 We initially calculated the average distance between wetland patches using the wetland model, representing a baseline measure of natural connectivity in the absence of infrastructure (i.e., the reference condition).
 Subsequently, the infrastructure index was introduced, portraying the extent of human-made structural elements within the landscape.
-The scaled indicator is the quotient of the values.
+-->
+<!-- The scaled indicator is the quotient of the values. -->
 
-### Uncertainty {#uncertainty-con}
-The current state of the indicator is as a geospatial data set with one scaled indicator value per mire polygon, and no errors around this value.
+<!-- ### Uncertainty {#uncertainty-con} -->
+<!-- The current state of the indicator is as a geospatial data set with one scaled indicator value per mire polygon, and no errors around this value.
 Future improvement must include aggregating these indicator values to regions.
 The uncertainties around these aggregated indicator values could then be represented by the spatial variation in the indicator values.
+-->
 
-## Prepare export {#pre-export-con}
-You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.
+<!-- ## Prepare export {#pre-export-con} -->
+<!-- You must run the whole code below. There is a maximum of 5000 wetland polygons for each export. -->
 
-### JAVA code {#java-con}
+<!-- ### JAVA code {#java-con} -->
 
-Koden er utviklet i JAVA for Google Earth Engine:
+<!-- Koden er utviklet i JAVA for Google Earth Engine: -->
 
-```{}
-// Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå
-Map.setCenter(11.0534047, 62.14137, 12);
+<!-- ```{} -->
+<!-- // Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå -->
+<!-- Map.setCenter(11.0534047, 62.14137, 12); -->
 
-// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine Data
-var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND');
-var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
-var myr = myrpred2.mean();
+<!-- // Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine Data -->
+<!-- var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND'); -->
+<!-- var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153'); -->
+<!-- var myr = myrpred2.mean(); -->
 
-// Legg til myrdata som et lag på kartet med tilpassede visningsparametere
-Map.addLayer(myr, { min: 0, max: 1, palette: ['white', 'green'] }, 'myr');
-Map.addLayer(infrastruktur, { min: 0, max: 13, palette: ['blue', 'yellow', 'red'] }, 'infrastruktur');
+<!-- // Legg til myrdata som et lag på kartet med tilpassede visningsparametere -->
+<!-- Map.addLayer(myr, { min: 0, max: 1, palette: ['white', 'green'] }, 'myr'); -->
+<!-- Map.addLayer(infrastruktur, { min: 0, max: 13, palette: ['blue', 'yellow', 'red'] }, 'infrastruktur'); -->
 
-// Definer en bufferavstand (i meter) som bestemmer området som blir analysert rundt hvert myrpolygon
-var bufferDistance = 500;
+<!-- // Definer en bufferavstand (i meter) som bestemmer området som blir analysert rundt hvert myrpolygon -->
+<!-- var bufferDistance = 500; -->
 
-// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur basert på grenseverdien
-var calculateMeanDistance = function (feature) {
-  var threshold = 3;  // Definer din grenseverdi for infrastruktur
+<!-- // Funksjon for å beregne gjennomsnittlig avstand til infrastruktur basert på grenseverdien -->
+<!-- var calculateMeanDistance = function (feature) { -->
+<!--   var threshold = 3;  // Definer din grenseverdi for infrastruktur -->
 
-  // Filtrer infrastrukturbildet basert på grenseverdien
-  var filteredInfrastructure = infrastruktur.gt(threshold);
+<!--   // Filtrer infrastrukturbildet basert på grenseverdien -->
+<!--   var filteredInfrastructure = infrastruktur.gt(threshold); -->
 
-  // Hent myrpolygonets sentrum
-  var myrCenter = feature.geometry().centroid(100);
+<!--   // Hent myrpolygonets sentrum -->
+<!--   var myrCenter = feature.geometry().centroid(100); -->
 
-  // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon
-  var distanceWithInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
-    reducer: ee.Reducer.mean(),
-    geometry: feature.geometry().buffer(bufferDistance),
-    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
-    maxPixels: 1e13
-  }).get('filtered_infra');
+<!--   // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon -->
+<!--   var distanceWithInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({ -->
+<!--     reducer: ee.Reducer.mean(), -->
+<!--     geometry: feature.geometry().buffer(bufferDistance), -->
+<!--     scale: 10,  // Oppløsningen til datasettet (meters per piksel) -->
+<!--     maxPixels: 1e13 -->
+<!--   }).get('filtered_infra'); -->
 
-  var distanceWithoutInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
-    reducer: ee.Reducer.mean(),
-    geometry: myrCenter.buffer(bufferDistance),
-    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
-    maxPixels: 1e13
-  }).get('filtered_infra');
+<!--   var distanceWithoutInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({ -->
+<!--     reducer: ee.Reducer.mean(), -->
+<!--     geometry: myrCenter.buffer(bufferDistance), -->
+<!--     scale: 10,  // Oppløsningen til datasettet (meters per piksel) -->
+<!--     maxPixels: 1e13 -->
+<!--   }).get('filtered_infra'); -->
 
-  // Skalér brøkverdien til meter ved å multiplisere med bufferavstanden
-  var scaledDistanceWithInfrastructure = ee.Number(distanceWithInfrastructure).multiply(bufferDistance);
-  var scaledDistanceWithoutInfrastructure = ee.Number(distanceWithoutInfrastructure).multiply(bufferDistance);
+<!--   // Skalér brøkverdien til meter ved å multiplisere med bufferavstanden -->
+<!--   var scaledDistanceWithInfrastructure = ee.Number(distanceWithInfrastructure).multiply(bufferDistance); -->
+<!--   var scaledDistanceWithoutInfrastructure = ee.Number(distanceWithoutInfrastructure).multiply(bufferDistance); -->
 
-  // Beregn kvotienten (forholdet mellom avstand uten infrastruktur og avstand med infrastruktur)
-  var kvotient = scaledDistanceWithoutInfrastructure.divide(scaledDistanceWithInfrastructure);
+<!--   // Beregn kvotienten (forholdet mellom avstand uten infrastruktur og avstand med infrastruktur) -->
+<!--   var kvotient = scaledDistanceWithoutInfrastructure.divide(scaledDistanceWithInfrastructure); -->
 
-  var properties = {
-    'distinfra_meters': scaledDistanceWithInfrastructure,
-    'distnoinfra_meters': scaledDistanceWithoutInfrastructure,
-    'kvotient': kvotient
-  };
+<!--   var properties = { -->
+<!--     'distinfra_meters': scaledDistanceWithInfrastructure, -->
+<!--     'distnoinfra_meters': scaledDistanceWithoutInfrastructure, -->
+<!--     'kvotient': kvotient -->
+<!--   }; -->
 
-  return feature.set(properties);
-};
+<!--   return feature.set(properties); -->
+<!-- }; -->
 
-// Velg områder med myr (myrinnhold > 80%) og bruk dette som geometri for analysen
-var omraderMedMyr = myr.gt(0.8).selfMask();
+<!-- // Velg områder med myr (myrinnhold > 80%) og bruk dette som geometri for analysen -->
+<!-- var omraderMedMyr = myr.gt(0.8).selfMask(); -->
 
-// Hent geometriene for hvert område med myr
-var omraderGeometries = omraderMedMyr.reduceToVectors({
-  geometry: geometry,
-  scale: 10,  // Oppløsningen til datasettet (meters per piksel)
-  maxPixels: 1e13
-});
+<!-- // Hent geometriene for hvert område med myr -->
+<!-- var omraderGeometries = omraderMedMyr.reduceToVectors({ -->
+<!--   geometry: geometry, -->
+<!--   scale: 10,  // Oppløsningen til datasettet (meters per piksel) -->
+<!--   maxPixels: 1e13 -->
+<!-- }); -->
 
-// Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr
-var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance);
+<!-- // Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr -->
+<!-- var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance); -->
 
-// Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet
-Map.addLayer(omraderMedMyr, { min: 0, max: 1, palette: ['white', 'green'] }, 'Myrpolygoner');
+<!-- // Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet -->
+<!-- Map.addLayer(omraderMedMyr, { min: 0, max: 1, palette: ['white', 'green'] }, 'Myrpolygoner'); -->
 
-// Vis scenariet "Konnektivitet med infrastruktur" på kartet
-var visParams = {
-  min: 0,
-  max: 1, // Juster dette basert på din infrastrukturindeksens min og maks
-  palette: ['blue', 'yellow', 'red'],
-};
-Map.addLayer(omraderMedGjennomsnittligAvstand, visParams, 'Gjennomsnittlig avstand til infrastruktur');
+<!-- // Vis scenariet "Konnektivitet med infrastruktur" på kartet -->
+<!-- var visParams = { -->
+<!--   min: 0, -->
+<!--   max: 1, // Juster dette basert på din infrastrukturindeksens min og maks -->
+<!--   palette: ['blue', 'yellow', 'red'], -->
+<!-- }; -->
+<!-- Map.addLayer(omraderMedGjennomsnittligAvstand, visParams, 'Gjennomsnittlig avstand til infrastruktur'); -->
 
-// Eksporter resultatene som en Shapefile til Google Drive
-Export.table.toDrive({
-  collection: omraderMedGjennomsnittligAvstand,
-  description: 'Konnektivitet_med_infrastruktur',
-  fileFormat: 'SHP',
-});
+<!-- // Eksporter resultatene som en Shapefile til Google Drive -->
+<!-- Export.table.toDrive({ -->
+<!--   collection: omraderMedGjennomsnittligAvstand, -->
+<!--   description: 'Konnektivitet_med_infrastruktur', -->
+<!--   fileFormat: 'SHP', -->
+<!-- }); -->
 
-```
+<!-- ``` -->
 
-The scaled "Connectivity" values are created by dividing the result without infrastructure by the result that includes infrastructure.
+<!-- The scaled "Connectivity" values are created by dividing the result without infrastructure by the result that includes infrastructure.
 This ratio will give you a value between 0 and 1, where 0 represents complete disruption of connectivity due to infrastructure, and 1 represents undisturbed natural connectivity.
+-->
 
-This step is not yet done, but in R this is how one would do it.
-Calculate the reference value for connectivity
-```{r, eval=FALSE}
-reference_value <- selected_wetlands_without_infrastructure$mean_distance / selected_wetlands_with_infrastructure$mean_distance
-```
+<!-- This step is not yet done, but in R this is how one would do it. -->
+<!-- Calculate the reference value for connectivity -->
+<!-- ```{r, eval=FALSE} -->
+<!-- reference_value <- selected_wetlands_without_infrastructure$mean_distance / selected_wetlands_with_infrastructure$mean_distance -->
+<!-- ``` -->

### 2023-12-01 Connectivity.Rmd
new text and code

--- 2023-10-19 DRAFT_Connectivity.Rmd
@@
 knitr::opts_chunk$set(echo = TRUE)
+library(flextable)
+library(tidyverse)
 ```
@@
 4.	You must create a function (``calculateMeanDistance``) to calculate the mean distance to infrastructure within a buffer around each wetland polygon.
-5.	Then you filter and keep only the homogeneous wetlands with more than 80% mire coverage.
-6.	You have to apply the ``calculateMeanDistance`` function to these selected wetlands, both with and without infrastructure.
+5.	Then you filter and keep only the homogeneous wetlands with more than 80% probability in the AI model for mire coverage.
+6.	You have to apply the ``calculateMinimumnDistance`` function to these selected wetlands, both with and without infrastructure.
 7.	You export the results as shapefiles for both scenarios.
@@
 It is also possible to create a time series for this indicator, but this is not yet done. See [GitHub issue](https://github.com/NINAnor/ecosystemCondition/issues/145).
+
+```{r concon, fig.cap="Conceptual model for the calculation of the conectivity indicator.
+Two distances are calculated: one from the centre point of one mire polygon to the closest neighbouring polygon (green arrow = the reference value), and one from the same centre point to the closest infrastructure (red arrow  = area with infrastructure index value above 3).
+The scaled indicator is the ratio, or quotient, of the length of the red arrow divided by the length of the green arrow."}
+knitr::include_graphics("images/conectivity-concept.png")
+```
 
@@
 
-B. the wetland model (Bakkestuen et al. 2023, Bakkestuen et al. _In prep._)
+B. The wetland model (Bakkestuen et al. 2023, Bakkestuen et al. _In prep._)
 
@@
 ## Prepare export {#pre-export-con}
-You must run the whole code below. There is a maximum of 5000 wetland polygons for each export.
+You must run the whole code below. There is an approximately maximum of 50000 wetland polygons or 60-70 MB shapefile for each export. The whole of Norway is exported in 25 tiles.
 
@@
 ```{}
-// Senter kartet til en bestemt geografisk posisjon (bredde- og lengdegrad) og velg en zoomnivå
-Map.setCenter(11.0534047, 62.14137, 12);
+// Laste inn myr datasett fra Google Earth Engine Data
+var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr168NN');
+var myr = myrpred2.min();
 
-// Laste inn infrastrukturindeks og myr datasett fra Google Earth Engine Data
+// Laste inn infrastrukturindeks fra Google Earth Engine Data
 var infrastruktur = ee.Image('users/vegar/NY_INFRA_IND');
-var myrpred2 = ee.ImageCollection('users/vegarbakkestuen/Myr153');
-var myr = myrpred2.mean();
 
-// Legg til myrdata som et lag på kartet med tilpassede visningsparametere
-Map.addLayer(myr, { min: 0, max: 1, palette: ['white', 'green'] }, 'myr');
-Map.addLayer(infrastruktur, { min: 0, max: 13, palette: ['blue', 'yellow', 'red'] }, 'infrastruktur');
+// Konverter infrastruktur til en vektorbasert maske med terskelverdi 2
+var infrastrukturMask = infrastruktur.gt(2);
+
+// Masker infrastrukturen med myrpolygonene
+var maskedInfra = infrastrukturMask.mask(infrastrukturMask);
 
 // Definer en bufferavstand (i meter) som bestemmer området som blir analysert rundt hvert myrpolygon
-var bufferDistance = 500;
+var bufferDistance = 1000;
 
-// Funksjon for å beregne gjennomsnittlig avstand til infrastruktur basert på grenseverdien
-var calculateMeanDistance = function (feature) {
-  var threshold = 3;  // Definer din grenseverdi for infrastruktur
+// Funksjon for å beregne minimum avstand mellom myrer og infrastruktur innenfor samme datasett
+var calculateMinimumDistance = function (feature) {
+  // Beregn avstanden til nærmeste nabo-myrene ved å bruke map-funksjonen
+  var minDistanceToMyr = ee.FeatureCollection(omraderMedMyrFC)
+    .filter(ee.Filter.neq('system:index', feature.get('system:index')))
+    .map(function (neighbor) {
+      var distanceToMyr = feature.geometry().distance(neighbor.geometry(), bufferDistance);
+      return ee.Feature(null, {'distance_to_myr': distanceToMyr});
+    });
 
-  // Filtrer infrastrukturbildet basert på grenseverdien
-  var filteredInfrastructure = infrastruktur.gt(threshold);
+  // Finn den minimale avstanden til myr ved å sortere samlingen
+  minDistanceToMyr = minDistanceToMyr.sort('distance_to_myr');
 
-  // Hent myrpolygonets sentrum
-  var myrCenter = feature.geometry().centroid(100);
+  // Hent den minimale avstanden til myr fra den første funksjonen i samlingen
+  var minDistanceToMyrValue = ee.Number(minDistanceToMyr.first().get('distance_to_myr'));
 
-  // Beregn gjennomsnittlig avstand til infrastruktur ved å redusere regionen rundt hvert myrpolygon
-  var distanceWithInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
-    reducer: ee.Reducer.mean(),
-    geometry: feature.geometry().buffer(bufferDistance),
+  // Konverter infrastrukturen til en vektorbasert maske
+  var infraVector = maskedInfra.reduceToVectors({
+    geometry: feature.geometry().buffer(bufferDistance), // Bruk buffer rundt myrpolygonet
     scale: 10,  // Oppløsningen til datasettet (meters per piksel)
     maxPixels: 1e13
-  }).get('filtered_infra');
+  });
 
-  var distanceWithoutInfrastructure = filteredInfrastructure.rename('filtered_infra').reduceRegion({
-    reducer: ee.Reducer.mean(),
-    geometry: myrCenter.buffer(bufferDistance),
-    scale: 10,  // Oppløsningen til datasettet (meters per piksel)
-    maxPixels: 1e13
-  }).get('filtered_infra');
+  // Beregn avstanden til infrastruktur ved å finne avstanden til nærmeste nabo i infrastrukturen
+  var distanceToInfra = feature.geometry().distance(infraVector.geometry(), bufferDistance);
 
-  // Skalér brøkverdien til meter ved å multiplisere med bufferavstanden
-  var scaledDistanceWithInfrastructure = ee.Number(distanceWithInfrastructure).multiply(bufferDistance);
-  var scaledDistanceWithoutInfrastructure = ee.Number(distanceWithoutInfrastructure).multiply(bufferDistance);
+  // Legg til avstandsverdiene som egenskaper
+  feature = feature.set({
+    'min_myr_distance': minDistanceToMyrValue,
+    'min_infra_distance': distanceToInfra,
+    'kvotient': distanceToInfra.divide(minDistanceToMyrValue)
+  });
 
-  // Beregn kvotienten (forholdet mellom avstand uten infrastruktur og avstand med infrastruktur)
-  var kvotient = scaledDistanceWithoutInfrastructure.divide(scaledDistanceWithInfrastructure);
-
-  var properties = {
-    'distinfra_meters': scaledDistanceWithInfrastructure,
-    'distnoinfra_meters': scaledDistanceWithoutInfrastructure,
-    'kvotient': kvotient
-  };
-
-  return feature.set(properties);
+  return feature;
 };
@@
 // Hent geometriene for hvert område med myr
-var omraderGeometries = omraderMedMyr.reduceToVectors({
-  geometry: geometry,
+var omraderMedMyrFC = omraderMedMyr.reduceToVectors({
+  geometry: geometry, // Bruk den definerte geometrien
   scale: 10,  // Oppløsningen til datasettet (meters per piksel)
@@
 
-// Bruk funksjonen calculateMeanDistance for å beregne gjennomsnittlig avstand til infrastruktur for hvert område med myr
-var omraderMedGjennomsnittligAvstand = omraderGeometries.map(calculateMeanDistance);
+// Bruk funksjonen calculateMinimumDistance for å beregne minimum avstand til nærmeste nabo-myr og infrastruktur
+var omraderMedMinimumDistance = omraderMedMyrFC.map(calculateMinimumDistance);
 
-// Legg til lag som viser myrpolygonene og lag med gjennomsnittlig avstand til infrastruktur på kartet
+// Legg til lag som viser myrpolygonene og lag med minimum avstand til nærmeste nabo-myr og infrastruktur på kartet
 Map.addLayer(omraderMedMyr, { min: 0, max: 1, palette: ['white', 'green'] }, 'Myrpolygoner');
-
-// Vis scenariet "Konnektivitet med infrastruktur" på kartet
-var visParams = {
-  min: 0,
-  max: 1, // Juster dette basert på din infrastrukturindeksens min og maks
-  palette: ['blue', 'yellow', 'red'],
-};
-Map.addLayer(omraderMedGjennomsnittligAvstand, visParams, 'Gjennomsnittlig avstand til infrastruktur');
+Map.addLayer(infrastruktur, { min: 0, max: 1, palette: ['white', 'blue'] }, 'infra');
+Map.addLayer(omraderMedMinimumDistance, { min: 0, max: 500, palette: ['blue', 'yellow', 'red'] }, 'Minimum avstand til nærmeste nabo-myr');
 
@@
 Export.table.toDrive({
-  collection: omraderMedGjennomsnittligAvstand,
-  description: 'Konnektivitet_med_infrastruktur',
+  collection: omraderMedMinimumDistance,
+  description: 'Minimum_avstand_til_myr_infrastruktur',
   fileFormat: 'SHP',
@@
 This ratio will give you a value between 0 and 1, where 0 represents complete disruption of connectivity due to infrastructure, and 1 represents undisturbed natural connectivity.
-
-This step is not yet done, but in R this is how one would do it.
-Calculate the reference value for connectivity
-```{r, eval=FALSE}
-reference_value <- selected_wetlands_without_infrastructure$mean_distance / selected_wetlands_with_infrastructure$mean_distance
-```

### 2023-12-04 insectIndicators.Rmd
updated to lastest dev

@@
 <br />
-
-<!--# Notes on this draft
-Most things are done but some things needs updates or touching up.
-
-Incomplete list of things that will change:
-
-* Upgrade genetic diversity calculation. The current calculation is just a placeholder, awaiting some new data from Marie. Waiting for software installation by IT. Unsure of timeline.
-* Calculate NorIns indicators on 5 year interval. (Moving window?)
-* Fix remaining comments in this draft inside hard brackets.-->
 
@@
 
-<!-- Document you work below.
-Try not to change  the headers too much.
-Data can be stored on NINA server.
-Since the book is rendered on the R Server this works fine, but note that directory paths are different on the server compared to you local machine.
-If it is not too big you may store under /data/ on this repository -->
-
 ## Introduction
@@
 This workflow is shortened by putting most of the underlying code in separate R-packages, which are freely available through Github.
-The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators (**Not public yet**), and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.
+The relevant packages are `bombLepiSurv` https://github.com/jenast/bombLepiSurv for the bumblebee and butterfly indicators, and `Norimon` https://github.com/jenast/Norimon for the rest of the indicators.
 
@@
 indDataset <- c(rep("NMBB", 2), 
-             rep("NorIns", 6))
+             rep("NorIns", 7))
 
@@
                 "Quotent of Apocrita vs Symphyta species richness",
+                "Number of alien species",
                 "Intraspesific genetic variation")
@@
 nor_cat <- c("Funksjonelt viktige arter og biofysiske stukturer",
-             "Funksjonelt viktige arter og biofysiske stukturer",
+             "Biologisk mangfold",
              "Funksjonelt viktige arter og biofysiske stukturer",
@@
              "Funksjonelt viktige arter og biofysiske strukturer",
+             "Biologisk mangfold",
              "Biologisk mangfold"
@@
 
-ect_cat <- c("B1",
+ect_cat <- c("B2",
              "B1",
@@
              "B2",
+             "B1",
              "B1"
@@
 Geo_scope <- c(rep("Sørlandet, Østfold-Vestfold, Vestlandet, Trøndelag", 2), 
-               rep("Sørlandet, Østlandet, Trøndelag, Nord-Norge", 6))
+               rep("Sørlandet, Østlandet, Trøndelag, Nord-Norge", 7))
 
 Ind_data <- c(rep("Present", 2),
-              rep("Present", 6))
+              rep("Present", 7))
 
 Ref_val   <- c(rep("Present", 2), 
-               rep("Lacking", 6))
+               rep("Lacking", 7))
 ```
@@
 A relative decrease in predatory/parasitic wasps could indicate a degradation in insect prey communities, and a decrease in the potential for natural pest control.
+This indicator is optionally given as just the species richness of Apocrita.
+**Number of alien species** takes into account the number of alien species found in a region and optionally, the frequency of sampling events they are observed in.
+This is based on the alien species list from the Norwegian Biodiversity Information Center, which lists known alien species and an increasing list of door-knocker species.
+Note that due to large historical knowledge gaps of insect species distributions, there could potentially be many more alien species in Norway, which are not picked up on this alien species list.
 Lastly, **Intraspecific genetic variation** captures biodiversity that until now has been hidden, due to lack of practical technologies to measure it.
-The need to conserve genetic diversity is identified in Norwegian law, and decreased intraspecific diversity could signal population decreases, and decreased potential to adapt to environmental changes, before whole species are lost to an area.
+The need to conserve genetic diversity is identified in Norwegian law, and decreases in intraspecific diversity could signal population decreases, and a decreased potential to adapt to environmental changes, before whole species are lost to an area.
 
@@
 The data used here comes from the ongoing monitoring programs "Norwegian insect monitoring" ([NorIns](https://www.nina.no/Naturmangfold/Insekter/Overv%C3%A5king-av-insekter)) and "Norwegian monitoring of bumblebees and butterflies" ([NMBB](https://www.nina.no/V%C3%A5re-fagomr%C3%A5der/Milj%C3%B8overv%C3%A5king-p%C3%A5-land/Humler-og-dagsommerfugler)).
-Both programs are initiated and financed by the Environmental agency, with the aim to produce continuous, areal representative, time-series of insect community data.
+Both programs are initiated and financed by the Environmental agency, with the aim to produce continuous area representative time-series of insect community data.
 The **NorIns** program use malaise and window traps to sample a broad community of insect species, which are identified through metabarcoding of DNA.
@@
 
-These 5 regions represents the planned spatial precision of the survey.
+The monitoring program is designed to produce estimates with a spatial resolution of these 5 large regions.
 Finer resolution could be evaluated in the future, although it is unlikely to be feasible without more intense sampling.
-This program uses a staggered sampling scheme where we visit different locations each year, and return to each location every 5 years.
+The program uses a staggered sampling scheme where we visit different locations each year, and return to each location every 5 years.
 This sampling scheme, coupled with the strong yearly variation in insect abundances makes it reasonable to use a temporal resolution of 5 years.
-[How to deal with this now?
-average over 5 years?
-Add 5 year aggregation in bootstrap_value()]
+The indicator values from **NorIns** are therefore calculated on 5 year long rolling windows.
 
@@
 Both programs are designed to be area representative, within their respective habitat types and regions.
-Both use semi-randomized selection schemes of localities, the **NorIns** selects by random draws, and the **NMBB** by a predefined continuous network of survey squares.
-In both programs however, some subjective decisions are made to define habitat criteria and to discard unpractical survey locations.
+Both use stratified randomized selection of localities.
+The **NorIns** program selects localities by random draws within habitat strata, and the **NMBB** by a predefined continuous network of survey squares, within which individual transects are then manually placed.
+In both programs, some subjective decisions are made in the definition of habitat criteria and in discarding unpractical survey locations.
 
@@
 ### Norwegian standard
-Four of the indicators are tied to the *Økologisk egenskap* (Ecological characteristic) of **Funksjonelt viktige arter og biofysiske stukturer** (Funktionally important species and structures), and the other four to **Biologisk mangfold** (Biodiversity).
+Three of the indicators are tied to the *Økologisk egenskap* (Ecological characteristic) of **Funksjonelt viktige arter og biofysiske stukturer** (Funktionally important species and structures), and the other six to **Biologisk mangfold** (Biodiversity).
 
@@
 ## Collinearities with other indicators
-None measured so far, but it is reasonable to assume that some of these indicators will share some causal relationships with each other, and therefore correlate to some degree.
-They might also correlate to other indicators, not described here, as well.
+None measured so far, but it is reasonable to assume that some of these indicators will share some causal factors, and therefore correlate to some degree.
+They might also correlate to other indicators as well, not described here.
 
@@
 ### Reference condition
-Assessing the current state of insect communities may be a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability.
-Still, even more challenging is to establishing reference values, when the reference states are extinct or prohibitively difficult to measure empirically.
-We currently lack reference values for all indicators from the **NorIns** program.
-It is unlikely that this can be solved "simply" by surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely no longer exist.
+Assessing the current state of insect communities is a complicated task, due to their taxonomic and functional breadth and high temporal and spatial variability.
+Still, even more challenging is establishing reference values, when the reference states are extinct or prohibitively difficult to measure empirically.
+We currently lack reference values for all indicators from the **NorIns** program, except for the alien species indicators where the reference value is zero by definition.
+It is unlikely that this can be solved by "simply" surveying a state in a reference area, since intact reference areas of sufficient size and numbers likely no longer exist.
 The question of how to handle these reference values is currently unresolved, and we present these indicator value calculations with made up reference values as placeholders in the code.
 
-For the **NMBB** indicators, the communities are small enough and well known enough to identify reference communities by expert opinion.
+For the **NMBB** indicators, the communities are small enough and well-known enough to identify reference communities by expert opinion.
 These communities specify the expected rate of observing each species.
@@
 The specific reference communities for the **NMBB** indicators won't be spelled out here, see instead the scripts below.
-We have not specified a value for _good ecological condition_ as of yet, since there lack a straightforward empirical basis to do so.
+We have not specified a value for _good ecological condition_ as of yet, since we lack a straightforward empirical basis to do so.
+By default then, the values is set to 0.6 , following the general framework.
 
 ## Uncertainties
-The uncertainties for the indicator values is calculated by bootstrapping the average values for each locality within a year.
+The uncertainties for the indicator values is calculated by bootstrapping the values for each locality within a year.
 This takes the variability between the localities into account.
-For the **NorIns** indicators, this uncertainties are given as standard deviation, as well as confidence intervals.The uncertainties for the **NMBB** indicators are constituted by a distribution of discrete values, and therefore we only provide confidence intervals and not a standard deviation.
+For the **NorIns** indicators, these uncertainties are given as standard deviation, as well as confidence intervals.
+The uncertainties for the **NMBB** indicators are constituted by a distribution of discrete values, and therefore we only provide confidence intervals and not a standard deviation.
 
 ## References
+
+* French, C.
+M., Bertola, L.
+D., Carnaval, A.
+C., Economo, E.
+P., Kass, J.
+M., Lohman, D.
+J., … Hickerson, M.
+J.
+2023, juli 17.
+Global determinants of insect mitochondrial genetic diversity.
+bioRxiv.
+bioRxiv.
+https://doi.org/10.1101/2022.02.09.479762
+
+* Miraldo, A., Li, S., Borregaard, M.
+K., Flórez-Rodríguez, A., Gopalakrishnan, S., Rizvanovic, M., … Nogués-Bravo, D.
+2016.
+An Anthropocene map of genetic diversity.
+Science 353(6307): 1532–1535.
+https://doi.org/10.1126/science.aaf4381
+
+* Rohde, K. 1992. Latitudinal Gradients in Species Diversity: The Search for the Primary Cause. Oikos 65(3): 514. https://doi.org/10.2307/3545569
+
+* Theodoridis, S., Fordham, D.
+A., Brown, S.
+C., Li, S., Rahbek, C., & Nogues-Bravo, D.
+2020.
+Evolutionary history and past climate change shape the distribution of genetic diversity in terrestrial mammals.
+Nature Communications 11(1): 2557.
+https://doi.org/10.1038/s41467-020-16449-5
+
 * Åström, Jens; Birkemoe, Tone; Brandsegg, Hege; Dahle, Sondre; Davey, Marie Louise; Ekrem, Torbjørn; Fossøy, Frode; Hanssen, Oddvar; Laugsand, Arne Endre; Majaneva, Markus; Staverløkk, Arnstein; Sverdrup-Thygeson, Anne; Ødegaard, Frode.
@@
 
-* Connor M French, Laura D Bertola, Ana C Carnaval, Evan P Economo, Jamie M Kass, David J Lohman, Katharine A Marske, Rudolf Meier, Isaac Overcast, Andrew J.
-Rominger, Phillip Staniczenko, Michael J Hickerson.
-2023.
-Global determinants of insect mitochondrial genetic diversity.
-https://doi.org/10.1101/2022.02.09.479762
-
 * Öberg, S., Gjershaug, J. O., Diserud, O. & Ødegaard, F. 2011. Videreutvikling av metodikk for 
@@
 The goal is to choose a set of taxa that represent specific qualities of the community that can indicate the ecological status of the ecosystem.
-In addition to the selection of taxa, we have to decide on the level of aggregation, e.g.
-if we should aggregate the data on the region scale, locality scale, or even the individual sample occasion scale.
-For ecological indicators, we will aggregate the values to a year locality scale, where the variables summarises all catches in a locality in a year.[NOTE, implement 5 year/region aggregation]
+In addition to the selection of taxa, we have to decide on the level of spatial and temporal aggregation, e.g.
+if we should aggregate the data spatially on the region scale, habitat scale or locality scale, and if we should aggregate it temporally on yearly or even the individual sample occasions within years.
+For the ecological indicators, we will aggregate the raw observation values to the year locality scale, where the data summarize all catches in a locality in a year.
 
 The second step is to get an estimate of the uncertainty of the data, were we use the bootstrap method.
-This method is flexible, as we could bootstrap the samples on different sampling levels (samples within localities, regions within country).
-For the insect indicators, we will bootstrap and summarize the variables to regional values, for example expressing the mean values of insect biomass in Trøndelag in 2021, with uncertainty.
+This method is also flexible, as we could bootstrap the samples on different sampling levels (e.g.
+samples within localities, localities within a region, or regions within country).
+When working with raw data on the year locality level, the most granular bootstrap would be to summarize the variables to the yearly, habitat, and regional level, for example expressing the mean values of insect biomass in semi-natural land and forests in Østlandet in 2021, with uncertainty.
 
@@
 Taken together, this framework is meant to facilitate the calculation of an arbitrary set of insect indicators, based on the combination of choices in data to fetch, aggregation level, and reference comparison.
-We can work through an example with pollinators, to showcase the framework.
-
-In the example, we start with connecting to the database in order to fetch the data.
+We showcase the framework by working through an example with pollinators.
+
+In the example we start with connecting to the database, in order to fetch the data.
 
@@
 ```
-
+<br/>
 We then fetch the community data for these families through the `get_observations` function.
@@
 "Shannon diversity" doesn't really make sense without counts or amounts, etc.
-This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (Connor et al.
+This will be replaced by measurements of genetic diversity that are based on published peer-reviewed methodology (French et al.
 2023), in the (hopefully near) future.
@@
 Individually, they represent a single measurement of an insect community in a habitat type in a region.
-To get the sampling uncertainty for this representation, we can bootstrap the values by this simple process: we choose a random set of localities within a year and region (with replacement) and calculate the average values.
+To get the sampling uncertainty for this representation, we can bootstrap the values by this simple process: we choose a random set of localities within a year and region (with replacement) and calculate the average value.
 We then repeat this process a large number of times to get a bootstrap sample of values, which can be used to express the uncertainty in the dataset.
@@
 But bootstrapping works fairly well on small samples and has the advantage that is doesn't make assumptions of the statistical distribution of the errors.
-Note that once we get a larger dataset with repeating surveys every 5 years, aggregating and bootstrapping over a whole 5 year period.
+We will use this level to showcase the functionality and to visualize some variation between years.
+However, since the sampling scheme of the monitoring program semi-randomly selects 10 new localities for each of the 5 years the rotating scheme, each years estimate will include some randomness that is caused by variability between localities.
+This means that the values between individual years might appear more random than the underlying overall trend.
+To get robust estimates that averages over this variability, we will summarize the actual indicator values to 5 year periods, further described below.
 
@@
                                       groups = c("year",
-                                                 "region_name")
+                                                 "region_name"),
+                                      rolling_year_window = FALSE
                                       )
@@
 Such subtractions should be made on the set of bootstrapped values, followed by new summary statistics being calculated, preserving the uncertainty from the bootstrap.
-The `boot_stat` has its own subtraction method `-` do to just that.
+The `boot_stat` object class has its own subtraction method `-` do to just that.
 In this example, we set the reference value arbitrarily to 30 species.
@@
 A `boot_stat` class has a `/` function that divides each bootstrap value by a given value, truncates the highest values to 1, and recalculates the summary values.
-We can also simply divide by the reference state, to compare the values to the reference state, and normalize them in one go.
+We can similarly divide by a predefined reference state.
 Here, we exemplify the method by dividing the value by 30.
@@
 
-More plotting options are available, see the `Norimon` indicator workflow vignette.
+#### Time-series plots
+The default plots are good for showing the uncertainty distributions, but can be a bit difficult to track the time series trends this way.
+We can instad use the `ts_plot` function for a simple time-series plot.
+
+```{r}
+ts_plot(poll_richness_boot)
+```
 
 #### Map plots
-In addition to the density plots above, we can also display the values geographically. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.
+In addition to these plots, we can also display the values geographically. The `map_plot()` function takes a `boot_stat` object and plots the values according to its region names.
 
@@
 ```
-
 We can also choose a different palette, for example from the NinaR package, and visualize the uncertainty by setting the transparency of the colors from the bootstrap standard deviations.
-By the way, most of these functions can be piped as well:
+Most of these functions can be piped as well:
 
@@
 
+More plotting options are available, see the `Norimon` indicator workflow vignette.
+
+#### 5 year rolling window summaries
+As described above, to take the staggered 5 year survey scheme into account and to average over random yearly variation, it makes sense to summarize the state over a whole 5 year period.
+This does not need to adhere to predefined periods, since any 5 year long period will result in a complete data set.
+We can therefore calculate 5 year long summaries in rolling windows, where each focal year is surrounded by +- 2 years.
+The bootstrap_value does this by default, if we don't specify `rolling_year_average = FALSE`
+
+```{r}
+poll_richness_boot_5_year <- bootstrap_value(poll_loc_year,
+                                      value = no_species,
+                                      groups = c("year",
+                                                 "region_name"),
+                                      rolling_year_window = TRUE
+                                      )
+```
+Since we are only on the third year in the survey scheme however, the rolling windows are too large to show any differences between years. This will change, starting with the 2023 year data.
+
+```{r}
+plot(poll_richness_boot_5_year)
+```
+
+The averaging over 5-years also shows up in the time-series plots. Note the very wide confidence intervals of the Sørlandet region, with only 1-year worth of data.
+
+```{r}
+ts_plot(poll_richness_boot_5_year)
+```
+
 This concludes the tour of the `Norimon` functionality.
@@
 $$CI = \frac{SR-SC}{SR}$$
-The state of reference (SR) represents a community of species that can be expected to be observed in a given habitat type and region.
+The state of reference (SR) represents a community of species which can be expected to be observed in a given habitat type and region.
 It is calculated by assigning each species to class of expected commonality: common (C), sporadic (S) and rare (R) species.
@@
 where $n_{CS}$ is the number of expected common species (C) that is observed sporadically (S), and $w_{CS}$ is the weight of this change.
-Similarly, $n_{CR}$ is the number of expected species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on.
+Similarly, $n_{CR}$ is the number of expected common species that is observed rarely, $n_{RL}$ is the number of rare species that is lost, and so on.
 Potential increases in observation rates do not inform the indicator.
@@
 
-Plotting functions for a series of indicator values is shown below in the indicator calculations.
+Plotting functions for a series of indicator values is shown below in the indicator calculations. This concludes the tour of the `bombLepiSurv` package.
 
@@
 #### NorIns data
-The `Norimon` package has convenience functions to fetch data from the database, housed internally at NINA.
-This database is currently not available outside NINA, but we will implement a solution for this.
+The `Norimon` package has convenience functions to fetch data from the database.
+This database is currently not available outside NINA, but we will implement a solution for this in the future.
 Either we will make the database externally available, or we will provide an alternative route to the data from the GBIF export.
@@
 #### NMBB data
-The data for the bumblebee and butterfly indicators can be accessed through the R-package `bombLepiSurv`.
+The data for the bumblebee and butterfly indicators can similarly be accessed through the R-package `bombLepiSurv`.
 
@@
 
-The indicators from the NMBB program are similary connected to regions, with the exception that the south-east region only covers the old counties of Vestfold and Østfold, and that Nord-Norge isn't covered yet.
+The indicators from the NMBB program are similarly connected to regions, with the exception that the south-east region only covers the old counties of Vestfold and Østfold, and that Nord-Norge isn't covered yet.
+
+```{r, eval=FALSE}
+bombLepiSurv::humlesommerfConnect()
+```
+
+```{r, echo=FALSE}
+bombLepiSurv::humlesommerfConnect(username = Sys.getenv("BOMB_USER"), Sys.getenv("BOMB_PASSWORD"))
+```
 
@@
 #### Biomass of flying insects in semi-natural land
+```{r, eval=FALSE}
+connect_to_insect_db()
+```
+
+```{r, echo=FALSE}
+connect_to_insect_db(user=Sys.getenv("INSECT_USER"), password=Sys.getenv("INSECT_PASSWORD"))
+```
 
@@
 biomass_sn <- get_biomass(subset_year = 2020:2022,
-                          subset_region = NULL)
+                          subset_region = NULL,
+                          subset_habitat = "Semi-nat")
 ```
@@
 Compare values to a reference state.
-The reference state is here set uniformly for all regions, with a made-up value just to be able to continue the calculations.
+The reference state is here set uniformly for all regions.
 Alternatively, we could calculate the indicator values for each region separately, compare to individual reference states, and then put it all together again.
 
-We here use a single value as a reference state and normalize the values at the same time, using the `/` function.
+To exemplify, we here use a single "made-up" value as a reference state and normalize the values at the same time, using the `/` function.
 
@@
 
-```{r}
+```{r biomass_diff}
 plot(biomass_sn_diff)
@@
 ```{r}
-biomass_sn_to_exp <- biomass_sn_diff$bootstrap_summary
+biomass_sn_to_exp <- biomass_sn_boot$bootstrap_summary
 
 biomass_sn_to_exp <- biomass_sn_to_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
 
@@
 
-biomass_sn_to_exp
+insect_biomass_semi_nat <- biomass_sn_to_exp
 ```
@@
 richness_sn <- get_observations(subset_year = 2020:2022,
-                                subset_region = NULL)
+                                subset_region = NULL,
+                                subset_habitat = "Semi-nat")
 ```
@@
 ```{r}
-plot(richness_sn_diff)
+plot(richness_sn_diff,
+     limit = c(0, 1))
+```
+
+```{r richness}
+map_plot(richness_sn_diff,
+         whole_country = T,
+         limit = c(0, 1))
 ```
@@
 ```{r}
-richness_sn_to_exp <- richness_sn_diff$bootstrap_summary
+richness_sn_to_exp <- richness_sn_boot$bootstrap_summary
 
 richness_sn_to_exp <- richness_sn_to_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
 
@@
 
-richness_sn_to_exp
+insect_richness_semi_nat <- richness_sn_to_exp
 ```
@@
                                    subset_families = pollinator_fam,
-                                   subset_region = NULL)
+                                   subset_habitat = "Semi-nat")
 ```
@@
 ```{r}
-plot(pollinators_sn_diff)
+plot(pollinators_sn_diff,
+     limit = c(0, 1))
 ```
@@
 ```{r}
-pollinators_sn_to_exp <- pollinators_sn_diff$bootstrap_summary
+pollinators_sn_to_exp <- pollinators_sn_boot$bootstrap_summary
 
 pollinators_sn_to_exp <- pollinators_sn_to_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
 
@@
 
-pollinators_sn_to_exp
+insect_pollinators_semi_nat <- pollinators_sn_to_exp
 ```
@@
 #### Species richness of dung associated insects
-Note that the species list will probably be expanded, after a taxonomic review.
-
-Fetch the data.
-
-```{r}
-dung_fam <- c("Muscidae")
-
+We here note all observations of a set of species, known to be associated with dung.
+This species list can be expanded after a further taxonomic review.
+Here we take all observed species in the house fly/stable fly family Muscidae, together with a selection of 70 beetle species known to be associated with dung.
+
+```{r}
+muscidae_species <- tbl(con,
+                    Id(schema = "views",
+                       table = "observed_muscidae_species")) %>% 
+  pull()
+
+coleoptera_dung_species <- tbl(con,
+                    Id(schema = "lookup",
+                       table = "coleoptera_dung_species")) %>% 
+  select(species_latin) %>% 
+  pull()
+
+dung_species <- c(coleoptera_dung_species, muscidae_species)
+```
+
+```{r}
 dung_sn <- get_observations(subset_year = 2020:2022,
-                                   subset_families = dung_fam,
-                                   subset_region = NULL)
-```
-
-[This will be amended with data based on genus and species lists. Need to use the lowest taxonomic resolution. Will implement a species lookup list in the database]
+                            subset_species = dung_species,
+                            subset_habitat = "Semi-nat")
+```
 
@@
 ```{r}
-dung_ref <- 90
+dung_ref <- 100
 
@@
 ```{r}
-plot(dung_sn_diff)
+plot(dung_sn_diff,
+     limit = c(0, 1))
 ```
@@
 ```{r}
-dung_sn_to_exp <- dung_sn_diff$bootstrap_summary
+dung_sn_to_exp <- dung_sn_boot$bootstrap_summary
 
 dung_sn_to_exp <- dung_sn_to_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
 
@@
 
-dung_sn_to_exp
+insect_dung_assoc_semi_nat <- dung_sn_to_exp
 ```
@@
 #### Relationship between Symphyta and Parasitica in semi-natural land
+Parasitica and Symphyta wasps are here defined as a set of families.
 
@@
                                 subset_families = symphyta_fam,
-                                subset_region = NULL)
+                                subset_habitat = "Semi-nat")
 
@@
 
-[Note, this list needs to be checked over by e.g. Arnstein.]
 ```{r}
@@
                                 subset_families = parasitica_fam,
-                                subset_region = NULL)
+                                subset_habitat = "Semi-nat")
 
@@
 ```{r}
-par_sym_sn_exp <- par_sym_sn_diff$bootstrap_summary
+par_sym_sn_exp <- par_sym_sn_boot$bootstrap_summary
 
 par_sym_sn_exp <- par_sym_sn_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
 
@@
 
-par_sym_sn_exp
-```
-
-#### Genetic diversity evenness of flying insects in semi-natural land
-This is not fully implemented yet, and the method shown here functions as a placeholder, awaiting a proper calculation.
-Here, the intraspecific variation is based simply on the number of unique genetic variants (ASVs), and their relative abundance measured as number of DNA-reads in the samples.
-But to do this properly, we need to use the full COI sequence to calculate the true GDE (see Connor et al.
-2023).
-This temporarily postponed due to lack of software, but will be updated (and amended here) as soon as possible.
-
-Fetch the data.
-```{r}
-richness_sn <- get_observations(subset_year = 2020:2022,
-                                subset_region = NULL)
-```
-
-Calculate the indicator values for each region and year.
-
-```{r}
-gde_sn_boot <- bootstrap_value(df = richness_sn,
-                                   value = GDE_by_asv,
+insect_par_sym_semi_nat <- par_sym_sn_exp
+```
+
+**Also calculate and export only parasitica values**
+Could also do similar with symphyta, if needed.
+
+```{r}
+par_sn_boot <- bootstrap_value(par_sym,
+                                value = no_species_par,
+                                groups = c("year", 
+                                           "region_name"))
+```
+
+```{r}
+par_sn_exp <- par_sn_boot$bootstrap_summary
+
+par_sn_exp <- par_sn_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
+  select(year,
+         region = region_name,
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
+
+par_sn_exp <- norway_reg_NorIns %>% 
+  inner_join(par_sn_exp,
+            by = c("region" = "region"),
+            multiple = "all")
+
+insect_par_semi_nat <- par_sn_exp
+```
+
+#### Number of alien species in semi-natural land
+Here we measure the prevalence of alien species in the trap catches. We present two options for calculating the indicator, as it is not clear at this point which of them is superior.
+
+**The simplest** approach is to just sum up the number of alien species present in a region, i.e.
+the species richness of alien species.
+This can be related to the total species richness of the community, and expressed as the proportion of alien species in the community.
+The reference state of alien species is zero by definition, since we don't want any alien species.
+The value for good ecological condition could be set at a low (but arbitrary) proportion of alien species in the community.
+This is done for the indicator of alien plant species, where the value for good ecological condition is set to 0.05 (< 5 % of the plant community should be alien).
+For a hyper-diverse community such as insects, this is probably too high a value.
+In a community of say 20 000 insect species, 5 % would mean 1 000 alien species, which is hard to see as a good condition.
+It is not clear to us at present what level to set here, but we exemplify the calculations with a good condition set to 0.1 % (< 20 species in a 20 000 species community).
+It is also likely that the species richness of the entire community will increase, perhaps substantially, with increased sampling.
+It is not clear if the number of detected native and alien species will increase at the same rate, so it may be that the proportion of alien species will change, perhaps substantially, with increased sampling effort.
+It may be difficult therefore to interpret this frequency of alien insect species.
+
+In addition, this approach will not take into account the spread of the species, a measure that is especially relevant for rare but increasing species.
+A species found only once will count as much as a species found in every locality.
+**To take the prevalence of the species into account**, we can measure the proportion of localities each alien species is observed in.
+These proportions could then be summed together for at total score of alien species.
+Here, a species found in all localities will count much more than a species only found once.
+
+For each alien species, the observation frequency is calculated as the number of localities the species is observed in, divided by the total amount of localities in a geographical region and time period.
+$$p_i = \frac{1}{N} \sum_{i=1, j=1}^{N}{obs_{ij}},$$ where $obs_ij$ is an observation of species $i$ in locality $j$, and $N$ is the total number of possible observations (localities in the region and time period).
+These proportions are combined into a common score for all alien species simply by summing them together; $$alien~score =\sum_{i=1}^{S}{~p_{i}},$$ where $p_i$ is the observation frequency of species $i$, and $S$ is the set of all observed alien species.
+If all alien species are present in all localities, the indicator value simply becomes $S$.
+If the species is not present in all localities, it's influence on the indicator is weighted down linearly with it presence frequency.
+The uncertainty of both types of indicators is calculated by bootstrapping the set of localities within a temporal and spatial scope, similarly to the other indicators.
+
+**Fetch the species list.**
+
+The alien species observations are stored in a materialized view for convenience.
+This is simply a list of observations of the alien species of categories LO, PH, HI, or SE, together with the year localities and locality samplings they were observed in.
+For the purposes of this calculation, we will only consider observations aggregated to the year localities level (and not the prevalence of alien species within locality and season).
+So the first thing to do is to fetch this list, and aggregate it to the year locality level.
+This summary is based on the 2023 of the alien species evaluation, and we only consider species that are identified by metabarcoding with identification confidence "Moderate" or "High" (not Low or Poor).
+In addition, we filter out the forest sites.
+
+```{r}
+alien_obs <- tbl(con,
+                 Id(schema = "views",
+                    table = "alien_obs")) %>% 
+  filter(identification_confidence %in% c("MODERATE", "HIGH"),
+         grepl("Semi-nat", locality))
+```
+
+```{r}
+no_alien_obs_ls <- alien_obs %>% 
+  summarize(as.integer(n_distinct(ls_id))) %>% 
+  pull()
+
+no_alien_obs_yl <- alien_obs %>% 
+  summarize(as.integer(n_distinct(yl_id))) %>% 
+  pull()
+
+no_alien_spec <- alien_obs %>% 
+  summarize(as.integer(n_distinct(species_latin_fixed))) %>% 
+  pull()
+
+no_alien_cat <- alien_obs %>% 
+  summarize(as.integer(n_distinct(risikokategori_2023))) %>% 
+  pull()
+
+```
+
+We can have quick look at some statistics of the data.
+We see that the project thus far has identified `r no_alien_spec` alien species, representing `r no_alien_cat` risk-categories, and that the observations area spread across (all) `r no_alien_obs_yl` semi-natural localities.
+
+To continue, we aggregate the presences of alien species to locality years, meaning recording in how many localities each alien species were detected at least once.
+
+```{r}
+alien_obs_yl <- alien_obs %>% 
+  group_by(yl_id,
+           locality,
+           year,
+           region_name,
+           risikokategori_2023,
+           species_latin_fixed) %>% 
+  summarize(present = sum(present) > 0,
+            na.rm = TRUE,
+            .groups = "drop") %>% 
+  arrange(locality, species_latin_fixed) %>% 
+  collect(na.rm = TRUE)
+```
+
+```{r}
+alien_obs_yl
+```
+
+Next step is to bootstrap the alien observations to get the uncertainty measurement.
+This has to be done in a slightly different way than for the aggregated diversity or biomass values that we have used for the other indicators.
+Here we have a set of alien species presences for a locality, and when we bootstrap the localities, all these presences should go together, as they represent a coherent community.
+So we first have to take bootstrap samples of sets of localities (with replacement).
+Then we calculate the presence (or presence frequency) of each alien species $p_i$, before we can summarize all of these into the indicator value $H_{alien}$.
+This is implemented in a separate function `bootstrap_alien_obs` in the `Norimon` package, were we can optionally either sum the species numbers, or sum the species presence frequencies.
+Note that for all alien species $i$ recorded in at least one locality, we have recorded the presences and absences in all localities $j$.
+This is needed for his function to work.
+This function is more computer intensive than the bootstrap of already aggregated values, so we employ parallelization through the `parallel` package internally.
+
+**We first calculate the indicator based on the the number of alien species**
+
+```{r}
+alien_sn_boot <- bootstrap_alien_obs(alien_obs_yl,
+                                     groups = c("year",
+                                                "region_name"),
+                                     rolling_year_window = TRUE,
+                                     only_no_spec = TRUE)
+
+```
+
+```{r}
+plot(alien_sn_boot)
+```
+
+The jagged lines in the density plot comes from the discrete values of number of species.
+Østlandet has the most alien species, followed by Sørlandet, and lastly Trøndelag.
+Trøndelag has a fewer total number of species, but similar amount of alien species in each locality, leading to a very narrow confidence band from the bootstrap.
+
+```{r}
+ts_plot(alien_sn_boot)
+```
+
+As explained above, these values could be expressed as a proportion of the the total species richness.
+Since the total species richness is different for the different regions, we (currently) have to do this separately for each region.
+
+```{r}
+alien_obs_yl_sor <- alien_obs_yl %>% 
+  filter(region_name == "Sørlandet")
+
+alien_obs_yl_ost <- alien_obs_yl %>% 
+  filter(region_name == "Østlandet")
+
+alien_obs_yl_trond <- alien_obs_yl %>% 
+  filter(region_name == "Trøndelag")
+
+```
+
+```{r}
+alien_sn_sor_boot <- bootstrap_alien_obs(alien_obs_yl_sor,
+                                     groups = c("year",
+                                                "region_name"),
+                                     rolling_year_window = TRUE,
+                                     only_no_spec = TRUE)
+alien_sn_ost_boot <- bootstrap_alien_obs(alien_obs_yl_ost,
+                                     groups = c("year",
+                                                "region_name"),
+                                     rolling_year_window = TRUE,
+                                     only_no_spec = TRUE)
+alien_sn_trond_boot <- bootstrap_alien_obs(alien_obs_yl_trond,
+                                     groups = c("year",
+                                                "region_name"),
+                                     rolling_year_window = TRUE,
+                                     only_no_spec = TRUE)
+```
+
+Next, we fetch a record of the species richness found so far, in each region.
+
+```{r}
+region_richn <- get_observations(agg_level = "region_habitat")
+```
+```{r}
+region_richn
+```
+
+```{r}
+region_richn_sor <- region_richn %>% 
+  filter(region_name == "Sørlandet",
+         habitat_type == "Semi-nat") %>% 
+  select(no_species) %>% 
+  pull() %>% 
+  as.numeric()
+
+region_richn_ost <- region_richn %>% 
+  filter(region_name == "Østlandet",
+         habitat_type == "Semi-nat") %>% 
+  select(no_species) %>% 
+  pull() %>% 
+  as.numeric()
+
+region_richn_trond <- region_richn %>% 
+  filter(region_name == "Trøndelag",
+         habitat_type == "Semi-nat") %>% 
+  select(no_species) %>% 
+  pull() %>% 
+  as.numeric()
+
+```
+Next, we scale each indicator to the total number of observed species.
+
+```{r}
+alien_sn_sor_prop <- alien_sn_sor_boot / region_richn_sor
+alien_sn_ost_prop <- alien_sn_ost_boot / region_richn_ost
+alien_sn_trond_prop <- alien_sn_trond_boot / region_richn_trond
+```
+
+We could scale these values with a reference value of 0, and a breaking point of 0.001.
+eaTools has a function for that, but for this to work, we currently can only operate on the aggregate values of the indicator (not the bootstrap samples).
+
+```{r}
+alien_sn_prop <- alien_sn_sor_prop[[1]] %>% 
+  rbind(alien_sn_ost_prop[[1]]) %>% 
+  rbind(alien_sn_trond_prop[[1]]) %>% 
+  select(year,
+         region = region_name,
+         alien_score = boot_value,
+         boot_sd)
+```
+```{r}
+alien_sn_prop_sf <- norway_reg_NorIns  %>% 
+  inner_join(alien_sn_prop,
+            by = c("region" = "region"),
+            multiple = "all")
+```
+
+```{r}
+alien_sn_prop_sf$alien_score_scaled <- eaTools::ea_normalise(alien_sn_prop_sf, 
+                                              vector = "alien_score", 
+                                              upper_reference_level = 0.01,
+                                              lower_reference_level = 0,
+                                              break_point = 0.001, 
+                                              reverse = TRUE,
+                                              plot = FALSE)
+```
+
+This calculation is then an sf-object that can be exported.
+
+```{r}
+
+alien_sn_exp <- alien_sn_prop_sf %>% 
+  mutate(reference_high = 0.01,
+         reference_low = 0,
+         thr = 0.001) %>% 
+  select(year,
+         region,
+         v = alien_score,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i = alien_score_scaled
+         )
+
+insect_alien_prop_semi_nat <- alien_sn_exp
+```
+
+**Secondly, we calculate the indicator based on the distribution of the alien species**
+
+We call this alien_H, since the calculation resembles Hill numbers from calculations of Shannon information.
+
+```{r}
+alien_H_boot <- bootstrap_alien_obs(alien_obs_yl,
+                                    groups = c("year",
+                                               "region_name"),
+                                    rolling_year_window = TRUE,
+                                    only_no_spec = FALSE)
+
+```
+```{r}
+alien_H_boot
+```
+
+```{r}
+plot(alien_H_boot)
+```
+<br/>
+Interstingly, by this calculation method, the different regions have much more similar values.
+Specifically, the southern region has comparable values to the eastern region, while Trøndelag has lower values.
+This indicates that region Østlandet has markedly higher number of alien species than the Southern region, but with generally more limited distribution.
+While in comparison, the alien species in the south are more frequently distributed throughout the sampling localities.
+
+```{r}
+ts_plot(alien_H_boot)
+```
+
+This indicator also has a natural reference value, with zero presences of alien species, meaning an alien score of 0.
+We could scale this indicator similarly as with the indicator based on the pure number of alien species.
+Similar to that case, there is no upper limit to how many alien species there can be, but we set a high enough value, and use a breaking point for the scaling, this upper end point has limited influence.
+
+We don't have enough experience to determine an appropriate breakpoint for the scaling at this point, or an upper value, but exemplify the calculation with some subjective values
+
+```{r}
+
+alien_H_boot_sf <- alien_H_boot[[1]] %>% 
+  select(year,
+         region = region_name,
+         alien_score = boot_value,
+         boot_sd)
+
+alien_H_boot_sf <- norway_reg_NorIns  %>% 
+  inner_join(alien_H_boot_sf,
+            by = c("region" = "region"),
+            multiple = "all")
+
+alien_H_boot_sf$alien_score_scaled <- eaTools::ea_normalise(alien_H_boot_sf,
+                                                            vector = "alien_score", 
+                                                            upper_reference_level = 10,
+                                                            lower_reference_level = 0,
+                                                            break_point = 2, 
+                                                            reverse = TRUE,
+                                                            plot = FALSE)
+
+```
+
+This is then ready for export.
+```{r}
+alien_H_sn_exp <- alien_H_boot_sf %>% 
+  mutate(reference_high = 10,
+         reference_low = 0,
+         thr = 2) %>% 
+  select(year,
+         region,
+         v = alien_score,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i = alien_score_scaled
+         )
+
+insect_alien_H_semi_nat <- alien_H_sn_exp
+```
+
+#### Genetic diversity of flying insects in semi-natural land
+The genetic diversity (GD) of a species is a measurement of the differences between DNA variants within the same species, here calculated on the COI region of the mitochondrial DNA.
+It is a different measurement in principle than the total number of species in a community, and the number of distinct DNA variants in a community.
+It is an empirical question whether these measurement correlate on the global or local scale.
+For instance, Miraldo et al.
+(2016) found global patterns of decreasing GD near the poles, coinciding with global patterns of species richness.
+These patterns may share causes, such as higher mutation rates and shorter generation times in warmer climates (Rohde K.
+1992).
+However, it is not clear if and how these patterns are distributed within a smaller region, such as within Norway.
+Spatial and temporal trends in genetic diversity may indicate pressures on species with reduced effective population sizes, before the complete loss of a species.
+
+Following e.g.
+Miranda et al.
+(2016), we define GD as the average number of nucleotide differences per nucleotide site across all pairwise sequence comparisons within a species.
+Mathematically this can be expressed as: $$\hat{\prod}_s = \frac{1}{(\frac{n}{2})} \sum_{i=1}^{n-1}\sum_{j=i+1}^{n}\frac{k_{ij}}{m_{ij}},$$ where $s$ is a single species, $k_{ij}$ is the number of different nucleotides between sequence $i$ and sequence $j$, $m_{ij}$ is the number of shared base pairs between sequence $i$ and $j$, and $(\frac{n}{2})$ represents the number of pairwise comparisons made (Theodoridis et al 2020).
+Following French et al.
+(2023), we here considered only species that where found at least 3 separate times, to make sure we had enough samples to observe any intraspecific diversity.
+
+These species-specific measurements can be combined into single community measurements in several ways.
+Firstly, we can simply take the average GD of a community of species (GDM), or $$GDM = \frac{1}{S}\sum_{s=1}^{S}\hat{\prod}_s, $$ where $S$ is the total number of species, and $s$ is each species (Theodoridis et al.
+2020).
+The mean genetic diversity of a community is a natural starting point, but can - analogous to species richness -, hide important patterns in the relative contributions from individual species.
+In short, the distributions of GD might be skewed, so that a few species carries the most influence on the measurement.
+Therefore, French et al.
+(2023) has developed a genetic evenness measurement (GDE), that is analogous to Shannon's entropy for communities on the species level.
+This is defined as $$\frac{exp(\sum_{i=0}^{N} - \hat{\prod}_i ln(\hat{\prod}_i))}{N},$$ where the denominator $N$ is the number of species in the community, and $\hat{\prod}_i$ is the GD calculated above for a single species.
+The numerator is the Hill number of the Shannon entropy.
+This results in a measurement that can be compared across communities of different species richness.
+
+We have calculated both the GDM and GDE, but here show only the results of the GDM.
+The calculated GDE values showed a strong relationship with the number of species in the samples.
+This could indicate a possible error in the calculation routines, or a more fundamental problem, which needs to be checked before going forward with this measurement.
+Alternatively, the calculation is correct, but then the values are superfluous, since they give essentially the same information as the total richness, already calculated as a separate indicator.
+
+Therefore, we here present only the GDM as an indicator.
+Note that this is a relatively new measures, and this is the first time it is calculated for large insect communities in Norway.
+We need further experience with these measurements to tease out potential pro's and con's in the Norwegian setting.
+The raw-data on the DNA-sequences is as of yet not publicly available, and the calculations of GDM and GDE is not shown in detail.
+
+```{r}
+gdm <- readRDS("data/GDM_year_locality.rds")
+```
+
+The `gdm` contains the mean genetic diversity for the species found in every year locality combination. This is calculated after first filtering on species that is found at least 3 times.
+
+The `gdm` object contains the year_locality_ids as a key, which we can join to the relevant metadata, e.g. through the alien observation object used above.
+
+```{r}
+gdm_sn <- alien_obs_yl  %>% 
+  ungroup() %>% 
+  select(yl_id, locality, year, region_name) %>% 
+  distinct() %>% 
+  inner_join(gdm, by = c("yl_id"= "year_locality_id"))
+
+gdm_sn
+```
+
+Since these calculations are so novel, it can be worth spending some time exploring how the mean genetic diversity relates to other richness measurement at the locality level.
+Remember that globally, the species richness and genetic mean diversity show a general positive correlation (Miraldo et al.
+2016).
+
+```{r}
+richness_sn <- get_observations(agg_level = "year_locality",
+                                subset_habitat = "Semi-nat")
+```
+
+```{r}
+richness_gdm_join <- richness_sn %>% 
+  inner_join(gdm_sn,
+             by = c("locality" = "locality",
+                    "year" = "year",
+                    "region_name" = "region_name"))
+```
+
+```{r}
+ggplot(richness_gdm_join) +
+  geom_point(aes(x = no_species,
+                 y = GDM,
+                 color = region_name)) +
+  geom_smooth(aes(x = no_species,
+                 y = GDM,
+                 color = region_name),
+              method = lm) +
+  scale_color_nina(name = "Region") +
+  xlab("Antall arter")
+```
+<br/>
+Interestingly, and perhaps reassuringly, the number of species doesn't seem to be clearly related to the average genetic diversity. We can look at the mean number of ASVs as well.
+
+```{r}
+ggplot(richness_gdm_join) +
+  geom_point(aes(x = mean_no_asv_per_species,
+                 y = GDM,
+                 color = region_name)) +
+  geom_smooth(aes(x = mean_no_asv_per_species,
+                 y = GDM,
+                 color = region_name),
+              method = lm) +
+  scale_color_nina(name = "Region") +
+  xlab("Antall genetiske varianter per art")
+```
+<br/>
+Here we see a positive relationship in Trøndelag (although uncertain), but not the other regions.
+
+```{r}
+ggplot(richness_gdm_join) +
+  geom_point(aes(x = no_species,
+                 y = mean_no_asv_per_species,
+                 color = region_name)) +
+  geom_smooth(aes(x = no_species,
+                 y = mean_no_asv_per_species,
+                 color = region_name),
+              method = lm) +
+  scale_color_nina(name = "Region") +
+  xlab("Antall arter") +
+  ylab("Antall genetiske varianter per art")
+
+```
+<br/>
+While, for comparison, the number of ASVs (genetic variants) and the number of species per locality show more clear relationships.
+This was just a quick look at the genetic aspect of biodiversity, and we would require further analysis to interpret them robustly.
+In the meantime, we continue with the calculation example of the indicators.
+We bootstrap the mean genetic diversity like with the other indicators.
+
+```{r}
+gdm_sn_boot <- bootstrap_value(df = gdm_sn,
+                                   value = GDM,
                                    groups = c("year",
@@
                                    )
+gdm_sn_boot
+```
+
+```{r}
+plot(gdm_sn_boot)
+```
+
+So far, Trøndelag seem to have lower genetic diversity than Sørlandet and Østlandet.
+
+```{r}
+ts_plot(gdm_sn_boot)
+```
+
+Although, as with the other indicators, we see variation between years. This can be seen if we do not bootstrap on a rolling 5 year window.
+
+```{r}
+gdm_sn_yearly_boot <- bootstrap_value(df = gdm_sn,
+                                      value = GDM,
+                                      groups = c("year",
+                                                 "region_name"),
+                                      rolling_year_window = FALSE
+                                   )
+```
+
+```{r}
+ts_plot(gdm_sn_yearly_boot)
+
 ```
@@
 ```{r}
-gde_sn_ref <- 0.7
-
-gde_sn_diff <- gde_sn_boot / gde_sn_ref
+gdm_sn_ref <- 0.02
+
+gdm_sn_diff <- gdm_sn_boot / gdm_sn_ref
 ```
@@
 ```{r}
-gde_sn_diff
-```
-
-```{r}
-plot(gde_sn_diff)
+gdm_sn_diff
+```
+
+```{r}
+plot(gdm_sn_diff,
+     limit = c(0, 1))
 ```
@@
 ```{r}
-gde_sn_to_exp <- gde_sn_diff$bootstrap_summary
-
-gde_sn_to_exp <- gde_sn_to_exp %>% 
+gdm_sn_to_exp <- gdm_sn_boot$bootstrap_summary
+
+gdm_sn_to_exp <- gdm_sn_to_exp %>% 
+  mutate(reference_high = NA,
+         reference_low = 0,
+         thr = 0.6,
+         i = NA) %>% 
   select(year,
          region = region_name,
-         ind_value = boot_value,
-         ind_sd = boot_sd,
-         ind_lower2.5 = boot_lower2.5,
-         ind_upper97.5 = boot_upper97.5)
-
-gde_sn_to_exp <- norway_reg_NorIns  %>% 
-  inner_join(gde_sn_to_exp,
+         v = boot_value,
+         sd = boot_sd,
+         reference_high,
+         reference_low,
+         thr,
+         i
+         )
+
+gdm_sn_to_exp <- norway_reg_NorIns %>% 
+  inner_join(gdm_sn_to_exp,
             by = c("region" = "region"),
@@
 
-gde_sn_to_exp
+insect_genetic_div_semi_nat <- gdm_sn_to_exp
 ```
@@
 
-```{r bombusGressBarplot, fig.cap = "Bumblebees in grasslands 2021"}
+```{r bombusGressBarplot, fig.cap = "Bumblebees in grasslands 2022"}
 allBombusGressmark2022 <- getAllData(type = "bumblebees",
@@
 
-The calculations for earlier years are already present as datasets in the R-package and can be fetched individually.
+The calculations for earlier years are already present as datasets in the R-package and can be fetched individually with `data("dataset")`.
 Once this is done, we can plot the timeseries of values using the function `indPlot`.
 
-```{r}
+```{r, echo = F}
 data("sInd2022VestGress")
@@
 hGressIndVals_to_exp <- hGressIndVals %>% 
+  mutate(reference_high = 1,
+         reference_low = 0,
+         thr = 0.6,
+         v = NA) %>% 
   select(year = Year,
          region = Region,
-         ind_value = Indicator_value,
+         i = Indicator_value,
          ind_sd = NULL,
@@
 
-hGressIndVals_to_exp
+insect_bumblebees_semi_nat <- hGressIndVals_to_exp
 
@@
 sGressIndVals_to_exp <- sGressIndVals %>% 
+  mutate(reference_high = 1,
+         reference_low = 0,
+         thr = 0.6,
+         v = NA) %>% 
   select(year = Year,
          region = Region,
-         ind_value = Indicator_value,
+         i = Indicator_value,
          ind_sd = NULL,
@@
 
-sGressIndVals_to_exp
+insect_butterflies_semi_nat <- sGressIndVals_to_exp
 
@@
 
-TBD. Need to consult with others.
+```{r}
+to_export <- ls()[grepl("insect_", ls())] %>% 
+  as.list
+
+```
+
+```{r echo = T, results = "hide"}
+
+write_indicators <- function(x,
+                             out_dir = "indicators/indicatorMaps",
+                             suffix = ".shp"){
+
+  system(paste0("mkdir -p ", out_dir, "/", x))
+
+  write_sf(get(x[1]),
+           dsn = paste0(out_dir,"/", x, "/", x, suffix))
+
+}
+
+lapply(to_export, write_indicators)
+```

### 2023-12-05 insectIndicators.Rmd

@@
               method = lm) +
-  scale_color_nina(name = "Region") +
+  NinaR::scale_color_nina(name = "Region") +
   xlab("Antall arter")
@@
               method = lm) +
-  scale_color_nina(name = "Region") +
+  NinaR::scale_color_nina(name = "Region") +
   xlab("Antall genetiske varianter per art")
@@
               method = lm) +
-  scale_color_nina(name = "Region") +
+  NinaR::scale_color_nina(name = "Region") +
   xlab("Antall arter") +
@@
 
-```{r echo = T, results = "hide"}
+```{r echo = T, results = "hide", eval = FALSE}
 

### 2023-12-14 birdIndicators.Rmd

@@
 ## Introduction {#intro-bb}
-The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called TOV-E) was established between 2005 and 2010.
-The main goal of the programme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (available at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway.
-The "hekkefuglovervåking" is coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), and has replaced predecessor monitoring schemes that ran from 1996 to 2008.
+The Norwegian common breeding bird monitoring scheme ("hekkefuglovervåking" in Norwegian; formerly called "TOV-E") was established between 2005 and 2010.
+The main goal of the scheme is to gather data on the abundance of the most common terrestrial bird species to assess population trends and changes across Norway (results can be found at the website), and help proposing efficient management and conservation measures for these species and their habitats in Norway.
+The "hekkefuglovervåking" is coordinated by NINA and BirdLife Norway and supported financially by the Norwegian Environment Agency (Miljødirektoratet), and replaced a predecessor monitoring scheme that ran from 1996 to 2008.
 
-These data collected through "hekkefuglovervåkingen" form the basis for a variety of national and international assessments including nationally reported species population trends (https://hekkefuglovervakingen.nina.no/hekkefugl), updates to the Red List of Birds in Norway following the IUCN protocol (https://artsdatabanken.no/lister/rodlisteforarter), indicators for the Norwegian Nature Index (https://www.naturindeks.no/), and single- and multi-species indicators that are reported nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/) and internationally to e.g.
+The data collected through "hekkefuglovervåkingen" form the basis for a variety of national and international assessments including nationally reported species population trends (https://hekkefuglovervakingen.nina.no/hekkefugl), updates to the Red List of Birds in Norway following the IUCN protocol (https://artsdatabanken.no/lister/rodlisteforarter), indicators for the Norwegian Nature Index (https://www.naturindeks.no/), and single- and multi-species indicators that are reported nationally (https://miljostatus.miljodirektoratet.no/tema/arter/fugler/) and internationally to e.g.
 the Pan European Bird Monitoring Scheme (PECBMS, https://pecbms.info/) and the OECD.
 
-Workflows generating outputs and numbers for several of these reporting routes are already integrated thanks to coordination and standardization efforts by PECMBS, and the calculation of breeding bird indicators for use in ecosystem condition assessments should be incorporated into the same workflow.
+Workflows generating outputs and numbers from all these reporting routes are already integrated thanks to coordination and standardization efforts by PECMBS, and the calculation of breeding bird indicators for use in ecosystem condition assessments should be incorporated into the same workflow.
 We therefore suggest to use the same multi-species indicators in the framework for ecosystem condition assessment as are reported nationally and internationally: the established "Farmland bird index", "Mountain bird index", and "Forest bird index", as well as a corresponding index for wetland birds.
@@
 June (depending on latitude) and carry out the census in good weather conditions.
-The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route.
+The observers record any bird seen or heard during 5 minutes at each of the pre-selected counting points in the route (see below for further information on the sampling design).
 A total of 229 species are typically recorded during this census.
@@
 ### Representativity in time and space {#rep-in-time-and-space-bb}
-"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats present in Norway.
+"Hekkefuglovervåking" is a nation-wide monitoring scheme consisting of 492 fixed routes across all terrestrial habitats along both latitudinal and altitudinal gradients present in Norway.
 Each route has between 12 and 20 counting points (ca.
@@
 
-This monitoring scheme is therefore designed to gather representative samples in all (common) habitats along both latitudinal and altitudinal gradients.
-
-Even though implementation of the monitoring scheme started in 2005, full coverage at the national scale was not achieved until 2010.
+Even though implementation of the monitoring scheme started in 2005, full geographical coverage at the national scale was not achieved until 2010.
 
@@
 Those values are then converted into "number of pairs" by multiplying the number of individuals by a conversion factor, which is 2 or 5 depending on the species (for national reporting only).
-In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of the population in a given baseline year (in Norway this is typically 2008).
+In the first step of the workflow, the raw counts are converted into species-specific abundance indices, defined as % of change relative to the population in a given baseline year (in Norway this is typically 2008).
 The abundance indices then form the basis of further analyses and aggregation routines.
@@
 The "hekkefuglovervåkingen" was established in 2005 in central Norway and expanded from there to reach full national coverage by 2010.
-For many species, data is already of good coverage and high quality starting in 2008.
-The monitoring programme is still ongoing as of 2023.
+For many species, data are already of good coverage and high quality starting in 2008.
+The monitoring scheme is still ongoing as of 2023.
 
-Predecessor monitoring schemes existed prior to the establishment of "hekkefuglovervåkingen" and led to the collection of decent data in the period from 1996 up to and including 2008. 
+A predecessor monitoring scheme existed prior to the establishment of "hekkefuglovervåkingen" and led to the collection of decent data in the period from 1996 up to and including 2008. 
 Although the predecessor monitoring programmes and "hekkefuglovervåkingen" are not the same, the overlap for the years 2007 and 2008 makes it possible to combine abundance indices from across time using tools like "RSWAN", which is provided by PECBMS.
-For a subset of 55 species, longer time-series of abundance indices (1996-2022, as of now) are therefore available.
+For a subset of 55 species, longer time-series of abundance indices (1996-2023, as of now) are therefore available.
 
 ### Aditional comments about the dataset {#additional-comments-bb}
+NINA perfoems quality-checks of the data delivered by the observers to ensure that the new data entered to the database are reliable.
+In addition, NINA completes the annual detaset by entering the zeroes (undetected species) at surveyed routes and -1s (for not surveyed routes).
+This ensures that all species have data (either a positive count, a zero or a -1) for each route and year.
+
 From all species recorded within the Norwegian common breeding bird monitoring scheme (n = 229), 85 have sufficient data for abundance trend analyses.
@@
 ### Data sets {#datasets-bb}
-The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåkingen" and the calculated abundance indices from the predecessor monitoring schemes.
-Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow.
+The workflow relies on two major sources of data: the up-to-date raw data from "hekkefuglovervåking" and the calculated abundance indices from the predecessor monitoring scheme.
+Additionally, we also use a couple of lists holding information on which species (and years) are to be included in which parts of the analytical workflow as the different outcomes require different information.
 
-#### Monitoring data from "Hekkefuglovervåkingen"
-The route-based raw data from "hekkefuglovervåkingen" are compiled after each breeding season and stored in NINA's database.
-From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2022).
+#### Monitoring data from "Hekkefuglovervåking"
+The route-based raw data from "hekkefuglovervåking" are compiled after each breeding season and curated and stored in NINA's database.
+Before storing new data in the database, a thorough quality-check is performed.
+From that database we download the data relevant for the PECBMS Trim analyses and for the specified range of years (here from 2006 up to and including 2023).
 The function ``downloadData_TRIM`` does that for a specified year range (from ``minYear`` up to and including ``maxYear``), and requires specifying which driver is used for accessing the database (via argument ``DriverName``).
@@
 When running locally from NINA computers, we need to select "SQL Server".
-Note that downloading this data requires user-specific access privileges which can be requested from NINA.
+Note that downloading these data requires user-specific access privileges which can be requested from NINA IT department ('datahjelp').
 
@@
 minYear <- 2006
-maxYear <- 2022
+maxYear <- 2023
 
@@
 #### PECBMS Trim results from predecessor monitoring programme
-The abundance indices generated by the PECBMS RTRIM workflow using data from the predecessor scheme are stored in NINAs local file repositories.
-Unlike the Trim analyses for the data from "hekkefuglovervåkingen", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being added.
+The abundance indices generated by the PECBMS RTRIM workflow using data from the predecessor scheme are stored in NINA's local file repositories.
+Unlike the Trim analyses for the data from "hekkefuglovervåking", which are re-run every year after new data has been added, the old abundance indices remain unchanged as no new data are being collected.
 These data are needed for the RSWAN workflow, and they need to be present in the respective "working_folder".
-So we copy them from their location in NINA's local file storage into the correct folder for analyses.
+Hence, we copy the files from their location in NINA's local file storage into the correct folder for analyses.
 
@@
 
-The table ``Spp_selection`` contains information for each species on which reporting routes is should be included in.
+The table ``Spp_selection`` contains information for each species on which reporting routes should be included in.
 The multispecies indices are grouped into a single category ("MSI") here, but we can extract the species that go into the MSI for each ecosystem via the additionally provided lists.
@@
 
-As per now, the breeding bird indicators are calculated at the national scale only and the data is therefore not linked to any geofiles here.
+As per now, the breeding bird indicators are calculated at the national scale only and the data are therefore not linked to any geofiles here.
 
@@
 
-In the next step, the downloaded raw data get analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data", @R-rtrim2020).
-The RTRIM analyses are run for a subset of 85 species which have sufficient data for it and using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/).
+In the next step, the downloaded raw data is analysed and converted into species-specific abundance indices using the RTRIM approach ("**TR**ends and **I**ndices for **M**onitoring data", @R-rtrim2020).
+The RTRIM analyses are run for a subset of 85 species which have sufficient data for it using the standardized scripts provided by PECBMS (more information on https://pecbms.info/methods/software/trim/).
 For the application here, we have written wrapper function around the PECBMS code for each step of the RTRIM workflow.
@@
 These are delivered directly to PECBMS each year after the analyses have been run.
-Prior to delivery, they get copied into a designated folder:
+Prior to delivery, they are copied into a designated folder:
 
@@
 
-For 55 of the 85 species, RTRIM analyses have also been run using data from the predecessor monitoring schemes.
+For 55 of the 85 species with sufficient data, RTRIM analyses have also been run using data from the predecessor monitoring scheme.
 The RTRIM output files from both old and new analyses should now be present in the same folder.
@@
 
-THe RSWAN routine produces output files for the selected 55 species that are equivalent to those produced by the RTRIM workflow, but have a prefix "COMB" to indicate that they are based on combined time series.
+The RSWAN routine produces output files for the selected 55 species that are equivalent to those produced by the RTRIM workflow, but have a prefix "COMB" to indicate that they are based on combined time series.
 
@@
 
-The breeding bird indicators for ecosystem condition assessments are mutli-species indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.
+The breeding bird indicators for ecosystem condition assessments are mutlispecies indices (MSIs) calculated following the guidelines from PECBMS (https://www.cbs.nl/en-gb/society/nature-and-environment/indices-and-trends--trim--/msi-tool) and using the algorithm of @soldaat2017.
 
@@
 
-The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument ). We list these for all four ecosystem indices:
+The function also requires us to provide a name (argument) and a list of EURING codes of species to include (argument). We list these for all four ecosystem indices:
 
@@
 ```
+
 Notably, I have not added the ``i_YYYY` columns yet as these can only be calculated once we have discussed and come to an agreement on reference value (year), threshold value, and scaling function.

### 2024-09-27 gjengroingMetaModel.Rmd
Test

--- 2023-10-19 gjengroing.Rmd
@@
 
-Zander Venter
-
-September 2023
+Andrew Gray
+
+September 2024
 
@@
 library(flextable)
+library(terra)
+library(dplyr)
+library(parallel)
+library(stringr)
+library(httr)
+library(jsonlite)
 # Set global variable defining whether you want to run everything from scratch (very long runtime) or use pre-exported data (short runtime)
@@
 ECT       <- "Structural state characteristic" 
-Contact   <- "Zander Venter"
+Contact   <- "Andrew Gray"
 ```
@@
 The gjengroing indicator describes the regrowth of woody vegetation (trees and bushes) in open ecosystems (wetland, semi- and naturally open areas) across Norway.
-We will use a spatial reference approach where reference areas define good or optimal vegetation regrowth heights.
-
-The workflow spans two platforms including RStudio and Google Earth Engine (GEE). To reproduce this workflow you will need a GEE account and access to the NINA RStudio and R/GeoSpatialData servers.
+Here, we use meta's canopy height base model to predict canopy height within areas of interest using airborne imagery from norge i bilder.
+This approach will allow future monitoring of encroachment, as airborne imagery is scheduled to be collected every 5 - 10 years whereas national LiDAR data has no future schedule.
+It is worth noting that the yet-to-be-launched P-Band radar satellite, Biomass, may provide an more precise method for future monitoring once data is available.
+Here though, we will use a spatial reference approach where reference areas define good or optimal vegetation regrowth heights.
+
+The workflow spans two platforms including RStudio and Python.
 
@@
 
+##### Directories
+
+```{r}
+wms_base_url <- 'https://wms.geonorge.no/skwms1/wms.nib?SERVICE=WMS&VERSION=1.1.1&REQUEST=GetMap&SRS=EPSG:25833'
+output_dir <- "/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/tiles"
+metadata_url <- 'https://tjenester.norgeibilder.no/rest/projectMetadata.ashx'
+if (!dir.exists(output_dir)) {
+  dir.create(output_dir, recursive = TRUE)
+}
+```
+
+###### Querying the airborne imagery metadata
+
+This enables us to determine the date of the orthophoto tiles we will later download from the geonorge webserver.
+
+```{r}
+ query_metadata <- function(xmin, ymin, xmax, ymax, crs_code) {
+  bbox_coords <- sprintf("%f,%f;%f,%f;%f,%f;%f,%f", xmin, ymin, xmin, ymax, xmax, ymax, xmax, ymin)
+
+  param <- list(
+    Filter = "ortofototype in (1,2,3,4,5,7,8,9,10,11,12)",
+    Coordinates = bbox_coords,
+    InputWKID = as.character(crs_code),
+    ReturnMetadata = TRUE,
+    StopOnCover = FALSE
+  )
+
+  response <- httr::GET(
+    url = metadata_url,
+    query = list(request = jsonlite::toJSON(param, auto_unbox = TRUE))
+  )
+
+  response_content <- httr::content(response, "text", encoding = "UTF-8")
+
+  if (httr::http_status(response)$category != "Success") {
+    return(NULL)
+  }
+
+  if (response_content == "" || is.null(response_content)) {
+    return(NULL)
+  }
+
+  parsed_metadata <- tryCatch({
+    fromJSON(response_content, flatten = TRUE)
+  }, error = function(e) {
+    return(NULL)
+  })
+
+  return(parsed_metadata)
+}
+```
+The WMS provides the most recent image over each search area so we can pull the date from the metadata service to add to the tile filename.
+
+```{r}
+get_most_recent_metadata <- function(metadata) {
+  filtered_metadata <- metadata$ProjectMetadata %>%
+    filter(!is.na(properties.fotodato_date)) %>%
+    mutate(fotodato_date = as.Date(properties.fotodato_date)) %>%
+    arrange(desc(fotodato_date)) %>%
+    slice(1)  # Get the most recent one
+
+  if (nrow(filtered_metadata) == 0) {
+    return(NULL)
+  }
+
+  return(filtered_metadata)
+}
+```
+
+##### Defining the area of interest
+Here we include a shapefile of our area of intertest.
+The script will loop through all features in the shapefile, calculate their extent and query the WMS to download orthophotos within that extent in 256 x 256 pixel tiles.
+```{r}
+ROIs <- vect("/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/aoi.shp")
+RegionPolygonUTM33N <- st_as_sf(ROIs)
+if (st_crs(RegionPolygonUTM33N)$epsg != 25833) {
+  RegionPolygonUTM33N <- st_transform(RegionPolygonUTM33N, crs = 25833)
+}
+
+tile_size <- 256
+resolution <- 0.5  # 0.5 meter resolution
+```
+
+This is the function to get and download the orthophotos over our areas of interest. 
+```{r}
+fn.getOrtoImages <- function(geometry, id_field, output_dir) {
+  tryCatch({
+    if (!st_is_valid(geometry)) {
+      geometry <- st_make_valid(geometry)
+    }
+
+    geom_extent <- st_bbox(geometry)
+
+    # Query metadata to get the most recent layer and date
+    metadata <- query_metadata(geom_extent$xmin, geom_extent$ymin, geom_extent$xmax, geom_extent$ymax, 25833)
+
+    if (is.null(metadata) || is.null(metadata$ProjectMetadata)) {
+      return(NULL)
+    }
+
+    recent_project <- get_most_recent_metadata(metadata)
+
+    if (is.null(recent_project)) {
+      return(NULL)
+    }
+
+    recent_layer <- recent_project$properties.prosjektnavn
+    image_date <- recent_project$properties.fotodato_date
+
+    tile_width_meters <- tile_size * resolution
+    tile_height_meters <- tile_size * resolution
+
+    # Ensure tile size remains 256x256
+    x_min <- floor(geom_extent$xmin / tile_width_meters) * tile_width_meters
+    y_min <- floor(geom_extent$ymin / tile_height_meters) * tile_height_meters
+    x_max <- ceiling(geom_extent$xmax / tile_width_meters) * tile_width_meters
+    y_max <- ceiling(geom_extent$ymax / tile_height_meters) * tile_height_meters
+
+    # Calculate number of tiles in x and y directions
+    x_tiles <- ceiling((x_max - x_min) / tile_width_meters)
+    y_tiles <- ceiling((y_max - y_min) / tile_height_meters)
+
+    tile_index <- 1
+
+    for (i in 0:(x_tiles - 1)) {
+      for (j in 0:(y_tiles - 1)) {
+        xmin <- x_min + i * tile_width_meters
+        xmax <- xmin + tile_width_meters
+        ymin <- y_min + j * tile_height_meters
+        ymax <- ymin + tile_height_meters
+
+        # Ensure the tiles have consistent size
+        bbox_polygon <- st_polygon(list(matrix(c(
+          xmin, ymin,
+          xmin, ymax,
+          xmax, ymax,
+          xmax, ymin,
+          xmin, ymin
+        ), ncol = 2, byrow = TRUE)))
+        bbox_polygon <- st_sfc(bbox_polygon, crs = st_crs(geometry))
+
+        bbx <- st_bbox(bbox_polygon)
+
+        # Construct WMS URL dynamically with correct SRS, LAYERS, and BBOX in EPSG:25833
+        wms_url <- paste0(
+          'https://wms.geonorge.no/skwms1/wms.nib?',
+          'SERVICE=WMS&VERSION=1.1.1&REQUEST=GetMap',
+          '&LAYERS=', URLencode(recent_layer),
+          '&SRS=EPSG:25833',
+          '&BBOX=', paste(bbx[c("xmin", "ymin", "xmax", "ymax")], collapse = ","),
+          '&WIDTH=', tile_size,
+          '&HEIGHT=', tile_size,
+          '&FORMAT=image/tiff',
+          '&TRANSPARENT=TRUE'
+        )
+
+        # Output file path using the "id" field, tile index, and image date in the file name
+        t <- file.path(output_dir, paste0(image_date,"_", id_field, "_", tile_index, ".tif"))
+
+        # GDAL translate options for 0.5 meter pixel size
+        reso <- c('-tr', as.character(resolution), as.character(resolution), '-co', 'COMPRESS=NONE')
+
+        # Try downloading the tile
+        result <- try({
+          sf::gdal_utils('translate', source = wms_url, destination = t, options = reso)
+        }, silent = TRUE)
+
+        # Retry mechanism
+        for (k in 1:5) {
+          if (!file.exists(t)) {
+            Sys.sleep(1)
+            result <- try({
+              sf::gdal_utils('translate', source = wms_url, destination = t, options = reso)
+            }, silent = TRUE)
+          }
+        }
+
+        tile_index <- tile_index + 1
+      }
+    }
+
+  }, error = function(e) {
+    return(NULL)
+  })
+}
+```
+
+We can run the functions in parallel to speed up processing time adjust the number of cores to suit your setup.
+```{r}
+numCores <- 30
+
+# Use mclapply for parallel processing
+mclapply(1:nrow(RegionPolygonUTM33N), function(i) {
+  geometry <- RegionPolygonUTM33N[i, ]
+  id_field <- RegionPolygonUTM33N$id[i]  # Adjust field name if necessary
+  fn.getOrtoImages(geometry, id_field, output_dir)
+}, mc.cores = numCores)
+```
+
+######Zanders/Ida's script######
 #### Regions {#reg-enc}

### 2024-09-27 gjengroingMetaModel.Rmd
added functions to download airborne imagery

@@
 wms_base_url <- 'https://wms.geonorge.no/skwms1/wms.nib?SERVICE=WMS&VERSION=1.1.1&REQUEST=GetMap&SRS=EPSG:25833'
-output_dir <- "/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/tiles"
+output_dir <- "data/encroachment/metamodel/tiles"
 metadata_url <- 'https://tjenester.norgeibilder.no/rest/projectMetadata.ashx'
@@
 The script will loop through all features in the shapefile, calculate their extent and query the WMS to download orthophotos within that extent in 256 x 256 pixel tiles.
-```{r}
-ROIs <- vect("/data/P-Prosjekter2/412421_okologisk_tilstand_2024/Andrew/meta/HighResCanopyHeight-main/DataDownload/testHarstad/aoi.shp")
-RegionPolygonUTM33N <- st_as_sf(ROIs)
-if (st_crs(RegionPolygonUTM33N)$epsg != 25833) {
-  RegionPolygonUTM33N <- st_transform(RegionPolygonUTM33N, crs = 25833)
-}
+The AOI shape file should be in EPSG:25833
+```{r}
+
+AOIs <- sf::st_read("data/encroachment/metamodel/AOIs.shp")
 
@@
 # Use mclapply for parallel processing
-mclapply(1:nrow(RegionPolygonUTM33N), function(i) {
-  geometry <- RegionPolygonUTM33N[i, ]
-  id_field <- RegionPolygonUTM33N$id[i]  # Adjust field name if necessary
+mclapply(1:nrow(AOIs), function(i) {
+  geometry <- AOIs[i, ]
+  id_field <- AOIs$id[i]  # Adjust field name if necessary
   fn.getOrtoImages(geometry, id_field, output_dir)
@@
 ```
+
+OK, now that we have our tiles, we can pass them to the canopy height model please see https://github.com/facebookresearch/HighResCanopyHeight for more details on the model.
 

### 2024-09-27 gjengroingMetaModel.Rmd
Modified to just include data download script

@@
 
-The workflow spans two platforms including RStudio and Python.
-
-## About the underlying data {#about-data-enc}
-
-We rely on the following datasets:
-
--   [Nature type polygons](https://kartkatalog.miljodirektoratet.no/Dataset/Details/2031) are used to identify reference areas with good ecological condition.
--   NIBIO's [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6) is used as the population sample for each ecosystem type.
-The population sample are the areas used to determine ecological condition relative to the reference areas.
-We also use the AR5 data to identify forest surrounding the population polygons for defining the lower limit of ecological condition index (i.e..
-poor condition),
--   LiDAR-derived digital elevation model from Kartverket's [høydedata](https://hoydedata.no/LaserInnsyn/).
-This includes both a terrain (DTM) and surface model (DSM).
-We calculate the canopy height model (DSM - DTM) to get the height of objects above the ground.
-From this we remove buildings, and what remains is vegetation - mostly trees but also some bushes or smaller woody plants.
--   Kartverket's [national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df).
-We use this to stratify reference (good condition) and forest (poor condition) heights by elevational band.
-This is done in combination with bioclimatic zones - see next point.
--   Moen's [bioclimatic zones](https://artsdatabanken.no/Pages/181901/Bioklimatiske_soner).
-We use this to stratify reference (good condition) and forest (poor condition) heights by bioclimatic (also referred to as vegetation/climatic) zones.
--   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97) are used to isolate vegetation in the LiDAR height data.
--   A [European satellite-based map](https://www.nature.com/articles/s41893-020-00609-y) of forest clear cuts is used for identifying AR5 forest patches in near-climax successional stages.
--   The [SSB 10km grid](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) is used for visualization purposes.
--   The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.
-
-### Representativity in time and space {#rep-enc}
-
-The index will cover the mainland of Norway.
-The analysis will be stratified by (1) åpne and (2) våtmark ecosystems.
-The former includes both semi-naturlig mark and naturlig åpne områder under skoggrensa (due to the fact that AR5 does not differentiate these), and the latter includes våtmark only.
-The LiDAR data cover a range of years and therefore the indicator represents conditions for circa 2010 to 2021.
-
-### Original units {#units-enc}
-
-The original units for ecological condition are meters. This is the height of the vegetation within referecne, polygon and forest areas.
-
-### Temporal coverage {#temp-enc}
-
-Circa 2010 to 2021.
-This is a single snapshot and not a change analysis.
-In the future, when LiDAR data has been repeated across the country, it may be possible to do a change assessment.
-In addition, the use of optical or radar satellite imagery may serve as a proxy for vegetation height using machine learning.
-This will allow for annual updates, depending on the uncertainty in the satellite-based maps of vegetation height.
-
-### Additional comments about the dataset {#comm-enc}
-
-None.
-
-## Ecosystem characteristic
-
-### Norwegian standard
-
-Funksjonell sammensetning innen trofiske nivåer, reflecting the change in dominant growth form.
-
-## Collinearities with other indicators
-
-There is possibly a collinearity with the [primary production indicator](#NDVI-indicator-natopen) (primærproduksjon).
-The primary production indicator uses the normalized difference vegetation index (NDVI) as a proxy for vegetation production.
-NDVI can be correlated with vegetation height and consequently yield similar results to the LiDAR-based gjengroing indicator.
-
-## Reference state and values
-
-The methodology used to calculate the _gjengroing_ indicator is outlined in the schematic below.
-The workflow in the schematic is conducted for all reference and population polygons in Norway and repeated for each ecosystem type (åpne and våtmark), respectively.
-The indicator values are aggregated to a 50km grid and regional level at the end.
-The individual steps are discussed in turn in the following subsections.
-
-```{r enc-workflow, fig.cap="Schematic illustration of how the encroachment indicator is calculated.
-The lower bound for poor condition is set by LiDAR heights from mature forest surrounding the population polygon.
-The upper bound for good condition is set by regional median LiDAR heights within NiN polygons in the same bioclimatic-elevation zone as the population polygon.", out.width='70%', echo=FALSE}
-knitr::include_graphics("images/gjengroing_schematic.jpg")
-```
-
-### Reference state {#state-enc}
-
-The reference state is defined as one where åpne ecosystems are indeed open, with little to no woody encroachment. 
-For natural ecosystems this reflects a state where climate warming, wetland drainage, and alien species etc have not led to any net loss of these nature types and their internal functioning. 
-For semi natural ecosystems it reflects a state where traditional husbandry is keeping he ecosystems open due to grazing and hay making.
-
-### Reference values, thresholds for defining *good ecological condition*, minimum and/or maximum values {#ref-enc}
-The NiN polygons including "Naturlig åpne områder under skoggrensa", "Semi-naturlig mark", and "Våtmark" with good ecological condition are used to define a reference gjengroing state.
-We use the aggregated "Tilstand" variable assigned to each NiN polygon.
-The 50th percentile of LiDAR-derived vegetation heights within these polygons is used to define the upper limit (ie.
-1) of the scaled indicator value.
-We cannot define local reference values based on proximity, because the NiN polygons are spatially biased and not close to all AR5 population polygons.
-Therefore we calculate regional reference values using elevation (300m bands) and bioclimatic zones as stratification.
-We calculate the mean reference value for each unique combination of elevation-bioclimatic zone.
-When calculating the index for each population polygon, the reference value is inherited from the elevation-bioclimatic zone it falls within.
-
-Once we have the reference vegetation height for a given ecosystem type and elevation-bioclimatic zone, we need to define the minimum (or worst/bad) condition.
-We use the 50th percentile of LiDAR-derived vegetation heights within AR5 skog polygons to define a climax vegetation successional stage where gjengroing is at its most extreme.
-In order not to include forest patches that have recently been harvested, we mask out any forest which has been clear-cut since 1986.
-Here, 1986 is a hard limit defined by the clear-cut dataset which was based on Landsat imagery.
-Therefore we can be assured that we are measuring forest that is at least 35 years old.
-
-We use both local and regional "reference" approaches to defining poor ecological condition.
-To do this we calculate the forest heights in AR5 forest patches that fall within 200m of a population polygon (local reference).
-For population polygons that do not have forest within their proximity, we use the mean forest height within the unique elevation-bioblimatic zone the population falls within (regional reference).
-
-To define the ecological condition of the population polygon, we measure 50th percentile of LiDAR-derived vegetation height.
-As a reminder, we use AR5 polygons that have types that are approximately compatible with the NiN ecosystem types.
-For våtmark we use the AR5 polygons defined as "Myr".
-For åpne ecosystems we use the AR5 polygons defined as either "Åpent fastmark" or "Innmarksbeite".
-
-We acknowledge that the AR5 data set has several shortcomings in terms of lags in update frequency and with precision in the maps themselves. 
-This indicator will therefore benefit strongly from new ecosystem delineation maps, that update more frequently, and preferably once that can differentiate between semi-natural and naturally open ecosystems.
-
-The vegetation height percentiles are then scaled to between 0 and 1 using a sigmoid transformation [Oliver at al. (2021)](https://www.sciencedirect.com/science/article/pii/S1470160X21000066).
-
-## Uncertainties {#err-enc}
-
-For the indicator map at the polygon level there is no uncertainty associated with the indicator values.
-For aggregated indicator values (e.g.
-for 50km grid and regions), the uncertainty in the indicator value is calculated from the spatial variation in the polygon-level reference height values.
-We could use the alternative of calculating the spatial variation in the indicator values via bootstrapping using the EAtools package.
-However, given the extreme number of population polygons, the uncertainty values are extremely small and therefore we choose to use the standard deviation in reference vegetation heights, converted to the indicator scale.
-
-## References
-
-Links to data and resources are provided with hyperlinks in-line.
-
-## Analyses {#analyses-enc}
-
-### Data sets {#data-enc}
-
-There are several datasets which are used in GEE which will not be imported into the R session here.
-These datasets have been obtained from the source and ingested into GEE by Zander Venter or Vegard Bakkestuen with the help of Miljødata section at NINA.
-They include
-
--   [AR5](https://kartkatalog.geonorge.no/metadata/arealressurskart-fkb-ar5-arealtyper/280bbd7a-5ce9-4c83-9e15-ac162cabd8a6)
--   [LiDAR-derived digital elevation model from høydedata](https://hoydedata.no/LaserInnsyn/).
--   [Kartverket's national 10m elevation model](https://kartkatalog.geonorge.no/metadata/dtm-10-terrengmodell-utm32/fd851873-f363-46f9-9fc6-bb1b403575df)
--   [FKB building footprints](https://kartkatalog.geonorge.no/metadata/fkb-bygning/8b4304ea-4fb0-479c-a24d-fa225e2c6e97)
--   [European forest clear-cut map](https://www.nature.com/articles/s41893-020-00609-y)
-
-The remaining datasets will be imported into the R session.
+This script downloads image tiles to be used in the meta canopy height model (INSERT CROSSREF).
 
@@
 OK, now that we have our tiles, we can pass them to the canopy height model please see https://github.com/facebookresearch/HighResCanopyHeight for more details on the model.
-
-######Zanders/Ida's script######
-#### Regions {#reg-enc}
-
-The regional delineation for Norway (five regions) are used for aggregating and reporting gjengroing condition values.
-
-```{r}
-regions <- sf::st_read("data/regions.shp", options = "ENCODING=UTF8") %>%
-  mutate(region = factor(region))
-```
-
-#### SSB 10 and 50km grids
-
-The [SSB 10km and 50km grids](https://kartkatalog.geonorge.no/metadata/statistisk-rutenett-5000m/32ac0653-d95c-446c-8558-bf9b79f4934e) are used for visualizing the distribution of available data and for aggregating gjengroing index scores.
-
-```{r}
-ssb10km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_10km/ssb10km.shp')%>%
-  # mutate the id value so that it aligns with the data coming from GEE
-  mutate(SSBID = as.numeric(SSBID)/1000)
-# Transform to the correct projection
-ssb10km <- st_transform(ssb10km, st_crs(regions))
-
-ssb50km <- st_read('/data/R/GeoSpatialData/Population_demography/Norway_SSB/Original/ssb_50km/ssb50km.shp')%>%
-  # mutate the id value so that it aligns with the data coming from GEE
-  mutate(SSBID = as.numeric(SSBID)/1000)
-# Transform to the correct projection
-ssb50km <- st_transform(ssb50km, st_crs(regions))
-```
-
-#### Bioclimatic regions
-
-The [Moen's bioclimatic regions](https://data.artsdatabanken.no//Natur_i_Norge/Natursystem/Beskrivelsessystem/Regional_naturvariasjon/Bioklimatisk_sone) are imported from NINA R/GeoSpatialData/ server.
-
-```{r}
-bioclim <- st_read('/data/R/GeoSpatialData/BiogeographicalRegions/Norway_VegetationZones_Moen/Original/Vector/soner.shp') %>%
-  mutate(NAVN = ifelse(NAVN == 'S°rboreal', 'Sørboreal', NAVN))
-bioclim <- st_transform(bioclim, st_crs(regions))
-```
-
-#### NiN polygons
-
-[NiN polygons](https://kartkatalog.miljodirektoratet.no/dataset/Details/2031) are imported here from the NINA servers:
-
-```{r}
-nin <- st_read('/data/R/GeoSpatialData/Habitats_biotopes/Norway_Miljodirektoratet_Naturtyper_nin/Original/versjon20221231/Natur_Naturtyper_NiN_norge_med_svalbard_25833/Natur_Naturtyper_NiN_norge_med_svalbard_25833.gdb')%>%
-  # Simplify ecosystem names
-  mutate(hovedøkosystem = recode(hovedøkosystem, 
-                                 "Våtmark" = 'Våtmark',
-                                 "Skog" = "Skog",
-                                 "Ingen" = "Ingen",
-                                 "Fjell" = "Fjell",
-                                 "Semi-naturlig mark" = 'Semi-naturlig',
-                                 "Naturlig åpne områder i lavlandet" = 'Naturlig åpne',
-                                 "Naturlig åpne områder under skoggrensa" = 'Naturlig åpne')) %>% 
-  # Fix any invalid geometries
-  mutate(validGeo = st_is_valid(SHAPE)) %>%
-  filter(validGeo) %>%
-  # Filter out the ecosystems we are not interested in
-  filter(!hovedøkosystem %in% c('Skog','Ingen','Fjell')) %>%
-  # Get the main ecosystem codes in case we need them later
-  mutate(hovedtype = substr(ninkartleggingsenheter, 4, 6),
-         hovedtype = str_remove(hovedtype, '-'),
-         id = identifikasjon_lokalid) %>%
-  # Drop polygons with no condition score
-  drop_na(tilstand) %>%
-  dplyr::select(id, hovedøkosystem, hovedtype, naturtypekode, tilstand)
-```
-
--   **Export the cleaned NiN data and upload to GEE before proceeding**
-
-```{r}
-# Only run if you are repeating this workflow for the first time
-# nin %>%
-#   dplyr::select(id, hovedøkosystem, tilstand) %>%
-#   st_transform(st_crs(fylker_wgs84)) %>%
-#   st_write('nin_cleaned.shp', append=FALSE)
-```
-
--   **Now run the following GEE JavaScript code by clicking on the hyperlink:** [GEE script](https://code.earthengine.google.com/b2d06ad72c515fbbb3024f157987eedc) This script is also found in the SCR repository as "GEE_script.js".
-When you run the script, then click the Tasks tab in the GEE code editor and run the export tasks.
-The export files will appear in your Google Drive.
-Download the files from your Google Drive to the /data/gjengroing/From_GEE directory.
-This script extracts three CSV files:
-    -   'area_cover_grid.csv' - this gives the area coverage of NiN, AR5 and LiDAR data for each 10x10km SSB grid cell.
-    -   'elevation_stratified.shp' - this is a 300m elevation band stratification that is simply used for visualizing the strata here in R.
-    -   'vegHeights.zip' - this is a folder with hundreds of CSV files.
-These files contain the vegetation heights for reference, population, and forest polygons over Norway.
-The exports were split by 50km grid to prevent hitting user memory limits in GEE.
-Unzip the folder before you upload to ./DATA/From_GEE/
-
-#### Data from Google Earth Engine
-
-Once you have run the GEE script above, and downloaded the data to the `/gjengroing/R_project/DATA/From_GEE/` directory, you can proceed with this R workflow.
-
-Create functions to import vegetation height files and stratify data by elevation band and vegetation/climate zone
-
-```{r}
-# Function to read in multiple files resulting from GEE exports
-readVegHeightFiles <- function(uniqueString){
-  dir <- '/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/vegHeights/'
-  files <- list.files(dir)[str_detect(list.files(dir), uniqueString)]
-  #print(files)
-
-  dat <- tibble()
-  for (i in files){
-    dat <- dat %>% 
-      bind_rows(read_csv(paste0(dir, i)) %>%
-                  mutate(ssbid = substr(str_split(i, '_')[[1]][3], 1, 14)))
-  }
-  return (dat)
-}
-
-# Vegetation type lookup
-vegLookup <- tibble(
-  vegClimZone = c(1,2,3,4,5,6), 
-  vegClimZoneLab = c('Nemoral','Boreonemoral','Sørboreal','Mellomboreal','Nordboreal','Alpin')
-  )
-
-# Elevation stratification and vegetation type cleaning function
-cleanElevVegType <- function(data){
-
-  dataOut <- data %>%
-    mutate(e = elevation) %>%
-    mutate(elevation = ifelse(e < 300, '0-300',
-                              ifelse(e > 300 & e < 600, '300-600',
-                                     ifelse(e > 600 & e < 900, '600-900',
-                                     ifelse(e > 900 & e < 1200, '900-1200', '>1200'))))) %>%
-    mutate(vegClimZone = round(vegClimZone)) %>%
-    left_join(vegLookup, by = 'vegClimZone') %>%
-    dplyr::select(-e, -vegClimZone) %>%
-    drop_na(vegClimZoneLab, elevation)
-
-  return (dataOut)
-
-}
-
-```
-
-Import the CSV files generated in GEE related to area covers and elevation bands:
-
-```{r}
-# Import data coverages per SSB 10km grid cell
-areaCovers <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/area_cover_grid.csv') %>%
-  dplyr::select(-'.geo', -'system:index')
-
-# Calculate percentage cover relative to the land area in each cell
-areaCoversPerc <- areaCovers %>%
-  gather(key, val, aapne_ar5, vaatmark_ar5,aapne_nin, vaatmark_nin, lidarCover, skog) %>%
-  mutate(type = ifelse(str_detect(key, 'ar5'), 'ar5', 
-                       ifelse(str_detect(key, 'nin'), 'nin', 
-                              ifelse(key == 'lidarCover', 'lidar', NA)))) %>%
-  mutate(ecosystem = ifelse(str_detect(key, 'aapne'), 'Semi-/Naturlig åpne', 
-                       ifelse(str_detect(key, 'vaatmark'), 'Våtmark', NA))) %>%
-  mutate(areaPerc = val/land*100,
-         areaPerc = ifelse(is.na(areaPerc), 0, areaPerc)) %>%
-  dplyr::select(SSBID, key, type,ecosystem, areaPerc)
-
-elevStrata <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/From_GEE/elevation_stratified.shp') %>%
-  mutate(elevBand = ifelse(mode == 1, '0-300',
-                           ifelse(mode == 2, '300-600',
-                                  ifelse(mode == 3, '600-900',
-                                         ifelse(mode == 4, '900-1200', '>1200')))),
-         elevBand = factor(elevBand, levels=c('0-300','300-600','600-900','900-1200','>1200')))
-elevStrata <- st_transform(elevStrata, st_crs(regions))
-
-```
-
-Import the CSV files generated in GEE of vegetation heights (circa 15 min runtime):
-
-```{r}
-
-if (runFromScratch){
-
-  # Import median LiDAR heights for NiN polygons
-  refVaatmarkRaw <- readVegHeightFiles('vaatmark_ref') %>%
-    mutate(ref = chm) %>%
-    dplyr::select(-'.geo', -'system:index', -chm)
-  refAapneRaw <- readVegHeightFiles('aapne_ref') %>%
-    mutate(ref = chm) %>%
-    dplyr::select(-'.geo', -'system:index', -chm)
-
-  # Add elevation and veg/clim zones and summarise per strata
-    # this will form the upper limit of the ecological condition score (ie. good condition)
-  refVaatmark <- cleanElevVegType(refVaatmarkRaw) %>%
-    group_by(elevation, vegClimZoneLab) %>%
-    summarise(ref = mean(ref, na.rm=T))
-  refAapne <- cleanElevVegType(refAapneRaw) %>%
-    group_by(elevation, vegClimZoneLab) %>%
-    summarise(ref = mean(ref, na.rm=T))
-
-  # Export to file for reference
-  refVaatmark %>%
-    st_write('data/encroachment/refVaatmark.csv')
-  refAapne %>%
-    st_write('data/encroachment/refAapne.csv')
-
-  # Import median LiDAR heights for AR5 polygons for population measurement
-  popVaatmarkHt <- readVegHeightFiles('vaatmark_pop')  %>%
-    mutate(pop = chm)%>%
-    dplyr::select(-chm)
-  popAapneHt <- readVegHeightFiles('aapne_pop')  %>%
-    mutate(pop = chm)%>%
-    dplyr::select(-chm)
-
-  # Turn into spatial objects using the .geo column in the CSV
-  popVaatmarkHt <- st_as_sf(bind_cols(popVaatmarkHt,geojsonsf::geojson_sf(popVaatmarkHt$.geo)))
-  popAapneHt <- st_as_sf(bind_cols(popAapneHt,geojsonsf::geojson_sf(popAapneHt$.geo)))
-
-  # Import median LiDAR heights for forest (skog) within 200m of population polygons
-    # this will form the lower limit of the ecological condition score (ie. poor condition)
-  popVaatmarkSkog <- readVegHeightFiles('vaatmark_skog') %>%
-    mutate(skog = chm)%>%
-    dplyr::select(-chm, -'.geo', -ssbid)
-  popAapneSkog <- readVegHeightFiles('aapne_skog') %>%
-    mutate(skog = chm)%>%
-    dplyr::select(-chm, -'.geo', -ssbid)
-
-  # Combine the population heights with the forest heights
-  popVaatmark <- popVaatmarkHt %>%
-    left_join(popVaatmarkSkog, by = 'system:index')%>%
-    dplyr::select(-'.geo')
-  popAapne <- popAapneHt %>%
-    left_join(popAapneSkog, by = 'system:index')%>%
-    dplyr::select(-'.geo')
-
-  # Add elevation and veg/clim zones to the combined population and forest heights
-  popVaatmark <- cleanElevVegType(popVaatmark)
-  popAapne <- cleanElevVegType(popAapne)
-
-  # Get regional reference values for forest height
-    # this is to fill in for population polygons which did not have any forest within 200m
-  vaatmarkSkogRegionalRef <- popVaatmark %>%
-    as_tibble() %>%
-    group_by(elevation, vegClimZoneLab) %>%
-    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
-    drop_na(skogFill)
-  aapneSkogRegionalRef <- popAapne %>%
-    as_tibble() %>%
-    group_by(elevation, vegClimZoneLab) %>%
-    summarise(skogFill = mean(skog, na.rm=TRUE)) %>%
-    drop_na(skogFill)
-
-  # Combine all the above into one data frame
-  #set.seed(123)
-  vaatmarkHts <- popVaatmark %>%
-    #sample_n(100000) %>%
-    as_tibble() %>% dplyr::select(-geometry) %>%
-    left_join(refVaatmark, by=c('vegClimZoneLab', 'elevation')) %>%
-    gather(type, height, ref, pop, skog) %>%
-    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
-    summarise(height = mean(height)) %>%
-    pivot_wider(values_from = height, names_from = type) %>%
-    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled value will be 1
-    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
-    drop_na(pop, ref) %>%
-    left_join(vaatmarkSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
-    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
-    dplyr::select(-skogFill) %>%
-    filter(skog < 30)
-  #colSums(is.na(vaatmarkHts))
-
-  aapneHts <- popAapne %>%
-    #sample_n(100000) %>%
-    as_tibble() %>% dplyr::select(-geometry) %>%
-    left_join(refAapne, by=c('vegClimZoneLab', 'elevation')) %>%
-    gather(type, height, ref, pop, skog) %>%
-    group_by(`system:index`, elevation, vegClimZoneLab, type) %>%
-    summarise(height = mean(height)) %>%
-    pivot_wider(values_from = height, names_from = type) %>%
-    # If population height is less than the reference value, then it is automatically "good" condition - ie. it inherits the reference height so that the rescaled index will be 1
-    mutate(pop = ifelse(pop < ref, ref, pop)) %>%
-    drop_na(pop, ref) %>%
-    left_join(aapneSkogRegionalRef, by = c('vegClimZoneLab', 'elevation')) %>%
-    mutate(skog = ifelse(is.na(skog), skogFill, skog)) %>%
-    dplyr::select(-skogFill) %>%
-    filter(skog < 30)
-  #colSums(is.na(aapneHts))
-
-} else {
-
-  refVaatmark  <- read_csv('data/encroachment/refVaatmark.csv')
-  refAapne <- read_csv('data/encroachment/refAapne.csv')
-
-}
-
-```
-
-```{r ref-open-enc, fig.cap="Refenerence vegetation height in semi-natural and naturally open ecosystems under the forest line in Norway."}
-refAapne %>%
-  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
-         vegClimZoneLab = factor(vegClimZoneLab)) %>%
-  ggplot(aes(x = elevation, y = ref))+
-  geom_bar(stat="identity")+
-  theme_bw(base_size = 12)+
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
-  facet_wrap(.~vegClimZoneLab)+
-  labs(y = "Reference vegetation height (m)",
-       x = "Elevation (meters above sea level)")
-```
-
-```{r ref-wet-enc, fig.cap="Refenerence vegetation height in wetland ecosystems under the forest line in Norway."}
-refVaatmark %>%
-  mutate(elevation = factor(elevation, levels = c("0-300", "300-600", "600-900", "900-1200", ">1200")),
-         vegClimZoneLab = factor(vegClimZoneLab)) %>%
-  ggplot(aes(x = elevation, y = ref))+
-  geom_bar(stat="identity")+
-  theme_bw(base_size = 12)+
-  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 12))+
-  facet_wrap(.~vegClimZoneLab)+
-  labs(y = "Reference vegetation height (m)",
-       x = "Elevation (meters above sea level)")
-```
-
-### Explore data coverages
-
-Here we will make maps of NiN, AR5 and LiDAR coverage over Norway to get a feel for the data distributions.
-These areas were calculated in GEE and exported to CSV in the GEE script provided above.
-All we need to do now is join them with the SSB 10km grid and visualize.
-
-```{r coverage-map-enc, fig.height=5, fig.width=15, fig.cap="Map of Norway showing coverage of different data sets."}
-
-c1 <- ssb10km %>% 
-  left_join(areaCoversPerc %>%
-              filter(type == 'nin') %>%
-              group_by(SSBID) %>%
-              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
-  ggplot() +
-  geom_sf(aes(fill=areaPerc), color=NA) +
-  geom_sf(data=regions, fill=NA, size=0.3) +
-  scale_fill_gradientn(colours = c('black', 'orange'),
-                       name = '% cover',
-                       limits=c(0, 2),
-                       oob=scales::squish) +
-  theme_void() +
-  theme(legend.position = c(0.7, 0.4)) +
-  ggtitle('A) NiN polygons with good condition')
-#c1
-c2 <- ssb10km %>% 
-  left_join(areaCoversPerc %>%
-              filter(type == 'ar5') %>%
-              group_by(SSBID) %>%
-              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
-  ggplot() +
-  geom_sf(aes(fill=areaPerc), color=NA) +
-  geom_sf(data=regions, fill=NA, size=0.3) +
-  scale_fill_gradientn(colours = c('black', 'orange'),
-                       name = '% cover',
-                       limits=c(0, 100),
-                       oob=scales::squish) +
-  theme_void() +
-  theme(legend.position = c(0.7, 0.4)) +
-  ggtitle('B) AR5 polygons')
-#c2
-c3 <- ssb10km %>% 
-  left_join(areaCoversPerc %>%
-              filter(type == 'lidar') %>%
-              group_by(SSBID) %>%
-              summarise(areaPerc = sum(areaPerc, na.rm=T)))  %>%
-  ggplot() +
-  geom_sf(aes(fill=areaPerc), color=NA) +
-  geom_sf(data=regions, fill=NA, size=0.3) +
-  scale_fill_gradientn(colours = c('black', 'orange'),
-                       name = '% cover',
-                       limits=c(0, 100),
-                       oob=scales::squish) +
-  theme_void() +
-  theme(legend.position = c(0.7, 0.4)) +
-  ggtitle('C) LiDAR canopy height model')
-#c3
-
-coverageFig <- grid.arrange(c1, c2, c3, ncol=3, widths=c(1,1,1), padding = unit(0, "line"), newpage = T)
-
-```
-
-### Bioclimatic zone and elevation strata for regional reference values
-
-We will visualize the elevation bands and bioclimatic zones that are used to define regional reference values for poor (forest vegetation height) and good (NiN vegetation height) ecological condition.
-For each unique spatial combination of these strata (n = 21), we calculate a mean reference value later on in the analysis.
-Note that poor ecological condition is preferably calculated using a local reference (forest height within 200m of population polygon), but a regional reference is used where there are no nearby forest patches.
-
-```{r sones-and-bands, fig.cap="Bioclimatic sones and elevation bands used to define strata in the enchroachment indicator."}
-
-bioclimPlot <- bioclim %>%
-  ggplot() + 
-  geom_sf(aes(fill = NAVN), color=NA)+
-    scale_fill_manual(values=brewer.pal(10, 'Paired'), name = 'Zone')+
-    ggtitle('A) Bioclimatic zones') +
-    theme_void()
-
-elevPlot <- elevStrata %>%
-  ggplot() +
-  geom_sf(aes(fill = elevBand), color=NA, size=0.001)+
-    scale_fill_manual(values=brewer.pal(5, 'Pastel1'), name = 'HASL (M)')+
-    ggtitle('B) Elevation bands')+
-    theme_void()
-
-strataPlot <- grid.arrange(bioclimPlot, elevPlot, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)
-
-```
-
-We can see that while AR5 polygons are well distributed across the country, the NiN polygons are clustered and biased toward low elevation regions of the country.
-The LiDAR data covers nearly the entire country with some exceptions in the north.
-Future iterations of this work will need to attempt to quantify the biases introduced by the relative distributions of reference and population polygons and by the datasets used to quantify gjengroing (in this case LiDAR).
-
-### Scaled indicator values at polygon level (circa 120 min runtime)
-
-We will calculate the gjengroing index for each population polygon over Norway using a Sigmoid scaling function.
-
-```{r}
-
-# Define Sigmoid scaling function
-scaleSigmoid <- function(variable){
-  refLow <- min(variable)
-  refHigh <- max(variable)
-  thr <- (refHigh-refLow)/2
-  indicator_LowHigh <- (variable - refLow)/(refHigh - refLow)
-  indicator_LowHigh[indicator_LowHigh[] < 0] <- 0
-  indicator_LowHigh[indicator_LowHigh[] > 1] <- 1
-  indicator_sigmoid <- 100.68*(1-exp(-5*(indicator_LowHigh)^2.5))/100
-  return (round(indicator_sigmoid, 4))
-}
-
-if (runFromScratch){
-
-  # Calcualte gjengroing index values
-  vaatmarkIndexPoly <- vaatmarkHts  %>%
-    # Gather into long format so taht we can mutate the scaleSigmoid() function
-    gather(type, height, ref, pop, skog) %>%
-    group_by(`system:index`) %>%
-    mutate(index = scaleSigmoid(height)) %>%
-    # Because short vegetation is good condition, we must invert the scaled indicator
-    mutate(index = 1-index) %>%
-    # Get the indicator values for the population polygons only (ref and skog will be 1 and 0)
-    filter(type == 'pop') %>%
-    dplyr::select(-type, -height) %>%
-    # Join with original data to get the actual median vegetaiton heights for each indicator score
-    left_join(vaatmarkHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
-    left_join(popVaatmark %>% dplyr::select("system:index", ssbid))
-
-  # do the same for åpne polygons - will not comment this because same as above
-  aapneIndexPoly <- aapneHts  %>%
-    gather(type, height, ref, pop, skog) %>%
-    group_by(`system:index`) %>%
-    mutate(index = scaleSigmoid(height)) %>%
-    mutate(index = 1-index) %>%
-    filter(type == 'pop') %>%
-    dplyr::select(-type, -height) %>%
-    left_join(aapneHts %>% dplyr::select("system:index", pop, ref, skog)) %>%
-    left_join(popAapne %>% dplyr::select("system:index", ssbid))
-
-  # Make into spatial objects
-  vaatmarkIndexPolySpat <- popVaatmark %>%
-    dplyr::select("system:index") %>%
-    left_join(vaatmarkIndexPoly) %>%
-    filter(!is.na(index))
-  aapneIndexPolySpat <- popAapne %>%
-    dplyr::select("system:index") %>%
-    left_join(aapneIndexPoly) %>%
-    filter(!is.na(index))
-
-  # Transform to correct projection
-  vaatmarkIndexPolySpat <- st_transform(vaatmarkIndexPolySpat, st_crs(regions))
-  aapneIndexPolySpat <- st_transform(aapneIndexPolySpat, st_crs(regions))
-
-} else {
-
-  vaatmarkIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp')%>%
-    mutate(`system:index` = id,
-           elevation = elev, 
-           vegClimZoneLab = vegZn) %>%
-    dplyr::select(-id, -elev, -vegZn)
-  aapneIndexPolySpat <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp')%>%
-    mutate(`system:index` = id,
-           elevation = elev, 
-           vegClimZoneLab = vegZn) %>%
-    dplyr::select(-id, -elev, -vegZn)
-
-}
-
-```
-
-Here we will visualize the spatial variation in scaled gjengroings indicator values, as well as the vegetation heights for reference (good condition), indicator, and mature forest (poor condition) over the country.
-We could plot all the polygons, but it is computationally intensive and very difficult to visualize spatial variations over the entire country.
-Therefore we will calculate the area-weighted mean in values for each bioclimatic-elevation strata (n = 21) over the country.
-
-```{r multiple-maps-enc, fig.width=10, fig.cap="Maps of scaled and unscaled indicator values, as well as reference values.
-For the unscaled variables the legend is truncated at 5 m to ensure the same colour gradient for all maps."}
-
-# Create intersection of the two strata layers
-#elevStrata <- st_transform(elevStrata, st_crs(regions))
-bioClimElev <- bioclim %>%
-  st_intersection(elevStrata) %>%
-  mutate(elevation = elevBand,
-         vegClimZoneLab = NAVN)%>%
-  dplyr::select(elevation, vegClimZoneLab) %>%
-  rowid_to_column("ID")
-
-if (runFromScratch){
-
-  # Get area-weighted mean for reference, forest and population vegetation heights 
-  # for each unique bioclimatic-elevation zone combination
-  vaatmarkIndexStrata <- bioClimElev %>% 
-    #sample_n(10000) %>%
-    st_join(vaatmarkIndexPolySpat %>% 
-                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
-    group_by(ID) %>%
-    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
-                as_tibble() %>% 
-                dplyr::select(-geometry)
-
-  aapneIndexStrata <- bioClimElev %>% 
-    #sample_n(10000) %>%
-    st_join(aapneIndexPolySpat %>% 
-                mutate(area = as.numeric(st_area(geometry))) ,st_intersects) %>%
-    group_by(ID) %>%
-    summarise_at(vars(pop, ref, skog, index), funs(weighted.mean(., w=area, na.rm=TRUE)))%>% 
-                as_tibble() %>% 
-                dplyr::select(-geometry)
-
-  # Write out for import later
-  vaatmarkIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata %>% write_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
-
-} else {
-
-  vaatmarkIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/vaatmarkIndexStrata.csv')
-  aapneIndexStrata <- read_csv('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/DATA/aapneIndexStrata.csv')
-
-}
-
-# Make plots of indicator heights and scaled condition values for each strata polygon
-
-theme_legend_1 <- theme(legend.key.size = unit(0.25, 'cm'), #change legend key size
-                        legend.key.height = unit(0.25, 'cm'), #change legend key height
-                        legend.key.width = unit(0.25, 'cm'), #change legend key width
-                        legend.title = element_text(size=6), #change legend title font size
-                        legend.text = element_text(size=6)) #change legend text font size
-
-makeStrataHeightMap <- function(data, var, title, limits){
-  p <- bioClimElev %>%
-    left_join(data, by = c('ID'))  %>%
-    mutate(response = .[[var]]) %>%
-    ggplot() + 
-    geom_sf(aes(fill = response), color=NA)+
-      scale_fill_gradientn(colors=viridis::viridis(10),
-                           limits = limits,
-                           oob = scales::squish,
-                           name = 'Veg ht (m)')+
-      ggtitle(title) +
-      theme_void() +
-    theme(plot.title = element_text(size=9)) +
-    theme_legend_1
-
-  return (p)
-}
-
-makeStrataIndicatorMap <- function(data, title){
-  p <- bioClimElev %>%
-    left_join(data, by = c('ID'))  %>%
-    ggplot() + 
-    geom_sf(aes(fill = index), color=NA)+
-      scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
-                           limits = c(0,1),
-                           name = 'Index')+
-      ggtitle(title) +
-      theme_void()+
-    theme(plot.title = element_text(size=9)) +
-    theme_legend_1
-
-  return (p)
-}
-
-a1 <- makeStrataIndicatorMap(aapneIndexStrata,  'A) Semi-/Naturlig-åpne scaled indicator')
-a2 <- makeStrataHeightMap(aapneIndexStrata, "ref", 'B) Semi-/Naturlig-åpne reference height', c(0,5))
-a3 <- makeStrataHeightMap(aapneIndexStrata, "pop", 'C) Semi-/Naturlig-åpne indicator height', c(0,5))
-a4 <- makeStrataHeightMap(aapneIndexStrata, "skog", 'D) Semi-/Naturlig-åpne forest height', c(0,5))
-
-v1 <- makeStrataIndicatorMap(vaatmarkIndexStrata,  'E) Våtmark scaled indicator')
-v2 <- makeStrataHeightMap(vaatmarkIndexStrata, "ref", 'F) Våtmark reference height', c(0,5))
-v3 <- makeStrataHeightMap(vaatmarkIndexStrata, "pop", 'G) Våtmark indicator height', c(0,5))
-v4 <- makeStrataHeightMap(vaatmarkIndexStrata, "skog", 'H) Mature forest height', c(0,5))
-
-strataHeightPlot <- grid.arrange(a1,a2,a3,a4, v1,v2,v3,v4, ncol=4, widths=c(1,1,1,1), padding = unit(0, "line"), newpage = T)
-
-```
-
-We can see that overall, the condition scores for both ecosystem types are poor along the coastal areas and lower-altitude zones of the country (A and E).
-At higher altitudes the gjengroing condition scores are higher.
-
-When comparing våtmark to semi-/naturlig-åpne ecosystems, we can see that the reference heights for våtmark are generally higher in terms of good (B and F) and poor (D and H) ecological condition.
-This makes ecological sense because wetlands can and do naturally occur under forest and therefore våtmark in good condition can have tree cover without necessarily being gjengroing.
-
-### Scaled indicator values at 50km grid level (circa 45 min runtime)
-
-Using the polygon-level indicator scores, we will aggregate to 50km grid level to explore the spatial variation over Norway.
-We are doing this as an alternative way of visualizing the broad-scale patterns as presented in the section above where we used bioclimatic-elevation strata.
-
-Here we will use a weighted average, but test out the eaTools::ea_spread function! It takes some time to run so best to do this overnight, or to parallelize in some way.
-
-```{r}
-
-if (runFromScratch){
-
-  # Aggregate polygon index scores to grid level using eaTools ea_spread function
-  vaatmarkIndexGrid <- eaTools::ea_spread(indicator_data = vaatmarkIndexPolySpat,
-                                            indicator = index,
-                                            regions = ssb50km,
-                                            groups = SSBID,
-                                            threshold = 1) %>%
-    mutate(index = w_mean, ssbid = ID) %>%
-    dplyr::select(ssbid, index, sd)
-
-  aapneIndexGrid <- eaTools::ea_spread(indicator_data = aapneIndexPolySpat,
-                                            indicator = index,
-                                            regions = ssb50km,
-                                            groups = SSBID,
-                                            threshold = 1) %>%
-    mutate(index = w_mean, ssbid = ID) %>%
-    dplyr::select(ssbid, index, sd)
-
-} else {
-
-  print('Will import pre-exported data')
-
-  vaatmarkIndexGrid <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp')
-  aapneIndexGrid <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp')
-
-}
-
-```
-
-Visualize the spatial distribution of gjengroing indicator scores at grid level.
-
-```{r grid-level-enc, fig.cap="Maps displaying scaled encroachment indicator values (A, E) along with the raw vegetation heights defining upper limits for good (B, F), and bad (D, H) condition, and observed population vegetation heights (C, G).
-Values are aggregaed to polygons defining unique bioclimatic-elevation strata over the country."}
-
-vGrid1 <- aapneIndexGrid %>%
-  ggplot() +
-  geom_sf(aes(fill=index), color=NA) +
-  geom_sf(data=regions, fill=NA, size=0.5) +
-  scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
-                       limits = c(0,1),
-                       name = 'Index') +
-  theme_void() +
-  theme(legend.position = c(0.7, 0.4)) +
-  ggtitle('A) Semi-/Naturlig gjengroing condition')
-#vGrid1
-
-vGrid2 <- vaatmarkIndexGrid %>%
-  ggplot() +
-  geom_sf(aes(fill=index), color=NA) +
-  geom_sf(data=regions, fill=NA, size=0.5) +
-  scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
-                       limits = c(0,1),
-                       name = 'Index') +
-  theme_void() +
-  theme(legend.position = c(0.7, 0.4)) +
-  ggtitle('B) Våtmark gjengroing condition')
-#vGrid2
-
-indexGridFig <- grid.arrange(vGrid1, vGrid2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)
-
-```
-
-These patterns mirror those in the previous section with higher condition scores at higher elevations and lower scores along the coastal zone and low-lying parts of the country.
-
-We see some very low condition scores for a few coastal 50km grid cells - this may be an artifact of creating an average with very few population polygons and therefore vulnerable to skewed average indicator scores.
-
-### Scaled indicator values at regional level (circa 120 min runtime)
-
-Now we will aggregate up to the regional level for final reporting.
-We will use the same method as in the section above, except here we will use a different method to calculate the uncertainty value around the indicator values.
-The ea_spread function in eaTools uses a bootstrapping method to calculate the standard error in indicator values when aggregating from the polygon to regional level.
-Because we have hundreds of thousands of polygons per region, the resulting standard errors are extremely small.
-Therefore we will use the standard deviation in reference values per region.
-The standard deviation, as a percentage of the average reference value is used to calculate the uncertainty in the scaled indicator value.
-
-```{r}
-
-if (runFromScratch){
-
-  # Get the mean population, reference and forest heights per region for interpreting index values
-  # The standard deviation is used later to define uncertainty around the indicator score
-  vaatmarkHeightsRegion <- regions %>%
-    st_join(vaatmarkIndexPolySpat) %>%
-    as_tibble() %>%
-    drop_na(index) %>%
-    group_by(region) %>%
-    summarise(pop_mean = mean(pop),
-              skog_mean = mean(skog),
-              ref_mean = mean(ref),
-              sd_ref = sd(ref))
-  aapneHeightsRegion <- regions %>%
-    st_join(aapneIndexPolySpat) %>%
-    as_tibble() %>%
-    drop_na(index) %>%
-    group_by(region)  %>%
-    summarise(pop_mean = mean(pop),
-              skog_mean = mean(skog),
-              ref_mean = mean(ref),
-              sd_ref = sd(ref))
-
-  # Aggregate polygon index scores to regional level using eaTools
-  vaatmarkIndexRegion <- eaTools::ea_spread(indicator_data = vaatmarkIndexPolySpat,
-                                            indicator = index,
-                                            regions = regions,
-                                            groups = region,
-                                            threshold = 1) %>%
-    mutate(index = w_mean, region = ID) %>%
-    dplyr::select(region, index, sd) %>%
-    left_join(vaatmarkHeightsRegion)
-
-  # Only run on first iteration - takes a long time
-  aapneIndexRegion <- eaTools::ea_spread(indicator_data = aapneIndexPolySpat,
-                                            indicator = index,
-                                            regions = regions,
-                                            groups = region,
-                                            threshold = 1) %>%
-    mutate(index = w_mean, region = ID) %>%
-    dplyr::select(region, index, sd) %>%
-    left_join(aapneHeightsRegion)
-
-} else {
-
-  print('Will import pre-exported data')
-
-  vaatmarkIndexRegion <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp')
-  aapneIndexRegion <- st_read('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp')
-
-}
-
-```
-
-Present the indicator values as table, plots, and maps:
-
-```{r}
-
-# Join into a table to present values
-aapentTable <- aapneIndexRegion %>% as_tibble() 
-vaatmarkTable <- vaatmarkIndexRegion %>% as_tibble()
-
-# Join ecosystem types and calculate uncertainty
-indicatorTable <- aapentTable %>% mutate(Ecosystem = 'Semi-/Naturlig åpne') %>%
-  bind_rows(vaatmarkTable %>% mutate(Ecosystem = 'Våtmark')) %>%
-  # Calculate the error based on the standard deviation in reference polygon heights
-  # as a percentage of the reference height, and multiply by index score
-  mutate(sd = (sd_ref/ref_mean)*index)
-
-# Format the table for presentation
-indicatorTable_formatted <- regions %>%
-  left_join(indicatorTable) %>% as_tibble() %>%
-  dplyr::select(-geometry)  %>%
-  dplyr::select( region, Ecosystem, index, sd, pop_mean, ref_mean, skog_mean) %>%
-  mutate_if(is.numeric, round, 2)
-names(indicatorTable_formatted) <- c( "Region", "Ecosystem",
-                                      "Scaled indicator",  "Scaled Std.dev",
-                                      "Indicator height","Reference height",
-                                      "Mature forest height")
-
-# Use knitr to visualize
-knitr::kable(indicatorTable_formatted)
-
-# Make plots of regional index values
-getIndicatorPlot <- function(label){
-  regions %>%
-    left_join(indicatorTable %>%
-                filter(Ecosystem == label)) %>%
-    mutate(lowError = index-sd, 
-           upError = index+sd,
-           upError = ifelse(upError > 1, 1, upError)) %>%
-    ggplot(aes(y=region, x=index)) +
-    geom_point() +
-    geom_segment(aes(yend=region, x=lowError, xend=upError)) +
-    coord_cartesian(xlim=c(0,1)) +
-    geom_vline(xintercept = 1) +
-    geom_vline(xintercept = 0) +
-    labs(x = 'Tilstandsverdi') +
-    theme(axis.title.y=element_blank()) +
-    ggtitle(label) 
-}
-i1 <- getIndicatorPlot('Semi-/Naturlig åpne')
-i2 <- getIndicatorPlot('Våtmark')
-indicatorGraphFig <- grid.arrange(i1, i2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T)
-
-# Make maps of regional index values
-getIndicatorMap <- function(data, label){
-  regions %>%
-    left_join(data) %>%
-    ggplot() +
-    geom_sf(aes(fill=index)) +
-    scale_fill_gradientn(colors=brewer.pal(10, 'Spectral'),
-                         limits = c(0,1))+
-    ggtitle(label)+
-    theme_void()
-}
-m1 <- getIndicatorMap(aapentTable, 'A) Semi-/Naturlig åpne')
-m2 <- getIndicatorMap(vaatmarkTable, 'B) Våtmark')
-indicatorGMapFig <- grid.arrange(m1, m2, ncol=2, widths=c(1,1), padding = unit(0, "line"), newpage = T) 
-```
-
-From these tables, figures and maps we can see that the gjengroing condition indicator scores are on average above 0.8 and therefore indicate a good overall condition in Norway.
-Østlandet comes out best for våtmark and semi-naturlig åpne ecosystems.
-Nord-Norge has poorest condition for semi-naturlig åpne ecosystems, while vestlandet has the poorest scores for våtmark.
-
-### Uncertainty and validation
-
-The calculation of the indicator involves many small decisions which can have significant effects on the resulting conditions cores.
-Therefore in the future it would be good to perform sensitivity analyses with each decision to test the effect on the outcome.
-The decisions which would benefit from further exploration include:
-
--   **Selection of the reference and population areas.** Here we were limited to using NiN and AR5 because these are arguably the best datasets currently available for this exercise.
-However, both NiN and AR5 are spatially and thematically biased in some ways.
-Therefore, it may be worth exploring alternative datasets in the future.
-NiN polygons are biased towards low lying areas which are generally warmer and more productive.
-For a given bioclimatic-elevation strata, this could lead to higher reference values for vegetation height than one would expect with a representative reference sample.
-This could consequently result in high elevation population polygons appearing to have very good condition because they are being compared with (relatively) high productivity polygons in the same bioclimatic-elevation strata.
-This is a possible explanation for why the high elevation parts of Norway appear to be in very good ecological condition.
-
--   **Definition of good condition states in NiN data.** Here we used the overall "tilstand" score for each NiN reference polygon.
-However, the overall score is a combination of several sub-scores that may or may not be relevant for ecological condition.
-For instance, ditches or wheel marks in wetlands is a common condition score in NiN, however it is debatable as to whether this should be considered a part of ecological condition.
-Therefore it may be beneficial to attempt repeating the indicator calculation using sub-categories of "tilstand" in the NiN data.
-
--   **Definition of climax successional stage for tree regrowth.** Here we use the 50th percentile of LiDAR-derived tree heights in AR5 forest polygons that have not been harvested since 1986.
-This defines the zero on the scaled indicator score.
-The resulting indicator scores will vary widely depending on how you define this zero-point.
-Therefore, testing different definitions (i.e.
-tree height percentiles) may be important.
-
--   **Quantifying regional vs local references for poor and good condition.** For good condition, we used NiN polygons and were restricted to these due to lack of better data.
-However, for poor condition we use climax successional stage for trees (as explained above) within 200m of each population polygon.
-For polygons with no surrounding forest, we impute with regional reference values (mean of forest height in elevation-bioclamic zone).
-The choice of 200m buffer zone was largely based on computational limitations.
-Calculating the average LiDAR vegetation height within 200m of nearly 2 million complex polygons takes time, even in GEE.
-It may make more ecological sense to define a larger buffer zone to use for defining reference for poor condition.
-However, longer processing times need to be expected.
-
--   **Quantification of uncertainty around indicator scores.** Here we used the standard deviation in the vegetation heights for reference (NiN) polygons within each region to quantify the uncertainty around scaled indicator scores.
-We did this because the method used in `eaTools::ea_spread()` function is a bootstrapping approach which resulted in extremely small uncertainty estimates due to the large number of polygons included in our analysis.
-[In the future, a better method for quantifying uncertainty is needed.](https://github.com/NINAnor/eaTools/issues/17) Perhaps by running sensitivity analyses that test several variations of the points mentioned above can form the basis for quantifying an error margin around indicator values.
-
-#### Validation of random polygons
-
-After exporting the final polygon-level indicator scores, we imported to QGIS and explored how they look at the landscape scale.
-We chose random locations spread across the 5 regions in Norway.
-The figures below show våtmark and åpne (semi- and naturlig-åpne) ecosystem polygons from AR5.
-The polygons have jagged edges because we chose a low resolution geometry export from GEE to save computing time and minimize storage space.
-Each polygon is colored based on the final scaled gjengroing indicator score (0 to 1).
-The polygons are also labelled with the median LiDAR-derived forest height.
-An orthophoto from Norgeibilder is displayed in the background for reference.
-
-In the first example from midt-norge, we see a våtmark polygon colored red (ie.
-in poor condition) due to the growth of a forest in the northern section.
-It looks like a plantation forest and possibly occurred after the wetland was drained.
-The height of the forest vegetation equates to a median of 2.4m which is higher than the reference value for this region (1.09m) and almost as high as the mature forest height reference of 2.8m.
-Therefore it gets a scaled indicator score of 0.1.
-In contrast the wetland polygon in the middle of the image is in good condition because it has a median vegetation height of 1.1m which is only fractionally higher than the regional reference for good ecological condition.
-
-![](images/val_1.jpg){width="50%"}
-
-The second example from Sørlandet shows three AR5 semi-naturlig åpne polygons with different indicator values.
-The lower polygon has poor ecological condition because its median vegetation height is 6.2m which is close to the local reference value for mature forest (7.9m).
-Although the central polygon appears to have greater coverage of forest in the orthophoto, the LiDAR data tells us that it is mostly covered by short trees and results in a median vegetation height of 2.8 - therefore good condition.
-
-![](images/val_2.jpg){width="50%"}
-
-The third example from Østlandet shows a range of indicator values for våtmark polygons.
-Although many of the polygons are partly covered by trees, their indicator values show good condition because the local reference for mature forest (zero on the indicator scale) is very high in this landscape of productive forest.
-
-![](images/val_3.jpg){width="50%"}
-
-The fourth example from Nord-norge shows AR5 semi-naturlig åpne polygons with low and high condition scores.
-The large polygon with poor condition score has a median vegetation height of 2.7 which is nearly as high as the local reference for mature forest (3.8m).
-
-![](images/val_4.jpg){width="50%"}
-
-The fifth and final example from further north in Nord-norge shows two våtmark polygons with contrasting condition scores, yet exactly the same median vegetation heights of 0.6m.
-The difference in the scaled indicator value is due to different local reference values for surrounding forest.
-The polygon on the left has surrounding forest (within 200m buffer zone) with median height 0.5m, while the polygon on the right has surrounding forest of 0.7m.
-This creates a contrast in the final indicator value and illustrates the effect that a local reference value can have on defining good and poor ecological condition.
-
-![](images/val_5.jpg){width="50%"}
-
-### Eksport file (final product) {#exp-enc}
-
-We will export gjengroing indicator values at three levels: - polygon level - 50km grid level - regional level
-
-```{r}
-
-if (runFromScratch){
-  # Polygon level
-  vaatmarkIndexPolySpat  %>%
-    #mutate(area = as.numeric(st_area(geometry)))%>% 
-    #filter(area > 7500) %>%
-    mutate(id = `system:index`,
-           elev = elevation, 
-           vegZn = vegClimZoneLab) %>%
-    dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index.shp', append=FALSE)
-
-  aapneIndexPolySpat %>% 
-    #mutate(area = as.numeric(st_area(geometry)))%>% 
-    #filter(area > 7500) %>%
-    mutate(id = `system:index`,
-           elev = elevation, 
-           vegZn = vegClimZoneLab) %>%
-    dplyr::select(-"system:index", -vegClimZoneLab, -elevation) %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index.shp', append=FALSE)
-
-  # Grid level
-  vaatmarkIndexGrid %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_grid.shp', append = FALSE)
-  aapneIndexGrid %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_grid.shp', append = FALSE)
-
-  # Region level
-  vaatmarkIndexRegion %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/vaatmark_index_region.shp', append = FALSE)
-  aapneIndexRegion %>%
-    st_write('/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/gjengroing/R_project/OUTPUT/aapne_index_region.shp', append = FALSE)
-}
-
-```
+Please follow (INSERT CROSSREF)
